### 2020-12-04 Outline.Rmd
intial push

---
title: "Remote sensing the gas transfer velocity using SWOT observations"
author: "Craig Brinkerhoff"
date: "December 4, 2020"
output:
 # pdf_document:
#    latex_engine: xelatex
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
options(tinytex.verbose = TRUE)
```

## Background

1)	CO2 emissions from rivers are significant (duh)
2)	Estimating this flux is done by estimating in-stream CO2 and the gas transfer velocity, or ‘evasion rate’ or k600 of the river channel
3)	K600 is incredibly cumbersome to directly measure, and impractical when performing upscaling to entire river networks
4)	Previous work has stressed that this is largely a f(turbulence) driven by river hydraulics/geomorphology
    a.	Empirical scaling functions have been developed to use hydraulics to model k600 for entire river systems but are prone to significant uncertainties (Raymond et al. 2012)
    b.	They are also fundamentally dependent on modeled or measured flow velocities which themselves are prone to significant uncertainties (or poorly informed hydraulic geometry models…)
    c.	What is needed is a simple way to directly observe the gas transfer velocity at scale and high temporal resolution without needing explicit flow velocity estimates. We argue that the upcoming NASA SWOT mission provides exactly the data required to make this a reality and remotely sense (RS) k600 at unprecedented temporal resolution

## Research Goals

1)	Formulate a new empirical scaling model to predict k600 using only RS-able geomorphology
2)	Build algorithm to run the new empirical model using SWOT observations
3)	?????????
    a.	Biogeochemists might be interested in a Merritt style test of this algorithm at field-scale (to circumvent velocity measurements…)
    b.	Pepsi2-style comparisons across temporal sampling?
    c. MORE

## Goal 1
#### Theory
1) Raymond 2012 show that k600 can be scaled using hydraulics b/c k600 generally correlates with the turbulent dissipation rate (eD = gSV)
2) Ulseth et al 2019 show two distinct scaling regimes for k600 with eD using an expanded dataset including steeper streams than those in Raymond 2012.
3) Here, we take the Ulseth et al. 2019 dataset and develop a k600~eD scaling function that accounts for this regime shift but is entirely remotely sensible (i.e. not velocity dependent)
	  a. k_600=a(gSV)b where a and b are differentially determined by river size. Thus, from a river’s width and slope alone (i.e. SWOT observations) we can choose the most reasonable scaling function.

![**Figure 1** *eD~k600 scaling by river width/slope. This is how we are able to scale k600 using just SWOT observables*](outputs//k600//width_eD_theory.jpg)

4) To test this, we then combine these five scaling functions into a single 'rule-based' regression model, train it on 80% of the data and validate it on the remaining 20% (below). The version ultimately implemented in our algorithm is trained on the entire dataset however.

![**Figure 2** *Model validation on 20% of data, withheld from training*](outputs//k600//validation.jpg)

#### Quantifying Uncertainity in Model Estimates
1) We then quantified the total uncertainity implicit in the model output. Uncertainty can stem from three sources: 1) error in using Manning's equation to approximate flow velocity, 2) error in using a k600~eD scaling model to approximate k600, and 3) error implicit in our remote sensing algorithm. This third error source will be explicitly accounted (via Bayesian inference) in the RS algorithm output, however we also wanted to quantify total uncertainty from the model being used in the RS algorithm, i.e. from the other two sources. To do this, we used Monte Carlo simulations to propogate uncertanites through the entire model.
    a. 8,000 sets of river hydraulic measurements from across the United States were taken from the Brinkerhoff etal (2019) dataset. For each one I ran a 10,000 run Monte Carlo simulation (sampling from the three parameters a, b, and "Manning's velocity") and obtained 8,000 different distributions quantifying model estimate uncertainity. Log uncertainity for the "Manning's velocity"" was assigned 0.25 following Hagemann et al. (2017).

![**Figure 3** *Model Uncertainity estimates: a) histogram of SDs for all 8,000 MC simulations b) Three example MC simulations of 10,000 samples each c) will be a map of the 8,000 sets of hydraulic measurements*](outputs//MonteCarlo//MCsimulations.jpg)

## Goal 2
1) We propose the BIGER (Bayesian Inference/Inversion of Gas Evasion Rate) algorithm to remotely sense k600 from river width/height alone.
2) We implement the k600~eD model from Q1 as a Bayesian likelihood function following methods developed for ungauged RSQ (Hagemann et al. 2017). Below is the derivation of the likelihood function used in the Bayesian inference model.
	  a. $k_{600}=a(gSV)^{b}$
	  b. $k_{600}=a(gS[\frac{1}{n} d^{2/3} S^{1/2}])^{b}$
	  c. $k_{600}=a(gS[\frac{1}{n} (A_{0}+dA)^{2/3} W^{-2/3} S^{1/2}])^{b}$
	  d. $k_{600}(\frac{1}{n}(A_0+dA)^{2/3})^{-b}=a(g W^{2/3}S^{3/2})^{b}$
	    i. Model parameters on left-hand side (need priors), data on right-hand side
	    ii. priors on k600, n, A0
	      -geoBAM  Brinkerhoff etal 2020
	      -k600 priors obtained similarly using Ulseth etal 2019
3) For the sake of validating the algorithm, we assume that the scaling model is ‘truth’ and see if we can infer these values using only SWOT observables (river width and height)
    a. Therefore, in the validation setup, posterior uncertainty is only Manning’s uncertainty (0.25) and not the total model uncertainty obtained in the MC simulations earlier. However, this total uncertainity would be implemented if running the algorithm for basic science objectives.
    b. We validated on 10-day pepsi 2 data. Plotted are means and 95% CIs of the posterior k600 values versus the ‘observed’ k600 calculated from observed velocity and slope for all timesteps across all 32 rivers

![**Figure 4** *Validation of remote sensing algorithm for 32 rivers with 10day sampling intervals*](outputs//validation//validation.jpg)

Below are boxplots of performance metrics for the 32 rivers, i.e. how well do individual rivers do? NRMSE, RRMSE, and rBIAS are optimally zero while KGE is optimally 1. A KGE greater than -0.41 indicates our model outperforms a mean estimate, akin to an NSE of 0 for the hydroglogists out there (Knoben etal 2019).

![**Figure 5** *Pepsi validation by river*](outputs//validation//validation_by_river.jpg)

The result is great KGEs, NRMSEs, and RRMSEs (espically for aboslutely no in situ information) however a pretty high rBIAS. This bias is borne out in the scatterplot earlier.

##Goal 3
Who knows! Any interesting questions you can think of?

### 2020-12-04 scripts/Figures_pepsi.R

@@ -1,0 +1,116 @@
+#Summer 2020
+#Craig Brinkerhoff
+#Produce figures for pepsi validation for RSK600 algorithm
+
+library(tidyr)
+library(ggplot2)
+library(cowplot)
+library(RColorBrewer)
+library(dplyr)
+library(colorspace)
+library(ggtext)
+theme_set(theme_cowplot())
+
+# Sum of squares function------------------------------
+sumsq <- function(x) sum(x^2)
+
+#read in results----------------------------------
+full_output <- read.csv('inputs//validation//results_10day.csv')
+r2 <- round(summary(lm(log(full_output$kest_mean)~log(full_output$kobs)))$r.squared, 2)
+t <- drop_na(full_output)
+rmse <- round(exp(Metrics::rmse(log(t$kest_mean), log(t$kobs))), 2)
+
+#Model Validation across all rivers and timesteps--------------------------------------------------------
+valPlot <- ggplot(full_output, aes(x=(kobs), y=(kest_mean))) +
+  geom_abline(size=2, linetype='dashed', color='grey') +
+ # geom_point(size=5, pch=21, color='black', fill='#1b9e77') +
+  geom_pointrange(aes(ymin = kest_low, ymax = kest_high), fatten=10, fill='#1b9e77', pch=21, color='black') +
+  geom_smooth(size=2, color='black', method='lm', se=F)+
+  xlab('Observed k [m/dy]') +
+  ylab('Remotely Sensed k [m/dy]') +
+    scale_y_log10(
+      breaks = scales::trans_breaks("log10", function(x) 10^x),
+      labels = scales::trans_format("log10", scales::math_format(10^.x)))+
+   scale_x_log10(
+      breaks = scales::trans_breaks("log10", function(x) 10^x),
+      labels = scales::trans_format("log10", scales::math_format(10^.x)))+
+  coord_cartesian(xlim=c(10^-1, 10^1.3), ylim=c(10^-1,10^1.3))+
+  scale_color_discrete_qualitative(palette = 'Harmonic') +
+  theme(legend.position = "none",
+        axis.text=element_text(size=20),
+        axis.title=element_text(size=24,face="bold"),
+        legend.text = element_text(size=17),
[…]

### 2020-12-04 scripts/k600_model.R

@@ -1,0 +1,249 @@
+#Modeling k600 from hydraulics
+#dataset is combing data used in Raymond et al 2012, Bealiue at eal 2012, and Ulseth et al 2019 (hopefully)
+#Fall 2020
+
+library(tidyverse)
+library(cowplot)
+library(Metrics)
+library(colorspace)
+library(ggfortify)
+library(ggtext)
+
+theme_set(theme_cowplot())
+options(scipen = 999)
+
+#Ulseth etal 2019------------------------------------------------------------
+ulseth_data <- read.csv('inputs//k600//Ulseth_etal_2019.csv') #contains data from 5 studies, including the infamous raymond etal 2012
+ulseth_data <- ulseth_data[,-8]
+colnames(ulseth_data)[colnames(ulseth_data) == 'ï..k600'] <- 'k600'
+ulseth_data$data <- ifelse(ulseth_data$data == 'This Study', 'Ulseth et al. 2019', as.character(ulseth_data$data))
+ulseth_data$data <- ifelse(ulseth_data$data == 'Raymond et al. 2012', 'Raymond et al. 2012 [5 studies]', as.character(ulseth_data$data))
+
+data <- ulseth_data
+data <- filter(data, is.na(width)==0) #filter out no width measurements
+
+data$eD <- 9.8 * data$slope * data$Vms #water column turbulent dissipation rate m2/s3
+data$eS <- sqrt(9.8*data$depth*data$slope) / data$depth #turbulent dissipation rate from bed friction
+
+#width regime models for k600-----------------------------------------------------------------------
+  #Width regimes are necessary because we can't remotely sense eD directly (hence the entire Bayesian algorithm!)
+    #Thus, we can't know the eD threhsold of 0.02 a priori. BUT, we do know width a priori
+ data$widthRegime <- ifelse(data$width < 10 & data$slope < 0.05, '1',
+                            ifelse(data$width < 10 & data$slope > 0.05, '1.5',
+                                ifelse(data$width < 50, '2',
+                                   ifelse(data$width < 100, '3','4'))))
+
+#Check linearity assumptions within each width class. If you cycle through the different classes, everything holds
[…]

### 2020-12-07 Outline.Rmd
Updates to Outline and Figures

@@
 author: "Craig Brinkerhoff"
-date: "December 4, 2020"
+date: "`r Sys.Date()`"
 output:
@@
     b. Pepsi2-style comparisons across temporal sampling?
-    c. MORE
+    c. Some sort of k600 timeseries analysis? But these are all simulated k so I dunno.
+    d. MORE
 
@@
 2) Ulseth et al 2019 show two distinct scaling regimes for k600 with eD using an expanded dataset including steeper streams than those in Raymond 2012.
-3) Here, we take the Ulseth et al. 2019 dataset and develop a k600~eD scaling function that accounts for this regime shift but is entirely remotely sensible (i.e. not velocity dependent)
-	  a. k_600=a(gSV)b where a and b are differentially determined by river size. Thus, from a river’s width and slope alone (i.e. SWOT observations) we can choose the most reasonable scaling function.
+3) Here, we take the Ulseth et al.
+2019 dataset and develop a k600~eD scaling function that accounts for this regime shift and the fact that scaling relationships differ as a f(river size).
+BUT, we only use SWOT observables to determine which scaling relationship to use.
+    a. i.e. river width and slope alone must be used to choose which scaling model to use.
+	    i. $k_{600}=a(gSV)^{b}$ where a and b are differentially determined by river size.
+	  b. Small rivers had to be broken into high/low slope to maintain all linear regression assumptions.
 
@@
 #### Quantifying Uncertainity in Model Estimates
-1) We then quantified the total uncertainity implicit in the model output.
-Uncertainty can stem from three sources: 1) error in using Manning's equation to approximate flow velocity, 2) error in using a k600~eD scaling model to approximate k600, and 3) error implicit in our remote sensing algorithm.
-This third error source will be explicitly accounted (via Bayesian inference) in the RS algorithm output, however we also wanted to quantify total uncertainty from the model being used in the RS algorithm, i.e.
+1) In order to remotely sense k600, we need a way to approximate flow velocity as it is not easily RS-able.
+We use a variant of Manning's equation, THE classic open channel flow depth~velocity relation, to do this (Durand et al.
+2014; Hagemann et al.
+2017 used to do the same thing from getting discharge from SWOT data).
+	  a. $k_{600}=a(gSV)^{b}$ Original Model
+	  b. $k_{600}=a(gS[\frac{1}{n} d^{2/3} S^{1/2}])^{b}$ Classical Manning's equation introduced
+	  c.
+$k_{600}=a(gS[\frac{1}{n} (A_{0}+dA)^{2/3} W^{-2/3} S^{1/2}])^{b}$ Durand et al.
+(2014) formulation of Manning's equation.
+(The handling of n and A0 in a RS framework will be addressed under Goal 2).
+2) We then quantified the total uncertainity implicit in the model output.
+Uncertainty can stem from three sources: 1) error from Manning's equation, 2) error in using a k600~eD scaling model, and 3) error implicit in our remote sensing algorithm.
+This third error source will be explicitly accounted for (via Bayesian inference) in the RS algorithm output, however we also wanted to quantify total uncertainty from the model being used in the RS algorithm, i.e.
 from the other two sources.
@@
 8,000 sets of river hydraulic measurements from across the United States were taken from the Brinkerhoff etal (2019) dataset.
-For each one I ran a 10,000 run Monte Carlo simulation (sampling from the three parameters a, b, and "Manning's velocity") and obtained 8,000 different distributions quantifying model estimate uncertainity.
+For each one I ran a 10,000 run Monte Carlo simulation (randomly sampling from the three parameters a, b, and "Manning's velocity") and obtained 8,000 different distributions quantifying model estimate uncertainity.
 Log uncertainity for the "Manning's velocity"" was assigned 0.25 following Hagemann et al.
-(2017).
+(2017)'s value for Manning's equation uncertainity.
 
-![**Figure 3** *Model Uncertainity estimates: a) histogram of SDs for all 8,000 MC simulations b) Three example MC simulations of 10,000 samples each c) will be a map of the 8,000 sets of hydraulic measurements*](outputs//MonteCarlo//MCsimulations.jpg)
+![**Figure 3** *Model uncertainity estimates: a) histogram of SDs for all 8,000 MC simulations b) Three example MC simulations of 10,000 samples each c) Map of the 8,000 sets of hydraulic measurements used for this analysis*](outputs//MonteCarlo//MCsimulations.jpg)
 
@@
 Below is the derivation of the likelihood function used in the Bayesian inference model.
-	  a. $k_{600}=a(gSV)^{b}$
-	  b. $k_{600}=a(gS[\frac{1}{n} d^{2/3} S^{1/2}])^{b}$
-	  c. $k_{600}=a(gS[\frac{1}{n} (A_{0}+dA)^{2/3} W^{-2/3} S^{1/2}])^{b}$
-	  d. $k_{600}(\frac{1}{n}(A_0+dA)^{2/3})^{-b}=a(g W^{2/3}S^{3/2})^{b}$
-	    i. Model parameters on left-hand side (need priors), data on right-hand side
-	    ii. priors on k600, n, A0
-	      -geoBAM  Brinkerhoff etal 2020
-	      -k600 priors obtained similarly using Ulseth etal 2019
-3) For the sake of validating the algorithm, we assume that the scaling model is ‘truth’ and see if we can infer these values using only SWOT observables (river width and height)
+	  a. $k_{600}=a(gS[\frac{1}{n} (A_{0}+dA)^{2/3} W^{-2/3} S^{1/2}])^{b}$
+	  b. $k_{600}(\frac{1}{n}(A_0+dA)^{2/3})^{-b}=a(g W^{2/3}S^{3/2})^{b}$
+	    i.
+Model parameters are now on left-hand side (i.e.
+need priors) and the data (i.e.
+SWOT observables) are on right-hand side.
+This allows us to, given the constant rhs value, sample from the left in the same manner that a traditional Bayesian sampling model samples from a set of model parameters (see Hagemann etal 2017 & Durand etal 2014).
+Therefore, the 'model parameters' are k600, n, and A0.
+In a Bayesian framework, these need priors assigned to them.
+	    ii. Brinkerhoff etal 2020 priors are used for n and A0. k600 priors are obtained similarly using Ulseth etal 2019 as the training data. <b> ELABORATE </b>
+3) For the sake of validating the algorithm, we assume that the scaling model is ‘truth’ and see if we can infer these values using only SWOT observables (river width and height/slope).
     a.
 Therefore, in the validation setup, posterior uncertainty is only Manning’s uncertainty (0.25) and not the total model uncertainty obtained in the MC simulations earlier.
-However, this total uncertainity would be implemented if running the algorithm for basic science objectives.
+However, this total uncertainity would be implemented if running the algorithm for general science objectives.
     b.
@@
 
+4.
 Below are boxplots of performance metrics for the 32 rivers, i.e.
@@
 NRMSE, RRMSE, and rBIAS are optimally zero while KGE is optimally 1.
-A KGE greater than -0.41 indicates our model outperforms a mean estimate, akin to an NSE of 0 for the hydroglogists out there (Knoben etal 2019).
+A KGE greater than -0.41 indicates our model outperforms a mean estimate, akin to an NSE of 0 for the hydrologists out there (Knoben etal 2019).
+Also see Knoben et al (2019) for why NSE completely blows up for k600.
+    a. The result is great KGEs, NRMSEs, and RRMSEs (espically for aboslutely no in situ information) however a pretty high rBIAS. This bias is borne out in the scatterplot earlier.
 
@@
 
-The result is great KGEs, NRMSEs, and RRMSEs (espically for aboslutely no in situ information) however a pretty high rBIAS. This bias is borne out in the scatterplot earlier.
+5.
+Below are timeseries plots of observed versus modeled k600 for three rivers randomly sampled from the three tertiles of the by-river kge scores.
+a) is a 'good' river, b is a 'reasonabley performing' river, and c is a 'bad' performing river.
+
+![**Figure 6** *Pepsi validation timeseries for three rivers, where the model results include the posterior means and 95% CIs: a) randomly selected from upper tertile of kge scores b) randomly selected from middle tertile, c) randomly selected from worse tertile*](outputs//validation//validation_timeseries.jpg)
 

### 2020-12-07 main.R

@@ -5,9 +5,11 @@
 source("scripts//k600_model.R")
 rm(list = ls())
 print('k model done')
+
 source("scripts//monte_carlo_analysis.R")
 rm(list = ls())
 print('MC simulations done')
+
 source("scripts//Figures_pepsi.R")
 rm(list = ls())
 print('Validation figures done. Hopefully you had the validation results stored in the project!!')

### 2020-12-07 scripts/Figures_pepsi.R

@@ -3,10 +3,9 @@
 #Produce figures for pepsi validation for RSK600 algorithm
 
 library(tidyr)
-library(ggplot2)
+library(tidyverse)
 library(cowplot)
 library(RColorBrewer)
-library(dplyr)
 library(colorspace)
 library(ggtext)
 theme_set(theme_cowplot())
@@ -26,7 +25,7 @@ valPlot <- ggplot(full_output, aes(x=(kobs), y=(kest_mean))) +
  # geom_point(size=5, pch=21, color='black', fill='#1b9e77') +
   geom_pointrange(aes(ymin = kest_low, ymax = kest_high), fatten=10, fill='#1b9e77', pch=21, color='black') +
   geom_smooth(size=2, color='black', method='lm', se=F)+
-  xlab('Observed k [m/dy]') +
+  xlab('Scaling Model k [m/dy]') +
   ylab('Remotely Sensed k [m/dy]') +
     scale_y_log10(
       breaks = scales::trans_breaks("log10", function(x) 10^x),
@@ -67,50 +66,53 @@ plotSWOTreaches <- ggplot(plot_stats, aes(x=key, y=value, fill=key)) +
   ylab('Value') +
   coord_cartesian(ylim = c(-1,1))+
   scale_fill_brewer(palette = 'Dark2', name='') +
-  theme(legend.position = 'none')
+  theme(legend.position = "none",
+        axis.text=element_text(size=20),
+        axis.title=element_text(size=24,face="bold"),
+        legend.text = element_text(size=17),
+        legend.title = element_text(size=17, face='bold'))
 ggsave('outputs//validation//validation_by_river.jpg', plotSWOTreaches, width=8, height=8)
 
-# 
+ 
 # #example timeseries plot------------------------------------------------------------------------------------
-# set.seed(213)
-# 
-# kge_bins <- quantile(stats_by_reach$kge, c(0.33, 0.66))
-# kge_bad <- filter(stats_by_reach$kge <= kge_bins[1]) %>% select(SWOTreach)
-# kge_good <- filter(stats_by_reach, test == 2 & kge >= kge_bins[2]) %>% select(SWOTreach)
-# kge_eh <- filter(stats_by_reach, test == 2 & kge > kge_bins[1] & kge < kge_bins[2]) %>% select(SWOTreach)
-# 
[…]

### 2020-12-07 Outline.Rmd
Figures update

@@
 	  b. Small rivers had to be broken into high/low slope to maintain all linear regression assumptions.
-
-![**Figure 1** *eD~k600 scaling by river width/slope. This is how we are able to scale k600 using just SWOT observables*](outputs//k600//width_eD_theory.jpg)
-
 4) To test this, we then combine these five scaling functions into a single 'rule-based' regression model, train it on 80% of the data and validate it on the remaining 20% (below).
@@
 
-![**Figure 2** *Model validation on 20% of data, withheld from training*](outputs//k600//validation.jpg)
+![**Figure 2** *a) eD~k600 scaling by river width/slope.
+This is how we are able to scale k600 using just SWOT observables.
+b) Overall model validation on 20% of data, withheld from training*](outputs//k600//validation.jpg)
 
@@
 
-![**Figure 5** *Pepsi validation by river*](outputs//validation//validation_by_river.jpg)
-
-5.
-Below are timeseries plots of observed versus modeled k600 for three rivers randomly sampled from the three tertiles of the by-river kge scores.
-a) is a 'good' river, b is a 'reasonabley performing' river, and c is a 'bad' performing river.
-
-![**Figure 6** *Pepsi validation timeseries for three rivers, where the model results include the posterior means and 95% CIs: a) randomly selected from upper tertile of kge scores b) randomly selected from middle tertile, c) randomly selected from worse tertile*](outputs//validation//validation_timeseries.jpg)
+![**Figure 5** *a) Performance metrics by river.
+b)-d): validation timeseries for three rivers, where the model results include the posterior means and 95% CIs: b) randomly selected from upper tertile of kge scores c) randomly selected from middle tertile, d) randomly selected from worse tertile*](outputs//validation//validation_by_river.jpg)
 

### 2020-12-07 scripts/Figures_pepsi.R

@@ -71,8 +71,6 @@ plotSWOTreaches <- ggplot(plot_stats, aes(x=key, y=value, fill=key)) +
         axis.title=element_text(size=24,face="bold"),
         legend.text = element_text(size=17),
         legend.title = element_text(size=17, face='bold'))
-ggsave('outputs//validation//validation_by_river.jpg', plotSWOTreaches, width=8, height=8)
-
  
 # #example timeseries plot------------------------------------------------------------------------------------
 set.seed(215)
@@ -114,5 +112,8 @@ goodRiverPlot <- ggplot(goodRiver, aes(x=time, y=value, color=key)) + #model
 # extract the legend from one of the plots
 legend <- get_legend(  goodRiverPlot + theme(legend.box.margin = margin(0, 0, 0, 100)))
 
-timeseriesPlot <- plot_grid(goodRiverPlot + theme(legend.position = 'none'), ehRiverPlot, badRiverPlot, legend, ncol=2, labels=c('a','b','c',NA), label_size = 18)
-ggsave('outputs//validation//validation_timeseries.jpg', timeseriesPlot, width=10, height=8)
+timeseriesPlot <- plot_grid(goodRiverPlot + theme(legend.position = 'none'), ehRiverPlot, badRiverPlot, legend, ncol=2, labels=c('b','c','d',NA), label_size = 18)
+
+#bring it alllllll together---------------------------
+plot2 <- plot_grid(plotSWOTreaches, timeseriesPlot, ncol=2, labels=c('a', NA), label_size = 18)
+ggsave('outputs//validation//validation_by_river.jpg', plot2, width=14, height=8)

### 2020-12-07 scripts/k600_model.R

@@ -81,7 +81,6 @@ test_vali <- ggplot(data_test, aes(x=(k600), y=(k600_pred), color=data)) +
     breaks = scales::trans_breaks("log10", function(x) 10^x),
     labels = scales::trans_format("log10", scales::math_format(10^.x))
   )
-ggsave('outputs//k600//validation.jpg', test_vali, width=8, height=5)
 
 #retrain model on all data for actual implementation in BIGEER algorithm (OOBtrain/test split)----------------------------------------
 theory_plot <- ggplot(data, aes(x=(eD), y=(k600), color=factor(widthRegime))) +
@@ -89,7 +88,7 @@ theory_plot <- ggplot(data, aes(x=(eD), y=(k600), color=factor(widthRegime))) +
   geom_smooth(method='lm', size=3, se=F) +
   xlab('eD [m2/s3]') +
   ylab('k600 [m/dy]') +
-  scale_color_discrete_qualitative(palette = 'Harmonic', name='River Width [m]', label=c('0-10\n(slope < 0.05)', '0-10\n(slope > 0.05)','10-50', '50-100', '100+'))+
+  scale_color_discrete_qualitative(palette = 'Harmonic', name='River Width [m]', label=c('0-10 (slope < 0.05)', '0-10 (slope > 0.05)','10-50', '50-100', '100+'))+
   scale_y_log10(
     breaks = scales::trans_breaks("log10", function(x) 10^x),
     labels = scales::trans_format("log10", scales::math_format(10^.x))
@@ -97,9 +96,15 @@ theory_plot <- ggplot(data, aes(x=(eD), y=(k600), color=factor(widthRegime))) +
   scale_x_log10(
     breaks = scales::trans_breaks("log10", function(x) 10^x),
     labels = scales::trans_format("log10", scales::math_format(10^.x))
-  ) +
-  theme(legend.position = 'bottom')
-ggsave('outputs//k600//width_eD_theory.jpg', theory_plot, width=8, height=5)
+  )
+
+#plot these two together--------------------------------
+legend1 <- get_legend(theory_plot + theme(legend.box.margin = margin(0, 0, 0, 50)))
+legend2 <- get_legend(test_vali + theme(legend.box.margin = margin(0, 0, 0, 50)))
+
+
[…]

### 2020-12-09 Outline.Rmd
new model runs

@@
 ---
-title: "Remote sensing the gas transfer velocity using SWOT observations"
+title: "Remote sensing the gas transfer velocity using SWOT observations and no in situ information"
 author: "Craig Brinkerhoff"
@@
     a.	Empirical scaling functions have been developed to use hydraulics to model k600 for entire river systems but are prone to significant uncertainties (Raymond et al. 2012)
-    b.	They are also fundamentally dependent on modeled or measured flow velocities which themselves are prone to significant uncertainties (or poorly informed hydraulic geometry models…)
-    c.	What is needed is a simple way to directly observe the gas transfer velocity at scale and high temporal resolution without needing explicit flow velocity estimates.
-We argue that the upcoming NASA SWOT mission provides exactly the data required to make this a reality and remotely sense (RS) k600 at unprecedented temporal resolution
+5) Further, in situ data for hydraulics is often unavaible (esp. in remote rivers) and so k models are at the mercy of modeled hydraulics, themselves highly uncertain.
+6)	What is needed is a simple way to observe the gas transfer velocity at large scales and high temporal resolution without relying on situ flow velocity data or modeled velocities.
+We argue that the upcoming SWOT mission provides exactly the data required to make this a reality (at least in large rivers) and remotely sense (RS) k600 as a function of river hydraulics.
 
@@
 ## Goal 2
-1) We propose the BIGER (Bayesian Inference/Inversion of Gas Evasion Rate) algorithm to remotely sense k600 from river width/height alone.
+1) We propose the BIGER (Bayesian Inference/Inversion of the Gas Evasion Rate) algorithm to remotely sense k600 from river width/height alone.
 2) We implement the k600~eD model from Q1 as a Bayesian likelihood function following methods developed for ungauged RSQ (Hagemann et al.
@@
 
-![**Figure 4** *Validation of remote sensing algorithm for 32 rivers with 10day sampling intervals*](outputs//validation//validation.jpg)
+![**Figure 4** *Validation of remote sensing algorithm for 32 rivers with 10day sampling intervals.
+Black bars are 95% CIs for the modeled values.
+Black line is linear regression and dashed grey line is 1:1 line.*](outputs//validation//validation.jpg)
 
@@
 NRMSE, RRMSE, and rBIAS are optimally zero while KGE is optimally 1.
-A KGE greater than -0.41 indicates our model outperforms a mean estimate, akin to an NSE of 0 for the hydrologists out there (Knoben etal 2019).
+A KGE greater than -0.41 indicates our model outperforms a mean estimate, akin to an NSE of 0 (Knoben etal 2019).
 Also see Knoben et al (2019) for why NSE completely blows up for k600.
-    a. The result is great KGEs, NRMSEs, and RRMSEs (espically for aboslutely no in situ information) however a pretty high rBIAS. This bias is borne out in the scatterplot earlier.
 
@@
 Who knows! Any interesting questions you can think of?
+
+##References
+Durand, M., Neal, J., Rodríguez, E., Andreadis, K.
+M., Smith, L.
+C., & Yoon, Y.
+(2014).
+Estimating reach-averaged discharge for the River Severn from measurements of river water surface elevation and slope.
+Journal of Hydrology, 511, 92–104.
+https://doi.org/10.1016/j.jhydrol.2013.12.050
+
+Engram, M., Walter Anthony, K.
+M., Sachs, T., Kohnert, K., Serafimovich, A., Grosse, G., & Meyer, F.
+J.
+(2020).
+Remote sensing northern lake methane ebullition.
+Nature Climate Change, 1–7.
+https://doi.org/10.1038/s41558-020-0762-8
+
+Hagemann, M.
+W., Gleason, C.
+J., & Durand, M.
+T.
+(2017).
+BAM: Bayesian AMHG-Manning Inference of Discharge Using Remotely Sensed Stream Width, Slope, and Height: BAM FLOW USING STREAM WIDTH SLOPE HEIGHT.
+Water Resources Research, 53(11), 9692–9707.
+https://doi.org/10.1002/2017WR021626
+
+Knoben, W.
+J.
+M., Freer, J.
+E., & Woods, R.
+A.
+(2019).
+Technical note: Inherent benchmark or not?
+Comparing Nash–Sutcliffe and Kling–Gupta efficiency scores.
+Hydrology and Earth System Sciences, 23(10), 4323–4331.
+https://doi.org/10.5194/hess-23-4323-2019

### 2020-12-09 main.R

@@ -10,6 +10,6 @@ source("scripts//monte_carlo_analysis.R")
 rm(list = ls())
 print('MC simulations done')
 
-source("scripts//Figures_pepsi.R")
+source("scripts//validation_figures.R")
 rm(list = ls())
-print('Validation figures done. Hopefully you had the validation results stored in the project!!')
+print('Validation figures done. Hopefully you had the correct validation results stored in the output directory!!')

### 2020-12-09 scripts/runBIGER.R

@@ -1,0 +1,133 @@
+#Creator: Craig Brinkerhoff
+#Date: Fall 2020
+#Description: run BIGER algorithm on pepsi rivers
+
+library(ncdf4)
+library(tidyverse)
+library(BIGER)
+
+#model settings-------------------------------
+g <- 9.8 #m/s2
+mannings_uncertainity <- 0.25 #see monte_carlo_analysis.R for actual uncertainty
+  #For these tests, should assume only error is from manning's equation (0.21), NOT k600 model b/c we are assuming this model is 'true'
+  #In actual implementation, we want posterior uncertainity to reflect both (which is ~0.30 per our Monte Carlo Analysis)
+
+#set working directory----------------------------------------------------------
+setwd('/home/cbrinkerhoff_umass_edu/RS_evasion/BIGER/')
+
+#BAM-stle functions------------------------------------------------------
+#' @param w Matrix of widths
+#' @param h Matrix of heights(FROM MARK)
+calcdA_mat <- function(w, h) {
+  stopifnot(all(dim(w) == dim(h)))
+  dA <- w
+  for (i in 1:nrow(dA)) {
+    dA[i, ] <- calcdA_vec(w[i, ], h[i, ])
+  }
+  
+  dA
+}
+
+#' Calculate partial cross-section area from width and height vectors (time series)
+#' @param w vector of widths
+#' @param h vector of heights(FROM MARK)
+calcdA_vec <- function(w, h) {
+  words <- order(w)
+  warr <- w[words]
+  harr <- h[words]
+  delh <- c(0, diff(harr))
+  delA <- cumsum(warr * delh)
+  dA <- 1:length(w)
+  dA[words] <- delA
+  dA
+}
+
+#load in SWOT observables-------------------------------------------
+files = list.files('/home/cbrinkerhoff_umass_edu/RS_evasion/pepsi_rivers/10day/', pattern="*.nc", full.names = TRUE)
+
+#run model-------------------------------------------------------
+output <- data.frame('river'= NA, 'time'=NA, 'kobs'=NA, 'kest_mean'=NA, 'kest_low'=NA, 'kest_high'=NA, 'kest_sd'=NA)
+for (currPepsi in files[1:length(files)]){
[…]

### 2020-12-18 Outline.Rmd
many updates

@@
 
+## Craig's things to do
+1) Address bias from k600 prior
+2) redo fig 6 with sigma_man not kest_sd
+2) Working on 3rd research question (courtesy of Merritt)
+
+--------------------------------------------------------------------------
+
 ## Background
@@
 5) Further, in situ data for hydraulics is often unavaible (esp. in remote rivers) and so k models are at the mercy of modeled hydraulics, themselves highly uncertain.
-6)	What is needed is a simple way to observe the gas transfer velocity at large scales and high temporal resolution without relying on situ flow velocity data or modeled velocities.
-We argue that the upcoming SWOT mission provides exactly the data required to make this a reality (at least in large rivers) and remotely sense (RS) k600 as a function of river hydraulics.
+6)	We argue that the upcoming SWOT mission will allow for addressing these limitations by remotely sensing (RS) k600 as a function of river hydraulics and, through a Bayesian framework, explicitly quantifying estimate uncertainity from multiple sources.
 
@@
 2)	Build algorithm to run the new empirical model using SWOT observations
-3)	?????????
-    a. Biogeochemists might be interested in a Merritt style test of this algorithm at field-scale (to circumvent velocity measurements…)
-    b. Pepsi2-style comparisons across temporal sampling?
-    c. Some sort of k600 timeseries analysis? But these are all simulated k so I dunno.
-    d. MORE
+3)	Explore the GHG flux implications of using this algorithm versus standard upscaling functions
 
@@
 #### Theory
-1) Raymond 2012 show that k600 can be scaled using hydraulics b/c k600 generally correlates with the turbulent dissipation rate (eD = gSV)
-2) Ulseth et al 2019 show two distinct scaling regimes for k600 with eD using an expanded dataset including steeper streams than those in Raymond 2012.
-3) Here, we take the Ulseth et al.
-2019 dataset and develop a k600~eD scaling function that accounts for this regime shift and the fact that scaling relationships differ as a f(river size).
-BUT, we only use SWOT observables to determine which scaling relationship to use.
-    a. i.e. river width and slope alone must be used to choose which scaling model to use.
-	    i. $k_{600}=a(gSV)^{b}$ where a and b are differentially determined by river size.
-	  b. Small rivers had to be broken into high/low slope to maintain all linear regression assumptions.
-4) To test this, we then combine these five scaling functions into a single 'rule-based' regression model, train it on 80% of the data and validate it on the remaining 20% (below).
-The version ultimately implemented in our algorithm is trained on the entire dataset however.
+1) Raymond 2012 show that k600 can be scaled using hydraulics b/c k600 generally correlates with the water column turbulent dissipation rate ($eD \approx gSV$)
+2) Ulseth et al 2019 show two distinct scaling regimes for k600 with eD using an expanded dataset including steeper streams than those in Raymond etal 2012 (all data used here were collected in the field and/or from previous studies by Ulseth etal 2019)
+3) Here, we extend the Ulseth et al.
+2019 model and develop an RS-able variant of it that continues to account for the regime shift presented in that paper but also uses only SWOT observables to determine which scaling relationship to use (as opposed to using an eD threshold, which can't be known a priori from SWOT).
+The theory behind this rule-based regression if plotted in Figure 1 with the Ulseth etal (2019) dataset.
+    a. i.e. river width and slope alone are used to choose which scaling model to use.
+	    i. $k_{600}=a_i(gSV)^{b_i}$ for parameters a and b and 'characteristic river size thresholds' i (i = [10m, 50m, 100m])
+	  b. Small rivers are further broken into high/low slope to maintain all linear regression assumptions.
 
-![**Figure 2** *a) eD~k600 scaling by river width/slope.
-This is how we are able to scale k600 using just SWOT observables.
-b) Overall model validation on 20% of data, withheld from training*](outputs//k600//validation.jpg)
+![**Figure 1** *Extension of the Ulseth etal 2019 eD~k600 scaling model that is completely RS-able. This is how we are able to scale k600 using just SWOT observables.*](outputs//k600//theory_plot.jpg)
 
-#### Quantifying Uncertainity in Model Estimates
-1) In order to remotely sense k600, we need a way to approximate flow velocity as it is not easily RS-able.
-We use a variant of Manning's equation, THE classic open channel flow depth~velocity relation, to do this (Durand et al.
-2014; Hagemann et al.
-2017 used to do the same thing from getting discharge from SWOT data).
-	  a. $k_{600}=a(gSV)^{b}$ Original Model
-	  b. $k_{600}=a(gS[\frac{1}{n} d^{2/3} S^{1/2}])^{b}$ Classical Manning's equation introduced
-	  c.
-$k_{600}=a(gS[\frac{1}{n} (A_{0}+dA)^{2/3} W^{-2/3} S^{1/2}])^{b}$ Durand et al.
-(2014) formulation of Manning's equation.
-(The handling of n and A0 in a RS framework will be addressed under Goal 2).
+4) We then compare this RS-able extension of the Ulseth model against the Ulseth model itself, as well as the three models from Raymond etal 2012 that use only slope and velocity as predictors (i.e.
+RS-able).
+The Raymond models were retrained on the expanded Ulseth dataset, which includes new data in addition to the Raymond 2012 dataset.
+We see identical predictive ability between the Ulseth model and our RS-able variant of that model.
+We also see that Raymond Eq.
+5 predicts negative k values, which can't be plotted in log-space and highlight the limitations of a linear model for this problem.
+
+![**Figure 2** *Validation of five k600 upscaling models on 20% of the Ulseth etal 2019 dataset witheld for independent testing.
+Raymond Eq.
+5 predicts negative k600 values, which are not plotted here and highlight the limitation of a linear model for this problem.*](outputs//k600//validation.jpg)
+
+#### Quantifying Uncertainity in Scaling Model + Manning's Equation
+1) In order to remotely sense k600, we need a way to approximate average flow velocity as it is not strictly RS-able.
+We use Manning's equation, the classic open channel flow depth~velocity relation, to do this.
+	  a. $k_{600}=a_i(gSV)^{b_i}$ Original Model
+	  b. $k_{600}=a_i(gS[\frac{1}{n} d^{2/3} S^{1/2}])^{b_i}$
 2) We then quantified the total uncertainity implicit in the model output.
 Uncertainty can stem from three sources: 1) error from Manning's equation, 2) error in using a k600~eD scaling model, and 3) error implicit in our remote sensing algorithm.
-This third error source will be explicitly accounted for (via Bayesian inference) in the RS algorithm output, however we also wanted to quantify total uncertainty from the model being used in the RS algorithm, i.e.
+This third error source will be explicitly accounted for (via Bayesian inference) in the RS algorithm output (Goal 2), however we also wanted to quantify total uncertainty from the model being used in the RS algorithm, i.e.
 from the other two sources.
@@
     a.
-8,000 sets of river hydraulic measurements from across the United States were taken from the Brinkerhoff etal (2019) dataset.
-For each one I ran a 10,000 run Monte Carlo simulation (randomly sampling from the three parameters a, b, and "Manning's velocity") and obtained 8,000 different distributions quantifying model estimate uncertainity.
+8,000 sets of river hydraulic measurements from across the United States were taken from the Brinkerhoff etal (2019) dataset (Fig 3c).
+For each one I ran a 10,000 run Monte Carlo simulation (randomly sampling from the three parameters a, b, and "Manning's velocity") and obtained 8,000 different distributions (Fig 3b for some examples) quantifying model estimate uncertainity.
 Log uncertainity for the "Manning's velocity"" was assigned 0.25 following Hagemann et al.
@@
 
+The average lnSD (i.e. k600 estimate uncertainity) is 1.31 (~3.7 m/dy).
+
 ## Goal 2
-1) We propose the BIGER (Bayesian Inference/Inversion of the Gas Evasion Rate) algorithm to remotely sense k600 from river width/height alone.
-2) We implement the k600~eD model from Q1 as a Bayesian likelihood function following methods developed for ungauged RSQ (Hagemann et al.
-2017).
-Below is the derivation of the likelihood function used in the Bayesian inference model.
-	  a. $k_{600}=a(gS[\frac{1}{n} (A_{0}+dA)^{2/3} W^{-2/3} S^{1/2}])^{b}$
-	  b. $k_{600}(\frac{1}{n}(A_0+dA)^{2/3})^{-b}=a(g W^{2/3}S^{3/2})^{b}$
+1) We propose the BIKER algorithm (<b>B</b>ayesian <b>I</b>nference/<b>I</b>nversion of the <b>k</b>600 <b>E</b>vasion <b>R</b>ate) to remotely sense k600 from river width/height alone.
+2) First, we use a variant of the Manning's equation that is defined using the SWOT observable variables width W, slope S, and change in channel area dA (the latter two are implicitly observed through river height H).
+We draw on Durand et al.
+2014 and Hagemann et al.
+2017 who do something similar to infer discharge from SWOT data.
+	  c. $k_{600}=a_i(gS[\frac{1}{n} (A_{0}+dA)^{2/3} W^{-2/3} S^{1/2}])^{b_i}$ Hagemann etal (2017) formulation of Manning's equation.
+3) We then implement this as a Bayesian likelihood function following methods developed for ungauged RSQ (Hagemann et al. 2017).
+	  a. $k_{600}=a_i(gS[\frac{1}{n} (A_{0}+dA)^{2/3} W^{-2/3} S^{1/2}])^{b_i}$
+	  b. $k_{600}(\frac{1}{n}(A_0+dA)^{2/3})^{-b_i}=a_i(g W^{-2/3}S^{3/2})^{b_i}$
 	    i.
@@
 Therefore, the 'model parameters' are k600, n, and A0.
-In a Bayesian framework, these need priors assigned to them.
-	    ii. Brinkerhoff etal 2020 priors are used for n and A0. k600 priors are obtained similarly using Ulseth etal 2019 as the training data. <b> ELABORATE </b>
-3) For the sake of validating the algorithm, we assume that the scaling model is ‘truth’ and see if we can infer these values using only SWOT observables (river width and height/slope).
+In a Bayesian framework, these need priors assigned to them, each defined by a set of prior hyperparameters.
+	    ii.
+Brinkerhoff etal 2020 prior hyperparameters are used for n and A0.
+k600 prior is calculatd using slope and trained onUlseth data.
+<b>The key here is that all hyperparameters are assigned using only river width and/or slope, meaning that absoultely no a priori information is needed about the river.
+Thus, no field measurements are required, no velocities need to be modeled, etc.</b>
+4) For the sake of validating the algorithm, we assume that the scaling model is ‘truth’ and see if we can infer these values using only SWOT observables (river width and height/slope).
+<b> This is thus largely a test of whether we can invert Manning's equation and define a reasoinable k600 prior.
+</b>
     a.
 Therefore, in the validation setup, posterior uncertainty is only Manning’s uncertainty (0.25) and not the total model uncertainty obtained in the MC simulations earlier.
-However, this total uncertainity would be implemented if running the algorithm for general science objectives.
+However, this total uncertainity would be used in practice.
     b.
-We validated on 10-day pepsi 2 data.
+We validated on SWOT simulated rivers, every 11 days, replicating the average SWOT overpass.
 Plotted are means and 95% CIs of the posterior k600 values versus the ‘observed’ k600 calculated from observed velocity and slope for all timesteps across all 32 rivers
 
-![**Figure 4** *Validation of remote sensing algorithm for 32 rivers with 10day sampling intervals.
+![**Figure 4** *Validation of remote sensing algorithm for 22 rivers with 11 day sampling intervals.
 Black bars are 95% CIs for the modeled values.
@@
 
+6.
+We also compare a priori model uncertainity (i.e.
+from Manning's equation) against posterior model uncertainity to see how much uncertianity is added by using only observed W and H, i.e.
+a test of the manifestation of the priors and equifinality.
+
+THIS NEEDS TO BE REDONE WITH SIGMA_MAN
+
+![**Figure 6** *A) Histogram of BIKER k600 uncertainities (across all rivers and timesteps) compared against a priori uncertainity from Manning's equation (0.25- dark blue line).
+Red line is the mean of the BIKER uncertainities (0.32).
+B) The same plot, but mean uncertaintities for the 22 rivers (red line at 0.33).*](outputs//validation//validation_uncertainity.jpg)
+
+I'm still trying to figure out why there's like five rivers that have crazy increased uncertianity (causing the bimodal distributions).
+
 ##Goal 3
-Who knows! Any interesting questions you can think of?
+#### Current Approach
+1) Used the 22 SWOT rivers in conjunction with biweekly in-stream CO2 data for a similarly large river (Ohio River- Beauliu etal 2012).
+<b> NOTE: </b> there are only 29 CO2 values in this dataset so FOR THIS TEST ONLY we ignored all timesteps beyond 29 (only a few rivers go to 34, but its fine for this test)
+    a. We assume these CO2 data aligned with the SWOT observations, which isn't true but is fine for this test as we care only about relative FCO2 differences
+    b. Also, keeping CO2 data constant across rivers allows us to explore the sentivity of the k model only
+
+2) We calculated k600 using three different methods
+    a. Ulseth etal 2019 model
+    b. Raymond etal. 2012 model that is often used in global upscaling (Raymond etal. 2012; Lauerwald etal 2015; Liu etal in review)
+    c. BIKER algorithm.
+
+3) kco2 was backed out using a Schmidt number for CO2 and 25C temp following Raymond etal 2012 and Wanninkof (1992)
+
+4) air-side pCO2 was assumed to be 390 uatm
+
+#### Current Results
+Figure 6 shows the resulting timeseries of FCO2 [$g/m2*dy$]).
+Basically, we reproduce scaling model dynamics and magntiude, but there is some scatter around this line.
+Basically, works in some rivers and not others.
+Need to parse this out.
+
+![**Figure 7** *Left: FCO2 from BIKER versus using observed average flow velocity for all timesteps for 22 SWOT rivers.
+right: timeseries plots for three example rivers.*](outputs//flux_implications//FCO2_plot.jpg)
 
 ##References
+Beaulieu, J.
+J., Shuster, W.
+D., & Rebholz, J.
+A.
+(2012).
+Controls on gas transfer velocities in a large river.
+Journal of Geophysical Research: Biogeosciences, 117(G2).
+https://doi.org/10.1029/2011JG001794
+
+Brinkerhoff, C.
+B., Gleason, C.
+J., Feng, D., & Lin, P.
+(2020).
+Constraining Remote River Discharge Estimation Using Reach-Scale Geomorphology.
+Water Resources Research, 56(11), e2020WR027949.
+https://doi.org/10.1029/2020WR027949
+
 Durand, M., Neal, J., Rodríguez, E., Andreadis, K.
@@
 https://doi.org/10.1038/s41558-020-0762-8
+
+Frasson, R.
+P.
+de M., Durand, M.
+T., & Rodríguez, E.
+(2019).
+Compilation of hydraulic models for the study of the spatial averaging on flow laws [Data set].
+Zenodo.
+https://doi.org/10.5281/zenodo.3463541
 
@@
 https://doi.org/10.5194/hess-23-4323-2019
+
+Lauerwald, R., Laruelle, G.
+G., Hartmann, J., Ciais, P., & Regnier, P.
+A.
+G.
+(2015).
+Spatial patterns in CO2 evasion from the global river network.
+Global Biogeochemical Cycles, 29(5), 534–554.
+https://doi.org/10.1002/2014GB004941
+
+Raymond, P.
+A., Zappa, C.
+J., Butman, D., Bott, T.
+L., Potter, J., Mulholland, P., et al.
+(2012).
+Scaling the gas transfer velocity and hydraulic geometry in streams and small rivers.
+Limnology and Oceanography: Fluids and Environments, 2(1), 41–53.
+https://doi.org/10.1215/21573689-1597669
+
+Raymond, P.
+A., Hartmann, J., Lauerwald, R., Sobek, S., McDonald, C., Hoover, M., et al.
+(2013).
+Global carbon dioxide emissions from inland waters.
+Nature, 503(7476), 355–359.
+https://doi.org/10.1038/nature12760
+
+Rodríguez, E., Durand, M., & Frasson, R. P. de M. (2020). Observing Rivers With Varying Spatial Scales. Water Resources Research, 56(9), e2019WR026476. https://doi.org/10.1029/2019WR026476
+
+Ulseth, A.
+J., Hall, R.
+O., Boix Canadell, M., Madinger, H.
+L., Niayifar, A., & Battin, T.
+J.
+(2019).
+Distinct air–water gas exchange regimes in low- and high-energy streams.
+Nature Geoscience, 12(4), 259–263.
+https://doi.org/10.1038/s41561-019-0324-8
+
+Wanninkhof, R. (1992). Relationship between wind speed and gas exchange over the ocean. Journal of Geophysical Research: Oceans, 97(C5), 7373–7382. https://doi.org/10.1029/92JC00188

### 2020-12-18 main.R

@@ -2,14 +2,21 @@
 #Does not actually implement the algorithm (i.e. runBIGEER.R)
   #but the results of those implmentations are included in the inputs directory.
 
+#k600 model development
 source("scripts//k600_model.R")
 rm(list = ls())
 print('k model done')
 
+#Uncertainity analysis
 source("scripts//monte_carlo_analysis.R")
 rm(list = ls())
 print('MC simulations done')
 
+#Produce validation figures
 source("scripts//validation_figures.R")
 rm(list = ls())
 print('Validation figures done. Hopefully you had the correct validation results stored in the output directory!!')
+
+#Knit markdown notes
+rmarkdown::render("Outline.Rmd")
+print('Project report, or Outline.html, is generated.')

### 2020-12-18 scripts/k600_model.R

@@ -8,6 +8,7 @@ library(Metrics)
 library(colorspace)
 library(ggfortify)
 library(ggtext)
+library(segmented)
 
 theme_set(theme_cowplot())
 options(scipen = 999)
@@ -22,19 +23,28 @@ data <- ulseth_data
 data <- filter(data, is.na(width)==0) #filter out no width measurements
 
 data$eD <- 9.8 * data$slope * data$Vms #water column turbulent dissipation rate m2/s3
-data$eS <- sqrt(9.8*data$depth*data$slope) / data$depth #turbulent dissipation rate from bed friction
 
 #width regime models for k600-----------------------------------------------------------------------
   #Width regimes are necessary because we can't remotely sense eD directly (hence the entire Bayesian algorithm!)
     #Thus, we can't know the eD threhsold of 0.02 a priori. BUT, we do know width a priori
- data$widthRegime <- ifelse(data$width < 10 & data$slope < 0.05, '1',
+data$widthRegime <- ifelse(data$width < 10 & data$slope < 0.05, '1',
                             ifelse(data$width < 10 & data$slope > 0.05, '1.5',
                                 ifelse(data$width < 50, '2',
                                    ifelse(data$width < 100, '3','4'))))
 
+data$log_eD <- log(data$eD)
+data$log_k600 <- log(data$k600)
+data$log_slope <- log(data$slope)
+data$raymond <- data$slope*data$Vms #for raymonds 2012 model
+
+lmSlope <- lm(log_k600~log_slope, data=data)
+segLmSlope <- segmented(lmSlope, npsi=1) #r2 0.55
+summary(segLmSlope)
+plot(segLmSlope)
+
 #Check linearity assumptions within each width class. If you cycle through the different classes, everything holds
-lm <- lm(log(k600)~log(eD), data=filter(data, widthRegime==1))
-autoplot(lm)
+#lm <- lm(log_k600~log_eD, data=filter(data, widthRegime==1))
+#autoplot(lm)
 
 #80/20 training and test split----------------------------------------
   #This is just for validation of method. In implementation I train model on entire dataset (see below)
@@ -44,6 +54,16 @@ sample = sample.split(data$k600, SplitRatio = .80)
[…]

### 2021-01-28 manuscript_v1.Rmd
big push

---
title: "Remote sensing the gas exchange velocity and carbon efflux using the SWOT Satellite"
author: "CB Brinkerhoff, CJ Gleason, PA Raymond, MH Harlan"
date: "`r Sys.Date()`"
output:
  word_document: 
      reference_docx: word-styles-reference.docx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##1 Introduction
Greenhouse gas (GHG) emissions from rivers and lakes are significant (Cole et al. 2007). Rivers are usually supersaturated with dissolved GHGs and present a net flux of these gases from water to air (Cole & Caraco 2001; Jones et al. 2003). This flux $[M/L^2T]$ is calculated, with in situ knowledge of the gas concentration gradient between the water $C_{water} [M/L^3]$ and the air $C_{air} [M/L^3]$ and the gas exchange velocity $k [L/T]$, as equation 1.

$Flux=k(C_{water}-C_{air})$ (1)

k is largely a function of water column turbulence, at least in fluvial settings. This is because k scales with diffusivity as normalized by the boundary-layer thickness, and in rivers turbulence defines the thickness of the boundary-layer (Hall & Ulseth, 2020 for a thorough review of gas exchange velocity in rivers). Due to its reliance on diffusivity (as well as kinematic viscosity), k is defined differently for different GHGs (Raymond et al. 2012; Wanninkof 1992). It is also often measured using different gases than those being studied, so a variant of k that is normalized to a Schmidt number of 600 ($k_{600} [L/T]$) is generally used to be GHG and temperature agnostic (Hall & Ulseth, 2020). This is synonymous with a k for carbon dioxide ($k_{CO2}$) at 20 degrees Celsius.

It is infeasible to measure $k_{600}$ directly when working at the river network scale and so ‘upscaling functions’ are used that predict $k_{600}$ from readily available river geomorphology (e.g. O’Connor & Dobbins, 1958; Palumbo & Brown, 2014; Raymond et al. 2012). Following the theory above, many upscaling functions are based on the basic concept that k is empirically correlated with the turbulent energy dissipation e at the air-water interface (Zappa et al. 2007). This is convenient for upscaling because in rivers and streams e can be equated to the turbulent energy dissipation rate eD $[L^2/T^3]$ (Ulseth et al. 2019; Raymond et al. 2012), which is estimated using easily calculated hydraulic parameters (eD≈gSV where g is gravitational acceleration, S is channel slope, and V is average flow velocity- Tsivoglou & Neal, 1976). It has been observed that this scaling relationship is particularly strong in higher energy streams where bubble-mediated gas exchange occurs, though the general scaling function holds across all rivers and streams (Ulseth et al. 2019). Despite reasonable predictive accuracy, these upscaling models suffer from residual variation indicating large predictive error (Wallin et al. 2011). Still, these are the current best options for estimating $k_{600}$ in rivers. It has been suggested that exploring at-a-station hydraulic variation (and thus $k_{600}$ variation) is a promising approach to reducing the residual variation in upscaling models (Hall & Ulseth, 2020).

Because channel slope is readily available in any hydrographic data product, most efforts to upscale $k_{600}$ to river networks are limited by estimates of the reach-averaged flow velocity of the river channel. Therefore, network-scale GHG flux estimates exist in some part at the mercy of the quality of average flow velocity data. Ideally, a worker can directly measure velocity at some permanently or temporarily gauged river in their system and build velocity~discharge rating curves to upscale to the entire river network. Rating curves are also referred to as hydraulic geometry (HG) models: simple power law relations between hydraulic properties and discharge (Leopold & Maddock, 1953). However, this is not feasible in ungauged basins (whether physically or politically ungauged- Gleason & Durand, 2020) where no in situ hydraulic information is available and fieldwork is impractical. Thus, in ungauged settings, present understandings of GHG fluxes from rivers are likely limited by our access to velocity and/or discharge data.

To circumvent this problem in ungauged basins/at massive spatial scales, researchers have used ‘global-scope’ HG models that use a single set of model parameters trained on large hydraulics datasets. This is the approach used in upscaling average flow velocity to continental/global river networks and has proven quite useful in those settings (e.g. Horgby et al. 2019; Borges et al. 2015; Lauerwald et al. 2015; Raymond et al. 2013). However, it has long been established that HG parameters exhibit high spatial variability across landscapes and vary significantly from river to river (Park, 1977). Because of this, a large body of geomorphology work has attempted to parse out process-based explanations for HG parameters (e.g. Dingman 2007; Parker, et al. 2007; Singh, 2003; Ferguson, 1986) and an open research question is how to best prescribe a set of HG parameters to a given river (particularly at the global scale). A potential alternative to this approach, in ungauged basins at least, is to directly estimate a river’s hydraulic properties from remote sensing (RS) data.

The problem of hydraulic estimation in ungauged basins is itself a burgeoning subfield within fluvial hydrology and an analogue to the problem presented here is the ungauged remote sensing of river discharge (RSQ- see Gleason & Durand, 2020 for a thorough review). A specific subset of these ungauged RSQ methods are termed ‘McFLIs’ or ‘Mass Conserved Flow Law Inversion’ algorithms (e.g. Andreadis et al. 2020; Brinkerhoff et al. 2020a; Durand et al. 2014; Garambois & Monnier, 2015; Hagemann et al. 2017). They use basic geomorphic theories, rather than full hydrologic and/or hydraulic models, and the concept of ‘prior river knowledge’ to estimate discharge from RS data where not all hydraulic terms are RS-able (Gleason & Durand, 2020; Gleason et al. 2017). McFLIs have largely been developed in the context of the NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) mission, which will launch in 2022 and provide the world’s first global measurements of water surface extent and elevation at novel temporal resolutions (Biancamaria et al. 2016). On average SWOT will sample rivers every 11 days and provide a unique opportunity to measure river and lake hydraulics at the global scale. While the raw performance of McFLIs is lower than a more sophisticated approach, they are globally flexible, easily implementable in any river that SWOT can sample, and improve our understandings of ungauged rivers when little to no information was previously available (Brinkerhoff et al. 2020a; Durand et al. 2016). Recently, McFLIs have also shown promise in providing additional and beneficial information to traditional hydrologic modeling via data assimilation (Ishitsuka et al. 2020).

McFLIs often employ Bayesian inference. Bayesian techniques have been previously used to estimate $k_{600}$ concurrently with stream metabolism from dissolved oxygen (O2) datasets (Appling et al. 2018; Grace et al. 2015; Holtgrieve et al. 2010). While this approach requires detailed field data and is thus not applicable to $k_{600}$ estimation in ungauged rivers, it does suggest that Bayesian inference could be useful for estimating $k_{600}$ when parameter equifinality is a problem (at it is with McFLIs).

Within this context, we sought to explore estimating $k_{600}$ in ungauged basins with no in situ information. We hypothesize that remote sensing $k_{600}$ is possible by combining 1) hydraulic geometry theory, 2) $k_{600}$ upscaling theory, 3) Bayesian inference (in a manner similar to McFLI), and 4) SWOT data. More specifically, this chapter aims to answer the following questions:
	1) Is upscaling the gas exchange velocity with no a priori knowledge of eD possible?
	2) Is ungauged RS of gas exchange velocity possible?
	3) How might this algorithm influence upscaling of riverine carbon efflux?

To answer these questions, we build a RS algorithm that ingests SWOT data and produces $k_{600}$ estimates (with fully quantified uncertainty within a Bayesian framework) that requires no in situ information about a river’s hydraulic properties. We validate the algorithm on simulated SWOT data for 22 rivers. Finally, we compare our algorithm against established methods: we use previously published carbon dioxide ($CO_2$) data to compare the bulk carbon efflux (via $CO_2$ evasion) from the 22 rivers as calculated using both our new algorithm and established velocity rating curves and in situ discharge records.

##2 Methods
We name the RS of $k_{600}$ algorithm BIKER, or the ‘Bayesian Inversion/Inference of the $k_{600}$ Rate’ algorithm. To build BIKER, two steps are taken: first, we develop a $k_{600}$ upscaling model that does not rely on a priori knowledge of eD (section 2.1) and then we use that upscaling model to build the actual RS algorithm (section 2.2). Following these methods, the validation setup is described (section 2.3), as is the comparison of estimated bulk carbon effluxes from a suite of average flow velocity models (section 2.4). A flowchart detailing the complete workflow is included (Figure 1).

![Figure 1. Flowchart of methodology developed in this chapter. ‘Extended Upscaling Model’ is detailed in section 2.1. ‘BIKER’ is detailed in section 2.2 and ‘Flux Estimation’ is detailed in section 2.4. See section 2.3 for the validation setup.](outputs/flowchart.jpg)

###2.1 Developing an RS-able $k_{600}$ upscaling model
Raymond et al. (2012) first used river hydraulics and the theoretical relationship between the turbulent dissipation rate (eD) and gas exchange to upscale $k_{600}$ from average flow velocity and channel slope, extending earlier work predicting the reaeration coefficient (e.g. O’Connor & Dobbins, 1958; Tsivoglou & Neal, 1976). Later, Ulseth et al. (2019) built upon Raymond et al. (2012)’s work by expanding their dataset to include measurements from steeper rivers and finding that two distinct upscaling regimes exist in low and high energy rivers. More specifically, Ulseth et al. (2019) scale $k_{600}$ using eD, where the resulting upscaling model parameters are significantly different whether eD is high or low. Their upscaling function is reprinted as equation 2.

$ln(k_{600})=3.10 + 0.35ln(eD) if eD < 0.02$  (2)

$ln(k_{600})=6.43+1.18ln(eD) if eD > 0.02$

While equation 2 would be ideal for the purposes outlined for this manuscript and provides best-to-date predictive accuracy for upscaling $k_{600}$, it requires an a priori knowledge of eD to assign its differential model parameters. We do not have this a priori knowledge in an ungauged setting and only have access to SWOT-observed river widths and heights (and therefore water surface slopes). Thus, we need to extend the Ulseth et al. (2019) model to assign the two model parameters using only RS-able hydraulics. To do this, we use the basic premise that the $k_{600}\sim eD$ scaling relationship is different for different river sizes. We use river width as a proxy for river size and train different upscaling functions for different groups of data with similar river sizes (Figure 2, trained on the Ulseth et al. 2019 dataset). In doing so, we find that narrow rivers (< 10m wide) need to be further broken into low energy (slope < 0.05) and high energy (slope > 0.05) regimes in order to meet the assumptions necessary to fit a linear regression model (specifically, normally distributed residuals of the predictions).

![Figure 2: Theoretical basis for the RS-able extension of the Ulseth et al. (2019) k600 upscaling model. Data from Ulseth et al. (2019).](outputs/k600/theory_plot.jpg)

We combined these five regression models into a ‘rule-based regression model’ (equation 3) for river size i and model parameters a and b. Per Figure 2, there are five possible sets of a and b parameters, as defined for rivers 10-50m wide, 50-100m wide, 100+m wide, low energy rivers 0-10m wide (slope < 0.05), and high energy rivers 0-10m wide (slope > 0.05).

$k_{600}=a_i(eD)^{b_i}$ (3)

This model therefore assigns ai and bi a priori using just river width and slope to define i. Another implication of this extension of the Ulseth et al. (2019) model is that we now use a and b parameters that implicitly reflect the distinct geomorphology of different sized rivers.
Finally, we tested whether this model can reasonably estimate $k_{600}$ relative to the original Ulseth et al. (2019) model, with the goal of at least replicating their performance. We compared it against the various eD-based upscaling models from Raymond et al. (2012) that are commonly used in the literature. This was done by training all upscaling functions on 80% of the Ulseth et al. (2019) data and comparing their performance on the remaining 20% of the data. Because the Raymond et al. (2012) models were originally trained on a subset of this dataset, we deemed it fairer to refit their parameters to this expanded dataset (but keep model structure identical).

###2.2 Developing the BIKER algorithm
With a validated upscaling model that requires no a priori knowledge of eD (results presented in section 3.1) we turn to developing the algorithm to RS $k_{600}$. The approach used here is heavily informed by Hagemann et al. (2017)’s algorithm for ungauged RSQ via Bayesian inference. Consult that paper for the logic behind conceptualizing the equation for discharge as a Bayesian inference problem, which we largely follow here to conceptualize upscaling $k_{600}$ as a Bayesian inference problem.

SWOT will provide river width and height (and thus water surface slope $S_h$) at nodes/cross-sections along a set of river reaches. Thus, our ‘data’ for the Bayesian inference are river width and water surface slope while all other terms are treated as statistical parameters. We also obtain a ‘change in channel area’ from the river height and width and assuming a rectangular channel geometry.

In order to use Bayesian inference to solve for $k_{600}$, we need to derive a sampling model/likelihood function. Following Tsivoglou & Neal (1976), $eD \approx gS_hV$ for unsteady flow conditions (channel slope $S_0 \neq S_h$- see text S1 for this model derivation and why assuming non-uniform flow is convenient for our application). Using this, as well as Manning’s equation to relate average flow velocity to channel geometry, we rewrite equation 3 as equation 4. In equation 4, we use the Manning’s formulation from Hagemann et al. (2017) which formulates depth via river width and area (in order to use SWOT-observables). $k_{600}$ is now defined as a function of $S_h$, channel roughness n $[T/L^{\frac{1}{3}}]$, channel width W [L], and channel area $A=A_0+dA [L^2]$ for the change in channel area dA). It is also defined by the upscaling parameters $a_i$ and $b_i$.

$k_{600}=a_i(gS_h^{\frac{3}{2}}\frac{1}{n}(A_0+dA)^\frac{2}{3}W^\frac{-2}{3})^{b_i}$  (4)

Equation 4 is then reworked into a Bayesian sampling model with the ‘data’ on the left-hand side sampled from the unknown model parameters on the right-hand side (equation 5). Note that in the formal likelihood specification each term in equation 5 is log-transformed such that equation 5 is a lognormal distribution, however they are written here without a transformation for succinctness. $σ_{upscale}^2$ refers to the uncertainty inherent in the $k_{600}$ estimates themselves (equation 4).

$a_i^{-1}(gS_h^\frac{-3}{2}W^\frac{2}{3})^{b_i} \sim normal(k_{600}^{-1}\frac{1}{n}(A_0+dA)^\frac{2}{3})^{b_i}, \sigma^2_{upscale})$  (5)

Equation five necessitates that Bayesian priors be assigned to model parameters n, $A_0$, and $k_{600}$. Bayesian priors formularize the a priori estimates (and uncertainties) for these parameters. Or more intuitively, they represent our ‘prior river knowledge’ of what n, A0, and $k_{600}$ probably are for some river given that they cannot be remotely sensed. All priors are formalized within the model as truncated normal distributions of the ln-transformed terms such that $ln(X) \sim N(\mu, \sigma^2)$ for λ<X<ε, using prior hyperparameters mean (μ), standard deviation (σ), and upper (ε) and lower bounds (λ) for parameter X = n, $A_0$, or $k_{600}$. In order to avoid relying on in situ information, we assign prior hyperparameters using SWOT data only. n and $A_0$ prior hyperparameters were assigned following the method developed by Brinkerhoff et al. (2020a). They developed a set of river channel prior hyperparameters for McFLI algorithms that are entirely RS-able and reflect differential channel hydraulics as a function of river geomorphology. With this method, they showed substantial improvement in the accuracy of their ungauged RSQ estimates with no changes to the algorithms or data used. We assigned the $k_{600}$ prior hyperparameters using a simple slope regression model trained on the Ulseth et al. (2019) dataset to assign μ (equation 6). $σ^2$ was set to the standard error of equation 6: 1.023. λ and ε were set to -3 and 2.7, respectively.

$\mu_{k600}=3.22+0.347ln(S_h) if ln(S_h) < -4.634$  (6)

$\mu_{k600}=6.85+1.13ln(S_h) if ln(S_h)>-4.634$

Finally, we estimated $σ_{upscale}^2$ using Monte Carlo (MC) methods to approximate total uncertainty from equation 4. Uncertainty in equation 4 can stem from two sources: 1) error baked into the upscaling model (ai and bi parameter uncertainty) and 2) error in Manning’s approximation of the average flow velocity. To do this, we ran MC simulations on 8,000 sets of field measurements of river channel hydraulics from the dataset built by Brinkerhoff et al. (2019). Each MC simulation was itself 10,000 runs, sampling from the normal distributions for $a_i$, $b_i$, and ln-transformed velocity. Estimate uncertainty for velocity followed Hagemann et al. (2017), who estimated ln-transformed Manning’s equation error to be 0.25. This resulted in 8,000 discrete distributions for $k_{600}$ from which the average uncertainty term is extracted and used for $σ_{upscale}^2$.

With the sampling model (equation 5) and prior distributions described, a joint posterior distribution conditional on the SWOT data is derived. To approximate this distribution, BIKER uses a Markov Chain Monte Carlo (MCMC) algorithm. Because it is written in the Stan probabilistic programming language, BIKER’s MCMC of choice is a Hamiltonian Monte Carlo which reduces computation time relative to other MCMC approaches (https://mc-stan.org/docs/2_26/reference-manual/hamiltonian-monte-carlo.html).

###2.3 BIKER validation setup
We validated the BIKER algorithm on 22 rivers using observed average flow velocity and observed slope. First, we detail the 22 rivers, and then we detail how ‘observed’ $k_{600}$ is calculated.

Because SWOT has yet to launch, it is standard practice to benchmark SWOT-related algorithms on ‘SWOT-like data’. There are three current types of SWOT-like data: 1) AirSWOT, which is an airborne Ka-band inSAR currently limited to five rivers globally, 2) simulated rivers that mimic the type of data SWOT will provide, and 3) the SWOT simulator, which introduces errors to these simulated rivers (mostly from radar layover error and random noise- Oubanas et al. 2018a). Because we are principally interested in algorithm performance, we limit our validation setup to simulated rivers in order to benchmark across as many rivers as possible. These simulated rivers are simply reach-averaged hydraulic model outputs where the water surface heights and widths are labelled as ‘RS observations’ and are used as the sole inputs to the BIKER algorithm. Simulated rivers mimic perfect measurement conditions (i.e. no observation errors) and represent the best-case scenarios of what SWOT will provide to hydrologists. Here, we use the 22 rivers archived in Frasson et al. (2019) and used by Rodriquez et al. (2020) to develop reach-averaged Saint Vernant equations within the SWOT context. We omitted the Arial Khan river in Bangladesh due to known problems with that hydraulic model and we further sampled the models for only the observations every 11 days to mimic the average SWOT overpass frequency (this also approximately algins the SWOT observations with the bi-weekly CO2 data used later- Section 2.4). All told, this yielded 503 sets of SWOT observations to validate BIKER with.

Regardless of the validation data used, we do not have observed $k_{600}$ data for these rivers, and to our knowledge no field dataset of $k_{600}$ exists in the type of temporal and spatial frequency that SWOT (and therefore the BIKER algorithm) provides. Further, we are principally interested in our ability to upscale $k_{600}$ from RS observations to avoid relying on velocity rating curves. We are less concerend with the actual accuracy of the upscaling model itself (which can be validated using existing datasets). Therefore, we take the best performing upscaling model from sections 2.1 and 3.1 and use that to calculate the ‘observed’ $k_{600}$ that BIKER is validated against. This is calculated using equation 7, where $V_{obs} [L/T]$ is observed river discharge divided by observed channel area. With this setup, we can directly explore our ability to infer observed average flow velocity and $k_{600}$ from river width and height alone. It also means that, for a fair validation scheme, $\sigma_{upscale}^2$ must be set to reflect only error from Manning’s equation (and not $a_i$ and $b_i$). Thus, $\sigma_{upscale}^2$ is set to 0.25 for this validation. However in practice, it should reflect the total uncertainty calculated in section 2.2.

$k_{600, obs}=a_i(gS_hV_{obs})^{b_i}$ (7)

Validation metrics take two forms (and are detailed in Table 1). To validate across all rivers and timesteps, we used the coefficient of determination r2 and the root mean square error (RMSE). Four normalized metrics were used for by-river validation to compare across rivers: RRMSE and NRMSE are normalized root mean square errors that have been normalized by the observed value and the mean observed value, respectively, to compare across rivers. rBIAS is a measure of prediction bias that is normalized by the mean observed value to compare across rivers. KGE is a standard metric used in streamflow prediction with an intuitive basis: a value greater than -0.41 indicates a model outperforms a uniform prediction of the mean value (Knoben et al. 2019) and a value greater than 0 is often interpreted as a useful prediction in ungauged settings.

*Table 1: Validation metrics used in this study, where r is the correlation coefficient, Nt is number of observations and i is the specific observation. σ refers to the variance of the sample and μ refers to the mean of the sample.*
|Description|Acronym|Definition|Ideal Score|Possible Range|Validation Scheme|
|-----------|-------|----------|-----------|--------------|-----------------|
|Coefficient of determination|$r^2$| Eq|1|0 to 1|	All rivers and all timesteps|
|Root-mean-square-error|RMSE| Eq|0|0 to ∞|All rivers and all timesteps|
|Relative root-mean-square error|RRMSE|Eq |0|0 to ∞|By river|
|Normalized root-mean-square error|NRMSE|Eq|0|0 to ∞|By river|
|Relative bias|rBIAS|Eq|0|-∞ to ∞|By river|
|Kling-Gupta efficiency|KGE|Eq|1|-∞ to 1|	By river|

###2.4 Upscaling to CO2 evasion and bulk carbon efflux
It is one thing to be able to predict $k_{600}$, but researchers are usually more interested in the gas fluxes from rivers and ultimately the bulk carbon efflux and not just the gas exchange velocity. Therefore, we explored 1) the ability of BIKER to reproduce hypothetical carbon dioxide evasion fluxes ($FCO_2 [M/L^2T]$) from these 22 rivers, and 2) the sensitivity of the estimated bulk carbon efflux (via $CO_2$ evasion) to the method used to estimate average flow velocity.

First, we calculated $FCO_2$. To do this, we paired the modeled $k_{CO2}$ values (obtained from the $k_{600}$ estimates) with field-measurements of water-side $CO_2$ concentrations. 29 bi-weekly $CO_2$ samples were made by Beaulieu et al. (2012- Figure S1) at one location in the Ohio River. We paired these 29 values with each 11-day-repeat set of SWOT observations, ignoring the timesteps beyond 29 (only 7% of the SWOT observations were ignored here and this was deemed acceptable). This $CO_2$ data is for the Ohio River only but was applied to all 22 rivers (which includes the Ohio River). Because we are exclusively interested in the relative differences between $FCO_2$ estimates and not the raw fluxes themselves, any $CO_2$ data representative of SWOT-observable rivers was deemed acceptable. Atmospheric $CO_2$ was assumed 390 uatm and water temperature was assumed 25 degrees Celsius. The Schmidt number, used to obtain kCO2 from $k_{600}$, was calculated following Raymond et al. (2012) and Wanninkof et al. (1992).

Then, we calculated bulk carbon efflux using four models for average flow velocity:  BIKER and three rating curves previously used for $CO_2$ upscaling (Lauerwald et al. 2015; Raymond et al. 2013; Raymond et al. 2012- see Table S1 for their definitions). It is worth stressing that the Lauerwald, et al. (2015) model was developed as one of two components of the Raymond et al. (2013) velocity model, and so was actually developed by the latter’s authors. The names used here refer to the final implementation used in each study’s upscaling exercises. For this study, all rating-curve methods use the in-situ discharge record while BIKER does not. This allows us to explore the relative differences in bulk carbon efflux estimates if implementing a wholly ungauged method versus two gauged methods (that also use global-scope HG parameters). Finally, we express the bulk carbon efflux (via $CO_2$ evasion) [M/T] as the average mass flow rate of carbon per year, i.e. multiplied by the surface area of all 22 rivers.

##3 Results
First, we present the results from the $k_{600}$ upscaling experiment (section 3.1). We then take the best performing upscaling function, implement it within BIKER, and validate it on the 22 SWOT rivers (section 3.2). Finally, we use $k_{600}$ as modeled by both BIKER and gauge-based rating curves to compare bulk carbon efflux estimates from the 22 rivers (section 3.3).

###3.1 Validating the $k_{600}$ upscaling model
Figure 3 plots the validations for the five upscaling functions tested. Both the original Ulseth et al. (2019) model, and our RS-able extension, perform the best on this independent set of validation data, with near identical performance. The RS-able Ulseth et al. (2019) model achieves an $r^2$ of `r results_3_1[results_3_1$Model == 'RS-able',]$r2` and an RMSE of `r results_3_1[results_3_1$Model == 'RS-able',]$rmse` m/day, which is just marginally better than the original Ulseth et al. (2019) model (`r results_3_1[results_3_1$Model == 'Ulseth',]$r2` and `r results_3_1[results_3_1$Model == 'Ulseth',]$rmse` m/day, respectively). The two power-law based Raymond models (Raymond et al. (2012) equations 4 and 3) achieve slightly worse performance, with $r^2$ in the 0.70s and their RMSE scores slightly higher (`r results_3_1[results_3_1$Model == 'Raymond_2',]$rmse` and `r results_3_1[results_3_1$Model == 'Raymond_3',]$rmse` m/day, respectively). Raymond et al. (2012) equation 5, which is linear in structure, worked well on their dataset but fails to capture the non-linear relationship evident in this expanded and more geomorphically diverse dataset. In fact, it predicts negative $k_{600}$ values and error metrics cannot be calculated because the prediction residuals are not normally distributed. Because we successfully reproduced the Ulseth et al. (2019) model, the rest of this manuscript deals exclusively with this RS-able variant of the model.

![Figure 3. Validation of five k600 upscaling models on 20% of the Ulseth et al (2019) dataset, withheld for independent testing. Raymond et al. (2012) Eq. 5 predicts negative k600 values, which are not plotted here and highlight a limitation of using a linear model for a non-linear problem (though it did work well on their smaller dataset).](outputs/k600/validation.jpg)

Uncertainties were then propagated through the RS-able Ulseth et l. (2019) model via MC simulations (section 2.2) for the 8,000 hydraulic measurements mapped in Figure 4c. Figure 4a plots a histogram of the 8,000 uncertainty terms extracted from those distributions (Figure 4b plots three of the distributions used to build Figure 4a). Across the 8,000 tests, we quantified average ln-transformed uncertainty in equation 4 to be `r round(results_3_1_MC$meanSigma,2)` (~`r round(exp(results_3_1_MC$meanSigma), 2)` m/day- dashed blue line in Figure S2a). Uncertainty solely from the upscaling model is 1.30, confirming that using Manning’s equation to approximate average flow velocity is introducing very little additional uncertainty to the model (Figure S2). This also confirms the large residuals inherent in $k_{600}$ upscaling models as described by Hall & Ulseth, (2020).

![Figure 4 Model uncertainty estimates: a: histogram of SDs for all 8,000 MC simulations with blue dashed line denoting the mean (1.32). b: Three example MC simulation results of 10,000 samples each. c: Map of the 8,000 sets of hydraulic measurements used for this analysis.](outputs/MonteCarlo/MCsimulations.jpg)

###3.2 Validating the BIKER algorithm
Figure 5a plots the validation results for $k_{600}$ across all 22 rivers and all timesteps. For BIKER, the points are the posterior means while the black lines are the 95% confidence intervals (CIs) for the predictions. $k_{600}$ is strongly correlated with the $k_{600}$ predicted by BIKER, capturing the general magnitude of the predictions and with most points falling on or near the 1:1 line. However, the 95% prediction intervals (dashed grey lines) highlight relatively large residuals for many predictions. Regardless, the RMSE for the BIKER predictions is only `r round(10^results_3_2_all$rmse,2)` m/day across all predictions. This is substantially less than the RMSEs for the upscaling models (Figure 3) and the MC-simulated uncertainties (Figure 4a), indicating that recovering upscaled $k_{600}$ from SWOT data is much more accurate than the upscaling process itself (which is logical).

![Figure 5. a: Validation of remote sensing algorithm for 22 rivers. Black bars are 95% CIs for the modeled values. Grey line is linear regression (and 95% prediction intervals are dashed) and dashed black line is 1:1 line. b: Cumulative density functions (CDFs) of the same results: dashed black line are observed values, green line are BIKER posterior means, and blue lines are BIKER 95% CIs.](outputs/validation/validation.jpg)

Figure 5a suggests a flip in the bias of the $k_{600}$ predictions, which is confirmed in Figure 5b. Figure 5b plots the cumulative density functions (CDFs) of observed and predicted $k_{600}$ where the green line is the BIKER posterior mean and the blue lines are the 95% CIs for the BIKER posterior. We see, more clearly than in Figure 5a, that the posterior mean is systematically underestimated below the median $k_{600}$ values and systemically overestimated above the median $k_{600}$ However, the uncertainty in these estimates captures $k_{600}$ very well: nearly the entire observed CDF falls between, on, or just outside of the 95% CIs (Figure 5b). This highlights the benefit of using a Bayesian predictive framework to fully propagate all parameter uncertainties through to the posterior. While these CIs are large and thus BIKER estimates are reasonably uncertain, we do improve upon our baseline understanding of $k_{600}$ in these rivers. Put another way, we reasonably capture $k_{600}$ with no in situ information about the river while simultaneously and explicitly accounting for the large uncertainties inherent in our $k_{600}$ estimates.

Figure 6a plots validation metrics calculated for each river (i.e. the boxplots are composed of metric scores for the 22 rivers- see Table 1 for metric definitions). Median rBIAS is `r round(median(results_3_2_rivs$rBIAS, na.rm=T),2)`, indicating nearly no bias in most rivers’ predictions. However, some rivers are substantially biased in both directions, further supporting the visual evidence in Figure 5 that sometimes BIKER is substantially under/overestimating $k_{600}$ and that this is river-specific. Median KGE is `r round(median(results_3_2_rivs$kge, na.rm=T),2)`, which is excellent given that absolutely no in situ information is being used to predict $k_{600}$. NRMSE and RMSE have median scores of `r round(median(results_3_2_rivs$nrmse, na.rm=T),2)` and `r round(median(results_3_2_rivs$rrmse, na.rm=T),2)`, respectively. While median KGE and rBIAS scores were strong, the ranges of these scores were somewhat large (standard deviation for KGE of `r round(sd(results_3_2_rivs$kge, na.rm=T),2)` and for rBIAS of `r round(sd(results_3_2_rivs$rBIAS, na.rm=T),2)`). This suggests that BIKER predicts $k_{600}$ very well in some rivers, but not well in other rivers.

![Figure 6. a: Performance metrics by river. See Table 1 for metric definitions. Dashed lines denote scores of 1, 0, and -0.41 for KGE (Knoben et al. 2019). b-d: validation timeseries for three rivers representative of good, reasonable, and poor performance: b was randomly selected from upper tertile of KGE scores, c was randomly selected from the middle tertile, and d from the worst tertile. Model results include the posterior means and 95% CIs.](outputs/validation/validation_by_river.jpg)

Figure 6b-d are representative timeseries plots of predicted and observed $k_{600}$ for three rivers chosen randomly from those with ‘good’ KGE scores (b), ‘okay’ KGE scores (c), and ‘bad’ KGE scores (d). See the Figure 5.6 caption for how this was calculated. For the Kanawha River, the entire timeseries of $k_{600}$ is correctly predicted, while in the Ohio Section 2 River there is a negative bias in the estimates. However, temporal dynamics are still correctly recovered. In the downstream Mississippi River, there is significant bias in the estimates as well as significant uncertainty (per the 95% CIs). The temporal dynamics are also quite off.

###3.3 Implications for carbon cycle estimates
Finally, we explore our ability to 1) use BIKER-produced $k_{600}$ to estimate $FCO_2$ from rivers and then 2) to estimate bulk carbon efflux from $CO_2$ evasion (section 2.4).

In Figure 7a, there is a strong fit to the observed $FCO_2$ data, with an RMSE of `r round(10^results_3_3_all$rmse,2)` $g/m^2dy$. The $r^2$ is slightly lower than that for $k_{600}$ (Figure 7a) due to some abnormally large residuals on a handful of points. There is also a slight underestimation bias for the lowest of the observed $FCO_2$, though this is smaller than the bias in the $k_{600}$ predictions (Figure 5a). $FCO_2$ prediction intervals also have a similar relative magnitude to those presented in Figure 5. Figure 7b-d includes subplots for the same rivers as Figure 4b-d, however with $FCO_2$ instead of $k_{600}$ There is functionally perfect recovery of $FCO_2$ in the Kanawha River, very good recovery (with some positive bias in peak evasion events) for the Ohio Section 2 River, and substantial negative bias in the Downstream Mississippi river. Again, the temporal dynamics are accurately modeled however there is some bias in the magnitude of the predictions. However, $FCO_2$ is so low in the Mississippi river (Figure 7d) that the absolute bias is not as large as it looks in these timeseries plots. This suggests that the BIKER model can easily estimate the changing dynamics of evasion but is at the mercy of the prior on $k_{600}$ to accurately estimate the magnitude of the evasion.

![Figure 7. a: FCO2 from BIKER versus using observed average flow velocity for all timesteps for 22 SWOT rivers (grey lines are linear regression and 95% prediction intervals, while black dashed line is the 1:1 line). b-d: timeseries plots for three example rivers from Figure 5.6.](outputs/flux_implications/FCO2_plot.jpg)

Compared against Figures 5a and 6b-d, there is an arguably stronger recovery of FCO2 than $k_{600}$ (Figure 7). This is presumably due to the structure of equation 1, of which $FCO_2$ is a function of both the observed $CO_2$ concentration gradients and $k_{600}$ This means that the errors in the BIKER $k_{600}$ estimates are diminished in relative importance when multiplied by the $CO_2$ concentration gradients.

Finally, we compare the bulk carbon efflux (via CO2 evasion) from the 22 rivers using BIKER posterior means and three gauge-based rating curves (Figure 8). In Figure 5.8a are barplots of the bulk carbon efflux across all rivers in gigagrams of carbon per year. The BIKER bulk carbon efflux (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_BIKER',]$sumFCO2, 0)` gG-C/yr) is nearly identical to the bulk carbon efflux calculated using observed velocity (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_ulset',]$sumFCO2, 0)` gG-C/yr) while the rating curves underestimate the efflux (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_raymond2012',]$sumFCO2, 0)`, `r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_raymond2013',]$sumFCO2, 0)`, and `r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_Lauerwald2015',]$sumFCO2, 0)` gG-C/yr for ‘Raymond 2012’, ‘Raymond 2013’, and ‘Lauerwald 2015’, respectively). BIKER functionally perfectly captures the estimated efflux from these rivers purely from RS data. This is borne out in the cumulative density functions (CDFs) in Figure 8b, which plot the distribution of all FCO2 estimates across all rivers and timesteps for each velocity model. BIKER nearly perfectly captures FCO2 estimates above approximately the 60th percentile, while the Lauerwald, et al. (2015) model nearly perfectly captures FCO2 below the 60th percentile. Because the bulk efflux is more sensitive to peak evasion events, at the bulk properties level BIKER more accurately models carbon efflux (Figure 8a) than the rating curves do. The Raymond et al. (2012) model, which was mostly trained on small streams, uniformly underestimates FCO2 and thus bulk carbon efflux. The Raymond et al. (2013) model, which is an average of the other two rating curve estimates (Table S1), is in between.

![Figure 8. a: Bulk carbon efflux, per year, from the 22 SWOT rivers as calculated using five different average flow velocity estimates: 1) BIKER, 2) through 4) average flow velocity as calculated using rating curves from the literature, and 5) observed average flow velocity. b: Cumulative density functions (CDFs) of all FCO2 estimates across all timesteps and rivers for the same four velocity models. The median of the distributions is noted with a black line.](outputs/flux_implications/FCO2_models.jpg)

By-river performance scores for $FCO_2$ (Figure S3) show the Lauerwald et al. (2015) model as the best performing in KGE, NRMSE and RRMSE, while BIKER arguably has the least prediction bias (rBIAS) of the models and is approximately the same in KGE as the Raymond et al. (2013) model. Again, the high performance of the Lauerwald et al. (2015) and Raymond et al. (2013) models makes sense as the rating curves rely on the in-situ discharge record. Across all four metrics, the Raymond et al. (2012) model is the worst, performing marginally worse than BIKER.

##4 Discussion
###4.1 Ungauged estimation of gas exchange velocity
BIKER is therefore successful at 1) reproducing $FCO_2$ without any in situ river information for $k_{600}$ (Figure 7) and 2) nearly perfectly reproducing the bulk carbon efflux without any in situ information for $k_{600}$ (Figure 5.8). It is important to stress that, unlike BIKER, the rating curves in Figure 8 rely on the observed discharge record. This means that Figure 8 represents the best performance that the rating curves could ever have; if ran using modeled discharge their accuracy would necessarily decrease. However, this decrease would not be much due to the high accuracy of those data. Thus, these results suggest that BIKER will be useful in two settings: upscaling in ungauged rivers as hypothesized, but also in potentially improving our bulk efflux understandings at gauged sites too. Future work should systematically quantify prediction error from using both global-scope rating curves and modeled discharge, like is the default approach used to date in upscaling studies.

However, Figures 8 and S3 still confirm that rating curves with a streamgauge are preferred when possible, particularly when interested in properties finer than the bulk efflux (Figure S3). Further, methods like the Raymond et al. (2013) and Lauerwald et al. (2015) models are implementable in rivers of any size, while BIKER is limited to the largest rivers on Earth due to SWOT’s coarse spatial resolution. However, our results suggest that, upon SWOT's launch, a BIKER-style approach to estimating gas exchange could be coupled with existing upscaling workflows to improve gas flux predictions where gauges are unavailable.

###4.2 Exploring spatiotemporal dynamics of $k_{600}$
To date, most field-scale studies of riverine gas exchange have focused on 1) its relationship with wind speed (e.g. Beaulieu et al. 2012; Borges et al. 2004; Zappa et al. 2007), 2) average flow velocity (e.g. Alin et al. 2011; Beaulieu et al 2012; Schelker et al. 2016), 3) discharge (e.g. Roberts et al. 2007; Uehlinger & Naegeli, 1998), or 4) oxygen diel curves (where gas exchange is still assumed constant in time- e.g. Appling et al. 2018; Demars et al. 2015). This leaves the spatiotemporal dynamics of gas exchange weakly constrained. A few studies have investigated these dynamics, but these studies have been limited to individual rivers and/or limited field seasons (e.g. Hall et al. 2012; Sand-Jensen & Staher, 2012). Wallin et al. 2011 performed a preliminary analysis in northern Sweden relating river specific (‘at-a-station’) temporal variability in gas exchange with channel slope, but they were limited to an average of only 8 measurements per river in a single watershed. Thus, there is a fundamental knowledge gap in understandings the geomorphic and hydraulic drivers of riverine gas exchange. This limitation is due both to a lack of process-level understanding (Hall & Ulseth, 2020) but also due to a lack of measurements. Authors have argued that the key to explaining the large residual variation in upscaling models is to explore at-a-station temporal variability in $k_{600}$ (Hall & Ulseth, 2020).

Therefore, to explore the spatiotemporal dynamics of $k_{600}$ at fine temporal resolution and but also at the global-scale, SWOT data is an attractive option for upscaling (providing daily hydraulic measurements for a 3 month fast sampling period for calibration and validation and sampling thereafter between 1 and 7 days per 21 day repeat cycle thereafter- Biancarma, et al. 2016). BIKER's success on simulated SWOT rivers bodes well for its implementation on real SWOT data, and functionally opens the door for daily estimation of riverine gas exchange globally once SWOT launches.

###4.3 Biases in upscaling k600
Systematic sampling biases in the data used to develop $k_{600}$ upscaling models (Ulseth et al. 2019; Raymond et al. 2012).

Finally, in reasonably wide rivers, wind begins to exert a non-trivial influence on gas exchange. It is well established that in lakes and the ocean, wind controls near-surface turbulence, and thus gas exchange (Beaulieu et al. 2012; Read et al. 2012). Authors have argued that large rivers are a hybrid of the hydraulics-driven turbulence in small rivers and the wind-driven turbulence in lakes (Beaulieu et al. 2012). As SWOT will measure only rivers wider than 50m, it follows that wind is likely exerting some influence on gas exchange in SWOT-observable reaches. However, we opted to ignore wind effects in our upscaling model and in BIKER to favor global scalability and implementability for two reasons: 1) current upscaling efforts do not account for wind in their estimation of $k_{600}$ and 2) it is infeasible to parameterize every set of SWOT data with local wind data, and relying on an in situ understanding of wind defeats the purpose of BIKER for ungauged settings. Rather, we argue that

##5 Conclusions

##6 Acknowledgements
BIKER is available at https://github.com/craigbrinkerhoff/BIKER. All results and code to generate results, figures, and the manuscript are archived at zenodod something something. Data used in this study is available in Frasson et al. (2019) and Ulseth et al. (2019). This project benefitted from early comments from Tamlin Pavelsky, as well as the large body of existing knowledge from the SWOT River Discharge Working Group.

### 2021-01-28 manuscript_vc/Outline.Rmd
big push

--- 2020-12-18 Outline.Rmd
@@
 ## Craig's things to do
-1) Address bias from k600 prior
-2) redo fig 6 with sigma_man not kest_sd
-2) Working on 3rd research question (courtesy of Merritt)
+1) Run with slope errors from Renato
 
@@
 Black bars are 95% CIs for the modeled values.
-Black line is linear regression and dashed grey line is 1:1 line.*](outputs//validation//validation.jpg)
+Grey line is linear regression (and 95% predicition intervals) and dashed black line is 1:1 line.*](outputs//validation//validation.jpg)
 
@@
 
-6.
-We also compare a priori model uncertainity (i.e.
-from Manning's equation) against posterior model uncertainity to see how much uncertianity is added by using only observed W and H, i.e.
-a test of the manifestation of the priors and equifinality.
-
-THIS NEEDS TO BE REDONE WITH SIGMA_MAN
-
-![**Figure 6** *A) Histogram of BIKER k600 uncertainities (across all rivers and timesteps) compared against a priori uncertainity from Manning's equation (0.25- dark blue line).
-Red line is the mean of the BIKER uncertainities (0.32).
-B) The same plot, but mean uncertaintities for the 22 rivers (red line at 0.33).*](outputs//validation//validation_uncertainity.jpg)
-
-I'm still trying to figure out why there's like five rivers that have crazy increased uncertianity (causing the bimodal distributions).
-
 ##Goal 3
-#### Current Approach
+#### Method
 1) Used the 22 SWOT rivers in conjunction with biweekly in-stream CO2 data for a similarly large river (Ohio River- Beauliu etal 2012).
@@
 
-![**Figure 7** *Left: FCO2 from BIKER versus using observed average flow velocity for all timesteps for 22 SWOT rivers.
+![**Figure 6** *Left: FCO2 from BIKER versus using observed average flow velocity for all timesteps for 22 SWOT rivers (grey lines are linear regression and 95% prediciton intervals, while black dashed line is the 1:1 line).
 right: timeseries plots for three example rivers.*](outputs//flux_implications//FCO2_plot.jpg)
 
-##References
+Finally, we wanted to see how our completely ungauged, no in situ information algorithm compares against gauged approaches for estimating average flow velocity.
+We used two V~Q rating curves used for global CO2 upscaling efforts (Liu et al.
+in review; Raymond etal.
+2013).
+
+![**Figure 7** *Left: barplots of total flux, per day and m2, of CO2 off of the 22 SWOT rivers for BIKER and threevelocity models: observed velocity and two hydraulic geomtery models (using observed discharge) used for global CO2 upscaling.
+Right: Cummulative density functions (CDFs) of all FCO2 estimates across all timesteps and rivers for the same set of velocity models.
+*](outputs//flux_implications//FCO2_models.jpg)
+
+## References
 Beaulieu, J.

### 2021-01-28 supp_v1.Rmd
big push

---
title: "Supplemental Information"
author: "Craig Brinkerhoff"
date: "January 28, 2021"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Supplemental Information
Text S1
Tsivoglou & Neal (1976) outlined an energy dissipation model for the reaeration coefficient (K2, or k without depth correction) that is frequently generalized and used to scale k600 (Raymond et al. 2012; Ulseth et al. 2019). They showed that K2 scales linearly with the energy dissipation rate eD. Their derivation explicitly shows that eD=dH/dT for some segment of stream where H is water surface elevation and T is hydraulic residence time. For a stream segment with residence time T, the loss in water surface elevation is synonymous equivalent to the downward component of flow velocity V in the h dimension over the stream segment length (i.e. Sh). So, we can generalize this to dH/dT=V_h S_h≈VS_h. Finally, this can be converted to an energy dissipation rate per unit weight of water such that eD [L^2/T^3 ]=g dH/dT≈〖gVS〗_h. This model is convenient for our use because SWOT will only measure Sh and not channel slope S0. Further, Manning’s equation under non-uniform flow conditions requires Sh and not S¬0 and so we can conveniently merge the slope terms across equations and directly measure them from SWOT.

To my knowledge, most k600 upscaling studies to date have assumed uniform flow conditions (i.e. S_0=S_h) to train their upscaling parameters. This is presumably because channel slope is readily available in hydrographic data products while water surface slope is not. Therefore, future work should look at the influence of non-uniform flow on the upscaling of k600. Once SWOT launches, this will be easy to compare as SWOT data products will have both channel slope and water surface slope.

![Figure S1 Timeseries of the biweekly CO2 data from Beaulieu et al. (2012). Sampling took place 2008-2009 in the Ohio River (upstream of Cincinnati, OH). Each point here was joined to the 11-day SWOT observations used in this study (section 5.2.4).](outputs/flux_implications/Beaulieu_timeseries.jpg)

![Figure S2 Histogram of equation 3 uncertainty terms extracted from 8,000 different Monte Carlo simulations (following that outlined in section 5.2.2). This indicates estimate uncertainty from the ai and bi parameters (and not Manning’s equation). Dashed blue line indicates the mean uncertainty (1.30 in ln-space, or ~3.67 m/day).](outputs/MonteCarlo/noMannings.jpeg)

![Figure S3 By-river performance metrics for FCO2 using the metrics outlined in Table 1. Dashed lines denote scores of 1, 0, and -0.41 for KGE (Knoben et al. 2019). Recall that the Raymond and Lauerwald models use the in-situ discharge record while BIKER does not.](outputs/flux_implications/FCO2_by_river.jpg)

Table S1 Details on the 3 velocity rating curve models used to estimate bulk carbon efflux from the SWOT rivers (section 5.2.4).
	Equation	Description
Raymond et al. (2012)	lnV=-1.64+0.285lnQ	Trained on Raymond et al. (2012) dataset
Raymond et al. (2013)	lnV=mean{█(-1.64+0.285lnQ,@-1.06+0.12lnQ  )}	Average of Raymond et al. (2012) model and one trained on 9,811 USGS streamgauges
Lauerwald et al. (2015)	lnV=-1.06+0.12lnQ	Borrowed from ½ of the Raymond et al. (2013) calculation (trained on 9,811 USGS streamgauges)

### 2021-01-28 main.R

@@ -17,6 +17,15 @@ source("scripts//validation_figures.R")
 rm(list = ls())
 print('Validation figures done. Hopefully you had the correct validation results stored in the output directory!!')
 
-#Knit markdown notes
-rmarkdown::render("Outline.Rmd")
+#Move all results to global environment
+results_3_1 <- read.csv('outputs//k600//results.csv')
+results_3_1_MC <- read.csv('outputs//MonteCarlo//results.csv')
+results_3_2_rivs <- read.csv('outputs//validation//results_by_riv.csv')
+results_3_2_all <- read.csv('outputs//validation//results_all_riv.csv')
+results_3_3_all <- read.csv('outputs//flux_implications//fco2_stats_all.csv')
+results_3_3_bulk <- read.csv('outputs//flux_implications//bulkFluxes.csv')
+results_3_3_bulk$key <- as.character(results_3_3_bulk$key)
+
+#knit manuscript
+rmarkdown::render("manuscript_v1.Rmd")
 print('Project report, or Outline.html, is generated.')

### 2021-01-28 scripts/MonteCarlo_analysis.R

@@ -147,6 +147,6 @@ combinedPlot <- plot_grid(combinedPlot, g1, ncol=1, labels = c(NA, 'c'), label_s
 #combine subplots-------------------------------------------------
 ggsave('outputs//MonteCarlo//MCsimulations.jpg', combinedPlot, width=10, height=6)
 
-#uncertainity for algorithm---------------------------------------
-meanSD
-medianSD
+#Save uncertainity to file for ms---------------------------------------
+results <- data.frame('meanSigma'=meanSD, 'medianSigma'=medianSD)
+write.csv(results, 'outputs//MonteCarlo//results.csv')

### 2021-01-28 scripts/k600_model.R

@@ -25,23 +25,16 @@ data <- filter(data, is.na(width)==0) #filter out no width measurements
 data$eD <- 9.8 * data$slope * data$Vms #water column turbulent dissipation rate m2/s3
 
 #width regime models for k600-----------------------------------------------------------------------
-  #Width regimes are necessary because we can't remotely sense eD directly (hence the entire Bayesian algorithm!)
-    #Thus, we can't know the eD threhsold of 0.02 a priori. BUT, we do know width a priori
-data$widthRegime <- ifelse(data$width < 10 & data$slope < 0.05, '1',
+data$widthRegime <- ifelse(data$width < 10 & data$slope < 0.05, '1', #meters
                             ifelse(data$width < 10 & data$slope > 0.05, '1.5',
                                 ifelse(data$width < 50, '2',
                                    ifelse(data$width < 100, '3','4'))))
 
-#log transform some variables
+#log transform some variables-----------------------------------------------------------
 data$log_eD <- log(data$eD)
 data$log_k600 <- log(data$k600)
 data$log_slope <- log(data$slope)
 data$raymond <- data$slope*data$Vms #for raymonds 2012 model
-
-data <- filter(data, data == 'Raymond et al. 2012 [5 studies]')
-lm <- lm(log(data$Vms)~log(data$Qm3s))
-summary(lm)
-confint(lm)
 
 #Check linearity assumptions within each width class. If you cycle through the different classes, everything holds
 #lm <- lm(log_k600~log_eD, data=filter(data, widthRegime==1))
@@ -88,16 +81,16 @@ data_test$k600_pred_raymond_3 <- exp(predict(lm_raymond_3, data_test))
 data_test$k600_pred_ulseth <- exp(predict(lm_ulseth, data_test))
 
 #Plot validation - me
-r2 <- round(summary(lm(log10(data_test$k600) ~ log10(data_test$k600_pred_me)))$r.squared,2)
-rmse <- round(10^(Metrics::rmse(log10(data_test$k600), log10(data_test$k600_pred_me))),2)
[…]

### 2021-02-03 manuscript_v1.Rmd
ms updates

@@
 date: "`r Sys.Date()`"
+bibliography: References.bib
 output:
@@
 ##1 Introduction
-Greenhouse gas (GHG) emissions from rivers and lakes are significant (Cole et al.
-2007).
-Rivers are usually supersaturated with dissolved GHGs and present a net flux of these gases from water to air (Cole & Caraco 2001; Jones et al.
-2003).
+Greenhouse gas (GHG) emissions from rivers and lakes are significant [@colePlumbingGlobalCarbon2007a].
+Rivers are usually supersaturated with dissolved GHGs and present a net flux of these gases from water to air [@coleCarbonCatchmentsConnecting2001a; @jonesLongtermDeclineCarbon2003].
 This flux $[M/L^2T]$ is calculated, with in situ knowledge of the gas concentration gradient between the water $C_{water} [M/L^3]$ and the air $C_{air} [M/L^3]$ and the gas exchange velocity $k [L/T]$, as equation 1.
 
-$Flux=k(C_{water}-C_{air})$ (1)
+$$Flux=k(C_{water}-C_{air}) (1)$$
 
 k is largely a function of water column turbulence, at least in fluvial settings.
-This is because k scales with diffusivity as normalized by the boundary-layer thickness, and in rivers turbulence defines the thickness of the boundary-layer (Hall & Ulseth, 2020 for a thorough review of gas exchange velocity in rivers).
-Due to its reliance on diffusivity (as well as kinematic viscosity), k is defined differently for different GHGs (Raymond et al.
-2012; Wanninkof 1992).
-It is also often measured using different gases than those being studied, so a variant of k that is normalized to a Schmidt number of 600 ($k_{600} [L/T]$) is generally used to be GHG and temperature agnostic (Hall & Ulseth, 2020).
+This is because k scales with diffusivity as normalized by the boundary-layer thickness, and in rivers turbulence defines the thickness of the boundary-layer (@hallGasExchangeStreams2020 for a thorough review of gas exchange velocity in rivers).
+Due to its reliance on diffusivity (as well as kinematic viscosity), k is defined differently for different GHGs [@raymondScalingGasTransfer2012a; @wanninkhofRelationshipWindSpeed1992a].
+It is also often measured using different gases than those being studied, so a variant of k that is normalized to a Schmidt number of 600 ($k_{600} [L/T]$) is generally used to be GHG and temperature agnostic [@hallGasExchangeStreams2020].
 This is synonymous with a k for carbon dioxide ($k_{CO2}$) at 20 degrees Celsius.
 
-It is infeasible to measure $k_{600}$ directly when working at the river network scale and so ‘upscaling functions’ are used that predict $k_{600}$ from readily available river geomorphology (e.g.
-O’Connor & Dobbins, 1958; Palumbo & Brown, 2014; Raymond et al.
-2012).
-Following the theory above, many upscaling functions are based on the basic concept that k is empirically correlated with the turbulent energy dissipation e at the air-water interface (Zappa et al.
-2007).
-This is convenient for upscaling because in rivers and streams e can be equated to the turbulent energy dissipation rate eD $[L^2/T^3]$ (Ulseth et al.
-2019; Raymond et al.
-2012), which is estimated using easily calculated hydraulic parameters (eD≈gSV where g is gravitational acceleration, S is channel slope, and V is average flow velocity- Tsivoglou & Neal, 1976).
-It has been observed that this scaling relationship is particularly strong in higher energy streams where bubble-mediated gas exchange occurs, though the general scaling function holds across all rivers and streams (Ulseth et al.
-2019).
+It is infeasible to measure $k_{600}$ directly when working at the river network scale and so ‘upscaling functions’ are used that predict $k_{600}$ from readily available river geomorphology [e.g.
+@oconnorMechanismReaerationNatural1958; @palumboAssessingPerformanceReaeration2014; @raymondScalingGasTransfer2012a].
+Following the theory above, many upscaling functions are based on the basic concept that k is empirically correlated with the turbulent energy dissipation e at the air-water interface [@zappaEnvironmentalTurbulentMixing2007].
+This is convenient for upscaling because in rivers and streams e can be equated to the turbulent energy dissipation rate eD $[L^2/T^3]$ [@ulsethDistinctAirWater2019a; @raymondScalingGasTransfer2012a], which is estimated using easily calculated hydraulic parameters (eD≈gSV where g is gravitational acceleration, S is channel slope, and V is average flow velocity- @tsivoglouTracerMeasurementReaeration1976).
+It has been observed that this scaling relationship is particularly strong in higher energy streams where bubble-mediated gas exchange occurs, though the general scaling function holds across all rivers and streams [@ulsethDistinctAirWater2019a].
 Despite reasonable predictive accuracy, these upscaling models suffer from residual variation indicating large predictive error (Wallin et al.
@@
 Still, these are the current best options for estimating $k_{600}$ in rivers.
-It has been suggested that exploring at-a-station hydraulic variation (and thus $k_{600}$ variation) is a promising approach to reducing the residual variation in upscaling models (Hall & Ulseth, 2020).
+It has been suggested that exploring at-a-station hydraulic variation (and thus $k_{600}$ variation) is a promising approach to reducing the residual variation in upscaling models [@hallGasExchangeStreams2020].
 
@@
 Ideally, a worker can directly measure velocity at some permanently or temporarily gauged river in their system and build velocity~discharge rating curves to upscale to the entire river network.
-Rating curves are also referred to as hydraulic geometry (HG) models: simple power law relations between hydraulic properties and discharge (Leopold & Maddock, 1953).
-However, this is not feasible in ungauged basins (whether physically or politically ungauged- Gleason & Durand, 2020) where no in situ hydraulic information is available and fieldwork is impractical.
+Rating curves are also referred to as hydraulic geometry (HG) models: simple power law relations between hydraulic properties and discharge [@leopoldHydraulicGeometryStream1953a].
+However, this is not feasible in ungauged basins (whether physically or politically ungauged- @gleasonRemoteSensingRiver2020) where no in situ hydraulic information is available and fieldwork is impractical.
 Thus, in ungauged settings, present understandings of GHG fluxes from rivers are likely limited by our access to velocity and/or discharge data.
@@
 This is the approach used in upscaling average flow velocity to continental/global river networks and has proven quite useful in those settings (e.g.
-Horgby et al.
-2019; Borges et al.
-2015; Lauerwald et al.
-2015; Raymond et al.
-2013).
-However, it has long been established that HG parameters exhibit high spatial variability across landscapes and vary significantly from river to river (Park, 1977).
+@horgbyUnexpectedLargeEvasion2019; @borgesVariabilityGasTransfer2004; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a).
+However, it has long been established that HG parameters exhibit high spatial variability across landscapes and vary significantly from river to river [@parkWorldwideVariationsHydraulic1977].
 Because of this, a large body of geomorphology work has attempted to parse out process-based explanations for HG parameters (e.g.
-Dingman 2007; Parker, et al.
-2007; Singh, 2003; Ferguson, 1986) and an open research question is how to best prescribe a set of HG parameters to a given river (particularly at the global scale).
+@dingmanAnalyticalDerivationAtastation2007b; @parkerPhysicalBasisQuasiuniversal2007b; @singhTHEORIESHYDRAULICGEOMETRY2003a; @fergusonHydraulicsHydraulicGeometry1986a) and an open research question is how to best prescribe a set of HG parameters to a given river (particularly at the global scale).
 A potential alternative to this approach, in ungauged basins at least, is to directly estimate a river’s hydraulic properties from remote sensing (RS) data.
 
-The problem of hydraulic estimation in ungauged basins is itself a burgeoning subfield within fluvial hydrology and an analogue to the problem presented here is the ungauged remote sensing of river discharge (RSQ- see Gleason & Durand, 2020 for a thorough review).
+The problem of hydraulic estimation in ungauged basins is itself a burgeoning subfield within fluvial hydrology and an analogue to the problem presented here is the ungauged remote sensing of river discharge (RSQ- see @gleasonRemoteSensingRiver2020 for a thorough review).
 A specific subset of these ungauged RSQ methods are termed ‘McFLIs’ or ‘Mass Conserved Flow Law Inversion’ algorithms (e.g.
-Andreadis et al.
-2020; Brinkerhoff et al.
-2020a; Durand et al.
-2014; Garambois & Monnier, 2015; Hagemann et al.
-2017).
-They use basic geomorphic theories, rather than full hydrologic and/or hydraulic models, and the concept of ‘prior river knowledge’ to estimate discharge from RS data where not all hydraulic terms are RS-able (Gleason & Durand, 2020; Gleason et al.
-2017).
-McFLIs have largely been developed in the context of the NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) mission, which will launch in 2022 and provide the world’s first global measurements of water surface extent and elevation at novel temporal resolutions (Biancamaria et al.
-2016).
+@andreadisConstrainingAssimilationSWOT2020d; @brinkerhoffConstrainingRemoteRiver2020; @durandEstimatingReachaveragedDischarge2014a; HIVDI; @hagemannBAMBayesianAMHGManning2017a).
+They use basic geomorphic theories, rather than full hydrologic and/or hydraulic models, and the concept of ‘prior river knowledge’ to estimate discharge from RS data where not all hydraulic terms are RS-able [@gleasonRemoteSensingRiver2020; @gleasonTrackingRiverFlows2017].
+McFLIs have largely been developed in the context of the NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) mission, which will launch in 2022 and provide the world’s first global measurements of water surface extent and elevation at novel temporal resolutions [@biancamariaSWOTMissionIts2016].
 On average SWOT will sample rivers every 11 days and provide a unique opportunity to measure river and lake hydraulics at the global scale.
-While the raw performance of McFLIs is lower than a more sophisticated approach, they are globally flexible, easily implementable in any river that SWOT can sample, and improve our understandings of ungauged rivers when little to no information was previously available (Brinkerhoff et al.
-2020a; Durand et al.
-2016).
-Recently, McFLIs have also shown promise in providing additional and beneficial information to traditional hydrologic modeling via data assimilation (Ishitsuka et al.
-2020).
+While the raw performance of McFLIs is lower than a more sophisticated approach, they are globally flexible, easily implementable in any river that SWOT can sample, and improve our understandings of ungauged rivers when little to no information was previously available [@brinkerhoffConstrainingRemoteRiver2020; @durandIntercomparisonRemoteSensing2016b].
+Recently, McFLIs have also shown promise in providing additional and beneficial information to traditional hydrologic modeling via data assimilation [@ishitsukaCombiningOpticalRemote].
 
 McFLIs often employ Bayesian inference.
-Bayesian techniques have been previously used to estimate $k_{600}$ concurrently with stream metabolism from dissolved oxygen (O2) datasets (Appling et al.
-2018; Grace et al.
-2015; Holtgrieve et al.
-2010).
+Bayesian techniques have been previously used to estimate $k_{600}$ concurrently with stream metabolism from dissolved oxygen (O2) datasets [@applingOvercomingEquifinalityLeveraging2018; @graceFastProcessingDiel2015; @holtgrieveSimultaneousQuantificationAquatic2010].
 While this approach requires detailed field data and is thus not applicable to $k_{600}$ estimation in ungauged rivers, it does suggest that Bayesian inference could be useful for estimating $k_{600}$ when parameter equifinality is a problem (at it is with McFLIs).
@@
 ###2.1 Developing an RS-able $k_{600}$ upscaling model
-Raymond et al.
-(2012) first used river hydraulics and the theoretical relationship between the turbulent dissipation rate (eD) and gas exchange to upscale $k_{600}$ from average flow velocity and channel slope, extending earlier work predicting the reaeration coefficient (e.g.
-O’Connor & Dobbins, 1958; Tsivoglou & Neal, 1976).
-Later, Ulseth et al.
-(2019) built upon Raymond et al.
-(2012)’s work by expanding their dataset to include measurements from steeper rivers and finding that two distinct upscaling regimes exist in low and high energy rivers.
-More specifically, Ulseth et al.
-(2019) scale $k_{600}$ using eD, where the resulting upscaling model parameters are significantly different whether eD is high or low.
+[@raymondScalingGasTransfer2012a] first used river hydraulics and the theoretical relationship between the turbulent dissipation rate (eD) and gas exchange to upscale $k_{600}$ from average flow velocity and channel slope, extending earlier work predicting the reaeration coefficient (e.g.
+@oconnorMechanismReaerationNatural1958; @tsivoglouTracerMeasurementReaeration1976).
+Later, [@ulsethDistinctAirWater2019a] built upon [@raymondScalingGasTransfer2012a]’s work by expanding their dataset to include measurements from steeper rivers and finding that two distinct upscaling regimes exist in low and high energy rivers.
+More specifically, [@ulsethDistinctAirWater2019a] scale $k_{600}$ using eD, where the resulting upscaling model parameters are significantly different whether eD is high or low.
 Their upscaling function is reprinted as equation 2.
 
-$ln(k_{600})=3.10 + 0.35ln(eD) if eD < 0.02$  (2)
+$$ln(k_{600})=3.10 + 0.35ln(eD) if eD < 0.02  (2) \\
 
-$ln(k_{600})=6.43+1.18ln(eD) if eD > 0.02$
+ln(k_{600})=6.43+1.18ln(eD) if eD > 0.02$$
 
@@
 We do not have this a priori knowledge in an ungauged setting and only have access to SWOT-observed river widths and heights (and therefore water surface slopes).
-Thus, we need to extend the Ulseth et al.
-(2019) model to assign the two model parameters using only RS-able hydraulics.
+Thus, we need to extend the [@ulsethDistinctAirWater2019a] model to assign the two model parameters using only RS-able hydraulics.
 To do this, we use the basic premise that the $k_{600}\sim eD$ scaling relationship is different for different river sizes.
-We use river width as a proxy for river size and train different upscaling functions for different groups of data with similar river sizes (Figure 2, trained on the Ulseth et al.
-2019 dataset).
+We use river width as a proxy for river size and train different upscaling functions for different groups of data with similar river sizes (Figure 2, trained on the [@ulsethDistinctAirWater2019a] dataset).
 In doing so, we find that narrow rivers (< 10m wide) need to be further broken into low energy (slope < 0.05) and high energy (slope > 0.05) regimes in order to meet the assumptions necessary to fit a linear regression model (specifically, normally distributed residuals of the predictions).
@@
 
-$k_{600}=a_i(eD)^{b_i}$ (3)
+$$k_{600}=a_i(eD)^{b_i} (3)$$
 
 This model therefore assigns ai and bi a priori using just river width and slope to define i.
-Another implication of this extension of the Ulseth et al.
-(2019) model is that we now use a and b parameters that implicitly reflect the distinct geomorphology of different sized rivers.
-Finally, we tested whether this model can reasonably estimate $k_{600}$ relative to the original Ulseth et al.
-(2019) model, with the goal of at least replicating their performance.
-We compared it against the various eD-based upscaling models from Raymond et al.
-(2012) that are commonly used in the literature.
-This was done by training all upscaling functions on 80% of the Ulseth et al.
-(2019) data and comparing their performance on the remaining 20% of the data.
-Because the Raymond et al.
-(2012) models were originally trained on a subset of this dataset, we deemed it fairer to refit their parameters to this expanded dataset (but keep model structure identical).
+Another implication of this extension of the [@ulsethDistinctAirWater2019a] model is that we now use a and b parameters that implicitly reflect the distinct geomorphology of different sized rivers.
+
+Finally, we tested whether this model can reasonably estimate $k_{600}$ relative to the original [@ulsethDistinctAirWater2019a] model, with the goal of at least replicating their performance.
+We compared it against the various eD-based upscaling models from [@raymondScalingGasTransfer2012a] that are commonly used in the literature.
+This was done by training all upscaling functions on 80% of the [@ulsethDistinctAirWater2019a] data and comparing their performance on the remaining 20% of the data.
+Because the [@raymondScalingGasTransfer2012a] models were originally trained on a subset of this dataset, we deemed it fairer to refit their parameters to this expanded dataset (but keep model structure identical).
 
@@
 With a validated upscaling model that requires no a priori knowledge of eD (results presented in section 3.1) we turn to developing the algorithm to RS $k_{600}$.
-The approach used here is heavily informed by Hagemann et al.
-(2017)’s algorithm for ungauged RSQ via Bayesian inference.
+The approach used here is heavily informed by [@hagemannBAMBayesianAMHGManning2017a]’s algorithm for ungauged RSQ via Bayesian inference.
 Consult that paper for the logic behind conceptualizing the equation for discharge as a Bayesian inference problem, which we largely follow here to conceptualize upscaling $k_{600}$ as a Bayesian inference problem.
@@
 In order to use Bayesian inference to solve for $k_{600}$, we need to derive a sampling model/likelihood function.
-Following Tsivoglou & Neal (1976), $eD \approx gS_hV$ for unsteady flow conditions (channel slope $S_0 \neq S_h$- see text S1 for this model derivation and why assuming non-uniform flow is convenient for our application).
+Following [@tsivoglouTracerMeasurementReaeration1976], $eD \approx gS_hV$ for unsteady flow conditions (channel slope $S_0 \neq S_h$- see text S1 for this model derivation and why assuming non-uniform flow is convenient for our application).
 Using this, as well as Manning’s equation to relate average flow velocity to channel geometry, we rewrite equation 3 as equation 4.
@@
 
-$k_{600}=a_i(gS_h^{\frac{3}{2}}\frac{1}{n}(A_0+dA)^\frac{2}{3}W^\frac{-2}{3})^{b_i}$  (4)
+$$k_{600}=a_i(gS_h^{\frac{3}{2}}\frac{1}{n}(A_0+dA)^\frac{2}{3}W^\frac{-2}{3})^{b_i} (4)$$
 
@@
 
-$a_i^{-1}(gS_h^\frac{-3}{2}W^\frac{2}{3})^{b_i} \sim normal(k_{600}^{-1}\frac{1}{n}(A_0+dA)^\frac{2}{3})^{b_i}, \sigma^2_{upscale})$  (5)
+$$a_i^{-1}(gS_h^\frac{-3}{2}W^\frac{2}{3})^{b_i} \sim normal(k_{600}^{-1}\frac{1}{n}(A_0+dA)^\frac{2}{3})^{b_i}, \sigma^2_{upscale})  (5)$$
 
@@
 In order to avoid relying on in situ information, we assign prior hyperparameters using SWOT data only.
-n and $A_0$ prior hyperparameters were assigned following the method developed by Brinkerhoff et al.
-(2020a).
+n and $A_0$ prior hyperparameters were assigned following the method developed by [@brinkerhoffConstrainingRemoteRiver2020].
 They developed a set of river channel prior hyperparameters for McFLI algorithms that are entirely RS-able and reflect differential channel hydraulics as a function of river geomorphology.
 With this method, they showed substantial improvement in the accuracy of their ungauged RSQ estimates with no changes to the algorithms or data used.
-We assigned the $k_{600}$ prior hyperparameters using a simple slope regression model trained on the Ulseth et al.
-(2019) dataset to assign μ (equation 6).
+We assigned the $k_{600}$ prior hyperparameters using a simple slope regression model trained on the [@ulsethDistinctAirWater2019a] dataset to assign μ (equation 6).
 $σ^2$ was set to the standard error of equation 6: 1.023.
@@
 
-$\mu_{k600}=3.22+0.347ln(S_h) if ln(S_h) < -4.634$  (6)
+$$\mu_{k600}=3.22+0.347ln(S_h) if ln(S_h) < -4.634   (6) \\
 
-$\mu_{k600}=6.85+1.13ln(S_h) if ln(S_h)>-4.634$
+\mu_{k600}=6.85+1.13ln(S_h) if ln(S_h)>-4.634$$
 
@@
 Uncertainty in equation 4 can stem from two sources: 1) error baked into the upscaling model (ai and bi parameter uncertainty) and 2) error in Manning’s approximation of the average flow velocity.
-To do this, we ran MC simulations on 8,000 sets of field measurements of river channel hydraulics from the dataset built by Brinkerhoff et al.
-(2019).
+To do this, we ran MC simulations on 8,000 sets of field measurements of river channel hydraulics from the dataset built by [@brinkerhoffReconcilingAtaStationAtManyStations2019].
 Each MC simulation was itself 10,000 runs, sampling from the normal distributions for $a_i$, $b_i$, and ln-transformed velocity.
-Estimate uncertainty for velocity followed Hagemann et al.
-(2017), who estimated ln-transformed Manning’s equation error to be 0.25.
+Estimate uncertainty for velocity followed [@hagemannBAMBayesianAMHGManning2017a], who estimated ln-transformed Manning’s equation error to be 0.25.
 This resulted in 8,000 discrete distributions for $k_{600}$ from which the average uncertainty term is extracted and used for $σ_{upscale}^2$.
@@
 Because SWOT has yet to launch, it is standard practice to benchmark SWOT-related algorithms on ‘SWOT-like data’.
-There are three current types of SWOT-like data: 1) AirSWOT, which is an airborne Ka-band inSAR currently limited to five rivers globally, 2) simulated rivers that mimic the type of data SWOT will provide, and 3) the SWOT simulator, which introduces errors to these simulated rivers (mostly from radar layover error and random noise- Oubanas et al.
-2018a).
+There are three current types of SWOT-like data: 1) AirSWOT, which is an airborne Ka-band inSAR currently limited to five rivers globally, 2) simulated rivers that mimic the type of data SWOT will provide, and 3) the SWOT simulator, which introduces errors to these simulated rivers (mostly from radar layover error and random noise- @oubanasDischargeEstimationUngauged2018).
 Because we are principally interested in algorithm performance, we limit our validation setup to simulated rivers in order to benchmark across as many rivers as possible.
@@
 no observation errors) and represent the best-case scenarios of what SWOT will provide to hydrologists.
-Here, we use the 22 rivers archived in Frasson et al.
-(2019) and used by Rodriquez et al.
-(2020) to develop reach-averaged Saint Vernant equations within the SWOT context.
+Here, we use the 22 rivers archived in [@frassonCompilationHydraulicModels2019a] and used by [@rodriguezObservingRiversVarying2020] to develop reach-averaged Saint Vernant equations within the SWOT context.
 We omitted the Arial Khan river in Bangladesh due to known problems with that hydraulic model and we further sampled the models for only the observations every 11 days to mimic the average SWOT overpass frequency (this also approximately algins the SWOT observations with the bi-weekly CO2 data used later- Section 2.4).
@@
 
-$k_{600, obs}=a_i(gS_hV_{obs})^{b_i}$ (7)
+$$k_{600, obs}=a_i(gS_hV_{obs})^{b_i} (7)$$
 
@@
 rBIAS is a measure of prediction bias that is normalized by the mean observed value to compare across rivers.
-KGE is a standard metric used in streamflow prediction with an intuitive basis: a value greater than -0.41 indicates a model outperforms a uniform prediction of the mean value (Knoben et al.
-2019) and a value greater than 0 is often interpreted as a useful prediction in ungauged settings.
+KGE is a standard metric used in streamflow prediction with an intuitive basis: a value greater than -0.41 indicates a model outperforms a uniform prediction of the mean value [@knobenTechnicalNoteInherent2019a] and a value greater than 0 is often interpreted as a useful prediction in ungauged settings.
 
@@
 To do this, we paired the modeled $k_{CO2}$ values (obtained from the $k_{600}$ estimates) with field-measurements of water-side $CO_2$ concentrations.
-29 bi-weekly $CO_2$ samples were made by Beaulieu et al.
-(2012- Figure S1) at one location in the Ohio River.
+29 bi-weekly $CO_2$ samples were made by [@beaulieuControlsGasTransfer2012a]- Figure S1) at one location in the Ohio River.
 We paired these 29 values with each 11-day-repeat set of SWOT observations, ignoring the timesteps beyond 29 (only 7% of the SWOT observations were ignored here and this was deemed acceptable).
@@
 Atmospheric $CO_2$ was assumed 390 uatm and water temperature was assumed 25 degrees Celsius.
-The Schmidt number, used to obtain kCO2 from $k_{600}$, was calculated following Raymond et al.
-(2012) and Wanninkof et al.
-(1992).
+The Schmidt number, used to obtain kCO2 from $k_{600}$, was calculated following [@raymondScalingGasTransfer2012a] and [@wanninkhofRelationshipWindSpeed1992a].
 
-Then, we calculated bulk carbon efflux using four models for average flow velocity:  BIKER and three rating curves previously used for $CO_2$ upscaling (Lauerwald et al.
-2015; Raymond et al.
-2013; Raymond et al.
-2012- see Table S1 for their definitions).
-It is worth stressing that the Lauerwald, et al.
-(2015) model was developed as one of two components of the Raymond et al.
-(2013) velocity model, and so was actually developed by the latter’s authors.
+Then, we calculated bulk carbon efflux using four models for average flow velocity:  BIKER and three rating curves previously used for $CO_2$ upscaling [@lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a; @raymondScalingGasTransfer2012a].
+See Table S1 for their definitions.
+It is worth stressing that the [@lauerwaldSpatialPatternsCO22015] model was developed as one of two components of the [@raymondScalingGasTransfer2012a] velocity model, and so was actually developed by the latter’s authors.
 The names used here refer to the final implementation used in each study’s upscaling exercises.
@@
 Figure 3 plots the validations for the five upscaling functions tested.
-Both the original Ulseth et al.
-(2019) model, and our RS-able extension, perform the best on this independent set of validation data, with near identical performance.
-The RS-able Ulseth et al.
-(2019) model achieves an $r^2$ of `r results_3_1[results_3_1$Model == 'RS-able',]$r2` and an RMSE of `r results_3_1[results_3_1$Model == 'RS-able',]$rmse` m/day, which is just marginally better than the original Ulseth et al.
-(2019) model (`r results_3_1[results_3_1$Model == 'Ulseth',]$r2` and `r results_3_1[results_3_1$Model == 'Ulseth',]$rmse` m/day, respectively).
-The two power-law based Raymond models (Raymond et al.
-(2012) equations 4 and 3) achieve slightly worse performance, with $r^2$ in the 0.70s and their RMSE scores slightly higher (`r results_3_1[results_3_1$Model == 'Raymond_2',]$rmse` and `r results_3_1[results_3_1$Model == 'Raymond_3',]$rmse` m/day, respectively).
-Raymond et al.
-(2012) equation 5, which is linear in structure, worked well on their dataset but fails to capture the non-linear relationship evident in this expanded and more geomorphically diverse dataset.
+Both the original [@ulsethDistinctAirWater2019a] model, and our RS-able extension, perform the best on this independent set of validation data, with near identical performance.
+The RS-able [@ulsethDistinctAirWater2019a] model achieves an $r^2$ of `r results_3_1[results_3_1$Model == 'RS-able',]$r2` and an RMSE of `r results_3_1[results_3_1$Model == 'RS-able',]$rmse` m/day, which is just marginally better than the original [@ulsethDistinctAirWater2019a] model (`r results_3_1[results_3_1$Model == 'Ulseth',]$r2` and `r results_3_1[results_3_1$Model == 'Ulseth',]$rmse` m/day, respectively).
+The two power-law based Raymond models ([@raymondScalingGasTransfer2012a] equations 4 and 3) achieve slightly worse performance, with $r^2$ in the 0.70s and their RMSE scores slightly higher (`r results_3_1[results_3_1$Model == 'Raymond_2',]$rmse` and `r results_3_1[results_3_1$Model == 'Raymond_3',]$rmse` m/day, respectively).
+[@raymondScalingGasTransfer2012a] equation 5, which is linear in structure, worked well on their dataset but fails to capture the non-linear relationship evident in this expanded and more geomorphically diverse dataset.
 In fact, it predicts negative $k_{600}$ values and error metrics cannot be calculated because the prediction residuals are not normally distributed.
-Because we successfully reproduced the Ulseth et al.
-(2019) model, the rest of this manuscript deals exclusively with this RS-able variant of the model.
+Because we successfully reproduced the [@ulsethDistinctAirWater2019a] model, the rest of this manuscript deals exclusively with this RS-able variant of the model.
 
@@
 
-Uncertainties were then propagated through the RS-able Ulseth et l.
-(2019) model via MC simulations (section 2.2) for the 8,000 hydraulic measurements mapped in Figure 4c.
+Uncertainties were then propagated through the RS-able [@ulsethDistinctAirWater2019a] model via MC simulations (section 2.2) for the 8,000 hydraulic measurements mapped in Figure 4c.
 Figure 4a plots a histogram of the 8,000 uncertainty terms extracted from those distributions (Figure 4b plots three of the distributions used to build Figure 4a).
@@
 Uncertainty solely from the upscaling model is 1.30, confirming that using Manning’s equation to approximate average flow velocity is introducing very little additional uncertainty to the model (Figure S2).
-This also confirms the large residuals inherent in $k_{600}$ upscaling models as described by Hall & Ulseth, (2020).
+This also confirms the large residuals inherent in $k_{600}$ upscaling models as described by [@hallGasExchangeStreams2020].
 
@@
 This is borne out in the cumulative density functions (CDFs) in Figure 8b, which plot the distribution of all FCO2 estimates across all rivers and timesteps for each velocity model.
-BIKER nearly perfectly captures FCO2 estimates above approximately the 60th percentile, while the Lauerwald, et al.
-(2015) model nearly perfectly captures FCO2 below the 60th percentile.
+BIKER nearly perfectly captures FCO2 estimates above approximately the 60th percentile, while the [@lauerwaldSpatialPatternsCO22015] model nearly perfectly captures FCO2 below the 60th percentile.
 Because the bulk efflux is more sensitive to peak evasion events, at the bulk properties level BIKER more accurately models carbon efflux (Figure 8a) than the rating curves do.
-The Raymond et al.
-(2012) model, which was mostly trained on small streams, uniformly underestimates FCO2 and thus bulk carbon efflux.
-The Raymond et al.
-(2013) model, which is an average of the other two rating curve estimates (Table S1), is in between.
+The [@raymondScalingGasTransfer2012a] model, which was mostly trained on small streams, uniformly underestimates FCO2 and thus bulk carbon efflux.
+The [@raymondGlobalCarbonDioxide2013a] model, which is an average of the other two rating curve estimates (Table S1), is in between.
 
@@
 
-By-river performance scores for $FCO_2$ (Figure S3) show the Lauerwald et al.
-(2015) model as the best performing in KGE, NRMSE and RRMSE, while BIKER arguably has the least prediction bias (rBIAS) of the models and is approximately the same in KGE as the Raymond et al.
-(2013) model.
-Again, the high performance of the Lauerwald et al.
-(2015) and Raymond et al.
-(2013) models makes sense as the rating curves rely on the in-situ discharge record.
-Across all four metrics, the Raymond et al.
-(2012) model is the worst, performing marginally worse than BIKER.
+By-river performance scores for $FCO_2$ (Figure S3) show the [@lauerwaldSpatialPatternsCO22015] model as the best performing in KGE, NRMSE and RRMSE, while BIKER arguably has the least prediction bias (rBIAS) of the models and is approximately the same in KGE as the [@raymondGlobalCarbonDioxide2013a] model.
+Again, the high performance of the [@lauerwaldSpatialPatternsCO22015] and [@raymondGlobalCarbonDioxide2013a] models makes sense as the rating curves rely on the in-situ discharge record.
+Across all four metrics, the [@raymondScalingGasTransfer2012a] model is the worst, performing marginally worse than BIKER.
 
@@
 However, Figures 8 and S3 still confirm that rating curves with a streamgauge are preferred when possible, particularly when interested in properties finer than the bulk efflux (Figure S3).
-Further, methods like the Raymond et al.
-(2013) and Lauerwald et al.
-(2015) models are implementable in rivers of any size, while BIKER is limited to the largest rivers on Earth due to SWOT’s coarse spatial resolution.
+Further, methods like the [@raymondGlobalCarbonDioxide2013a] and [@lauerwaldSpatialPatternsCO22015] models are implementable in rivers of any size, while BIKER is limited to the largest rivers on Earth due to SWOT’s coarse spatial resolution.
 However, our results suggest that, upon SWOT's launch, a BIKER-style approach to estimating gas exchange could be coupled with existing upscaling workflows to improve gas flux predictions where gauges are unavailable.
@@
 ###4.2 Exploring spatiotemporal dynamics of $k_{600}$
-To date, most field-scale studies of riverine gas exchange have focused on 1) its relationship with wind speed (e.g.
-Beaulieu et al.
-2012; Borges et al.
-2004; Zappa et al.
-2007), 2) average flow velocity (e.g.
-Alin et al.
-2011; Beaulieu et al 2012; Schelker et al.
-2016), 3) discharge (e.g.
-Roberts et al.
-2007; Uehlinger & Naegeli, 1998), or 4) oxygen diel curves (where gas exchange is still assumed constant in time- e.g.
-Appling et al.
-2018; Demars et al.
-2015).
+To date, most field-scale studies of riverine gas exchange have focused on 1) its relationship with wind speed [e.g @beaulieuControlsGasTransfer2012a; @borgesVariabilityGasTransfer2004; @zappaEnvironmentalTurbulentMixing2007], 2) average flow velocity [e.g.
+@alinPhysicalControlsCarbon2011; @beaulieuControlsGasTransfer2012a; @schelkerCO2EvasionSteep2016a], 3) discharge [e.g.
+@robertsMultipleScalesTemporal2007; @uehlingerEcosystemMetabolismDisturbance1998], or 4) oxygen diel curves (where gas exchange is still assumed constant in time- e.g.
+@applingOvercomingEquifinalityLeveraging2018; @demarsStreamMetabolismOpen2015).
 This leaves the spatiotemporal dynamics of gas exchange weakly constrained.
-A few studies have investigated these dynamics, but these studies have been limited to individual rivers and/or limited field seasons (e.g.
-Hall et al.
-2012; Sand-Jensen & Staher, 2012).
-Wallin et al.
-2011 performed a preliminary analysis in northern Sweden relating river specific (‘at-a-station’) temporal variability in gas exchange with channel slope, but they were limited to an average of only 8 measurements per river in a single watershed.
+A few studies have investigated these dynamics, but these studies have been limited to individual rivers and/or limited field seasons [@hallAirWaterOxygen2012; @sand-jensenCO2DynamicsDanish2012].
+[@wallinSpatiotemporalVariabilityGas2011] performed a preliminary analysis in northern Sweden relating river specific (‘at-a-station’) temporal variability in gas exchange with channel slope, but they were limited to an average of only 8 measurements per river in a single watershed.
 Thus, there is a fundamental knowledge gap in understandings the geomorphic and hydraulic drivers of riverine gas exchange.
-This limitation is due both to a lack of process-level understanding (Hall & Ulseth, 2020) but also due to a lack of measurements.
-Authors have argued that the key to explaining the large residual variation in upscaling models is to explore at-a-station temporal variability in $k_{600}$ (Hall & Ulseth, 2020).
+This limitation is due both to a lack of process-level understanding [@hallGasExchangeStreams2020] but also due to a lack of measurements.
+Authors have argued that the key to explaining the large residual variation in upscaling models is to explore at-a-station temporal variability in $k_{600}$ [@hallGasExchangeStreams2020].
 
-Therefore, to explore the spatiotemporal dynamics of $k_{600}$ at fine temporal resolution and but also at the global-scale, SWOT data is an attractive option for upscaling (providing daily hydraulic measurements for a 3 month fast sampling period for calibration and validation and sampling thereafter between 1 and 7 days per 21 day repeat cycle thereafter- Biancarma, et al.
-2016).
+Therefore, to explore the spatiotemporal dynamics of $k_{600}$ at fine temporal resolution and but also at the global-scale, SWOT data is an attractive option for upscaling because it provides daily hydraulic measurements for a 3 month fast sampling period for calibration and validation and sampling thereafter between 1 and 7 days per 21 day repeat cycle thereafter [@biancamariaSWOTMissionIts2016].
 BIKER's success on simulated SWOT rivers bodes well for its implementation on real SWOT data, and functionally opens the door for daily estimation of riverine gas exchange globally once SWOT launches.
@@
 ###4.3 Biases in upscaling k600
-Systematic sampling biases in the data used to develop $k_{600}$ upscaling models (Ulseth et al. 2019; Raymond et al. 2012).
+Systematic sampling biases in the data used to develop $k_{600}$ upscaling models [@ulsethDistinctAirWater2019a; @raymondScalingGasTransfer2012a].
 
 Finally, in reasonably wide rivers, wind begins to exert a non-trivial influence on gas exchange.
-It is well established that in lakes and the ocean, wind controls near-surface turbulence, and thus gas exchange (Beaulieu et al.
-2012; Read et al.
-2012).
-Authors have argued that large rivers are a hybrid of the hydraulics-driven turbulence in small rivers and the wind-driven turbulence in lakes (Beaulieu et al.
-2012).
+It is well established that in lakes and the ocean, wind controls near-surface turbulence, and thus gas exchange [@beaulieuControlsGasTransfer2012a; @readLakesizeDependencyWind2012].
+Authors have argued that large rivers are a hybrid of the hydraulics-driven turbulence in small rivers and the wind-driven turbulence in lakes [@beaulieuControlsGasTransfer2012a].
 As SWOT will measure only rivers wider than 50m, it follows that wind is likely exerting some influence on gas exchange in SWOT-observable reaches.
@@
 BIKER is available at https://github.com/craigbrinkerhoff/BIKER.
-All results and code to generate results, figures, and the manuscript are archived at zenodod something something.
+All results and code to generate results, figures, and the manuscript are archived at zenodo something something.
 Data used in this study is available in Frasson et al.
@@
 (2019).
-This project benefitted from early comments from Tamlin Pavelsky, as well as the large body of existing knowledge from the SWOT River Discharge Working Group.
+This manuscript benefitted from early comments from Tamlin Pavelsky, as well as the large body of existing knowledge from the SWOT River Discharge Working Group.
+
+## 7 References

### 2021-02-12 manuscript_v1.Rmd
message

@@
 ---
-title: "Remote sensing the gas exchange velocity and carbon efflux using the SWOT Satellite"
+title: "Remote sensing riverine gas exchange from SWOT and implications for estimating carbon efflux from rivers"
 author: "CB Brinkerhoff, CJ Gleason, PA Raymond, MH Harlan"
@@
 bibliography: References.bib
+csl: american-geophysical-union.csl
 output:
@@
 
+##Key Points
+- We introduce the BIKER algorithm to remotely sense riverine gas exchange using the SWOT satellite and no in situ information
+- BIKER can nearly replicate estimates of $CO_2$ fluxes made via hydraulic geometry and an in situ streamgauge
+- Expected SWOT measurement errors exert a trivial influence on BIKER's estimates
+- BIKER's accuracy is limited by process-level uncertainties in gas exchange from rivers and not by its algorithmic setup
+
+##Abstract
+Go BIKER!!
+x
+x
+x
+x
+x
+x
+x
+x
+x
+x
+
 ##1 Introduction
-Greenhouse gas (GHG) emissions from rivers and lakes are significant [@colePlumbingGlobalCarbon2007a].
-Rivers are usually supersaturated with dissolved GHGs and present a net flux of these gases from water to air [@coleCarbonCatchmentsConnecting2001a; @jonesLongtermDeclineCarbon2003].
+The inland water drainage network both transports and transforms various constiutents as they move from the landscape to the ocean and a byproduct of many of these processes is riverine greenhouse gas (GHG) production [@colePlumbingGlobalCarbon2007a].
+Rivers are usually supersaturated with dissolved GHGs and present a significant flux of these gases from water to air [@raymondGlobalCarbonDioxide2013a].
 This flux $[M/L^2T]$ is calculated, with in situ knowledge of the gas concentration gradient between the water $C_{water} [M/L^3]$ and the air $C_{air} [M/L^3]$ and the gas exchange velocity $k [L/T]$, as equation 1.
+Note that *k* is generally studied as normalized by a Schmidt number of 600 ($k_{600}$) to be gas and temperature agnostic [see @hallGasExchangeStreams2020 for a thorough review of gas exchange in rivers].
 
-$$Flux=k(C_{water}-C_{air}) (1)$$
+$$Flux=k(C_{water}-C_{air}) \text{(1)}$$
 
-k is largely a function of water column turbulence, at least in fluvial settings.
-This is because k scales with diffusivity as normalized by the boundary-layer thickness, and in rivers turbulence defines the thickness of the boundary-layer (@hallGasExchangeStreams2020 for a thorough review of gas exchange velocity in rivers).
-Due to its reliance on diffusivity (as well as kinematic viscosity), k is defined differently for different GHGs [@raymondScalingGasTransfer2012a; @wanninkhofRelationshipWindSpeed1992a].
-It is also often measured using different gases than those being studied, so a variant of k that is normalized to a Schmidt number of 600 ($k_{600} [L/T]$) is generally used to be GHG and temperature agnostic [@hallGasExchangeStreams2020].
-This is synonymous with a k for carbon dioxide ($k_{CO2}$) at 20 degrees Celsius.
+It is infeasible to measure $k_{600}$ directly when working at the drainage-network-scale and so ‘upscaling functions’ are used that predict $k_{600}$ from readily available river geomorphology [e.g.
+@oconnorMechanismReaerationNatural1958; @palumboAssessingPerformanceReaeration2014; @raymondScalingGasTransfer2012a].
+Many upscaling functions are based on the basic theory that *k* is empirically correlated with the turbulent energy dissipation *e* at the air-water interface [@zappaEnvironmentalTurbulentMixing2007].
+This is convenient for upscaling because in rivers and streams *e* can be equated to the turbulent energy dissipation rate *eD* $[L^2/T^3]$ [@ulsethDistinctAirWater2019a; @raymondScalingGasTransfer2012a], which is estimated using easily calculated hydraulic parameters: $eD \approx gSV$ where *g* is gravitational acceleration $[L/T^2]$, *S* is channel slope, and *V* is average flow velocity $[L/T]$ [@tsivoglouTracerMeasurementReaeration1976].
+Because *S* is readily available in any hydrographic data product, most efforts to upscale $k_{600}$ to drainage networks are limited by estimates of the *V* of the river channel.
+Thus, drainage-network-scale GHG emissions estimates are sensitive to their *V* data.
+This is exasperated in ungauged basins where *V* models cannot be trained because no in situ hydraulic information is available and fieldwork is impractical [@gleasonRemoteSensingRiver2020].
 
-It is infeasible to measure $k_{600}$ directly when working at the river network scale and so ‘upscaling functions’ are used that predict $k_{600}$ from readily available river geomorphology [e.g.
-@oconnorMechanismReaerationNatural1958; @palumboAssessingPerformanceReaeration2014; @raymondScalingGasTransfer2012a].
-Following the theory above, many upscaling functions are based on the basic concept that k is empirically correlated with the turbulent energy dissipation e at the air-water interface [@zappaEnvironmentalTurbulentMixing2007].
-This is convenient for upscaling because in rivers and streams e can be equated to the turbulent energy dissipation rate eD $[L^2/T^3]$ [@ulsethDistinctAirWater2019a; @raymondScalingGasTransfer2012a], which is estimated using easily calculated hydraulic parameters (eD≈gSV where g is gravitational acceleration, S is channel slope, and V is average flow velocity- @tsivoglouTracerMeasurementReaeration1976).
-It has been observed that this scaling relationship is particularly strong in higher energy streams where bubble-mediated gas exchange occurs, though the general scaling function holds across all rivers and streams [@ulsethDistinctAirWater2019a].
-Despite reasonable predictive accuracy, these upscaling models suffer from residual variation indicating large predictive error (Wallin et al.
-2011).
-Still, these are the current best options for estimating $k_{600}$ in rivers.
-It has been suggested that exploring at-a-station hydraulic variation (and thus $k_{600}$ variation) is a promising approach to reducing the residual variation in upscaling models [@hallGasExchangeStreams2020].
+To circumvent this problem in ungauged basins/at drainage-network-scales, to date researchers have used ‘global-scope’ hydraulic geometry (HG) models: simple power law relations between hydraulic properties and discharge [@leopoldHydraulicGeometryStream1953a] that are trained on extensive hydraulics datasets.
+This is the approach used in upscaling *V* to continental/global river networks and has proven useful in those settings [e.g.
+@horgbyUnexpectedLargeEvasion2019; @borgesVariabilityGasTransfer2004; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a].
+However, it has long been established that HG parameters exhibit high spatial variability across landscapes and vary significantly from river to river [@parkWorldwideVariationsHydraulic1977].
+Because of this, a large body of geomorphology work has attempted to parse out process-based explanations for HG parameters [e.g.
+@dingmanAnalyticalDerivationAtastation2007b; @parkerPhysicalBasisQuasiuniversal2007b; @singhTHEORIESHYDRAULICGEOMETRY2003a; @fergusonHydraulicsHydraulicGeometry1986a] and an open research question is how to best prescribe a set of HG parameters to a given river (particularly at the global scale).
+A potential alternative to this approach is to directly estimate a river’s hydraulic properties using remote sensing (RS) data.
 
-Because channel slope is readily available in any hydrographic data product, most efforts to upscale $k_{600}$ to river networks are limited by estimates of the reach-averaged flow velocity of the river channel.
-Therefore, network-scale GHG flux estimates exist in some part at the mercy of the quality of average flow velocity data.
-Ideally, a worker can directly measure velocity at some permanently or temporarily gauged river in their system and build velocity~discharge rating curves to upscale to the entire river network.
-Rating curves are also referred to as hydraulic geometry (HG) models: simple power law relations between hydraulic properties and discharge [@leopoldHydraulicGeometryStream1953a].
-However, this is not feasible in ungauged basins (whether physically or politically ungauged- @gleasonRemoteSensingRiver2020) where no in situ hydraulic information is available and fieldwork is impractical.
-Thus, in ungauged settings, present understandings of GHG fluxes from rivers are likely limited by our access to velocity and/or discharge data.
+Remote hydraulic estimation is itself a burgeoning subfield within remote sensing of hydrology and an analogue to the problem presented here is the remote sensing of river discharge (RSQ- see @gleasonRemoteSensingRiver2020 for a thorough review).
+There are many ways to perform RSQ, from calibrating hydrologic and hydraulic models [] to methods that use no in situ information ('ungauged approaches') [].
+Many of these ungauged approaches were developed in the context of the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission.
 
-To circumvent this problem in ungauged basins/at massive spatial scales, researchers have used ‘global-scope’ HG models that use a single set of model parameters trained on large hydraulics datasets.
-This is the approach used in upscaling average flow velocity to continental/global river networks and has proven quite useful in those settings (e.g.
-@horgbyUnexpectedLargeEvasion2019; @borgesVariabilityGasTransfer2004; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a).
-However, it has long been established that HG parameters exhibit high spatial variability across landscapes and vary significantly from river to river [@parkWorldwideVariationsHydraulic1977].
-Because of this, a large body of geomorphology work has attempted to parse out process-based explanations for HG parameters (e.g.
-@dingmanAnalyticalDerivationAtastation2007b; @parkerPhysicalBasisQuasiuniversal2007b; @singhTHEORIESHYDRAULICGEOMETRY2003a; @fergusonHydraulicsHydraulicGeometry1986a) and an open research question is how to best prescribe a set of HG parameters to a given river (particularly at the global scale).
-A potential alternative to this approach, in ungauged basins at least, is to directly estimate a river’s hydraulic properties from remote sensing (RS) data.
-
-The problem of hydraulic estimation in ungauged basins is itself a burgeoning subfield within fluvial hydrology and an analogue to the problem presented here is the ungauged remote sensing of river discharge (RSQ- see @gleasonRemoteSensingRiver2020 for a thorough review).
+SWOT will launch in 2022 and provide the world’s first global measurements of water surface extent and elevation at novel temporal resolutions.
+SWOT is a wide swath Ka-band radar interferometer and will sample rivers every 1 to 7 days per 21 day repeat cycle.
+It will measure rivers wider than 100m with a goal of expanding this to rivers at least 50m wide [@biancamariaSWOTMissionIts2016].
+In the context of RSQ, a decade of SWOT work has explored the multi-parameter problem of estimating roughness and bathymetry to produce the SWOT discharge product [e.g.
+@andreadisConstrainingAssimilationSWOT2020d; @brinkerhoffConstrainingRemoteRiver2020; @durandEstimatingReachaveragedDischarge2014a; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014a; @hagemannBAMBayesianAMHGManning2017a; @larnierRiverDischargeBathymetry2020a; @oubanasDischargeEstimationUngauged2018].
 A specific subset of these ungauged RSQ methods are termed ‘McFLIs’ or ‘Mass Conserved Flow Law Inversion’ algorithms (e.g.
-@andreadisConstrainingAssimilationSWOT2020d; @brinkerhoffConstrainingRemoteRiver2020; @durandEstimatingReachaveragedDischarge2014a; HIVDI; @hagemannBAMBayesianAMHGManning2017a).
+@andreadisConstrainingAssimilationSWOT2020d; @brinkerhoffConstrainingRemoteRiver2020; @durandEstimatingReachaveragedDischarge2014a).
 They use basic geomorphic theories, rather than full hydrologic and/or hydraulic models, and the concept of ‘prior river knowledge’ to estimate discharge from RS data where not all hydraulic terms are RS-able [@gleasonRemoteSensingRiver2020; @gleasonTrackingRiverFlows2017].
-McFLIs have largely been developed in the context of the NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) mission, which will launch in 2022 and provide the world’s first global measurements of water surface extent and elevation at novel temporal resolutions [@biancamariaSWOTMissionIts2016].
-On average SWOT will sample rivers every 11 days and provide a unique opportunity to measure river and lake hydraulics at the global scale.
-While the raw performance of McFLIs is lower than a more sophisticated approach, they are globally flexible, easily implementable in any river that SWOT can sample, and improve our understandings of ungauged rivers when little to no information was previously available [@brinkerhoffConstrainingRemoteRiver2020; @durandIntercomparisonRemoteSensing2016b].
-Recently, McFLIs have also shown promise in providing additional and beneficial information to traditional hydrologic modeling via data assimilation [@ishitsukaCombiningOpticalRemote].
+While the raw performance of McFLIs is lower than the more sophisticated approaches mentioned above, they are globally flexible, easily implementable in any river that SWOT can sample, and improve our hydrological understandings of ungauged rivers when little to no information was previously available [@brinkerhoffConstrainingRemoteRiver2020; @durandIntercomparisonRemoteSensing2016b].
+Recently, McFLIs have also shown promise in providing additional and beneficial information to traditional hydrologic modeling via data assimilation [@ishitsukaCombiningOpticalRemote2020].
 
 McFLIs often employ Bayesian inference.
-Bayesian techniques have been previously used to estimate $k_{600}$ concurrently with stream metabolism from dissolved oxygen (O2) datasets [@applingOvercomingEquifinalityLeveraging2018; @graceFastProcessingDiel2015; @holtgrieveSimultaneousQuantificationAquatic2010].
-While this approach requires detailed field data and is thus not applicable to $k_{600}$ estimation in ungauged rivers, it does suggest that Bayesian inference could be useful for estimating $k_{600}$ when parameter equifinality is a problem (at it is with McFLIs).
+Bayesian techniques have been previously used to estimate $k_{600}$ concurrently with stream metabolism from dissolved oxygen ($O_2$) datasets [@applingOvercomingEquifinalityLeveraging2018; @graceFastProcessingDiel2015; @holtgrieveSimultaneousQuantificationAquatic2010].
+While this approach requires detailed field data and is thus not applicable to $k_{600}$ estimation in ungauged rivers, it does suggest that Bayesian techniques could be useful for estimating $k_{600}$ when parameter equifinality is a problem (as it is with McFLIs).
 
-Within this context, we sought to explore estimating $k_{600}$ in ungauged basins with no in situ information.
-We hypothesize that remote sensing $k_{600}$ is possible by combining 1) hydraulic geometry theory, 2) $k_{600}$ upscaling theory, 3) Bayesian inference (in a manner similar to McFLI), and 4) SWOT data.
-More specifically, this chapter aims to answer the following questions:
-	1) Is upscaling the gas exchange velocity with no a priori knowledge of eD possible?
-	2) Is ungauged RS of gas exchange velocity possible?
-	3) How might this algorithm influence upscaling of riverine carbon efflux?
+Thus, to advance our understanding of riverine GHG fluxes in ungauged basins and/or at drainage-network-scales, we sought to estimate $k_{600}$ using only RS data and no in situ information.
+We hypothesize that remote sensing $k_{600}$ is possible by coupling 1) hydraulic geometry theory, 2) $k_{600}$ upscaling theory, 3) Bayesian remote sensing (in a manner similar to McFLI), and 4) SWOT data.
+More specifically, this manuscript aims to answer two questions:
+	1) Is ungauged RS of the gas exchange velocity possible?
+	2) How might this algorithm influence upscaled estimates of riverine carbon efflux to the atmosphere?
 
 To answer these questions, we build a RS algorithm that ingests SWOT data and produces $k_{600}$ estimates (with fully quantified uncertainty within a Bayesian framework) that requires no in situ information about a river’s hydraulic properties.
-We validate the algorithm on simulated SWOT data for 22 rivers.
-Finally, we compare our algorithm against established methods: we use previously published carbon dioxide ($CO_2$) data to compare the bulk carbon efflux (via $CO_2$ evasion) from the 22 rivers as calculated using both our new algorithm and established velocity rating curves and in situ discharge records.
+We validate the algorithm on simulated SWOT data for 22 rivers (as SWOT has yet to launch, simulated data is necessary- Section 2.3).
+We also quantify the influence of the expected measurement errors in SWOT data on our $k_{600}$ estimates.
+Finally, we compare our algorithm against published methods: we use previously published dissolved carbon dioxide ($CO_2$) data to compare the bulk carbon efflux (via $CO_2$ evasion) from the 22 rivers as calculated using both our new algorithm and published velocity rating curves and in situ discharge records.
 
 ##2 Methods
-We name the RS of $k_{600}$ algorithm BIKER, or the ‘Bayesian Inversion/Inference of the $k_{600}$ Rate’ algorithm.
-To build BIKER, two steps are taken: first, we develop a $k_{600}$ upscaling model that does not rely on a priori knowledge of eD (section 2.1) and then we use that upscaling model to build the actual RS algorithm (section 2.2).
+We name the RS of $k_{600}$ algorithm BIKER, or the ‘Bayesian Inversion of the $k_{600}$ Rate’ algorithm.
+To build BIKER, two steps are taken: first, we develop a $k_{600}$ upscaling model that does not rely on a priori knowledge of the river (section 2.1) and then we use that upscaling model to build the actual RS algorithm (section 2.2).
 Following these methods, the validation setup is described (section 2.3), as is the comparison of estimated bulk carbon effluxes from a suite of average flow velocity models (section 2.4).
@@
 ###2.1 Developing an RS-able $k_{600}$ upscaling model
-[@raymondScalingGasTransfer2012a] first used river hydraulics and the theoretical relationship between the turbulent dissipation rate (eD) and gas exchange to upscale $k_{600}$ from average flow velocity and channel slope, extending earlier work predicting the reaeration coefficient (e.g.
+[@raymondScalingGasTransfer2012a] first used river hydraulics and the theoretical relationship between *eD* and gas exchange to upscale $k_{600}$ from average flow velocity and channel slope, extending earlier work predicting the reaeration coefficient (e.g.
 @oconnorMechanismReaerationNatural1958; @tsivoglouTracerMeasurementReaeration1976).
 Later, [@ulsethDistinctAirWater2019a] built upon [@raymondScalingGasTransfer2012a]’s work by expanding their dataset to include measurements from steeper rivers and finding that two distinct upscaling regimes exist in low and high energy rivers.
-More specifically, [@ulsethDistinctAirWater2019a] scale $k_{600}$ using eD, where the resulting upscaling model parameters are significantly different whether eD is high or low.
+More specifically, [@ulsethDistinctAirWater2019a] scale $k_{600}$ using eD, where the resulting upscaling model parameters are significantly different whether *eD* is high or low.
 Their upscaling function is reprinted as equation 2.
 
-$$ln(k_{600})=3.10 + 0.35ln(eD) if eD < 0.02  (2) \\
+$$ln(k_{600})=3.10 + 0.35ln(eD) if eD < 0.02  \text{(2)}$$
+$$ln(k_{600})=6.43+1.18ln(eD) if eD > 0.02$$
 
-ln(k_{600})=6.43+1.18ln(eD) if eD > 0.02$$
-
-While equation 2 would be ideal for the purposes outlined for this manuscript and provides best-to-date predictive accuracy for upscaling $k_{600}$, it requires an a priori knowledge of eD to assign its differential model parameters.
+While equation 2 would be ideal for the purposes outlined for this manuscript and provides best-to-date predictive accuracy for upscaling $k_{600}$, it requires an a priori knowledge of *eD* to assign its differential model parameters.
 We do not have this a priori knowledge in an ungauged setting and only have access to SWOT-observed river widths and heights (and therefore water surface slopes).
@@
 
-We combined these five regression models into a ‘rule-based regression model’ (equation 3) for river size i and model parameters a and b.
-Per Figure 2, there are five possible sets of a and b parameters, as defined for rivers 10-50m wide, 50-100m wide, 100+m wide, low energy rivers 0-10m wide (slope < 0.05), and high energy rivers 0-10m wide (slope > 0.05).
+We combined these five regression models into a ‘rule-based regression model’ (equation 3) for river size *i* and model parameters *a* and *b*.
+Per Figure 2, there are five possible sets of *a* and *b* parameters, as defined for rivers 10-50m wide, 50-100m wide, 100+m wide, low energy rivers 0-10m wide (slope < 0.05), and high energy rivers 0-10m wide (slope > 0.05).
 
-$$k_{600}=a_i(eD)^{b_i} (3)$$
+$$k_{600}=a_i(eD)^{b_i} \text{(3)}$$
 
-This model therefore assigns ai and bi a priori using just river width and slope to define i.
-Another implication of this extension of the [@ulsethDistinctAirWater2019a] model is that we now use a and b parameters that implicitly reflect the distinct geomorphology of different sized rivers.
+This model therefore assigns $a_i$ and $b_i$ a priori using just river width and slope to define *i*.
+Another implication of this extension of the [@ulsethDistinctAirWater2019a] model is that we now use *a* and *b* parameters that implicitly reflect the distinct geomorphology of different rivers.
 
 Finally, we tested whether this model can reasonably estimate $k_{600}$ relative to the original [@ulsethDistinctAirWater2019a] model, with the goal of at least replicating their performance.
-We compared it against the various eD-based upscaling models from [@raymondScalingGasTransfer2012a] that are commonly used in the literature.
+We compared it against the various *eD*-based upscaling models from [@raymondScalingGasTransfer2012a] that are commonly used in the literature.
 This was done by training all upscaling functions on 80% of the [@ulsethDistinctAirWater2019a] data and comparing their performance on the remaining 20% of the data.
@@
 ###2.2 Developing the BIKER algorithm
-With a validated upscaling model that requires no a priori knowledge of eD (results presented in section 3.1) we turn to developing the algorithm to RS $k_{600}$.
+With a validated upscaling model that requires no a priori knowledge of *eD* (results presented in section 3.1) we turn to developing the algorithm to RS $k_{600}$.
 The approach used here is heavily informed by [@hagemannBAMBayesianAMHGManning2017a]’s algorithm for ungauged RSQ via Bayesian inference.
@@
 (2017) which formulates depth via river width and area (in order to use SWOT-observables).
-$k_{600}$ is now defined as a function of $S_h$, channel roughness n $[T/L^{\frac{1}{3}}]$, channel width W [L], and channel area $A=A_0+dA [L^2]$ for the change in channel area dA).
+$k_{600}$ is now defined as a function of $S_h$, channel roughness n $[T/L^{\frac{1}{3}}]$, channel width *W* [L], and channel area $A=A_0+dA [L^2]$ for the change in channel area *dA*).
 It is also defined by the upscaling parameters $a_i$ and $b_i$.
 
-$$k_{600}=a_i(gS_h^{\frac{3}{2}}\frac{1}{n}(A_0+dA)^\frac{2}{3}W^\frac{-2}{3})^{b_i} (4)$$
+$$k_{600}=a_i(gS_h^{\frac{3}{2}}\frac{1}{n}(A_0+dA)^\frac{2}{3}W^\frac{-2}{3})^{b_i} \text{(4)}$$
 
@@
 
-$$a_i^{-1}(gS_h^\frac{-3}{2}W^\frac{2}{3})^{b_i} \sim normal(k_{600}^{-1}\frac{1}{n}(A_0+dA)^\frac{2}{3})^{b_i}, \sigma^2_{upscale})  (5)$$
+$$a_i^{-1}(gS_h^\frac{-3}{2}W^\frac{2}{3})^{b_i} \sim normal(k_{600}^{-1}\frac{1}{n}(A_0+dA)^\frac{2}{3})^{b_i}, \sigma^2_{upscale})  \text{(5)}$$
 
@@
 Bayesian priors formularize the a priori estimates (and uncertainties) for these parameters.
-Or more intuitively, they represent our ‘prior river knowledge’ of what n, A0, and $k_{600}$ probably are for some river given that they cannot be remotely sensed.
-All priors are formalized within the model as truncated normal distributions of the ln-transformed terms such that $ln(X) \sim N(\mu, \sigma^2)$ for λ<X<ε, using prior hyperparameters mean (μ), standard deviation (σ), and upper (ε) and lower bounds (λ) for parameter X = n, $A_0$, or $k_{600}$.
+Or more intuitively, they represent our ‘prior river knowledge’ of what *n*, $A_0$, and $k_{600}$ probably are for some river given that they cannot be remotely sensed.
+All priors are formalized within the model as truncated normal distributions of the ln-transformed terms such that $ln(X) \sim N(\mu, \sigma^2)$ for λ<X<ε, using prior hyperparameters mean ($\mu$), standard deviation ($\sigma$), and upper (ε) and lower bounds (λ) for parameter X = n, $A_0$, or $k_{600}$.
 In order to avoid relying on in situ information, we assign prior hyperparameters using SWOT data only.
-n and $A_0$ prior hyperparameters were assigned following the method developed by [@brinkerhoffConstrainingRemoteRiver2020].
+*n* and $A_0$ prior hyperparameters were assigned following the method developed by [@brinkerhoffConstrainingRemoteRiver2020].
 They developed a set of river channel prior hyperparameters for McFLI algorithms that are entirely RS-able and reflect differential channel hydraulics as a function of river geomorphology.
 With this method, they showed substantial improvement in the accuracy of their ungauged RSQ estimates with no changes to the algorithms or data used.
-We assigned the $k_{600}$ prior hyperparameters using a simple slope regression model trained on the [@ulsethDistinctAirWater2019a] dataset to assign μ (equation 6).
+We assigned the $k_{600}$ prior hyperparameters using a simple slope regression model trained on the [@ulsethDistinctAirWater2019a] dataset to assign $\mu$ (equation 6).
 $σ^2$ was set to the standard error of equation 6: 1.023.
@@
 
-$$\mu_{k600}=3.22+0.347ln(S_h) if ln(S_h) < -4.634   (6) \\
-
-\mu_{k600}=6.85+1.13ln(S_h) if ln(S_h)>-4.634$$
+$$\mu_{k600}=3.22+0.347ln(S_h) if ln(S_h) < -4.634   \text{(6)}$$
+$$\mu_{k600}=6.85+1.13ln(S_h) if ln(S_h)>-4.634$$
 
@@
 ###2.3 BIKER validation setup
-We validated the BIKER algorithm on 22 rivers using observed average flow velocity and observed slope. First, we detail the 22 rivers, and then we detail how ‘observed’ $k_{600}$ is calculated.
+We validated the BIKER algorithm on 22 rivers using observed $V$ and observed $S_h$ First, we detail the 22 rivers, and then we detail how ‘observed’ $k_{600}$ is calculated.
 
 Because SWOT has yet to launch, it is standard practice to benchmark SWOT-related algorithms on ‘SWOT-like data’.
-There are three current types of SWOT-like data: 1) AirSWOT, which is an airborne Ka-band inSAR currently limited to five rivers globally, 2) simulated rivers that mimic the type of data SWOT will provide, and 3) the SWOT simulator, which introduces errors to these simulated rivers (mostly from radar layover error and random noise- @oubanasDischargeEstimationUngauged2018).
+There are three current types of SWOT-like data: 1) AirSWOT, which is an airborne Ka-band inSAR currently limited to five rivers globally, 2) simulated rivers that mimic the type of data SWOT will provide, and 3) the SWOT simulator, which introduces measurement errors to these simulated rivers.
 Because we are principally interested in algorithm performance, we limit our validation setup to simulated rivers in order to benchmark across as many rivers as possible.
@@
 Simulated rivers mimic perfect measurement conditions (i.e.
-no observation errors) and represent the best-case scenarios of what SWOT will provide to hydrologists.
+no measurement errors) and represent the best-case scenarios of what SWOT will provide to hydrologists.
 Here, we use the 22 rivers archived in [@frassonCompilationHydraulicModels2019a] and used by [@rodriguezObservingRiversVarying2020] to develop reach-averaged Saint Vernant equations within the SWOT context.
-We omitted the Arial Khan river in Bangladesh due to known problems with that hydraulic model and we further sampled the models for only the observations every 11 days to mimic the average SWOT overpass frequency (this also approximately algins the SWOT observations with the bi-weekly CO2 data used later- Section 2.4).
+We omitted the Arial Khan river in Bangladesh due to known problems with that hydraulic model and we further sampled the models for only the observations every 11 days to mimic the average SWOT overpass frequency (this also approximately algins the SWOT observations with the bi-weekly $CO_2$ data used later in this study- Section 2.4).
 All told, this yielded 503 sets of SWOT observations to validate BIKER with.
 
+We also introduced measurement error into these simulated SWOT observations to assess the performance degradation due to errors inherent in SWOT measurements.
+Error in SWOT-measured river heights and water surface slopes will come from tolerances internal to the satellite, i.e.
+the SWOT error budget, as well as radar layover error.
+River width errors are assumed here to be neglible [@durandHowWillRadar2020a].
+Layover error is the phenomenon when radar returns arrive at the sensor at the same time, generally in areas of high topographic variation [@reesPhysicalPrinciplesRemote2013].
+While the SWOT simulator mentioned above is ideal for this analysis, it is cumbersome to run on many rivers and so here we coarsely model measurement error using the results from a recent global anlaysis of SWOT radar layover errors [@durandHowWillRadar2020a].
+They developed a first-principles model for layover error from SWOT observations and, using global hydrography, quantified SWOT measurement error from both error tolerances and radar layover for 220,924 river reaches.
+The 68th percentiles of these reach-scale errors were 1.7cm/km and 10.4cm for water surface slope and height, respectively.
+To apply this model to our 22 rivers, we generate new river heights *H* and water surface slopes $S_h$ by sampling every observation from a normal distribution as defined as $S_{err_{i,t}} \sim normal(S_{obs_{i,t}}, 0.000017^{2})$ and $H_{err_{i,t}} \sim normal(H_{obs_{i,t}}, 0.104^{2})$ (in meters).
+Because BIKER ingests *dA*, rather than *H*, reach-scale height errors were converted to *dA* errors, again following [@durandHowWillRadar2020a]: $\sigma_{dA} \approx W\sigma_H\sqrt{2}$ given our rectangular channel assumptions.
+
 Regardless of the validation data used, we do not have observed $k_{600}$ data for these rivers, and to our knowledge no field dataset of $k_{600}$ exists in the type of temporal and spatial frequency that SWOT (and therefore the BIKER algorithm) provides.
-Further, we are principally interested in our ability to upscale $k_{600}$ from RS observations to avoid relying on velocity rating curves.
-We are less concerend with the actual accuracy of the upscaling model itself (which can be validated using existing datasets).
+Further, we are principally interested in our ability to upscale $k_{600}$ from RS observations and are less concerend with the actual accuracy of the upscaling model itself (which can be validated using existing datasets).
 Therefore, we take the best performing upscaling model from sections 2.1 and 3.1 and use that to calculate the ‘observed’ $k_{600}$ that BIKER is validated against.
 This is calculated using equation 7, where $V_{obs} [L/T]$ is observed river discharge divided by observed channel area.
-With this setup, we can directly explore our ability to infer observed average flow velocity and $k_{600}$ from river width and height alone.
+With this setup, we can directly explore our ability to infer observed *V* and $k_{600}$ from river width and height alone.
 It also means that, for a fair validation scheme, $\sigma_{upscale}^2$ must be set to reflect only error from Manning’s equation (and not $a_i$ and $b_i$).
@@
 
-$$k_{600, obs}=a_i(gS_hV_{obs})^{b_i} (7)$$
+$$k_{600, obs}=a_i(gS_hV_{obs})^{b_i} \text{(7)}$$
 
 Validation metrics take two forms (and are detailed in Table 1).
-To validate across all rivers and timesteps, we used the coefficient of determination r2 and the root mean square error (RMSE).
-Four normalized metrics were used for by-river validation to compare across rivers: RRMSE and NRMSE are normalized root mean square errors that have been normalized by the observed value and the mean observed value, respectively, to compare across rivers.
+To validate across all rivers and timesteps, we used the coefficient of determination $r^2$ and the root mean square error (RMSE).
+Four normalized metrics were used for by-river validation: RRMSE and NRMSE are normalized root mean square errors that have been normalized by the observed value and the mean observed value, respectively, to compare across rivers.
 rBIAS is a measure of prediction bias that is normalized by the mean observed value to compare across rivers.
@@
 σ refers to the variance of the sample and μ refers to the mean of the sample.*
-|Description|Acronym|Definition|Ideal Score|Possible Range|Validation Scheme|
+
+|**Description**|**Acronym**|**Definition**|**Ideal Score**|**Possible Range**|**Validation Scheme**|
 |-----------|-------|----------|-----------|--------------|-----------------|
-|Coefficient of determination|$r^2$| Eq|1|0 to 1|	All rivers and all timesteps|
-|Root-mean-square-error|RMSE| Eq|0|0 to ∞|All rivers and all timesteps|
-|Relative root-mean-square error|RRMSE|Eq |0|0 to ∞|By river|
-|Normalized root-mean-square error|NRMSE|Eq|0|0 to ∞|By river|
-|Relative bias|rBIAS|Eq|0|-∞ to ∞|By river|
-|Kling-Gupta efficiency|KGE|Eq|1|-∞ to 1|	By river|
+|Coefficient of determination|$r^2$| $1-(\frac{\sum\limits_{i=1}^{N_t}{(k_{600, i}-\hat{k_{600}})^2}}{\sum{(k_{600, i}-\bar{k_{600}})^2}})$ |1|0 to 1|	All rivers and all timesteps|
+|Root-mean-square-error|RMSE| $\sqrt{\frac{1}{N_t}\sum\limits_{i=1}^{N_t}(\hat{k_{600, i}}-k_{600, i})^2}$ |0|0 to ∞|All rivers and all timesteps|
+|Relative root-mean-square error|RRMSE| $\sqrt{\frac{1}{N_t}\sum\limits_{i=1}^{N_t}(\frac{\hat{k_{600, i}}-k_{600, i}}{k_{600, i}})^2}$ |0|0 to ∞|By river|
+|Normalized root-mean-square error|NRMSE| $\sqrt{\frac{1}{N_t}\sum\limits_{i=1}^{N_t}(\frac{\hat{k_{600, i}}-k_{600, i}}{\bar{k_{600, i}}})^2}$ |0|0 to ∞|By river|
+|Relative bias|rBIAS| $\frac{1}{N_t}\sum\limits_{i=1}^{N_t}(\frac{\hat{k_{600, i}}-k_{600, i}}{\bar{k_{600, i}}})$ |0|-∞ to ∞|By river|
+|Kling-Gupta efficiency|KGE| $1-\sqrt{(r-1)^2 + (\frac{\hat{\sigma_{k600}}}{\sigma_{k600}}-1)^2 + (\frac{\hat{\mu_{k600}}}{\mu_{k600}}-1)^2}$ |1|-∞ to 1|	By river|
 
 ###2.4 Upscaling to CO2 evasion and bulk carbon efflux
-It is one thing to be able to predict $k_{600}$, but researchers are usually more interested in the gas fluxes from rivers and ultimately the bulk carbon efflux and not just the gas exchange velocity.
-Therefore, we explored 1) the ability of BIKER to reproduce hypothetical carbon dioxide evasion fluxes ($FCO_2 [M/L^2T]$) from these 22 rivers, and 2) the sensitivity of the estimated bulk carbon efflux (via $CO_2$ evasion) to the method used to estimate average flow velocity.
+It is one thing to to predict $k_{600}$, but researchers are usually more interested in the gas fluxes from rivers and ultimately the bulk carbon efflux from river to atmosphere.
+Therefore, we explored 1) the ability of BIKER to reproduce hypothetical carbon dioxide evasion fluxes ($FCO_2 [M/L^2T]$) from these 22 rivers, and 2) the sensitivity of the estimated bulk carbon efflux (via $CO_2$ evasion) to the method used to estimate *V*.
 
 First, we calculated $FCO_2$.
-To do this, we paired the modeled $k_{CO2}$ values (obtained from the $k_{600}$ estimates) with field-measurements of water-side $CO_2$ concentrations.
-29 bi-weekly $CO_2$ samples were made by [@beaulieuControlsGasTransfer2012a]- Figure S1) at one location in the Ohio River.
-We paired these 29 values with each 11-day-repeat set of SWOT observations, ignoring the timesteps beyond 29 (only 7% of the SWOT observations were ignored here and this was deemed acceptable).
-This $CO_2$ data is for the Ohio River only but was applied to all 22 rivers (which includes the Ohio River).
+To do this, we paired the modeled $k_{CO2}$ values (obtained from the $k_{600}$ estimates) with field-measurements of water-side $CO_2$ concentrations and water temperatures.
+26 bi-weekly samples were made by [@beaulieuControlsGasTransfer2012a]- Figure S1) at one location in the Ohio River for one calender year from 2008-2009.
+We paired these 26 values with each 11-day-repeat set of SWOT observations by date, ignoring the timesteps beyond 26 (only ~15% of the SWOT observations were ignored here and this was deemed acceptable).
+Not all of the SWOT rivers have observations for a full year, and when simulation dates were not available they were assumed to start on January 1st.
+Note that this $CO_2$ data is for the Ohio River only but was applied to all 22 rivers (which includes the Ohio River).
 Because we are exclusively interested in the relative differences between $FCO_2$ estimates and not the raw fluxes themselves, any $CO_2$ data representative of SWOT-observable rivers was deemed acceptable.
-Atmospheric $CO_2$ was assumed 390 uatm and water temperature was assumed 25 degrees Celsius.
-The Schmidt number, used to obtain kCO2 from $k_{600}$, was calculated following [@raymondScalingGasTransfer2012a] and [@wanninkhofRelationshipWindSpeed1992a].
+Atmospheric $CO_2$ was assumed 390 uatm.
+The Schmidt number, used to obtain $k_{CO2}$ from $k_{600}$, was calculated following [@raymondScalingGasTransfer2012a] and [@wanninkhofRelationshipWindSpeed1992a].
 
-Then, we calculated bulk carbon efflux using four models for average flow velocity:  BIKER and three rating curves previously used for $CO_2$ upscaling [@lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a; @raymondScalingGasTransfer2012a].
+Then, we estimated bulk carbon efflux using four models for average flow velocity:  BIKER and three HG models previously used for $CO_2$ upscaling [@lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a; @raymondScalingGasTransfer2012a].
 See Table S1 for their definitions.
-It is worth stressing that the [@lauerwaldSpatialPatternsCO22015] model was developed as one of two components of the [@raymondScalingGasTransfer2012a] velocity model, and so was actually developed by the latter’s authors.
+It is worth stressing that the [@lauerwaldSpatialPatternsCO22015] model was developed as one of two components of the [@raymondGlobalCarbonDioxide2013a] model, and so was actually developed by the latter’s authors.
 The names used here refer to the final implementation used in each study’s upscaling exercises.
 For this study, all rating-curve methods use the in-situ discharge record while BIKER does not.
-This allows us to explore the relative differences in bulk carbon efflux estimates if implementing a wholly ungauged method versus two gauged methods (that also use global-scope HG parameters).
-Finally, we express the bulk carbon efflux (via $CO_2$ evasion) [M/T] as the average mass flow rate of carbon per year, i.e.
+This allows us to explore the relative differences in bulk carbon efflux estimates if implementing a wholly ungauged method (BIKER) versus gauged methods (all three HG models)).
+Finally, we express the bulk carbon efflux (via $CO_2$ evasion) $[M/T]$ as the average mass flow rate of carbon per year, i.e.
 multiplied by the surface area of all 22 rivers.
@@
 We then take the best performing upscaling function, implement it within BIKER, and validate it on the 22 SWOT rivers (section 3.2).
-Finally, we use $k_{600}$ as modeled by both BIKER and gauge-based rating curves to compare bulk carbon efflux estimates from the 22 rivers (section 3.3).
+Finally, we use $k_{600}$ as modeled by both BIKER and gauge-based HG models to compare $FCO_2$ and bulk carbon efflux estimates from the 22 rivers (section 3.3).
 
@@
 Figure 4a plots a histogram of the 8,000 uncertainty terms extracted from those distributions (Figure 4b plots three of the distributions used to build Figure 4a).
-Across the 8,000 tests, we quantified average ln-transformed uncertainty in equation 4 to be `r round(results_3_1_MC$meanSigma,2)` (~`r round(exp(results_3_1_MC$meanSigma), 2)` m/day- dashed blue line in Figure S2a).
+Across the 8,000 tests, we quantified average ln-transformed uncertainty in equation 4 to be `r round(results_3_1_MC$meanSigma,2)` (~`r round(exp(results_3_1_MC$meanSigma), 2)` m/day- dashed blue line in Figure 4a).
 Uncertainty solely from the upscaling model is 1.30, confirming that using Manning’s equation to approximate average flow velocity is introducing very little additional uncertainty to the model (Figure S2).
@@
 ###3.2 Validating the BIKER algorithm
-Figure 5a plots the validation results for $k_{600}$ across all 22 rivers and all timesteps.
+Figure 5a plots the validation results for $k_{600}$ (with no measurement error) across all 22 rivers and all timesteps.
 For BIKER, the points are the posterior means while the black lines are the 95% confidence intervals (CIs) for the predictions.
@@
 Regardless, the RMSE for the BIKER predictions is only `r round(10^results_3_2_all$rmse,2)` m/day across all predictions.
-This is substantially less than the RMSEs for the upscaling models (Figure 3) and the MC-simulated uncertainties (Figure 4a), indicating that recovering upscaled $k_{600}$ from SWOT data is much more accurate than the upscaling process itself (which is logical).
+This is substantially less than the RMSEs for the upscaling models (Figure 3) and the MC-simulated uncertainties (Figure 4a), indicating that recovering upscaled $k_{600}$ from SWOT data is much more accurate than the upscaling process itself.
 
@@
 Figure 6a plots validation metrics calculated for each river (i.e.
-the boxplots are composed of metric scores for the 22 rivers- see Table 1 for metric definitions).
+the boxplots are composed of metric scores for the 22 rivers with and without measurement error- see Table 1 for metric definitions).
+SWOT measurement uncertainties exert a neglible influence on 3/4 performance metrics and slightly decrease KGE across the 22 rivers (Figure 6a).
+Thus, the results presented for the rest of the manuscript assume no measurement error.
 Median rBIAS is `r round(median(results_3_2_rivs$rBIAS, na.rm=T),2)`, indicating nearly no bias in most rivers’ predictions.
@@
 While median KGE and rBIAS scores were strong, the ranges of these scores were somewhat large (standard deviation for KGE of `r round(sd(results_3_2_rivs$kge, na.rm=T),2)` and for rBIAS of `r round(sd(results_3_2_rivs$rBIAS, na.rm=T),2)`).
-This suggests that BIKER predicts $k_{600}$ very well in some rivers, but not well in other rivers.
+This suggests that BIKER predicts $k_{600}$ very well in some rivers, but poorly in other rivers.
 
@@
 In the downstream Mississippi River, there is significant bias in the estimates as well as significant uncertainty (per the 95% CIs).
-The temporal dynamics are also quite off.
+The temporal dynamics are also incorrectly predicted.
 
 ###3.3 Implications for carbon cycle estimates
-Finally, we explore our ability to 1) use BIKER-produced $k_{600}$ to estimate $FCO_2$ from rivers and then 2) to estimate bulk carbon efflux from $CO_2$ evasion (section 2.4).
+Finally, we explore our ability to 1) use BIKER-produced $k_{600}$ to estimate $FCO_2$ and estimate bulk carbon efflux from $CO_2$ evasion (section 2.4).
 
 In Figure 7a, there is a strong fit to the observed $FCO_2$ data, with an RMSE of `r round(10^results_3_3_all$rmse,2)` $g/m^2dy$.
-The $r^2$ is slightly lower than that for $k_{600}$ (Figure 7a) due to some abnormally large residuals on a handful of points.
-There is also a slight underestimation bias for the lowest of the observed $FCO_2$, though this is smaller than the bias in the $k_{600}$ predictions (Figure 5a).
+The $r^2$ is identical to that for $k_{600}$ (Figure 7a).
+There is also less underestimation bias for the lowest of the observed $FCO_2$ than in the $k_{600}$ predictions (Figure 5a).
+This is presumably due to the sturcture of the $FCO_2$ equation, which reduces the relative importance of errors in the gas exchange values.
 $FCO_2$ prediction intervals also have a similar relative magnitude to those presented in Figure 5.
-Figure 7b-d includes subplots for the same rivers as Figure 4b-d, however with $FCO_2$ instead of $k_{600}$ There is functionally perfect recovery of $FCO_2$ in the Kanawha River, very good recovery (with some positive bias in peak evasion events) for the Ohio Section 2 River, and substantial negative bias in the Downstream Mississippi river.
+Figure 7b-d includes subplots for the same rivers as Figure 4b-d, however with $FCO_2$ instead of $k_{600}$.
+There is functionally perfect recovery of $FCO_2$ in the Kanawha River, very good recovery (with some positive bias in peak evasion events) for the Ohio Section 2 River, and far less negative bias in the Downstream Mississippi river than is present in the $k_{600}$ estiamtes (Figure 5b).
 Again, the temporal dynamics are accurately modeled however there is some bias in the magnitude of the predictions.
-However, $FCO_2$ is so low in the Mississippi river (Figure 7d) that the absolute bias is not as large as it looks in these timeseries plots.
-This suggests that the BIKER model can easily estimate the changing dynamics of evasion but is at the mercy of the prior on $k_{600}$ to accurately estimate the magnitude of the evasion.
+This suggests that the BIKER model can easily estimate the changing dynamics of gas exchange but is at the mercy of the prior on $k_{600}$ to accurately estimate the magnitude of the evasion.
 
@@
 
-Compared against Figures 5a and 6b-d, there is an arguably stronger recovery of FCO2 than $k_{600}$ (Figure 7).
-This is presumably due to the structure of equation 1, of which $FCO_2$ is a function of both the observed $CO_2$ concentration gradients and $k_{600}$ This means that the errors in the BIKER $k_{600}$ estimates are diminished in relative importance when multiplied by the $CO_2$ concentration gradients.
-
-Finally, we compare the bulk carbon efflux (via CO2 evasion) from the 22 rivers using BIKER posterior means and three gauge-based rating curves (Figure 8).
-In Figure 5.8a are barplots of the bulk carbon efflux across all rivers in gigagrams of carbon per year.
-The BIKER bulk carbon efflux (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_BIKER',]$sumFCO2, 0)` gG-C/yr) is nearly identical to the bulk carbon efflux calculated using observed velocity (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_ulset',]$sumFCO2, 0)` gG-C/yr) while the rating curves underestimate the efflux (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_raymond2012',]$sumFCO2, 0)`, `r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_raymond2013',]$sumFCO2, 0)`, and `r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_Lauerwald2015',]$sumFCO2, 0)` gG-C/yr for ‘Raymond 2012’, ‘Raymond 2013’, and ‘Lauerwald 2015’, respectively).
-BIKER functionally perfectly captures the estimated efflux from these rivers purely from RS data.
-This is borne out in the cumulative density functions (CDFs) in Figure 8b, which plot the distribution of all FCO2 estimates across all rivers and timesteps for each velocity model.
-BIKER nearly perfectly captures FCO2 estimates above approximately the 60th percentile, while the [@lauerwaldSpatialPatternsCO22015] model nearly perfectly captures FCO2 below the 60th percentile.
-Because the bulk efflux is more sensitive to peak evasion events, at the bulk properties level BIKER more accurately models carbon efflux (Figure 8a) than the rating curves do.
-The [@raymondScalingGasTransfer2012a] model, which was mostly trained on small streams, uniformly underestimates FCO2 and thus bulk carbon efflux.
-The [@raymondGlobalCarbonDioxide2013a] model, which is an average of the other two rating curve estimates (Table S1), is in between.
+We then compare the $FCO_2$ and bulk carbon efflux (via $CO_2$ evasion) from the 22 rivers using BIKER posterior means and three gauge-based HG models (Figure 8).
+In Figure 8a are barplots of the bulk carbon efflux across all rivers in gigagrams of carbon per year.
+The BIKER bulk carbon efflux (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_BIKER',]$sumFCO2, 0)` gG-C/yr) is slightly overestimated relative to to the bulk carbon efflux calculated using observed velocity (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_ulset',]$sumFCO2, 0)` gG-C/yr).
+However, it is a better estimate than those using observed streamflow and HG (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_raymond2012',]$sumFCO2, 0)`, `r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_raymond2013',]$sumFCO2, 0)`, and `r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_Lauerwald2015',]$sumFCO2, 0)` gG-C/yr for ‘Raymond 2012’, ‘Raymond 2013’, and ‘Lauerwald 2015’, respectively).
+Thus, BIKER reasonably captures the estimated efflux from these rivers purely from RS data, and surprisingly outperforms the gauge-based approaches.
 
@@
 
-By-river performance scores for $FCO_2$ (Figure S3) show the [@lauerwaldSpatialPatternsCO22015] model as the best performing in KGE, NRMSE and RRMSE, while BIKER arguably has the least prediction bias (rBIAS) of the models and is approximately the same in KGE as the [@raymondGlobalCarbonDioxide2013a] model.
-Again, the high performance of the [@lauerwaldSpatialPatternsCO22015] and [@raymondGlobalCarbonDioxide2013a] models makes sense as the rating curves rely on the in-situ discharge record.
-Across all four metrics, the [@raymondScalingGasTransfer2012a] model is the worst, performing marginally worse than BIKER.
+Figure 8b plots the by-river performance scores for $FCO_2$.
+For KGE, 'Lauerwald 2015' is the best performing (median KGE: `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_Lauerwald2015',]$kge, na.rm=T),2)`) while BIKER (median KGE: `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_BIKER',]$kge, na.rm=T),2)`) slightly outperforms the other two gauge-based methods (median KGEs: `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_raymond2013',]$kge, na.rm=T),2)` and `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_raymond2012',]$kge, na.rm=T),2)` for 'Raymond 2013' and 'Raymond 2012', respectively).
+However, all four scores are very similar.
+For rBIAS, BIKER indicates the least bias in some rivers and similar bias to the other models in other rivers, as indicated by the scores' inter-quartile range (`r round(IQR(results_3_3_rivs[results_3_3_rivs$key=='FCO2_BIKER',]$rBIAS, na.rm=T),2)`) and median (`r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_BIKER',]$rBIAS, na.rm=T),2)`).
+Meanwhile, all three rating curves uniformly produce significant negative bias in their estimates (median scores of `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_Lauerwald2015',]$rBIAS, na.rm=T),2)`, `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_raymond2013',]$rBIAS, na.rm=T),2)`, and `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_raymond2012',]$rBIAS, na.rm=T),2)` for 'Lauerwald 2015', 'Raymond 2013', and 'Raymond 2012', respectively).
+NRMSE and RRMSE have similar BIKER performance, which is generally worse than the rating curves.
+But, BIKER NRMSE is approximately on par with 'Raymond 2013 and 'Raymond 2012' but BIKER RRMSE (median RRMSE: `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_BIKER',]$rrmse, na.rm=T),2)`) is significantly worse than the gauge-based approaches.
+RRMSE is highly sensitive to prediction errors in small values and so this suggests that BIKER performs poorly when predicting small $CO_2$ fluxes.
+In summary, BIKER performance for KGE, NRMSE, and rBIAS is either similar or negligibly worse than the gauge-based upscaling approaches tested here.
+This is despite relying on absolutely no in situ information like all three other models do.
 
 ##4 Discussion
-###4.1 Ungauged estimation of gas exchange velocity
-BIKER is therefore successful at 1) reproducing $FCO_2$ without any in situ river information for $k_{600}$ (Figure 7) and 2) nearly perfectly reproducing the bulk carbon efflux without any in situ information for $k_{600}$ (Figure 5.8).
-It is important to stress that, unlike BIKER, the rating curves in Figure 8 rely on the observed discharge record.
-This means that Figure 8 represents the best performance that the rating curves could ever have; if ran using modeled discharge their accuracy would necessarily decrease.
-However, this decrease would not be much due to the high accuracy of those data.
-Thus, these results suggest that BIKER will be useful in two settings: upscaling in ungauged rivers as hypothesized, but also in potentially improving our bulk efflux understandings at gauged sites too.
-Future work should systematically quantify prediction error from using both global-scope rating curves and modeled discharge, like is the default approach used to date in upscaling studies.
-
-However, Figures 8 and S3 still confirm that rating curves with a streamgauge are preferred when possible, particularly when interested in properties finer than the bulk efflux (Figure S3).
-Further, methods like the [@raymondGlobalCarbonDioxide2013a] and [@lauerwaldSpatialPatternsCO22015] models are implementable in rivers of any size, while BIKER is limited to the largest rivers on Earth due to SWOT’s coarse spatial resolution.
-However, our results suggest that, upon SWOT's launch, a BIKER-style approach to estimating gas exchange could be coupled with existing upscaling workflows to improve gas flux predictions where gauges are unavailable.
-
-###4.2 Exploring spatiotemporal dynamics of $k_{600}$
+###4.1 Towards exploring global spatiotemporal dynamics of $k_{600}$
 To date, most field-scale studies of riverine gas exchange have focused on 1) its relationship with wind speed [e.g @beaulieuControlsGasTransfer2012a; @borgesVariabilityGasTransfer2004; @zappaEnvironmentalTurbulentMixing2007], 2) average flow velocity [e.g.
@@
 @applingOvercomingEquifinalityLeveraging2018; @demarsStreamMetabolismOpen2015).
-This leaves the spatiotemporal dynamics of gas exchange weakly constrained.
-A few studies have investigated these dynamics, but these studies have been limited to individual rivers and/or limited field seasons [@hallAirWaterOxygen2012; @sand-jensenCO2DynamicsDanish2012].
+However, the spatiotemporal dynamics of gas exchange are still weakly constrained.
+A few studies have investigated these dynamics, but they have been limited to individual rivers and/or limited field seasons [@hallAirWaterOxygen2012; @sand-jensenCO2DynamicsDanish2012].
 [@wallinSpatiotemporalVariabilityGas2011] performed a preliminary analysis in northern Sweden relating river specific (‘at-a-station’) temporal variability in gas exchange with channel slope, but they were limited to an average of only 8 measurements per river in a single watershed.
-Thus, there is a fundamental knowledge gap in understandings the geomorphic and hydraulic drivers of riverine gas exchange.
-This limitation is due both to a lack of process-level understanding [@hallGasExchangeStreams2020] but also due to a lack of measurements.
+This limited knowledge of large-scale $k_{600}$ spatiotemporal dynamics is due both to a lack of process-level understanding [@hallGasExchangeStreams2020] but also a lack of measurements.
 Authors have argued that the key to explaining the large residual variation in upscaling models is to explore at-a-station temporal variability in $k_{600}$ [@hallGasExchangeStreams2020].
 
-Therefore, to explore the spatiotemporal dynamics of $k_{600}$ at fine temporal resolution and but also at the global-scale, SWOT data is an attractive option for upscaling because it provides daily hydraulic measurements for a 3 month fast sampling period for calibration and validation and sampling thereafter between 1 and 7 days per 21 day repeat cycle thereafter [@biancamariaSWOTMissionIts2016].
-BIKER's success on simulated SWOT rivers bodes well for its implementation on real SWOT data, and functionally opens the door for daily estimation of riverine gas exchange globally once SWOT launches.
+Therefore, estimating $k_{600}$ from SWOT data is an attractive option for exploring the spatiotemporal dynamics of $k_{600}$ at fine temporal resolution and at the global-scale.
+SWOT will provide daily hydraulic measurements for a 3 month fast sampling period for calibration and validation and sampling thereafter between 1 and 7 days per 21 day repeat cycle [@biancamariaSWOTMissionIts2016].
+BIKER's success in 1) infering $k_{600}$ using simulated SWOT data (Figures 5 and 6) and 2) being robust to measurement errors internal to that SWOT data (Figure 6a) bode well for BIKER's eventual implementation on real SWOT observations.
+The results presented here functionally open the door for daily estimation of riverine gas exchange globally once SWOT launches.
 
-###4.3 Biases in upscaling k600
-Systematic sampling biases in the data used to develop $k_{600}$ upscaling models [@ulsethDistinctAirWater2019a; @raymondScalingGasTransfer2012a].
+###4.2 Estimating bulk carbon efflux from SWOT
+Section 3.3 confirms that BIKER is successful at 1) reproducing $FCO_2$ without any in situ river information for $k_{600}$ (Figure 7) and 2) reproducing the bulk carbon efflux without any in situ information for $k_{600}$ (Figure 5.8).
+It is important to stress that, unlike BIKER, the HG models in Figure 8 rely on the observed discharge record.
+This means that Figure 8 represents the best performance that those models could ever have; if ran using modeled discharge their accuracy would necessarily decrease.
+However, this decrease would not be too much due to the generally high accuracy of those data (e.g.
+@linGlobalReconstructionNaturalized2019a for globally routed discharge).
+Thus, these results suggest that BIKER will be useful in two settings: upscaling in ungauged rivers as hypothesized, but also in potentially improving our bulk efflux understandings at gauged sites too.
+Future work should systematically quantify prediction error from using both global-scope HG and modeled discharge, like is the default approach used to date in recent upscaling studies (e.g.
+[@lauerwaldSpatialPatternsCO22015, @horgbyUnexpectedLargeEvasion2019]).
 
-Finally, in reasonably wide rivers, wind begins to exert a non-trivial influence on gas exchange.
-It is well established that in lakes and the ocean, wind controls near-surface turbulence, and thus gas exchange [@beaulieuControlsGasTransfer2012a; @readLakesizeDependencyWind2012].
+Figure 8 also confirms that the training data used for HG models exerts a significant influence on upscaled carbon emissions from rivers.
+The 'Lauerwald 2015' model, which was trained on larger rivers than the data used in either the 'Raymond 2013' or 'Raymond 2012' models, should perform best on SWOT-observable rivers (which are greater than 50m wide).
+Figure 8 confirms this.
+Meanwhile, BIKER has no similar reliance on hydraulic parameters trained on different river sizes and only assumes that Manning's equation is viable in the river [which is largely the case in rivers wide enough to be SWOT-observable- @fergusonTimeAbandonManning2010].
+Upon SWOT's launch, the BIKER approach to estimating gas exchange could be coupled with existing upscaling workflows to improve riverine gas flux predictions where gauges are unavailable.
+
+###4.2 $K_{600}$ estimate uncertainity from physical processes verus from RS
+Throughout the BIKER validation, we have assumed no parameter uncertainity in the upscaling parameters $a_i$ and $b_i$.
+Figure 4 highlights the significant influence that those large parameter uncertainties have on estimates of $k_{600}$.
+Concurrently, BIKER estimate uncertainty (Figure 5) is small realitive to the Figure 4 uncertainities.
+Therefore, BIKER provides similar estimates of $k_{600}$ to those produced by current upscaling models and most of the total uncertainity stems from the upscaling model itself.
+We argue that BIKER's estimate uncertainity is limited by current process-level knowledge of $k_{600}$ upscaling and not by the RS algorithm itself, suggesting that BIKER's predictive performance can only improve from advancing our understandings of the processes governing gas exchange from drainage networks.
+In that context, we next explore some considerations for future work on $k_{600}$ upscaling theory in the context of SWOT-observable rivers.
+
+####4.2.1 Gas exhange under uniform and non-uniform flow conditions
+Section 2.2 highlights that most upscaling studies to date have assumed uniform flow conditions (i.e.
+$S_h \neq S_0$) in order to train upscaling models using readily available *S* data.
+However, the first-principles model previously used to define *eD* [@tsivoglouTracerMeasurementReaeration1976] by [@ulsethDistinctAirWater2019a] and [@raymondScalingGasTransfer2012a] does not make this simplifying assumption.
+Therefore, it is an open research question whether parameterizing $k_{600}$ upscaling models via $S_h$ can account for some of the unexplained residual variation in current upscaling models [@hallGasExchangeStreams2020].
+Conveniently, SWOT will explicitly measure $S_h$ at unprecendeted spatial and temporal resolutions and will be sptially joined to hydrography that provides $S_0$.
+Thus, future researchers can use BIKER and SWOT in conjunction to directly compare $k_{600}$ as estimated using $S_0$ and $S_h$.
+
+####4.2.2 Bed roughness and gas exchange in SWOT rivers
+Channel bed roughness effects riverine gas exchange mostly in high energy streams where *S* is sufficiently steep for bubble-induced gas exchange to occur [@hallGasExchangeStreams2020].
+[@ulsethDistinctAirWater2019a] showed bed roughness coarsely scales with $k_{600}$ in steep Alpine streams.
+However, they coarsely estimated bed roughness from arial imagery and to date most similar work has focused on labratory exercises [e.g.
+@chansonStreamReaerationNonuniform; moogStreamReaerationNonuniform].
+We argue that bed roughness is not controlling gas exchange in SWOT-observable rivers because they are so large (and therefore flat per river network geomorphology).
+Instead, gas exchange in SWOT-observable rivers is presumably dominated by water-column, rather than bed friction, turbulence.
+We show in Figure S3 with the [@ulsethDistinctAirWater2019a] data that the 'effective bed roughness height' scales with $k_{600}$ only in extremely steep streams (see Text S2 for the calculation of this roughness term).
+This relationship fundamentally breaks down in less steep rivers, and such steep slopes are functionally impossible in SWOT-observable rivers that are over 50m wide.
+This promising initial result indicates bed roughness controls some aspects of gas exchange and should be explicitly explored in future work.
+This is particularly important because small, steep rivers dominate global river networks due to the fractal nature of river systems [@tarbotonFractalNatureRiver1988].
+
+####4.2.3 Wind-driven gas exchange in SWOT rivers
+In wide rivers like those that SWOT will observe, wind begins to exert a non-trivial influence on gas exchange.
+It is well established that in lakes and the ocean, wind controls near-surface turbulence and thus gas exchange [@beaulieuControlsGasTransfer2012a; @readLakesizeDependencyWind2012].
 Authors have argued that large rivers are a hybrid of the hydraulics-driven turbulence in small rivers and the wind-driven turbulence in lakes [@beaulieuControlsGasTransfer2012a].
 As SWOT will measure only rivers wider than 50m, it follows that wind is likely exerting some influence on gas exchange in SWOT-observable reaches.
-However, we opted to ignore wind effects in our upscaling model and in BIKER to favor global scalability and implementability for two reasons: 1) current upscaling efforts do not account for wind in their estimation of $k_{600}$ and 2) it is infeasible to parameterize every set of SWOT data with local wind data, and relying on an in situ understanding of wind defeats the purpose of BIKER for ungauged settings.
-Rather, we argue that
+Here, we opted to ignore wind effects in our upscaling model and in BIKER to favor global scalability and implementability for two reasons: 1) current upscaling efforts do not account for wind in their estimation of $k_{600}$ either and 2) it is infeasible to parameterize every set of SWOT measurements with local wind data.
+Further, relying on an in situ understanding of wind defeats the purpose of BIKER for ungauged settings.
+Future work should explore the feasibility of assigning a Bayesian prior on wind speed for BIKER.
 
 ##5 Conclusions
+Efforts to upscale gas exchange velocities from river networks generally do so using river channel slope and average flow velocity.
+Therefore, gas evasion estimates are senstivie to availble data on average flow velocity.
+In ungauged basins, this poses a problem because a velocity~discharge rating curve can not be built.
+To circumvent this problem, we propose using established techniques from the ungauged remote sensing of river discharge literature and applying them to  gas exchange velocity in ungauged rivers.
+This is formalized as the BIKER algorithm and uses data from the upcoming SWOT satellite to estimate gas exchange velocity solely from river width and water surface elevation.
+We validated BIKER on 22 'rivers' of simulated SWOT data [@frassonCompilationHydraulicModels2019a], obtaining an RMSE of `r round(10^results_3_2_all$rmse,2)` after accounting for upscaling parameter uncertainity.
+When generalized to estimate bulk carbon efflux from these 22 rivers, BIKER reasonably captures the observed efflux (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_BIKER',]$sumFCO2, 0)` gG-C/yr versus `r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_ulset',]$sumFCO2, 0)` gG-C/yr, respectively).
+FUrther, BIKER's $FCO_2$ estimates across the 22 rivers are neglibly worse than those made using a streamgauge and the global-scope HG models frequently used in the literature.
+This suggests that BIKER will be useful in two settings: upscaling in ungauged rivers as hypothesized, but also in potentially improving our understandings at gauged sites too.
+As river evasion models gain in sophistication and can continue to ingest higher temporal resolution data, BIKER and SWOT should prove useful in providing such information.
 
@@
 BIKER is available at https://github.com/craigbrinkerhoff/BIKER.
-All results and code to generate results, figures, and the manuscript are archived at zenodo something something.
+All code to build and generate results, figures, and the manuscript are archived at zenodo something something.
 Data used in this study is available in Frasson et al.
@@
 (2019).
-This manuscript benefitted from early comments from Tamlin Pavelsky, as well as the large body of existing knowledge from the SWOT River Discharge Working Group.
+This manuscript benefitted extensively from the decade plus body of work generated by the SWOT Discharge Algorithm Working Group.
 

### 2021-02-12 supp_v1.Rmd
message

@@
 title: "Supplemental Information"
-author: "Craig Brinkerhoff"
-date: "January 28, 2021"
-output: word_document
+author: "CB Brinkerhoff, CJ Gleason, PA Raymond, MH Harlan"
+date: "`r Sys.Date()`"
+csl: APA.csl
+output:
+  word_document: 
+      reference_docx: word-styles-reference.docx
+bibliography: References.bib
+
 ---
@@
 
-#Supplemental Information
-Text S1
-Tsivoglou & Neal (1976) outlined an energy dissipation model for the reaeration coefficient (K2, or k without depth correction) that is frequently generalized and used to scale k600 (Raymond et al.
-2012; Ulseth et al.
-2019).
-They showed that K2 scales linearly with the energy dissipation rate eD.
-Their derivation explicitly shows that eD=dH/dT for some segment of stream where H is water surface elevation and T is hydraulic residence time.
-For a stream segment with residence time T, the loss in water surface elevation is synonymous equivalent to the downward component of flow velocity V in the h dimension over the stream segment length (i.e.
-Sh).
-So, we can generalize this to dH/dT=V_h S_h≈VS_h.
-Finally, this can be converted to an energy dissipation rate per unit weight of water such that eD [L^2/T^3 ]=g dH/dT≈〖gVS〗_h.
-This model is convenient for our use because SWOT will only measure Sh and not channel slope S0.
-Further, Manning’s equation under non-uniform flow conditions requires Sh and not S¬0 and so we can conveniently merge the slope terms across equations and directly measure them from SWOT.
+##Text S1
+[@tsivoglouTracerMeasurementReaeration1976] outlined an energy dissipation model for the reaeration coefficient ($K_2$, or k without depth correction) that is frequently generalized and used to scale $k_{600}$ [@ulsethDistinctAirWater2019a; @raymondScalingGasTransfer2012a].
+They showed that $K_2$ scales linearly with the energy dissipation rate eD.
+Their derivation explicitly shows that $eD=\frac{dH}{dT}$ for some segment of stream where H is water surface elevation and T is hydraulic residence time.
+For a stream segment with residence time T, the loss in water surface elevation is synonymous with the downward component of flow velocity V in the h dimension over the reach length (i.e.
+$S_h$).
+So, we can generalize this to $\frac{dH}{dT}=V_hS_h≈VS_h$.
+Finally, this can be converted to an energy dissipation rate per unit weight of water such that $eD  [\frac{L^2}{T^3}]=g\frac{dH}{dT}≈gVS_h$.
+This model is convenient for our use because SWOT will only measure $S_h$ and not channel slope $S_0$.
+Further, Manning’s equation under non-uniform flow conditions requires $S_h$ and not $S_0$ and so we can conveniently merge the slope terms across equations and directly measure them from SWOT.
 
-To my knowledge, most k600 upscaling studies to date have assumed uniform flow conditions (i.e.
-S_0=S_h) to train their upscaling parameters.
-This is presumably because channel slope is readily available in hydrographic data products while water surface slope is not.
-Therefore, future work should look at the influence of non-uniform flow on the upscaling of k600.
-Once SWOT launches, this will be easy to compare as SWOT data products will have both channel slope and water surface slope.
+##Text S2
+In Figure S4 we show that bed roughness only exerts an influence on $k_{600}$ when rivers are incredibly steep.
+We do this using the 'Keulegan effective roughness height' for the channel bed, which is detailed in equation S1 below (in the form form [@dingmanAnalyticalDerivationAtastation2007b]).
+This model was first outlined in the 1930s by Keulegan and outlined explicitly in [@fergusonHydraulicsHydraulicGeometry1986a] and [@dingmanAnalyticalDerivationAtastation2007b] in the context of hydraulic geometry (noting that it is a more generalized form of the Manning's and Chezy's velocity~depth relations commonly used in the literature).
+The model arises from depth integrating the Prandl-von Karman equation for velocity in a turbulent boundary layer ($\frac{u}{u_\star}=2.5ln(\frac{30y}{k_s})$) [@[@fergusonHydraulicsHydraulicGeometry1986a]].
+
+$$V=2.5(gDS)^{1/2}ln(\frac{11D}{y_r})   (S1)$$
+$y_r [L]$ is the effective bed roughness height we solved for using the data from [@ulsethDistinctAirWater2019a] where D is channel depth [L], S is channel slope, g is gravitational acceleration $[M/T^2]$, and V is average flow velocity $[M/T]$.
+Because it is an 'effective/equivalent height', this does not necessairly reflect actual roughness elements in the river channel but is a reasonable way to quantify relative differences in bed roughness across data.
+It is convenient to use here because it provides an actual roughness height and is not simply a coefficient like the conductance coefficients in Manning's and Chezy's velocity~depth relations.
+
+In Figure S4, this effective roughness height was plotted against $k_{600}$ and the data was progressively filtered for steeper and steeper rivers.
 
@@
 
-![Figure S3 By-river performance metrics for FCO2 using the metrics outlined in Table 1.
-Dashed lines denote scores of 1, 0, and -0.41 for KGE (Knoben et al.
-2019).
-Recall that the Raymond and Lauerwald models use the in-situ discharge record while BIKER does not.](outputs/flux_implications/FCO2_by_river.jpg)
+![Figure S3.
+Scaling $k_{600}$ using an effective bed roughness term (see Text S2).
+This scaling relationship only exists for rivers with very steep slopes, here presented as greater than 0.1](outputs/k600/roughness.jpg)
 
@@
 Lauerwald et al. (2015)	lnV=-1.06+0.12lnQ	Borrowed from ½ of the Raymond et al. (2013) calculation (trained on 9,811 USGS streamgauges)
+
+##References

### 2021-02-12 scripts/FCO2_implications.R

@@ -8,18 +8,18 @@ library(tidyverse)
 library(ncdf4)
 library(cowplot)
 library(ggtext)
+library(lubridate)
 theme_set(theme_cowplot())
 
 munge <- 0 #set to 1 if you need to generate results and individual river plots, otherwise 0
 
 #constants-------------------
-molarMass <- 44.01 #g/mol
-t <- 25 #C
-Sc_co2 <- 1742 + (-91.24*t) + (2.208*t^2) + (-0.0219*t^3) #Raymond etal 2012
+molarMass <- 44.01 #g/mol for CO2
+Sc_co2_func <- function(t){1742 + (-91.24*t) + (2.208*t^2) + (-0.0219*t^3)} #Raymond etal 2012
 pCO2_a <- 390 #uatm
-henrys_law <- -60.2409+93.4517*(100/(t+273.15))+23.3585*log((t+273.15)/100) #Weiss 1974
+henrys_law_func <- function(t) {-60.2409+93.4517*(100/(t+273.15))+23.3585*log((t+273.15)/100)} #Weiss 1974
 g <- 9.8 #m/s2
-sa <- 0 #m2 total surface area of 22 rivers
+sa <- 0 #m2 total surface area of 22 rivers. DUmmy favlue starts at 0
 
 #legend labels------------------
 legend_labels <- c('BIKER', 
@@ -74,15 +74,18 @@ s <- data.frame('river'=NA, 'meanS'=NA)
 if (munge == 1) {
   #read in co2 data (Beauliu etal 2012)--------------------------------
   data <- read.csv('inputs/flux_implications/Beauliu2012.csv', header = TRUE)
-  data <- data[2:30,]
-  Data <- select(data, c(Date, CO2))
-  #Data$Date <- as.Date(as.character(Data$Date), '%m/%d/%Y')
-  colnames(Data) <- c('Date', 'CO2_umol_L')
+  data <- data[5:30,] #only keep 26 biweekly sample so they reflect a single year (fall 2009-winter/spring/summer 2008). This amounts to 
+  Data <- select(data, c(Date, CO2, Temperature))
+  Data$Date <- as.Date(as.character(Data$Date), '%m/%d/%y')
+  Data$Date <- format(Data$Date, format="%m-%d")
+  Data <- Data[order(Data$Date),]
+  colnames(Data) <- c('Date', 'CO2_umol_L', 'Water_temp_C')
+  Data$Water_temp_C <- as.numeric(as.character(Data$Water_temp_C))
[…]

### 2021-02-12 scripts/k600_model.R

@@ -261,6 +261,88 @@ biker_k_prior <- group_by(data, slopeRegime) %>%
             n=n())
 write.csv(biker_k_prior, 'outputs//k600//k_priors.csv')
 
+#Figure S4: bed roughness vs. k600-------------------------------------------
+data$keulegan <- 11*data$depth*(1/exp(data$Vms/(2.5*(9.8*data$depth*data$slope)^(1/2))))
+
+lm_roughness <- lm(log10(data$k600)~log10(data$keulegan))
+r2_roughness <- round(summary(lm_roughness)$r.squared, 2)
+roughness_0 <- ggplot(data, aes(x=keulegan, y=k600)) +
+  geom_point() +
+  geom_smooth(method='lm', se=F)+
+  geom_richtext(aes(x=10^-1.5, y=10^3), label=paste0('r<sup>2</sup>: ', r2_roughness), color='black') +
+  coord_cartesian(x=c(10^-2.5, 10^1.5), c(10^-2, 10^5)) +
+  ylab('K600 [m/dy]') +
+  xlab('') +
+  ggtitle('All Data')+
+  scale_y_log10(
+    breaks = scales::trans_breaks("log10", function(x) 10^x),
+    labels = scales::trans_format("log10", scales::math_format(10^.x))
+  ) +
+  scale_x_log10(
+    breaks = scales::trans_breaks("log10", function(x) 10^x),
+    labels = scales::trans_format("log10", scales::math_format(10^.x))
+  )
+
+lm_roughness <- lm(log10(k600)~log10(keulegan), data = filter(data, slope > 0.001))
+r2_roughness <- round(summary(lm_roughness)$r.squared, 2)
+roughness_1 <- ggplot(filter(data, slope > 0.001), aes(x=keulegan, y=k600)) +
+  geom_point() +
+  geom_smooth(method='lm', se=F)+
+  geom_richtext(aes(x=10^-1.5, y=10^3), label=paste0('r<sup>2</sup>: ', r2_roughness), color='black') +
+  coord_cartesian(x=c(10^-2.5, 10^1.5), c(10^-2, 10^5)) +
+  ylab('') +
+  xlab('') +
+  ggtitle('Slope > 0.001')+
+  scale_y_log10(
+    breaks = scales::trans_breaks("log10", function(x) 10^x),
+    labels = scales::trans_format("log10", scales::math_format(10^.x))
+  ) +
+  scale_x_log10(
[…]

### 2021-02-12 scripts/monte_carlo_analysis.R

@@ -1,153 +1,0 @@
-#Monte Carlo analysis for quantifying uncertainity of CO2 evasion model
-  #This quantifies model error for two scenarios: 1) uncertainity associated with Manning's equation, and
-    #2) uncertainity associated with Manning's equation AND the gas transfer velocity model from Raymond et al (2012)
-    #The former is used when we are assuming that there is no error in the gas transfer velocity equation.
-    #We do this to calculate Flux using the same model to test the use of a Bayesian framework
-#Creator: Craig Brinkerhoff
-#Summer 2020
-
-library(cowplot)
-library(tidyverse)
-library(tmap)
-library(st)
-library(sf)
-library(grid)
-theme_set(theme_cowplot())
-
-#random samplers: M sets of reasonable observations for model
-M <- 8000 # number of sets of hydraulic 'measurements'
-set.seed(455)
-
-#Pull in hydraulic measurements from Brinkerhoff etal 2019 to test parameter uncertainity------------------------------------------
-hydraulics_init <- read.csv('inputs//MonteCarlo//field_measurements.csv')
-nhd_med <- read.csv('inputs//MonteCarlo//NHD_join_table.csv')
-
-#add reach slopes from NHD
-hydraulics <- left_join(hydraulics_init, select(nhd_med, SLOPE, SOURCE_FEA, LatSite, LonSite), by = c("site_no" = "SOURCE_FEA"))
-
-#intial cleaning of field measurements
-hydraulics <- filter(hydraulics, is.finite(chan_width) ==1) %>%
-  filter(is.finite(chan_velocity)==1) %>%
-  filter(is.finite(chan_discharge)==1) %>%
-  filter(is.finite(chan_area)==1) %>%
-  filter(chan_width > 0) %>%
-  filter(chan_velocity > 0) %>%
-  filter(chan_discharge > 0) %>%
-  filter(chan_area > 0) %>%
-  filter(measured_rating_diff != 'Poor') %>%
-  filter(measured_rating_diff != 'POOR') %>%
-  filter(is.finite(SLOPE)==1)
-
-#imperial to metric
-hydraulics$area <- hydraulics$chan_area * 0.092903 #ft2 to m2
-hydraulics$width <- hydraulics$chan_width*0.305 #m
-hydraulics$velocity <- hydraulics$chan_velocity*0.305 #m/s
-hydraulics$slope <- hydraulics$SLOPE
[…]

### 2021-02-13 supp_v1.Rmd
message

@@
 date: "`r Sys.Date()`"
-csl: APA.csl
+csl: american-geophysical-union.csl
 output:

### 2021-02-17 manuscript_v1.Rmd
manuscript edits

@@
 ---
-title: "Remote sensing riverine gas exchange from SWOT and implications for estimating carbon efflux from rivers"
+title: "Remote sensing riverine gas exchange and carbon efflux using SWOT observations"
 author: "CB Brinkerhoff, CJ Gleason, PA Raymond, MH Harlan"
@@
 bibliography: References.bib
-csl: american-geophysical-union.csl
+csl: rse_citation_style.csl
 output:
@@
 
-##Key Points
-- We introduce the BIKER algorithm to remotely sense riverine gas exchange using the SWOT satellite and no in situ information
-- BIKER can nearly replicate estimates of $CO_2$ fluxes made via hydraulic geometry and an in situ streamgauge
+##Key Points (85 characters each)
+- We introduce the BIKER algorithm to remotely sense riverine gas exchange using the SWOT satellite
+- BIKER replicates estimates of $CO_2$ fluxes when coupled with in situ concentration data
 - Expected SWOT measurement errors exert a trivial influence on BIKER's estimates
-- BIKER's accuracy is limited by process-level uncertainties in gas exchange from rivers and not by its algorithmic setup
+- BIKER's accuracy is limited by process-level uncertainties and not by its algorithmic setup
 
@@
 Go BIKER!!
-x
+400 words
 x
@@
 ###2.1 Developing an RS-able $k_{600}$ upscaling model
-[@raymondScalingGasTransfer2012a] first used river hydraulics and the theoretical relationship between *eD* and gas exchange to upscale $k_{600}$ from average flow velocity and channel slope, extending earlier work predicting the reaeration coefficient (e.g.
-@oconnorMechanismReaerationNatural1958; @tsivoglouTracerMeasurementReaeration1976).
+[@raymondScalingGasTransfer2012a] first used river hydraulics and the theoretical relationship between *eD* and gas exchange to upscale $k_{600}$ from average flow velocity and channel slope, extending earlier work predicting the reaeration coefficient [e.g.
+@oconnorMechanismReaerationNatural1958; @tsivoglouTracerMeasurementReaeration1976].
 Later, [@ulsethDistinctAirWater2019a] built upon [@raymondScalingGasTransfer2012a]’s work by expanding their dataset to include measurements from steeper rivers and finding that two distinct upscaling regimes exist in low and high energy rivers.
-More specifically, [@ulsethDistinctAirWater2019a] scale $k_{600}$ using eD, where the resulting upscaling model parameters are significantly different whether *eD* is high or low.
+More specifically, [@ulsethDistinctAirWater2019a] scale $k_{600}$ using *eD*, where the resulting upscaling model parameters are significantly different whether *eD* is high or low.
 Their upscaling function is reprinted as equation 2.
@@
 To do this, we use the basic premise that the $k_{600}\sim eD$ scaling relationship is different for different river sizes.
-We use river width as a proxy for river size and train different upscaling functions for different groups of data with similar river sizes (Figure 2, trained on the [@ulsethDistinctAirWater2019a] dataset).
+We use river width as a proxy for river size and train different upscaling functions for different groups of data with similar river sizes (Figure 2, using data from [@ulsethDistinctAirWater2019a] dataset).
 In doing so, we find that narrow rivers (< 10m wide) need to be further broken into low energy (slope < 0.05) and high energy (slope > 0.05) regimes in order to meet the assumptions necessary to fit a linear regression model (specifically, normally distributed residuals of the predictions).
@@
 With a validated upscaling model that requires no a priori knowledge of *eD* (results presented in section 3.1) we turn to developing the algorithm to RS $k_{600}$.
-The approach used here is heavily informed by [@hagemannBAMBayesianAMHGManning2017a]’s algorithm for ungauged RSQ via Bayesian inference.
-Consult that paper for the logic behind conceptualizing the equation for discharge as a Bayesian inference problem, which we largely follow here to conceptualize upscaling $k_{600}$ as a Bayesian inference problem.
+The approach used here is strongly informed by [@hagemannBAMBayesianAMHGManning2017a]’s algorithm for ungauged RSQ via Bayesian inference.
+Consult that paper for the logic behind conceptualizing the equation for discharge as a Bayesian inference problem, which we largely follow here to conceptualize remote sensing $k_{600}$ as a Bayesian inference problem.
 
@@
 In order to use Bayesian inference to solve for $k_{600}$, we need to derive a sampling model/likelihood function.
-Following [@tsivoglouTracerMeasurementReaeration1976], $eD \approx gS_hV$ for unsteady flow conditions (channel slope $S_0 \neq S_h$- see text S1 for this model derivation and why assuming non-uniform flow is convenient for our application).
+Following @tsivoglouTracerMeasurementReaeration1976, $eD \approx gS_hV$ for unsteady flow conditions (channel slope $S_0 \neq S_h$- see text S1 for this model derivation and why assuming non-uniform flow is convenient for our application).
 Using this, as well as Manning’s equation to relate average flow velocity to channel geometry, we rewrite equation 3 as equation 4.
@@
 (2017) which formulates depth via river width and area (in order to use SWOT-observables).
-$k_{600}$ is now defined as a function of $S_h$, channel roughness n $[T/L^{\frac{1}{3}}]$, channel width *W* [L], and channel area $A=A_0+dA [L^2]$ for the change in channel area *dA*).
+$k_{600}$ is now defined as a function of $S_h$, channel roughness n $[T/L^{\frac{1}{3}}]$, channel width *W* [L], and channel area $A_0+dA [L^2]$ for the change in channel area *dA* and median channel area $A_0$).
 It is also defined by the upscaling parameters $a_i$ and $b_i$.
@@
 Equation 4 is then reworked into a Bayesian sampling model with the ‘data’ on the left-hand side sampled from the unknown model parameters on the right-hand side (equation 5).
-Note that in the formal likelihood specification each term in equation 5 is log-transformed such that equation 5 is a lognormal distribution, however they are written here without a transformation for succinctness.
+Note that in the formal likelihood specification each term in equation 5 is log-transformed to yield a normal sampling model, however they are written here without a transformation for succinctness.
 $σ_{upscale}^2$ refers to the uncertainty inherent in the $k_{600}$ estimates themselves (equation 4).
@@
 
-Equation five necessitates that Bayesian priors be assigned to model parameters n, $A_0$, and $k_{600}$.
+Equation five necessitates that Bayesian priors be assigned to model parameters *n*, $A_0$, and $k_{600}$.
 Bayesian priors formularize the a priori estimates (and uncertainties) for these parameters.
 Or more intuitively, they represent our ‘prior river knowledge’ of what *n*, $A_0$, and $k_{600}$ probably are for some river given that they cannot be remotely sensed.
-All priors are formalized within the model as truncated normal distributions of the ln-transformed terms such that $ln(X) \sim N(\mu, \sigma^2)$ for λ<X<ε, using prior hyperparameters mean ($\mu$), standard deviation ($\sigma$), and upper (ε) and lower bounds (λ) for parameter X = n, $A_0$, or $k_{600}$.
+All priors are formalized within the model as truncated normal distributions of the ln-transformed terms such that $ln(X) \sim N(\mu, \sigma^2)$ for $\lambda < x < \epsilon$, using prior hyperparameters mean ($\mu$), standard deviation ($\sigma$), and upper ($\epsilon$) and lower bounds ($\lambda$) for parameter $X=n$, $A_0$, or $k_{600}$.
 In order to avoid relying on in situ information, we assign prior hyperparameters using SWOT data only.
@@
 Finally, we estimated $σ_{upscale}^2$ using Monte Carlo (MC) methods to approximate total uncertainty from equation 4.
-Uncertainty in equation 4 can stem from two sources: 1) error baked into the upscaling model (ai and bi parameter uncertainty) and 2) error in Manning’s approximation of the average flow velocity.
-To do this, we ran MC simulations on 8,000 sets of field measurements of river channel hydraulics from the dataset built by [@brinkerhoffReconcilingAtaStationAtManyStations2019].
-Each MC simulation was itself 10,000 runs, sampling from the normal distributions for $a_i$, $b_i$, and ln-transformed velocity.
-Estimate uncertainty for velocity followed [@hagemannBAMBayesianAMHGManning2017a], who estimated ln-transformed Manning’s equation error to be 0.25.
-This resulted in 8,000 discrete distributions for $k_{600}$ from which the average uncertainty term is extracted and used for $σ_{upscale}^2$.
+We specifically explore three uncertainty scenarios to assess which error source is most influential.
+Uncertainty in equation 4 can stem from three sources: 1) error baked into the upscaling model ($a_i$ and $b_i$ parameter uncertainty), 2) error in Manning’s approximation of the average flow velocity, and 3) measurement error.
+Our interest is using BIKER for satellite remote sensing and so we consider measurement errors from SWOT-derived water surface slopes.
+To do this, we ran three sets of MC simulations on 5,000 sets of field measurements of river channel hydraulics [@brinkerhoffReconcilingAtaStationAtManyStations2019].
+Each MC simulation was itself 10,000 runs, sampling from the normal distributions for $a_i$ and $b_i$ (uncertainty scenario 1), $a_i$, $b_i$, and ln-transformed velocity (uncertainty scenario 2), and $a_i$, $b_i$, ln-transformed velocity, and water surface slope (uncertainty scenario 3).
+Estimate uncertainty for velocity followed @hagemannBAMBayesianAMHGManning2017a, who estimated ln-transformed Manning’s equation error to be 0.25.
+The specifics of the slope-error model used are elaborated in Section 2.3, but in short we account for both the SWOT error tolerances and radar layover error [@durandHowWillRadar2020a].
+This resulted in three sets of 5,000 discrete distributions for $k_{600}$ from which the median uncertainty term is extracted and used for $σ_{upscale}^2$.
 
-With the sampling model (equation 5) and prior distributions described, a joint posterior distribution conditional on the SWOT data is derived.
+With the sampling model (equation 5), prior distributions, and the $σ_{upscale}^2$ parameter described, a joint posterior distribution conditional on the SWOT data is derived.
 To approximate this distribution, BIKER uses a Markov Chain Monte Carlo (MCMC) algorithm.
-Because it is written in the Stan probabilistic programming language, BIKER’s MCMC of choice is a Hamiltonian Monte Carlo which reduces computation time relative to other MCMC approaches (https://mc-stan.org/docs/2_26/reference-manual/hamiltonian-monte-carlo.html).
+Because it is written in the Stan probabilistic programming language, BIKER’s MCMC of choice is a Hamiltonian Monte Carlo which reduces computation time relative to other sampling algorithms (https://mc-stan.org/docs/2_26/reference-manual/hamiltonian-monte-carlo.html).
 
@@
 no measurement errors) and represent the best-case scenarios of what SWOT will provide to hydrologists.
-Here, we use the 22 rivers archived in [@frassonCompilationHydraulicModels2019a] and used by [@rodriguezObservingRiversVarying2020] to develop reach-averaged Saint Vernant equations within the SWOT context.
+Here, we use the 22 rivers archived by @frassonCompilationHydraulicModels2019a and used by @rodriguezObservingRiversVarying2020 to develop reach-averaged Saint Vernant equations within the SWOT context.
 We omitted the Arial Khan river in Bangladesh due to known problems with that hydraulic model and we further sampled the models for only the observations every 11 days to mimic the average SWOT overpass frequency (this also approximately algins the SWOT observations with the bi-weekly $CO_2$ data used later in this study- Section 2.4).
@@
 The 68th percentiles of these reach-scale errors were 1.7cm/km and 10.4cm for water surface slope and height, respectively.
-To apply this model to our 22 rivers, we generate new river heights *H* and water surface slopes $S_h$ by sampling every observation from a normal distribution as defined as $S_{err_{i,t}} \sim normal(S_{obs_{i,t}}, 0.000017^{2})$ and $H_{err_{i,t}} \sim normal(H_{obs_{i,t}}, 0.104^{2})$ (in meters).
+To apply this model to our 22 rivers, we generate new river heights *H* and water surface slopes $S_h$ by sampling every observation from a normal distribution as by equations 7 and 8.
 Because BIKER ingests *dA*, rather than *H*, reach-scale height errors were converted to *dA* errors, again following [@durandHowWillRadar2020a]: $\sigma_{dA} \approx W\sigma_H\sqrt{2}$ given our rectangular channel assumptions.
 
+$$S_{err_{i,t}} \sim normal(S_{obs_{i,t}}, 0.000017^{2}) (7)$$
+
+$$H_{err_{i,t}} \sim normal(H_{obs_{i,t}}, 0.104^{2}) (8)$$
+
 Regardless of the validation data used, we do not have observed $k_{600}$ data for these rivers, and to our knowledge no field dataset of $k_{600}$ exists in the type of temporal and spatial frequency that SWOT (and therefore the BIKER algorithm) provides.
-Further, we are principally interested in our ability to upscale $k_{600}$ from RS observations and are less concerend with the actual accuracy of the upscaling model itself (which can be validated using existing datasets).
+Further, we are principally interested in our ability to reproduce upscaled $k_{600}$ from RS observations and are less concerend with the actual accuracy of the upscaling model itself (which can be validated using existing datasets).
 Therefore, we take the best performing upscaling model from sections 2.1 and 3.1 and use that to calculate the ‘observed’ $k_{600}$ that BIKER is validated against.
-This is calculated using equation 7, where $V_{obs} [L/T]$ is observed river discharge divided by observed channel area.
+This is calculated using equation 9, where $V_{obs} [L/T]$ is observed river discharge divided by observed channel area.
 With this setup, we can directly explore our ability to infer observed *V* and $k_{600}$ from river width and height alone.
@@
 
-$$k_{600, obs}=a_i(gS_hV_{obs})^{b_i} \text{(7)}$$
+$$k_{600, obs}=a_i(gS_hV_{obs})^{b_i} \text{(9)}$$
 
@@
 To validate across all rivers and timesteps, we used the coefficient of determination $r^2$ and the root mean square error (RMSE).
-Four normalized metrics were used for by-river validation: RRMSE and NRMSE are normalized root mean square errors that have been normalized by the observed value and the mean observed value, respectively, to compare across rivers.
+Four normalized metrics were used for by-river validation: RRMSE and NRMSE are normalized root mean square errors that have been normalized by the observed value and the mean observed value (respectively) to compare across rivers.
 rBIAS is a measure of prediction bias that is normalized by the mean observed value to compare across rivers.
-KGE is a standard metric used in streamflow prediction with an intuitive basis: a value greater than -0.41 indicates a model outperforms a uniform prediction of the mean value [@knobenTechnicalNoteInherent2019a] and a value greater than 0 is often interpreted as a useful prediction in ungauged settings.
+KGE is a standard metric used in streamflow prediction with an intuitive basis: a value greater than 0 is often interpreted as a useful prediction in ungauged settings, and a value greater than -0.41 indicates a model outperforms a uniform prediction of the mean value [@knobenTechnicalNoteInherent2019a].
 
@@
 It is one thing to to predict $k_{600}$, but researchers are usually more interested in the gas fluxes from rivers and ultimately the bulk carbon efflux from river to atmosphere.
-Therefore, we explored 1) the ability of BIKER to reproduce hypothetical carbon dioxide evasion fluxes ($FCO_2 [M/L^2T]$) from these 22 rivers, and 2) the sensitivity of the estimated bulk carbon efflux (via $CO_2$ evasion) to the method used to estimate *V*.
+Therefore, we explored 1) the ability of BIKER to reproduce hypothetical carbon dioxide evasion fluxes ($FCO_2 [M/L^2T]$) from these 22 rivers, and 2) the sensitivity of the estimated bulk carbon efflux (via $CO_2$ evasion) to the method used to estimate reach averaged flow velocity.
 
@@
 To do this, we paired the modeled $k_{CO2}$ values (obtained from the $k_{600}$ estimates) with field-measurements of water-side $CO_2$ concentrations and water temperatures.
-26 bi-weekly samples were made by [@beaulieuControlsGasTransfer2012a]- Figure S1) at one location in the Ohio River for one calender year from 2008-2009.
+26 bi-weekly samples were made by @beaulieuControlsGasTransfer2012a- Figure S1) at one location in the Ohio River for one calender year from 2008-2009.
 We paired these 26 values with each 11-day-repeat set of SWOT observations by date, ignoring the timesteps beyond 26 (only ~15% of the SWOT observations were ignored here and this was deemed acceptable).
@@
 
-Then, we estimated bulk carbon efflux using four models for average flow velocity:  BIKER and three HG models previously used for $CO_2$ upscaling [@lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a; @raymondScalingGasTransfer2012a].
+Then, we estimated bulk carbon efflux using four models for average flow velocity:  BIKER and three gauge-based HG models previously used for $CO_2$ upscaling [@lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a; @raymondScalingGasTransfer2012a].
 See Table S1 for their definitions.
 It is worth stressing that the [@lauerwaldSpatialPatternsCO22015] model was developed as one of two components of the [@raymondGlobalCarbonDioxide2013a] model, and so was actually developed by the latter’s authors.
-The names used here refer to the final implementation used in each study’s upscaling exercises.
-For this study, all rating-curve methods use the in-situ discharge record while BIKER does not.
-This allows us to explore the relative differences in bulk carbon efflux estimates if implementing a wholly ungauged method (BIKER) versus gauged methods (all three HG models)).
-Finally, we express the bulk carbon efflux (via $CO_2$ evasion) $[M/T]$ as the average mass flow rate of carbon per year, i.e.
-multiplied by the surface area of all 22 rivers.
+The names used here refer to the specific implementation used in each study.
+For our study, all HG methods use the in-situ discharge record while BIKER does not.
+This allows us to explore the relative differences in bulk carbon efflux estimates if implementing a wholly ungauged method (BIKER) versus gauged methods (all three HG models).
+Finally, we express the bulk carbon efflux (via $CO_2$ evasion) $[M/T]$ as the average mass flow rate of carbon per year from the 22 rivers, accounting for total river surface area.
 
@@
 
-Uncertainties were then propagated through the RS-able [@ulsethDistinctAirWater2019a] model via MC simulations (section 2.2) for the 8,000 hydraulic measurements mapped in Figure 4c.
-Figure 4a plots a histogram of the 8,000 uncertainty terms extracted from those distributions (Figure 4b plots three of the distributions used to build Figure 4a).
-Across the 8,000 tests, we quantified average ln-transformed uncertainty in equation 4 to be `r round(results_3_1_MC$meanSigma,2)` (~`r round(exp(results_3_1_MC$meanSigma), 2)` m/day- dashed blue line in Figure 4a).
-Uncertainty solely from the upscaling model is 1.30, confirming that using Manning’s equation to approximate average flow velocity is introducing very little additional uncertainty to the model (Figure S2).
-This also confirms the large residuals inherent in $k_{600}$ upscaling models as described by [@hallGasExchangeStreams2020].
+Uncertainties were then propagated through the RS-able [@ulsethDistinctAirWater2019a] model via MC simulations for three error scenarios (section 2.2) for the 5,000 hydraulic measurements.
+Figures 4a, 4c, and 4e plot histograms of the 5,000 uncertainty terms extracted from those distributions (Figures 4b, 4d, and 4f plot three of the distributions used to build the histograms).
+Across the 5,000 tests for the fullest error scenario (Figure 4e), we quantified average ln-transformed uncertainty in equation 4 to be `r round(results_3_1_MC$medianSigma,2)` (~`r round(exp(results_3_1_MC$medianSigma), 2)` m/day- dashed blue line in Figure 4a).
+Uncertainty solely from the upscaling model is also 1.27, confirming that functionally all uncertainity in the equation 4 estimates is due to upscaling and not due to the assumptions made for BIKER (specifically, using Manning’s equation to approximate average flow velocity and SWOT measuurement error).
+This also confirms the large residuals inherent in $k_{600}$ upscaling models as described by @hallGasExchangeStreams2020.
 
-![Figure 4 Model uncertainty estimates: a: histogram of SDs for all 8,000 MC simulations with blue dashed line denoting the mean (1.32).
-b: Three example MC simulation results of 10,000 samples each.
-c: Map of the 8,000 sets of hydraulic measurements used for this analysis.](outputs/MonteCarlo/MCsimulations.jpg)
+![Figure 4 Model uncertainty estimates for three error scenarios, as detailed.
+Left: histogram of error terms (SD) for all 5,000 MC simulations with blue dashed line denoting the median.
+Right: Three example MC simulation results of 10,000 samples each.](outputs/MonteCarlo/MC_results.jpg)
 
@@
 Regardless, the RMSE for the BIKER predictions is only `r round(10^results_3_2_all$rmse,2)` m/day across all predictions.
-This is substantially less than the RMSEs for the upscaling models (Figure 3) and the MC-simulated uncertainties (Figure 4a), indicating that recovering upscaled $k_{600}$ from SWOT data is much more accurate than the upscaling process itself.
+This is substantially less than the RMSEs for the upscaling models (Figure 3) and the MC-simulated uncertainties (Figure 4a), indicating that inferring upscaled $k_{600}$ from SWOT data is much more accurate than the upscaling process itself.
 
@@
 We see, more clearly than in Figure 5a, that the posterior mean is systematically underestimated below the median $k_{600}$ values and systemically overestimated above the median $k_{600}$ However, the uncertainty in these estimates captures $k_{600}$ very well: nearly the entire observed CDF falls between, on, or just outside of the 95% CIs (Figure 5b).
-This highlights the benefit of using a Bayesian predictive framework to fully propagate all parameter uncertainties through to the posterior.
+This highlights one benefit of using Bayesian inference to fully propagate all prior uncertainties through to the posterior.
 While these CIs are large and thus BIKER estimates are reasonably uncertain, we do improve upon our baseline understanding of $k_{600}$ in these rivers.
@@
 Figure 6a plots validation metrics calculated for each river (i.e.
-the boxplots are composed of metric scores for the 22 rivers with and without measurement error- see Table 1 for metric definitions).
+the boxplots are composed of scores for the 22 rivers with and without measurement error- see Table 1 for metric definitions).
 SWOT measurement uncertainties exert a neglible influence on 3/4 performance metrics and slightly decrease KGE across the 22 rivers (Figure 6a).
-Thus, the results presented for the rest of the manuscript assume no measurement error.
+With SWOT's errors not exerting a significant influence on BIKER, the results presented for the rest of the manuscript assume no measurement error.
 Median rBIAS is `r round(median(results_3_2_rivs$rBIAS, na.rm=T),2)`, indicating nearly no bias in most rivers’ predictions.
@@
 
-###3.3 Implications for carbon cycle estimates
-Finally, we explore our ability to 1) use BIKER-produced $k_{600}$ to estimate $FCO_2$ and estimate bulk carbon efflux from $CO_2$ evasion (section 2.4).
+###3.3 Influences on carbon efflux estimates
+Finally, we explore our ability to 1) use BIKER-produced $k_{600}$ to estimate $FCO_2$ and 2) estimate bulk carbon efflux from $CO_2$ evasion (section 2.4).
 
@@
 a: FCO2 from BIKER versus using observed average flow velocity for all timesteps for 22 SWOT rivers (grey lines are linear regression and 95% prediction intervals, while black dashed line is the 1:1 line).
-b-d: timeseries plots for three example rivers from Figure 5.6.](outputs/flux_implications/FCO2_plot.jpg)
+b-d: timeseries plots for three example rivers from Figure 5.6.](outputs/FCO2/FCO2_plot.jpg)
 
@@
 b: Cumulative density functions (CDFs) of all FCO2 estimates across all timesteps and rivers for the same four velocity models.
-The median of the distributions is noted with a black line.](outputs/flux_implications/FCO2_models.jpg)
+The median of the distributions is noted with a black line.](outputs/FCO2/FCO2_models.jpg)
 
@@
 Section 3.3 confirms that BIKER is successful at 1) reproducing $FCO_2$ without any in situ river information for $k_{600}$ (Figure 7) and 2) reproducing the bulk carbon efflux without any in situ information for $k_{600}$ (Figure 5.8).
-It is important to stress that, unlike BIKER, the HG models in Figure 8 rely on the observed discharge record.
+This encouraging result has two main implications for future work.
+First, we can couple BIKER with in situ gas concentration loggers to produce $FCO_2$ estimates at novel temporal resolutions, given that to date the temporal resolution of $k_600$ has arguably lagged behind recent advances in gas concentration loggers (CIte KElly??- don't think this is published yet).
+BIKER can likely also be ran at the field scale using arrays of pressure transducers to estimate water surface slope following recent work doing the same using the BAM RSQ algorithm (Harlan et al.
+2021).
+This opens the door to field-scale $FCO_2$ estimates at unprecented temporal resolution.
+
+Secondly, it is important to stress that, unlike BIKER, the HG models in Figure 8 rely on the observed discharge record.
 This means that Figure 8 represents the best performance that those models could ever have; if ran using modeled discharge their accuracy would necessarily decrease.
@@
 Thus, these results suggest that BIKER will be useful in two settings: upscaling in ungauged rivers as hypothesized, but also in potentially improving our bulk efflux understandings at gauged sites too.
-Future work should systematically quantify prediction error from using both global-scope HG and modeled discharge, like is the default approach used to date in recent upscaling studies (e.g.
+Future work should systematically quantify prediction error from using both global-scope HG and modeled discharge, like is the default workflow used in recent upscaling studies (e.g.
 [@lauerwaldSpatialPatternsCO22015, @horgbyUnexpectedLargeEvasion2019]).
@@
 Instead, gas exchange in SWOT-observable rivers is presumably dominated by water-column, rather than bed friction, turbulence.
-We show in Figure S3 with the [@ulsethDistinctAirWater2019a] data that the 'effective bed roughness height' scales with $k_{600}$ only in extremely steep streams (see Text S2 for the calculation of this roughness term).
+We show in Figure S2 with the [@ulsethDistinctAirWater2019a] data that the 'effective bed roughness height' scales with $k_{600}$ only in extremely steep streams (see Text S2 for the calculation of this roughness term).
 This relationship fundamentally breaks down in less steep rivers, and such steep slopes are functionally impossible in SWOT-observable rivers that are over 50m wide.
@@
 ##6 Acknowledgements
+C.B.
+Brinkerhoff was funded on xxxxxxxxxxxxx.
 BIKER is available at https://github.com/craigbrinkerhoff/BIKER.
-All code to build and generate results, figures, and the manuscript are archived at zenodo something something.
-Data used in this study is available in Frasson et al.
-(2019) and Ulseth et al.
-(2019).
+All code to build and generate results, figures, and the manuscript are available at https://github.com/craigbrinkerhoff/RSK600.
+Data used in this study was made generously avaible by Frasson et al.
+(2019), Ulseth et al.
+(2019), and Beaulieu et al.
+(2012).
 This manuscript benefitted extensively from the decade plus body of work generated by the SWOT Discharge Algorithm Working Group.
 
-## 7 References
+## References

### 2021-02-17 supp_v1.Rmd
manuscript edits

--- 2021-02-12 supp_v1.Rmd
@@
 ---
-title: "Supplemental Information"
+title: "Supplemental Information: remote sensing riverine gas exchange and carbon efflux using SWOT observations"
 author: "CB Brinkerhoff, CJ Gleason, PA Raymond, MH Harlan"
 date: "`r Sys.Date()`"
-csl: APA.csl
+csl: rse_citation_style.csl
 output:
@@
 This model was first outlined in the 1930s by Keulegan and outlined explicitly in [@fergusonHydraulicsHydraulicGeometry1986a] and [@dingmanAnalyticalDerivationAtastation2007b] in the context of hydraulic geometry (noting that it is a more generalized form of the Manning's and Chezy's velocity~depth relations commonly used in the literature).
-The model arises from depth integrating the Prandl-von Karman equation for velocity in a turbulent boundary layer ($\frac{u}{u_\star}=2.5ln(\frac{30y}{k_s})$) [@[@fergusonHydraulicsHydraulicGeometry1986a]].
+The model arises from depth integrating the Prandl-von Karman equation for velocity in a turbulent boundary layer ($\frac{u}{u_\star}=2.5ln(\frac{30y}{k_s})$) [@fergusonHydraulicsHydraulicGeometry1986a].
 
@@
 Sampling took place 2008-2009 in the Ohio River (upstream of Cincinnati, OH).
-Each point here was joined to the 11-day SWOT observations used in this study (section 5.2.4).](outputs/flux_implications/Beaulieu_timeseries.jpg)
+Each point here was joined to the 11-day SWOT observations used in this study (section 5.2.4).](outputs/FCO2/Beaulieu_timeseries.jpg)
 
-![Figure S2 Histogram of equation 3 uncertainty terms extracted from 8,000 different Monte Carlo simulations (following that outlined in section 5.2.2).
-This indicates estimate uncertainty from the ai and bi parameters (and not Manning’s equation).
-Dashed blue line indicates the mean uncertainty (1.30 in ln-space, or ~3.67 m/day).](outputs/MonteCarlo/noMannings.jpeg)
-
-![Figure S3.
+![Figure S2.
 Scaling $k_{600}$ using an effective bed roughness term (see Text S2).

### 2021-02-17 main.R

@@ -22,9 +22,9 @@ results_3_1 <- read.csv('outputs//k600//results.csv')
 results_3_1_MC <- read.csv('outputs//MonteCarlo//results.csv')
 results_3_2_rivs <- read.csv('outputs//validation//results_by_riv.csv')
 results_3_2_all <- read.csv('outputs//validation//results_all_riv.csv')
-results_3_3_all <- read.csv('outputs//flux_implications//fco2_stats_all.csv')
-results_3_3_rivs <- read.csv('outputs//flux_implications//fco2_stats_by_river.csv')
-results_3_3_bulk <- read.csv('outputs//flux_implications//bulkFluxes.csv')
+results_3_3_all <- read.csv('outputs//FCO2//fco2_stats_all.csv')
+results_3_3_rivs <- read.csv('outputs//FCO2//fco2_stats_by_river.csv')
+results_3_3_bulk <- read.csv('outputs//FCO2//bulkFluxes.csv')
 results_3_3_bulk$key <- as.character(results_3_3_bulk$key)
 
 #knit manuscript

### 2021-02-18 manuscript_v1.Rmd
manuscript edits

@@
 ---
-title: "Remote sensing riverine gas exchange and carbon efflux using SWOT observations"
+title: "Remote sensing riverine gas exchange and estimating carbon efflux using SWOT observations"
 author: "CB Brinkerhoff, CJ Gleason, PA Raymond, MH Harlan"
@@
 Go BIKER!!
-400 words
+max 400 words
 x
@@
 Because *S* is readily available in any hydrographic data product, most efforts to upscale $k_{600}$ to drainage networks are limited by estimates of the *V* of the river channel.
-Thus, drainage-network-scale GHG emissions estimates are sensitive to their *V* data.
+Thus, drainage-network-scale GHG emissions estimates are highly sensitive to their *V* data.
 This is exasperated in ungauged basins where *V* models cannot be trained because no in situ hydraulic information is available and fieldwork is impractical [@gleasonRemoteSensingRiver2020].
@@
 This is the approach used in upscaling *V* to continental/global river networks and has proven useful in those settings [e.g.
-@horgbyUnexpectedLargeEvasion2019; @borgesVariabilityGasTransfer2004; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a].
-However, it has long been established that HG parameters exhibit high spatial variability across landscapes and vary significantly from river to river [@parkWorldwideVariationsHydraulic1977].
+@borgesVariabilityGasTransfer2004; @horgbyUnexpectedLargeEvasion2019; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a].
+However, it has long been established that HG parameters exhibit high spatial variability across landscapes and vary significantly from river to river [@parkWorldwideVariationsHydraulic1977; @rhodesBfmDiagramGraphical1977].
 Because of this, a large body of geomorphology work has attempted to parse out process-based explanations for HG parameters [e.g.
@@
 Remote hydraulic estimation is itself a burgeoning subfield within remote sensing of hydrology and an analogue to the problem presented here is the remote sensing of river discharge (RSQ- see @gleasonRemoteSensingRiver2020 for a thorough review).
-There are many ways to perform RSQ, from calibrating hydrologic and hydraulic models [] to methods that use no in situ information ('ungauged approaches') [].
+There are many ways to perform RSQ, from calibrating to channel hydraulics [e.g.
+@brakenridgeOrbitalMicrowaveMeasurement2007; @pavelskyUsingWidthbasedRating2014; @pavelskyRemoteSensingSuspended2009; @tarpanelliEstimationRiverDischarge2013] to calibrating hydraulic/hydrologic models [e.g.
+@bjerklieEstimatingDischargeRivers2005; @chandanpurkarSatelliteReanalysisBasedMass2017; @linGlobalReconstructionNaturalized2019a; @nealDataAssimilationApproach2009] to methods that use no in situ information as inputs to  hydraulic/hydrologic models (termed 'ungauged approaches').
 Many of these ungauged approaches were developed in the context of the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission.
@@
 In the context of RSQ, a decade of SWOT work has explored the multi-parameter problem of estimating roughness and bathymetry to produce the SWOT discharge product [e.g.
-@andreadisConstrainingAssimilationSWOT2020d; @brinkerhoffConstrainingRemoteRiver2020; @durandEstimatingReachaveragedDischarge2014a; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014a; @hagemannBAMBayesianAMHGManning2017a; @larnierRiverDischargeBathymetry2020a; @oubanasDischargeEstimationUngauged2018].
-A specific subset of these ungauged RSQ methods are termed ‘McFLIs’ or ‘Mass Conserved Flow Law Inversion’ algorithms (e.g.
-@andreadisConstrainingAssimilationSWOT2020d; @brinkerhoffConstrainingRemoteRiver2020; @durandEstimatingReachaveragedDischarge2014a).
+@andreadisConstrainingAssimilationSWOT2020d; @brinkerhoffConstrainingRemoteRiver2020; @durandEstimatingReachaveragedDischarge2014a; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014a; @garamboisVariationalEstimationEffective2020; @hagemannBAMBayesianAMHGManning2017a; @larnierRiverDischargeBathymetry2020a; @oubanasDischargeEstimationUngauged2018].
+A specific subset of these ungauged RSQ methods are termed ‘McFLIs’ or ‘Mass Conserved Flow Law Inversion’ algorithms [e.g.
+@andreadisConstrainingAssimilationSWOT2020d; @brinkerhoffConstrainingRemoteRiver2020; @hagemannBAMBayesianAMHGManning2017a; @durandEstimatingReachaveragedDischarge2014a].
 They use basic geomorphic theories, rather than full hydrologic and/or hydraulic models, and the concept of ‘prior river knowledge’ to estimate discharge from RS data where not all hydraulic terms are RS-able [@gleasonRemoteSensingRiver2020; @gleasonTrackingRiverFlows2017].
@@
 
-Thus, to advance our understanding of riverine GHG fluxes in ungauged basins and/or at drainage-network-scales, we sought to estimate $k_{600}$ using only RS data and no in situ information.
-We hypothesize that remote sensing $k_{600}$ is possible by coupling 1) hydraulic geometry theory, 2) $k_{600}$ upscaling theory, 3) Bayesian remote sensing (in a manner similar to McFLI), and 4) SWOT data.
+Thus, we hypothesize that SWOT data will improve our understanding of riverine GHG fluxes in ungauged basins and/or at drainage-network-scales, after techniques have been devleoped to ingest its novel measurements of water surface width and height.
+Therefore, we sought to develop a method to estimate $k_{600}$ using only SWOT measurements and no in situ information.
+We hypothesize that remote sensing $k_{600}$ is possible by coupling Bayesian remote sensing techniques (in a manner similar to McFLI) and SWOT data with hydraulic geometry and gas exchange theory.
 More specifically, this manuscript aims to answer two questions:
-	1) Is ungauged RS of the gas exchange velocity possible?
+	1) Is ungauged RS of the gas exchange velocity possible using soon-to-be-available SWOT river measurements?
 	2) How might this algorithm influence upscaled estimates of riverine carbon efflux to the atmosphere?
 
-To answer these questions, we build a RS algorithm that ingests SWOT data and produces $k_{600}$ estimates (with fully quantified uncertainty within a Bayesian framework) that requires no in situ information about a river’s hydraulic properties.
+To answer these questions, we develop a RS algorithm that ingests SWOT data and produces $k_{600}$ estimates and associated uncertainties that requires only water surface width and slope and therefore no in situ information about a river’s hydraulic properties.
 We validate the algorithm on simulated SWOT data for 22 rivers (as SWOT has yet to launch, simulated data is necessary- Section 2.3).
-We also quantify the influence of the expected measurement errors in SWOT data on our $k_{600}$ estimates.
-Finally, we compare our algorithm against published methods: we use previously published dissolved carbon dioxide ($CO_2$) data to compare the bulk carbon efflux (via $CO_2$ evasion) from the 22 rivers as calculated using both our new algorithm and published velocity rating curves and in situ discharge records.
+We also quantify the sensitivity of our algorithm to the expected measurement error implicit in SWOT data.
+Finally, we compare our algorithm against published methods for estimating carbon dioxide evasion ($FCO_2$) from rivers: we use previously published dissolved $CO_2$ data to compare the bulk carbon efflux (via $CO_2$ evasion) from the 22 rivers as calculated using both our new algorithm and published gauged techniques.
 
@@
 We name the RS of $k_{600}$ algorithm BIKER, or the ‘Bayesian Inversion of the $k_{600}$ Rate’ algorithm.
-To build BIKER, two steps are taken: first, we develop a $k_{600}$ upscaling model that does not rely on a priori knowledge of the river (section 2.1) and then we use that upscaling model to build the actual RS algorithm (section 2.2).
+To build BIKER, two steps are taken: first, we develop a $k_{600}$ upscaling model that relies on no a priori knowledge of the river (section 2.1) and then we couple that upscaling model with remote sensing techniques to build BIKER (section 2.2).
 Following these methods, the validation setup is described (section 2.3), as is the comparison of estimated bulk carbon effluxes from a suite of average flow velocity models (section 2.4).
@@
 ![Figure 1.
-Flowchart of methodology developed in this chapter.
+Flowchart of methodology used in this study.
 ‘Extended Upscaling Model’ is detailed in section 2.1.
@@
 ###2.1 Developing an RS-able $k_{600}$ upscaling model
-[@raymondScalingGasTransfer2012a] first used river hydraulics and the theoretical relationship between *eD* and gas exchange to upscale $k_{600}$ from average flow velocity and channel slope, extending earlier work predicting the reaeration coefficient [e.g.
+[@raymondScalingGasTransfer2012a] first used river hydraulics and the theoretical relationship between *eD* and gas exchange to upscale $k_{600}$ from average flow velocity and channel slope, extending earlier work predicting the reaeration coefficient in a similar manner [e.g.
 @oconnorMechanismReaerationNatural1958; @tsivoglouTracerMeasurementReaeration1976].
@@
 While equation 2 would be ideal for the purposes outlined for this manuscript and provides best-to-date predictive accuracy for upscaling $k_{600}$, it requires an a priori knowledge of *eD* to assign its differential model parameters.
-We do not have this a priori knowledge in an ungauged setting and only have access to SWOT-observed river widths and heights (and therefore water surface slopes).
+We do not have this a priori knowledge in an ungauged, RS-based setting and only have access to SWOT-observed river widths and heights (and therefore water surface slopes).
 Thus, we need to extend the [@ulsethDistinctAirWater2019a] model to assign the two model parameters using only RS-able hydraulics.
 To do this, we use the basic premise that the $k_{600}\sim eD$ scaling relationship is different for different river sizes.
-We use river width as a proxy for river size and train different upscaling functions for different groups of data with similar river sizes (Figure 2, using data from [@ulsethDistinctAirWater2019a] dataset).
+We use river width to define river size and train different upscaling functions for different groups of data with similar river sizes [Figure 2, using the data from @ulsethDistinctAirWater2019a].
 In doing so, we find that narrow rivers (< 10m wide) need to be further broken into low energy (slope < 0.05) and high energy (slope > 0.05) regimes in order to meet the assumptions necessary to fit a linear regression model (specifically, normally distributed residuals of the predictions).
@@
 
-This model therefore assigns $a_i$ and $b_i$ a priori using just river width and slope to define *i*.
+This model therefore assigns $a_i$ and $b_i$ a priori using just SWOT measurements of water surface width and slope to define *i*.
 Another implication of this extension of the [@ulsethDistinctAirWater2019a] model is that we now use *a* and *b* parameters that implicitly reflect the distinct geomorphology of different rivers.
@@
 Finally, we estimated $σ_{upscale}^2$ using Monte Carlo (MC) methods to approximate total uncertainty from equation 4.
-We specifically explore three uncertainty scenarios to assess which error source is most influential.
+Specifically, We explore three uncertainty scenarios to assess which source is most influential.
 Uncertainty in equation 4 can stem from three sources: 1) error baked into the upscaling model ($a_i$ and $b_i$ parameter uncertainty), 2) error in Manning’s approximation of the average flow velocity, and 3) measurement error.
 Our interest is using BIKER for satellite remote sensing and so we consider measurement errors from SWOT-derived water surface slopes.
-To do this, we ran three sets of MC simulations on 5,000 sets of field measurements of river channel hydraulics [@brinkerhoffReconcilingAtaStationAtManyStations2019].
-Each MC simulation was itself 10,000 runs, sampling from the normal distributions for $a_i$ and $b_i$ (uncertainty scenario 1), $a_i$, $b_i$, and ln-transformed velocity (uncertainty scenario 2), and $a_i$, $b_i$, ln-transformed velocity, and water surface slope (uncertainty scenario 3).
+To do this, we run 15,000 unique MC simulations on 5,000 sets of field measurements of river channel hydraulics under three different uncertainty scenarios [@brinkerhoffReconcilingAtaStationAtManyStations2019].
+Each MC simulation is itself 10,000 runs, sampling from the normal distributions for $a_i$ and $b_i$ (uncertainty scenario 1), $a_i$, $b_i$, and ln-transformed velocity (uncertainty scenario 2), and $a_i$, $b_i$, ln-transformed velocity, and water surface slope (uncertainty scenario 3).
 Estimate uncertainty for velocity followed @hagemannBAMBayesianAMHGManning2017a, who estimated ln-transformed Manning’s equation error to be 0.25.
-The specifics of the slope-error model used are elaborated in Section 2.3, but in short we account for both the SWOT error tolerances and radar layover error [@durandHowWillRadar2020a].
-This resulted in three sets of 5,000 discrete distributions for $k_{600}$ from which the median uncertainty term is extracted and used for $σ_{upscale}^2$.
+The specifics of the slope-error model used are elaborated in Section 2.3 [@durandHowWillRadar2020a].
+This results in three sets of 5,000 discrete distributions for $k_{600}$ from which the median uncertainty term is extracted and used for $σ_{upscale}^2$.
 
-With the sampling model (equation 5), prior distributions, and the $σ_{upscale}^2$ parameter described, a joint posterior distribution conditional on the SWOT data is derived.
+With the sampling model (equation 5), prior distributions, and the $σ_{upscale}^2$ parameter described, a joint posterior distribution conditional on the SWOT data is described.
 To approximate this distribution, BIKER uses a Markov Chain Monte Carlo (MCMC) algorithm.
@@
 Simulated rivers mimic perfect measurement conditions (i.e.
-no measurement errors) and represent the best-case scenarios of what SWOT will provide to hydrologists.
-Here, we use the 22 rivers archived by @frassonCompilationHydraulicModels2019a and used by @rodriguezObservingRiversVarying2020 to develop reach-averaged Saint Vernant equations within the SWOT context.
-We omitted the Arial Khan river in Bangladesh due to known problems with that hydraulic model and we further sampled the models for only the observations every 11 days to mimic the average SWOT overpass frequency (this also approximately algins the SWOT observations with the bi-weekly $CO_2$ data used later in this study- Section 2.4).
+no measurement errors).
+Here, we use the 22 rivers archived by @frassonCompilationHydraulicModels2019a and used by @rodriguezObservingRiversVarying2020 to develop reach-averaged Saint Vernant equations for the SWOT mission.
+We omitted the Arial Khan river in Bangladesh due to known problems with that hydraulic model and we further sampled the models for only the observations every 11 days to mimic the average SWOT overpass frequency (this also approximately aligns the SWOT observations with the bi-weekly $CO_2$ data used later in this study- Section 2.4).
 All told, this yielded 503 sets of SWOT observations to validate BIKER with.
 
-We also introduced measurement error into these simulated SWOT observations to assess the performance degradation due to errors inherent in SWOT measurements.
-Error in SWOT-measured river heights and water surface slopes will come from tolerances internal to the satellite, i.e.
-the SWOT error budget, as well as radar layover error.
-River width errors are assumed here to be neglible [@durandHowWillRadar2020a].
-Layover error is the phenomenon when radar returns arrive at the sensor at the same time, generally in areas of high topographic variation [@reesPhysicalPrinciplesRemote2013].
+We also introduced measurement error into these simulated SWOT observations to assess BIKER's performance degradation due to errors inherent in SWOT measurements.
+Error in SWOT measurements will come from both the error tolerances intrinsic in the satellite data product as well as radar layover error.
+Here, we assume errors in river width are negligible [@durandHowWillRadar2020a] and solely focus on measurement errors in river height and slope.
+Layover error is the phenomenon when radar returns from different places arrive at the sensor at the same time, SOMETHING HERE.
+This generally occurs in areas of high topographic variation [@reesPhysicalPrinciplesRemote2013] where the land surface 'obscures' the water surface relative to SWOT's incidence angle.
 While the SWOT simulator mentioned above is ideal for this analysis, it is cumbersome to run on many rivers and so here we coarsely model measurement error using the results from a recent global anlaysis of SWOT radar layover errors [@durandHowWillRadar2020a].
@@
 The 68th percentiles of these reach-scale errors were 1.7cm/km and 10.4cm for water surface slope and height, respectively.
-To apply this model to our 22 rivers, we generate new river heights *H* and water surface slopes $S_h$ by sampling every observation from a normal distribution as by equations 7 and 8.
-Because BIKER ingests *dA*, rather than *H*, reach-scale height errors were converted to *dA* errors, again following [@durandHowWillRadar2020a]: $\sigma_{dA} \approx W\sigma_H\sqrt{2}$ given our rectangular channel assumptions.
+To apply this model to our 22 rivers, we generate new river heights *H* and water surface slopes $S_h$ by sampling every observation from normal distributions following equations 7 and 8.
+Because BIKER ingests *dA*, rather than *H*, reach-scale height errors were converted to *dA* errors, again following [@durandHowWillRadar2020a] and our rectangular channel assumptions: $\sigma_{dA} \approx W\sigma_H\sqrt{2}$.
 
@@
 Regardless of the validation data used, we do not have observed $k_{600}$ data for these rivers, and to our knowledge no field dataset of $k_{600}$ exists in the type of temporal and spatial frequency that SWOT (and therefore the BIKER algorithm) provides.
-Further, we are principally interested in our ability to reproduce upscaled $k_{600}$ from RS observations and are less concerend with the actual accuracy of the upscaling model itself (which can be validated using existing datasets).
+Further, we are principally interested in our ability to reproduce upscaled $k_{600}$ from RS observations and are less concerend with the actual accuracy of the upscaling model itself [which can be validated using existing datasets like previously done by @raymondScalingGasTransfer2012a and @ulsethDistinctAirWater2019a].
 Therefore, we take the best performing upscaling model from sections 2.1 and 3.1 and use that to calculate the ‘observed’ $k_{600}$ that BIKER is validated against.
-This is calculated using equation 9, where $V_{obs} [L/T]$ is observed river discharge divided by observed channel area.
-With this setup, we can directly explore our ability to infer observed *V* and $k_{600}$ from river width and height alone.
+This is done using equation 9, where $V_{obs} [L/T]$ is observed river discharge divided by observed channel area.
+With this setup, we can directly explore BIKER's ability to infer observed *V* and $k_{600}$ from river width and height alone.
 It also means that, for a fair validation scheme, $\sigma_{upscale}^2$ must be set to reflect only error from Manning’s equation (and not $a_i$ and $b_i$).
+**PROBABLY NEED TO ACCOUNT FOR SLOPE UNCERTAINITY TOO**.
 Thus, $\sigma_{upscale}^2$ is set to 0.25 for this validation.
-However in practice, it should reflect the total uncertainty calculated in section 2.2.
+However in practice, it should reflect the total uncertainty calculated in section 2.2 from all three sources.
 
@@
 We then take the best performing upscaling function, implement it within BIKER, and validate it on the 22 SWOT rivers (section 3.2).
-Finally, we use $k_{600}$ as modeled by both BIKER and gauge-based HG models to compare $FCO_2$ and bulk carbon efflux estimates from the 22 rivers (section 3.3).
+Finally, we use $k_{600}$ as modeled by both BIKER and gauge-based HG models to compare $FCO_2$ and bulk carbon efflux estimates from the 22 rivers across gauged and ungauged methods (section 3.3).
 
@@
 Figure 3 plots the validations for the five upscaling functions tested.
-Both the original [@ulsethDistinctAirWater2019a] model, and our RS-able extension, perform the best on this independent set of validation data, with near identical performance.
+Both the original [@ulsethDistinctAirWater2019a] model and our RS-able extension perform the best on this independent set of validation data, with near identical performance.
 The RS-able [@ulsethDistinctAirWater2019a] model achieves an $r^2$ of `r results_3_1[results_3_1$Model == 'RS-able',]$r2` and an RMSE of `r results_3_1[results_3_1$Model == 'RS-able',]$rmse` m/day, which is just marginally better than the original [@ulsethDistinctAirWater2019a] model (`r results_3_1[results_3_1$Model == 'Ulseth',]$r2` and `r results_3_1[results_3_1$Model == 'Ulseth',]$rmse` m/day, respectively).
-The two power-law based Raymond models ([@raymondScalingGasTransfer2012a] equations 4 and 3) achieve slightly worse performance, with $r^2$ in the 0.70s and their RMSE scores slightly higher (`r results_3_1[results_3_1$Model == 'Raymond_2',]$rmse` and `r results_3_1[results_3_1$Model == 'Raymond_3',]$rmse` m/day, respectively).
+The two power-law based [@raymondScalingGasTransfer2012a] models (equations 4 and 3 in that study) achieve slightly worse performance, with $r^2$ in the 0.70s and their RMSE scores slightly higher (`r results_3_1[results_3_1$Model == 'Raymond_2',]$rmse` and `r results_3_1[results_3_1$Model == 'Raymond_3',]$rmse` m/day, respectively).
 [@raymondScalingGasTransfer2012a] equation 5, which is linear in structure, worked well on their dataset but fails to capture the non-linear relationship evident in this expanded and more geomorphically diverse dataset.
 In fact, it predicts negative $k_{600}$ values and error metrics cannot be calculated because the prediction residuals are not normally distributed.
-Because we successfully reproduced the [@ulsethDistinctAirWater2019a] model, the rest of this manuscript deals exclusively with this RS-able variant of the model.
+**SHOULD I INCLUDE THIS MODEL??** Because we successfully reproduced the [@ulsethDistinctAirWater2019a] model, the rest of this manuscript deals exclusively with this RS-able variant of the model.
 
@@
 For BIKER, the points are the posterior means while the black lines are the 95% confidence intervals (CIs) for the predictions.
-$k_{600}$ is strongly correlated with the $k_{600}$ predicted by BIKER, capturing the general magnitude of the predictions and with most points falling on or near the 1:1 line.
-However, the 95% prediction intervals (dashed grey lines) highlight relatively large residuals for many predictions.
+$k_{600}$ is strongly correlated with the $k_{600}$ predicted by BIKER.
+BIKER captures the general magnitude of the predictions and most points fall on or near the 1:1 line.
+However, the 95% prediction intervals (dashed grey lines) highlight reasonably large residuals for many predictions.
 Regardless, the RMSE for the BIKER predictions is only `r round(10^results_3_2_all$rmse,2)` m/day across all predictions.
-This is substantially less than the RMSEs for the upscaling models (Figure 3) and the MC-simulated uncertainties (Figure 4a), indicating that inferring upscaled $k_{600}$ from SWOT data is much more accurate than the upscaling process itself.
+Again, this is substantially less than the RMSEs for the upscaling models versus field measurements (Figure 3) and the MC-simulated uncertainties (Figure 4a, 4c, and 4e).
 
@@
 Figure 5a suggests a flip in the bias of the $k_{600}$ predictions, which is confirmed in Figure 5b.
-Figure 5b plots the cumulative density functions (CDFs) of observed and predicted $k_{600}$ where the green line is the BIKER posterior mean and the blue lines are the 95% CIs for the BIKER posterior.
-We see, more clearly than in Figure 5a, that the posterior mean is systematically underestimated below the median $k_{600}$ values and systemically overestimated above the median $k_{600}$ However, the uncertainty in these estimates captures $k_{600}$ very well: nearly the entire observed CDF falls between, on, or just outside of the 95% CIs (Figure 5b).
+Figure 5b plots the cumulative density functions (CDFs) of observed and predicted $k_{600}$ where the green line is the BIKER posterior mean and the blue lines are the CDFs of the BIKER posterior 95% CIs.
+We see, more clearly than in Figure 5a, that the posterior mean is systematically underestimated for values less than the median $k_{600}$ and systemically overestimated for values above the median $k_{600}$.
+However, the uncertainty in these estimates captures $k_{600}$ very well: nearly the entire observed CDF falls between, on, or just outside of the 95% CIs (Figure 5b).
 This highlights one benefit of using Bayesian inference to fully propagate all prior uncertainties through to the posterior.
 While these CIs are large and thus BIKER estimates are reasonably uncertain, we do improve upon our baseline understanding of $k_{600}$ in these rivers.
-Put another way, we reasonably capture $k_{600}$ with no in situ information about the river while simultaneously and explicitly accounting for the large uncertainties inherent in our $k_{600}$ estimates.
+Put another way, we reasonably capture $k_{600}$ with no in situ information about the river while simultaneously and explicitly accounting for the reasonably large uncertainties inherent in our $k_{600}$ estimates.
 
@@
 The $r^2$ is identical to that for $k_{600}$ (Figure 7a).
-There is also less underestimation bias for the lowest of the observed $FCO_2$ than in the $k_{600}$ predictions (Figure 5a).
-This is presumably due to the sturcture of the $FCO_2$ equation, which reduces the relative importance of errors in the gas exchange values.
+There is also less underestimation bias for the smallest of the observed $FCO_2$ than in the $k_{600}$ predictions (Figure 5a).
+This is presumably due to the structure of the $FCO_2$ equation, which reduces the relative importance of errors in the gas exchange values given that the $CO_2$ data is measured in situ.
 $FCO_2$ prediction intervals also have a similar relative magnitude to those presented in Figure 5.
 Figure 7b-d includes subplots for the same rivers as Figure 4b-d, however with $FCO_2$ instead of $k_{600}$.
-There is functionally perfect recovery of $FCO_2$ in the Kanawha River, very good recovery (with some positive bias in peak evasion events) for the Ohio Section 2 River, and far less negative bias in the Downstream Mississippi river than is present in the $k_{600}$ estiamtes (Figure 5b).
+There is functionally perfect recovery of $FCO_2$ in the Kanawha River, very good recovery (with some positive bias in peak evasion events) for the Ohio Section 2 River, and far less negative bias in the Downstream Mississippi river than is present in the $k_{600}$ estimates (Figure 5b).
 Again, the temporal dynamics are accurately modeled however there is some bias in the magnitude of the predictions.
@@
 a: FCO2 from BIKER versus using observed average flow velocity for all timesteps for 22 SWOT rivers (grey lines are linear regression and 95% prediction intervals, while black dashed line is the 1:1 line).
-b-d: timeseries plots for three example rivers from Figure 5.6.](outputs/FCO2/FCO2_plot.jpg)
+b-d: timeseries plots for the three example rivers from Figure 5.6.](outputs/FCO2/FCO2_plot.jpg)
 
 We then compare the $FCO_2$ and bulk carbon efflux (via $CO_2$ evasion) from the 22 rivers using BIKER posterior means and three gauge-based HG models (Figure 8).
-In Figure 8a are barplots of the bulk carbon efflux across all rivers in gigagrams of carbon per year.
+Figure 8a are barplots of the bulk carbon efflux (via $CO_2$ evasion) across all rivers in gigagrams of carbon per year.
 The BIKER bulk carbon efflux (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_BIKER',]$sumFCO2, 0)` gG-C/yr) is slightly overestimated relative to to the bulk carbon efflux calculated using observed velocity (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_ulset',]$sumFCO2, 0)` gG-C/yr).
-However, it is a better estimate than those using observed streamflow and HG (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_raymond2012',]$sumFCO2, 0)`, `r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_raymond2013',]$sumFCO2, 0)`, and `r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_Lauerwald2015',]$sumFCO2, 0)` gG-C/yr for ‘Raymond 2012’, ‘Raymond 2013’, and ‘Lauerwald 2015’, respectively).
-Thus, BIKER reasonably captures the estimated efflux from these rivers purely from RS data, and surprisingly outperforms the gauge-based approaches.
+However, it is a better estimate than those using a streamgauge and HG models (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_raymond2012',]$sumFCO2, 0)`, `r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_raymond2013',]$sumFCO2, 0)`, and `r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_Lauerwald2015',]$sumFCO2, 0)` gG-C/yr for ‘Raymond 2012’, ‘Raymond 2013’, and ‘Lauerwald 2015’, respectively).
+Thus, BIKER reasonably captures the estimated efflux from these rivers purely from RS data, and surprisingly outperforms the gauge-based approaches when assessiung bulk flux properties (Figure 8a).
 
@@
 Meanwhile, all three rating curves uniformly produce significant negative bias in their estimates (median scores of `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_Lauerwald2015',]$rBIAS, na.rm=T),2)`, `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_raymond2013',]$rBIAS, na.rm=T),2)`, and `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_raymond2012',]$rBIAS, na.rm=T),2)` for 'Lauerwald 2015', 'Raymond 2013', and 'Raymond 2012', respectively).
-NRMSE and RRMSE have similar BIKER performance, which is generally worse than the rating curves.
-But, BIKER NRMSE is approximately on par with 'Raymond 2013 and 'Raymond 2012' but BIKER RRMSE (median RRMSE: `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_BIKER',]$rrmse, na.rm=T),2)`) is significantly worse than the gauge-based approaches.
+BIKER has similar NRMSE and RRMSE performance, which is generally worse than the rating curves.
+But, BIKER NRMSE is only modestly worse than 'Raymond 2013 and 'Raymond 2012' but BIKER RRMSE (median RRMSE: `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_BIKER',]$rrmse, na.rm=T),2)`) is significantly worse than the gauge-based approaches.
 RRMSE is highly sensitive to prediction errors in small values and so this suggests that BIKER performs poorly when predicting small $CO_2$ fluxes.
@@
 ##4 Discussion
-###4.1 Towards exploring global spatiotemporal dynamics of $k_{600}$
+###4.1 Towards remote sensing global spatiotemporal dynamics of $k_{600}$
 To date, most field-scale studies of riverine gas exchange have focused on 1) its relationship with wind speed [e.g @beaulieuControlsGasTransfer2012a; @borgesVariabilityGasTransfer2004; @zappaEnvironmentalTurbulentMixing2007], 2) average flow velocity [e.g.
@@
 SWOT will provide daily hydraulic measurements for a 3 month fast sampling period for calibration and validation and sampling thereafter between 1 and 7 days per 21 day repeat cycle [@biancamariaSWOTMissionIts2016].
-BIKER's success in 1) infering $k_{600}$ using simulated SWOT data (Figures 5 and 6) and 2) being robust to measurement errors internal to that SWOT data (Figure 6a) bode well for BIKER's eventual implementation on real SWOT observations.
-The results presented here functionally open the door for daily estimation of riverine gas exchange globally once SWOT launches.
+BIKER's success in 1) infering $k_{600}$ using simulated SWOT data over a wide range of rivers (Figures 5 and 6) and 2) being robust to measurement errors internal to the SWOT data (Figure 6a) bode well for BIKER's eventual implementation on real SWOT observations.
+The results presented here suggest that daily estimation of riverine gas exchange globally could be possible once SWOT launches.
 
@@
 This encouraging result has two main implications for future work.
-First, we can couple BIKER with in situ gas concentration loggers to produce $FCO_2$ estimates at novel temporal resolutions, given that to date the temporal resolution of $k_600$ has arguably lagged behind recent advances in gas concentration loggers (CIte KElly??- don't think this is published yet).
-BIKER can likely also be ran at the field scale using arrays of pressure transducers to estimate water surface slope following recent work doing the same using the BAM RSQ algorithm (Harlan et al.
-2021).
-This opens the door to field-scale $FCO_2$ estimates at unprecented temporal resolution.
+First, we can likely couple BIKER with in situ gas concentration loggers to produce $FCO_2$ estimates at novel temporal resolutions, given that the temporal resolution of $k_600$ has arguably lagged behind recent advances in gas concentration data **get citations from Kelly**.
+BIKER can likely also be ran at the field scale using arrays of pressure transducers to estimate water surface slope following recent work doing the same using the [@hagemannBAMBayesianAMHGManning2017a] RSQ algorithm [@harlanDischargeEstimationDense].
 
-Secondly, it is important to stress that, unlike BIKER, the HG models in Figure 8 rely on the observed discharge record.
-This means that Figure 8 represents the best performance that those models could ever have; if ran using modeled discharge their accuracy would necessarily decrease.
-However, this decrease would not be too much due to the generally high accuracy of those data (e.g.
-@linGlobalReconstructionNaturalized2019a for globally routed discharge).
+Secondly, it is important to stress that, unlike BIKER, the HG models in Figure 8 rely on an in situ streamgauge.
+This means that Figure 8 represents the best performance that those models could ever have; if ran using modeled discharge their accuracy would necessarily decrease, though this decrease would not be too much due to the generally good performance of those data.
 Thus, these results suggest that BIKER will be useful in two settings: upscaling in ungauged rivers as hypothesized, but also in potentially improving our bulk efflux understandings at gauged sites too.
-Future work should systematically quantify prediction error from using both global-scope HG and modeled discharge, like is the default workflow used in recent upscaling studies (e.g.
-[@lauerwaldSpatialPatternsCO22015, @horgbyUnexpectedLargeEvasion2019]).
+Future work should systematically quantify prediction error from using both global-scope HG and modeled discharge, like is the default workflow used in recent upscaling studies [e.g.
+@lauerwaldSpatialPatternsCO22015, @horgbyUnexpectedLargeEvasion2019].
+**Iassume Shaoda's paper will still be in review and can't cite here **
 
@@
 Throughout the BIKER validation, we have assumed no parameter uncertainity in the upscaling parameters $a_i$ and $b_i$.
-Figure 4 highlights the significant influence that those large parameter uncertainties have on estimates of $k_{600}$.
-Concurrently, BIKER estimate uncertainty (Figure 5) is small realitive to the Figure 4 uncertainities.
-Therefore, BIKER provides similar estimates of $k_{600}$ to those produced by current upscaling models and most of the total uncertainity stems from the upscaling model itself.
-We argue that BIKER's estimate uncertainity is limited by current process-level knowledge of $k_{600}$ upscaling and not by the RS algorithm itself, suggesting that BIKER's predictive performance can only improve from advancing our understandings of the processes governing gas exchange from drainage networks.
-In that context, we next explore some considerations for future work on $k_{600}$ upscaling theory in the context of SWOT-observable rivers.
+We have shown that BIKER estimate uncertainty is significantly less than $a_i$ and $b_i$ uncertainity (Figures 4 and 5).
+Therefore, BIKER provides similar estimates of $k_{600}$ to those produced by current upscaling models and most of the total uncertainity stems from the upscaling model itself (Figure 4).
+Therefore, we argue that BIKER's estimate uncertainity is limited by current process-level knowledge of $k_{600}$ upscaling and not by BIKER or SWOT, suggesting that BIKER's predictive performance can only improve from advancing our understandings of the processes governing gas exchange from SWOT-observable rivers.
+Validating the BIKER algorithm has provided a unique lense to view $k_600$ upscaling through, and in that context we next explore some considerations for future work on $k_{600}$ upscaling in SWOT-observable rivers.
 
@@
 Section 2.2 highlights that most upscaling studies to date have assumed uniform flow conditions (i.e.
-$S_h \neq S_0$) in order to train upscaling models using readily available *S* data.
-However, the first-principles model previously used to define *eD* [@tsivoglouTracerMeasurementReaeration1976] by [@ulsethDistinctAirWater2019a] and [@raymondScalingGasTransfer2012a] does not make this simplifying assumption.
+$S_h \neq S_0$) in order to train upscaling models using readily available slope data.
+However, the first-principles model previously used by [@ulsethDistinctAirWater2019a] and [@raymondScalingGasTransfer2012a] to define *eD* [@tsivoglouTracerMeasurementReaeration1976] does not make this simplifying assumption.
 Therefore, it is an open research question whether parameterizing $k_{600}$ upscaling models via $S_h$ can account for some of the unexplained residual variation in current upscaling models [@hallGasExchangeStreams2020].
-Conveniently, SWOT will explicitly measure $S_h$ at unprecendeted spatial and temporal resolutions and will be sptially joined to hydrography that provides $S_0$.
-Thus, future researchers can use BIKER and SWOT in conjunction to directly compare $k_{600}$ as estimated using $S_0$ and $S_h$.
+Conveniently, SWOT will explicitly measure $S_h$ at unprecendeted spatial and temporal resolutions and will be spatially joined to hydrography that provides $S_0$.
+Thus, future researchers can use BIKER and SWOT in conjunction to directly explore this question.
 
 ####4.2.2 Bed roughness and gas exchange in SWOT rivers
-Channel bed roughness effects riverine gas exchange mostly in high energy streams where *S* is sufficiently steep for bubble-induced gas exchange to occur [@hallGasExchangeStreams2020].
-[@ulsethDistinctAirWater2019a] showed bed roughness coarsely scales with $k_{600}$ in steep Alpine streams.
+Channel bed roughness effects riverine gas exchange mostly in high energy streams where slope is sufficiently steep for bubble-induced gas exchange to occur [@hallGasExchangeStreams2020].
+[@ulsethDistinctAirWater2019a] showed bed roughness loosely scales with $k_{600}$ in steep Alpine streams.
 However, they coarsely estimated bed roughness from arial imagery and to date most similar work has focused on labratory exercises [e.g.
@@
 Instead, gas exchange in SWOT-observable rivers is presumably dominated by water-column, rather than bed friction, turbulence.
-We show in Figure S2 with the [@ulsethDistinctAirWater2019a] data that the 'effective bed roughness height' scales with $k_{600}$ only in extremely steep streams (see Text S2 for the calculation of this roughness term).
+We show in Figure S2 with the [@ulsethDistinctAirWater2019a] data that the 'effective bed roughness height' scales with $k_{600}$ only in extremely steep streams (see Text S2 for the calculation of this bed roughness term).
 This relationship fundamentally breaks down in less steep rivers, and such steep slopes are functionally impossible in SWOT-observable rivers that are over 50m wide.
@@
 Efforts to upscale gas exchange velocities from river networks generally do so using river channel slope and average flow velocity.
-Therefore, gas evasion estimates are senstivie to availble data on average flow velocity.
+Therefore, gas evasion estimates are sensitive to available data on average flow velocity.
 In ungauged basins, this poses a problem because a velocity~discharge rating curve can not be built.
 To circumvent this problem, we propose using established techniques from the ungauged remote sensing of river discharge literature and applying them to gas exchange velocity in ungauged rivers.
-This is formalized as the BIKER algorithm and uses data from the upcoming SWOT satellite to estimate gas exchange velocity solely from river width and water surface elevation.
-We validated BIKER on 22 'rivers' of simulated SWOT data [@frassonCompilationHydraulicModels2019a], obtaining an RMSE of `r round(10^results_3_2_all$rmse,2)` after accounting for upscaling parameter uncertainity.
+This is formalized as the BIKER algorithm and uses data from the upcoming SWOT satellite to remotely sense gas exchange velocity solely from river width and water surface height.
+We validate BIKER on 22 'rivers' of simulated SWOT data [@frassonCompilationHydraulicModels2019a], obtaining an RMSE of `r round(10^results_3_2_all$rmse,2)` after accounting for upscaling parameter uncertainity.
 When generalized to estimate bulk carbon efflux from these 22 rivers, BIKER reasonably captures the observed efflux (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_BIKER',]$sumFCO2, 0)` gG-C/yr versus `r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_ulset',]$sumFCO2, 0)` gG-C/yr, respectively).
-FUrther, BIKER's $FCO_2$ estimates across the 22 rivers are neglibly worse than those made using a streamgauge and the global-scope HG models frequently used in the literature.
-This suggests that BIKER will be useful in two settings: upscaling in ungauged rivers as hypothesized, but also in potentially improving our understandings at gauged sites too.
-As river evasion models gain in sophistication and can continue to ingest higher temporal resolution data, BIKER and SWOT should prove useful in providing such information.
+Further, BIKER's $FCO_2$ estimates across the 22 rivers are only modestly worse than those made using a streamgauge and the global-scope HG models frequently used in the literature.
+This suggests that BIKER will be useful not only for upscaling in ungauged rivers, but also in potentially improving our understandings in gauged rivers too.
+As drainage network models of GHG evasion gain in sophistication and continue to ingest higher and higher temporal resolution data, BIKER and SWOT should prove useful in providing such information.
 
@@
 BIKER is available at https://github.com/craigbrinkerhoff/BIKER.
-All code to build and generate results, figures, and the manuscript are available at https://github.com/craigbrinkerhoff/RSK600.
-Data used in this study was made generously avaible by Frasson et al.
+All code to build and generate results, figures, and the manuscript is available at https://github.com/craigbrinkerhoff/RSK600.
+Data used in this study was made generously available by Frasson et al.
 (2019), Ulseth et al.

### 2021-02-18 supp_v1.Rmd
manuscript edits

@@
 ---
-title: "Supplemental Information: remote sensing riverine gas exchange and carbon efflux using SWOT observations"
+title: "Supplemental Information: remote sensing riverine gas exchange and estimating carbon efflux using SWOT observations"
 author: "CB Brinkerhoff, CJ Gleason, PA Raymond, MH Harlan"
@@
 In Figure S4 we show that bed roughness only exerts an influence on $k_{600}$ when rivers are incredibly steep.
-We do this using the 'Keulegan effective roughness height' for the channel bed, which is detailed in equation S1 below (in the form form [@dingmanAnalyticalDerivationAtastation2007b]).
+We do this using the 'Keulegan effective roughness height' for the channel bed, which is detailed in equation S1 below [in the form from @dingmanAnalyticalDerivationAtastation2007b].
 This model was first outlined in the 1930s by Keulegan and outlined explicitly in [@fergusonHydraulicsHydraulicGeometry1986a] and [@dingmanAnalyticalDerivationAtastation2007b] in the context of hydraulic geometry (noting that it is a more generalized form of the Manning's and Chezy's velocity~depth relations commonly used in the literature).
-The model arises from depth integrating the Prandl-von Karman equation for velocity in a turbulent boundary layer ($\frac{u}{u_\star}=2.5ln(\frac{30y}{k_s})$) [@fergusonHydraulicsHydraulicGeometry1986a].
+The model arises from depth integrating the Prandl-von Karman equation for velocity in a turbulent boundary layer [@fergusonHydraulicsHydraulicGeometry1986a].
 
@@
 
-Table S1 Details on the 3 velocity rating curve models used to estimate bulk carbon efflux from the SWOT rivers (section 5.2.4).
-	Equation	Description
-Raymond et al. (2012)	lnV=-1.64+0.285lnQ	Trained on Raymond et al. (2012) dataset
-Raymond et al. (2013)	lnV=mean{█(-1.64+0.285lnQ,@-1.06+0.12lnQ  )}	Average of Raymond et al. (2012) model and one trained on 9,811 USGS streamgauges
-Lauerwald et al. (2015)	lnV=-1.06+0.12lnQ	Borrowed from ½ of the Raymond et al. (2013) calculation (trained on 9,811 USGS streamgauges)
+*Table S1: Details on the 3 velocity rating curve models used to estimate bulk carbon efflux from the SWOT rivers (section 5.2.4).*
+
+|**Name**|**Equation**|**Description**|**Reference**|
+|-------------|-------------|-------------|-------------|-------------|
+|Raymond 2012|$ln(V)=-1.64+0.285ln(Q)$|Trained on their dataset|@raymondScalingGasTransfer2012a|
+|Raymond 2013|$ln(V)=mean(-1.64+0.285ln(Q),  -1.06+0.12ln(Q))$ |Average of the Raymond 2012 equation and one trained on 9,811 USGS streamgauges|@raymondGlobalCarbonDioxide2013a|
+|Lauerwald 2015|$ln(V)=-1.06+0.12ln(Q)$ |Borrowed from ½ of the Raymond 2013 calculation (trained on 9,811 USGS streamgauges)|@lauerwaldSpatialPatternsCO22015|
 

### 2021-02-19 manuscript_v1.Rmd
manuscript edits

@@
 
-$$Flux=k(C_{water}-C_{air}) \text{(1)}$$
+$$Flux=k(C_{water}-C_{air}) \mathbf{(1)}$$
 
@@
 This is convenient for upscaling because in rivers and streams *e* can be equated to the turbulent energy dissipation rate *eD* $[L^2/T^3]$ [@ulsethDistinctAirWater2019a; @raymondScalingGasTransfer2012a], which is estimated using easily calculated hydraulic parameters: $eD \approx gSV$ where *g* is gravitational acceleration $[L/T^2]$, *S* is channel slope, and *V* is average flow velocity $[L/T]$ [@tsivoglouTracerMeasurementReaeration1976].
-Because *S* is readily available in any hydrographic data product, most efforts to upscale $k_{600}$ to drainage networks are limited by estimates of the *V* of the river channel.
-Thus, drainage-network-scale GHG emissions estimates are highly sensitive to their *V* data.
+Because *S* is readily available in any hydrographic data product, most efforts to upscale $k_{600}$ to drainage networks are limited by the quality of their estimates of the *V* of the river channel.
 This is exasperated in ungauged basins where *V* models cannot be trained because no in situ hydraulic information is available and fieldwork is impractical [@gleasonRemoteSensingRiver2020].
@@
 Remote hydraulic estimation is itself a burgeoning subfield within remote sensing of hydrology and an analogue to the problem presented here is the remote sensing of river discharge (RSQ- see @gleasonRemoteSensingRiver2020 for a thorough review).
-There are many ways to perform RSQ, from calibrating to channel hydraulics [e.g.
-@brakenridgeOrbitalMicrowaveMeasurement2007; @pavelskyUsingWidthbasedRating2014; @pavelskyRemoteSensingSuspended2009; @tarpanelliEstimationRiverDischarge2013] to calibrating hydraulic/hydrologic models [e.g.
-@bjerklieEstimatingDischargeRivers2005; @chandanpurkarSatelliteReanalysisBasedMass2017; @linGlobalReconstructionNaturalized2019a; @nealDataAssimilationApproach2009] to methods that use no in situ information as inputs to  hydraulic/hydrologic models (termed 'ungauged approaches').
-Many of these ungauged approaches were developed in the context of the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission.
+There are many ways to perform RSQ, from calibrating RS data to local channel hydraulics [e.g.
+@brakenridgeOrbitalMicrowaveMeasurement2007; @pavelskyUsingWidthbasedRating2014; @pavelskyRemoteSensingSuspended2009; @tarpanelliEstimationRiverDischarge2013] to calibrating hydraulic/hydrologic models with both in situ and RS data [e.g.
+@bjerklieEstimatingDischargeRivers2005; @chandanpurkarSatelliteReanalysisBasedMass2017; @linGlobalReconstructionNaturalized2019a; @nealDataAssimilationApproach2009] to methods that use no in situ information in their hydraulic/hydrologic models (termed 'ungauged approaches').
+Many, but not all, of these ungauged approaches were developed in the context of the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission.
+**I don't know how if introducing SWOT via RSQ is the best idea.
+Maybe I should just introduce SWOT on its own?**
 
@@
 
-Thus, we hypothesize that SWOT data will improve our understanding of riverine GHG fluxes in ungauged basins and/or at drainage-network-scales, after techniques have been devleoped to ingest its novel measurements of water surface width and height.
+In this context, we hypothesize that SWOT data will improve our understanding of riverine GHG fluxes in ungauged basins and/or at drainage-network-scales after techniques have been developed to ingest SWOT's novel measurements of water surface width and height.
 Therefore, we sought to develop a method to estimate $k_{600}$ using only SWOT measurements and no in situ information.
@@
 We name the RS of $k_{600}$ algorithm BIKER, or the ‘Bayesian Inversion of the $k_{600}$ Rate’ algorithm.
-To build BIKER, two steps are taken: first, we develop a $k_{600}$ upscaling model that relies on no a priori knowledge of the river (section 2.1) and then we couple that upscaling model with remote sensing techniques to build BIKER (section 2.2).
-Following these methods, the validation setup is described (section 2.3), as is the comparison of estimated bulk carbon effluxes from a suite of average flow velocity models (section 2.4).
-A flowchart detailing the complete workflow is included (Figure 1).
+To build BIKER, two steps are taken: first, we develop a $k_{600}$ upscaling model that relies on no a priori knowledge of river hydraulics (section 2.1) and then we couple that upscaling model with remote sensing techniques to build BIKER (section 2.2).
+Following these methods, the validation setup is described (section 2.3), as is the workflow for comparing estimated bulk carbon effluxes from a suite of average flow velocity models (section 2.4).
+A flowchart detailing the entire study is included (Figure 1).
 
@@
 ###2.1 Developing an RS-able $k_{600}$ upscaling model
-[@raymondScalingGasTransfer2012a] first used river hydraulics and the theoretical relationship between *eD* and gas exchange to upscale $k_{600}$ from average flow velocity and channel slope, extending earlier work predicting the reaeration coefficient in a similar manner [e.g.
+[@raymondScalingGasTransfer2012a] first used river hydraulics and the theoretical relationship between *eD* and gas exchange to upscale $k_{600}$ from average flow velocity and channel slope.
+This extended earlier work predicting the reaeration coefficient in a similar manner [e.g.
 @oconnorMechanismReaerationNatural1958; @tsivoglouTracerMeasurementReaeration1976].
-Later, [@ulsethDistinctAirWater2019a] built upon [@raymondScalingGasTransfer2012a]’s work by expanding their dataset to include measurements from steeper rivers and finding that two distinct upscaling regimes exist in low and high energy rivers.
+Later, [@ulsethDistinctAirWater2019a] built upon [@raymondScalingGasTransfer2012a] by expanding their dataset to include measurements from steeper rivers and finding that two distinct upscaling regimes exist in low and high energy rivers.
 More specifically, [@ulsethDistinctAirWater2019a] scale $k_{600}$ using *eD*, where the resulting upscaling model parameters are significantly different whether *eD* is high or low.
@@
 
-$$ln(k_{600})=3.10 + 0.35ln(eD) if eD < 0.02  \text{(2)}$$
+$$ln(k_{600})=3.10 + 0.35ln(eD) if eD < 0.02  \mathbf{(2)}$$
 $$ln(k_{600})=6.43+1.18ln(eD) if eD > 0.02$$
 
-While equation 2 would be ideal for the purposes outlined for this manuscript and provides best-to-date predictive accuracy for upscaling $k_{600}$, it requires an a priori knowledge of *eD* to assign its differential model parameters.
+While equation 2 would be ideal for the goal of this manuscript and provides best-to-date predictive accuracy for upscaling $k_{600}$, it requires an a priori knowledge of *eD* to assign its differential model parameters.
 We do not have this a priori knowledge in an ungauged, RS-based setting and only have access to SWOT-observed river widths and heights (and therefore water surface slopes).
@@
 
-$$k_{600}=a_i(eD)^{b_i} \text{(3)}$$
+$$k_{600}=a_i(eD)^{b_i} \mathbf{(3)}$$
 
@@
 ###2.2 Developing the BIKER algorithm
-With a validated upscaling model that requires no a priori knowledge of *eD* (results presented in section 3.1) we turn to developing the algorithm to RS $k_{600}$.
+With a validated upscaling model that requires no a priori knowledge of *eD* (results presented in section 3.1) we turn to developing the algorithm to remotely sense $k_{600}$.
 The approach used here is strongly informed by [@hagemannBAMBayesianAMHGManning2017a]’s algorithm for ungauged RSQ via Bayesian inference.
-Consult that paper for the logic behind conceptualizing the equation for discharge as a Bayesian inference problem, which we largely follow here to conceptualize remote sensing $k_{600}$ as a Bayesian inference problem.
+Consult that paper for the logic behind conceptualizing the equation for discharge as a Bayesian inference problem, which we largely follow here to conceptualize remote sensing $k_{600}$ as a Bayesian inference problem that can be solved using SWOT data.
 
@@
 In order to use Bayesian inference to solve for $k_{600}$, we need to derive a sampling model/likelihood function.
-Following @tsivoglouTracerMeasurementReaeration1976, $eD \approx gS_hV$ for unsteady flow conditions (channel slope $S_0 \neq S_h$- see text S1 for this model derivation and why assuming non-uniform flow is convenient for our application).
+Following @tsivoglouTracerMeasurementReaeration1976, $eD \approx gS_hV$ for unsteady flow conditions (channel slope $S_0 \neq S_h$- see text S1 for this model derivation and why assuming non-uniform flow is convenient for ingesting SWOT data).
 Using this, as well as Manning’s equation to relate average flow velocity to channel geometry, we rewrite equation 3 as equation 4.
@@
 (2017) which formulates depth via river width and area (in order to use SWOT-observables).
-$k_{600}$ is now defined as a function of $S_h$, channel roughness n $[T/L^{\frac{1}{3}}]$, channel width *W* [L], and channel area $A_0+dA [L^2]$ for the change in channel area *dA* and median channel area $A_0$).
+$k_{600}$ is now defined as a function of $S_h$, channel roughness n $[T/L^{\frac{1}{3}}]$, channel width *W* [L], and channel area $A_0+dA [L^2]$ (for the change in channel area *dA* and median channel area $A_0$).
 It is also defined by the upscaling parameters $a_i$ and $b_i$.
 
-$$k_{600}=a_i(gS_h^{\frac{3}{2}}\frac{1}{n}(A_0+dA)^\frac{2}{3}W^\frac{-2}{3})^{b_i} \text{(4)}$$
+$$k_{600}=a_i(gS_h^{\frac{3}{2}}\frac{1}{n}(A_0+dA)^\frac{2}{3}W^\frac{-2}{3})^{b_i} \mathbf{(4)}$$
 
@@
 Note that in the formal likelihood specification each term in equation 5 is log-transformed to yield a normal sampling model, however they are written here without a transformation for succinctness.
-$σ_{upscale}^2$ refers to the uncertainty inherent in the $k_{600}$ estimates themselves (equation 4).
+The $σ_{upscale}^2$ parameter refers to the uncertainty inherent in the $k_{600}$ estimates themselves (equation 4).
 
@@
 Bayesian priors formularize the a priori estimates (and uncertainties) for these parameters.
-Or more intuitively, they represent our ‘prior river knowledge’ of what *n*, $A_0$, and $k_{600}$ probably are for some river given that they cannot be remotely sensed.
+Or more intuitively for these physcial quantities, they represent our ‘prior river knowledge’ of what *n*, $A_0$, and $k_{600}$ probably are for some river given that they cannot be remotely sensed.
 All priors are formalized within the model as truncated normal distributions of the ln-transformed terms such that $ln(X) \sim N(\mu, \sigma^2)$ for $\lambda < x < \epsilon$, using prior hyperparameters mean ($\mu$), standard deviation ($\sigma$), and upper ($\epsilon$) and lower bounds ($\lambda$) for parameter $X=n$, $A_0$, or $k_{600}$.
 In order to avoid relying on in situ information, we assign prior hyperparameters using SWOT data only.
-*n* and $A_0$ prior hyperparameters were assigned following the method developed by [@brinkerhoffConstrainingRemoteRiver2020].
+*n* and $A_0$ prior hyperparameters were assigned following the method developed by @brinkerhoffConstrainingRemoteRiver2020.
 They developed a set of river channel prior hyperparameters for McFLI algorithms that are entirely RS-able and reflect differential channel hydraulics as a function of river geomorphology.
 With this method, they showed substantial improvement in the accuracy of their ungauged RSQ estimates with no changes to the algorithms or data used.
-We assigned the $k_{600}$ prior hyperparameters using a simple slope regression model trained on the [@ulsethDistinctAirWater2019a] dataset to assign $\mu$ (equation 6).
+We assigned the $k_{600}$ prior hyperparameters using a simple slope regression model trained on the @ulsethDistinctAirWater2019a dataset to assign $\mu$ (equation 6).
 $σ^2$ was set to the standard error of equation 6: 1.023.
@@
 
-$$\mu_{k600}=3.22+0.347ln(S_h) if ln(S_h) < -4.634   \text{(6)}$$
+$$\mu_{k600}=3.22+0.347ln(S_h) if ln(S_h) < -4.634   \mathbf{(6)}$$
 $$\mu_{k600}=6.85+1.13ln(S_h) if ln(S_h)>-4.634$$
@@
 Finally, we estimated $σ_{upscale}^2$ using Monte Carlo (MC) methods to approximate total uncertainty from equation 4.
-Specifically, We explore three uncertainty scenarios to assess which source is most influential.
+Specifically, we explore three uncertainty scenarios to assess which source of error is most influential.
 Uncertainty in equation 4 can stem from three sources: 1) error baked into the upscaling model ($a_i$ and $b_i$ parameter uncertainty), 2) error in Manning’s approximation of the average flow velocity, and 3) measurement error.
 Our interest is using BIKER for satellite remote sensing and so we consider measurement errors from SWOT-derived water surface slopes.
-To do this, we run 15,000 unique MC simulations on 5,000 sets of field measurements of river channel hydraulics under three different uncertainty scenarios [@brinkerhoffReconcilingAtaStationAtManyStations2019].
+To do this, we run 15,000 unique MC simulations on 5,000 sets of field measurements of river channel hydraulics [@brinkerhoffReconcilingAtaStationAtManyStations2019] under three different uncertainty scenarios.
 Each MC simulation is itself 10,000 runs, sampling from the normal distributions for $a_i$ and $b_i$ (uncertainty scenario 1), $a_i$, $b_i$, and ln-transformed velocity (uncertainty scenario 2), and $a_i$, $b_i$, ln-transformed velocity, and water surface slope (uncertainty scenario 3).
 Estimate uncertainty for velocity followed @hagemannBAMBayesianAMHGManning2017a, who estimated ln-transformed Manning’s equation error to be 0.25.
-The specifics of the slope-error model used are elaborated in Section 2.3 [@durandHowWillRadar2020a].
-This results in three sets of 5,000 discrete distributions for $k_{600}$ from which the median uncertainty term is extracted and used for $σ_{upscale}^2$.
+The specifics of the slope-error model that we used are elaborated on in Section 2.3 and rely on a recent analysis of SWOT slope measurement errors [@durandHowWillRadar2020a].
+This results in three sets of 5,000 discrete distributions for $k_{600}$ from which the median uncertainty term for each uncertainity scenario is extracted and used for $σ_{upscale}^2$.
 
@@
 ###2.3 BIKER validation setup
-We validated the BIKER algorithm on 22 rivers using observed $V$ and observed $S_h$ First, we detail the 22 rivers, and then we detail how ‘observed’ $k_{600}$ is calculated.
+We validated BIKER on 22 rivers using observed $V$ and observed $S_h$ First, we detail the 22 rivers, and then we detail how ‘observed’ $k_{600}$ is calculated.
 
@@
 This generally occurs in areas of high topographic variation [@reesPhysicalPrinciplesRemote2013] where the land surface 'obscures' the water surface relative to SWOT's incidence angle.
-While the SWOT simulator mentioned above is ideal for this analysis, it is cumbersome to run on many rivers and so here we coarsely model measurement error using the results from a recent global anlaysis of SWOT radar layover errors [@durandHowWillRadar2020a].
+While the SWOT simulator mentioned above is ideal for this error nalysis, it is cumbersome to run on many rivers and so here we coarsely model measurement error using the results from a recent global anlaysis of SWOT water surface slope/height errors [@durandHowWillRadar2020a].
 They developed a first-principles model for layover error from SWOT observations and, using global hydrography, quantified SWOT measurement error from both error tolerances and radar layover for 220,924 river reaches.
 The 68th percentiles of these reach-scale errors were 1.7cm/km and 10.4cm for water surface slope and height, respectively.
-To apply this model to our 22 rivers, we generate new river heights *H* and water surface slopes $S_h$ by sampling every observation from normal distributions following equations 7 and 8.
+To apply this model to our 22 rivers, we generate new river heights and water surface slopes by sampling every observation from normal distributions following equations 7 and 8 for node *i* and timestep *t*.
 Because BIKER ingests *dA*, rather than *H*, reach-scale height errors were converted to *dA* errors, again following [@durandHowWillRadar2020a] and our rectangular channel assumptions: $\sigma_{dA} \approx W\sigma_H\sqrt{2}$.
 
-$$S_{err_{i,t}} \sim normal(S_{obs_{i,t}}, 0.000017^{2}) (7)$$
+$$S_{err_{i,t}} \sim normal(S_{obs_{i,t}}, 0.000017^{2}) \mathbf{(7)}$$
 
-$$H_{err_{i,t}} \sim normal(H_{obs_{i,t}}, 0.104^{2}) (8)$$
+$$H_{err_{i,t}} \sim normal(H_{obs_{i,t}}, 0.104^{2}) \mathbf{(8)}$$
 
@@
 
-$$k_{600, obs}=a_i(gS_hV_{obs})^{b_i} \text{(9)}$$
+$$k_{600, obs}=a_i(gS_hV_{obs})^{b_i} \mathbf{(9)}$$
 
@@
 Atmospheric $CO_2$ was assumed 390 uatm.
-The Schmidt number, used to obtain $k_{CO2}$ from $k_{600}$, was calculated following [@raymondScalingGasTransfer2012a] and [@wanninkhofRelationshipWindSpeed1992a].
+The Schmidt number, used to obtain $k_{CO2}$ from $k_{600}$, was calculated following @raymondScalingGasTransfer2012a and @wanninkhofRelationshipWindSpeed1992a.
 
@@
 See Table S1 for their definitions.
-It is worth stressing that the [@lauerwaldSpatialPatternsCO22015] model was developed as one of two components of the [@raymondGlobalCarbonDioxide2013a] model, and so was actually developed by the latter’s authors.
+It is worth stressing that the @lauerwaldSpatialPatternsCO22015 model was developed as one of two components of the @raymondGlobalCarbonDioxide2013a model, and so was actually developed by the latter’s authors.
 The names used here refer to the specific implementation used in each study.
@@
 We then take the best performing upscaling function, implement it within BIKER, and validate it on the 22 SWOT rivers (section 3.2).
-Finally, we use $k_{600}$ as modeled by both BIKER and gauge-based HG models to compare $FCO_2$ and bulk carbon efflux estimates from the 22 rivers across gauged and ungauged methods (section 3.3).
+Finally, we use $k_{600}$ as modeled by both BIKER and gauge-based HG models to compare $FCO_2$ and bulk carbon efflux estimates made using both gauged and ungauged methods (section 3.3).
 
@@
 Figure 3 plots the validations for the five upscaling functions tested.
-Both the original [@ulsethDistinctAirWater2019a] model and our RS-able extension perform the best on this independent set of validation data, with near identical performance.
-The RS-able [@ulsethDistinctAirWater2019a] model achieves an $r^2$ of `r results_3_1[results_3_1$Model == 'RS-able',]$r2` and an RMSE of `r results_3_1[results_3_1$Model == 'RS-able',]$rmse` m/day, which is just marginally better than the original [@ulsethDistinctAirWater2019a] model (`r results_3_1[results_3_1$Model == 'Ulseth',]$r2` and `r results_3_1[results_3_1$Model == 'Ulseth',]$rmse` m/day, respectively).
-The two power-law based [@raymondScalingGasTransfer2012a] models (equations 4 and 3 in that study) achieve slightly worse performance, with $r^2$ in the 0.70s and their RMSE scores slightly higher (`r results_3_1[results_3_1$Model == 'Raymond_2',]$rmse` and `r results_3_1[results_3_1$Model == 'Raymond_3',]$rmse` m/day, respectively).
-[@raymondScalingGasTransfer2012a] equation 5, which is linear in structure, worked well on their dataset but fails to capture the non-linear relationship evident in this expanded and more geomorphically diverse dataset.
+Both the original @ulsethDistinctAirWater2019a model and our RS-able extension perform the best on this independent set of validation data, with near identical performance.
+The RS-able @ulsethDistinctAirWater2019a model achieves an $r^2$ of `r results_3_1[results_3_1$Model == 'RS-able',]$r2` and an RMSE of `r results_3_1[results_3_1$Model == 'RS-able',]$rmse` m/day, which is just marginally better than the original @ulsethDistinctAirWater2019a model (`r results_3_1[results_3_1$Model == 'Ulseth',]$r2` and `r results_3_1[results_3_1$Model == 'Ulseth',]$rmse` m/day, respectively).
+The two power-law based @raymondScalingGasTransfer2012a models (equations 4 and 3 in that study) achieve slightly worse performance, with $r^2$ in the 0.70s and their RMSE scores slightly higher (`r results_3_1[results_3_1$Model == 'Raymond_2',]$rmse` and `r results_3_1[results_3_1$Model == 'Raymond_3',]$rmse` m/day, respectively).
+@raymondScalingGasTransfer2012a equation 5, which is linear in structure, worked well on their dataset but fails to capture the non-linear relationship evident in this expanded and more geomorphically diverse dataset.
 In fact, it predicts negative $k_{600}$ values and error metrics cannot be calculated because the prediction residuals are not normally distributed.
-**SHOULD I INCLUDE THIS MODEL??** Because we successfully reproduced the [@ulsethDistinctAirWater2019a] model, the rest of this manuscript deals exclusively with this RS-able variant of the model.
+**SHOULD I INCLUDE THIS MODEL or not mention it??** Because we successfully reproduced the @ulsethDistinctAirWater2019a model, the rest of this manuscript deals exclusively with this RS-able variant of the model.
 
@@
 
-Uncertainties were then propagated through the RS-able [@ulsethDistinctAirWater2019a] model via MC simulations for three error scenarios (section 2.2) for the 5,000 hydraulic measurements.
+Uncertainties were then propagated through the RS-able @ulsethDistinctAirWater2019a model via MC simulations for the three error scenarios (section 2.2) for the 5,000 hydraulic measurements.
 Figures 4a, 4c, and 4e plot histograms of the 5,000 uncertainty terms extracted from those distributions (Figures 4b, 4d, and 4f plot three of the distributions used to build the histograms).
-Across the 5,000 tests for the fullest error scenario (Figure 4e), we quantified average ln-transformed uncertainty in equation 4 to be `r round(results_3_1_MC$medianSigma,2)` (~`r round(exp(results_3_1_MC$medianSigma), 2)` m/day- dashed blue line in Figure 4a).
-Uncertainty solely from the upscaling model is also 1.27, confirming that functionally all uncertainity in the equation 4 estimates is due to upscaling and not due to the assumptions made for BIKER (specifically, using Manning’s equation to approximate average flow velocity and SWOT measuurement error).
-This also confirms the large residuals inherent in $k_{600}$ upscaling models as described by @hallGasExchangeStreams2020.
+Across the 5,000 tests for the fullest error scenario (Figure 4e), we quantified median ln-transformed uncertainty in equation 4 to be `r round(results_3_1_MC$medianSigma,2)` (~`r round(exp(results_3_1_MC$medianSigma), 2)` m/day- dashed blue line in Figure 4a).
+Uncertainty solely from the upscaling model is also 1.27, confirming that functionally all uncertainity in the equation 4 estimates is due to upscaling and not due to the assumptions made for BIKER (specifically, using Manning’s equation to approximate average flow velocity and SWOT measurement error).
+Figures 4b, 4d, and 4f are also functionally identical, further confirming this result.
 
@@
 Regardless, the RMSE for the BIKER predictions is only `r round(10^results_3_2_all$rmse,2)` m/day across all predictions.
-Again, this is substantially less than the RMSEs for the upscaling models versus field measurements (Figure 3) and the MC-simulated uncertainties (Figure 4a, 4c, and 4e).
+Again, this is substantially less than the RMSEs for the upscaling models versus field measurements (Figure 3) and the MC-simulated uncertainties (Figure 4).
 
@@
 While median KGE and rBIAS scores were strong, the ranges of these scores were somewhat large (standard deviation for KGE of `r round(sd(results_3_2_rivs$kge, na.rm=T),2)` and for rBIAS of `r round(sd(results_3_2_rivs$rBIAS, na.rm=T),2)`).
-This suggests that BIKER predicts $k_{600}$ very well in some rivers, but poorly in other rivers.
 
@@
 2019).
-b-d: validation timeseries for three rivers representative of good, reasonable, and poor performance: b was randomly selected from upper tertile of KGE scores, c was randomly selected from the middle tertile, and d from the worst tertile.
+b-d: validation timeseries for three rivers representative of good, reasonable, and poor performance.
+b was randomly selected from the upper tertile of KGE scores, c was randomly selected from the middle tertile, and d from the worst tertile.
 Model results include the posterior means and 95% CIs.](outputs/validation/validation_by_river.jpg)
@@
 There is functionally perfect recovery of $FCO_2$ in the Kanawha River, very good recovery (with some positive bias in peak evasion events) for the Ohio Section 2 River, and far less negative bias in the Downstream Mississippi river than is present in the $k_{600}$ estimates (Figure 5b).
-Again, the temporal dynamics are accurately modeled however there is some bias in the magnitude of the predictions.
+Again, the temporal dynamics are accurately modeled in the Kanawha and Ohio Rivers, however there is some bias in the magnitude of the predictions.
 This suggests that the BIKER model can easily estimate the changing dynamics of gas exchange but is at the mercy of the prior on $k_{600}$ to accurately estimate the magnitude of the evasion.

### 2021-02-19 scripts/FCO2_analysis.R

@@ -303,7 +303,7 @@ plotRivs <- ggplot(plot_stats, aes(x=key2, y=value, fill=key)) +
   geom_hline(yintercept=-0.41, linetype='dashed') +
   xlab('Metric') +
   ylab('Value') +
-  coord_cartesian(ylim = c(-1,1))+
+  coord_cartesian(ylim = c(-1,1.5))+
   scale_fill_brewer(palette = 'Set1', name='') +
   theme(legend.position = "none",
         axis.text=element_text(size=20),

### 2021-02-19 scripts/validation.R

@@ -12,7 +12,7 @@ sumsq <- function(x) sum(x^2)
 
 #read in results----------------------------------
 results <- read.csv('outputs//validation//BIKER_validation_results.csv')
-full_output <- filter(results, errFlag == 0)
+full_output <- filter(results, errFlag == 0) #remove results with SWOT measurement error
 
 lm_kfit <- lm(log10(full_output$kest_mean)~log10(full_output$kobs))
 r2 <- round(summary(lm_kfit)$r.squared, 2)
@@ -64,21 +64,17 @@ k600_cdfs <- ggplot(t, aes(x=value, color=key, linetype=factor(flag))) +
         legend.text = element_text(size=17),
         legend.title = element_text(size=17, face='bold'))
 
-valPlot <- plot_grid(valPlot, k600_cdfs, ncol=2, labels=c('a', 'b'))
+valPlot <- plot_grid(valPlot, k600_cdfs, ncol=2, labels=c('a', 'b'), label_size=18)
 ggsave('outputs//validation//validation.jpg', valPlot, width=12, height=7)
 
 #by river metrics---------------------------------------------
-#errors <- full_output_errors$kest_mean
-#full_output2 <- cbind(full_output, errors)
-#full_output2 <- gather(results, key=key, value=value, c('kest_mean', 'errFlag'))
 stats_by_reach <- group_by(results, river, errFlag) %>%
   summarise(kge = hydroGOF::KGE(kest_mean, kobs),
-            nse =   hydroGOF::NSE(kest_mean, kobs),
             nrmse = sqrt(mean((kobs - kest_mean)^2)) / mean(kobs, na.rm=T),
             rBIAS =   mean(kest_mean- kobs) / mean(kobs, na.rm=T),
             rrmse =   sqrt(mean((kest_mean- kobs)^2 / kobs^2)))
 
-plot_stats <- gather(stats_by_reach, key=key, value=value, c('nse', 'nrmse', 'rBIAS', 'rrmse', 'kge'))
+plot_stats <- gather(stats_by_reach, key=key, value=value, c('nrmse', 'rBIAS', 'rrmse', 'kge'))
 plot_stats <- filter(plot_stats, key %in% c('nrmse', 'rrmse', 'rBIAS', 'kge'))
 
 plotSWOTreaches <- ggplot(plot_stats, aes(x=key, y=value, fill=factor(errFlag))) +
[…]

### 2021-02-19 manuscript_v1.Rmd
min obs slope bug

@@
 - We introduce the BIKER algorithm to remotely sense riverine gas exchange using the SWOT satellite
-- BIKER replicates estimates of $CO_2$ fluxes when coupled with in situ concentration data
+- BIKER reproduces estimates of $CO_2$ fluxes when coupled with in situ gas concentration data
 - Expected SWOT measurement errors exert a trivial influence on BIKER's estimates
@@
 
-$$a_i^{-1}(gS_h^\frac{-3}{2}W^\frac{2}{3})^{b_i} \sim normal(k_{600}^{-1}\frac{1}{n}(A_0+dA)^\frac{2}{3})^{b_i}, \sigma^2_{upscale})  \text{(5)}$$
+$$a_i^{-1}(gS_h^\frac{-3}{2}W^\frac{2}{3})^{b_i} \sim normal(k_{600}^{-1}\frac{1}{n}(A_0+dA)^\frac{2}{3})^{b_i}, \sigma^2_{upscale})  \mathbf{(5)}$$
 
@@
 A few studies have investigated these dynamics, but they have been limited to individual rivers and/or limited field seasons [@hallAirWaterOxygen2012; @sand-jensenCO2DynamicsDanish2012].
-[@wallinSpatiotemporalVariabilityGas2011] performed a preliminary analysis in northern Sweden relating river specific (‘at-a-station’) temporal variability in gas exchange with channel slope, but they were limited to an average of only 8 measurements per river in a single watershed.
+@wallinSpatiotemporalVariabilityGas2011 performed a preliminary analysis in northern Sweden relating river specific (‘at-a-station’) temporal variability in gas exchange with channel slope, but they were limited to an average of only 8 measurements per river in a single watershed.
 This limited knowledge of large-scale $k_{600}$ spatiotemporal dynamics is due both to a lack of process-level understanding [@hallGasExchangeStreams2020] but also a lack of measurements.
@@
 
-###4.2 Estimating bulk carbon efflux from SWOT
+###4.2 Estimating bulk carbon efflux using the SWOT satellite
 Section 3.3 confirms that BIKER is successful at 1) reproducing $FCO_2$ without any in situ river information for $k_{600}$ (Figure 7) and 2) reproducing the bulk carbon efflux without any in situ information for $k_{600}$ (Figure 5.8).
 This encouraging result has two main implications for future work.
-First, we can likely couple BIKER with in situ gas concentration loggers to produce $FCO_2$ estimates at novel temporal resolutions, given that the temporal resolution of $k_600$ has arguably lagged behind recent advances in gas concentration data **get citations from Kelly**.
+First, section 2.4 confirms that we can couple BIKER and SWOT with in situ gas concentration loggers to produce $FCO_2$ estimates at novel temporal resolutions in SWOT-observable rivers.
+This is particualry useful given recent advances in high temporal resoultion in situ gas concentration loggers **get citations from Kelly**.
 BIKER can likely also be ran at the field scale using arrays of pressure transducers to estimate water surface slope following recent work doing the same using the [@hagemannBAMBayesianAMHGManning2017a] RSQ algorithm [@harlanDischargeEstimationDense].
 
-Secondly, it is important to stress that, unlike BIKER, the HG models in Figure 8 rely on an in situ streamgauge.
+Secondly, it is important to stress that unlike BIKER, the HG models in Figure 8 rely on an in situ streamgauge.
 This means that Figure 8 represents the best performance that those models could ever have; if ran using modeled discharge their accuracy would necessarily decrease, though this decrease would not be too much due to the generally good performance of those data.
 Thus, these results suggest that BIKER will be useful in two settings: upscaling in ungauged rivers as hypothesized, but also in potentially improving our bulk efflux understandings at gauged sites too.
-Future work should systematically quantify prediction error from using both global-scope HG and modeled discharge, like is the default workflow used in recent upscaling studies [e.g.
+Future work should systematically quantify prediction error from coupling global-scope HG models with modeled discharge, as is the default workflow used in recent upscaling studies [e.g.
 @lauerwaldSpatialPatternsCO22015, @horgbyUnexpectedLargeEvasion2019].
-**Iassume Shaoda's paper will still be in review and can't cite here **
+**I assume Shaoda's paper will still be in review and can't cite here **
 
@@
 
-###4.2 $K_{600}$ estimate uncertainity from physical processes verus from RS
+###4.2 Sources of BIKER estimate uncertainit: process-level or remote sensing?
 Throughout the BIKER validation, we have assumed no parameter uncertainity in the upscaling parameters $a_i$ and $b_i$.
@@
 Therefore, BIKER provides similar estimates of $k_{600}$ to those produced by current upscaling models and most of the total uncertainity stems from the upscaling model itself (Figure 4).
-Therefore, we argue that BIKER's estimate uncertainity is limited by current process-level knowledge of $k_{600}$ upscaling and not by BIKER or SWOT, suggesting that BIKER's predictive performance can only improve from advancing our understandings of the processes governing gas exchange from SWOT-observable rivers.
-Validating the BIKER algorithm has provided a unique lense to view $k_600$ upscaling through, and in that context we next explore some considerations for future work on $k_{600}$ upscaling in SWOT-observable rivers.
+We argue that BIKER's estimate uncertainity is limited by current process-level knowledge of $k_{600}$ upscaling and not by BIKER or SWOT, suggesting that BIKER's predictive performance can only improve from advancing our understandings of the physical processes governing gas exchange from SWOT-observable rivers.
+Developing and validating the BIKER algorithm has provided a unique lense in which we can probe current understandings of riverine gas exchange theory.
+In that context, we next explore some considerations for future work on $k_{600}$ upscaling in SWOT-observable rivers.
 
-####4.2.1 Gas exhange under uniform and non-uniform flow conditions
+####4.2.1 Gas exchange under uniform and non-uniform flow conditions
 Section 2.2 highlights that most upscaling studies to date have assumed uniform flow conditions (i.e.
@@
 All code to build and generate results, figures, and the manuscript is available at https://github.com/craigbrinkerhoff/RSK600.
-Data used in this study was made generously available by Frasson et al.
+Data used in this study was made generously available by the authors of Frasson et al.
 (2019), Ulseth et al.

### 2021-02-19 main.R

@@ -1,21 +1,27 @@
-#Runs the three analysis scripts for this project.
-#Does not actually implement the algorithm (i.e. runBIGEER.R)
-  #but the results of those implmentations are included in the inputs directory.
+#################
+#uncomment this code if you want to run the project on your local machine. Optimized for parallel computing on an HPC so only the strongest machines will survive!!!
+#################
 
-#k600 model development
-source("scripts//k600_model.R")
-rm(list = ls())
-print('k model done')
 
-#Uncertainity analysis
-source("scripts//monte_carlo_analysis.R")
-rm(list = ls())
-print('MC simulations done')
-
-#Produce validation figures
-source("scripts//validation_figures.R")
-rm(list = ls())
-print('Validation figures done. Hopefully you had the correct validation results stored in the output directory!!')
+# #k600 model development
+# source("scripts//k600_analysis.R")
+# rm(list = ls())
+# 
+# #Uncertainity analysis
+# source("scripts//MonteCarlo_analysis.R")
+# rm(list = ls())
+# 
+# #run BIKER
+# source("scripts//runBIKER.R")
+# rm(list = ls())
+# 
+# #Produce validation figures
+# source("scripts//validation.R")
+# rm(list = ls())
+# 
+# #FCo2 analysis
+# source("scripts//FCO2_analysis.R")
+# rm(list = ls())
 
 #Move all results to global environment
 results_3_1 <- read.csv('outputs//k600//results.csv')

### 2021-02-19 scripts/runBIKER.R

@@ -55,7 +55,7 @@ run_BIKER <- function(currPepsi, errFlag) {
 
   #prep pepsi rivers------------------------------------------------
   S_obs[S_obs<=0]=NA
-  S_obs[is.na(S_obs)] = 0.000001 #min obs SWOT slope
+  S_obs[is.na(S_obs)] = 0.000017 #min obs SWOT slope Biancarma etal 2016
   W_obs[W_obs<0]=NA
   H_obs[H_obs<0]=NA
   area[area<0]=NA
@@ -124,8 +124,8 @@ run_BIKER <- function(currPepsi, errFlag) {
     set.seed(100)
     for(i in 1:nrow(S_obs)) {
       for(j in 1:ncol(S_obs)){
-        S_obs[i,j] <- rnorm(1,S_obs[i,j], 1.7e-5) #km/km
-        S_obs[i,j] <- ifelse(S_obs[i,j] <= 0, 0.000001, S_obs[i,j]) #min obs slope Biancarma 2016
+        S_obs[i,j] <- rnorm(1,S_obs[i,j], 1.7e-5) #Durand etal 2020 [km/km]
+        S_obs[i,j] <- ifelse(S_obs[i,j] <= 0, 0.000017, S_obs[i,j]) #min obs slope Biancarma 2016
       }
     }
 

### 2021-02-22 manuscript_v1.Rmd
manuscript

@@
 - BIKER reproduces estimates of $CO_2$ fluxes when coupled with in situ gas concentration data
-- Expected SWOT measurement errors exert a trivial influence on BIKER's estimates
+- Expected SWOT measurement errors exert a marginal influence on BIKER's estimates
 - BIKER's accuracy is limited by process-level uncertainties and not by its algorithmic setup
@@
 ##1 Introduction
-The inland water drainage network both transports and transforms various constiutents as they move from the landscape to the ocean and a byproduct of many of these processes is riverine greenhouse gas (GHG) production [@colePlumbingGlobalCarbon2007a].
-Rivers are usually supersaturated with dissolved GHGs and present a significant flux of these gases from water to air [@raymondGlobalCarbonDioxide2013a].
-This flux $[M/L^2T]$ is calculated, with in situ knowledge of the gas concentration gradient between the water $C_{water} [M/L^3]$ and the air $C_{air} [M/L^3]$ and the gas exchange velocity $k [L/T]$, as equation 1.
+The inland water drainage network both transports and transforms various constiutents as they move from the landscape to the ocean [@colePlumbingGlobalCarbon2007a].
+As a result, rivers and lakes are usually supersaturated with dissolved greenhouse gases (GHGs) and present a significant flux of these gases from water to air [@raymondGlobalCarbonDioxide2013a].
+This flux $[M/L^2T]$ is calculated as equation 1 for the gas concentration gradient between the water $C_{water} [M/L^3]$ and the air $C_{air} [M/L^3]$ and the gas exchange velocity $k [L/T]$.
 Note that *k* is generally studied as normalized by a Schmidt number of 600 ($k_{600}$) to be gas and temperature agnostic [see @hallGasExchangeStreams2020 for a thorough review of gas exchange in rivers].
@@
 This is convenient for upscaling because in rivers and streams *e* can be equated to the turbulent energy dissipation rate *eD* $[L^2/T^3]$ [@ulsethDistinctAirWater2019a; @raymondScalingGasTransfer2012a], which is estimated using easily calculated hydraulic parameters: $eD \approx gSV$ where *g* is gravitational acceleration $[L/T^2]$, *S* is channel slope, and *V* is average flow velocity $[L/T]$ [@tsivoglouTracerMeasurementReaeration1976].
-Because *S* is readily available in any hydrographic data product, most efforts to upscale $k_{600}$ to drainage networks are limited by the quality of their estimates of the *V* of the river channel.
-This is exasperated in ungauged basins where *V* models cannot be trained because no in situ hydraulic information is available and fieldwork is impractical [@gleasonRemoteSensingRiver2020].
+Because *S* is readily available in any hydrographic data product, most efforts to upscale $k_{600}$ to drainage networks are limited by the quality of their estimates of the velocity of the river channel.
+This is exasperated in ungauged basins where velocity models cannot be trained because no in situ hydraulic information is available and fieldwork is impractical [@gleasonRemoteSensingRiver2020].
 
@@
 Many, but not all, of these ungauged approaches were developed in the context of the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission.
-**I don't know how if introducing SWOT via RSQ is the best idea.
+**I don't know if introducing SWOT via RSQ is the best idea.
 Maybe I should just introduce SWOT on its own?**
@@
 We hypothesize that remote sensing $k_{600}$ is possible by coupling Bayesian remote sensing techniques (in a manner similar to McFLI) and SWOT data with hydraulic geometry and gas exchange theory.
-More specifically, this manuscript aims to answer two questions:
-	1) Is ungauged RS of the gas exchange velocity possible using soon-to-be-available SWOT river measurements?
-	2) How might this algorithm influence upscaled estimates of riverine carbon efflux to the atmosphere?
+More specifically, this manuscript aims to answer two questions: <br>
+	1) is ungauged RS of the gas exchange velocity possible using soon-to-be-available SWOT river measurements? <br>
+	2) how might this algorithm influence upscaled estimates of riverine carbon efflux to the atmosphere?
 
@@
 We validate the algorithm on simulated SWOT data for 22 rivers (as SWOT has yet to launch, simulated data is necessary- Section 2.3).
-We also quantify the sensitivity of our algorithm to the expected measurement error implicit in SWOT data.
-Finally, we compare our algorithm against published methods for estimating carbon dioxide evasion ($FCO_2$) from rivers: we use previously published dissolved $CO_2$ data to compare the bulk carbon efflux (via $CO_2$ evasion) from the 22 rivers as calculated using both our new algorithm and published gauged techniques.
+We also quantify the sensitivity of our algorithm to the expected measurement errors implicit in SWOT measurements.
+Finally, we compare our algorithm against published methods for estimating carbon dioxide evasion ($FCO_2$) from rivers: we use previously published dissolved $CO_2$ data to compare the bulk carbon efflux (via $CO_2$ evasion) from the 22 rivers as calculated using both our new algorithm and published gauge-based techniques.
 
@@
 ###2.1 Developing an RS-able $k_{600}$ upscaling model
-[@raymondScalingGasTransfer2012a] first used river hydraulics and the theoretical relationship between *eD* and gas exchange to upscale $k_{600}$ from average flow velocity and channel slope.
+@raymondScalingGasTransfer2012a first used river hydraulics and the theoretical relationship between *eD* and gas exchange to upscale $k_{600}$ from average flow velocity and channel slope.
 This extended earlier work predicting the reaeration coefficient in a similar manner [e.g.
 @oconnorMechanismReaerationNatural1958; @tsivoglouTracerMeasurementReaeration1976].
-Later, [@ulsethDistinctAirWater2019a] built upon [@raymondScalingGasTransfer2012a] by expanding their dataset to include measurements from steeper rivers and finding that two distinct upscaling regimes exist in low and high energy rivers.
-More specifically, [@ulsethDistinctAirWater2019a] scale $k_{600}$ using *eD*, where the resulting upscaling model parameters are significantly different whether *eD* is high or low.
+Later, @ulsethDistinctAirWater2019a built upon @raymondScalingGasTransfer2012a by expanding their dataset to include measurements from steeper rivers and finding that two distinct upscaling regimes exist in low and high energy rivers.
+More specifically, @ulsethDistinctAirWater2019a scale $k_{600}$ using *eD*, where the resulting upscaling model parameters are significantly different whether *eD* is high or low.
 Their upscaling function is reprinted as equation 2.
@@
 This model therefore assigns $a_i$ and $b_i$ a priori using just SWOT measurements of water surface width and slope to define *i*.
-Another implication of this extension of the [@ulsethDistinctAirWater2019a] model is that we now use *a* and *b* parameters that implicitly reflect the distinct geomorphology of different rivers.
+Another implication of this extension of the @ulsethDistinctAirWater2019a model is that we now use *a* and *b* parameters that implicitly reflect the distinct geomorphology of different rivers.
 
-Finally, we tested whether this model can reasonably estimate $k_{600}$ relative to the original [@ulsethDistinctAirWater2019a] model, with the goal of at least replicating their performance.
-We compared it against the various *eD*-based upscaling models from [@raymondScalingGasTransfer2012a] that are commonly used in the literature.
-This was done by training all upscaling functions on 80% of the [@ulsethDistinctAirWater2019a] data and comparing their performance on the remaining 20% of the data.
-Because the [@raymondScalingGasTransfer2012a] models were originally trained on a subset of this dataset, we deemed it fairer to refit their parameters to this expanded dataset (but keep model structure identical).
+Finally, we tested whether this model can reasonably estimate $k_{600}$ relative to the original @ulsethDistinctAirWater2019a model, with the goal of at least replicating their performance.
+We compared it against the various *eD*-based upscaling models from @raymondScalingGasTransfer2012a that are commonly used in the literature.
+This was done by training all upscaling functions on 80% of the @ulsethDistinctAirWater2019a data and comparing their performance on the remaining 20% of the data.
+Because the @raymondScalingGasTransfer2012a models were originally trained on a subset of this dataset, we deemed it fairer to refit their parameters to this expanded dataset (but keep model structure identical).
 
@@
 With a validated upscaling model that requires no a priori knowledge of *eD* (results presented in section 3.1) we turn to developing the algorithm to remotely sense $k_{600}$.
-The approach used here is strongly informed by [@hagemannBAMBayesianAMHGManning2017a]’s algorithm for ungauged RSQ via Bayesian inference.
+The approach used here is strongly informed by @hagemannBAMBayesianAMHGManning2017a’s algorithm for ungauged RSQ via Bayesian inference.
 Consult that paper for the logic behind conceptualizing the equation for discharge as a Bayesian inference problem, which we largely follow here to conceptualize remote sensing $k_{600}$ as a Bayesian inference problem that can be solved using SWOT data.
@@
 SWOT will provide river width and height (and thus water surface slope $S_h$) at nodes/cross-sections along a set of river reaches.
-Thus, our ‘data’ for the Bayesian inference are river width and water surface slope while all other terms are treated as statistical parameters.
+Thus, our ‘data’ for the Bayesian inference are SWOT measurements of river width and water surface slope while all other terms are treated as statistical parameters.
 We also obtain a ‘change in channel area’ from the river height and width and assuming a rectangular channel geometry.
@@
 In order to use Bayesian inference to solve for $k_{600}$, we need to derive a sampling model/likelihood function.
-Following @tsivoglouTracerMeasurementReaeration1976, $eD \approx gS_hV$ for unsteady flow conditions (channel slope $S_0 \neq S_h$- see text S1 for this model derivation and why assuming non-uniform flow is convenient for ingesting SWOT data).
+Following @tsivoglouTracerMeasurementReaeration1976, $eD \approx gS_hV$ for unsteady flow conditions (channel slope $S_0 \neq S_h$- see text S1 for this model derivation and why assuming non-uniform flow is convenient for an algorithm ingesting SWOT data).
 Using this, as well as Manning’s equation to relate average flow velocity to channel geometry, we rewrite equation 3 as equation 4.
@@
 (2017) which formulates depth via river width and area (in order to use SWOT-observables).
-$k_{600}$ is now defined as a function of $S_h$, channel roughness n $[T/L^{\frac{1}{3}}]$, channel width *W* [L], and channel area $A_0+dA [L^2]$ (for the change in channel area *dA* and median channel area $A_0$).
+$k_{600}$ is now defined as a function of $S_h$, channel roughness n $[T/L^{\frac{1}{3}}]$, channel width *W* [L], and channel area $A_0+dA [L^2]$ (discretized via the change in channel area *dA* and median channel area $A_0$).
 It is also defined by the upscaling parameters $a_i$ and $b_i$.
@@
 Equation 4 is then reworked into a Bayesian sampling model with the ‘data’ on the left-hand side sampled from the unknown model parameters on the right-hand side (equation 5).
-Note that in the formal likelihood specification each term in equation 5 is log-transformed to yield a normal sampling model, however they are written here without a transformation for succinctness.
-The $σ_{upscale}^2$ parameter refers to the uncertainty inherent in the $k_{600}$ estimates themselves (equation 4).
+Note that in the formal likelihood specification both sides of equation 5 are ln-transformed to produce the normal sampling model presented here, however they are written here without that transformation for succinctness.
+The $σ_{upscale}^2$ parameter refers to the uncertainty inherent in the upscaled $k_{600}$ estimates themselves (equation 4).
 
@@
 Or more intuitively for these physcial quantities, they represent our ‘prior river knowledge’ of what *n*, $A_0$, and $k_{600}$ probably are for some river given that they cannot be remotely sensed.
-All priors are formalized within the model as truncated normal distributions of the ln-transformed terms such that $ln(X) \sim N(\mu, \sigma^2)$ for $\lambda < x < \epsilon$, using prior hyperparameters mean ($\mu$), standard deviation ($\sigma$), and upper ($\epsilon$) and lower bounds ($\lambda$) for parameter $X=n$, $A_0$, or $k_{600}$.
+All priors are formalized within the model as truncated normal distributions of the ln-transformed terms such that $ln(X) \sim N(\mu, \sigma^2)$ for $\lambda < X < \epsilon$, using prior hyperparameters mean ($\mu$), standard deviation ($\sigma$), and upper ($\epsilon$) and lower bounds ($\lambda$) for parameter *X*.
 In order to avoid relying on in situ information, we assign prior hyperparameters using SWOT data only.
@@
 
-Finally, we estimated $σ_{upscale}^2$ using Monte Carlo (MC) methods to approximate total uncertainty from equation 4.
-Specifically, we explore three uncertainty scenarios to assess which source of error is most influential.
+Finally, we estimate $σ_{upscale}^2$ using Monte Carlo (MC) methods to approximate total uncertainty from equation 4.
+Specifically, we explore three uncertainty scenarios to assess the realtive inlfuences of the different sources of uncertainity.
 Uncertainty in equation 4 can stem from three sources: 1) error baked into the upscaling model ($a_i$ and $b_i$ parameter uncertainty), 2) error in Manning’s approximation of the average flow velocity, and 3) measurement error.
@@
 Estimate uncertainty for velocity followed @hagemannBAMBayesianAMHGManning2017a, who estimated ln-transformed Manning’s equation error to be 0.25.
-The specifics of the slope-error model that we used are elaborated on in Section 2.3 and rely on a recent analysis of SWOT slope measurement errors [@durandHowWillRadar2020a].
-This results in three sets of 5,000 discrete distributions for $k_{600}$ from which the median uncertainty term for each uncertainity scenario is extracted and used for $σ_{upscale}^2$.
+The slope-error model that we used is explained in detail in Section 2.3 and relies on a recent analysis of SWOT slope measurement errors [@durandHowWillRadar2020a].
+This results in three sets of 5,000 discrete distributions for $k_{600}$ from which the median uncertainty term for each  scenario is extracted and used for $σ_{upscale}^2$.
 
-With the sampling model (equation 5), prior distributions, and the $σ_{upscale}^2$ parameter described, a joint posterior distribution conditional on the SWOT data is described.
+With the sampling model (equation 5), prior distributions, and the $σ_{upscale}^2$ parameter described, a joint posterior distribution conditional on the SWOT data is obtained.
 To approximate this distribution, BIKER uses a Markov Chain Monte Carlo (MCMC) algorithm.
@@
 Here, we use the 22 rivers archived by @frassonCompilationHydraulicModels2019a and used by @rodriguezObservingRiversVarying2020 to develop reach-averaged Saint Vernant equations for the SWOT mission.
-We omitted the Arial Khan river in Bangladesh due to known problems with that hydraulic model and we further sampled the models for only the observations every 11 days to mimic the average SWOT overpass frequency (this also approximately aligns the SWOT observations with the bi-weekly $CO_2$ data used later in this study- Section 2.4).
+We omitted the Arial Khan river in Bangladesh due to known problems with that hydraulic model (**is that sufficient to say?**) and we further sampled the models for only the observations every 11 days to mimic the average SWOT overpass frequency (this also approximately aligns the SWOT observations with the bi-weekly $CO_2$ data used later in this study- Section 2.4).
 All told, this yielded 503 sets of SWOT observations to validate BIKER with.
@@
 Here, we assume errors in river width are negligible [@durandHowWillRadar2020a] and solely focus on measurement errors in river height and slope.
-Layover error is the phenomenon when radar returns from different places arrive at the sensor at the same time, SOMETHING HERE.
-This generally occurs in areas of high topographic variation [@reesPhysicalPrinciplesRemote2013] where the land surface 'obscures' the water surface relative to SWOT's incidence angle.
-While the SWOT simulator mentioned above is ideal for this error nalysis, it is cumbersome to run on many rivers and so here we coarsely model measurement error using the results from a recent global anlaysis of SWOT water surface slope/height errors [@durandHowWillRadar2020a].
+Layover error is the phenomenon when radar returns from different places arrive at the sensor at the same time, leading to taller landscape features appearing closer to the sensor than shorter landscape features that are the same horizontal distance from the sensor.
+This will occur when the landscape slope is steeper than the radar's incidence angle [@durandHowWillRadar2020a] and so generally occurs in areas of high topographic variation [@reesPhysicalPrinciplesRemote2013].
+
+To explore the effect of SWOT measurement errors on BIKER, the SWOT simulator mentioned above is the ideal tool.
+However, it is cumbersome to run on many rivers and so here we coarsely model measurement error using the results from a recent global anlaysis of SWOT water surface slope/height errors [@durandHowWillRadar2020a].
 They developed a first-principles model for layover error from SWOT observations and, using global hydrography, quantified SWOT measurement error from both error tolerances and radar layover for 220,924 river reaches.
@@
 Because BIKER ingests *dA*, rather than *H*, reach-scale height errors were converted to *dA* errors, again following [@durandHowWillRadar2020a] and our rectangular channel assumptions: $\sigma_{dA} \approx W\sigma_H\sqrt{2}$.
+Note that equation 7 is also used to define slope measurement error in the MC analysis (section 2.2).
 
@@
 It also means that, for a fair validation scheme, $\sigma_{upscale}^2$ must be set to reflect only error from Manning’s equation (and not $a_i$ and $b_i$).
-**PROBABLY NEED TO ACCOUNT FOR SLOPE UNCERTAINITY TOO**.
+**MIGHT NEED TO ACCOUNT FOR SLOPE UNCERTAINITY TOO?
+I can probably justify not doing this though**.
 Thus, $\sigma_{upscale}^2$ is set to 0.25 for this validation.
@@
 It is one thing to to predict $k_{600}$, but researchers are usually more interested in the gas fluxes from rivers and ultimately the bulk carbon efflux from river to atmosphere.
-Therefore, we explored 1) the ability of BIKER to reproduce hypothetical carbon dioxide evasion fluxes ($FCO_2 [M/L^2T]$) from these 22 rivers, and 2) the sensitivity of the estimated bulk carbon efflux (via $CO_2$ evasion) to the method used to estimate reach averaged flow velocity.
+Therefore, we explored 1) the ability of BIKER to reproduce hypothetical carbon dioxide evasion fluxes ($FCO_2 [M/L^2T]$) from these 22 rivers, and 2) the sensitivity of the estimated bulk carbon efflux $[M/T]$ (via $CO_2$ evasion) to the method used to estimate reach averaged flow velocity.
 
@@
 Not all of the SWOT rivers have observations for a full year, and when simulation dates were not available they were assumed to start on January 1st.
-Note that this $CO_2$ data is for the Ohio River only but was applied to all 22 rivers (which includes the Ohio River).
+Note that this $CO_2$ data is for the Ohio River only but was applied to all 22 rivers (which includes multiple sections of the Ohio River).
 Because we are exclusively interested in the relative differences between $FCO_2$ estimates and not the raw fluxes themselves, any $CO_2$ data representative of SWOT-observable rivers was deemed acceptable.
@@
 This allows us to explore the relative differences in bulk carbon efflux estimates if implementing a wholly ungauged method (BIKER) versus gauged methods (all three HG models).
-Finally, we express the bulk carbon efflux (via $CO_2$ evasion) $[M/T]$ as the average mass flow rate of carbon per year from the 22 rivers, accounting for total river surface area.
+Finally, we express the bulk carbon efflux as the average mass flow rate of carbon (via $CO_2$ evasion) per year from the 22 rivers after accounting for total river surface area.
 
@@
 We then take the best performing upscaling function, implement it within BIKER, and validate it on the 22 SWOT rivers (section 3.2).
-Finally, we use $k_{600}$ as modeled by both BIKER and gauge-based HG models to compare $FCO_2$ and bulk carbon efflux estimates made using both gauged and ungauged methods (section 3.3).
+Finally, we use $k_{600}$ as modeled by both BIKER and gauge-based HG models to compare $FCO_2$ and bulk carbon efflux estimates made using gauged and ungauged methods (section 3.3).
 
@@
 
-Uncertainties were then propagated through the RS-able @ulsethDistinctAirWater2019a model via MC simulations for the three error scenarios (section 2.2) for the 5,000 hydraulic measurements.
-Figures 4a, 4c, and 4e plot histograms of the 5,000 uncertainty terms extracted from those distributions (Figures 4b, 4d, and 4f plot three of the distributions used to build the histograms).
+Uncertainties were then propagated through the RS-able @ulsethDistinctAirWater2019a model via MC simulations for the three uncertainity scenarios (section 2.2) 5,000 times.
+Figures 4a, 4c, and 4e plot histograms of the 5,000 uncertainty terms extracted from those distributions (Figures 4b, 4d, and 4f plot three of the distributions used to build the histograms as examples).
 Across the 5,000 tests for the fullest error scenario (Figure 4e), we quantified median ln-transformed uncertainty in equation 4 to be `r round(results_3_1_MC$medianSigma,2)` (~`r round(exp(results_3_1_MC$medianSigma), 2)` m/day- dashed blue line in Figure 4a).
-Uncertainty solely from the upscaling model is also 1.27, confirming that functionally all uncertainity in the equation 4 estimates is due to upscaling and not due to the assumptions made for BIKER (specifically, using Manning’s equation to approximate average flow velocity and SWOT measurement error).
-Figures 4b, 4d, and 4f are also functionally identical, further confirming this result.
+While the cases when estimate uncertainity is extremely high get larger under the fullest uncertainity scenario (Figure 4e right-hand tail), median uncertainty solely from the upscaling model is also 1.27, confirming that functionally all uncertainity in the equation 4 estimates is due to upscaling and not due to the assumptions made for BIKER.
+Figures 4b, 4d, and 4f are also functionally identical, further corroborating this result.
 
@@
 However, the 95% prediction intervals (dashed grey lines) highlight reasonably large residuals for many predictions.
-Regardless, the RMSE for the BIKER predictions is only `r round(10^results_3_2_all$rmse,2)` m/day across all predictions.
-Again, this is substantially less than the RMSEs for the upscaling models versus field measurements (Figure 3) and the MC-simulated uncertainties (Figure 4).
+Regardless, the RMSE for the BIKER predictions is only `r round(results_3_2_all$rmse,2)` (or `r round(10^results_3_2_all$rmse,2)`) m/day across all predictions.
+Again, this is substantially less than the RMSEs for the upscaling models (Figure 3) and the MC-propogated uncertainties (Figure 4).
 
 ![Figure 5.
-a: Validation of remote sensing algorithm for 22 rivers.
+a: Validation of BIKER for 22 SWOT rivers.
 Black bars are 95% CIs for the modeled values.
@@
 Figure 5a suggests a flip in the bias of the $k_{600}$ predictions, which is confirmed in Figure 5b.
-Figure 5b plots the cumulative density functions (CDFs) of observed and predicted $k_{600}$ where the green line is the BIKER posterior mean and the blue lines are the CDFs of the BIKER posterior 95% CIs.
+Figure 5b plots the cumulative density functions (CDFs) of observed and predicted $k_{600}$ where the green line is the CDF of the BIKER posterior means and the blue lines are the CDFs of the BIKER posterior 95% CIs.
 We see, more clearly than in Figure 5a, that the posterior mean is systematically underestimated for values less than the median $k_{600}$ and systemically overestimated for values above the median $k_{600}$.
@@
 
-Figure 6a plots validation metrics calculated for each river (i.e.
-the boxplots are composed of scores for the 22 rivers with and without measurement error- see Table 1 for metric definitions).
-SWOT measurement uncertainties exert a neglible influence on 3/4 performance metrics and slightly decrease KGE across the 22 rivers (Figure 6a).
-With SWOT's errors not exerting a significant influence on BIKER, the results presented for the rest of the manuscript assume no measurement error.
-Median rBIAS is `r round(median(results_3_2_rivs$rBIAS, na.rm=T),2)`, indicating nearly no bias in most rivers’ predictions.
+Figure 6a plots validation metrics calculated for each river with and without SWOT measurement error (i.e.
+the boxplots are composed of scores for the 22 rivers- see Table 1 for metric definitions).
+SWOT measurement uncertainties slightly degrade performance in KGE, neglibily degrade rBIAS and RRMSE, and actually improve performance in NRMSE (Figure 6a).
+Aside from the modest decrease in KGE, SWOT measurement error does not exert a significant influence on BIKER and so the results presented for the rest of the manuscript assume no measurement error.
+Median rBIAS is `r round(median(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$rBIAS, na.rm=T),2)`, indicating nearly no bias in most rivers’ predictions.
 However, some rivers are substantially biased in both directions, further supporting the visual evidence in Figure 5 that sometimes BIKER is substantially under/overestimating $k_{600}$ and that this is river-specific.
-Median KGE is `r round(median(results_3_2_rivs$kge, na.rm=T),2)`, which is excellent given that absolutely no in situ information is being used to predict $k_{600}$.
-NRMSE and RMSE have median scores of `r round(median(results_3_2_rivs$nrmse, na.rm=T),2)` and `r round(median(results_3_2_rivs$rrmse, na.rm=T),2)`, respectively.
-While median KGE and rBIAS scores were strong, the ranges of these scores were somewhat large (standard deviation for KGE of `r round(sd(results_3_2_rivs$kge, na.rm=T),2)` and for rBIAS of `r round(sd(results_3_2_rivs$rBIAS, na.rm=T),2)`).
+Median KGE is `r round(median(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$kge, na.rm=T),2)`, which is excellent given that absolutely no in situ information is being used to predict $k_{600}$.
+NRMSE and RMSE have median scores of `r round(median(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$nrmse, na.rm=T),2)` and `r round(median(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$rrmse, na.rm=T),2)`, respectively.
+While median KGE and rBIAS scores were strong, the ranges of these scores were somewhat large (standard deviation for KGE of `r round(sd(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$kge, na.rm=T),2)` and for rBIAS of `r round(sd(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$rBIAS, na.rm=T),2)`).
 
@@
 Figure 6b-d are representative timeseries plots of predicted and observed $k_{600}$ for three rivers chosen randomly from those with ‘good’ KGE scores (b), ‘okay’ KGE scores (c), and ‘bad’ KGE scores (d).
-See the Figure 5.6 caption for how this was calculated.
+See the Figure 6 caption for how this was calculated.
 For the Kanawha River, the entire timeseries of $k_{600}$ is correctly predicted, while in the Ohio Section 2 River there is a negative bias in the estimates.
@@
 BIKER has similar NRMSE and RRMSE performance, which is generally worse than the rating curves.
-But, BIKER NRMSE is only modestly worse than 'Raymond 2013 and 'Raymond 2012' but BIKER RRMSE (median RRMSE: `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_BIKER',]$rrmse, na.rm=T),2)`) is significantly worse than the gauge-based approaches.
+But, BIKER NRMSE is only slightly worse than 'Raymond 2013 and 'Raymond 2012' but BIKER RRMSE (median RRMSE: `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_BIKER',]$rrmse, na.rm=T),2)`) is significantly worse than the gauge-based approaches.
 RRMSE is highly sensitive to prediction errors in small values and so this suggests that BIKER performs poorly when predicting small $CO_2$ fluxes.
-In summary, BIKER performance for KGE, NRMSE, and rBIAS is either similar or negligibly worse than the gauge-based upscaling approaches tested here.
+In summary, BIKER performance for KGE, NRMSE, and rBIAS is either better, similar, or negligibly worse than the gauge-based upscaling approaches tested here and notably worse for RRMSE.
 This is despite relying on absolutely no in situ information like all three other models do.
@@
 @alinPhysicalControlsCarbon2011; @beaulieuControlsGasTransfer2012a; @schelkerCO2EvasionSteep2016a], 3) discharge [e.g.
-@robertsMultipleScalesTemporal2007; @uehlingerEcosystemMetabolismDisturbance1998], or 4) oxygen diel curves (where gas exchange is still assumed constant in time- e.g.
-@applingOvercomingEquifinalityLeveraging2018; @demarsStreamMetabolismOpen2015).
-However, the spatiotemporal dynamics of gas exchange are still weakly constrained.
+@robertsMultipleScalesTemporal2007; @uehlingerEcosystemMetabolismDisturbance1998], or 4) oxygen diel curves [where gas exchange is still assumed constant in time- e.g.
+@applingOvercomingEquifinalityLeveraging2018; @demarsStreamMetabolismOpen2015].
+However, the spatiotemporal dynamics of riverine gas exchange are still weakly constrained.
 A few studies have investigated these dynamics, but they have been limited to individual rivers and/or limited field seasons [@hallAirWaterOxygen2012; @sand-jensenCO2DynamicsDanish2012].
@@
 SWOT will provide daily hydraulic measurements for a 3 month fast sampling period for calibration and validation and sampling thereafter between 1 and 7 days per 21 day repeat cycle [@biancamariaSWOTMissionIts2016].
-BIKER's success in 1) infering $k_{600}$ using simulated SWOT data over a wide range of rivers (Figures 5 and 6) and 2) being robust to measurement errors internal to the SWOT data (Figure 6a) bode well for BIKER's eventual implementation on real SWOT observations.
+BIKER's success in 1) infering $k_{600}$ using simulated SWOT data over a wide range of rivers (Figures 5 and 6) and 2) being robust to measurement errors internal to the SWOT data (Figures 4 and 6a) bode well for BIKER's eventual implementation on real SWOT observations.
 The results presented here suggest that daily estimation of riverine gas exchange globally could be possible once SWOT launches.
@@
 This encouraging result has two main implications for future work.
-First, section 2.4 confirms that we can couple BIKER and SWOT with in situ gas concentration loggers to produce $FCO_2$ estimates at novel temporal resolutions in SWOT-observable rivers.
+First, section 2.4 confirms that we can couple BIKER and SWOT with in situ gas concentration loggers to produce $FCO_2$ estimates at novel temporal resolution in SWOT-observable rivers.
 This is particualry useful given recent advances in high temporal resoultion in situ gas concentration loggers **get citations from Kelly**.
-BIKER can likely also be ran at the field scale using arrays of pressure transducers to estimate water surface slope following recent work doing the same using the [@hagemannBAMBayesianAMHGManning2017a] RSQ algorithm [@harlanDischargeEstimationDense].
+BIKER can likely also be ran at the field scale using arrays of pressure transducers to estimate water surface slope following recent work doing the same using the @hagemannBAMBayesianAMHGManning2017a RSQ algorithm [@harlanDischargeEstimationDense].
 
@@
 This means that Figure 8 represents the best performance that those models could ever have; if ran using modeled discharge their accuracy would necessarily decrease, though this decrease would not be too much due to the generally good performance of those data.
-Thus, these results suggest that BIKER will be useful in two settings: upscaling in ungauged rivers as hypothesized, but also in potentially improving our bulk efflux understandings at gauged sites too.
+Thus, these results suggest that BIKER will be useful in two settings: 1) upscaling in ungauged rivers as hypothesized, but also in 2) potentially improving our bulk efflux understandings at gauged sites.
 Future work should systematically quantify prediction error from coupling global-scope HG models with modeled discharge, as is the default workflow used in recent upscaling studies [e.g.
 @lauerwaldSpatialPatternsCO22015, @horgbyUnexpectedLargeEvasion2019].
-**I assume Shaoda's paper will still be in review and can't cite here **
+**I assume Shaoda's paper will still be in review and can't cite here...**
 
@@
 Figure 8 confirms this.
-Meanwhile, BIKER has no similar reliance on hydraulic parameters trained on different river sizes and only assumes that Manning's equation is viable in the river [which is largely the case in rivers wide enough to be SWOT-observable- @fergusonTimeAbandonManning2010].
-Upon SWOT's launch, the BIKER approach to estimating gas exchange could be coupled with existing upscaling workflows to improve riverine gas flux predictions where gauges are unavailable.
+Meanwhile, BIKER has no similar reliance on hydraulic parameters trained on different river sizes and only assumes that Manning's equation is viable in the river [which is generally the case in rivers wide enough to be SWOT-observable- @fergusonTimeAbandonManning2010].
+Upon SWOT's launch, the BIKER approach to estimating gas exchange could be coupled with existing upscaling workflows to improve riverine gas flux predictions where gauges are unavailable but SWOT measurements are.
 
-###4.2 Sources of BIKER estimate uncertainit: process-level or remote sensing?
+###4.2 Sources of BIKER uncertainty: process-level or remote sensing?
 Throughout the BIKER validation, we have assumed no parameter uncertainity in the upscaling parameters $a_i$ and $b_i$.
 We have shown that BIKER estimate uncertainty is significantly less than $a_i$ and $b_i$ uncertainity (Figures 4 and 5).
-Therefore, BIKER provides similar estimates of $k_{600}$ to those produced by current upscaling models and most of the total uncertainity stems from the upscaling model itself (Figure 4).
-We argue that BIKER's estimate uncertainity is limited by current process-level knowledge of $k_{600}$ upscaling and not by BIKER or SWOT, suggesting that BIKER's predictive performance can only improve from advancing our understandings of the physical processes governing gas exchange from SWOT-observable rivers.
-Developing and validating the BIKER algorithm has provided a unique lense in which we can probe current understandings of riverine gas exchange theory.
+BIKER provides similar estimates of $k_{600}$ to those produced by current upscaling models and most of the total uncertainity stems from the upscaling model itself (Figure 4).
+We argue that BIKER's estimate uncertainity is therefore limited by our current process-level understanding of $k_{600}$ upscaling and not by BIKER or SWOT, suggesting that BIKER's predictive performance can only improve from advancing our understandings of the physical processes governing gas exchange from SWOT-observable rivers.
+Developing and validating the BIKER algorithm has provided a unique lense through which we can study riverine gas exchange theory.
 In that context, we next explore some considerations for future work on $k_{600}$ upscaling in SWOT-observable rivers.
@@
 $S_h \neq S_0$) in order to train upscaling models using readily available slope data.
-However, the first-principles model previously used by [@ulsethDistinctAirWater2019a] and [@raymondScalingGasTransfer2012a] to define *eD* [@tsivoglouTracerMeasurementReaeration1976] does not make this simplifying assumption.
+However, the first-principles model previously used by @ulsethDistinctAirWater2019a and @raymondScalingGasTransfer2012a to define *eD* [@tsivoglouTracerMeasurementReaeration1976] does not make this simplifying assumption.
 Therefore, it is an open research question whether parameterizing $k_{600}$ upscaling models via $S_h$ can account for some of the unexplained residual variation in current upscaling models [@hallGasExchangeStreams2020].
 Conveniently, SWOT will explicitly measure $S_h$ at unprecendeted spatial and temporal resolutions and will be spatially joined to hydrography that provides $S_0$.
-Thus, future researchers can use BIKER and SWOT in conjunction to directly explore this question.
+Thus, future researchers can use BIKER and SWOT in conjunction to directly answer this question.
 
 ####4.2.2 Bed roughness and gas exchange in SWOT rivers
-Channel bed roughness effects riverine gas exchange mostly in high energy streams where slope is sufficiently steep for bubble-induced gas exchange to occur [@hallGasExchangeStreams2020].
-[@ulsethDistinctAirWater2019a] showed bed roughness loosely scales with $k_{600}$ in steep Alpine streams.
+Channel bed roughness affects riverine gas exchange mostly in high energy streams where slope is sufficiently steep for bubble-induced gas exchange to occur [@hallGasExchangeStreams2020].
+@ulsethDistinctAirWater2019a showed bed roughness loosely scales with $k_{600}$ in steep Alpine streams.
 However, they coarsely estimated bed roughness from arial imagery and to date most similar work has focused on labratory exercises [e.g.
 @chansonStreamReaerationNonuniform; moogStreamReaerationNonuniform].
-We argue that bed roughness is not controlling gas exchange in SWOT-observable rivers because they are so large (and therefore flat per river network geomorphology).
+We argue that bed roughness is not controlling gas exchange in SWOT-observable rivers because they are so large and therefore flat.
 Instead, gas exchange in SWOT-observable rivers is presumably dominated by water-column, rather than bed friction, turbulence.
-We show in Figure S2 with the [@ulsethDistinctAirWater2019a] data that the 'effective bed roughness height' scales with $k_{600}$ only in extremely steep streams (see Text S2 for the calculation of this bed roughness term).
+We show in Figure S2 with the @ulsethDistinctAirWater2019a data that the 'effective bed roughness height' scales with $k_{600}$ only in extremely steep streams (see Text S2 for the calculation of this bed roughness term).
 This relationship fundamentally breaks down in less steep rivers, and such steep slopes are functionally impossible in SWOT-observable rivers that are over 50m wide.
 This promising initial result indicates bed roughness controls some aspects of gas exchange and should be explicitly explored in future work.
-This is particularly important because small, steep rivers dominate global river networks due to the fractal nature of river systems [@tarbotonFractalNatureRiver1988].
+This is particularly important because small, steep rivers dominate global river networks and their GHG evasion due to the fractal nature of river systems [@horgbyUnexpectedLargeEvasion2019; @tarbotonFractalNatureRiver1988].
 

### 2021-03-02 BIKER.Rmd
ms update

---
title: "Introduction to BIKER"
author: "Craig Brinkerhoff"
date: "February 25, 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

BIKER enables estimation of the normalized riverine gas exchange velocity from measurements of river surface width and slope. It does so using Bayesian inference and a Hamiltonian Monte Carlo sampler to generate a posterior distribution for a gas exchange upscaling model. BIKER was developed in the context of the NASA/CNES/UKSA/CSA SWOT mission (https://swot.jpl.nasa.gov/mission/overview/). The academic manuscript associated with this model development and validation is in preperation.

To run BIKER, the following workflow is ideal. Note that BIKER follows standard hydrology workflow for the discretization of the river observations along a mass-conserved river reach. This means that matrix rows represent spatial steps along the reach while matrix columns represent timesteps for these observations.

## Installation
BIKER is always under active development and the working version is available on GitHub. Follow [this link](https://github.com/craigbrinkerhoff/BIKER) to download.

## Workflow
#### Inputs
BIKER requires 3 inputs: <br>
- $Wobs$: a matrix of water surface widths <br>
- $Sobs$: a matrix of water surface slopes <br> 
- $dAobs$: a matrix of 'change in cross-sectional channel areas'. This must be approximated by the user. We generally assume a rectangular river channel so that this can be calculated as the following: $width*\delta H$.

#### Run BIKER
The following is the series of fuctions that need to be run to use BIKER. See the references tab for more information about the function options. <br>

```
reach_data <- biker_data(w=Wobs, s=Wobs, da=dAobs) #collect data into object the algorithm can read
reach_priors <- biker_priors(reach_data) #estimate prior hyperparameters for Bayesian inference
k600_estimates <- biker_estimate(reach_data, reach_priors) #sample from joint posterior distribution
```

#### Manually specifying prior hyperparameters
Priors can also be overwritten manually if you so desire. This is done as following, for example, for the A0_hat prior which has a value for every spatial step in the river reach:
```
reach_priors <- biker_priors(reach_data, logA0_hat = rep(5000, ncol(W_obs)))
```

If you want the full list of customizable prior hyperparameters, run the following:
```
prior_settings()
```

### 2021-03-05 manuscript_vc/manuscript_v1.Rmd
manuscript v2

--- 2021-02-22 manuscript_v1.Rmd
@@
 
-##Key Points (85 characters each)
-- We introduce the BIKER algorithm to remotely sense riverine gas exchange using the SWOT satellite
-- BIKER reproduces estimates of $CO_2$ fluxes when coupled with in situ gas concentration data
+##Highlights (3-5 points, 85 characters each w/ spaces)
+- The BIKER algorithm can remotely sense riverine gas exchange using the SWOT satellite
+- BIKER predicts $CO_2$ fluxes when coupled with in situ gas concentration data
+- Predicted $CO_2$ fluxes are similar/marginally worse than when using in-situ methods
 - Expected SWOT measurement errors exert a marginal influence on BIKER's estimates
-- BIKER's accuracy is limited by process-level uncertainties and not by its algorithmic setup
+- Accuracy is limited by process-level uncertainties and not by the algorithmic setup
+
+##Keywords
+gas exchange, SWOT, remote sensing, carbon, ungauged basin, biogeochemistry
 
@@
 The inland water drainage network both transports and transforms various constiutents as they move from the landscape to the ocean [@colePlumbingGlobalCarbon2007a].
-As a result, rivers and lakes are usually supersaturated with dissolved greenhouse gases (GHGs) and present a significant flux of these gases from water to air [@raymondGlobalCarbonDioxide2013a].
+As a result some of these processes, rivers and lakes are usually supersaturated with dissolved greenhouse gases (GHGs) and present a significant flux of these gases from water to air [@raymondGlobalCarbonDioxide2013a].
 This flux $[M/L^2T]$ is calculated as equation 1 for the gas concentration gradient between the water $C_{water} [M/L^3]$ and the air $C_{air} [M/L^3]$ and the gas exchange velocity $k [L/T]$.
@@
 This is convenient for upscaling because in rivers and streams *e* can be equated to the turbulent energy dissipation rate *eD* $[L^2/T^3]$ [@ulsethDistinctAirWater2019a; @raymondScalingGasTransfer2012a], which is estimated using easily calculated hydraulic parameters: $eD \approx gSV$ where *g* is gravitational acceleration $[L/T^2]$, *S* is channel slope, and *V* is average flow velocity $[L/T]$ [@tsivoglouTracerMeasurementReaeration1976].
-Because *S* is readily available in any hydrographic data product, most efforts to upscale $k_{600}$ to drainage networks are limited by the quality of their estimates of the velocity of the river channel.
+Because *S* is readily available in any hydrographic data product, most efforts to upscale $k_{600}$ to drainage networks are limited by the quality of their estimates of the reach-averaged flow velocity.
 This is exasperated in ungauged basins where velocity models cannot be trained because no in situ hydraulic information is available and fieldwork is impractical [@gleasonRemoteSensingRiver2020].
@@
 Bayesian techniques have been previously used to estimate $k_{600}$ concurrently with stream metabolism from dissolved oxygen ($O_2$) datasets [@applingOvercomingEquifinalityLeveraging2018; @graceFastProcessingDiel2015; @holtgrieveSimultaneousQuantificationAquatic2010].
-While this approach requires detailed field data and is thus not applicable to $k_{600}$ estimation in ungauged rivers, it does suggest that Bayesian techniques could be useful for estimating $k_{600}$ when parameter equifinality is a problem (as it is with McFLIs).
+While this approach requires detailed field data and is thus not applicable to $k_{600}$ estimation in ungauged rivers, it does suggest that Bayesian techniques could be useful for remotely sensing $k_{600}$ when parameter equifinality is a problem as it is with McFLIs [@andreadisConstrainingAssimilationSWOT2020d].
+Equinality issues refer to under-constrained systems that can have multiple solutions to the same discharge problem [@garamboisInferenceEffectiveRiver2015].
 
-In this context, we hypothesize that SWOT data will improve our understanding of riverine GHG fluxes in ungauged basins and/or at drainage-network-scales after techniques have been developed to ingest SWOT's novel measurements of water surface width and height.
+In this context, we hypothesize that SWOT data will improve our understanding of riverine GHG fluxes in ungauged basins and/or at drainage-network-scales we develop techniques to ingest SWOT's novel measurements of water surface width and height.
 Therefore, we sought to develop a method to estimate $k_{600}$ using only SWOT measurements and no in situ information.
 We hypothesize that remote sensing $k_{600}$ is possible by coupling Bayesian remote sensing techniques (in a manner similar to McFLI) and SWOT data with hydraulic geometry and gas exchange theory.
-More specifically, this manuscript aims to answer two questions: <br>
-	1) is ungauged RS of the gas exchange velocity possible using soon-to-be-available SWOT river measurements? <br>
-	2) how might this algorithm influence upscaled estimates of riverine carbon efflux to the atmosphere?
+More specifically, this manuscript aims to answer two questions:
+
+- Is ungauged RS of the gas exchange velocity possible using soon-to-be-available SWOT river measurements?
+- How might this algorithm influence upscaled estimates of riverine carbon efflux to the atmosphere?
 
@@
 While equation 2 would be ideal for the goal of this manuscript and provides best-to-date predictive accuracy for upscaling $k_{600}$, it requires an a priori knowledge of *eD* to assign its differential model parameters.
-We do not have this a priori knowledge in an ungauged, RS-based setting and only have access to SWOT-observed river widths and heights (and therefore water surface slopes).
-Thus, we need to extend the [@ulsethDistinctAirWater2019a] model to assign the two model parameters using only RS-able hydraulics.
+We do not have this a priori knowledge in an ungauged, RS-oriented setting and only have access to SWOT-observed river widths and heights (and therefore water surface slopes).
+Thus, we need to extend the @ulsethDistinctAirWater2019a model to assign the two model parameters using only RS-able hydraulics.
 To do this, we use the basic premise that the $k_{600}\sim eD$ scaling relationship is different for different river sizes.
@@
 The approach used here is strongly informed by @hagemannBAMBayesianAMHGManning2017a’s algorithm for ungauged RSQ via Bayesian inference.
-Consult that paper for the logic behind conceptualizing the equation for discharge as a Bayesian inference problem, which we largely follow here to conceptualize remote sensing $k_{600}$ as a Bayesian inference problem that can be solved using SWOT data.
+Consult that paper for the logic behind conceptualizing the equation for discharge as a Bayesian remote sensing problem, which we largely follow here to conceptualize $k_{600}$ as a Bayesian remote sensing problem that can be solved using SWOT data.
 
 SWOT will provide river width and height (and thus water surface slope $S_h$) at nodes/cross-sections along a set of river reaches.
-Thus, our ‘data’ for the Bayesian inference are SWOT measurements of river width and water surface slope while all other terms are treated as statistical parameters.
-We also obtain a ‘change in channel area’ from the river height and width and assuming a rectangular channel geometry.
+Thus, our ‘data’ for the Bayesian inference are SWOT measurements of river width and water surface slope while all other terms are treated as statistical parameters that cannot be directly observed from SWOT.
+We also obtain a ‘change in channel area’ *dA* from the river height and width and assuming a rectangular channel geometry.
 
-In order to use Bayesian inference to solve for $k_{600}$, we need to derive a sampling model/likelihood function.
+In order to use Bayesian inference to solve for $k_{600}$, we need a sampling model/likelihood function that estimates the unobservable statisitcal parameters.
 Following @tsivoglouTracerMeasurementReaeration1976, $eD \approx gS_hV$ for unsteady flow conditions (channel slope $S_0 \neq S_h$- see text S1 for this model derivation and why assuming non-uniform flow is convenient for an algorithm ingesting SWOT data).
@@
 (2017) which formulates depth via river width and area (in order to use SWOT-observables).
-$k_{600}$ is now defined as a function of $S_h$, channel roughness n $[T/L^{\frac{1}{3}}]$, channel width *W* [L], and channel area $A_0+dA [L^2]$ (discretized via the change in channel area *dA* and median channel area $A_0$).
+$k_{600}$ is now defined as a function of $S_h$, channel roughness *n* $[T/L^{\frac{1}{3}}]$, channel width *W* [L], and channel area $A_0+dA [L^2]$ (discretized via median channel area $A_0$ and the previously introduced *dA*).
 It is also defined by the upscaling parameters $a_i$ and $b_i$.
 
-$$k_{600}=a_i(gS_h^{\frac{3}{2}}\frac{1}{n}(A_0+dA)^\frac{2}{3}W^\frac{-2}{3})^{b_i} \mathbf{(4)}$$
+$$k_{600}=a_i\left[gS_h^{\frac{3}{2}}\frac{1}{n}(A_0+dA)^\frac{2}{3}W^\frac{-2}{3}\right]^{b_i} \mathbf{(4)}$$
 
-Equation 4 is then reworked into a Bayesian sampling model with the ‘data’ on the left-hand side sampled from the unknown model parameters on the right-hand side (equation 5).
-Note that in the formal likelihood specification both sides of equation 5 are ln-transformed to produce the normal sampling model presented here, however they are written here without that transformation for succinctness.
-The $σ_{upscale}^2$ parameter refers to the uncertainty inherent in the upscaled $k_{600}$ estimates themselves (equation 4).
+Equation 4 is then reworked into a Bayesian sampling model to have all of the ‘data’ on the left-hand side sampled from all the unknown model parameters on the right-hand side (equation 5, where *x* is the SWOT observables and $\Theta$ are a set of statistical parameters).
+Note that in the formal likelihood specification both sides of equation 5 are ln-transformed to produce the normal sampling model presented here, however they are written in equation 5 without that transformation for succinctness.
+The $σ_{upscale}^2$ parameter refers to the uncertainty inherent in the upscaling model itself (equation 4).
+This will be explained in detail below.
 
-$$a_i^{-1}(gS_h^\frac{-3}{2}W^\frac{2}{3})^{b_i} \sim normal(k_{600}^{-1}\frac{1}{n}(A_0+dA)^\frac{2}{3})^{b_i}, \sigma^2_{upscale})  \mathbf{(5)}$$
+$$f(x|\Theta) = \left[a_i^{-1}(gS_h^\frac{-3}{2}W^\frac{2}{3})^{b_i}\right] \sim normal\left(\left[k_{600}^{-1}\frac{1}{n}(A_0+dA)^\frac{2}{3})^{b_i}\right], \sigma^2_{upscale}\right)  \mathbf{(5)}$$
 
 Equation five necessitates that Bayesian priors be assigned to model parameters *n*, $A_0$, and $k_{600}$.
-Bayesian priors formularize the a priori estimates (and uncertainties) for these parameters.
+Bayesian priors formularize the a priori estimates (and uncertainties) for the non-remotely-sensed parameters.
 Or more intuitively for these physcial quantities, they represent our ‘prior river knowledge’ of what *n*, $A_0$, and $k_{600}$ probably are for some river given that they cannot be remotely sensed.
@@
 Finally, we estimate $σ_{upscale}^2$ using Monte Carlo (MC) methods to approximate total uncertainty from equation 4.
-Specifically, we explore three uncertainty scenarios to assess the realtive inlfuences of the different sources of uncertainity.
-Uncertainty in equation 4 can stem from three sources: 1) error baked into the upscaling model ($a_i$ and $b_i$ parameter uncertainty), 2) error in Manning’s approximation of the average flow velocity, and 3) measurement error.
+Specifically, we explore three uncertainty scenarios to assess the realtive influences of different sources of uncertainity.
+Uncertainty in equation 4 can stem from 1) error baked into the upscaling model ($a_i$ and $b_i$ parameter uncertainty), 2) error in Manning’s approximation of the average flow velocity, and 3) measurement error.
 Our interest is using BIKER for satellite remote sensing and so we consider measurement errors from SWOT-derived water surface slopes.
-To do this, we run 15,000 unique MC simulations on 5,000 sets of field measurements of river channel hydraulics [@brinkerhoffReconcilingAtaStationAtManyStations2019] under three different uncertainty scenarios.
+To do this, we run 15,000 unique MC simulations on 5,000 sets of field measurements of river channel hydraulics from @brinkerhoffReconcilingAtaStationAtManyStations2019 under three different uncertainty scenarios.
 Each MC simulation is itself 10,000 runs, sampling from the normal distributions for $a_i$ and $b_i$ (uncertainty scenario 1), $a_i$, $b_i$, and ln-transformed velocity (uncertainty scenario 2), and $a_i$, $b_i$, ln-transformed velocity, and water surface slope (uncertainty scenario 3).
 Estimate uncertainty for velocity followed @hagemannBAMBayesianAMHGManning2017a, who estimated ln-transformed Manning’s equation error to be 0.25.
-The slope-error model that we used is explained in detail in Section 2.3 and relies on a recent analysis of SWOT slope measurement errors [@durandHowWillRadar2020a].
-This results in three sets of 5,000 discrete distributions for $k_{600}$ from which the median uncertainty term for each  scenario is extracted and used for $σ_{upscale}^2$.
+The slope-error model that we used is explained in detail in Section 2.3 and relies on a recent analysis of SWOT water surface slope measurement errors [@durandHowWillRadar2020a].
+This results in three sets of 5,000 discrete distributions for $k_{600}$ from which the median uncertainty term for any scenario is extracted and used for $σ_{upscale}^2$.
+**a table might help here to layout the three tests**
 
@@
 Here, we use the 22 rivers archived by @frassonCompilationHydraulicModels2019a and used by @rodriguezObservingRiversVarying2020 to develop reach-averaged Saint Vernant equations for the SWOT mission.
-We omitted the Arial Khan river in Bangladesh due to known problems with that hydraulic model (**is that sufficient to say?**) and we further sampled the models for only the observations every 11 days to mimic the average SWOT overpass frequency (this also approximately aligns the SWOT observations with the bi-weekly $CO_2$ data used later in this study- Section 2.4).
+We omitted the Arial Khan river in Bangladesh due to known problems with that hydraulic model **is that sufficient to say?** and we further sampled the models for only the observations every 11 days to mimic the average SWOT overpass frequency (this also approximately aligns the SWOT observations with the bi-weekly $CO_2$ data used later in this study- Section 2.4).
 All told, this yielded 503 sets of SWOT observations to validate BIKER with.
@@
 To explore the effect of SWOT measurement errors on BIKER, the SWOT simulator mentioned above is the ideal tool.
-However, it is cumbersome to run on many rivers and so here we coarsely model measurement error using the results from a recent global anlaysis of SWOT water surface slope/height errors [@durandHowWillRadar2020a].
+However, it is cumbersome to run on many rivers **is this sufficient?** and so here we coarsely model measurement error using the results from a recent global anlaysis of SWOT water surface slope/height errors [@durandHowWillRadar2020a].
 They developed a first-principles model for layover error from SWOT observations and, using global hydrography, quantified SWOT measurement error from both error tolerances and radar layover for 220,924 river reaches.
 The 68th percentiles of these reach-scale errors were 1.7cm/km and 10.4cm for water surface slope and height, respectively.
-To apply this model to our 22 rivers, we generate new river heights and water surface slopes by sampling every observation from normal distributions following equations 7 and 8 for node *i* and timestep *t*.
-Because BIKER ingests *dA*, rather than *H*, reach-scale height errors were converted to *dA* errors, again following [@durandHowWillRadar2020a] and our rectangular channel assumptions: $\sigma_{dA} \approx W\sigma_H\sqrt{2}$.
+To apply this model to our 22 rivers, we generate new SWOT observations by sampling every observation from normal distributions following equations 7 and 8 for node *i* and timestep *t*.
+Note that because BIKER ingests *dA*, rather than *H*, reach-scale height errors were converted to *dA* errors again following @durandHowWillRadar2020a and our rectangular channel assumptions: $\sigma_{dA} \approx W\sigma_H\sqrt{2}$.
 Note that equation 7 is also used to define slope measurement error in the MC analysis (section 2.2).
@@
 
-$$H_{err_{i,t}} \sim normal(H_{obs_{i,t}}, 0.104^{2}) \mathbf{(8)}$$
+$$dA_{err_{i,t}} \sim normal(dA_{obs_{i,t}}, \left[W_{obs_{i,t}}*\sqrt{2}*0.104\right]^{2}) \mathbf{(8)}$$
 
 Regardless of the validation data used, we do not have observed $k_{600}$ data for these rivers, and to our knowledge no field dataset of $k_{600}$ exists in the type of temporal and spatial frequency that SWOT (and therefore the BIKER algorithm) provides.
-Further, we are principally interested in our ability to reproduce upscaled $k_{600}$ from RS observations and are less concerend with the actual accuracy of the upscaling model itself [which can be validated using existing datasets like previously done by @raymondScalingGasTransfer2012a and @ulsethDistinctAirWater2019a].
+Further, we are principally interested in our ability to reproduce upscaled $k_{600}$ from RS observations and are less concerend with the actual accuracy of the upscaling model itself [which can be validated using existing datasets like previously done by @raymondScalingGasTransfer2012a; @ulsethDistinctAirWater2019a].
 Therefore, we take the best performing upscaling model from sections 2.1 and 3.1 and use that to calculate the ‘observed’ $k_{600}$ that BIKER is validated against.
@@
 It also means that, for a fair validation scheme, $\sigma_{upscale}^2$ must be set to reflect only error from Manning’s equation (and not $a_i$ and $b_i$).
-**MIGHT NEED TO ACCOUNT FOR SLOPE UNCERTAINITY TOO?
-I can probably justify not doing this though**.
 Thus, $\sigma_{upscale}^2$ is set to 0.25 for this validation.
@@
 Validation metrics take two forms (and are detailed in Table 1).
-To validate across all rivers and timesteps, we used the coefficient of determination $r^2$ and the root mean square error (RMSE).
+To validate across all rivers and timesteps, we used the coefficient of determination $r^2$ and the root mean square error RMSE.
 Four normalized metrics were used for by-river validation: RRMSE and NRMSE are normalized root mean square errors that have been normalized by the observed value and the mean observed value (respectively) to compare across rivers.
 rBIAS is a measure of prediction bias that is normalized by the mean observed value to compare across rivers.
-KGE is a standard metric used in streamflow prediction with an intuitive basis: a value greater than 0 is often interpreted as a useful prediction in ungauged settings, and a value greater than -0.41 indicates a model outperforms a uniform prediction of the mean value [@knobenTechnicalNoteInherent2019a].
+KGE is a standard metric used in streamflow prediction with an intuitive basis: a value greater than 0 is often interpreted as a useful prediction in ungauged settings, and a value greater than -0.41 indicates a model outperforms a uniform prediction of the mean observed value [@knobenTechnicalNoteInherent2019a].
 
 *Table 1: Validation metrics used in this study, where r is the correlation coefficient, Nt is number of observations and i is the specific observation.
-σ refers to the variance of the sample and μ refers to the mean of the sample.*
+σ refers to the variance of the sample and μ refers to the mean of the sample.
+As is standard, a carrot accent indicates the predicted value.*
 
@@
 It is one thing to to predict $k_{600}$, but researchers are usually more interested in the gas fluxes from rivers and ultimately the bulk carbon efflux from river to atmosphere.
-Therefore, we explored 1) the ability of BIKER to reproduce hypothetical carbon dioxide evasion fluxes ($FCO_2 [M/L^2T]$) from these 22 rivers, and 2) the sensitivity of the estimated bulk carbon efflux $[M/T]$ (via $CO_2$ evasion) to the method used to estimate reach averaged flow velocity.
+Therefore, we explored 1) the ability of BIKER to reproduce hypothetical carbon dioxide evasion fluxes ($FCO_2$ $[M/L^2T]$) from these 22 rivers, and 2) the sensitivity of the estimated bulk carbon efflux $[M/T]$ via $CO_2$ evasion to the method used to estimate reach averaged flow velocity.
 
 First, we calculated $FCO_2$.
-To do this, we paired the modeled $k_{CO2}$ values (obtained from the $k_{600}$ estimates) with field-measurements of water-side $CO_2$ concentrations and water temperatures.
+To do this, we paired the modeled $k_{CO2}$ values (obtained from $k_{600}$) with field-measurements of water-side $CO_2$ concentrations and water temperatures.
 26 bi-weekly samples were made by @beaulieuControlsGasTransfer2012a- Figure S1) at one location in the Ohio River for one calender year from 2008-2009.
-We paired these 26 values with each 11-day-repeat set of SWOT observations by date, ignoring the timesteps beyond 26 (only ~15% of the SWOT observations were ignored here and this was deemed acceptable).
+We paired these 26 values with each 11-day-repeat set of SWOT observations by date, ignoring the timesteps beyond 26 (only ~15% of the SWOT observations were ignored here and we deemed this acceptable).
 Not all of the SWOT rivers have observations for a full year, and when simulation dates were not available they were assumed to start on January 1st.
-Note that this $CO_2$ data is for the Ohio River only but was applied to all 22 rivers (which includes multiple sections of the Ohio River).
+Note that this $CO_2$ data is for the Ohio River only but was applied to all 22 rivers (which includes multiple sections of the Ohio River- @frassonCompilationHydraulicModels2019a).
 Because we are exclusively interested in the relative differences between $FCO_2$ estimates and not the raw fluxes themselves, any $CO_2$ data representative of SWOT-observable rivers was deemed acceptable.
 Atmospheric $CO_2$ was assumed 390 uatm.
-The Schmidt number, used to obtain $k_{CO2}$ from $k_{600}$, was calculated following @raymondScalingGasTransfer2012a and @wanninkhofRelationshipWindSpeed1992a.
+The Schmidt number, used to backcalculate $k_{CO2}$ from $k_{600}$, was calculated following @raymondScalingGasTransfer2012a and @wanninkhofRelationshipWindSpeed1992a.
 
@@
 The names used here refer to the specific implementation used in each study.
-For our study, all HG methods use the in-situ discharge record while BIKER does not.
+For this study, all HG methods use the in-situ discharge record while BIKER does not.
 This allows us to explore the relative differences in bulk carbon efflux estimates if implementing a wholly ungauged method (BIKER) versus gauged methods (all three HG models).
@@
 Uncertainties were then propagated through the RS-able @ulsethDistinctAirWater2019a model via MC simulations for the three uncertainity scenarios (section 2.2) 5,000 times.
-Figures 4a, 4c, and 4e plot histograms of the 5,000 uncertainty terms extracted from those distributions (Figures 4b, 4d, and 4f plot three of the distributions used to build the histograms as examples).
+Figures 4a, 4c, and 4e plot histograms of the 5,000 uncertainty terms extracted from those distributions.
+Figures 4b, 4d, and 4f plot three of the distributions used to build the histograms as examples.
 Across the 5,000 tests for the fullest error scenario (Figure 4e), we quantified median ln-transformed uncertainty in equation 4 to be `r round(results_3_1_MC$medianSigma,2)` (~`r round(exp(results_3_1_MC$medianSigma), 2)` m/day- dashed blue line in Figure 4a).
-While the cases when estimate uncertainity is extremely high get larger under the fullest uncertainity scenario (Figure 4e right-hand tail), median uncertainty solely from the upscaling model is also 1.27, confirming that functionally all uncertainity in the equation 4 estimates is due to upscaling and not due to the assumptions made for BIKER.
-Figures 4b, 4d, and 4f are also functionally identical, further corroborating this result.
+Median uncertainty solely from the upscaling model is 1.27 (~`r exp(1.27)` m/day), confirming that functionally all uncertainity in the equation 4 estimates is due to upscaling and not due to the assumptions made for BIKER.
+The only notable difference between these histograms is in the Figure 4e right-hand tail, where the cases when estimate uncertainity is extremely high get larger.
+Figures 4b, 4d, and 4f are also functionally identical, further corroborating our conclusion.
 
@@
 ###3.2 Validating the BIKER algorithm
-Figure 5a plots the validation results for $k_{600}$ (with no measurement error) across all 22 rivers and all timesteps.
+Figure 5a plots the validation results for $k_{600}$ (with no SWOT measurement error) across all 22 rivers and all timesteps.
 For BIKER, the points are the posterior means while the black lines are the 95% confidence intervals (CIs) for the predictions.
@@
 However, the 95% prediction intervals (dashed grey lines) highlight reasonably large residuals for many predictions.
-Regardless, the RMSE for the BIKER predictions is only `r round(results_3_2_all$rmse,2)` (or `r round(10^results_3_2_all$rmse,2)`) m/day across all predictions.
+Regardless, the RMSE for the BIKER predictions is only `r round(results_3_2_all$rmse,2)` in log10 space (~`r round(10^results_3_2_all$rmse,2)` m/day) across all predictions.
 Again, this is substantially less than the RMSEs for the upscaling models (Figure 3) and the MC-propogated uncertainties (Figure 4).
@@
 This highlights one benefit of using Bayesian inference to fully propagate all prior uncertainties through to the posterior.
-While these CIs are large and thus BIKER estimates are reasonably uncertain, we do improve upon our baseline understanding of $k_{600}$ in these rivers.
+In summary, Figure 5 confirms that we improve upon our baseline understanding of $k_{600}$ in these rivers.
 Put another way, we reasonably capture $k_{600}$ with no in situ information about the river while simultaneously and explicitly accounting for the reasonably large uncertainties inherent in our $k_{600}$ estimates.
 
-Figure 6a plots validation metrics calculated for each river with and without SWOT measurement error (i.e.
-the boxplots are composed of scores for the 22 rivers- see Table 1 for metric definitions).
+Figure 6a plots validation metrics calculated for each river with and without SWOT measurement error.
+The boxplots are composed of scores for the 22 rivers- see Table 1 for metric definitions.
 SWOT measurement uncertainties slightly degrade performance in KGE, neglibily degrade rBIAS and RRMSE, and actually improve performance in NRMSE (Figure 6a).
-Aside from the modest decrease in KGE, SWOT measurement error does not exert a significant influence on BIKER and so the results presented for the rest of the manuscript assume no measurement error.
+Aside from the modest decrease in KGE, we deem that SWOT measurement error does not exert a significant influence on BIKER and so the results presented for the rest of the manuscript assume no measurement error.
 Median rBIAS is `r round(median(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$rBIAS, na.rm=T),2)`, indicating nearly no bias in most rivers’ predictions.
-However, some rivers are substantially biased in both directions, further supporting the visual evidence in Figure 5 that sometimes BIKER is substantially under/overestimating $k_{600}$ and that this is river-specific.
+However, some rivers are substantially biased in both directions, further supporting the visual evidence in Figure 5 that sometimes BIKER is substantially under/overestimating the magnitude of $k_{600}$ and that this is river-specific.
 Median KGE is `r round(median(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$kge, na.rm=T),2)`, which is excellent given that absolutely no in situ information is being used to predict $k_{600}$.
@@
 See Table 1 for metric definitions.
-Dashed lines denote scores of 1, 0, and -0.41 for KGE (Knoben et al.
-2019).
+Dashed lines denote scores of 1, 0, and -0.41 for KGE (section 3.2).
 b-d: validation timeseries for three rivers representative of good, reasonable, and poor performance.
-b was randomly selected from the upper tertile of KGE scores, c was randomly selected from the middle tertile, and d from the worst tertile.
+b) was randomly selected from the upper tertile of KGE scores, c) was randomly selected from the middle tertile, and d) from the worst tertile.
 Model results include the posterior means and 95% CIs.](outputs/validation/validation_by_river.jpg)
@@
 In the downstream Mississippi River, there is significant bias in the estimates as well as significant uncertainty (per the 95% CIs).
-The temporal dynamics are also incorrectly predicted.
+The temporal dynamics are also incorrectly predicted at times.
 
@@
 In Figure 7a, there is a strong fit to the observed $FCO_2$ data, with an RMSE of `r round(10^results_3_3_all$rmse,2)` $g/m^2dy$.
-The $r^2$ is identical to that for $k_{600}$ (Figure 7a).
+The $r^2$ is slightly higher than $k_{600}$ (Figure 7a).
 There is also less underestimation bias for the smallest of the observed $FCO_2$ than in the $k_{600}$ predictions (Figure 5a).
@@
 $FCO_2$ prediction intervals also have a similar relative magnitude to those presented in Figure 5.
-Figure 7b-d includes subplots for the same rivers as Figure 4b-d, however with $FCO_2$ instead of $k_{600}$.
+Figure 7b-d includes subplots for the same rivers as Figure 4b-d, however with $FCO_2$ plotted instead of $k_{600}$.
 There is functionally perfect recovery of $FCO_2$ in the Kanawha River, very good recovery (with some positive bias in peak evasion events) for the Ohio Section 2 River, and far less negative bias in the Downstream Mississippi river than is present in the $k_{600}$ estimates (Figure 5b).
@@
 The BIKER bulk carbon efflux (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_BIKER',]$sumFCO2, 0)` gG-C/yr) is slightly overestimated relative to to the bulk carbon efflux calculated using observed velocity (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_ulset',]$sumFCO2, 0)` gG-C/yr).
-However, it is a better estimate than those using a streamgauge and HG models (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_raymond2012',]$sumFCO2, 0)`, `r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_raymond2013',]$sumFCO2, 0)`, and `r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_Lauerwald2015',]$sumFCO2, 0)` gG-C/yr for ‘Raymond 2012’, ‘Raymond 2013’, and ‘Lauerwald 2015’, respectively).
-Thus, BIKER reasonably captures the estimated efflux from these rivers purely from RS data, and surprisingly outperforms the gauge-based approaches when assessiung bulk flux properties (Figure 8a).
+However, it is a better estimate than those using a streamgauge and HG (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_raymond2012',]$sumFCO2, 0)`, `r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_raymond2013',]$sumFCO2, 0)`, and `r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_Lauerwald2015',]$sumFCO2, 0)` gG-C/yr for ‘Raymond 2012’, ‘Raymond 2013’, and ‘Lauerwald 2015’, respectively).
+Thus, BIKER reasonably captures the estimated efflux from these rivers purely from RS data, and surprisingly outperforms the gauge-based approaches when assessing this bulk flux property (Figure 8a).
 
 ![Figure 8.
-a: Bulk carbon efflux, per year, from the 22 SWOT rivers as calculated using five different average flow velocity estimates: 1) BIKER, 2) through 4) average flow velocity as calculated using rating curves from the literature, and 5) observed average flow velocity.
-b: Cumulative density functions (CDFs) of all FCO2 estimates across all timesteps and rivers for the same four velocity models.
-The median of the distributions is noted with a black line.](outputs/FCO2/FCO2_models.jpg)
+a: Bulk carbon efflux, per year, from the 22 SWOT rivers as calculated using four different average flow velocity estimates: 1) BIKER, 2) through 4) average flow velocity as calculated using HG models from the literature.
+b: Histograms of by-river performance in estimating FCO2 across all timesteps and rivers for the same four velocity models.
+Dashed lines are identical to Figure 6a.](outputs/FCO2/FCO2_models.jpg)
 
@@
 For rBIAS, BIKER indicates the least bias in some rivers and similar bias to the other models in other rivers, as indicated by the scores' inter-quartile range (`r round(IQR(results_3_3_rivs[results_3_3_rivs$key=='FCO2_BIKER',]$rBIAS, na.rm=T),2)`) and median (`r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_BIKER',]$rBIAS, na.rm=T),2)`).
-Meanwhile, all three rating curves uniformly produce significant negative bias in their estimates (median scores of `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_Lauerwald2015',]$rBIAS, na.rm=T),2)`, `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_raymond2013',]$rBIAS, na.rm=T),2)`, and `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_raymond2012',]$rBIAS, na.rm=T),2)` for 'Lauerwald 2015', 'Raymond 2013', and 'Raymond 2012', respectively).
-BIKER has similar NRMSE and RRMSE performance, which is generally worse than the rating curves.
-But, BIKER NRMSE is only slightly worse than 'Raymond 2013 and 'Raymond 2012' but BIKER RRMSE (median RRMSE: `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_BIKER',]$rrmse, na.rm=T),2)`) is significantly worse than the gauge-based approaches.
+Meanwhile, all three HG models uniformly produce significant negative bias in their estimates (median scores of `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_Lauerwald2015',]$rBIAS, na.rm=T),2)`, `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_raymond2013',]$rBIAS, na.rm=T),2)`, and `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_raymond2012',]$rBIAS, na.rm=T),2)` for 'Lauerwald 2015', 'Raymond 2013', and 'Raymond 2012', respectively).
+BIKER has similar performance for NRMSE and RRMSE, which is generally worse than the HG models.
+But, BIKER NRMSE is only slightly worse than 'Raymond 2013 and 'Raymond 2012' while BIKER RRMSE (median RRMSE: `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_BIKER',]$rrmse, na.rm=T),2)`) is significantly worse than the gauge-based approaches.
 RRMSE is highly sensitive to prediction errors in small values and so this suggests that BIKER performs poorly when predicting small $CO_2$ fluxes.
-In summary, BIKER performance for KGE, NRMSE, and rBIAS is either better, similar, or negligibly worse than the gauge-based upscaling approaches tested here and notably worse for RRMSE.
+In summary, BIKER performance for KGE, NRMSE, and rBIAS is either similar or negligibly worse than the gauge-based upscaling approaches tested here and notably worse for RRMSE.
 This is despite relying on absolutely no in situ information like all three other models do.
@@
 To date, most field-scale studies of riverine gas exchange have focused on 1) its relationship with wind speed [e.g @beaulieuControlsGasTransfer2012a; @borgesVariabilityGasTransfer2004; @zappaEnvironmentalTurbulentMixing2007], 2) average flow velocity [e.g.
-@alinPhysicalControlsCarbon2011; @beaulieuControlsGasTransfer2012a; @schelkerCO2EvasionSteep2016a], 3) discharge [e.g.
-@robertsMultipleScalesTemporal2007; @uehlingerEcosystemMetabolismDisturbance1998], or 4) oxygen diel curves [where gas exchange is still assumed constant in time- e.g.
-@applingOvercomingEquifinalityLeveraging2018; @demarsStreamMetabolismOpen2015].
+@alinPhysicalControlsCarbon2011; @beaulieuControlsGasTransfer2012a; @schelkerCO2EvasionSteep2016a], or 3) discharge [e.g.
+@robertsMultipleScalesTemporal2007; @uehlingerEcosystemMetabolismDisturbance1998].
 However, the spatiotemporal dynamics of riverine gas exchange are still weakly constrained.
@@
 ###4.2 Estimating bulk carbon efflux using the SWOT satellite
-Section 3.3 confirms that BIKER is successful at 1) reproducing $FCO_2$ without any in situ river information for $k_{600}$ (Figure 7) and 2) reproducing the bulk carbon efflux without any in situ information for $k_{600}$ (Figure 5.8).
+Section 3.3 confirms that BIKER is successful at two things without any in situ information for $k_{600}$: 1) reproducing $FCO_2$ (Figure 7) and 2) reproducing the bulk carbon efflux (Figure 5.8).
 This encouraging result has two main implications for future work.
-First, section 2.4 confirms that we can couple BIKER and SWOT with in situ gas concentration loggers to produce $FCO_2$ estimates at novel temporal resolution in SWOT-observable rivers.
-This is particualry useful given recent advances in high temporal resoultion in situ gas concentration loggers **get citations from Kelly**.
+First, section 3.3 confirms that we can couple BIKER and SWOT with in situ gas concentration loggers to produce $FCO_2$ estimates at novel temporal resolution in SWOT-observable rivers.
+This is particualry useful given recent advances in high temporal resolution in situ gas concentration measurements **get citations from Kelly**.
 BIKER can likely also be ran at the field scale using arrays of pressure transducers to estimate water surface slope following recent work doing the same using the @hagemannBAMBayesianAMHGManning2017a RSQ algorithm [@harlanDischargeEstimationDense].
@@
 Future work should systematically quantify prediction error from coupling global-scope HG models with modeled discharge, as is the default workflow used in recent upscaling studies [e.g.
-@lauerwaldSpatialPatternsCO22015, @horgbyUnexpectedLargeEvasion2019].
-**I assume Shaoda's paper will still be in review and can't cite here...**
+@lauerwaldSpatialPatternsCO22015; @horgbyUnexpectedLargeEvasion2019].
+**I assume Shaoda's paper will still be in review and can't cite here but this is exactly what they do...**
 
@@
 Figure 8 confirms this.
-Meanwhile, BIKER has no similar reliance on hydraulic parameters trained on different river sizes and only assumes that Manning's equation is viable in the river [which is generally the case in rivers wide enough to be SWOT-observable- @fergusonTimeAbandonManning2010].
+Meanwhile, BIKER has no similar reliance on hydraulic parameters trained on different river sizes and only assumes that Manning's equation is viable in the river at hand [which is generally the case in rivers large enough to be SWOT-observable- @fergusonTimeAbandonManning2010].
 Upon SWOT's launch, the BIKER approach to estimating gas exchange could be coupled with existing upscaling workflows to improve riverine gas flux predictions where gauges are unavailable but SWOT measurements are.
@@
 We argue that BIKER's estimate uncertainity is therefore limited by our current process-level understanding of $k_{600}$ upscaling and not by BIKER or SWOT, suggesting that BIKER's predictive performance can only improve from advancing our understandings of the physical processes governing gas exchange from SWOT-observable rivers.
-Developing and validating the BIKER algorithm has provided a unique lense through which we can study riverine gas exchange theory.
+Developing and validating the BIKER algorithm has provided a unique lense through which we can discuss riverine gas exchange theory.
 In that context, we next explore some considerations for future work on $k_{600}$ upscaling in SWOT-observable rivers.
@@
 However, they coarsely estimated bed roughness from arial imagery and to date most similar work has focused on labratory exercises [e.g.
-@chansonStreamReaerationNonuniform; moogStreamReaerationNonuniform].
+@chansonStreamReaerationNonuniform; @moogStreamReaerationNonuniform].
 We argue that bed roughness is not controlling gas exchange in SWOT-observable rivers because they are so large and therefore flat.
@@
 This relationship fundamentally breaks down in less steep rivers, and such steep slopes are functionally impossible in SWOT-observable rivers that are over 50m wide.
-This promising initial result indicates bed roughness controls some aspects of gas exchange and should be explicitly explored in future work.
-This is particularly important because small, steep rivers dominate global river networks and their GHG evasion due to the fractal nature of river systems [@horgbyUnexpectedLargeEvasion2019; @tarbotonFractalNatureRiver1988].
+This promising initial result indicates bed roughness controls some aspects of gas exchange and should be explicitly explored in future work, but is less relevant for BIKER's application to SWOT data.
+This is particularly important because small, steep rivers dominate global river networks and their GHG evasion [@horgbyUnexpectedLargeEvasion2019] due to the fractal nature of river systems [@tarbotonFractalNatureRiver1988].
 
@@
 Therefore, gas evasion estimates are sensitive to available data on average flow velocity.
-In ungauged basins, this poses a problem because a velocity~discharge rating curve can not be built.
+In ungauged basins, this poses a problem because a velocity~discharge relationships can not be built.
 To circumvent this problem, we propose using established techniques from the ungauged remote sensing of river discharge literature and applying them to gas exchange velocity in ungauged rivers.
 This is formalized as the BIKER algorithm and uses data from the upcoming SWOT satellite to remotely sense gas exchange velocity solely from river width and water surface height.
-We validate BIKER on 22 'rivers' of simulated SWOT data [@frassonCompilationHydraulicModels2019a], obtaining an RMSE of `r round(10^results_3_2_all$rmse,2)` after accounting for upscaling parameter uncertainity.
+We validate BIKER on 22 'rivers' of simulated SWOT data [@frassonCompilationHydraulicModels2019a], obtaining an RMSE of `r round(10^results_3_2_all$rmse,2)` ,/day after accounting for upscaling parameter uncertainity.
 When generalized to estimate bulk carbon efflux from these 22 rivers, BIKER reasonably captures the observed efflux (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_BIKER',]$sumFCO2, 0)` gG-C/yr versus `r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_ulset',]$sumFCO2, 0)` gG-C/yr, respectively).
-Further, BIKER's $FCO_2$ estimates across the 22 rivers are only modestly worse than those made using a streamgauge and the global-scope HG models frequently used in the literature.
+Further, BIKER's $FCO_2$ estimates across the 22 rivers are either functionally the same or modestly worse than those made using a streamgauge and the HG models frequently used in the literature.
 This suggests that BIKER will be useful not only for upscaling in ungauged rivers, but also in potentially improving our understandings in gauged rivers too.
@@
 C.B.
-Brinkerhoff was funded on xxxxxxxxxxxxx.
+Brinkerhoff was funded on **xxxxxxxxxxxxx**.
 BIKER is available at https://github.com/craigbrinkerhoff/BIKER.
 All code to build and generate results, figures, and the manuscript is available at https://github.com/craigbrinkerhoff/RSK600.
-Data used in this study was made generously available by the authors of Frasson et al.
+Data used in this study was generously made available by the authors of Frasson et al.
 (2019), Ulseth et al.

### 2021-03-05 main.R

@@ -34,7 +34,7 @@ results_3_3_bulk <- read.csv('outputs//FCO2//bulkFluxes.csv')
 results_3_3_bulk$key <- as.character(results_3_3_bulk$key)
 
 #knit manuscript
-rmarkdown::render("manuscript_v1.Rmd")
+rmarkdown::render("manuscript_v2.Rmd")
 print('Manuscript generated.')
 
 rmarkdown::render("supp_v1.Rmd")

### 2021-05-05 docs/manuscript_v2.Rmd
complete overhaul of repo

--- 2021-03-05 manuscript_vc/manuscript_v1.Rmd
@@
 ---
-title: "Remote sensing riverine gas exchange and estimating carbon efflux using SWOT observations"
-author: "CB Brinkerhoff, CJ Gleason, PA Raymond, MH Harlan"
+title: "Potential for remotely sensing riverine gas exchange and carbon efflux from SWOT observations"
+author: 'CB Brinkerhoff, CJ Gleason, PA Raymond, MH Harlan'
 date: "`r Sys.Date()`"
@@
 ##Highlights (3-5 points, 85 characters each w/ spaces)
-- The BIKER algorithm can remotely sense riverine gas exchange using the SWOT satellite
-- BIKER predicts $CO_2$ fluxes when coupled with in situ gas concentration data
-- Predicted $CO_2$ fluxes are similar/marginally worse than when using in-situ methods
+- The SWOT satellite should be able to indirectly remotely sense riverine gas exchange velocity
+- The BIKER algorithm predicts $CO_2$ fluxes when coupled with in situ gas concentration data
+- Predicted BIKER $CO_2$ fluxes compare well with in-situ methods
 - Expected SWOT measurement errors exert a marginal influence on BIKER's estimates
-- Accuracy is limited by process-level uncertainties and not by the algorithmic setup
+- Errors are dominated by process-level uncertainties, not parameter inversion
 
@@
 ##1 Introduction
-The inland water drainage network both transports and transforms various constiutents as they move from the landscape to the ocean [@colePlumbingGlobalCarbon2007a].
-As a result some of these processes, rivers and lakes are usually supersaturated with dissolved greenhouse gases (GHGs) and present a significant flux of these gases from water to air [@raymondGlobalCarbonDioxide2013a].
-This flux $[M/L^2T]$ is calculated as equation 1 for the gas concentration gradient between the water $C_{water} [M/L^3]$ and the air $C_{air} [M/L^3]$ and the gas exchange velocity $k [L/T]$.
+Inland waters transport and transform various constiutents as they move from the landscape to the ocean [@colePlumbingGlobalCarbon2007a].
+**Too big a jump.
+What is the journal?
+‘constituents’ will need to be defined, as will ‘processes’**As a result some of these processes, rivers and lakes are usually supersaturated with dissolved greenhouse gases (GHGs) and thus present a significant flux of these gases from water to air [@raymondGlobalCarbonDioxide2013a].
+This flux $[M/L^2T]$ is expressed as equation 1 given the gas concentration gradient between the water $C_{water} [M/L^3]$ and the air $C_{air} [M/L^3]$ and the gas exchange velocity $k [L/T]$.
 Note that *k* is generally studied as normalized by a Schmidt number of 600 ($k_{600}$) to be gas and temperature agnostic [see @hallGasExchangeStreams2020 for a thorough review of gas exchange in rivers].
@@
 
-It is infeasible to measure $k_{600}$ directly when working at the drainage-network-scale and so ‘upscaling functions’ are used that predict $k_{600}$ from readily available river geomorphology [e.g.
+To directly measure $k_{600}$ in a given stream, scientists use a range of methods mostly involving floating domes or eddy covariance towers to measure at a point in the river, or using injections of tracer gases to estimate reach-scale gas exchange [again, see @hallGasExchangeStreams2020].
+Obiously, these types of of measuremnets are preferred but are infeasible when working at the network-scale across potentially tens of thousands of rivers, and so upscaling functions are used to predict $k_{600}$ from readily available river geomorphology parameters [e.g.
 @oconnorMechanismReaerationNatural1958; @palumboAssessingPerformanceReaeration2014; @raymondScalingGasTransfer2012a].
 Many upscaling functions are based on the basic theory that *k* is empirically correlated with the turbulent energy dissipation *e* at the air-water interface [@zappaEnvironmentalTurbulentMixing2007].
-This is convenient for upscaling because in rivers and streams *e* can be equated to the turbulent energy dissipation rate *eD* $[L^2/T^3]$ [@ulsethDistinctAirWater2019a; @raymondScalingGasTransfer2012a], which is estimated using easily calculated hydraulic parameters: $eD \approx gSV$ where *g* is gravitational acceleration $[L/T^2]$, *S* is channel slope, and *V* is average flow velocity $[L/T]$ [@tsivoglouTracerMeasurementReaeration1976].
-Because *S* is readily available in any hydrographic data product, most efforts to upscale $k_{600}$ to drainage networks are limited by the quality of their estimates of the reach-averaged flow velocity.
-This is exasperated in ungauged basins where velocity models cannot be trained because no in situ hydraulic information is available and fieldwork is impractical [@gleasonRemoteSensingRiver2020].
+This is convenient for upscaling because in rivers and streams *e* can be equated to the turbulent energy dissipation rate *eD* $[L^2/T^3]$ [@ulsethDistinctAirWater2019a; @raymondScalingGasTransfer2012a], which is estimated using easily calculated hydraulic parameters: $eD \approx gSV$ where *g* is gravitational acceleration $[L/T^2]$, *S* is channel slope $[L/L]$, and *V* is average flow velocity $[L/T]$ [@tsivoglouTracerMeasurementReaeration1976].
+Because channel slope is readily available in any hydrographic data product, most efforts to upscale $k_{600}$ to drainage networks are thereby limited by the quality of the final parameter *V*.
+Since *V* must be measured in the field or itself estimated by other models/empirical methods, uncertainty in *V* must dominates errors in $k_{600}$ estimates.
+This is exacerbated in ungauged basins that cover large sections of the earth system, especially in the carbon-rich Arctic inland waters, as no in situ hydraulic information is available and fieldwork is impractical [@gleasonRemoteSensingRiver2020].
 
-To circumvent this problem in ungauged basins/at drainage-network-scales, to date researchers have used ‘global-scope’ hydraulic geometry (HG) models: simple power law relations between hydraulic properties and discharge [@leopoldHydraulicGeometryStream1953a] that are trained on extensive hydraulics datasets.
-This is the approach used in upscaling *V* to continental/global river networks and has proven useful in those settings [e.g.
-@borgesVariabilityGasTransfer2004; @horgbyUnexpectedLargeEvasion2019; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a].
-However, it has long been established that HG parameters exhibit high spatial variability across landscapes and vary significantly from river to river [@parkWorldwideVariationsHydraulic1977; @rhodesBfmDiagramGraphical1977].
+To circumvent this problem in ungauged basins and at network-scales, researchers have used ‘global-scope’ hydraulic geometry (HG) models: simple power law relations between in this case *V* and discharge that are trained on extensive hydraulics datasets [@leopoldHydraulicGeometryStream1953a; @borgesVariabilityGasTransfer2004; @horgbyUnexpectedLargeEvasion2019; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a].
+However, it has long been established that HG parameters are highly variable across landscapes and from river to river [@gleasonHydraulicGeometryNatural2015; @parkWorldwideVariationsHydraulic1977; @rhodesBfmDiagramGraphical1977].
 Because of this, a large body of geomorphology work has attempted to parse out process-based explanations for HG parameters [e.g.
 @dingmanAnalyticalDerivationAtastation2007b; @parkerPhysicalBasisQuasiuniversal2007b; @singhTHEORIESHYDRAULICGEOMETRY2003a; @fergusonHydraulicsHydraulicGeometry1986a] and an open research question is how to best prescribe a set of HG parameters to a given river (particularly at the global scale).
-A potential alternative to this approach is to directly estimate a river’s hydraulic properties using remote sensing (RS) data.
+**Xxx give me a sentence or two on the problems with this is approach here.
+You bury the problem with the sentence “however, it has long” but then cover it up.
+The transition doesn’t work without closing this paragraph with a supported (i.e.
+with citations) statement explicitly defining why we can’t do the global scope HG approach.
+xxx **
 
-Remote hydraulic estimation is itself a burgeoning subfield within remote sensing of hydrology and an analogue to the problem presented here is the remote sensing of river discharge (RSQ- see @gleasonRemoteSensingRiver2020 for a thorough review).
+A potential alternative to this ‘global-scope’ HG approach is to directly estimate a river’s hydraulic properties (e.g.
+V) from remote sensing (RS) data.
+Remote sensing of river hydraulics is a burgeoning subfield within remote sensing of hydrology, typically in service of remote sensing of river discharge (RSQ- see @gleasonRemoteSensingRiver2020 for a thorough review).
 There are many ways to perform RSQ, from calibrating RS data to local channel hydraulics [e.g.
@@
 Many, but not all, of these ungauged approaches were developed in the context of the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission.
-**I don't know if introducing SWOT via RSQ is the best idea.
-Maybe I should just introduce SWOT on its own?**
+**CRAIG: might need to add unaguged citations here or rewrite perspective**
 
-SWOT will launch in 2022 and provide the world’s first global measurements of water surface extent and elevation at novel temporal resolutions.
-SWOT is a wide swath Ka-band radar interferometer and will sample rivers every 1 to 7 days per 21 day repeat cycle.
-It will measure rivers wider than 100m with a goal of expanding this to rivers at least 50m wide [@biancamariaSWOTMissionIts2016].
-In the context of RSQ, a decade of SWOT work has explored the multi-parameter problem of estimating roughness and bathymetry to produce the SWOT discharge product [e.g.
+SWOT is expected to launch in 2022 and provide the world’s first global measurements of water surface extent and elevation at novel temporal resolutions.
+SWOT is a wide swath Ka-band radar interferometer and will sample rivers every 1 to 7 days per 21 day repeat cycle that will measure rivers wider than 100m with a goal of expanding this to rivers at least 50m wide [@biancamariaSWOTMissionIts2016].
+A decade of SWOT work has explored the multi-parameter problem of estimating the river hydraulic parameters of roughness and bathymetry from remote sensing to produce the SWOT discharge product [e.g.
 @andreadisConstrainingAssimilationSWOT2020d; @brinkerhoffConstrainingRemoteRiver2020; @durandEstimatingReachaveragedDischarge2014a; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014a; @garamboisVariationalEstimationEffective2020; @hagemannBAMBayesianAMHGManning2017a; @larnierRiverDischargeBathymetry2020a; @oubanasDischargeEstimationUngauged2018].
-A specific subset of these ungauged RSQ methods are termed ‘McFLIs’ or ‘Mass Conserved Flow Law Inversion’ algorithms [e.g.
+A specific subset of these ungauged methods are termed ‘McFLIs’ or ‘Mass Conserved Flow Law Inversion’ algorithms [e.g.
 @andreadisConstrainingAssimilationSWOT2020d; @brinkerhoffConstrainingRemoteRiver2020; @hagemannBAMBayesianAMHGManning2017a; @durandEstimatingReachaveragedDischarge2014a].
-They use basic geomorphic theories, rather than full hydrologic and/or hydraulic models, and the concept of ‘prior river knowledge’ to estimate discharge from RS data where not all hydraulic terms are RS-able [@gleasonRemoteSensingRiver2020; @gleasonTrackingRiverFlows2017].
-While the raw performance of McFLIs is lower than the more sophisticated approaches mentioned above, they are globally flexible, easily implementable in any river that SWOT can sample, and improve our hydrological understandings of ungauged rivers when little to no information was previously available [@brinkerhoffConstrainingRemoteRiver2020; @durandIntercomparisonRemoteSensing2016b].
+These use basic geomorphic theories and the concept of ‘prior river knowledge’ to estimate discharge from RS data where not all hydraulic terms are RS-able [@gleasonRemoteSensingRiver2020; @gleasonTrackingRiverFlows2017].
+McFLIs are globally flexible, easily implementable in any river that SWOT can sample, and improve our hydrological understandings of ungauged rivers when little to no information was previously available [@brinkerhoffConstrainingRemoteRiver2020; @durandIntercomparisonRemoteSensing2016b].
 Recently, McFLIs have also shown promise in providing additional and beneficial information to traditional hydrologic modeling via data assimilation [@ishitsukaCombiningOpticalRemote2020].
 
-McFLIs often employ Bayesian inference.
+While all McFLIs to date have been developed in the context of RSQ, there is no reason their logic cannot be used to estimate $k_{600}$, particularly as McFLIs often employ Bayesian inference for equifinal inverse problems.
+Equifinality refers to an under-constrained mathematical system that have essentially infinite parameter combinations that can produce the same result: there are in essence more unknowns than equations [@garamboisInferenceEffectiveRiver2015], and this problem is common to both RSQ and $K_{600}$ estimation.
 Bayesian techniques have been previously used to estimate $k_{600}$ concurrently with stream metabolism from dissolved oxygen ($O_2$) datasets [@applingOvercomingEquifinalityLeveraging2018; @graceFastProcessingDiel2015; @holtgrieveSimultaneousQuantificationAquatic2010].
-While this approach requires detailed field data and is thus not applicable to $k_{600}$ estimation in ungauged rivers, it does suggest that Bayesian techniques could be useful for remotely sensing $k_{600}$ when parameter equifinality is a problem as it is with McFLIs [@andreadisConstrainingAssimilationSWOT2020d].
-Equinality issues refer to under-constrained systems that can have multiple solutions to the same discharge problem [@garamboisInferenceEffectiveRiver2015].
+While these studies are field based and are not applicable to $k_{600}$ estimation in ungauged rivers, they suggest that Bayesian techniques could be useful for remotely sensing $k_{600}$ when parameter equifinality is a problem, which matches McFLI logic precisely [@andreadisConstrainingAssimilationSWOT2020d].
 
-In this context, we hypothesize that SWOT data will improve our understanding of riverine GHG fluxes in ungauged basins and/or at drainage-network-scales we develop techniques to ingest SWOT's novel measurements of water surface width and height.
-Therefore, we sought to develop a method to estimate $k_{600}$ using only SWOT measurements and no in situ information.
+In this context, we hypothesize that a combination of SWOT data, a Bayesian McFLI paradigm, and gas exchange theory can successfully estimate $K_{600}$.
+This would in turn improve our understanding of riverine GHG fluxes in ungauged basins.
+Therefore, we developed a method to estimate $k_{600}$ solely from SWOT measurements, using hydrologic model data to produce SWOT-like data as SWOT has not yet launched [standard practice in the SWOT community- e,g, @durandIntercomparisonRemoteSensing2016b].
 We hypothesize that remote sensing $k_{600}$ is possible by coupling Bayesian remote sensing techniques (in a manner similar to McFLI) and SWOT data with hydraulic geometry and gas exchange theory.
@@
 
-- Is ungauged RS of the gas exchange velocity possible using soon-to-be-available SWOT river measurements?
-- How might this algorithm influence upscaled estimates of riverine carbon efflux to the atmosphere?
+- Is ungauged RS of the gas exchange velocity ($K_{600}$) possible using soon-to-be-available SWOT river measurements?
+- If possible, how will this RS method change estimates of riverine carbon efflux at network scales?
 
-To answer these questions, we develop a RS algorithm that ingests SWOT data and produces $k_{600}$ estimates and associated uncertainties that requires only water surface width and slope and therefore no in situ information about a river’s hydraulic properties.
-We validate the algorithm on simulated SWOT data for 22 rivers (as SWOT has yet to launch, simulated data is necessary- Section 2.3).
-We also quantify the sensitivity of our algorithm to the expected measurement errors implicit in SWOT measurements.
-Finally, we compare our algorithm against published methods for estimating carbon dioxide evasion ($FCO_2$) from rivers: we use previously published dissolved $CO_2$ data to compare the bulk carbon efflux (via $CO_2$ evasion) from the 22 rivers as calculated using both our new algorithm and published gauge-based techniques.
+To answer these questions, we developed a new McFLI algorithm that ingests SWOT data and produces $k_{600}$ estimates and its explicit Bayesian uncertainity that requires no in situ inputs of any kind (although in situ data can be ingested and will improve results).
+We name the RS of $k_{600}$ algorithm BIKER, or the ‘Bayesian Inversion of the $k_{600}$ Rate’.
+We validate BIKER on simulated SWOT data for 22 rivers.
+We also quantify BIKER's sensitivity to the expected SWOT measurement errors: while SWOT data represent a sea change in inland water monitoring, SWOT is expected to have an approximately 10cm error in water surface elevation [@biancamariaSWOTMissionIts2016].
+Finally, we use previously published dissolved $CO_2$ data to compare the bulk carbon efflux (via $CO_2$ evasion) from the 22 rivers as calculated using BIKER and previously published in situ techniques to show how BIKER’s $k_{600}$ estimates translate into carbon efflux.
 
 ##2 Methods
-We name the RS of $k_{600}$ algorithm BIKER, or the ‘Bayesian Inversion of the $k_{600}$ Rate’ algorithm.
-To build BIKER, two steps are taken: first, we develop a $k_{600}$ upscaling model that relies on no a priori knowledge of river hydraulics (section 2.1) and then we couple that upscaling model with remote sensing techniques to build BIKER (section 2.2).
-Following these methods, the validation setup is described (section 2.3), as is the workflow for comparing estimated bulk carbon effluxes from a suite of average flow velocity models (section 2.4).
-A flowchart detailing the entire study is included (Figure 1).
+To build BIKER, we first develop a mathematical $k_{600}$ model that relies on no a priori knowledge of river hydraulics (section 2.1) and then we set that model as the core of a McFLI (section 2.2).
+Following method descriptions, we describe the validation (section 2.3) and the workflow for comparing estimated bulk carbon effluxes from a suite of average flow velocity models (section 2.4).
+A flowchart detailing the entire study is given as Figure 1.
 
@@
 Flowchart of methodology used in this study.
-‘Extended Upscaling Model’ is detailed in section 2.1.
-‘BIKER’ is detailed in section 2.2 and ‘Flux Estimation’ is detailed in section 2.4.
+We begin by deriving a mathmatical model for k600 that uses only river width and slope to assign its parameters (section 2.1).
+Then, we implement this model within a McFLI ('BIKER') to estimate k600 solely from river width and water surface slope (section 2.2).
+Finally, we couple BIKER with in situ CO2 data to compare BIKER-estimated gas fluxes with established in situ methods (section 2.4).
 See section 2.3 for the validation setup.](outputs/flowchart.jpg)
 
-###2.1 Developing an RS-able $k_{600}$ upscaling model
-@raymondScalingGasTransfer2012a first used river hydraulics and the theoretical relationship between *eD* and gas exchange to upscale $k_{600}$ from average flow velocity and channel slope.
-This extended earlier work predicting the reaeration coefficient in a similar manner [e.g.
+###2.1 Developing a $k_{600}$ model for RS
+@raymondScalingGasTransfer2012a first used river hydraulics and the theoretical relationship between *eD* and gas exchange to estimate $k_{600}$ from average flow velocity and channel slope.
+This extended earlier work predicting the dissolved oxygen reaeration coefficient in a similar manner [e.g.
 @oconnorMechanismReaerationNatural1958; @tsivoglouTracerMeasurementReaeration1976].
-Later, @ulsethDistinctAirWater2019a built upon @raymondScalingGasTransfer2012a by expanding their dataset to include measurements from steeper rivers and finding that two distinct upscaling regimes exist in low and high energy rivers.
-More specifically, @ulsethDistinctAirWater2019a scale $k_{600}$ using *eD*, where the resulting upscaling model parameters are significantly different whether *eD* is high or low.
-Their upscaling function is reprinted as equation 2.
+Later, @ulsethDistinctAirWater2019a built upon @raymondScalingGasTransfer2012a by including measurements from steeper rivers and finding that two distinct $k_{600}$ regimes exist in low and high energy rivers.
+More specifically, @ulsethDistinctAirWater2019a scale $k_{600}$ using *eD*, where the resulting model parameters are significantly different whether *eD* is high or low.
+The general form of their $k_{600}$ function is reprinted as equation 2, with the two energy regimes represented by subscript *r*.
+To date, upscaling studies have assumed uniform flow ($S_0=S_h$) in order to train their parameters across large datasets that only have information on $S_0$.
 
-$$ln(k_{600})=3.10 + 0.35ln(eD) if eD < 0.02  \mathbf{(2)}$$
-$$ln(k_{600})=6.43+1.18ln(eD) if eD > 0.02$$
+$$k_{600}=a_r(eD)^{b_r}  \mathbf{(2)}$$
 
-While equation 2 would be ideal for the goal of this manuscript and provides best-to-date predictive accuracy for upscaling $k_{600}$, it requires an a priori knowledge of *eD* to assign its differential model parameters.
-We do not have this a priori knowledge in an ungauged, RS-oriented setting and only have access to SWOT-observed river widths and heights (and therefore water surface slopes).
-Thus, we need to extend the @ulsethDistinctAirWater2019a model to assign the two model parameters using only RS-able hydraulics.
-To do this, we use the basic premise that the $k_{600}\sim eD$ scaling relationship is different for different river sizes.
-We use river width to define river size and train different upscaling functions for different groups of data with similar river sizes [Figure 2, using the data from @ulsethDistinctAirWater2019a].
-In doing so, we find that narrow rivers (< 10m wide) need to be further broken into low energy (slope < 0.05) and high energy (slope > 0.05) regimes in order to meet the assumptions necessary to fit a linear regression model (specifically, normally distributed residuals of the predictions).
+While equation 2 as trained by @ulsethDistinctAirWater2019a dataset achieves the goal of this manuscript and provides a best-to-date predictive accuracy for $k_{600}$, it requires a priori knowledge of *eD* to assign its differential model parameters.
+Because $eD \approx gS_hV$ [@tsivoglouTracerMeasurementReaeration1976], we therefore need a priori knowledge of *V* to use equation 2 in a McFLI.
+So, we must rederive the @ulsethDistinctAirWater2019a $k_{600}$ model such that model parameters are solely assigned using the SWOT observables *W* and $S_h$.
 
-![Figure 2: Theoretical basis for the RS-able extension of the Ulseth et al. (2019) k600 upscaling model. Data from Ulseth et al. (2019).](outputs/k600/theory_plot.jpg)
+Using the @tsivoglouTracerMeasurementReaeration1976 equation for *eD* for unsteady flow conditions (channel slope $S_0 \neq S_h$- see text S1 for this model derivation and why assuming non-uniform flow is convenient for an algorithm ingesting SWOT data) in conjunction with Manning’s equation to relate average flow velocity to channel geometry, we rewrite equation 2 as equation 3.
+In the far-right form of equation 3, we use the Manning’s formulation from @hagemannBAMBayesianAMHGManning2017a which formulates depth via river width and area (in order to use SWOT-observables).
+$k_{600}$ is now defined as a function of $S_h$, channel roughness *n* $[T/L^{\frac{1}{3}}]$, channel width *W* [L], and channel area $A_0+dA [L^2]$ (discretized via median channel area $A_0$ and the change in channel area *dA*).
+Note that we estimate *dA* by assuming a rectangular river channel so that $dA \approx \delta H*W$ and is also solely a function of *W* and $S_h$.
+Finally, equation 3 is also defined by the parameters $a_i$ and $b_i$, which are described next.
 
-We combined these five regression models into a ‘rule-based regression model’ (equation 3) for river size *i* and model parameters *a* and *b*.
-Per Figure 2, there are five possible sets of *a* and *b* parameters, as defined for rivers 10-50m wide, 50-100m wide, 100+m wide, low energy rivers 0-10m wide (slope < 0.05), and high energy rivers 0-10m wide (slope > 0.05).
+**COLIN ABOUT DA: sensitivity analysis? "We investigated the sensitivity of BIKER to this rectangular assumption and found it negligible (xxx section x.x if appropriate xxx)".**
 
-$$k_{600}=a_i(eD)^{b_i} \mathbf{(3)}$$
+$$k_{600} = a_i(eD)^{b_i} \approx a_i(gSV)^{b_i} \approx a_i\left[gS_h^{\frac{3}{2}}\frac{1}{n}(A_0+dA)^\frac{2}{3}W^\frac{-2}{3}\right]^{b_i} \mathbf{(3)}$$
 
-This model therefore assigns $a_i$ and $b_i$ a priori using just SWOT measurements of water surface width and slope to define *i*.
+As will be explained in section 2.2, the only observations in the BIKER formulation of equation 3 are *W* and $S_h$.
+So, we must assign $a_i$ and $b_i$ using just *W* and $S_h$.
+To do this, we use the basic premise that the equation 2 relationship is different for different river sizes *i*.
+We use *W* and *S* to define *i* and train different $k_{600}$ functions for rivers with the same *i*.
+We fit this model to the @ulsethDistinctAirWater2019a data (Figure 2) for five possible sets of $a_i$ and $b_i$ parameters, as defined for rivers 10-50m wide, 50-100m wide, 100+m wide, low energy rivers 0-10m wide (*S* < 0.05), and high energy rivers 0-10m wide (slope > 0.05).
+We find that narrow rivers (< 10m wide) are best modeled by further breaking them down into low energy (*S* < 0.05) and high energy (*S* > 0.05) regimes in order to meet the assumptions necessary to fit a linear regression model as proposed by @raymondScalingGasTransfer2012a and @ulsethDistinctAirWater2019a (specifically, normally distributed residuals of the predictions).
 Another implication of this extension of the @ulsethDistinctAirWater2019a model is that we now use *a* and *b* parameters that implicitly reflect the distinct geomorphology of different rivers.
 
+![Figure 2: Theoretical basis for the RS-able extension of the Ulseth et al.
+(2019) k600 upscaling model.
+We use the differenial relationship between k600 and eD for different river sizes to define a 'rule-based' regression model.
+This model assigns differeniable model parameters using just river width and slope, following the form in equation 3.
+Data from Ulseth et al.
+(2019).](outputs/k600/theory_plot.jpg)
+
+**these two paragraphs need to make it clear before you get to the end that you believe the relationship to be linear, but you will not limit yourself to raymond and/or ulseth.
+Rather, you wish to find any number of linear relationships that best fit the k600 data as statistically and hydraulicly defined.
+THIS MIGHT NEED TO BE A NON-LINEAR FUNCTION.
+THE OTHER PROBLEM IS THAT ALL WE ARE JUST HIGHLIGHTING THAT ALL UPSCALING RESEARCH HAS BASICALLY ASSUMED UNIFORM FLOW....**
+
 Finally, we tested whether this model can reasonably estimate $k_{600}$ relative to the original @ulsethDistinctAirWater2019a model, with the goal of at least replicating their performance.
-We compared it against the various *eD*-based upscaling models from @raymondScalingGasTransfer2012a that are commonly used in the literature.
-This was done by training all upscaling functions on 80% of the @ulsethDistinctAirWater2019a data and comparing their performance on the remaining 20% of the data.
-Because the @raymondScalingGasTransfer2012a models were originally trained on a subset of this dataset, we deemed it fairer to refit their parameters to this expanded dataset (but keep model structure identical).
+We compared it against the various *eD*-based $k_{600}$ models from @raymondScalingGasTransfer2012a that are commonly used in the literature.
+This validation was done using a classic 80/20 split of the @ulsethDistinctAirWater2019a dataset, where 80% of the data was used to fit these models and 20% was used for independent testing of their predictive performance.
+This meant that the parameter values were slightly different than those published in both papers, but model structures were identical.
+This was done because the @raymondScalingGasTransfer2012a models were originally trained on a subset of this dataset and we deemed it fairer to compare model peformances using identical data.
 
-###2.2 Developing the BIKER algorithm
-With a validated upscaling model that requires no a priori knowledge of *eD* (results presented in section 3.1) we turn to developing the algorithm to remotely sense $k_{600}$.
-The approach used here is strongly informed by @hagemannBAMBayesianAMHGManning2017a’s algorithm for ungauged RSQ via Bayesian inference.
-Consult that paper for the logic behind conceptualizing the equation for discharge as a Bayesian remote sensing problem, which we largely follow here to conceptualize $k_{600}$ as a Bayesian remote sensing problem that can be solved using SWOT data.
+###2.2 Developing BIKER
+With a validated equation 3 in hand, we turn next to remotely sensing $k_{600}$.
+The approach used here is strongly informed by @hagemannBAMBayesianAMHGManning2017a’s McFLI algorithm for ungauged RSQ via Bayesian inference, further explored in more recent work [@brinkerhoffConstrainingRemoteRiver2020].
+These papers conceptualize discharge as a Bayesian remote sensing problem, which we largely follow here to conceptualize $k_{600}$ as a Bayesian remote sensing problem that can be solved using SWOT data.
+Below, we translate @hagemannBAMBayesianAMHGManning2017a’s ideas from discharge to $k_{600}$.
 
-SWOT will provide river width and height (and thus water surface slope $S_h$) at nodes/cross-sections along a set of river reaches.
-Thus, our ‘data’ for the Bayesian inference are SWOT measurements of river width and water surface slope while all other terms are treated as statistical parameters that cannot be directly observed from SWOT.
-We also obtain a ‘change in channel area’ *dA* from the river height and width and assuming a rectangular channel geometry.
+SWOT will measure river width and height (and thus water surface slope $S_h$) at approximately 200m intervals in pre-defined mass-conserved river reaches (**What is the PDD?**; SWORD).
+Following @hagemannBAMBayesianAMHGManning2017a, the observations needed for Bayesian inference are SWOT measurements of *W* and $S_h$, and all other terms in equation 3 (*n*, $A_0$, and $k_{600}$) are treated as statistical parameters that cannot be directly observed from SWOT.
 
-In order to use Bayesian inference to solve for $k_{600}$, we need a sampling model/likelihood function that estimates the unobservable statisitcal parameters.
-Following @tsivoglouTracerMeasurementReaeration1976, $eD \approx gS_hV$ for unsteady flow conditions (channel slope $S_0 \neq S_h$- see text S1 for this model derivation and why assuming non-uniform flow is convenient for an algorithm ingesting SWOT data).
-Using this, as well as Manning’s equation to relate average flow velocity to channel geometry, we rewrite equation 3 as equation 4.
-In equation 4, we use the Manning’s formulation from Hagemann et al.
-(2017) which formulates depth via river width and area (in order to use SWOT-observables).
-$k_{600}$ is now defined as a function of $S_h$, channel roughness *n* $[T/L^{\frac{1}{3}}]$, channel width *W* [L], and channel area $A_0+dA [L^2]$ (discretized via median channel area $A_0$ and the previously introduced *dA*).
-It is also defined by the upscaling parameters $a_i$ and $b_i$.
-
-$$k_{600}=a_i\left[gS_h^{\frac{3}{2}}\frac{1}{n}(A_0+dA)^\frac{2}{3}W^\frac{-2}{3}\right]^{b_i} \mathbf{(4)}$$
-
-Equation 4 is then reworked into a Bayesian sampling model to have all of the ‘data’ on the left-hand side sampled from all the unknown model parameters on the right-hand side (equation 5, where *x* is the SWOT observables and $\Theta$ are a set of statistical parameters).
-Note that in the formal likelihood specification both sides of equation 5 are ln-transformed to produce the normal sampling model presented here, however they are written in equation 5 without that transformation for succinctness.
-The $σ_{upscale}^2$ parameter refers to the uncertainty inherent in the upscaling model itself (equation 4).
+We therefore rework equation 3 into a Bayesian sampling model to have all of the ‘observations’ on the left-hand side sampled from all the unknown model parameters on the right-hand side (equation 4, where *x* are SWOT observables given $\Theta$, a set of statistical parameters).
+Note that in this formal likelihood specification both sides of equation 4 are log-transformed (natural log) to produce the normal sampling model presented here, however they are written in equation 4 without that transformation for succinctness.
+The $σ_{k600}^2$ parameter refers to the uncertainty inherent in equation 3's estimates.
 This will be explained in detail below.
 
-$$f(x|\Theta) = \left[a_i^{-1}(gS_h^\frac{-3}{2}W^\frac{2}{3})^{b_i}\right] \sim normal\left(\left[k_{600}^{-1}\frac{1}{n}(A_0+dA)^\frac{2}{3})^{b_i}\right], \sigma^2_{upscale}\right)  \mathbf{(5)}$$
+$$f(x|\Theta) = \left[a_i^{-1}(gS_h^\frac{-3}{2}W^\frac{2}{3})^{b_i}\right] \sim normal\left(\left[k_{600}^{-1}\frac{1}{n}(A_0+dA)^\frac{2}{3})^{b_i}\right], \sigma^2_{k600}\right)  \mathbf{(4)}$$
 
-Equation five necessitates that Bayesian priors be assigned to model parameters *n*, $A_0$, and $k_{600}$.
-Bayesian priors formularize the a priori estimates (and uncertainties) for the non-remotely-sensed parameters.
-Or more intuitively for these physcial quantities, they represent our ‘prior river knowledge’ of what *n*, $A_0$, and $k_{600}$ probably are for some river given that they cannot be remotely sensed.
-All priors are formalized within the model as truncated normal distributions of the ln-transformed terms such that $ln(X) \sim N(\mu, \sigma^2)$ for $\lambda < X < \epsilon$, using prior hyperparameters mean ($\mu$), standard deviation ($\sigma$), and upper ($\epsilon$) and lower bounds ($\lambda$) for parameter *X*.
+Equation four necessitates that Bayesian priors be assigned to model parameters *n*, $A_0$, and $k_{600}$.
+Bayesian priors formalize the a priori estimates (and uncertainties) for the non-remotely-sensed parameters.
+To state that more intuitively for these physical quantities, priors represent our ‘prior river knowledge’ of what *n*, $A_0$, and $k_{600}$ probably are for some river given that they cannot be remotely sensed.
+All priors are formalized within the model as truncated normal distributions of the log-transformed terms such that $log(X) \sim N(\mu, \sigma^2)$ for $\lambda < X < \epsilon$, using prior hyperparameters mean ($\mu$), standard deviation ($\sigma$), and upper ($\epsilon$) and lower bounds ($\lambda$) for parameter *X*.
 In order to avoid relying on in situ information, we assign prior hyperparameters using SWOT data only.
@@
 They developed a set of river channel prior hyperparameters for McFLI algorithms that are entirely RS-able and reflect differential channel hydraulics as a function of river geomorphology.
-With this method, they showed substantial improvement in the accuracy of their ungauged RSQ estimates with no changes to the algorithms or data used.
+They used an extensive database of field measurements and machine learning to identify patterns that associate the river width with river geomorphic properties, so these non-remotely sensible properties may be assigned to rivers given existing remotely sensed data (namely, river width from optical imagery).
+With this method, they showed substantial improvement in the accuracy of ungauged RSQ estimates with no changes to the algorithms or observations.
+This leaves only $k_{600}$ hyperparameters to be defined.
 We assigned the $k_{600}$ prior hyperparameters using a simple slope regression model trained on the @ulsethDistinctAirWater2019a dataset to assign $\mu$ (equation 6).
@@
 
-$$\mu_{k600}=3.22+0.347ln(S_h) if ln(S_h) < -4.634   \mathbf{(6)}$$
-$$\mu_{k600}=6.85+1.13ln(S_h) if ln(S_h)>-4.634$$
+$$\mu_{k600}=3.22+0.347ln(S_h) if ln(S_h) < -4.634   \mathbf{(6a)}$$
+$$\mu_{k600}=6.85+1.13ln(S_h) if ln(S_h)>-4.634 \mathbf{(6b)}$$
 
-Finally, we estimate $σ_{upscale}^2$ using Monte Carlo (MC) methods to approximate total uncertainty from equation 4.
-Specifically, we explore three uncertainty scenarios to assess the realtive influences of different sources of uncertainity.
-Uncertainty in equation 4 can stem from 1) error baked into the upscaling model ($a_i$ and $b_i$ parameter uncertainty), 2) error in Manning’s approximation of the average flow velocity, and 3) measurement error.
-Our interest is using BIKER for satellite remote sensing and so we consider measurement errors from SWOT-derived water surface slopes.
-To do this, we run 15,000 unique MC simulations on 5,000 sets of field measurements of river channel hydraulics from @brinkerhoffReconcilingAtaStationAtManyStations2019 under three different uncertainty scenarios.
-Each MC simulation is itself 10,000 runs, sampling from the normal distributions for $a_i$ and $b_i$ (uncertainty scenario 1), $a_i$, $b_i$, and ln-transformed velocity (uncertainty scenario 2), and $a_i$, $b_i$, ln-transformed velocity, and water surface slope (uncertainty scenario 3).
-Estimate uncertainty for velocity followed @hagemannBAMBayesianAMHGManning2017a, who estimated ln-transformed Manning’s equation error to be 0.25.
-The slope-error model that we used is explained in detail in Section 2.3 and relies on a recent analysis of SWOT water surface slope measurement errors [@durandHowWillRadar2020a].
-This results in three sets of 5,000 discrete distributions for $k_{600}$ from which the median uncertainty term for any scenario is extracted and used for $σ_{upscale}^2$.
-**a table might help here to layout the three tests**
+Finally, we estimate $σ_{k600}^2$ using Monte Carlo (MC) methods to approximate total uncertainty from equation 3.
+Uncertainty in equation 3 can stem from 1) $a_i$ and $b_i$ parameter uncertainty, 2) error in Manning’s approximation of the average flow velocity, and 3) measurement error.
+Total equation 3 uncertainty is a function of all three uncertainity sources, so we need to propogate the three uncertanties through the algebra of equation 3.
+Here, we use MC simulations to do this.
+MC simulations repeately sample from parameter distributions defined by their individual expectations and uncertainities to produce a distribution of model estimates from which an overall uncertainity term can be extracted.
+Here, we run 5,000 unique MC simulations on 5,000 sets of field measurements of river channel hydraulics from @brinkerhoffReconcilingAtaStationAtManyStations2019 and use the median uncertainity term across those 5,000 distributions as $\sigma_{k600}$.
+Each MC simulation is itself 10,000 runs, sampling from the normal distributions for $a_i$, $b_i$, the log-transformed flow velocity, and $S_h$.
+Uncertainty for velocity followed @hagemannBAMBayesianAMHGManning2017a, who estimated ln-transformed Manning’s equation error in discharge to be 0.25 (we assume that the 0.25 value is due soley from Manning's equation and not measurements of cross-sectional area in order to use here as the velocity error).
+The slope-error model that we used is explained in detail in Section 2.3 and relies on a recent analysis of expected errors in SWOT $S_h$ measurements [@durandHowWillRadar2020a].
 
-With the sampling model (equation 5), prior distributions, and the $σ_{upscale}^2$ parameter described, a joint posterior distribution conditional on the SWOT data is obtained.
-To approximate this distribution, BIKER uses a Markov Chain Monte Carlo (MCMC) algorithm.
-Because it is written in the Stan probabilistic programming language, BIKER’s MCMC of choice is a Hamiltonian Monte Carlo which reduces computation time relative to other sampling algorithms (https://mc-stan.org/docs/2_26/reference-manual/hamiltonian-monte-carlo.html).
+Finally, we repeat the above workflow three times under three different uncertainty scenarios to assess the overall influence of each uncertainity source on the equation 3 estimates.
+Uncertainty scenario 1 propogates errors in $a_i$ and $b_i$ only, uncertainty scenario 2 additionally propogates errors in log-transformed velocity, and uncertainty scenario 3 additionally propogates errors in $S_h$.
+
+With the sampling model (equation 4), prior distributions, and the $σ_{k600}^2$ parameter described, a joint posterior distribution conditional on the SWOT data is obtained.
+To approximate this distribution, we use a Markov Chain Monte Carlo (MCMC) algorithm.
+Because it is written in the Stan probabilistic programming language, we use a Hamiltonian Monte Carlo which reduces computation time relative to other sampling algorithms [again following @hagemannBAMBayesianAMHGManning2017a].
 

### 2021-05-05 docs/supp_v1.Rmd
complete overhaul of repo

--- 2021-02-18 supp_v1.Rmd
@@
 ##Text S1
-[@tsivoglouTracerMeasurementReaeration1976] outlined an energy dissipation model for the reaeration coefficient ($K_2$, or k without depth correction) that is frequently generalized and used to scale $k_{600}$ [@ulsethDistinctAirWater2019a; @raymondScalingGasTransfer2012a].
-They showed that $K_2$ scales linearly with the energy dissipation rate eD.
-Their derivation explicitly shows that $eD=\frac{dH}{dT}$ for some segment of stream where H is water surface elevation and T is hydraulic residence time.
-For a stream segment with residence time T, the loss in water surface elevation is synonymous with the downward component of flow velocity V in the h dimension over the reach length (i.e.
-$S_h$).
+@tsivoglouTracerMeasurementReaeration1976 outlined an energy dissipation model for the reaeration coefficient ($K_2$, or *k* without depth correction) that is frequently generalized and used to scale $k_{600}$ [@ulsethDistinctAirWater2019a; @raymondScalingGasTransfer2012a].
+They showed that $K_2$ scales linearly with the energy dissipation rate *eD*.
+Their derivation explicitly shows that $eD=\frac{dH}{dT}$ for some segment of stream where *H* is water surface height and *dT* is hydraulic residence time.
+For a stream segment with some residence time *T*, the loss in water surface height is synonymous with the downward component of flow velocity *V* in the *H* dimension over the reach length (i.e.
+water surface slope $S_h$).
 So, we can generalize this to $\frac{dH}{dT}=V_hS_h≈VS_h$.
@@
 This model is convenient for our use because SWOT will only measure $S_h$ and not channel slope $S_0$.
-Further, Manning’s equation under non-uniform flow conditions requires $S_h$ and not $S_0$ and so we can conveniently merge the slope terms across equations and directly measure them from SWOT.
+Further, Manning’s equation under non-uniform flow conditions requires $S_h$ and not $S_0$ and so we can conveniently merge the various slope terms across equations to yield equation 4 in the main text and then directly measure $S_h$ from SWOT.
 
 ##Text S2
-In Figure S4 we show that bed roughness only exerts an influence on $k_{600}$ when rivers are incredibly steep.
-We do this using the 'Keulegan effective roughness height' for the channel bed, which is detailed in equation S1 below [in the form from @dingmanAnalyticalDerivationAtastation2007b].
-This model was first outlined in the 1930s by Keulegan and outlined explicitly in [@fergusonHydraulicsHydraulicGeometry1986a] and [@dingmanAnalyticalDerivationAtastation2007b] in the context of hydraulic geometry (noting that it is a more generalized form of the Manning's and Chezy's velocity~depth relations commonly used in the literature).
+In Figure S2 we show that bed roughness only exerts an influence on $k_{600}$ when rivers are incredibly steep.
+We do this using the 'Keulegan effective roughness height' for the channel bed, which is detailed in equation S1 below [in the form as written by @dingmanAnalyticalDerivationAtastation2007b].
+This model was first outlined in the 1930s by Keulegan and outlined explicitly in @fergusonHydraulicsHydraulicGeometry1986a and @dingmanAnalyticalDerivationAtastation2007b in the context of hydraulic geometry (noting that it is a more generalized form of the Manning's and Chezy's velocity~depth relations commonly used in the literature).
 The model arises from depth integrating the Prandl-von Karman equation for velocity in a turbulent boundary layer [@fergusonHydraulicsHydraulicGeometry1986a].
@@
 $$V=2.5(gDS)^{1/2}ln(\frac{11D}{y_r})   (S1)$$
-$y_r [L]$ is the effective bed roughness height we solved for using the data from [@ulsethDistinctAirWater2019a] where D is channel depth [L], S is channel slope, g is gravitational acceleration $[M/T^2]$, and V is average flow velocity $[M/T]$.
-Because it is an 'effective/equivalent height', this does not necessairly reflect actual roughness elements in the river channel but is a reasonable way to quantify relative differences in bed roughness across data.
+$y_r [L]$ is the effective bed roughness height we solved for using the data from @ulsethDistinctAirWater2019a where *D* is channel depth $[L]$, *S* is channel slope, *g* is gravitational acceleration $[M/T^2]$, and *V* is average flow velocity $[M/T]$.
+Because $y_r$ is an 'effective/equivalent height', this does not necessairly reflect the actual size of roughness elements in the river channel but is a reasonable way to quantify relative differences in bed roughness across data.
 It is convenient to use here because it provides an actual roughness height and is not simply a coefficient like the conductance coefficients in Manning's and Chezy's velocity~depth relations.
 
-In Figure S4, this effective roughness height was plotted against $k_{600}$ and the data was progressively filtered for steeper and steeper rivers.
+In Figure S2, this effective roughness height was plotted against $k_{600}$ and the data was progressively filtered for steeper and steeper river reaches.
 
@@
 Sampling took place 2008-2009 in the Ohio River (upstream of Cincinnati, OH).
-Each point here was joined to the 11-day SWOT observations used in this study (section 5.2.4).](outputs/FCO2/Beaulieu_timeseries.jpg)
+Each point here was joined to the 11-day SWOT observations used in this study (section 2.4).](outputs/FCO2/Beaulieu_timeseries.jpg)
 
@@
 Scaling $k_{600}$ using an effective bed roughness term (see Text S2).
-This scaling relationship only exists for rivers with very steep slopes, here presented as greater than 0.1](outputs/k600/roughness.jpg)
+This scaling relationship only exists for rivers with very steep slopes, here defined as greater than 0.1](outputs/k600/roughness.jpg)
 
-*Table S1: Details on the 3 velocity rating curve models used to estimate bulk carbon efflux from the SWOT rivers (section 5.2.4).*
+*Table S1: Details on the 3 velocity hydraulic geometry models used to estimate FCO2 from the SWOT rivers (section 2.4).*
 

### 2021-05-05 generateMS.R

@@ -1,0 +1,20 @@
+#################
+#This will generate and store all results in your current RStudio environment so that you can knit the manuscript correctly, and then it'' knit the manuscript too.
+#################
+
+#Move all results to global environment
+results_3_1 <- read.csv('cache//k600//results.csv')
+results_3_1_MC <- read.csv('cache//MonteCarlo//results.csv')
+results_3_2_rivs <- read.csv('cache//validation//results_by_riv.csv')
+results_3_2_all <- read.csv('cache//validation//results_all_riv.csv')
+results_3_3_all <- read.csv('cache//FCO2//fco2_stats_all.csv')
+results_3_3_rivs <- read.csv('cache//FCO2//fco2_stats_by_river.csv')
+results_3_3_bulk <- read.csv('cache//FCO2//bulkFluxes.csv')
+results_3_3_bulk$key <- as.character(results_3_3_bulk$key)
+
+#knit manuscript
+rmarkdown::render("docs//manuscript_v2.Rmd")
+print('Manuscript generated.')
+
+rmarkdown::render("docs//supp_v1.Rmd")
+print('Supplemental Information generated.')

### 2021-05-05 run.sh

@@ -1,6 +1,14 @@
-Rscript scripts/roughness_k600.R
-#Rscript scripts/k600_analysis.R
-#Rscript scripts/MonteCarlo_analysis.R
-#Rscript scripts/runBIKER.R
-#Rscript scripts/validation.R
-#Rscript scripts/FCO2_analysis.R
+#!/bin/bash
+#SBATCH -c 8  # Number of Cores per Task
+#SBATCH -p cee_water_cjgleason
+#SBATCH --mem=25000  # Requested Memory
+#SBATCH -t 500:00:00  # Job time limit
+#SBATCH -o out.txt  # %j = job ID
+#SBATCH -e err.txt
+
+source ~/.bashrc
+
+module load R
+module load gcc/9.3.0
+
+Rscript src/main.R

### 2021-05-05 scripts/FCO2_analysis.R

@@ -1,393 +1,0 @@
-###########################
-#Description: Uses BIKER and rating curves to calculate FCO2 and bulk C fluxes from the SWOT simulated rivers
-#Craig Brinkerhoff
-#Winter 2021
-##########################
-
-#load model settings and packages-------------------------------
-source(here::here('scripts' , 'inputs.R'))
-
-#legend labels------------------
-legend_labels <- c('BIKER',
-                   'Model Using Observed velocity')
-
-#Functions----------------------------------------
-Sc_co2_func <- function(t){1742 + (-91.24*t) + (2.208*t^2) + (-0.0219*t^3)} #Raymond etal 2012
-
-henrys_law_func <- function(t) {-60.2409+93.4517*(100/(t+273.15))+23.3585*log((t+273.15)/100)} #Weiss 1974
-
-calcdA_mat <- function(w, h) {
-  stopifnot(all(dim(w) == dim(h)))
-  dA <- w
-  for (i in 1:nrow(dA)) {
-    dA[i, ] <- calcdA_vec(w[i, ], h[i, ])
-  }
-
-  dA
-}
-
-calcdA_vec <- function(w, h) {
-  words <- order(w)
-  warr <- w[words]
-  harr <- h[words]
-  delh <- c(0, diff(harr))
-  delA <- cumsum(warr * delh)
-  dA <- 1:length(w)
-  dA[words] <- delA
-  dA
-}
-
-k600_craig <- function(w, s, v) {
-  widthRegime <- ifelse(mean(w, na.rm = T) < 10 & mean(s) < 0.05, 1,
-                        ifelse(mean(w, na.rm = T) < 10 & mean(s, na.rm=T) >= 0.05, 5,
-                               ifelse(mean(w, na.rm = T) < 50, 2,
-                                      ifelse(mean(w, na.rm = T)< 100, 3, 4))))
-  if (widthRegime == 1){
-    k <- (111.58121 * (g*s * (v))^0.6488131)
-  }
-  if(widthRegime ==2 ){
-    k <- (109.04977 * (g*s * (v))^0.6624354)
-  }
-  if(widthRegime == 3){
-    k <- (31.84344 * (g*s * (v))^0.4366114)
-  }
-  if(widthRegime == 4) {
[…]

### 2021-05-05 scripts/MonteCarlo_analysis.R

@@ -1,247 +1,0 @@
-#Monte Carlo analysis for quantifying uncertainty of k600 model and BIKER's assumptions
-#Creator: Craig Brinkerhoff
-#Summer 2020
-
-#load model settings and packages-------------------------------
-source(here::here('scripts' , 'inputs.R'))
-set.seed(455)
-
-#Pull in hydraulic measurements from Brinkerhoff etal 2019 to test parameter uncertainty------------------------------------------
-if(munge == 1){
-  hydraulics_init <- read.csv('inputs/MonteCarlo/field_measurements.csv')
-  nhd_med <- read.csv('inputs//MonteCarlo//NHD_join_table.csv')
-
-  #add reach slopes from NHD
-  hydraulics <- left_join(hydraulics_init, select(nhd_med, SLOPE, SOURCE_FEA, LatSite, LonSite), by = c("site_no" = "SOURCE_FEA"))
-
-  #intial cleaning of field measurements
-  hydraulics <- filter(hydraulics, is.finite(chan_width) ==1) %>%
-    filter(is.finite(chan_velocity)==1) %>%
-    filter(is.finite(chan_discharge)==1) %>%
-    filter(is.finite(chan_area)==1) %>%
-    filter(chan_width > 0) %>%
-    filter(chan_velocity > 0) %>%
-    filter(chan_discharge > 0) %>%
-    filter(chan_area > 0) %>%
-    filter(measured_rating_diff != 'Poor') %>%
-    filter(measured_rating_diff != 'POOR') %>%
-    filter(is.finite(SLOPE)==1)
-
-    #imperial to metric
-    hydraulics$area <- hydraulics$chan_area * 0.092903 #ft2 to m2
-    hydraulics$width <- hydraulics$chan_width*0.305 #m
-    hydraulics$velocity <- hydraulics$chan_velocity*0.305 #m/s
-    hydraulics$slope <- hydraulics$SLOPE
-    hydraulics$depth <- hydraulics$area / hydraulics$width #m
-    hydraulics$n <- (hydraulics$depth^(2/3)*hydraulics$SLOPE^(1/2))/hydraulics$velocity
-
-    #many slopes were saved with floor values of 1e-5, so I just remove these because they influence the k600 model to provide artificial results
[…]

### 2021-05-05 scripts/runBIKER.R

@@ -1,170 +1,0 @@
-######################
-#Description: Runs the BIKER algorithm on SWOT-simulated rivers
-#Creator: Craig Brinkerhoff
-#Date: Winter 2021
-#####################
-
-#load model settings and packages-------------------------------
-source(here::here('scripts' , 'inputs.R'))
-
-#functions to calculate dA from W and H------------------------------------------------------
-#' @param w Matrix of widths
-#' @param h Matrix of heights(FROM MARK)
-calcdA_mat <- function(w, h) {
-  stopifnot(all(dim(w) == dim(h)))
-  dA <- w
-  for (i in 1:nrow(dA)) {
-    dA[i, ] <- calcdA_vec(w[i, ], h[i, ])
-  }
-
-  dA
-}
-
-#' Calculate partial cross-section area from width and height vectors (time series)
-#' @param w vector of widths
-#' @param h vector of heights(FROM MARK)
-calcdA_vec <- function(w, h) {
-  words <- order(w)
-  warr <- w[words]
-  harr <- h[words]
-  delh <- c(0, diff(harr))
-  delA <- cumsum(warr * delh)
-  dA <- 1:length(w)
-  dA[words] <- delA
-  dA
-}
-
-#load in SWOT observables-------------------------------------------
-#Frasson etal 2019 & Rodriquez etal 2020
-files <- list.files('inputs/Frasson_etal_2019', pattern="*.nc", full.names = TRUE)
-files <- files[-1] #remove Arial Khan
-
-#function for parallelizing BIKER runs------------------------------
-run_BIKER <- function(currPepsi, errFlag) {
-  #run model-------------------------------------------------------
-  name <- substr(currPepsi, 26, nchar(currPepsi))
-  name <- substr(name,1,nchar(name)-3)
-  set.seed(12)
-  data_in =nc_open(currPepsi)
-
-  W_obs=ncvar_get(data_in,'Reach_Timeseries/W')
-  H_obs=ncvar_get(data_in,'Reach_Timeseries/H')
-  S_obs = ncvar_get(data_in, 'Reach_Timeseries/S')
-  area = ncvar_get(data_in, 'Reach_Timeseries/A')
-  Q_obs = ncvar_get(data_in, 'Reach_Timeseries/Q')
-
[…]

### 2021-05-05 scripts/scraps.R

@@ -1,241 +1,0 @@
-#scrap-------------------------------------------------
-
-#select measurements that have ~similar slopes and consider them all at the 'same xs'
-swing <- 0.10 #percent
-data$widthRegime <- ifelse(data$slope >= 0.0004 & data$slope <= 0.0004+(0.0004*swing), '1',
-                              ifelse(data$slope >= 0.001 & data$slope <= 0.001+(0.001*swing), '2',
-                                ifelse(data$slope >= 0.003 & data$slope <= 0.003+(0.003*swing), '3',
-                                  ifelse(data$slope >= 0.005 & data$slope <= 0.005+(0.005*swing), '4',
-                                    ifelse(data$slope >= 0.008 & data$slope <= 0.008+(0.008*swing), '5',
-                                      ifelse(data$slope >= 0.01 & data$slope <= 0.01+(0.01*swing), '6',
-                                        ifelse(data$slope >= 0.01 & data$slope <= 0.01+(0.01*swing), '7',
-                                          ifelse(data$slope >= 0.03 & data$slope <= 0.03+(0.03*swing), '8',
-                                            ifelse(data$slope >= 0.05 & data$slope <= 0.05+(0.05*swing), '9',
-                                              ifelse(data$slope >= 0.1 & data$slope <= 0.1+(0.1*swing), '10',
-                                                ifelse(data$slope >= 0.13 & data$slope <= 0.13+(0.13*swing), '11',NA)))))))))))
-
-
-                                                #amhg test...-----------------
-                                                r2_amhg <- round(summary(lm(log10(k_model$a)~k_model$b))$r.squared,2)
-                                                amhg_plot <- ggplot(k_model, aes(x=a, y=b, color=widthRegime)) +
-                                                  geom_point(size=8)+
-                                                  geom_smooth(method='lm', se=F, color='black') +
-                                                  xlab('Rating Curve Coefficient') +
-                                                  ylab('Rating Curve exponent')+
-                                                  scale_color_discrete_qualitative(palette='Dark2')+
-                                                  geom_richtext(aes(x=10^1.3, y=1), label=paste0('r<sup>2</sup>: ', r2_amhg), color='black') +
-                                                  scale_x_log10(
[…]

### 2021-05-05 scripts/validation.R

@@ -1,150 +1,0 @@
-########################
-#Description: Produce figures for BIKER runs on SWOT-simulated rivers
-#Creator: Craig Brinkerhoff
-#Date: Fall 2020
-######################
-
-#load model settings and packages-------------------------------
-source(here::here('scripts' , 'inputs.R'))
-
-# Sum of squares function------------------------------
-sumsq <- function(x) sum(x^2)
-
-#read in results----------------------------------
-results <- read.csv('outputs//validation//BIKER_validation_results.csv')
-full_output <- filter(results, errFlag == 0) #remove results with SWOT measurement error
-
-lm_kfit <- lm(log10(full_output$kest_mean)~log10(full_output$kobs))
-r2 <- round(summary(lm_kfit)$r.squared, 2)
-full_output <- drop_na(full_output)
-rmse <- round((Metrics::rmse(log10(full_output$kest_mean), log10(full_output$kobs))), 2)
-
-predInts <- predict(lm_kfit, interval='prediction')
-full_output <- cbind(full_output, predInts)
-
-#Model Validation across all rivers and timesteps--------------------------------------------------------
-valPlot <- ggplot(full_output, aes(x=(kobs), y=(kest_mean))) +
-  geom_abline(size=2, linetype='dashed', color='black') +
-  geom_pointrange(aes(ymin = kest_low, ymax = kest_high), fatten=10, fill='#1b9e77', pch=21, color='black') +
-  geom_smooth(size=2, color='darkgrey', method='lm', se=F)+
-  geom_line(aes(y=10^(lwr)), color='darkgrey', linetype='dashed', size=1.75) +
-  geom_line(aes(y=10^(upr)), color='darkgrey', linetype='dashed', size=1.75) +
-  xlab('k600 via observed velocity [m/dy]') +
-  ylab('BIKER k600 [m/dy]') +
-  scale_y_log10(
-      breaks = scales::trans_breaks("log10", function(x) 10^x),
-      labels = scales::trans_format("log10", scales::math_format(10^.x)))+
-  scale_x_log10(
-      breaks = scales::trans_breaks("log10", function(x) 10^x),
-      labels = scales::trans_format("log10", scales::math_format(10^.x)))+
[…]

### 2021-05-05 src/MonteCarlo_analysis.R

@@ -1,0 +1,247 @@
+#Monte Carlo analysis for quantifying uncertainty of k600 model and BIKER's assumptions
+#Creator: Craig Brinkerhoff
+#Summer 2020
+
+print('staring MC analysis...')
+
+set.seed(455)
+
+#Pull in hydraulic measurements from Brinkerhoff etal 2019 to test parameter uncertainty------------------------------------------
+if(munge == 1){
+  hydraulics_init <- read.csv('cache/MonteCarlo/field_measurements.csv')
+  nhd_med <- read.csv('cache/MonteCarlo/NHD_join_table.csv')
+
+  #add reach slopes from NHD
+  hydraulics <- left_join(hydraulics_init, select(nhd_med, SLOPE, SOURCE_FEA, LatSite, LonSite), by = c("site_no" = "SOURCE_FEA"))
+
+  #intial cleaning of field measurements
+  hydraulics <- filter(hydraulics, is.finite(chan_width) ==1) %>%
+    filter(is.finite(chan_velocity)==1) %>%
+    filter(is.finite(chan_discharge)==1) %>%
+    filter(is.finite(chan_area)==1) %>%
+    filter(chan_width > 0) %>%
+    filter(chan_velocity > 0) %>%
+    filter(chan_discharge > 0) %>%
+    filter(chan_area > 0) %>%
+    filter(measured_rating_diff != 'Poor') %>%
+    filter(measured_rating_diff != 'POOR') %>%
+    filter(is.finite(SLOPE)==1)
+
+    #imperial to metric
+    hydraulics$area <- hydraulics$chan_area * 0.092903 #ft2 to m2
+    hydraulics$width <- hydraulics$chan_width*0.305 #m
+    hydraulics$velocity <- hydraulics$chan_velocity*0.305 #m/s
+    hydraulics$slope <- hydraulics$SLOPE
+    hydraulics$depth <- hydraulics$area / hydraulics$width #m
+    hydraulics$n <- (hydraulics$depth^(2/3)*hydraulics$SLOPE^(1/2))/hydraulics$velocity
+
+    #many slopes were saved with floor values of 1e-5, so I just remove these because they influence the k600 model to provide artificial results
+    hydraulics <- filter(hydraulics, slope > 1e-5)
+
[…]

### 2021-05-05 src/main.R

@@ -1,0 +1,18 @@
+###############
+#Set up project
+###############
+.libPaths('/nas/cee-water/cjgleason/r-lib')
+library(ProjectTemplate, lib.loc = "/nas/cee-water/cjgleason/r-lib/")
+load.project()
+
+theme_set(theme_cowplot())
+
+#Run analysis-------------------------------
+#source(here::here('src' , 'roughness_k600.R'))
+#source(here::here('src' , 'MonteCarlo_analysis.R'))
+source(here::here('src' , 'runBIKER.R'))
+source(here::here('src' , 'validation.R'))
+#source(here::here('src' , 'FCO2_analysis.R'))
+
+
+#OUT OF DATE I THINK source(here::here('src' , 'k600_analysis.R'))

### 2021-05-05 src/runBIKER.R

@@ -1,0 +1,147 @@
+######################
+#Description: Runs the BIKER algorithm on SWOT-simulated rivers
+#Creator: Craig Brinkerhoff
+#Date: Winter 2021
+#####################
+
+print('running BIKER...')
+
+#load in SWOT observables-------------------------------------------
+#Frasson etal 2019 & Rodriquez etal 2020
+files <- list.files('data/Frasson_etal_2019', pattern="*.nc", full.names = TRUE)
+files <- files[-1] #remove Arial Khan
+
+#function for parallelizing BIKER runs------------------------------
+run_BIKER <- function(currPepsi, errFlag, kmodel) {
+  #run model-------------------------------------------------------
+  name <- substr(currPepsi, 24, nchar(currPepsi))
+  name <- substr(name,1,nchar(name)-3)
+  set.seed(12)
+  data_in =nc_open(currPepsi)
+
+  W_obs=ncvar_get(data_in,'Reach_Timeseries/W')
+  H_obs=ncvar_get(data_in,'Reach_Timeseries/H')
+  S_obs = ncvar_get(data_in, 'Reach_Timeseries/S')
+  area = ncvar_get(data_in, 'Reach_Timeseries/A')
+  Q_obs = ncvar_get(data_in, 'Reach_Timeseries/Q')
+
+  #prep pepsi rivers------------------------------------------------
+  S_obs[S_obs<=0]=NA
+  S_obs[is.na(S_obs)] = 0.000017 #min obs SWOT slope Biancarma etal 2016
+  W_obs[W_obs<0]=NA
+  H_obs[H_obs<0]=NA
+  area[area<0]=NA
+  Q_obs[Q_obs<0]=NA
+
+  #Sample every 11 days (SWOT-style)------------------------
+  W_obs = W_obs[,seq(1, ncol(W_obs), 11)]
+  S_obs = S_obs[,seq(1, ncol(S_obs), 11)]
+  H_obs = H_obs[,seq(1, ncol(H_obs), 11)]
+  Q_obs = Q_obs[,seq(1, ncol(Q_obs), 11)]
+  area = area[,seq(1, ncol(area), 11)]
+
+  #Some NA handling in slopes
+  if (any(apply(S_obs,2,sum,na.rm=TRUE) ==0)){
+    remove_index =  which((apply(S_obs,2,sum,na.rm=TRUE) ==0) ==TRUE)
+
+    W_obs=W_obs[,-remove_index]
+    H_obs=H_obs[,-remove_index]
[…]

### 2021-05-18 manuscript_v3.Rmd
first rewrite of methods using ustar model

--- 2021-05-05 docs/manuscript_v2.Rmd
@@
 ##2 Methods
-To build BIKER, we first develop a mathematical $k_{600}$ model that relies on no a priori knowledge of river hydraulics (section 2.1) and then we set that model as the core of a McFLI (section 2.2).
-Following method descriptions, we describe the validation (section 2.3) and the workflow for comparing estimated bulk carbon effluxes from a suite of average flow velocity models (section 2.4).
+To build BIKER, we join a process-based model for $k_{o2}$ with a McFLI framework for inverting SWOT measurements (section 2.1).
+Following the description of that process, we describe the validation (section 2.3) and the workflow for comparing estimated bulk carbon effluxes from a suite of hydraulic geometry models (section 2.4).
 A flowchart detailing the entire study is given as Figure 1.
@@
 Flowchart of methodology used in this study.
-We begin by deriving a mathmatical model for k600 that uses only river width and slope to assign its parameters (section 2.1).
-Then, we implement this model within a McFLI ('BIKER') to estimate k600 solely from river width and water surface slope (section 2.2).
+We begin by deriving a process-based model for ko2 that uses only river width and slope to assign its parameters (section 2.1).
+Then, we implement this model within a McFLI ('BIKER') to estimate ko2 solely from river width and water surface slope (section 2.2).
 Finally, we couple BIKER with in situ CO2 data to compare BIKER-estimated gas fluxes with established in situ methods (section 2.4).
-See section 2.3 for the validation setup.](outputs/flowchart.jpg)
+See section 2.3 for the validation setup.](cache/flowchart.jpg)
 
-###2.1 Developing a $k_{600}$ model for RS
-@raymondScalingGasTransfer2012a first used river hydraulics and the theoretical relationship between *eD* and gas exchange to estimate $k_{600}$ from average flow velocity and channel slope.
-This extended earlier work predicting the dissolved oxygen reaeration coefficient in a similar manner [e.g.
-@oconnorMechanismReaerationNatural1958; @tsivoglouTracerMeasurementReaeration1976].
-Later, @ulsethDistinctAirWater2019a built upon @raymondScalingGasTransfer2012a by including measurements from steeper rivers and finding that two distinct $k_{600}$ regimes exist in low and high energy rivers.
-More specifically, @ulsethDistinctAirWater2019a scale $k_{600}$ using *eD*, where the resulting model parameters are significantly different whether *eD* is high or low.
-The general form of their $k_{600}$ function is reprinted as equation 2, with the two energy regimes represented by subscript *r*.
-To date, upscaling studies have assumed uniform flow ($S_0=S_h$) in order to train their parameters across large datasets that only have information on $S_0$.
+###2.1 Developing BIKER
+To use the McFLI framework, a physical model for the parameter of interest is necessary.
+In this case we need a physical model for *k*.
+First, we detail which model for *k* we chose to implement, and then we explain how the implementation works.
 
-$$k_{600}=a_r(eD)^{b_r}  \mathbf{(2)}$$
+There is a long tradition of water quality engineers and biogeochemists developing predictive models for different forms of *k*, going back to at least the 1950s [@wangPhysicallyBasedScaling2021].
+The goals of these models vary, from predicting oxygen reaeration coefficients to predicting $CO_2$ gas transfer velocities.
+Most of these models operate under the basic assumption that river channel hydraulics govern gas exchange and so simple statisical relationships can be generated between forms of *k* and hydraulic properties.
+However, many of them are not rooted in process-based mass transport theory and rather are simply empirical relations.
+This means that many of these models exist at the mercy of the limited training data available to fit said models and are highly sensitive to which data are used.
 
-While equation 2 as trained by @ulsethDistinctAirWater2019a dataset achieves the goal of this manuscript and provides a best-to-date predictive accuracy for $k_{600}$, it requires a priori knowledge of *eD* to assign its differential model parameters.
-Because $eD \approx gS_hV$ [@tsivoglouTracerMeasurementReaeration1976], we therefore need a priori knowledge of *V* to use equation 2 in a McFLI.
-So, we must rederive the @ulsethDistinctAirWater2019a $k_{600}$ model such that model parameters are solely assigned using the SWOT observables *W* and $S_h$.
+Recently, many of these models were rexplored in rivers and streams by @wangPhysicallyBasedScaling2021.
+They significantly expanded training datasets of field-measured $k_{o_2}$ (*k* for dissolved oxygen DO at 20 degrees Celsius) by using inverse modeling of DO timeseries.
+This apprroach was briefly mentioned in the introduction of this manuscript.
+These methods calibrate $k_{o_2}$, along with respiration and primary production, to high-fidelity, in situ datasets of DO with the goal of minimizing the difference between the observed DO and the mass-balance DO [e.g.
+@applingOvercomingEquifinalityLeveraging2018; @chapraDeltaMethodEstimating1991].
+@wangPhysicallyBasedScaling2021 used this technique at 35 rivers across the United States and showed that $k_{o_2}$ models deduced from the classic, process-based 'thin-film' and 'surface renewal' theories yield approximately identical parameters when fit to field measurements or this new dataset of simulated $k_{o_2}$.
+This suggests that there are uniform scaling realtionships between certain hydraulic properties and $k_{o_2}$ that exist regardless of the training data used.
+See [@hallGasExchangeStreams2020 and @wangPhysicallyBasedScaling2021] for more in-depth explanations of the thin-film and surface renewal theories with respect to rivers.
+These are the standard models for understanding gas exchange in many aquatic environments from lakes to esutraries to rivers.
+The better fit, and more parsimonious, of these two models is reprinted as equations 2 and 3 where $U_\star$ is the shear velocity of the river channel [L/T] ($U_\star = \sqrt{gR_hS_h}$).
+Equation 2 was fit to field measurements of $k_{o_2}$ while equation 3 was fit to the simulated $k_{o_2}$ for 35 rivers.
+$$k_{o2} = 47.67U_\star \mathbf{(2)}$$
+$$k_{o2} = 49.71U_\star \mathbf{(3)}$$
 
-Using the @tsivoglouTracerMeasurementReaeration1976 equation for *eD* for unsteady flow conditions (channel slope $S_0 \neq S_h$- see text S1 for this model derivation and why assuming non-uniform flow is convenient for an algorithm ingesting SWOT data) in conjunction with Manning’s equation to relate average flow velocity to channel geometry, we rewrite equation 2 as equation 3.
-In the far-right form of equation 3, we use the Manning’s formulation from @hagemannBAMBayesianAMHGManning2017a which formulates depth via river width and area (in order to use SWOT-observables).
-$k_{600}$ is now defined as a function of $S_h$, channel roughness *n* $[T/L^{\frac{1}{3}}]$, channel width *W* [L], and channel area $A_0+dA [L^2]$ (discretized via median channel area $A_0$ and the change in channel area *dA*).
-Note that we estimate *dA* by assuming a rectangular river channel so that $dA \approx \delta H*W$ and is also solely a function of *W* and $S_h$.
-Finally, equation 3 is also defined by the parameters $a_i$ and $b_i$, which are described next.
+This model was implemented in BIKER (using 48 as the approximate parameter across equations 2 and 3) for a few reasons.
+First, the analysis performed by [@wangPhysicallyBasedScaling2021] provides compelling evidence that a statistical parameter of approximately 48 is uniform across rivers and streams, regardless of the training data used.
+This is encouraging for use within BIKER as BIKER is specifically designed to be as river-agnostic as possible.
+Second, it yields a simple linear relationship across all rivers, regardless of their size or steepeness.
+This is not necessairly true of *k* models based on the turbulent dissipation rate *eD* ($eD=gS_hV$), which is another common approach [@hallGasExchangeStreams2020; e.g.
+@raymondScalingGasTransfer2012a; @oconnorMechanismReaerationNatural1958; @tsivoglouTracerMeasurementReaeration1976], where recent work has shown that *k* does not scale with $eD$ via the same statistical parameters across all rivers [@ulsethDistinctAirWater2019a].
+We further explored the relationship presented in that paper and found that the $k~eD$ relationship fundamentally breaks down in rivers with low *eD*, which corresponds to nearly all rivers that SWOT will observe (**supp figure**).
+Therefore, an $eD$ based predictive model is of little use in SWOT rivers.
+Third, we sought to explicitly predict the gas transfer velocity for DO, as its use extends beyond constraining GHG fluxes from rivers.
+For example **stuff here**.
+To convert from $k_o2$ to *k* for any other GHG (as will be done later in this manuscript to calculate $CO_2$ evasion fluxes), we use the Schmidt number to convert between water temperatures and gases of interest following equation 4.
+530 is the Schmidt number for oxygen at 20 degrees Celsius and *Sc* is the Schmidt number at one's desired temperature for some dissolved gas.
+$$k_{GHG, temp}=k_{02}*(Sc/530)^{-1/2} \mathbf{(4)}$$
+
+We stress that most other predictive models for *k* could be implemented within BIKER.
+However, our goal of using BIKER on any river that is observable by SWOT necessitates that we pick an equation that is applicable across as many rivers as possible.
+Future work should look at using a river-specific *k* equation to 'calibrate' BIKER to a specific river of interest.
+
+With a predictive model for $k_{o2}$ chosen, we now implement it within a McFLI framework.
+The approach used here is strongly informed by the @hagemannBAMBayesianAMHGManning2017a McFLI algorithm for ungauged RSQ, further explored in more recent work [@brinkerhoffConstrainingRemoteRiver2020].
+These papers conceptualize discharge as a Bayesian remote sensing problem, which we largely follow here to conceptualize $k_{o2}$ as a Bayesian remote sensing problem that can be solved using SWOT data.
+
+First, the model needs to be rewritten as a function of SWOT-observables river width *W* and height (and thus water surface slope $S_h$).
+SWOT will measure these at approximately 200m intervals in pre-defined mass-conserved river reaches (**What is the PDD?**; SWORD).
+To do this, we follow @hagemannBAMBayesianAMHGManning2017a to formulate depth via river width and area.
+Area is further split into the SWOT-observable portion *dA* and the unobservable portion $A_0$ following @durandEstimatingReachaveragedDischarge2014a.
+$A_0$ is the median channel area and *dA* is the change in channel area.
+We estimate *dA* by assuming a rectangular river channel so that $dA \approx \delta H*W$ and is also solely a function of SWOT-observables *W* and $S_h$.
 
@@
 
-$$k_{600} = a_i(eD)^{b_i} \approx a_i(gSV)^{b_i} \approx a_i\left[gS_h^{\frac{3}{2}}\frac{1}{n}(A_0+dA)^\frac{2}{3}W^\frac{-2}{3}\right]^{b_i} \mathbf{(3)}$$
-
-As will be explained in section 2.2, the only observations in the BIKER formulation of equation 3 are *W* and $S_h$.
-So, we must assign $a_i$ and $b_i$ using just *W* and $S_h$.
-To do this, we use the basic premise that the equation 2 relationship is different for different river sizes *i*.
-We use *W* and *S* to define *i* and train different $k_{600}$ functions for rivers with the same *i*.
-We fit this model to the @ulsethDistinctAirWater2019a data (Figure 2) for five possible sets of $a_i$ and $b_i$ parameters, as defined for rivers 10-50m wide, 50-100m wide, 100+m wide, low energy rivers 0-10m wide (*S* < 0.05), and high energy rivers 0-10m wide (slope > 0.05).
-We find that narrow rivers (< 10m wide) are best modeled by further breaking them down into low energy (*S* < 0.05) and high energy (*S* > 0.05) regimes in order to meet the assumptions necessary to fit a linear regression model as proposed by @raymondScalingGasTransfer2012a and @ulsethDistinctAirWater2019a (specifically, normally distributed residuals of the predictions).
-Another implication of this extension of the @ulsethDistinctAirWater2019a model is that we now use *a* and *b* parameters that implicitly reflect the distinct geomorphology of different rivers.
-
-![Figure 2: Theoretical basis for the RS-able extension of the Ulseth et al.
-(2019) k600 upscaling model.
-We use the differenial relationship between k600 and eD for different river sizes to define a 'rule-based' regression model.
-This model assigns differeniable model parameters using just river width and slope, following the form in equation 3.
-Data from Ulseth et al.
-(2019).](outputs/k600/theory_plot.jpg)
-
-**these two paragraphs need to make it clear before you get to the end that you believe the relationship to be linear, but you will not limit yourself to raymond and/or ulseth.
-Rather, you wish to find any number of linear relationships that best fit the k600 data as statistically and hydraulicly defined.
-THIS MIGHT NEED TO BE A NON-LINEAR FUNCTION.
-THE OTHER PROBLEM IS THAT ALL WE ARE JUST HIGHLIGHTING THAT ALL UPSCALING RESEARCH HAS BASICALLY ASSUMED UNIFORM FLOW....**
-
-Finally, we tested whether this model can reasonably estimate $k_{600}$ relative to the original @ulsethDistinctAirWater2019a model, with the goal of at least replicating their performance.
-We compared it against the various *eD*-based $k_{600}$ models from @raymondScalingGasTransfer2012a that are commonly used in the literature.
-This validation was done using a classic 80/20 split of the @ulsethDistinctAirWater2019a dataset, where 80% of the data was used to fit these models and 20% was used for independent testing of their predictive performance.
-This meant that the parameter values were slightly different than those published in both papers, but model structures were identical.
-This was done because the @raymondScalingGasTransfer2012a models were originally trained on a subset of this dataset and we deemed it fairer to compare model peformances using identical data.
-
-###2.2 Developing BIKER
-With a validated equation 3 in hand, we turn next to remotely sensing $k_{600}$.
-The approach used here is strongly informed by @hagemannBAMBayesianAMHGManning2017a’s McFLI algorithm for ungauged RSQ via Bayesian inference, further explored in more recent work [@brinkerhoffConstrainingRemoteRiver2020].
-These papers conceptualize discharge as a Bayesian remote sensing problem, which we largely follow here to conceptualize $k_{600}$ as a Bayesian remote sensing problem that can be solved using SWOT data.
-Below, we translate @hagemannBAMBayesianAMHGManning2017a’s ideas from discharge to $k_{600}$.
-
-SWOT will measure river width and height (and thus water surface slope $S_h$) at approximately 200m intervals in pre-defined mass-conserved river reaches (**What is the PDD?**; SWORD).
-Following @hagemannBAMBayesianAMHGManning2017a, the observations needed for Bayesian inference are SWOT measurements of *W* and $S_h$, and all other terms in equation 3 (*n*, $A_0$, and $k_{600}$) are treated as statistical parameters that cannot be directly observed from SWOT.
-
-We therefore rework equation 3 into a Bayesian sampling model to have all of the ‘observations’ on the left-hand side sampled from all the unknown model parameters on the right-hand side (equation 4, where *x* are SWOT observables given $\Theta$, a set of statistical parameters).
-Note that in this formal likelihood specification both sides of equation 4 are log-transformed (natural log) to produce the normal sampling model presented here, however they are written in equation 4 without that transformation for succinctness.
-The $σ_{k600}^2$ parameter refers to the uncertainty inherent in equation 3's estimates.
+$$k_{o2} = 48*U_\star=48\sqrt{gR_hS_h}=48g^{1/2}S_h^{1/2}(\frac{A_0+dA}{W})^{1/2} \mathbf{(5)}$$
+Next, we rework equation 5 into a Bayesian sampling model to have all of the observations (*W* and $S_h$) on the left-hand side sampled from all of the unknown model parameters ($A_0$ and $k_{o_2}$) on the right-hand side.
+This is equation 6, where *x* are SWOT observables given $\Theta$, a set of statistical parameters.
+Note that in this formal likelihood specification both sides of equation 6 are log-transformed (natural log) to produce the normal sampling model presented here, however they are written in equation 6 without that transformation for succinctness.
+The $\sigma_{k_{o2}}^2$ parameter refers to the uncertainty inherent in equation 6's estimates.
 This will be explained in detail below.
 
-$$f(x|\Theta) = \left[a_i^{-1}(gS_h^\frac{-3}{2}W^\frac{2}{3})^{b_i}\right] \sim normal\left(\left[k_{600}^{-1}\frac{1}{n}(A_0+dA)^\frac{2}{3})^{b_i}\right], \sigma^2_{k600}\right)  \mathbf{(4)}$$
+$$f(x|\Theta) = \left[48^{-1}(\frac{W}{gS_h})^{1/2}\right] \sim normal\left(\left[k_{o2}^{-1}(A_0+dA)^{1/2}\right], \sigma^2_{k_{o2}}\right)  \mathbf{(6)}$$
 
-Equation four necessitates that Bayesian priors be assigned to model parameters *n*, $A_0$, and $k_{600}$.
+Equation six necessitates that Bayesian priors be assigned to parameters $A_0$ and $k_{o_2}$.
 Bayesian priors formalize the a priori estimates (and uncertainties) for the non-remotely-sensed parameters.
-To state that more intuitively for these physical quantities, priors represent our ‘prior river knowledge’ of what *n*, $A_0$, and $k_{600}$ probably are for some river given that they cannot be remotely sensed.
+To state that more intuitively for these physical quantities, priors represent our prior river knowledge of what $A_0$ and $k_{o_2}$ probably are for some river given that they cannot be remotely sensed.
 All priors are formalized within the model as truncated normal distributions of the log-transformed terms such that $log(X) \sim N(\mu, \sigma^2)$ for $\lambda < X < \epsilon$, using prior hyperparameters mean ($\mu$), standard deviation ($\sigma$), and upper ($\epsilon$) and lower bounds ($\lambda$) for parameter *X*.
 In order to avoid relying on in situ information, we assign prior hyperparameters using SWOT data only.
-*n* and $A_0$ prior hyperparameters were assigned following the method developed by @brinkerhoffConstrainingRemoteRiver2020.
+$A_0$ prior hyperparameters were assigned following the method developed by @brinkerhoffConstrainingRemoteRiver2020.
 They developed a set of river channel prior hyperparameters for McFLI algorithms that are entirely RS-able and reflect differential channel hydraulics as a function of river geomorphology.
-They used an extensive database of field measurements and machine learning to identify patterns that associate the river width with river geomorphic properties, so these non-remotely sensible properties may be assigned to rivers given existing remotely sensed data (namely, river width from optical imagery).
-With this method, they showed substantial improvement in the accuracy of ungauged RSQ estimates with no changes to the algorithms or observations.
-This leaves only $k_{600}$ hyperparameters to be defined.
-We assigned the $k_{600}$ prior hyperparameters using a simple slope regression model trained on the @ulsethDistinctAirWater2019a dataset to assign $\mu$ (equation 6).
-$σ^2$ was set to the standard error of equation 6: 1.023.
-λ and ε were set to -3 and 2.7, respectively.
+They used an extensive database of field measurements and machine learning to identify patterns that associate the river width with river geomorphic properties, so these non-remotely sensible properties may be assigned to rivers using only the existing remotely sensed data.
+This leaves the $k_{o_2}$ hyperparameters to be defined.
+We assigned the $k_{o_2}$ prior hyperparameters using a simple slope regression model trained on the @raymondScalingGasTransfer2012a dataset to assign $\mu$ (equation 7).
+$\sigma^2$ was set to the standard error of equation 7: 0.77.
+$\lambda$ and $\epsilon$ were set to log(0.0001) and log(60), respectively.
 
-$$\mu_{k600}=3.22+0.347ln(S_h) if ln(S_h) < -4.634   \mathbf{(6a)}$$
-$$\mu_{k600}=6.85+1.13ln(S_h) if ln(S_h)>-4.634 \mathbf{(6b)}$$
+$$\mu_{k_{o_2}}=3.3492 +0.3661ln(S_h)   \mathbf{(7)}$$
 
-Finally, we estimate $σ_{k600}^2$ using Monte Carlo (MC) methods to approximate total uncertainty from equation 3.
+**FIGURE OUT Finally, we estimate $σ_{k600}^2$ using Monte Carlo (MC) methods to approximate total uncertainty from equation 3.
 Uncertainty in equation 3 can stem from 1) $a_i$ and $b_i$ parameter uncertainty, 2) error in Manning’s approximation of the average flow velocity, and 3) measurement error.
@@
 Uncertainty for velocity followed @hagemannBAMBayesianAMHGManning2017a, who estimated ln-transformed Manning’s equation error in discharge to be 0.25 (we assume that the 0.25 value is due soley from Manning's equation and not measurements of cross-sectional area in order to use here as the velocity error).
-The slope-error model that we used is explained in detail in Section 2.3 and relies on a recent analysis of expected errors in SWOT $S_h$ measurements [@durandHowWillRadar2020a].
+The slope-error model that we used is explained in detail in Section 2.3 and relies on a recent analysis of expected errors in SWOT $S_h$ measurements [@durandHowWillRadar2020a].**
 
-Finally, we repeat the above workflow three times under three different uncertainty scenarios to assess the overall influence of each uncertainity source on the equation 3 estimates.
-Uncertainty scenario 1 propogates errors in $a_i$ and $b_i$ only, uncertainty scenario 2 additionally propogates errors in log-transformed velocity, and uncertainty scenario 3 additionally propogates errors in $S_h$.
-
-With the sampling model (equation 4), prior distributions, and the $σ_{k600}^2$ parameter described, a joint posterior distribution conditional on the SWOT data is obtained.
+With the sampling model (equation 4), prior distributions, and the $\sigma_{k_{o_2}}^2$ parameter described, a joint posterior distribution conditional on the SWOT data is obtained.
 To approximate this distribution, we use a Markov Chain Monte Carlo (MCMC) algorithm.
@@
 ###2.3 BIKER validation setup
-We validated BIKER on 22 rivers using observed $V$ and observed $S_h$ First, we detail the 22 rivers, and then we detail how ‘observed’ $k_{600}$ is calculated.
+We validated BIKER on 50 rivers using observed $R_h$ and observed $S_h$ First, we detail the 50 rivers, and then we detail how observed $k_{o_2}$ is calculated.
 
-Because SWOT has yet to launch, it is standard practice to benchmark SWOT-related algorithms on ‘SWOT-like data’.
+Because SWOT has yet to launch, it is standard practice to benchmark SWOT-related algorithms on SWOT-like data.
 There are three current types of SWOT-like data: 1) AirSWOT, which is an airborne Ka-band inSAR currently limited to five rivers globally, 2) simulated rivers that mimic the type of data SWOT will provide, and 3) the SWOT simulator, which introduces measurement errors to these simulated rivers.
 Because we are principally interested in algorithm performance, we limit our validation setup to simulated rivers in order to benchmark across as many rivers as possible.
-These simulated rivers are simply reach-averaged hydraulic model outputs where the water surface heights and widths are labelled as ‘RS observations’ and are used as the sole inputs to the BIKER algorithm.
+These simulated rivers are simply reach-averaged hydraulic model outputs where the water surface heights and widths are labelled as RS observations and are used as the sole inputs to the BIKER algorithm.
 Simulated rivers mimic perfect measurement conditions (i.e.
 no measurement errors).
-Here, we use the 22 rivers archived by @frassonCompilationHydraulicModels2019a and used by @rodriguezObservingRiversVarying2020 to develop reach-averaged Saint Vernant equations for the SWOT mission.
-We omitted the Arial Khan river in Bangladesh due to known problems with that hydraulic model **is that sufficient to say?** and we further sampled the models for only the observations every 11 days to mimic the average SWOT overpass frequency (this also approximately aligns the SWOT observations with the bi-weekly $CO_2$ data used later in this study- Section 2.4).
-All told, this yielded 503 sets of SWOT observations to validate BIKER with.
+Here, we use the 50 rivers developed by @frassonExploringFactorsControlling and @durandIntercomparisonRemoteSensing2016b.
+These are the two benchmarking studies that have explored RSQ algorithm performance for the SWOT mission.
+Consult both of those papers for maps of the river's locations, though the 50 rivers are spread across the United States, Canada, France, Italy, United Kingdom, India, and Bangladesh.
+We omitted the Arial Khan river in Bangladesh due to known problems with that hydraulic model **is that sufficient to say?**.
 
-We also introduced measurement error into these simulated SWOT observations to assess BIKER's performance degradation due to errors inherent in SWOT measurements.
-Error in SWOT measurements will come from both the error tolerances intrinsic in the satellite data product as well as radar layover error.
-Here, we assume errors in river width are negligible [@durandHowWillRadar2020a] and solely focus on measurement errors in river height and slope.
+We also assess the influence of measurement error on BIKER's performance.
+Error in SWOT measurements will come from both the error tolerances intrinsic in the satellite data product as well as radar layover error and width measurement error.
 Layover error is the phenomenon when radar returns from different places arrive at the sensor at the same time, leading to taller landscape features appearing closer to the sensor than shorter landscape features that are the same horizontal distance from the sensor.
@@
 
-To explore the effect of SWOT measurement errors on BIKER, the SWOT simulator mentioned above is the ideal tool.
-However, it is cumbersome to run on many rivers **is this sufficient?** and so here we coarsely model measurement error using the results from a recent global anlaysis of SWOT water surface slope/height errors [@durandHowWillRadar2020a].
-They developed a first-principles model for layover error from SWOT observations and, using global hydrography, quantified SWOT measurement error from both error tolerances and radar layover for 220,924 river reaches.
-The 68th percentiles of these reach-scale errors were 1.7cm/km and 10.4cm for water surface slope and height, respectively.
-To apply this model to our 22 rivers, we generate new SWOT observations by sampling every observation from normal distributions following equations 7 and 8 for node *i* and timestep *t*.
-Note that because BIKER ingests *dA*, rather than *H*, reach-scale height errors were converted to *dA* errors again following @durandHowWillRadar2020a and our rectangular channel assumptions: $\sigma_{dA} \approx W\sigma_H\sqrt{2}$.
-Note that equation 7 is also used to define slope measurement error in the MC analysis (section 2.2).
+For river heights and slopes, we use the SWOT error model devolped by @durandHowWillRadar2020a and implemented on 16 of the rivers in this study by @frassonExploringFactorsControlling.
+Systematic height errors are assumed constant across the river but variable in time.
+Width errors were derived using a model built by @frassonExploringFactorsControlling based on SWOT-simulator runs for the Sacramento and Po rivers.
+Width errors due to poor water classification are ignored here.
+Please consult @frassonExploringFactorsControlling for the specifics of how these realistic errors were incorporated into the hydraulic models.
+Ultimately, we reran BIKER on these 16 rivers with the SWOT error model corrupting the SWOT-observables *W* and *S_h*.
 
-$$S_{err_{i,t}} \sim normal(S_{obs_{i,t}}, 0.000017^{2}) \mathbf{(7)}$$
-
-$$dA_{err_{i,t}} \sim normal(dA_{obs_{i,t}}, \left[W_{obs_{i,t}}*\sqrt{2}*0.104\right]^{2}) \mathbf{(8)}$$
-
-Regardless of the validation data used, we do not have observed $k_{600}$ data for these rivers, and to our knowledge no field dataset of $k_{600}$ exists in the type of temporal and spatial frequency that SWOT (and therefore the BIKER algorithm) provides.
-Further, we are principally interested in our ability to reproduce upscaled $k_{600}$ from RS observations and are less concerend with the actual accuracy of the upscaling model itself [which can be validated using existing datasets like previously done by @raymondScalingGasTransfer2012a; @ulsethDistinctAirWater2019a].
-Therefore, we take the best performing upscaling model from sections 2.1 and 3.1 and use that to calculate the ‘observed’ $k_{600}$ that BIKER is validated against.
-This is done using equation 9, where $V_{obs} [L/T]$ is observed river discharge divided by observed channel area.
-With this setup, we can directly explore BIKER's ability to infer observed *V* and $k_{600}$ from river width and height alone.
-It also means that, for a fair validation scheme, $\sigma_{upscale}^2$ must be set to reflect only error from Manning’s equation (and not $a_i$ and $b_i$).
+Regardless of the validation data or SWOT error budget used, we do not have observed $k_{o_2}$ data for these rivers, and to our knowledge no field dataset of $k_{o_2}$ exists in the type of temporal and spatial frequency that SWOT (and therefore the BIKER algorithm) provides.
+Further, we are principally interested in BIKER's ability to reproduce $k_{o_2}$ from SWOT observations and are less concerned with the actual accuracy of the predictive model itself [which can be validated using existing datasets like previously done by @wangPhysicallyBasedScaling2021].
+Therefore, we take the model outlined in section 2.1 and use that to calculate the observed $k_{o_2}$ that BIKER is validated against.
+This is done using equation 8, where $R_{h, obs} [L/T]$ is observed river channel area divided by observed channel width.
+With this setup, we can directly explore BIKER's ability to infer observed $U_\star$ and $k_{o_2}$ from *W* and $S_h$ alone.
+It also means that, for a fair validation scheme, $\sigma_{upscale}^2$ must be set to reflect only error from Manning's equation (and not the parameter 48 in equation 8).
 Thus, $\sigma_{upscale}^2$ is set to 0.25 for this validation.
@@
 
-$$k_{600, obs}=a_i(gS_hV_{obs})^{b_i} \mathbf{(9)}$$
+$$k_{o_2, obs}=48\sqrt{gS_hR_{h, obs}} \mathbf{(8)}$$
 
@@
 To validate across all rivers and timesteps, we used the coefficient of determination $r^2$ and the root mean square error RMSE.
-Four normalized metrics were used for by-river validation: RRMSE and NRMSE are normalized root mean square errors that have been normalized by the observed value and the mean observed value (respectively) to compare across rivers.
-rBIAS is a measure of prediction bias that is normalized by the mean observed value to compare across rivers.
+Four normalized metrics were used for by-river validation: RRMSE and NRMSE are normalized root mean square errors that have been normalized by the observed value and the mean observed value (respectively).
+rBIAS is a measure of prediction bias that is normalized by the mean observed value.
 KGE is a standard metric used in streamflow prediction with an intuitive basis: a value greater than 0 is often interpreted as a useful prediction in ungauged settings, and a value greater than -0.41 indicates a model outperforms a uniform prediction of the mean observed value [@knobenTechnicalNoteInherent2019a].
@@
 ###2.4 Upscaling to CO2 evasion and bulk carbon efflux
-It is one thing to to predict $k_{600}$, but researchers are usually more interested in the gas fluxes from rivers and ultimately the bulk carbon efflux from river to atmosphere.
+It is one thing to to predict $k_{o_2}$, but researchers are usually more interested in the gas fluxes from rivers and ultimately the bulk carbon efflux from river to atmosphere.
 Therefore, we explored 1) the ability of BIKER to reproduce hypothetical carbon dioxide evasion fluxes ($FCO_2$ $[M/L^2T]$) from these 22 rivers, and 2) the sensitivity of the estimated bulk carbon efflux $[M/T]$ via $CO_2$ evasion to the method used to estimate reach averaged flow velocity.
@@
 
-###3.1 Validating the $k_{600}$ upscaling model
-Figure 3 plots the validations for the five upscaling functions tested.
-Both the original @ulsethDistinctAirWater2019a model and our RS-able extension perform the best on this independent set of validation data, with near identical performance.
-The RS-able @ulsethDistinctAirWater2019a model achieves an $r^2$ of `r results_3_1[results_3_1$Model == 'RS-able',]$r2` and an RMSE of `r results_3_1[results_3_1$Model == 'RS-able',]$rmse` m/day, which is just marginally better than the original @ulsethDistinctAirWater2019a model (`r results_3_1[results_3_1$Model == 'Ulseth',]$r2` and `r results_3_1[results_3_1$Model == 'Ulseth',]$rmse` m/day, respectively).
-The two power-law based @raymondScalingGasTransfer2012a models (equations 4 and 3 in that study) achieve slightly worse performance, with $r^2$ in the 0.70s and their RMSE scores slightly higher (`r results_3_1[results_3_1$Model == 'Raymond_2',]$rmse` and `r results_3_1[results_3_1$Model == 'Raymond_3',]$rmse` m/day, respectively).
-@raymondScalingGasTransfer2012a equation 5, which is linear in structure, worked well on their dataset but fails to capture the non-linear relationship evident in this expanded and more geomorphically diverse dataset.
-In fact, it predicts negative $k_{600}$ values and error metrics cannot be calculated because the prediction residuals are not normally distributed.
-**SHOULD I INCLUDE THIS MODEL or not mention it??** Because we successfully reproduced the @ulsethDistinctAirWater2019a model, the rest of this manuscript deals exclusively with this RS-able variant of the model.
-
-![Figure 3.
-Validation of five k600 upscaling models on 20% of the Ulseth et al (2019) dataset, withheld for independent testing.
-Raymond et al.
-(2012) Eq.
-5 predicts negative k600 values, which are not plotted here and highlight a limitation of using a linear model for a non-linear problem (though it did work well on their smaller dataset).](outputs/k600/validation.jpg)
-
-Uncertainties were then propagated through the RS-able @ulsethDistinctAirWater2019a model via MC simulations for the three uncertainity scenarios (section 2.2) 5,000 times.
-Figures 4a, 4c, and 4e plot histograms of the 5,000 uncertainty terms extracted from those distributions.
-Figures 4b, 4d, and 4f plot three of the distributions used to build the histograms as examples.
-Across the 5,000 tests for the fullest error scenario (Figure 4e), we quantified median ln-transformed uncertainty in equation 4 to be `r round(results_3_1_MC$medianSigma,2)` (~`r round(exp(results_3_1_MC$medianSigma), 2)` m/day- dashed blue line in Figure 4a).
-Median uncertainty solely from the upscaling model is 1.27 (~`r exp(1.27)` m/day), confirming that functionally all uncertainity in the equation 4 estimates is due to upscaling and not due to the assumptions made for BIKER.
-The only notable difference between these histograms is in the Figure 4e right-hand tail, where the cases when estimate uncertainity is extremely high get larger.
-Figures 4b, 4d, and 4f are also functionally identical, further corroborating our conclusion.
-
-![Figure 4 Model uncertainty estimates for three error scenarios, as detailed.
-Left: histogram of error terms (SD) for all 5,000 MC simulations with blue dashed line denoting the median.
-Right: Three example MC simulation results of 10,000 samples each.](outputs/MonteCarlo/MC_results.jpg)
-
 ###3.2 Validating the BIKER algorithm
@@
 Grey line is linear regression (and 95% prediction intervals are dashed) and dashed black line is 1:1 line.
-b: Cumulative density functions (CDFs) of the same results: dashed black line are observed values, green line are BIKER posterior means, and blue lines are BIKER 95% CIs.](outputs/validation/validation.jpg)
+b: Cumulative density functions (CDFs) of the same results: dashed black line are observed values, green line are BIKER posterior means, and blue lines are BIKER 95% CIs.](cache/validation/validation.jpg)
 
@@
 b) was randomly selected from the upper tertile of KGE scores, c) was randomly selected from the middle tertile, and d) from the worst tertile.
-Model results include the posterior means and 95% CIs.](outputs/validation/validation_by_river.jpg)
+Model results include the posterior means and 95% CIs.](cache/validation/validation_by_river.jpg)
 
@@
 a: FCO2 from BIKER versus using observed average flow velocity for all timesteps for 22 SWOT rivers (grey lines are linear regression and 95% prediction intervals, while black dashed line is the 1:1 line).
-b-d: timeseries plots for the three example rivers from Figure 5.6.](outputs/FCO2/FCO2_plot.jpg)
+b-d: timeseries plots for the three example rivers from Figure 5.6.](cache/FCO2/FCO2_plot.jpg)
 
@@
 b: Histograms of by-river performance in estimating FCO2 across all timesteps and rivers for the same four velocity models.
-Dashed lines are identical to Figure 6a.](outputs/FCO2/FCO2_models.jpg)
+Dashed lines are identical to Figure 6a.](cache/FCO2/FCO2_models.jpg)
 

### 2021-05-18 generateMS.R

@@ -2,9 +2,11 @@
 #This will generate and store all results in your current RStudio environment so that you can knit the manuscript correctly, and then it'' knit the manuscript too.
 #################
 
+setwd('C:\\Users\\craig\\Documents\\GitHub\\RSK600')
+
 #Move all results to global environment
-results_3_1 <- read.csv('cache//k600//results.csv')
-results_3_1_MC <- read.csv('cache//MonteCarlo//results.csv')
+#results_3_1 <- read.csv('cache//k600//results.csv')
+#results_3_1_MC <- read.csv('cache//MonteCarlo//results.csv')
 results_3_2_rivs <- read.csv('cache//validation//results_by_riv.csv')
 results_3_2_all <- read.csv('cache//validation//results_all_riv.csv')
 results_3_3_all <- read.csv('cache//FCO2//fco2_stats_all.csv')
@@ -13,7 +15,7 @@ results_3_3_bulk <- read.csv('cache//FCO2//bulkFluxes.csv')
 results_3_3_bulk$key <- as.character(results_3_3_bulk$key)
 
 #knit manuscript
-rmarkdown::render("docs//manuscript_v2.Rmd")
+rmarkdown::render("manuscript_v3.Rmd")
 print('Manuscript generated.')
 
 rmarkdown::render("docs//supp_v1.Rmd")

### 2021-05-24 manuscript_v3.Rmd
big ms edits

@@
 ##Highlights (3-5 points, 85 characters each w/ spaces)
-- The SWOT satellite should be able to indirectly remotely sense riverine gas exchange velocity
-- The BIKER algorithm predicts $CO_2$ fluxes when coupled with in situ gas concentration data
-- Predicted BIKER $CO_2$ fluxes compare well with in-situ methods
+- The BIKER algorithm indirectly remotely senses riverine gas exchange velocity
+- Predicted gas exchange velocities compare well with in-situ methods
+- BIKER predicts $CO_2$ fluxes when coupled with in situ gas concentration data
 - Expected SWOT measurement errors exert a marginal influence on BIKER's estimates
@@
 Go BIKER!!
-max 400 words
-x
-x
-x
-x
-x
-x
-x
-x
-x
+*Note: this is for Remote Sensing of Environment (I think). So 400 words here.*
 
 ##1 Introduction
-Inland waters transport and transform various constiutents as they move from the landscape to the ocean [@colePlumbingGlobalCarbon2007a].
-**Too big a jump.
-What is the journal?
-‘constituents’ will need to be defined, as will ‘processes’**As a result some of these processes, rivers and lakes are usually supersaturated with dissolved greenhouse gases (GHGs) and thus present a significant flux of these gases from water to air [@raymondGlobalCarbonDioxide2013a].
-This flux $[M/L^2T]$ is expressed as equation 1 given the gas concentration gradient between the water $C_{water} [M/L^3]$ and the air $C_{air} [M/L^3]$ and the gas exchange velocity $k [L/T]$.
-Note that *k* is generally studied as normalized by a Schmidt number of 600 ($k_{600}$) to be gas and temperature agnostic [see @hallGasExchangeStreams2020 for a thorough review of gas exchange in rivers].
+The Earth's carbon cycle transports and manipulates carbon between the atmosphere, geosphere, hydrosphere, and biosphere via a complex set of interconnected biogeochemical processes.
+The primary connection between the terrestrial and oceanic biospheres is the global river network, which exports much carbon from the continents to the oceans [@colePlumbingGlobalCarbon2007a].
+However en route to the oceans, this river network additionally transports and transforms sediments, nutrients, organic matter, dissolved gases, and other constituents.
+Organic matter and inorganic nutrients are also produced and/or consumed within rivers via ecosystem metabolism [@odumPrimaryProductionFlowing1956].
+The net effect of all of this is the global river network is usually supersaturated with dissolved gases and evades substantial amounts of carbon from its water surface to the atmosphere [@colePlumbingGlobalCarbon2007a].
+
+Current estimates of total carbon dioxide ($CO_2$) evasion to the atmosphere vary from 650-1800 Tg C/yr [@lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a], with 167 Tg-C/yr coming from mountain streams alone [@horgbyUnexpectedLargeEvasion2019].
+Further, in-river metabolism is increasingly understood to contribute to this flux [@hotchkissSourcesProcessesControlling2015].
+Equation 1 represents this flux for any sparingly soluble gas $[M/L^2T]$ given the gas concentration gradient between the water $C_{water} [M/L^3]$ and the air $C_{air} [M/L^3]$ and the gas exchange velocity $k [L/T]$.
+Given the structure of equation 1, calculations of this flux are highly sensitive to our measurements/estimates of *k*.
+Note that *k* is different for different gases and water temperatures [see @hallGasExchangeStreams2020 for a thorough review of gas exchange in rivers].
 
@@
 
-To directly measure $k_{600}$ in a given stream, scientists use a range of methods mostly involving floating domes or eddy covariance towers to measure at a point in the river, or using injections of tracer gases to estimate reach-scale gas exchange [again, see @hallGasExchangeStreams2020].
-Obiously, these types of of measuremnets are preferred but are infeasible when working at the network-scale across potentially tens of thousands of rivers, and so upscaling functions are used to predict $k_{600}$ from readily available river geomorphology parameters [e.g.
+*k* is also vital to understanding freshwater ecology through bulk river metabolism [gross primary production and ecological respiration- @bernhardtMetabolicRegimesFlowing2018].
+The general mass-balance equation for river metabolism (*M* $[M/L^2T]$) is provided as equation 2, where *DO* is the dissolved oxygen concentration $[M/L^3]$, *D* is river depth [L], and $k_{o_2}$ is *k* for oxygen at some temperature.
+When viewing equations 1 and 2 in aggregate, it is clear that *k* plays a critical role in two of the fundamental biogeochemical processes that occur in river networks.
+Therefore, both closing the global carbon cycle and further understanding freshwater ecology rely on accurate and high-fidelity assessments of *k*.
+$$\frac{dDO}{dt}=\frac{M}{D}+\frac{k_{o_2}}{D}(DO_{air}-DO_{water}) \mathbf{(2)}$$
+Researchers use a range of methods to measure *k*, from floating domes to eddy covariance towers to measure at a specific point in the river.
+They also inject tracer gases to estimate reach-scale gas exchange [@hallGasExchangeStreams2020].
+These approaches are ideal but are infeasible when working at the river network scale across potentially tens of thousands of rivers.
+So, predictive models are used to estimate *k* from readily available river geomorphology [e.g.
 @oconnorMechanismReaerationNatural1958; @palumboAssessingPerformanceReaeration2014; @raymondScalingGasTransfer2012a].
-Many upscaling functions are based on the basic theory that *k* is empirically correlated with the turbulent energy dissipation *e* at the air-water interface [@zappaEnvironmentalTurbulentMixing2007].
-This is convenient for upscaling because in rivers and streams *e* can be equated to the turbulent energy dissipation rate *eD* $[L^2/T^3]$ [@ulsethDistinctAirWater2019a; @raymondScalingGasTransfer2012a], which is estimated using easily calculated hydraulic parameters: $eD \approx gSV$ where *g* is gravitational acceleration $[L/T^2]$, *S* is channel slope $[L/L]$, and *V* is average flow velocity $[L/T]$ [@tsivoglouTracerMeasurementReaeration1976].
-Because channel slope is readily available in any hydrographic data product, most efforts to upscale $k_{600}$ to drainage networks are thereby limited by the quality of the final parameter *V*.
-Since *V* must be measured in the field or itself estimated by other models/empirical methods, uncertainty in *V* must dominates errors in $k_{600}$ estimates.
-This is exacerbated in ungauged basins that cover large sections of the earth system, especially in the carbon-rich Arctic inland waters, as no in situ hydraulic information is available and fieldwork is impractical [@gleasonRemoteSensingRiver2020].
+These models often, though not alwyas, predict $k_{600}$, or *k* that has been normalized to a Schmidt number of 600 ($CO_2$ at 20 degrees Celsius or $O_2$ at 17.5 degrees).
 
-To circumvent this problem in ungauged basins and at network-scales, researchers have used ‘global-scope’ hydraulic geometry (HG) models: simple power law relations between in this case *V* and discharge that are trained on extensive hydraulics datasets [@leopoldHydraulicGeometryStream1953a; @borgesVariabilityGasTransfer2004; @horgbyUnexpectedLargeEvasion2019; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a].
-However, it has long been established that HG parameters are highly variable across landscapes and from river to river [@gleasonHydraulicGeometryNatural2015; @parkWorldwideVariationsHydraulic1977; @rhodesBfmDiagramGraphical1977].
-Because of this, a large body of geomorphology work has attempted to parse out process-based explanations for HG parameters [e.g.
-@dingmanAnalyticalDerivationAtastation2007b; @parkerPhysicalBasisQuasiuniversal2007b; @singhTHEORIESHYDRAULICGEOMETRY2003a; @fergusonHydraulicsHydraulicGeometry1986a] and an open research question is how to best prescribe a set of HG parameters to a given river (particularly at the global scale).
-**Xxx give me a sentence or two on the problems with this is approach here.
-You bury the problem with the sentence “however, it has long” but then cover it up.
-The transition doesn’t work without closing this paragraph with a supported (i.e.
-with citations) statement explicitly defining why we can’t do the global scope HG approach.
-xxx **
+There is a long tradition of water quality engineers and biogeochemists developing predictive models for different forms of *k*, going back to at least the 1950s [@hallGasExchangeStreams2020] and extending the original work of @streeterMeasuresNaturalOxidation1935 and @phelpsStudiesSelfPurificationStreams1914.
+The goals of these models vary, from predicting oxygen reaeration coefficients to predicting $CO_2$ gas transfer velocities.
+Most of these models operate under the basic assumption that river channel hydraulics govern gas exchange and so simple statisical relationships can be generated between *k* and hydraulic properties.
+However, many of them are not based on fundamental mass transport theory and rather are simple empirical relations.
+This means that those models' parameters are sensitive to the training data used [@wangPhysicallyBasedScaling2021].
 
-A potential alternative to this ‘global-scope’ HG approach is to directly estimate a river’s hydraulic properties (e.g.
-V) from remote sensing (RS) data.
-Remote sensing of river hydraulics is a burgeoning subfield within remote sensing of hydrology, typically in service of remote sensing of river discharge (RSQ- see @gleasonRemoteSensingRiver2020 for a thorough review).
+Multiple predictive models are based on the observation that *k* is generally correlated with both the turbulent energy dissipation rate *eD* $[L^2/T^3]$ [@ulsethDistinctAirWater2019a; @raymondScalingGasTransfer2012a] and the water surface surface shear velocity $U_\star [L/T]$ [@wangPhysicallyBasedScaling2021; @katulStructureFunctionModel2018].
+This is convenient because both *eD* and $U_\star$ are calculated using easily estimated hydraulic parameters: $eD \approx gS_hV$ [@tsivoglouTracerMeasurementReaeration1976] and $U_\star=\sqrt{gS_hR_h}$, where *g* is gravitational acceleration $[L/T^2]$, $S_h$ is water surface slope $[L/L]$, *V* is average flow velocity $[L/T]$ and $R_h$ is the hydraulic radius [L] (which often approximates *D*).
+Because slope is readily available in any hydrographic data product, most efforts to predict *k* across thousands of rivers are thereby limited by the quality of the final parameters *V* or $R_h$.
+Since both must be measured in the field or themselves estimated by other empirical methods, uncertainty in *V* or $R_h$ must dominate any errors in *k* estimates.
+This is exacerbated in ungauged basins that cover large areas, especially in the carbon-rich Arctic inland waters, as no in situ hydraulic information is available and fieldwork is impractical [@gleasonRemoteSensingRiver2020].
+
+To circumvent this problem in ungauged basins and across thousands of rivers, researchers have used 'global-scope' hydraulic geometry (HG) models that extend HG beyond its traditional use.
+Traditional HG models are simple power law relations between streamflow and channel width, depth, or velocity at a specific cross-section in a river [@leopoldHydraulicGeometryStream1953a] and a large body of geomorphology work has attempted to parse out process-based explanations for observed HG parameters [e.g.
+@dingmanAnalyticalDerivationAtastation2007b; @parkerPhysicalBasisQuasiuniversal2007b; @singhTHEORIESHYDRAULICGEOMETRY2003a; @fergusonHydraulicsHydraulicGeometry1986a].
+This has been extended to 'downstream HG' which relates a characteristic streamflow (i.e.
+mean annual or bankful discharge) to characterisitc hydraulics along a specific river's course [@leopoldHydraulicGeometryStream1953a].
+
+For the purposes of predicting *k* from *V* or $R_h$, biogeochemists have further extended HG models by training them on large hydraulics datasets for thousands of rivers and under all streamflow conditions [e.g @borgesGloballySignificantGreenhousegas2015a; @horgbyUnexpectedLargeEvasion2019; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a].
+However, it has long been established that HG parameters are either cross-section specific and highly variable from river to river (traditional HG) or they are discharge specific [downstream HG- @gleasonHydraulicGeometryNatural2015; @parkWorldwideVariationsHydraulic1977; @rhodesBfmDiagramGraphical1977].
+This means that the 'global-scope' HG models that are usually used are highly reliant on their training data and can potentially yield pretty different hydraulics for the same streamflow.
+It is currently not well understood how sensitive global estimates of riverine gas evasion are to the specific 'global-scope' HG model that is employed by the worker.
+
+A potential alternative to this 'global-scope' HG approach is to directly estimate a river's hydraulic properties from remote sensing (RS) data.
+Remote sensing of river hydraulics is a burgeoning subfield within remote sensing of hydrology, often in service of remote sensing of river discharge (RSQ- see @gleasonRemoteSensingRiver2020 for a recent review).
 There are many ways to perform RSQ, from calibrating RS data to local channel hydraulics [e.g.
 @brakenridgeOrbitalMicrowaveMeasurement2007; @pavelskyUsingWidthbasedRating2014; @pavelskyRemoteSensingSuspended2009; @tarpanelliEstimationRiverDischarge2013] to calibrating hydraulic/hydrologic models with both in situ and RS data [e.g.
-@bjerklieEstimatingDischargeRivers2005; @chandanpurkarSatelliteReanalysisBasedMass2017; @linGlobalReconstructionNaturalized2019a; @nealDataAssimilationApproach2009] to methods that use no in situ information in their hydraulic/hydrologic models (termed 'ungauged approaches').
+@bjerklieEstimatingDischargeRivers2005; @chandanpurkarSatelliteReanalysisBasedMass2017; @linGlobalReconstructionNaturalized2019a; @nealDataAssimilationApproach2009] to methods that use no in situ information in their hydraulic/hydrologic models (termed 'ungauged approaches' and detailed below).
 Many, but not all, of these ungauged approaches were developed in the context of the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission.
-**CRAIG: might need to add unaguged citations here or rewrite perspective**
 
-SWOT is expected to launch in 2022 and provide the world’s first global measurements of water surface extent and elevation at novel temporal resolutions.
-SWOT is a wide swath Ka-band radar interferometer and will sample rivers every 1 to 7 days per 21 day repeat cycle that will measure rivers wider than 100m with a goal of expanding this to rivers at least 50m wide [@biancamariaSWOTMissionIts2016].
+SWOT is expected to launch in 2022 and provide the world's first global measurements of water surface extent and elevation at novel temporal resolutions.
+SWOT is a wide swath Ka-band radar interferometer and will sample rivers every 1 to 7 days per 21 day repeat cycle.
+This yields an average sampling resolution of 11 days.
+It will measure rivers wider than 100m with a goal of expanding this to rivers at least 50m wide [@biancamariaSWOTMissionIts2016].
 A decade of SWOT work has explored the multi-parameter problem of estimating the river hydraulic parameters of roughness and bathymetry from remote sensing to produce the SWOT discharge product [e.g.
-@andreadisConstrainingAssimilationSWOT2020d; @brinkerhoffConstrainingRemoteRiver2020; @durandEstimatingReachaveragedDischarge2014a; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014a; @garamboisVariationalEstimationEffective2020; @hagemannBAMBayesianAMHGManning2017a; @larnierRiverDischargeBathymetry2020a; @oubanasDischargeEstimationUngauged2018].
-A specific subset of these ungauged methods are termed ‘McFLIs’ or ‘Mass Conserved Flow Law Inversion’ algorithms [e.g.
-@andreadisConstrainingAssimilationSWOT2020d; @brinkerhoffConstrainingRemoteRiver2020; @hagemannBAMBayesianAMHGManning2017a; @durandEstimatingReachaveragedDischarge2014a].
-These use basic geomorphic theories and the concept of ‘prior river knowledge’ to estimate discharge from RS data where not all hydraulic terms are RS-able [@gleasonRemoteSensingRiver2020; @gleasonTrackingRiverFlows2017].
-McFLIs are globally flexible, easily implementable in any river that SWOT can sample, and improve our hydrological understandings of ungauged rivers when little to no information was previously available [@brinkerhoffConstrainingRemoteRiver2020; @durandIntercomparisonRemoteSensing2016b].
-Recently, McFLIs have also shown promise in providing additional and beneficial information to traditional hydrologic modeling via data assimilation [@ishitsukaCombiningOpticalRemote2020].
+@andreadisConstrainingAssimilationSWOT2020d; @brinkerhoffConstrainingRemoteRiver2020a; @durandEstimatingReachaveragedDischarge2014a; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014a; @garamboisVariationalEstimationEffective2020; @hagemannBAMBayesianAMHGManning2017a; @larnierRiverDischargeBathymetry2020a; @oubanasDischargeEstimationUngauged2018].
+A specific subset of these ungauged methods are termed McFLIs or 'Mass Conserved Flow Law Inversion' algorithms [e.g.
+@andreadisConstrainingAssimilationSWOT2020d; @brinkerhoffConstrainingRemoteRiver2020a; @hagemannBAMBayesianAMHGManning2017a; @durandEstimatingReachaveragedDischarge2014a].
+These use basic geomorphic theories and the concept of 'prior river knowledge' to estimate discharge from RS data where not all hydraulic terms are RS-able [@gleasonRemoteSensingRiver2020; @gleasonTrackingRiverFlows2017].
+McFLIs are readily implemented in any river that SWOT can observe and improve our hydrological understandings of ungauged rivers when little to no information was previously available [@brinkerhoffConstrainingRemoteRiver2020a; @durandIntercomparisonRemoteSensing2016b].
+Recently, McFLIs have also shown promise in providing additional and beneficial information to traditional hydrologic modeling via data assimilation of McFLIs into a traditional hydrologic model [@ishitsukaCombiningOpticalRemote2020].
 
-While all McFLIs to date have been developed in the context of RSQ, there is no reason their logic cannot be used to estimate $k_{600}$, particularly as McFLIs often employ Bayesian inference for equifinal inverse problems.
-Equifinality refers to an under-constrained mathematical system that have essentially infinite parameter combinations that can produce the same result: there are in essence more unknowns than equations [@garamboisInferenceEffectiveRiver2015], and this problem is common to both RSQ and $K_{600}$ estimation.
-Bayesian techniques have been previously used to estimate $k_{600}$ concurrently with stream metabolism from dissolved oxygen ($O_2$) datasets [@applingOvercomingEquifinalityLeveraging2018; @graceFastProcessingDiel2015; @holtgrieveSimultaneousQuantificationAquatic2010].
-While these studies are field based and are not applicable to $k_{600}$ estimation in ungauged rivers, they suggest that Bayesian techniques could be useful for remotely sensing $k_{600}$ when parameter equifinality is a problem, which matches McFLI logic precisely [@andreadisConstrainingAssimilationSWOT2020d].
+While all McFLIs to date have been developed in the context of RSQ, there is no reason their logic cannot be used to estimate *k*, particularly as McFLIs often employ Bayesian inference for equifinal inverse problems.
+Equifinality refers to an under-constrained mathematical system that has essentially infinite parameter combinations that can produce the same result: there are in essence more unknowns than equations [@garamboisInferenceEffectiveRiver2015], and this problem is common to both RSQ and RS of *k*.
+Bayesian techniques have been previously used to concurrently solve for *k* and *M* in equation 2 [@applingOvercomingEquifinalityLeveraging2018; @graceFastProcessingDiel2015; @holtgrieveSimultaneousQuantificationAquatic2010].
+While these studies are field based, require high-fidelity in situ data, and are not applicable to *k* estimation in ungauged rivers, they suggest that Bayesian techniques could be useful for remotely sensing *k* when parameter equifinality is a problem.
+This matches McFLI logic [@andreadisConstrainingAssimilationSWOT2020d].
 
-In this context, we hypothesize that a combination of SWOT data, a Bayesian McFLI paradigm, and gas exchange theory can successfully estimate $K_{600}$.
-This would in turn improve our understanding of riverine GHG fluxes in ungauged basins.
-Therefore, we developed a method to estimate $k_{600}$ solely from SWOT measurements, using hydrologic model data to produce SWOT-like data as SWOT has not yet launched [standard practice in the SWOT community- e,g, @durandIntercomparisonRemoteSensing2016b].
-We hypothesize that remote sensing $k_{600}$ is possible by coupling Bayesian remote sensing techniques (in a manner similar to McFLI) and SWOT data with hydraulic geometry and gas exchange theory.
-More specifically, this manuscript aims to answer two questions:
+In this context, we hypothesize that a combination of SWOT data, a Bayesian McFLI paradigm, and gas exchange theory can successfully estimate *k*.
+This would in turn improve our understanding of riverine gas fluxes in ungauged basins (equation 1), as well as potentially help in parameterizing stream metabolism models (equation 2).
+Therefore, this manuscript aims to answer two questions:
 
-- Is ungauged RS of the gas exchange velocity ($K_{600}$) possible using soon-to-be-available SWOT river measurements?
-- If possible, how will this RS method change estimates of riverine carbon efflux at network scales?
+- Is ungauged RS of the gas exchange velocity (*k*) possible using soon-to-be-available SWOT river measurements?
+- If possible, how will this RS method effect estimates of riverine carbon efflux from rivers?
 
-To answer these questions, we developed a new McFLI algorithm that ingests SWOT data and produces $k_{600}$ estimates and its explicit Bayesian uncertainity that requires no in situ inputs of any kind (although in situ data can be ingested and will improve results).
-We name the RS of $k_{600}$ algorithm BIKER, or the ‘Bayesian Inversion of the $k_{600}$ Rate’.
-We validate BIKER on simulated SWOT data for 22 rivers.
-We also quantify BIKER's sensitivity to the expected SWOT measurement errors: while SWOT data represent a sea change in inland water monitoring, SWOT is expected to have an approximately 10cm error in water surface elevation [@biancamariaSWOTMissionIts2016].
-Finally, we use previously published dissolved $CO_2$ data to compare the bulk carbon efflux (via $CO_2$ evasion) from the 22 rivers as calculated using BIKER and previously published in situ techniques to show how BIKER’s $k_{600}$ estimates translate into carbon efflux.
+To answer these questions, we developed a new McFLI algorithm that ingests SWOT data and produces $k_{o_2}$ estimates and its explicit Bayesian uncertainity that requires no in situ inputs of any kind (although in situ data can be ingested and will improve results).
+We name the RS of $k_{o_2}$ algorithm BIKER, or the 'Bayesian Inversion of the $k_{o_2}$ Evasion Rate'.
+We validate BIKER on simulated SWOT data for 49 SWOT-observable rivers from around the world, using hydraulic models  to produce SWOT-like data as SWOT has not yet launched.
+This is standard practice in the SWOT community [@frassonExploringFactorsControlling2021; @durandIntercomparisonRemoteSensing2016b].
+We also quantify BIKER's sensitivity to the expected SWOT measurement errors on 17 of those rivers: while SWOT data represent a sea change in inland water monitoring, SWOT is expected to have an approximately 10cm error in water surface elevation [@biancamariaSWOTMissionIts2016], as well as river width errors [@frassonExploringFactorsControlling2021].
+Finally, we use previously published dissolved $CO_2$ data to compare the bulk carbon efflux (via $CO_2$ evasion) from the 49 rivers as calculated using BIKER and previously published 'global-scope' HG models to compare BIKER with established in situ techniques.
 
 ##2 Methods
-To build BIKER, we join a process-based model for $k_{o2}$ with a McFLI framework for inverting SWOT measurements (section 2.1).
-Following the description of that process, we describe the validation (section 2.3) and the workflow for comparing estimated bulk carbon effluxes from a suite of hydraulic geometry models (section 2.4).
-A flowchart detailing the entire study is given as Figure 1.
+To build BIKER, we join a process-based model for $k_{o_2}$ (section 2.1) with a McFLI framework for inverting SWOT measurements (section 2.2).
+Following the description of that process, we describe the validation (section 2.3) and the workflow for comparing estimated bulk carbon effluxes from a suite of 'global-scope' HG models (section 2.4).
+A flowchart detailing the entire study is provided as Figure 1.
 
@@
 Flowchart of methodology used in this study.
-We begin by deriving a process-based model for ko2 that uses only river width and slope to assign its parameters (section 2.1).
-Then, we implement this model within a McFLI ('BIKER') to estimate ko2 solely from river width and water surface slope (section 2.2).
+We begin by choosing a process-based model for k (section 2.1) that is a function of river hydraulics measurable by SWOT.
+Then, we implement this model within a McFLI ('BIKER') to estimate k solely from river width and water surface slope (section 2.2).
 Finally, we couple BIKER with in situ CO2 data to compare BIKER-estimated gas fluxes with established in situ methods (section 2.4).
@@
 
-###2.1 Developing BIKER
-To use the McFLI framework, a physical model for the parameter of interest is necessary.
-In this case we need a physical model for *k*.
-First, we detail which model for *k* we chose to implement, and then we explain how the implementation works.
+###2.1 Choosing a predictive model for *k*
+To predict k from just the SWOT observables,river width *W* and water surface elevation *H*, a physical model for *k* is first selected.
+As mentioned in section 1, there are nearly two dozen predictive models for *k* that have been devolped since the 1950s [@wangPhysicallyBasedScaling2021; @hallGasExchangeStreams2020].
+Recently, many of these models were reexplored in rivers and streams by @wangPhysicallyBasedScaling2021.
+They significantly expanded the existing training datasets of field-measured $k_{o_2}$ by using inverse modeling of DO timeseries.
+This approach was also briefly mentioned in section 1.
+These methods calibrate $k_{o_2}$, along with *M* (discretized as seperate respiration and gross primary production terms), to high-fidelity, in situ datasets of DO with the goal of minimizing the difference between the observed DO and the modeled DO [e.g.
+@applingOvercomingEquifinalityLeveraging2018; @chapraDeltaMethodEstimating1991].
+@wangPhysicallyBasedScaling2021 used the *streamMetabolizer* technique [@applingOvercomingEquifinalityLeveraging2018] at 35 rivers across the United States and showed that $k_{o_2}$ models deduced from the classic, process-based 'thin-film' and 'surface renewal' theories for mass transport yield approximately identical parameters when fit to field measurements or this new dataset of simulated $k_{o_2}$.
+They also showed through cross-validation that these models are robust to overfitting on specific sets of data (a concern with many of the aforementioned predictive models- section 1).
+This suggests that there are uniform scaling realtionships between certain hydraulic properties and $k_{o_2}$ that exist regardless of the training data used.
+See @hallGasExchangeStreams2020 and @wangPhysicallyBasedScaling2021 for more in-depth explanations of the thin-film and surface renewal theories with respect to rivers.
+In short, these are the standard physical models for conceptualizing gas exchange in various aquatic environments from lakes to estuaries to rivers.
+The better fit, and more parsimonious, of the two models from @wangPhysicallyBasedScaling2021 is reprinted as equation 3(with reported coefficients of determination $r^2$) where $\alpha$ is the fitted parameter in the linear regression.
+Equation 3, 2nd line was fit to 588 field measurements of $k_{o_2}$ while equation 3, third line was fit to 3,919 simulated $k_{o_2}$ values at 35 rivers.
+$$\begin{eqnarray} k_{o_2} &=& \alpha U_\star \mathbf{(3)} \\ &=& 47.67U_\star (r^2:0.53) \\ &=& 49.71U_\star (r^2: 0.76)  \end{eqnarray}$$
 
-There is a long tradition of water quality engineers and biogeochemists developing predictive models for different forms of *k*, going back to at least the 1950s [@wangPhysicallyBasedScaling2021].
-The goals of these models vary, from predicting oxygen reaeration coefficients to predicting $CO_2$ gas transfer velocities.
-Most of these models operate under the basic assumption that river channel hydraulics govern gas exchange and so simple statisical relationships can be generated between forms of *k* and hydraulic properties.
-However, many of them are not rooted in process-based mass transport theory and rather are simply empirical relations.
-This means that many of these models exist at the mercy of the limited training data available to fit said models and are highly sensitive to which data are used.
-
-Recently, many of these models were rexplored in rivers and streams by @wangPhysicallyBasedScaling2021.
-They significantly expanded training datasets of field-measured $k_{o_2}$ (*k* for dissolved oxygen DO at 20 degrees Celsius) by using inverse modeling of DO timeseries.
-This apprroach was briefly mentioned in the introduction of this manuscript.
-These methods calibrate $k_{o_2}$, along with respiration and primary production, to high-fidelity, in situ datasets of DO with the goal of minimizing the difference between the observed DO and the mass-balance DO [e.g.
-@applingOvercomingEquifinalityLeveraging2018; @chapraDeltaMethodEstimating1991].
-@wangPhysicallyBasedScaling2021 used this technique at 35 rivers across the United States and showed that $k_{o_2}$ models deduced from the classic, process-based 'thin-film' and 'surface renewal' theories yield approximately identical parameters when fit to field measurements or this new dataset of simulated $k_{o_2}$.
-This suggests that there are uniform scaling realtionships between certain hydraulic properties and $k_{o_2}$ that exist regardless of the training data used.
-See [@hallGasExchangeStreams2020 and @wangPhysicallyBasedScaling2021] for more in-depth explanations of the thin-film and surface renewal theories with respect to rivers.
-These are the standard models for understanding gas exchange in many aquatic environments from lakes to esutraries to rivers.
-The better fit, and more parsimonious, of these two models is reprinted as equations 2 and 3 where $U_\star$ is the shear velocity of the river channel [L/T] ($U_\star = \sqrt{gR_hS_h}$).
-Equation 2 was fit to field measurements of $k_{o_2}$ while equation 3 was fit to the simulated $k_{o_2}$ for 35 rivers.
-$$k_{o2} = 47.67U_\star \mathbf{(2)}$$
-$$k_{o2} = 49.71U_\star \mathbf{(3)}$$
-
-This model was implemented in BIKER (using 48 as the approximate parameter across equations 2 and 3) for a few reasons.
-First, the analysis performed by [@wangPhysicallyBasedScaling2021] provides compelling evidence that a statistical parameter of approximately 48 is uniform across rivers and streams, regardless of the training data used.
-This is encouraging for use within BIKER as BIKER is specifically designed to be as river-agnostic as possible.
+We implement this model within BIKER (using 48 as a reasonable value for $\alpha$ given equation 3) for a few reasons.
+First, as suggested earlier @wangPhysicallyBasedScaling2021 provides compelling evidence that an $\alpha$ of approximately 48 is uniform across rivers and streams, regardless of the training data used.
+This is encouraging for use within BIKER as BIKER is specifically designed to be as river-agnostic as possible and so we sought a fully generalized equation.
 Second, it yields a simple linear relationship across all rivers, regardless of their size or steepeness.
-This is not necessairly true of *k* models based on the turbulent dissipation rate *eD* ($eD=gS_hV$), which is another common approach [@hallGasExchangeStreams2020; e.g.
-@raymondScalingGasTransfer2012a; @oconnorMechanismReaerationNatural1958; @tsivoglouTracerMeasurementReaeration1976], where recent work has shown that *k* does not scale with $eD$ via the same statistical parameters across all rivers [@ulsethDistinctAirWater2019a].
-We further explored the relationship presented in that paper and found that the $k~eD$ relationship fundamentally breaks down in rivers with low *eD*, which corresponds to nearly all rivers that SWOT will observe (**supp figure**).
-Therefore, an $eD$ based predictive model is of little use in SWOT rivers.
-Third, we sought to explicitly predict the gas transfer velocity for DO, as its use extends beyond constraining GHG fluxes from rivers.
-For example **stuff here**.
-To convert from $k_o2$ to *k* for any other GHG (as will be done later in this manuscript to calculate $CO_2$ evasion fluxes), we use the Schmidt number to convert between water temperatures and gases of interest following equation 4.
-530 is the Schmidt number for oxygen at 20 degrees Celsius and *Sc* is the Schmidt number at one's desired temperature for some dissolved gas.
-$$k_{GHG, temp}=k_{02}*(Sc/530)^{-1/2} \mathbf{(4)}$$
+This is not necessairly true of other *k* models, for example those based on *eD* (section 1).
+Recent work has shown that *k* does not scale with *eD* via the same statistical parameters across all river geomorphologies [@ulsethDistinctAirWater2019a].
+We further explored the relationship presented in that paper and found that the $k \sim eD$ relationship fundamentally breaks down in rivers with low *eD*, which corresponds to nearly all rivers that SWOT will observe (**supp figure, need to do**).
+Therefore, an *eD* based predictive model is less useful in SWOT rivers.
+Third, we sought to explicitly predict $k_{o_2}$, as its use extends beyond constraining gas fluxes from rivers and into parameterizing river metabolism models (equation 2).
 
@@
 However, our goal of using BIKER on any river that is observable by SWOT necessitates that we pick an equation that is applicable across as many rivers as possible.
-Future work should look at using a river-specific *k* equation to 'calibrate' BIKER to a specific river of interest.
+Future work should look at using a river-specific *k* equation to pseudo-calibrate BIKER to a specific river.
 
-With a predictive model for $k_{o2}$ chosen, we now implement it within a McFLI framework.
-The approach used here is strongly informed by the @hagemannBAMBayesianAMHGManning2017a McFLI algorithm for ungauged RSQ, further explored in more recent work [@brinkerhoffConstrainingRemoteRiver2020].
-These papers conceptualize discharge as a Bayesian remote sensing problem, which we largely follow here to conceptualize $k_{o2}$ as a Bayesian remote sensing problem that can be solved using SWOT data.
+###2.2 Developing BIKER
+With a predictive model for $k_{o_2}$ chosen, we now implement it within the McFLI framework.
+The approach used here is informed by the @hagemannBAMBayesianAMHGManning2017a McFLI algorithm for ungauged RSQ, further explored in more recent work [@brinkerhoffConstrainingRemoteRiver2020a].
+These papers conceptualize discharge as a Bayesian remote sensing problem, which we largely follow here to conceptualize $k_{o_2}$ as a Bayesian remote sensing problem that can be solved using SWOT data.
 
-First, the model needs to be rewritten as a function of SWOT-observables river width *W* and height (and thus water surface slope $S_h$).
-SWOT will measure these at approximately 200m intervals in pre-defined mass-conserved river reaches (**What is the PDD?**; SWORD).
-To do this, we follow @hagemannBAMBayesianAMHGManning2017a to formulate depth via river width and area.
-Area is further split into the SWOT-observable portion *dA* and the unobservable portion $A_0$ following @durandEstimatingReachaveragedDischarge2014a.
-$A_0$ is the median channel area and *dA* is the change in channel area.
-We estimate *dA* by assuming a rectangular river channel so that $dA \approx \delta H*W$ and is also solely a function of SWOT-observables *W* and $S_h$.
+BIKER and Bayesian inference in general starts from Bayes rule, after acknowleding that it needs only to be specified as a proportionality to approximate the full posterior distribution (equation 4).
+In equation 4, $\Theta$ is some set of parameters we want to solve for (including $k_{o_2}$), *x* is the observed data, $f(x|\Theta)$ is the 'likelihood function' or sampling model conditional on the parameters, and $p(\Theta)$ is the joint prior distribution of the parameters.
+For BIKER, *x* is the SWOT-observables *W* and *H*.
+First, we detail the likelihood specification and then the prior specifications.
 
-**COLIN ABOUT DA: sensitivity analysis? "We investigated the sensitivity of BIKER to this rectangular assumption and found it negligible (xxx section x.x if appropriate xxx)".**
+$$p(\Theta|x) \propto f(x|\Theta)p(\Theta) \mathbf{(4)}$$
 
-$$k_{o2} = 48*U_\star=48\sqrt{gR_hS_h}=48g^{1/2}S_h^{1/2}(\frac{A_0+dA}{W})^{1/2} \mathbf{(5)}$$
-Next, we rework equation 5 into a Bayesian sampling model to have all of the observations (*W* and $S_h$) on the left-hand side sampled from all of the unknown model parameters ($A_0$ and $k_{o_2}$) on the right-hand side.
-This is equation 6, where *x* are SWOT observables given $\Theta$, a set of statistical parameters.
-Note that in this formal likelihood specification both sides of equation 6 are log-transformed (natural log) to produce the normal sampling model presented here, however they are written in equation 6 without that transformation for succinctness.
-The $\sigma_{k_{o2}}^2$ parameter refers to the uncertainty inherent in equation 6's estimates.
+####2.2.1 BIKER likelihood
+To concieve of equation 3 as a likelihood function, $k_{o_2}$ first needs to be written as a function of SWOT-observables *W* and *H* (which provides water surface slope $S_h$).
+This algebra is carried out in equation 5, where we assume that $R_h \approx \frac{A}{W}$ where *A* is wetted channel area.
+*A* is further split into the SWOT-observable portion *dA* and the unobservable portion $A_0$ following @durandEstimatingReachaveragedDischarge2014a.
+*dA* is estimated by assuming a rectangular river channel so that $dA \approx \delta H*W$.
+
+*COLIN COMMENT ABOUT DA: sensitivity analysis? "We investigated the sensitivity of BIKER to this rectangular assumption and found it negligible (xxx section x.x if appropriate xxx)".*
+
+$$\begin{eqnarray} k_{o_2} &=& 48\sqrt{gR_hS_h} \mathbf{(5)} \\ &=& 48g^{1/2}S_h^{1/2}(\frac{A_0+dA}{W})^{1/2} \end{eqnarray}$$
+
+Next, equation 5 is written as a Bayesian sampling model to have all of the SWOT observations on the left-hand side sampled from the unknown model parameters ($A_0$ and $k_{o_2}$) on the right-hand side.
+This is equation 6.
+The $\sigma_{k_{o2}}^2$ parameter refers to the uncertainty inherent in equation 5's estimates.
 This will be explained in detail below.
+Note that in the formal model specification, equation 6 is written as a normal distribution of its log-transformed quantities.
 
-$$f(x|\Theta) = \left[48^{-1}(\frac{W}{gS_h})^{1/2}\right] \sim normal\left(\left[k_{o2}^{-1}(A_0+dA)^{1/2}\right], \sigma^2_{k_{o2}}\right)  \mathbf{(6)}$$
+$$ f(x|\Theta) = \left[48^{-1}(\frac{W}{gS_h})^{1/2}\right] \sim lognormal\left(\left[k_{o2}^{-1}(A_0+dA)^{1/2}\right], \sigma^2_{k_{o2}}\right)  \mathbf{(6)}$$
 
-Equation six necessitates that Bayesian priors be assigned to parameters $A_0$ and $k_{o_2}$.
-Bayesian priors formalize the a priori estimates (and uncertainties) for the non-remotely-sensed parameters.
-To state that more intuitively for these physical quantities, priors represent our prior river knowledge of what $A_0$ and $k_{o_2}$ probably are for some river given that they cannot be remotely sensed.
-All priors are formalized within the model as truncated normal distributions of the log-transformed terms such that $log(X) \sim N(\mu, \sigma^2)$ for $\lambda < X < \epsilon$, using prior hyperparameters mean ($\mu$), standard deviation ($\sigma$), and upper ($\epsilon$) and lower bounds ($\lambda$) for parameter *X*.
-In order to avoid relying on in situ information, we assign prior hyperparameters using SWOT data only.
-$A_0$ prior hyperparameters were assigned following the method developed by @brinkerhoffConstrainingRemoteRiver2020.
-They developed a set of river channel prior hyperparameters for McFLI algorithms that are entirely RS-able and reflect differential channel hydraulics as a function of river geomorphology.
-They used an extensive database of field measurements and machine learning to identify patterns that associate the river width with river geomorphic properties, so these non-remotely sensible properties may be assigned to rivers using only the existing remotely sensed data.
+####2.2.2 Prior Specifications
+Equation 6 necessitates that we specify prior distributions for parameters $A_0$, $k_{o_2}$.
+Prior distributions formalize the a priori estimates (and uncertainties) for the non-remotely-sensed parameters.
+More intuitively, BIKER priors represent our 'prior river knowledge' of what $A_0$ and $k_{o_2}$ probably are for some river since they cannot be remotely sensed.
+All priors are formalized within the model as truncated normal distributions of the log-transformed terms such that $log(X) \sim N(\mu, \sigma^2)$ for $\lambda < X < \epsilon$, using prior hyperparameters mean ($\mu$), standard deviation ($\sigma$), and upper ($\epsilon$) and lower bounds ($\lambda$) for any parameter *X*.
+In order to avoid relying on in situ information, as is the goal of BIKER and SWOT, we assign prior hyperparameters using SWOT data only.
+$A_0$ prior hyperparameters were assigned following the method developed by @brinkerhoffConstrainingRemoteRiver2020a They developed a set of river channel prior hyperparameters for McFLI algorithms that are entirely RS-able and reflect differential channel hydraulics as a function of river geomorphology.
+They used an extensive database of field measurements and machine learning to identify patterns that associate the river width with river geomorphic properties, so prior hyperparameters may be assigned to rivers using only the existing remotely sensed data.
 This leaves the $k_{o_2}$ hyperparameters to be defined.
-We assigned the $k_{o_2}$ prior hyperparameters using a simple slope regression model trained on the @raymondScalingGasTransfer2012a dataset to assign $\mu$ (equation 7).
-$\sigma^2$ was set to the standard error of equation 7: 0.77.
+We assigned those using a simple slope regression model trained on an available subset of the @raymondScalingGasTransfer2012a dataset of field-measured $k_{o_2}$: equation 7 assigns $\mu$ while $\sigma^2$ was set to equation 7's standard error: 0.77.
 $\lambda$ and $\epsilon$ were set to log(0.0001) and log(60), respectively.
@@
 
-**FIGURE OUT Finally, we estimate $σ_{k600}^2$ using Monte Carlo (MC) methods to approximate total uncertainty from equation 3.
-Uncertainty in equation 3 can stem from 1) $a_i$ and $b_i$ parameter uncertainty, 2) error in Manning’s approximation of the average flow velocity, and 3) measurement error.
-Total equation 3 uncertainty is a function of all three uncertainity sources, so we need to propogate the three uncertanties through the algebra of equation 3.
-Here, we use MC simulations to do this.
-MC simulations repeately sample from parameter distributions defined by their individual expectations and uncertainities to produce a distribution of model estimates from which an overall uncertainity term can be extracted.
-Here, we run 5,000 unique MC simulations on 5,000 sets of field measurements of river channel hydraulics from @brinkerhoffReconcilingAtaStationAtManyStations2019 and use the median uncertainity term across those 5,000 distributions as $\sigma_{k600}$.
-Each MC simulation is itself 10,000 runs, sampling from the normal distributions for $a_i$, $b_i$, the log-transformed flow velocity, and $S_h$.
-Uncertainty for velocity followed @hagemannBAMBayesianAMHGManning2017a, who estimated ln-transformed Manning’s equation error in discharge to be 0.25 (we assume that the 0.25 value is due soley from Manning's equation and not measurements of cross-sectional area in order to use here as the velocity error).
-The slope-error model that we used is explained in detail in Section 2.3 and relies on a recent analysis of expected errors in SWOT $S_h$ measurements [@durandHowWillRadar2020a].**
+Finally, we estimate $\sigma_{k_{o_2}}^2$ using Monte Carlo (MC) methods to approximate total uncertainty in equation 5's estimates.
+Uncertainty in equation 5 can stem from 1) parameter uncertainty in $\alpha$ or 2) error in the assumption that $R_h \approx \frac{A}{W}$.
+Total equation 5 uncertainty is a function of both uncertainity sources, so we need to propogate those uncertanties through the algebra of equation 5.
+We use MC simulations to do this.
+MC simulations repeately sample from parameter distributions defined by their individual means and uncertainities to produce a distribution of model predictions from which an uncertainity term can be extracted.
+See **supp figure** for a flowchart of the entire methodology, but in short we run 8,000 different MC simulations on 8,000 sets of field measurements of river channel hydraulics taken from the @brinkerhoffReconcilingAtaStationAtManyStations2019 dataset and use the median uncertainity term across those 8,000 distributions as $\sigma_{k_{o_2}}$.
+Each of the 8,000 simulations are themselves 10,000 runs, sampling from the normal distributions for $\alpha$ and $R_h$.
+We defined the distribution of the $R_h$ prediction again using the @brinkerhoffReconcilingAtaStationAtManyStations2019 dataset (n = 500,000+): we fit a linear regression between $R_h$ and $\frac{A}{W}$ and use that model's log-transformed standard error as the uncertainity in the MC simulations.
+The resulting value was 0.15 and we further inflated it to 0.20 to account for potential errors in assuming $dA \approx \delta H*W$.
 
-With the sampling model (equation 4), prior distributions, and the $\sigma_{k_{o_2}}^2$ parameter described, a joint posterior distribution conditional on the SWOT data is obtained.
-To approximate this distribution, we use a Markov Chain Monte Carlo (MCMC) algorithm.
-Because it is written in the Stan probabilistic programming language, we use a Hamiltonian Monte Carlo which reduces computation time relative to other sampling algorithms [again following @hagemannBAMBayesianAMHGManning2017a].
+With the sampling model (equation 6), prior distributions, and the $\sigma_{k_{o_2}}^2$ parameter described, a joint posterior distribution conditional on the SWOT observations is specified.
+To approximate this distribution, we use a Markov Chain Monte Carlo (MCMC) algorithm implemented using the Stan probabilistic programming language.
+Specifically, BIKER uses a Hamiltonian Monte Carlo sampler which reduces computation time relative to other sampling algorithms.
 
 ###2.3 BIKER validation setup
-We validated BIKER on 50 rivers using observed $R_h$ and observed $S_h$ First, we detail the 50 rivers, and then we detail how observed $k_{o_2}$ is calculated.
+We validated BIKER on 49 rivers using observed $R_h$ and observed $S_h$ First, we detail the 49 rivers, and then we detail how observed $k_{o_2}$ is 'measured'.
 
 Because SWOT has yet to launch, it is standard practice to benchmark SWOT-related algorithms on SWOT-like data.
-There are three current types of SWOT-like data: 1) AirSWOT, which is an airborne Ka-band inSAR currently limited to five rivers globally, 2) simulated rivers that mimic the type of data SWOT will provide, and 3) the SWOT simulator, which introduces measurement errors to these simulated rivers.
+There are three current types of SWOT-like data: 1) AirSWOT, which is an airborne Ka-band inSAR currently limited to five rivers globally [e.g.
+of its use: @tuozzoloEstimatingRiverDischarge2019], 2) simulated rivers that mimic the type of data SWOT will provide, and 3) the SWOT simulator, which introduces measurement errors to these simulated rivers [e.g.
+of its use: @oubanasDischargeEstimationUngauged2018].
 Because we are principally interested in algorithm performance, we limit our validation setup to simulated rivers in order to benchmark across as many rivers as possible.
-These simulated rivers are simply reach-averaged hydraulic model outputs where the water surface heights and widths are labelled as RS observations and are used as the sole inputs to the BIKER algorithm.
-Simulated rivers mimic perfect measurement conditions (i.e.
-no measurement errors).
-Here, we use the 50 rivers developed by @frassonExploringFactorsControlling and @durandIntercomparisonRemoteSensing2016b.
+These simulated rivers are simply reach-averaged hydraulic model outputs where the water surface heights and widths are labelled as RS observations and are used as the sole inputs to BIKER.
+Here, we use the 49 rivers collected by @frassonExploringFactorsControlling2021 and @durandIntercomparisonRemoteSensing2016b.
 These are the two benchmarking studies that have explored RSQ algorithm performance for the SWOT mission.
-Consult both of those papers for maps of the river's locations, though the 50 rivers are spread across the United States, Canada, France, Italy, United Kingdom, India, and Bangladesh.
-We omitted the Arial Khan river in Bangladesh due to known problems with that hydraulic model **is that sufficient to say?**.
+Please consult both of those papers for all of the the rivers' locations as well as hydraulic model specifications.
+Ultimately, the 49 rivers are spread across the United States, France, Italy, the United Kingdom, and Bangladesh.
+We omit both models for the Saint Lawrence River, Canada from @durandIntercomparisonRemoteSensing2016b because they lack enough hydraulic information to calculate $U_\star$.
 
@@
 This will occur when the landscape slope is steeper than the radar's incidence angle [@durandHowWillRadar2020a] and so generally occurs in areas of high topographic variation [@reesPhysicalPrinciplesRemote2013].
-
-For river heights and slopes, we use the SWOT error model devolped by @durandHowWillRadar2020a and implemented on 16 of the rivers in this study by @frassonExploringFactorsControlling.
-Systematic height errors are assumed constant across the river but variable in time.
-Width errors were derived using a model built by @frassonExploringFactorsControlling based on SWOT-simulator runs for the Sacramento and Po rivers.
-Width errors due to poor water classification are ignored here.
-Please consult @frassonExploringFactorsControlling for the specifics of how these realistic errors were incorporated into the hydraulic models.
-Ultimately, we reran BIKER on these 16 rivers with the SWOT error model corrupting the SWOT-observables *W* and *S_h*.
+For river heights and slopes, we use the error model devolped by @durandHowWillRadar2020a and implemented on 17/49 of the rivers by @frassonExploringFactorsControlling2021.
+Width errors were derived using a model built by @frassonExploringFactorsControlling2021 based on SWOT-simulator runs for the Sacramento river, California [@frassonAutomatedRiverReach2017b] and the Po river, Italy [@domeneghettiCharacterizingWaterSurface2018].
+Those two simulations allowed for explicit characterization of width errors that were then extrapolated onto the other 15 rivers.
+Width errors due to poor water classification are ignored as they were by @frassonExploringFactorsControlling2021.
+Please consult @frassonExploringFactorsControlling2021 for the specifics of how these realistic errors were incorporated into the hydraulic models.
+Ultimately, we reran BIKER on these 17 rivers with the SWOT error model corrupting the SWOT-observables *W* and *H*.
 
 Regardless of the validation data or SWOT error budget used, we do not have observed $k_{o_2}$ data for these rivers, and to our knowledge no field dataset of $k_{o_2}$ exists in the type of temporal and spatial frequency that SWOT (and therefore the BIKER algorithm) provides.
-Further, we are principally interested in BIKER's ability to reproduce $k_{o_2}$ from SWOT observations and are less concerned with the actual accuracy of the predictive model itself [which can be validated using existing datasets like previously done by @wangPhysicallyBasedScaling2021].
+Further, we are principally interested in BIKER's ability to reproduce $k_{o_2}$ from SWOT observations and are less concerned with the actual accuracy of the predictive model itself, which can be validated using existing datasets as has been previously done [e.g.
+@wangPhysicallyBasedScaling2021; @ulsethDistinctAirWater2019a; @raymondScalingGasTransfer2012a].
 Therefore, we take the model outlined in section 2.1 and use that to calculate the observed $k_{o_2}$ that BIKER is validated against.
 This is done using equation 8, where $R_{h, obs} [L/T]$ is observed river channel area divided by observed channel width.
-With this setup, we can directly explore BIKER's ability to infer observed $U_\star$ and $k_{o_2}$ from *W* and $S_h$ alone.
-It also means that, for a fair validation scheme, $\sigma_{upscale}^2$ must be set to reflect only error from Manning's equation (and not the parameter 48 in equation 8).
-Thus, $\sigma_{upscale}^2$ is set to 0.25 for this validation.
-However in practice, it should reflect the total uncertainty calculated in section 2.2 from all three sources.
+With this setup, we can directly explore BIKER's ability to infer observed $U_\star$ and $k_{o_2}$ from *W* and *H* alone.
+It also means that, for a fair validation scheme, $\sigma_{k_{o_2}}^2$ must be set to reflect only error from our assumptions about $R_h$ and not the parameter uncertainity inherent in $\alpha$ (the 48 in equation 8).
+Thus, $\sigma_{k_{o_2}}^2$ is set to 0.20 for this validation as suggested in section 2.2.
+However once SWOT launches, it should reflect the total uncertainty calculated in section 2.2 from both sources.
 
@@
 
+Validation is performed using the BIKER posterior means.
 Validation metrics take two forms (and are detailed in Table 1).
@@
 ###2.4 Upscaling to CO2 evasion and bulk carbon efflux
-It is one thing to to predict $k_{o_2}$, but researchers are usually more interested in the gas fluxes from rivers and ultimately the bulk carbon efflux from river to atmosphere.
-Therefore, we explored 1) the ability of BIKER to reproduce hypothetical carbon dioxide evasion fluxes ($FCO_2$ $[M/L^2T]$) from these 22 rivers, and 2) the sensitivity of the estimated bulk carbon efflux $[M/T]$ via $CO_2$ evasion to the method used to estimate reach averaged flow velocity.
+It is one thing to accurately predict $k_{o_2}$, but researchers are often most interested in the actual gas fluxes from rivers and ultimately the carbon emitted from river to atmosphere.
+Therefore, we explore 1) the ability of BIKER to reproduce carbon dioxide evasion fluxes ($FCO_2$ $[M/L^2T]$) from these 49 rivers, and 2) the sensitivity of the carbon efflux efflux $[M/T]$ due to the chosen method to approximate $R_h$ and ultiamtely $U_\star$.
 
-First, we calculated $FCO_2$.
-To do this, we paired the modeled $k_{CO2}$ values (obtained from $k_{600}$) with field-measurements of water-side $CO_2$ concentrations and water temperatures.
-26 bi-weekly samples were made by @beaulieuControlsGasTransfer2012a- Figure S1) at one location in the Ohio River for one calender year from 2008-2009.
-We paired these 26 values with each 11-day-repeat set of SWOT observations by date, ignoring the timesteps beyond 26 (only ~15% of the SWOT observations were ignored here and we deemed this acceptable).
+First, we calculate $FCO_2$.
+To do this, we pair the modeled $k_{CO_2}$ values (obtained from $k_{O_2}$ and equation 9) with field-measurements of water-side $CO_2$ concentrations and water temperatures.
+In equation 9, 530 is the Schmidt number for oxygen at 20 degrees Celsius (which BIKER predicts) and *Sc* is the Schmidt number at one's desired temperature for some dissolved gas.
+26 bi-weekly dissolved $CO_2$ samples were made by @beaulieuControlsGasTransfer2012a (Figure S1) at one location in the Ohio River for one calender year from 2008-2009.
+We paired these 26 values with every 11th SWOT observation by date, ignoring the timesteps beyond 26 (only ~15% of the SWOT observations were ignored here and we deemed this acceptable).
+We chose to sample every 11 days as this is the average sampling resolution for SWOT.
 Not all of the SWOT rivers have observations for a full year, and when simulation dates were not available they were assumed to start on January 1st.
-Note that this $CO_2$ data is for the Ohio River only but was applied to all 22 rivers (which includes multiple sections of the Ohio River- @frassonCompilationHydraulicModels2019a).
+Note that this $CO_2$ data is for the Ohio River only but was applied to all 49 rivers (which includes multiple sections of the Ohio River).
 Because we are exclusively interested in the relative differences between $FCO_2$ estimates and not the raw fluxes themselves, any $CO_2$ data representative of SWOT-observable rivers was deemed acceptable.
 Atmospheric $CO_2$ was assumed 390 uatm.
-The Schmidt number, used to backcalculate $k_{CO2}$ from $k_{600}$, was calculated following @raymondScalingGasTransfer2012a and @wanninkhofRelationshipWindSpeed1992a.
+The Schmidt number, used in equation 9, was calculated following @raymondScalingGasTransfer2012a and @wanninkhofRelationshipWindSpeed1992a.
 
-Then, we estimated bulk carbon efflux using four models for average flow velocity:  BIKER and three gauge-based HG models previously used for $CO_2$ upscaling [@lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a; @raymondScalingGasTransfer2012a].
+$$k_{GHG, temp}=k_{02}*(Sc/530)^{-1/2} \mathbf{(9)}$$
+
+Next, we estimate bulk carbon efflux using four models for average channel depth (used to calculate $U_\star$):  BIKER's posterior means and three gauge-based HG models previously used for $CO_2$ upscaling [@lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a; @raymondScalingGasTransfer2012a].
 See Table S1 for their definitions.
-It is worth stressing that the @lauerwaldSpatialPatternsCO22015 model was developed as one of two components of the @raymondGlobalCarbonDioxide2013a model, and so was actually developed by the latter’s authors.
+It is worth stressing that the @lauerwaldSpatialPatternsCO22015 model is simply one of two components of the @raymondGlobalCarbonDioxide2013a model, and so was actually developed by the latter's authors.
 The names used here refer to the specific implementation used in each study.
 For this study, all HG methods use the in-situ discharge record while BIKER does not.
-This allows us to explore the relative differences in bulk carbon efflux estimates if implementing a wholly ungauged method (BIKER) versus gauged methods (all three HG models).
-Finally, we express the bulk carbon efflux as the average mass flow rate of carbon (via $CO_2$ evasion) per year from the 22 rivers after accounting for total river surface area.
+This allows us to assess whether BIKER's estimates (wholly unguaged) are comparable to gauged methods (all three HG models).
+Finally, we express the bulk carbon efflux as the average mass flow rate of carbon (via $CO_2$ evasion) per year from the 49 rivers after accounting for total river surface area.
 
 ##3 Results
-First, we present the results from the $k_{600}$ upscaling experiment (section 3.1).
-We then take the best performing upscaling function, implement it within BIKER, and validate it on the 22 SWOT rivers (section 3.2).
-Finally, we use $k_{600}$ as modeled by both BIKER and gauge-based HG models to compare $FCO_2$ and bulk carbon efflux estimates made using gauged and ungauged methods (section 3.3).
+First, we present the results from the BIKER validation on 49 SWOT rivers (section 3.1).
+Next, we compare BIKER inversion errors to the errors inherent in equations 2 and 3, as well as the rivers' hydraulic properties (section 3.2).
+Finally, we use 4 different models for $k_{CO_2}$ to compare $FCO_2$ and bulk carbon efflux estimates made using gauged and ungauged methods (section 3.3).
 
-###3.2 Validating the BIKER algorithm
-Figure 5a plots the validation results for $k_{600}$ (with no SWOT measurement error) across all 22 rivers and all timesteps.
-For BIKER, the points are the posterior means while the black lines are the 95% confidence intervals (CIs) for the predictions.
-$k_{600}$ is strongly correlated with the $k_{600}$ predicted by BIKER.
+###3.1 Validating the BIKER algorithm
+Figure 2a plots the validation results for $k_{o_2}$ (with no SWOT measurement error) across all 49 rivers and all timesteps.
+The points are the posterior means while the black lines are the 95% confidence intervals (CIs) for the predictions.
+$k_{o_2}$ is strongly correlated with the $k_{o_2}$ predicted by BIKER.
 BIKER captures the general magnitude of the predictions and most points fall on or near the 1:1 line.
-However, the 95% prediction intervals (dashed grey lines) highlight reasonably large residuals for many predictions.
-Regardless, the RMSE for the BIKER predictions is only `r round(results_3_2_all$rmse,2)` in log10 space (~`r round(10^results_3_2_all$rmse,2)` m/day) across all predictions.
-Again, this is substantially less than the RMSEs for the upscaling models (Figure 3) and the MC-propogated uncertainties (Figure 4).
+However, the 95% prediction intervals (dashed grey lines) highlight a slight underestimation bias for many predictions and a handful of predictions that are significanyl underestimated.
+The RMSE for the BIKER predictions is only `r round(results_3_2_all$rmse,2)` m/day) across all predictions.
 
-![Figure 5.
+![Figure 2.
 a: Validation of BIKER for 22 SWOT rivers.
@@
 
-Figure 5a suggests a flip in the bias of the $k_{600}$ predictions, which is confirmed in Figure 5b.
-Figure 5b plots the cumulative density functions (CDFs) of observed and predicted $k_{600}$ where the green line is the CDF of the BIKER posterior means and the blue lines are the CDFs of the BIKER posterior 95% CIs.
-We see, more clearly than in Figure 5a, that the posterior mean is systematically underestimated for values less than the median $k_{600}$ and systemically overestimated for values above the median $k_{600}$.
-However, the uncertainty in these estimates captures $k_{600}$ very well: nearly the entire observed CDF falls between, on, or just outside of the 95% CIs (Figure 5b).
-This highlights one benefit of using Bayesian inference to fully propagate all prior uncertainties through to the posterior.
-In summary, Figure 5 confirms that we improve upon our baseline understanding of $k_{600}$ in these rivers.
-Put another way, we reasonably capture $k_{600}$ with no in situ information about the river while simultaneously and explicitly accounting for the reasonably large uncertainties inherent in our $k_{600}$ estimates.
+Figure 2a highlights a handful of values that are grossly underestimated by BIKER, which is confirmed in Figure 2b.
+Figure 2b plots the cumulative density functions (CDFs) of observed and predicted $k_{o_2}$ where the green line is the CDF of the BIKER posterior means, the blue lines are the CDFs of the BIKER posterior 95% CIs, and the dashed black line is the CDF of observed $k_{o_2}$.
+We see, more clearly than in Figure 2a, that BIKER captures $k_{o_2}$ very well: most of the  observed CDF falls between the 95% CIs or on the posterior mean (Figure 2b).
+If we define a 'hit rate' as the oberved value following between the CIs, BIKER yields a hit rate of **get hit rate**.
+Most of the  **get inverse of hit rate**% are extremes: the uppermost quartile is systematically overestimated by BIKER, and a handful of the smallest $k_{o_2}$ in this data are underestimated.
+This highlights one benefit of using Bayesian inference to fully propagate prior and model uncertainties through to the posterior.
+In summary, Figure 2 confirms that we improve upon our baseline understanding of $k_{o_2}$ in these river: we accurately capture $k_{o_2}$ with no in situ information about the river while simultaneously and explicitly accounting for the uncertainties inherent in our $k_{o_2}$ estimates.
 
-Figure 6a plots validation metrics calculated for each river with and without SWOT measurement error.
-The boxplots are composed of scores for the 22 rivers- see Table 1 for metric definitions.
-SWOT measurement uncertainties slightly degrade performance in KGE, neglibily degrade rBIAS and RRMSE, and actually improve performance in NRMSE (Figure 6a).
-Aside from the modest decrease in KGE, we deem that SWOT measurement error does not exert a significant influence on BIKER and so the results presented for the rest of the manuscript assume no measurement error.
-Median rBIAS is `r round(median(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$rBIAS, na.rm=T),2)`, indicating nearly no bias in most rivers’ predictions.
-However, some rivers are substantially biased in both directions, further supporting the visual evidence in Figure 5 that sometimes BIKER is substantially under/overestimating the magnitude of $k_{600}$ and that this is river-specific.
-Median KGE is `r round(median(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$kge, na.rm=T),2)`, which is excellent given that absolutely no in situ information is being used to predict $k_{600}$.
+Figure 3a plots validation metrics calculated for each river with and without SWOT measurement error (green and purple, respectively).
+The boxplots are composed of scores for either the 49 or 17 rivers- see Table 1 for metric definitions.
+SWOT measurement uncertainties slightly degrade performance across all four error metrics (Figure 3a), though caution should be used in over-interepting boxplots with n = 17.
+Therefore, we deem that SWOT measurement error does not exert a significant influence on BIKER and so the results presented for the rest of the manuscript assume no measurement error in order to use all 49 rivers.
+
+Median KGE is `r round(median(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$kge, na.rm=T),2)`, which is excellent given that absolutely no in situ information is being used to predict $k_{o_2}$.
+`r sum(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$kge > -0.41, na.rm=T)`/49 rivers score a KGE > -0.41 and `r sum(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$kge > 0, na.rm=T)`/49 are greater than 0, indicating that BIKER performs better than an estimate of the mean $k_{0_2}$.
+Median rBIAS is `r round(median(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$rBIAS, na.rm=T),2)`, indicating nearly no bias in most rivers' predictions.
+However, some rivers are substantially biased in both directions (particuarly positive bias, where `r sum(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$rBIAS > 1, na.rm=T)` rivers have scores > 1).
+This further supports the visual evidence in Figure 2 that sometimes BIKER is substantially under/overestimating the magnitude of $k_{o_2}$ and that this is river-specific.
 NRMSE and RMSE have median scores of `r round(median(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$nrmse, na.rm=T),2)` and `r round(median(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$rrmse, na.rm=T),2)`, respectively.
@@
 
-![Figure 6.
+![Figure 3.
 a: Performance metrics by river.
@@
 Dashed lines denote scores of 1, 0, and -0.41 for KGE (section 3.2).
-b-d: validation timeseries for three rivers representative of good, reasonable, and poor performance.
+b-d: validation timeseries for three rivers representative of good, reasonable, and poor BIKER performance.
 b) was randomly selected from the upper tertile of KGE scores, c) was randomly selected from the middle tertile, and d) from the worst tertile.
@@
 
-Figure 6b-d are representative timeseries plots of predicted and observed $k_{600}$ for three rivers chosen randomly from those with ‘good’ KGE scores (b), ‘okay’ KGE scores (c), and ‘bad’ KGE scores (d).
-See the Figure 6 caption for how this was calculated.
-For the Kanawha River, the entire timeseries of $k_{600}$ is correctly predicted, while in the Ohio Section 2 River there is a negative bias in the estimates.
-However, temporal dynamics are still correctly recovered.
-In the downstream Mississippi River, there is significant bias in the estimates as well as significant uncertainty (per the 95% CIs).
-The temporal dynamics are also incorrectly predicted at times.
+Figure 3b-d are representative timeseries plots of predicted and observed $k_{o_2}$ for three rivers chosen randomly from those with 'good' KGE scores (b), 'okay' KGE scores (c), and 'bad' KGE scores (d).
+See the Figure 3 caption for how this was determined.
+For the Missouri Downstream River, the entire timeseries of $k_{o_2}$ is correctly predicted, while in the Ohio Section 8 River there is a positive bias in the later estimates.
+Interestingly, there is near perfect recovery of $k_{o_2}$ for the first approximately 150 days.
+In the Connecticut River, there is significant positive bias in the estimates as well as large uncertainty (per the 95% CIs) and temporal dynamics are largely missing from BIKER's predicitions.
+
+###3.2 BIKER performance versus *k* model performance
+Next, we compare error introduced by BIKER's Bayesian inversion to error introduced by the use of $\alpha$ in equations 2 and 3.
+Recall that the BIKER validation has already had parameter error accounted for (section 2.3) and so BIKER errors reflect error introduced only from Bayesian inference and not from the use $\alpha$.
+Figure 4 compares BIKER's performance per river (via NRMSE) to the reported cross-validated model performance for equations 2 and 3 [@wangPhysicallyBasedScaling2021].
+Note that in that paper they use the same metrics but refer to it as 'relative error' and so we do too in Figure 4.
+43/49 rivers have lower inversion errors than the cross-validated error in equations 2 and 3 (~57%), with most rivers' scores far less than 57%.
+Recall that the median NRMSE is `r round(median(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$nrmse, na.rm=T),2)` which is less than 1/2 of the reported error for $\alpha$.
+For the six rivers which inversion errors greater than the parameter error, 4/6 are approaching or over 100% error.
+This means that BIKER fundamentally does not work in these rivers, while in all the other rivers the inversion errors are nearly always much less than the parameter error.
+It should also be stressed that NRMSE is one of the worst perfoming metrics for BIKER (figure 3a) and so a similar comparison using KGE or rBIAS would likely yield even stronger results.
+Therefore, BIKER introduces a small amount of additional error into the physical model for $k_{o_2}$ and any errors in the final $k_{o_2}$ estimates are dominated by process-level uncertainity in predicting $k_{o_2}$ rather than uncertainity in inverting SWOT observations to estimate $k_{o_2}$.
+
+![Figure 4: NRMSE (referred to as relative error in @wangPhysicallyBasedScaling2021.) for all 49 rivers.
+Dashed line is the cross-validation relative error for the ko2 model used in this study (~57%), reprinted from @wangPhysicallyBasedScaling2021.
+We see that errors introduced by BIKER are nearly always much less than prediction uncertainity stemming from the ko2 model itself.
+NRMSE is also one of the worst perfoming metrics for BIKER (figure 3a).](cache/validation/relativeErr.jpg)
+
+Figure 4 shows that BIKER performs very poorly on a small subset of these 49 rivers.
+So, we also compared BIKER performance against river hydraulic properties, specifically mean observed $k_{o_2}$, *W*.
+and $S_h$.
+We found no clear patterns between these properties and KGE, NRMSE, or RRMSE (Figures Sx, Sy, Sz respectively).
+However, Figure 5 plots the patterns betwwen these hydraulic properties and rBIAS.
+There is an apparent uniform and positive bias in BIKER's predicitions in the narrowest and steepest of these 49 rivers.
+For these 49 rivers, these are approximately those with slopes > 0.001 (Figure 5c) and widths < 100m (Figure 5b).
+Following basic gas exchange theory [@ulsethDistinctAirWater2019a], these are also the rivers with the greatest $k_{o_2}$ values (Figure 5a).
+
+![Figure 5: BIKER performance (defined by rBIAS- Table 1) versus (a) mean river ko2, (b) mean river width, and (c) mean river slope.
+An rBIAS of zero (dashed black line) indicates no bias in the predictions.](cache/validation/riverProperties_rbias.jpg)
 
 ###3.3 Influences on carbon efflux estimates
-Finally, we explore our ability to 1) use BIKER-produced $k_{600}$ to estimate $FCO_2$ and 2) estimate bulk carbon efflux from $CO_2$ evasion (section 2.4).
+Finally, we explore our ability to 1) use BIKER-produced $k_{o_2}$ to estimate $FCO_2$ and 2) estimate bulk carbon efflux from $CO_2$ evasion (section 2.4).
 
-In Figure 7a, there is a strong fit to the observed $FCO_2$ data, with an RMSE of `r round(10^results_3_3_all$rmse,2)` $g/m^2dy$.
-The $r^2$ is slightly higher than $k_{600}$ (Figure 7a).
-There is also less underestimation bias for the smallest of the observed $FCO_2$ than in the $k_{600}$ predictions (Figure 5a).
-This is presumably due to the structure of the $FCO_2$ equation, which reduces the relative importance of errors in the gas exchange values given that the $CO_2$ data is measured in situ.
-$FCO_2$ prediction intervals also have a similar relative magnitude to those presented in Figure 5.
-Figure 7b-d includes subplots for the same rivers as Figure 4b-d, however with $FCO_2$ plotted instead of $k_{600}$.
-There is functionally perfect recovery of $FCO_2$ in the Kanawha River, very good recovery (with some positive bias in peak evasion events) for the Ohio Section 2 River, and far less negative bias in the Downstream Mississippi river than is present in the $k_{600}$ estimates (Figure 5b).
-Again, the temporal dynamics are accurately modeled in the Kanawha and Ohio Rivers, however there is some bias in the magnitude of the predictions.
-This suggests that the BIKER model can easily estimate the changing dynamics of gas exchange but is at the mercy of the prior on $k_{600}$ to accurately estimate the magnitude of the evasion.
+In Figure 6a, there is a strong fit to the observed $FCO_2$ data, with an RMSE of `r round(10^results_3_3_all$rmse,2)` $g/m^2dy$.
+The $r^2$ is slightly lower than $k_{o_2}$ (Figure 6a), likely due the the handful of extremely underestimated $FCO_2$ predictions.
+Overall, there is less systematic bias in the predicitons across all 49 rivers than in the $k_{o_2}$ predictions (Figure 3a).
+This is presumably due to the structure of the $FCO_2$ equation, which reduces the relative importance of errors in *k* given that the $CO_2$ data is measured in situ.
+$FCO_2$ prediction intervals are wider than those presented in Figure 2a.
+Figure 6b-d includes subplots for the same rivers as Figure 3b-d, however with $FCO_2$ plotted instead of $k_{o_2}$.
+There is very good recovery of $FCO_2$ in both the Missouri Downstream River and the Ohio Section 8 river (Figures 6b and 6c), with both magnitude and temporal dynamics modeled quite well.
+The Connecticut river is systematically over estimated (Figure 6d), though BIKER's CIs do reflect reasonably uncertain estimates and the temporal dynamics are not far from the observed dynamics.
+This suggests that BIKER can easily estimate the changing dynamics of gas exchange but is at the mercy of the prior on $k_{o_2}$ to accurately estimate the magnitude of the evasion.
+
+![Figure 6.
+a: FCO2 via BIKER versus via equations 2 and 3 for every 11th timestep for the 49 rivers (grey lines are linear regression and 95% prediction intervals, while black dashed line is the 1:1 line).
+b-d: timeseries plots for the three example rivers from Figure 3b, 3c, and 3d.](cache/FCO2/FCO2_plot.jpg)
+
+Finally, we compare $FCO_2$ and bulk carbon efflux (via $CO_2$ evasion) from the 49 rivers using BIKER posterior means and three gauge-based HG models (Figure 7).
+Figure 7a are barplots of the bulk carbon efflux (via $CO_2$ evasion) across the 49 rivers in gigagrams of carbon per year.
+The BIKER bulk carbon efflux (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_BIKER',]$sumFCO2, 0)` gG-C/yr) is nearly identical to that predicted by the 'raymond 2013' model(`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_raymond2013',]$sumFCO2, 0)`), though both are somewhat overestimated relative to the observed flux (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_ulset',]$sumFCO2, 0)`).
+The 'raymond 2012' model is the closest (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_raymond2012',]$sumFCO2, 0)`), while the 'Lauerwald 2015' model grossly overestimates this bulk efflux (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_Lauerwald2015',]$sumFCO2, 0)`).
+Thus, despite BIKER using absolutely no in situ data, it provides similar estimates of the carbon efflux to two of the in-situ approaches and surprisingly outperforms the 3rd in-situ method (Figure 7a).
 
 ![Figure 7.
-a: FCO2 from BIKER versus using observed average flow velocity for all timesteps for 22 SWOT rivers (grey lines are linear regression and 95% prediction intervals, while black dashed line is the 1:1 line).
-b-d: timeseries plots for the three example rivers from Figure 5.6.](cache/FCO2/FCO2_plot.jpg)
-
-We then compare the $FCO_2$ and bulk carbon efflux (via $CO_2$ evasion) from the 22 rivers using BIKER posterior means and three gauge-based HG models (Figure 8).
-Figure 8a are barplots of the bulk carbon efflux (via $CO_2$ evasion) across all rivers in gigagrams of carbon per year.
-The BIKER bulk carbon efflux (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_BIKER',]$sumFCO2, 0)` gG-C/yr) is slightly overestimated relative to to the bulk carbon efflux calculated using observed velocity (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_ulset',]$sumFCO2, 0)` gG-C/yr).
-However, it is a better estimate than those using a streamgauge and HG (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_raymond2012',]$sumFCO2, 0)`, `r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_raymond2013',]$sumFCO2, 0)`, and `r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_Lauerwald2015',]$sumFCO2, 0)` gG-C/yr for ‘Raymond 2012’, ‘Raymond 2013’, and ‘Lauerwald 2015’, respectively).
-Thus, BIKER reasonably captures the estimated efflux from these rivers purely from RS data, and surprisingly outperforms the gauge-based approaches when assessing this bulk flux property (Figure 8a).
-
-![Figure 8.
 a: Bulk carbon efflux, per year, from the 22 SWOT rivers as calculated using four different average flow velocity estimates: 1) BIKER, 2) through 4) average flow velocity as calculated using HG models from the literature.
@@
 
-Figure 8b plots the by-river performance scores for $FCO_2$.
-For KGE, 'Lauerwald 2015' is the best performing (median KGE: `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_Lauerwald2015',]$kge, na.rm=T),2)`) while BIKER (median KGE: `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_BIKER',]$kge, na.rm=T),2)`) slightly outperforms the other two gauge-based methods (median KGEs: `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_raymond2013',]$kge, na.rm=T),2)` and `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_raymond2012',]$kge, na.rm=T),2)` for 'Raymond 2013' and 'Raymond 2012', respectively).
-However, all four scores are very similar.
+Figure 7b plots the by-river performance scores for $FCO_2$.
+In line with the Figure 7a results, 'Lauerwald 2015' is the worst performing (median KGE: `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_Lauerwald2015',]$kge, na.rm=T),2)`) while BIKER (median KGE: `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_BIKER',]$kge, na.rm=T),2)`) is only slightly worse than the other two in-situ methods (median KGEs: `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_raymond2013',]$kge, na.rm=T),2)` and `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_raymond2012',]$kge, na.rm=T),2)` for 'Raymond 2013' and 'Raymond 2012', respectively).
 For rBIAS, BIKER indicates the least bias in some rivers and similar bias to the other models in other rivers, as indicated by the scores' inter-quartile range (`r round(IQR(results_3_3_rivs[results_3_3_rivs$key=='FCO2_BIKER',]$rBIAS, na.rm=T),2)`) and median (`r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_BIKER',]$rBIAS, na.rm=T),2)`).
-Meanwhile, all three HG models uniformly produce significant negative bias in their estimates (median scores of `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_Lauerwald2015',]$rBIAS, na.rm=T),2)`, `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_raymond2013',]$rBIAS, na.rm=T),2)`, and `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_raymond2012',]$rBIAS, na.rm=T),2)` for 'Lauerwald 2015', 'Raymond 2013', and 'Raymond 2012', respectively).
-BIKER has similar performance for NRMSE and RRMSE, which is generally worse than the HG models.
-But, BIKER NRMSE is only slightly worse than 'Raymond 2013 and 'Raymond 2012' while BIKER RRMSE (median RRMSE: `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_BIKER',]$rrmse, na.rm=T),2)`) is significantly worse than the gauge-based approaches.
-RRMSE is highly sensitive to prediction errors in small values and so this suggests that BIKER performs poorly when predicting small $CO_2$ fluxes.
-In summary, BIKER performance for KGE, NRMSE, and rBIAS is either similar or negligibly worse than the gauge-based upscaling approaches tested here and notably worse for RRMSE.
-This is despite relying on absolutely no in situ information like all three other models do.
+Meanwhile, both 'Lauerwald 2015' and 'Raymond 2013' produce significant positive bias in their estimates (median scores of `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_Lauerwald2015',]$rBIAS, na.rm=T),2)` and `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_raymond2013',]$rBIAS, na.rm=T),2)`, respectively).
+Again, the 'Raymond 2012' model performs the best with virtually no bias (`r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_raymond2012',]$rBIAS, na.rm=T),2)`).
+BIKER and 'Lauerwald 2015' have similar performance for NRMSE and RRMSE, which is generally worse than the other two models.
+In summary, BIKER performance across all four metrics is either similar or slightly worse than two of the in-situ approaches tested here and better than the 'Lauerwald 2015' model.
+This is despite relying on absolutely no streamgauge data like all three other models do.
 
 ##4 Discussion
-###4.1 Towards remote sensing global spatiotemporal dynamics of $k_{600}$
+###4.1 Towards remote sensing of global spatiotemporal dynamics of *k*
 To date, most field-scale studies of riverine gas exchange have focused on 1) its relationship with wind speed [e.g @beaulieuControlsGasTransfer2012a; @borgesVariabilityGasTransfer2004; @zappaEnvironmentalTurbulentMixing2007], 2) average flow velocity [e.g.
 @alinPhysicalControlsCarbon2011; @beaulieuControlsGasTransfer2012a; @schelkerCO2EvasionSteep2016a], or 3) discharge [e.g.
-@robertsMultipleScalesTemporal2007; @uehlingerEcosystemMetabolismDisturbance1998].
+@wangPhysicallyBasedScaling2021; @robertsMultipleScalesTemporal2007; @uehlingerEcosystemMetabolismDisturbance1998].
 However, the spatiotemporal dynamics of riverine gas exchange are still weakly constrained.
 A few studies have investigated these dynamics, but they have been limited to individual rivers and/or limited field seasons [@hallAirWaterOxygen2012; @sand-jensenCO2DynamicsDanish2012].
-@wallinSpatiotemporalVariabilityGas2011 performed a preliminary analysis in northern Sweden relating river specific (‘at-a-station’) temporal variability in gas exchange with channel slope, but they were limited to an average of only 8 measurements per river in a single watershed.
-This limited knowledge of large-scale $k_{600}$ spatiotemporal dynamics is due both to a lack of process-level understanding [@hallGasExchangeStreams2020] but also a lack of measurements.
-Authors have argued that the key to explaining the large residual variation in upscaling models is to explore at-a-station temporal variability in $k_{600}$ [@hallGasExchangeStreams2020].
+@wallinSpatiotemporalVariabilityGas2011 performed a preliminary analysis in northern Sweden relating cross-section specific temporal variability in gas exchange with channel slope, but they were limited to an average of only 8 measurements per river in a single watershed.
+This limited knowledge of large-scale *k* spatiotemporal dynamics is due both to a lack of process-level understanding [@hallGasExchangeStreams2020] but also a lack of measurements.
+Authors have argued that the key to explaining the large residual variation in upscaling models is to explore at-a-station temporal variability in *k* [@hallGasExchangeStreams2020].
 
-Therefore, estimating $k_{600}$ from SWOT data is an attractive option for exploring the spatiotemporal dynamics of $k_{600}$ at fine temporal resolution and at the global-scale.
+Therefore, estimating *k* from SWOT data is an attractive option for exploring its spatiotemporal dynamics at fine temporal resolution and at the global-scale.
 SWOT will provide daily hydraulic measurements for a 3 month fast sampling period for calibration and validation and sampling thereafter between 1 and 7 days per 21 day repeat cycle [@biancamariaSWOTMissionIts2016].
-BIKER's success in 1) infering $k_{600}$ using simulated SWOT data over a wide range of rivers (Figures 5 and 6) and 2) being robust to measurement errors internal to the SWOT data (Figures 4 and 6a) bode well for BIKER's eventual implementation on real SWOT observations.
+BIKER's success in 1) infering $k_{o_2}$ using simulated SWOT data over a wide range of rivers (Figures 2 and 3) and 2) being robust to measurement errors internal to the SWOT data (Figures 2) bode well for BIKER's eventual implementation on real SWOT observations.
 The results presented here suggest that daily estimation of riverine gas exchange globally could be possible once SWOT launches.
 
-###4.2 Estimating bulk carbon efflux using the SWOT satellite
-Section 3.3 confirms that BIKER is successful at two things without any in situ information for $k_{600}$: 1) reproducing $FCO_2$ (Figure 7) and 2) reproducing the bulk carbon efflux (Figure 5.8).
-This encouraging result has two main implications for future work.
+###4.2 Estimating bulk carbon efflux using SWOT
+Section 3.3 confirms that BIKER is successful, without any in situ information, at predicting 1) $FCO_2$ (Figure 6) and 2) the bulk carbon efflux (Figure 7).
+This encouraging result has two main implications for future work coupling remote sensing via SWOT and in situ data.
 First, section 3.3 confirms that we can couple BIKER and SWOT with in situ gas concentration loggers to produce $FCO_2$ estimates at novel temporal resolution in SWOT-observable rivers.
-This is particualry useful given recent advances in high temporal resolution in situ gas concentration measurements **get citations from Kelly**.
-BIKER can likely also be ran at the field scale using arrays of pressure transducers to estimate water surface slope following recent work doing the same using the @hagemannBAMBayesianAMHGManning2017a RSQ algorithm [@harlanDischargeEstimationDense].
+This is particualry useful given recent advances in high temporal resolution in situ gas concentration measurements [@ahoHighestRatesGross2021].
+BIKER can likely also be ran at the field scale using arrays of pressure transducers to estimate water surface slope (rather than using RS techniques) following recent work doing the same using the @hagemannBAMBayesianAMHGManning2017a RSQ algorithm [@harlanDischargeEstimationDense].
 
-Secondly, it is important to stress that unlike BIKER, the HG models in Figure 8 rely on an in situ streamgauge.
-This means that Figure 8 represents the best performance that those models could ever have; if ran using modeled discharge their accuracy would necessarily decrease, though this decrease would not be too much due to the generally good performance of those data.
-Thus, these results suggest that BIKER will be useful in two settings: 1) upscaling in ungauged rivers as hypothesized, but also in 2) potentially improving our bulk efflux understandings at gauged sites.
+Secondly, it is important to stress that unlike BIKER, the HG models in Figure 7 rely on an in situ streamgauge.
+This means that Figure 7 represents the best performance that those models could ever have; if ran using modeled discharge their accuracy would necessarily decrease.
+Thus, these results suggest that BIKER will be useful in two settings: 1) upscaling in ungauged rivers as hypothesized, but also in 2) potentially improving our carbon efflux understandings at gauged sites.
 Future work should systematically quantify prediction error from coupling global-scope HG models with modeled discharge, as is the default workflow used in recent upscaling studies [e.g.
@@
 
-Figure 8 also confirms that the training data used for HG models exerts a significant influence on upscaled carbon emissions from rivers.
+Figure 7 also confirms that the training data used for HG models exerts a significant influence on upscaled carbon emissions from rivers.
 The 'Lauerwald 2015' model, which was trained on larger rivers than the data used in either the 'Raymond 2013' or 'Raymond 2012' models, should perform best on SWOT-observable rivers (which are greater than 50m wide).
 Figure 8 confirms this.
-Meanwhile, BIKER has no similar reliance on hydraulic parameters trained on different river sizes and only assumes that Manning's equation is viable in the river at hand [which is generally the case in rivers large enough to be SWOT-observable- @fergusonTimeAbandonManning2010].
-Upon SWOT's launch, the BIKER approach to estimating gas exchange could be coupled with existing upscaling workflows to improve riverine gas flux predictions where gauges are unavailable but SWOT measurements are.
+Meanwhile, BIKER has no similar reliance on hydraulic parameters trained on different river sizes and only assumes that the channel's hydraulic radius can be approximated by mean flow depth (which is generally the case in rivers large enough to be SWOT-observable- text Sx and Table Sx).
+Upon SWOT's launch, the BIKER approach to estimating *k* could be coupled with ethier existing upscaling workflows or even explicit $CO_2$ transport models [@brinkerhoffLakeMorphometryRiver2021; **Saccardi & Winnick in review**] to improve riverine gas flux predictions where gauges are unavailable but SWOT measurements are.
+This coupling could potentially be done using data assimilation techniques, which have proven very useful for similar objectives in recent RSQ work [@ishitsukaCombiningOpticalRemote2020].
 
 ###4.2 Sources of BIKER uncertainty: process-level or remote sensing?
-Throughout the BIKER validation, we have assumed no parameter uncertainity in the upscaling parameters $a_i$ and $b_i$.
-We have shown that BIKER estimate uncertainty is significantly less than $a_i$ and $b_i$ uncertainity (Figures 4 and 5).
-BIKER provides similar estimates of $k_{600}$ to those produced by current upscaling models and most of the total uncertainity stems from the upscaling model itself (Figure 4).
-We argue that BIKER's estimate uncertainity is therefore limited by our current process-level understanding of $k_{600}$ upscaling and not by BIKER or SWOT, suggesting that BIKER's predictive performance can only improve from advancing our understandings of the physical processes governing gas exchange from SWOT-observable rivers.
-Developing and validating the BIKER algorithm has provided a unique lense through which we can discuss riverine gas exchange theory.
-In that context, we next explore some considerations for future work on $k_{600}$ upscaling in SWOT-observable rivers.
+Throughout the BIKER validation, we have assumed no parameter uncertainity in the upscaling parameters $\alpha$.
+We have shown that BIKER estimate uncertainty is almost always less than $\alpha$ uncertainity, and usually far less (Figure 4).
+Implicit measurement errors in SWOT data also exert a trivial influence on BIKER accuracy (Figure 2).
+Therefore, most of the total BIKER uncertainity stems from the predictive model itself.
+We argue that BIKER's total uncertainity is therefore limited by current process-level understanding of riverine *k* and not by SWOT measurement errors or the BIKER inversion process.
+This suggests that BIKER's predictive performance can only improve as we continue advancing our understandings of the physical processes governing gas exchange from SWOT-observable rivers.
 
-####4.2.1 Gas exchange under uniform and non-uniform flow conditions
-Section 2.2 highlights that most upscaling studies to date have assumed uniform flow conditions (i.e.
-$S_h \neq S_0$) in order to train upscaling models using readily available slope data.
-However, the first-principles model previously used by @ulsethDistinctAirWater2019a and @raymondScalingGasTransfer2012a to define *eD* [@tsivoglouTracerMeasurementReaeration1976] does not make this simplifying assumption.
-Therefore, it is an open research question whether parameterizing $k_{600}$ upscaling models via $S_h$ can account for some of the unexplained residual variation in current upscaling models [@hallGasExchangeStreams2020].
-Conveniently, SWOT will explicitly measure $S_h$ at unprecendeted spatial and temporal resolutions and will be spatially joined to hydrography that provides $S_0$.
-Thus, future researchers can use BIKER and SWOT in conjunction to directly answer this question.
-
-####4.2.2 Bed roughness and gas exchange in SWOT rivers
-Channel bed roughness affects riverine gas exchange mostly in high energy streams where slope is sufficiently steep for bubble-induced gas exchange to occur [@hallGasExchangeStreams2020].
-@ulsethDistinctAirWater2019a showed bed roughness loosely scales with $k_{600}$ in steep Alpine streams.
-However, they coarsely estimated bed roughness from arial imagery and to date most similar work has focused on labratory exercises [e.g.
-@chansonStreamReaerationNonuniform; @moogStreamReaerationNonuniform].
-We argue that bed roughness is not controlling gas exchange in SWOT-observable rivers because they are so large and therefore flat.
-Instead, gas exchange in SWOT-observable rivers is presumably dominated by water-column, rather than bed friction, turbulence.
-We show in Figure S2 with the @ulsethDistinctAirWater2019a data that the 'effective bed roughness height' scales with $k_{600}$ only in extremely steep streams (see Text S2 for the calculation of this bed roughness term).
-This relationship fundamentally breaks down in less steep rivers, and such steep slopes are functionally impossible in SWOT-observable rivers that are over 50m wide.
-This promising initial result indicates bed roughness controls some aspects of gas exchange and should be explicitly explored in future work, but is less relevant for BIKER's application to SWOT data.
-This is particularly important because small, steep rivers dominate global river networks and their GHG evasion [@horgbyUnexpectedLargeEvasion2019] due to the fractal nature of river systems [@tarbotonFractalNatureRiver1988].
-
-####4.2.3 Wind-driven gas exchange in SWOT rivers
-In wide rivers like those that SWOT will observe, wind begins to exert a non-trivial influence on gas exchange.
-It is well established that in lakes and the ocean, wind controls near-surface turbulence and thus gas exchange [@beaulieuControlsGasTransfer2012a; @readLakesizeDependencyWind2012].
-Authors have argued that large rivers are a hybrid of the hydraulics-driven turbulence in small rivers and the wind-driven turbulence in lakes [@beaulieuControlsGasTransfer2012a].
-As SWOT will measure only rivers wider than 50m, it follows that wind is likely exerting some influence on gas exchange in SWOT-observable reaches.
-Here, we opted to ignore wind effects in our upscaling model and in BIKER to favor global scalability and implementability for two reasons: 1) current upscaling efforts do not account for wind in their estimation of $k_{600}$ either and 2) it is infeasible to parameterize every set of SWOT measurements with local wind data.
-Further, relying on an in situ understanding of wind defeats the purpose of BIKER for ungauged settings.
-Future work should explore the feasibility of assigning a Bayesian prior on wind speed for BIKER.
+We speculate that for BIKER to be useful on actual SWOT data post-launch, we will need to begin 'closing the gap' on process-level uncertainitites in predicting *k* from river hydraulics.
+Future work should focus on two areas.
+First, we must expand theoretical work on predicting *k* from hydraulics in large, SWOT-observable rivers where wind exerts a non-trivial influence on *k* [@beaulieuControlsGasTransfer2012a].
+While the predictive models for *k* that have been discussed in this manuscript are useful, they are limited by their inability to account for wind-induced turbulence that is common in large rivers and the primary controller of *k* in lakes and estuaries.
+Second, future work must compare BIKER-generated *k* values to field measured values to explicitly validate the algorithm.
+This will be possible once SWOT launches and those data are available.
 
 ##5 Conclusions
-Efforts to upscale gas exchange velocities from river networks generally do so using river channel slope and average flow velocity.
-Therefore, gas evasion estimates are sensitive to available data on average flow velocity.
-In ungauged basins, this poses a problem because a velocity~discharge relationships can not be built.
+Efforts to predict gas exchange velocities from river networks generally do so using river channel hydraulics.
+Therefore, gas evasion estimates are sensitive to available in situ hydraulics data.
+In ungauged basins, this poses a problem because hydraulics~discharge relationships can not be built.
 To circumvent this problem, we propose using established techniques from the ungauged remote sensing of river discharge literature and applying them to gas exchange velocity in ungauged rivers.
-This is formalized as the BIKER algorithm and uses data from the upcoming SWOT satellite to remotely sense gas exchange velocity solely from river width and water surface height.
-We validate BIKER on 22 'rivers' of simulated SWOT data [@frassonCompilationHydraulicModels2019a], obtaining an RMSE of `r round(10^results_3_2_all$rmse,2)` ,/day after accounting for upscaling parameter uncertainity.
-When generalized to estimate bulk carbon efflux from these 22 rivers, BIKER reasonably captures the observed efflux (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_BIKER',]$sumFCO2, 0)` gG-C/yr versus `r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_ulset',]$sumFCO2, 0)` gG-C/yr, respectively).
-Further, BIKER's $FCO_2$ estimates across the 22 rivers are either functionally the same or modestly worse than those made using a streamgauge and the HG models frequently used in the literature.
+This is formalized as the BIKER algorithm, which uses data from the upcoming SWOT satellite to remotely sense gas exchange velocity (for dissolved oxygen) solely from river width and water surface height.
+We validate BIKER on 49 'rivers' of simulated SWOT data [@frassonExploringFactorsControlling2021; @durandIntercomparisonRemoteSensing2016b], obtaining an RMSE of `r round(10^results_3_2_all$rmse,2)`/day after accounting for upscaling parameter uncertainity.
+When generalized to estimate bulk carbon efflux (via $CO_2$ evasion) from these 49 rivers, BIKER reasonably captures the observed efflux (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_BIKER',]$sumFCO2, 0)` gG-C/yr versus `r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_ulset',]$sumFCO2, 0)` gG-C/yr, respectively).
+Further, BIKER's estimates across the 49 rivers are only modestly worse than those made using a streamgauge and the hydraulic geometry models frequently used in the literature.
 This suggests that BIKER will be useful not only for upscaling in ungauged rivers, but also in potentially improving our understandings in gauged rivers too.
@@
 C.B.
-Brinkerhoff was funded on **xxxxxxxxxxxxx**.
+Brinkerhoff was funded on **FINESST babyyyy**.
 BIKER is available at https://github.com/craigbrinkerhoff/BIKER.
 All code to build and generate results, figures, and the manuscript is available at https://github.com/craigbrinkerhoff/RSK600.
-Data used in this study was generously made available by the authors of Frasson et al.
-(2019), Ulseth et al.
-(2019), and Beaulieu et al.
-(2012).
+Data used in this study was generously made available by the authors of @frassonExploringFactorsControlling2021, @wangPhysicallyBasedScaling2021, @beaulieuControlsGasTransfer2012a, @ulsethDistinctAirWater2019a, and @raymondScalingGasTransfer2012a.
 This manuscript benefitted extensively from the decade plus body of work generated by the SWOT Discharge Algorithm Working Group.

### 2021-05-26 manuscript_v3.Rmd
more ms updates with supp

@@
 ---
-title: "Potential for remotely sensing riverine gas exchange and carbon efflux from SWOT observations"
+title: "Remote sensing riverine gas exchange and carbon efflux in ungauged rivers from SWOT observations"
 author: 'CB Brinkerhoff, CJ Gleason, PA Raymond, MH Harlan'
@@
 ##1 Introduction
-The Earth's carbon cycle transports and manipulates carbon between the atmosphere, geosphere, hydrosphere, and biosphere via a complex set of interconnected biogeochemical processes.
+**I don't like this opening paragraph** Vital to sustaining life, the Earth's carbon cycle moves carbon between the atmosphere, geosphere, hydrosphere, and biosphere via a complex set of interconnected biogeochemical processes.
 The primary connection between the terrestrial and oceanic biospheres is the global river network, which exports much carbon from the continents to the oceans [@colePlumbingGlobalCarbon2007a].
@@
 Organic matter and inorganic nutrients are also produced and/or consumed within rivers via ecosystem metabolism [@odumPrimaryProductionFlowing1956].
-The net effect of all of this is the global river network is usually supersaturated with dissolved gases and evades substantial amounts of carbon from its water surface to the atmosphere [@colePlumbingGlobalCarbon2007a].
+One net effect of all of this is the global river network is usually supersaturated with dissolved gases and evades substantial amounts of carbon from its water surface to the atmosphere [@colePlumbingGlobalCarbon2007a].
 
@@
 Equation 1 represents this flux for any sparingly soluble gas $[M/L^2T]$ given the gas concentration gradient between the water $C_{water} [M/L^3]$ and the air $C_{air} [M/L^3]$ and the gas exchange velocity $k [L/T]$.
-Given the structure of equation 1, calculations of this flux are highly sensitive to our measurements/estimates of *k*.
+Given the structure of equation 1, calculations of this flux are very sensitive to one's measurements/estimates of *k*.
 Note that *k* is different for different gases and water temperatures [see @hallGasExchangeStreams2020 for a thorough review of gas exchange in rivers].
 
-$$Flux=k(C_{water}-C_{air}) \mathbf{(1)}$$
+$$\mathbf{(1)} Flux=k(C_{water}-C_{air})$$
 
-*k* is also vital to understanding freshwater ecology through bulk river metabolism [gross primary production and ecological respiration- @bernhardtMetabolicRegimesFlowing2018].
+*k* is also vital to understanding freshwater ecology through bulk river metabolism [more specifically, gross primary production and ecological respiration- @bernhardtMetabolicRegimesFlowing2018].
 The general mass-balance equation for river metabolism (*M* $[M/L^2T]$) is provided as equation 2, where *DO* is the dissolved oxygen concentration $[M/L^3]$, *D* is river depth [L], and $k_{o_2}$ is *k* for oxygen at some temperature.
-When viewing equations 1 and 2 in aggregate, it is clear that *k* plays a critical role in two of the fundamental biogeochemical processes that occur in river networks.
-Therefore, both closing the global carbon cycle and further understanding freshwater ecology rely on accurate and high-fidelity assessments of *k*.
-$$\frac{dDO}{dt}=\frac{M}{D}+\frac{k_{o_2}}{D}(DO_{air}-DO_{water}) \mathbf{(2)}$$
+When viewing equations 1 and 2 in aggregate, it is clear that *k* plays a critical role in two of the core, fundamental biogeochemical processes that occur in river networks.
+Both are reliant on accurate and high-fidelity assessments of *k*, as well as a rigorous understanding of the physical mechanisms driving *k*.
+$$\mathbf{(2)} \frac{dDO}{dt}=\frac{M}{D}+\frac{k_{o_2}}{D}(DO_{air}-DO_{water})$$
 Researchers use a range of methods to measure *k*, from floating domes to eddy covariance towers to measure at a specific point in the river.
@@
 @oconnorMechanismReaerationNatural1958; @palumboAssessingPerformanceReaeration2014; @raymondScalingGasTransfer2012a].
-These models often, though not alwyas, predict $k_{600}$, or *k* that has been normalized to a Schmidt number of 600 ($CO_2$ at 20 degrees Celsius or $O_2$ at 17.5 degrees).
+These models often, though not always, predict $k_{600}$ or *k* that has been normalized to a Schmidt number of 600 ($CO_2$ at 20 degrees Celsius or $O_2$ at 17.5 degrees).
 
@@
 However, many of them are not based on fundamental mass transport theory and rather are simple empirical relations.
-This means that those models' parameters are sensitive to the training data used [@wangPhysicallyBasedScaling2021].
+This means that those models' parameter values are sensitive to the training data used [@wangPhysicallyBasedScaling2021].
 
-Multiple predictive models are based on the observation that *k* is generally correlated with both the turbulent energy dissipation rate *eD* $[L^2/T^3]$ [@ulsethDistinctAirWater2019a; @raymondScalingGasTransfer2012a] and the water surface surface shear velocity $U_\star [L/T]$ [@wangPhysicallyBasedScaling2021; @katulStructureFunctionModel2018].
+Many of these predictive models are based on the observation that *k* is generally correlated with both the turbulent energy dissipation rate *eD* $[L^2/T^3]$ [@ulsethDistinctAirWater2019a; @raymondScalingGasTransfer2012a] and the water surface surface shear velocity $U_\star [L/T]$ [@wangPhysicallyBasedScaling2021; @katulStructureFunctionModel2018].
 This is convenient because both *eD* and $U_\star$ are calculated using easily estimated hydraulic parameters: $eD \approx gS_hV$ [@tsivoglouTracerMeasurementReaeration1976] and $U_\star=\sqrt{gS_hR_h}$, where *g* is gravitational acceleration $[L/T^2]$, $S_h$ is water surface slope $[L/L]$, *V* is average flow velocity $[L/T]$ and $R_h$ is the hydraulic radius [L] (which often approximates *D*).
 Because slope is readily available in any hydrographic data product, most efforts to predict *k* across thousands of rivers are thereby limited by the quality of the final parameters *V* or $R_h$.
-Since both must be measured in the field or themselves estimated by other empirical methods, uncertainty in *V* or $R_h$ must dominate any errors in *k* estimates.
+Since both must be measured in the field or themselves estimated by empirical models, uncertainty in *V* or $R_h$ dominates errors in *k* estimates.
 This is exacerbated in ungauged basins that cover large areas, especially in the carbon-rich Arctic inland waters, as no in situ hydraulic information is available and fieldwork is impractical [@gleasonRemoteSensingRiver2020].
 
-To circumvent this problem in ungauged basins and across thousands of rivers, researchers have used 'global-scope' hydraulic geometry (HG) models that extend HG beyond its traditional use.
-Traditional HG models are simple power law relations between streamflow and channel width, depth, or velocity at a specific cross-section in a river [@leopoldHydraulicGeometryStream1953a] and a large body of geomorphology work has attempted to parse out process-based explanations for observed HG parameters [e.g.
+To circumvent this problem in ungauged basins and across thousands of rivers, researchers have used 'global-scope' hydraulic geometry (HG) models that extend HG beyond its 'at-a-station' use.
+At-a-station HG models are simple power law relations between streamflow and channel width, depth, or velocity at a specific cross-section in a river [@leopoldHydraulicGeometryStream1953a] and a large body of geomorphology work has attempted to parse out process-based explanations for at-a-station HG parameters [e.g.
 @dingmanAnalyticalDerivationAtastation2007b; @parkerPhysicalBasisQuasiuniversal2007b; @singhTHEORIESHYDRAULICGEOMETRY2003a; @fergusonHydraulicsHydraulicGeometry1986a].
 This has been extended to 'downstream HG' which relates a characteristic streamflow (i.e.
-mean annual or bankful discharge) to characterisitc hydraulics along a specific river's course [@leopoldHydraulicGeometryStream1953a].
+mean annual or bankful discharge) to characteristic hydraulics along a specific river's course [@leopoldHydraulicGeometryStream1953a].
 
 For the purposes of predicting *k* from *V* or $R_h$, biogeochemists have further extended HG models by training them on large hydraulics datasets for thousands of rivers and under all streamflow conditions [e.g @borgesGloballySignificantGreenhousegas2015a; @horgbyUnexpectedLargeEvasion2019; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a].
-However, it has long been established that HG parameters are either cross-section specific and highly variable from river to river (traditional HG) or they are discharge specific [downstream HG- @gleasonHydraulicGeometryNatural2015; @parkWorldwideVariationsHydraulic1977; @rhodesBfmDiagramGraphical1977].
-This means that the 'global-scope' HG models that are usually used are highly reliant on their training data and can potentially yield pretty different hydraulics for the same streamflow.
+However, it has long been established that HG parameters are either cross-section specific and highly variable from river to river (at-a-station HG) or they are discharge specific [downstream HG- @gleasonHydraulicGeometryNatural2015; @parkWorldwideVariationsHydraulic1977; @rhodesBfmDiagramGraphical1977].
+This suggests that the 'global-scope' HG models that are usually used are also sensitive to their training data and can potentially yield pretty different hydraulics for the same streamflow.
 It is currently not well understood how sensitive global estimates of riverine gas evasion are to the specific 'global-scope' HG model that is employed by the worker.
@@
 A decade of SWOT work has explored the multi-parameter problem of estimating the river hydraulic parameters of roughness and bathymetry from remote sensing to produce the SWOT discharge product [e.g.
-@andreadisConstrainingAssimilationSWOT2020d; @brinkerhoffConstrainingRemoteRiver2020a; @durandEstimatingReachaveragedDischarge2014a; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014a; @garamboisVariationalEstimationEffective2020; @hagemannBAMBayesianAMHGManning2017a; @larnierRiverDischargeBathymetry2020a; @oubanasDischargeEstimationUngauged2018].
+@andreadisConstrainingAssimilationSWOT2020d; @brinkerhoffConstrainingRemoteRiver2020a; @durandEstimatingReachaveragedDischarge2014a; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014a; @garamboisVariationalEstimationEffective2020; @brissetAssimilationAltimetricData2018; @hagemannBAMBayesianAMHGManning2017a; @larnierRiverDischargeBathymetry2020a; @oubanasDischargeEstimationUngauged2018].
 A specific subset of these ungauged methods are termed McFLIs or 'Mass Conserved Flow Law Inversion' algorithms [e.g.
@@
 McFLIs are readily implemented in any river that SWOT can observe and improve our hydrological understandings of ungauged rivers when little to no information was previously available [@brinkerhoffConstrainingRemoteRiver2020a; @durandIntercomparisonRemoteSensing2016b].
-Recently, McFLIs have also shown promise in providing additional and beneficial information to traditional hydrologic modeling via data assimilation of McFLIs into a traditional hydrologic model [@ishitsukaCombiningOpticalRemote2020].
+Recently, McFLIs have also shown promise in providing additional and beneficial information via data assimilation of McFLIs into a traditional hydrologic model [@ishitsukaCombiningOpticalRemote2020].
 
@@
 Equifinality refers to an under-constrained mathematical system that has essentially infinite parameter combinations that can produce the same result: there are in essence more unknowns than equations [@garamboisInferenceEffectiveRiver2015], and this problem is common to both RSQ and RS of *k*.
-Bayesian techniques have been previously used to concurrently solve for *k* and *M* in equation 2 [@applingOvercomingEquifinalityLeveraging2018; @graceFastProcessingDiel2015; @holtgrieveSimultaneousQuantificationAquatic2010].
-While these studies are field based, require high-fidelity in situ data, and are not applicable to *k* estimation in ungauged rivers, they suggest that Bayesian techniques could be useful for remotely sensing *k* when parameter equifinality is a problem.
+Bayesian techniques have been previously used to concurrently solve for *k* and *M* for different forms of equation 2 [@applingOvercomingEquifinalityLeveraging2018; @graceFastProcessingDiel2015; @holtgrieveSimultaneousQuantificationAquatic2010].
+While these studies require high-fidelity in situ data and are not applicable to *k* estimation in ungauged rivers, they suggest that Bayesian techniques could be useful for remotely sensing *k* when parameter equifinality is a problem.
 This matches McFLI logic [@andreadisConstrainingAssimilationSWOT2020d].
 
-In this context, we hypothesize that a combination of SWOT data, a Bayesian McFLI paradigm, and gas exchange theory can successfully estimate *k*.
+In this context, we hypothesize that a combination of SWOT data, a Bayesian McFLI paradigm, and gas exchange theory can successfully estimate *k* using absolutely no in situ information.
 This would in turn improve our understanding of riverine gas fluxes in ungauged basins (equation 1), as well as potentially help in parameterizing stream metabolism models (equation 2).
@@
 
-To answer these questions, we developed a new McFLI algorithm that ingests SWOT data and produces $k_{o_2}$ estimates and its explicit Bayesian uncertainity that requires no in situ inputs of any kind (although in situ data can be ingested and will improve results).
-We name the RS of $k_{o_2}$ algorithm BIKER, or the 'Bayesian Inversion of the $k_{o_2}$ Evasion Rate'.
-We validate BIKER on simulated SWOT data for 49 SWOT-observable rivers from around the world, using hydraulic models  to produce SWOT-like data as SWOT has not yet launched.
+To answer these questions, we developed a new McFLI algorithm that ingests SWOT data and produces $k_{o_2}$ estimates and their explicit Bayesian uncertainities that requires no in situ inputs of any kind (although in situ data can be ingested and will improve results).
+We name the RS of $k_{o_2}$ algorithm BIKER, or the 'Bayesian Inference/Inversion of the $k_{o_2}$ Evasion Rate'.
+We validate BIKER on simulated SWOT data for 49 SWOT-observable rivers from around the world using hydraulic models to produce SWOT-like data as SWOT has not yet launched.
 This is standard practice in the SWOT community [@frassonExploringFactorsControlling2021; @durandIntercomparisonRemoteSensing2016b].
-We also quantify BIKER's sensitivity to the expected SWOT measurement errors on 17 of those rivers: while SWOT data represent a sea change in inland water monitoring, SWOT is expected to have an approximately 10cm error in water surface elevation [@biancamariaSWOTMissionIts2016], as well as river width errors [@frassonExploringFactorsControlling2021].
-Finally, we use previously published dissolved $CO_2$ data to compare the bulk carbon efflux (via $CO_2$ evasion) from the 49 rivers as calculated using BIKER and previously published 'global-scope' HG models to compare BIKER with established in situ techniques.
+We also quantify BIKER's sensitivity to the expected SWOT measurement errors on 17 of those rivers: while SWOT data represent a sea change in inland water monitoring, it is expected to have an approximately 10cm error in water surface elevation [@biancamariaSWOTMissionIts2016], as well as river width errors [@frassonExploringFactorsControlling2021].
+Finally, we use previously published dissolved $CO_2$ data to represent a hypotetical in situ $CO_2$ sensor and compare the bulk carbon efflux from the 49 rivers as calculated using BIKER and previously published in situ techniques.
 
-##2 Methods
-To build BIKER, we join a process-based model for $k_{o_2}$ (section 2.1) with a McFLI framework for inverting SWOT measurements (section 2.2).
-Following the description of that process, we describe the validation (section 2.3) and the workflow for comparing estimated bulk carbon effluxes from a suite of 'global-scope' HG models (section 2.4).
+##2 Data
+Numerous datasets were used in this study to develop and validate BIKER and are detailed below.
+
+BIKER validation (section 3.3) was performed on 49 SWOT-simulated rivers.
+Because SWOT has yet to launch, it is standard practice to benchmark SWOT-related algorithms on SWOT-like data.
+There are three current types of SWOT-like data: 1) AirSWOT, which is an airborne Ka-band inSAR currently limited to five rivers globally, 2) simulated rivers that mimic the type of data SWOT will provide, and 3) the SWOT simulator, which introduces measurement errors to these simulated rivers.
+Because we are principally interested in algorithm performance, we limit our validation setup to simulated rivers in order to benchmark across as many rivers as possible.
+These simulated rivers are simply reach-averaged hydraulic model outputs where the water surface heights and widths are labelled as RS observations and are used as the sole inputs to BIKER.
+Here, we use 49/51 rivers collected by @frassonExploringFactorsControlling2021 and @durandIntercomparisonRemoteSensing2016b.
+These are the two benchmarking studies that have explored RSQ algorithm performance for the SWOT mission.
+Please consult both of those papers for all of the rivers' locations as well as hydraulic model specifications.
+Ultimately, the 49 rivers are spread across the United States, France, Italy, the United Kingdom, and Bangladesh.
+We omit both models for the Saint Lawrence River, Canada from @durandIntercomparisonRemoteSensing2016b because they lack enough hydraulic information to calculate $U_\star$.
+
+We also assess the influence of measurement error on BIKER's performance.
+Error in SWOT measurements will come from both the error tolerances intrinsic in the satellite data product as well as radar layover error and width measurement error.
+Layover error is the phenomenon when radar returns from different places arrive at the sensor at the same time, leading to taller landscape features appearing closer to the sensor than shorter landscape features that are the same horizontal distance from the sensor.
+For river heights and slopes, we use the error model devolped by @durandHowWillRadar2020a and implemented on 17/49 of the rivers by @frassonExploringFactorsControlling2021.
+Width errors were derived using a model built by @frassonExploringFactorsControlling2021 based on SWOT-simulator runs for the Sacramento river, California [@frassonAutomatedRiverReach2017b] and the Po river, Italy [@domeneghettiCharacterizingWaterSurface2018].
+Those two simulations allowed for explicit characterization of width errors that were then extrapolated onto the other 15 rivers.
+Width errors due to poor water classification are ignored as they were in @frassonExploringFactorsControlling2021.
+Please consult @frassonExploringFactorsControlling2021 for the specifics of how these realistic errors were incorporated into the hydraulic models.
+
+BIKER prior specifications (section 3.2.2) require the use of a training dataset of field-measured river hydraulics.
+To do this, we use a previously published compilation of field measurements that were originally made to calibrate United States Geological Survey (USGS) streamgauge rating curves [@brinkerhoffReconcilingAtaStationAtManyStations2019].
+That dataset contains over 530,000 unique measurements of average channel velocity, width, depth, and discharge from across the continental United States.
+
+For the $CO_2$ evasion and carbon efflux calculations (section 3.4), we use 26 bi-weekly dissolved $CO_2$ samples made by @beaulieuControlsGasTransfer2012a at one location in the Ohio River for one calender year from 2008-2009 (Figure S1).
+Note that this $CO_2$ data is for the Ohio River only but was applied to all 49 rivers (which includes multiple sections of the Ohio River).
+Because we are exclusively interested in the relative differences between $FCO_2$ estimates and not the raw fluxes themselves, any $CO_2$ data representative of SWOT-observable rivers was deemed acceptable.
+**Kelly just published hourly CO2 data for the CT, maybe we should use that instead to get mean daily values?**
+
+Finally, we also made extensive use of data and models developed by workers to predict gas evasion rates from river hydraulics.
+Specifically, we used the data measured, collected, or simulated by @raymondScalingGasTransfer2012a, @ulsethDistinctAirWater2019a, and @wangPhysicallyBasedScaling2021.
+
+##3 Methods
+To build BIKER, we join a process-based model for $k_{o_2}$ (section 3.1) with a McFLI framework for inverting SWOT measurements via Bayesian inference (section 3.2).
+Following the description of that process, we describe the validation setup (section 3.3) and the workflow for comparing estimated bulk carbon effluxes from a suite of 'global-scope' HG models (section 3.4).
 A flowchart detailing the entire study is provided as Figure 1.
@@
 Flowchart of methodology used in this study.
-We begin by choosing a process-based model for k (section 2.1) that is a function of river hydraulics measurable by SWOT.
-Then, we implement this model within a McFLI ('BIKER') to estimate k solely from river width and water surface slope (section 2.2).
-Finally, we couple BIKER with in situ CO2 data to compare BIKER-estimated gas fluxes with established in situ methods (section 2.4).
-See section 2.3 for the validation setup.](cache/flowchart.jpg)
+We begin by choosing a process-based model for k (section 3.1) that is a function of river hydraulics measurable by SWOT.
+Then, we implement this model within a McFLI ('BIKER') to estimate k solely from river width and water surface slope (section 3.2).
+Finally, we couple BIKER with in situ CO2 data to compare BIKER-estimated gas fluxes with established in situ methods (section 3.4).
+See section 3.3 for the validation setup.](cache/flowchart.jpg)
 
-###2.1 Choosing a predictive model for *k*
-To predict k from just the SWOT observables,river width *W* and water surface elevation *H*, a physical model for *k* is first selected.
+###3.1 Choosing a predictive model for *k*
+To predict *k* from just the SWOT observables, a physical model for *k* is first selected.
 As mentioned in section 1, there are nearly two dozen predictive models for *k* that have been devolped since the 1950s [@wangPhysicallyBasedScaling2021; @hallGasExchangeStreams2020].
-Recently, many of these models were reexplored in rivers and streams by @wangPhysicallyBasedScaling2021.
-They significantly expanded the existing training datasets of field-measured $k_{o_2}$ by using inverse modeling of DO timeseries.
-This approach was also briefly mentioned in section 1.
-These methods calibrate $k_{o_2}$, along with *M* (discretized as seperate respiration and gross primary production terms), to high-fidelity, in situ datasets of DO with the goal of minimizing the difference between the observed DO and the modeled DO [e.g.
-@applingOvercomingEquifinalityLeveraging2018; @chapraDeltaMethodEstimating1991].
-@wangPhysicallyBasedScaling2021 used the *streamMetabolizer* technique [@applingOvercomingEquifinalityLeveraging2018] at 35 rivers across the United States and showed that $k_{o_2}$ models deduced from the classic, process-based 'thin-film' and 'surface renewal' theories for mass transport yield approximately identical parameters when fit to field measurements or this new dataset of simulated $k_{o_2}$.
+Recently, many of these models were reexplored in rivers and streams by @wangPhysicallyBasedScaling2021, who significantly expanded the existing training datasets of field-measured $k_{o_2}$ by using the *streamMetabolizer* model to calibrate a form of equation 2 to high-fidelity in situ DO datasets [@applingOvercomingEquifinalityLeveraging2018] at 35 rivers across the United States.
+They showed that two $k_{o_2}$ models deduced from classic, process-based theories for mass transport yield approximately identical parameters when fit to ethier field measurements or this new dataset of simulated $k_{o_2}$.
 They also showed through cross-validation that these models are robust to overfitting on specific sets of data (a concern with many of the aforementioned predictive models- section 1).
 This suggests that there are uniform scaling realtionships between certain hydraulic properties and $k_{o_2}$ that exist regardless of the training data used.
-See @hallGasExchangeStreams2020 and @wangPhysicallyBasedScaling2021 for more in-depth explanations of the thin-film and surface renewal theories with respect to rivers.
-In short, these are the standard physical models for conceptualizing gas exchange in various aquatic environments from lakes to estuaries to rivers.
-The better fit, and more parsimonious, of the two models from @wangPhysicallyBasedScaling2021 is reprinted as equation 3(with reported coefficients of determination $r^2$) where $\alpha$ is the fitted parameter in the linear regression.
-Equation 3, 2nd line was fit to 588 field measurements of $k_{o_2}$ while equation 3, third line was fit to 3,919 simulated $k_{o_2}$ values at 35 rivers.
-$$\begin{eqnarray} k_{o_2} &=& \alpha U_\star \mathbf{(3)} \\ &=& 47.67U_\star (r^2:0.53) \\ &=& 49.71U_\star (r^2: 0.76)  \end{eqnarray}$$
+@wangPhysicallyBasedScaling2021's better fit, and more parsimonious, of the two models is reprinted as equation 3 (with that paper's reported coefficients of determination $r^2$).
+$\alpha$ is the fitted parameter in the linear regression.
+Equation 3-2nd line was fit to 588 field measurements of $k_{o_2}$ while equation 3-third line was fit to 3,919 simulated $k_{o_2}$ values at 35 rivers.
+$$\mathbf{(3)} \begin{eqnarray} k_{o_2} &=& \alpha U_\star \\ &=& 47.67U_\star (r^2:0.53, measured) \\ &=& 49.71U_\star (r^2: 0.76, simulated)  \end{eqnarray}$$
 
-We implement this model within BIKER (using 48 as a reasonable value for $\alpha$ given equation 3) for a few reasons.
-First, as suggested earlier @wangPhysicallyBasedScaling2021 provides compelling evidence that an $\alpha$ of approximately 48 is uniform across rivers and streams, regardless of the training data used.
-This is encouraging for use within BIKER as BIKER is specifically designed to be as river-agnostic as possible and so we sought a fully generalized equation.
+We implement this model within BIKER (using 48 as a reasonable value for $\alpha$ given both forms of equation 3) for a few reasons.
+First, @wangPhysicallyBasedScaling2021 provides compelling evidence that an $\alpha$ of approximately 48 is uniform across rivers and streams, regardless of the training data used.
+This is encouraging for use within BIKER as BIKER is specifically designed to be as river-agnostic as possible and so we sought as generalized an equation as possible.
 Second, it yields a simple linear relationship across all rivers, regardless of their size or steepeness.
@@
 Recent work has shown that *k* does not scale with *eD* via the same statistical parameters across all river geomorphologies [@ulsethDistinctAirWater2019a].
-We further explored the relationship presented in that paper and found that the $k \sim eD$ relationship fundamentally breaks down in rivers with low *eD*, which corresponds to nearly all rivers that SWOT will observe (**supp figure, need to do**).
+We further explored the relationship presented in that paper and found that the $k \sim eD$ relationship fundamentally breaks down in rivers with low *eD*, which corresponds to nearly all rivers that SWOT will observe (**Figure S2, need to do**).
 Therefore, an *eD* based predictive model is less useful in SWOT rivers.
 Third, we sought to explicitly predict $k_{o_2}$, as its use extends beyond constraining gas fluxes from rivers and into parameterizing river metabolism models (equation 2).
-
 We stress that most other predictive models for *k* could be implemented within BIKER.
@@
 
-###2.2 Developing BIKER
-With a predictive model for $k_{o_2}$ chosen, we now implement it within the McFLI framework.
-The approach used here is informed by the @hagemannBAMBayesianAMHGManning2017a McFLI algorithm for ungauged RSQ, further explored in more recent work [@brinkerhoffConstrainingRemoteRiver2020a].
+###3.2 Developing BIKER
+With a predictive model for $k_{o_2}$ chosen, we now implement it within a McFLI framework.
+The approach used here is informed by the @hagemannBAMBayesianAMHGManning2017a McFLI algorithm for ungauged RSQ, further explored in more recent work by @brinkerhoffConstrainingRemoteRiver2020a.
 These papers conceptualize discharge as a Bayesian remote sensing problem, which we largely follow here to conceptualize $k_{o_2}$ as a Bayesian remote sensing problem that can be solved using SWOT data.
 
-BIKER and Bayesian inference in general starts from Bayes rule, after acknowleding that it needs only to be specified as a proportionality to approximate the full posterior distribution (equation 4).
-In equation 4, $\Theta$ is some set of parameters we want to solve for (including $k_{o_2}$), *x* is the observed data, $f(x|\Theta)$ is the 'likelihood function' or sampling model conditional on the parameters, and $p(\Theta)$ is the joint prior distribution of the parameters.
+BIKER, and Bayesian inference in general, starts from Bayes rule (equation 4).
+In equation 4, $\Theta$ is some set of non-remotely-sensible parameters we want to solve for (including $k_{o_2}$), *x* is the observed data, $f(x|\Theta)$ is the 'likelihood function' or sampling model conditional on the parameters, and $p(\Theta)$ is the joint prior distribution of the parameters.
+Therefore, we are interested in solving for $p(\Theta|x)$, or the 'posterior' distribution.
 For BIKER, *x* is the SWOT-observables *W* and *H*.
-First, we detail the likelihood specification and then the prior specifications.
+Note that $p(x)$ is usually computationally intractable to integrate, so Bayesian inference tools require only the proportionality to be specified: $p(\Theta|x) \propto f(x|\Theta)p(x)$.
+Sampling algorithms are then used to approximate the posterior distribution, as is done in BIKER and detailed below.
+First, we detail our likelihood specification and then our prior specifications.
 
-$$p(\Theta|x) \propto f(x|\Theta)p(\Theta) \mathbf{(4)}$$
+$$\mathbf{(4)} p(\Theta|x) = \frac{f(x|\Theta)p(\Theta)}{p(x)}$$
 
-####2.2.1 BIKER likelihood
+####3.2.1 BIKER sampling model
 To concieve of equation 3 as a likelihood function, $k_{o_2}$ first needs to be written as a function of SWOT-observables *W* and *H* (which provides water surface slope $S_h$).
-This algebra is carried out in equation 5, where we assume that $R_h \approx \frac{A}{W}$ where *A* is wetted channel area.
+This algebra is carried out and results in equation 5, where we assume that $R_h \approx \frac{A}{W}$ where *A* is wetted channel area.
 *A* is further split into the SWOT-observable portion *dA* and the unobservable portion $A_0$ following @durandEstimatingReachaveragedDischarge2014a.
-*dA* is estimated by assuming a rectangular river channel so that $dA \approx \delta H*W$.
+*dA* is estimated assuming a rectangular river channel so that $dA \approx \delta H*W$.
 
-*COLIN COMMENT ABOUT DA: sensitivity analysis? "We investigated the sensitivity of BIKER to this rectangular assumption and found it negligible (xxx section x.x if appropriate xxx)".*
+**COLIN COMMENT from old ms: sensitivity analysis? "We investigated the sensitivity of BIKER to this rectangular assumption and found it negligible (xxx section x.x if appropriate xxx)".**
 
-$$\begin{eqnarray} k_{o_2} &=& 48\sqrt{gR_hS_h} \mathbf{(5)} \\ &=& 48g^{1/2}S_h^{1/2}(\frac{A_0+dA}{W})^{1/2} \end{eqnarray}$$
+$$\mathbf{(5)} \begin{eqnarray} k_{o_2} = 48g^{1/2}S_h^{1/2}(\frac{A_0+dA}{W})^{1/2} \end{eqnarray}$$
 
@@
 
-$$ f(x|\Theta) = \left[48^{-1}(\frac{W}{gS_h})^{1/2}\right] \sim lognormal\left(\left[k_{o2}^{-1}(A_0+dA)^{1/2}\right], \sigma^2_{k_{o2}}\right)  \mathbf{(6)}$$
+$$\mathbf{(6)} f(x|\Theta) = \left[48^{-1}(\frac{W}{gS_h})^{1/2}\right] \sim lognormal\left(\left[k_{o2}^{-1}(A_0+dA)^{1/2}\right], \sigma^2_{k_{o2}}\right)$$
 
-####2.2.2 Prior Specifications
-Equation 6 necessitates that we specify prior distributions for parameters $A_0$, $k_{o_2}$.
+####3.2.2 Prior Specifications
+Equation 6 necessitates that we specify prior distributions for parameters $A_0$ and $k_{o_2}$.
 Prior distributions formalize the a priori estimates (and uncertainties) for the non-remotely-sensed parameters.
 More intuitively, BIKER priors represent our 'prior river knowledge' of what $A_0$ and $k_{o_2}$ probably are for some river since they cannot be remotely sensed.
-All priors are formalized within the model as truncated normal distributions of the log-transformed terms such that $log(X) \sim N(\mu, \sigma^2)$ for $\lambda < X < \epsilon$, using prior hyperparameters mean ($\mu$), standard deviation ($\sigma$), and upper ($\epsilon$) and lower bounds ($\lambda$) for any parameter *X*.
+All priors are formalized within the model as truncated normal distributions of the log-transformed terms such that $log(X) \sim N(\mu, \sigma^2)$ for $\lambda < log(X) < \epsilon$, using prior hyperparameters mean ($\mu$), standard deviation ($\sigma$), and upper ($\epsilon$) and lower bounds ($\lambda$) for any parameter *X*.
 In order to avoid relying on in situ information, as is the goal of BIKER and SWOT, we assign prior hyperparameters using SWOT data only.
 $A_0$ prior hyperparameters were assigned following the method developed by @brinkerhoffConstrainingRemoteRiver2020a They developed a set of river channel prior hyperparameters for McFLI algorithms that are entirely RS-able and reflect differential channel hydraulics as a function of river geomorphology.
-They used an extensive database of field measurements and machine learning to identify patterns that associate the river width with river geomorphic properties, so prior hyperparameters may be assigned to rivers using only the existing remotely sensed data.
+They used an extensive database of field measurements and machine learning to identify patterns that associate river width with the hydraulic priors needed to run McFLIs so that prior hyperparameters may be assigned to rivers using only the existing remotely sensed data.
 This leaves the $k_{o_2}$ hyperparameters to be defined.
-We assigned those using a simple slope regression model trained on an available subset of the @raymondScalingGasTransfer2012a dataset of field-measured $k_{o_2}$: equation 7 assigns $\mu$ while $\sigma^2$ was set to equation 7's standard error: 0.77.
+We assigned those using a simple slope regression model trained on an available subset of the @raymondScalingGasTransfer2012a dataset of field-measured $k_{o_2}$: equation 7 assigns $\mu$ while $\sigma^2$ was set to equation 7's log-transformed standard error: 0.77.
 $\lambda$ and $\epsilon$ were set to log(0.0001) and log(60), respectively.
 
-$$\mu_{k_{o_2}}=3.3492 +0.3661ln(S_h)   \mathbf{(7)}$$
+$$\mathbf{(7)} \mu_{k_{o_2}}=3.3492 +0.3661ln(S_h)$$
 
@@
 Uncertainty in equation 5 can stem from 1) parameter uncertainty in $\alpha$ or 2) error in the assumption that $R_h \approx \frac{A}{W}$.
-Total equation 5 uncertainty is a function of both uncertainity sources, so we need to propogate those uncertanties through the algebra of equation 5.
+Because total equation 5 uncertainty is a function of both uncertainity sources, we need to propogate those uncertanties through the algebra of equation 5.
 We use MC simulations to do this.
 MC simulations repeately sample from parameter distributions defined by their individual means and uncertainities to produce a distribution of model predictions from which an uncertainity term can be extracted.
-See **supp figure** for a flowchart of the entire methodology, but in short we run 8,000 different MC simulations on 8,000 sets of field measurements of river channel hydraulics taken from the @brinkerhoffReconcilingAtaStationAtManyStations2019 dataset and use the median uncertainity term across those 8,000 distributions as $\sigma_{k_{o_2}}$.
+See Figure S3 for a flowchart of the entire methodology, but in short we run 8,000 different MC simulations on 8,000 sets of field measurements from the @brinkerhoffReconcilingAtaStationAtManyStations2019 dataset and use the median uncertainity term across those 8,000 distributions as $\sigma_{k_{o_2}}$.
 Each of the 8,000 simulations are themselves 10,000 runs, sampling from the normal distributions for $\alpha$ and $R_h$.
-We defined the distribution of the $R_h$ prediction again using the @brinkerhoffReconcilingAtaStationAtManyStations2019 dataset (n = 500,000+): we fit a linear regression between $R_h$ and $\frac{A}{W}$ and use that model's log-transformed standard error as the uncertainity in the MC simulations.
+We set the distribution of the $R_h$ prediction again using the @brinkerhoffReconcilingAtaStationAtManyStations2019 dataset: we fit a linear regression between $R_h$ and $\frac{A}{W}$ and use that model's log-transformed standard error as the uncertainity in the MC simulations.
 The resulting value was 0.15 and we further inflated it to 0.20 to account for potential errors in assuming $dA \approx \delta H*W$.
 
-With the sampling model (equation 6), prior distributions, and the $\sigma_{k_{o_2}}^2$ parameter described, a joint posterior distribution conditional on the SWOT observations is specified.
+**Still need to run this MC analysis because I don't have uncertainity terms for Wang et al (2021)'s parameters... This won't effect the validation though**
+
+With the likelihood function (equation 6), prior distributions, and the $\sigma_{k_{o_2}}^2$ parameter described (section 3.2.2), a joint posterior distribution conditional on the SWOT observations is specified.
 To approximate this distribution, we use a Markov Chain Monte Carlo (MCMC) algorithm implemented using the Stan probabilistic programming language.
-Specifically, BIKER uses a Hamiltonian Monte Carlo sampler which reduces computation time relative to other sampling algorithms.
+Specifically, BIKER uses a Hamiltonian Monte Carlo sampler which reduces computation time relative to other sampling algorithms [@hagemannBAMBayesianAMHGManning2017a].
 
-###2.3 BIKER validation setup
-We validated BIKER on 49 rivers using observed $R_h$ and observed $S_h$ First, we detail the 49 rivers, and then we detail how observed $k_{o_2}$ is 'measured'.
-
-Because SWOT has yet to launch, it is standard practice to benchmark SWOT-related algorithms on SWOT-like data.
-There are three current types of SWOT-like data: 1) AirSWOT, which is an airborne Ka-band inSAR currently limited to five rivers globally [e.g.
-of its use: @tuozzoloEstimatingRiverDischarge2019], 2) simulated rivers that mimic the type of data SWOT will provide, and 3) the SWOT simulator, which introduces measurement errors to these simulated rivers [e.g.
-of its use: @oubanasDischargeEstimationUngauged2018].
-Because we are principally interested in algorithm performance, we limit our validation setup to simulated rivers in order to benchmark across as many rivers as possible.
-These simulated rivers are simply reach-averaged hydraulic model outputs where the water surface heights and widths are labelled as RS observations and are used as the sole inputs to BIKER.
-Here, we use the 49 rivers collected by @frassonExploringFactorsControlling2021 and @durandIntercomparisonRemoteSensing2016b.
-These are the two benchmarking studies that have explored RSQ algorithm performance for the SWOT mission.
-Please consult both of those papers for all of the the rivers' locations as well as hydraulic model specifications.
-Ultimately, the 49 rivers are spread across the United States, France, Italy, the United Kingdom, and Bangladesh.
-We omit both models for the Saint Lawrence River, Canada from @durandIntercomparisonRemoteSensing2016b because they lack enough hydraulic information to calculate $U_\star$.
-
-We also assess the influence of measurement error on BIKER's performance.
-Error in SWOT measurements will come from both the error tolerances intrinsic in the satellite data product as well as radar layover error and width measurement error.
-Layover error is the phenomenon when radar returns from different places arrive at the sensor at the same time, leading to taller landscape features appearing closer to the sensor than shorter landscape features that are the same horizontal distance from the sensor.
-This will occur when the landscape slope is steeper than the radar's incidence angle [@durandHowWillRadar2020a] and so generally occurs in areas of high topographic variation [@reesPhysicalPrinciplesRemote2013].
-For river heights and slopes, we use the error model devolped by @durandHowWillRadar2020a and implemented on 17/49 of the rivers by @frassonExploringFactorsControlling2021.
-Width errors were derived using a model built by @frassonExploringFactorsControlling2021 based on SWOT-simulator runs for the Sacramento river, California [@frassonAutomatedRiverReach2017b] and the Po river, Italy [@domeneghettiCharacterizingWaterSurface2018].
-Those two simulations allowed for explicit characterization of width errors that were then extrapolated onto the other 15 rivers.
-Width errors due to poor water classification are ignored as they were by @frassonExploringFactorsControlling2021.
-Please consult @frassonExploringFactorsControlling2021 for the specifics of how these realistic errors were incorporated into the hydraulic models.
-Ultimately, we reran BIKER on these 17 rivers with the SWOT error model corrupting the SWOT-observables *W* and *H*.
-
-Regardless of the validation data or SWOT error budget used, we do not have observed $k_{o_2}$ data for these rivers, and to our knowledge no field dataset of $k_{o_2}$ exists in the type of temporal and spatial frequency that SWOT (and therefore the BIKER algorithm) provides.
+###3.3 Validation setup
+We validated BIKER on the aforementioned 49 SWOT-simulated rivers (section 2) using daily observed $R_h$ and observed $S_h$.
+We also re-validated BIKER on the 17 rivers with the SWOT error model corrupting the SWOT-observables *W* and *H*.
+However, regardless of the validation data or SWOT error budget used, we do not have observed $k_{o_2}$ data for these rivers, and to our knowledge no field dataset of $k_{o_2}$ exists in the type of temporal and spatial frequency that SWOT (and therefore the BIKER algorithm) provides.
 Further, we are principally interested in BIKER's ability to reproduce $k_{o_2}$ from SWOT observations and are less concerned with the actual accuracy of the predictive model itself, which can be validated using existing datasets as has been previously done [e.g.
 @wangPhysicallyBasedScaling2021; @ulsethDistinctAirWater2019a; @raymondScalingGasTransfer2012a].
-Therefore, we take the model outlined in section 2.1 and use that to calculate the observed $k_{o_2}$ that BIKER is validated against.
+Therefore, we take the model outlined in section 3.1 and use that to calculate the observed $k_{o_2}$ that BIKER is validated against.
 This is done using equation 8, where $R_{h, obs} [L/T]$ is observed river channel area divided by observed channel width.
@@
 It also means that, for a fair validation scheme, $\sigma_{k_{o_2}}^2$ must be set to reflect only error from our assumptions about $R_h$ and not the parameter uncertainity inherent in $\alpha$ (the 48 in equation 8).
-Thus, $\sigma_{k_{o_2}}^2$ is set to 0.20 for this validation as suggested in section 2.2.
-However once SWOT launches, it should reflect the total uncertainty calculated in section 2.2 from both sources.
+Thus, $\sigma_{k_{o_2}}^2$ is set to 0.20 for this validation as suggested in section 3.2.
+However once SWOT launches, it should reflect the total uncertainty calculated in section 3.2 from both sources.
 
-$$k_{o_2, obs}=48\sqrt{gS_hR_{h, obs}} \mathbf{(8)}$$
+$$\mathbf{(8)} k_{o_2, obs}=48\sqrt{gS_hR_{h, obs}}$$
 
@@
 
-###2.4 Upscaling to CO2 evasion and bulk carbon efflux
+###3.4 Upscaling to CO2 evasion and bulk carbon efflux
 It is one thing to accurately predict $k_{o_2}$, but researchers are often most interested in the actual gas fluxes from rivers and ultimately the carbon emitted from river to atmosphere.
-Therefore, we explore 1) the ability of BIKER to reproduce carbon dioxide evasion fluxes ($FCO_2$ $[M/L^2T]$) from these 49 rivers, and 2) the sensitivity of the carbon efflux efflux $[M/T]$ due to the chosen method to approximate $R_h$ and ultiamtely $U_\star$.
+Therefore, we explore 1) the ability of BIKER to reproduce $CO_2$ evasion fluxes ($FCO_2$ $[M/L^2T]$) from these 49 rivers, and 2) the sensitivity of the carbon efflux efflux $[M/T]$ due to the chosen method to approximate $R_h$ and ultiamtely $U_\star$.
 
 First, we calculate $FCO_2$.
-To do this, we pair the modeled $k_{CO_2}$ values (obtained from $k_{O_2}$ and equation 9) with field-measurements of water-side $CO_2$ concentrations and water temperatures.
-In equation 9, 530 is the Schmidt number for oxygen at 20 degrees Celsius (which BIKER predicts) and *Sc* is the Schmidt number at one's desired temperature for some dissolved gas.
-26 bi-weekly dissolved $CO_2$ samples were made by @beaulieuControlsGasTransfer2012a (Figure S1) at one location in the Ohio River for one calender year from 2008-2009.
-We paired these 26 values with every 11th SWOT observation by date, ignoring the timesteps beyond 26 (only ~15% of the SWOT observations were ignored here and we deemed this acceptable).
+To do this, we pair the aforementioned 26 $CO_2$ samples (section 2) with every 11th SWOT observation by date, ignoring the timesteps beyond 26 (only ~15% of the SWOT observations were ignored here and we deemed this acceptable).
 We chose to sample every 11 days as this is the average sampling resolution for SWOT.
 Not all of the SWOT rivers have observations for a full year, and when simulation dates were not available they were assumed to start on January 1st.
-Note that this $CO_2$ data is for the Ohio River only but was applied to all 49 rivers (which includes multiple sections of the Ohio River).
-Because we are exclusively interested in the relative differences between $FCO_2$ estimates and not the raw fluxes themselves, any $CO_2$ data representative of SWOT-observable rivers was deemed acceptable.
+We also pair the modeled $k_{CO_2}$ values (obtained from $k_{o_2}$ and equation 9) with these water-side $CO_2$ concentrations and water temperatures.
+In equation 9, 530 is the Schmidt number for oxygen at 20 degrees Celsius (which BIKER predicts) and *Sc* is the Schmidt number at one's desired temperature for some dissolved gas.
 Atmospheric $CO_2$ was assumed 390 uatm.
 The Schmidt number, used in equation 9, was calculated following @raymondScalingGasTransfer2012a and @wanninkhofRelationshipWindSpeed1992a.
+$FCO_2$ validation was performed using the same metrics as $k_{o_2}$ validation (Table 1).
 
-$$k_{GHG, temp}=k_{02}*(Sc/530)^{-1/2} \mathbf{(9)}$$
+$$\mathbf{(9)} k_{GHG, temp}=k_{o_2}*(Sc/530)^{-1/2}$$
 
@@
 
-##3 Results
-First, we present the results from the BIKER validation on 49 SWOT rivers (section 3.1).
-Next, we compare BIKER inversion errors to the errors inherent in equations 2 and 3, as well as the rivers' hydraulic properties (section 3.2).
-Finally, we use 4 different models for $k_{CO_2}$ to compare $FCO_2$ and bulk carbon efflux estimates made using gauged and ungauged methods (section 3.3).
+##4 Results
+First, we present the results from the BIKER validation on 49 SWOT rivers (section 4.1).
+Next, we compare BIKER inversion errors to the errors inherent in equations 2 and 3, as well as the rivers' hydraulic properties (section 4.2).
+Finally, we use 4 different models for $k_{CO_2}$ to compare $FCO_2$ and bulk carbon efflux estimates made using gauged and ungauged methods (section 4.3).
 
-###3.1 Validating the BIKER algorithm
+###4.1 Validating the BIKER algorithm
 Figure 2a plots the validation results for $k_{o_2}$ (with no SWOT measurement error) across all 49 rivers and all timesteps.
@@
 
-###3.2 BIKER performance versus *k* model performance
-Next, we compare error introduced by BIKER's Bayesian inversion to error introduced by the use of $\alpha$ in equations 2 and 3.
-Recall that the BIKER validation has already had parameter error accounted for (section 2.3) and so BIKER errors reflect error introduced only from Bayesian inference and not from the use $\alpha$.
-Figure 4 compares BIKER's performance per river (via NRMSE) to the reported cross-validated model performance for equations 2 and 3 [@wangPhysicallyBasedScaling2021].
-Note that in that paper they use the same metrics but refer to it as 'relative error' and so we do too in Figure 4.
-43/49 rivers have lower inversion errors than the cross-validated error in equations 2 and 3 (~57%), with most rivers' scores far less than 57%.
-Recall that the median NRMSE is `r round(median(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$nrmse, na.rm=T),2)` which is less than 1/2 of the reported error for $\alpha$.
-For the six rivers which inversion errors greater than the parameter error, 4/6 are approaching or over 100% error.
-This means that BIKER fundamentally does not work in these rivers, while in all the other rivers the inversion errors are nearly always much less than the parameter error.
+###4.2 BIKER performance versus *k* model performance
+The MC simulations described in section 3.2.2 yielded an overall BIKER uncertainity of **must calculate** (**Figure S4, need to do**).
+While this will be useful for running BIKER on actual SWOT data, recall that the validation presented here (section 4.1) accounts for parameter error for a fair comparison against the 'observed' $k_{o_2}$ values (section 3.3).
+In that context, figure 4 compares BIKER inversion error per river and via NRMSE to the error generated by process-level uncertainities.
+This is the dashed black line in Figure 4 and is the reported cross-validated model performance for equation 3 [@wangPhysicallyBasedScaling2021]: approximately 57%.
+Note that in that paper they use the same metric but refer to it as 'relative error' and so we do too in this section.
+43/49 rivers have lower inversion errors than the cross-validated model error for equation 3 (~57%), with most rivers' scores are far less than 57%.
+The median relative error in BIKER inversion is `r round(median(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$nrmse, na.rm=T),2)` which is less than 1/2 of the reported error for $\alpha$.
+For the six rivers with inversion errors greater than the parameter error, 4/6 are approaching or over 100% error.
+This means that BIKER inversion errors are nearly always far less than the parameter error, however occasionally the inversion is very poor and introduces much more error than equation 3 does.
 It should also be stressed that NRMSE is one of the worst perfoming metrics for BIKER (figure 3a) and so a similar comparison using KGE or rBIAS would likely yield even stronger results.
-Therefore, BIKER introduces a small amount of additional error into the physical model for $k_{o_2}$ and any errors in the final $k_{o_2}$ estimates are dominated by process-level uncertainity in predicting $k_{o_2}$ rather than uncertainity in inverting SWOT observations to estimate $k_{o_2}$.
+Therefore, BIKER introduces a small amount of additional error into the physical model for $k_{o_2}$ and any errors in the final $k_{o_2}$ estimates are dominated by process-level uncertainity in predicting $k_{o_2}$ via equation 3 rather than uncertainity in inverting SWOT observations to estimate $k_{o_2}$.
 
-![Figure 4: NRMSE (referred to as relative error in @wangPhysicallyBasedScaling2021.) for all 49 rivers.
+![Figure 4: NRMSE (referred to as relative error in @wangPhysicallyBasedScaling2021) for all 49 rivers.
 Dashed line is the cross-validation relative error for the ko2 model used in this study (~57%), reprinted from @wangPhysicallyBasedScaling2021.
@@
 
-Figure 4 shows that BIKER performs very poorly on a small subset of these 49 rivers.
-So, we also compared BIKER performance against river hydraulic properties, specifically mean observed $k_{o_2}$, *W*.
+Figure 4 shows that BIKER performs extremely poorly on 6/49 rivers.
+To explore why, we also compared BIKER performance against river hydraulic properties mean observed $k_{o_2}$, *W*.
 and $S_h$.
-We found no clear patterns between these properties and KGE, NRMSE, or RRMSE (Figures Sx, Sy, Sz respectively).
+We found no clear patterns between these properties and KGE, NRMSE, or RRMSE (Figures S5, **S6, and S7 to do**, respectively).
 However, Figure 5 plots the patterns betwwen these hydraulic properties and rBIAS.
@@
 Following basic gas exchange theory [@ulsethDistinctAirWater2019a], these are also the rivers with the greatest $k_{o_2}$ values (Figure 5a).
+**Check if this is the sam 6 rivers....**.
 
@@
 
-###3.3 Influences on carbon efflux estimates
-Finally, we explore our ability to 1) use BIKER-produced $k_{o_2}$ to estimate $FCO_2$ and 2) estimate bulk carbon efflux from $CO_2$ evasion (section 2.4).
+###4.3 Influences on carbon efflux estimates
+Finally, we explore our ability to 1) use BIKER-produced $k_{o_2}$ to estimate $FCO_2$ and 2) estimate bulk carbon efflux from $CO_2$ evasion (section 3.4).
 
@@
 
-##4 Discussion
-###4.1 Towards remote sensing of global spatiotemporal dynamics of *k*
+##5 Discussion
+###5.1 Towards remote sensing of global spatiotemporal dynamics of *k*
 To date, most field-scale studies of riverine gas exchange have focused on 1) its relationship with wind speed [e.g @beaulieuControlsGasTransfer2012a; @borgesVariabilityGasTransfer2004; @zappaEnvironmentalTurbulentMixing2007], 2) average flow velocity [e.g.
@@
 
-###4.2 Estimating bulk carbon efflux using SWOT
-Section 3.3 confirms that BIKER is successful, without any in situ information, at predicting 1) $FCO_2$ (Figure 6) and 2) the bulk carbon efflux (Figure 7).
+###5.2 Estimating bulk carbon efflux using SWOT
+Section 4.3 confirms that BIKER is successful, without any in situ information, at predicting 1) $FCO_2$ (Figure 6) and 2) the bulk carbon efflux (Figure 7).
 This encouraging result has two main implications for future work coupling remote sensing via SWOT and in situ data.
-First, section 3.3 confirms that we can couple BIKER and SWOT with in situ gas concentration loggers to produce $FCO_2$ estimates at novel temporal resolution in SWOT-observable rivers.
+First, section 4.3 confirms that we can couple BIKER and SWOT with in situ gas concentration loggers to produce $FCO_2$ estimates at novel temporal resolution in SWOT-observable rivers.
 This is particualry useful given recent advances in high temporal resolution in situ gas concentration measurements [@ahoHighestRatesGross2021].
@@
 
-###4.2 Sources of BIKER uncertainty: process-level or remote sensing?
+###5.2 Sources of BIKER uncertainty: process-level or remote sensing?
 Throughout the BIKER validation, we have assumed no parameter uncertainity in the upscaling parameters $\alpha$.
@@
 
-##5 Conclusions
+##6 Conclusions
 Efforts to predict gas exchange velocities from river networks generally do so using river channel hydraulics.
@@
 
-##6 Acknowledgements
+##7 Acknowledgements
 C.B.
@@
 All code to build and generate results, figures, and the manuscript is available at https://github.com/craigbrinkerhoff/RSK600.
-Data used in this study was generously made available by the authors of @frassonExploringFactorsControlling2021, @wangPhysicallyBasedScaling2021, @beaulieuControlsGasTransfer2012a, @ulsethDistinctAirWater2019a, and @raymondScalingGasTransfer2012a.
+Data used in this study was generously made available by the authors of @frassonExploringFactorsControlling2021, @durandIntercomparisonRemoteSensing2016b, @wangPhysicallyBasedScaling2021, @beaulieuControlsGasTransfer2012a, @ulsethDistinctAirWater2019a, and @raymondScalingGasTransfer2012a.
 This manuscript benefitted extensively from the decade plus body of work generated by the SWOT Discharge Algorithm Working Group.

### 2021-05-26 supp_v3.Rmd
more ms updates with supp

---
title: "Supplemental Information: remote sensing riverine gas exchange and carbon efflux in ungauged rivers from SWOT observations"
author: "CB Brinkerhoff, CJ Gleason, PA Raymond, MH Harlan"
date: "`r Sys.Date()`"
csl: rse_citation_style.csl
output:
  word_document: 
      reference_docx: word-styles-reference.docx
bibliography: References.bib

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Contents
This supplementary information contains 7 figures and 1 table. Please consult https://github.com/craigbrinkerhoff/RSK600 for all code to build and generate results, figures, and the manuscript.

##Figures
![Figure S1 Timeseries of the biweekly CO2 data from @beaulieuControlsGasTransfer2012a. Sampling took place 2008-2009 in the Ohio River (upstream of Cincinnati, Ohio, United States). Each point here was joined to the 11-day SWOT observations used in this study (section 2.4).](cache/FCO2/Beaulieu_timeseries.jpg)

![Figure S2. k600 versus the turbulent dissipation rate eD using the data from @ulsethDistinctAirWater2019a. Colors correspond to whether a measurement is SWOT-observable, i.e. the river's width is > 50m. The k600 ~ eD model fundamentally breaks down in SWOT-observable rivers](cache/k600/roughness_theory/eM_k600.jpg)

![Figure S3: Flowchart detailing the workflow to approximate estimate uncertainity for equation 5 in the main text. Uncertainity can stem from 1) parameter uncertainity and 2) error in our assumption that the hydraulic radius approximates the channel area divided by width. We performed 8,000 different Monte carlo simulations (themselves 10,000 runs each) on 8,000 river channel measurements in the dataset from @brinkerhoffReconcilingAtaStationAtManyStations2019](cache/suppFlowchart.jpg)

![Figure S4: plot of overall BIKER uncertainity](cache/validation/riverProperties_kge.jpg)

![Figure S5: BIKER performance, defined by KGE (see Table 1 in the main text for the metric definition) versus mean river ko2, width, and slope.](cache/validation/riverProperties_kge.jpg)

![Figure S6: BIKER performance, defined by NRMSE (see Table 1 in the main text for the metric definition) versus mean river ko2, width, and slope.](cache/validation/riverProperties_kge.jpg)

![Figure S7: BIKER performance, defined by RRMSE (see Table 1 in the main text for the metric definition) versus mean river ko2, width, and slope.](cache/validation/riverProperties_kge.jpg)

##Tables
*Table S1: Details on the 3 velocity hydraulic geometry models used to estimate FCO2 from the SWOT rivers (section 2.4).*

|**Name**|**Equation**|**Description**|**Reference**|
|-------------|-------------|-------------|-------------|-------------|
|Raymond 2012|$ln(V)=-1.64+0.285ln(Q)$|Trained on their dataset|@raymondScalingGasTransfer2012a|
|Raymond 2013|$ln(V)=mean(-1.64+0.285ln(Q),  -1.06+0.12ln(Q))$ |Average of the Raymond 2012 equation and one trained on 9,811 USGS streamgauges|@raymondGlobalCarbonDioxide2013a|
|Lauerwald 2015|$ln(V)=-1.06+0.12ln(Q)$ |Borrowed from 1/2 of the Raymond 2013 calculation (trained on 9,811 USGS streamgauges)|@lauerwaldSpatialPatternsCO22015|

##References

### 2021-06-04 manuscript_v3.Rmd
some ms updates

@@
 
-##Highlights (3-5 points, 85 characters each w/ spaces)
+##Highlights
 - The BIKER algorithm indirectly remotely senses riverine gas exchange velocity
@@
 Go BIKER!!
-*Note: this is for Remote Sensing of Environment (I think). So 400 words here.*
+*Note: this is for Remote Sensing of Environment (I think). So 400 words here. also for highlights: (3-5 points, 85 characters each w/ spaces)*
 
 ##1 Introduction
-**I don't like this opening paragraph** Vital to sustaining life, the Earth's carbon cycle moves carbon between the atmosphere, geosphere, hydrosphere, and biosphere via a complex set of interconnected biogeochemical processes.
-The primary connection between the terrestrial and oceanic biospheres is the global river network, which exports much carbon from the continents to the oceans [@colePlumbingGlobalCarbon2007a].
-However en route to the oceans, this river network additionally transports and transforms sediments, nutrients, organic matter, dissolved gases, and other constituents.
+**I don't like this opening paragraph** The Earth's carbon cycle moves carbon between the atmosphere, geosphere, hydrosphere, and biosphere via a complex set of interconnected biogeochemical processes.
+The primary connection between the terrestrial and oceanic biospheres is the global river network, which exports a significant amount of carbon from the continents to the oceans [@colePlumbingGlobalCarbon2007a], while additionally transporting and transforming sediments, nutrients, organic matter, dissolved gases, and other constituents.
 Organic matter and inorganic nutrients are also produced and/or consumed within rivers via ecosystem metabolism [@odumPrimaryProductionFlowing1956].
@@
 Further, in-river metabolism is increasingly understood to contribute to this flux [@hotchkissSourcesProcessesControlling2015].
-Equation 1 represents this flux for any sparingly soluble gas $[M/L^2T]$ given the gas concentration gradient between the water $C_{water} [M/L^3]$ and the air $C_{air} [M/L^3]$ and the gas exchange velocity $k [L/T]$.
+Equation 1 represents this flux for any sparingly soluble gas $[M/L^2T]$ given the gas concentration gradient between the water $C_{water}$ $[M/L^3]$ and the air $C_{air}$ $[M/L^3]$ and the gas exchange velocity *k [L/T]*.
 Given the structure of equation 1, calculations of this flux are very sensitive to one's measurements/estimates of *k*.
@@
 They also inject tracer gases to estimate reach-scale gas exchange [@hallGasExchangeStreams2020].
-These approaches are ideal but are infeasible when working at the river network scale across potentially tens of thousands of rivers.
+These approaches are ideal but are infeasible when working in ungauged rivers or at the river network scale across thousands of rivers.
 So, predictive models are used to estimate *k* from readily available river geomorphology [e.g.
@@
 
-Many of these predictive models are based on the observation that *k* is generally correlated with both the turbulent energy dissipation rate *eD* $[L^2/T^3]$ [@ulsethDistinctAirWater2019a; @raymondScalingGasTransfer2012a] and the water surface surface shear velocity $U_\star [L/T]$ [@wangPhysicallyBasedScaling2021; @katulStructureFunctionModel2018].
+Many of these predictive models rely on the correlation between *k* both the turbulent energy dissipation rate *eD* $[L^2/T^3]$ [@ulsethDistinctAirWater2019a; @raymondScalingGasTransfer2012a] and the water surface surface shear velocity $U_\star [L/T]$ [@wangPhysicallyBasedScaling2021; @katulStructureFunctionModel2018].
 This is convenient because both *eD* and $U_\star$ are calculated using easily estimated hydraulic parameters: $eD \approx gS_hV$ [@tsivoglouTracerMeasurementReaeration1976] and $U_\star=\sqrt{gS_hR_h}$, where *g* is gravitational acceleration $[L/T^2]$, $S_h$ is water surface slope $[L/L]$, *V* is average flow velocity $[L/T]$ and $R_h$ is the hydraulic radius [L] (which often approximates *D*).
@@
 Since both must be measured in the field or themselves estimated by empirical models, uncertainty in *V* or $R_h$ dominates errors in *k* estimates.
-This is exacerbated in ungauged basins that cover large areas, especially in the carbon-rich Arctic inland waters, as no in situ hydraulic information is available and fieldwork is impractical [@gleasonRemoteSensingRiver2020].
+This is exacerbated in ungauged basins that cover large areas, especially in the carbon-rich Arctic inland waters, as little in situ hydraulic information is available and fieldwork is impractical [@gleasonRemoteSensingRiver2020].
 
@@
 For the purposes of predicting *k* from *V* or $R_h$, biogeochemists have further extended HG models by training them on large hydraulics datasets for thousands of rivers and under all streamflow conditions [e.g @borgesGloballySignificantGreenhousegas2015a; @horgbyUnexpectedLargeEvasion2019; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a].
-However, it has long been established that HG parameters are either cross-section specific and highly variable from river to river (at-a-station HG) or they are discharge specific [downstream HG- @gleasonHydraulicGeometryNatural2015; @parkWorldwideVariationsHydraulic1977; @rhodesBfmDiagramGraphical1977].
+However, it has long been established that HG parameters are either cross-section specific and highly variable from river to river as in at-a-station HG or they are discharge specific as in downstream HG [@gleasonHydraulicGeometryNatural2015; @parkWorldwideVariationsHydraulic1977; @rhodesBfmDiagramGraphical1977].
 This suggests that the 'global-scope' HG models that are usually used are also sensitive to their training data and can potentially yield pretty different hydraulics for the same streamflow.
 It is currently not well understood how sensitive global estimates of riverine gas evasion are to the specific 'global-scope' HG model that is employed by the worker.
+Further, these approaches rely on ethier in situ discharge records or modeled streamflow which introduces additional uncertainity.
 
@@
 We also quantify BIKER's sensitivity to the expected SWOT measurement errors on 17 of those rivers: while SWOT data represent a sea change in inland water monitoring, it is expected to have an approximately 10cm error in water surface elevation [@biancamariaSWOTMissionIts2016], as well as river width errors [@frassonExploringFactorsControlling2021].
-Finally, we use previously published dissolved $CO_2$ data to represent a hypotetical in situ $CO_2$ sensor and compare the bulk carbon efflux from the 49 rivers as calculated using BIKER and previously published in situ techniques.
+Finally, we use previously published dissolved $CO_2$ data to represent a hypotetical in situ $CO_2$ sensor and compare the bulk carbon efflux from the 49 rivers as calculated using BIKER and previously published in situ techniques for predicting $k_{CO_2}$.
 
 ##2 Data
-Numerous datasets were used in this study to develop and validate BIKER and are detailed below.
+Numerous datasets were used in this study to develop and validate BIKER and are detailed below. Please see Figure 1 for a map of the approximate locations for the data used in this study.
+
+![Figure 1: Map of the 49 hydraulic models and 1 timeseries of CO2 samples used in this study.
+Note that hydraulic model locations are approximate as most of the models are not geo-referenced.
+We additionally used over 530,000 discrete measurements of river channel hydraulics from across the continetal United States (Brinkerhoff et al.
+2019) that are not mapped here.](cache/map_fin.jpg)
 
@@
 These are the two benchmarking studies that have explored RSQ algorithm performance for the SWOT mission.
-Please consult both of those papers for all of the rivers' locations as well as hydraulic model specifications.
+The approximate locations of these rivers are plotted in FIgure 1, and please consult both of those papers for all of the hydraulic model specifications.
 Ultimately, the 49 rivers are spread across the United States, France, Italy, the United Kingdom, and Bangladesh.
-We omit both models for the Saint Lawrence River, Canada from @durandIntercomparisonRemoteSensing2016b because they lack enough hydraulic information to calculate $U_\star$.
+We omit both models for the Saint Lawrence River, Canada from @durandIntercomparisonRemoteSensing2016b because they lack enough hydraulic information to calculate $U_\star$, which will be necessary in section 3.2.
 
@@
 That dataset contains over 530,000 unique measurements of average channel velocity, width, depth, and discharge from across the continental United States.
+It has been previously used to inform prior specifications for RSQ algorithms [@andreadisConstrainingAssimilationSWOT2020d; @brinkerhoffConstrainingRemoteRiver2020a].
+We acknowledge that our priors will therefore be biased towards the types of rivers in the continental United States and miss key hydroclimatic regions like Arctic and Tropical rivers.
+However, previous success in using these data in unrepresented regions [@brinkerhoffConstrainingRemoteRiver2020a] bodes well for their use here.
+Further, 14/19 rivers are from outside of this region (Figure 1).
 
@@
 Because we are exclusively interested in the relative differences between $FCO_2$ estimates and not the raw fluxes themselves, any $CO_2$ data representative of SWOT-observable rivers was deemed acceptable.
-**Kelly just published hourly CO2 data for the CT, maybe we should use that instead to get mean daily values?**
 
@@
 Specifically, we used the data measured, collected, or simulated by @raymondScalingGasTransfer2012a, @ulsethDistinctAirWater2019a, and @wangPhysicallyBasedScaling2021.
+Note that much of this data was collated from many other datasets from many other authors.
 
 ##3 Methods
-To build BIKER, we join a process-based model for $k_{o_2}$ (section 3.1) with a McFLI framework for inverting SWOT measurements via Bayesian inference (section 3.2).
+To build BIKER, we join a predictive model for $k_{o_2}$ (section 3.1) with a McFLI framework for inverting SWOT measurements via Bayesian inference (section 3.2).
 Following the description of that process, we describe the validation setup (section 3.3) and the workflow for comparing estimated bulk carbon effluxes from a suite of 'global-scope' HG models (section 3.4).
-A flowchart detailing the entire study is provided as Figure 1.
+A flowchart detailing the entire study is provided as Figure 2.
 
-![Figure 1.
+![Figure 2.
 Flowchart of methodology used in this study.
@@
 To predict *k* from just the SWOT observables, a physical model for *k* is first selected.
-As mentioned in section 1, there are nearly two dozen predictive models for *k* that have been devolped since the 1950s [@wangPhysicallyBasedScaling2021; @hallGasExchangeStreams2020].
-Recently, many of these models were reexplored in rivers and streams by @wangPhysicallyBasedScaling2021, who significantly expanded the existing training datasets of field-measured $k_{o_2}$ by using the *streamMetabolizer* model to calibrate a form of equation 2 to high-fidelity in situ DO datasets [@applingOvercomingEquifinalityLeveraging2018] at 35 rivers across the United States.
+As mentioned in section 1, there are dozens of predictive models for *k* that have been devolped since the 1950s [@hallGasExchangeStreams2020].
+Recently, many of these models were re-explored in rivers and streams by @wangPhysicallyBasedScaling2021, who significantly expanded the existing training datasets of field-measured $k_{o_2}$ by using the *streamMetabolizer* model to calibrate a form of equation 2 to high-fidelity in situ DO datasets [@applingOvercomingEquifinalityLeveraging2018] at 35 rivers across the United States.
 They showed that two $k_{o_2}$ models deduced from classic, process-based theories for mass transport yield approximately identical parameters when fit to ethier field measurements or this new dataset of simulated $k_{o_2}$.
-They also showed through cross-validation that these models are robust to overfitting on specific sets of data (a concern with many of the aforementioned predictive models- section 1).
-This suggests that there are uniform scaling realtionships between certain hydraulic properties and $k_{o_2}$ that exist regardless of the training data used.
+They also showed through cross-validation that these models are more robust to overfitting on specific sets of data than many of the other models tested.
+While these two models still exhibit large errors, they provide the best fit of those tested and potentially suggest that there are uniform scaling realtionships between certain hydraulic properties and $k_{o_2}$.
 @wangPhysicallyBasedScaling2021's better fit, and more parsimonious, of the two models is reprinted as equation 3 (with that paper's reported coefficients of determination $r^2$).
@@
 
+**I'm struggling with this paragraph, and also don't know how to write it diplomatically**
 We implement this model within BIKER (using 48 as a reasonable value for $\alpha$ given both forms of equation 3) for a few reasons.
-First, @wangPhysicallyBasedScaling2021 provides compelling evidence that an $\alpha$ of approximately 48 is uniform across rivers and streams, regardless of the training data used.
+First, @wangPhysicallyBasedScaling2021 suggest that an $\alpha$ of approximately 48 is uniform across rivers and streams, regardless of the training data used.
 This is encouraging for use within BIKER as BIKER is specifically designed to be as river-agnostic as possible and so we sought as generalized an equation as possible.
@@
 This is not necessairly true of other *k* models, for example those based on *eD* (section 1).
-Recent work has shown that *k* does not scale with *eD* via the same statistical parameters across all river geomorphologies [@ulsethDistinctAirWater2019a].
-We further explored the relationship presented in that paper and found that the $k \sim eD$ relationship fundamentally breaks down in rivers with low *eD*, which corresponds to nearly all rivers that SWOT will observe (**Figure S2, need to do**).
+The relationship between *eD* and *k* is not necessairly linear [@raymondScalingGasTransfer2012a], suggesting poorer performance for small *k* values.
+Further, recent work has shown that *k* does not scale with *eD* via the same statistical parameters across all river geomorphologies [@ulsethDistinctAirWater2019a].
+We further explored the relationship presentedby [@ulsethDistinctAirWater2019a] and found that the $k \sim eD$ relationship fundamentally breaks down in rivers with low *eD*, which corresponds to nearly all rivers that SWOT will observe (Figure S2).
 Therefore, an *eD* based predictive model is less useful in SWOT rivers.
@@
 
-BIKER, and Bayesian inference in general, starts from Bayes rule (equation 4).
-In equation 4, $\Theta$ is some set of non-remotely-sensible parameters we want to solve for (including $k_{o_2}$), *x* is the observed data, $f(x|\Theta)$ is the 'likelihood function' or sampling model conditional on the parameters, and $p(\Theta)$ is the joint prior distribution of the parameters.
+BIKER, and Bayesian inference in general, starts from Bayes rule (equation 4), where $\Theta$ is some set of non-remotely-sensible parameters we want to solve for (including $k_{o_2}$), *x* is the observed data, $f(x|\Theta)$ is the 'likelihood function' or sampling model conditional on the parameters, and $p(\Theta)$ is the joint prior distribution of the parameters.
 Therefore, we are interested in solving for $p(\Theta|x)$, or the 'posterior' distribution.
@@
 
-####3.2.1 BIKER sampling model
+####3.2.1 BIKER likelihood
 To concieve of equation 3 as a likelihood function, $k_{o_2}$ first needs to be written as a function of SWOT-observables *W* and *H* (which provides water surface slope $S_h$).
@@
 
-Next, equation 5 is written as a Bayesian sampling model to have all of the SWOT observations on the left-hand side sampled from the unknown model parameters ($A_0$ and $k_{o_2}$) on the right-hand side.
+Next, equation 5 is written as a Bayesian sampling model, in which all of the SWOT observations are sampled from the unknown model parameters ($A_0 + dA$ and $k_{o_2}$).
 This is equation 6.
@@
 Note that in the formal model specification, equation 6 is written as a normal distribution of its log-transformed quantities.
+Here, it is written as a lognormal distribution for succinctness.
 
@@
 ####3.2.2 Prior Specifications
-Equation 6 necessitates that we specify prior distributions for parameters $A_0$ and $k_{o_2}$.
+Equations 4 and 6 necessitate that we specify prior distributions for the parameters $A_0 + dA$ and $k_{o_2}$.
 Prior distributions formalize the a priori estimates (and uncertainties) for the non-remotely-sensed parameters.
-More intuitively, BIKER priors represent our 'prior river knowledge' of what $A_0$ and $k_{o_2}$ probably are for some river since they cannot be remotely sensed.
+More intuitively, BIKER priors represent our 'prior river knowledge' of what $A_0 + dA$ and $k_{o_2}$ probably are for some river since they cannot be remotely sensed.
 All priors are formalized within the model as truncated normal distributions of the log-transformed terms such that $log(X) \sim N(\mu, \sigma^2)$ for $\lambda < log(X) < \epsilon$, using prior hyperparameters mean ($\mu$), standard deviation ($\sigma$), and upper ($\epsilon$) and lower bounds ($\lambda$) for any parameter *X*.
 In order to avoid relying on in situ information, as is the goal of BIKER and SWOT, we assign prior hyperparameters using SWOT data only.
-$A_0$ prior hyperparameters were assigned following the method developed by @brinkerhoffConstrainingRemoteRiver2020a They developed a set of river channel prior hyperparameters for McFLI algorithms that are entirely RS-able and reflect differential channel hydraulics as a function of river geomorphology.
+$A_0$ prior hyperparameters were assigned following the method developed by @brinkerhoffConstrainingRemoteRiver2020a.
+They developed a set of river channel prior hyperparameters for McFLI algorithms that are entirely RS-able and reflect differential channel hydraulics as a function of river geomorphology.
 They used an extensive database of field measurements and machine learning to identify patterns that associate river width with the hydraulic priors needed to run McFLIs so that prior hyperparameters may be assigned to rivers using only the existing remotely sensed data.
@@
 We use MC simulations to do this.
-MC simulations repeately sample from parameter distributions defined by their individual means and uncertainities to produce a distribution of model predictions from which an uncertainity term can be extracted.
+MC simulations repeately sample from the parameter distributions to produce a distribution of model predictions from which an uncertainity term can be extracted.
 See Figure S3 for a flowchart of the entire methodology, but in short we run 8,000 different MC simulations on 8,000 sets of field measurements from the @brinkerhoffReconcilingAtaStationAtManyStations2019 dataset and use the median uncertainity term across those 8,000 distributions as $\sigma_{k_{o_2}}$.
 Each of the 8,000 simulations are themselves 10,000 runs, sampling from the normal distributions for $\alpha$ and $R_h$.
-We set the distribution of the $R_h$ prediction again using the @brinkerhoffReconcilingAtaStationAtManyStations2019 dataset: we fit a linear regression between $R_h$ and $\frac{A}{W}$ and use that model's log-transformed standard error as the uncertainity in the MC simulations.
-The resulting value was 0.15 and we further inflated it to 0.20 to account for potential errors in assuming $dA \approx \delta H*W$.
+We quantified the error in $R_h \sim \frac{A}{W}$ again using the @brinkerhoffReconcilingAtaStationAtManyStations2019 dataset.
 
-**Still need to run this MC analysis because I don't have uncertainity terms for Wang et al (2021)'s parameters... This won't effect the validation though**
+**Still need to run this MC analysis because I don't have uncertainity terms for Wang et al (2021)'s parameters...
+This won't effect the paper results though b/c i don't actually use it in the validation**
 
@@
 ###3.3 Validation setup
-We validated BIKER on the aforementioned 49 SWOT-simulated rivers (section 2) using daily observed $R_h$ and observed $S_h$.
-We also re-validated BIKER on the 17 rivers with the SWOT error model corrupting the SWOT-observables *W* and *H*.
+We validate BIKER on the aforementioned 49 SWOT-simulated rivers (section 2) using daily observed $R_h$ and observed $S_h$.
+We also re-validate BIKER on the 17 rivers with the SWOT error model corrupting the SWOT-observables *W* and *H*.
 However, regardless of the validation data or SWOT error budget used, we do not have observed $k_{o_2}$ data for these rivers, and to our knowledge no field dataset of $k_{o_2}$ exists in the type of temporal and spatial frequency that SWOT (and therefore the BIKER algorithm) provides.
@@
 This is done using equation 8, where $R_{h, obs} [L/T]$ is observed river channel area divided by observed channel width.
-With this setup, we can directly explore BIKER's ability to infer observed $U_\star$ and $k_{o_2}$ from *W* and *H* alone.
+This means that $k_{o_2, obs}$ relies on in situ river shear velocity, while BIKER is predicting shear velocity from *W* and $S_h$ alone.
+Therefore, we are directly exploring BIKER's ability to infer observed $U_\star$ and $k_{o_2}$ from *W* and *H* alone.
 It also means that, for a fair validation scheme, $\sigma_{k_{o_2}}^2$ must be set to reflect only error from our assumptions about $R_h$ and not the parameter uncertainity inherent in $\alpha$ (the 48 in equation 8).
@@
 It is one thing to accurately predict $k_{o_2}$, but researchers are often most interested in the actual gas fluxes from rivers and ultimately the carbon emitted from river to atmosphere.
-Therefore, we explore 1) the ability of BIKER to reproduce $CO_2$ evasion fluxes ($FCO_2$ $[M/L^2T]$) from these 49 rivers, and 2) the sensitivity of the carbon efflux efflux $[M/T]$ due to the chosen method to approximate $R_h$ and ultiamtely $U_\star$.
+Therefore, we explore 1) the ability of BIKER to reproduce $CO_2$ evasion fluxes ($FCO_2$ $[M/L^2T]$) from these 49 rivers, and 2) the sensitivity of the carbon efflux $[M/T]$ due to the chosen method to approximate $R_h$ and ultiamtely $U_\star$.
 
 First, we calculate $FCO_2$.
-To do this, we pair the aforementioned 26 $CO_2$ samples (section 2) with every 11th SWOT observation by date, ignoring the timesteps beyond 26 (only ~15% of the SWOT observations were ignored here and we deemed this acceptable).
-We chose to sample every 11 days as this is the average sampling resolution for SWOT.
+To do this, we pair the biweekly 26 $CO_2$ samples (section 2) with every 11th SWOT observation by date, ignoring the timesteps beyond 26 (only ~15% of the SWOT observations were ignored here and we deemed this acceptable).
+We chose to sample every 11 days as this is the average sampling resolution for SWOT and the $CO_2$ data is only every 14 days.
 Not all of the SWOT rivers have observations for a full year, and when simulation dates were not available they were assumed to start on January 1st.
 We also pair the modeled $k_{CO_2}$ values (obtained from $k_{o_2}$ and equation 9) with these water-side $CO_2$ concentrations and water temperatures.
-In equation 9, 530 is the Schmidt number for oxygen at 20 degrees Celsius (which BIKER predicts) and *Sc* is the Schmidt number at one's desired temperature for some dissolved gas.
+In equation 9, 530 is the Schmidt number for oxygen at 20 degrees Celsius (which is the temperature at which BIKER predicts *k*) and *Sc* is the Schmidt number at one's desired temperature for some dissolved gas.
 Atmospheric $CO_2$ was assumed 390 uatm.
@@
 
-Next, we estimate bulk carbon efflux using four models for average channel depth (used to calculate $U_\star$):  BIKER's posterior means and three gauge-based HG models previously used for $CO_2$ upscaling [@lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a; @raymondScalingGasTransfer2012a].
-See Table S1 for their definitions.
-It is worth stressing that the @lauerwaldSpatialPatternsCO22015 model is simply one of two components of the @raymondGlobalCarbonDioxide2013a model, and so was actually developed by the latter's authors.
+Next, we estimate bulk carbon efflux using five models for average channel depth (used to calculate $U_\star$):  BIKER's posterior means and four gauge-based HG models previously used for $CO_2$ upscaling [@horgbyUnexpectedLargeEvasion2019; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a; @raymondScalingGasTransfer2012a].
+See  for their definitions.
+It is worth stressing that the @lauerwaldSpatialPatternsCO22015 model is actually one of two components of the @raymondGlobalCarbonDioxide2013a model, and so was developed by the latter's authors.
 The names used here refer to the specific implementation used in each study.
+The @horgbyUnexpectedLargeEvasion2019 was developed explicitly for steep mountain streams, of which SWOT will not sample.
+We include it here to explore just how sensitive efflux estimates are to the HG model that is employed, noting that it will likely perform poorly.
 For this study, all HG methods use the in-situ discharge record while BIKER does not.
-This allows us to assess whether BIKER's estimates (wholly unguaged) are comparable to gauged methods (all three HG models).
+This allows us to assess whether BIKER's estimates (wholly ungauged) are comparable to gauged methods (all four HG models).
 Finally, we express the bulk carbon efflux as the average mass flow rate of carbon (via $CO_2$ evasion) per year from the 49 rivers after accounting for total river surface area.
@@
 ##4 Results
-First, we present the results from the BIKER validation on 49 SWOT rivers (section 4.1).
+First, we present the results from the BIKER validation on 49 SWOT rivers and on 17 SWOT rivers with measurement errors (section 4.1).
 Next, we compare BIKER inversion errors to the errors inherent in equations 2 and 3, as well as the rivers' hydraulic properties (section 4.2).
-Finally, we use 4 different models for $k_{CO_2}$ to compare $FCO_2$ and bulk carbon efflux estimates made using gauged and ungauged methods (section 4.3).
+Finally, we use five different models for $k_{CO_2}$ to compare $FCO_2$ and bulk carbon efflux estimates made using gauged and ungauged methods (section 4.3).
 
 ###4.1 Validating the BIKER algorithm
-Figure 2a plots the validation results for $k_{o_2}$ (with no SWOT measurement error) across all 49 rivers and all timesteps.
+Figure 3a plots the validation results for $k_{o_2}$ (with no SWOT measurement error) across all 49 rivers and all timesteps.
 The points are the posterior means while the black lines are the 95% confidence intervals (CIs) for the predictions.
-$k_{o_2}$ is strongly correlated with the $k_{o_2}$ predicted by BIKER.
-BIKER captures the general magnitude of the predictions and most points fall on or near the 1:1 line.
-However, the 95% prediction intervals (dashed grey lines) highlight a slight underestimation bias for many predictions and a handful of predictions that are significanyl underestimated.
+$k_{o_2}$ is strongly correlated with the BIKER-predicted $k_{o_2}$.
+Using absolutely no in situ information, BIKER captures the magnitude of the predictions and most points fall on or near the 1:1 line.
+However, the 95% prediction intervals (dashed grey lines) highlight a slight underestimation bias for many predictions and a small number of predictions that are significantly underestimated.
 The RMSE for the BIKER predictions is only `r round(results_3_2_all$rmse,2)` m/day) across all predictions.
 
-![Figure 2.
+![Figure 3.
 a: Validation of BIKER for 22 SWOT rivers.
@@
 
-Figure 2a highlights a handful of values that are grossly underestimated by BIKER, which is confirmed in Figure 2b.
-Figure 2b plots the cumulative density functions (CDFs) of observed and predicted $k_{o_2}$ where the green line is the CDF of the BIKER posterior means, the blue lines are the CDFs of the BIKER posterior 95% CIs, and the dashed black line is the CDF of observed $k_{o_2}$.
-We see, more clearly than in Figure 2a, that BIKER captures $k_{o_2}$ very well: most of the  observed CDF falls between the 95% CIs or on the posterior mean (Figure 2b).
-If we define a 'hit rate' as the oberved value following between the CIs, BIKER yields a hit rate of **get hit rate**.
-Most of the  **get inverse of hit rate**% are extremes: the uppermost quartile is systematically overestimated by BIKER, and a handful of the smallest $k_{o_2}$ in this data are underestimated.
-This highlights one benefit of using Bayesian inference to fully propagate prior and model uncertainties through to the posterior.
-In summary, Figure 2 confirms that we improve upon our baseline understanding of $k_{o_2}$ in these river: we accurately capture $k_{o_2}$ with no in situ information about the river while simultaneously and explicitly accounting for the uncertainties inherent in our $k_{o_2}$ estimates.
+Figure 3a highlights a small number of values that are grossly underestimated by BIKER, which is confirmed in Figure 3b.
+Figure 3b plots the cumulative density functions (CDFs) of observed and predicted $k_{o_2}$ where the green line is the CDF of the BIKER posterior means, the blue lines are the CDFs of the BIKER posterior 95% CIs, and the dashed black line is the CDF of observed $k_{o_2}$.
+We see, more clearly than in Figure 3a, that BIKER captures $k_{o_2}$ very well: most of the  observed CDF falls between the 95% CIs or on the posterior mean (Figure 3b).
+Most of the values poorly captured by BIKER are extremes: the uppermost quartile is systematically overestimated by BIKER, and a handful of the smallest $k_{o_2}$ in this data are underestimated (also visible in Figure 3a).
+In summary, Figure 3 confirms that we improve upon our baseline understanding of $k_{o_2}$ in these rivers: we accurately capture $k_{o_2}$ with no in situ information about the river while simultaneously and explicitly accounting for the uncertainties inherent in our $k_{o_2}$ estimates.
 
-Figure 3a plots validation metrics calculated for each river with and without SWOT measurement error (green and purple, respectively).
-The boxplots are composed of scores for either the 49 or 17 rivers- see Table 1 for metric definitions.
-SWOT measurement uncertainties slightly degrade performance across all four error metrics (Figure 3a), though caution should be used in over-interepting boxplots with n = 17.
-Therefore, we deem that SWOT measurement error does not exert a significant influence on BIKER and so the results presented for the rest of the manuscript assume no measurement error in order to use all 49 rivers.
+Figure 4a plots validation metrics calculated for each river with and without SWOT measurement error (green and purple, respectively).
+The boxplots are composed of validation scores for either the 49 or 17 rivers- see Table 1 for metric definitions.
+SWOT measurement uncertainties slightly degrade performance across all four error metrics (Figure 4a), though caution should be used in over-interepting boxplots with a sample size of only 17.
+We deem that SWOT measurement error does not exert a significant influence on BIKER and so the results presented for the rest of the manuscript assume no measurement error in order to use all 49 rivers.
 
 Median KGE is `r round(median(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$kge, na.rm=T),2)`, which is excellent given that absolutely no in situ information is being used to predict $k_{o_2}$.
-`r sum(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$kge > -0.41, na.rm=T)`/49 rivers score a KGE > -0.41 and `r sum(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$kge > 0, na.rm=T)`/49 are greater than 0, indicating that BIKER performs better than an estimate of the mean $k_{0_2}$.
-Median rBIAS is `r round(median(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$rBIAS, na.rm=T),2)`, indicating nearly no bias in most rivers' predictions.
+`r sum(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$kge > -0.41, na.rm=T)`/49 rivers score a KGE > -0.41 and `r sum(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$kge > 0, na.rm=T)`/49 are greater than 0, indicating that BIKER provides additional meaningful information in an ungauged setting.
+Median rBIAS is `r round(median(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$rBIAS, na.rm=T),2)`, suggesting nearly no bias in most rivers' predictions.
 However, some rivers are substantially biased in both directions (particuarly positive bias, where `r sum(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$rBIAS > 1, na.rm=T)` rivers have scores > 1).
-This further supports the visual evidence in Figure 2 that sometimes BIKER is substantially under/overestimating the magnitude of $k_{o_2}$ and that this is river-specific.
+This further supports the visual evidence in Figure 3 that sometimes BIKER is substantially under/overestimating the magnitude of $k_{o_2}$ and that this is river-specific.
 NRMSE and RMSE have median scores of `r round(median(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$nrmse, na.rm=T),2)` and `r round(median(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$rrmse, na.rm=T),2)`, respectively.
@@
 
-![Figure 3.
+![Figure 4.
 a: Performance metrics by river.
@@
 
-Figure 3b-d are representative timeseries plots of predicted and observed $k_{o_2}$ for three rivers chosen randomly from those with 'good' KGE scores (b), 'okay' KGE scores (c), and 'bad' KGE scores (d).
+Figure 4b-d are representative timeseries plots of predicted and observed $k_{o_2}$ for three rivers chosen randomly from those with 'good' KGE scores (b), 'okay' KGE scores (c), and 'bad' KGE scores (d).
 See the Figure 3 caption for how this was determined.
@@
 Interestingly, there is near perfect recovery of $k_{o_2}$ for the first approximately 150 days.
-In the Connecticut River, there is significant positive bias in the estimates as well as large uncertainty (per the 95% CIs) and temporal dynamics are largely missing from BIKER's predicitions.
+In the Connecticut River, there is significant positive bias in the estimates but also massive uncertainty (per the 95% CIs) in those estimates.
+Temporal dynamics are also largely missing from BIKER's Connecticut River predicitions.
 
 ###4.2 BIKER performance versus *k* model performance
-The MC simulations described in section 3.2.2 yielded an overall BIKER uncertainity of **must calculate** (**Figure S4, need to do**).
+The MC simulations described in section 3.2.2 yielded an overall BIKER uncertainity of **must update alpha uncertainty** (Figure S4).
 While this will be useful for running BIKER on actual SWOT data, recall that the validation presented here (section 4.1) accounts for parameter error for a fair comparison against the 'observed' $k_{o_2}$ values (section 3.3).
-In that context, figure 4 compares BIKER inversion error per river and via NRMSE to the error generated by process-level uncertainities.
-This is the dashed black line in Figure 4 and is the reported cross-validated model performance for equation 3 [@wangPhysicallyBasedScaling2021]: approximately 57%.
-Note that in that paper they use the same metric but refer to it as 'relative error' and so we do too in this section.
-43/49 rivers have lower inversion errors than the cross-validated model error for equation 3 (~57%), with most rivers' scores are far less than 57%.
+In that context, figure 5 compares BIKER inversion error per river and via NRMSE to the error generated by process-level uncertainities.
+This is the dashed black line in Figure 5 and is the reported cross-validated model performance for equation 3 [@wangPhysicallyBasedScaling2021].
+This is approximately 57%.
+Note that @wangPhysicallyBasedScaling2021 calculate the NRMSE but refer to it as 'relative error'.
+43/49 rivers have lower inversion errors than the cross-validated model error for equation 3 (~57%), with most rivers' scores far less than 57% (Figure 5).
 The median relative error in BIKER inversion is `r round(median(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$nrmse, na.rm=T),2)` which is less than 1/2 of the reported error for $\alpha$.
 For the six rivers with inversion errors greater than the parameter error, 4/6 are approaching or over 100% error.
-This means that BIKER inversion errors are nearly always far less than the parameter error, however occasionally the inversion is very poor and introduces much more error than equation 3 does.
+This means that BIKER inversion errors are nearly always far less than the parameter error, however occasionally the inversion is very poor and introduces significantly more error than equation 3 does.
 It should also be stressed that NRMSE is one of the worst perfoming metrics for BIKER (figure 3a) and so a similar comparison using KGE or rBIAS would likely yield even stronger results.
-Therefore, BIKER introduces a small amount of additional error into the physical model for $k_{o_2}$ and any errors in the final $k_{o_2}$ estimates are dominated by process-level uncertainity in predicting $k_{o_2}$ via equation 3 rather than uncertainity in inverting SWOT observations to estimate $k_{o_2}$.
+In summary, BIKER introduces a small amount of additional error into the predictive model for $k_{o_2}$ and any errors in the final $k_{o_2}$ estimates are dominated by process-level uncertainity in predicting $k_{o_2}$ via equation 3 rather than uncertainity in inverting SWOT observations to solve equation 3.
 
-![Figure 4: NRMSE (referred to as relative error in @wangPhysicallyBasedScaling2021) for all 49 rivers.
+![Figure 5: NRMSE, referred to as relative error by @wangPhysicallyBasedScaling2021, for all 49 rivers.
 Dashed line is the cross-validation relative error for the ko2 model used in this study (~57%), reprinted from @wangPhysicallyBasedScaling2021.
@@
 
-Figure 4 shows that BIKER performs extremely poorly on 6/49 rivers.
-To explore why, we also compared BIKER performance against river hydraulic properties mean observed $k_{o_2}$, *W*.
-and $S_h$.
-We found no clear patterns between these properties and KGE, NRMSE, or RRMSE (Figures S5, **S6, and S7 to do**, respectively).
-However, Figure 5 plots the patterns betwwen these hydraulic properties and rBIAS.
-There is an apparent uniform and positive bias in BIKER's predicitions in the narrowest and steepest of these 49 rivers.
-For these 49 rivers, these are approximately those with slopes > 0.001 (Figure 5c) and widths < 100m (Figure 5b).
-Following basic gas exchange theory [@ulsethDistinctAirWater2019a], these are also the rivers with the greatest $k_{o_2}$ values (Figure 5a).
-**Check if this is the sam 6 rivers....**.
+**This paragraph needs some work**
+Figures 4 and 5 show that BIKER performs extremely poorly on 6/49 rivers.
+To explore this, and more broadly why some rivers perform well and others do not, we also compared BIKER performance against the following river hydraulic properties: mean observed $k_{o_2}$, *W*, and $S_h$.
+We found two patterns that emerged from this analysis.
+First, rBIAS (figure 6) exhibited no dataset-wide correlations with hydraulic properties, however we did find a uniform, positive bias in BIKER's predictions in the narrowest and steepest rivers.
+For the 49 rivers used in this study, these are rivers with slopes approximately > 0.0005 (Figure 5c) and widths < 100m (Figure 6b).
+Following basic gas exchange theory, these are also the rivers with the greatest $k_{o_2}$ values (Figure 6a).
+Second, KGE, NRMSE, and RRMSE (Figures S5, S6, and S7, respectively) similarly showed no dataset-wide correlations with the hydraulic properties.
+However for the seven widest rivers tested (> 1000m wide), not a single river performed extremely poorly, as subjectively interpreted from Figures S5, S6, and S7.
+Taken in aggregate, this means that a river's hydraulic properties alone cannot predict how BIKER will perform, however narrow/steep rivers will likely have positive bias in their predictions and the widest rivers will likely perform adequately (not not necessiarly great).
 
-![Figure 5: BIKER performance (defined by rBIAS- Table 1) versus (a) mean river ko2, (b) mean river width, and (c) mean river slope.
+![Figure 6: BIKER performance (defined by rBIAS- Table 1) versus (a) mean river ko2, (b) mean river width, and (c) mean river slope.
 An rBIAS of zero (dashed black line) indicates no bias in the predictions.](cache/validation/riverProperties_rbias.jpg)
 
-###4.3 Influences on carbon efflux estimates
+###4.3 Predicting carbon efflux from large rivers
 Finally, we explore our ability to 1) use BIKER-produced $k_{o_2}$ to estimate $FCO_2$ and 2) estimate bulk carbon efflux from $CO_2$ evasion (section 3.4).
 
-In Figure 6a, there is a strong fit to the observed $FCO_2$ data, with an RMSE of `r round(10^results_3_3_all$rmse,2)` $g/m^2dy$.
-The $r^2$ is slightly lower than $k_{o_2}$ (Figure 6a), likely due the the handful of extremely underestimated $FCO_2$ predictions.
+In Figure 7a, there is a very strong fit to the observed $FCO_2$ data, with an RMSE of `r round(results_3_3_all$rmse,2)` $g/m^2dy$.
+The $r^2$ is just slightly lower than $k_{o_2}$ (Figure 6a).
 Overall, there is less systematic bias in the predicitons across all 49 rivers than in the $k_{o_2}$ predictions (Figure 3a).
 This is presumably due to the structure of the $FCO_2$ equation, which reduces the relative importance of errors in *k* given that the $CO_2$ data is measured in situ.
-$FCO_2$ prediction intervals are wider than those presented in Figure 2a.
-Figure 6b-d includes subplots for the same rivers as Figure 3b-d, however with $FCO_2$ plotted instead of $k_{o_2}$.
-There is very good recovery of $FCO_2$ in both the Missouri Downstream River and the Ohio Section 8 river (Figures 6b and 6c), with both magnitude and temporal dynamics modeled quite well.
-The Connecticut river is systematically over estimated (Figure 6d), though BIKER's CIs do reflect reasonably uncertain estimates and the temporal dynamics are not far from the observed dynamics.
+$FCO_2$ prediction intervals are wider than those presented in Figure 3a.
+Figure 7b-d includes subplots for the same rivers as Figure 3b-d, however with $FCO_2$ plotted instead of $k_{o_2}$.
+There is near perfect recovery of $FCO_2$ in the Missouri Downstream River and very good recovery in the Ohio Section 8 river (Figures 7b and 7c, respectively).
+Both the magnitude and temporal dynamics are modeled very well, with the Ohio Section 8 River experiencing a small amount of bias in some later predictions.
+The Connecticut river is systematically over estimated aside from the last four points (Figure 7d), though BIKER's CIs do reflect quite uncertain estimates and the temporal dynamics are not far from the observed dynamics.
 This suggests that BIKER can easily estimate the changing dynamics of gas exchange but is at the mercy of the prior on $k_{o_2}$ to accurately estimate the magnitude of the evasion.
 
-![Figure 6.
-a: FCO2 via BIKER versus via equations 2 and 3 for every 11th timestep for the 49 rivers (grey lines are linear regression and 95% prediction intervals, while black dashed line is the 1:1 line).
+![Figure 7: a: FCO2 via BIKER versus via equations 2 and 3 for every 11th timestep for the 49 rivers (grey lines are linear regression and 95% prediction intervals, while black dashed line is the 1:1 line).
 b-d: timeseries plots for the three example rivers from Figure 3b, 3c, and 3d.](cache/FCO2/FCO2_plot.jpg)
 
-Finally, we compare $FCO_2$ and bulk carbon efflux (via $CO_2$ evasion) from the 49 rivers using BIKER posterior means and three gauge-based HG models (Figure 7).
-Figure 7a are barplots of the bulk carbon efflux (via $CO_2$ evasion) across the 49 rivers in gigagrams of carbon per year.
-The BIKER bulk carbon efflux (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_BIKER',]$sumFCO2, 0)` gG-C/yr) is nearly identical to that predicted by the 'raymond 2013' model(`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_raymond2013',]$sumFCO2, 0)`), though both are somewhat overestimated relative to the observed flux (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_ulset',]$sumFCO2, 0)`).
-The 'raymond 2012' model is the closest (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_raymond2012',]$sumFCO2, 0)`), while the 'Lauerwald 2015' model grossly overestimates this bulk efflux (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_Lauerwald2015',]$sumFCO2, 0)`).
-Thus, despite BIKER using absolutely no in situ data, it provides similar estimates of the carbon efflux to two of the in-situ approaches and surprisingly outperforms the 3rd in-situ method (Figure 7a).
+Finally, we compare $FCO_2$ and bulk carbon efflux (via $CO_2$ evasion) from the 49 rivers using BIKER posterior means and four gauge-based HG models (Figure 8).
+Figure 8a are barplots of the bulk carbon efflux (via $CO_2$ evasion) across the 49 rivers in gigagrams of carbon per year.
+Both the BIKER bulk carbon efflux (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_BIKER',]$sumFCO2, 0)` gG-C/yr) and the 'Raymond 2012' estimate (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_raymond2012',]$sumFCO2, 0)`) are quite close to the observed flux (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_obs',]$sumFCO2, 0)`), though the former is overestimated and the latter is underestimated.
+The 'raymond 2013' model is slightly more overestimated (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_raymond2013',]$sumFCO2, 0)`), while the 'Lauerwald 2015' model grossly overestimates this bulk efflux (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_lauerwald2015',]$sumFCO2, 0)`).
+As expected, the 'Horgby 2019' model is ill-suited for rivers this large and produces signifiantly wrong estimates (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_horgby2019',]$sumFCO2, 0)`).
+Thus, despite BIKER using absolutely no in situ data, it provides similar estimates of the carbon efflux to one of the in-situ approaches and the observed efflux itself (Figure 8a).
 
-![Figure 7.
+![Figure 8.
 a: Bulk carbon efflux, per year, from the 22 SWOT rivers as calculated using four different average flow velocity estimates: 1) BIKER, 2) through 4) average flow velocity as calculated using HG models from the literature.
-b: Histograms of by-river performance in estimating FCO2 across all timesteps and rivers for the same four velocity models.
+b: Boxplots of by-river performance in estimating FCO2 across all timesteps and rivers for the same five velocity models.
 Dashed lines are identical to Figure 6a.](cache/FCO2/FCO2_models.jpg)
 
-Figure 7b plots the by-river performance scores for $FCO_2$.
-In line with the Figure 7a results, 'Lauerwald 2015' is the worst performing (median KGE: `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_Lauerwald2015',]$kge, na.rm=T),2)`) while BIKER (median KGE: `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_BIKER',]$kge, na.rm=T),2)`) is only slightly worse than the other two in-situ methods (median KGEs: `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_raymond2013',]$kge, na.rm=T),2)` and `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_raymond2012',]$kge, na.rm=T),2)` for 'Raymond 2013' and 'Raymond 2012', respectively).
-For rBIAS, BIKER indicates the least bias in some rivers and similar bias to the other models in other rivers, as indicated by the scores' inter-quartile range (`r round(IQR(results_3_3_rivs[results_3_3_rivs$key=='FCO2_BIKER',]$rBIAS, na.rm=T),2)`) and median (`r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_BIKER',]$rBIAS, na.rm=T),2)`).
-Meanwhile, both 'Lauerwald 2015' and 'Raymond 2013' produce significant positive bias in their estimates (median scores of `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_Lauerwald2015',]$rBIAS, na.rm=T),2)` and `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_raymond2013',]$rBIAS, na.rm=T),2)`, respectively).
-Again, the 'Raymond 2012' model performs the best with virtually no bias (`r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_raymond2012',]$rBIAS, na.rm=T),2)`).
-BIKER and 'Lauerwald 2015' have similar performance for NRMSE and RRMSE, which is generally worse than the other two models.
-In summary, BIKER performance across all four metrics is either similar or slightly worse than two of the in-situ approaches tested here and better than the 'Lauerwald 2015' model.
+Figure 8b plots the by-river performance scores for $FCO_2$.
+In line with the Figure 7a results, 'Lauerwald 2015' and 'Horgby 2019' are the worst performing (median KGEs, respectively: `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_lauerwald2015',]$kge, na.rm=T),2)` and `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_horgby2019',]$kge, na.rm=T),2)`) while BIKER (median KGE: `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_BIKER',]$kge, na.rm=T),2)`) is on par with the other two in-situ methods (median KGEs: `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_raymond2013',]$kge, na.rm=T),2)` and `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_raymond2012',]$kge, na.rm=T),2)` for 'Raymond 2013' and 'Raymond 2012', respectively).
+For rBIAS, both BIKER and 'Raymond 2012' indicate the least bias in some rivers and similar bias to the other models in other rivers.
+For BIKER, this is indicated by the scores' inter-quartile range (`r round(IQR(results_3_3_rivs[results_3_3_rivs$key=='FCO2_BIKER',]$rBIAS, na.rm=T),2)`) and median (`r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_BIKER',]$rBIAS, na.rm=T),2)`).
+Meanwhile, both 'Lauerwald 2015' and 'Horgby 2019' produce significant bias in their estimates: 'Lauerwald 2015' has a median rBIAS of `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_lauerwald2015',]$rBIAS, na.rm=T),2)`, while 'Horgby 2019' had `r round(median(results_3_3_rivs[results_3_3_rivs$key=='FCO2_horgby2019',]$rBIAS, na.rm=T),2)` .
+Similar patterns play out for NRMSE and RRMSE.
+In summary, BIKER performance across all four metrics is either similar or slightly worse than one of the in-situ approaches tested here and better than the the other three in situ approaches.
 This is despite relying on absolutely no streamgauge data like all three other models do.
@@
 SWOT will provide daily hydraulic measurements for a 3 month fast sampling period for calibration and validation and sampling thereafter between 1 and 7 days per 21 day repeat cycle [@biancamariaSWOTMissionIts2016].
-BIKER's success in 1) infering $k_{o_2}$ using simulated SWOT data over a wide range of rivers (Figures 2 and 3) and 2) being robust to measurement errors internal to the SWOT data (Figures 2) bode well for BIKER's eventual implementation on real SWOT observations.
+BIKER's success in 1) infering $k_{o_2}$ using simulated SWOT data over a wide range of rivers (Figures 3 and 4) and 2) being robust to measurement errors internal to the SWOT data (Figure 3) bode well for BIKER's eventual implementation on real SWOT observations.
 The results presented here suggest that daily estimation of riverine gas exchange globally could be possible once SWOT launches.
@@
 ###5.2 Estimating bulk carbon efflux using SWOT
-Section 4.3 confirms that BIKER is successful, without any in situ information, at predicting 1) $FCO_2$ (Figure 6) and 2) the bulk carbon efflux (Figure 7).
+Section 4.3 confirms that BIKER is successful, without any in situ information, at predicting 1) $FCO_2$ (Figure 7) and 2) the bulk carbon efflux (Figure 8).
 This encouraging result has two main implications for future work coupling remote sensing via SWOT and in situ data.
@@
 
-Secondly, it is important to stress that unlike BIKER, the HG models in Figure 7 rely on an in situ streamgauge.
-This means that Figure 7 represents the best performance that those models could ever have; if ran using modeled discharge their accuracy would necessarily decrease.
+Secondly, it is important to stress that unlike BIKER, the HG models in Figure 8 rely on an in situ streamgauge.
+This means that Figure 8 represents the best performance that those models could ever have; if ran using modeled discharge their accuracy would necessarily decrease.
 Thus, these results suggest that BIKER will be useful in two settings: 1) upscaling in ungauged rivers as hypothesized, but also in 2) potentially improving our carbon efflux understandings at gauged sites.
@@
 
-Figure 7 also confirms that the training data used for HG models exerts a significant influence on upscaled carbon emissions from rivers.
+Figure 8 also confirms that the training data used for HG models exerts a significant influence on upscaled carbon emissions from rivers.
 The 'Lauerwald 2015' model, which was trained on larger rivers than the data used in either the 'Raymond 2013' or 'Raymond 2012' models, should perform best on SWOT-observable rivers (which are greater than 50m wide).
 Figure 8 confirms this.
-Meanwhile, BIKER has no similar reliance on hydraulic parameters trained on different river sizes and only assumes that the channel's hydraulic radius can be approximated by mean flow depth (which is generally the case in rivers large enough to be SWOT-observable- text Sx and Table Sx).
+Meanwhile, BIKER has no similar reliance on hydraulic parameters trained on different river sizes and only assumes that the channel's hydraulic radius can be approximated by mean flow depth (which is generally the case in rivers large enough to be SWOT-observable- **text Sx**).
 Upon SWOT's launch, the BIKER approach to estimating *k* could be coupled with ethier existing upscaling workflows or even explicit $CO_2$ transport models [@brinkerhoffLakeMorphometryRiver2021; **Saccardi & Winnick in review**] to improve riverine gas flux predictions where gauges are unavailable but SWOT measurements are.
@@
 Throughout the BIKER validation, we have assumed no parameter uncertainity in the upscaling parameters $\alpha$.
-We have shown that BIKER estimate uncertainty is almost always less than $\alpha$ uncertainity, and usually far less (Figure 4).
-Implicit measurement errors in SWOT data also exert a trivial influence on BIKER accuracy (Figure 2).
+We have shown that BIKER estimate uncertainty is almost always less than $\alpha$ uncertainity, and usually far less (Figure 5).
+Implicit measurement errors in SWOT data also exert a trivial influence on BIKER accuracy (Figure 4a).
 Therefore, most of the total BIKER uncertainity stems from the predictive model itself.

### 2021-06-04 supp_v3.Rmd
some ms updates

@@
 Colors correspond to whether a measurement is SWOT-observable, i.e.
-the river's width is > 50m.
-The k600 ~ eD model fundamentally breaks down in SWOT-observable rivers](cache/k600/roughness_theory/eM_k600.jpg)
+the river's width is > 100m.
+The k600 ~ eD model fundamentally breaks down in SWOT-observable rivers](cache/k_eD/eD_k600.jpg)
 
@@
 
-![Figure S4: plot of overall BIKER uncertainity](cache/validation/riverProperties_kge.jpg)
+![Figure S4: plot of overall BIKER uncertainity. THESE ARE STAND IN RESULTS AND NEED TO BE UPDATED](cache/MonteCarlo/MC_results.jpg)
 
@@
 
-![Figure S6: BIKER performance, defined by NRMSE (see Table 1 in the main text for the metric definition) versus mean river ko2, width, and slope.](cache/validation/riverProperties_kge.jpg)
+![Figure S6: BIKER performance, defined by NRMSE (see Table 1 in the main text for the metric definition) versus mean river ko2, width, and slope.](cache/validation/riverProperties_nrmse.jpg)
 
-![Figure S7: BIKER performance, defined by RRMSE (see Table 1 in the main text for the metric definition) versus mean river ko2, width, and slope.](cache/validation/riverProperties_kge.jpg)
+![Figure S7: BIKER performance, defined by RRMSE (see Table 1 in the main text for the metric definition) versus mean river ko2, width, and slope.](cache/validation/riverProperties_rrmse.jpg)
 
@@
 |-------------|-------------|-------------|-------------|-------------|
-|Raymond 2012|$ln(V)=-1.64+0.285ln(Q)$|Trained on their dataset|@raymondScalingGasTransfer2012a|
-|Raymond 2013|$ln(V)=mean(-1.64+0.285ln(Q),  -1.06+0.12ln(Q))$ |Average of the Raymond 2012 equation and one trained on 9,811 USGS streamgauges|@raymondGlobalCarbonDioxide2013a|
-|Lauerwald 2015|$ln(V)=-1.06+0.12ln(Q)$ |Borrowed from 1/2 of the Raymond 2013 calculation (trained on 9,811 USGS streamgauges)|@lauerwaldSpatialPatternsCO22015|
+|Raymond 2012|$D=0.409Q^{0.294}$|1,026 measurements across the United States|@raymondScalingGasTransfer2012a|
+|Raymond 2013|$D=0.409Q^{0.294},  D=0.449Q^{0.37}$ |Average of the Raymond 2012 equation and one using 9,811 measurements at US streamgauges|@raymondGlobalCarbonDioxide2013a|
+|Lauerwald 2015|$D=0.449Q^{0.37}$ |Borrowed from 1/2 of the Raymond 2013 calculation (trained on 9,811 measurements at US streamgauges).
+This was used on all rivers of at least stream order 4, which we assume applies to all of these SWOT-observable rivers|@lauerwaldSpatialPatternsCO22015|
+|Horgby 2019|$D=0.298Q^{0.222}$ |141 measurements made in 12 mountain streams in the Swiss Alps|@horgbyUnexpectedLargeEvasion2019|
 

### 2021-06-04 src/buildMap.R

@@ -1,0 +1,42 @@
+#Craig Brinkerhoff
+#pepsi 2 river locations map
+#Summer 2021
+
+library(tidyverse)
+library(tmap)
+library(cowplot)
+library(RColorBrewer)
+library(sf)
+library(grid)
+theme_set(theme_cowplot())
+
+setwd('C:\\Users\\craig\\Documents\\GitHub\\RSK600')
+
+##########
+#Read in locations and rivers----------------
+#########
+data <- read.csv('data\\dataLocations.csv', header = T)
+colnames(data) <- c('River', 'lat', 'lon', 'Data')
+#data <- data[-1,]
+data$lat <- as.numeric(as.character(data$lat))
+data$lon <- as.numeric(as.character(data$lon))
+
+global_rivs <- st_read('data\\ne_10m_rivers_lake_centerlines\\ne_10m_rivers_lake_centerlines.shp')
+global_land <- st_read('data\\ne_50m_land\\ne_50m_land.shp')
+data <- st_as_sf(data, coords = c('lon', 'lat'))
+st_crs(data) <- st_crs(global_rivs)
+
+##########
+#main map------------
+##########
+base <- tm_shape(global_land) +
+  tm_polygons(col = '#99d8c9') +
+  tm_compass(position = c('RIGHT', 'TOP')) +
+  tm_scale_bar(width=1/10, position=c('RIGHT', 'BOTTOM'), lwd=2)
+map <- tm_shape(global_rivs)+
+  tm_lines(col='darkblue', scale=0.75)
+rivers <- tm_shape(data) +
+  tm_symbols(col =  "Data", size=0.5, shape=21, palette = c('#e31a1c', '#33a02c', '#1f78b4'), border.col = 'black', border.lwd = 1)
+map <- base + map + rivers
+
+tmap_save(map, 'cache/map.jpg')

### 2021-06-04 src/validation.R

@@ -179,7 +179,7 @@ write.csv(stats_by_reach, 'cache/validation/results_by_riv.csv')
 ################
 ##RBIAS VS HYDRAULIC PROPERTIES---------------------------------------
 ################
-riverPropertiesPlot_k <- ggplot(stats_by_reach, aes(y=rBIAS, x=meanKobs)) +
+riverPropertiesPlot_k <- ggplot(stats_by_reach[stats_by_reach$errFlag==0,], aes(y=rBIAS, x=meanKobs)) +
   geom_point(size=3) +
   geom_hline(yintercept=0, linetype='dashed') +
   xlab('Mean observed k [m/dy]') +
@@ -191,7 +191,7 @@ riverPropertiesPlot_k <- ggplot(stats_by_reach, aes(y=rBIAS, x=meanKobs)) +
         axis.title=element_text(size=24,face="bold"),
         legend.text = element_text(size=17),
         legend.title = element_text(size=17, face='bold'))
-riverPropertiesPlot_W <- ggplot(stats_by_reach, aes(y=rBIAS, x=meanWobs)) +
+riverPropertiesPlot_W <- ggplot(stats_by_reach[stats_by_reach$errFlag==0,], aes(y=rBIAS, x=meanWobs)) +
   geom_point(size=3) +
   geom_hline(yintercept=0, linetype='dashed') +
   xlab('Mean observed width [m]') +
@@ -204,7 +204,7 @@ riverPropertiesPlot_W <- ggplot(stats_by_reach, aes(y=rBIAS, x=meanWobs)) +
         axis.title=element_text(size=24,face="bold"),
         legend.text = element_text(size=17),
         legend.title = element_text(size=17, face='bold'))
-riverPropertiesPlot_S <- ggplot(stats_by_reach, aes(y=rBIAS, x=meanSobs)) +
+riverPropertiesPlot_S <- ggplot(stats_by_reach[stats_by_reach$errFlag==0,], aes(y=rBIAS, x=meanSobs)) +
   geom_point(size=3) +
   geom_hline(yintercept=0, linetype='dashed') +
   xlab('Mean observed slope') +
@@ -228,7 +228,7 @@ ggsave('cache/validation/riverProperties_rbias.jpg', riverPropertiesPlot, width=
 #########################
 ##KGE VS HYDRAULIC properties-------------------------------------------------------
 #########################
[…]

### 2021-07-19 manuscript_v4.Rmd
ms draft 4 updates

---
title: "Gas exchange in large rivers controlled by largest turbulent eddies: implications for remotely sensing gas exchange via SWOT"
author: 'CB Brinkerhoff, CJ Gleason, PA Raymond, MH Harlan'
date: "`r Sys.Date()`"
bibliography: References.bib
csl: agu_citation_style.csl
output:
  word_document: 
      reference_docx: word-styles-reference.docx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Key Points
- In large rivers, there exists a local equilibrium between production and dissipation of turbulent kinetic energy at the free surface
- Gas exchange in large rivers is equivalently parameterized across multiple turbulence scales in open-channel flows
- BIKER algorithm exploits this theory, predicting gas exchange velocity and $CO_2$ fluxes from simulated SWOT satellite data and in situ $CO_2$ data

##Keywords
gas exchange, open-channel flow, remote sensing, SWOT, biogeochemistry, ungauged basin

##Abstract
Go BIKER!!
*AGU Advances. So 250 words here*
*AGU advances is 8000 words total, though the format of the ones I've looked at vary quite a bit from long letter to short article. Many are clearly shorter than 8000 I think*

##Plain Language Summary
Necessary for AGU advances (200 words)

##1 Introduction
Since the publication of @colePlumbingGlobalCarbon2007a, the global river network is recognized to emit substantial amounts of carbon in addition to exporting to the oceans. Current estimates of total carbon dioxide ($CO_2$) evasion to the atmosphere from the global river network vary from 650-1800 Tg C/yr [@lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a], with 167 Tg-C/yr coming from mountain streams alone [@horgbyUnexpectedLargeEvasion2019]. Despite its incredibly small percentage of the global land surface, this is on par with the total oceanic $CO_2$ uptake rate [@gruberOceanicSinkAnthropogenic2019] and the global forest carbon uptake rate [@panLargePersistentCarbon2011]. Though this flux is still relatively poorly constrained, it plays a critical role in the global carbon cycle and is of great interst to aquatic biogeochemists. Equation 1 represents this flux for any sparingly soluble gas given the gas concentration gradient between the water $C_{water}$ and the air $C_{air}$ and the gas exchange velocity *k*.

$$\mathbf{(1)} Flux=k(C_{water}-C_{air})$$

The structure of equation 1 necessitates that calculations of this flux are highly sensitive to one's measurements/estimates of *k*. More broadly, *k* is also of significant importanceto aquatic ecologists modeling stream metabolism [@bernhardtMetabolicRegimesFlowing2018], water quality engineers [@chapraSurfaceWaterQualityModeling2008], and general interfacial mass transport [@lorkeUnifiedScalingRelation2006]. Decades of work have focused on elucidating the physical mechanisms behind *k* in oceans and lakes and, to a lesser extent, fluvial gas exchange [see @wanninkhofAdvancesQuantifyingAirSea2009 and @hallGasExchangeStreams2020 for reviews on gas exchange in in oceans and rivers, respectively]. Given that *k* should scale with turbulence in a fluid flow, a generalized model for *k* (equation 2) relies on the characteristic turbulence velocity scale $v_K$, which itself is a function of the characteristic length and time scales [@katulStructureFunctionModel2018]. Extensive field and labratory experiments have confirmed the validity of this model when $v_K$ is set to the Kolmogorov velocity scale [e.g. @lamontEddyCellModel1970; @zappaEnvironmentalTurbulentMixing2007; @lorkeUnifiedScalingRelation2006; @zappaVariationSurfaceTurbulence2003 **MORE**], where the idea is that the turbulent kinetic energy (TKE) cascade necessitates that TKE is produced by the largest, integral-scale turbulent eddies in the flow and then passed down the cascade to progressively smaller eddies until it is dissipated as heat at the Kolmogorov-scale: the smallest eddies in a turbulent flow. Equation 2 via Kolmogorov-scale eddies can also be derived from both classic thin-film and surface renewal theories for gas exchange [@wangPhysicallyBasedScaling2021].
$$\mathbf{(2)} k\sim Sc^{-n}v_K$$

While this model works reasonably well in non-fluvial environments, there is considerable uncertainity in how it applies to fluvial systems. Specifically, the influence of the riverbed and its large roughness elements in small streams (i.e. rapids) complicates scaling fluvial *k* and might lead to substantially higher *k* in streams with whitewater [e.g. @hallAirWaterOxygen2012; @ulsethDistinctAirWater2019a]. Conversely, even less attention has been paid to the other side of the stream-to-river continumn, where large rivers' hydraulic properties begin to mimic small lakes and estuaries. To date, the handful of existing field studies of large-river *k* have suggested that their *k* properties begin to mimic small lakes with respect to wind dynamics, though little else is well-established [@alinPhysicalControlsCarbon2011; @beaulieuControlsGasTransfer2012a; @wangPhysicallyBasedScaling2021]. Further, these mechanistic uncertainties are additionally limited by a large dearth of field-measured fluvial *k*. @wangPhysicallyBasedScaling2021 attempted to address this by simulating *k* in 35 rivers using stream metabolism modeling [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets, though this is still limited by the biases inherent in the DO data. This is all exacerbated in ungauged basins that cover large areas, especially in the carbon-rich Arctic inland waters, where little in situ information is available and fieldwork is impractical [@gleasonRemoteSensingRiver2020].

These mechanistic uncertainties are then propogated through upscaling workflows when biogeochemists predict *k* across thousands of rivers [e.g @borgesGloballySignificantGreenhousegas2015a; @horgbyUnexpectedLargeEvasion2019; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a] using equation 3 coupled with hydraulic geometry (HG), or the scaling relationships between streamflow and river channel hydraulics [@leopoldHydraulicGeometryStream1953a]. It is currently not well understood how sensitive global estimates of fluvial gas evasion are to the specific hydraulic geometry model that is employed by the worker. Further, these approaches rely on either in situ discharge records or modeled streamflow which introduces either additional uncertainity or the problems associated with ungauged rivers.

A potential alternative to this upscaling approach is to directly estimate a river's hydraulic properties from remote sensing (RS) data. Remote sensing of river hydraulics is a burgeoning subfield within remote sensing of hydrology, often in service of remote sensing of river discharge (RSQ- see @gleasonRemoteSensingRiver2020 for a recent review). This is accomplished via two general approaches: 'gauged' methods which rely on in situ river data to calibrate one's method to the river(s) at hand and 'ungauged' techniques which focus on generalizability and global implementation in the service of merely improving existing knowledge in data-poor domains [@gleasonRemoteSensingRiver2020]. Many, but not all, of these ungauged approaches were developed in the context of the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission. SWOT is expected to launch in 2022 and provide the world's first global measurements of water surface extent and elevation at novel temporal resolutions. SWOT is a wide swath radar interferometer and will sample rivers every 1 to 7 days per 21 day repeat cycle. This yields an average sampling resolution of 11 days. It will measure rivers wider than 100m with a goal of expanding this to rivers at least 50m wide [@biancamariaSWOTMissionIts2016]. A decade of SWOT work has explored the multi-parameter problem of estimating the river hydraulic parameters of roughness and bathymetry from remote sensing to produce the SWOT discharge product [e.g. @andreadisConstrainingAssimilationSWOT2020d; @brinkerhoffConstrainingRemoteRiver2020a; @durandEstimatingReachaveragedDischarge2014a; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014a; @garamboisVariationalEstimationEffective2020; @brissetAssimilationAltimetricData2018; @hagemannBAMBayesianAMHGManning2017a; @larnierRiverDischargeBathymetry2020a; @oubanasDischargeEstimationUngauged2018].

In this context, here we revist the fundamental mechanisms behind gas exchange in large rivers using a simple TKE budget balance and one of the largest available datasets of field-measured river hydraulics [@brinkerhoffReconcilingAtaStationAtManyStations2019] to answer the following question: does *k* behave fundamentally differently in large rivers versus other streams and how does this relate to the classical model used to explain gas evasion in non-fluvial aquatic systems (equation 2, section 3)? We exploit the findings from this simple analysis to develop a novel methodology that predicts *k* and its explicit uncertainity solely using SWOT observations (section 4). It requires no in situ inputs of any kind (although in situ data can be ingested and will improve results). We name the RS of $k_{600}$ algorithm BIKER, or the 'Bayesian Inference/Inversion of the $k_{600}$ Evasion Rate' and validate it for 47 SWOT-observable rivers from around the world using hydraulic models to produce SWOT-like data (as SWOT has not yet launched- section 2). We also quantify BIKER's sensitivity to the expected SWOT measurement errors on 17 of those rivers: while SWOT data represent a sea change in inland water monitoring, it is expected to have an approximately 10cm error in water surface elevation [@biancamariaSWOTMissionIts2016] as well as river width errors [@frassonExploringFactorsControlling2021]. Finally, we use previously published dissolved $CO_2$ data to represent a hypotetical in situ $CO_2$ sensor and compare the bulk carbon efflux from the 47 rivers as calculated using BIKER and previously published in situ techniques for predicting $k_{600}$.

##2 Data
Numerous datasets were used in this study. Please see Figure 1 for a map of the approximate locations for the data used in this study. We also provide a flowchart detailing the entire study as Figure S1.

![Figure 1: Map of the 47 hydraulic models and 1 timeseries of CO2 samples used in this study. Note that hydraulic model locations are approximate as most of the models are not geo-referenced. We additionally used over 530,000 discrete measurements of river channel hydraulics from across the continetal United States (Brinkerhoff et al. 2019) and over 700 gas exchange velocity measurements (Ulseth et al. 2019) that are not mapped here.](cache/maps/map_fin.jpg)

TKE and gas exchange theory (section 2) are explored using two datasets. First, we use a previously published compilation of field measurements that were originally made to calibrate United States Geological Survey (USGS) streamgauge rating curves and collected by @brinkerhoffReconcilingAtaStationAtManyStations2019. That dataset contains over 530,000 unique measurements of river channel velocity, width, depth, and discharge from across the continental United States. This is to our knowledge the largest collection of field measurements of river channel hydraulics and serves as the best dataset to calculate water surface TKE budgets. Second, we use the dataset from [@ulsethDistinctAirWater2019a] who measured and/or collected (from the literature) over 700 measurements of stream hydraulics and *k* and is, to our knowledge, the largest collection of field-measured river and stream *k*. This is also the dataset used to validate our *k* predictive models.

BIKER validation (section 4) was performed on 47 SWOT-simulated rivers. Because SWOT has yet to launch, it is standard practice to benchmark SWOT-related algorithms on "SWOT-like" data. These simulated rivers are simply river-reach-averaged hydraulic model outputs where the water surface heights and widths are labelled as RS observations and are used as the sole inputs to BIKER. Here, we use 47/51 rivers collected by @frassonExploringFactorsControlling2021 and @durandIntercomparisonRemoteSensing2016b. These are the two benchmarking studies that have explored RSQ algorithm performance for the SWOT mission. The approximate locations of these rivers are plotted in Figure 1, and please consult both of those papers for all of the hydraulic model specifications. Ultimately, the 47 rivers are spread across the United States, France, Italy, the United Kingdom, and Bangladesh. We omit three models from @durandIntercomparisonRemoteSensing2016b because they lack enough hydraulic information to calculate $U_\star$, which is necessary for algorithm validation (both Saint Lawrence River models and the Tanana River model).

To assess the influence of measurement error on BIKER's performance (section 4), we use the error model developed by @durandHowWillRadar2020a and implemented on 17/47 of the rivers by @frassonExploringFactorsControlling2021. Error in SWOT measurements will come from both the error tolerances intrinsic in the satellite data product as well as radar layover error. Layover error is the phenomenon when radar returns from different places arrive at the sensor at the same time, leading to taller landscape features appearing closer to the sensor than shorter landscape features that are the same horizontal distance from the sensor [@durandHowWillRadar2020a]. Width errors due to poor water classification are ignored as they were in @frassonExploringFactorsControlling2021.

For the $CO_2$ evasion and carbon efflux calculations (section 4), we use 26 bi-weekly dissolved $CO_2$ samples made by @beaulieuControlsGasTransfer2012a at one location in the Ohio River for one calender year from 2008-2009 (Figure S2). Note that this $CO_2$ data is for the Ohio River only but was applied to all 47 rivers (which includes multiple sections of the Ohio River). Because we are exclusively interested in the relative differences between $FCO_2$ estimates and not the raw fluxes themselves, any $CO_2$ data representative of SWOT-observable rivers was deemed acceptable.

##3 The TKE budget in large rivers and implications for scaling river gas exchange velocity
Turbulent flow is fundamental to fluvial geomorphology and open-channel flow and has been the subject of extensive research across scientific and engineering disciplines for over a century. Turbulent kinetic energy (TKE) is the energy generated and dissipated by the eddies created from the chaotic nature of turbulent flow. @nakagawaTurbulenceOpenChannel1993 outline the standard energy budget for TKE, at some depth *h*, for a 2-dimensional open-channel flow (equation 3). *G* is the TKE produced, $\epsilon$ is the TKE dissipated, $T_d$ is TKE transported from the riverbed via turbulent diffusion, and $V_d$ is TKE transported via viscous diffusion. See Appendix A for all variable descriptions and notation.
$$\mathbf{(3)} G+T_d+V_d=\epsilon$$

$\epsilon$ is frequently described as $\epsilon(h)=\frac{U_*^3}{h}$. However in rivers, which exhibit additional depth-scale shear due to downslope flow over 'macroroughness elements' (channel geometry, bars, meanders, etc.), energy dissipation occurs much more evenly over the flow depth than this parameterization allows for. It has been shown that, for rivers at least, a more appropriate dissipation model considers stream power (per unit weight water: $\epsilon_D = gSU$), which represents the energy dissipated by the total frictional resistance of the river reach's geometry [@moogStreamReaerationNonuniform1999]. This is often called 'energy dissipation via form-drag' in the literature, and is generally used to upscale *k* in rivers and streams [i.e. @raymondScalingGasTransfer2012a; @ulsethDistinctAirWater2019a]. All three of these studies have confirmed that $\epsilon_D$ is far greater than $\epsilon$ in rivers, therefore generating the majority of fluvial TKE dissipation.

We express *G* at the free surface in two different ways (equation 4) following @nakagawaTurbulenceOpenChannel1993 on the left and our derivation (detailed in Text S1) on the right. Looking at equation 4, it is readily apparent that $\frac{R_h}{H}$ will cancel out if they are equal. This occurs in most large and wide rivers and is a common assumption in hydraulic modeling. In this scenario, the following three statements must be true: 1) $G \approx gSU \approx \epsilon_D \approx \epsilon$, 2) $\epsilon(H) \neq \frac{U_*^3}{H}$, and 3) $T_d \approx 0$. If this "large river" condition is met, then additonal TKE transported from the riverbed is functionally zero and the TKE budget at the free surface is just stream power (per unit weight water). It also implies a local equilibrium bewteen TKE produced and TKE dissipated at the free surface. 
$$\mathbf{(4)} G=\frac{UU_*^2}{H}=gSU\frac{R_h}{H} $$

We test this hypothesis using the data from @brinkerhoffReconcilingAtaStationAtManyStations2019 and @ulsethDistinctAirWater2019a by plotting boxplots of $\frac{R_h}{H}$ ratios for over 531,000 different measurements of river hydraulics versus $\epsilon_D - G$. *G* is calculated as $\frac{UU_*^2}{H}$. Note that this term is equivalent to $T_d$ in equation 1 assuming that $V_d$ is neglibile at the free surface [@nakagawaTurbulenceOpenChannel1993]. In order to calculate when $R_h=H$, we round all ratios to 3 significant figures and set all ratios greater than or equal to 0.995 as $R_h=H$. If our hypothesis holds, then $\epsilon_D - G$ should be functionally zero only when $R_h=H$.

![Figure 2: Hydraulic radius/mean depth ratios (Rh/H) versus stream power per unit weight water - TKE production (eD-G, or equivalently, Td at the free surface) for our two datasets of field-measured river channel hydraulics. eD-G is functionally zero only when the Rh = H. This indicates that, when a river is large enough, the TKE energy budget is simply governed by stream power and TKE produced is equal to TKE dissipated. A local equilibrium of TKE is acheived, where energy generated by large-scale eddies is nearly equal to energy dissipated by the smallest scale eddies. In this figure, Rh/H bins contain between 92 and 138,307 measurements. ](cache/k600_theory/turbulence.jpg)

Figure 2, which plots this test, shows that $T_d$, or $\epsilon_D-G$, is generally three to four orders of magnitude less when $R_h=H$ than it is when $\frac{R_h}{H}$ is between 90-99% (median $T_d$ of  $10^{-7.29}$ versus $10^{3.84} \frac{J}{kg*s}$, respectively). When the $\frac{R_h}{H}$ ratio is very small (0-9%), this rises to a median value of $10^{-2.56} \frac{J}{kg*s}$. We deem the $\epsilon_D - G$ values when $R_h=H$ (which are generally $10^{-6}$ to $10^{-8} \frac{J}{kg*s}$) to be functionally zero. This indicates that, when a river is large enough, the TKE energy budget at the free surface is simply stream power per unit weight water and a local equilibrium of free-surface TKE is acheived, where energy produced via the largest scale (turbulent) eddies is equivalent to energy dissipated via the smallest scale (Kolmorogov) eddies. This idea of a local equilibrium is certainly not new, however this is to our knowledge the first confirmation that it happens naturally in most large rivers. From here on we refer to the $R_h=H$ river scenario as the 'large river' condition.

This finding as a significant impact on how we model *k* in these large rivers. Because the local equilibrium acheived during the 'large river condition' necessitates that $\epsilon$ is equivalent to *G* (Figure 2), we can presumably set $v_K$ equal to the largest ('integral-scale') eddies in the system, which are much more easily remotely-sensible than the Kolmogorov velocity scale. Otherwise, *k* should continue to scale only using the Kolmogorov-scale (which is similar across all turbulent flows by its definition).

To test this hypothesis, we fit the following three models to the @ulsethDistinctAirWater2019a dataset of *K* measurements all normalized to a Schmidt number of 600: $k_{600}=\beta U_*$, $k_{600}=\alpha \epsilon^{1/4}$, and $k_{600}=a (\epsilon_D)^b$ under two scenarios: $R_h=H$ and $R_h =/= H$. Note that because all data are normalized to a Schmidt number of 600, both the Schmidt numbers and viscosity terms are constant and are implicit in the statistical coefficient. The model structures (and the fit parameters) are detailed in Table 1. The third model tested (rows 5 and 6 of Table 1), was fit to see which slope coefficient arrives empirically by fitting the generalized scaling function to the data. Therfore, the $r2$ for these models is reflective of the log-transformed $k_{600}$ (unlike the other two models). This was the approach used by @ulsethDistinctAirWater2019a, who argue that bubble-mediated gas exchange yields steeper slopes (in high-energy streams) than 0.25.

![Figure 3: Six k600 predictive models tested on the dataset from Ulseth et al. (2019). a-c are fit to the 'large river condition' while d-f are not. Red coefficients of determination refer to log-tranformed quantities while blue coefficients of determination refer to the natural quantities.](cache/k600_theory/k600Plot.jpg)

The results of this test are presented as Figure 3 and Table 1, where our hypothesis is confirmed: both the $U_*$ and $\epsilon^{1/4}$ models explain ~70% of variance in *k* under the 'large river' condition while the $U_*$ model does not work when this condition is not met ($r^2$:`r round(models_ustar[2,]$r2,2)` versus $r^2$=`r round(models_ustar[1,]$r2,2)`). In fact, Figures 3a and 3b are nearly identical in their scaling dynamics, while Figures 3d and 3e are very different, lending further evidence that these two scaling models are synonmous only when the 'large river' condition is met. @wangPhysicallyBasedScaling2021 also found similar predictive performance for their $k\sim U_*$ model on simulated data for 35 rivers where they assumed $R_h \approx H$ (reported $r^2$:0.76). Further, the 1/4 slope parameter is statistically recovered in the 'large river' condition (row 3, Table 1: slope = `r round(models_ulseth[2,]$slope,2)`). We note that the large residuals in Figure 3c are a byproduct of the model existing in log-space and that panels 3b and 3c are functionally identical (see Table 1 for their nearly identical intercept and slope parameters). The top row of Figure 3 confirms that under the 'large river condition' *k* can be scaled using either integral or Kolmogorov turbulence scales and that they are synonymous. This opens the door for remotely sensing *k* using almost entirely remotely-sensed parameters (section 4).

*Table 1: Model performance and parameter values for the six models tested*

|**Figure 3 Panel**|**River Regime**|**Theoretical Model**|**Model Structure**|**r2**|**Intercept Parameter**|**Slope Parameter**|
|---------------------|------------------|-------------------|------|---------------|---------------|------------------|
|3a|$R_h=H$|$k \sim Sc^{-n}U_*$|$k_{600}=\beta U_*$|`r round(models_ustar[2,]$r2, 2)`|`r round(models_ustar[2,]$int, 2)`|`r round(models_ustar[2,]$slope, 2)`|
|3d|$R_h=/=H$|$k \sim Sc^{-n}U_*$|$k_{600}=\beta U_*$|`r round(models_ustar[1,]$r2, 2)`|`r round(models_ustar[1,]$int, 2)`|`r round(models_ustar[1,]$slope, 2)`|
|3b|$R_h=H$|$k\sim Sc^{-n}(\epsilon)^{1/4}$|$k_{600}=\alpha(\epsilon_D)^{1/4}$|`r round(models_eD4[2,]$r2, 2)`|`r round(models_eD4[2,]$int, 2)`|`r round(models_eD4[2,]$slope, 2)`|
|3e|$R_h=/=H$|$k\sim Sc^{-n}(v\epsilon)^{1/4}$|$k_{600}=\alpha(\epsilon_D)^{1/4}$|`r round(models_eD4[1,]$r2, 2)`|`r round(models_eD4[1,]$int, 2)`|`r round(models_eD4[1,]$slope, 2)`|
|3c|$R_h=H$|$k\sim Sc^{-n}(v\epsilon)^b$|$k_{600}=a(\epsilon_D)^{b}$|`r round(models_ulseth[2,]$r2, 2)`|`r round(models_ulseth[2,]$int, 2)`|`r round(models_ulseth[2,]$slope, 2)`|
|3f|$R_h=/=H$|$k\sim Sc^{-n}(v\epsilon)^b$|$k_{600}=a(\epsilon_D)^{b}$|`r round(models_ulseth[1,]$r2, 2)`|`r round(models_ulseth[1,]$int, 2)`|`r round(models_ulseth[1,]$slope, 2)`|

The 1/4 exponent is clearly not valid when $R_h=/=H$ (Figure 3b). However, the generalized form of this model explains 73% of the variation in log-transformed *k* and was deemed a strong scaling model for fluvial *k* when rivers do not meet the 'large river' condition. That model yields a slope of `r round(models_ulseth[1,]$slope,2)` which corroborates other models with slopes steeper than 1/4 in rivers [@wangPhysicallyBasedScaling2021; @ulsethDistinctAirWater2019a; @raymondScalingGasTransfer2012a]. Further theoretical modeling is needed to parse out why fluvial *k* in smaller, steeper systems generally scales at a steeper rate than in non-fluvial systems, and how this relates to the breakpointed model proposed by @ulsethDistinctAirWater2019a.

##4 Extending this finding to remotely sense riverine gas exchange velocity via the BIKER algorithm
We have shown that scaling *k* by either integral or Kolmogorov turbulence scales explains ~70% of the variation in $k_{600}$ in large rivers. The integral-scale model relies solely on the shear velocity $U_*$. Because of the structure of $U_*$ (Appendix A), we can easily reduce the 'large river' $k_{600}$ model (Figure 3a, row 1 Table 1) to an equation consisting solely of remotely-sensible river hydraulics and an inferred estimate of the channel area. By scaling *k* this way, we significantly reduce potential equifinality issues by having just two unknowns: $k_{600}$ and the median channel area $A_0$. Equifinality refers to an under-constrained mathematical system that has essentially infinite parameter combinations that can produce the same result: there are in essence more unknowns than equations [@garamboisInferenceEffectiveRiver2015]. This problem is experienced by both SWOT RSQ algorithms and existing in situ techniques that concurrently solve for *k* and stream metabolism [@applingOvercomingEquifinalityLeveraging2018; @graceFastProcessingDiel2015; @holtgrieveSimultaneousQuantificationAquatic2010]. In both of these domains, the other parameters are quite difficult to estimate (bed roughness and stream metabolism, respectively) while for BIKER, median channel area is relatively easy to estimate from the SWOT-observable river width [@brinkerhoffConstrainingRemoteRiver2020a].

That said, this all assumes that the 'large river' condition is met. Because SWOT will be limited to the widest rivers on Earth (minimum average width of 100m), the $R_h=H$ assumption is nearly always valid. First, we verified that this is true using both the @brinkerhoffReconcilingAtaStationAtManyStations2019 and @ulsethDistinctAirWater2019a datasets (Figure S3), finding that the vast majority of rivers with a width of at least 100m exhibit this $R_h=H$ behavior. Thus, our theoretical findings in sections 2 and 3 should hold in nearly all rivers that SWOT will observe and we should be able to remotely estimate $k_{600}$. BIKER is informed by the @hagemannBAMBayesianAMHGManning2017a algorithm for ungauged RSQ, further explored in more recent work by @brinkerhoffConstrainingRemoteRiver2020a. These papers conceptualize discharge as a Bayesian remote sensing problem, which we largely follow here to conceptualize $k_{600}$ as a Bayesian remote sensing problem that can be solved using SWOT data.

BIKER, and Bayesian inference in general, starts from Bayes rule (equation 5), where $\Theta$ is some set of non-remotely-sensible parameters we want to solve for (including $k_{600}$), *x* is the observed data, $f(x|\Theta)$ is the 'likelihood function' or sampling model conditional on the parameters, and $p(\Theta)$ is the joint prior distribution of the parameters. Therefore, we are interested in solving for $p(\Theta|x)$, or the 'posterior' distribution. For BIKER, *x* is the SWOT-observables: river width *W* and water surface elevation *H* (which is used to calculate the water surface slope $_h$). Note that $p(x)$ is usually computationally intractable to integrate, so Bayesian inference tools require only the proportionality to be specified: $p(\Theta|x) \propto f(x|\Theta)p(x)$. Sampling algorithms are then used to approximate the posterior distribution, as is done in BIKER.

$$\mathbf{(5)} p(\Theta|x) = \frac{f(x|\Theta)p(\Theta)}{p(x)}$$

The heart of BIKER is its reformulation of the $k_{600}$ model as a Bayesian sampling model that is conditional only on the data that SWOT will provide. To do this, $k_{600}$ first needs to be written as a function of SWOT-observables *W* and $S_h$. This algebra is carried out using the model parameter from row 1 of Table 1 and yields equation 6, where we follow our earlier assumption that $R_h = H =\frac{A}{W}$ where *A* is wetted channel area. *A* is further split into the SWOT-observable portion *dA* and the unobservable portion $A_0$ following @durandEstimatingReachaveragedDischarge2014a. *dA* is estimated assuming a rectangular river channel (usually valid when $R_h=H$) so that $dA \approx \delta H*W$. As mentioned earlier, equifinality issues are limited in this model as there is only one unknown besides $k_{600}$, and $A_0$ is reasonably easy to approximate from river width [@brinkerhoffConstrainingRemoteRiver2020a]. There are no reliances on bed roughness, flow velocity, or any other terms that are difficult to infer from river width.

$$\mathbf{(6)} \begin{eqnarray} k_{600} = \bigg(56.0294 S_h(\frac{A_0+dA}{W})\bigg)^{1/2} \end{eqnarray}$$

Next, equation 6 is written as a Bayesian sampling model, in which all of the SWOT observations are sampled from the unknown model parameters ($A_0$ and $k_{600}$). This is equation 7 after log-transforming all of the variables. The $\sigma_{k_{600}}^2$ parameter refers to the uncertainty inherent in equation 6's estimates. Equation 7 also necessitate that we specify prior distributions for the parameters $A_0$ and $k_{600}$. Prior distributions, defined by their hyperparameters, formalize the a priori estimates (and uncertainties) for the non-remotely-sensed terms. More intuitively, BIKER priors represent our 'prior river knowledge' of what $A_0$ and $k_{600}$ probably are for some river since they cannot be remotely sensed. Hyperparameter specifications are detailed in Text S2, however the goal was to rely on absolutely no in situ information such that we could run this method on any river on Earth that is SWOT-observable.

$$\mathbf{(7)} \left[log(56.0294) + \frac{1}{2}log(g) + \frac{1}{2}log(S_h) - \frac{1}{2}log(W)\right] \sim normal\left(\left[log(k_{600})-\frac{1}{2}log(A_0+dA)\right], \sigma^2_{k_{600}}\right)$$

With the sampling model (equation 7) and hyperparameters described (Text S2), a joint posterior distribution conditional on the SWOT observations is specified. To approximate this distribution, we use a Markov Chain Monte Carlo (MCMC) algorithm implemented using the Stan probabilistic programming language. Specifically, Stan uses a Hamiltonian Monte Carlo sampler which reduces computation time relative to other sampling algorithms [@hagemannBAMBayesianAMHGManning2017a].

We validate BIKER on 47 SWOT-simulated rivers using daily observed $R_h$ and observed $S_h$. We also re-validate BIKER on the 17 rivers with the SWOT error model corrupting the SWOT-observables *W* and *H* (this workflow and the metrics used are detailed in Text S4). Regardless of the validation setup, we do not have observed $k_{600}$ data for these rivers, and to our knowledge no field dataset of $k_{600}$ exists in the type of temporal and spatial frequency that SWOT (and therefore the BIKER algorithm) provides. Therefore, we take the model outlined in row 1 of Table 1 and use that to calculate the observed $k_{600}$ that BIKER is validated against. With this setup, we are directly exploring BIKER's ability to infer observed $A_0$ and $k_{600}$ from *W* and *H* alone, as the model has already been sucessfully validated (Table 1, Figure 3).

Figure 4a plots the validation results for $k_{600}$ (with no SWOT measurement error) across all  rivers and all timesteps. The points are the posterior means while the black lines are the 95% confidence intervals (CIs) for the predictions. $k_{600}$ is strongly correlated with the BIKER-predicted $k_{600}$ ($r^2$ of 0.76). Using absolutely no in situ information, BIKER captures the magnitude of the predictions and most points fall on or near the 1:1 line. The regression of the estimates (solid grey line) nearly perfectly recovers the 1:1 line (dashed black), but there is an overestimation bias in the largest $k_{600}$ values. The RMSE for the BIKER predictions is `r round(results_3_2_all$rmse,2)` m/day) across all observations.

![Figure 4. a: Validation of BIKER for 47 SWOT rivers. Black bars are 95% CIs for the modeled values. Grey line is linear regression (and 95% prediction intervals are dashed) and dashed black line is 1:1 line. b: Cumulative density functions (CDFs) of the same results: dashed black line are observed values, green line are BIKER posterior means, and blue lines are BIKER 95% CIs.](cache/validation/validation.jpg)

Figure 4a highlights a overestimation bias by BIKER when $k_{600}$ is larger, which is confirmed in Figure 4b. Figure 4b plots the cumulative density functions (CDFs) of observed and predicted $k_{600}$ where the green line is the CDF of the BIKER posterior means, the blue lines are the CDFs of the BIKER posterior 95% CIs, and the dashed black line is the CDF of observed $k_{600}$. We see, more clearly than in Figure 4a, that BIKER captures $k_{600}$ very well: most of the  observed CDF falls nearly exactly on or adjacent to the posterior means (Figure 4b). However, the uppermost quartile is systematically overestimated by BIKER (also visible in Figure 4a). We propose that this is due to the model used to assign the $k_{600}$ prior hyperparameters being less representative for these rivers. In line with this, the rivers with the poorest BIKER performance uniformally are the narrowest and steepest in the set (Figure S4). In summary, Figure 3 confirms that we improve upon our baseline understanding of $k_{600}$ in these rivers: we accurately capture $k_{600}$ with no in situ information about the river while simultaneously and explicitly accounting for the uncertainties inherent in our $k_{600}$ estimates.

Figure 5a plots validation metrics calculated for each river with and without SWOT measurement error (green and purple, respectively). The boxplots are composed of validation scores for either the 47 or 17 rivers- see Table S1 for metric definitions. SWOT measurement errors neglibly influence BIKER's performance across all four error metrics (Figure 5a), though caution should be used in over-interepting boxplots with a sample size of only 17. Given Figure 5a, we deem that SWOT measurement error does not exert a significant influence on BIKER and so the results presented for the rest of the manuscript assume no measurement error in order to use all 47 rivers.

![Figure 5. a: Performance metrics by river. See Table 1 for metric definitions. Dashed lines denote scores of 1, 0, and -0.41 for r2 (section 3.2). b-d: validation timeseries for three rivers representative of good, reasonable, and poor BIKER performance. b) was randomly selected from the upper tertile of NRMSE scores, c) was randomly selected from the middle tertile, and d) from the worst tertile. Model results include the posterior means and 95% CIs.](cache/validation/validation_by_river.jpg)

Median r2 is `r round(median(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$r2, na.rm=T),2)`, which is excellent given that absolutely no in situ information is being used to predict $k_{600}$. NRMSE and RRMSE have median scores of `r round(median(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$nrmse, na.rm=T),2)` and `r round(median(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$rrmse, na.rm=T),2)`, respectively, which are again good for a completely ungauged method. Median rBIAS is `r round(median(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$rBIAS, na.rm=T),2)`, highlighting a  positive bias in most rivers' predictions. This further supports the visual evidence in Figure 4 that sometimes BIKER is overestimating the magnitude of $k_{600}$ and that this might be river-specific. While median r2 and rBIAS scores were strong, the ranges of these scores were somewhat large (standard deviation for r2 of `r round(sd(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$r2, na.rm=T),2)` and for rBIAS of `r round(sd(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$rBIAS, na.rm=T),2)`).

Figure 5b-d are representative timeseries plots of predicted and observed $k_{600}$ for three rivers chosen randomly from those with 'good' NRMSE scores (b), 'okay' NRMSE scores (c), and 'bad' NRMSE scores (d). See the Figure 5 caption for how this was determined. For the Ohio Section 2 River, the entire timeseries of $k_{600}$ is nearly perfectly predicted, while in the Seine Upstream River the dynamics (peaks and valleys) are reasonably captured but they are magnified to be far larger than the observed dynamics. Generally, though, mean $k_{600}$ is actually reasonably recovered for the Seine Upstream, as confirmed visually. In the Connecticut River, there is significant positive bias in the estimates but also massive uncertainty (per the 95% CIs) in those estimates, indicating that BIKER is highly uncertain about its output (and rightfully so). Correct temporal dynamics are also largely missing from BIKER's Connecticut River predicitions.

##5 Remotely sensing carbon emissions from rivers
It is one thing to accurately predict *k*, but researchers are often most interested in the actual gas fluxes from rivers and ultimately the carbon emitted from river to atmosphere. Therefore, we also explore 1) BIKER's ability to reproduce $CO_2$ evasion fluxes ($FCO_2$ $[M/L^2T]$) from these 47 rivers, and 2) a comparison of the representative carbon efflux $[M/T]$ from BIKER with eastbalished in situ methods. The details of this workflow are in Text S4, but broadly we pair the biweekly 26 $CO_2$ samples (section 2, figure S2) from @beaulieuControlsGasTransfer2012a with a subset of SWOT observations (as the data are not daily). We then calculate $FCO_2$ using the equation 7 and assuming atmospheric $CO_{2_{air}}$ is 390 uatm. Finally, we estimate a median bulk carbon efflux using BIKER's posterior means and three other in situ models for average channel depth (used to calculate $U_\star$) used for upscaling: one trained on the @brinkerhoffReconcilingAtaStationAtManyStations2019 dataset, and two previously published models [@raymondGlobalCarbonDioxide2013a; @raymondScalingGasTransfer2012a]. See Table S2 for their definitions. This allows us to assess whether BIKER's estimates (wholly ungauged) are comparable to gauged methods (all four HG models). Finally, we express the bulk carbon efflux as the median mass flow rate of carbon (via $CO_2$ evasion) per year from the 47 rivers after accounting for total river surface area.

$$\mathbf{(7)} FCO_2 = k_{CO_2}([CO_{2_{water}} - [CO_{2_{air}}])$$

In Figure 6a, there is a very strong fit to the observed $FCO_2$ data, with an RMSE of `r round(results_3_3_all$rmse,2)` $g/m^2dy$. The performance is notably better than for $k_{600}$ alone (Figure 4a) and there is no systematic bias in the predicitons across all 47 rivers. This is presumably due to the structure of the $FCO_2$ equation, which reduces the relative importance of errors in *k* given that the $CO_2$ data is measured in situ. $FCO_2$ prediction intervals are slightly narrower than those presented in Figure 4a.

![Figure 6: a: FCO2 via BIKER versus via equations 2 and 3 for every 11th timestep for the 49 rivers (grey lines are linear regression and 95% prediction intervals, while black dashed line is the 1:1 line). b-d: timeseries plots for the three example rivers from Figure 3b, 3c, and 3d.](cache/FCO2/FCO2_models.jpg)

Figure 6b compares the bulk carbon efflux (via $CO_2$ evasion) from the 47 rivers using BIKER posterior means and four streamgauge-based HG models. Both the BIKER bulk carbon efflux (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_BIKER',]$sumFCO2, 0)` gG-C/yr) and the 'Brinkerhoff 2019' estimate (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_brinkerhoff2019',]$sumFCO2, 0)`) are quite close to the observed flux (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_obs',]$sumFCO2, 0)`), with the latter nearly perfectly recovering the observed value. BIKER overestimates the efflux somewhat. The 'raymond 2013' model is even more overestimated (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_raymond2013',]$sumFCO2, 0)`), while the 'Raymond 2012' model appreciably underestimates this bulk efflux (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_raymond2012',]$sumFCO2, 0)`). Thus, despite BIKER using absolutely no in situ data like the other 4 methods do, it provides a reasonable estimate of the carbon efflux and better a estimate than 2 of the approaches that rely on a streamgauge (Figure 6b).

##6 Discussion
###6.1 Towards a mechanistic understanding of gas exchange in large rivers
Field studies of gas exchange in wide rivers have suggested that their *k* properties behave differently than in steeper and smaller rivers [@ulsethDistinctAirWater2019a; @beaulieuControlsGasTransfer2012a; @alinPhysicalControlsCarbon2011; @wangPhysicallyBasedScaling2021]. In sections 2 and 3 we have shown that this must be the case given the algebraic structure of the TKE budget in open-channel flows [@nakagawaTurbulenceOpenChannel1993] and the assumptions underlying turbulent scaling with interfacial fluxes [@katulStructureFunctionModel2018; @lorkeUnifiedScalingRelation2006]. This result profoundly amends how we scale *k* in large rivers and directly addresses a long-standing problem of upscaling fluvial *k*: that our current methods perform poorer in larger rivers than small ones.

We also statistically recovered the 1/4 scaling exponent in the classic $k \sim \epsilon^{1/4}$ model (Figure 3c) for gas exchange that has been repeadely validated in non-fluvial aquatic systems. This again suggests that *k* in large rivers behaves more similarly to lakes, esturaries, and the ocean than to steeper rivers where bubble-mediated gas exchange might occur [@ulsethDistinctAirWater2019a]. We also suggest that the TKE budget in smaller rivers and streams is additionally complicated by diffusive transport of TKE from the bed to the free surface, but this TKE production mechanism is functionally zero if the 'large river' condition is met (Figure 2). Channel bed roughness effects riverine gas exchange mostly in high energy streams where slope is sufficiently steep for bubble-induced gas exchange to occur [@hallGasExchangeStreams2020]. @ulsethDistinctAirWater2019a showed bed roughness loosely correlates with $k_{600}$ in steep Alpine streams. However, they coarsely estimated bed roughness from arial imagery, only studied extremely small mountain streams, and to date most similar work has focused on labratory exercises [e.g. @chansonStreamReaerationNonuniform; @moogStreamReaerationNonuniform1999]. In this study, we confirm that bed roughness is exerting functionally no influence on free-surface TKE in large rivers (Section 2 and figure 2) and therefore *k*. We show that free-surface TKE must be entirely produced and dissipated by form-drag and not bottom friction or roughness. However, this is only true in the largest of rivers. It is likely that the steeper slope parameter in Figure 3d and row six, Table 1 is due to the influence of bottom roughness on free-surface TKE in smaller streams.

###6.2 Towards remote sensing of global spatiotemporal dynamics of *k* in large rivers
To date, most field-scale studies of riverine gas exchange have focused on 1) its relationship with wind speed [e.g @beaulieuControlsGasTransfer2012a; @borgesVariabilityGasTransfer2004; @zappaEnvironmentalTurbulentMixing2007], 2) average flow velocity [e.g. @alinPhysicalControlsCarbon2011; @beaulieuControlsGasTransfer2012a; @schelkerCO2EvasionSteep2016a], or 3) discharge [e.g. @wangPhysicallyBasedScaling2021; @robertsMultipleScalesTemporal2007; @uehlingerEcosystemMetabolismDisturbance1998]. However, the spatiotemporal dynamics of riverine gas exchange are still weakly constrained. A few studies have investigated these dynamics, but they have been limited to individual rivers and/or limited field seasons [@hallAirWaterOxygen2012; @sand-jensenCO2DynamicsDanish2012]. @wallinSpatiotemporalVariabilityGas2011 performed a preliminary analysis in northern Sweden relating cross-section specific temporal variability in gas exchange with channel slope, but they were limited to an average of only 8 measurements per river in a single watershed. This limited knowledge of large-scale *k* spatiotemporal dynamics is due both to a lack of process-level understanding [@hallGasExchangeStreams2020] but also a lack of measurements. Authors have argued that the key to explaining the large residual variation in upscaling models is to explore at-a-station temporal variability in *k* [@hallGasExchangeStreams2020].

Therefore, estimating *k* from SWOT data is an attractive option for exploring its spatiotemporal dynamics at fine temporal resolution and at the global-scale. SWOT will provide daily hydraulic measurements for a 3 month fast sampling period for calibration and validation and sampling thereafter between 1 and 7 days per 21 day repeat cycle [@biancamariaSWOTMissionIts2016]. BIKER's success in 1) infering $k_{600}$ using simulated SWOT data over a wide range of rivers (Figures 4 and 5) and 2) being robust to measurement errors internal to the SWOT data (Figure 3) bode well for BIKER's eventual implementation on real SWOT observations.  The results presented here suggest that daily estimation of riverine gas exchange globally could be possible once SWOT launches.

###6.3 Towards remotely sensing riverine carbon emissions using SWOT
Figure 6 confirms that BIKER is quite successful, without any in situ information aside from a $CO_2$ logger, at predicting 1) $FCO_2$ (Figure 6a) and 2) the bulk carbon efflux (Figure 6b). This encouraging result has two main implications for future work coupling remote sensing via SWOT with in situ data. First, it confirms that we can couple BIKER and SWOT with in situ gas concentration loggers to produce $FCO_2$ estimates at novel temporal resolution in SWOT-observable rivers. This is particualry useful given recent advances in high temporal resolution in situ gas concentration measurements [@ahoHighestRatesGross2021] but no such similar advances in modeling *k* at high temporal resolutions. BIKER can likely also be ran at the field scale using arrays of pressure transducers to estimate water surface slope (rather than using satellite-based altimeters like SWOT) following recent work doing the same using the @hagemannBAMBayesianAMHGManning2017a RSQ algorithm [@harlanDischargeEstimationDense].

Secondly, it is important to stress that unlike BIKER, the HG models in Figure 6b rely on an in situ streamgauge. This means that Figure 6b represents the best performance that those models could ever have; if ran using modeled discharge their accuracy would necessarily decrease. Thus, these results suggest that BIKER will be useful in two settings: 1) upscaling in ungauged rivers as hypothesized, but also in 2) potentially improving our carbon efflux understandings at gauged sites. Future work should systematically quantify prediction error from coupling global-scope HG models with modeled discharge, as is the default workflow used in fluvial $FCO_2$ upscaling studies [e.g. @lauerwaldSpatialPatternsCO22015; @horgbyUnexpectedLargeEvasion2019; @raymondGlobalCarbonDioxide2013a]. **I assume Shaoda's paper will still be in review and can't cite here but this is exactly what they do...**

Figure 6b also confirms that the training data used for HG models exerts a significant influence on upscaled carbon emissions from rivers. In Figure 6b there is a nearly 0.5 Tg-C range between estimates, which is significant and nearly entirely a function of the data used to fit these depth HG models. In this context, the 'Brinkerhoff 2019' model likely outperforms both 'Raymond models' because the training data is orders of magnitude larger and more geomorphically varied than those used in the 'Raymond' models (530,945 measurements versus 1,026 and 10837 measurements). Meanwhile, BIKER has no similar reliance on hydraulic parameters trained on different datasets and only assumes that *dA* can be calculated by assuming a rectangular river channel. Upon SWOT's launch, the BIKER approach to estimating *k* could be coupled with ethier existing upscaling workflows or even explicit $CO_2$ transport models [@brinkerhoffLakeMorphometryRiver2021; **Saccardi & Winnick in review**] to improve riverine gas flux predictions where gauges are unavailable but SWOT measurements are. This coupling could potentially be done using data assimilation techniques, which have proven very useful for similar objectives in recent RSQ work [@ishitsukaCombiningOpticalRemote2020].

##7 Conclusions

##8 Acknowledgements
C.B. Brinkerhoff was funded on **FINESST babyyyy**. BIKER is available at https://github.com/craigbrinkerhoff/BIKER. All code to build and generate results, figures, and the manuscript is available at https://github.com/craigbrinkerhoff/RSK600. Data used in this study was generously made available by the authors of @frassonExploringFactorsControlling2021, @durandIntercomparisonRemoteSensing2016b, @beaulieuControlsGasTransfer2012a, @ulsethDistinctAirWater2019a, and @raymondScalingGasTransfer2012a.

##9 Apendix A
*Table A1: Variable description and notation for this study. TKE refers to 'turbulent kinetic energy'*

|**Notation**|**Description**|**Calculation (if applicable)**|**Units**|
|------------|---------------|-------------------------------|---------|
|*Bulk carbon efflux*|carbon mass transport rate from river to atmosphere|NA|$[\frac{M}{T}]$|
|$C_{water}$|Water-side gas concentration|NA|$[L]$|
|$C_{air}$|Air-side gas concentration|NA|$[L]$|
|$\epsilon$|TKE dissipation rate|$\frac{U_*^3}{H}$|$[\frac{E}{M*T}]$|
|$\epsilon_D$|TKE dissipation rate via form-drag|$gSU$|$[\frac{E}{M*T}]$|
|$FCO_2$|$CO_2$ evasion flux from river to atmosphere|NA|$[\frac{M}{L^2T}]$|
|$g$|gravitational acceleration|9.8|$[\frac{L}{T^2}]$|
|$G$|TKE production rate|$\frac{UU_*}{H}$|$[\frac{E}{M*T}]$|
|$h$|Elevation above the bed|NA|$[L$|
|$H$|Mean flow depth|NA|$[L]$|
|$k$|gas exchange velocity|NA|$[\frac{L}{T}]$|
|$k_{600}$|gas exchange velocity normalized to Sc=600|NA|$[\frac{L}{T}]$|
|$R_h$|Hydraulic radius|$\frac{AW}{2H+W}$|$[L]$|
|$S$|River slope|NA|$[\frac{L}{L}]$|
|$Sc$|Schmidt number|NA|$unitless$|
|$T_d$|TKE turbulent diffusion rate|NA|$[\frac{E}{M*T}]$|
|$U$|Reach-averaged flow velocity|NA|$[\frac{L}{T}]$|
|$U_*$|Shear velocity|$\sqrt{gSR_h}$|$[\frac{L}{T}]$|
|$v$|kinematic viscosity|NA|$[\frac{L^2}{T}]$|
|$V_d$|TKE viscous diffusion rate|NA|$[\frac{E}{M*T}]$|
|$v_K$|turbulence characteristic velocity scale|NA|$[\frac{L}{T}]$|
|$W$|Flow width|NA|$[L]$|

## References

### 2021-07-19 supp_v4.Rmd
ms draft 4 updates

---
title: "Gas exchange in big rivers controlled by stream power and turbulence scales: implications for remotely sensing gas exchange"
author: "CB Brinkerhoff, CJ Gleason, PA Raymond, MH Harlan"
date: "`r Sys.Date()`"
csl: agu_citation_style.csl
output:
  word_document: 
      reference_docx: word-styles-reference.docx
bibliography: References.bib

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Contents
This supplementary information contains 7 figures and 1 table. Please consult https://github.com/craigbrinkerhoff/RSK600 for all code to build and generate results, figures, and the manuscript.

## Text S1: Deriving two expressions for turbulent kinetic energy production
Below we detail how we arrived at our two expressions for a cross-section-averaged turbulent kinetic energy production (*G*), or equation 2 in the main text.

@nakagawaTurbulenceOpenChannel1993 outline the standard energy budget for turbulent kinetic energy (TKE), at some distance from the bed *h*, for a 2-dimensional open-channel flow (equation S1). *G* is the TKE produced, $\epsilon$ is the TKE dissipated, $T_d$ is TKE transported from the riverbed via turbulent diffusion, and $V_d$ is TKE transported via viscous diffusion. $T_d$ should always be positive at the free surface, but this is increasingly less true as one moves closer to the riverbed. Since we are interested in the TKE budget at the river surface, $V_d$ can be assumed neglible [@nakagawaTurbulenceOpenChannel1993]. 
$$\mathbf{(S1)} G+T_d+V_d=\epsilon$$

@nakagawaTurbulenceOpenChannel1993 further define *G* and the direct energy dissipation *E*, as averaged over the entire river cross-section and integrated over the entire water column, as equation S2 where *H* is the full water-column depth, *U* is the cross-sectional average flow velocity, and $U_*$ is the friction velocity. $U_*$ is defined by the water surface slope *S*, gravitational acceleration *g* and the channel hydraulic radius $R_h$ as $U_*=\sqrt{gSR_h}$.
$$\mathbf{(S2)} \int_{0}^{H}{Gdh} + \int_{0}^{H}{Edh}=UU_*^2$$

$\int_{0}^{H}{Edh}$ can also be assumed neglible at the free surface [@nakagawaTurbulenceOpenChannel1993]. We then integrate over the full water-column depth *H* to arrive at an expression for *G* at the free surface that reduces to equation S5. Equations S3 to S4 show the algebra.
$$\mathbf{(S3)} \int_{0}^{H}{Gdh} = U\sqrt{gSR_h}^2$$
$$\mathbf{(S4)} GH = UgSR_h$$
$$\mathbf{(S5)} G=gSU\frac{R_h}{H}$$
Alternatively, we can derive another expression for *G* from equation S2. This is provided as equation S8 with the necessary algebra in equations S6-S7.
$$\mathbf{(S6)} \int_{0}^{H}{Gdh}=UU_*^2$$
$$\mathbf{(S7)} GH=UU_*^2$$
$$\mathbf{(S8)} G=\frac{UU_*^2}{H}$$

## Text S2: BIKER hyperparameterization
We assign prior hyperparameters using SWOT data only. All priors are formalized within the model as truncated normal distributions of the log-transformed terms such that $log(X) \sim  \mathcal{N}(\mu, \sigma^2)$ for $\lambda < log(X) < \gamma$, using prior hyperparameters mean ($\mu$), standard deviation ($\sigma$), and upper ($\gamma$) and lower bounds ($\lambda$) for any parameter *X*.

$A_0$ prior hyperparameters were assigned following an updated version of the method developed by @brinkerhoffConstrainingRemoteRiver2020a. They developed a set of river channel prior hyperparameters for McFLI algorithms that are entirely RS-able and reflect differential channel hydraulics as a function of river geomorphology. They used an extensive database of field measurements and machine learning to identify patterns that associate river width with the hydraulic priors needed to run McFLIs so that prior hyperparameters may be assigned to rivers using only the existing remotely sensed data. For this study, we extracted $\lambda$ and $\gamma$ as the Xth percentile values rather than the absoulte maximum and minimum values to avoid physically impossible bounds on $A_0$. This leaves the $k_{600}$ hyperparameters to be defined. We assigned those by using a simple slope regression model to predict $U_*$, trained on the @ulsethDistinctAirWater2019a: equation S9 assigns $\mu$ while $\sigma^2$ was set to equation S9's log-transformed standard error after propogation of errors for both equation S9 and equation 4 in the main text: 1.12. $\lambda$ and $\gamma$ were set to log(0.001) m/day and log(500) m/day, respectively. Finally, we estimate $\sigma_{k_{600}}$ using the standard model error from equation 4 in the main text and as fit on the data shown in Figure 3a.

$$\mathbf{(S9)} \mu_{k_{600}}=56.0294 * a(ln(S_h))^b$$

## Text S3: BIKER validation setup
Regardless of the validation setup or SWOT error budget used, we do not have observed $k_{600}$ data for these rivers, and to our knowledge no field dataset of $k_{600}$ exists in the type of temporal and spatial frequency that SWOT (and therefore the BIKER algorithm) provides. Therefore, we take the model outlined in row 1 of Table 1 and use that to calculate the observed $k_{600}$, that BIKER is validated against (equation S10).

$$\mathbf{(S10)} k_{600, obs}=56.0294\sqrt{gS_hR_{h, obs}}$$

It also means that, for a fair validation scheme, $\sigma_{k_{600}}^2$ in equation 5 in the main text must be set to reflect only error from our assumptions about calculating *dA* and not the parameter uncertainity inherent in the model coefficient (i.e. the 56.0294 in equation 4 in the main text). Because this model is only theoretically valid when $R_h=H$, we convientantly do not need to account for uncertanties associated with assuming that $R_h = H$. So for this validation, $\sigma_{k_{600}}^2$ is set to 0.20 (natural-log space) to reflect a small degree of uncertainity that could arise from assuming a rectangular river channel. However once SWOT launches, it should reflect the total uncertainty described in Text S3 (1.12).

Validation is performed using the BIKER posterior means. Validation metrics take two forms (and are detailed in Table S1). To validate across all rivers and timesteps, we used the coefficient of determination $r^2$ and the root mean square error RMSE. Three normalized metrics were used for by-river validation: RRMSE and NRMSE are normalized root mean square errors that have been normalized by the observed value and the mean observed value (respectively). rBIAS is a measure of prediction bias that is normalized by the mean observed value. r2 is again used on the by-river case.

## Text S4: BIKER FCO2 calculations
To calculate $CO_2$ fluxes and carbon efflux, we pair @beaulieuControlsGasTransfer2012a's biweekly 26 $CO_2$ and water temperature samples with every 11th SWOT observation by date, ignoring the timesteps beyond 26. This amounts to only ~15% of the SWOT observations and we deem this acceptable. We sample every 11 days as this is the average sampling resolution for SWOT and the $CO_2$ and water temperature data are approximately the same at a 14 day resolution. Not all of the SWOT rivers have observations for a full year, and when simulation dates are not available they are assumed to start on January 1st. We then pair the modeled $k_{CO_2}$ values (obtained from $k_{600}$ and equation S11) with these water-side $CO_2$ concentrations and water temperatures. In equation S11, *Sc* is the Schmidt number at the observed water temperature. This was calculated following @raymondScalingGasTransfer2012a and @wanninkhofRelationshipWindSpeed1992a. Atmospheric $CO_2$ was assumed 390 uatm. $FCO_2$ validation was performed using the same metrics as $k_{600}$ validation (Table 1).

$$\mathbf{(S11)} k_{CO2, temp}=k_{600}*(Sc/600)^{-1/2}$$
To calculate the bulk carbon efflux, we multiply the median $FCO_2$ value across all rivers and timesteps (in Tg-C/yr) by the total average surface area of all 47 rivers.

##Figures
![Figure S1: Flowchart of the entire study's workflow](cache/flowchart.jpg)

![Figure S2 Timeseries of the biweekly CO2 data from @beaulieuControlsGasTransfer2012a. Sampling took place 2008-2009 in the Ohio River (upstream of Cincinnati, Ohio, United States). Each point here was joined to the 11-day SWOT observations used in this study (section 2.4).](cache/FCO2/Beaulieu_timeseries.jpg)

![Figure S3 Stuff](cache/k600_theory/swotPlot.jpg)

![Figure S4 Mean river properties (observed k600, width, and slope) for the 47 rivers versus their BIKER performance as defined by NRMSE (see Table S1 for this metric's definition). The dashed black line is the median NRMSE score. The poorly performing rivers are universally the narrowest and steepest of the set.](cache/validation/riverProperties_nrmse.jpg)

##Tables
*Table S1: Validation metrics used in this study, where r is the correlation coefficient, Nt is number of observations and i is the specific observation. σ refers to the variance of the sample and μ refers to the mean of the sample. As is standard, a carrot accent indicates the predicted value.*

|**Description**|**Acronym**|**Definition**|**Ideal Score**|**Possible Range**|**Validation Scheme**|
|-----------|-------|----------|-----------|--------------|-----------------|
|Coefficient of determination|$r^2$| $1-(\frac{\sum\limits_{i=1}^{N_t}{(k_{600, i}-\hat{k_{600}})^2}}{\sum{(k_{600, i}-\bar{k_{600}})^2}})$ |1|0 to 1|	All rivers and all timesteps|
|Root-mean-square-error|RMSE| $\sqrt{\frac{1}{N_t}\sum\limits_{i=1}^{N_t}(\hat{k_{600, i}}-k_{600, i})^2}$ |0|0 to ∞|All rivers and all timesteps|
|Relative root-mean-square error|RRMSE| $\sqrt{\frac{1}{N_t}\sum\limits_{i=1}^{N_t}(\frac{\hat{k_{600, i}}-k_{600, i}}{k_{600, i}})^2}$ |0|0 to ∞|By river|
|Normalized root-mean-square error|NRMSE| $\sqrt{\frac{1}{N_t}\sum\limits_{i=1}^{N_t}(\frac{\hat{k_{600, i}}-k_{600, i}}{\bar{k_{600, i}}})^2}$ |0|0 to ∞|By river|
|Relative bias|rBIAS| $\frac{1}{N_t}\sum\limits_{i=1}^{N_t}(\frac{\hat{k_{600, i}}-k_{600, i}}{\bar{k_{600, i}}})$ |0|-∞ to ∞|By river|

*Table S2: Details on the 3 depth hydraulic geometry models used to estimate FCO2 from the SWOT rivers (section 2.4).*

|**Name**|**Equation**|**Description**|**Reference**|
|-------------|-------------|-------------|-------------|-------------|
|Brinkerhoff 2019|$D=0.258Q^{0.395}$|530,945 measurements made across the United States at streamgauges|this study; @brinkerhoffReconcilingAtaStationAtManyStations2019|
|Raymond 2012|$D=0.409Q^{0.294}$|1,026 measurements across the United States|@raymondScalingGasTransfer2012a|
|Raymond 2013|$D=0.409Q^{0.294},  D=0.449Q^{0.37}$ |Average of the Raymond 2012 equation and one using 9,811 measurements at US streamgauges|@raymondGlobalCarbonDioxide2013a|

##References

### 2021-07-19 generateMS.R

@@ -5,8 +5,9 @@
 setwd('C:\\Users\\craig\\Documents\\GitHub\\RSK600')
 
 #Move all results to global environment
-#results_3_1 <- read.csv('cache//k600//results.csv')
-#results_3_1_MC <- read.csv('cache//MonteCarlo//results.csv')
+models_ulseth <- read.csv('cache\\k600_theory\\ulseth_models.csv')
+models_ustar <- read.csv('cache\\k600_theory\\ustar_models.csv')
+models_eD4 <- read.csv('cache\\k600_theory\\eD_models.csv')
 results_3_2_rivs <- read.csv('cache//validation//results_by_riv.csv')
 results_3_2_all <- read.csv('cache//validation//results_all_riv.csv')
 results_3_3_all <- read.csv('cache//FCO2//fco2_stats_all.csv')
@@ -15,8 +16,8 @@ results_3_3_bulk <- read.csv('cache//FCO2//bulkFluxes.csv')
 results_3_3_bulk$key <- as.character(results_3_3_bulk$key)
 
 #knit manuscript
-rmarkdown::render("manuscript_v3.Rmd")
+rmarkdown::render("manuscript_v4.Rmd")
 print('Manuscript generated.')
 
-rmarkdown::render("supp_v3.Rmd")
+rmarkdown::render("supp_v4.Rmd")
 print('Supplemental Information generated.')

### 2021-07-19 src/buildMap.R

@@ -39,4 +39,4 @@ rivers <- tm_shape(data) +
   tm_symbols(col =  "Data", size=0.5, shape=21, palette = c('#e31a1c', '#33a02c', '#1f78b4'), border.col = 'black', border.lwd = 1)
 map <- base + map + rivers
 
-tmap_save(map, 'cache/map.jpg')
+tmap_save(map, 'cache/maps/map.jpg')

### 2021-07-19 src/swot_k_model.R

@@ -14,12 +14,10 @@ setwd('C:\\Users\\craig\\Documents\\GitHub\\RSK600\\')
 
 #some constants
 g <- 9.8
-cf <- 0.100 #Henderson found empirically that Cf=0.113, if we use the empirical Strickler relation for n (1923- different rivers), we get Cf=0.09. So, we used Cf=0.100. See the notes!
-Sg <- 2.65 #specific gravity
 
 #########READ IN HYDRAULICS DATA FROM BRINKERHOFF ETAL 2019------------------------------
-data <- read.csv('C:\\Users\\craig\\Documents\\OneDrive - University of Massachusetts\\Published\\2019_AMHG_GRL\\working\\final\\field_measurements.csv')
-nhd_med <- read.csv('C:\\Users\\craig\\Documents\\OneDrive - University of Massachusetts\\Published\\2019_AMHG_GRL\\working\\final\\NHD_join_table.csv')
+data <- read.csv('data\\Brinkerhoff_etal_2019\\field_measurements.csv')
+nhd_med <- read.csv('data\\Brinkerhoff_etal_2019\\NHD_join_table.csv')
 data <- left_join(data, select(nhd_med, SLOPE, SOURCE_FEA), by = c("site_no" = "SOURCE_FEA"))
 
 data <- filter(data, is.finite(chan_width) ==1) %>%
@@ -41,8 +39,7 @@ data$Vms <- data$chan_velocity*0.305 #m/s
 data$depth <- data$chan_area / data$width
 data$slope <- data$SLOPE
 
-
-######READ IN DATA WITH K600-------------------
+######READ IN FIELD DATA WITH K600-------------------
 #Ulseth etal 2019
 ulseth_data <- read.csv('data/Ulseth_etal_2019.csv', fileEncoding="UTF-8-BOM") #contains data from 5 studies
 ulseth_data <- ulseth_data[,-8]
@@ -63,10 +60,10 @@ data$eD <- g * data$slope * data$Vms #dissipation rate of surface turbulence ori
 data$Rh <- (data$depth*data$width)/(data$width + 2*data$depth) #hydraulic radius [m]
 data$ustar <- sqrt(g*data$Rh*data$slope)
 data$eS <- (data$ustar^3)/(0.41*data$depth) #dissipation rate of surface turbulence originating from bed friction [W/kg]
-data$Gtotal <- data$Vms*data$ustar^2
+data$Gtotal <- (data$Vms*data$ustar^2)/data$depth
 
 ###########SOME FLAGS-------------------------------
[…]

### 2021-07-26 manuscript_v4.Rmd
rewriting

@@
 ##Key Points
-- In large rivers, there exists a local equilibrium between production and dissipation of turbulent kinetic energy at the free surface
-- Gas exchange in large rivers is equivalently parameterized across multiple turbulence scales in open-channel flows
+- In large rivers, there exists a local equilibrium between the production and dissipation of turbulent kinetic energy at the free surface
+- Gas exchange in large rivers is equivalently parameterized across multiple turbulence scales, with the classical Kolmogorov model recovered empirically
 - BIKER algorithm exploits this theory, predicting gas exchange velocity and $CO_2$ fluxes from simulated SWOT satellite data and in situ $CO_2$ data
@@
 ##Abstract
-Go BIKER!!
-*AGU Advances. So 250 words here*
-*AGU advances is 8000 words total, though the format of the ones I've looked at vary quite a bit from long letter to short article. Many are clearly shorter than 8000 I think*
+*AGU Advances. So 250 words here* <br>
+*AGU advances is 8000 words total, the format seems to be a long letter so I've tried to right this in that style. Might be a bold organization but we'll see*
 
 ##Plain Language Summary
-Necessary for AGU advances (200 words)
+*Necessary for AGU advances (200 words)*
 
 ##1 Introduction
+Natural systems play a fundamental role in the budgeting and accounting of the global carbon cycle under climate change.
 Since the publication of @colePlumbingGlobalCarbon2007a, the global river network is recognized to emit substantial amounts of carbon in addition to exporting to the oceans.
-Current estimates of total carbon dioxide ($CO_2$) evasion to the atmosphere from the global river network vary from 650-1800 Tg C/yr [@lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a], with 167 Tg-C/yr coming from mountain streams alone [@horgbyUnexpectedLargeEvasion2019].
-Despite its incredibly small percentage of the global land surface, this is on par with the total oceanic $CO_2$ uptake rate [@gruberOceanicSinkAnthropogenic2019] and the global forest carbon uptake rate [@panLargePersistentCarbon2011].
-Though this flux is still relatively poorly constrained, it plays a critical role in the global carbon cycle and is of great interst to aquatic biogeochemists.
-Equation 1 represents this flux for any sparingly soluble gas given the gas concentration gradient between the water $C_{water}$ and the air $C_{air}$ and the gas exchange velocity *k*.
+Current estimates of total carbon dioxide evasion ($FCO_2$) to the atmosphere from the global river network vary from 650-1800 Tg C/yr [@lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a] **maybe add Shaoda here**, with 167 Tg-C/yr coming from mountain streams alone [@horgbyUnexpectedLargeEvasion2019].
+Despite its incredibly small percentage of the global land surface [0.47%- @raymondGlobalCarbonDioxide2013a], this flux is on par with the total oceanic $CO_2$ uptake rate [@gruberOceanicSinkAnthropogenic2019; @horgbyUnexpectedLargeEvasion2019] and the global forest carbon uptake rate [@panLargePersistentCarbon2011].
+It is still relatively poorly constrained but is clearly a critical component of the global carbon cycle.
+Equation 1 represents this flux given the gas concentration gradient between the water $CO_{2_{water}}$ and the air $CO_{2_{air}}$ and the gas exchange velocity *k*.
 
-$$\mathbf{(1)} Flux=k(C_{water}-C_{air})$$
+$$\mathbf{(1)} FCO_2=k(CO_{2_{water}}-CO_{2_{air}})$$
 
-The structure of equation 1 necessitates that calculations of this flux are highly sensitive to one's measurements/estimates of *k*.
-More broadly, *k* is also of significant importanceto aquatic ecologists modeling stream metabolism [@bernhardtMetabolicRegimesFlowing2018], water quality engineers [@chapraSurfaceWaterQualityModeling2008], and general interfacial mass transport [@lorkeUnifiedScalingRelation2006].
-Decades of work have focused on elucidating the physical mechanisms behind *k* in oceans and lakes and, to a lesser extent, fluvial gas exchange [see @wanninkhofAdvancesQuantifyingAirSea2009 and @hallGasExchangeStreams2020 for reviews on gas exchange in in oceans and rivers, respectively].
-Given that *k* should scale with turbulence in a fluid flow, a generalized model for *k* (equation 2) relies on the characteristic turbulence velocity scale $v_K$, which itself is a function of the characteristic length and time scales [@katulStructureFunctionModel2018].
+The structure of equation 1 necessitates that calculations of this flux are highly sensitive to one's measurements/estimates of *k* [@hallGasExchangeStreams2020].
+Even more broadly, fluvial *k* is of significant importance to aquatic ecologists modeling stream metabolism [e.g.
+@bernhardtMetabolicRegimesFlowing2018] and water quality engineers modeling river responses to pollutant loadings [among other things- @chapraSurfaceWaterQualityModeling2008].
+Decades of work have focused on elucidating the physical mechanisms behind *k* in oceans and lakes [@wanninkhofAdvancesQuantifyingAirSea2009] and, to a lesser extent, fluvial gas exchange [@hallGasExchangeStreams2020].
+Given that *k* should scale with turbulence in a turbulent flow, a generalized model for *k* (equation 2) relies on the characteristic turbulence velocity scale $v_K$, which itself is a function of the characteristic length and time scales [@katulStructureFunctionModel2018].
 Extensive field and labratory experiments have confirmed the validity of this model when $v_K$ is set to the Kolmogorov velocity scale [e.g.
-@lamontEddyCellModel1970; @zappaEnvironmentalTurbulentMixing2007; @lorkeUnifiedScalingRelation2006; @zappaVariationSurfaceTurbulence2003 **MORE**], where the idea is that the turbulent kinetic energy (TKE) cascade necessitates that TKE is produced by the largest, integral-scale turbulent eddies in the flow and then passed down the cascade to progressively smaller eddies until it is dissipated as heat at the Kolmogorov-scale: the smallest eddies in a turbulent flow.
+@lamontEddyCellModel1970; @zappaEnvironmentalTurbulentMixing2007; @lorkeUnifiedScalingRelation2006; @zappaVariationSurfaceTurbulence2003; @moogStreamReaerationNonuniform1999; @wangCoefficientsSmallEddy2015; @vachonRelationshipNearsurfaceTurbulence2010], where the idea is that turbulent kinetic energy (TKE) is produced by the largest, integral-scale turbulent eddies in the flow and then passed down the energy cascade to progressively smaller eddies until eddy motion is stable and TKE is dissipated as heat at the Kolmogorov-scale: the smallest eddies in a turbulent flow.
 Equation 2 via Kolmogorov-scale eddies can also be derived from both classic thin-film and surface renewal theories for gas exchange [@wangPhysicallyBasedScaling2021].
@@
 Specifically, the influence of the riverbed and its large roughness elements in small streams (i.e.
-rapids) complicates scaling fluvial *k* and might lead to substantially higher *k* in streams with whitewater [e.g.
+bedforms, logjams, sediment out of suspension) complicates scaling fluvial *k* and might lead to substantially higher *k* in streams with whitewater [e.g.
 @hallAirWaterOxygen2012; @ulsethDistinctAirWater2019a].
-Conversely, even less attention has been paid to the other side of the stream-to-river continumn, where large rivers' hydraulic properties begin to mimic small lakes and estuaries.
-To date, the handful of existing field studies of large-river *k* have suggested that their *k* properties begin to mimic small lakes with respect to wind dynamics, though little else is well-established [@alinPhysicalControlsCarbon2011; @beaulieuControlsGasTransfer2012a; @wangPhysicallyBasedScaling2021].
-Further, these mechanistic uncertainties are additionally limited by a large dearth of field-measured fluvial *k*.
-@wangPhysicallyBasedScaling2021 attempted to address this by simulating *k* in 35 rivers using stream metabolism modeling [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets, though this is still limited by the biases inherent in the DO data.
+Conversely, even less attention has been paid to the other side of the stream-to-river continumn, which is often hypothesized, in relation to *k*, as a hybrid condition of both fluvial and non-fluvial (i.e.
+lentic or estuarine) hydraulic properties.
+To date, the handful of existing field studies of large-river *k* have suggested that *k* begins to be notably influenced by wind dynamics, though little else is well-established [@alinPhysicalControlsCarbon2011; @beaulieuControlsGasTransfer2012a; @wangPhysicallyBasedScaling2021].
+These mechanistic uncertainties are additionally limited by a large dearth of field-measured fluvial *k*.
+@wangPhysicallyBasedScaling2021 attempted to address this by simulating *k* in 35 rivers using stream metabolism modeling [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets, finding that equation 2 is valid in their simulated rivers and that *k*~streamflow relationships are different in large and small rivers.
+However, they did not directly compare against *k* data in extremely small systems [@ulsethDistinctAirWater2019a] and stopped short of parsing out any hydraulic explanations for why this happens.
+
+Further, these mechanistic uncertainties are then propogated through upscaling workflows when biogeochemists predict *k* across thousands of rivers [e.g @borgesGloballySignificantGreenhousegas2015a; @horgbyUnexpectedLargeEvasion2019; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a] via equation 1 coupled with hydraulic geometry [HG: the scaling relationships between streamflow and river channel hydraulics- @leopoldHydraulicGeometryStream1953a], and it is currently not well understood how sensitive global estimates of fluvial gas evasion are to the specific HG model that is employed by the worker.
+Further, these approaches rely on either in situ discharge records or modeled streamflow which introduces additional uncertainities.
 This is all exacerbated in ungauged basins that cover large areas, especially in the carbon-rich Arctic inland waters, where little in situ information is available and fieldwork is impractical [@gleasonRemoteSensingRiver2020].
 
-These mechanistic uncertainties are then propogated through upscaling workflows when biogeochemists predict *k* across thousands of rivers [e.g @borgesGloballySignificantGreenhousegas2015a; @horgbyUnexpectedLargeEvasion2019; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a] using equation 3 coupled with hydraulic geometry (HG), or the scaling relationships between streamflow and river channel hydraulics [@leopoldHydraulicGeometryStream1953a].
-It is currently not well understood how sensitive global estimates of fluvial gas evasion are to the specific hydraulic geometry model that is employed by the worker.
-Further, these approaches rely on either in situ discharge records or modeled streamflow which introduces either additional uncertainity or the problems associated with ungauged rivers.
-
 A potential alternative to this upscaling approach is to directly estimate a river's hydraulic properties from remote sensing (RS) data.
-Remote sensing of river hydraulics is a burgeoning subfield within remote sensing of hydrology, often in service of remote sensing of river discharge (RSQ- see @gleasonRemoteSensingRiver2020 for a recent review).
-This is accomplished via two general approaches: 'gauged' methods which rely on in situ river data to calibrate one's method to the river(s) at hand and 'ungauged' techniques which focus on generalizability and global implementation in the service of merely improving existing knowledge in data-poor domains [@gleasonRemoteSensingRiver2020].
-Many, but not all, of these ungauged approaches were developed in the context of the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission.
+Remote sensing of river hydraulics is a burgeoning subfield within remote sensing of hydrology, often in service of remote sensing of river discharge [RSQ- @gleasonRemoteSensingRiver2020].
+This is accomplished via two general approaches: 'gauged' methods which rely on in situ river data to calibrate one's method to the river(s) at hand and 'ungauged' techniques which focus on hydraulic generalizability in the service of merely improving existing knowledge in data-poor domains [@gleasonRemoteSensingRiver2020].
+Many, but not all, of these ungauged approaches are developed in the context of the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission.
 SWOT is expected to launch in 2022 and provide the world's first global measurements of water surface extent and elevation at novel temporal resolutions.
-SWOT is a wide swath radar interferometer and will sample rivers every 1 to 7 days per 21 day repeat cycle.
-This yields an average sampling resolution of 11 days.
-It will measure rivers wider than 100m with a goal of expanding this to rivers at least 50m wide [@biancamariaSWOTMissionIts2016].
+SWOT is a wide swath radar interferometer and will sample rivers every 1 to 7 days per 21 day repeat cycle and will measure rivers wider than 100m with a goal of expanding this to rivers at least 50m wide [@biancamariaSWOTMissionIts2016].
 A decade of SWOT work has explored the multi-parameter problem of estimating the river hydraulic parameters of roughness and bathymetry from remote sensing to produce the SWOT discharge product [e.g.
@@
 
-In this context, here we revist the fundamental mechanisms behind gas exchange in large rivers using a simple TKE budget balance and one of the largest available datasets of field-measured river hydraulics [@brinkerhoffReconcilingAtaStationAtManyStations2019] to answer the following question: does *k* behave fundamentally differently in large rivers versus other streams and how does this relate to the classical model used to explain gas evasion in non-fluvial aquatic systems (equation 2, section 3)?
-We exploit the findings from this simple analysis to develop a novel methodology that predicts *k* and its explicit uncertainity solely using SWOT observations (section 4).
-It requires no in situ inputs of any kind (although in situ data can be ingested and will improve results).
-We name the RS of $k_{600}$ algorithm BIKER, or the 'Bayesian Inference/Inversion of the $k_{600}$ Evasion Rate' and validate it for 47 SWOT-observable rivers from around the world using hydraulic models to produce SWOT-like data (as SWOT has not yet launched- section 2).
+In this context, here we revist the fundamental mechanisms behind gas exchange and turbulence in large rivers using one of the largest available datasets of field-measured river hydraulics [@brinkerhoffReconcilingAtaStationAtManyStations2019] to answer the following question: does *k* behave fundamentally differently in large rivers than in other streams and how does this relate to the classical model of gas evasion in aquatic systems (equation 2)?
+We exploit the findings from this simple analysis to develop a novel methodology that predicts $k_{600}$ (*k* normalized to a Schmidt number of 600) and its explicit uncertainity solely using SWOT observations.
+The method requires no in situ inputs of any kind (although in situ data can be ingested and will improve results).
+We name the RS of $k_{600}$ algorithm BIKER, or the 'Bayesian Inference/Inversion of the $k_{600}$ Evasion Rate' and validate it for 47 SWOT-observable rivers from around the world using hydraulic models to produce SWOT-like data (as SWOT has not yet launched).
 We also quantify BIKER's sensitivity to the expected SWOT measurement errors on 17 of those rivers: while SWOT data represent a sea change in inland water monitoring, it is expected to have an approximately 10cm error in water surface elevation [@biancamariaSWOTMissionIts2016] as well as river width errors [@frassonExploringFactorsControlling2021].
-Finally, we use previously published dissolved $CO_2$ data to represent a hypotetical in situ $CO_2$ sensor and compare the bulk carbon efflux from the 47 rivers as calculated using BIKER and previously published in situ techniques for predicting $k_{600}$.
+Finally, we use previously published dissolved $CO_2$ data to represent a hypothetical in situ $CO_2$ sensor and compare the bulk carbon efflux from the 47 rivers as calculated using BIKER and previously published in situ techniques for predicting $k_{600}$.
 
@@
 ![Figure 1: Map of the 47 hydraulic models and 1 timeseries of CO2 samples used in this study.
-Note that hydraulic model locations are approximate as most of the models are not geo-referenced.
-We additionally used over 530,000 discrete measurements of river channel hydraulics from across the continetal United States (Brinkerhoff et al.
+Note that hydraulic model locations are approximate as some of the models are not geo-referenced.
+Not mapped here are over 530,000 discrete measurements of river channel hydraulics from across the continetal United States (Brinkerhoff et al.
 2019) and over 700 gas exchange velocity measurements (Ulseth et al.
-2019) that are not mapped here.](cache/maps/map_fin.jpg)
+2019).](cache/maps/map_fin.jpg)
 
 TKE and gas exchange theory (section 2) are explored using two datasets.
-First, we use a previously published compilation of field measurements that were originally made to calibrate United States Geological Survey (USGS) streamgauge rating curves and collected by @brinkerhoffReconcilingAtaStationAtManyStations2019.
+First, we use a previously published compilation of field measurements that were originally made to calibrate United States Geological Survey (USGS) streamgauge rating curves and gathered by @brinkerhoffReconcilingAtaStationAtManyStations2019.
 That dataset contains over 530,000 unique measurements of river channel velocity, width, depth, and discharge from across the continental United States.
 This is to our knowledge the largest collection of field measurements of river channel hydraulics and serves as the best dataset to calculate water surface TKE budgets.
-Second, we use the dataset from [@ulsethDistinctAirWater2019a] who measured and/or collected (from the literature) over 700 measurements of stream hydraulics and *k* and is, to our knowledge, the largest collection of field-measured river and stream *k*.
-This is also the dataset used to validate our *k* predictive models.
+Second, we use the dataset from @ulsethDistinctAirWater2019a who measured and/or collected from the literature over 700 measurements of stream hydraulics and $k_{600}$ and is, to our knowledge, the largest collection of field-measured river and stream $k_{600}$.
+This is also the dataset used to validate our *k* scaling models.
 
@@
 Ultimately, the 47 rivers are spread across the United States, France, Italy, the United Kingdom, and Bangladesh.
-We omit three models from @durandIntercomparisonRemoteSensing2016b because they lack enough hydraulic information to calculate $U_\star$, which is necessary for algorithm validation (both Saint Lawrence River models and the Tanana River model).
+We omit three models from @durandIntercomparisonRemoteSensing2016b because they lack enough hydraulic information to calculate $U_\star$, which is necessary for algorithm validation.
+These are both Saint Lawrence River models and the Tanana River model.
 
@@
 
-##3 The TKE budget in large rivers and implications for scaling river gas exchange velocity
-Turbulent flow is fundamental to fluvial geomorphology and open-channel flow and has been the subject of extensive research across scientific and engineering disciplines for over a century.
+##3 The TKE budget in large rivers and implications for scaling fluvial gas exchange velocity
+Turbulence is fundamental to fluvial geomorphology and open-channel flows and has been the subject of extensive research across scientific and engineering disciplines for over a century.
 Turbulent kinetic energy (TKE) is the energy generated and dissipated by the eddies created from the chaotic nature of turbulent flow.
@@
 
-$\epsilon$ is frequently described as $\epsilon(h)=\frac{U_*^3}{h}$.
-However in rivers, which exhibit additional depth-scale shear due to downslope flow over 'macroroughness elements' (channel geometry, bars, meanders, etc.), energy dissipation occurs much more evenly over the flow depth than this parameterization allows for.
+$\epsilon$ at the free surface is frequently described as $\epsilon(H)=\frac{U_*^3}{H}$.
+However in rivers, which exhibit additional depth-scale shear due to downslope flow over 'macroroughness elements' (i.e.
+channel geometry, bars, meanders, etc.), energy dissipation occurs much more evenly over the flow depth than this parameterization allows for.
 It has been shown that, for rivers at least, a more appropriate dissipation model considers stream power (per unit weight water: $\epsilon_D = gSU$), which represents the energy dissipated by the total frictional resistance of the river reach's geometry [@moogStreamReaerationNonuniform1999].
@@
 @raymondScalingGasTransfer2012a; @ulsethDistinctAirWater2019a].
-All three of these studies have confirmed that $\epsilon_D$ is far greater than $\epsilon$ in rivers, therefore generating the majority of fluvial TKE dissipation.
+Both @raymondScalingGasTransfer2012a and @moogStreamReaerationNonuniform1999 confirmed that $\epsilon_D$ is far greater than $\epsilon$ in open-channel flows, therefore generating the majority of fluvial TKE dissipation.
 
-We express *G* at the free surface in two different ways (equation 4) following @nakagawaTurbulenceOpenChannel1993 on the left and our derivation (detailed in Text S1) on the right.
+We express *G* at the free surface in two different ways (equation 4) following @nakagawaTurbulenceOpenChannel1993 on the left and our derivation on the right.
+Text S1 details the necessary algebra to arrive at both expressions of *G*.
 Looking at equation 4, it is readily apparent that $\frac{R_h}{H}$ will cancel out if they are equal.
@@
 In this scenario, the following three statements must be true: 1) $G \approx gSU \approx \epsilon_D \approx \epsilon$, 2) $\epsilon(H) \neq \frac{U_*^3}{H}$, and 3) $T_d \approx 0$.
-If this "large river" condition is met, then additonal TKE transported from the riverbed is functionally zero and the TKE budget at the free surface is just stream power (per unit weight water).
-It also implies a local equilibrium bewteen TKE produced and TKE dissipated at the free surface.
+More intuitively, if this "large river" condition is met then additonal TKE transported from the riverbed is functionally zero and the TKE budget at the free surface is just stream power (per unit weight water).
+This implies a local equilibrium between TKE produced and TKE dissipated at the free surface.
 $$\mathbf{(4)} G=\frac{UU_*^2}{H}=gSU\frac{R_h}{H} $$
 
-We test this hypothesis using the data from @brinkerhoffReconcilingAtaStationAtManyStations2019 and @ulsethDistinctAirWater2019a by plotting boxplots of $\frac{R_h}{H}$ ratios for over 531,000 different measurements of river hydraulics versus $\epsilon_D - G$.
+We test this hypothesis using the data from @brinkerhoffReconcilingAtaStationAtManyStations2019 and @ulsethDistinctAirWater2019a by plotting boxplots of $\frac{R_h}{H}$ ratios for over 531,000 different measurements of river hydraulics versus $\epsilon_D - G$, which is equivalent to $T_d$ in equation 3 assuming that $V_d$ is neglibile away from the wall [@nakagawaTurbulenceOpenChannel1993].
 *G* is calculated as $\frac{UU_*^2}{H}$.
-Note that this term is equivalent to $T_d$ in equation 1 assuming that $V_d$ is neglibile at the free surface [@nakagawaTurbulenceOpenChannel1993].
 In order to calculate when $R_h=H$, we round all ratios to 3 significant figures and set all ratios greater than or equal to 0.995 as $R_h=H$.
-If our hypothesis holds, then $\epsilon_D - G$ should be functionally zero only when $R_h=H$.
+If our hypothesis holds, then $\epsilon_D - G$ should be functionally zero only when the 'large river' condition is met.
 
-![Figure 2: Hydraulic radius/mean depth ratios (Rh/H) versus stream power per unit weight water - TKE production (eD-G, or equivalently, Td at the free surface) for our two datasets of field-measured river channel hydraulics.
-eD-G is functionally zero only when the Rh = H.
-This indicates that, when a river is large enough, the TKE energy budget is simply governed by stream power and TKE produced is equal to TKE dissipated.
-A local equilibrium of TKE is acheived, where energy generated by large-scale eddies is nearly equal to energy dissipated by the smallest scale eddies.
-In this figure, Rh/H bins contain between 92 and 138,307 measurements.
-](cache/k600_theory/turbulence.jpg)
+![Figure 2: Hydraulic radius/mean depth ratios (Rh/H) versus stream power per unit weight water - TKE production (eD-G at the free surface) for our two datasets of field-measured river channel hydraulics.
+eD-G is functionally zero only when a river is sufficiently large.
+This indicates that a local equilibrium of TKE is acheived and energy generated by large-scale eddies is nearly equal to energy dissipated by the smallest scale eddies.
+In this figure, Rh/H bins contain between 573 and 138,307 measurements.](cache/k600_theory/turbulence.jpg)
 
-Figure 2, which plots this test, shows that $T_d$, or $\epsilon_D-G$, is generally three to four orders of magnitude less when $R_h=H$ than it is when $\frac{R_h}{H}$ is between 90-99% (median $T_d$ of  $10^{-7.29}$ versus $10^{3.84} \frac{J}{kg*s}$, respectively).
-When the $\frac{R_h}{H}$ ratio is very small (0-9%), this rises to a median value of $10^{-2.56} \frac{J}{kg*s}$.
-We deem the $\epsilon_D - G$ values when $R_h=H$ (which are generally $10^{-6}$ to $10^{-8} \frac{J}{kg*s}$) to be functionally zero.
-This indicates that, when a river is large enough, the TKE energy budget at the free surface is simply stream power per unit weight water and a local equilibrium of free-surface TKE is acheived, where energy produced via the largest scale (turbulent) eddies is equivalent to energy dissipated via the smallest scale (Kolmorogov) eddies.
-This idea of a local equilibrium is certainly not new, however this is to our knowledge the first confirmation that it happens naturally in most large rivers.
+Figure 2, which plots this test, shows that $\epsilon_D-G$, or equivalently $T_d$, is nearly two orders of magnitude less when $R_h=H$ than it is when $\frac{R_h}{H}$ is between 90-99% (median $T_d$ across both datasets of $10^{-5.39}$ versus $10^{-3.83} \frac{J}{kg*s}$, respectively).
+When the $\frac{R_h}{H}$ ratio is very small (0-9%), this rises to a median value of $10^{-2.56} \frac{J}{kg*s}$, or nearly three orders of magnitude.
+We deem the $\epsilon_D - G$ values when $R_h=H$ (which are generally $10^{-5}$ to $10^{-8} \frac{J}{kg*s}$) to be functionally zero.
+This means that a local equilibrium of free-surface TKE is achieved, where energy produced via the largest scale (turbulent) eddies is equivalent to energy dissipated via the smallest scale (Kolmorogov) eddies.
+This idea of a local equilibrium is certainly not new, however this is to our knowledge the first confirmation that it happens naturally in most large rivers and that is arises algebraically from the TKE budget in open-channel flows.
 From here on we refer to the $R_h=H$ river scenario as the 'large river' condition.
 
-This finding as a significant impact on how we model *k* in these large rivers.
+This finding as a significant impact on how we model *k* in large rivers.
 Because the local equilibrium acheived during the 'large river condition' necessitates that $\epsilon$ is equivalent to *G* (Figure 2), we can presumably set $v_K$ equal to the largest ('integral-scale') eddies in the system, which are much more easily remotely-sensible than the Kolmogorov velocity scale.
-Otherwise, *k* should continue to scale only using the Kolmogorov-scale (which is similar across all turbulent flows by its definition).
+Otherwise, *k* should continue to scale only using the Kolmogorov-scale (which is consistent across all turbulent flows by its definition).
 
-To test this hypothesis, we fit the following three models to the @ulsethDistinctAirWater2019a dataset of *K* measurements all normalized to a Schmidt number of 600: $k_{600}=\beta U_*$, $k_{600}=\alpha \epsilon^{1/4}$, and $k_{600}=a (\epsilon_D)^b$ under two scenarios: $R_h=H$ and $R_h =/= H$.
-Note that because all data are normalized to a Schmidt number of 600, both the Schmidt numbers and viscosity terms are constant and are implicit in the statistical coefficient.
+To test this hypothesis, we fit the following three models to the @ulsethDistinctAirWater2019a dataset of *k* measurements normalized to a Schmidt number of 600: $k_{600}=\beta U_*$, $k_{600}=\alpha \epsilon^{1/4}$, and $k_{600}=a (\epsilon_D)^b$ under two scenarios: $R_h=H$ and $R_h =/= H$.
+Note that because all data are normalized to a Schmidt number of 600, both the Schmidt numbers and viscosity terms are constant and are implicit in the statistical coefficients.
 The model structures (and the fit parameters) are detailed in Table 1.
-The third model tested (rows 5 and 6 of Table 1), was fit to see which slope coefficient arrives empirically by fitting the generalized scaling function to the data.
+The third model tested (rows 5 and 6 of Table 1), was fit to see which slope coefficient is retrieved empirically by fitting the generalized scaling function to the data.
 Therfore, the $r2$ for these models is reflective of the log-transformed $k_{600}$ (unlike the other two models).
-This was the approach used by @ulsethDistinctAirWater2019a, who argue that bubble-mediated gas exchange yields steeper slopes (in high-energy streams) than 0.25.
+This was the approach used by @ulsethDistinctAirWater2019a, who observed slopes steeper than 0.25 in both regimes of their breakpointed regression model.
 
@@
 The results of this test are presented as Figure 3 and Table 1, where our hypothesis is confirmed: both the $U_*$ and $\epsilon^{1/4}$ models explain ~70% of variance in *k* under the 'large river' condition while the $U_*$ model does not work when this condition is not met ($r^2$:`r round(models_ustar[2,]$r2,2)` versus $r^2$=`r round(models_ustar[1,]$r2,2)`).
-In fact, Figures 3a and 3b are nearly identical in their scaling dynamics, while Figures 3d and 3e are very different, lending further evidence that these two scaling models are synonmous only when the 'large river' condition is met.
-@wangPhysicallyBasedScaling2021 also found similar predictive performance for their $k\sim U_*$ model on simulated data for 35 rivers where they assumed $R_h \approx H$ (reported $r^2$:0.76).
+In fact, Figures 3a and 3b are nearly visually identical in their scaling dynamics, while Figures 3d and 3e are very different, lending further evidence that these two scaling models are synonmous only when the 'large river' condition is met.
+This is in line with @wangPhysicallyBasedScaling2021, who found similar predictive performance and parameter values for these two models on 35 simulated rivers which they assumed large enough that $R_h \approx H$.
 Further, the 1/4 slope parameter is statistically recovered in the 'large river' condition (row 3, Table 1: slope = `r round(models_ulseth[2,]$slope,2)`).
@@
 The top row of Figure 3 confirms that under the 'large river condition' *k* can be scaled using either integral or Kolmogorov turbulence scales and that they are synonymous.
-This opens the door for remotely sensing *k* using almost entirely remotely-sensed parameters (section 4).
+This opens the door for remotely sensing *k* using almost entirely remotely-sensed parameters.
+This is elaborated on in section 4.
 
@@
 
-|**Figure 3 Panel**|**River Regime**|**Theoretical Model**|**Model Structure**|**r2**|**Intercept Parameter**|**Slope Parameter**|
+|**Figure 3 Panel**|**River Regime**|**Model Structure**|**r2**|**Intercept Parameter**|**Slope Parameter**|
 |---------------------|------------------|-------------------|------|---------------|---------------|------------------|
-|3a|$R_h=H$|$k \sim Sc^{-n}U_*$|$k_{600}=\beta U_*$|`r round(models_ustar[2,]$r2, 2)`|`r round(models_ustar[2,]$int, 2)`|`r round(models_ustar[2,]$slope, 2)`|
-|3d|$R_h=/=H$|$k \sim Sc^{-n}U_*$|$k_{600}=\beta U_*$|`r round(models_ustar[1,]$r2, 2)`|`r round(models_ustar[1,]$int, 2)`|`r round(models_ustar[1,]$slope, 2)`|
-|3b|$R_h=H$|$k\sim Sc^{-n}(\epsilon)^{1/4}$|$k_{600}=\alpha(\epsilon_D)^{1/4}$|`r round(models_eD4[2,]$r2, 2)`|`r round(models_eD4[2,]$int, 2)`|`r round(models_eD4[2,]$slope, 2)`|
-|3e|$R_h=/=H$|$k\sim Sc^{-n}(v\epsilon)^{1/4}$|$k_{600}=\alpha(\epsilon_D)^{1/4}$|`r round(models_eD4[1,]$r2, 2)`|`r round(models_eD4[1,]$int, 2)`|`r round(models_eD4[1,]$slope, 2)`|
-|3c|$R_h=H$|$k\sim Sc^{-n}(v\epsilon)^b$|$k_{600}=a(\epsilon_D)^{b}$|`r round(models_ulseth[2,]$r2, 2)`|`r round(models_ulseth[2,]$int, 2)`|`r round(models_ulseth[2,]$slope, 2)`|
-|3f|$R_h=/=H$|$k\sim Sc^{-n}(v\epsilon)^b$|$k_{600}=a(\epsilon_D)^{b}$|`r round(models_ulseth[1,]$r2, 2)`|`r round(models_ulseth[1,]$int, 2)`|`r round(models_ulseth[1,]$slope, 2)`|
+|3a|$R_h=H$|$k_{600}=\beta U_*$|`r round(models_ustar[2,]$r2, 2)`|`r signif(models_ustar[2,]$int, 3)`|`r round(models_ustar[2,]$slope, 2)`|
+|3d|$R_h=/=H$|$k_{600}=\beta U_*$|`r round(models_ustar[1,]$r2, 2)`|`r signif(models_ustar[1,]$int, 3)`|`r round(models_ustar[1,]$slope, 2)`|
+|3b|$R_h=H$|$k_{600}=\alpha(\epsilon_D)^{1/4}$|`r round(models_eD4[2,]$r2, 2)`|`r signif(models_eD4[2,]$int, 3)`|`r round(models_eD4[2,]$slope, 2)`|
+|3e|$R_h=/=H$|$k_{600}=\alpha(\epsilon_D)^{1/4}$|`r round(models_eD4[1,]$r2, 2)`|`r signif(models_eD4[1,]$int, 3)`|`r round(models_eD4[1,]$slope, 2)`|
+|3c|$R_h=H$|$k_{600}=a(\epsilon_D)^{b}$|`r round(models_ulseth[2,]$r2, 2)`|`r signif(models_ulseth[2,]$int, 3)`|`r round(models_ulseth[2,]$slope, 2)`|
+|3f|$R_h=/=H$|$k_{600}=a(\epsilon_D)^{b}$|`r round(models_ulseth[1,]$r2, 2)`|`r signif(models_ulseth[1,]$int, 3)`|`r round(models_ulseth[1,]$slope, 2)`|
 
-The 1/4 exponent is clearly not valid when $R_h=/=H$ (Figure 3b).
-However, the generalized form of this model explains 73% of the variation in log-transformed *k* and was deemed a strong scaling model for fluvial *k* when rivers do not meet the 'large river' condition.
+The 1/4 exponent is clearly not valid when $R_h=/=H$ (Figure 3e).
+However, the generalized form of this model explains `r 100*round(models_ulseth[1,]$r2,2)`% of the variation in log-transformed *k* and was deemed a strong scaling model for fluvial *k* when rivers do not meet the 'large river' condition.
 That model yields a slope of `r round(models_ulseth[1,]$slope,2)` which corroborates other models with slopes steeper than 1/4 in rivers [@wangPhysicallyBasedScaling2021; @ulsethDistinctAirWater2019a; @raymondScalingGasTransfer2012a].
-Further theoretical modeling is needed to parse out why fluvial *k* in smaller, steeper systems generally scales at a steeper rate than in non-fluvial systems, and how this relates to the breakpointed model proposed by @ulsethDistinctAirWater2019a.
+Further theoretical modeling is needed to parse out why fluvial *k* in systems that do not meet the 'large river condition' generally scale at a steeper rate than in non-fluvial systems, and how this relates to the breakpointed model proposed by @ulsethDistinctAirWater2019a.
+The TKE budget in equation 3 suggests that it is a function of turbulent diffusion of additional TKE from the bed to the free surface.
 
-##4 Extending this finding to remotely sense riverine gas exchange velocity via the BIKER algorithm
-We have shown that scaling *k* by either integral or Kolmogorov turbulence scales explains ~70% of the variation in $k_{600}$ in large rivers.
+##4 Exploiting the 'large river' model to remotely sense gas exchange velocity via the BIKER algorithm
+We have shown that scaling $k_{600}$ by either integral or Kolmogorov turbulence scales explains ~70% of the variation in $k_{600}$ in large rivers and that their scaling dynamics are virtually identical.
 The integral-scale model relies solely on the shear velocity $U_*$.
@@
 This problem is experienced by both SWOT RSQ algorithms and existing in situ techniques that concurrently solve for *k* and stream metabolism [@applingOvercomingEquifinalityLeveraging2018; @graceFastProcessingDiel2015; @holtgrieveSimultaneousQuantificationAquatic2010].
-In both of these domains, the other parameters are quite difficult to estimate (bed roughness and stream metabolism, respectively) while for BIKER, median channel area is relatively easy to estimate from the SWOT-observable river width [@brinkerhoffConstrainingRemoteRiver2020a].
+In both of these domains, the other parameters are quite difficult to estimate (bed roughness and stream metabolism, respectively) while for BIKER, median channel area is relatively easy to approximate from the SWOT-observable river width [@brinkerhoffConstrainingRemoteRiver2020a].
 
@@
 First, we verified that this is true using both the @brinkerhoffReconcilingAtaStationAtManyStations2019 and @ulsethDistinctAirWater2019a datasets (Figure S3), finding that the vast majority of rivers with a width of at least 100m exhibit this $R_h=H$ behavior.
-Thus, our theoretical findings in sections 2 and 3 should hold in nearly all rivers that SWOT will observe and we should be able to remotely estimate $k_{600}$.
+Thus, our theoretical findings in sections 2 and 3 should hold in nearly all rivers that SWOT will observe and we should be able to remotely estimate $k_{600}$ via SWOT.
 BIKER is informed by the @hagemannBAMBayesianAMHGManning2017a algorithm for ungauged RSQ, further explored in more recent work by @brinkerhoffConstrainingRemoteRiver2020a.
@@
 Therefore, we are interested in solving for $p(\Theta|x)$, or the 'posterior' distribution.
-For BIKER, *x* is the SWOT-observables: river width *W* and water surface elevation *H* (which is used to calculate the water surface slope $_h$).
+For BIKER, *x* is the SWOT-observables: river width *W* and water surface elevation *H* (which is used to calculate the water surface slope $S_h$).
 Note that $p(x)$ is usually computationally intractable to integrate, so Bayesian inference tools require only the proportionality to be specified: $p(\Theta|x) \propto f(x|\Theta)p(x)$.
@@
 *dA* is estimated assuming a rectangular river channel (usually valid when $R_h=H$) so that $dA \approx \delta H*W$.
-As mentioned earlier, equifinality issues are limited in this model as there is only one unknown besides $k_{600}$, and $A_0$ is reasonably easy to approximate from river width [@brinkerhoffConstrainingRemoteRiver2020a].
-There are no reliances on bed roughness, flow velocity, or any other terms that are difficult to infer from river width.
+Thus, there are no reliances on bed roughness, flow velocity, or any other terms that are difficult to infer from river width in BIKER.
 
@@
 More intuitively, BIKER priors represent our 'prior river knowledge' of what $A_0$ and $k_{600}$ probably are for some river since they cannot be remotely sensed.
-Hyperparameter specifications are detailed in Text S2, however the goal was to rely on absolutely no in situ information such that we could run this method on any river on Earth that is SWOT-observable.
+Hyperparameter specifications are detailed in Text S2, however the goal was to rely on absolutely no in situ information such that we could run this method on any river on Earth solely using SWOT observations.
 
@@
 We validate BIKER on 47 SWOT-simulated rivers using daily observed $R_h$ and observed $S_h$.
-We also re-validate BIKER on the 17 rivers with the SWOT error model corrupting the SWOT-observables *W* and *H* (this workflow and the metrics used are detailed in Text S4).
+We also re-validate BIKER on the 17 rivers with the SWOT error model corrupting the SWOT-observables *W* and *H* (this workflow and the metrics used are detailed in Text S4 and Table S1).
 Regardless of the validation setup, we do not have observed $k_{600}$ data for these rivers, and to our knowledge no field dataset of $k_{600}$ exists in the type of temporal and spatial frequency that SWOT (and therefore the BIKER algorithm) provides.
 Therefore, we take the model outlined in row 1 of Table 1 and use that to calculate the observed $k_{600}$ that BIKER is validated against.
-With this setup, we are directly exploring BIKER's ability to infer observed $A_0$ and $k_{600}$ from *W* and *H* alone, as the model has already been sucessfully validated (Table 1, Figure 3).
+With this setup, we are directly exploring BIKER's ability to infer observed $A_0$ and $k_{600}$ from *W* and *H* alone, as the *k* scaling model has already been sucessfully validated (Table 1, Figure 3).
 
@@
 We see, more clearly than in Figure 4a, that BIKER captures $k_{600}$ very well: most of the  observed CDF falls nearly exactly on or adjacent to the posterior means (Figure 4b).
-However, the uppermost quartile is systematically overestimated by BIKER (also visible in Figure 4a).
+However, the uppermost quartile is systematically overestimated by BIKER.
 We propose that this is due to the model used to assign the $k_{600}$ prior hyperparameters being less representative for these rivers.
-In line with this, the rivers with the poorest BIKER performance uniformally are the narrowest and steepest in the set (Figure S4).
-In summary, Figure 3 confirms that we improve upon our baseline understanding of $k_{600}$ in these rivers: we accurately capture $k_{600}$ with no in situ information about the river while simultaneously and explicitly accounting for the uncertainties inherent in our $k_{600}$ estimates.
+In line with this, the rivers with the poorest BIKER performance uniformally are the narrowest and steepest in the dataset (Figure S4).
+In summary, Figure 3 confirms that we improve upon our baseline understanding of $k_{600}$ in these rivers: we accurately capture $k_{600}$ using no in situ information about the river while simultaneously and explicitly accounting for the uncertainties inherent in our $k_{600}$ estimates.
 
@@
 See Table 1 for metric definitions.
-Dashed lines denote scores of 1, 0, and -0.41 for r2 (section 3.2).
+Dashed lines denote scores of 0, 0.50, and 1.
 b-d: validation timeseries for three rivers representative of good, reasonable, and poor BIKER performance.
 b) was randomly selected from the upper tertile of NRMSE scores, c) was randomly selected from the middle tertile, and d) from the worst tertile.
+See Figure S5 for all other rivers.
+TO DO.
 Model results include the posterior means and 95% CIs.](cache/validation/validation_by_river.jpg)
 
-Median r2 is `r round(median(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$r2, na.rm=T),2)`, which is excellent given that absolutely no in situ information is being used to predict $k_{600}$.
+Median river-specific $r^2$ is `r round(median(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$r2, na.rm=T),2)`, which is excellent given that absolutely no in situ information is being used to predict $k_{600}$.
 NRMSE and RRMSE have median scores of `r round(median(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$nrmse, na.rm=T),2)` and `r round(median(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$rrmse, na.rm=T),2)`, respectively, which are again good for a completely ungauged method.
+These are comparable to the best NRMSE scores achieved by the SWOT RSQ algorithms [@frassonExploringFactorsControlling2021].
 Median rBIAS is `r round(median(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$rBIAS, na.rm=T),2)`, highlighting a  positive bias in most rivers' predictions.
 This further supports the visual evidence in Figure 4 that sometimes BIKER is overestimating the magnitude of $k_{600}$ and that this might be river-specific.
-While median r2 and rBIAS scores were strong, the ranges of these scores were somewhat large (standard deviation for r2 of `r round(sd(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$r2, na.rm=T),2)` and for rBIAS of `r round(sd(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$rBIAS, na.rm=T),2)`).
+While median $r^2$ and rBIAS scores were strong, the ranges of these scores were somewhat large (standard deviation for $r^2$ of `r round(sd(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$r2, na.rm=T),2)` and for rBIAS of `r round(sd(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$rBIAS, na.rm=T),2)`).
 
 Figure 5b-d are representative timeseries plots of predicted and observed $k_{600}$ for three rivers chosen randomly from those with 'good' NRMSE scores (b), 'okay' NRMSE scores (c), and 'bad' NRMSE scores (d).
-See the Figure 5 caption for how this was determined.
+See the Figure 5 caption for how this was determined and see Figure S5 for the other river timeseries plots (**TO DO**).
 For the Ohio Section 2 River, the entire timeseries of $k_{600}$ is nearly perfectly predicted, while in the Seine Upstream River the dynamics (peaks and valleys) are reasonably captured but they are magnified to be far larger than the observed dynamics.
@@
 It is one thing to accurately predict *k*, but researchers are often most interested in the actual gas fluxes from rivers and ultimately the carbon emitted from river to atmosphere.
-Therefore, we also explore 1) BIKER's ability to reproduce $CO_2$ evasion fluxes ($FCO_2$ $[M/L^2T]$) from these 47 rivers, and 2) a comparison of the representative carbon efflux $[M/T]$ from BIKER with eastbalished in situ methods.
-The details of this workflow are in Text S4, but broadly we pair the biweekly 26 $CO_2$ samples (section 2, figure S2) from @beaulieuControlsGasTransfer2012a with a subset of SWOT observations (as the data are not daily).
-We then calculate $FCO_2$ using the equation 7 and assuming atmospheric $CO_{2_{air}}$ is 390 uatm.
+Therefore, we also explore 1) BIKER's ability to reproduce $FCO_2$ $[M/L^2T]$ (equation 1) from these 47 rivers, and 2) a comparison of the representative carbon efflux $[M/T]$ from BIKER with established in situ methods.
+The details of this workflow are in Text S4, but broadly we pair the biweekly 26 $CO_2$ and water temperature samples (section 2, figure S2) from @beaulieuControlsGasTransfer2012a with a subset of SWOT observations (as the data are not daily).
+We then calculate $FCO_2$ using equation 1 and assuming atmospheric $CO_{2_{air}}$ is 390 uatm.
 Finally, we estimate a median bulk carbon efflux using BIKER's posterior means and three other in situ models for average channel depth (used to calculate $U_\star$) used for upscaling: one trained on the @brinkerhoffReconcilingAtaStationAtManyStations2019 dataset, and two previously published models [@raymondGlobalCarbonDioxide2013a; @raymondScalingGasTransfer2012a].
@@
 This allows us to assess whether BIKER's estimates (wholly ungauged) are comparable to gauged methods (all four HG models).
-Finally, we express the bulk carbon efflux as the median mass flow rate of carbon (via $CO_2$ evasion) per year from the 47 rivers after accounting for total river surface area.
-
-$$\mathbf{(7)} FCO_2 = k_{CO_2}([CO_{2_{water}} - [CO_{2_{air}}])$$
 
 In Figure 6a, there is a very strong fit to the observed $FCO_2$ data, with an RMSE of `r round(results_3_3_all$rmse,2)` $g/m^2dy$.
-The performance is notably better than for $k_{600}$ alone (Figure 4a) and there is no systematic bias in the predicitons across all 47 rivers.
+The performance is notably better than for $k_{600}$ alone (Figure 4a) and there is no systematic bias in the predictions across all 47 rivers.
 This is presumably due to the structure of the $FCO_2$ equation, which reduces the relative importance of errors in *k* given that the $CO_2$ data is measured in situ.
-$FCO_2$ prediction intervals are slightly narrower than those presented in Figure 4a.
+$FCO_2$ prediction intervals are slightly narrower than those presented in Figure 4a as well.
 
@@
 
-Figure 6b compares the bulk carbon efflux (via $CO_2$ evasion) from the 47 rivers using BIKER posterior means and four streamgauge-based HG models.
-Both the BIKER bulk carbon efflux (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_BIKER',]$sumFCO2, 0)` gG-C/yr) and the 'Brinkerhoff 2019' estimate (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_brinkerhoff2019',]$sumFCO2, 0)`) are quite close to the observed flux (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_obs',]$sumFCO2, 0)`), with the latter nearly perfectly recovering the observed value.
+Figure 6b compares the bulk carbon efflux (via $F_{CO_2}$) from the 47 rivers using BIKER posterior means and four streamgauge-based HG models.
+Both the BIKER bulk carbon efflux (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_BIKER',]$sumFCO2, 2)` gG-C/yr) and the 'Brinkerhoff 2019' estimate (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_brinkerhoff2019',]$sumFCO2, 2)`) are quite close to the observed flux (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_obs',]$sumFCO2, 2)`), with the latter nearly perfectly recovering the observed value.
 BIKER overestimates the efflux somewhat.
-The 'raymond 2013' model is even more overestimated (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_raymond2013',]$sumFCO2, 0)`), while the 'Raymond 2012' model appreciably underestimates this bulk efflux (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_raymond2012',]$sumFCO2, 0)`).
+The 'raymond 2013' model is even more overestimated (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_raymond2013',]$sumFCO2, 2)`), while the 'Raymond 2012' model appreciably underestimates this bulk efflux (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_raymond2012',]$sumFCO2, 2)`).
 Thus, despite BIKER using absolutely no in situ data like the other 4 methods do, it provides a reasonable estimate of the carbon efflux and better a estimate than 2 of the approaches that rely on a streamgauge (Figure 6b).
@@
 In sections 2 and 3 we have shown that this must be the case given the algebraic structure of the TKE budget in open-channel flows [@nakagawaTurbulenceOpenChannel1993] and the assumptions underlying turbulent scaling with interfacial fluxes [@katulStructureFunctionModel2018; @lorkeUnifiedScalingRelation2006].
-This result profoundly amends how we scale *k* in large rivers and directly addresses a long-standing problem of upscaling fluvial *k*: that our current methods perform poorer in larger rivers than small ones.
+This result significantly amends how we can scale *k* in large rivers and directly addresses a long-standing problem of upscaling fluvial *k*: that our current methods perform poorer in larger rivers than small ones.
 
-We also statistically recovered the 1/4 scaling exponent in the classic $k \sim \epsilon^{1/4}$ model (Figure 3c) for gas exchange that has been repeadely validated in non-fluvial aquatic systems.
-This again suggests that *k* in large rivers behaves more similarly to lakes, esturaries, and the ocean than to steeper rivers where bubble-mediated gas exchange might occur [@ulsethDistinctAirWater2019a].
-We also suggest that the TKE budget in smaller rivers and streams is additionally complicated by diffusive transport of TKE from the bed to the free surface, but this TKE production mechanism is functionally zero if the 'large river' condition is met (Figure 2).
-Channel bed roughness effects riverine gas exchange mostly in high energy streams where slope is sufficiently steep for bubble-induced gas exchange to occur [@hallGasExchangeStreams2020].
+We also statistically recovered the 1/4 scaling exponent in the classic $k \sim \epsilon^{1/4}$ model (Figure 3c, row 5 Table 1) for gas exchange that has been repeadely validated in non-fluvial aquatic systems.
+This again suggests that *k* in large rivers behaves more similarly to lakes, esturaries, and the ocean than to steeper rivers.
+We also suggest that the TKE budget in smaller rivers and streams is additionally complicated by diffusive transport of TKE from the bed to the free surface, and that $T_d$ in equation 3 might explain why fluvial $k_{600}$ does not scale to the 1/4 power in smaller rivers.
+The presence of an appreciable $T_d$ term also increases the influence of channel bed roughness.
 @ulsethDistinctAirWater2019a showed bed roughness loosely correlates with $k_{600}$ in steep Alpine streams.
@@
 @chansonStreamReaerationNonuniform; @moogStreamReaerationNonuniform1999].
-In this study, we confirm that bed roughness is exerting functionally no influence on free-surface TKE in large rivers (Section 2 and figure 2) and therefore *k*.
+In this study and under the 'large river' condition, we confirm that bed roughness is exerting functionally no influence on free-surface TKE (Section 2 and figure 2) and therefore *k*.
 We show that free-surface TKE must be entirely produced and dissipated by form-drag and not bottom friction or roughness.
@@
 This limited knowledge of large-scale *k* spatiotemporal dynamics is due both to a lack of process-level understanding [@hallGasExchangeStreams2020] but also a lack of measurements.
-Authors have argued that the key to explaining the large residual variation in upscaling models is to explore at-a-station temporal variability in *k* [@hallGasExchangeStreams2020].
 
@@
 SWOT will provide daily hydraulic measurements for a 3 month fast sampling period for calibration and validation and sampling thereafter between 1 and 7 days per 21 day repeat cycle [@biancamariaSWOTMissionIts2016].
-BIKER's success in 1) infering $k_{600}$ using simulated SWOT data over a wide range of rivers (Figures 4 and 5) and 2) being robust to measurement errors internal to the SWOT data (Figure 3) bode well for BIKER's eventual implementation on real SWOT observations.
+BIKER's success in 1) infering $k_{600}$ using simulated SWOT data over a wide range of rivers (Figures 4 and 5) and 2) being robust to measurement errors internal to the SWOT data (Figure 5) bode well for BIKER's eventual implementation on real SWOT observations.
 The results presented here suggest that daily estimation of riverine gas exchange globally could be possible once SWOT launches.
@@
 ###6.3 Towards remotely sensing riverine carbon emissions using SWOT
-Figure 6 confirms that BIKER is quite successful, without any in situ information aside from a $CO_2$ logger, at predicting 1) $FCO_2$ (Figure 6a) and 2) the bulk carbon efflux (Figure 6b).
+Figure 6 confirms that BIKER is quite successful, without any in situ information aside from a $CO_2$ logger, at predicting both $FCO_2$ (Figure 6a) and 2) the bulk carbon efflux (Figure 6b).
 This encouraging result has two main implications for future work coupling remote sensing via SWOT with in situ data.
-First, it confirms that we can couple BIKER and SWOT with in situ gas concentration loggers to produce $FCO_2$ estimates at novel temporal resolution in SWOT-observable rivers.
-This is particualry useful given recent advances in high temporal resolution in situ gas concentration measurements [@ahoHighestRatesGross2021] but no such similar advances in modeling *k* at high temporal resolutions.
+First, it confirms that we can couple BIKER and SWOT with in situ gas concentration loggers to produce $FCO_2$ estimates at novel temporal resolutions in SWOT-observable rivers.
+This is particualry useful given recent advances in in situ gas concentration loggers [@ahoHighestRatesGross2021] but no such similar advances in modeling *k* at equivalent temporal resolutions.
 BIKER can likely also be ran at the field scale using arrays of pressure transducers to estimate water surface slope (rather than using satellite-based altimeters like SWOT) following recent work doing the same using the @hagemannBAMBayesianAMHGManning2017a RSQ algorithm [@harlanDischargeEstimationDense].
@@
 ##7 Conclusions
+Gas exchange from aquatic systems has been studied for nearly a century, with a robust model (equation 2) that has been repeatedly verified across many non-fluvial environments.
+Despite renewed interest in fluvial gas exchange in the last decade or so, there are considerable uncertainties in how we model the gas exchange velocity in river environments and particularly in large rivers where wind increasingly influences *k*.
+Here, we show algebraically that TKE budgets in big rivers exist in a local equilbrium at the free surface and that we can equivalently scale *k* using any characteristic turbulence scale.
+We validate this using over 530,000 measurements of river channel hydraulics and over 700 discrete measurements of the normalized gas exchange velocity $k_{600}$.
+We then exploit this finding to remotely sense $k_{600}$ using simulated data that will be provided by the forthcoming SWOT satellite, showing good performance and relying on absoutely no on-the-ground information.
+Finally, we pair this algorithm (named 'BIKER') with an in situ $CO_2$ logger and show strong performance in reproducing $CO_2$ evasion fluxes and carbon efflux from the rivers' surfaces.
+
+These strong results functionally open the door for global-scale, near daily estimates of fluvial gas exchange velocity once SWOT launches in 2022.
+This unprecedented amount of data should allow for significant insights into both the mechanistic and temporal dynamics of fluvial gas exchange in large rivers around the world.
+This, in turn, should allow us to better parameterize upscaling workflows such that the global fluvial carbon flux is better constrained.
 
@@
 |*Bulk carbon efflux*|carbon mass transport rate from river to atmosphere|NA|$[\frac{M}{T}]$|
-|$C_{water}$|Water-side gas concentration|NA|$[L]$|
-|$C_{air}$|Air-side gas concentration|NA|$[L]$|
+|$CO_{2_{water}}$|Water-side $CO_2$ concentration|NA|$[L]$|
+|$CO_{2_{air}}$|Atmospheric-side $CO_2$ concentration|NA|$[L]$|
 |$\epsilon$|TKE dissipation rate|$\frac{U_*^3}{H}$|$[\frac{E}{M*T}]$|

### 2021-07-26 supp_v4.Rmd
rewriting

@@
 ---
-title: "Gas exchange in big rivers controlled by stream power and turbulence scales: implications for remotely sensing gas exchange"
+title: "Supplemental Information: Gas exchange in large rivers controlled by largest turbulent eddies: implications for remotely sensing gas exchange via SWOT"
 author: "CB Brinkerhoff, CJ Gleason, PA Raymond, MH Harlan"
@@
 ## Text S1: Deriving two expressions for turbulent kinetic energy production
-Below we detail how we arrived at our two expressions for a cross-section-averaged turbulent kinetic energy production (*G*), or equation 2 in the main text.
+Below we detail how we arrived at our two expressions for a cross-section-averaged turbulent kinetic energy production (*G*), or equation 3 in the main text.
+Please see Appendix A for variable descriptions and notation.
 
@@
 They used an extensive database of field measurements and machine learning to identify patterns that associate river width with the hydraulic priors needed to run McFLIs so that prior hyperparameters may be assigned to rivers using only the existing remotely sensed data.
-For this study, we extracted $\lambda$ and $\gamma$ as the Xth percentile values rather than the absoulte maximum and minimum values to avoid physically impossible bounds on $A_0$.
+For this study, we extracted $\lambda_{A_0}$ and $\gamma_{A_0}$ as the **Xth** percentile values rather than the absolute maximum and minimum values to avoid physically impossible bounds on $A_0$.
+
 This leaves the $k_{600}$ hyperparameters to be defined.
-We assigned those by using a simple slope regression model to predict $U_*$, trained on the @ulsethDistinctAirWater2019a: equation S9 assigns $\mu$ while $\sigma^2$ was set to equation S9's log-transformed standard error after propogation of errors for both equation S9 and equation 4 in the main text: 1.12.
-$\lambda$ and $\gamma$ were set to log(0.001) m/day and log(500) m/day, respectively.
-Finally, we estimate $\sigma_{k_{600}}$ using the standard model error from equation 4 in the main text and as fit on the data shown in Figure 3a.
+We assigned those by using a simple slope regression model to predict $U_*$ solely from a SWOT-observable value.
+It was trained on the @ulsethDistinctAirWater2019a dataset.
+Equation S9 assigns $\mu_{k_{600}}$ while $\sigma^2_{k_{600}}$ was set to equation S9's log-transformed standard error after propogation of errors for both equation S9 and equation 6 in the main text.
+This amounted to a value of 1.12.
+$\lambda_{k_{600}}$ and $\gamma_{k_600}}$ were set to log(0.001) m/day and log(500) m/day, respectively.
+Finally, we estimate $\sigma_{k_{600}}$ using the standard model error from equation 6 in the main text and as fit on the data shown in Figure 3a.
 
@@
 Regardless of the validation setup or SWOT error budget used, we do not have observed $k_{600}$ data for these rivers, and to our knowledge no field dataset of $k_{600}$ exists in the type of temporal and spatial frequency that SWOT (and therefore the BIKER algorithm) provides.
-Therefore, we take the model outlined in row 1 of Table 1 and use that to calculate the observed $k_{600}$, that BIKER is validated against (equation S10).
+Therefore, we take the model outlined in row 1 of Table 1 in the main text and use that to calculate the observed $k_{600}$, that BIKER is validated against (equation S10).
 
@@
 
-It also means that, for a fair validation scheme, $\sigma_{k_{600}}^2$ in equation 5 in the main text must be set to reflect only error from our assumptions about calculating *dA* and not the parameter uncertainity inherent in the model coefficient (i.e.
-the 56.0294 in equation 4 in the main text).
+It also means that, for a fair validation scheme, $\sigma_{k_{600}}^2$ in equation 7 in the main text must be set to reflect only error from our assumptions about calculating *dA* and not the parameter uncertainity inherent in the model coefficient (i.e.
+the 56.0294 in equation 6 in the main text).
 Because this model is only theoretically valid when $R_h=H$, we convientantly do not need to account for uncertanties associated with assuming that $R_h = H$.
@@
 To validate across all rivers and timesteps, we used the coefficient of determination $r^2$ and the root mean square error RMSE.
-Three normalized metrics were used for by-river validation: RRMSE and NRMSE are normalized root mean square errors that have been normalized by the observed value and the mean observed value (respectively).
+Four normalized metrics were used for by-river validation: RRMSE and NRMSE are normalized root mean square errors that have been normalized by the observed value and the mean observed value (respectively).
 rBIAS is a measure of prediction bias that is normalized by the mean observed value.
-r2 is again used on the by-river case.
+r2 is again used on the by-river case too.
 
-## Text S4: BIKER FCO2 calculations
+## Text S4: BIKER FCO2 workflow
 To calculate $CO_2$ fluxes and carbon efflux, we pair @beaulieuControlsGasTransfer2012a's biweekly 26 $CO_2$ and water temperature samples with every 11th SWOT observation by date, ignoring the timesteps beyond 26.
@@
 
-![Figure S3 Stuff](cache/k600_theory/swotPlot.jpg)
+![Figure S3 Cummulative density functions (CDFs) of river width measurements, as binned by their dataset and and whether or not they meet the 'large river' condition (Rh=H).
+The dashed black line indicates the minimum river width that the SWOT satellite will observe (100m).
+This brief analysis indicates that the vast majority of SWOT-observable rivers should meet the 'large river' condition.](cache/k600_theory/swotPlot.jpg)
 

### 2021-07-26 generateMS.R

@@ -5,6 +5,7 @@
 setwd('C:\\Users\\craig\\Documents\\GitHub\\RSK600')
 
 #Move all results to global environment
+turbulence_stats <- read.csv('cache\\k600_theory\\turbulence_stats.csv')
 models_ulseth <- read.csv('cache\\k600_theory\\ulseth_models.csv')
 models_ustar <- read.csv('cache\\k600_theory\\ustar_models.csv')
 models_eD4 <- read.csv('cache\\k600_theory\\eD_models.csv')

### 2021-07-26 src/swot_k_model.R

@@ -59,7 +59,7 @@ data <- rbind(data, ulseth_data_s)
 data$eD <- g * data$slope * data$Vms #dissipation rate of surface turbulence originating from depth-scale form drag [W/kg]
 data$Rh <- (data$depth*data$width)/(data$width + 2*data$depth) #hydraulic radius [m]
 data$ustar <- sqrt(g*data$Rh*data$slope)
-data$eS <- (data$ustar^3)/(0.41*data$depth) #dissipation rate of surface turbulence originating from bed friction [W/kg]
+data$eS <- (data$ustar^3)/(data$depth) #dissipation rate of surface turbulence originating from bed friction [W/kg]
 data$Gtotal <- (data$Vms*data$ustar^2)/data$depth
 
 ###########SOME FLAGS-------------------------------
@@ -87,15 +87,13 @@ data$figureFlag <- ifelse(round(data$Rh/data$depth, 3) >= 0.995, 'Rh=H',
                                                                     ifelse(round(data$Rh/data$depth, 3) >=0.20, '20-29%',
                                                                            ifelse(round(data$Rh/data$depth, 3) >= 0.10, '10-19%', '0-9%'))))))))))
 
-#useful stats
-data$w_d <- data$width/data$depth
-group_by(data, figureFlag) %>% 
+#useful stats across both datasets tested
+turbulence_stats <- group_by(data, figureFlag) %>% 
   summarise(mean = mean(Td),
             median = median(Td),
             sd = sd(Td),
-            n=n(),
-            meanW = mean(w_d),
-            medianW = median(w_d))
+            n=n())
+write.csv(turbulence_stats, 'cache\\k600_theory\\turbulence_stats.csv')
 
 #create boxplots
 boxes <- ggplot(data, aes(x=figureFlag, y=Td, fill=study)) +
@@ -250,21 +248,6 @@ text2 <- ggdraw() + draw_label("All Other Rivers", fontface = 'bold', x = 0.5, y
 k600Plot <- plot_grid(text1, ustar_k600_wide, eD4_k600_wide, ulseth_k600_wide, text2, ustar_k600_narrow, eD4_k600_narrow, ulseth_k600_narrow, labels=c(NA, 'a', 'b', 'c', NA, 'd', 'e', 'f'), ncol=4)
[…]

### 2021-07-28 manuscript_v4.Rmd
pete notes and ms updates

@@
 Natural systems play a fundamental role in the budgeting and accounting of the global carbon cycle under climate change.
-Since the publication of @colePlumbingGlobalCarbon2007a, the global river network is recognized to emit substantial amounts of carbon in addition to exporting to the oceans.
+Since the publication of @colePlumbingGlobalCarbon2007a, the global river network is recognized to emit substantial amounts of carbon to the atmosphere, in addition to exporting it to the oceans.
 Current estimates of total carbon dioxide evasion ($FCO_2$) to the atmosphere from the global river network vary from 650-1800 Tg C/yr [@lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a] **maybe add Shaoda here**, with 167 Tg-C/yr coming from mountain streams alone [@horgbyUnexpectedLargeEvasion2019].
@@
 The structure of equation 1 necessitates that calculations of this flux are highly sensitive to one's measurements/estimates of *k* [@hallGasExchangeStreams2020].
-Even more broadly, fluvial *k* is of significant importance to aquatic ecologists modeling stream metabolism [e.g.
+Even more broadly, fluvial *k* for various dissolved gases are of significant importance to aquatic ecologists modeling stream metabolism [e.g.
 @bernhardtMetabolicRegimesFlowing2018] and water quality engineers modeling river responses to pollutant loadings [among other things- @chapraSurfaceWaterQualityModeling2008].
 Decades of work have focused on elucidating the physical mechanisms behind *k* in oceans and lakes [@wanninkhofAdvancesQuantifyingAirSea2009] and, to a lesser extent, fluvial gas exchange [@hallGasExchangeStreams2020].
-Given that *k* should scale with turbulence in a turbulent flow, a generalized model for *k* (equation 2) relies on the characteristic turbulence velocity scale $v_K$, which itself is a function of the characteristic length and time scales [@katulStructureFunctionModel2018].
+Given that *k* should scale with turbulence in a turbulent flow [@hallGasExchangeStreams2020], a generalized model for *k* (equation 2) relies on the characteristic turbulence velocity scale $v_K$, which itself is a function of the characteristic length and time scales [@katulStructureFunctionModel2018].
 Extensive field and labratory experiments have confirmed the validity of this model when $v_K$ is set to the Kolmogorov velocity scale [e.g.
@@
 @hallAirWaterOxygen2012; @ulsethDistinctAirWater2019a].
-Conversely, even less attention has been paid to the other side of the stream-to-river continumn, which is often hypothesized, in relation to *k*, as a hybrid condition of both fluvial and non-fluvial (i.e.
-lentic or estuarine) hydraulic properties.
+Less attention has been paid to the other side of the stream-to-river continumn, which is often conceptualized in relation to *k* as a hybrid condition of both fluvial and non-fluvial hydraulic properties.
 To date, the handful of existing field studies of large-river *k* have suggested that *k* begins to be notably influenced by wind dynamics, though little else is well-established [@alinPhysicalControlsCarbon2011; @beaulieuControlsGasTransfer2012a; @wangPhysicallyBasedScaling2021].
 These mechanistic uncertainties are additionally limited by a large dearth of field-measured fluvial *k*.
-@wangPhysicallyBasedScaling2021 attempted to address this by simulating *k* in 35 rivers using stream metabolism modeling [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets, finding that equation 2 is valid in their simulated rivers and that *k*~streamflow relationships are different in large and small rivers.
-However, they did not directly compare against *k* data in extremely small systems [@ulsethDistinctAirWater2019a] and stopped short of parsing out any hydraulic explanations for why this happens.
+@wangPhysicallyBasedScaling2021 attempted to address this by simulating *k* in 35 rivers using stream metabolism modeling [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets, finding that equation 2 is valid in their simulated rivers and that *k*~streamflow relationships breakdown in large rivers.
+However, they did not directly compare against *k* data in extremely small systems [@ulsethDistinctAirWater2019a] and stopped short of parsing out hydraulic explanations for why this happens.
 
-Further, these mechanistic uncertainties are then propogated through upscaling workflows when biogeochemists predict *k* across thousands of rivers [e.g @borgesGloballySignificantGreenhousegas2015a; @horgbyUnexpectedLargeEvasion2019; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a] via equation 1 coupled with hydraulic geometry [HG: the scaling relationships between streamflow and river channel hydraulics- @leopoldHydraulicGeometryStream1953a], and it is currently not well understood how sensitive global estimates of fluvial gas evasion are to the specific HG model that is employed by the worker.
+Further, these mechanistic uncertainties are then propogated through upscaling workflows when biogeochemists predict *k* across thousands of rivers [e.g @borgesGloballySignificantGreenhousegas2015a; @horgbyUnexpectedLargeEvasion2019; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a] via equation 1 coupled with hydraulic geometry [HG: the scaling relationships between streamflow and river channel hydraulics- @leopoldHydraulicGeometryStream1953a].
+It is currently not well understood how sensitive global estimates of fluvial gas evasion are to the specific HG model that is employed by the worker.
 Further, these approaches rely on either in situ discharge records or modeled streamflow which introduces additional uncertainities.
@@
 That dataset contains over 530,000 unique measurements of river channel velocity, width, depth, and discharge from across the continental United States.
-This is to our knowledge the largest collection of field measurements of river channel hydraulics and serves as the best dataset to calculate water surface TKE budgets.
+This is to our knowledge the largest collection of field measurements of river channel hydraulics.
 Second, we use the dataset from @ulsethDistinctAirWater2019a who measured and/or collected from the literature over 700 measurements of stream hydraulics and $k_{600}$ and is, to our knowledge, the largest collection of field-measured river and stream $k_{600}$.
@@
 Ultimately, the 47 rivers are spread across the United States, France, Italy, the United Kingdom, and Bangladesh.
-We omit three models from @durandIntercomparisonRemoteSensing2016b because they lack enough hydraulic information to calculate $U_\star$, which is necessary for algorithm validation.
-These are both Saint Lawrence River models and the Tanana River model.
+We omit three models from @durandIntercomparisonRemoteSensing2016b because they lack enough hydraulic information to calculate the shear velocity $U_\star$, which is necessary for algorithm validation (see Appendix A for all variable descriptions and notation used in this study).
+These are two models for the Saint Lawrence River and one for the Tanana River.
 
@@
 *G* is the TKE produced, $\epsilon$ is the TKE dissipated, $T_d$ is TKE transported from the riverbed via turbulent diffusion, and $V_d$ is TKE transported via viscous diffusion.
-See Appendix A for all variable descriptions and notation.
 $$\mathbf{(3)} G+T_d+V_d=\epsilon$$
 
-$\epsilon$ at the free surface is frequently described as $\epsilon(H)=\frac{U_*^3}{H}$.
-However in rivers, which exhibit additional depth-scale shear due to downslope flow over 'macroroughness elements' (i.e.
-channel geometry, bars, meanders, etc.), energy dissipation occurs much more evenly over the flow depth than this parameterization allows for.
+$\epsilon$ at the free surface is frequently described as $\epsilon(H)=\frac{U_*^3}{H}$ where *H* is the flow depth.
+However in rivers, which exhibit additional depth-scale shear due to downslope flow against the channel geometry, bars, and meanders, energy dissipation occurs much more evenly over the flow depth than this parameterization allows for.
 It has been shown that, for rivers at least, a more appropriate dissipation model considers stream power (per unit weight water: $\epsilon_D = gSU$), which represents the energy dissipated by the total frictional resistance of the river reach's geometry [@moogStreamReaerationNonuniform1999].
@@
 *G* is calculated as $\frac{UU_*^2}{H}$.
-In order to calculate when $R_h=H$, we round all ratios to 3 significant figures and set all ratios greater than or equal to 0.995 as $R_h=H$.
+In order to determine when $R_h=H$, we round all ratios to 3 significant figures and set all ratios greater than or equal to 0.995 as $R_h=H$.
 If our hypothesis holds, then $\epsilon_D - G$ should be functionally zero only when the 'large river' condition is met.
 
-![Figure 2: Hydraulic radius/mean depth ratios (Rh/H) versus stream power per unit weight water - TKE production (eD-G at the free surface) for our two datasets of field-measured river channel hydraulics.
+![Figure 2: Rh/H versus eD-G at the free surface for our two datasets of field-measured river channel hydraulics (see Appendix A for variable descriptions).
 eD-G is functionally zero only when a river is sufficiently large.
@@
 
-Figure 2, which plots this test, shows that $\epsilon_D-G$, or equivalently $T_d$, is nearly two orders of magnitude less when $R_h=H$ than it is when $\frac{R_h}{H}$ is between 90-99% (median $T_d$ across both datasets of $10^{-5.39}$ versus $10^{-3.83} \frac{J}{kg*s}$, respectively).
-When the $\frac{R_h}{H}$ ratio is very small (0-9%), this rises to a median value of $10^{-2.56} \frac{J}{kg*s}$, or nearly three orders of magnitude.
-We deem the $\epsilon_D - G$ values when $R_h=H$ (which are generally $10^{-5}$ to $10^{-8} \frac{J}{kg*s}$) to be functionally zero.
+Figure 2, which plots this test, shows that $\epsilon_D-G$ is nearly two orders of magnitude less when $R_h=H$ than it is when $\frac{R_h}{H}$ is between 90-99% (median value across both datasets of $10^{-5.39}$ versus $10^{-3.83} \frac{J}{kg*s}$, respectively).
+When the $\frac{R_h}{H}$ ratio is very small (0-9%), this rises to a median value of $10^{-2.56} \frac{J}{kg*s}$, or nearly three orders of magnitude larger.
+The $\epsilon_D - G$ values when $R_h=H$ (which are generally $10^{-5}$ to $10^{-8} \frac{J}{kg*s}$) are functionally zero.
 This means that a local equilibrium of free-surface TKE is achieved, where energy produced via the largest scale (turbulent) eddies is equivalent to energy dissipated via the smallest scale (Kolmorogov) eddies.
@@
 
-This finding as a significant impact on how we model *k* in large rivers.
-Because the local equilibrium acheived during the 'large river condition' necessitates that $\epsilon$ is equivalent to *G* (Figure 2), we can presumably set $v_K$ equal to the largest ('integral-scale') eddies in the system, which are much more easily remotely-sensible than the Kolmogorov velocity scale.
+This finding has a significant impact on how we model *k* in large rivers.
+Because the local equilibrium acheived during the 'large river condition' necessitates that $\epsilon$ is equivalent to *G* (Figure 2), we can presumably set $v_K$ in equation 2 equal to the largest ('integral-scale') eddies in the system, which are much more easily remotely-sensible than the Kolmogorov velocity scale.
 Otherwise, *k* should continue to scale only using the Kolmogorov-scale (which is consistent across all turbulent flows by its definition).
@@
 Therfore, the $r2$ for these models is reflective of the log-transformed $k_{600}$ (unlike the other two models).
-This was the approach used by @ulsethDistinctAirWater2019a, who observed slopes steeper than 0.25 in both regimes of their breakpointed regression model.
+This was the approach used by @ulsethDistinctAirWater2019a, who observed slopes steeper than 1/4 in both regimes of their breakpointed regression model.
 
@@
 In fact, Figures 3a and 3b are nearly visually identical in their scaling dynamics, while Figures 3d and 3e are very different, lending further evidence that these two scaling models are synonmous only when the 'large river' condition is met.
-This is in line with @wangPhysicallyBasedScaling2021, who found similar predictive performance and parameter values for these two models on 35 simulated rivers which they assumed large enough that $R_h \approx H$.
+This is in line with @wangPhysicallyBasedScaling2021, who found similar predictive performance and parameter values for these two models on 35 simulated riverswhich they assumed were large enough that $R_h \approx H$.
 Further, the 1/4 slope parameter is statistically recovered in the 'large river' condition (row 3, Table 1: slope = `r round(models_ulseth[2,]$slope,2)`).
-We note that the large residuals in Figure 3c are a byproduct of the model existing in log-space and that panels 3b and 3c are functionally identical (see Table 1 for their nearly identical intercept and slope parameters).
+We note that the large residuals in Figure 3c are a byproduct of the model existing in log-space and that panels 3b and 3c are functionally identical models (see Table 1 for their nearly identical intercept and slope parameters).
 The top row of Figure 3 confirms that under the 'large river condition' *k* can be scaled using either integral or Kolmogorov turbulence scales and that they are synonymous.
@@
 Further theoretical modeling is needed to parse out why fluvial *k* in systems that do not meet the 'large river condition' generally scale at a steeper rate than in non-fluvial systems, and how this relates to the breakpointed model proposed by @ulsethDistinctAirWater2019a.
-The TKE budget in equation 3 suggests that it is a function of turbulent diffusion of additional TKE from the bed to the free surface.
+That said, the TKE budget in equation 3 suggests that it is a function of turbulent diffusion of additional TKE from the bed to the free surface.
 
@@
 The integral-scale model relies solely on the shear velocity $U_*$.
-Because of the structure of $U_*$ (Appendix A), we can easily reduce the 'large river' $k_{600}$ model (Figure 3a, row 1 Table 1) to an equation consisting solely of remotely-sensible river hydraulics and an inferred estimate of the channel area.
+Because of the structure of $U_*$ (Appendix A), we can easily reduce the 'large river' $k_{600}$ model (Figure 3a, row 1 Table 1) to an equation consisting solely of remotely-sensible river hydraulics and an intial estimate of the channel area.
 By scaling *k* this way, we significantly reduce potential equifinality issues by having just two unknowns: $k_{600}$ and the median channel area $A_0$.
 Equifinality refers to an under-constrained mathematical system that has essentially infinite parameter combinations that can produce the same result: there are in essence more unknowns than equations [@garamboisInferenceEffectiveRiver2015].
-This problem is experienced by both SWOT RSQ algorithms and existing in situ techniques that concurrently solve for *k* and stream metabolism [@applingOvercomingEquifinalityLeveraging2018; @graceFastProcessingDiel2015; @holtgrieveSimultaneousQuantificationAquatic2010].
-In both of these domains, the other parameters are quite difficult to estimate (bed roughness and stream metabolism, respectively) while for BIKER, median channel area is relatively easy to approximate from the SWOT-observable river width [@brinkerhoffConstrainingRemoteRiver2020a].
+This problem is experienced by both SWOT RSQ algorithms and in situ tools that concurrently solve for *k* and stream metabolism [@applingOvercomingEquifinalityLeveraging2018; @graceFastProcessingDiel2015; @holtgrieveSimultaneousQuantificationAquatic2010].
+In both of these domains, the other unknown parameters are often difficult to estimate (bed roughness and stream metabolism, respectively) while for BIKER, median channel area is relatively easy to approximate from the SWOT-observable river width [@brinkerhoffConstrainingRemoteRiver2020a].
 
@@
 Thus, our theoretical findings in sections 2 and 3 should hold in nearly all rivers that SWOT will observe and we should be able to remotely estimate $k_{600}$ via SWOT.
-BIKER is informed by the @hagemannBAMBayesianAMHGManning2017a algorithm for ungauged RSQ, further explored in more recent work by @brinkerhoffConstrainingRemoteRiver2020a.
+Therefore, we develop the BIKER algorithm, which is informed by the @hagemannBAMBayesianAMHGManning2017a algorithm for ungauged RSQ and further explored in more recent work by @brinkerhoffConstrainingRemoteRiver2020a.
 These papers conceptualize discharge as a Bayesian remote sensing problem, which we largely follow here to conceptualize $k_{600}$ as a Bayesian remote sensing problem that can be solved using SWOT data.
@@
 The heart of BIKER is its reformulation of the $k_{600}$ model as a Bayesian sampling model that is conditional only on the data that SWOT will provide.
-To do this, $k_{600}$ first needs to be written as a function of SWOT-observables *W* and $S_h$.
+To do this, $k_{600}$ is first written as a function of SWOT-observables *W* and $S_h$.
 This algebra is carried out using the model parameter from row 1 of Table 1 and yields equation 6, where we follow our earlier assumption that $R_h = H =\frac{A}{W}$ where *A* is wetted channel area.
@@
 *dA* is estimated assuming a rectangular river channel (usually valid when $R_h=H$) so that $dA \approx \delta H*W$.
-Thus, there are no reliances on bed roughness, flow velocity, or any other terms that are difficult to infer from river width in BIKER.
+Thus, there are no reliances on bed roughness, flow velocity, or any other terms that are difficult to infer from river width.
 
@@
 The points are the posterior means while the black lines are the 95% confidence intervals (CIs) for the predictions.
-$k_{600}$ is strongly correlated with the BIKER-predicted $k_{600}$ ($r^2$ of 0.76).
+$k_{600}$ is strongly correlated with the BIKER-predicted $k_{600}$ ($r^2$ of `r round(results_3_2_all$r2, 2)`).
 Using absolutely no in situ information, BIKER captures the magnitude of the predictions and most points fall on or near the 1:1 line.
-The regression of the estimates (solid grey line) nearly perfectly recovers the 1:1 line (dashed black), but there is an overestimation bias in the largest $k_{600}$ values.
+The regression of the estimates (solid grey line) nearly recovers the 1:1 line (dashed black), but there is an over/underestimation bias in the largest/smallest $k_{600}$ values.
 The RMSE for the BIKER predictions is `r round(results_3_2_all$rmse,2)` m/day) across all observations.
@@
 Grey line is linear regression (and 95% prediction intervals are dashed) and dashed black line is 1:1 line.
-b: Cumulative density functions (CDFs) of the same results: dashed black line are observed values, green line are BIKER posterior means, and blue lines are BIKER 95% CIs.](cache/validation/validation.jpg)
+b-d: validation timeseries for three rivers representative of good, reasonable, and poor BIKER performance.
+b) was randomly selected from the upper tertile of NRMSE scores, c) was randomly selected from the middle tertile, and d) from the worst tertile.
+See Figure S5 for all other rivers.
+Model results include the posterior means and 95%.](cache/validation/validation.jpg)
 
-Figure 4a highlights a overestimation bias by BIKER when $k_{600}$ is larger, which is confirmed in Figure 4b.
-Figure 4b plots the cumulative density functions (CDFs) of observed and predicted $k_{600}$ where the green line is the CDF of the BIKER posterior means, the blue lines are the CDFs of the BIKER posterior 95% CIs, and the dashed black line is the CDF of observed $k_{600}$.
-We see, more clearly than in Figure 4a, that BIKER captures $k_{600}$ very well: most of the  observed CDF falls nearly exactly on or adjacent to the posterior means (Figure 4b).
-However, the uppermost quartile is systematically overestimated by BIKER.
-We propose that this is due to the model used to assign the $k_{600}$ prior hyperparameters being less representative for these rivers.
-In line with this, the rivers with the poorest BIKER performance uniformally are the narrowest and steepest in the dataset (Figure S4).
-In summary, Figure 3 confirms that we improve upon our baseline understanding of $k_{600}$ in these rivers: we accurately capture $k_{600}$ using no in situ information about the river while simultaneously and explicitly accounting for the uncertainties inherent in our $k_{600}$ estimates.
+Figure 4b-d are representative timeseries plots of predicted and observed $k_{600}$ for three rivers chosen randomly from those with 'good' NRMSE scores (b), 'okay' NRMSE scores (c), and 'bad' NRMSE scores (d).
+See Table S1 for the definition of NRMSE, the Figure 4 caption for how this was determined, and Figure S5 for the other river timeseries plots.
+For the Missouri Midsection River, the entire timeseries of $k_{600}$ is nearly perfectly predicted, while in the Seine River the dynamics (peaks and valleys) are reasonably captured but they are magnified to be far larger than the observed dynamics.
+Generally, though, mean $k_{600}$ is reasonably recovered for the Seine, as confirmed visually.
+In the Fresno River, there is significant positive bias in the estimates but also massive uncertainty (per the 95% CIs) in those estimates, indicating that BIKER is highly uncertain about its output (and rightfully so).
+Correct temporal dynamics are also largely missing from BIKER's Fresno River predicitions.
 
 Figure 5a plots validation metrics calculated for each river with and without SWOT measurement error (green and purple, respectively).
-The boxplots are composed of validation scores for either the 47 or 17 rivers- see Table S1 for metric definitions.
+The points making up these boxplots (47 and 17, respectively) are overlain atop the boxplots.
+See Table S1 for metric definitions.
 SWOT measurement errors neglibly influence BIKER's performance across all four error metrics (Figure 5a), though caution should be used in over-interepting boxplots with a sample size of only 17.
@@
 See Table 1 for metric definitions.
+b: NRMSE scores (per river) versus the number of SWOT observations used in the Bayesian inversion.
+As expected, performance improves substantially with large amounts of data and degrades significantly with small amounts of data.
 Dashed lines denote scores of 0, 0.50, and 1.
-b-d: validation timeseries for three rivers representative of good, reasonable, and poor BIKER performance.
-b) was randomly selected from the upper tertile of NRMSE scores, c) was randomly selected from the middle tertile, and d) from the worst tertile.
-See Figure S5 for all other rivers.
-TO DO.
-Model results include the posterior means and 95% CIs.](cache/validation/validation_by_river.jpg)
+CIs.](cache/validation/validation_by_river.jpg)
 
@@
 
-Figure 5b-d are representative timeseries plots of predicted and observed $k_{600}$ for three rivers chosen randomly from those with 'good' NRMSE scores (b), 'okay' NRMSE scores (c), and 'bad' NRMSE scores (d).
-See the Figure 5 caption for how this was determined and see Figure S5 for the other river timeseries plots (**TO DO**).
-For the Ohio Section 2 River, the entire timeseries of $k_{600}$ is nearly perfectly predicted, while in the Seine Upstream River the dynamics (peaks and valleys) are reasonably captured but they are magnified to be far larger than the observed dynamics.
-Generally, though, mean $k_{600}$ is actually reasonably recovered for the Seine Upstream, as confirmed visually.
-In the Connecticut River, there is significant positive bias in the estimates but also massive uncertainty (per the 95% CIs) in those estimates, indicating that BIKER is highly uncertain about its output (and rightfully so).
-Correct temporal dynamics are also largely missing from BIKER's Connecticut River predicitions.
+Figure 5b highlights one benefit of using Bayesian inference to estimate $k_{600}$: because the posterior is conditional on the SWOT observations, performance should improve with more data.
+Figure 5b plots by-river NRMSE scores versus the number of SWOT observations.
+While performance varies considerably when observations are up to ~400, the three rivers with nearly 600 observations universally show excellent BIKER performance, and the worst BIKER performance is universally in rivers with only 12 observations.
 
@@
 Therefore, we also explore 1) BIKER's ability to reproduce $FCO_2$ $[M/L^2T]$ (equation 1) from these 47 rivers, and 2) a comparison of the representative carbon efflux $[M/T]$ from BIKER with established in situ methods.
-The details of this workflow are in Text S4, but broadly we pair the biweekly 26 $CO_2$ and water temperature samples (section 2, figure S2) from @beaulieuControlsGasTransfer2012a with a subset of SWOT observations (as the data are not daily).
+The details of this workflow are in Text S4, but broadly we pair the 26 biweekly $CO_2$ and water temperature samples (section 2, figure S2) from @beaulieuControlsGasTransfer2012a with a subset of SWOT observations (as the data are not daily).
 We then calculate $FCO_2$ using equation 1 and assuming atmospheric $CO_{2_{air}}$ is 390 uatm.
@@
 Figure 6b compares the bulk carbon efflux (via $F_{CO_2}$) from the 47 rivers using BIKER posterior means and four streamgauge-based HG models.
-Both the BIKER bulk carbon efflux (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_BIKER',]$sumFCO2, 2)` gG-C/yr) and the 'Brinkerhoff 2019' estimate (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_brinkerhoff2019',]$sumFCO2, 2)`) are quite close to the observed flux (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_obs',]$sumFCO2, 2)`), with the latter nearly perfectly recovering the observed value.
-BIKER overestimates the efflux somewhat.
-The 'raymond 2013' model is even more overestimated (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_raymond2013',]$sumFCO2, 2)`), while the 'Raymond 2012' model appreciably underestimates this bulk efflux (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_raymond2012',]$sumFCO2, 2)`).
-Thus, despite BIKER using absolutely no in situ data like the other 4 methods do, it provides a reasonable estimate of the carbon efflux and better a estimate than 2 of the approaches that rely on a streamgauge (Figure 6b).
+The results are largely equivalent for both the mean and median carbon efflux.
+Both the BIKER estimate (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_BIKER',]$sumFCO2, 2)` gG-C/yr) and the 'Raymond 2013' estimate (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_raymond2013',]$sumFCO2, 2)`) are somewhat overestimated relative to the observed flux (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_obs',]$sumFCO2, 2)`).
+The 'Brinkerhoff 2019' model nearly perfectly recovers the observed value (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_brinkerhoff2019',]$sumFCO2, 2)`).
+Finally, the 'Raymond 2012' model slightly underestimates this bulk efflux (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_raymond2012',]$sumFCO2, 2)`).
+Thus, despite BIKER using absolutely no streamgauge or other in situ data like the other 3 methods do, it provides a reasonable estimate of the carbon efflux (Figure 6b).
 
@@
 BIKER's success in 1) infering $k_{600}$ using simulated SWOT data over a wide range of rivers (Figures 4 and 5) and 2) being robust to measurement errors internal to the SWOT data (Figure 5) bode well for BIKER's eventual implementation on real SWOT observations.
-The results presented here suggest that daily estimation of riverine gas exchange globally could be possible once SWOT launches.
+Further, BIKER's improved performance with longer datasets (Figure 5b) bodes well for future SWOT implementation, as the three-year campaign will provide well north of 600 observations for many rivers, suggesting even better BIKER inversions than those presented here.
+All of this suggests that daily estimation of riverine gas exchange globally could be possible once SWOT launches.
 
@@
 |------------|---------------|-------------------------------|---------|
+|*A*|Channel cross-sectional area|NA|$[L^2]$|
 |*Bulk carbon efflux*|carbon mass transport rate from river to atmosphere|NA|$[\frac{M}{T}]$|
@@
 |$CO_{2_{air}}$|Atmospheric-side $CO_2$ concentration|NA|$[L]$|
+|$D_m$|Molecular diffusion coefficient|NA|$[\frac{L^2}{T}]$|
 |$\epsilon$|TKE dissipation rate|$\frac{U_*^3}{H}$|$[\frac{E}{M*T}]$|
@@
 |$G$|TKE production rate|$\frac{UU_*}{H}$|$[\frac{E}{M*T}]$|
-|$h$|Elevation above the bed|NA|$[L$|
+|$h$|Elevation above the bed|NA|$[L]$|
 |$H$|Mean flow depth|NA|$[L]$|
 |$k$|gas exchange velocity|NA|$[\frac{L}{T}]$|
-|$k_{600}$|gas exchange velocity normalized to Sc=600|NA|$[\frac{L}{T}]$|
+|$k_{600}$|gas exchange velocity normalized to Sc=600|$(\frac{600}{Sc_k})^{-0.5}k$|$[\frac{L}{T}]$|
 |$R_h$|Hydraulic radius|$\frac{AW}{2H+W}$|$[L]$|
 |$S$|River slope|NA|$[\frac{L}{L}]$|
-|$Sc$|Schmidt number|NA|$unitless$|
+|$Sc$|Schmidt number|$\frac{v}{D_m}$|$unitless$|
 |$T_d$|TKE turbulent diffusion rate|NA|$[\frac{E}{M*T}]$|

### 2021-07-28 generateMS.R

@@ -12,7 +12,6 @@ models_eD4 <- read.csv('cache\\k600_theory\\eD_models.csv')
 results_3_2_rivs <- read.csv('cache//validation//results_by_riv.csv')
 results_3_2_all <- read.csv('cache//validation//results_all_riv.csv')
 results_3_3_all <- read.csv('cache//FCO2//fco2_stats_all.csv')
-results_3_3_rivs <- read.csv('cache//FCO2//fco2_stats_by_river.csv')
 results_3_3_bulk <- read.csv('cache//FCO2//bulkFluxes.csv')
 results_3_3_bulk$key <- as.character(results_3_3_bulk$key)
 

### 2021-07-28 src/validation.R

@@ -110,8 +110,8 @@ plotSWOTreaches <- ggplot(plot_stats, aes(x=key, y=value, fill=factor(errFlag)))
 ############
 ##NRMSE VS AMOUNT OF DATA-----------------------
 ############
-nrmse_vs_n <- ggplot(stats_by_reach, aes(y=nrmse, x=n_data, color=factor(errFlag))) +
-  geom_point(size=5, aes(pch=factor(errFlag)),show.legend=FALSE) +
+nrmse_vs_n <- ggplot(stats_by_reach[stats_by_reach$errFlag == 0,], aes(y=nrmse, x=n_data, color=factor(errFlag))) +
+  geom_point(size=5,show.legend=FALSE) +
   geom_hline(yintercept=1, linetype='dashed') +
   geom_hline(yintercept=0, linetype='dashed') +
   geom_hline(yintercept=0.50, linetype='dashed') +

### 2021-09-09 manuscript_v4.Rmd
big push for reframing as a finding of channel efficiency

@@
 title: "Gas exchange in large rivers controlled by largest turbulent eddies: implications for remotely sensing gas exchange via SWOT"
-author: 'CB Brinkerhoff, CJ Gleason, PA Raymond, MH Harlan'
+author: 'CB Brinkerhoff, CJ Gleason, CJ Zappa, PA Raymond, MH Harlan'
 date: "`r Sys.Date()`"

### 2021-09-09 src/swot_k_model.R

@@ -179,7 +179,6 @@ plots_2 <- plot_grid(wdPlot, tke_depthPlot, ncol=2, labels=c('b','c'), label_siz
 plot_1 <- plot_grid(boxes_TKE, plots_2, ncol=1, labels=c('a',NA), label_size = 20)
 ggsave('cache\\k600_theory\\turbulence.jpg', plot_1, width=14, height=15)
 
-
 #confirm that depth-scale and Kolmogorov-scale turbulence is ~similar when Rh=H (used later in the k600 plot but need in memory now)--------------
 ratio_stats <- group_by(data, figureFlag) %>% 
   summarise(meanratio = mean(((1e-6*eD)^0.25)/ustar),
@@ -187,7 +186,7 @@ ratio_stats <- group_by(data, figureFlag) %>%
             n=n())
 write.csv(ratio_stats, 'cache\\k600_theory\\V_ratio_stats.csv')
 
-velocityScalePlot <- ggplot(data, aes(x=flag_depth, y=(((1e-6*eD)^0.25)/ustar), fill=flag_depth)) + #kinematic viscosity is assumed 1e-6 m2/s because no water temperature data is available
+velocityScalePlot <- ggplot(data, aes(x=flag_depth, y=(((1e-6*eS)^0.25)/ustar), fill=flag_depth)) + #kinematic viscosity is assumed 1e-6 m2/s because no water temperature data is available
   geom_boxplot(size=1.2) +
   scale_fill_brewer(palette='Accent', name='')+
   geom_hline(yintercept = 1, linetype='dashed', size=2) +
@@ -206,24 +205,24 @@ velocityScalePlot <- ggplot(data, aes(x=flag_depth, y=(((1e-6*eD)^0.25)/ustar), 
         legend.title = element_text(size=17, face='bold'),
         legend.position = 'none')
 
-#frac(v*eD ^ {1/4}, U*)
-
 #Plot distributions of Rh/H ratios for SWOT-observable and non-SWOT observable rivers
-swotPlot <- ggplot(data, aes(x=width, linetype=flag_depth)) +
-  stat_ecdf(size=2, color='darkorange') +
-  scale_linetype_manual(values=c("dotted", "solid"), labels=c('Rh =/= H', 'Rh = H'), name='River Regime')+
-  geom_vline(xintercept = 50, linetype='dashed', size=2)+
-  geom_vline(xintercept = 100, linetype='dashed', size=2)+
[…]

### 2021-09-10 docs/supp_v4.Rmd
first ms push with hydraulic geometry stuff

--- 2021-07-26 supp_v4.Rmd
@@
 ##Contents
-This supplementary information contains 7 figures and 1 table. Please consult https://github.com/craigbrinkerhoff/RSK600 for all code to build and generate results, figures, and the manuscript.
+This supplementary information contains 4 texts, 5 figures and 2 tables.
+Please consult https://github.com/craigbrinkerhoff/RSK600 for all code to build and generate results, figures, and the manuscript.
 
@@
 They used an extensive database of field measurements and machine learning to identify patterns that associate river width with the hydraulic priors needed to run McFLIs so that prior hyperparameters may be assigned to rivers using only the existing remotely sensed data.
-For this study, we extracted $\lambda_{A_0}$ and $\gamma_{A_0}$ as the **Xth** percentile values rather than the absolute maximum and minimum values to avoid physically impossible bounds on $A_0$.
+For this study, we extracted $\lambda_{A_0}$ and $\gamma_{A_0}$ as the 5th and 95th percentile values rather than the absolute maximum and minimum values to avoid physically impossible bounds on $A_0$.
 
@@
 
+![Figure S5 All timeseries of k per river](cache/validation/timeseries_noerr.jpg)
+
 ##Tables

### 2021-09-10 manuscript_v5.Rmd
first ms push with hydraulic geometry stuff

--- 2021-07-28 manuscript_v4.Rmd
@@
 ---
-title: "Gas exchange in large rivers controlled by largest turbulent eddies: implications for remotely sensing gas exchange via SWOT"
-author: 'CB Brinkerhoff, CJ Gleason, PA Raymond, MH Harlan'
+title: "Gas exchange in large rivers controlled by hydraulic geometry: implications for remotely sensing gas exchange via the SWOT satellite"
+author: 'CB Brinkerhoff, CJ Gleason, CJ Zappa, PA Raymond, MH Harlan'
 date: "`r Sys.Date()`"
@@
 ##Key Points
-- In large rivers, there exists a local equilibrium between the production and dissipation of turbulent kinetic energy at the free surface
-- Gas exchange in large rivers is equivalently parameterized across multiple turbulence scales, with the classical Kolmogorov model recovered empirically
+- River channel efficiency effec the relationship between gas exchange and form-drag dissipation of turbulence
+- Large, less efficient channel geometry necessitates that near-surface turbulence is dominated by form-drag rather than bed friction
 - BIKER algorithm exploits this theory, predicting gas exchange velocity and $CO_2$ fluxes from simulated SWOT satellite data and in situ $CO_2$ data
@@
 ##Keywords
-gas exchange, open-channel flow, remote sensing, SWOT, biogeochemistry, ungauged basin
+gas exchange, fluvial geomorphology, open-channel flow, remote sensing, SWOT, biogeochemistry, ungauged basin
 
@@
 Equation 1 represents this flux given the gas concentration gradient between the water $CO_{2_{water}}$ and the air $CO_{2_{air}}$ and the gas exchange velocity *k*.
+Consult Appendix A for variable nomenclature used in this study.
 
@@
 Decades of work have focused on elucidating the physical mechanisms behind *k* in oceans and lakes [@wanninkhofAdvancesQuantifyingAirSea2009] and, to a lesser extent, fluvial gas exchange [@hallGasExchangeStreams2020].
-Given that *k* should scale with turbulence in a turbulent flow [@hallGasExchangeStreams2020], a generalized model for *k* (equation 2) relies on the characteristic turbulence velocity scale $v_K$, which itself is a function of the characteristic length and time scales [@katulStructureFunctionModel2018].
-Extensive field and labratory experiments have confirmed the validity of this model when $v_K$ is set to the Kolmogorov velocity scale [e.g.
-@lamontEddyCellModel1970; @zappaEnvironmentalTurbulentMixing2007; @lorkeUnifiedScalingRelation2006; @zappaVariationSurfaceTurbulence2003; @moogStreamReaerationNonuniform1999; @wangCoefficientsSmallEddy2015; @vachonRelationshipNearsurfaceTurbulence2010], where the idea is that turbulent kinetic energy (TKE) is produced by the largest, integral-scale turbulent eddies in the flow and then passed down the energy cascade to progressively smaller eddies until eddy motion is stable and TKE is dissipated as heat at the Kolmogorov-scale: the smallest eddies in a turbulent flow.
-Equation 2 via Kolmogorov-scale eddies can also be derived from both classic thin-film and surface renewal theories for gas exchange [@wangPhysicallyBasedScaling2021].
-$$\mathbf{(2)} k\sim Sc^{-n}v_K$$
+Given that *k* should scale with near-surface turbulence in a turbulent flow [@hallGasExchangeStreams2020], extensive field and labratory experiments have converged on the 'small-eddy model' as proposed by @lamontEddyCellModel1970, which scales *k* via the smallest-scale turbulent eddies [e.g.
+@zappaEnvironmentalTurbulentMixing2007; @lorkeUnifiedScalingRelation2006; @zappaVariationSurfaceTurbulence2003; @moogStreamReaerationNonuniform1999; @wangCoefficientsSmallEddy2015; @vachonRelationshipNearsurfaceTurbulence2010].
+This model is provided as equation 2, where $\epsilon_S$ is the dissipation rate of near-surface turbulence and $v$ is the kinematic viscosity and *Sc* is the Schmidt number.
+$$\mathbf{(2)} k\sim Sc^{-n}(v\epsilon_S)^{1/4}$$
 
-While this model works reasonably well in non-fluvial environments, there is considerable uncertainity in how it applies to fluvial systems.
-Specifically, the influence of the riverbed and its large roughness elements in small streams (i.e.
-bedforms, logjams, sediment out of suspension) complicates scaling fluvial *k* and might lead to substantially higher *k* in streams with whitewater [e.g.
-@hallAirWaterOxygen2012; @ulsethDistinctAirWater2019a].
-Less attention has been paid to the other side of the stream-to-river continumn, which is often conceptualized in relation to *k* as a hybrid condition of both fluvial and non-fluvial hydraulic properties.
-To date, the handful of existing field studies of large-river *k* have suggested that *k* begins to be notably influenced by wind dynamics, though little else is well-established [@alinPhysicalControlsCarbon2011; @beaulieuControlsGasTransfer2012a; @wangPhysicallyBasedScaling2021].
+While this model works reasonably well in non-fluvial environments where turbulence production is largely confined to bottom shear or wind shear [e.g.
+@lorkeUnifiedScalingRelation2006; @zappaEnvironmentalTurbulentMixing2007], there is considerable uncertainity in how it applies to fluvial systems.
+First, 'bubble-mediated gas exchange' in whitewater might lead to substantially higher *k* in small mountain streams with very large bedload [@hallAirWaterOxygen2012; @ulsethDistinctAirWater2019a].
+Additionally, 'form-drag shear' via river channel banks, meanders, bars, etc.
+further complicate scaling fluvial *k* [@moogStreamReaerationNonuniform1999].
+Those authors proposed modeling fluvial *k* via 'form-drag dissipation' $\epsilon_D$, assuming that total reach stream power is responsible for all turbulence production and therefore dissipation.
+Numerous authors have since shown that this model ($\epsilon_D=gS\overline{U}$) reasonably predicts *k* in rivers [@raymondScalingGasTransfer2012a; @ulsethDistinctAirWater2019a; @moogStreamReaerationNonuniform1999], though it is not synonmous with equation 2 as $\epsilon_D$ refers to a depth-scale disispation rate and is not necessairily indicative of the near-surface dissipation rate.
+Likewise, $k\sim\epsilon_D$ usually has a slope much steeper than 1/4.
+
+Relatively less attention has been paid to gas exchange in very large rivers, which is often conceptualized as a hybrid condition of both fluvial and non-fluvial hydraulic properties.
+To date, the handful of existing field studies of large-river *k* have suggested that *k* begins to be influenced by wind shear, though little else is well-established [@alinPhysicalControlsCarbon2011; @beaulieuControlsGasTransfer2012a; @wangPhysicallyBasedScaling2021].
 These mechanistic uncertainties are additionally limited by a large dearth of field-measured fluvial *k*.
-@wangPhysicallyBasedScaling2021 attempted to address this by simulating *k* in 35 rivers using stream metabolism modeling [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets, finding that equation 2 is valid in their simulated rivers and that *k*~streamflow relationships breakdown in large rivers.
+@wangPhysicallyBasedScaling2021 attempted to address this by simulating *k* in 35 rivers using stream metabolism modeling [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets, finding that equation 2 + $\epsilon$ modeled via the log-law-of-the-wall is valid in their simulated rivers and that *k*~streamflow relationships breakdown in large rivers.
 However, they did not directly compare against *k* data in extremely small systems [@ulsethDistinctAirWater2019a] and stopped short of parsing out hydraulic explanations for why this happens.
@@
 
-In this context, here we revist the fundamental mechanisms behind gas exchange and turbulence in large rivers using one of the largest available datasets of field-measured river hydraulics [@brinkerhoffReconcilingAtaStationAtManyStations2019] to answer the following question: does *k* behave fundamentally differently in large rivers than in other streams and how does this relate to the classical model of gas evasion in aquatic systems (equation 2)?
+In this context, here we revist the fundamental mechanisms behind gas exchange and turbulence in large rivers to answer the following question: does *k*'s relationship with form-drag fundamentally change in large rivers and how does this relate to one of the classical model's of gas evasion in aquatic systems (equation 2)?
 We exploit the findings from this simple analysis to develop a novel methodology that predicts $k_{600}$ (*k* normalized to a Schmidt number of 600) and its explicit uncertainity solely using SWOT observations.
@@
 
-TKE and gas exchange theory (section 2) are explored using two datasets.
-First, we use a previously published compilation of field measurements that were originally made to calibrate United States Geological Survey (USGS) streamgauge rating curves and gathered by @brinkerhoffReconcilingAtaStationAtManyStations2019.
+Gas exchange theory (section 2) is explored using the dataset from @ulsethDistinctAirWater2019a who measured and/or collected from the literature over 700 measurements of stream hydraulics and $k_{600}$ and is, to our knowledge, the largest collection of field-measured river and stream $k_{600}$.
+To assess general hydraulic conditions in rivers across geographic contexts, we additionally use a previously published compilation of field hydraulics measurements that were originally made to calibrate United States Geological Survey (USGS) streamgauge rating curves [@brinkerhoffReconcilingAtaStationAtManyStations2019].
 That dataset contains over 530,000 unique measurements of river channel velocity, width, depth, and discharge from across the continental United States.
 This is to our knowledge the largest collection of field measurements of river channel hydraulics.
-Second, we use the dataset from @ulsethDistinctAirWater2019a who measured and/or collected from the literature over 700 measurements of stream hydraulics and $k_{600}$ and is, to our knowledge, the largest collection of field-measured river and stream $k_{600}$.
-This is also the dataset used to validate our *k* scaling models.
+Ultimately, this dataset was filtered down to 171,553 discrete measurements (see Text S1 for our filtering protocol).
 
@@
 
-##3 The TKE budget in large rivers and implications for scaling fluvial gas exchange velocity
+##3 Form-drag in large rivers and implications for scaling fluvial gas exchange velocity
 Turbulence is fundamental to fluvial geomorphology and open-channel flows and has been the subject of extensive research across scientific and engineering disciplines for over a century.
-Turbulent kinetic energy (TKE) is the energy generated and dissipated by the eddies created from the chaotic nature of turbulent flow.
-@nakagawaTurbulenceOpenChannel1993 outline the standard energy budget for TKE, at some depth *h*, for a 2-dimensional open-channel flow (equation 3).
-*G* is the TKE produced, $\epsilon$ is the TKE dissipated, $T_d$ is TKE transported from the riverbed via turbulent diffusion, and $V_d$ is TKE transported via viscous diffusion.
-$$\mathbf{(3)} G+T_d+V_d=\epsilon$$
+Turbulence in river flows is usually generated by shear forces within the river channel (wind shear's influence is usually limited to only the absolute widest rivers).
+This within-bank shear comes from the riverbed, as well as 'form-drag' via riverbanks, meanders, bars, suspended sediment, etc.
+Both @raymondScalingGasTransfer2012a and @moogStreamReaerationNonuniform1999 confirmed that $\epsilon_D$ is far greater than $\epsilon_S$ in open-channel flows, therefore generating the majority of fluvial TKE dissipation.
 
-$\epsilon$ at the free surface is frequently described as $\epsilon(H)=\frac{U_*^3}{H}$ where *H* is the flow depth.
-However in rivers, which exhibit additional depth-scale shear due to downslope flow against the channel geometry, bars, and meanders, energy dissipation occurs much more evenly over the flow depth than this parameterization allows for.
-It has been shown that, for rivers at least, a more appropriate dissipation model considers stream power (per unit weight water: $\epsilon_D = gSU$), which represents the energy dissipated by the total frictional resistance of the river reach's geometry [@moogStreamReaerationNonuniform1999].
-This is often called 'energy dissipation via form-drag' in the literature, and is generally used to upscale *k* in rivers and streams [i.e.
-@raymondScalingGasTransfer2012a; @ulsethDistinctAirWater2019a].
-Both @raymondScalingGasTransfer2012a and @moogStreamReaerationNonuniform1999 confirmed that $\epsilon_D$ is far greater than $\epsilon$ in open-channel flows, therefore generating the majority of fluvial TKE dissipation.
+'Form-drag' shear exists at the depth-scale and therefore is scale-dependent: while small roughness elements near the bed produce form-drag, if they are small relative to the total flow depth they will exert little influence on near-surface dissipation [@moogStreamReaerationNonuniform1999].
+This means that, if all other properties of a river flow are held constant but the channel is progressively made wider and shallower, the depth-scale will decrease and form-drag will compose a progressively larger component of near surface dissipation.
+This concept is illustrated graphically in Figure 2 and represents the geomorphic concept of 'river channel efficiency': all else being constant, a less efficient channel (wide/shallow) will have a longer wetted perimeter and therefore a larger shear surface area than the same flow through a narrow/deep channel [@chowOpenChannelHydraulics1959a].
+This means that the inefficient channel will 1) exhibit more form-drag and thus turbulence, and 2) turbulence generated via form-drag will more readily reach the surface.
+Note that to compare wetted perimeters across many rivers with different channel areas, we use the hydraulic radius $R_h$ (channel area divided by the wetted perimeter) as compared to the mean flow depth $H$.
+An efficient channel will have a low $\frac{R_h}{H}$ ratio while an inefficient channel will have a high $\frac{R_h}{H}$ ratio that approaches 1.
 
-We express *G* at the free surface in two different ways (equation 4) following @nakagawaTurbulenceOpenChannel1993 on the left and our derivation on the right.
-Text S1 details the necessary algebra to arrive at both expressions of *G*.
-Looking at equation 4, it is readily apparent that $\frac{R_h}{H}$ will cancel out if they are equal.
-This occurs in most large and wide rivers and is a common assumption in hydraulic modeling.
-In this scenario, the following three statements must be true: 1) $G \approx gSU \approx \epsilon_D \approx \epsilon$, 2) $\epsilon(H) \neq \frac{U_*^3}{H}$, and 3) $T_d \approx 0$.
-More intuitively, if this "large river" condition is met then additonal TKE transported from the riverbed is functionally zero and the TKE budget at the free surface is just stream power (per unit weight water).
-This implies a local equilibrium between TKE produced and TKE dissipated at the free surface.
-$$\mathbf{(4)} G=\frac{UU_*^2}{H}=gSU\frac{R_h}{H} $$
+![Figure 2: Conceptual model relating river channel efficiency to the turbulent dissipation rate.
+All else being constant, a less efficient channel shape (wide/shallow) will 1) have more form-drag and thus turbulence and 2) turbulence generated via form-drag will more readily reach the surface.](cache/k600_theory/conceptual_model.jpg)
 
-We test this hypothesis using the data from @brinkerhoffReconcilingAtaStationAtManyStations2019 and @ulsethDistinctAirWater2019a by plotting boxplots of $\frac{R_h}{H}$ ratios for over 531,000 different measurements of river hydraulics versus $\epsilon_D - G$, which is equivalent to $T_d$ in equation 3 assuming that $V_d$ is neglibile away from the wall [@nakagawaTurbulenceOpenChannel1993].
-*G* is calculated as $\frac{UU_*^2}{H}$.
-In order to determine when $R_h=H$, we round all ratios to 3 significant figures and set all ratios greater than or equal to 0.995 as $R_h=H$.
-If our hypothesis holds, then $\epsilon_D - G$ should be functionally zero only when the 'large river' condition is met.
+This all suggests that, for incredibly inefficient river channels, $\epsilon_S \approx \epsilon_D$.
+Put another way, the channel produces so much turbulence from form-drag, and the flow is so shallow relative to its width, that depth-scale dissipation is likely synonmous with near-surface dissipation.
+This would mean that equation 2 could be equivalently written as $k\sim Sc^{-n}(v\epsilon_D)^{1/4}$ when a channel is sufficiently inefficient.
+To explore when, if at all, this point is reached we fit models using 701 river $k_{600}$ measurements from @ulsethDistinctAirWater2019a.
+We defined 'channel inefficiency' using minimum $\frac{R_h}{H}$ thresholds from 0.75 to 0.995 and then we fit the two models outlined in equations 3 and 4.
+Because all of the data are normalized to Schmidt numbers of 600, we can convienantly fold the viscosity and Schmidt numbers into the model coefficients.
+The goal of this analysis is to find at what $\frac{R_h}{H}$ threshold does 1) equation 3's b parameter approximate 1/4 and 2) equation 4's goodness-of-fit (via the coefficient of determination $r^2$) reach a suitably high value.
+$$\mathbf{(3)} k_{600}=a(\epsilon_D)^{b}$$
+$$\mathbf{(4)} k_{600}=\beta(\epsilon_S)^{1/4}$$
 
-![Figure 2: Rh/H versus eD-G at the free surface for our two datasets of field-measured river channel hydraulics (see Appendix A for variable descriptions).
-eD-G is functionally zero only when a river is sufficiently large.
-This indicates that a local equilibrium of TKE is acheived and energy generated by large-scale eddies is nearly equal to energy dissipated by the smallest scale eddies.
-In this figure, Rh/H bins contain between 573 and 138,307 measurements.](cache/k600_theory/turbulence.jpg)
+The results of this analysis are plotted in Figure 3a, along with channel geometry schematics to highlight the progressive shift from efficient to inefficient river channels.
+Equation 3's b parameter only begins to approximate 1/4 when $R_h$ is within 1% of the mean flow depth.
+At this point, the model's goodness-of-fit also reaches a reasonable value of 0.68.
+This indicates that a river is sufficiently inefficient when its hyradulic radius approximates the mean flow depth, and at that point $\epsilon_S \approx \epsilon_D$.
+Figures 3b-4d plot the model fits for this final scenario where $\frac{R_h}{H} >= 0.995$ As Figure 3c confirms, the small-eddy model is viable when a channel is so inefficient that dissipation via form-drag approximates the near-surface dissipation rate in equation 2.
+Figure 4d further confirms this, where the slope of the generalized model (equation 3) equals `r round(models_ulseth[models_ulseth$name == 'Rh=H',]$slope, 2)` which is nearly equivalent to 1/4.
 
-Figure 2, which plots this test, shows that $\epsilon_D-G$ is nearly two orders of magnitude less when $R_h=H$ than it is when $\frac{R_h}{H}$ is between 90-99% (median value across both datasets of $10^{-5.39}$ versus $10^{-3.83} \frac{J}{kg*s}$, respectively).
-When the $\frac{R_h}{H}$ ratio is very small (0-9%), this rises to a median value of $10^{-2.56} \frac{J}{kg*s}$, or nearly three orders of magnitude larger.
-The $\epsilon_D - G$ values when $R_h=H$ (which are generally $10^{-5}$ to $10^{-8} \frac{J}{kg*s}$) are functionally zero.
-This means that a local equilibrium of free-surface TKE is achieved, where energy produced via the largest scale (turbulent) eddies is equivalent to energy dissipated via the smallest scale (Kolmorogov) eddies.
-This idea of a local equilibrium is certainly not new, however this is to our knowledge the first confirmation that it happens naturally in most large rivers and that is arises algebraically from the TKE budget in open-channel flows.
-From here on we refer to the $R_h=H$ river scenario as the 'large river' condition.
+![Figure 3: Model development for predicitng k from form-drag dissipation in rivers.
+(a) Rh/H thresholds versus fitted model results.
+Dark blue is the fitted b parameter in equation 3 and dark green is the goodness-of-fit of equation 4 as defined by the coefficient of determination.
+Once a channel is maximally inefficient, i.e.
+Rh=H, equation 4 is a viable model for k and equation 3 empirically recovers the theoretical value of 1/4 as outlined in equation 2.
+(b-d) Fitted model results for inefficient channels (defined as Rh/H >= 0.995) predicting k via the shear velocity U*, equation 4, and equation 3 (respectively).
+Note that colors align to the same models for subpanels a, c, and d.
+Further note that the coefficient of determination in subpanel (d) refers to log-transformed residuals and hence is not directly comparable to those in subpanels (b) and (c).](cache/k600_theory/k600Plot.jpg)
 
-This finding has a significant impact on how we model *k* in large rivers.
-Because the local equilibrium acheived during the 'large river condition' necessitates that $\epsilon$ is equivalent to *G* (Figure 2), we can presumably set $v_K$ in equation 2 equal to the largest ('integral-scale') eddies in the system, which are much more easily remotely-sensible than the Kolmogorov velocity scale.
-Otherwise, *k* should continue to scale only using the Kolmogorov-scale (which is consistent across all turbulent flows by its definition).
+Finally, becuase $\epsilon_D$ refers to a depth-scale quantity, we tested whether a more parsimonious model is possible by removing the 'middle-man', i.e.
+the Kolmogorov velocity scale, and directly predicting *k* from depth-scale turbulence.
+Figure 3b plots fitted model results for the $k=\beta_1U_*$ model.
+For this data, $U_*$ explains 70% of variation in $k_{600}$ and, as expected, the fitted model is nearly identical to Figure 4c in scaling dynamics (as confirmed visually and via the goodness-of-fit).
+This all indicates that depth-scale turbulent production/dissipation dominates the surface dissipation signal in maximally inefficient river channels.
+These findings build off of @ulsethDistinctAirWater2019a's intial result that channel slope exerts a drammatic influence on *k* scaling dynamics, further suggesting that fluvial *k* is as much a function of hydraulic geometry as the turbulence properties of the flow.
 
-To test this hypothesis, we fit the following three models to the @ulsethDistinctAirWater2019a dataset of *k* measurements normalized to a Schmidt number of 600: $k_{600}=\beta U_*$, $k_{600}=\alpha \epsilon^{1/4}$, and $k_{600}=a (\epsilon_D)^b$ under two scenarios: $R_h=H$ and $R_h =/= H$.
-Note that because all data are normalized to a Schmidt number of 600, both the Schmidt numbers and viscosity terms are constant and are implicit in the statistical coefficients.
-The model structures (and the fit parameters) are detailed in Table 1.
-The third model tested (rows 5 and 6 of Table 1), was fit to see which slope coefficient is retrieved empirically by fitting the generalized scaling function to the data.
-Therfore, the $r2$ for these models is reflective of the log-transformed $k_{600}$ (unlike the other two models).
-This was the approach used by @ulsethDistinctAirWater2019a, who observed slopes steeper than 1/4 in both regimes of their breakpointed regression model.
+With all of that said, how often is a river channel so inefficient that $R_h \approx H$?
+To answer this question, we join the 701 measurements from @ulsethDistinctAirWater2019a to 171,553 discrete river hydraulics measurements from across the continetal United States [@brinkerhoffReconcilingAtaStationAtManyStations2019].
+We find that `r round(percs_overall[percs_overall$flag_depth == 'Rh=H',]$perc,0)`% of hydraulic radii are within 5% of their mean flow depth.
+Because these measurements were made exclusively at streamgauges where rivers are somewhat channelized and width variability is decreased, we opted for a slightly more liberal ratio of 5%.
+Because larger rivers are more likely to have inefficient channels, we then flag the measurements that are SWOT-observable, i.e.
+width > 100m.
+We find that while `r round(percs[percs$flag_depth == 'Rh=H',]$perc,0)[1]`% of non-SWOT observable rivers meet this inefficient condition, `r round(percs[percs$flag_depth == 'Rh=H',]$perc,0)[2]`% of SWOT-observable rivers do.
+This significant difference, and near universality of SWOT-observable rivers being inefficient, suggests that our theoretical findings in Figure 3 should hold in nearly all rivers that SWOT will observe.
+This opens the door for exploiting our 'inefficient river k model' for remote sensing *k* via the SWOT satellite.
 
-![Figure 3: Six k600 predictive models tested on the dataset from Ulseth et al.
-(2019).
-a-c are fit to the 'large river condition' while d-f are not.
-Red coefficients of determination refer to log-tranformed quantities while blue coefficients of determination refer to the natural quantities.](cache/k600_theory/k600Plot.jpg)
-
-The results of this test are presented as Figure 3 and Table 1, where our hypothesis is confirmed: both the $U_*$ and $\epsilon^{1/4}$ models explain ~70% of variance in *k* under the 'large river' condition while the $U_*$ model does not work when this condition is not met ($r^2$:`r round(models_ustar[2,]$r2,2)` versus $r^2$=`r round(models_ustar[1,]$r2,2)`).
-In fact, Figures 3a and 3b are nearly visually identical in their scaling dynamics, while Figures 3d and 3e are very different, lending further evidence that these two scaling models are synonmous only when the 'large river' condition is met.
-This is in line with @wangPhysicallyBasedScaling2021, who found similar predictive performance and parameter values for these two models on 35 simulated riverswhich they assumed were large enough that $R_h \approx H$.
-Further, the 1/4 slope parameter is statistically recovered in the 'large river' condition (row 3, Table 1: slope = `r round(models_ulseth[2,]$slope,2)`).
-We note that the large residuals in Figure 3c are a byproduct of the model existing in log-space and that panels 3b and 3c are functionally identical models (see Table 1 for their nearly identical intercept and slope parameters).
-The top row of Figure 3 confirms that under the 'large river condition' *k* can be scaled using either integral or Kolmogorov turbulence scales and that they are synonymous.
-This opens the door for remotely sensing *k* using almost entirely remotely-sensed parameters.
-This is elaborated on in section 4.
-
-*Table 1: Model performance and parameter values for the six models tested*
-
-|**Figure 3 Panel**|**River Regime**|**Model Structure**|**r2**|**Intercept Parameter**|**Slope Parameter**|
-|---------------------|------------------|-------------------|------|---------------|---------------|------------------|
-|3a|$R_h=H$|$k_{600}=\beta U_*$|`r round(models_ustar[2,]$r2, 2)`|`r signif(models_ustar[2,]$int, 3)`|`r round(models_ustar[2,]$slope, 2)`|
-|3d|$R_h=/=H$|$k_{600}=\beta U_*$|`r round(models_ustar[1,]$r2, 2)`|`r signif(models_ustar[1,]$int, 3)`|`r round(models_ustar[1,]$slope, 2)`|
-|3b|$R_h=H$|$k_{600}=\alpha(\epsilon_D)^{1/4}$|`r round(models_eD4[2,]$r2, 2)`|`r signif(models_eD4[2,]$int, 3)`|`r round(models_eD4[2,]$slope, 2)`|
-|3e|$R_h=/=H$|$k_{600}=\alpha(\epsilon_D)^{1/4}$|`r round(models_eD4[1,]$r2, 2)`|`r signif(models_eD4[1,]$int, 3)`|`r round(models_eD4[1,]$slope, 2)`|
-|3c|$R_h=H$|$k_{600}=a(\epsilon_D)^{b}$|`r round(models_ulseth[2,]$r2, 2)`|`r signif(models_ulseth[2,]$int, 3)`|`r round(models_ulseth[2,]$slope, 2)`|
-|3f|$R_h=/=H$|$k_{600}=a(\epsilon_D)^{b}$|`r round(models_ulseth[1,]$r2, 2)`|`r signif(models_ulseth[1,]$int, 3)`|`r round(models_ulseth[1,]$slope, 2)`|
-
-The 1/4 exponent is clearly not valid when $R_h=/=H$ (Figure 3e).
-However, the generalized form of this model explains `r 100*round(models_ulseth[1,]$r2,2)`% of the variation in log-transformed *k* and was deemed a strong scaling model for fluvial *k* when rivers do not meet the 'large river' condition.
-That model yields a slope of `r round(models_ulseth[1,]$slope,2)` which corroborates other models with slopes steeper than 1/4 in rivers [@wangPhysicallyBasedScaling2021; @ulsethDistinctAirWater2019a; @raymondScalingGasTransfer2012a].
-Further theoretical modeling is needed to parse out why fluvial *k* in systems that do not meet the 'large river condition' generally scale at a steeper rate than in non-fluvial systems, and how this relates to the breakpointed model proposed by @ulsethDistinctAirWater2019a.
-That said, the TKE budget in equation 3 suggests that it is a function of turbulent diffusion of additional TKE from the bed to the free surface.
-
-##4 Exploiting the 'large river' model to remotely sense gas exchange velocity via the BIKER algorithm
-We have shown that scaling $k_{600}$ by either integral or Kolmogorov turbulence scales explains ~70% of the variation in $k_{600}$ in large rivers and that their scaling dynamics are virtually identical.
-The integral-scale model relies solely on the shear velocity $U_*$.
-Because of the structure of $U_*$ (Appendix A), we can easily reduce the 'large river' $k_{600}$ model (Figure 3a, row 1 Table 1) to an equation consisting solely of remotely-sensible river hydraulics and an intial estimate of the channel area.
+##4 Exploiting the 'inefficient river model' to remotely sense gas exchange velocity via the BIKER algorithm
+We have shown that scaling $k_{600}$ via the depth-scale shear-velocity ($U_*$) explains ~70% of the variation in $k_{600}$ in inefficient rivers and that nearly all SWOT-observable rivers are inefficient.
+Further, because of the structure of $U_*$ (Appendix A) we can easily reduce the 'inefficient river' $k_{600}$ model (Figure 3b) to an equation consisting solely of remotely-sensible river hydraulics and an intial estimate of the channel area.
 By scaling *k* this way, we significantly reduce potential equifinality issues by having just two unknowns: $k_{600}$ and the median channel area $A_0$.
@@
 In both of these domains, the other unknown parameters are often difficult to estimate (bed roughness and stream metabolism, respectively) while for BIKER, median channel area is relatively easy to approximate from the SWOT-observable river width [@brinkerhoffConstrainingRemoteRiver2020a].
-
-That said, this all assumes that the 'large river' condition is met.
-Because SWOT will be limited to the widest rivers on Earth (minimum average width of 100m), the $R_h=H$ assumption is nearly always valid.
-First, we verified that this is true using both the @brinkerhoffReconcilingAtaStationAtManyStations2019 and @ulsethDistinctAirWater2019a datasets (Figure S3), finding that the vast majority of rivers with a width of at least 100m exhibit this $R_h=H$ behavior.
-Thus, our theoretical findings in sections 2 and 3 should hold in nearly all rivers that SWOT will observe and we should be able to remotely estimate $k_{600}$ via SWOT.
 Therefore, we develop the BIKER algorithm, which is informed by the @hagemannBAMBayesianAMHGManning2017a algorithm for ungauged RSQ and further explored in more recent work by @brinkerhoffConstrainingRemoteRiver2020a.
@@
 To do this, $k_{600}$ is first written as a function of SWOT-observables *W* and $S_h$.
-This algebra is carried out using the model parameter from row 1 of Table 1 and yields equation 6, where we follow our earlier assumption that $R_h = H =\frac{A}{W}$ where *A* is wetted channel area.
+This algebra is carried out using the model parameter from Figure 3b (56.0294) and yields equation 6, where we follow our earlier assumption that $R_h = H =\frac{A}{W}$ where *A* is wetted channel area.
 *A* is further split into the SWOT-observable portion *dA* and the unobservable portion $A_0$ following @durandEstimatingReachaveragedDischarge2014a.
@@
 We validate BIKER on 47 SWOT-simulated rivers using daily observed $R_h$ and observed $S_h$.
-We also re-validate BIKER on the 17 rivers with the SWOT error model corrupting the SWOT-observables *W* and *H* (this workflow and the metrics used are detailed in Text S4 and Table S1).
+We also re-validate BIKER on the 17 rivers with the SWOT error model corrupting the SWOT-observables *W* and *H* (this workflow and the metrics used are detailed in Text S3 and Table S1).
 Regardless of the validation setup, we do not have observed $k_{600}$ data for these rivers, and to our knowledge no field dataset of $k_{600}$ exists in the type of temporal and spatial frequency that SWOT (and therefore the BIKER algorithm) provides.
-Therefore, we take the model outlined in row 1 of Table 1 and use that to calculate the observed $k_{600}$ that BIKER is validated against.
-With this setup, we are directly exploring BIKER's ability to infer observed $A_0$ and $k_{600}$ from *W* and *H* alone, as the *k* scaling model has already been sucessfully validated (Table 1, Figure 3).
+Therefore, we take the model outlined in Figure 3b and use that to calculate the observed $k_{600}$ that BIKER is validated against.
+With this setup, we are directly exploring BIKER's ability to infer observed $A_0$ and $k_{600}$ from *W* and *H* alone, as the *k* scaling model has already been sucessfully validated (Figure 3).
 
@@
 b) was randomly selected from the upper tertile of NRMSE scores, c) was randomly selected from the middle tertile, and d) from the worst tertile.
-See Figure S5 for all other rivers.
+See Figure S3 for all other rivers.
 Model results include the posterior means and 95%.](cache/validation/validation.jpg)
@@
 Figure 4b-d are representative timeseries plots of predicted and observed $k_{600}$ for three rivers chosen randomly from those with 'good' NRMSE scores (b), 'okay' NRMSE scores (c), and 'bad' NRMSE scores (d).
-See Table S1 for the definition of NRMSE, the Figure 4 caption for how this was determined, and Figure S5 for the other river timeseries plots.
+See Table S1 for the definition of NRMSE, the Figure 4 caption for how this was determined, and Figure S3 for the other river timeseries plots.
 For the Missouri Midsection River, the entire timeseries of $k_{600}$ is nearly perfectly predicted, while in the Seine River the dynamics (peaks and valleys) are reasonably captured but they are magnified to be far larger than the observed dynamics.
@@
 a: Performance metrics by river.
-See Table 1 for metric definitions.
+See Table S1 for metric definitions.
 b: NRMSE scores (per river) versus the number of SWOT observations used in the Bayesian inversion.
@@
 
-We also statistically recovered the 1/4 scaling exponent in the classic $k \sim \epsilon^{1/4}$ model (Figure 3c, row 5 Table 1) for gas exchange that has been repeadely validated in non-fluvial aquatic systems.
+We also statistically recovered the 1/4 scaling exponent in the classic $k \sim \epsilon^{1/4}$ model (Figure 3d) for gas exchange that has been repeadely validated in non-fluvial aquatic systems.
 This again suggests that *k* in large rivers behaves more similarly to lakes, esturaries, and the ocean than to steeper rivers.
@@
 However, this is only true in the largest of rivers.
-It is likely that the steeper slope parameter in Figure 3d and row six, Table 1 is due to the influence of bottom roughness on free-surface TKE in smaller streams.
 

### 2021-09-10 supp_v5.Rmd
first ms push with hydraulic geometry stuff

--- 2021-09-10 docs/supp_v4.Rmd
@@
 ---
-title: "Supplemental Information: Gas exchange in large rivers controlled by largest turbulent eddies: implications for remotely sensing gas exchange via SWOT"
-author: "CB Brinkerhoff, CJ Gleason, PA Raymond, MH Harlan"
+title: "Supplemental Information: Gas exchange in large rivers controlled by hydraulic geometry: implications for remotely sensing gas exchange via the SWOT satellite"
+author: "CB Brinkerhoff, CJ Gleason, CJ Zappa, PA Raymond, MH Harlan"
 date: "`r Sys.Date()`"
@@
 ##Contents
-This supplementary information contains 4 texts, 5 figures and 2 tables.
+This supplementary information contains 4 texts, 4 figures and 2 tables.
 Please consult https://github.com/craigbrinkerhoff/RSK600 for all code to build and generate results, figures, and the manuscript.
 
-## Text S1: Deriving two expressions for turbulent kinetic energy production
-Below we detail how we arrived at our two expressions for a cross-section-averaged turbulent kinetic energy production (*G*), or equation 3 in the main text.
-Please see Appendix A for variable descriptions and notation.
+## Test S1: Filtering procedure for hydraulics data
+To quantify how frequently SWOT-observable rivers are also inefficient river channels (section 2 of main text), we used the dataset of field-measured river hydraulics in @brinkerhoffReconcilingAtaStationAtManyStations2019.
+That dataset has over 500,000 discrete measurements of river width, velocity, area, and discharge that were made by the United States Geological Survey (USGS) to calibrate streamgauge rating curves.
+Here, we descirbe how this dataset was filtered down to 171,553 measurements to meet the assumptions in our conceptual model.
 
-@nakagawaTurbulenceOpenChannel1993 outline the standard energy budget for turbulent kinetic energy (TKE), at some distance from the bed *h*, for a 2-dimensional open-channel flow (equation S1).
-*G* is the TKE produced, $\epsilon$ is the TKE dissipated, $T_d$ is TKE transported from the riverbed via turbulent diffusion, and $V_d$ is TKE transported via viscous diffusion.
-$T_d$ should always be positive at the free surface, but this is increasingly less true as one moves closer to the riverbed.
-Since we are interested in the TKE budget at the river surface, $V_d$ can be assumed neglible [@nakagawaTurbulenceOpenChannel1993].
-$$\mathbf{(S1)} G+T_d+V_d=\epsilon$$
-
-@nakagawaTurbulenceOpenChannel1993 further define *G* and the direct energy dissipation *E*, as averaged over the entire river cross-section and integrated over the entire water column, as equation S2 where *H* is the full water-column depth, *U* is the cross-sectional average flow velocity, and $U_*$ is the friction velocity.
-$U_*$ is defined by the water surface slope *S*, gravitational acceleration *g* and the channel hydraulic radius $R_h$ as $U_*=\sqrt{gSR_h}$.
-$$\mathbf{(S2)} \int_{0}^{H}{Gdh} + \int_{0}^{H}{Edh}=UU_*^2$$
-
-$\int_{0}^{H}{Edh}$ can also be assumed neglible at the free surface [@nakagawaTurbulenceOpenChannel1993].
-We then integrate over the full water-column depth *H* to arrive at an expression for *G* at the free surface that reduces to equation S5.
-Equations S3 to S4 show the algebra.
-$$\mathbf{(S3)} \int_{0}^{H}{Gdh} = U\sqrt{gSR_h}^2$$
-$$\mathbf{(S4)} GH = UgSR_h$$
-$$\mathbf{(S5)} G=gSU\frac{R_h}{H}$$
-Alternatively, we can derive another expression for *G* from equation S2. This is provided as equation S8 with the necessary algebra in equations S6-S7.
-$$\mathbf{(S6)} \int_{0}^{H}{Gdh}=UU_*^2$$
-$$\mathbf{(S7)} GH=UU_*^2$$
-$$\mathbf{(S8)} G=\frac{UU_*^2}{H}$$
+First, we removed all measurements missing data, measurements tagged by the USGS as 'poor', measurements with impossible values, or measurements of 0.
+While this would indicate a dry channel, our hydraulic geometry model necessitates within-bank flow.
+Likewise, because hydraulic geometry only applies to within-bank flows and not flood events, we had to remove all overbank flows.
+This was done by first filtering for sites with at least 20 measurements (to build robust estimates of bankfull hydraulics) and then calculating bankfull width and depth as the width or depth with a return period of two years.
+While the only true way to calculate bankfull hydraulics is manually in the field, this was ovbiously impractical.
+A two year return period is a standard approximation for determining out-of-bank flow in single-channel meandering rivers and was the method used by @brinkerhoffReconcilingAtaStationAtManyStations2019.
+We then removed all measurements with a width or depth beyond their respective at-a-station 2 year values.
 
@@
 
-![Figure S3 Cummulative density functions (CDFs) of river width measurements, as binned by their dataset and and whether or not they meet the 'large river' condition (Rh=H).
-The dashed black line indicates the minimum river width that the SWOT satellite will observe (100m).
-This brief analysis indicates that the vast majority of SWOT-observable rivers should meet the 'large river' condition.](cache/k600_theory/swotPlot.jpg)
-
-![Figure S4 Mean river properties (observed k600, width, and slope) for the 47 rivers versus their BIKER performance as defined by NRMSE (see Table S1 for this metric's definition).
-The dashed black line is the median NRMSE score.
-The poorly performing rivers are universally the narrowest and steepest of the set.](cache/validation/riverProperties_nrmse.jpg)
-
-![Figure S5 All timeseries of k per river](cache/validation/timeseries_noerr.jpg)
+![Figure S3 All timeseries of k per river](cache/validation/timeseries_noerr.jpg)
 

### 2021-09-14 manuscript_v5.Rmd
big ms_v5 push

@@
 ---
-title: "Gas exchange in large rivers controlled by hydraulic geometry: implications for remotely sensing gas exchange via the SWOT satellite"
-author: 'CB Brinkerhoff, CJ Gleason, CJ Zappa, PA Raymond, MH Harlan'
-date: "`r Sys.Date()`"
+title: "Gas exchange in large rivers influenced by hydraulic geometry: implications for remotely sensing gas exchange via the SWOT satellite"
+author:
+  - Craig B Brinkerhoff:
+      email: cbrinkerhoff@umass.edu
+      institute: UMass
+      correspondence: true
+  - Colin J Gleason:
+      institute: UMass
+  - Christopher J Zappa:
+      institute: Lamont
+  - Peter A Raymond:
+      institute: Yale
+  - Merritt H Harlan:
+      institute: UMass
+institute:
+  - UMass: Department of Civil & Environmental Engineering, University of Massachusetts
+  - Lamont: Lamont-Doherty Earth Observatory, Columbia University
+  - Yale: School of the Environment, Yale University
 bibliography: References.bib
@@
   word_document: 
+      pandoc_args:
+        - '--lua-filter=scholarly-metadata.lua'
+        - '--lua-filter=author-info-blocks.lua'
       reference_docx: word-styles-reference.docx
@@
 
-##Key Points
-- River channel efficiency effec the relationship between gas exchange and form-drag dissipation of turbulence
-- Large, less efficient channel geometry necessitates that near-surface turbulence is dominated by form-drag rather than bed friction
+## Key Points
+- River channel efficiency effects the relationship between gas exchange and dissipation of turbulence
+- Wide, less "efficient" river channels exert greater influence on near-surface turbulence than narrower, more efficient ones
 - BIKER algorithm exploits this theory, predicting gas exchange velocity and $CO_2$ fluxes from simulated SWOT satellite data and in situ $CO_2$ data
 
-##Keywords
+## Keywords
 gas exchange, fluvial geomorphology, open-channel flow, remote sensing, SWOT, biogeochemistry, ungauged basin
 
-##Abstract
+## Abstract
 *AGU Advances. So 250 words here* <br>
-*AGU advances is 8000 words total, the format seems to be a long letter so I've tried to right this in that style. Might be a bold organization but we'll see*
+*AGU advances is 8000 words total, the format seems to be a long letter so I've tried to right this in that style.*
 
-##Plain Language Summary
+## Plain Language Summary
 *Necessary for AGU advances (200 words)*
 
-##1 Introduction
+## 1 Introduction
 Natural systems play a fundamental role in the budgeting and accounting of the global carbon cycle under climate change.
@@
 The structure of equation 1 necessitates that calculations of this flux are highly sensitive to one's measurements/estimates of *k* [@hallGasExchangeStreams2020].
-Even more broadly, fluvial *k* for various dissolved gases are of significant importance to aquatic ecologists modeling stream metabolism [e.g.
+More broadly, fluvial *k* for various dissolved gases are of significant importance to aquatic ecologists modeling stream metabolism [e.g.
 @bernhardtMetabolicRegimesFlowing2018] and water quality engineers modeling river responses to pollutant loadings [among other things- @chapraSurfaceWaterQualityModeling2008].
-Decades of work have focused on elucidating the physical mechanisms behind *k* in oceans and lakes [@wanninkhofAdvancesQuantifyingAirSea2009] and, to a lesser extent, fluvial gas exchange [@hallGasExchangeStreams2020].
+Because of this far-reaching importance, decades of work have focused on elucidating the physical mechanisms behind *k* in oceans and lakes [@wanninkhofAdvancesQuantifyingAirSea2009] and, to a lesser extent, fluvial gas exchange [@hallGasExchangeStreams2020].
 Given that *k* should scale with near-surface turbulence in a turbulent flow [@hallGasExchangeStreams2020], extensive field and labratory experiments have converged on the 'small-eddy model' as proposed by @lamontEddyCellModel1970, which scales *k* via the smallest-scale turbulent eddies [e.g.
-@zappaEnvironmentalTurbulentMixing2007; @lorkeUnifiedScalingRelation2006; @zappaVariationSurfaceTurbulence2003; @moogStreamReaerationNonuniform1999; @wangCoefficientsSmallEddy2015; @vachonRelationshipNearsurfaceTurbulence2010].
-This model is provided as equation 2, where $\epsilon_S$ is the dissipation rate of near-surface turbulence and $v$ is the kinematic viscosity and *Sc* is the Schmidt number.
+@lorkeUnifiedScalingRelation2006; @zappaEnvironmentalTurbulentMixing2007; @zappaVariationSurfaceTurbulence2003; @moogStreamReaerationNonuniform1999; @wangCoefficientsSmallEddy2015; @vachonRelationshipNearsurfaceTurbulence2010].
+This model is provided as equation 2, where $\epsilon_S$ is the dissipation rate of near-surface turbulence, $v$ is the kinematic viscosity and *Sc* is the Schmidt number.
+Note that we directly measuring $\epsilon_S$ is non-trivial and though a frequent model is to impose the log-law-of-the-wall to scale bottom shear to the surface, this necessiarly assumes that bed shear alone effects turbulence which is often not the case.
 $$\mathbf{(2)} k\sim Sc^{-n}(v\epsilon_S)^{1/4}$$
 
-While this model works reasonably well in non-fluvial environments where turbulence production is largely confined to bottom shear or wind shear [e.g.
+While this model works reasonably well in non-fluvial environments where turbulence production is mostly limited to bottom shear or wind shear [e.g.
 @lorkeUnifiedScalingRelation2006; @zappaEnvironmentalTurbulentMixing2007], there is considerable uncertainity in how it applies to fluvial systems.
@@
 further complicate scaling fluvial *k* [@moogStreamReaerationNonuniform1999].
-Those authors proposed modeling fluvial *k* via 'form-drag dissipation' $\epsilon_D$, assuming that total reach stream power is responsible for all turbulence production and therefore dissipation.
-Numerous authors have since shown that this model ($\epsilon_D=gS\overline{U}$) reasonably predicts *k* in rivers [@raymondScalingGasTransfer2012a; @ulsethDistinctAirWater2019a; @moogStreamReaerationNonuniform1999], though it is not synonmous with equation 2 as $\epsilon_D$ refers to a depth-scale disispation rate and is not necessairily indicative of the near-surface dissipation rate.
-Likewise, $k\sim\epsilon_D$ usually has a slope much steeper than 1/4.
+Those authors proposed modeling fluvial *k* via form-drag dissipation $\epsilon_D$, assuming that total reach stream power is responsible for all turbulence production and therefore dissipation.
+Numerous authors have since shown that this model ($\epsilon_D=gS\overline{U}$) reasonably predicts *k* in rivers [@raymondScalingGasTransfer2012a; @ulsethDistinctAirWater2019a; @moogStreamReaerationNonuniform1999], though it should be stressed that $\epsilon_S \neq \epsilon_D$.
+This is because $\epsilon_D$ refers to a depth-scale disispation rate and is not necessairily indicative of the near-surface dissipation rate.
+Consequently, the, $k\sim\epsilon_D$ relationship usually has a slope much steeper than the 1/4 in equation 2.
 
-Relatively less attention has been paid to gas exchange in very large rivers, which is often conceptualized as a hybrid condition of both fluvial and non-fluvial hydraulic properties.
-To date, the handful of existing field studies of large-river *k* have suggested that *k* begins to be influenced by wind shear, though little else is well-established [@alinPhysicalControlsCarbon2011; @beaulieuControlsGasTransfer2012a; @wangPhysicallyBasedScaling2021].
+Compared to small streams, relatively little attention has been paid to gas exchange in very large rivers.
+Large rivers are often conceptualized as a hybrid condition of both fluvial and non-fluvial *k* dynamics.
+To date, the handful of existing field studies of large-river *k* suggest that *k* begins to be influenced by wind shear once the water surface is less protected, though little else is well-established [@alinPhysicalControlsCarbon2011; @beaulieuControlsGasTransfer2012a; @wangPhysicallyBasedScaling2021].
+This is also presumably limited to forested environments that provide shelter from wind.
+
 These mechanistic uncertainties are additionally limited by a large dearth of field-measured fluvial *k*.
-@wangPhysicallyBasedScaling2021 attempted to address this by simulating *k* in 35 rivers using stream metabolism modeling [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets, finding that equation 2 + $\epsilon$ modeled via the log-law-of-the-wall is valid in their simulated rivers and that *k*~streamflow relationships breakdown in large rivers.
-However, they did not directly compare against *k* data in extremely small systems [@ulsethDistinctAirWater2019a] and stopped short of parsing out hydraulic explanations for why this happens.
+@wangPhysicallyBasedScaling2021 attempted to address this by simulating *k* in 35 rivers using a stream metabolism model [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets.
+They found that equation 2, coupled with $\epsilon$ modeled via the log-law-of-the-wall, is valid in their simulated rivers and that *k*~streamflow relationships break down in large rivers.
+However, they did not directly compare against *k* measurements made in extremely small systems [@ulsethDistinctAirWater2019a] and stopped short of parsing out hydraulic explanations for their results.
 
-Further, these mechanistic uncertainties are then propogated through upscaling workflows when biogeochemists predict *k* across thousands of rivers [e.g @borgesGloballySignificantGreenhousegas2015a; @horgbyUnexpectedLargeEvasion2019; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a] via equation 1 coupled with hydraulic geometry [HG: the scaling relationships between streamflow and river channel hydraulics- @leopoldHydraulicGeometryStream1953a].
+Finally, these mechanistic uncertainties are then propogated through upscaling workflows when biogeochemists predict *k* across thousands of rivers [e.g @borgesGloballySignificantGreenhousegas2015a; @horgbyUnexpectedLargeEvasion2019; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a] via equation 1 coupled with hydraulic geometry [HG: the scaling relationships between streamflow and river channel hydraulics- @leopoldHydraulicGeometryStream1953a].
 It is currently not well understood how sensitive global estimates of fluvial gas evasion are to the specific HG model that is employed by the worker.
@@
 
-In this context, here we revist the fundamental mechanisms behind gas exchange and turbulence in large rivers to answer the following question: does *k*'s relationship with form-drag fundamentally change in large rivers and how does this relate to one of the classical model's of gas evasion in aquatic systems (equation 2)?
+In this context, here we revist the fundamental mechanisms behind gas exchange and turbulence in large rivers to answer the following question: does *k*'s relationship with form-drag fundamentally change along the stream-to-river continumn and how does this relate to one of the classical models of gas evasion in aquatic systems (equation 2)?
 We exploit the findings from this simple analysis to develop a novel methodology that predicts $k_{600}$ (*k* normalized to a Schmidt number of 600) and its explicit uncertainity solely using SWOT observations.
@@
 
-##2 Data
+## 2 Data
 Numerous datasets were used in this study.
@@
 Note that hydraulic model locations are approximate as some of the models are not geo-referenced.
-Not mapped here are over 530,000 discrete measurements of river channel hydraulics from across the continetal United States (Brinkerhoff et al.
+Not mapped here are over 170,000 discrete measurements of river channel hydraulics from across the continetal United States (Brinkerhoff et al.
 2019) and over 700 gas exchange velocity measurements (Ulseth et al.
@@
 
-Gas exchange theory (section 2) is explored using the dataset from @ulsethDistinctAirWater2019a who measured and/or collected from the literature over 700 measurements of stream hydraulics and $k_{600}$ and is, to our knowledge, the largest collection of field-measured river and stream $k_{600}$.
-To assess general hydraulic conditions in rivers across geographic contexts, we additionally use a previously published compilation of field hydraulics measurements that were originally made to calibrate United States Geological Survey (USGS) streamgauge rating curves [@brinkerhoffReconcilingAtaStationAtManyStations2019].
-That dataset contains over 530,000 unique measurements of river channel velocity, width, depth, and discharge from across the continental United States.
+Gas exchange theory (section 3) is explored using the dataset from @ulsethDistinctAirWater2019a who measured and/or collected from the literature over 700 measurements of stream hydraulics and $k_{600}$ and is, to our knowledge, the largest collection of field-measured river and stream $k_{600}$.
+To assess general hydraulic conditions in rivers across geographic contexts, we additionally use a previously published compilation of field hydraulics measurements [@brinkerhoffReconcilingAtaStationAtManyStations2019].
+These were originally made to calibrate United States Geological Survey (USGS) streamgauge rating curves.
+That dataset contains over 530,000 unique measurements of river channel velocity, width, and discharge from across the continental United States.
 This is to our knowledge the largest collection of field measurements of river channel hydraulics.
-Ultimately, this dataset was filtered down to 171,553 discrete measurements (see Text S1 for our filtering protocol).
+Ultimately, this dataset was filtered down to 171,553 discrete measurements for this study (see Text S1 for our filtering protocol).
 
@@
 Ultimately, the 47 rivers are spread across the United States, France, Italy, the United Kingdom, and Bangladesh.
-We omit three models from @durandIntercomparisonRemoteSensing2016b because they lack enough hydraulic information to calculate the shear velocity $U_\star$, which is necessary for algorithm validation (see Appendix A for all variable descriptions and notation used in this study).
+We omit three models from @durandIntercomparisonRemoteSensing2016b because they lack enough hydraulic information to calculate the shear velocity $U_\star$, which is necessary for algorithm validation and is explained in section 4.
 These are two models for the Saint Lawrence River and one for the Tanana River.
@@
 
-For the $CO_2$ evasion and carbon efflux calculations (section 4), we use 26 bi-weekly dissolved $CO_2$ samples made by @beaulieuControlsGasTransfer2012a at one location in the Ohio River for one calender year from 2008-2009 (Figure S2).
+For the $CO_2$ evasion and carbon efflux calculations (section 5), we use 26 bi-weekly dissolved $CO_2$ samples made by @beaulieuControlsGasTransfer2012a at one location in the Ohio River for one calender year from 2008-2009 (Figure S2).
 Note that this $CO_2$ data is for the Ohio River only but was applied to all 47 rivers (which includes multiple sections of the Ohio River).
-Because we are exclusively interested in the relative differences between $FCO_2$ estimates and not the raw fluxes themselves, any $CO_2$ data representative of SWOT-observable rivers was deemed acceptable.
+Because we are exclusively interested in the relative differences between $FCO_2$ estimates and not the raw fluxes themselves, any $CO_2$ data representative of SWOT-observable rivers was deemed acceptable for this validation exercise.
 
-##3 Form-drag in large rivers and implications for scaling fluvial gas exchange velocity
+## 3 Form-drag in large rivers and implications for scaling fluvial gas exchange velocity
 Turbulence is fundamental to fluvial geomorphology and open-channel flows and has been the subject of extensive research across scientific and engineering disciplines for over a century.
-Turbulence in river flows is usually generated by shear forces within the river channel (wind shear's influence is usually limited to only the absolute widest rivers).
-This within-bank shear comes from the riverbed, as well as 'form-drag' via riverbanks, meanders, bars, suspended sediment, etc.
-Both @raymondScalingGasTransfer2012a and @moogStreamReaerationNonuniform1999 confirmed that $\epsilon_D$ is far greater than $\epsilon_S$ in open-channel flows, therefore generating the majority of fluvial TKE dissipation.
+As noted in the introduction, turbulence in river flows is usually generated by shear forces at the riverbed, but also across the entire flow depth via 'form-drag' from riverbanks, meanders, bars, suspended sediment, etc [@moogStreamReaerationNonuniform1999].
+This concept is synonymous with the geomorphic idea of available stream power, or the total energy dissipated from a river reach via the liquid's potential energy along its slope [originally derived from first-principles by @bagnoldSedimentDischargeStream1960; @bagnoldApproachSedimentTransport1966].
+Both @raymondScalingGasTransfer2012a and @moogStreamReaerationNonuniform1999 confirmed that $\epsilon_D$ (simply the available stream power normalized by the water mass) is far greater than $\epsilon_S$ in open-channel flows, therefore generating the majority of fluvial TKE dissipation.
 
-'Form-drag' shear exists at the depth-scale and therefore is scale-dependent: while small roughness elements near the bed produce form-drag, if they are small relative to the total flow depth they will exert little influence on near-surface dissipation [@moogStreamReaerationNonuniform1999].
-This means that, if all other properties of a river flow are held constant but the channel is progressively made wider and shallower, the depth-scale will decrease and form-drag will compose a progressively larger component of near surface dissipation.
-This concept is illustrated graphically in Figure 2 and represents the geomorphic concept of 'river channel efficiency': all else being constant, a less efficient channel (wide/shallow) will have a longer wetted perimeter and therefore a larger shear surface area than the same flow through a narrow/deep channel [@chowOpenChannelHydraulics1959a].
-This means that the inefficient channel will 1) exhibit more form-drag and thus turbulence, and 2) turbulence generated via form-drag will more readily reach the surface.
-Note that to compare wetted perimeters across many rivers with different channel areas, we use the hydraulic radius $R_h$ (channel area divided by the wetted perimeter) as compared to the mean flow depth $H$.
+While form-drag dissipation is effectively an alternative representation of reach-scale stream power, little attention has been paid to the resulting scale-dependecy of $\epsilon_D$.
+Available stream power, and thus 'form-drag' shear, exists at the depth-scale and is scale-dependent: while small roughness elements near the bed produce form-drag, if they are small relative to the total flow depth they will exert little influence on near-surface dissipation [@moogStreamReaerationNonuniform1999].
+This means that, if all other properties of a river flow are held constant but the channel is progressively made wider and shallower, its depth will decrease and its wetted perimeter (and therefore total frictional resistance) will increase.
+This concept is illustrated graphically in Figure 2 and relates the geomorphic concept of 'river channel efficiency' to turbulence: all else being constant, a less efficient channel (wide/shallow) will have greater frictional resistance and therefore form-drag (Chow, 1959).
+As shown via the orange arrows in Figure 2, the inefficient channel should 1) produce more turbulence, and 2) turbulence generated via form-drag should more readily reach the surface (the red arrows in Figure 2).
+Note that to compare wetted perimeters across many rivers with different channel areas, we use the hydraulic radius $R_h$ (channel area divided by the wetted perimeter) as compared to the mean flow depth $H$ (channel area divided by the width).
 An efficient channel will have a low $\frac{R_h}{H}$ ratio while an inefficient channel will have a high $\frac{R_h}{H}$ ratio that approaches 1.
@@
 ![Figure 2: Conceptual model relating river channel efficiency to the turbulent dissipation rate.
-All else being constant, a less efficient channel shape (wide/shallow) will 1) have more form-drag and thus turbulence and 2) turbulence generated via form-drag will more readily reach the surface.](cache/k600_theory/conceptual_model.jpg)
+All else being constant, a less efficient channel shape (wide/shallow) should 1) have more form-drag and thus turbulence and 2) turbulence generated via form-drag should more readily reach the surface.
+This is illustrated by the orange arrows, where the inefficient channel arrows more frequently reach the surface (and influence surface turbulence- red arrows).](cache/k600_theory/conceptual_model.jpg)
 
 This all suggests that, for incredibly inefficient river channels, $\epsilon_S \approx \epsilon_D$.
-Put another way, the channel produces so much turbulence from form-drag, and the flow is so shallow relative to its width, that depth-scale dissipation is likely synonmous with near-surface dissipation.
-This would mean that equation 2 could be equivalently written as $k\sim Sc^{-n}(v\epsilon_D)^{1/4}$ when a channel is sufficiently inefficient.
-To explore when, if at all, this point is reached we fit models using 701 river $k_{600}$ measurements from @ulsethDistinctAirWater2019a.
-We defined 'channel inefficiency' using minimum $\frac{R_h}{H}$ thresholds from 0.75 to 0.995 and then we fit the two models outlined in equations 3 and 4.
-Because all of the data are normalized to Schmidt numbers of 600, we can convienantly fold the viscosity and Schmidt numbers into the model coefficients.
-The goal of this analysis is to find at what $\frac{R_h}{H}$ threshold does 1) equation 3's b parameter approximate 1/4 and 2) equation 4's goodness-of-fit (via the coefficient of determination $r^2$) reach a suitably high value.
+The channel produces so much turbulence from form-drag, and the flow is so shallow relative to its width, that depth-scale dissipation should be nearly synonymous with near-surface dissipation and equation 2 could be equivalently written as $k\sim Sc^{-n}(v\epsilon_D)^{1/4}$.
+To explore when (if at all) this occurs in rivers, we fit models using 701 river $k_{600}$ measurements from @ulsethDistinctAirWater2019a.
+First, we defined 'channel inefficiency' using minimum $\frac{R_h}{H}$ thresholds ranging from 0.75 to 0.995 and then we fit the two models outlined in equations 3 and 4 (because all of the data are normalized to Schmidt numbers of 600, we convienantly fold the viscosity and Schmidt numbers into the model coefficients).
+The goal of this analysis is to find at what $\frac{R_h}{H}$ threshold does 1) equation 3's b parameter approximate 1/4 and thus mirror equation 2 and 2) equation 4's goodness-of-fit (via the coefficient of determination $r^2$) reach a suitably high value that it is deemed physically valid.
 $$\mathbf{(3)} k_{600}=a(\epsilon_D)^{b}$$
-$$\mathbf{(4)} k_{600}=\beta(\epsilon_S)^{1/4}$$
+$$\mathbf{(4)} k_{600}=\beta(\epsilon_D)^{1/4}$$
 
@@
 Equation 3's b parameter only begins to approximate 1/4 when $R_h$ is within 1% of the mean flow depth.
-At this point, the model's goodness-of-fit also reaches a reasonable value of 0.68.
+At this point, the model's goodness-of-fit also reaches a reasonable value of `round(models_eD4[models_eD4$name == 'Rh=H',]$r2,2)`.
 This indicates that a river is sufficiently inefficient when its hyradulic radius approximates the mean flow depth, and at that point $\epsilon_S \approx \epsilon_D$.
-Figures 3b-4d plot the model fits for this final scenario where $\frac{R_h}{H} >= 0.995$ As Figure 3c confirms, the small-eddy model is viable when a channel is so inefficient that dissipation via form-drag approximates the near-surface dissipation rate in equation 2.
-Figure 4d further confirms this, where the slope of the generalized model (equation 3) equals `r round(models_ulseth[models_ulseth$name == 'Rh=H',]$slope, 2)` which is nearly equivalent to 1/4.
+Figures 3b and 3c plot the model fits for this final scenario where $\frac{R_h}{H} >= 0.995$ As Figure 3c confirms, the small-eddy model is viable when a channel is so inefficient that dissipation via form-drag approximates the near-surface dissipation rate in equation 2.
+Figure 3b further confirms this, where the slope of the generalized model (equation 3) equals `r round(models_ulseth[models_ulseth$name == 'Rh=H',]$slope, 2)` which is nearly equivalent to 1/4.
 
@@
 Finally, becuase $\epsilon_D$ refers to a depth-scale quantity, we tested whether a more parsimonious model is possible by removing the 'middle-man', i.e.
-the Kolmogorov velocity scale, and directly predicting *k* from depth-scale turbulence.
-Figure 3b plots fitted model results for the $k=\beta_1U_*$ model.
-For this data, $U_*$ explains 70% of variation in $k_{600}$ and, as expected, the fitted model is nearly identical to Figure 4c in scaling dynamics (as confirmed visually and via the goodness-of-fit).
-This all indicates that depth-scale turbulent production/dissipation dominates the surface dissipation signal in maximally inefficient river channels.
+the Kolmogorov velocity scale that the small-eddy model is based on [@lamontEddyCellModel1970], and directly predicting *k* from depth-scale turbulence.
+Figure 3d plots the fitted $k=\beta_1U_*$ model.
+For this data, $U_*$ explains 70% of variation in $k_{600}$ and, as expected, the fitted model is nearly identical to Figure 3c in scaling dynamics and variance explained.
+This all indicates that depth-scale turbulent production/dissipation dominates the surface dissipation signal in extremely inefficient river channels.
 These findings build off of @ulsethDistinctAirWater2019a's intial result that channel slope exerts a drammatic influence on *k* scaling dynamics, further suggesting that fluvial *k* is as much a function of hydraulic geometry as the turbulence properties of the flow.
@@
 With all of that said, how often is a river channel so inefficient that $R_h \approx H$?
-To answer this question, we join the 701 measurements from @ulsethDistinctAirWater2019a to 171,553 discrete river hydraulics measurements from across the continetal United States [@brinkerhoffReconcilingAtaStationAtManyStations2019].
+To answer this question, we join the 701 measurements from @ulsethDistinctAirWater2019a to 171,553 discrete river hydraulics measurements from across the continental United States [@brinkerhoffReconcilingAtaStationAtManyStations2019].
 We find that `r round(percs_overall[percs_overall$flag_depth == 'Rh=H',]$perc,0)`% of hydraulic radii are within 5% of their mean flow depth.
 Because these measurements were made exclusively at streamgauges where rivers are somewhat channelized and width variability is decreased, we opted for a slightly more liberal ratio of 5%.
-Because larger rivers are more likely to have inefficient channels, we then flag the measurements that are SWOT-observable, i.e.
+Further, because larg rivers are more likely to have inefficient channels, we identify the measurements that are SWOT-observable, i.e.
 width > 100m.
-We find that while `r round(percs[percs$flag_depth == 'Rh=H',]$perc,0)[1]`% of non-SWOT observable rivers meet this inefficient condition, `r round(percs[percs$flag_depth == 'Rh=H',]$perc,0)[2]`% of SWOT-observable rivers do.
-This significant difference, and near universality of SWOT-observable rivers being inefficient, suggests that our theoretical findings in Figure 3 should hold in nearly all rivers that SWOT will observe.
+We find that while `r round(percs[percs$flag_depth == 'Rh=H',]$perc,0)[1]`% of non-SWOT observable rivers are inefficient, `r round(percs[percs$flag_depth == 'Rh=H',]$perc,0)[2]`% of SWOT-observable rivers are inefficient.
+This significant difference, and near universality of SWOT-observable rivers being inefficient, suggests that our theoretical findings in Figure 3 should reasonably hold in nearly all rivers that SWOT will observe.
 This opens the door for exploiting our 'inefficient river k model' for remote sensing *k* via the SWOT satellite.
 
-##4 Exploiting the 'inefficient river model' to remotely sense gas exchange velocity via the BIKER algorithm
-We have shown that scaling $k_{600}$ via the depth-scale shear-velocity ($U_*$) explains ~70% of the variation in $k_{600}$ in inefficient rivers and that nearly all SWOT-observable rivers are inefficient.
-Further, because of the structure of $U_*$ (Appendix A) we can easily reduce the 'inefficient river' $k_{600}$ model (Figure 3b) to an equation consisting solely of remotely-sensible river hydraulics and an intial estimate of the channel area.
-By scaling *k* this way, we significantly reduce potential equifinality issues by having just two unknowns: $k_{600}$ and the median channel area $A_0$.
-Equifinality refers to an under-constrained mathematical system that has essentially infinite parameter combinations that can produce the same result: there are in essence more unknowns than equations [@garamboisInferenceEffectiveRiver2015].
+## 4 Exploiting the 'inefficient river model' to remotely sense gas exchange velocity via the BIKER algorithm
+We have shown that scaling $k_{600}$ via the shear velocity ($U_*$) explains 70% of the variation in $k_{600}$ in inefficient rivers and that nearly all SWOT-observable rivers are inefficient.
+This is incredibly convenient for remote sensing because the structure of $U_*$ (Appendix A) reduces the model to two non-remotely-sensible terms: $k_{600}$ and an estimate of the channel area.
+This is explained below.
+By scaling *k* this way, we significantly reduce equifinality issues, where equifinality refers to an under-constrained mathematical system that has essentially infinite parameter combinations that can produce the same result: there are more unknowns than equations [@garamboisInferenceEffectiveRiver2015].
 This problem is experienced by both SWOT RSQ algorithms and in situ tools that concurrently solve for *k* and stream metabolism [@applingOvercomingEquifinalityLeveraging2018; @graceFastProcessingDiel2015; @holtgrieveSimultaneousQuantificationAquatic2010].
 In both of these domains, the other unknown parameters are often difficult to estimate (bed roughness and stream metabolism, respectively) while for BIKER, median channel area is relatively easy to approximate from the SWOT-observable river width [@brinkerhoffConstrainingRemoteRiver2020a].
-Therefore, we develop the BIKER algorithm, which is informed by the @hagemannBAMBayesianAMHGManning2017a algorithm for ungauged RSQ and further explored in more recent work by @brinkerhoffConstrainingRemoteRiver2020a.
+
+Taking all of this together, we develop the BIKER algorithm.
+BIKER is informed by the @hagemannBAMBayesianAMHGManning2017a algorithm for ungauged RSQ and further explored in more recent work by @brinkerhoffConstrainingRemoteRiver2020a.
 These papers conceptualize discharge as a Bayesian remote sensing problem, which we largely follow here to conceptualize $k_{600}$ as a Bayesian remote sensing problem that can be solved using SWOT data.
-
-BIKER, and Bayesian inference in general, starts from Bayes rule (equation 5), where $\Theta$ is some set of non-remotely-sensible parameters we want to solve for (including $k_{600}$), *x* is the observed data, $f(x|\Theta)$ is the 'likelihood function' or sampling model conditional on the parameters, and $p(\Theta)$ is the joint prior distribution of the parameters.
+BIKER, and Bayesian inference in general, starts from Bayes rule (equation 5), where $\Theta$ is some set of non-remotely-sensible parameters we want to solve for (including $k_{600}$), *x* is the observed data, $f(x|\Theta)$ is the sampling model where data are conditional on the parameters, and $p(\Theta)$ is the joint prior distribution of the parameters.
 Therefore, we are interested in solving for $p(\Theta|x)$, or the 'posterior' distribution.
@@
 To do this, $k_{600}$ is first written as a function of SWOT-observables *W* and $S_h$.
-This algebra is carried out using the model parameter from Figure 3b (56.0294) and yields equation 6, where we follow our earlier assumption that $R_h = H =\frac{A}{W}$ where *A* is wetted channel area.
+This algebra is carried out using the model parameter from Figure 3d (56.0294) and yields equation 6, assuming the channel is inefficient ($R_h = H =\frac{A}{W}$).
+Here, *A* is wetted channel area.
 *A* is further split into the SWOT-observable portion *dA* and the unobservable portion $A_0$ following @durandEstimatingReachaveragedDischarge2014a.
-*dA* is estimated assuming a rectangular river channel (usually valid when $R_h=H$) so that $dA \approx \delta H*W$.
-Thus, there are no reliances on bed roughness, flow velocity, or any other terms that are difficult to infer from river width.
+*dA* is estimated assuming a rectangular river channel so that $dA \approx \delta H*W$.
+Thus, there are no reliances on bed roughness, flow velocity, or any other terms that are difficult to infer from river width (if we implemented the nearly identical model in Figure 3b, we would need to simultaneously estimate roughness and/or velocity which becomes a much more difficult problem).
 
@@
 Next, equation 6 is written as a Bayesian sampling model, in which all of the SWOT observations are sampled from the unknown model parameters ($A_0$ and $k_{600}$).
-This is equation 7 after log-transforming all of the variables.
-The $\sigma_{k_{600}}^2$ parameter refers to the uncertainty inherent in equation 6's estimates.
+This is equation 7 after log-transforming all of the variables to describe them as normal distributions.
+$\sigma_{k_{600}}^2$ refers to the uncertainty inherent in equation 6's estimates.
 Equation 7 also necessitate that we specify prior distributions for the parameters $A_0$ and $k_{600}$.
@@
 Hyperparameter specifications are detailed in Text S2, however the goal was to rely on absolutely no in situ information such that we could run this method on any river on Earth solely using SWOT observations.
+Priors could in theory be improved if there is available a priori information about the river.
 
@@
 
-We validate BIKER on 47 SWOT-simulated rivers using daily observed $R_h$ and observed $S_h$.
-We also re-validate BIKER on the 17 rivers with the SWOT error model corrupting the SWOT-observables *W* and *H* (this workflow and the metrics used are detailed in Text S3 and Table S1).
+We validate BIKER on 47 SWOT-simulated rivers using daily observed hydraulics and the metrics described in Table S1.
+We also re-validate BIKER on the 17 rivers with the SWOT error model that corrupts the river width and slope.
 Regardless of the validation setup, we do not have observed $k_{600}$ data for these rivers, and to our knowledge no field dataset of $k_{600}$ exists in the type of temporal and spatial frequency that SWOT (and therefore the BIKER algorithm) provides.
@@
 With this setup, we are directly exploring BIKER's ability to infer observed $A_0$ and $k_{600}$ from *W* and *H* alone, as the *k* scaling model has already been sucessfully validated (Figure 3).
+This is elaborated on in Text S3.
 
@@
 
-##5 Remotely sensing carbon emissions from rivers
+## 5 Remotely sensing carbon emissions from rivers
 It is one thing to accurately predict *k*, but researchers are often most interested in the actual gas fluxes from rivers and ultimately the carbon emitted from river to atmosphere.
@@
 
-##6 Discussion
-###6.1 Towards a mechanistic understanding of gas exchange in large rivers
+## 6 Discussion
+### 6.1 Gas exchange along the stream-to-ocean continumn
 Field studies of gas exchange in wide rivers have suggested that their *k* properties behave differently than in steeper and smaller rivers [@ulsethDistinctAirWater2019a; @beaulieuControlsGasTransfer2012a; @alinPhysicalControlsCarbon2011; @wangPhysicallyBasedScaling2021].
-In sections 2 and 3 we have shown that this must be the case given the algebraic structure of the TKE budget in open-channel flows [@nakagawaTurbulenceOpenChannel1993] and the assumptions underlying turbulent scaling with interfacial fluxes [@katulStructureFunctionModel2018; @lorkeUnifiedScalingRelation2006].
-This result significantly amends how we can scale *k* in large rivers and directly addresses a long-standing problem of upscaling fluvial *k*: that our current methods perform poorer in larger rivers than small ones.
+While it has been shown that the small-eddy model is reasonably valid across most rivers [@wangPhysicallyBasedScaling2021], it has yet to be seen why and when form-drag dissipation is similar to near-surface dissipation in the context of *k*.
+In section 3 we show that the geomorphic concept of channel efficiency can explain when this is the case or not (Figure 3a).
+In extremely inefficient channels, the 'small-eddy' model is physically valid in flowing water environments as confirmed in Figures 3b and 3c.
+This has significant implications for how we understand shear and turbulence in rivers and their influence on *k*, as compared to environments where bottom-shear dominates (like lakes or oceans).
 
-We also statistically recovered the 1/4 scaling exponent in the classic $k \sim \epsilon^{1/4}$ model (Figure 3d) for gas exchange that has been repeadely validated in non-fluvial aquatic systems.
-This again suggests that *k* in large rivers behaves more similarly to lakes, esturaries, and the ocean than to steeper rivers.
-We also suggest that the TKE budget in smaller rivers and streams is additionally complicated by diffusive transport of TKE from the bed to the free surface, and that $T_d$ in equation 3 might explain why fluvial $k_{600}$ does not scale to the 1/4 power in smaller rivers.
-The presence of an appreciable $T_d$ term also increases the influence of channel bed roughness.
-@ulsethDistinctAirWater2019a showed bed roughness loosely correlates with $k_{600}$ in steep Alpine streams.
-However, they coarsely estimated bed roughness from arial imagery, only studied extremely small mountain streams, and to date most similar work has focused on labratory exercises [e.g.
-@chansonStreamReaerationNonuniform; @moogStreamReaerationNonuniform1999].
-In this study and under the 'large river' condition, we confirm that bed roughness is exerting functionally no influence on free-surface TKE (Section 2 and figure 2) and therefore *k*.
-We show that free-surface TKE must be entirely produced and dissipated by form-drag and not bottom friction or roughness.
-However, this is only true in the largest of rivers.
+@wangPhysicallyBasedScaling2021 proposed a conceptual model for fluvial *k* where *k* is dominated by bottom shear in small streams while in large rivers *k* is also affected by wind, biota, etc.
+This builds on the [@ulsethDistinctAirWater2019a] finding that steep, 'high-energy' streams have a fundamentally different $k_{600} \sim \epsilon_D$ relationship than less steep rivers and this may be due to the presence of bubble-mediated gas exchange and massive roughness elements in small mountain streams.
+Taking both of these models, along with our results, in aggregate, we suggest a combined conceptual model that follows the geomorphic concept of the 'river longitudinal profile' and is outlined in Figure 7.
+Small streams are steep, rough, and highly efficient channels that evacuate all but the largest sediment leaving a very rough riverbed.
+These small streams likely have strong *k*~hydraulics relationships and surface turbulence is dominated by their huge bed roughness.
+However, as one moves down the stream-to-ocean continumn, river channels get wider, flatter, relatively more frictional, and far less efficient at evacuating sediment.
+This means there is a concurrent fining of the riverbed and an increase in form-drag.
+In these largest rivers, form-drag starts dominating the *k*~hydraulics relationship rather than bed roughness.
+While wind certainly also plays a role in *k* in these large rivers [particuarly at higher winds- @beaulieuControlsGasTransfer2012a], we show here that for maximally inefficient river channels form-drag alone can reasonably predict $k_{600}$.
+The relationship between form-drag dissipation and wind is left to future work.
 
-###6.2 Towards remote sensing of global spatiotemporal dynamics of *k* in large rivers
+![Figure 7: Conceptual model of the river longitudinal profile as it pertains to gas exchange velocity.
+As rivers grow in size and one moves downstream, their channels become wider, flatter, less efficient at transporting sediment, and have greater relative frictional resistance.
+This suggests a shift in k~hydraulics relationships from one dominated by bed roughness in small, steep, and efficient systems versus one dominated by form-drag in large, flat, and inefficient systems.](cache/k600_theory/longProfile.jpg)
+
+### 6.2 Towards remote sensing of global spatiotemporal dynamics of *k* in large rivers
 To date, most field-scale studies of riverine gas exchange have focused on 1) its relationship with wind speed [e.g @beaulieuControlsGasTransfer2012a; @borgesVariabilityGasTransfer2004; @zappaEnvironmentalTurbulentMixing2007], 2) average flow velocity [e.g.
@@
 
-###6.3 Towards remotely sensing riverine carbon emissions using SWOT
+### 6.3 Towards remotely sensing riverine carbon emissions using SWOT
 Figure 6 confirms that BIKER is quite successful, without any in situ information aside from a $CO_2$ logger, at predicting both $FCO_2$ (Figure 6a) and 2) the bulk carbon efflux (Figure 6b).
@@
 This is particualry useful given recent advances in in situ gas concentration loggers [@ahoHighestRatesGross2021] but no such similar advances in modeling *k* at equivalent temporal resolutions.
-BIKER can likely also be ran at the field scale using arrays of pressure transducers to estimate water surface slope (rather than using satellite-based altimeters like SWOT) following recent work doing the same using the @hagemannBAMBayesianAMHGManning2017a RSQ algorithm [@harlanDischargeEstimationDense].
+BIKER can likely also be ran at the field scale using arrays of pressure transducers to estimate water surface slope (rather than using satellite-based altimeters like SWOT) following recent work doing the same using the @hagemannBAMBayesianAMHGManning2017a RSQ algorithm [@harlanDischargeEstimationDense2021].
 
@@
 Future work should systematically quantify prediction error from coupling global-scope HG models with modeled discharge, as is the default workflow used in fluvial $FCO_2$ upscaling studies [e.g.
-@lauerwaldSpatialPatternsCO22015; @horgbyUnexpectedLargeEvasion2019; @raymondGlobalCarbonDioxide2013a].
-**I assume Shaoda's paper will still be in review and can't cite here but this is exactly what they do...**
+@lauerwaldSpatialPatternsCO22015; @horgbyUnexpectedLargeEvasion2019; @raymondGlobalCarbonDioxide2013a; **Liu et al in reivew**].
 
@@
 
-##7 Conclusions
+## 7 Conclusions
 Gas exchange from aquatic systems has been studied for nearly a century, with a robust model (equation 2) that has been repeatedly verified across many non-fluvial environments.
@@
 
-##8 Acknowledgements
+## 8 Acknowledgements
 C.B.
@@
 All code to build and generate results, figures, and the manuscript is available at https://github.com/craigbrinkerhoff/RSK600.
-Data used in this study was generously made available by the authors of @frassonExploringFactorsControlling2021, @durandIntercomparisonRemoteSensing2016b, @beaulieuControlsGasTransfer2012a, @ulsethDistinctAirWater2019a, and @raymondScalingGasTransfer2012a.
+We thank Renato Frasson, Michael Durand, Amber Ulseth, and Jake Beaulieu for generously making their data available for this study.
+We also thank the entire SWOT discharge working group for their decade-plus body of work which inspired this study.
 
-##9 Apendix A
-*Table A1: Variable description and notation for this study. TKE refers to 'turbulent kinetic energy'*
+## 9 Apendix A
+*Table A1: Variable description and notation for this study.*
 
@@
 |$D_m$|Molecular diffusion coefficient|NA|$[\frac{L^2}{T}]$|
-|$\epsilon$|TKE dissipation rate|$\frac{U_*^3}{H}$|$[\frac{E}{M*T}]$|
-|$\epsilon_D$|TKE dissipation rate via form-drag|$gSU$|$[\frac{E}{M*T}]$|
+|$\epsilon_S$|Dissipation rate of near-surface turbulence|log-law-of-the-wall model: $\frac{U_*^3}{H}$|$[\frac{E}{M*T}]$|
+|$\epsilon_D$|Form-drag turbulent dissipation rate|$gSU$|$[\frac{E}{M*T}]$|
 |$FCO_2$|$CO_2$ evasion flux from river to atmosphere|NA|$[\frac{M}{L^2T}]$|
@@
 |$G$|TKE production rate|$\frac{UU_*}{H}$|$[\frac{E}{M*T}]$|
-|$h$|Elevation above the bed|NA|$[L]$|
-|$H$|Mean flow depth|NA|$[L]$|
+|$H$|Mean flow depth|$A/W$|$[L]$|
 |$k$|gas exchange velocity|NA|$[\frac{L}{T}]$|
-|$k_{600}$|gas exchange velocity normalized to Sc=600|$(\frac{600}{Sc_k})^{-0.5}k$|$[\frac{L}{T}]$|
+|$k_{600}$|gas exchange velocity normalized to $Sc=600$|$(\frac{600}{Sc_k})^{-0.5}k$|$[\frac{L}{T}]$|
 |$R_h$|Hydraulic radius|$\frac{AW}{2H+W}$|$[L]$|
@@
 |$v$|kinematic viscosity|NA|$[\frac{L^2}{T}]$|
-|$V_d$|TKE viscous diffusion rate|NA|$[\frac{E}{M*T}]$|
-|$v_K$|turbulence characteristic velocity scale|NA|$[\frac{L}{T}]$|
 |$W$|Flow width|NA|$[L]$|

### 2021-09-14 supp_v5.Rmd
big ms_v5 push

@@
 title: "Supplemental Information: Gas exchange in large rivers controlled by hydraulic geometry: implications for remotely sensing gas exchange via the SWOT satellite"
-author: "CB Brinkerhoff, CJ Gleason, CJ Zappa, PA Raymond, MH Harlan"
-date: "`r Sys.Date()`"
+author:
+  - Craig B Brinkerhoff:
+      email: cbrinkerhoff@umass.edu
+      institute: UMass
+      correspondence: true
+  - Colin J Gleason:
+      institute: UMass
+  - Christopher J Zappa:
+      institute: Lamont
+  - Peter A Raymond:
+      institute: Yale
+  - Merritt H Harlan:
+      institute: UMass
+institute:
+  - UMass: Department of Civil & Environmental Engineering, University of Massachusetts, Amherst, MA
+  - Lamont: Lamont-Doherty Earth Observatory, Columbia University, Palisades, NY
+  - Yale: School of the Environment, Yale University, New Haven, CT
+bibliography: References.bib
 csl: agu_citation_style.csl
@@
   word_document:
+      pandoc_args:
+        - '--lua-filter=scholarly-metadata.lua'
+        - '--lua-filter=author-info-blocks.lua'
       reference_docx: word-styles-reference.docx
-bibliography: References.bib
 
@@
 
-##Contents
-This supplementary information contains 4 texts, 4 figures and 2 tables.
+## Contents
+This supplementary information contains 4 texts, 3 figures and 2 tables.
 Please consult https://github.com/craigbrinkerhoff/RSK600 for all code to build and generate results, figures, and the manuscript.
@@
 
-##Figures
+## Figure S1
 ![Figure S1: Flowchart of the entire study's workflow](cache/flowchart.jpg)
 
+## Figure S2
 ![Figure S2 Timeseries of the biweekly CO2 data from @beaulieuControlsGasTransfer2012a.
@@
 
+## Figure S3
 ![Figure S3 All timeseries of k per river](cache/validation/timeseries_noerr.jpg)
 
-##Tables
+## Table S1
 *Table S1: Validation metrics used in this study, where r is the correlation coefficient, Nt is number of observations and i is the specific observation.
@@
 
+## Table S2
 *Table S2: Details on the 3 depth hydraulic geometry models used to estimate FCO2 from the SWOT rivers (section 2.4).*
@@
 
-##References
+## References

### 2021-09-30 docs/scraps.Rmd
model update to the chainsaw model

--- 2021-09-14 manuscript_v5.Rmd
@@
----
-title: "Gas exchange in large rivers influenced by hydraulic geometry: implications for remotely sensing gas exchange via the SWOT satellite"
-author:
-  - Craig B Brinkerhoff:
-      email: cbrinkerhoff@umass.edu
-      institute: UMass
-      correspondence: true
-  - Colin J Gleason:
-      institute: UMass
-  - Christopher J Zappa:
-      institute: Lamont
-  - Peter A Raymond:
-      institute: Yale
-  - Merritt H Harlan:
-      institute: UMass
-institute:
-  - UMass: Department of Civil & Environmental Engineering, University of Massachusetts
-  - Lamont: Lamont-Doherty Earth Observatory, Columbia University
-  - Yale: School of the Environment, Yale University
-bibliography: References.bib
-csl: agu_citation_style.csl
-output:
-  word_document: 
-      pandoc_args:
-        - '--lua-filter=scholarly-metadata.lua'
-        - '--lua-filter=author-info-blocks.lua'
-      reference_docx: word-styles-reference.docx
----
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-```
-
-## Key Points
-- River channel efficiency effects the relationship between gas exchange and dissipation of turbulence
-- Wide, less "efficient" river channels exert greater influence on near-surface turbulence than narrower, more efficient ones
-- BIKER algorithm exploits this theory, predicting gas exchange velocity and $CO_2$ fluxes from simulated SWOT satellite data and in situ $CO_2$ data
-
-## Keywords
-gas exchange, fluvial geomorphology, open-channel flow, remote sensing, SWOT, biogeochemistry, ungauged basin
-
-## Abstract
-*AGU Advances. So 250 words here* <br>
-*AGU advances is 8000 words total, the format seems to be a long letter so I've tried to right this in that style.*
-
-## Plain Language Summary
-*Necessary for AGU advances (200 words)*
-
-## 1 Introduction
-Natural systems play a fundamental role in the budgeting and accounting of the global carbon cycle under climate change.
-Since the publication of @colePlumbingGlobalCarbon2007a, the global river network is recognized to emit substantial amounts of carbon to the atmosphere, in addition to exporting it to the oceans.
-Current estimates of total carbon dioxide evasion ($FCO_2$) to the atmosphere from the global river network vary from 650-1800 Tg C/yr [@lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a] **maybe add Shaoda here**, with 167 Tg-C/yr coming from mountain streams alone [@horgbyUnexpectedLargeEvasion2019].
-Despite its incredibly small percentage of the global land surface [0.47%- @raymondGlobalCarbonDioxide2013a], this flux is on par with the total oceanic $CO_2$ uptake rate [@gruberOceanicSinkAnthropogenic2019; @horgbyUnexpectedLargeEvasion2019] and the global forest carbon uptake rate [@panLargePersistentCarbon2011].
-It is still relatively poorly constrained but is clearly a critical component of the global carbon cycle.
-Equation 1 represents this flux given the gas concentration gradient between the water $CO_{2_{water}}$ and the air $CO_{2_{air}}$ and the gas exchange velocity *k*.
-Consult Appendix A for variable nomenclature used in this study.
-
-$$\mathbf{(1)} FCO_2=k(CO_{2_{water}}-CO_{2_{air}})$$
-
-The structure of equation 1 necessitates that calculations of this flux are highly sensitive to one's measurements/estimates of *k* [@hallGasExchangeStreams2020].
-More broadly, fluvial *k* for various dissolved gases are of significant importance to aquatic ecologists modeling stream metabolism [e.g.
-@bernhardtMetabolicRegimesFlowing2018] and water quality engineers modeling river responses to pollutant loadings [among other things- @chapraSurfaceWaterQualityModeling2008].
-Because of this far-reaching importance, decades of work have focused on elucidating the physical mechanisms behind *k* in oceans and lakes [@wanninkhofAdvancesQuantifyingAirSea2009] and, to a lesser extent, fluvial gas exchange [@hallGasExchangeStreams2020].
-Given that *k* should scale with near-surface turbulence in a turbulent flow [@hallGasExchangeStreams2020], extensive field and labratory experiments have converged on the 'small-eddy model' as proposed by @lamontEddyCellModel1970, which scales *k* via the smallest-scale turbulent eddies [e.g.
-@lorkeUnifiedScalingRelation2006; @zappaEnvironmentalTurbulentMixing2007; @zappaVariationSurfaceTurbulence2003; @moogStreamReaerationNonuniform1999; @wangCoefficientsSmallEddy2015; @vachonRelationshipNearsurfaceTurbulence2010].
-This model is provided as equation 2, where $\epsilon_S$ is the dissipation rate of near-surface turbulence, $v$ is the kinematic viscosity and *Sc* is the Schmidt number.
-Note that we directly measuring $\epsilon_S$ is non-trivial and though a frequent model is to impose the log-law-of-the-wall to scale bottom shear to the surface, this necessiarly assumes that bed shear alone effects turbulence which is often not the case.
-$$\mathbf{(2)} k\sim Sc^{-n}(v\epsilon_S)^{1/4}$$
-
-While this model works reasonably well in non-fluvial environments where turbulence production is mostly limited to bottom shear or wind shear [e.g.
-@lorkeUnifiedScalingRelation2006; @zappaEnvironmentalTurbulentMixing2007], there is considerable uncertainity in how it applies to fluvial systems.
-First, 'bubble-mediated gas exchange' in whitewater might lead to substantially higher *k* in small mountain streams with very large bedload [@hallAirWaterOxygen2012; @ulsethDistinctAirWater2019a].
-Additionally, 'form-drag shear' via river channel banks, meanders, bars, etc.
-further complicate scaling fluvial *k* [@moogStreamReaerationNonuniform1999].
-Those authors proposed modeling fluvial *k* via form-drag dissipation $\epsilon_D$, assuming that total reach stream power is responsible for all turbulence production and therefore dissipation.
-Numerous authors have since shown that this model ($\epsilon_D=gS\overline{U}$) reasonably predicts *k* in rivers [@raymondScalingGasTransfer2012a; @ulsethDistinctAirWater2019a; @moogStreamReaerationNonuniform1999], though it should be stressed that $\epsilon_S \neq \epsilon_D$.
-This is because $\epsilon_D$ refers to a depth-scale disispation rate and is not necessairily indicative of the near-surface dissipation rate.
-Consequently, the, $k\sim\epsilon_D$ relationship usually has a slope much steeper than the 1/4 in equation 2.
-
-Compared to small streams, relatively little attention has been paid to gas exchange in very large rivers.
-Large rivers are often conceptualized as a hybrid condition of both fluvial and non-fluvial *k* dynamics.
-To date, the handful of existing field studies of large-river *k* suggest that *k* begins to be influenced by wind shear once the water surface is less protected, though little else is well-established [@alinPhysicalControlsCarbon2011; @beaulieuControlsGasTransfer2012a; @wangPhysicallyBasedScaling2021].
-This is also presumably limited to forested environments that provide shelter from wind.
-
-These mechanistic uncertainties are additionally limited by a large dearth of field-measured fluvial *k*.
-@wangPhysicallyBasedScaling2021 attempted to address this by simulating *k* in 35 rivers using a stream metabolism model [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets.
-They found that equation 2, coupled with $\epsilon$ modeled via the log-law-of-the-wall, is valid in their simulated rivers and that *k*~streamflow relationships break down in large rivers.
-However, they did not directly compare against *k* measurements made in extremely small systems [@ulsethDistinctAirWater2019a] and stopped short of parsing out hydraulic explanations for their results.
-
-Finally, these mechanistic uncertainties are then propogated through upscaling workflows when biogeochemists predict *k* across thousands of rivers [e.g @borgesGloballySignificantGreenhousegas2015a; @horgbyUnexpectedLargeEvasion2019; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a] via equation 1 coupled with hydraulic geometry [HG: the scaling relationships between streamflow and river channel hydraulics- @leopoldHydraulicGeometryStream1953a].
-It is currently not well understood how sensitive global estimates of fluvial gas evasion are to the specific HG model that is employed by the worker.
-Further, these approaches rely on either in situ discharge records or modeled streamflow which introduces additional uncertainities.
-This is all exacerbated in ungauged basins that cover large areas, especially in the carbon-rich Arctic inland waters, where little in situ information is available and fieldwork is impractical [@gleasonRemoteSensingRiver2020].
-
-A potential alternative to this upscaling approach is to directly estimate a river's hydraulic properties from remote sensing (RS) data.
-Remote sensing of river hydraulics is a burgeoning subfield within remote sensing of hydrology, often in service of remote sensing of river discharge [RSQ- @gleasonRemoteSensingRiver2020].
-This is accomplished via two general approaches: 'gauged' methods which rely on in situ river data to calibrate one's method to the river(s) at hand and 'ungauged' techniques which focus on hydraulic generalizability in the service of merely improving existing knowledge in data-poor domains [@gleasonRemoteSensingRiver2020].
-Many, but not all, of these ungauged approaches are developed in the context of the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission.
-SWOT is expected to launch in 2022 and provide the world's first global measurements of water surface extent and elevation at novel temporal resolutions.
-SWOT is a wide swath radar interferometer and will sample rivers every 1 to 7 days per 21 day repeat cycle and will measure rivers wider than 100m with a goal of expanding this to rivers at least 50m wide [@biancamariaSWOTMissionIts2016].
-A decade of SWOT work has explored the multi-parameter problem of estimating the river hydraulic parameters of roughness and bathymetry from remote sensing to produce the SWOT discharge product [e.g.
-@andreadisConstrainingAssimilationSWOT2020d; @brinkerhoffConstrainingRemoteRiver2020a; @durandEstimatingReachaveragedDischarge2014a; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014a; @garamboisVariationalEstimationEffective2020; @brissetAssimilationAltimetricData2018; @hagemannBAMBayesianAMHGManning2017a; @larnierRiverDischargeBathymetry2020a; @oubanasDischargeEstimationUngauged2018].
-
-In this context, here we revist the fundamental mechanisms behind gas exchange and turbulence in large rivers to answer the following question: does *k*'s relationship with form-drag fundamentally change along the stream-to-river continumn and how does this relate to one of the classical models of gas evasion in aquatic systems (equation 2)?
-We exploit the findings from this simple analysis to develop a novel methodology that predicts $k_{600}$ (*k* normalized to a Schmidt number of 600) and its explicit uncertainity solely using SWOT observations.
-The method requires no in situ inputs of any kind (although in situ data can be ingested and will improve results).
-We name the RS of $k_{600}$ algorithm BIKER, or the 'Bayesian Inference/Inversion of the $k_{600}$ Evasion Rate' and validate it for 47 SWOT-observable rivers from around the world using hydraulic models to produce SWOT-like data (as SWOT has not yet launched).
-We also quantify BIKER's sensitivity to the expected SWOT measurement errors on 17 of those rivers: while SWOT data represent a sea change in inland water monitoring, it is expected to have an approximately 10cm error in water surface elevation [@biancamariaSWOTMissionIts2016] as well as river width errors [@frassonExploringFactorsControlling2021].
-Finally, we use previously published dissolved $CO_2$ data to represent a hypothetical in situ $CO_2$ sensor and compare the bulk carbon efflux from the 47 rivers as calculated using BIKER and previously published in situ techniques for predicting $k_{600}$.
-
-## 2 Data
-Numerous datasets were used in this study.
-Please see Figure 1 for a map of the approximate locations for the data used in this study.
-We also provide a flowchart detailing the entire study as Figure S1.
-
-![Figure 1: Map of the 47 hydraulic models and 1 timeseries of CO2 samples used in this study.
-Note that hydraulic model locations are approximate as some of the models are not geo-referenced.
-Not mapped here are over 170,000 discrete measurements of river channel hydraulics from across the continetal United States (Brinkerhoff et al.
-2019) and over 700 gas exchange velocity measurements (Ulseth et al.
-2019).](cache/maps/map_fin.jpg)
-
-Gas exchange theory (section 3) is explored using the dataset from @ulsethDistinctAirWater2019a who measured and/or collected from the literature over 700 measurements of stream hydraulics and $k_{600}$ and is, to our knowledge, the largest collection of field-measured river and stream $k_{600}$.
-To assess general hydraulic conditions in rivers across geographic contexts, we additionally use a previously published compilation of field hydraulics measurements [@brinkerhoffReconcilingAtaStationAtManyStations2019].
-These were originally made to calibrate United States Geological Survey (USGS) streamgauge rating curves.
-That dataset contains over 530,000 unique measurements of river channel velocity, width, and discharge from across the continental United States.
-This is to our knowledge the largest collection of field measurements of river channel hydraulics.
-Ultimately, this dataset was filtered down to 171,553 discrete measurements for this study (see Text S1 for our filtering protocol).
-
-BIKER validation (section 4) was performed on 47 SWOT-simulated rivers.
-Because SWOT has yet to launch, it is standard practice to benchmark SWOT-related algorithms on "SWOT-like" data.
-These simulated rivers are simply river-reach-averaged hydraulic model outputs where the water surface heights and widths are labelled as RS observations and are used as the sole inputs to BIKER.
-Here, we use 47/51 rivers collected by @frassonExploringFactorsControlling2021 and @durandIntercomparisonRemoteSensing2016b.
-These are the two benchmarking studies that have explored RSQ algorithm performance for the SWOT mission.
-The approximate locations of these rivers are plotted in Figure 1, and please consult both of those papers for all of the hydraulic model specifications.
-Ultimately, the 47 rivers are spread across the United States, France, Italy, the United Kingdom, and Bangladesh.
-We omit three models from @durandIntercomparisonRemoteSensing2016b because they lack enough hydraulic information to calculate the shear velocity $U_\star$, which is necessary for algorithm validation and is explained in section 4.
-These are two models for the Saint Lawrence River and one for the Tanana River.
-
-To assess the influence of measurement error on BIKER's performance (section 4), we use the error model developed by @durandHowWillRadar2020a and implemented on 17/47 of the rivers by @frassonExploringFactorsControlling2021.
-Error in SWOT measurements will come from both the error tolerances intrinsic in the satellite data product as well as radar layover error.
-Layover error is the phenomenon when radar returns from different places arrive at the sensor at the same time, leading to taller landscape features appearing closer to the sensor than shorter landscape features that are the same horizontal distance from the sensor [@durandHowWillRadar2020a].
-Width errors due to poor water classification are ignored as they were in @frassonExploringFactorsControlling2021.
-
-For the $CO_2$ evasion and carbon efflux calculations (section 5), we use 26 bi-weekly dissolved $CO_2$ samples made by @beaulieuControlsGasTransfer2012a at one location in the Ohio River for one calender year from 2008-2009 (Figure S2).
-Note that this $CO_2$ data is for the Ohio River only but was applied to all 47 rivers (which includes multiple sections of the Ohio River).
-Because we are exclusively interested in the relative differences between $FCO_2$ estimates and not the raw fluxes themselves, any $CO_2$ data representative of SWOT-observable rivers was deemed acceptable for this validation exercise.
-
-## 3 Form-drag in large rivers and implications for scaling fluvial gas exchange velocity
 Turbulence is fundamental to fluvial geomorphology and open-channel flows and has been the subject of extensive research across scientific and engineering disciplines for over a century.
-As noted in the introduction, turbulence in river flows is usually generated by shear forces at the riverbed, but also across the entire flow depth via 'form-drag' from riverbanks, meanders, bars, suspended sediment, etc [@moogStreamReaerationNonuniform1999].
-This concept is synonymous with the geomorphic idea of available stream power, or the total energy dissipated from a river reach via the liquid's potential energy along its slope [originally derived from first-principles by @bagnoldSedimentDischargeStream1960; @bagnoldApproachSedimentTransport1966].
+As noted in the introduction, turbulence in river flows is usually generated by shear forces at the riverbed, but also across the entire flow depth via 'form-drag' from riverbanks, meanders, bars, suspended sediment, etc.
+[@moogStreamReaerationNonuniform1999].
+This concept is synonymous with the geomorphic idea of available stream power, or the total energy dissipated from a river reach via the liquid's potential energy while it flows downslope [@bagnoldSedimentDischargeStream1960; @bagnoldApproachSedimentTransport1966].
 Both @raymondScalingGasTransfer2012a and @moogStreamReaerationNonuniform1999 confirmed that $\epsilon_D$ (simply the available stream power normalized by the water mass) is far greater than $\epsilon_S$ in open-channel flows, therefore generating the majority of fluvial TKE dissipation.
@@
 
-![Figure 3: Model development for predicitng k from form-drag dissipation in rivers.
-(a) Rh/H thresholds versus fitted model results.
-Dark blue is the fitted b parameter in equation 3 and dark green is the goodness-of-fit of equation 4 as defined by the coefficient of determination.
-Once a channel is maximally inefficient, i.e.
-Rh=H, equation 4 is a viable model for k and equation 3 empirically recovers the theoretical value of 1/4 as outlined in equation 2.
-(b-d) Fitted model results for inefficient channels (defined as Rh/H >= 0.995) predicting k via the shear velocity U*, equation 4, and equation 3 (respectively).
-Note that colors align to the same models for subpanels a, c, and d.
-Further note that the coefficient of determination in subpanel (d) refers to log-transformed residuals and hence is not directly comparable to those in subpanels (b) and (c).](cache/k600_theory/k600Plot.jpg)
-
 Finally, becuase $\epsilon_D$ refers to a depth-scale quantity, we tested whether a more parsimonious model is possible by removing the 'middle-man', i.e.
@@
 These findings build off of @ulsethDistinctAirWater2019a's intial result that channel slope exerts a drammatic influence on *k* scaling dynamics, further suggesting that fluvial *k* is as much a function of hydraulic geometry as the turbulence properties of the flow.
-
-With all of that said, how often is a river channel so inefficient that $R_h \approx H$?
-To answer this question, we join the 701 measurements from @ulsethDistinctAirWater2019a to 171,553 discrete river hydraulics measurements from across the continental United States [@brinkerhoffReconcilingAtaStationAtManyStations2019].
-We find that `r round(percs_overall[percs_overall$flag_depth == 'Rh=H',]$perc,0)`% of hydraulic radii are within 5% of their mean flow depth.
-Because these measurements were made exclusively at streamgauges where rivers are somewhat channelized and width variability is decreased, we opted for a slightly more liberal ratio of 5%.
-Further, because larg rivers are more likely to have inefficient channels, we identify the measurements that are SWOT-observable, i.e.
-width > 100m.
-We find that while `r round(percs[percs$flag_depth == 'Rh=H',]$perc,0)[1]`% of non-SWOT observable rivers are inefficient, `r round(percs[percs$flag_depth == 'Rh=H',]$perc,0)[2]`% of SWOT-observable rivers are inefficient.
-This significant difference, and near universality of SWOT-observable rivers being inefficient, suggests that our theoretical findings in Figure 3 should reasonably hold in nearly all rivers that SWOT will observe.
-This opens the door for exploiting our 'inefficient river k model' for remote sensing *k* via the SWOT satellite.
-
-## 4 Exploiting the 'inefficient river model' to remotely sense gas exchange velocity via the BIKER algorithm
-We have shown that scaling $k_{600}$ via the shear velocity ($U_*$) explains 70% of the variation in $k_{600}$ in inefficient rivers and that nearly all SWOT-observable rivers are inefficient.
-This is incredibly convenient for remote sensing because the structure of $U_*$ (Appendix A) reduces the model to two non-remotely-sensible terms: $k_{600}$ and an estimate of the channel area.
-This is explained below.
-By scaling *k* this way, we significantly reduce equifinality issues, where equifinality refers to an under-constrained mathematical system that has essentially infinite parameter combinations that can produce the same result: there are more unknowns than equations [@garamboisInferenceEffectiveRiver2015].
-This problem is experienced by both SWOT RSQ algorithms and in situ tools that concurrently solve for *k* and stream metabolism [@applingOvercomingEquifinalityLeveraging2018; @graceFastProcessingDiel2015; @holtgrieveSimultaneousQuantificationAquatic2010].
-In both of these domains, the other unknown parameters are often difficult to estimate (bed roughness and stream metabolism, respectively) while for BIKER, median channel area is relatively easy to approximate from the SWOT-observable river width [@brinkerhoffConstrainingRemoteRiver2020a].
-
-Taking all of this together, we develop the BIKER algorithm.
-BIKER is informed by the @hagemannBAMBayesianAMHGManning2017a algorithm for ungauged RSQ and further explored in more recent work by @brinkerhoffConstrainingRemoteRiver2020a.
-These papers conceptualize discharge as a Bayesian remote sensing problem, which we largely follow here to conceptualize $k_{600}$ as a Bayesian remote sensing problem that can be solved using SWOT data.
-BIKER, and Bayesian inference in general, starts from Bayes rule (equation 5), where $\Theta$ is some set of non-remotely-sensible parameters we want to solve for (including $k_{600}$), *x* is the observed data, $f(x|\Theta)$ is the sampling model where data are conditional on the parameters, and $p(\Theta)$ is the joint prior distribution of the parameters.
-Therefore, we are interested in solving for $p(\Theta|x)$, or the 'posterior' distribution.
-For BIKER, *x* is the SWOT-observables: river width *W* and water surface elevation *H* (which is used to calculate the water surface slope $S_h$).
-Note that $p(x)$ is usually computationally intractable to integrate, so Bayesian inference tools require only the proportionality to be specified: $p(\Theta|x) \propto f(x|\Theta)p(x)$.
-Sampling algorithms are then used to approximate the posterior distribution, as is done in BIKER.
-
-$$\mathbf{(5)} p(\Theta|x) = \frac{f(x|\Theta)p(\Theta)}{p(x)}$$
-
-The heart of BIKER is its reformulation of the $k_{600}$ model as a Bayesian sampling model that is conditional only on the data that SWOT will provide.
-To do this, $k_{600}$ is first written as a function of SWOT-observables *W* and $S_h$.
-This algebra is carried out using the model parameter from Figure 3d (56.0294) and yields equation 6, assuming the channel is inefficient ($R_h = H =\frac{A}{W}$).
-Here, *A* is wetted channel area.
-*A* is further split into the SWOT-observable portion *dA* and the unobservable portion $A_0$ following @durandEstimatingReachaveragedDischarge2014a.
-*dA* is estimated assuming a rectangular river channel so that $dA \approx \delta H*W$.
-Thus, there are no reliances on bed roughness, flow velocity, or any other terms that are difficult to infer from river width (if we implemented the nearly identical model in Figure 3b, we would need to simultaneously estimate roughness and/or velocity which becomes a much more difficult problem).
-
-$$\mathbf{(6)} \begin{eqnarray} k_{600} = \bigg(56.0294 S_h(\frac{A_0+dA}{W})\bigg)^{1/2} \end{eqnarray}$$
-
-Next, equation 6 is written as a Bayesian sampling model, in which all of the SWOT observations are sampled from the unknown model parameters ($A_0$ and $k_{600}$).
-This is equation 7 after log-transforming all of the variables to describe them as normal distributions.
-$\sigma_{k_{600}}^2$ refers to the uncertainty inherent in equation 6's estimates.
-Equation 7 also necessitate that we specify prior distributions for the parameters $A_0$ and $k_{600}$.
-Prior distributions, defined by their hyperparameters, formalize the a priori estimates (and uncertainties) for the non-remotely-sensed terms.
-More intuitively, BIKER priors represent our 'prior river knowledge' of what $A_0$ and $k_{600}$ probably are for some river since they cannot be remotely sensed.
-Hyperparameter specifications are detailed in Text S2, however the goal was to rely on absolutely no in situ information such that we could run this method on any river on Earth solely using SWOT observations.
-Priors could in theory be improved if there is available a priori information about the river.
-
-$$\mathbf{(7)} \left[log(56.0294) + \frac{1}{2}log(g) + \frac{1}{2}log(S_h) - \frac{1}{2}log(W)\right] \sim normal\left(\left[log(k_{600})-\frac{1}{2}log(A_0+dA)\right], \sigma^2_{k_{600}}\right)$$
-
-With the sampling model (equation 7) and hyperparameters described (Text S2), a joint posterior distribution conditional on the SWOT observations is specified.
-To approximate this distribution, we use a Markov Chain Monte Carlo (MCMC) algorithm implemented using the Stan probabilistic programming language.
-Specifically, Stan uses a Hamiltonian Monte Carlo sampler which reduces computation time relative to other sampling algorithms [@hagemannBAMBayesianAMHGManning2017a].
-
-We validate BIKER on 47 SWOT-simulated rivers using daily observed hydraulics and the metrics described in Table S1.
-We also re-validate BIKER on the 17 rivers with the SWOT error model that corrupts the river width and slope.
-Regardless of the validation setup, we do not have observed $k_{600}$ data for these rivers, and to our knowledge no field dataset of $k_{600}$ exists in the type of temporal and spatial frequency that SWOT (and therefore the BIKER algorithm) provides.
-Therefore, we take the model outlined in Figure 3b and use that to calculate the observed $k_{600}$ that BIKER is validated against.
-With this setup, we are directly exploring BIKER's ability to infer observed $A_0$ and $k_{600}$ from *W* and *H* alone, as the *k* scaling model has already been sucessfully validated (Figure 3).
-This is elaborated on in Text S3.
-
-Figure 4a plots the validation results for $k_{600}$ (with no SWOT measurement error) across all  rivers and all timesteps.
-The points are the posterior means while the black lines are the 95% confidence intervals (CIs) for the predictions.
-$k_{600}$ is strongly correlated with the BIKER-predicted $k_{600}$ ($r^2$ of `r round(results_3_2_all$r2, 2)`).
-Using absolutely no in situ information, BIKER captures the magnitude of the predictions and most points fall on or near the 1:1 line.
-The regression of the estimates (solid grey line) nearly recovers the 1:1 line (dashed black), but there is an over/underestimation bias in the largest/smallest $k_{600}$ values.
-The RMSE for the BIKER predictions is `r round(results_3_2_all$rmse,2)` m/day) across all observations.
-
-![Figure 4.
-a: Validation of BIKER for 47 SWOT rivers.
-Black bars are 95% CIs for the modeled values.
-Grey line is linear regression (and 95% prediction intervals are dashed) and dashed black line is 1:1 line.
-b-d: validation timeseries for three rivers representative of good, reasonable, and poor BIKER performance.
-b) was randomly selected from the upper tertile of NRMSE scores, c) was randomly selected from the middle tertile, and d) from the worst tertile.
-See Figure S3 for all other rivers.
-Model results include the posterior means and 95%.](cache/validation/validation.jpg)
-
-Figure 4b-d are representative timeseries plots of predicted and observed $k_{600}$ for three rivers chosen randomly from those with 'good' NRMSE scores (b), 'okay' NRMSE scores (c), and 'bad' NRMSE scores (d).
-See Table S1 for the definition of NRMSE, the Figure 4 caption for how this was determined, and Figure S3 for the other river timeseries plots.
-For the Missouri Midsection River, the entire timeseries of $k_{600}$ is nearly perfectly predicted, while in the Seine River the dynamics (peaks and valleys) are reasonably captured but they are magnified to be far larger than the observed dynamics.
-Generally, though, mean $k_{600}$ is reasonably recovered for the Seine, as confirmed visually.
-In the Fresno River, there is significant positive bias in the estimates but also massive uncertainty (per the 95% CIs) in those estimates, indicating that BIKER is highly uncertain about its output (and rightfully so).
-Correct temporal dynamics are also largely missing from BIKER's Fresno River predicitions.
-
-Figure 5a plots validation metrics calculated for each river with and without SWOT measurement error (green and purple, respectively).
-The points making up these boxplots (47 and 17, respectively) are overlain atop the boxplots.
-See Table S1 for metric definitions.
-SWOT measurement errors neglibly influence BIKER's performance across all four error metrics (Figure 5a), though caution should be used in over-interepting boxplots with a sample size of only 17.
-Given Figure 5a, we deem that SWOT measurement error does not exert a significant influence on BIKER and so the results presented for the rest of the manuscript assume no measurement error in order to use all 47 rivers.
-
-![Figure 5.
-a: Performance metrics by river.
-See Table S1 for metric definitions.
-b: NRMSE scores (per river) versus the number of SWOT observations used in the Bayesian inversion.
-As expected, performance improves substantially with large amounts of data and degrades significantly with small amounts of data.
-Dashed lines denote scores of 0, 0.50, and 1.
-CIs.](cache/validation/validation_by_river.jpg)
-
-Median river-specific $r^2$ is `r round(median(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$r2, na.rm=T),2)`, which is excellent given that absolutely no in situ information is being used to predict $k_{600}$.
-NRMSE and RRMSE have median scores of `r round(median(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$nrmse, na.rm=T),2)` and `r round(median(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$rrmse, na.rm=T),2)`, respectively, which are again good for a completely ungauged method.
-These are comparable to the best NRMSE scores achieved by the SWOT RSQ algorithms [@frassonExploringFactorsControlling2021].
-Median rBIAS is `r round(median(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$rBIAS, na.rm=T),2)`, highlighting a  positive bias in most rivers' predictions.
-This further supports the visual evidence in Figure 4 that sometimes BIKER is overestimating the magnitude of $k_{600}$ and that this might be river-specific.
-While median $r^2$ and rBIAS scores were strong, the ranges of these scores were somewhat large (standard deviation for $r^2$ of `r round(sd(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$r2, na.rm=T),2)` and for rBIAS of `r round(sd(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$rBIAS, na.rm=T),2)`).
-
-Figure 5b highlights one benefit of using Bayesian inference to estimate $k_{600}$: because the posterior is conditional on the SWOT observations, performance should improve with more data.
-Figure 5b plots by-river NRMSE scores versus the number of SWOT observations.
-While performance varies considerably when observations are up to ~400, the three rivers with nearly 600 observations universally show excellent BIKER performance, and the worst BIKER performance is universally in rivers with only 12 observations.
-
-## 5 Remotely sensing carbon emissions from rivers
-It is one thing to accurately predict *k*, but researchers are often most interested in the actual gas fluxes from rivers and ultimately the carbon emitted from river to atmosphere.
-Therefore, we also explore 1) BIKER's ability to reproduce $FCO_2$ $[M/L^2T]$ (equation 1) from these 47 rivers, and 2) a comparison of the representative carbon efflux $[M/T]$ from BIKER with established in situ methods.
-The details of this workflow are in Text S4, but broadly we pair the 26 biweekly $CO_2$ and water temperature samples (section 2, figure S2) from @beaulieuControlsGasTransfer2012a with a subset of SWOT observations (as the data are not daily).
-We then calculate $FCO_2$ using equation 1 and assuming atmospheric $CO_{2_{air}}$ is 390 uatm.
-Finally, we estimate a median bulk carbon efflux using BIKER's posterior means and three other in situ models for average channel depth (used to calculate $U_\star$) used for upscaling: one trained on the @brinkerhoffReconcilingAtaStationAtManyStations2019 dataset, and two previously published models [@raymondGlobalCarbonDioxide2013a; @raymondScalingGasTransfer2012a].
-See Table S2 for their definitions.
-This allows us to assess whether BIKER's estimates (wholly ungauged) are comparable to gauged methods (all four HG models).
-
-In Figure 6a, there is a very strong fit to the observed $FCO_2$ data, with an RMSE of `r round(results_3_3_all$rmse,2)` $g/m^2dy$.
-The performance is notably better than for $k_{600}$ alone (Figure 4a) and there is no systematic bias in the predictions across all 47 rivers.
-This is presumably due to the structure of the $FCO_2$ equation, which reduces the relative importance of errors in *k* given that the $CO_2$ data is measured in situ.
-$FCO_2$ prediction intervals are slightly narrower than those presented in Figure 4a as well.
-
-![Figure 6: a: FCO2 via BIKER versus via equations 2 and 3 for every 11th timestep for the 49 rivers (grey lines are linear regression and 95% prediction intervals, while black dashed line is the 1:1 line).
-b-d: timeseries plots for the three example rivers from Figure 3b, 3c, and 3d.](cache/FCO2/FCO2_models.jpg)
-
-Figure 6b compares the bulk carbon efflux (via $F_{CO_2}$) from the 47 rivers using BIKER posterior means and four streamgauge-based HG models.
-The results are largely equivalent for both the mean and median carbon efflux.
-Both the BIKER estimate (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_BIKER',]$sumFCO2, 2)` gG-C/yr) and the 'Raymond 2013' estimate (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_raymond2013',]$sumFCO2, 2)`) are somewhat overestimated relative to the observed flux (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_obs',]$sumFCO2, 2)`).
-The 'Brinkerhoff 2019' model nearly perfectly recovers the observed value (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_brinkerhoff2019',]$sumFCO2, 2)`).
-Finally, the 'Raymond 2012' model slightly underestimates this bulk efflux (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_raymond2012',]$sumFCO2, 2)`).
-Thus, despite BIKER using absolutely no streamgauge or other in situ data like the other 3 methods do, it provides a reasonable estimate of the carbon efflux (Figure 6b).
-
-## 6 Discussion
-### 6.1 Gas exchange along the stream-to-ocean continumn
-Field studies of gas exchange in wide rivers have suggested that their *k* properties behave differently than in steeper and smaller rivers [@ulsethDistinctAirWater2019a; @beaulieuControlsGasTransfer2012a; @alinPhysicalControlsCarbon2011; @wangPhysicallyBasedScaling2021].
-While it has been shown that the small-eddy model is reasonably valid across most rivers [@wangPhysicallyBasedScaling2021], it has yet to be seen why and when form-drag dissipation is similar to near-surface dissipation in the context of *k*.
-In section 3 we show that the geomorphic concept of channel efficiency can explain when this is the case or not (Figure 3a).
-In extremely inefficient channels, the 'small-eddy' model is physically valid in flowing water environments as confirmed in Figures 3b and 3c.
-This has significant implications for how we understand shear and turbulence in rivers and their influence on *k*, as compared to environments where bottom-shear dominates (like lakes or oceans).
-
-@wangPhysicallyBasedScaling2021 proposed a conceptual model for fluvial *k* where *k* is dominated by bottom shear in small streams while in large rivers *k* is also affected by wind, biota, etc.
-This builds on the [@ulsethDistinctAirWater2019a] finding that steep, 'high-energy' streams have a fundamentally different $k_{600} \sim \epsilon_D$ relationship than less steep rivers and this may be due to the presence of bubble-mediated gas exchange and massive roughness elements in small mountain streams.
-Taking both of these models, along with our results, in aggregate, we suggest a combined conceptual model that follows the geomorphic concept of the 'river longitudinal profile' and is outlined in Figure 7.
-Small streams are steep, rough, and highly efficient channels that evacuate all but the largest sediment leaving a very rough riverbed.
-These small streams likely have strong *k*~hydraulics relationships and surface turbulence is dominated by their huge bed roughness.
-However, as one moves down the stream-to-ocean continumn, river channels get wider, flatter, relatively more frictional, and far less efficient at evacuating sediment.
-This means there is a concurrent fining of the riverbed and an increase in form-drag.
-In these largest rivers, form-drag starts dominating the *k*~hydraulics relationship rather than bed roughness.
-While wind certainly also plays a role in *k* in these large rivers [particuarly at higher winds- @beaulieuControlsGasTransfer2012a], we show here that for maximally inefficient river channels form-drag alone can reasonably predict $k_{600}$.
-The relationship between form-drag dissipation and wind is left to future work.
-
-![Figure 7: Conceptual model of the river longitudinal profile as it pertains to gas exchange velocity.
-As rivers grow in size and one moves downstream, their channels become wider, flatter, less efficient at transporting sediment, and have greater relative frictional resistance.
-This suggests a shift in k~hydraulics relationships from one dominated by bed roughness in small, steep, and efficient systems versus one dominated by form-drag in large, flat, and inefficient systems.](cache/k600_theory/longProfile.jpg)
-
-### 6.2 Towards remote sensing of global spatiotemporal dynamics of *k* in large rivers
-To date, most field-scale studies of riverine gas exchange have focused on 1) its relationship with wind speed [e.g @beaulieuControlsGasTransfer2012a; @borgesVariabilityGasTransfer2004; @zappaEnvironmentalTurbulentMixing2007], 2) average flow velocity [e.g.
-@alinPhysicalControlsCarbon2011; @beaulieuControlsGasTransfer2012a; @schelkerCO2EvasionSteep2016a], or 3) discharge [e.g.
-@wangPhysicallyBasedScaling2021; @robertsMultipleScalesTemporal2007; @uehlingerEcosystemMetabolismDisturbance1998].
-However, the spatiotemporal dynamics of riverine gas exchange are still weakly constrained.
-A few studies have investigated these dynamics, but they have been limited to individual rivers and/or limited field seasons [@hallAirWaterOxygen2012; @sand-jensenCO2DynamicsDanish2012].
-@wallinSpatiotemporalVariabilityGas2011 performed a preliminary analysis in northern Sweden relating cross-section specific temporal variability in gas exchange with channel slope, but they were limited to an average of only 8 measurements per river in a single watershed.
-This limited knowledge of large-scale *k* spatiotemporal dynamics is due both to a lack of process-level understanding [@hallGasExchangeStreams2020] but also a lack of measurements.
-
-Therefore, estimating *k* from SWOT data is an attractive option for exploring its spatiotemporal dynamics at fine temporal resolution and at the global-scale.
-SWOT will provide daily hydraulic measurements for a 3 month fast sampling period for calibration and validation and sampling thereafter between 1 and 7 days per 21 day repeat cycle [@biancamariaSWOTMissionIts2016].
-BIKER's success in 1) infering $k_{600}$ using simulated SWOT data over a wide range of rivers (Figures 4 and 5) and 2) being robust to measurement errors internal to the SWOT data (Figure 5) bode well for BIKER's eventual implementation on real SWOT observations.
-Further, BIKER's improved performance with longer datasets (Figure 5b) bodes well for future SWOT implementation, as the three-year campaign will provide well north of 600 observations for many rivers, suggesting even better BIKER inversions than those presented here.
-All of this suggests that daily estimation of riverine gas exchange globally could be possible once SWOT launches.
-
-### 6.3 Towards remotely sensing riverine carbon emissions using SWOT
-Figure 6 confirms that BIKER is quite successful, without any in situ information aside from a $CO_2$ logger, at predicting both $FCO_2$ (Figure 6a) and 2) the bulk carbon efflux (Figure 6b).
-This encouraging result has two main implications for future work coupling remote sensing via SWOT with in situ data.
-First, it confirms that we can couple BIKER and SWOT with in situ gas concentration loggers to produce $FCO_2$ estimates at novel temporal resolutions in SWOT-observable rivers.
-This is particualry useful given recent advances in in situ gas concentration loggers [@ahoHighestRatesGross2021] but no such similar advances in modeling *k* at equivalent temporal resolutions.
-BIKER can likely also be ran at the field scale using arrays of pressure transducers to estimate water surface slope (rather than using satellite-based altimeters like SWOT) following recent work doing the same using the @hagemannBAMBayesianAMHGManning2017a RSQ algorithm [@harlanDischargeEstimationDense2021].
-
-Secondly, it is important to stress that unlike BIKER, the HG models in Figure 6b rely on an in situ streamgauge.
-This means that Figure 6b represents the best performance that those models could ever have; if ran using modeled discharge their accuracy would necessarily decrease.
-Thus, these results suggest that BIKER will be useful in two settings: 1) upscaling in ungauged rivers as hypothesized, but also in 2) potentially improving our carbon efflux understandings at gauged sites.
-Future work should systematically quantify prediction error from coupling global-scope HG models with modeled discharge, as is the default workflow used in fluvial $FCO_2$ upscaling studies [e.g.
-@lauerwaldSpatialPatternsCO22015; @horgbyUnexpectedLargeEvasion2019; @raymondGlobalCarbonDioxide2013a; **Liu et al in reivew**].
-
-Figure 6b also confirms that the training data used for HG models exerts a significant influence on upscaled carbon emissions from rivers.
-In Figure 6b there is a nearly 0.5 Tg-C range between estimates, which is significant and nearly entirely a function of the data used to fit these depth HG models.
-In this context, the 'Brinkerhoff 2019' model likely outperforms both 'Raymond models' because the training data is orders of magnitude larger and more geomorphically varied than those used in the 'Raymond' models (530,945 measurements versus 1,026 and 10837 measurements).
-Meanwhile, BIKER has no similar reliance on hydraulic parameters trained on different datasets and only assumes that *dA* can be calculated by assuming a rectangular river channel.
-Upon SWOT's launch, the BIKER approach to estimating *k* could be coupled with ethier existing upscaling workflows or even explicit $CO_2$ transport models [@brinkerhoffLakeMorphometryRiver2021; **Saccardi & Winnick in review**] to improve riverine gas flux predictions where gauges are unavailable but SWOT measurements are.
-This coupling could potentially be done using data assimilation techniques, which have proven very useful for similar objectives in recent RSQ work [@ishitsukaCombiningOpticalRemote2020].
-
-## 7 Conclusions
-Gas exchange from aquatic systems has been studied for nearly a century, with a robust model (equation 2) that has been repeatedly verified across many non-fluvial environments.
-Despite renewed interest in fluvial gas exchange in the last decade or so, there are considerable uncertainties in how we model the gas exchange velocity in river environments and particularly in large rivers where wind increasingly influences *k*.
-Here, we show algebraically that TKE budgets in big rivers exist in a local equilbrium at the free surface and that we can equivalently scale *k* using any characteristic turbulence scale.
-We validate this using over 530,000 measurements of river channel hydraulics and over 700 discrete measurements of the normalized gas exchange velocity $k_{600}$.
-We then exploit this finding to remotely sense $k_{600}$ using simulated data that will be provided by the forthcoming SWOT satellite, showing good performance and relying on absoutely no on-the-ground information.
-Finally, we pair this algorithm (named 'BIKER') with an in situ $CO_2$ logger and show strong performance in reproducing $CO_2$ evasion fluxes and carbon efflux from the rivers' surfaces.
-
-These strong results functionally open the door for global-scale, near daily estimates of fluvial gas exchange velocity once SWOT launches in 2022.
-This unprecedented amount of data should allow for significant insights into both the mechanistic and temporal dynamics of fluvial gas exchange in large rivers around the world.
-This, in turn, should allow us to better parameterize upscaling workflows such that the global fluvial carbon flux is better constrained.
-
-## 8 Acknowledgements
-C.B.
-Brinkerhoff was funded on **FINESST babyyyy**.
-BIKER is available at https://github.com/craigbrinkerhoff/BIKER.
-All code to build and generate results, figures, and the manuscript is available at https://github.com/craigbrinkerhoff/RSK600.
-We thank Renato Frasson, Michael Durand, Amber Ulseth, and Jake Beaulieu for generously making their data available for this study.
-We also thank the entire SWOT discharge working group for their decade-plus body of work which inspired this study.
-
-## 9 Apendix A
-*Table A1: Variable description and notation for this study.*
-
-|**Notation**|**Description**|**Calculation (if applicable)**|**Units**|
-|------------|---------------|-------------------------------|---------|
-|*A*|Channel cross-sectional area|NA|$[L^2]$|
-|*Bulk carbon efflux*|carbon mass transport rate from river to atmosphere|NA|$[\frac{M}{T}]$|
-|$CO_{2_{water}}$|Water-side $CO_2$ concentration|NA|$[L]$|
-|$CO_{2_{air}}$|Atmospheric-side $CO_2$ concentration|NA|$[L]$|
-|$D_m$|Molecular diffusion coefficient|NA|$[\frac{L^2}{T}]$|
-|$\epsilon_S$|Dissipation rate of near-surface turbulence|log-law-of-the-wall model: $\frac{U_*^3}{H}$|$[\frac{E}{M*T}]$|
-|$\epsilon_D$|Form-drag turbulent dissipation rate|$gSU$|$[\frac{E}{M*T}]$|
-|$FCO_2$|$CO_2$ evasion flux from river to atmosphere|NA|$[\frac{M}{L^2T}]$|
-|$g$|gravitational acceleration|9.8|$[\frac{L}{T^2}]$|
-|$G$|TKE production rate|$\frac{UU_*}{H}$|$[\frac{E}{M*T}]$|
-|$H$|Mean flow depth|$A/W$|$[L]$|
-|$k$|gas exchange velocity|NA|$[\frac{L}{T}]$|
-|$k_{600}$|gas exchange velocity normalized to $Sc=600$|$(\frac{600}{Sc_k})^{-0.5}k$|$[\frac{L}{T}]$|
-|$R_h$|Hydraulic radius|$\frac{AW}{2H+W}$|$[L]$|
-|$S$|River slope|NA|$[\frac{L}{L}]$|
-|$Sc$|Schmidt number|$\frac{v}{D_m}$|$unitless$|
-|$T_d$|TKE turbulent diffusion rate|NA|$[\frac{E}{M*T}]$|
-|$U$|Reach-averaged flow velocity|NA|$[\frac{L}{T}]$|
-|$U_*$|Shear velocity|$\sqrt{gSR_h}$|$[\frac{L}{T}]$|
-|$v$|kinematic viscosity|NA|$[\frac{L^2}{T}]$|
-|$W$|Flow width|NA|$[L]$|
-
-## References

### 2021-09-30 manuscript_v5.Rmd
model update to the chainsaw model

@@
 ---
-title: "Gas exchange in large rivers influenced by hydraulic geometry: implications for remotely sensing gas exchange via the SWOT satellite"
+title: "Gas exchange in large rivers influenced by largest turbulent eddies: implications for remotely sensing gas exchange via the SWOT satellite"
 author:
@@
 ## Key Points
-- River channel efficiency effects the relationship between gas exchange and dissipation of turbulence
-- Wide, less "efficient" river channels exert greater influence on near-surface turbulence than narrower, more efficient ones
+- Both large and small scale eddies control gas exchange in hydraulically wide rivers
 - BIKER algorithm exploits this theory, predicting gas exchange velocity and $CO_2$ fluxes from simulated SWOT satellite data and in situ $CO_2$ data
+- BIKER is also robust to measurement errors internal to the SWOT river measurements
 
@@
 ## Abstract
-*AGU Advances. So 250 words here* <br>
-*AGU advances is 8000 words total, the format seems to be a long letter so I've tried to right this in that style.*
+UPDATE
+
+Extensive research over the past two decades has shown that the global river network emits significant amounts of carbon dioxide ($CO_2$) to the atmosphere via gas evasion, a diffusive flux across the air-water interface.
+This global flux is still relatively poorly constrained, with much ambiguity coming from mechanistic and predictive uncertainties in the gas exchange velocity (the rate of this diffusion).
+We propose that the NASA/CNES/USKA/CSA SWOT satellite, set to usher in a new age of river hydrology research at the global scale, will provide similarly transformative insights to fluvial gas exchange modeling.
+Here, we present work predicting gas exchange velocity using SWOT observations.
+We exploit the unique geomorphology of SWOT-observable rivers to develop a process-based model that is nearly entirely remotely sensible.
+We then couple this model with established Bayesian techniques to develop BIKER, or the 'Bayesian Inference/Inversion of the k600 Exchange Rat' algorithm.
+We validate BIKER on 47 SWOT-simulated rivers and over 700 discrete measurements of river gas exchange, yielding 1) a theoretical model that explains 64% of variation in gas exchange in SWOT-observable rivers and 2) an algorithm that predicts this theoretical model solely from SWOT observations with a median normalized RMSE of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$nrmse), 2)`.
+Additionally, by-river performance improves as the amount of available SWOT data increases.
+BIKER is also robust to measurement errors implicit in the SWOT data, with a median normalize RMSE of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==1,]$nrmse), 2)`.
+With SWOT set to launch in 2022, we suggest that BIKER might address the lack of data on gas exchange in large rivers and help to better constrain current and future fluvial gas evasion models at the global scale.
 
@@
 Natural systems play a fundamental role in the budgeting and accounting of the global carbon cycle under climate change.
-Since the publication of @colePlumbingGlobalCarbon2007a, the global river network is recognized to emit substantial amounts of carbon to the atmosphere, in addition to exporting it to the oceans.
-Current estimates of total carbon dioxide evasion ($FCO_2$) to the atmosphere from the global river network vary from 650-1800 Tg C/yr [@lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a] **maybe add Shaoda here**, with 167 Tg-C/yr coming from mountain streams alone [@horgbyUnexpectedLargeEvasion2019].
-Despite its incredibly small percentage of the global land surface [0.47%- @raymondGlobalCarbonDioxide2013a], this flux is on par with the total oceanic $CO_2$ uptake rate [@gruberOceanicSinkAnthropogenic2019; @horgbyUnexpectedLargeEvasion2019] and the global forest carbon uptake rate [@panLargePersistentCarbon2011].
+Since the publication of @colePlumbingGlobalCarbon2007, the global river network is recognized to emit substantial amounts of carbon to the atmosphere, in addition to exporting it to the oceans.
+Current estimates of total carbon dioxide evasion ($FCO_2$) to the atmosphere from the global river network vary from 650-1800 Tg C/yr [@lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013] (*Liu etal in review*), with 167 Tg-C/yr coming from mountain streams alone [@horgbyUnexpectedLargeEvasion2019].
+Despite its incredibly small percentage of the global land surface [0.47%- @raymondGlobalCarbonDioxide2013], this flux is on par with the total oceanic $CO_2$ uptake rate [@gruberOceanicSinkAnthropogenic2019; @horgbyUnexpectedLargeEvasion2019] and the global forest carbon uptake rate [@panLargePersistentCarbon2011].
 It is still relatively poorly constrained but is clearly a critical component of the global carbon cycle.
@@
 @bernhardtMetabolicRegimesFlowing2018] and water quality engineers modeling river responses to pollutant loadings [among other things- @chapraSurfaceWaterQualityModeling2008].
-Because of this far-reaching importance, decades of work have focused on elucidating the physical mechanisms behind *k* in oceans and lakes [@wanninkhofAdvancesQuantifyingAirSea2009] and, to a lesser extent, fluvial gas exchange [@hallGasExchangeStreams2020].
+Because of this far-reaching importance, decades of work have focused on elucidating the physical mechanisms behind *k* in oceans and lakes [@wanninkhofAdvancesQuantifyingAirSea2009 for a review] and, to a lesser extent, fluvial gas exchange [@hallGasExchangeStreams2020 for a review].
 Given that *k* should scale with near-surface turbulence in a turbulent flow [@hallGasExchangeStreams2020], extensive field and labratory experiments have converged on the 'small-eddy model' as proposed by @lamontEddyCellModel1970, which scales *k* via the smallest-scale turbulent eddies [e.g.
-@lorkeUnifiedScalingRelation2006; @zappaEnvironmentalTurbulentMixing2007; @zappaVariationSurfaceTurbulence2003; @moogStreamReaerationNonuniform1999; @wangCoefficientsSmallEddy2015; @vachonRelationshipNearsurfaceTurbulence2010].
-This model is provided as equation 2, where $\epsilon_S$ is the dissipation rate of near-surface turbulence, $v$ is the kinematic viscosity and *Sc* is the Schmidt number.
-Note that we directly measuring $\epsilon_S$ is non-trivial and though a frequent model is to impose the log-law-of-the-wall to scale bottom shear to the surface, this necessiarly assumes that bed shear alone effects turbulence which is often not the case.
-$$\mathbf{(2)} k\sim Sc^{-n}(v\epsilon_S)^{1/4}$$
+@lorkeUnifiedScalingRelation2006; @zappaEnvironmentalTurbulentMixing2007; @zappaVariationSurfaceTurbulence2003; @moogStreamReaerationNonuniform1999; @vachonRelationshipNearsurfaceTurbulence2010].
+This model is provided as equation 2, where $\epsilon_S$ is the dissipation rate of near-surface turbulence, $v$ is the kinematic viscosity, *Sc* is the Schmidt number.
+Note that directly measuring $\epsilon_S$ is non-trivial and a frequent model is to impose the log-law-of-the-wall to scale bottom shear to the surface (Appendix A).
+
+$$\mathbf{(2)} k\sim Sc^{-1/2}(v\epsilon_S)^{1/4}$$
 
@@
 @lorkeUnifiedScalingRelation2006; @zappaEnvironmentalTurbulentMixing2007], there is considerable uncertainity in how it applies to fluvial systems.
-First, 'bubble-mediated gas exchange' in whitewater might lead to substantially higher *k* in small mountain streams with very large bedload [@hallAirWaterOxygen2012; @ulsethDistinctAirWater2019a].
-Additionally, 'form-drag shear' via river channel banks, meanders, bars, etc.
-further complicate scaling fluvial *k* [@moogStreamReaerationNonuniform1999].
-Those authors proposed modeling fluvial *k* via form-drag dissipation $\epsilon_D$, assuming that total reach stream power is responsible for all turbulence production and therefore dissipation.
-Numerous authors have since shown that this model ($\epsilon_D=gS\overline{U}$) reasonably predicts *k* in rivers [@raymondScalingGasTransfer2012a; @ulsethDistinctAirWater2019a; @moogStreamReaerationNonuniform1999], though it should be stressed that $\epsilon_S \neq \epsilon_D$.
-This is because $\epsilon_D$ refers to a depth-scale disispation rate and is not necessairily indicative of the near-surface dissipation rate.
-Consequently, the, $k\sim\epsilon_D$ relationship usually has a slope much steeper than the 1/4 in equation 2.
-
+First, 'bubble-mediated gas exchange' in whitewater might lead to substantially higher *k* in small mountain streams with very large bedload [@hallAirWaterOxygen2012; @ulsethDistinctAirWater2019].
+Another approach is to model *k* via 'form-drag shear' (i.e.
+total stream power per unit mass water) that captures frictional resistance via river channel banks, meanders, bars, etc.
+[@moogStreamReaerationNonuniform1999].
+Authors have since shown that this model reasonably predicts *k* in rivers [@raymondScalingGasTransfer2012; @ulsethDistinctAirWater2019], though it requires good estimates of the average flow velocity.
 Compared to small streams, relatively little attention has been paid to gas exchange in very large rivers.
 Large rivers are often conceptualized as a hybrid condition of both fluvial and non-fluvial *k* dynamics.
-To date, the handful of existing field studies of large-river *k* suggest that *k* begins to be influenced by wind shear once the water surface is less protected, though little else is well-established [@alinPhysicalControlsCarbon2011; @beaulieuControlsGasTransfer2012a; @wangPhysicallyBasedScaling2021].
-This is also presumably limited to forested environments that provide shelter from wind.
+The handful of existing field studies of large-river *k* suggest that *k* begins to be influenced by wind shear once the water surface is less protected, though little else is well-established [@alinPhysicalControlsCarbon2011; @beaulieuControlsGasTransfer2012; @wangPhysicallyBasedScaling2021].
+These mechanistic uncertainties are additionally limited by a dearth of field-measured fluvial *k*.
+@wangPhysicallyBasedScaling2021 attempted to address this by simulating *k* in 35 rivers using a stream metabolism model [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets to implicitly solve for *k*.
+They found that equation 2, coupled with $\epsilon$ modeled via the log-law-of-the-wall, is valid in their simulated rivers and that *k*~streamflow relationships break down in large rivers.
+However, their simulated rivers were limited by a lack of hydraulic measurements and had to rely on hydraulic geomtery models to estimate river depth and velocity.
+Further, they did not assess model fit in very steep systems [@ulsethDistinctAirWater2019] and stopped short of parsing out hydraulic explanations for their results.
 
-These mechanistic uncertainties are additionally limited by a large dearth of field-measured fluvial *k*.
-@wangPhysicallyBasedScaling2021 attempted to address this by simulating *k* in 35 rivers using a stream metabolism model [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets.
-They found that equation 2, coupled with $\epsilon$ modeled via the log-law-of-the-wall, is valid in their simulated rivers and that *k*~streamflow relationships break down in large rivers.
-However, they did not directly compare against *k* measurements made in extremely small systems [@ulsethDistinctAirWater2019a] and stopped short of parsing out hydraulic explanations for their results.
-
-Finally, these mechanistic uncertainties are then propogated through upscaling workflows when biogeochemists predict *k* across thousands of rivers [e.g @borgesGloballySignificantGreenhousegas2015a; @horgbyUnexpectedLargeEvasion2019; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a] via equation 1 coupled with hydraulic geometry [HG: the scaling relationships between streamflow and river channel hydraulics- @leopoldHydraulicGeometryStream1953a].
+These mechanistic uncertainties are then propogated through upscaling workflows when biogeochemists predict *k* across thousands of rivers [e.g @borgesGloballySignificantGreenhousegas2015; @horgbyUnexpectedLargeEvasion2019; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013] via equation 1 coupled with hydraulic geometry [HG: the scaling relationships between streamflow and river channel hydraulics- @leopoldHydraulicGeometryStream1953].
 It is currently not well understood how sensitive global estimates of fluvial gas evasion are to the specific HG model that is employed by the worker.
 Further, these approaches rely on either in situ discharge records or modeled streamflow which introduces additional uncertainities.
-This is all exacerbated in ungauged basins that cover large areas, especially in the carbon-rich Arctic inland waters, where little in situ information is available and fieldwork is impractical [@gleasonRemoteSensingRiver2020].
+This is all exacerbated in ungauged basins that cover large areas, especially in the carbon-rich Arctic inland waters, where little in situ information is available and fieldwork is impractical [@gleasonRemoteSensingRiver2019].
 
 A potential alternative to this upscaling approach is to directly estimate a river's hydraulic properties from remote sensing (RS) data.
-Remote sensing of river hydraulics is a burgeoning subfield within remote sensing of hydrology, often in service of remote sensing of river discharge [RSQ- @gleasonRemoteSensingRiver2020].
-This is accomplished via two general approaches: 'gauged' methods which rely on in situ river data to calibrate one's method to the river(s) at hand and 'ungauged' techniques which focus on hydraulic generalizability in the service of merely improving existing knowledge in data-poor domains [@gleasonRemoteSensingRiver2020].
+Remote sensing of river hydraulics is a burgeoning subfield within remote sensing of hydrology, often in service of remote sensing of river discharge [RSQ- @gleasonRemoteSensingRiver2019].
+This is accomplished via two general approaches: 'gauged' methods which rely on in situ river data to calibrate one's method to the river(s) at hand and 'ungauged' techniques which focus on hydraulic generalizability in the service of merely improving existing knowledge in data-poor domains [@gleasonRemoteSensingRiver2019].
 Many, but not all, of these ungauged approaches are developed in the context of the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission.
@@
 A decade of SWOT work has explored the multi-parameter problem of estimating the river hydraulic parameters of roughness and bathymetry from remote sensing to produce the SWOT discharge product [e.g.
-@andreadisConstrainingAssimilationSWOT2020d; @brinkerhoffConstrainingRemoteRiver2020a; @durandEstimatingReachaveragedDischarge2014a; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014a; @garamboisVariationalEstimationEffective2020; @brissetAssimilationAltimetricData2018; @hagemannBAMBayesianAMHGManning2017a; @larnierRiverDischargeBathymetry2020a; @oubanasDischargeEstimationUngauged2018].
+@andreadisConstrainingAssimilationSWOT2020; @brinkerhoffConstrainingRemoteRiver2020; @durandEstimatingReachaveragedDischarge2014; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014a; @garamboisVariationalEstimationEffective2020; @brissetAssimilationAltimetricData2018; @hagemannBAMBayesianAMHGManning2017; @larnierRiverDischargeBathymetry2020a; @oubanasDischargeEstimationUngauged2018].
 
-In this context, here we revist the fundamental mechanisms behind gas exchange and turbulence in large rivers to answer the following question: does *k*'s relationship with form-drag fundamentally change along the stream-to-river continumn and how does this relate to one of the classical models of gas evasion in aquatic systems (equation 2)?
-We exploit the findings from this simple analysis to develop a novel methodology that predicts $k_{600}$ (*k* normalized to a Schmidt number of 600) and its explicit uncertainity solely using SWOT observations.
+In this context, here we borrow tools from fluvial geomorphology and revist the fundamental mechanisms behind gas exchange in large rivers to answer the following question: does the small-eddy model work in large rivers like those that SWOT will observe?
+We exploit the findings from this analysis to develop a novel methodology that predicts $k_{600}$ (*k* normalized to a Schmidt number of 600) and its explicit uncertainity solely using SWOT observations.
 The method requires no in situ inputs of any kind (although in situ data can be ingested and will improve results).
@@
 Not mapped here are over 170,000 discrete measurements of river channel hydraulics from across the continetal United States (Brinkerhoff et al.
-2019) and over 700 gas exchange velocity measurements (Ulseth et al.
-2019).](cache/maps/map_fin.jpg)
+2019) and 763 gas exchange velocity measurements (Ulseth et al.
+2019; Raymond et al.
+2012; Churchill et al.
+1964; Owens et al.
+1964).](cache/maps/map_fin.jpg)
 
-Gas exchange theory (section 3) is explored using the dataset from @ulsethDistinctAirWater2019a who measured and/or collected from the literature over 700 measurements of stream hydraulics and $k_{600}$ and is, to our knowledge, the largest collection of field-measured river and stream $k_{600}$.
+Gas exchange theory (section 3) is explored using multiple datasets which measured and/or collected from the literature 763 measurements of stream hydraulics and $k_{600}$.
+This is, to our knowledge, the largest collection of field-measured river and stream $k_{600}$.
+701 of these measurements were previously collected from other studies by @raymondScalingGasTransfer2012 and @ulsethDistinctAirWater2019.
+The remaining measurements were added by us but measured by @churchillPredictionStreamReaeration1964 and @owensReaerationStudiesStreams1964.
+See Table S1 for a complete list of the studies that collected these measurements.
 To assess general hydraulic conditions in rivers across geographic contexts, we additionally use a previously published compilation of field hydraulics measurements [@brinkerhoffReconcilingAtaStationAtManyStations2019].
@@
 These simulated rivers are simply river-reach-averaged hydraulic model outputs where the water surface heights and widths are labelled as RS observations and are used as the sole inputs to BIKER.
-Here, we use 47/51 rivers collected by @frassonExploringFactorsControlling2021 and @durandIntercomparisonRemoteSensing2016b.
+Here, we use 47/51 rivers collected by @frassonExploringFactorsControlling2021 and @durandIntercomparisonRemoteSensing2016.
 These are the two benchmarking studies that have explored RSQ algorithm performance for the SWOT mission.
@@
 Ultimately, the 47 rivers are spread across the United States, France, Italy, the United Kingdom, and Bangladesh.
-We omit three models from @durandIntercomparisonRemoteSensing2016b because they lack enough hydraulic information to calculate the shear velocity $U_\star$, which is necessary for algorithm validation and is explained in section 4.
+We omit three models from @durandIntercomparisonRemoteSensing2016 because they lack enough hydraulic information to calculate the shear velocity $U_\star$, which is necessary for algorithm validation and is explained in section 4.
 These are two models for the Saint Lawrence River and one for the Tanana River.
 
-To assess the influence of measurement error on BIKER's performance (section 4), we use the error model developed by @durandHowWillRadar2020a and implemented on 17/47 of the rivers by @frassonExploringFactorsControlling2021.
+To assess the influence of measurement error on BIKER's performance (section 4), we use the error model developed by @durandHowWillRadar2020 and implemented on 17/47 of the rivers by @frassonExploringFactorsControlling2021.
 Error in SWOT measurements will come from both the error tolerances intrinsic in the satellite data product as well as radar layover error.
-Layover error is the phenomenon when radar returns from different places arrive at the sensor at the same time, leading to taller landscape features appearing closer to the sensor than shorter landscape features that are the same horizontal distance from the sensor [@durandHowWillRadar2020a].
+Layover error is the phenomenon when radar returns from different places arrive at the sensor at the same time, leading to taller landscape features appearing closer to the sensor than shorter landscape features that are the same horizontal distance from the sensor [@durandHowWillRadar2020].
 Width errors due to poor water classification are ignored as they were in @frassonExploringFactorsControlling2021.
 
-For the $CO_2$ evasion and carbon efflux calculations (section 5), we use 26 bi-weekly dissolved $CO_2$ samples made by @beaulieuControlsGasTransfer2012a at one location in the Ohio River for one calender year from 2008-2009 (Figure S2).
+For the $CO_2$ evasion and carbon efflux calculations (section 5), we use 26 bi-weekly dissolved $CO_2$ samples made by @beaulieuControlsGasTransfer2012 at one location in the Ohio River for one calender year from 2008-2009 (Figure S2).
 Note that this $CO_2$ data is for the Ohio River only but was applied to all 47 rivers (which includes multiple sections of the Ohio River).
@@
 
-## 3 Form-drag in large rivers and implications for scaling fluvial gas exchange velocity
-Turbulence is fundamental to fluvial geomorphology and open-channel flows and has been the subject of extensive research across scientific and engineering disciplines for over a century.
-As noted in the introduction, turbulence in river flows is usually generated by shear forces at the riverbed, but also across the entire flow depth via 'form-drag' from riverbanks, meanders, bars, suspended sediment, etc [@moogStreamReaerationNonuniform1999].
-This concept is synonymous with the geomorphic idea of available stream power, or the total energy dissipated from a river reach via the liquid's potential energy along its slope [originally derived from first-principles by @bagnoldSedimentDischargeStream1960; @bagnoldApproachSedimentTransport1966].
-Both @raymondScalingGasTransfer2012a and @moogStreamReaerationNonuniform1999 confirmed that $\epsilon_D$ (simply the available stream power normalized by the water mass) is far greater than $\epsilon_S$ in open-channel flows, therefore generating the majority of fluvial TKE dissipation.
+## 3 Hydraulic geometry in large rivers and implications for scaling fluvial gas exchange velocity
+Rivers and streams change predictably along their longitudinal profile from headwater to ocean.
+In general, as river size increases, channels become more rectangular, their shapes elongate (becoming wider quicker than they become deeper) and their hydraulic radii begin to approximate their mean flow depth (i.e.
+$R_h \approx H$).
+This is a common assumption in hydraulic and geomorphic modeling of large rivers, and occurs in `r round(percs[percs$flag_swot == 'SWOT' & percs$flag_hydraulicWide == 'Rh=H',]$perc,0)`% of our SWOT-observable river measurements but only `r round(percs_overall[percs_overall$flag_hydraulicWide == 'Rh=H',]$perc,0)`% of all `r sum(percs$n)` river measurements (see Text S1 for how we arrived at these numbers).
+These channels are often referred to as 'hydraulically wide'.
 
-While form-drag dissipation is effectively an alternative representation of reach-scale stream power, little attention has been paid to the resulting scale-dependecy of $\epsilon_D$.
-Available stream power, and thus 'form-drag' shear, exists at the depth-scale and is scale-dependent: while small roughness elements near the bed produce form-drag, if they are small relative to the total flow depth they will exert little influence on near-surface dissipation [@moogStreamReaerationNonuniform1999].
-This means that, if all other properties of a river flow are held constant but the channel is progressively made wider and shallower, its depth will decrease and its wetted perimeter (and therefore total frictional resistance) will increase.
-This concept is illustrated graphically in Figure 2 and relates the geomorphic concept of 'river channel efficiency' to turbulence: all else being constant, a less efficient channel (wide/shallow) will have greater frictional resistance and therefore form-drag (Chow, 1959).
-As shown via the orange arrows in Figure 2, the inefficient channel should 1) produce more turbulence, and 2) turbulence generated via form-drag should more readily reach the surface (the red arrows in Figure 2).
-Note that to compare wetted perimeters across many rivers with different channel areas, we use the hydraulic radius $R_h$ (channel area divided by the wetted perimeter) as compared to the mean flow depth $H$ (channel area divided by the width).
-An efficient channel will have a low $\frac{R_h}{H}$ ratio while an inefficient channel will have a high $\frac{R_h}{H}$ ratio that approaches 1.
+To model gas exchange velocity in hydraulically wide channels, we make some simplifying assumptions to equation 2.
+First, we impose the log-law-of-the-wall model for $\epsilon_S$ (Appendix A) on to the small-eddy model (equation 2) and simplify $R_H = H$, arriving at equation 3.
+Equation 4 further simplifies this for *k* normalized to a Schmidt number of 600, and with statistical parameter $\alpha$ reflecting both the scaling as well as the constant Schmidt number and viscosity [@wangPhysicallyBasedScaling2021].
+Equation 4 thus defines gas exchange velocity solely as a function of water surface slope and average flow depth without a reliance on shear velocity, mean flow velocity, or $\epsilon_S$ at all.
+Again, this is hypothetically only valid in a wide channel.
 
-![Figure 2: Conceptual model relating river channel efficiency to the turbulent dissipation rate.
-All else being constant, a less efficient channel shape (wide/shallow) should 1) have more form-drag and thus turbulence and 2) turbulence generated via form-drag should more readily reach the surface.
-This is illustrated by the orange arrows, where the inefficient channel arrows more frequently reach the surface (and influence surface turbulence- red arrows).](cache/k600_theory/conceptual_model.jpg)
+$$\mathbf{(3)} k \sim Sc^{-1/2}v^{1/4}(gS)^{3/8}H^{1/8}$$
+$$\mathbf{(4)} k_{600}=\alpha(gS)^{3/8}H^{1/8}$$
 
-This all suggests that, for incredibly inefficient river channels, $\epsilon_S \approx \epsilon_D$.
-The channel produces so much turbulence from form-drag, and the flow is so shallow relative to its width, that depth-scale dissipation should be nearly synonymous with near-surface dissipation and equation 2 could be equivalently written as $k\sim Sc^{-n}(v\epsilon_D)^{1/4}$.
-To explore when (if at all) this occurs in rivers, we fit models using 701 river $k_{600}$ measurements from @ulsethDistinctAirWater2019a.
-First, we defined 'channel inefficiency' using minimum $\frac{R_h}{H}$ thresholds ranging from 0.75 to 0.995 and then we fit the two models outlined in equations 3 and 4 (because all of the data are normalized to Schmidt numbers of 600, we convienantly fold the viscosity and Schmidt numbers into the model coefficients).
-The goal of this analysis is to find at what $\frac{R_h}{H}$ threshold does 1) equation 3's b parameter approximate 1/4 and thus mirror equation 2 and 2) equation 4's goodness-of-fit (via the coefficient of determination $r^2$) reach a suitably high value that it is deemed physically valid.
-$$\mathbf{(3)} k_{600}=a(\epsilon_D)^{b}$$
-$$\mathbf{(4)} k_{600}=\beta(\epsilon_D)^{1/4}$$
+Alternatively, @moogAirWaterGasTransfer1999 noted that low-turbulence uniform channel flows often do not exhibit homogenous surface dissipation at their air-water interface.
+They proposed a conceptual 'chainsaw model' that relies on both small-scale eddies and large, depth-scale ones.
+It limits surface dissipation to only 'active zones' within which a homogenous dissipation rate occurs and the small-eddy model is valid.
+But, the fraction of the interface that are active zones is proportional to the larger, depth-scale eddies.
+This means that if the entire interface is treated as an active zone, the model would simply reduce to the small-eddy model.
+We chose to also test this model given that SWOT-observable rivers are generally large and low-turbulence (here, low-turbulence is realitive to the highly-turbulent water surfaces of small, steep streams with significant whitewater).
+Further, to our knowledge the chainsaw model has never been empirically tested, and so we explore it here.
+To do this, we impose a hydraulically wide channel on the chainsaw model just like we did to the small-eddy model and arrive at equation 5 (with statistical coefficient $\beta$).
+The result is identical to the hydraulically-wide small-eddy model (equation 4), except for the exponents, which are both steeper.
 
-The results of this analysis are plotted in Figure 3a, along with channel geometry schematics to highlight the progressive shift from efficient to inefficient river channels.
-Equation 3's b parameter only begins to approximate 1/4 when $R_h$ is within 1% of the mean flow depth.
-At this point, the model's goodness-of-fit also reaches a reasonable value of `round(models_eD4[models_eD4$name == 'Rh=H',]$r2,2)`.
-This indicates that a river is sufficiently inefficient when its hyradulic radius approximates the mean flow depth, and at that point $\epsilon_S \approx \epsilon_D$.
-Figures 3b and 3c plot the model fits for this final scenario where $\frac{R_h}{H} >= 0.995$ As Figure 3c confirms, the small-eddy model is viable when a channel is so inefficient that dissipation via form-drag approximates the near-surface dissipation rate in equation 2.
-Figure 3b further confirms this, where the slope of the generalized model (equation 3) equals `r round(models_ulseth[models_ulseth$name == 'Rh=H',]$slope, 2)` which is nearly equivalent to 1/4.
+$$\mathbf{(5)} k_{600}=\beta(gS)^{9/16}H^{11/16}$$
 
-![Figure 3: Model development for predicitng k from form-drag dissipation in rivers.
-(a) Rh/H thresholds versus fitted model results.
-Dark blue is the fitted b parameter in equation 3 and dark green is the goodness-of-fit of equation 4 as defined by the coefficient of determination.
-Once a channel is maximally inefficient, i.e.
-Rh=H, equation 4 is a viable model for k and equation 3 empirically recovers the theoretical value of 1/4 as outlined in equation 2.
-(b-d) Fitted model results for inefficient channels (defined as Rh/H >= 0.995) predicting k via the shear velocity U*, equation 4, and equation 3 (respectively).
-Note that colors align to the same models for subpanels a, c, and d.
-Further note that the coefficient of determination in subpanel (d) refers to log-transformed residuals and hence is not directly comparable to those in subpanels (b) and (c).](cache/k600_theory/k600Plot.jpg)
+We test the performances of equations 4 and 5 on our dataset of field-measured $k_{600}$ in Figure 3.
+Figure 3a fits the hydraulically-wide small-eddy model (equation 4) to hydraulically-wide measurements: only the data where $R_h$ is within 1% percent of $H$.
+Figure 3b fits the hydraulically wide chainsaw model (equation 5) to the same subset of data.
 
-Finally, becuase $\epsilon_D$ refers to a depth-scale quantity, we tested whether a more parsimonious model is possible by removing the 'middle-man', i.e.
-the Kolmogorov velocity scale that the small-eddy model is based on [@lamontEddyCellModel1970], and directly predicting *k* from depth-scale turbulence.
-Figure 3d plots the fitted $k=\beta_1U_*$ model.
-For this data, $U_*$ explains 70% of variation in $k_{600}$ and, as expected, the fitted model is nearly identical to Figure 3c in scaling dynamics and variance explained.
-This all indicates that depth-scale turbulent production/dissipation dominates the surface dissipation signal in extremely inefficient river channels.
-These findings build off of @ulsethDistinctAirWater2019a's intial result that channel slope exerts a drammatic influence on *k* scaling dynamics, further suggesting that fluvial *k* is as much a function of hydraulic geometry as the turbulence properties of the flow.
+![Figure 3: Model development for predicting gas exchange velocity from SWOT-observable variables.
+(a) Empirical confirmation of the hydraulically wide small-eddy model.
+(b) Empirical confirmation of the hydraulically-wide chainsaw model.
+(c) Cummulative density function (CDF) of the hydraulic radius/mean flow depth ratios for all SWOT-observable measurements.
+Nearly all of these measurements would be considered hydraulically wide.](cache/k600_theory/k600Plot.jpg)
 
-With all of that said, how often is a river channel so inefficient that $R_h \approx H$?
-To answer this question, we join the 701 measurements from @ulsethDistinctAirWater2019a to 171,553 discrete river hydraulics measurements from across the continental United States [@brinkerhoffReconcilingAtaStationAtManyStations2019].
-We find that `r round(percs_overall[percs_overall$flag_depth == 'Rh=H',]$perc,0)`% of hydraulic radii are within 5% of their mean flow depth.
-Because these measurements were made exclusively at streamgauges where rivers are somewhat channelized and width variability is decreased, we opted for a slightly more liberal ratio of 5%.
-Further, because larg rivers are more likely to have inefficient channels, we identify the measurements that are SWOT-observable, i.e.
-width > 100m.
-We find that while `r round(percs[percs$flag_depth == 'Rh=H',]$perc,0)[1]`% of non-SWOT observable rivers are inefficient, `r round(percs[percs$flag_depth == 'Rh=H',]$perc,0)[2]`% of SWOT-observable rivers are inefficient.
-This significant difference, and near universality of SWOT-observable rivers being inefficient, suggests that our theoretical findings in Figure 3 should reasonably hold in nearly all rivers that SWOT will observe.
-This opens the door for exploiting our 'inefficient river k model' for remote sensing *k* via the SWOT satellite.
+The small-eddy model in Figure 3a produces clearly biased estimates, systematically underestimating low gas exchange velocities and overestimating higher ones.
+However, when this model is expanded to account for a differential dissipation rate at the interface via the chainsaw model [@moogAirWaterGasTransfer1999], the dynamics are accurately captured (Figure 3b).
+The coefficients of variation are similar (`r round(modelsWIDE[modelsWIDE$name == 'Small-eddy',]$r2,2)` and `r round(modelsWIDE[modelsWIDE$name == 'Chainsaw',]$r2,2)`, respectively), however the small-eddy model is clearly misrepresenting the scaling dynamics of gas exchange velocity in hydraulically-wide rivers.
+Given the algebraic structure of these two models, the discrepancy is in the exponents applied to the river hydraulic terms.
+More intuitively, this means that in hydraulically wide rivers where turbulence is generally relatively low, large-scale eddies exert a significant influence on surface dissipation rates: ultimately under the chainsaw model, gas exchange velocities are more sensitive to slope and mean depth than under the small-eddy model (Figure 3b).
+Figure 3c plots the cummulative density function (CDF) of all SWOT-observable measurements in our hydraulics dataset (Section 2), confirming that nearly all SWOT-observable measurements are hydraulically-wide.
+As stated previously, `r round(percs[percs$flag_swot == 'SWOT' & percs$flag_hydraulicWide == 'Rh=H',]$perc,0)`% of SWOT-observable measurements are within 5% of being perfectly hydraulically wide.
+Finally, this is to our knowledge the first empirical confirmation that the chainsaw model is valid in rivers (at least hydraulically-wide ones).
 
-## 4 Exploiting the 'inefficient river model' to remotely sense gas exchange velocity via the BIKER algorithm
-We have shown that scaling $k_{600}$ via the shear velocity ($U_*$) explains 70% of the variation in $k_{600}$ in inefficient rivers and that nearly all SWOT-observable rivers are inefficient.
-This is incredibly convenient for remote sensing because the structure of $U_*$ (Appendix A) reduces the model to two non-remotely-sensible terms: $k_{600}$ and an estimate of the channel area.
-This is explained below.
-By scaling *k* this way, we significantly reduce equifinality issues, where equifinality refers to an under-constrained mathematical system that has essentially infinite parameter combinations that can produce the same result: there are more unknowns than equations [@garamboisInferenceEffectiveRiver2015].
-This problem is experienced by both SWOT RSQ algorithms and in situ tools that concurrently solve for *k* and stream metabolism [@applingOvercomingEquifinalityLeveraging2018; @graceFastProcessingDiel2015; @holtgrieveSimultaneousQuantificationAquatic2010].
-In both of these domains, the other unknown parameters are often difficult to estimate (bed roughness and stream metabolism, respectively) while for BIKER, median channel area is relatively easy to approximate from the SWOT-observable river width [@brinkerhoffConstrainingRemoteRiver2020a].
+The success of the chainsaw model in hydraulically wide channels means that we can reasonably predict gas exchange velocities in hydraulically wide channels using just water surface slope and mean flow depth.
+These easily obtainable hydraulics parameters can also be directly measured or estimated from satellite remote sensing platforms, effectively opening the door for remotely sensing gas exhange via equation 5.
+This is explored next.
 
-Taking all of this together, we develop the BIKER algorithm.
-BIKER is informed by the @hagemannBAMBayesianAMHGManning2017a algorithm for ungauged RSQ and further explored in more recent work by @brinkerhoffConstrainingRemoteRiver2020a.
+## 4 Exploiting the hydraulically wide channels to remotely sense gas exchange velocity
+We have shown that scaling $k_{600}$ via equation 5 explains `r round(modelsWIDE[modelsWIDE$name == 'Chainsaw',]$r2,2)`% of the variation in $k_{600}$ in hydraulically wide rivers, which includes nearly all SWOT-observable rivers.
+This is incredibly convenient for remote sensing because there are only two non-remotely-sensible terms: $k_{600}$ and mean flow depth.
+By scaling *k* this way, we significantly reduce the equifinality issues that plague the more complicated SWOT discharge algorithms.
+Equifinality refers to an under-constrained mathematical system that has essentially infinite parameter combinations that can produce the same result: there are more unknowns than equations [@garamboisInferenceEffectiveRiver2015].
+This is also a problem for the in situ tools that concurrently solve for *k* and stream metabolism [@applingOvercomingEquifinalityLeveraging2018; @graceFastProcessingDiel2015; @holtgrieveSimultaneousQuantificationAquatic2010].
+In both of these domains, the other unknown parameters are often difficult to estimate (bed roughness and stream metabolism, respectively) while for BIKER, mean flow depth is relatively easy to approximate from the SWOT-observable river width [@brinkerhoffConstrainingRemoteRiver2020].
+This is explained in detail later.
+
+Taking all of this together, we develop the BIKER algorithm to remotely sense gas exchange velocities from the SWOT satellite.
+BIKER is informed by the @hagemannBAMBayesianAMHGManning2017 algorithm for ungauged RSQ and further explored in more recent work by @brinkerhoffConstrainingRemoteRiver2020.
 These papers conceptualize discharge as a Bayesian remote sensing problem, which we largely follow here to conceptualize $k_{600}$ as a Bayesian remote sensing problem that can be solved using SWOT data.
@@
 Therefore, we are interested in solving for $p(\Theta|x)$, or the 'posterior' distribution.
-For BIKER, *x* is the SWOT-observables: river width *W* and water surface elevation *H* (which is used to calculate the water surface slope $S_h$).
+For BIKER, *x* is the SWOT-observables: river width *W* and water surface elevation $H_e$ (which is used to calculate the water surface slope $S_h$).
 Note that $p(x)$ is usually computationally intractable to integrate, so Bayesian inference tools require only the proportionality to be specified: $p(\Theta|x) \propto f(x|\Theta)p(x)$.
@@
 To do this, $k_{600}$ is first written as a function of SWOT-observables *W* and $S_h$.
-This algebra is carried out using the model parameter from Figure 3d (56.0294) and yields equation 6, assuming the channel is inefficient ($R_h = H =\frac{A}{W}$).
+This algebra is carried out using equation 4 and the model parameter from Figure 3a (`r round(modelsWIDE[modelsWIDE$name == 'Chainsaw',]$coef,2)`) and yields equation 6, assuming the channel is hydraulically wide ($R_h = H =\frac{A}{W}$).
 Here, *A* is wetted channel area.
-*A* is further split into the SWOT-observable portion *dA* and the unobservable portion $A_0$ following @durandEstimatingReachaveragedDischarge2014a.
-*dA* is estimated assuming a rectangular river channel so that $dA \approx \delta H*W$.
-Thus, there are no reliances on bed roughness, flow velocity, or any other terms that are difficult to infer from river width (if we implemented the nearly identical model in Figure 3b, we would need to simultaneously estimate roughness and/or velocity which becomes a much more difficult problem).
+*A* is further split into the SWOT-observable portion *dA* and the unobservable portion $A_0$ following @durandEstimatingReachaveragedDischarge2014.
+*dA* is estimated assuming a rectangular river channel so that $dA \approx \delta H_e*W$.
+Thus, there is no reliance on bed roughness, flow velocity, or any other terms that are more difficult to infer from river width.
 
+**UPDATE**
 $$\mathbf{(6)} \begin{eqnarray} k_{600} = \bigg(56.0294 S_h(\frac{A_0+dA}{W})\bigg)^{1/2} \end{eqnarray}$$
@@
 Next, equation 6 is written as a Bayesian sampling model, in which all of the SWOT observations are sampled from the unknown model parameters ($A_0$ and $k_{600}$).
-This is equation 7 after log-transforming all of the variables to describe them as normal distributions.
+This is equation 7 after describing them as log-normal distributions.
 $\sigma_{k_{600}}^2$ refers to the uncertainty inherent in equation 6's estimates.
@@
 Priors could in theory be improved if there is available a priori information about the river.
+We could also assign a prior to $\beta$ in equations 4 and 6 because there is an inherent uncertainty associated with it.
+For the sake of the validation presented here (and elaborated on below), we assumed $\beta$ is known a priori and did not treat it as a parameter.
 
@@
 To approximate this distribution, we use a Markov Chain Monte Carlo (MCMC) algorithm implemented using the Stan probabilistic programming language.
-Specifically, Stan uses a Hamiltonian Monte Carlo sampler which reduces computation time relative to other sampling algorithms [@hagemannBAMBayesianAMHGManning2017a].
+Specifically, Stan uses a Hamiltonian Monte Carlo sampler which reduces computation time relative to other sampling algorithms [@hagemannBAMBayesianAMHGManning2017].
 
-We validate BIKER on 47 SWOT-simulated rivers using daily observed hydraulics and the metrics described in Table S1.
-We also re-validate BIKER on the 17 rivers with the SWOT error model that corrupts the river width and slope.
+We validate BIKER on 47 SWOT-simulated rivers using daily observed hydraulics and the metrics described in Table S2.
+We also re-validate BIKER on the 17 rivers with the SWOT error model which corrupts the river width and slope.
 Regardless of the validation setup, we do not have observed $k_{600}$ data for these rivers, and to our knowledge no field dataset of $k_{600}$ exists in the type of temporal and spatial frequency that SWOT (and therefore the BIKER algorithm) provides.
-Therefore, we take the model outlined in Figure 3b and use that to calculate the observed $k_{600}$ that BIKER is validated against.
+Therefore, we take the model outlined in equation 4 and Figure 3 and use that to calculate the observed $k_{600}$ that BIKER is validated against.
+This is why equation 3's $\beta$ is trated as known and not a parameter.
 With this setup, we are directly exploring BIKER's ability to infer observed $A_0$ and $k_{600}$ from *W* and *H* alone, as the *k* scaling model has already been sucessfully validated (Figure 3).
-This is elaborated on in Text S3.
+This workflow is elaborated on in Text S3.
 
@@
 The points are the posterior means while the black lines are the 95% confidence intervals (CIs) for the predictions.
-$k_{600}$ is strongly correlated with the BIKER-predicted $k_{600}$ ($r^2$ of `r round(results_3_2_all$r2, 2)`).
+$k_{600}$ is strongly correlated with the BIKER-predicted $k_{600}$ ($r^2$ of `r round(results_k600_all$r2, 2)`).
 Using absolutely no in situ information, BIKER captures the magnitude of the predictions and most points fall on or near the 1:1 line.
 The regression of the estimates (solid grey line) nearly recovers the 1:1 line (dashed black), but there is an over/underestimation bias in the largest/smallest $k_{600}$ values.
-The RMSE for the BIKER predictions is `r round(results_3_2_all$rmse,2)` m/day) across all observations.
+The RMSE for the BIKER predictions is `r round(results_k600_all$rmse,2)` m/day) across all observations.
 
@@
 Figure 4b-d are representative timeseries plots of predicted and observed $k_{600}$ for three rivers chosen randomly from those with 'good' NRMSE scores (b), 'okay' NRMSE scores (c), and 'bad' NRMSE scores (d).
-See Table S1 for the definition of NRMSE, the Figure 4 caption for how this was determined, and Figure S3 for the other river timeseries plots.
+See Table S2 for the definition of NRMSE, the Figure 4 caption for how this was determined, and Figure S3 for the other river timeseries plots.
 For the Missouri Midsection River, the entire timeseries of $k_{600}$ is nearly perfectly predicted, while in the Seine River the dynamics (peaks and valleys) are reasonably captured but they are magnified to be far larger than the observed dynamics.
@@
 The points making up these boxplots (47 and 17, respectively) are overlain atop the boxplots.
-See Table S1 for metric definitions.
+See Table S2 for metric definitions.
 SWOT measurement errors neglibly influence BIKER's performance across all four error metrics (Figure 5a), though caution should be used in over-interepting boxplots with a sample size of only 17.
@@
 a: Performance metrics by river.
-See Table S1 for metric definitions.
+See Table S2 for metric definitions.
 b: NRMSE scores (per river) versus the number of SWOT observations used in the Bayesian inversion.
@@
 
-Median river-specific $r^2$ is `r round(median(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$r2, na.rm=T),2)`, which is excellent given that absolutely no in situ information is being used to predict $k_{600}$.
-NRMSE and RRMSE have median scores of `r round(median(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$nrmse, na.rm=T),2)` and `r round(median(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$rrmse, na.rm=T),2)`, respectively, which are again good for a completely ungauged method.
+Median river-specific $r^2$ is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$r2, na.rm=T),2)`, which is excellent given that absolutely no in situ information is being used to predict $k_{600}$.
+NRMSE and RRMSE have median scores of `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$nrmse, na.rm=T),2)` and `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$rrmse, na.rm=T),2)`, respectively, which are again good for a completely ungauged method.
 These are comparable to the best NRMSE scores achieved by the SWOT RSQ algorithms [@frassonExploringFactorsControlling2021].
-Median rBIAS is `r round(median(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$rBIAS, na.rm=T),2)`, highlighting a  positive bias in most rivers' predictions.
+Median rBIAS is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$rBIAS, na.rm=T),2)`, highlighting a  positive bias in most rivers' predictions.
 This further supports the visual evidence in Figure 4 that sometimes BIKER is overestimating the magnitude of $k_{600}$ and that this might be river-specific.
-While median $r^2$ and rBIAS scores were strong, the ranges of these scores were somewhat large (standard deviation for $r^2$ of `r round(sd(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$r2, na.rm=T),2)` and for rBIAS of `r round(sd(results_3_2_rivs[results_3_2_rivs$errFlag == 0,]$rBIAS, na.rm=T),2)`).
+While median $r^2$ and rBIAS scores were strong, the ranges of these scores were somewhat large (standard deviation for $r^2$ of `r round(sd(results_k600_rivs[results_k600_rivs$errFlag == 0,]$r2, na.rm=T),2)` and for rBIAS of `r round(sd(results_k600_rivs[results_k600_rivs$errFlag == 0,]$rBIAS, na.rm=T),2)`).
 
@@
 Therefore, we also explore 1) BIKER's ability to reproduce $FCO_2$ $[M/L^2T]$ (equation 1) from these 47 rivers, and 2) a comparison of the representative carbon efflux $[M/T]$ from BIKER with established in situ methods.
-The details of this workflow are in Text S4, but broadly we pair the 26 biweekly $CO_2$ and water temperature samples (section 2, figure S2) from @beaulieuControlsGasTransfer2012a with a subset of SWOT observations (as the data are not daily).
+The details of this workflow are in Text S4, but broadly we pair the 26 biweekly $CO_2$ and water temperature samples (section 2, figure S2) from @beaulieuControlsGasTransfer2012 with a subset of SWOT observations (as the data are not daily).
 We then calculate $FCO_2$ using equation 1 and assuming atmospheric $CO_{2_{air}}$ is 390 uatm.
-Finally, we estimate a median bulk carbon efflux using BIKER's posterior means and three other in situ models for average channel depth (used to calculate $U_\star$) used for upscaling: one trained on the @brinkerhoffReconcilingAtaStationAtManyStations2019 dataset, and two previously published models [@raymondGlobalCarbonDioxide2013a; @raymondScalingGasTransfer2012a].
-See Table S2 for their definitions.
+Finally, we estimate a median bulk carbon efflux using BIKER's posterior means and three other in situ models for average channel depth (used to calculate $U_\star$) used for upscaling: one trained on the @brinkerhoffReconcilingAtaStationAtManyStations2019 dataset, and two previously published models [@raymondGlobalCarbonDioxide2013; @raymondScalingGasTransfer2012].
+See Table S3 for their definitions.
 This allows us to assess whether BIKER's estimates (wholly ungauged) are comparable to gauged methods (all four HG models).
 
-In Figure 6a, there is a very strong fit to the observed $FCO_2$ data, with an RMSE of `r round(results_3_3_all$rmse,2)` $g/m^2dy$.
+In Figure 6a, there is a very strong fit to the observed $FCO_2$ data, with an RMSE of `r round(results_fco2_all$rmse,2)` $g/m^2dy$.
 The performance is notably better than for $k_{600}$ alone (Figure 4a) and there is no systematic bias in the predictions across all 47 rivers.
@@
 The results are largely equivalent for both the mean and median carbon efflux.
-Both the BIKER estimate (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_BIKER',]$sumFCO2, 2)` gG-C/yr) and the 'Raymond 2013' estimate (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_raymond2013',]$sumFCO2, 2)`) are somewhat overestimated relative to the observed flux (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_obs',]$sumFCO2, 2)`).
-The 'Brinkerhoff 2019' model nearly perfectly recovers the observed value (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_brinkerhoff2019',]$sumFCO2, 2)`).
-Finally, the 'Raymond 2012' model slightly underestimates this bulk efflux (`r round(results_3_3_bulk[results_3_3_bulk$key == 'FCO2_raymond2012',]$sumFCO2, 2)`).
+Both the BIKER estimate (`r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_BIKER',]$sumFCO2, 2)` gG-C/yr) and the 'Raymond 2013' estimate (`r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_raymond2013',]$sumFCO2, 2)`) are somewhat overestimated relative to the observed flux (`r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_obs',]$sumFCO2, 2)`).
+The 'Brinkerhoff 2019' model nearly perfectly recovers the observed value (`r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_brinkerhoff2019',]$sumFCO2, 2)`).
+Finally, the 'Raymond 2012' model slightly underestimates this bulk efflux (`r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_raymond2012',]$sumFCO2, 2)`).
 Thus, despite BIKER using absolutely no streamgauge or other in situ data like the other 3 methods do, it provides a reasonable estimate of the carbon efflux (Figure 6b).
@@
 ### 6.1 Gas exchange along the stream-to-ocean continumn
-Field studies of gas exchange in wide rivers have suggested that their *k* properties behave differently than in steeper and smaller rivers [@ulsethDistinctAirWater2019a; @beaulieuControlsGasTransfer2012a; @alinPhysicalControlsCarbon2011; @wangPhysicallyBasedScaling2021].
+Field studies of gas exchange in wide rivers have suggested that their *k* properties behave differently than in steeper and smaller rivers [@ulsethDistinctAirWater2019; @beaulieuControlsGasTransfer2012; @alinPhysicalControlsCarbon2011; @wangPhysicallyBasedScaling2021].
 While it has been shown that the small-eddy model is reasonably valid across most rivers [@wangPhysicallyBasedScaling2021], it has yet to be seen why and when form-drag dissipation is similar to near-surface dissipation in the context of *k*.
@@
 @wangPhysicallyBasedScaling2021 proposed a conceptual model for fluvial *k* where *k* is dominated by bottom shear in small streams while in large rivers *k* is also affected by wind, biota, etc.
-This builds on the [@ulsethDistinctAirWater2019a] finding that steep, 'high-energy' streams have a fundamentally different $k_{600} \sim \epsilon_D$ relationship than less steep rivers and this may be due to the presence of bubble-mediated gas exchange and massive roughness elements in small mountain streams.
+This builds on the [@ulsethDistinctAirWater2019] finding that steep, 'high-energy' streams have a fundamentally different $k_{600} \sim \epsilon_D$ relationship than less steep rivers and this may be due to the presence of bubble-mediated gas exchange and massive roughness elements in small mountain streams.
 Taking both of these models, along with our results, in aggregate, we suggest a combined conceptual model that follows the geomorphic concept of the 'river longitudinal profile' and is outlined in Figure 7.
@@
 In these largest rivers, form-drag starts dominating the *k*~hydraulics relationship rather than bed roughness.
-While wind certainly also plays a role in *k* in these large rivers [particuarly at higher winds- @beaulieuControlsGasTransfer2012a], we show here that for maximally inefficient river channels form-drag alone can reasonably predict $k_{600}$.
+While wind certainly also plays a role in *k* in these large rivers [particuarly at higher winds- @beaulieuControlsGasTransfer2012], we show here that for maximally inefficient river channels form-drag alone can reasonably predict $k_{600}$.
 The relationship between form-drag dissipation and wind is left to future work.
 
-![Figure 7: Conceptual model of the river longitudinal profile as it pertains to gas exchange velocity.
-As rivers grow in size and one moves downstream, their channels become wider, flatter, less efficient at transporting sediment, and have greater relative frictional resistance.
-This suggests a shift in k~hydraulics relationships from one dominated by bed roughness in small, steep, and efficient systems versus one dominated by form-drag in large, flat, and inefficient systems.](cache/k600_theory/longProfile.jpg)
+Finally, this is to our knowledge the first empirical confirmation that the chainsaw model is valid in rivers (at least hydraulically-wide ones).
+Future tests should be performed to explore it's feasibility in smaller streams and/or other aquatic environments where turbulence is generally lower.
+We are not accounting for dissipation generated via wind shear, as none of these hydraulics-driven gas exchange models do.
+We suspect that this model will be less representative in less-sheletered rivers where wind becomes a driving factor of gas exchange rates, however this model setup is a good start towards coupling hydraulics-driven gas exchange with wind-driven gas exchange.
+Such a model would likely take a form similar to the conceptual model proposed by @wangPhysicallyBasedScaling2021.
 
 ### 6.2 Towards remote sensing of global spatiotemporal dynamics of *k* in large rivers
-To date, most field-scale studies of riverine gas exchange have focused on 1) its relationship with wind speed [e.g @beaulieuControlsGasTransfer2012a; @borgesVariabilityGasTransfer2004; @zappaEnvironmentalTurbulentMixing2007], 2) average flow velocity [e.g.
-@alinPhysicalControlsCarbon2011; @beaulieuControlsGasTransfer2012a; @schelkerCO2EvasionSteep2016a], or 3) discharge [e.g.
+To date, most field-scale studies of riverine gas exchange have focused on 1) its relationship with wind speed [e.g @beaulieuControlsGasTransfer2012; @borgesVariabilityGasTransfer2004; @zappaEnvironmentalTurbulentMixing2007], 2) average flow velocity [e.g.
+@alinPhysicalControlsCarbon2011; @beaulieuControlsGasTransfer2012; @schelkerCO2EvasionSteep2016], or 3) discharge [e.g.
 @wangPhysicallyBasedScaling2021; @robertsMultipleScalesTemporal2007; @uehlingerEcosystemMetabolismDisturbance1998].
@@
 This is particualry useful given recent advances in in situ gas concentration loggers [@ahoHighestRatesGross2021] but no such similar advances in modeling *k* at equivalent temporal resolutions.
-BIKER can likely also be ran at the field scale using arrays of pressure transducers to estimate water surface slope (rather than using satellite-based altimeters like SWOT) following recent work doing the same using the @hagemannBAMBayesianAMHGManning2017a RSQ algorithm [@harlanDischargeEstimationDense2021].
+BIKER can likely also be ran at the field scale using arrays of pressure transducers to estimate water surface slope (rather than using satellite-based altimeters like SWOT) following recent work doing the same using the @hagemannBAMBayesianAMHGManning2017 RSQ algorithm [@harlanDischargeEstimationDense2021].
 
@@
 Future work should systematically quantify prediction error from coupling global-scope HG models with modeled discharge, as is the default workflow used in fluvial $FCO_2$ upscaling studies [e.g.
-@lauerwaldSpatialPatternsCO22015; @horgbyUnexpectedLargeEvasion2019; @raymondGlobalCarbonDioxide2013a; **Liu et al in reivew**].
+@lauerwaldSpatialPatternsCO22015; @horgbyUnexpectedLargeEvasion2019; @raymondGlobalCarbonDioxide2013; **Liu et al in reivew**].
 
@@
 |$\epsilon_S$|Dissipation rate of near-surface turbulence|log-law-of-the-wall model: $\frac{U_*^3}{H}$|$[\frac{E}{M*T}]$|
-|$\epsilon_D$|Form-drag turbulent dissipation rate|$gSU$|$[\frac{E}{M*T}]$|
 |$FCO_2$|$CO_2$ evasion flux from river to atmosphere|NA|$[\frac{M}{L^2T}]$|
 |$g$|gravitational acceleration|9.8|$[\frac{L}{T^2}]$|
-|$G$|TKE production rate|$\frac{UU_*}{H}$|$[\frac{E}{M*T}]$|
 |$H$|Mean flow depth|$A/W$|$[L]$|

### 2021-09-30 supp_v5.Rmd
model update to the chainsaw model

--- 2021-09-10 supp_v5.Rmd
@@
 ---
-title: "Supplemental Information: Gas exchange in large rivers controlled by hydraulic geometry: implications for remotely sensing gas exchange via the SWOT satellite"
-author: "CB Brinkerhoff, CJ Gleason, CJ Zappa, PA Raymond, MH Harlan"
-date: "`r Sys.Date()`"
+title: "Supplemental Information: Gas exchange in large rivers influenced by largest turbulent eddies: implications for remotely sensing gas exchange via the SWOT satellite"
+author:
+  - Craig B Brinkerhoff:
+      email: cbrinkerhoff@umass.edu
+      institute: UMass
+      correspondence: true
+  - Colin J Gleason:
+      institute: UMass
+  - Christopher J Zappa:
+      institute: Lamont
+  - Peter A Raymond:
+      institute: Yale
+  - Merritt H Harlan:
+      institute: UMass
+institute:
+  - UMass: Department of Civil & Environmental Engineering, University of Massachusetts, Amherst, MA
+  - Lamont: Lamont-Doherty Earth Observatory, Columbia University, Palisades, NY
+  - Yale: School of the Environment, Yale University, New Haven, CT
+bibliography: References.bib
 csl: agu_citation_style.csl
@@
 
-##Contents
-This supplementary information contains 4 texts, 4 figures and 2 tables.
+## Contents
+This supplementary information contains 4 texts, 3 figures and 3 tables.
 Please consult https://github.com/craigbrinkerhoff/RSK600 for all code to build and generate results, figures, and the manuscript.
 
-## Test S1: Filtering procedure for hydraulics data
-To quantify how frequently SWOT-observable rivers are also inefficient river channels (section 2 of main text), we used the dataset of field-measured river hydraulics in @brinkerhoffReconcilingAtaStationAtManyStations2019.
+## Test S1: Estimating how many SWOT rivers are hydraulically wide
+To quantify how frequently SWOT-observable rivers are also hydraulically wide river channels (section 3 of main text), we used the dataset of field-measured river hydraulics in @brinkerhoffReconcilingAtaStationAtManyStations2019.
 That dataset has over 500,000 discrete measurements of river width, velocity, area, and discharge that were made by the United States Geological Survey (USGS) to calibrate streamgauge rating curves.
-Here, we descirbe how this dataset was filtered down to 171,553 measurements to meet the assumptions in our conceptual model.
+Here, we describe how this dataset was filtered down to 171,553 measurements and how we quantified what was 'hydraulically wide'.
 
-First, we removed all measurements missing data, measurements tagged by the USGS as 'poor', measurements with impossible values, or measurements of 0.
+First, we removed all measurements tagged by the USGS as 'poor', measurements with impossible values, or measurements of 0.
 While this would indicate a dry channel, our hydraulic geometry model necessitates within-bank flow.
-Likewise, because hydraulic geometry only applies to within-bank flows and not flood events, we had to remove all overbank flows.
+Likewise, because hydraulic geometry only applies to within-bank flows and not flood events, we remove all overbank flows.
 This was done by first filtering for sites with at least 20 measurements (to build robust estimates of bankfull hydraulics) and then calculating bankfull width and depth as the width or depth with a return period of two years.
-While the only true way to calculate bankfull hydraulics is manually in the field, this was ovbiously impractical.
+While the only true way to calculate bankfull hydraulics is manually in the field, this is obviously impractical here.
 A two year return period is a standard approximation for determining out-of-bank flow in single-channel meandering rivers and was the method used by @brinkerhoffReconcilingAtaStationAtManyStations2019.
@@
 
+After joining this dataset with the hydraulics measurements provided with the 763 $k_{600}$ measurements, we classified each measurement as hydraulically wide if its hydraulic radius was within 4% of its mean flow depth.
+Hydraulic radius was calculated assuming a rectangular channel such that $R_H=2H+W$.
+Because these measurements were almost entirely made at streamgauges, which generally have a location bias favoring stable channels near bridges and other structures that yield lower width dynamics than those observed away from gauges [@allenPatternsRiverWidth2015; @parkWorldwideVariationsHydraulic1977], we used a wider threshold of 4%.
+In the 'width-limited' scenario common at streamgauges, the hydraulic radius is likely further from the mean flow depth than would occur naturally.
+The slightly more liberal threshold of 4% allowed for implicit accounting for this sampling bias.
+For gas exchange velocity model fitting however, we continued to use the more conservative (and theoretically defensibl) threshold of 1%.
+
 ## Text S2: BIKER hyperparameterization
+In this text we explain in detail how BIKER's hyperparameter values were set for each river.
+
 We assign prior hyperparameters using SWOT data only.
@@
 
-$A_0$ prior hyperparameters were assigned following an updated version of the method developed by @brinkerhoffConstrainingRemoteRiver2020a.
+$A_0$ prior hyperparameters were assigned following an updated version of the method developed by @brinkerhoffConstrainingRemoteRiver2020.
 They developed a set of river channel prior hyperparameters for McFLI algorithms that are entirely RS-able and reflect differential channel hydraulics as a function of river geomorphology.
@@
 We assigned those by using a simple slope regression model to predict $U_*$ solely from a SWOT-observable value.
-It was trained on the @ulsethDistinctAirWater2019a dataset.
+It was trained on the @ulsethDistinctAirWater2019 dataset.
 Equation S9 assigns $\mu_{k_{600}}$ while $\sigma^2_{k_{600}}$ was set to equation S9's log-transformed standard error after propogation of errors for both equation S9 and equation 6 in the main text.
 This amounted to a value of 1.12.
-$\lambda_{k_{600}}$ and $\gamma_{k_600}}$ were set to log(0.001) m/day and log(500) m/day, respectively.
+$\lambda_{k_{600}}$ and $\gamma_{{k_600}}$ were set to log(0.001) m/day and log(500) m/day, respectively.
 Finally, we estimate $\sigma_{k_{600}}$ using the standard model error from equation 6 in the main text and as fit on the data shown in Figure 3a.
 
+**NEEDS TO BE REDONE**
 $$\mathbf{(S9)} \mu_{k_{600}}=56.0294 * a(ln(S_h))^b$$
@@
 ## Text S3: BIKER validation setup
+In this text, we explain the BIKER validation setup in detail.
+
 Regardless of the validation setup or SWOT error budget used, we do not have observed $k_{600}$ data for these rivers, and to our knowledge no field dataset of $k_{600}$ exists in the type of temporal and spatial frequency that SWOT (and therefore the BIKER algorithm) provides.
@@
 ## Text S4: BIKER FCO2 workflow
-To calculate $CO_2$ fluxes and carbon efflux, we pair @beaulieuControlsGasTransfer2012a's biweekly 26 $CO_2$ and water temperature samples with every 11th SWOT observation by date, ignoring the timesteps beyond 26.
+Here, we explain how $CO_2$ fluxes were calcualted for each simulated SWOT river.
+
+To calculate $CO_2$ fluxes and carbon efflux, we pair @beaulieuControlsGasTransfer2012's biweekly 26 $CO_2$ and water temperature samples with every 11th SWOT observation by date, ignoring the timesteps beyond 26.
 This amounts to only ~15% of the SWOT observations and we deem this acceptable.
@@
 In equation S11, *Sc* is the Schmidt number at the observed water temperature.
-This was calculated following @raymondScalingGasTransfer2012a and @wanninkhofRelationshipWindSpeed1992a.
+This was calculated following @raymondScalingGasTransfer2012 and @wanninkhofRelationshipWindSpeed1992.
 Atmospheric $CO_2$ was assumed 390 uatm.
@@
 
-##Figures
+## Figure S1
+**UPDATE**
 ![Figure S1: Flowchart of the entire study's workflow](cache/flowchart.jpg)
 
-![Figure S2 Timeseries of the biweekly CO2 data from @beaulieuControlsGasTransfer2012a.
+## Figure S2
+![Figure S2 Timeseries of the biweekly CO2 data from @beaulieuControlsGasTransfer2012.
 Sampling took place 2008-2009 in the Ohio River (upstream of Cincinnati, Ohio, United States).
@@
 
+## Figure S3
 ![Figure S3 All timeseries of k per river](cache/validation/timeseries_noerr.jpg)
 
-##Tables
-*Table S1: Validation metrics used in this study, where r is the correlation coefficient, Nt is number of observations and i is the specific observation.
+## Table S1
+*Table S1: Studies that gas exchange velocity measurements come from.
+'Study' refers to the paper from which these measurements were taken.
+Any data wrangling was done by those authors.
+'Field Workers' refers to who actually made the measurements.
+The Raymond et al.
+(2012) dataset is itself a meta-analysis.
+Please see that paper for how those measurements were collected.
+We used the set of measurements ultimately published by Ulseth et al.
+2019.*
+
+|**Study**|**Workers**|**Number of measurements**|
+|---------|-------------|--------------------------|
+|@ulsethDistinctAirWater2019|@ulsethDistinctAirWater2019||
+|@ulsethDistinctAirWater2019|@hallUseArgonMeasure2018||
+|@ulsethDistinctAirWater2019|@schelkerCO2EvasionSteep2016||
+|@ulsethDistinctAirWater2019|@mauriceInfluenceFlowBed2017||
+|@ulsethDistinctAirWater2019|@raymondScalingGasTransfer2012||
+<!-- |@raymondScalingGasTransfer2012|@melchingReaerationEquationsDerived| -->
+<!-- |@raymondScalingGasTransfer2012|@tsivoglouCharacterizationStreamReaeration1972| -->
+<!-- |@raymondScalingGasTransfer2012|Stroud Water Center| -->
+<!-- |@raymondScalingGasTransfer2012|@bernotInterregionalComparisonLanduse2010| -->
+<!-- |@raymondScalingGasTransfer2012|@mulhollandInterbiomeComparisonFactors2001| -->
+<!-- |@churchillPredictionStreamReaeration1964|@churchillPredictionStreamReaeration1964|| -->
+<!-- |@owensReaerationStudiesStreams1964|@owensReaerationStudiesStreams1964|| -->
+
+## Table S2
+*Table S2: Validation metrics used in this study, where r is the correlation coefficient, Nt is number of observations and i is the specific observation.
 σ refers to the variance of the sample and μ refers to the mean of the sample.
@@
 
-*Table S2: Details on the 3 depth hydraulic geometry models used to estimate FCO2 from the SWOT rivers (section 2.4).*
+## Table S3
+*Table S3: Details on the 3 depth hydraulic geometry models used to estimate FCO2 from the SWOT rivers (section 2.4).*
 
@@
 |Brinkerhoff 2019|$D=0.258Q^{0.395}$|530,945 measurements made across the United States at streamgauges|this study; @brinkerhoffReconcilingAtaStationAtManyStations2019|
-|Raymond 2012|$D=0.409Q^{0.294}$|1,026 measurements across the United States|@raymondScalingGasTransfer2012a|
-|Raymond 2013|$D=0.409Q^{0.294},  D=0.449Q^{0.37}$ |Average of the Raymond 2012 equation and one using 9,811 measurements at US streamgauges|@raymondGlobalCarbonDioxide2013a|
+|Raymond 2012|$D=0.409Q^{0.294}$|1,026 measurements across the United States|@raymondScalingGasTransfer2012|
+|Raymond 2013|$D=0.409Q^{0.294},  D=0.449Q^{0.37}$ |Average of the Raymond 2012 equation and one using 9,811 measurements at US streamgauges|@raymondGlobalCarbonDioxide2013|
 
-##References
+## References

### 2021-09-30 generateMS.R

@@ -4,23 +4,25 @@
 
 library(tidyverse)
 
-setwd("C:/Users/craig/Documents/OneDrive - University of Massachusetts/Ongoing Projects/RSK600")
+setwd("C:/Users/cbrinkerhoff/OneDrive - University of Massachusetts/Ongoing Projects/RSK600")
+#setwd("C:/Users/craig/Documents/OneDrive - University of Massachusetts/Ongoing Projects/RSK600")
 
 #Move all results to global environment
 #turbulence_stats <- read.csv('cache\\k600_theory\\turbulence_stats.csv')
-models_ulseth <- read.csv('cache\\k600_theory\\ulseth_models.csv')
-models_ustar <- read.csv('cache\\k600_theory\\ustar_models.csv')
-models_eD4 <- read.csv('cache\\k600_theory\\eD_models.csv')
+#models_ulseth <- read.csv('cache\\k600_theory\\ulseth_models.csv')
+#models_ustar <- read.csv('cache\\k600_theory\\ustar_models.csv')
+#models_eD4 <- read.csv('cache\\k600_theory\\eD_models.csv')
+modelsWIDE <- read.csv('cache\\k600_theory\\hydraulicWide_models.csv')
 percs <- read.csv('cache\\k600_theory\\Rh_H_percents.csv')
-percs_overall <- dplyr::group_by(percs, flag_depth) %>% 
+percs_overall <- dplyr::group_by(percs, flag_hydraulicWide) %>% 
   dplyr::summarise(total = sum(n)) %>%
   dplyr::mutate(perc = 100 * total/sum(total))
 
-results_3_2_rivs <- read.csv('cache//validation//results_by_riv.csv')
-results_3_2_all <- read.csv('cache//validation//results_all_riv.csv')
-results_3_3_all <- read.csv('cache//FCO2//fco2_stats_all.csv')
-results_3_3_bulk <- read.csv('cache//FCO2//bulkFluxes.csv')
-results_3_3_bulk$key <- as.character(results_3_3_bulk$key)
+results_k600_rivs <- read.csv('cache//validation//results_by_riv.csv')
+results_k600_all <- read.csv('cache//validation//results_all_riv.csv')
+results_fco2_all <- read.csv('cache//FCO2//fco2_stats_all.csv')
+results_fco2_bulk <- read.csv('cache//FCO2//bulkFluxes.csv')
[…]

### 2021-09-30 src/old/swot_k_model_v2.R

@@ -11,15 +11,13 @@ library(cowplot)
 library(segmented)
 theme_set(theme_cowplot())
 
-setwd('C:\\Users\\craig\\Documents\\OneDrive - University of Massachusetts\\Ongoing Projects\\RSK600')
+setwd('C:\\Users\\cbrinkerhoff\\OneDrive - University of Massachusetts\\Ongoing Projects\\RSK600')
 
 #some constants
 g <- 9.8 #gravitational acceleration [m/s2]
 
 #######READ IN HYDRAULICS DATA FROM BRINKERHOFF ETAL 2019 TO GET SWOT VERSUS INEFFICIENT CHANNELS------------------------------
 data <- read.csv('data\\Brinkerhoff_etal_2019\\field_measurements.csv')
-#nhd_med <- read.csv('data\\Brinkerhoff_etal_2019\\NHD_join_table.csv')
-#data <- left_join(data, select(nhd_med, SLOPE, SOURCE_FEA), by = c("site_no" = "SOURCE_FEA"))
 
 #some necessary filtering
 data <- filter(data, is.finite(chan_width) ==1) %>%
@@ -30,8 +28,7 @@ data <- filter(data, is.finite(chan_width) ==1) %>%
   filter(chan_velocity > 0) %>%
   filter(chan_discharge > 0) %>%
   filter(chan_area > 0) %>%
-  filter(measured_rating_diff %in% c('Excellent', 'EXCL', 'GOOD', 'Good', 'Fair', 'FAIR'))# %>%
- # filter(is.finite(SLOPE)==1)
+  filter(measured_rating_diff %in% c('Excellent', 'EXCL', 'GOOD', 'Good', 'Fair', 'FAIR'))
 
 #imperial to metric
 data$area <- data$chan_area * 0.092903 #ft2 to m2
@@ -64,18 +61,6 @@ data <- left_join(data, bankfullD, 'site_no')
 data <- filter(data, width <= bankful_width)
 data <- filter(data, depth <= bankful_depth)
 
-#get Dingman r channel shape (f / b in AHG parlance) and only keep measurements made at sites with robust r estimates > 1 (Dingman 2007)
-# ahg <- group_by(data, site_no) %>%
-#   do(model_w = lm(log(width)~log(Qm3s), data=.),
-#      model_d = lm(log(depth)~log(Qm3s), data=.)) %>%
-#   mutate(b = model_w$coefficient[2],
-#          r2_w = summary(model_w)$r.squared,
[…]

### 2021-09-30 src/swot_k_model_v3.R

@@ -1,0 +1,205 @@
+#################
+##Description: Developing gas exchange model for 'hydraulically wide channels'
+              #It also estimates how many SWOT-observable rivers meet this hydraulically-wide condition
+##Creator: Craig Brinkerhoff
+##Date: Fall 2021
+
+##TO RUN:
+#  1) set the working directory below to the '~/RSK600' directory
+#  2) run script!
+#################
+
+#######SETUP-----------------------------
+library(tidyverse)
+library(colorspace)
+library(cowplot)
+library(segmented)
+theme_set(theme_cowplot())
+
+setwd('C:\\Users\\cbrinkerhoff\\OneDrive - University of Massachusetts\\Ongoing Projects\\RSK600')
+#setwd('C:\\Users\\craig\\Documents\\OneDrive - University of Massachusetts\\Ongoing Projects\\RSK600')
+
+#some constants
+g <- 9.8 #gravitational acceleration [m/s2]
+
+#######READ IN HYDRAULICS DATA FROM BRINKERHOFF ETAL 2019 TO GET SWOT VERSUS INEFFICIENT CHANNELS------------------------------
+data <- read.csv('data\\Brinkerhoff_etal_2019\\field_measurements.csv')
+
+#some necessary filtering
+data <- filter(data, is.finite(chan_width) ==1) %>%
+  filter(is.finite(chan_velocity)==1) %>%
+  filter(is.finite(chan_discharge)==1) %>%
+  filter(is.finite(chan_area)==1) %>%
+  filter(chan_width > 0) %>%
+  filter(chan_velocity > 0) %>%
+  filter(chan_discharge > 0) %>%
+  filter(chan_area > 0) %>%
+  filter(measured_rating_diff %in% c('Excellent', 'EXCL', 'GOOD', 'Good', 'Fair', 'FAIR'))
+
+#imperial to metric
+data$area <- data$chan_area * 0.092903 #ft2 to m2
+data$width <- data$chan_width*0.305 #m
+data$Vms <- data$chan_velocity*0.305 #m/s
+data$depth <- data$area / data$width
+data$slope <- NA #data$SLOPE
+data$Qm3s <- data$chan_discharge * 0.0283 #ft3/s to m3/s
+
+#calculate bankfull width so we can throw out out-of-bank events
+bankfullW = group_by(data, site_no) %>%
+  filter(n() >= 20) %>%
[…]

### 2021-10-01 manuscript_v5.Rmd
revalidation of biker using chainsaw model

@@
 We name the RS of $k_{600}$ algorithm BIKER, or the 'Bayesian Inference/Inversion of the $k_{600}$ Evasion Rate' and validate it for 47 SWOT-observable rivers from around the world using hydraulic models to produce SWOT-like data (as SWOT has not yet launched).
-We also quantify BIKER's sensitivity to the expected SWOT measurement errors on 17 of those rivers: while SWOT data represent a sea change in inland water monitoring, it is expected to have an approximately 10cm error in water surface elevation [@biancamariaSWOTMissionIts2016] as well as river width errors [@frassonExploringFactorsControlling2021].
+We also quantify BIKER's sensitivity to the expected SWOT measurement errors on 16 of those rivers: while SWOT data represent a sea change in inland water monitoring, it is expected to have an approximately 10cm error in water surface elevation [@biancamariaSWOTMissionIts2016] as well as river width errors [@frassonExploringFactorsControlling2021].
 Finally, we use previously published dissolved $CO_2$ data to represent a hypothetical in situ $CO_2$ sensor and compare the bulk carbon efflux from the 47 rivers as calculated using BIKER and previously published in situ techniques for predicting $k_{600}$.
@@
 We validate BIKER on 47 SWOT-simulated rivers using daily observed hydraulics and the metrics described in Table S2.
-We also re-validate BIKER on the 17 rivers with the SWOT error model which corrupts the river width and slope.
+We also re-validate BIKER on the 16 rivers with the SWOT error model which corrupts the river width and slope.
 Regardless of the validation setup, we do not have observed $k_{600}$ data for these rivers, and to our knowledge no field dataset of $k_{600}$ exists in the type of temporal and spatial frequency that SWOT (and therefore the BIKER algorithm) provides.

### 2021-10-01 lib/functions.R

@@ -39,7 +39,12 @@ calcdA_vec <- function(w, h) {
 }
 
 #k600 function using craig's model
-k600_ustar_craig <- function(Rh, slope) {
-  Ustar <- sqrt(g*Rh*slope)
-  return(56.0294*Ustar)
+#k600_ustar_craig <- function(Rh, slope) {
+#  Ustar <- sqrt(g*Rh*slope)
+#  return(56.0294*Ustar)
+#}
+
+#Calculate 'observed' k600 via the hydraulically-wide chainsaw model (see '~src\swot_k_model.R' for its validation)
+k600_model <- function(depth, slope) {
+  return(76.4*(g*slope)^(9/16)*(depth)^(11/16))
 }

### 2021-10-01 lib/inputs.R

@@ -8,9 +8,9 @@
 g <- 9.8 #gravitational acceleration [m/s2]
 molarMass <- 12.01 #[g/mol] for C
 pCO2_a <- 390 #[uatm]
-uncertainity <- 0.20 #see regression model for actual uncertainity
+uncertainity <- 0.30 #see regression model for actual uncertainity
 
 ###############
 #Implementation settings
 ###############
-cores <- 8 #num cores to run parallel processes on (depends on HPC session setup)
+cores <- 47 #num cores to run parallel processes on (depends on HPC session setup)

### 2021-10-01 src/FCO2_analysis.R

@@ -1,14 +1,14 @@
 ###########################
 #Description: Uses BIKER and rating curves to calculate FCO2 and median C fluxes from the SWOT simulated rivers
 #Craig Brinkerhoff
-#Winter 2021
+#Fall 2021
 ##########################
 
 print('starting FCO2...')
 
 #legend labels------------------
 legend_labels <- c('BIKER',
-                   'Model Using Observed \nshear velocity')
+                   'Model using observed \ndepth')
 
 ###################
 ##FUNCTION TO GENERATE FCO2 FROM ALL DEPTH MODELS AND ALL SWOT RIVERS--------------------------------
@@ -20,56 +20,49 @@ fun_FCO2_analysis <- function(river) {
   data_in =nc_open(river)
 
   W_obs=ncvar_get(data_in,'Reach_Timeseries/W')
-  H_obs=ncvar_get(data_in,'Reach_Timeseries/H')
   S_obs = ncvar_get(data_in, 'Reach_Timeseries/S')
   area = ncvar_get(data_in, 'Reach_Timeseries/A')
   Q_obs = ncvar_get(data_in, 'Reach_Timeseries/Q')
   reaches <- ncvar_get(data_in, 'River_Info/rch_bnd')
 
-  #Account for models known to not start on 1/1/XXXX. Seine starts in june and ___ starts in Feburary
-  start_date <- ncvar_get(data_in, 'Reach_Timeseries/t')
-  start_date <- ifelse(start_date[1] !=1, start_date[1], 1)
-  river_start_date <- format(as.Date('0000/01/01') + start_date, format='%m')
+  #Account for models known to not start on 1/1/XXXX.
+  start_date <- ncvar_get(data_in, 'Reach_Timeseries/t') #grab first date for river
+  start_date <- ifelse(start_date[1] !=1, start_date[1], 1) #if this isn't 1/1 , then use that first provided value. Otherwise, start in janurary.
+  river_start_date <- format(as.Date('0000/01/01') + start_date, format='%m') #convert to date formats via lubridate package
 
-  #prep pepsi rivers
+  #prep river hydraulic models
   S_obs[S_obs<=0]=NA
-  S_obs[is.na(S_obs)] = 0.000001 #min obs SWOT slope
[…]

### 2021-10-01 src/figures_k600.R

@@ -1,0 +1,290 @@
+########################
+#Description: Produce figures for BIKER runs on SWOT-simulated rivers
+#Creator: Craig Brinkerhoff
+#Date: Fall 2021
+######################
+
+print('validating BIKER...')
+
+`%notin%` <- Negate(`%in%`)
+
+################
+##READ IN RESULTS----------------------------------
+###############
+results <- read.csv('cache/validation/BIKER_validation_results.csv')
+full_output <- filter(results, errFlag == 0) #remove results with SWOT measurement error
+
+#Calculate r2 and rmse metrics
+lm_kfit <- lm(log10(full_output$kest_mean)~log10(full_output$kobs))
+r2 <- round(summary(lm_kfit)$r.squared, 2)
+full_output <- drop_na(full_output)
+rmse <- round((Metrics::rmse((full_output$kest_mean), (full_output$kobs))), 2)
+
+#Calculate prediction intervals for plot
+predInts <- predict(lm_kfit, interval='prediction')
+full_output <- cbind(full_output, predInts)
+
+####################
+##MODEL VALIDATION ACROSS ALL RIVERS AND TIMESTEPS--------------------------------------------------------
+####################
+valPlot <- ggplot(full_output, aes(x=(kobs), y=(kest_mean))) +
+  geom_abline(size=2, linetype='dashed', color='black') +
+  geom_pointrange(aes(ymin = kest_low, ymax = kest_high), fatten=10, fill='#1b9e77', pch=21, color='black') +
+  geom_smooth(size=2, color='darkgrey', method='lm', se=F)+
+  geom_line(aes(y=10^(lwr)), color='darkgrey', linetype='dashed', size=1.75) +
+  geom_line(aes(y=10^(upr)), color='darkgrey', linetype='dashed', size=1.75) +
+  xlab('k600 via observed \ndepth [m/dy]') +
+  ylab('BIKER k600 [m/dy]') +
+  scale_y_log10(
+      breaks = scales::trans_breaks("log10", function(x) 10^x),
+      labels = scales::trans_format("log10", scales::math_format(10^.x)))+
+  scale_x_log10(
+      breaks = scales::trans_breaks("log10", function(x) 10^x),
+      labels = scales::trans_format("log10", scales::math_format(10^.x)))+
+  scale_color_discrete_qualitative(palette = 'Harmonic') +
+  theme(legend.position = "none",
[…]

### 2021-10-01 src/figures_k600.R.R

@@ -1,0 +1,290 @@
+########################
+#Description: Produce figures for BIKER runs on SWOT-simulated rivers
+#Creator: Craig Brinkerhoff
+#Date: Fall 2021
+######################
+
+print('validating BIKER...')
+
+`%notin%` <- Negate(`%in%`)
+
+################
+##READ IN RESULTS----------------------------------
+###############
+results <- read.csv('cache/validation/BIKER_validation_results.csv')
+full_output <- filter(results, errFlag == 0) #remove results with SWOT measurement error
+
+#Calculate r2 and rmse metrics
+lm_kfit <- lm(log10(full_output$kest_mean)~log10(full_output$kobs))
+r2 <- round(summary(lm_kfit)$r.squared, 2)
+full_output <- drop_na(full_output)
+rmse <- round((Metrics::rmse((full_output$kest_mean), (full_output$kobs))), 2)
+
+#Calculate prediction intervals for plot
+predInts <- predict(lm_kfit, interval='prediction')
+full_output <- cbind(full_output, predInts)
+
+####################
+##MODEL VALIDATION ACROSS ALL RIVERS AND TIMESTEPS--------------------------------------------------------
+####################
+valPlot <- ggplot(full_output, aes(x=(kobs), y=(kest_mean))) +
+  geom_abline(size=2, linetype='dashed', color='black') +
+  geom_pointrange(aes(ymin = kest_low, ymax = kest_high), fatten=10, fill='#1b9e77', pch=21, color='black') +
+  geom_smooth(size=2, color='darkgrey', method='lm', se=F)+
+  geom_line(aes(y=10^(lwr)), color='darkgrey', linetype='dashed', size=1.75) +
+  geom_line(aes(y=10^(upr)), color='darkgrey', linetype='dashed', size=1.75) +
+  xlab('k600 via observed \ndepth [m/dy]') +
+  ylab('BIKER k600 [m/dy]') +
+  scale_y_log10(
+      breaks = scales::trans_breaks("log10", function(x) 10^x),
+      labels = scales::trans_format("log10", scales::math_format(10^.x)))+
+  scale_x_log10(
+      breaks = scales::trans_breaks("log10", function(x) 10^x),
+      labels = scales::trans_format("log10", scales::math_format(10^.x)))+
+  scale_color_discrete_qualitative(palette = 'Harmonic') +
+  theme(legend.position = "none",
[…]

### 2021-10-01 src/main.R

@@ -8,6 +8,7 @@ load.project()
 theme_set(theme_cowplot())
 
 #Run analysis-------------------------------
-#source(here::here('src' , 'runBIKER.R'))
-#source(here::here('src' , 'validation.R'))
+source(here::here('src' , 'runBIKER.R'))
 source(here::here('src' , 'FCO2_analysis.R'))
+source(here::here('src' , 'figures_k600.R'))
+source(here::here('src' , 'figures_FCO2.R'))

### 2021-10-05 lib/functions.R
first push for running validation on eD+chainsaw model

@@ -45,6 +45,7 @@ calcdA_vec <- function(w, h) {
 #}
 
 #Calculate 'observed' k600 via the hydraulically-wide chainsaw model (see '~src\swot_k_model.R' for its validation)
-k600_model <- function(depth, slope) {
-  return(76.4*(g*slope)^(9/16)*(depth)^(11/16))
+k600_model <- function(depth, slope, vel) {
+  #return(76.4*(g*slope)^(9/16)*(depth)^(11/16)) #CHAINSAW MODEL FOR ES
+  return(62.82*(g*slope)^(7/16)*vel^(1/4)*D_obs^(9/16)) #CHAINSAW MODEL FOR ED
 }

### 2021-10-05 src/figures_FCO2.R

@@ -4,6 +4,12 @@
 #Date: Fall 2021
 ######################
 
+#for local runs
+library(tidyverse)
+library(cowplot)
+theme_set(theme_cowplot())
+setwd('C:\\Users\\craig\\Documents\\OneDrive - University of Massachusetts\\Ongoing Projects\\RSK600')
+
 print('validating BIKER + CO2 data...')
 
 `%notin%` <- Negate(`%in%`)
@@ -130,3 +136,42 @@ ggsave('cache/FCO2/FCO2_models.jpg', FCO2_models, width=14, height=8)
 #################
 results_all_rivs <- data.frame('rmse'=rmse, 'r2'=lmr2)
 write.csv(results_all_rivs, 'cache/FCO2/fco2_stats_all.csv')
+
+
+
+
+
+#testing--------------------------
+massFluxes_byRiver <- group_by(output, river) %>%
+  filter(n() >= 3) %>% #only keep rivers with at least 3 measurements to get by-river medians
+  gather(key=key, value=value, c(FCO2_BIKER, FCO2_raymond2012, FCO2_raymond2013, FCO2_brinkerhoff2019)) %>%
+  group_by(river, key) %>%
+  summarise(sumFCO2 = mean(value * 1e-12 * 365, na.rm=T)*total_sa, #tG-C/yr from all rivers
+            medianFCO2 = median(value * 1e-12 * 365, na.rm=T)*total_sa,  #tG-C/yr from all rivers
+            obsFCO2 = mean(FCO2_obs * 1e-12 * 365, na.rm=T)*total_sa)
+
+stats <- group_by(massFluxes_byRiver, key) %>%
+  summarise(rmse = Metrics::rmse(obsFCO2, medianFCO2),
+            r2 = summary(lm(medianFCO2~obsFCO2))$r.squared,
+            bias =   Metrics::bias(obsFCO2, medianFCO2))
+
+massFluxes_models <- group_by(massFluxes_byRiver, key) %>%
+  do(model = lm((medianFCO2)~(obsFCO2), data=.)) %>%
+  mutate(r2 = summary(model)$r.squared,
+         slope = model$coefficients[2],
+         int = (model$coefficients[1]))
+
+ggplot(massFluxes_byRiver[massFluxes_byRiver$key == 'FCO2_brinkerhoff2019',], aes(x= obsFCO2, y=medianFCO2, color=key)) +
+  geom_point(size=5, alpha=0.50)+
+  geom_smooth(method='lm', se=F, size=5) +
[…]

### 2021-10-05 src/runBIKER.R

@@ -21,12 +21,14 @@ run_BIKER <- function(currPepsi, errFlag) {
   H_obs=ncvar_get(data_in,'Reach_Timeseries/H')
   S_obs = ncvar_get(data_in, 'Reach_Timeseries/S')
   area = ncvar_get(data_in, 'Reach_Timeseries/A')
+  Q_obs=ncvar_get(data_in,'Reach_Timeseries/Q')
 
   #prep river hydraulics-----------------------------------------------------------
   S_obs[S_obs<=0]=NA
   S_obs[is.na(S_obs)] = 0.000017 #min obs SWOT slope Biancarma etal 2016
   W_obs[W_obs<0]=NA
   H_obs[H_obs<0]=NA
+  Q_obs[Q_obs<0]=NA
   area[area<0]=NA
 
   #Some NA handling in slopes (borrowed from Frassion etal 2021 BAM runs)-----------------------------------------------------------
@@ -36,6 +38,7 @@ run_BIKER <- function(currPepsi, errFlag) {
     W_obs=W_obs[,-remove_index]
     H_obs=H_obs[,-remove_index]
     S_obs=S_obs[,-remove_index]
+    Q_obs=Q_obs[,-remove_index]
     area=area[,-remove_index]
   }
 
@@ -45,6 +48,7 @@ run_BIKER <- function(currPepsi, errFlag) {
     W_obs=W_obs[-remove_index,]
     H_obs=H_obs[-remove_index,]
     S_obs=S_obs[-remove_index,]
+    Q_obs=Q_obs[,-remove_index]
     area=area[,-remove_index]
   }
 
@@ -74,16 +78,18 @@ run_BIKER <- function(currPepsi, errFlag) {
 
     #calculate observed k600 with no measurement error
     Dtrue <- area/Wtrue
+    V_obs <- Q_obs/area
     #Rhtrue <- area / (Wtrue + 2*Dtrue)
-    k_obs <- k600_model(Dtrue, Strue) #k600 equation
+    k_obs <- k600_model(Dtrue, Strue, V_obs) #k600 equation
     k_obs <- colMeans(k_obs, na.rm=T)
   }
 
   #Calculate observed k600 with no measurement error
   else {
-    D_obs <- area/W_obs
+    D_obs <- area/W_obs #[m]
+    V_obs <- Q_obs/area #[m/s]
     #Rh_obs <- area / (W_obs + 2*D_obs)
[…]

### 2021-11-16 scraps.Rmd
'manuscript_v6'

Beyond $CO_2$ emissions, *k* for various dissolved gases are of significant importance to aquatic ecologists modeling stream metabolism [e.g. @bernhardtMetabolicRegimesFlowing2018] and water quality engineers modeling river responses to pollutant loadings [@chapraSurfaceWaterQualityModeling2008]. Because of this far-reaching importance, decades of work have focused on elucidating the physical mechanisms behind *k* in oceans and lakes [@wanninkhofAdvancesQuantifyingAirSea2009 for a review] and, to a lesser extent, fluvial gas exchange [see @hallGasExchangeStreams2020 for a review].

All of these mechanistic uncertainties are then propagated through upscaling workflows when biogeochemists predict *k* across thousands of rivers [e.g @borgesGloballySignificantGreenhousegas2015; @horgbyUnexpectedLargeEvasion2019; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013] most often via equation 1 coupled with hydraulic geometry (HG). HG is the scaling relationships between streamflow and river channel hydraulics [@leopoldHydraulicGeometryStream1953; @gleasonHydraulicGeometryNatural2015] that allow biogeochemists to predict river depth and velocity from streamflow models and/or streamgauges . It is currently not well understood how sensitive global estimates of fluvial gas evasion are to the specific HG model that is employed by the worker. Further, these approaches rely on either in situ discharge records or modeled streamflow which further contributes uncertainty. This is all exacerbated in ungauged basins that cover large areas, especially in the carbon-rich Arctic inland waters, where little in situ information is available and fieldwork is impractical [@gleasonRemoteSensingRiver2020].

from remote sensing (RS) data. Remote sensing of river hydraulics is a burgeoning subfield within remote sensing of hydrology, often in service of remote sensing of river discharge [RSQ- @gleasonRemoteSensingRiver2020]. This is accomplished via two general approaches: 'gauged' methods which rely on in situ river data to calibrate a method to the river(s) at hand and 'ungauged' techniques which focus on hydraulic generalizability in the service of merely improving existing knowledge in data-poor domains [@gleasonRemoteSensingRiver2020]. Many, but not all, of these ungauged approaches are developed in the context of the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission. SWOT is expected to launch in 2022 and provide the world's first global measurements of water surface extent and elevation at novel temporal resolutions. SWOT is a wide swath radar interferometer and will sample rivers every 1 to 7 days per 21 day repeat cycle and will measure rivers wider than 100m with a goal of expanding this to rivers at least 50m wide [@biancamariaSWOTMissionIts2016]. A decade of SWOT work has explored the multi-parameter problem of estimating the river hydraulic parameters of roughness and bathymetry from remote sensing to produce the SWOT discharge product [e.g. @andreadisConstrainingAssimilationSWOT2020; @brinkerhoffConstrainingRemoteRiver2020; @durandEstimatingReachaveragedDischarge2014; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014a; @garamboisVariationalEstimationEffective2020; @brissetAssimilationAltimetricData2018; @hagemannBAMBayesianAMHGManning2017; @larnierRiverDischargeBathymetry2020a; @oubanasDischargeEstimationUngauged2018].

-------------------------------------------------

Given that *k* can only be directly measured via a known gas concentration gradient, eddy-covariance measurements, or tracer additions to the stream [@hallGasExchangeStreams2020], SWOT data will not provide the necessary parameters to directly measure *k*. Further, these techniques are entirely infeasible at the global scale where our goal with BIKER is primarily to better understand spatiotemporal patterns of *k* rather than obtain absolute measurements of it. Therefore, we need a simple yet physically-based model for *k* to directly relate channel hydraulics to gas exchange in the large rivers that SWOT will measure. We hypothesize that this can be done by leveraging the unique geomorphology of the very wide rivers that SWOT will observe. In this section, we develop and test process-based models for *k*, selecting one that is then implemented within BIKER (section 5).

 In particular, there is uncertainty as to which models are most appropriate at different locations along the stream-to-river continuum given that river hydraulics change predictably along a river's longitudinal profile [@wangPhysicallyBasedScaling2021]. In small and steep headwaters, 'bubble-mediated gas exchange' in whitewater might lead to substantially elevated $k_{fluvial}$ where bedload is large and general channel roughness is extremely high [@mauriceInfluenceFlowBed2017; @ulsethDistinctAirWater2019].

### 2021-11-16 generateMS.R

@@ -2,31 +2,31 @@
 #This will grab and store all results in your current RStudio environment so that you can knit the manuscript correctly. It then knits the manuscript too.
 #################
 
-library(tidyverse)
+library(dplyr)
 
-setwd("C:/Users/cbrinkerhoff/OneDrive - University of Massachusetts/Ongoing Projects/RSK600")
-#setwd("C:/Users/craig/Documents/OneDrive - University of Massachusetts/Ongoing Projects/RSK600")
+#setwd("C:/Users/cbrinkerhoff/OneDrive - University of Massachusetts/Ongoing Projects/RSK600")
+setwd("C:/Users/craig/Documents/OneDrive - University of Massachusetts/Ongoing Projects/RSK600")
 
 #Move all results to global environment
-#turbulence_stats <- read.csv('cache\\k600_theory\\turbulence_stats.csv')
-#models_ulseth <- read.csv('cache\\k600_theory\\ulseth_models.csv')
-#models_ustar <- read.csv('cache\\k600_theory\\ustar_models.csv')
-#models_eD4 <- read.csv('cache\\k600_theory\\eD_models.csv')
 modelsWIDE <- read.csv('cache\\k600_theory\\hydraulicWide_models.csv')
 percs <- read.csv('cache\\k600_theory\\Rh_H_percents.csv')
-percs_overall <- dplyr::group_by(percs, flag_hydraulicWide) %>% 
-  dplyr::summarise(total = sum(n)) %>%
-  dplyr::mutate(perc = 100 * total/sum(total))
+percs_overall <- group_by(percs, flag_hydraulicWide) %>% 
+  summarise(total = sum(n)) %>%
+  mutate(perc = 100 * total/sum(total))
 
 results_k600_rivs <- read.csv('cache//validation//results_by_riv.csv')
-results_k600_all <- read.csv('cache//validation//results_all_riv.csv')
-results_fco2_all <- read.csv('cache//FCO2//fco2_stats_all.csv')
-results_fco2_bulk <- read.csv('cache//FCO2//bulkFluxes.csv')
+results_fco2_rivs <- read.csv('cache//FCO2//CO2_results.csv')
+results_fco2_bulk <- read.csv('cache//FCO2//massFluxes.csv')
 results_fco2_bulk$key <- as.character(results_fco2_bulk$key)
 
 #knit manuscript
-rmarkdown::render("manuscript_v5.Rmd")
[…]

### 2021-11-16 src/figures_FCO2.R

@@ -5,10 +5,13 @@
 ######################
 
 #for local runs
-library(tidyverse)
+library(ggplot2)
+library(dplyr)
+library(tidyr)
 library(cowplot)
+library(hydroGOF)
 theme_set(theme_cowplot())
-#setwd('C:\\Users\\craig\\Documents\\OneDrive - University of Massachusetts\\Ongoing Projects\\RSK600')
+setwd('C:\\Users\\cbrinkerhoff\\OneDrive - University of Massachusetts\\Ongoing Projects\\RSK600')
 
 print('validating BIKER + CO2 data...')
 
@@ -115,14 +118,14 @@ stats <- gather(output, key=key, value=value, c(FCO2_BIKER, FCO2_raymond2012, FC
   group_by(river) %>%
   filter(n() >= 3) %>%
   group_by(river, key) %>%
-  summarise(r2 = summary(lm(value~FCO2_obs))$r.squared,
+  summarise(r = sqrt(summary(lm(value~FCO2_obs))$r.squared),
             NRMSE = sqrt(mean((FCO2_obs - value)^2, na.rm=T)) / mean(FCO2_obs, na.rm=T),
             rBIAS =   mean(value- FCO2_obs, na.rm=T) / mean(FCO2_obs, na.rm=T),
             KGE =   KGE(value, FCO2_obs), #sqrt(mean((kobs- kest_mean)^2 / kobs^2, na.rm=T)),
             n_data=n())
 
-plot_stats <- gather(stats, key=key2, value=value2, c('NRMSE', 'rBIAS', 'KGE', 'r2'))
-plot_stats <- filter(plot_stats, key2 %in% c('NRMSE', 'KGE', 'rBIAS', 'r2'))
+plot_stats <- gather(stats, key=key2, value=value2, c('NRMSE', 'rBIAS', 'KGE', 'r'))
+plot_stats <- filter(plot_stats, key2 %in% c('NRMSE', 'KGE', 'rBIAS', 'r'))
 lowKGE <- nrow(plot_stats[plot_stats$key2 == 'KGE' & plot_stats$value2 < -2,])
 
 boxPlots <- ggplot(plot_stats, aes(x=key2, y=value2, fill=key))+

### 2021-11-16 src/figures_k600.R

@@ -40,14 +40,12 @@ valPlot <- ggplot(full_output, aes(x=(kobs), y=(kest_mean))) +
   geom_abline(size=3, linetype='dashed', color='darkgrey') +
   xlab('k600 via observed \nhydraulics [m/dy]') +
   ylab('BIKER k600 [m/dy]') +
-  scale_y_log10(
-      limits = c(10^-1, 10^2),
-      breaks = scales::trans_breaks("log10", function(x) 10^x),
-      labels = scales::trans_format("log10", scales::math_format(10^.x)))+
-  scale_x_log10(
-      limits = c(10^-1, 10^2),
-      breaks = scales::trans_breaks("log10", function(x) 10^x),
-      labels = scales::trans_format("log10", scales::math_format(10^.x)))+
+  scale_y_log10(limits=c(10^-1,10^2),
+                breaks=c(0.1, 1, 10, 100),
+                labels=c('0.1', '1', '10', '100'))+
+  scale_x_log10(limits=c(10^-1,10^2),
+                breaks=c(0.1, 1, 10, 100),
+                labels=c('0.1', '1', '10', '100'))+
   scale_fill_discrete_qualitative(palette = 'Dark3') +
   annotate("text", x = 10^-0.4, y = 10^1.5, label = '(Colors correspond \nto rivers)', size=6)+
   theme(legend.position = "none",
@@ -55,28 +53,31 @@ valPlot <- ggplot(full_output, aes(x=(kobs), y=(kest_mean))) +
         axis.title=element_text(size=24,face="bold"),
         legend.text = element_text(size=17),
         legend.title = element_text(size=17, face='bold'))
+ggsave('cache/validation/validation.jpg', valPlot, width=8, height=8)
 
 ########################
 ##CALCULATE BY-RIVER ERROR METRICS---------------------------------------------
 ########################
 stats_by_river <- group_by(results, river, errFlag) %>%
-  summarise(r2 = summary(lm(kest_mean~kobs))$r.squared,
+  summarise(r = sqrt(summary(lm(kest_mean~kobs))$r.squared),
             NRMSE = sqrt(mean((kobs - kest_mean)^2, na.rm=T)) / mean(kobs, na.rm=T),
[…]

### 2021-11-16 src/swot_k_model_v3.R

@@ -10,14 +10,15 @@
 #################
 
 #######SETUP-----------------------------
-library(tidyverse)
+library(dplyr)
+library(tidyr)
+library(ggplot2)
 library(colorspace)
 library(cowplot)
-library(segmented)
 theme_set(theme_cowplot())
 
-setwd('C:\\Users\\cbrinkerhoff\\OneDrive - University of Massachusetts\\Ongoing Projects\\RSK600')
-#setwd('C:\\Users\\craig\\Documents\\OneDrive - University of Massachusetts\\Ongoing Projects\\RSK600')
+#setwd('C:\\Users\\cbrinkerhoff\\OneDrive - University of Massachusetts\\Ongoing Projects\\RSK600')
+setwd('C:\\Users\\craig\\Documents\\OneDrive - University of Massachusetts\\Ongoing Projects\\RSK600')
 
 #some constants
 g <- 9.8 #gravitational acceleration [m/s2]
@@ -105,151 +106,150 @@ data$flag_hydraulicWide <- ifelse(data$Rh/data$depth > 0.96, 'Rh=H', 'Rh=/=H') #
 percs <- group_by(data, flag_swot, flag_hydraulicWide) %>% summarise(n=n()) %>% group_by(flag_swot) %>% mutate(perc = 100 * n/sum(n))
 write.csv(percs, 'cache/k600_theory/Rh_H_percents.csv')
 
-##########PLOT CDF OF SWOT OBSERVATIONS
-swot_cdf <- ggplot(data[data$width > 100,], aes(x=Rh/depth*100)) +
-  stat_ecdf(size=3, color='#fc8d62') + 
-  xlim(0,100) +
-  xlab('Rh/H ratio [%]') +
-  ylab('Density')+
-  ggtitle('Swot-Observable Rivers')+
-  theme(axis.text=element_text(size=19),
-        axis.title=element_text(size=24,face="bold"),
-        legend.text = element_text(size=17),
-        legend.title = element_text(size=17, face='bold'),
-        legend.position = 'none')
-
 #######LOG TRANSFORM SOME VARIABLES-----------------------------------------------------------
 data$log_slope <- log(data$slope)
 
-######## K600 MODEL--------------------
+######## K600 MODELS--------------------
 #hydraulically-wide rivers
 data <- filter(data, study %in% c('Ulseth_etal_2019', 'Churchill_etal_1962', 'Owens_etal_1964'))
[…]

### 2022-01-07 docs/scraps.Rmd
fixed figure themes and some ms drafting

@@
+To assess the influence of SWOT's implicit measurement error on BIKER performance, we use the error models developed by @durandHowWillRadar2020 and @frassonExploringFactorsControlling2021, which the latter implemented on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])`/`r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` of the rivers.
+Error in SWOT measurements will come from both the error tolerances intrinsic in the satellite data product as well as radar layover error and river width error.
+Layover error is the phenomenon when radar returns from different places arrive at the sensor at the same time, leading to taller landscape features appearing closer to the sensor than shorter landscape features that are the same horizontal distance from the sensor [@durandHowWillRadar2020].
+Width errors due to poor water classification are ignored but width errors due to the cross-track distance are included [@frassonExploringFactorsControlling2021].
+The cross-track distance, reflective of the amount of the river cross-section that falls within SWOT's swath, controls the size of the pixels **help, I on't understand cross-track distance**.
+@frassonExploringFactorsControlling2021 tested a suite of measurement error scenarios, however here we test only their 'full uncertainty scenario' that most closely represents the measurement errors that will be implicit in SWOT data.
+*Should this be more elaborately explained now that these data descriptions are moved to the BIKER section?*
+
+To do this, we start from 'McFLI logic' as used in some SWOT RSQ algorithms, where hydraulic properties vary along cross-sections *i* within the reach but the entire mass of water moving through is conserved and thus only varies with timestep *t* [@gleasonTrackingRiverFlows2017].
+McFLI logic is laid out in Figure 4.
+This model implies that the reach-averaged $k_{600}$, which is ultimately a function of streamflow per equations 5-8 [@wangPhysicallyBasedScaling2021], will only vary temporally as well.
+Under this conceptual river reach model, BIKER predicts a time-varying, reach-averaged $k_{600_t}$.
+
 Turbulence is fundamental to fluvial geomorphology and open-channel flows and has been the subject of extensive research across scientific and engineering disciplines for over a century.

### 2022-01-07 docs/supp_v6.Rmd
fixed figure themes and some ms drafting

---
title: "Supplemental Information: Exploiting hydraulic geometry to remotely sense gas exchange velocity via the SWOT satellite"
author:
  - Craig B Brinkerhoff:
      email: cbrinkerhoff@umass.edu
      institute: UMass
      correspondence: true
  - Peter A Raymond:
      institute: Yale
  - Colin J Gleason:
      institute: UMass
  - Christopher J Zappa:
      institute: Lamont
  - Merritt H Harlan:
      institute: UMass
institute:
  - UMass: Department of Civil & Environmental Engineering, University of Massachusetts, Amherst, MA
  - Yale: School of the Environment, Yale University, New Haven, CT
  - Lamont: Lamont-Doherty Earth Observatory, Columbia University, Palisades, NY
bibliography: References.bib
csl: agu_citation_style.csl
output:
  word_document:
      pandoc_args:
        - '--lua-filter=scholarly-metadata.lua'
        - '--lua-filter=author-info-blocks.lua'
      reference_docx: word-styles-reference.docx

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Contents
This supplementary information contains 4 texts, 1 figure and 3 tables. Please consult https://github.com/craigbrinkerhoff/RSK600 for all code to build and generate results, figures, and the manuscript.

## Test S1: Estimating how many SWOT rivers are hydraulically wide
To quantify how frequently SWOT-observable rivers are also hydraulically wide river channels (section 3 of main text), we used the dataset of field-measured river hydraulics in @brinkerhoffReconcilingAtaStationAtManyStations2019. That dataset has over 500,000 discrete measurements of river width, velocity, area, and discharge that were made by the United States Geological Survey (USGS) to calibrate streamgauge rating curves. Here, we describe how this dataset was filtered down to 171,553 measurements and how we quantified what was 'hydraulically wide'.

First, we removed all measurements tagged by the USGS as 'poor', measurements with impossible values, or measurements of 0. While this would indicate a dry channel, our hydraulic geometry model necessitates within-bank flow. Likewise, because hydraulic geometry only applies to within-bank flows and not flood events, we remove all overbank flows. This was done by first filtering for sites with at least 20 measurements (to build robust estimates of bankfull hydraulics) and then calculating bankfull width and depth as the width or depth with a return period of two years. While the only true way to calculate bankfull hydraulics is manually in the field, this is obviously impractical here. A two year return period is a standard approximation for determining out-of-bank flow in single-channel meandering rivers and was the method used by @brinkerhoffReconcilingAtaStationAtManyStations2019. We then removed all measurements with a width or depth beyond their respective at-a-station 2 year values.

After joining this dataset with the hydraulics measurements provided with the 763 $k_{600}$ measurements, we classified each measurement as hydraulically wide if its hydraulic radius was within 4% of its mean flow depth. Hydraulic radius was calculated assuming a rectangular channel such that $R_H=2H+W$. Because these measurements were almost entirely made at streamgauges, which generally have a location bias favoring stable channels near bridges and other structures that yield lower width dynamics than those observed away from gauges [@allenPatternsRiverWidth2015; @parkWorldwideVariationsHydraulic1977], we used a wider threshold of 4%. In the 'width-limited' scenario common at streamgauges, the hydraulic radius is likely further from the mean flow depth than would occur naturally. The slightly more liberal threshold of 4% allowed for implicit accounting for this sampling bias.

## Text S2: Gas exchange model derivations
In this text we provide the full algebra to arrive at the four physically-based gas exchange models in the main text (equations 3-6). Equation 3 is derived via equations S1-S5. Equation 4 is derived via equations S6-S7. Equation 5 is derived via equations S8-S13. Equation 6 is derived via equations S14-S18. Consult Appendix A for all variable definitions. Note that $\alpha$ and $\beta$ are statistical parameters obtained via least squares regression for the small-eddy and chainsaw models, respectively. Subscripts denote variants of the same parameter, depending on which $\epsilon$ model is employed.

**Small-eddy models**

*log-law-of-the-wall dissipation*
$$k_{600}=\alpha(\epsilon_S)^{1/4} \mathbf{(S1)}$$
$$k_{600}=\alpha(\frac{U_*^3}{H})^{1/4} \mathbf{(S2)}$$
$$k_{600}=\alpha(\frac{\sqrt{gR_hS}^3}{H})^{1/4} \mathbf{(S3)}$$
$$k_{600}=\alpha(\frac{(gHS)^{3/2}}{H})^{1/4} \mathbf{(S4)}$$
$$k_{600}=\alpha(gS)^{3/8}H^{1/8} \mathbf{(S5)}$$

*form-drag dissipation*
$$k_{600}=\alpha_1(\epsilon_D)^{1/4} \mathbf{(S6)}$$
$$k_{600}=\alpha_1(gS\bar{U})^{1/4} \mathbf{(S7)}$$

**Chainsaw models**

*log-law-of-the-wall dissipation*
$$k_{600}=\beta_0(\epsilon_S)^{1/4}A_p^{3/4} \mathbf{(S8)}$$
$$k_{600}=\beta (\frac{U_*^3}{H})^{1/4}((U_*H)^{1/2})^{3/4} \mathbf{(S9)}$$
$$k_{600}=\beta U_*^{9/8}H^{1/8} \mathbf{(S10)}$$
$$k_{600}=\beta (gR_hS)^{9/16}H^{1/8} \mathbf{(S11)}$$
$$k_{600}=\beta (gHS)^{9/16}H^{1/8} \mathbf{(S12)}$$
$$k_{600}=\beta (gS)^{9/16}H^{11/16} \mathbf{(S13)}$$

*form-drag dissipation*
$$k_{600}=\beta_0(\epsilon_D)^{1/4}A_p^{3/4} \mathbf{(S14)}$$
$$k_{600}=\beta_1(gS\bar{U})^{1/4}((U_*H)^{1/2})^{3/4} \mathbf{(S15)}$$
$$k_{600}=\beta_1(gS\bar{U})^{1/4}(gR_hS)^{3/16}H^{3/8} \mathbf{(S16)}$$
$$k_{600}=\beta_1(gS\bar{U})^{1/4}(gHS)^{3/16}H^{3/8} \mathbf{(S17)}$$
$$k_{600}=\beta_1(gS)^{7/16}\bar{U}^{1/4}H^{9/16} \mathbf{(S18)}$$

## Text S3: BIKER hyperparameterization
In this text we explain in detail how BIKER's hyperparameter values were set for each river.

We assign prior hyperparameters using SWOT data only. All priors are formalized within the model as truncated normal distributions of the log-transformed terms such that $log(X) \sim  \mathcal{N}(\mu, \sigma^2)$ for $\lambda < log(X) < \gamma$, using prior hyperparameters mean ($\mu$), standard deviation ($\sigma$), and upper ($\gamma$) and lower bounds ($\lambda$) for any parameter *X*.

$A_0_i$ and $n_i$ prior hyperparameters were assigned following an updated version of the method developed by @brinkerhoffConstrainingRemoteRiver2020. They developed a set of river channel prior hyperparameters for McFLI algorithms that are entirely RS-able and reflect differential channel hydraulics as a function of river geomorphology. They used an extensive database of field measurements and machine learning to identify patterns that associate river width with the hydraulic priors needed to run McFLIs so that prior hyperparameters may be assigned to rivers using only the existing remotely sensed data. For this study, we extracted $A_0_{\lambda}$ and $A_0_{\gamma}$ as the 5th and 95th percentile values rather than the absolute maximum and minimum values to avoid physically impossible bounds on $A_0$.

This leaves the $k_{600_t}$ hyperparameters to be defined. $k_{600_{\mu}}$ is set by invoking the hydraulic geometry (HG) relationships developed in section 5.1.3 of the main text using the data from @brinkerhoffReconcilingAtaStationAtManyStations2019. We replaced both depth and velocity terms from our gas exchange model (equation 6 in the main text) with these HG models, resulting in equation S19 where Q is the mass-conserved streamflow for the river reach.

$$k_{600_{\mu}} = 62.82(gS_{i,t})^{7/16}(0.265Q_{t}^{0.395})^{9/16}*(0.284Q_{t}^{0.191})^{1/4} \mathbf{(S18)}$$

Obviously, we have no a priori information about Q. So, we use the Q prior that will be globally available when SWOT launches: a mean annual estimate from a water balance model. This is provided with the simulated SWOT data by both @durandIntercomparisonRemoteSensing2016 and @frassonExploringFactorsControlling2021. Again, using a temporally-invariant estimate of streamflow is the worst case scenario and BIKER's performance will improve with a more informed prior om streamflow (and therefore $k_{600}$). However as noted in the main text, our primary goal with this initial validation is to benchmark BIKER's worst case scenario for performance and so we do that here. Because BIKER treats every timestep as a different parameter, future work should investigate ways to assign temporally varying priors for $k_{600}$.

$k_{600_{\sigma}}$ is set to 0.30 (log-space). This corresponds to a coefficient of variation of approximately 0.30, which we took to reflect a reasonably strong agreement between the prior and the observed values. $k_{600_{\lambda}}$ was set to log(0.001) m/day. $k_{600_{\gamma}}$ was set to log(500) m/day. The

## Text S4: Specifiying the $\sigma_{k_{600}}$ parameter
Here, we detail how the $\sigma_{k_{600}}$ parameter is specified in this study, as well as how it should be specified once SWOT launches.

Recall that $\sigma_{k_{600}}$ refers to the total uncertainty inherent in equation 8, i.e. stemming from both the chainsaw/form-drag dissipation model and the Manning's model for $\bar{U}$. For the purposes of this study, we are validating BIKER against equation 10 in the main text and so all uncertainties associated with the $\beta_1$ parameter are ignored and we only need to reflect the Manning's uncertainty in our specification of $\sigma_{k_{600}}^2$. Therefore, we take @hagemannBAMBayesianAMHGManning2017's estimated uncertainty from Manning's equation to infer streamflow from SWOT observations (0.25) and inflate it slightly to also account for the hydraulically-wide channel assumption and arrive at 0.30. This is the$\sigma_{k_{600}}$ used in this study.

In the scenario that BIKER is run on real SWOT data, $\sigma_k_{600}$ must reflect the full uncertainty implicit in equation 8 in the main text. This means we must also account for uncertainty from the $\beta_1$ parameter. Assuming perfect measurements from are made by the SWOT satellite, the $k_{600}$ relative uncertainty is expressed for some set of hydraulic observations as equation S19. We use the 166 hydraulically-wide measurements in our field-measured $k_{600}$ dataset and run 166 different Monte Carlo simulations (each 10,000 runs) to obtain 166 different $\sigma_k_{600}$ terms. We then take the average of those to be a reasonable estimate of $\sigma_k_{600}$. This ultimately provided a value of 1.1 for the log-transformed $\sigma_k_{600}$ that should be used once SWOT launches.

$$log_{k_{600}} \sim normal \left[ log(\beta_1), log(\sigma_{\beta_1}) \right] + \frac{1}{4}normal\left[log(\bar{U}), log(\sigma_{mannings})\right] \frac{7}{16}log(gS) + \frac{9}{16}log(H)  \mathbf{(S19)}$$

## Figure S1
![Figure S1 Timeseries of the biweekly CO2 data from @beaulieuControlsGasTransfer2012. Sampling took place 2008-2009 in the Ohio River (upstream of Cincinnati, Ohio, United States). Each point here was joined to the 11-day SWOT observations used in this study (section 2.4).](cache/FCO2/Beaulieu_timeseries.jpg)

## Table S1
*Table S1: Studies that gas exchange velocity measurements come from. 'Study' refers to the paper from which these measurements were taken. Any data wrangling was done by those authors. 'Field Workers' refers to who actually made the measurements. The Raymond et al. (2012) dataset is itself a meta-analysis. Please see that paper for how those measurements were collected. We used the set of measurements ultimately published by Ulseth et al. 2019.*

|**Study**|**Workers**|**Number of measurements**|
|---------|-------------|--------------------------|
|@ulsethDistinctAirWater2019|@ulsethDistinctAirWater2019||
|@ulsethDistinctAirWater2019|@hallUseArgonMeasure2018||
|@ulsethDistinctAirWater2019|@schelkerCO2EvasionSteep2016||
|@ulsethDistinctAirWater2019|@mauriceInfluenceFlowBed2017||
|@ulsethDistinctAirWater2019|@raymondScalingGasTransfer2012||
|@churchillPredictionStreamReaeration1964|@churchillPredictionStreamReaeration1964||
|@owensReaerationStudiesStreams1964|@owensReaerationStudiesStreams1964||
<!-- |@raymondScalingGasTransfer2012|@melchingReaerationEquationsDerived| -->
<!-- |@raymondScalingGasTransfer2012|@tsivoglouCharacterizationStreamReaeration1972| -->
<!-- |@raymondScalingGasTransfer2012|Stroud Water Center| -->
<!-- |@raymondScalingGasTransfer2012|@bernotInterregionalComparisonLanduse2010| -->
<!-- |@raymondScalingGasTransfer2012|@mulhollandInterbiomeComparisonFactors2001| -->
<!-- |@churchillPredictionStreamReaeration1964|@churchillPredictionStreamReaeration1964|| -->
<!-- |@owensReaerationStudiesStreams1964|@owensReaerationStudiesStreams1964|| -->

## Table S2
*Table S2: Validation metrics used in this study, where Nt is number of observations and i is the specific observation. σ refers to the variance of the sample and μ refers to the mean of the sample. As is standard, a carrot accent indicates the predicted value.*

|**Description**|**Acronym**|**Definition**|**Ideal Score**|**Possible Range**|**Validation Scheme**|
|-----------|-------|----------|-----------|--------------|-----------------|
|Correlation Coefficient|$r$| $\sqrt{1-(\frac{\sum\limits_{i=1}^{N_t}{(k_{600, i}-\hat{k_{600}})^2}}{\sum{(k_{600, i}-\bar{k_{600}})^2}})}$ |1|-1 to 1|	All rivers and all timesteps|
|Root-mean-square-error|RMSE| $\sqrt{\frac{1}{N_t}\sum\limits_{i=1}^{N_t}(\hat{k_{600, i}}-k_{600, i})^2}$ |0|0 to ∞|All rivers and all timesteps|
|Relative root-mean-square error|RRMSE| $\sqrt{\frac{1}{N_t}\sum\limits_{i=1}^{N_t}(\frac{\hat{k_{600, i}}-k_{600, i}}{k_{600, i}})^2}$ |0|0 to ∞|By river|
|Normalized root-mean-square error|NRMSE| $\sqrt{\frac{1}{N_t}\sum\limits_{i=1}^{N_t}(\frac{\hat{k_{600, i}}-k_{600, i}}{\bar{k_{600, i}}})^2}$ |0|0 to ∞|By river|
|Relative bias|rBIAS| $\frac{1}{N_t}\sum\limits_{i=1}^{N_t}(\frac{\hat{k_{600, i}}-k_{600, i}}{\bar{k_{600, i}}})$ |0|-∞ to ∞|By river|

## Table S3
*Table S3: Details on the 3 depth hydraulic geometry models used to estimate FCO2 from the SWOT rivers (section 2.4).*

|**Name**|**Depth equation**|**Velocity equation**|**Description**|**Reference**|
|--------|------------------|---------------------|---------------|-------------|
|Brinkerhoff 2019|$D=0.265Q^{0.395}$|$\bar{U}=0.284Q^{0.191}$|530,945 measurements made across the United States at streamgauges|this study; @brinkerhoffReconcilingAtaStationAtManyStations2019|
|Raymond 2012|$D=0.409Q^{0.294}$|$\bar{U}=0.194Q^{0.285}$||1,026 measurements across the United States|@raymondScalingGasTransfer2012|
|Raymond 2013|$D=0.409Q^{0.294},  D=0.449Q^{0.37}$|$\bar{U}=0.194Q^{0.285}, \bar{U}=0.346Q^{0.120}$|Average of the Raymond 2012 equation and one using 9,811 measurements at US streamgauges|@raymondGlobalCarbonDioxide2013|

## References

### 2022-01-07 manuscript_v7.Rmd
fixed figure themes and some ms drafting

---
title: "Using hydraulic geometry to remotely sense gas exchange velocity via the SWOT satellite"
author:
  - Craig B Brinkerhoff:
      email: cbrinkerhoff@umass.edu
      institute: UMass
      correspondence: true
  - Colin J Gleason:
      institute: UMass
  - Christopher J Zappa:
      institute: Lamont
  - Peter A Raymond:
      institute: Yale
  - Merritt H Harlan:
      institute: UMass
institute:
  - UMass: Department of Civil & Environmental Engineering, University of Massachusetts-Amherst
  - Lamont: Lamont-Doherty Earth Observatory, Columbia University
  - Yale: School of the Environment, Yale University
bibliography: References.bib
csl: agu_citation_style.csl
output:
  word_document:
      pandoc_args:
        - '--lua-filter=scholarly-metadata.lua'
        - '--lua-filter=author-info-blocks.lua'
      reference_docx: word-styles-reference.docx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Key Points
- BIKER algorithm predicts gas exchange velocity and $CO_2$ fluxes from simulated SWOT data
- BIKER is robust to measurement errors implicit in SWOT river measurements
- BIKER will allow for unprecedented study of gas exchange spatiotemporal dynamics after SWOT's launch

## Keywords
gas exchange, fluvial geomorphology, remote sensing, open-channel flow, SWOT, biogeochemistry

## Abstract
Extensive research over the past two decades has shown that the global river network emits significant amounts of greenhouse gas via evasion. Despite much progress, there is still significant uncertainty in the temporal dynamics of gas exchange velocity (i.e. the speed of evasion) and thus carbon emissions to the atmosphere. We propose that the NASA/CNES/USKA/CSA SWOT satellite, set to usher in a new age of hydrology research at the global scale, can provide similarly transformative insights to fluvial gas exchange modeling upon launch in 2022. Here, we present work inferring gas exchange velocity from future SWOT observations without in situ calibration. We exploit the unique geomorphology of SWOT-observable rivers to develop a physical model of gas exchange that is nearly entirely remotely sensible and explains `r round(modelsWIDE[modelsWIDE$name == 'Chainsaw-eD',]$r2,2)*100`% of variation in gas exchange. We then couple this model with established Bayesian inference techniques to develop BIKER, or the 'Bayesian Inference of the $k_{600}$ Exchange Rate' algorithm. We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers and 763 discrete measurements of river gas exchange, yielding an algorithm that predicts this theoretical model solely from SWOT observations with a by-river median normalized RMSE of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$NRMSE), 2)*100`%. Like many fluvial Bayesian inference algorithms, BIKER is better at reproducing the temporal dynamics of gas exchange (median correlation coefficient of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$r), 2)`), rather than estimating the absolute rates of exchange. Finally, we show BIKER robust to measurement errors implicit in the SWOT data. With SWOT set to launch in late 2022, we suggest that BIKER can be deployed to map the global-scale spatiotemporal dynamics of fluvial gas exchange in large rivers.

## 1 Introduction
Natural systems play a critical role in the budgeting and accounting of the global carbon cycle under climate change. Following @colePlumbingGlobalCarbon2007, the global river network is recognized to emit substantial amounts of carbon to the atmosphere via evasion (gas exchange driven by a diffusion gradient and water turbulence), in addition to their long understood role of transporting carbon to the oceans. Current estimates of total carbon dioxide evasion ($FCO_2$) to the atmosphere from the global river network vary from 650-1800 Tg C/yr [@lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013] (*Liu etal in review- add number is paper if accepted in time*), with 167 Tg-C/yr coming from mountain streams alone [@horgbyUnexpectedLargeEvasion2019]. Despite rivers' incredibly small percentage of the global land surface [0.`r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])`%- @raymondGlobalCarbonDioxide2013], this C flux is on par with the total oceanic $CO_2$ uptake rate [@gruberOceanicSinkAnthropogenic2019; @horgbyUnexpectedLargeEvasion2019] and the global forest carbon uptake rate [@panLargePersistentCarbon2011].

River $CO_2$ evasion is increasingly better constrained and is clearly a critical component of the global carbon cycle. Equation 1 represents this riverine flux given $\delta CO_2$ (the gas concentration gradient between the water $CO_{2_{water}}$ and the air $CO_{2_{air}}$) and the gas exchange velocity *k*. Consult Appendix A for variable nomenclature used throughout this study.

$$FCO_2=k* \delta CO_2 \mathbf{(1)}$$
The structure of equation 1 necessitates that calculations of this flux are highly sensitive to measurements/estimates of *k*. However, *k* can only be directly calculated via a known gas concentration gradient, eddy-covariance measurements, or tracer additions to the stream [@hallGasExchangeStreams2020]. In trying to constrain the global fluvial $FCO_2$ flux across millions of rivers, this calculation is impossible, and necessitates the use of predictive models for *k* that are based on easily obtained river hydraulic properties. In that vein, there have been over 20 empirical models developed to predict *k*, generally using some combination of mean velocity $\bar{U}$, shear velocity $U_*$, width $W$, depth $H$, and slope $S$ as predictors [@wangPhysicallyBasedScaling2021]. Note that these models usually predict $k_{600}$, or *k* normalized by a Schmidt number of 600. This is to remove the effect of water temperature and gas type from predictive models, as warmer waters and lower *Sc* numbers both increase gas exchange rates. For example, $k_{600}$ reflects the $k_{CO_{2}}$ at 20 degrees Celsius. Through this normalization, these models exclusively perturb the hydraulic controls on gas exchange [@hallGasExchangeStreams2020].

These empirical models enable estimating a global $FCO_2$ flux ($FCO_2_{upscaled}$) by summing across millions of rivers, but they also change the base parameters that ultimately control that aggregate estimate. That is, by making *k* a function of hydraulics, $FCO_2_{upscaled}$ is now a direct function of river hydraulics. This functional relationship is described in equation 2. It suggests that $FCO_2$ estimates are not only at the mercy of the accuracy and spatiotemporal resolution of $\delta CO_2$, but also the accuracy and resolution of our river hydraulics estimates.

$$FCO_{2_{upscaled}}=f(k_{CO_2}, \delta CO_2)=f(\bar{U}, H, S, W, \delta CO_2) \mathbf{(2)}$$
For example, to upscale $FCO_2$ globally, @raymondGlobalCarbonDioxide2013, @lauerwaldSpatialPatternsCO22015, and @horgbyUnexpectedLargeEvasion2019 all relied on $k_{CO_2}$ values indirectly estimated using mean annual streamflow models and scaling equations to predict hydraulic terms, while @borgesGloballySignificantGreenhousegas2015 used a combination of the above method and a constant $k_{CO_2}$ in space and time to upscale over Africa. In all of these foundational studies, the temporal dynamics of the gas exchange velocity (and thus $FCO_2$) were ignored because of hydraulic data limitations. It has been shown at the field-scale that temporal dynamics of gas exchange can vary widely from site to site [@wallinSpatiotemporalVariabilityGas2011], but it has remained impractical to obtain temporally explicit $k_{CO_2}$ at continental-to-global scales. This knowledge gap is contrasted agaist the robust existing literature exploring spatiotemporal patterns in $CO_2_{water}$ [e.g. @crawfordCO2TimeSeries2017; @peterScalesDriversTemporal2014; @ranDynamicsRiverineCO22017; @rocher-rosLandscapeProcessDomains2019; @raymondAtmosphericCO2Evasion2000].

To that end, the field has made significant progress in estimating $CO_2_{water}$, where we now have river-reach explicit modeling at monthly temporal resolutions using global hydrography datasets [@horgbyUnexpectedLargeEvasion2019; @brinkerhoffLakeMorphometryRiver2021; @saccardiImprovingPredictionsStream2021; *Liu etal in review*], but an equivalently sophisticated representation of $k_{CO_2}$ is still lacking. As equations 1 and 2 dictate, these terms share the burden of calculating $FCO_2$. Therefore, these next-generation models will have to contend with this discrepancy, which is directly dictated by the lack of direct measurements of global river hydraulics.

@wangPhysicallyBasedScaling2021 recently attempted to address this problem by simulating *k* in 35 rivers of many sizes (widths ranging from 0.23–349m) using a stream metabolism model [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets to infer what *k* must have been. They then compared this simulated dataset against direct measurements of *k*, finding similar performance and parameter values for process-based models of gas evasion. However, they were still limited by a lack of direct hydraulic measurements and had to rely on scaling equations to estimate river depth and velocity. Even though approaches like this are incredibly useful for expanding our mechanistic understanding of gas exchange, it is less useful for global upscaling purposes as it relies on highly detailed in situ DO data.

In theory, this discrepancy between the quality of our $CO_2_{water}$ and $k_{CO_2}$ estimates could be alleviated if direct hydraulics measurements were available at the global scale at a sufficient temporal resolution. This is turn would also address the uncertainty regarding continental-to-global scale temporal dynamics of gas exchange noted earlier. Conveniently, these very data will soon be available via the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission. SWOT is expected to launch in 2022 and provide the world's first direct measurements of global water surface extent and elevation at novel temporal resolutions. SWOT is a wide swath radar interferometer and will sample rivers every 1 to 7 days per 21 day repeat cycle, measuring rivers wider than 100m with a goal of expanding this to rivers at least 50m wide [@biancamariaSWOTMissionIts2016]. Via its direct hydraulic measurements, SWOT is expected to usher in a sea change in global-scale hydrology, and could similarly transform fluvial biogeochemsitry if techniques are developed to ingest SWOT data and infer *k*. Here, we borrow tools from fluvial geomorphology and existing SWOT algorithms to answer the following two questions:

  - Can we leverage the hydraulic geometry of the large rivers SWOT will observe to develop a physically-based model for $k_{600}$ unique to these rivers?

  - Can we exploit such a model to infer $k_{600}$ (and its uncertainty) solely from SWOT observations?

To answer this first question, we use the specific hydraulic geometry of large, SWOT-observable rivers to develop a process-based model for large-river gas exchange. We then take these findings and explore question 2 by implementing the model within an RS of $k_{600}$ algorithm named BIKER ('Bayesian Inference of the $k_{600}$ Evasion Rate'). The goal with BIKER is to require no in situ inputs of any kind (although in situ data could be ingested and should improve results) such that it is completely globally implementable on any river on Earth that SWOT will measure. Finally, we also test whether BIKER can predict actual gas fluxes when paired with an in situ $CO_2$ sensor and compare it against established methods that are reliant on in situ hydraulic measurements.

This paper is split into two distinct parts: gas-exchange theory/model development (section 2) and BIKER algorithm development/validation (section 3). Section 3 is fundamentally dependent on the results presented in section 2, therefore section 2 presents both theory and results. Both sections detail the data used. We conclude with a discussion (section 4) that encompasses both gas exchange theory and remote sensing. We also provide a flowchart detailing the entire study as Figure 1.

![Figure 1: Flowchart detailing entire study. Dashed lines indicate workflow to calculate the observed data that we validate against. See Appendix A for variable nomenclature used in this study.](cache/flowchart.jpg)

## 2 Hydraulic geometry in large rivers and predicting fluvial gas exchange velocity
To predict *k* in the very large rivers that SWOT will observe, we start from established process-based models for *k*, impose hydraulic assumptions that are valid in large rivers, and then obtain a suite of models we can test. Following a description of the data used (section 2.1), we outline these established models (section 2.2) and then step through our hydraulic assumptions to arrive at equations that can be ingested by SWOT (section 2.2). Following this model development, all models are empirically tested (section 2.3)

### 2.1 Data
Our gas exchange theory is developed using multiple datasets of directly measured $k_{600}$ and directly measure stream hydraulics collected from the literature. In total, we have collected 763 measurements of $k_{600}$, with 701 of these measurements previously published by @raymondScalingGasTransfer2012 and @ulsethDistinctAirWater2019. These come from tracer studies and thus define $k_{600}$ and the rivers' mean hydraulic properties at the reach scale. The remaining measurements were measured by @churchillPredictionStreamReaeration1964 and @owensReaerationStudiesStreams1964. See Table S1 for a complete list of the studies that collected these measurements.

We expand our dataset of stream hydraulics using a previously published compilation of in situ hydraulic measurements [@brinkerhoffReconcilingAtaStationAtManyStations2019]. That dataset contains over 530,000 unique measurements of river channel velocity, width, and discharge from across the continental United States, originally taken to calibrate United States Geological Survey (USGS) streamgauge rating curves and made public by the USGS. Ultimately, this dataset was filtered to `r sum(percs$n)` discrete measurements for this study, in addition to the direct measurements in the gas exchange dataset (see Text S1 for our filtering protocol).

### 2.2 Process-based modeling of river *k*
*k* should scale with near-surface turbulence in a turbulent flow [@hallGasExchangeStreams2020] and thus extensive field and laboratory experiments have converged on the 'small-eddy model' as first proposed by @lamontEddyCellModel1970 *Zappa said this actually goes back further than this paper* to scale *k* via the smallest-scale turbulent eddies [e.g. @katulStructureFunctionModel2018; @lorkeUnifiedScalingRelation2006; @zappaEnvironmentalTurbulentMixing2007; @zappaVariationSurfaceTurbulence2003; @moogStreamReaerationNonuniform1999; @vachonRelationshipNearsurfaceTurbulence2010; @tokoroHighGastransferVelocity2008; @wangPhysicallyBasedScaling2021]. This model is provided as equation 3, where $\epsilon$ is the dissipation rate of near-surface turbulence, $v$ is the kinematic viscosity, and *Sc* is the Schmidt number.

$$k\propto Sc^{-1/2}(v\epsilon)^{1/4} \mathbf{(3)}$$
Further, some workers have noted that laboratory and field observations suggest open channel flows with small bed roughness (generally big rivers) do not exhibit homogeneous surface dissipation at their air-water interface [@moogAirWaterGasTransfer1999; @talkeTurbulentKineticEnergy2013]. Given this observation, @moogAirWaterGasTransfer1999 proposed an extension to the small-eddy model, additionally scaling $k_{600}$ using a shear Reynold's number formulation. *look up other shear reynolds ways to additionally scale k*. This is equation 4 and is referred to as the 'Chainsaw model'. The Chainsaw model is hypothetically useful in low-turbulence flows where a relative lack of large-scale eddies effectively 'filter out' the number of small-eddies that actually reach the interface and lead to gas exchange [@talkeTurbulentKineticEnergy2013]. In the context of BIKER, we chose to test this model because large, SWOT-observable rivers are generally the smoothest, least-turbulent flows along the stream-to-river continuum. Further, to our knowledge the chainsaw model has never been empirically tested in predicting *k*, aside from confirming that large-scale eddies differentially move turbulence to the surface in a hydraulically-wide river [@talkeTurbulentKineticEnergy2013].

$$k \propto Sc^{-1/2}(v\epsilon)^{1/4}(\frac{U_*H}{v})^{3/8} \mathbf{(4)}$$
Finally, both equations rely on measurements of $\epsilon$, of which measuring directly is non-trivial. When working at large scales, a commonly used $\epsilon$ model assumes that all turbulence is generated at the bed and transported to the air-water interface via the log-law-of-the-wall such that $\epsilon = \epsilon_S = \frac{U_*^3}{H}$ [@lorkeUnifiedScalingRelation2006; @nezuTurbulenceOpenChannel1993]. Another approach models *k* via 'form-drag dissipation' ($\epsilon = \epsilon_D = gS\bar{U}$) which is equivalently the total stream power per unit mass water. This normalized stream power captures the bulk frictional resistance (and thus energy dissipation) via channel banks, meanders, bars, etc. that is unique to fluvial systems [@moogStreamReaerationNonuniform1999]. Authors have since shown that this model can reasonably predict *k* in rivers and streams [@raymondScalingGasTransfer2012; @ulsethDistinctAirWater2019; @wangPhysicallyBasedScaling2021].

### 2.3 Imposing large-river hydraulic geometry
Rivers and streams change predictably along their longitudinal profile from headwater to ocean, and we can exploit the distinct hydraulic geometry of large rivers at the end of this continumn to estimate *k* in SWOT-observable systems. In general, as river size increases, channels become more rectangular, their shapes elongate (becoming wider quicker than they become deeper) and their hydraulic radii begin to approximate their mean flow depth, i.e. $R_h \approx H$ [@leopoldHydraulicGeometryStream1953]. This is a common assumption in hydraulic and geomorphic modeling of large rivers, and occurs in `r round(percs[percs$flag_swot == 'SWOT' & percs$flag_hydraulicWide == 'Rh=H',]$perc,0)`% of our SWOT-observable river measurements but only `r round(percs_overall[percs_overall$flag_hydraulicWide == 'Rh=H',]$perc,0)`% of non-SWOT observable river measurements (calculated using both datasets of stream hydraulics outlined in section 2.1- see Text S1 for full details). These channels are often referred to as 'hydraulically-wide'.

Assuming that all SWOT-observable rivers are hydraulically-wide, we test the performance of four models for predicting *k* in hydraulically-wide channels: the four unique combinations of equations 3-4 and the two $\epsilon$ models already outlined. While the complete model derivations for the small-eddy models (section 2.3.1) and the chainsaw models (section 2.3.2) are provided in Text S2, the broad goal was to reduce the equations down to their fundamental parameters, identifying which terms are SWOT observable and limiting the number of terms not directly measurable via SWOT.

#### 2.3.1 'Small-eddy' model for gas evasion
We derive equations starting from the small-eddy model (equation 3) for hydraulically-wide channels. If we assume $\epsilon=\epsilon_S$ and $R_H = H$ in equation 2, and finally solve for $k_{600}$ via statistical parameter $\alpha$, we eventually arrive at equation 5. If we do the same but assume $\epsilon=\epsilon_D$ and statistical parameter $\alpha_1$, we arrive at equation 6.

$$k_{600}=\alpha(gS)^{3/8}H^{1/8} \mathbf{(5)}$$

$$k_{600}=\alpha_1(gS\bar{U})^{1/4} \mathbf{(6)}$$
Equations 5 and 6 thus define gas exchange velocity solely as a function of slope and mean flow depth or mean flow velocity. This is theoretically valid only in a hydraulically-wide channel.

#### 2.3.2 'Chainsaw' model for gas evasion
We also impose $\epsilon=\epsilon_S$ and $R_H=H$ on the chainsaw model just like we did in section 2.3.1 and arrive at equation 7 (with statistical coefficient $\beta$). The result is identical to the hydraulically-wide small-eddy model (equation 5), except for the exponents which are both steeper. Equation 8 reflects the same derivation, but when $\epsilon=\epsilon_D$ and statistical coefficient $\beta_1$.

$$k_{600}=\beta(gS)^{9/16}H^{11/16} \mathbf{(7)}$$
$$k_{600}=\beta_1(gS)^{7/16}\bar{U}^{1/4}H^{9/16} \mathbf{(8)}$$
We now have four different models for $k_{600}$ in hydraulically-wide rivers, each with a different set of assumptions and parameters but all defined as functions of the water surface slope and other hydraulic terms that are not directly measurable via SWOT. All four models are summarized in Table 1.

*Table 1: Four physically-based models for $k_{600}$ each derived under a different set of mass transport assumptions but all valid in hydraulically-wide rivers. Their derivations are outlined in sections 3.2.1 and 3.2.2 and their empirical tests are shown in section 3.3.*

|**Description**|**Equation**|**References**|**Empirical test**|**Empirical fit via $r^2$**|
|---------------|------------|--------------|------------------|-----------------|
|Small-eddy, log-law-of-the-wall|$k_{600}=\alpha(gs)^{3/8}H^{1/8}$|@lamontEddyCellModel1970; log-law-of-the-wall|Figure 3a|`r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eS',]$r2, 2)`|
|Small-eddy, form-drag dissipation|$k_{600}=\alpha_1(gS\bar{U})^{1/4}$|@lamontEddyCellModel1970; @moogStreamReaerationNonuniform1999|Figure 3c|`r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eD',]$r2, 2)`|
|Chainsaw, log-law-of-the-wall|$k_{600}=\beta(gs)^{9/16}H^{11/16}$|@moogAirWaterGasTransfer1999; log-law-of-the-wall|Figure 3b|`r round(modelsWIDE[modelsWIDE$name == 'Chainsaw-eS',]$r2, 2)`|
|Chainsaw, form-drag dissipation|$k_{600}=\beta_1(gs)^{7/16}\bar{U}^{1/4}H^{9/16}$|@moogAirWaterGasTransfer1999; @moogStreamReaerationNonuniform1999|Figure 3d|`r round(modelsWIDE[modelsWIDE$name == 'Chainsaw-eD',]$r2, 2)`|

### 2.4 Model validation
With equations 5-8 defined, we now empirically test their strength of fit for hydraulically-wide channels. We validate on the dataset of direct measurements of $k_{600}$ and stream hydraulics, after filtering for measurements made in hydraulically-wide channels. This was defined here as those whose hydraulic radius was within 1% of their mean flow depth. All told, this amounts to 166 *double check* direct measurements of hydraulically-wide $k_{600}$ and stream hydraulics to test with. Models were assesed via the coefficient of determination ($r^2$) and results are plotted in Figure 3. Note that Figure 3 axes are plotted in logarithmic space just for visualization: model fit and validation (via $r^2$) were calculated in linear space as their models dictate.

![Figure 2: Empirical testing of four models for predicting gas exchange velocity in hydraulically-wide channels: (a) Small-eddy model with log-law-of-the-wall dissipation. (b) Chainsaw model with log-law-of-the-wall dissipation. (c) Small-eddy model with form-drag dissipation. (d) Chainsaw model with form-drag dissipation. Note that axes are plotted in logarithmic space just for visualization: model fit and validation were calculated in linear space.](cache/k600_theory/k600Plot.jpg)

Despite reasonable $r^2$ scores, both small-eddy models produce clearly biased estimates (Figures 3a, 3c). They systematically underestimate lower gas exchange velocities and overestimate higher ones, with the log-law-of-the-wall model (Figure 3a) performing marginally worse both visually and via $r^2$ (`r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eS',]$r2, 2)` versus `r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eD',]$r2, 2)` for Figure 3c). However, when this model is expanded to retard the dissipation rate in the smoothest flows via the Chainsaw model [equations 6 and 7- @moogAirWaterGasTransfer1999], the dynamics are more accurately captured (Figures 3b, 3d) and $r^2$ scores go up appreciably (`r round(modelsWIDE[modelsWIDE$name == 'Chainsaw-eS',]$r2, 2)` and `r round(modelsWIDE[modelsWIDE$name == 'Chainsaw-eD',]$r2, 2)` for the log-law-of-the-wall and form-drag dissipation, respectively). Modeling dissipation via total form-drag rather than just via bed shear and the log-law-of-the-wall improves model performance. This is likely because the majority of dissipation occurring in rivers comes not just from the streambed, but from bulk form drag via channel walls, bars, and meanders [@raymondScalingGasTransfer2012; @moogStreamReaerationNonuniform1999; @wangPhysicallyBasedScaling2021; @ulsethDistinctAirWater2019].

The success of the chainsaw model in hydraulically-wide channels, and in particular when coupled with the $\epsilon=\epsilon_D$ model (equation 8, Figure 3d), provides us with a strong physical-model for gas evasion. The river hydraulics terms in equation 8 can either be directly measured or reasonably inferred from SWOT measurements, effectively opening the door for remotely sensing the gas exchange velocity. This is explored next.

## 3 Exploiting equation 8 to remotely sense gas exchange velocity
We have shown that scaling $k_{600}$ via equation 8 explains `r round(modelsWIDE[modelsWIDE$name == 'Chainsaw-eD',]$r2,2)*100`% of the variation in $k_{600}$ in hydraulically-wide rivers, which includes nearly all SWOT-observable rivers. Further, equation 8 has only three non-remotely-sensible terms: $k_{600}$, mean flow depth, and mean flow velocity. Conveniently, techniques to simultaneously infer hydraulic parameters from SWOT data have been established over the last decade to infer streamflow from SWOT's measurements [e.g. @andreadisConstrainingAssimilationSWOT2020; @brinkerhoffConstrainingRemoteRiver2020; @durandEstimatingReachaveragedDischarge2014; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014a; @garamboisVariationalEstimationEffective2020; @brissetAssimilationAltimetricData2018; @hagemannBAMBayesianAMHGManning2017; @larnierRiverDischargeBathymetry2020a; @oubanasDischargeEstimationUngauged2018]. For BIKER, we follow the work developed by @durandEstimatingReachaveragedDischarge2014, @hagemannBAMBayesianAMHGManning2017, and @brinkerhoffConstrainingRemoteRiver2020 to infer mean flow velocity and depth from SWOT observations using a modified form of the Manning's equation for mean flow velocity. Following a description of the data used (section 3.1), we detail algorithm development and experimental design (section 3.2) and then we present the validation results (section 3.3).

### 3.1 Data
![Figure 3: Map of the 47 hydraulic models and 1 timeseries of CO2 samples used in this study. Note that hydraulic model locations are approximate as some of the models are not geo-referenced. (Table S1).](cache/maps/map3.jpg)

To validate our remote sensing solution for $k_{600}$ (BIKER), we cannot use SWOT measurements as it has yet to launch. In the hydrology literature, it is therefore standard practice to benchmark SWOT-related algorithms on "SWOT-like" data. We use 47 SWOT-simulated rivers for validation, where these simulated rivers are simply river-reach-averaged hydraulic model outputs where the water surface heights, slopes, and widths are labelled as RS observations and are used as the sole inputs to BIKER. These were published by @frassonExploringFactorsControlling2021 and @durandIntercomparisonRemoteSensing2016 and are the two benchmarking studies that explored remote sensing of discharge (RSQ) algorithm performance for the SWOT mission. They have prepared the data to mimic SWOT by adding realistic radar errors and sampling along the satellite’s future ground track. In essence, these authors have run standard hydraulic models forced with known inflows and measured bathymetry to model the hydraulic response of these river reaches, and then sampled these model outputs according to the timing of the satellite overpass. @frassonExploringFactorsControlling2021 also developed a sophisticated error model that mimics the effects of the atmosphere, topography, and water surface geometry on SWOT measurements (16/47 rivers) to add realistic errors that depend on the time and place of observation. The resulting data are the most realistic synthetic SWOT river data in the literature. The approximate locations of these rivers are plotted in Figure 1.

We use 26 bi-weekly dissolved $CO_2$ samples collected by @beaulieuControlsGasTransfer2012 at one location in the Ohio River for one calender year from 2008-2009 (Figure S1) for the $CO_2$ evasion and carbon efflux calculations (section 3.1.3). Note that this $CO_2$ data is for the Ohio River only but was applied to all rivers (which includes multiple sections of the Ohio River). Because we are exclusively interested in the relative differences between $FCO_2$ estimates and not the raw fluxes themselves, any $CO_2$ data representative of SWOT-observable rivers was deemed acceptable for this validation exercise.

### 3.2 Section 3 methods
#### 3.2.1 BIKER
BIKER, and Bayesian inference in general, starts from Bayes rule (equation 9), where $\Theta$ is some set of non-remotely-sensible parameters we want to solve for (including $k_{600}$), *x* is the observed data, $f(x|\Theta)$ is the sampling model where data are conditional on the parameters, and $p(\Theta)$ is the joint prior distribution of the parameters. Therefore, we are interested in solving for $p(\Theta|x)$, or the 'posterior' distribution. Note that $p(x)$ is usually computationally intractable to integrate exactly, but Bayesian inference tools require only the proportionality to be specified: $p(\Theta|x) \propto f(x|\Theta)p(x)$. Sampling algorithms are then used to approximate the posterior distribution, as is done in BIKER.

$$p(\Theta|x) = \frac{f(x|\Theta)p(\Theta)}{p(x)} \mathbf{(9)}$$
The heart of BIKER is its reformulation of the $k_{600}$ model (equation 8) as a Bayesian sampling model that is conditional on the non-remotely-sensible parameters (i.e. $f(x|\Theta)$). This is similar to the McFLI logic used in some SWOT RSQ algorithms [@gleasonTrackingRiverFlows2017]. To start, we write $k_{600}$ as a function of SWOT-observables $W$ and $S$. This algebra is carried out using equation 8, the fitted value for $\beta_1$ from Figure 3d (`r round(modelsWIDE[modelsWIDE$name == 'Chainsaw-eD',]$coef,2)`), and Manning's equation for mean flow velocity ($\bar{U}=\frac{1}{n}R_h^{2/3}S^{1/2}$). Following section 2.3, we also assume that the channel is hydraulically-wide ($R_h = H =\frac{A}{W}$). To leverage additional SWOT data, the wetted channel area *A* is further split into the SWOT-observable portion $dA$ and the unobservable portion $A_0$ following @durandEstimatingReachaveragedDischarge2014. *dA* is readily approximated using SWOT measurements if assuming a rectangular river channel ($dA \approx \delta H_e*W$), which is again generally valid in a hydraulically-wide channel.

All of this algebra simplifies to equation 10, after accounting for time-varying *t* terms and space-varying *i* terms within the reach. Conveniently, both $k_{600}$ is measured (and predicted by equation 8) at the reach-scale, so we can reduce it to a reach-averaged, time-varying quantity. This lowers the number of parameters BIKER must infer and makes the problem better constrained.

$$\begin{eqnarray} k_{600_t} = 62.82 g^{7/16}S_{it}^{9/16}\frac{1}{n_i}^{1/4}(\frac{A_0_i+dA_{it}}{W_{it}})^{35/48} \end{eqnarray} \mathbf{(10)}$$
Next, equation 10 is written as a Bayesian sampling model, in which all of the SWOT observations are sampled from the unknown model parameters ($A_0_i$, $k_{600_t}$, and $n_i$). This is equation 11 after describing them as log-normal distributions. $\sigma_{k_{600}}^2$ refers to the total uncertainty implicit in equation 10. This uncertainty is due to parameter uncertainty from equation 8, Manning's equation, and the rectangular channel assumption.

$$\left[log(62.82) + \frac{7}{16}log(g) + \frac{9}{16}log(S_{i,t}) - \frac{35}{48}log(W_{i,t})\right] \sim normal\left(\left[log(k_{600_t})+\frac{1}{4}log(n_i)-\frac{35}{48}log(A_0_i+dA_{i,t})\right], \sigma^2_{k_{600}}\right) \mathbf{(11)}$$
Equations 9 and 11 also necessitate that we specify prior distributions for the parameters $A_0_i$, $k_{600_t}$, and $n_i$. Prior distributions, described by their hyperparameters, formalize the a priori estimates (and uncertainties) for the non-remotely-sensed terms. More intuitively, BIKER priors represent our 'prior river knowledge' of what $A_0_i$, $k_{600_t}$, and $n_i$ probably are for some river since they cannot be directly remotely sensed, but educated guesses can be made using the SWOT data available to us. This is equivalent to the 'empirical Bayes method'. Our goal in prior specification was to rely on absolutely no in situ information such that we could run this method on any river on Earth solely using SWOT observations. In theory, more informed priors via various a priori information about a specific river would improve BIKER performance, but here we chose to test only the fully generalized algorithm. Therefore, the validation presented here is a 'worst-case scenario', wherein BIKER performance will improve with better prior information on the river. In that context, we used a variation of the prior specification method developed by @brinkerhoffConstrainingRemoteRiver2020, who developed geomorphic 'river types' with distinct prior hyperparameter sets for $A_0_i$ and $n_i$. These hyperparameters are assigned to a river solely using SWOT data *W* and *S*, therefore meeting our needs for complete global implementability. Hyperparameter assignment for $k_{600_t}$ was developed similarly (all hyperparameter specifications are elaborated on in Texts S3 and S4).

With the sampling model defined ($f(x|\Theta)$ = equation 11) and hyperparameters $p(x$) described (Texts S3 and S4), a joint posterior distribution conditional on the SWOT observations ($p(\Theta|x) \propto f(x|\Theta)p(x)$) is specified. To approximate this distribution, we use a Markov Chain Monte Carlo (MCMC) algorithm implemented using the Stan probabilistic programming language. Specifically, Stan uses a Hamiltonian Monte Carlo sampler which reduces computation time relative to other sampling algorithms [@hagemannBAMBayesianAMHGManning2017].

#### 3.2.2 BIKER validation
We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers (section 3.1) using daily observed hydraulics. We also re-validate BIKER on the `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` rivers with the SWOT error model which corrupts the SWOT measurements (widths, heights, and slopes).

We do not have observed $k_{600_t}$ data for these rivers and, to our knowledge, no field dataset of $k_{600}$ exists in the type of temporal and spatial frequency that SWOT (and therefore the BIKER algorithm) provides. Thus, we take the model outlined in equation 7 and Figure 3d and use that to calculate the observed $k_{600_t}$ that BIKER is validated against (equation 12). Here, reach-averaged flow velocity is directly calculated as discharge divided by channel area. This directly explores BIKER's ability to infer observed $A_0_i$, $k_{600_t}$, and $n_i$ simultaneously from $W_{i,t}$ and $H_{i,t}$ alone and was deemed acceptable given that the $k_{600}$ model (equation 7) has already been successfully validated in Figure 3d.

$$k_{600_{obs}}=62.82(gS)^{7/16}\bar{U}^{1/4}H^{9/16} \mathbf{(12)}$$
It also means that, for a fair validation scheme, the parameter specification for $\sigma^2_{k_{600}}$ in equation 11 should not include the uncertainty from equation 8. So for this validation only, $\sigma^2_{k_{600}}$ is set to 0.30 (natural-log space) to reflect uncertainty from Manning's equation and assuming a rectangular channel (Text S4). However, if BIKER is run on real SWOT observations after the satellite launches, $\sigma^2_{k_{600}}$ should reflect the total uncertainty, which is described in Text S4.

With the setup described by equations 11 and 12, validation is performed using the BIKER posterior means. Because we are most interested in BIKER's ability to correctly infer temporal dynamics in $k_{600}$, and each river's priors are described specifically for that river, we validate by river. These metrics are defined in Table S2 and consist of the correlation coefficient *r* to quantify accuracy of BIKER's temporal dynamics, the root mean square error normalized by the observed mean (NRMSE) and prediction bias normalized by the observed mean (rBIAS) to assess bias, and the Kling-Gupta Efficiency (KGE). KGE is frequently used to assess streamflow prediction and simultaneously assesses accuracy in both bias and dynamics. While a value greater than -0.41 means the model outperforms a uniform prediction of the mean [@knobenTechnicalNoteInherent2019], generally KGE scores are interpreted as being meaningful in ungauged settings if > 0.

#### 3.2.3 Carbon efflux validation
It is one thing to accurately model the temporal dynamics of $k_{600}$, but researchers are often most interested in the actual gas fluxes from rivers and ultimately the carbon emitted from river to atmosphere. Therefore, we also explore BIKER's ability to reproduce $FCO_2$ (equation 1) and the bulk carbon efflux from these `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers. To do this, we pair the 26 biweekly $CO_2$ and water temperature samples from @beaulieuControlsGasTransfer2012 (section 2, figure S1) with every 14th set of daily SWOT observations (as the $CO_2$ data is bi-weekly). Conveniently, this is very close to the nominal average sampling resolution for SWOT (11 days). Not all of the SWOT rivers have observations for a full year, and when simulation dates are not available they are assumed to start on January 1st. We then calculate $FCO_2$ using equation 1 and assuming atmospheric $CO_{2_{air}}$ is 390 uatm. Equation 13 is used to obtain a temperature and $CO_2$ specific gas exchange velocity from BIKER's output, where the Schmidt number *Sc* was estimated following @raymondScalingGasTransfer2012. We then estimate median and mean bulk carbon effluxes by multiplying the median or mean $FCO_2$ value across all rivers and timesteps (in Tg-C/yr) by the total average surface area of all 47 rivers.

$$k_{CO2, temp}=k_{600}*(Sc/600)^{-1/2} \mathbf{(13)}$$
Finally, we also compare BIKER's outputs with established in situ methods, whose specific details and parameters are provided in Table S3. Because streamflow measurements and models are vastly more frequent than measurements of river channel geometry, these in situ models are used to predict reach-averaged flow depth and flow velocity as functions of streamflow. Their outputs are in turn used to estimate *k* in upscaling workflows. Therefore, they rely on in situ streamflow records or high-quality streamflow routing outputs (unlike BIKER, which only requires a river be SWOT-observable). This allows us to assess whether BIKER's estimates (wholly ungauged) are comparable to gauged methods (all three HG models). Of the three models tested here, one is trained on the @brinkerhoffReconcilingAtaStationAtManyStations2019 dataset, and two are previously published models [@raymondGlobalCarbonDioxide2013; @raymondScalingGasTransfer2012].

### 3.3 Section 3 results
#### 3.3.1 BIKER
Figure 4 plots all `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` timeseries plots of predicted and observed $k_{600}$, again assuming no SWOT measurement error. Note that the y axis is normalized by maximum observed values to compare rivers. In general, temporal dynamics of $k_{600}$ are reproduced quite well by BIKER, however there is often positive bias in the estimates (e.g. Olentangy River). Some rivers are predicted nearly perfectly (e.g. Brahmaputra, Ohio Section 3, Wabash) while others perform more poorly (e.g. San Joaquin, Stanislaus), though again much of this error stems from errors in predicting the magnitude of $k_{600}$ rather than the temporal dynamics. Rivers where the temporal dynamics are more poorly inferred tend to have shorter timeseries (e.g. Merced River, Tuolumne River). Next, we explicitly quantify each rivers' performance.

![Figure 4. River timeseries plots of $k_{600}$. Orange is calculated via observed hydraulics, while green is BIKER. The green ribbon indicates the 95% CIs for the predicted values. Y axis is normalized by maximum observed values to compare visually.](cache/validation/timeseries_noerr.jpg)

We calculate performance metrics following section 3.2.2 and Table S2. These are presented in Figure 5a, which plots validation metrics calculated for each river with and without SWOT measurement error (green and purple boxplots, respectively). SWOT measurement errors negligibly influence BIKER's performance across all four error metrics (Figure 5a), though caution should be used in over-interpreting boxplots with a sample size of only `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])`. Given Figure 5a, we deem that SWOT measurement error does not exert a significant influence on BIKER and so the results presented for the rest of the manuscript assume no measurement error in order to use all `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers.

![Figure 5. a: Performance metrics by river. See Table S2 for metric definitions. b: *r* versus the number of SWOT observations used in the Bayesian inversion. c) *r* versus the mean river width. As expected, performance improves with more data. Dashed lines denote scores of 0, 0.50, and 1. CIs.](cache/validation/validation_by_river.jpg)

Median river-specific $KGE$ is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$KGE, na.rm=T),2)` and median river-specific *r* is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$r, na.rm=T),2)`. This indicates very strong inference of each river's temporal $k_{600}$ dynamics. These scores are excellent given that absolutely no in situ information is being used to predict $k_{600}$. NRMSE has a median score of `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$NRMSE, na.rm=T),2)`, which is reasonable for a completely ungauged method. These are comparable to the best NRMSE scores achieved by recent versions of the SWOT RSQ algorithms [@frassonExploringFactorsControlling2021]. Median rBIAS is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$rBIAS, na.rm=T),2)`, which is again excellent. However, the shape of the rBIAS boxplot in Figure 3a highlights a large portion of rivers which have a positive bias (Figure 4 also confirms this). Taken in aggregate, Figures 4-5 indicate that BIKER is quite good at capturing temporal dynamics in $k_600$, however there is often positive bias in its estimates.

Figure 5b highlights one benefit of using Bayesian inference to estimate $k_{600}$: because the posterior is conditional on the SWOT observations, performance should improve with more data. Figure 6b plots by-river *r* scores versus the number of SWOT observations. While performance does not directly increase from having more data, very poor performances disappear as the amount of available data increases. The vast majority of the extremely poorly performing rivers are limited to those with only 12 days of SWOT observations. This bodes well for future SWOT implementation, as SWOT has a nominal life-cycle of three years. A similar relationship also exists between mean observed river width and *r* (Figure 5c), where wider rivers have better BIKER performance. However, it is impossible to say whether this is a genuine trend or an artifact of the specific rivers being tested, where the larger rivers often have longer timeseries.

Given the promising results *r* results in Figure 5a, we further explored BIKER's ability to predict temporal trends in $k_{600}$. Figure 6a plots the regression lines between BIKER $k_{600}$ and observed $k_{600}$ for each river, making clear that some rivers have notable bias (regression lines sitting above and below the 1:1 dashed blue line).However regardless of coefficient error, nearly all rivers scale at a slope of approximately 1 (dashed blue line), indicating that temporal trends are strongly captured. This is explicitly confirmed in Figure 6b, where we plot a histogram of the slopes of Figure 6a's regression lines, showing that vast majority of by-river regression slopes approximate 1 (median slope: `r round(median(results_dynamics$Slope),2)`).

![Figure 6: a: Temporal scaling of BIKER predictions for the 47 rivers, compared to observed k600 and plotted as regression lines to aid in visualization. b: Histogram of the slopes of these by-river regression lines. Slopes equal 1 indicate perfect predicition of the temporal trends in daily k600 (regardless of bias in the estimates).](cache/validation/dynamics.jpg)

#### 3.3.2 Carbon efflux
Next, we compare BIKER's estimates of carbon fluxes against established in situ techniques. It is important to remember that only BIKER is completely globally implementable, while the other three models presented necessarily rely on having a streamflow record readily available. Therefore, the HG methods reflect their 'best-case scenarios' while BIKER reflects its worst case scenario, where priors are set entirely from SWOT observations and are generally the least informative they will ever be.

Figure 7a evaluates the $FCO_2$ estimates using the same 4 error metrics as Figure 6a. BIKER, the sole remote technique, is in red while the three in situ techniques are plotted as different shades of purple. We see that BIKER performs slightly worse than all three in situ methods across all metrics except *r*, where it performs similarly to the 'Raymond 2013' model. **NEED ACTUAL NUMBERS HERE**. However, across all 47 rivers BIKER has the least overall bias in its estimates (**add number here**). In summary, BIKER performs slightly worse than the in situ tools, as expected. However, the drop in performance is not substantial and is promising given that the in situ models represent their best case scenarios (perfect in situ discharge records) while BIKER represents its worst-case scenario (relatively uninformed priors).

![Figure 7: a) FCO2 validation for BIKER and the three in situ HG models. b) Average carbon emissions via CO2 evasion across all rivers and timesteps, accounting for surface area. Completely remotely-sensed methods are colored in red, in situ methods in purple, and the observed values in orange.](cache/FCO2/FCO2_models.jpg)

Figures 7b compares the average carbon emissions rate (via $F_{CO_2}$) from the `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers using BIKER posterior means and the three streamgauge-based HG models. All estimates are positively biased, with BIKER (`r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_BIKER',]$sumFCO2, 2)` gG-C/yr) and 'Raymond 2013' (`r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_raymond2013',]$sumFCO2, 2)` gG-C/yr) the most overestimated relative to the observed flux (`r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_obs',]$sumFCO2, 2)` gG-C/yr). *MORE* Thus, despite BIKER using absolutely no streamgauge or other in situ data like the other 3 methods do, it provides a reasonable upscaling estimate of the bulk carbon efflux across many rivers (Figure 6c).

## 4 Discussion
### 4.1 Gas exchange in hydraulically-wide rivers
Field studies of gas exchange in wide rivers have suggested that their *k* properties behave differently than in steeper and smaller rivers [@ulsethDistinctAirWater2019; @beaulieuControlsGasTransfer2012; @alinPhysicalControlsCarbon2011; @wangPhysicallyBasedScaling2021]. While much work has focused on the small-stream side of the stream-to-river continuum, comparitvely less work has been done in large systems. Here, we focus on the 'smooth-channel' end of the continuum, using a model for gas exchange that scales *k* by both the turbulent dissipation rate and the Shear Reynold's number specifically adapted for 'hydraulically-wide' rivers. This is empirically validated in Figure 3d. Scaling *k* this way (via both the turbulent dissipation rate and the shear Reynold's number) is frequently done in oceanic gas exchange studies [*EXAMPLES*] and has been occasionally explored in open-channel flows [@moogAirWaterGasTransfer1999; @wangPhysicallyBasedScaling2021]. However to our knowledge, @moogAirWaterGasTransfer1999's specific model has never been empirically validated in rivers. Figure 3d provides this empirical verification. Future tests should be performed to explore the model's feasibility in smaller streams and/or other aquatic environments where turbulence is also generally low. Future tests should also explore other shear Reynold's scaling relations for gas exchange in rivers.

Given the algebraic structure of the various dissipation and gas exchange models tested (Table 1), the performance differences observed in Figure 3 are largely due to the exponents applied to the river hydraulic terms (which are described in the subplot x axis labels). While the use of the form-drag dissipation model includes a mean flow velocity term and improves performance, the primary difference between the respective small-eddy and chainsaw equations is the exponents. Intuitively, this exponent discrepancy indicates that in hydraulically-wide rivers where turbulence is relatively low, large-scale eddies exert a noteworthy influence on surface dissipation rates and that the small-eddy model is not sufficient.

Finally, we are not accounting for gas exchange facilitated by air-side wind, which exerts a significant influence on lake and ocean gas exchange. Similarly, none of the existing hydraulics-driven gas exchange models do either, assuming that the rivers are sheltered enough that wind is of less concern. We suspect that our model will be less representative in extremely wide rivers where wind becomes a driving factor of gas exchange rates [@beaulieuControlsGasTransfer2012]. In that vein, BIKER's outputs can be interpreted as the 'gas exchange velocity in low-wind conditions'. That said, BIKER's flexible implementation is a good start towards eventually coupling hydraulics-driven gas exchange with wind-driven gas exchange. Such a model would likely take a form similar to the conceptual model proposed by @wangPhysicallyBasedScaling2021 that proposes calculating both a wind-driven *k* and a hydraulics-driven *k* and then weighting each term via parameters. In the case of BIKER, this would necessitate additional parameters that would need to be known a priori for specific rivers. This is left to future work.

### 4.2 Towards remote sensing of global spatiotemporal dynamics of *k* in large rivers
To date, the studies exploring the spatiotemporal dynamics of riverine gas exchange have arguably been held back by a lack of data. A few studies have investigated these dynamics, but they have been limited to individual rivers and/or limited field seasons [@hallAirWaterOxygen2012; @sand-jensenCO2DynamicsDanish2012]. For example, @wallinSpatiotemporalVariabilityGas2011 performed a preliminary analysis in northern Sweden relating cross-section specific temporal variability in gas exchange with channel slope, but they were limited to a mean of only 8 measurements per river in a single watershed.

Therefore, estimating *k* from SWOT data is an attractive option, at least for exploring its spatiotemporal dynamics at fine temporal resolution and at the global-scale. Section 5 shows that BIKER successfully infers the temporal dynamics of $k_{600}$ using simulated SWOT data over a wide range of rivers and with implicit measurement errors accounted for (Figures 4-6). This strong result bodes well for BIKER's eventual implementation on real SWOT observations, as does BIKER's improved performance with longer timeseries (Figure 6b). The nominal three-year SWOT campaign will provide hundreds of observations for most rivers, suggesting even better BIKER inversions than those presented here. All of this suggests that near-daily estimation of the spatiotemporal trends in global riverine gas exchange will be possible once SWOT launches.

With that said, Figures 4-7 highlight a substantial range of algorithm performances across rivers. These differences in performance are likely due to the representativeness of the priors used for that river. This all makes sense as Section 2 has effectively reduced $k_{600}$ to a function of hydraulics that are nearly all directly measurable by SWOT. Any resulting bias in BIKER's predictions is therefore attributable to bias in the priors used for the non-remotely sensed terms (equations 10-11). For SWOT discharge algorithms, authors have repeatedly shown that the 'quality' of prior hyperparameters plays a large role in the success of discharge inversions [@frassonExploringFactorsControlling2021; @andreadisConstrainingAssimilationSWOT2020; @brinkerhoffConstrainingRemoteRiver2020; @tuozzoloEstimatingRiverDischarge2019] and our results here further corroborate this finding. It should be stressed that a substantial portion of rivers from [@frassonExploringFactorsControlling2021] are canal-shaped in nature with different hydraulic properties than a natural river channel. Nearly universally, these rivers underperformed (Figure 5), however it is impossible to isolate whether that is due to channel geomorphology or the fact that these canals also usually had short timeseries (around 12 days).

### 4.3 Towards remotely sensing riverine carbon emissions using SWOT
Figure 7 confirms that BIKER is quite successful, without any in situ information aside from a $CO_2$ logger, at predicting both $FCO_2$ (Figure 6a) and 2) the bulk carbon efflux (Figure 7b-7c). This encouraging result has two main implications for future work coupling remote sensing via SWOT with in situ data. First, it confirms that we can couple BIKER and SWOT with in situ gas concentration loggers to produce reasonable $FCO_2$ estimates at novel temporal resolutions in SWOT-observable rivers. This is particularly useful given recent advances in in situ gas concentration loggers [e.g. @ahoHighestRatesGross2021] but no such similar advances in modeling *k* at equivalent temporal resolutions. BIKER can likely also be ran at the field scale using arrays of pressure transducers to estimate water surface slope (rather than using satellite-based altimeters like SWOT) following recent similar work for estimating streamflow [@harlanDischargeEstimationDense2021].

Secondly, it is important to again stress that unlike BIKER, the HG models in Figure 7 rely on an in situ streamgauge. This means that Figure 7 represents the best performance that those models could ever have, while BIKER represents the worse performance it could ever have. If the HG models are ran using modeled discharge, their accuracy would necessarily decrease. Thus, these results suggest that BIKER will be useful in not only assessing gas exchange velocities, but also in providing additional measurements to upscaling workflows that rely on modeled streamflow [e.g. **Liu etal in review**].

## 5 Conclusions
Gas exchange from aquatic systems has been studied for nearly a century and dozens of predictive equations have been developed to obtain *k* values from easily measured hydraulic properties rather than via direct measurement. Likewise, equation 3 has been repeatedly verified across many non-fluvial environments. However. despite renewed interest in fluvial gas exchange in the last decade or so, there are considerable uncertainties in how fluvial *k* varies with respect to time. This is generally due to a lack of direct measurements of $k_{600}$. In this paper, we propose that the soon-to-launch SWOT satellite will provide enough hydraulic measurements to analyse the temporal dynamics of $k_{600}$ using SWOT measurements, and therefore allow for a global-scale analysis of spatiotemporal trends in *k* once SWOT launches.

In preparation for SWOT's launch, we develop the BIKER algorithm to infer $k_{600}$ using absolutely no on-the-ground information. Validating on simulated SWOT rivers, we show show strong recovery of rivers' temporal $k_{600}$ dynamics. We also pair BIKER with an in situ $CO_2$ logger and show strong performance in reproducing $CO_2$ evasion fluxes and bulk carbon efflux from the rivers' combined surfaces. These strong results functionally open the door for global-scale, near daily estimates of fluvial gas exchange velocity once SWOT launches in 2022. This unprecedented amount of data should allow for significant insights into the temporal dynamics of fluvial gas exchange in large rivers around the world.

## 6 Acknowledgements
C.B. Brinkerhoff was funded on **FINESST**. The version of BIKER used in this study is available at **citable code here**. BIKER remains in active development and is available at https://github.com/craigbrinkerhoff/BIKER. All code to build and generate results, figures, and the manuscript is available at https://github.com/craigbrinkerhoff/RSK600. We thank Renato Frasson, Michael Durand, Amber Ulseth, and Jake Beaulieu for generously making their data available for this study. We also thank the entire SWOT discharge working group for their decade-plus body of work which inspired this study.

## 7 Apendix A
*Table A1: Variable description and notation for this study. ADD THE BAYES PARAMETERS YET*

|**Notation**|**Description**|**Calculation (if not directly measured)**|**Units**|
|------------|---------------|-------------------------------|---------|
|$A$|Channel cross-sectional area|NA|$[L^2]$|
|$A_0$|Non-SWOT-observable cross-sectional area|NA|$[L^2]$|
|$\alpha$|Statistical parameter for equation 5 scaling relation|NA|$\frac{m^{1/2}s^{3/4}}{dy}$|
|$\alpha_1$|Statistical parameter for equation 6 scaling relation|NA|$\frac{m^{1/2}s^{3/4}}{dy}$|
|*Bulk carbon efflux*|carbon mass transport rate from river to atmosphere|NA|$[\frac{M}{T}]$|
|$\beta$|Statistical parameter for equation 7 scaling relation|NA|$\frac{m^{-1/4}s^{9/8}}{dy}$|
|$\beta_1$|Statistical parameter for equation 8 scaling relation|NA|$\frac{m^{1/2}s^{3/4}}{dy}$|
|$CO_{2_{water}}$|Water-side $CO_2$ concentration|NA|$[L]$|
|$CO_{2_{air}}$|Atmospheric-side $CO_2$ concentration|NA|$[L]$|
|$dA$|change in cross-sectional area|$\delta H_e*W$|$[L]$|
|$D_m$|Molecular diffusion coefficient|NA|$[\frac{L^2}{T}]$|
|$\epsilon$|Dissipation rate of near-surface turbulence|NA|$[\frac{E}{M*T}]$|
|$\epsilon_S$|log-law-of-the-wall model for $\epsilon$|$\frac{U_*^3}{H}$|$[\frac{E}{M*T}]$|
|$\epsilon_D$|Form-drag model for $\epsilon$|$gS\bar{U}$|$[\frac{E}{M*T}]$
|$FCO_2$|$CO_2$ evasion flux from river to atmosphere|NA|$[\frac{M}{L^2T}]$|
|$FCO_{2_{upscaled}}$|Upscaling estimate of the  global $CO_2$ evasion flux from river to atmosphere|NA|$[\frac{M}{L^2T}]$|
|$g$|gravitational acceleration|9.8|$[\frac{L}{T^2}]$|
|$H$|Mean flow depth|$A/W$|$[L]$|
|$H_e$|Water surface elevation|NA|$[L]$|
|$i$|Cross-section discretization within river reach|NA|NA|
|$k$|gas exchange velocity|NA|$[\frac{L}{T}]$|
|$k_{600}$|gas exchange velocity normalized to $Sc=600$|$(\frac{600}{Sc_k})^{-0.5}k$|$[\frac{L}{T}]$|
|$k_{600_{obs}}$|"observed" gas exchange velocity normalized to $Sc=600$|$k_{600_{obs}}=62.82(gS)^{7/16}\bar{U}^{1/4}H^{9/16}$|$[\frac{L}{T}]$|
|$n$|Manning's roughness coefficient|$\frac{R_h^{2/3}S^{1/2}}{\bar{U}}$|$[\frac{T}{L^3}]$|
|$\uprho$|Density of water|NA|$[\frac{M}{L^3}]$|
|$Q$|River discharge|NA|$\frac{L^3}{T}$|
|$R_h$|Hydraulic radius|$\frac{AW}{2H+W}$|$[L]$|
|$S$|River slope|NA|$unitless$|
|$Sc$|Schmidt number|$\frac{v}{D_m}$|NA|
|$\sigma^2_{k_{600}}$|BIKER uncertainty (posterior distribution standard deviation)|NA|$[L/T]$|
|$t$|timestep discretization within river reach|NA|NA|
|$\Theta$|Bayesian parameter set|NA|NA
|$\bar{U}$|Cross-sectional average velocity|$\frac{Q}{A}$|$[\frac{L}{T}]$|
|$U_*$|Shear velocity|$\sqrt{gSR_h}$|$[\frac{L}{T}]$|
|$\mu$|Viscosity|NA|$[P*T]$|
|$v$|kinematic viscosity|$\frac{\mu}{\uprho}$|$[\frac{L^2}{T}]$|
|$W$|Flow width|NA|$[L]$|
|$x$|Bayesian data set|NA|NA|

## References

### 2022-01-07 generateMS.R

@@ -14,19 +14,15 @@ percs_overall <- group_by(percs, flag_hydraulicWide) %>%
   summarise(total = sum(n)) %>%
   mutate(perc = 100 * total/sum(total))
 
+results_dynamics <- read.csv('cache/validation/results_dynamics.csv')
 results_k600_rivs <- read.csv('cache//validation//results_by_riv.csv')
 results_fco2_rivs <- read.csv('cache//FCO2//CO2_results.csv')
 results_fco2_bulk <- read.csv('cache//FCO2//massFluxes.csv')
 results_fco2_bulk$key <- as.character(results_fco2_bulk$key)
 
 #knit manuscript
-rmarkdown::render("manuscript_v6.Rmd")
+rmarkdown::render("manuscript_v7.Rmd")
 print('Manuscript generated.')
 
-rmarkdown::render("supp_v6.Rmd")
-print('Supplemental Information generated.')
-
-
-#scraps
-#results_k600_all <- read.csv('cache//validation//results_all_riv.csv')
-#results_fco2_all <- read.csv('cache//FCO2//fco2_stats_all.csv')
+#rmarkdown::render("supp_v7.Rmd")
+#print('Supplemental Information generated.')

### 2022-01-07 src/buildMap.R

@@ -1,42 +1,0 @@
-#Craig Brinkerhoff
-#pepsi 2 river locations map
-#Summer 2021
-
-library(tidyverse)
-library(tmap)
-library(cowplot)
-library(RColorBrewer)
-library(sf)
-library(grid)
-theme_set(theme_cowplot())
-
-setwd('C:\\Users\\craig\\Documents\\GitHub\\RSK600')
-
-##########
-#Read in locations and rivers----------------
-#########
-data <- read.csv('data\\dataLocations.csv', header = T)
-colnames(data) <- c('River', 'lat', 'lon', 'Data')
-#data <- data[-1,]
-data$lat <- as.numeric(as.character(data$lat))
-data$lon <- as.numeric(as.character(data$lon))
-
-global_rivs <- st_read('data\\ne_10m_rivers_lake_centerlines\\ne_10m_rivers_lake_centerlines.shp')
-global_land <- st_read('data\\ne_50m_land\\ne_50m_land.shp')
-data <- st_as_sf(data, coords = c('lon', 'lat'))
-st_crs(data) <- st_crs(global_rivs)
-
-##########
-#main map------------
-##########
-base <- tm_shape(global_land) +
-  tm_polygons(col = '#99d8c9') +
-  tm_compass(position = c('RIGHT', 'TOP')) +
-  tm_scale_bar(width=1/10, position=c('RIGHT', 'BOTTOM'), lwd=2)
-map <- tm_shape(global_rivs)+
-  tm_lines(col='darkblue', scale=0.75)
-rivers <- tm_shape(data) +
-  tm_symbols(col =  "Data", size=0.5, shape=21, palette = c('#e31a1c', '#33a02c', '#1f78b4'), border.col = 'black', border.lwd = 1)
-map <- base + map + rivers
-
-tmap_save(map, 'cache/maps/map.jpg')

### 2022-01-07 src/old/buildMap.R

@@ -1,0 +1,152 @@
+#Craig Brinkerhoff
+#pepsi 2 river locations map
+#Summer 2021
+
+library(ggplot2)
+library(tmap)
+library(cowplot)
+library(RColorBrewer)
+library(sf)
+library(grid)
+theme_set(theme_cowplot())
+
+setwd('C:/Users/craig/Documents/OneDrive - University of Massachusetts/Ongoing Projects/RSK600')
+#setwd("C:/Users/cbrinkerhoff/OneDrive - University of Massachusetts/Ongoing Projects/RSK600")
+
+##########
+#Read in locations and rivers----------------
+#########
+data <- read.csv('data\\dataLocations.csv', header = T)
+colnames(data) <- c('River', 'lat', 'lon', 'Data')
+#data <- data[-1,]
+data$lat <- as.numeric(as.character(data$lat))
+data$lon <- as.numeric(as.character(data$lon))
+
+global_rivs <- st_read('data\\ne_10m_rivers_lake_centerlines\\ne_10m_rivers_lake_centerlines.shp')
+global_land <- st_read('data\\ne_50m_land\\ne_50m_land.shp')
+data <- st_as_sf(data, coords = c('lon', 'lat'))
+st_crs(data) <- st_crs(global_rivs)
+break
+##########
+#main map------------
+##########
+base <- tm_shape(global_land) +
+  tm_polygons(col = '#99d8c9') +
+  tm_compass(position = c('RIGHT', 'TOP')) +
+  tm_scale_bar(width=1/10, position=c('RIGHT', 'BOTTOM'), lwd=2)
+map <- tm_shape(global_rivs)+
+  tm_lines(col='darkblue', scale=0.75)
+rivers <- tm_shape(data) +
+  tm_symbols(col =  "Data", size=0.5, shape=21, palette = c('#e31a1c', '#33a02c', '#1f78b4'), border.col = 'black', border.lwd = 1)
+map <- base + map #+ rivers
+
+tmap_save(map, 'cache/maps/map.jpg')
+
+
+# 
+# 
+# 
+# #subpanel CA
+# ca <- filter(data, River %in% c('AshSlough',
+#                                 'Berenda Slough',
+#                                 'Chowchilla Canal',
+#                                 'Fresno River',
+#                                 'Grant Line Canal',
+#                                 'Mariposa Bypass',
+#                                 'Merced River',
+#                                 'Middle River',
+#                                 'San Joaquin',
+#                                 'San Joaquin 2',
[…]

### 2022-01-07 src/swot_k_model_v3.R

@@ -15,7 +15,7 @@ library(tidyr)
 library(ggplot2)
 library(colorspace)
 library(cowplot)
-theme_set(theme_cowplot())
+theme_set(theme_classic())
 
 #setwd('C:\\Users\\cbrinkerhoff\\OneDrive - University of Massachusetts\\Ongoing Projects\\RSK600')
 setwd('C:\\Users\\craig\\Documents\\OneDrive - University of Massachusetts\\Ongoing Projects\\RSK600')
@@ -237,6 +237,8 @@ models <- data.frame('name'=c('Chainsaw-eS', 'Small-eddy-eS', 'Chainsaw-eD', 'Sm
                      )
 write.csv(models, 'cache\\k600_theory\\hydraulicWide_models.csv')
 
+
+
 #######MONTE CARLO PROPOGATION OF UNCERTANTIES-------------------------
 beta_sigma <- summary(lm_hydraulicallyWide_chainsaw_eD)$sigma
 u_sigma <- 0.3
@@ -247,7 +249,7 @@ n <- 10000 #MC sample size
 output <- 1:nrow(hydraulicallyWide)
 for (i in output) {
   log_knowns <- (7/16)*log(g) + (7/16)*log(hydraulicallyWide[i,]$slope) + (9/16)*log(hydraulicallyWide[i,]$depth)
-  log_k600_pred <- rnorm(n, log(summary(lm_hydraulicallyWide_chainsaw_eD)$coefficient[1]), log(beta_sigma)) + knowns + (1/4)*rnorm(n, log(hydraulicallyWide[i,]$Vms), u_sigma)
+  log_k600_pred <- rnorm(n, log(summary(lm_hydraulicallyWide_chainsaw_eD)$coefficient[1]), log(beta_sigma)) + log_knowns + (1/4)*rnorm(n, log(hydraulicallyWide[i,]$Vms), u_sigma)
   output[i] <- sd(log_k600_pred)
 }
 

### 2022-01-14 manuscript_v7.Rmd
some ms changes

@@
 ---
-title: "Using hydraulic geometry to remotely sense gas exchange velocity via the SWOT satellite"
+title: "Using hydraulic geometry and the SWOT satellite to remotely sense river gas exchange velocity"
 author:
@@
 In that vein, there have been over 20 empirical models developed to predict *k*, generally using some combination of mean velocity $\bar{U}$, shear velocity $U_*$, width $W$, depth $H$, and slope $S$ as predictors [@wangPhysicallyBasedScaling2021].
-Note that these models usually predict $k_{600}$, or *k* normalized by a Schmidt number of 600.
-This is to remove the effect of water temperature and gas type from predictive models, as warmer waters and lower *Sc* numbers both increase gas exchange rates.
-For example, $k_{600}$ reflects the $k_{CO_{2}}$ at 20 degrees Celsius.
-Through this normalization, these models exclusively perturb the hydraulic controls on gas exchange [@hallGasExchangeStreams2020].
+These models usually predict $k_{600}$, or *k* normalized by a Schmidt number of 600.
+This is to remove the effect of water temperature and gas type from predictive models, as warmer waters and lower *Sc* numbers both increase gas exchange rates [@hallGasExchangeStreams2020], i.e.e $k_{600}$ reflects the $k_{CO_{2}}$ at 20 degrees Celsius.
+Through this normalization, these models exclusively perturb the geophysical controls on gas exchange [@hallGasExchangeStreams2020].
 
-These empirical models enable estimating a global $FCO_2$ flux ($FCO_2_{upscaled}$) by summing across millions of rivers, but they also change the base parameters that ultimately control that aggregate estimate.
+These empirical models enable estimating a global $FCO_2$ flux ($FCO_2_{upscaled}$) from millions of rivers, but they also change the base parameters that ultimately control that aggregate estimate.
 That is, by making *k* a function of hydraulics, $FCO_2_{upscaled}$ is now a direct function of river hydraulics.
@@
 This knowledge gap is contrasted agaist the robust existing literature exploring spatiotemporal patterns in $CO_2_{water}$ [e.g.
-@crawfordCO2TimeSeries2017; @peterScalesDriversTemporal2014; @ranDynamicsRiverineCO22017; @rocher-rosLandscapeProcessDomains2019; @raymondAtmosphericCO2Evasion2000].
+@ahoDistinctConcentrationdischargeDynamics2021; @ahoDifferentialResponseGreenhouse2019; @crawfordCO2TimeSeries2017; @peterScalesDriversTemporal2014; @ranDynamicsRiverineCO22017; @rocher-rosLandscapeProcessDomains2019; @raymondAtmosphericCO2Evasion2000; @liuHydrologicControlsPCO22018a].
 
 To that end, the field has made significant progress in estimating $CO_2_{water}$, where we now have river-reach explicit modeling at monthly temporal resolutions using global hydrography datasets [@horgbyUnexpectedLargeEvasion2019; @brinkerhoffLakeMorphometryRiver2021; @saccardiImprovingPredictionsStream2021; *Liu etal in review*], but an equivalently sophisticated representation of $k_{CO_2}$ is still lacking.
-As equations 1 and 2 dictate, these terms share the burden of calculating $FCO_2$.
-Therefore, these next-generation models will have to contend with this discrepancy, which is directly dictated by the lack of direct measurements of global river hydraulics.
+As equations 1 and 2 dictate, $CO_2_{water}$ and $k_{CO_2}$ share the burden of calculating $FCO_2$ and therefore these next-generation models will have to contend with this discrepancy, which is directly dictated by the lack of direct measurements of global river hydraulics.
 
@@
 This is turn would also address the uncertainty regarding continental-to-global scale temporal dynamics of gas exchange noted earlier.
-Conveniently, these very data will soon be available via the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission.
+Conveniently, these data will soon be available via the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission.
 SWOT is expected to launch in 2022 and provide the world's first direct measurements of global water surface extent and elevation at novel temporal resolutions.
@@
 Via its direct hydraulic measurements, SWOT is expected to usher in a sea change in global-scale hydrology, and could similarly transform fluvial biogeochemsitry if techniques are developed to ingest SWOT data and infer *k*.
-Here, we borrow tools from fluvial geomorphology and existing SWOT algorithms to answer the following two questions:
+In this context, we borrow tools from fluvial geomorphology and existing SWOT algorithms to answer the following two questions:
 
-  - Can we leverage the hydraulic geometry of the large rivers SWOT will observe to develop a physically-based model for $k_{600}$ unique to these rivers?
+  - Can we use hydraulic geometry to develop a physically-based model for $k_{600}$ unique to these rivers?
 
@@
 
-To answer this first question, we use the specific hydraulic geometry of large, SWOT-observable rivers to develop a process-based model for large-river gas exchange.
+To answer this first question, we use the specific hydraulic geometry, or the fundamental geomoprhic relationships between streamflow and channel shape [@leopoldHydraulicGeometryStream1953] of large, SWOT-observable rivers to develop a process-based model for large-river gas exchange.
 We then take these findings and explore question 2 by implementing the model within an RS of $k_{600}$ algorithm named BIKER ('Bayesian Inference of the $k_{600}$ Evasion Rate').
-The goal with BIKER is to require no in situ inputs of any kind (although in situ data could be ingested and should improve results) such that it is completely globally implementable on any river on Earth that SWOT will measure.
+The goal with BIKER is to require no in situ inputs of any kind (although in situ data could be ingested and should improve results) such that it is completely globally implementable on any SWOT-observable river on Earth.
 Finally, we also test whether BIKER can predict actual gas fluxes when paired with an in situ $CO_2$ sensor and compare it against established methods that are reliant on in situ hydraulic measurements.
@@
 
-![Figure 1: Flowchart detailing entire study.
+![Figure 1: Flowchart detailing the BIKER algorithm.
 Dashed lines indicate workflow to calculate the observed data that we validate against.
@@
 ## 2 Hydraulic geometry in large rivers and predicting fluvial gas exchange velocity
-To predict *k* in the very large rivers that SWOT will observe, we start from established process-based models for *k*, impose hydraulic assumptions that are valid in large rivers, and then obtain a suite of models we can test.
+To predict *k* in the very large rivers that SWOT will observe, we start from established process-based models for *k*, impose hydraulic assumptions that are valid in large rivers, and then obtain a suite of models we can empirically test.
 Following a description of the data used (section 2.1), we outline these established models (section 2.2) and then step through our hydraulic assumptions to arrive at equations that can be ingested by SWOT (section 2.2).
@@
 ### 2.1 Data
-Our gas exchange theory is developed using multiple datasets of directly measured $k_{600}$ and directly measure stream hydraulics collected from the literature.
+Our gas exchange theory is developed using multiple datasets of directly measured $k_{600}$ and directly measured stream hydraulics collected from the literature.
 In total, we have collected 763 measurements of $k_{600}$, with 701 of these measurements previously published by @raymondScalingGasTransfer2012 and @ulsethDistinctAirWater2019.
-These come from tracer studies and thus define $k_{600}$ and the rivers' mean hydraulic properties at the reach scale.
 The remaining measurements were measured by @churchillPredictionStreamReaeration1964 and @owensReaerationStudiesStreams1964.
 See Table S1 for a complete list of the studies that collected these measurements.
+All measurements come from tracer studies and thus define $k_{600}$ and the rivers' mean hydraulic properties at the reach scale.
 
-We expand our dataset of stream hydraulics using a previously published compilation of in situ hydraulic measurements [@brinkerhoffReconcilingAtaStationAtManyStations2019].
+To qunatify how many SWOT-obsrvable rivers meet our hydraulic assumptions, we expand our set of stream hydraulics (but not $k_{600}$) using a previously published compilation of in situ hydraulic measurements [@brinkerhoffReconcilingAtaStationAtManyStations2019].
 That dataset contains over 530,000 unique measurements of river channel velocity, width, and discharge from across the continental United States, originally taken to calibrate United States Geological Survey (USGS) streamgauge rating curves and made public by the USGS.
-Ultimately, this dataset was filtered to `r sum(percs$n)` discrete measurements for this study, in addition to the direct measurements in the gas exchange dataset (see Text S1 for our filtering protocol).
 
 ### 2.2 Process-based modeling of river *k*
-*k* should scale with near-surface turbulence in a turbulent flow [@hallGasExchangeStreams2020] and thus extensive field and laboratory experiments have converged on the 'small-eddy model' as first proposed by @lamontEddyCellModel1970 *Zappa said this actually goes back further than this paper* to scale *k* via the smallest-scale turbulent eddies [e.g.
+*k* should scale with near-surface turbulence in a turbulent flow [@hallGasExchangeStreams2020] and thus extensive field and laboratory experiments have converged on the 'small-eddy model' as derived by @lamontEddyCellModel1970 and empirically anticipated by @calderbankContinuousPhaseHeat1961.
+Based on classic surface-renewal theory, this model scales *k* via the smallest-scale (the 'Kolmogorov scale') turbulent eddies and has been repeatedly empirically validated in freshwater systems [e.g.
 @katulStructureFunctionModel2018; @lorkeUnifiedScalingRelation2006; @zappaEnvironmentalTurbulentMixing2007; @zappaVariationSurfaceTurbulence2003; @moogStreamReaerationNonuniform1999; @vachonRelationshipNearsurfaceTurbulence2010; @tokoroHighGastransferVelocity2008; @wangPhysicallyBasedScaling2021].
@@
 Given this observation, @moogAirWaterGasTransfer1999 proposed an extension to the small-eddy model, additionally scaling $k_{600}$ using a shear Reynold's number formulation.
-*look up other shear reynolds ways to additionally scale k*.
-This is equation 4 and is referred to as the 'Chainsaw model'.
-The Chainsaw model is hypothetically useful in low-turbulence flows where a relative lack of large-scale eddies effectively 'filter out' the number of small-eddies that actually reach the interface and lead to gas exchange [@talkeTurbulentKineticEnergy2013].
-In the context of BIKER, we chose to test this model because large, SWOT-observable rivers are generally the smoothest, least-turbulent flows along the stream-to-river continuum.
-Further, to our knowledge the chainsaw model has never been empirically tested in predicting *k*, aside from confirming that large-scale eddies differentially move turbulence to the surface in a hydraulically-wide river [@talkeTurbulentKineticEnergy2013].
+*HERE: Zappa's note that this is frequently done in air-ocean gas exchange studies*.
+This is equation 4 and is referred to as the 'Reynolds extension' model.
+The Reynolds model is hypothetically useful in low-turbulence flows where a relative lack of large-scale eddies effectively 'filter out' the number of small-eddies that actually reach the interface and lead to gas exchange [@talkeTurbulentKineticEnergy2013].
+In the context of BIKER, we chose to test this model because large, SWOT-observable rivers are generally the smoothest, least-turbulent flows along the stream-to-ocean continuum.
+Further, to our knowledge this Reynolds extension model has never been empirically tested in predicting river *k*, aside from confirming that large-scale eddies differentially move turbulence to the surface in a hydraulically-wide river [@talkeTurbulentKineticEnergy2013].
 
@@
 
-### 2.3 Imposing large-river hydraulic geometry
+### 2.3 Imposing large-river hydraulic geometry on *k* models
 Rivers and streams change predictably along their longitudinal profile from headwater to ocean, and we can exploit the distinct hydraulic geometry of large rivers at the end of this continumn to estimate *k* in SWOT-observable systems.
 In general, as river size increases, channels become more rectangular, their shapes elongate (becoming wider quicker than they become deeper) and their hydraulic radii begin to approximate their mean flow depth, i.e.
-$R_h \approx H$ [@leopoldHydraulicGeometryStream1953].
-This is a common assumption in hydraulic and geomorphic modeling of large rivers, and occurs in `r round(percs[percs$flag_swot == 'SWOT' & percs$flag_hydraulicWide == 'Rh=H',]$perc,0)`% of our SWOT-observable river measurements but only `r round(percs_overall[percs_overall$flag_hydraulicWide == 'Rh=H',]$perc,0)`% of non-SWOT observable river measurements (calculated using both datasets of stream hydraulics outlined in section 2.1- see Text S1 for full details).
-These channels are often referred to as 'hydraulically-wide'.
+$R_h \approx H$.
+This is a common assumption in hydraulic and geomorphic modeling of large rivers, where the average $\frac{R_h}{H}$ ratio across our `r HG_swot$n` SWOT-observable river flows is `r HG_swot$mean * 100`% (see Text S1 for how we built this dataset).
+We refer to these flows as 'hydraulically-wide'.
 
@@
 
-#### 2.3.2 'Chainsaw' model for gas evasion
-We also impose $\epsilon=\epsilon_S$ and $R_H=H$ on the chainsaw model just like we did in section 2.3.1 and arrive at equation 7 (with statistical coefficient $\beta$).
+#### 2.3.2 'Reynolds extension' model for gas evasion
+We also impose $\epsilon=\epsilon_S$ and $R_H=H$ on the Reynolds extension model just like we did in section 2.3.1 and arrive at equation 7 (with statistical coefficient $\beta$).
 The result is identical to the hydraulically-wide small-eddy model (equation 5), except for the exponents which are both steeper.
@@
 
-|**Description**|**Equation**|**References**|**Empirical test**|**Empirical fit via $r^2$**|
-|---------------|------------|--------------|------------------|-----------------|
-|Small-eddy, log-law-of-the-wall|$k_{600}=\alpha(gs)^{3/8}H^{1/8}$|@lamontEddyCellModel1970; log-law-of-the-wall|Figure 3a|`r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eS',]$r2, 2)`|
-|Small-eddy, form-drag dissipation|$k_{600}=\alpha_1(gS\bar{U})^{1/4}$|@lamontEddyCellModel1970; @moogStreamReaerationNonuniform1999|Figure 3c|`r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eD',]$r2, 2)`|
-|Chainsaw, log-law-of-the-wall|$k_{600}=\beta(gs)^{9/16}H^{11/16}$|@moogAirWaterGasTransfer1999; log-law-of-the-wall|Figure 3b|`r round(modelsWIDE[modelsWIDE$name == 'Chainsaw-eS',]$r2, 2)`|
-|Chainsaw, form-drag dissipation|$k_{600}=\beta_1(gs)^{7/16}\bar{U}^{1/4}H^{9/16}$|@moogAirWaterGasTransfer1999; @moogStreamReaerationNonuniform1999|Figure 3d|`r round(modelsWIDE[modelsWIDE$name == 'Chainsaw-eD',]$r2, 2)`|
+|**Description**|**Equation**|**Empirical test**|**Empirical fit via $r^2$**|
+|---------------|------------|------------------|-----------------|
+|Small-eddy, log-law-of-the-wall|$k_{600}=\alpha(gs)^{3/8}H^{1/8}$|Figure 3a|`r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eS',]$r2, 2)`|
+|Small-eddy, form-drag dissipation|$k_{600}=\alpha_1(gS\bar{U})^{1/4}$|Figure 3c|`r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eD',]$r2, 2)`|
+|Chainsaw, log-law-of-the-wall|$k_{600}=\beta(gs)^{9/16}H^{11/16}$|Figure 3b|`r round(modelsWIDE[modelsWIDE$name == 'reynolds-eS',]$r2, 2)`|
+|Chainsaw, form-drag dissipation|$k_{600}=\beta_1(gs)^{7/16}\bar{U}^{1/4}H^{9/16}$||Figure 3d|`r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2, 2)`|
 
 ### 2.4 Model validation
-With equations 5-8 defined, we now empirically test their strength of fit for hydraulically-wide channels.
-We validate on the dataset of direct measurements of $k_{600}$ and stream hydraulics, after filtering for measurements made in hydraulically-wide channels.
-This was defined here as those whose hydraulic radius was within 1% of their mean flow depth.
-All told, this amounts to 166 *double check* direct measurements of hydraulically-wide $k_{600}$ and stream hydraulics to test with.
-Models were assesed via the coefficient of determination ($r^2$) and results are plotted in Figure 3.
+With equations 5-8 derived, we now empirically test their strength of fit for hydraulically-wide river flows.
+We validate on the dataset of direct measurements of $k_{600}$ and stream hydraulics, after filtering for measurements made in hydraulically-wide channels (n= `r n`).
+As equations 5-8 necessitate that this assumption is true, we defined 'hydraulically-wide' as flows whose hydraulic radius was within 1% of their mean flow depth.
+All told, this amounts to `r n` direct measurements of hydraulically-wide $k_{600}$ and stream hydraulics to test with.
+Models were assessed via the coefficient of determination ($r^2$) and results are plotted in Figure 3.
 Note that Figure 3 axes are plotted in logarithmic space just for visualization: model fit and validation (via $r^2$) were calculated in linear space as their models dictate.
@@
 They systematically underestimate lower gas exchange velocities and overestimate higher ones, with the log-law-of-the-wall model (Figure 3a) performing marginally worse both visually and via $r^2$ (`r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eS',]$r2, 2)` versus `r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eD',]$r2, 2)` for Figure 3c).
-However, when this model is expanded to retard the dissipation rate in the smoothest flows via the Chainsaw model [equations 6 and 7- @moogAirWaterGasTransfer1999], the dynamics are more accurately captured (Figures 3b, 3d) and $r^2$ scores go up appreciably (`r round(modelsWIDE[modelsWIDE$name == 'Chainsaw-eS',]$r2, 2)` and `r round(modelsWIDE[modelsWIDE$name == 'Chainsaw-eD',]$r2, 2)` for the log-law-of-the-wall and form-drag dissipation, respectively).
+However, when this model is expanded to retard the dissipation rate in the smoothest flows via the Reynolds extension model [equations 6 and 7- @moogAirWaterGasTransfer1999], the dynamics are more accurately captured (Figures 3b, 3d) and $r^2$ scores go up appreciably (`r round(modelsWIDE[modelsWIDE$name == 'reynolds-eS',]$r2, 2)` and `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2, 2)` for the log-law-of-the-wall and form-drag dissipation, respectively).
 Modeling dissipation via total form-drag rather than just via bed shear and the log-law-of-the-wall improves model performance.
@@
 
-The success of the chainsaw model in hydraulically-wide channels, and in particular when coupled with the $\epsilon=\epsilon_D$ model (equation 8, Figure 3d), provides us with a strong physical-model for gas evasion.
+The success of the Reynolds extensions model in hydraulically-wide channels, and in particular when coupled with the $\epsilon=\epsilon_D$ model (equation 8, Figure 3d), provides us with a strong physical-model for gas evasion.
 The river hydraulics terms in equation 8 can either be directly measured or reasonably inferred from SWOT measurements, effectively opening the door for remotely sensing the gas exchange velocity.
@@
 ## 3 Exploiting equation 8 to remotely sense gas exchange velocity
-We have shown that scaling $k_{600}$ via equation 8 explains `r round(modelsWIDE[modelsWIDE$name == 'Chainsaw-eD',]$r2,2)*100`% of the variation in $k_{600}$ in hydraulically-wide rivers, which includes nearly all SWOT-observable rivers.
+We have shown that scaling $k_{600}$ via equation 8 explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2,2)*100`% of the variation in $k_{600}$ in hydraulically-wide rivers, which includes nearly all SWOT-observable rivers.
 Further, equation 8 has only three non-remotely-sensible terms: $k_{600}$, mean flow depth, and mean flow velocity.
@@
 
-To validate our remote sensing solution for $k_{600}$ (BIKER), we cannot use SWOT measurements as it has yet to launch.
+To validate BIKER, we cannot use actual SWOT measurements as SWOT has yet to launch.
 In the hydrology literature, it is therefore standard practice to benchmark SWOT-related algorithms on "SWOT-like" data.
-We use 47 SWOT-simulated rivers for validation, where these simulated rivers are simply river-reach-averaged hydraulic model outputs where the water surface heights, slopes, and widths are labelled as RS observations and are used as the sole inputs to BIKER.
+We use 47 SWOT-simulated rivers for validation, where these simulated rivers are simply reach-averaged hydraulic model outputs where the water surface heights, slopes, and widths are labelled as RS observations and are used as the sole inputs to BIKER.
 These were published by @frassonExploringFactorsControlling2021 and @durandIntercomparisonRemoteSensing2016 and are the two benchmarking studies that explored remote sensing of discharge (RSQ) algorithm performance for the SWOT mission.
-They have prepared the data to mimic SWOT by adding realistic radar errors and sampling along the satellite’s future ground track.
-In essence, these authors have run standard hydraulic models forced with known inflows and measured bathymetry to model the hydraulic response of these river reaches, and then sampled these model outputs according to the timing of the satellite overpass.
+They used standard hydraulic models forced with known inflows and measured bathymetry to model the hydraulic response of the rivers and then further prepared the data to mimic SWOT by adding realistic radar errors and sampling along the satellite’s future ground track.
 @frassonExploringFactorsControlling2021 also developed a sophisticated error model that mimics the effects of the atmosphere, topography, and water surface geometry on SWOT measurements (16/47 rivers) to add realistic errors that depend on the time and place of observation.
@@
 This encouraging result has two main implications for future work coupling remote sensing via SWOT with in situ data.
-First, it confirms that we can couple BIKER and SWOT with in situ gas concentration loggers to produce reasonable $FCO_2$ estimates at novel temporal resolutions in SWOT-observable rivers.
-This is particularly useful given recent advances in in situ gas concentration loggers [e.g.
-@ahoHighestRatesGross2021] but no such similar advances in modeling *k* at equivalent temporal resolutions.
-BIKER can likely also be ran at the field scale using arrays of pressure transducers to estimate water surface slope (rather than using satellite-based altimeters like SWOT) following recent similar work for estimating streamflow [@harlanDischargeEstimationDense2021].
+First, it confirms that we can couple BIKER with in situ gas concentration loggers to produce $FCO_2$ estimates at novel temporal resolutions in SWOT-observable rivers.
+High temporal fidelity datasets of dissolved $CO_2$ in SWOT-observable rivers are coming online [e.g.
+@ahoHighestRatesGross2021] but no such similar datasets for $k_{CO_2}$ at equivalent temporal resolutions exist.
+For rivers unobservable by SWOT, we further suggest that BIKER could be run at the field scale (rather than via satellite-based altimeters like SWOT) using arrays of in situ pressure transducers to estimate water surface slope following recent similar work for estimating streamflow [@harlanDischargeEstimationDense2021].
+Regardless, both approaches would produce $CO_2_{water}$ and $k_{CO_2}$ datasets at equivalent temporal resolution and allow us to directly calculate daily to sub-daily $CO_2$ fluxes from river reaches.
 
-Secondly, it is important to again stress that unlike BIKER, the HG models in Figure 7 rely on an in situ streamgauge.
-This means that Figure 7 represents the best performance that those models could ever have, while BIKER represents the worse performance it could ever have.
-If the HG models are ran using modeled discharge, their accuracy would necessarily decrease.
-Thus, these results suggest that BIKER will be useful in not only assessing gas exchange velocities, but also in providing additional measurements to upscaling workflows that rely on modeled streamflow [e.g.
-**Liu etal in review**].
+While this is all useful at the reach-scale, Figure 7 directly implies that BIKER will also be useful when coupled with large-scale $CO_{2_water}$ models [e.g.
+**Liu etal in review**; @saccardiImprovingPredictionsStream2021].
+These two models robustly predict dissolved $CO_2$ concentrations using two different approaches: machine learning [**Liu et al in review**] and process-based reactive transport modeling [@saccardiImprovingPredictionsStream2021] but both produce reach-scale $CO_2$ concentrations that are spatially and temporally consistent.
+Our promising results suggest that BIKER could provide additional (and directly observed) measurements to these models, thereby better informing model results through direct observations.
+This is likely accomplished via data assimilation which has proven useful for similar problems across geophysical modeling applications [@carrassiDataAssimilationGeosciences2018] and specifically for assimilating remotely sensed information into river routing models [@ishitsukaCombiningOpticalRemote2020].
 

### 2022-01-14 supp_v7.Rmd
some ms changes

--- 2022-01-07 docs/supp_v6.Rmd
@@
 ---
-title: "Supplemental Information: Exploiting hydraulic geometry to remotely sense gas exchange velocity via the SWOT satellite"
+title: "Supplemental Information: Using hydraulic geometry and the SWOT satellite to remotely sense river gas exchange velocity"
 author:
@@
       correspondence: true
-  - Peter A Raymond:
-      institute: Yale
   - Colin J Gleason:
@@
       institute: Lamont
+  - Peter A Raymond:
+      institute: Yale
   - Merritt H Harlan:
@@
   - UMass: Department of Civil & Environmental Engineering, University of Massachusetts, Amherst, MA
+  - Lamont: Lamont-Doherty Earth Observatory, Columbia University, Palisades, NY
   - Yale: School of the Environment, Yale University, New Haven, CT
-  - Lamont: Lamont-Doherty Earth Observatory, Columbia University, Palisades, NY
 bibliography: References.bib
@@
 
-After joining this dataset with the hydraulics measurements provided with the 763 $k_{600}$ measurements, we classified each measurement as hydraulically wide if its hydraulic radius was within 4% of its mean flow depth.
-Hydraulic radius was calculated assuming a rectangular channel such that $R_H=2H+W$.
-Because these measurements were almost entirely made at streamgauges, which generally have a location bias favoring stable channels near bridges and other structures that yield lower width dynamics than those observed away from gauges [@allenPatternsRiverWidth2015; @parkWorldwideVariationsHydraulic1977], we used a wider threshold of 4%.
-In the 'width-limited' scenario common at streamgauges, the hydraulic radius is likely further from the mean flow depth than would occur naturally.
-The slightly more liberal threshold of 4% allowed for implicit accounting for this sampling bias.
+Finally, we filtered this dataset to only measurements at least 100m wide ('SWOT-observable'). This left us with `r HG_swot$n` total river hydraulics measurements.
 
@@
 Consult Appendix A for all variable definitions.
-Note that $\alpha$ and $\beta$ are statistical parameters obtained via least squares regression for the small-eddy and chainsaw models, respectively.
+Note that $\alpha$ and $\beta$ are statistical parameters obtained via least squares regression for the small-eddy and Reynolds models, respectively.
 Subscripts denote variants of the same parameter, depending on which $\epsilon$ model is employed.
@@
 
-**Chainsaw models**
+**Reynolds extension models**
 
@@
 They developed a set of river channel prior hyperparameters for McFLI algorithms that are entirely RS-able and reflect differential channel hydraulics as a function of river geomorphology.
-They used an extensive database of field measurements and machine learning to identify patterns that associate river width with the hydraulic priors needed to run McFLIs so that prior hyperparameters may be assigned to rivers using only the existing remotely sensed data.
+They used an extensive database of field measurements and machine learning to identify patterns that associate river width with the hydraulic priors needed to run McFLIs so that prior hyperparameters are assigned to rivers using only the remotely sensed measurements.
 For this study, we extracted $A_0_{\lambda}$ and $A_0_{\gamma}$ as the 5th and 95th percentile values rather than the absolute maximum and minimum values to avoid physically impossible bounds on $A_0$.
@@
 This leaves the $k_{600_t}$ hyperparameters to be defined.
-$k_{600_{\mu}}$ is set by invoking the hydraulic geometry (HG) relationships developed in section 5.1.3 of the main text using the data from @brinkerhoffReconcilingAtaStationAtManyStations2019.
+$k_{600_{\mu}}$ is set by invoking the hydraulic geometry (HG) relationships developed in section 3.2.3 of the main text using the data from @brinkerhoffReconcilingAtaStationAtManyStations2019.
 We replaced both depth and velocity terms from our gas exchange model (equation 6 in the main text) with these HG models, resulting in equation S19 where Q is the mass-conserved streamflow for the river reach.
 
-$$k_{600_{\mu}} = 62.82(gS_{i,t})^{7/16}(0.265Q_{t}^{0.395})^{9/16}*(0.284Q_{t}^{0.191})^{1/4} \mathbf{(S18)}$$
+$$k_{600_{\mu}} = 62.82(gS_{i,t})^{7/16}(0.265Q_{t}^{0.395})^{9/16}*(0.284Q_{t}^{0.191})^{1/4} \mathbf{(S19)}$$
 
@@
 This is provided with the simulated SWOT data by both @durandIntercomparisonRemoteSensing2016 and @frassonExploringFactorsControlling2021.
-Again, using a temporally-invariant estimate of streamflow is the worst case scenario and BIKER's performance will improve with a more informed prior om streamflow (and therefore $k_{600}$).
+Using a temporally-invariant estimate of streamflow is the worst case scenario and BIKER's performance will improve with a more informed prior on streamflow (and therefore $k_{600}$).
+This was recently shown on many of these rivers for SWOT discharge algorithms [@frassonExploringFactorsControlling2021].
 However as noted in the main text, our primary goal with this initial validation is to benchmark BIKER's worst case scenario for performance and so we do that here.
-Because BIKER treats every timestep as a different parameter, future work should investigate ways to assign temporally varying priors for $k_{600}$.
 
-$k_{600_{\sigma}}$ is set to 0.30 (log-space).
+Finally, $k_{600_{\sigma}}$ is set to 0.30 (log-space).
 This corresponds to a coefficient of variation of approximately 0.30, which we took to reflect a reasonably strong agreement between the prior and the observed values.
@@
 $k_{600_{\gamma}}$ was set to log(500) m/day.
-The
 
@@
 Recall that $\sigma_{k_{600}}$ refers to the total uncertainty inherent in equation 8, i.e.
-stemming from both the chainsaw/form-drag dissipation model and the Manning's model for $\bar{U}$.
+stemming from both the Reynolds/form-drag dissipation model and the Manning's model for $\bar{U}$.
 For the purposes of this study, we are validating BIKER against equation 10 in the main text and so all uncertainties associated with the $\beta_1$ parameter are ignored and we only need to reflect the Manning's uncertainty in our specification of $\sigma_{k_{600}}^2$.
 Therefore, we take @hagemannBAMBayesianAMHGManning2017's estimated uncertainty from Manning's equation to infer streamflow from SWOT observations (0.25) and inflate it slightly to also account for the hydraulically-wide channel assumption and arrive at 0.30.
-This is the$\sigma_{k_{600}}$ used in this study.
+This is the $\sigma_{k_{600}}$ used in this study.
 
@@
 This means we must also account for uncertainty from the $\beta_1$ parameter.
-Assuming perfect measurements from are made by the SWOT satellite, the $k_{600}$ relative uncertainty is expressed for some set of hydraulic observations as equation S19.
+Assuming perfect, no-error measurements are made by the SWOT satellite, the $k_{600}$ relative uncertainty is expressed for some set of hydraulic observations as equation S20.
 We use the 166 hydraulically-wide measurements in our field-measured $k_{600}$ dataset and run 166 different Monte Carlo simulations (each 10,000 runs) to obtain 166 different $\sigma_k_{600}$ terms.
@@
 
-$$log_{k_{600}} \sim normal \left[ log(\beta_1), log(\sigma_{\beta_1}) \right] + \frac{1}{4}normal\left[log(\bar{U}), log(\sigma_{mannings})\right] \frac{7}{16}log(gS) + \frac{9}{16}log(H)  \mathbf{(S19)}$$
+$$log_{k_{600}} \sim normal \left[ log(\beta_1), log(\sigma_{\beta_1}) \right] + \frac{1}{4}normal\left[log(\bar{U}), log(\sigma_{mannings})\right] \frac{7}{16}log(gS) + \frac{9}{16}log(H)  \mathbf{(S20)}$$
 
@@
 |---------|-------------|--------------------------|
-|@ulsethDistinctAirWater2019|@ulsethDistinctAirWater2019||
-|@ulsethDistinctAirWater2019|@hallUseArgonMeasure2018||
-|@ulsethDistinctAirWater2019|@schelkerCO2EvasionSteep2016||
-|@ulsethDistinctAirWater2019|@mauriceInfluenceFlowBed2017||
-|@ulsethDistinctAirWater2019|@raymondScalingGasTransfer2012||
-|@churchillPredictionStreamReaeration1964|@churchillPredictionStreamReaeration1964||
-|@owensReaerationStudiesStreams1964|@owensReaerationStudiesStreams1964||
-<!-- |@raymondScalingGasTransfer2012|@melchingReaerationEquationsDerived| -->
-<!-- |@raymondScalingGasTransfer2012|@tsivoglouCharacterizationStreamReaeration1972| -->
-<!-- |@raymondScalingGasTransfer2012|Stroud Water Center| -->
-<!-- |@raymondScalingGasTransfer2012|@bernotInterregionalComparisonLanduse2010| -->
-<!-- |@raymondScalingGasTransfer2012|@mulhollandInterbiomeComparisonFactors2001| -->
-<!-- |@churchillPredictionStreamReaeration1964|@churchillPredictionStreamReaeration1964|| -->
-<!-- |@owensReaerationStudiesStreams1964|@owensReaerationStudiesStreams1964|| -->
+|@ulsethDistinctAirWater2019|@ulsethDistinctAirWater2019|
+|@ulsethDistinctAirWater2019|@hallUseArgonMeasure2018|
+|@ulsethDistinctAirWater2019|@schelkerCO2EvasionSteep2016|
+|@ulsethDistinctAirWater2019|@mauriceInfluenceFlowBed2017|
+|@ulsethDistinctAirWater2019|@raymondScalingGasTransfer2012|
+|@churchillPredictionStreamReaeration1964|@churchillPredictionStreamReaeration1964|
+|@owensReaerationStudiesStreams1964|@owensReaerationStudiesStreams1964|
 

### 2022-01-14 src/debugBIKER.R

@@ -1,0 +1,126 @@
+#packages--------------------------------------------------------------------
+library(ncdf4)
+library(dplyr)
+
+g <- 9.8
+uncertainity <- 0.30
+
+#load in rivers----------------------------------------------------------------
+files <- list.files('C:/Users/craig/Documents/OneDrive - University of Massachusetts/Ongoing Projects/RSK600/data/Frasson_etal_2021/IdealDataxxxxxx', pattern="*.nc", full.names = TRUE) #pepsi 2
+files2 <- list.files('C:/Users/craig/Documents/OneDrive - University of Massachusetts/Ongoing Projects/RSK600/data/Durand_etal_2016/xxxxxxxxxxxxxxxx', pattern="*.nc", full.names = TRUE) #pepsi 1
+files <- c(files, files2)
+
+#grab a dummy river
+currPepsi <- files[22]
+
+#extract river name-----------------------------------------------------------
+name <- substr(currPepsi, 40, nchar(currPepsi))
+name <- substr(name,1,nchar(name)-3)
+set.seed(12)
+data_in =nc_open(currPepsi)
+
+#extract reach-scale river hydraulics-----------------------------------------------------------
+W_obs=ncvar_get(data_in,'Reach_Timeseries/W')
+H_obs=ncvar_get(data_in,'Reach_Timeseries/H')
+S_obs = ncvar_get(data_in, 'Reach_Timeseries/S')
+area = ncvar_get(data_in, 'Reach_Timeseries/A')
+Q_obs=ncvar_get(data_in,'Reach_Timeseries/Q')
+priorQ <- ncvar_get(data_in, 'River_Info/QWBM')
+
+W_obs=W_obs[1:3,2:5]
+H_obs=H_obs[1:3,2:5]
+S_obs = S_obs[1:3,2:5]
+area = area[1:3,2:5]
+Q_obs= Q_obs[1:3,2:5]
+priorQ <- ncvar_get(data_in, 'River_Info/QWBM')
+
+#prep river hydraulics-----------------------------------------------------------
+S_obs[S_obs<=0]=NA
+S_obs[is.na(S_obs)] = 0.000017 #min obs SWOT slope Biancarma etal 2016
+W_obs[W_obs<0]=NA
+H_obs[H_obs<0]=NA
+Q_obs[Q_obs<0]=NA
+area[area<0]=NA
+
[…]

### 2022-01-14 src/runBIKER.R

@@ -97,7 +97,6 @@ run_BIKER <- function(currPepsi, errFlag) {
     D_obs <- area/W_obs #[m]
     V_obs <- Q_obs/area #[m/s]
     dA_obs <- calcdA_mat(W_obs,H_obs) #[m2]
-    #Rh_obs <- area / (W_obs + 2*D_obs)
 
     k_obs <- k600_model(D_obs, S_obs, V_obs) #k600 equation
     k_obs <- colMeans(k_obs, na.rm=T)

### 2022-01-14 src/swot_k_model_v3.R

@@ -35,14 +35,14 @@ data <- filter(data, is.finite(chan_width) ==1) %>%
   filter(chan_velocity > 0) %>%
   filter(chan_discharge > 0) %>%
   filter(chan_area > 0) %>%
-  filter(measured_rating_diff %in% c('Excellent', 'EXCL', 'GOOD', 'Good', 'Fair', 'FAIR'))
+  filter(measured_rating_diff %in% c('Excellent', 'EXCL', 'GOOD', 'Good'))#, 'Fair', 'FAIR'))
 
 #imperial to metric
 data$area <- data$chan_area * 0.092903 #ft2 to m2
 data$width <- data$chan_width*0.305 #m
 data$Vms <- data$chan_velocity*0.305 #m/s
 data$depth <- data$area / data$width
-data$slope <- NA #data$SLOPE
+data$slope <- NA 
 data$Qm3s <- data$chan_discharge * 0.0283 #ft3/s to m3/s
 
 #calculate bankfull width so we can throw out out-of-bank events
@@ -65,10 +65,12 @@ bankfullD = group_by(data, site_no) %>%
   summarise(bankful_depth = max(depth[temp]))
 data <- left_join(data, bankfullD, 'site_no')
 
-data <- filter(data, width <= bankful_width)
-data <- filter(data, depth <= bankful_depth)
+data <- filter(data, width < bankful_width)
+data <- filter(data, depth < bankful_depth)
 
 #######READ IN FIELD DATA COLLECTED FROM USGS REPORT-------------------
+  #Chuchill et al 1964
+  #Owens et al 1964
 usgs_data <- read.csv('data/additional_USGS_reaeration_data.csv')
 usgs_data$depth <- usgs_data$depth_ft * 0.3048 #ft to m
 usgs_data$width <- usgs_data$width_ft * 0.3048 #ft to m
@@ -100,11 +102,12 @@ data$ustar <- sqrt(g*data$Rh*data$slope) #friction/shear velocity [m/s]
 
 #######SOME FLAGS FOR SWOT-OBSERVABLE RIVERS AND WHEN RH=H-------------------------------
 data$flag_swot <- ifelse(data$width >= 100, 'SWOT', 'Small')
-data$flag_hydraulicWide <- ifelse(data$Rh/data$depth > 0.96, 'Rh=H', 'Rh=/=H') #set to 95% just to get idea of number of SWOT measurements that generally meet this condition
 
[…]

### 2022-01-21 manuscript_v7.Rmd
manuscript edits + related reruns

@@
 output:
-  word_document:
+  redoc::redoc:
+#output:
+#  word_document:
       pandoc_args:
@@
 Extensive research over the past two decades has shown that the global river network emits significant amounts of greenhouse gas via evasion.
-Despite much progress, there is still significant uncertainty in the temporal dynamics of gas exchange velocity (i.e.
-the speed of evasion) and thus carbon emissions to the atmosphere.
+Despite much progress, there is still large uncertainty in the temporal dynamics of gas exchange velocity and thus carbon emissions to the atmosphere.
 We propose that the NASA/CNES/USKA/CSA SWOT satellite, set to usher in a new age of hydrology research at the global scale, can provide similarly transformative insights to fluvial gas exchange modeling upon launch in 2022.
 Here, we present work inferring gas exchange velocity from future SWOT observations without in situ calibration.
-We exploit the unique geomorphology of SWOT-observable rivers to develop a physical model of gas exchange that is nearly entirely remotely sensible and explains `r round(modelsWIDE[modelsWIDE$name == 'Chainsaw-eD',]$r2,2)*100`% of variation in gas exchange.
+We exploit the unique geomorphology of SWOT-observable rivers to develop a physical model of gas exchange that is nearly entirely remotely sensible and explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2,2)*100`% of variation in gas exchange.
 We then couple this model with established Bayesian inference techniques to develop BIKER, or the 'Bayesian Inference of the $k_{600}$ Exchange Rate' algorithm.
-We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers and 763 discrete measurements of river gas exchange, yielding an algorithm that predicts this theoretical model solely from SWOT observations with a by-river median normalized RMSE of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$NRMSE), 2)*100`%.
-Like many fluvial Bayesian inference algorithms, BIKER is better at reproducing the temporal dynamics of gas exchange (median correlation coefficient of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$r), 2)`), rather than estimating the absolute rates of exchange.
+We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers and 166 discrete measurements of river gas exchange, yielding an algorithm that predicts this theoretical model solely from SWOT observations with a by-river median Kling-Gupta Efficiency of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$KGE), 2)`.
+Like many similar Bayesian remote sensing algorithms, BIKER is better at reproducing the temporal dynamics of gas exchange (median correlation coefficient of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$r), 2)`), than the absolute rates of exchange.
 Finally, we show BIKER robust to measurement errors implicit in the SWOT data.
-With SWOT set to launch in late 2022, we suggest that BIKER can be deployed to map the global-scale spatiotemporal dynamics of fluvial gas exchange in large rivers.
+With SWOT set to launch in late 2022, we suggest that BIKER will be useful in mapping the global-scale spatiotemporal dynamics of fluvial gas exchange in large rivers.
 
@@
 Following @colePlumbingGlobalCarbon2007, the global river network is recognized to emit substantial amounts of carbon to the atmosphere via evasion (gas exchange driven by a diffusion gradient and water turbulence), in addition to their long understood role of transporting carbon to the oceans.
-Current estimates of total carbon dioxide evasion ($FCO_2$) to the atmosphere from the global river network vary from 650-1800 Tg C/yr [@lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013] (*Liu etal in review- add number is paper if accepted in time*), with 167 Tg-C/yr coming from mountain streams alone [@horgbyUnexpectedLargeEvasion2019].
-Despite rivers' incredibly small percentage of the global land surface [0.`r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])`%- @raymondGlobalCarbonDioxide2013], this C flux is on par with the total oceanic $CO_2$ uptake rate [@gruberOceanicSinkAnthropogenic2019; @horgbyUnexpectedLargeEvasion2019] and the global forest carbon uptake rate [@panLargePersistentCarbon2011].
+Current estimates of total carbon dioxide evasion ($FCO_2$) to the atmosphere from the global river network vary from 650-2000 Tg C/yr [*Liu accepted*; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013], with 167 Tg-C/yr coming from mountain streams alone [@horgbyUnexpectedLargeEvasion2019].
+Despite rivers' incredibly small percentage of the global land surface [0.67%- *Liu accepted*], this C flux is on par with the total oceanic $CO_2$ uptake rate [@gruberOceanicSinkAnthropogenic2019; @horgbyUnexpectedLargeEvasion2019] and the global forest carbon uptake rate [@panLargePersistentCarbon2011].
 
@@
 
-To that end, the field has made significant progress in estimating $CO_2_{water}$, where we now have river-reach explicit modeling at monthly temporal resolutions using global hydrography datasets [@horgbyUnexpectedLargeEvasion2019; @brinkerhoffLakeMorphometryRiver2021; @saccardiImprovingPredictionsStream2021; *Liu etal in review*], but an equivalently sophisticated representation of $k_{CO_2}$ is still lacking.
+To that end, the field has made significant progress in estimating $CO_2_{water}$, where we now have river-reach explicit modeling at monthly temporal resolutions using global hydrography datasets [@horgbyUnexpectedLargeEvasion2019; @brinkerhoffLakeMorphometryRiver2021; @saccardiImprovingPredictionsStream2021; *Liu accepetd*], but an equivalently sophisticated representation of $k_{CO_2}$ is still lacking.
 As equations 1 and 2 dictate, $CO_2_{water}$ and $k_{CO_2}$ share the burden of calculating $FCO_2$ and therefore these next-generation models will have to contend with this discrepancy, which is directly dictated by the lack of direct measurements of global river hydraulics.
@@
 
-To answer this first question, we use the specific hydraulic geometry, or the fundamental geomoprhic relationships between streamflow and channel shape [@leopoldHydraulicGeometryStream1953] of large, SWOT-observable rivers to develop a process-based model for large-river gas exchange.
+To answer this first question, we use the hydraulic geometry [the fundamental geomorphic relationships between streamflow and channel shape- @leopoldHydraulicGeometryStream1953] of large, SWOT-observable rivers to develop a process-based model for large-river gas exchange.
 We then take these findings and explore question 2 by implementing the model within an RS of $k_{600}$ algorithm named BIKER ('Bayesian Inference of the $k_{600}$ Evasion Rate').
@@
 ## 2 Hydraulic geometry in large rivers and predicting fluvial gas exchange velocity
-To predict *k* in the very large rivers that SWOT will observe, we start from established process-based models for *k*, impose hydraulic assumptions that are valid in large rivers, and then obtain a suite of models we can empirically test.
+To predict *k* in the very large rivers that SWOT will observe, we start from established process-based models for *k*, impose hydraulic assumptions that are valid in SWOT-observable rivers, and then obtain a suite of models we empirically test.
 Following a description of the data used (section 2.1), we outline these established models (section 2.2) and then step through our hydraulic assumptions to arrive at equations that can be ingested by SWOT (section 2.2).
@@
 ### 2.1 Data
-Our gas exchange theory is developed using multiple datasets of directly measured $k_{600}$ and directly measured stream hydraulics collected from the literature.
-In total, we have collected 763 measurements of $k_{600}$, with 701 of these measurements previously published by @raymondScalingGasTransfer2012 and @ulsethDistinctAirWater2019.
-The remaining measurements were measured by @churchillPredictionStreamReaeration1964 and @owensReaerationStudiesStreams1964.
+We develop our gas exchange theory using multiple datasets of directly measured $k_{600}$ and directly measured stream hydraulics collected from the literature.
+In total, there are 763 measurements of $k_{600}$, with 701 of these measurements previously gathered by @raymondScalingGasTransfer2012 and @ulsethDistinctAirWater2019.
+The remaining measurements come from @churchillPredictionStreamReaeration1964 and @owensReaerationStudiesStreams1964.
 See Table S1 for a complete list of the studies that collected these measurements.
@@
 
-To qunatify how many SWOT-obsrvable rivers meet our hydraulic assumptions, we expand our set of stream hydraulics (but not $k_{600}$) using a previously published compilation of in situ hydraulic measurements [@brinkerhoffReconcilingAtaStationAtManyStations2019].
-That dataset contains over 530,000 unique measurements of river channel velocity, width, and discharge from across the continental United States, originally taken to calibrate United States Geological Survey (USGS) streamgauge rating curves and made public by the USGS.
+To quantify how many SWOT-observable rivers meet our hydraulic assumptions, we expand our set of stream hydraulics (but not $k_{600}$) using a previously published compilation of in situ hydraulic measurements [@brinkerhoffReconcilingAtaStationAtManyStations2019].
+That dataset contains over 530,000 unique measurements of river channel velocity, width, and discharge from across the continental United States, originally made to calibrate United States Geological Survey (USGS) streamgauge rating curves and made public by the USGS.
 
 ### 2.2 Process-based modeling of river *k*
-*k* should scale with near-surface turbulence in a turbulent flow [@hallGasExchangeStreams2020] and thus extensive field and laboratory experiments have converged on the 'small-eddy model' as derived by @lamontEddyCellModel1970 and empirically anticipated by @calderbankContinuousPhaseHeat1961.
-Based on classic surface-renewal theory, this model scales *k* via the smallest-scale (the 'Kolmogorov scale') turbulent eddies and has been repeatedly empirically validated in freshwater systems [e.g.
+*k* should scale with near-surface turbulence in turbulent streamflows [@hallGasExchangeStreams2020] and thus extensive field and laboratory experiments have converged on the 'small-eddy model' as derived by @lamontEddyCellModel1970 and empirically anticipated by @calderbankContinuousPhaseHeat1961.
+This model scales *k* via the smallest-scale turbulent eddies and has been repeatedly empirically validated in freshwater systems [e.g.
 @katulStructureFunctionModel2018; @lorkeUnifiedScalingRelation2006; @zappaEnvironmentalTurbulentMixing2007; @zappaVariationSurfaceTurbulence2003; @moogStreamReaerationNonuniform1999; @vachonRelationshipNearsurfaceTurbulence2010; @tokoroHighGastransferVelocity2008; @wangPhysicallyBasedScaling2021].
-This model is provided as equation 3, where $\epsilon$ is the dissipation rate of near-surface turbulence, $v$ is the kinematic viscosity, and *Sc* is the Schmidt number.
+The small-eddy model is provided as equation 3, where $\epsilon$ is the dissipation rate of near-surface turbulence, $v$ is the kinematic viscosity, and *Sc* is the Schmidt number.
 
 $$k\propto Sc^{-1/2}(v\epsilon)^{1/4} \mathbf{(3)}$$
-Further, some workers have noted that laboratory and field observations suggest open channel flows with small bed roughness (generally big rivers) do not exhibit homogeneous surface dissipation at their air-water interface [@moogAirWaterGasTransfer1999; @talkeTurbulentKineticEnergy2013].
+Some laboratory and field observations additionally suggest that open channel flows with small bed roughness do not exhibit homogeneous surface dissipation at their air-water interface [@moogAirWaterGasTransfer1999; @talkeTurbulentKineticEnergy2013].
 Given this observation, @moogAirWaterGasTransfer1999 proposed an extension to the small-eddy model, additionally scaling $k_{600}$ using a shear Reynold's number formulation.
-*HERE: Zappa's note that this is frequently done in air-ocean gas exchange studies*.
-This is equation 4 and is referred to as the 'Reynolds extension' model.
-The Reynolds model is hypothetically useful in low-turbulence flows where a relative lack of large-scale eddies effectively 'filter out' the number of small-eddies that actually reach the interface and lead to gas exchange [@talkeTurbulentKineticEnergy2013].
+This is equation 4 and is referred to here as the 'Reynolds extension' model.
+The Reynolds model is hypothetically useful in low-turbulence flows where a relative lack of large-scale eddies effectively 'filter out' the number of small-eddies that actually reach the interface and initiate gas exchange [@talkeTurbulentKineticEnergy2013].
+Scaling *k* via a shear Reynold's formulation is sometimes done to parameterize wave-breaking gas exchange models in the open ocean  [@brumerWaveRelatedReynoldsNumber2017; @zhaoDependenceWhitecapCoverage2001; @zhaoEffectWindWaves2003], though is infrequently done in rivers.
 In the context of BIKER, we chose to test this model because large, SWOT-observable rivers are generally the smoothest, least-turbulent flows along the stream-to-ocean continuum.
-Further, to our knowledge this Reynolds extension model has never been empirically tested in predicting river *k*, aside from confirming that large-scale eddies differentially move turbulence to the surface in a hydraulically-wide river [@talkeTurbulentKineticEnergy2013].
+Further, to our knowledge this Reynolds extension model has never been empirically tested in predicting river *k*, aside from confirming that large-scale eddies differentially move turbulence to the surface in a large river [@talkeTurbulentKineticEnergy2013].
 
 $$k \propto Sc^{-1/2}(v\epsilon)^{1/4}(\frac{U_*H}{v})^{3/8} \mathbf{(4)}$$
-Finally, both equations rely on measurements of $\epsilon$, of which measuring directly is non-trivial.
+Equations 3 and 4 both rely on $\epsilon$, of which measuring directly is non-trivial.
 When working at large scales, a commonly used $\epsilon$ model assumes that all turbulence is generated at the bed and transported to the air-water interface via the log-law-of-the-wall such that $\epsilon = \epsilon_S = \frac{U_*^3}{H}$ [@lorkeUnifiedScalingRelation2006; @nezuTurbulenceOpenChannel1993].
@@
 ### 2.3 Imposing large-river hydraulic geometry on *k* models
-Rivers and streams change predictably along their longitudinal profile from headwater to ocean, and we can exploit the distinct hydraulic geometry of large rivers at the end of this continumn to estimate *k* in SWOT-observable systems.
+Given the theoretical context privded in Section 2.2, we know turn to SWOT-observable systems specifically.
+Rivers and streams change predictably along their longitudinal profile from headwater to ocean, and we can exploit the hydraulic geometry of large rivers at the end of this continumn to estimate *k* in SWOT-observable systems.
 In general, as river size increases, channels become more rectangular, their shapes elongate (becoming wider quicker than they become deeper) and their hydraulic radii begin to approximate their mean flow depth, i.e.
 $R_h \approx H$.
-This is a common assumption in hydraulic and geomorphic modeling of large rivers, where the average $\frac{R_h}{H}$ ratio across our `r HG_swot$n` SWOT-observable river flows is `r HG_swot$mean * 100`% (see Text S1 for how we built this dataset).
+This is a common assumption in hydraulic and geomorphic modeling of large rivers, where the average SWOT-observable river flow has a $\frac{R_h}{H}$ ratio of `r HG_swot$mean * 100`% (n = `r HG_swot$n`; see Text S1 for how we built this dataset).
 We refer to these flows as 'hydraulically-wide'.
 
-Assuming that all SWOT-observable rivers are hydraulically-wide, we test the performance of four models for predicting *k* in hydraulically-wide channels: the four unique combinations of equations 3-4 and the two $\epsilon$ models already outlined.
-While the complete model derivations for the small-eddy models (section 2.3.1) and the chainsaw models (section 2.3.2) are provided in Text S2, the broad goal was to reduce the equations down to their fundamental parameters, identifying which terms are SWOT observable and limiting the number of terms not directly measurable via SWOT.
+Assuming that all SWOT-observable rivers are hydraulically-wide, we test the performance of four models for predicting *k* in hydraulically-wide channels: the four unique combinations of equations 3-4 and the two $\epsilon$ models outlined above.
+While the complete model derivations for the small-eddy models (section 2.3.1) and the reynolds-extension models (section 2.3.2) are provided in Text S2, the final equations are provided in Table 1.
+The overall goal was to reduce the equations down to their fundamental parameters, identifying which terms are SWOT observable and limiting the number of terms not directly measurable via SWOT.
+
+*Table 1: Four physically-based models for $k_{600}$ each derived under a different set of mass transport assumptions but all valid in hydraulically-wide rivers.
+Their derivations are outlined in sections 3.2.1 and 3.2.2 and their empirical tests are shown in section 3.3.*
+
+|**$k_{600}$ model**|**$\epsilon$ model**|**Final equation**|**Empirical test**|**Empirical fit ($r^2$)**|
+|-----------|--------------------|------------------|-----------------|----------------------------|
+|Small-eddy|Log-law-of-the-wall|$k_{600}=\alpha(gS)^{3/8}H^{1/8}$|Figure 3a|`r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eS',]$r2, 2)`|
+|Small-eddy|Form-drag dissipation|$k_{600}=\alpha_1(gS\bar{U})^{1/4}$|Figure 3c|`r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eD',]$r2, 2)`|
+|Reynolds-extension|Log-law-of-the-wall|$k_{600}=\beta(gS)^{9/16}H^{11/16}$|Figure 3b|`r round(modelsWIDE[modelsWIDE$name == 'reynolds-eS',]$r2, 2)`|
+|Reynolds-extension|Form-drag dissipation|$k_{600}=\beta_1(gS)^{7/16}\bar{U}^{1/4}H^{9/16}$|Figure 3d|`r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2, 2)`|
 
@@
 We derive equations starting from the small-eddy model (equation 3) for hydraulically-wide channels.
-If we assume $\epsilon=\epsilon_S$ and $R_H = H$ in equation 2, and finally solve for $k_{600}$ via statistical parameter $\alpha$, we eventually arrive at equation 5.
+If we assume $\epsilon=\epsilon_S$ and $R_H = H$ in equation 2, and solve for $k_{600}$ via statistical parameter $\alpha$, we eventually arrive at equation 5.
 If we do the same but assume $\epsilon=\epsilon_D$ and statistical parameter $\alpha_1$, we arrive at equation 6.
@@
 The result is identical to the hydraulically-wide small-eddy model (equation 5), except for the exponents which are both steeper.
-Equation 8 reflects the same derivation, but when $\epsilon=\epsilon_D$ and statistical coefficient $\beta_1$.
+Equation 8 reflects the same derivation, but with $\epsilon=\epsilon_D$ and statistical coefficient $\beta_1$.
 
@@
 We now have four different models for $k_{600}$ in hydraulically-wide rivers, each with a different set of assumptions and parameters but all defined as functions of the water surface slope and other hydraulic terms that are not directly measurable via SWOT.
-All four models are summarized in Table 1.
-
-*Table 1: Four physically-based models for $k_{600}$ each derived under a different set of mass transport assumptions but all valid in hydraulically-wide rivers.
-Their derivations are outlined in sections 3.2.1 and 3.2.2 and their empirical tests are shown in section 3.3.*
-
-|**Description**|**Equation**|**Empirical test**|**Empirical fit via $r^2$**|
-|---------------|------------|------------------|-----------------|
-|Small-eddy, log-law-of-the-wall|$k_{600}=\alpha(gs)^{3/8}H^{1/8}$|Figure 3a|`r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eS',]$r2, 2)`|
-|Small-eddy, form-drag dissipation|$k_{600}=\alpha_1(gS\bar{U})^{1/4}$|Figure 3c|`r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eD',]$r2, 2)`|
-|Chainsaw, log-law-of-the-wall|$k_{600}=\beta(gs)^{9/16}H^{11/16}$|Figure 3b|`r round(modelsWIDE[modelsWIDE$name == 'reynolds-eS',]$r2, 2)`|
-|Chainsaw, form-drag dissipation|$k_{600}=\beta_1(gs)^{7/16}\bar{U}^{1/4}H^{9/16}$||Figure 3d|`r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2, 2)`|
+Again, all four models are summarized in Table 1.
 
@@
 With equations 5-8 derived, we now empirically test their strength of fit for hydraulically-wide river flows.
-We validate on the dataset of direct measurements of $k_{600}$ and stream hydraulics, after filtering for measurements made in hydraulically-wide channels (n= `r n`).
-As equations 5-8 necessitate that this assumption is true, we defined 'hydraulically-wide' as flows whose hydraulic radius was within 1% of their mean flow depth.
+We validate on the dataset of in situ measurements of $k_{600}$, after filtering for measurements made in hydraulically-wide channels (n= `r n`), which was operationally defined as flows whose hydraulic radius was within 1% of their mean flow depth.
 All told, this amounts to `r n` direct measurements of hydraulically-wide $k_{600}$ and stream hydraulics to test with.
 Models were assessed via the coefficient of determination ($r^2$) and results are plotted in Figure 3.
-Note that Figure 3 axes are plotted in logarithmic space just for visualization: model fit and validation (via $r^2$) were calculated in linear space as their models dictate.
+Note that Figure 3 axes are plotted in logarithmic space only for visualization: model fit and validation (via $r^2$) were calculated in linear space as their models dictate.
 
 ![Figure 2: Empirical testing of four models for predicting gas exchange velocity in hydraulically-wide channels: (a) Small-eddy model with log-law-of-the-wall dissipation.
-(b) Chainsaw model with log-law-of-the-wall dissipation.
+(b) Reynolds-extension model with log-law-of-the-wall dissipation.
 (c) Small-eddy model with form-drag dissipation.
-(d) Chainsaw model with form-drag dissipation.
-Note that axes are plotted in logarithmic space just for visualization: model fit and validation were calculated in linear space.](cache/k600_theory/k600Plot.jpg)
+(d) Reynolds-extension model with form-drag dissipation.
+Note that axes are plotted in logarithmic space just for visualization: model fit and validation were calculated in linear space.](cache/k600_theory/k600Plot.jpg){width=1000px}
 
@@
 However, when this model is expanded to retard the dissipation rate in the smoothest flows via the Reynolds extension model [equations 6 and 7- @moogAirWaterGasTransfer1999], the dynamics are more accurately captured (Figures 3b, 3d) and $r^2$ scores go up appreciably (`r round(modelsWIDE[modelsWIDE$name == 'reynolds-eS',]$r2, 2)` and `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2, 2)` for the log-law-of-the-wall and form-drag dissipation, respectively).
-Modeling dissipation via total form-drag rather than just via bed shear and the log-law-of-the-wall improves model performance.
+Modeling dissipation via total form-drag rather than just via bed shear improves model performance regardless of $k_{600}$ equation used.
 This is likely because the majority of dissipation occurring in rivers comes not just from the streambed, but from bulk form drag via channel walls, bars, and meanders [@raymondScalingGasTransfer2012; @moogStreamReaerationNonuniform1999; @wangPhysicallyBasedScaling2021; @ulsethDistinctAirWater2019].
 
-The success of the Reynolds extensions model in hydraulically-wide channels, and in particular when coupled with the $\epsilon=\epsilon_D$ model (equation 8, Figure 3d), provides us with a strong physical-model for gas evasion.
+The success of the Reynolds-extension models in hydraulically-wide channels, and in particular when coupled with form-drag dissipation (equation 8, Figure 3d), provides us with a strong physical-model for gas evasion.
 The river hydraulics terms in equation 8 can either be directly measured or reasonably inferred from SWOT measurements, effectively opening the door for remotely sensing the gas exchange velocity.
@@
 @andreadisConstrainingAssimilationSWOT2020; @brinkerhoffConstrainingRemoteRiver2020; @durandEstimatingReachaveragedDischarge2014; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014a; @garamboisVariationalEstimationEffective2020; @brissetAssimilationAltimetricData2018; @hagemannBAMBayesianAMHGManning2017; @larnierRiverDischargeBathymetry2020a; @oubanasDischargeEstimationUngauged2018].
-For BIKER, we follow the work developed by @durandEstimatingReachaveragedDischarge2014, @hagemannBAMBayesianAMHGManning2017, and @brinkerhoffConstrainingRemoteRiver2020 to infer mean flow velocity and depth from SWOT observations using a modified form of the Manning's equation for mean flow velocity.
+For BIKER, we follow the work developed by @hagemannBAMBayesianAMHGManning2017, @brinkerhoffConstrainingRemoteRiver2020, and @durandEstimatingReachaveragedDischarge2014 to infer $k_{600}$, channel bathymetry, and roughness from SWOT observations using a modified form of Manning's equation for mean flow velocity.
 Following a description of the data used (section 3.1), we detail algorithm development and experimental design (section 3.2) and then we present the validation results (section 3.3).
@@
 They used standard hydraulic models forced with known inflows and measured bathymetry to model the hydraulic response of the rivers and then further prepared the data to mimic SWOT by adding realistic radar errors and sampling along the satellite’s future ground track.
-@frassonExploringFactorsControlling2021 also developed a sophisticated error model that mimics the effects of the atmosphere, topography, and water surface geometry on SWOT measurements (16/47 rivers) to add realistic errors that depend on the time and place of observation.
+@frassonExploringFactorsControlling2021 also developed a sophisticated error model that mimics the effects of the atmosphere, topography, and water surface geometry on SWOT measurements (16/47 rivers) to add realistic radar layover errors that depend on the time and place of observation.
 The resulting data are the most realistic synthetic SWOT river data in the literature.
@@
 
-We use 26 bi-weekly dissolved $CO_2$ samples collected by @beaulieuControlsGasTransfer2012 at one location in the Ohio River for one calender year from 2008-2009 (Figure S1) for the $CO_2$ evasion and carbon efflux calculations (section 3.1.3).
+For the $CO_2$ evasion and carbon efflux calculations (section 3.1.3), we use 26 bi-weekly dissolved $CO_2$ samples collected by @beaulieuControlsGasTransfer2012 at one location in the Ohio River for one calender year from 2008-2009 (Figure S1).
 Note that this $CO_2$ data is for the Ohio River only but was applied to all rivers (which includes multiple sections of the Ohio River).
@@
 $f(x|\Theta)$).
-This is similar to the McFLI logic used in some SWOT RSQ algorithms [@gleasonTrackingRiverFlows2017].
+This is similar to the 'McFLI' ('Mass-Conserved Flow Law Inversion') logic used in some SWOT RSQ algorithms [@gleasonTrackingRiverFlows2017].
 To start, we write $k_{600}$ as a function of SWOT-observables $W$ and $S$.
-This algebra is carried out using equation 8, the fitted value for $\beta_1$ from Figure 3d (`r round(modelsWIDE[modelsWIDE$name == 'Chainsaw-eD',]$coef,2)`), and Manning's equation for mean flow velocity ($\bar{U}=\frac{1}{n}R_h^{2/3}S^{1/2}$).
-Following section 2.3, we also assume that the channel is hydraulically-wide ($R_h = H =\frac{A}{W}$).
-To leverage additional SWOT data, the wetted channel area *A* is further split into the SWOT-observable portion $dA$ and the unobservable portion $A_0$ following @durandEstimatingReachaveragedDischarge2014.
-*dA* is readily approximated using SWOT measurements if assuming a rectangular river channel ($dA \approx \delta H_e*W$), which is again generally valid in a hydraulically-wide channel.
+This algebra is carried out using equation 8, the fitted value for $\beta_1$ from Figure 3d (`r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$coef,2)`), and Manning's equation for mean flow velocity ($\bar{U}=\frac{1}{n}R_h^{2/3}S^{1/2}$).
+Following section 2.3, we continue to assume that the channel is hydraulically-wide ($R_h = H =\frac{A}{W}$).
+To leverage additional SWOT data, the wetted channel area *A* is further split into the the SWOT-unobservable portion $A_0$ and SWOT-observable portion $dA$ following @durandEstimatingReachaveragedDischarge2014 and @hagemannBAMBayesianAMHGManning2017 where $dA_{it} = \sum_{t':W_{t'}\le W_t}{W_{t'} \delta H_{t'}}$ for cross-section *i* and timestep *t*.
 
-All of this algebra simplifies to equation 10, after accounting for time-varying *t* terms and space-varying *i* terms within the reach.
-Conveniently, both $k_{600}$ is measured (and predicted by equation 8) at the reach-scale, so we can reduce it to a reach-averaged, time-varying quantity.
-This lowers the number of parameters BIKER must infer and makes the problem better constrained.
+All of this algebra simplifies to equation 10.
+Conveniently, $k_{600}$ as measured by a tracer addition to a stream is inherently a reach-scale quantity.
+Therefore, equations 8 and 10 both yield a reach-scale $k_{600}$ (i.e.
+$k_{600_i}=k_{600}\forall i$).
+This lowers the number of parameters BIKER must infer and makes the problem much better constrained.
 
@@
 Next, equation 10 is written as a Bayesian sampling model, in which all of the SWOT observations are sampled from the unknown model parameters ($A_0_i$, $k_{600_t}$, and $n_i$).
-This is equation 11 after describing them as log-normal distributions.
+This is equation 11 after describing everything as normal distributions of the log-transformed terms.
 $\sigma_{k_{600}}^2$ refers to the total uncertainty implicit in equation 10.
-This uncertainty is due to parameter uncertainty from equation 8, Manning's equation, and the rectangular channel assumption.
+This uncertainty arises from parameter uncertainty in equation 8, Manning's equation, and the rectangular channel assumption.
 
@@
 Equations 9 and 11 also necessitate that we specify prior distributions for the parameters $A_0_i$, $k_{600_t}$, and $n_i$.
-Prior distributions, described by their hyperparameters, formalize the a priori estimates (and uncertainties) for the non-remotely-sensed terms.
+Prior distributions, described by their hyperparameters, formalize the a priori estimates and uncertainties for the non-remotely-sensed terms.
 More intuitively, BIKER priors represent our 'prior river knowledge' of what $A_0_i$, $k_{600_t}$, and $n_i$ probably are for some river since they cannot be directly remotely sensed, but educated guesses can be made using the SWOT data available to us.
-This is equivalent to the 'empirical Bayes method'.
+This is equivalent to the 'empirical Bayes approach' to Bayesian inference [cite textbook].
 Our goal in prior specification was to rely on absolutely no in situ information such that we could run this method on any river on Earth solely using SWOT observations.
-In theory, more informed priors via various a priori information about a specific river would improve BIKER performance, but here we chose to test only the fully generalized algorithm.
-Therefore, the validation presented here is a 'worst-case scenario', wherein BIKER performance will improve with better prior information on the river.
+In theory, more informed priors via various a priori information about a specific river would improve BIKER performance, but here we chose to test the fully generalized algorithm.
+Therefore, the validation presented here is a 'worst-case scenario', wherein BIKER performance should improve with better prior information on the river.
 In that context, we used a variation of the prior specification method developed by @brinkerhoffConstrainingRemoteRiver2020, who developed geomorphic 'river types' with distinct prior hyperparameter sets for $A_0_i$ and $n_i$.
@@
 
-With the sampling model defined ($f(x|\Theta)$ = equation 11) and hyperparameters $p(x$) described (Texts S3 and S4), a joint posterior distribution conditional on the SWOT observations ($p(\Theta|x) \propto f(x|\Theta)p(x)$) is specified.
+With the sampling model defined ($f(x|\Theta)$ = equation 11) and hyperparameters $p(x$) described (Texts S3 and S4), a joint posterior distribution conditional on the SWOT observations ($p(\Theta|x) \propto f(x|\Theta)p(x)$) is therefore also specified.
 To approximate this distribution, we use a Markov Chain Monte Carlo (MCMC) algorithm implemented using the Stan probabilistic programming language.
@@
 Thus, we take the model outlined in equation 7 and Figure 3d and use that to calculate the observed $k_{600_t}$ that BIKER is validated against (equation 12).
-Here, reach-averaged flow velocity is directly calculated as discharge divided by channel area.
+Here, reach-averaged flow velocity is calculated via mass conservation as discharge divided by channel area.
 This directly explores BIKER's ability to infer observed $A_0_i$, $k_{600_t}$, and $n_i$ simultaneously from $W_{i,t}$ and $H_{i,t}$ alone and was deemed acceptable given that the $k_{600}$ model (equation 7) has already been successfully validated in Figure 3d.
@@
 It also means that, for a fair validation scheme, the parameter specification for $\sigma^2_{k_{600}}$ in equation 11 should not include the uncertainty from equation 8.
-So for this validation only, $\sigma^2_{k_{600}}$ is set to 0.30 (natural-log space) to reflect uncertainty from Manning's equation and assuming a rectangular channel (Text S4).
-However, if BIKER is run on real SWOT observations after the satellite launches, $\sigma^2_{k_{600}}$ should reflect the total uncertainty, which is described in Text S4.
+So for this validation only, $\sigma^2_{k_{600}}$ is set to 0.30 (natural-log space).
+However, if BIKER is run on real SWOT observations after the satellite launches, $\sigma^2_{k_{600}}$ should reflect the total uncertainty, which is specified in Text S4 via Monte Carlo simulations as 1.1.
 
@@
 Because we are most interested in BIKER's ability to correctly infer temporal dynamics in $k_{600}$, and each river's priors are described specifically for that river, we validate by river.
-These metrics are defined in Table S2 and consist of the correlation coefficient *r* to quantify accuracy of BIKER's temporal dynamics, the root mean square error normalized by the observed mean (NRMSE) and prediction bias normalized by the observed mean (rBIAS) to assess bias, and the Kling-Gupta Efficiency (KGE).
+These metrics are defined in Table S2 and consist of the following:
+- Correlation coefficient *r* to quantify accuracy of BIKER's temporal dynamics
+- Root mean square error normalized by the observed mean (NRMSE) and prediction bias normalized by the observed mean (rBIAS) to assess bias
+- Kling-Gupta Efficiency (KGE).
 KGE is frequently used to assess streamflow prediction and simultaneously assesses accuracy in both bias and dynamics.
@@
 It is one thing to accurately model the temporal dynamics of $k_{600}$, but researchers are often most interested in the actual gas fluxes from rivers and ultimately the carbon emitted from river to atmosphere.
-Therefore, we also explore BIKER's ability to reproduce $FCO_2$ (equation 1) and the bulk carbon efflux from these `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers.
+Therefore, we also explore BIKER's ability to reproduce $FCO_2$ (equation 1) and the annual carbon emission rate from these `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers.
 To do this, we pair the 26 biweekly $CO_2$ and water temperature samples from @beaulieuControlsGasTransfer2012 (section 2, figure S1) with every 14th set of daily SWOT observations (as the $CO_2$ data is bi-weekly).
-Conveniently, this is very close to the nominal average sampling resolution for SWOT (11 days).
-Not all of the SWOT rivers have observations for a full year, and when simulation dates are not available they are assumed to start on January 1st.
+Not all of the SWOT rivers have at least 14 observations and those rivers were removed for this experiment.
+Similarly, not all have observations for a full year, and when simulation dates are not available they are assumed to start on January 1st.
 We then calculate $FCO_2$ using equation 1 and assuming atmospheric $CO_{2_{air}}$ is 390 uatm.
 Equation 13 is used to obtain a temperature and $CO_2$ specific gas exchange velocity from BIKER's output, where the Schmidt number *Sc* was estimated following @raymondScalingGasTransfer2012.
-We then estimate median and mean bulk carbon effluxes by multiplying the median or mean $FCO_2$ value across all rivers and timesteps (in Tg-C/yr) by the total average surface area of all 47 rivers.
+We then estimate a yearly average carbon emission rate (via $CO_2$ evasion) by applying each river's mean $FCO_2$ over the river's surface area and then summing all of these emission rates to obtain an emissions rate in carbon per year across all rivers.
 
 $$k_{CO2, temp}=k_{600}*(Sc/600)^{-1/2} \mathbf{(13)}$$
-Finally, we also compare BIKER's outputs with established in situ methods, whose specific details and parameters are provided in Table S3.
-Because streamflow measurements and models are vastly more frequent than measurements of river channel geometry, these in situ models are used to predict reach-averaged flow depth and flow velocity as functions of streamflow.
+Finally, we also compare BIKER's outputs with established in situ techniques, whose specific details and parameters are provided in Table S3.
+Streamflow is more readily modeled at the global scale than river channel geometry, and so these in situ models are used to predict $\bar{U}$ and *H* as functions of streamflow.
 Their outputs are in turn used to estimate *k* in upscaling workflows.
@@
 This allows us to assess whether BIKER's estimates (wholly ungauged) are comparable to gauged methods (all three HG models).
-Of the three models tested here, one is trained on the @brinkerhoffReconcilingAtaStationAtManyStations2019 dataset, and two are previously published models [@raymondGlobalCarbonDioxide2013; @raymondScalingGasTransfer2012].
+Of the three models tested here, 'Brinkerhoff2019' is trained on the dataset described in section 2.1, and two are previously published models [@raymondGlobalCarbonDioxide2013; @raymondScalingGasTransfer2012].
 
@@
 Figure 4 plots all `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` timeseries plots of predicted and observed $k_{600}$, again assuming no SWOT measurement error.
-Note that the y axis is normalized by maximum observed values to compare rivers.
-In general, temporal dynamics of $k_{600}$ are reproduced quite well by BIKER, however there is often positive bias in the estimates (e.g.
-Olentangy River).
-Some rivers are predicted nearly perfectly (e.g.
-Brahmaputra, Ohio Section 3, Wabash) while others perform more poorly (e.g.
-San Joaquin, Stanislaus), though again much of this error stems from errors in predicting the magnitude of $k_{600}$ rather than the temporal dynamics.
-Rivers where the temporal dynamics are more poorly inferred tend to have shorter timeseries (e.g.
-Merced River, Tuolumne River).
-Next, we explicitly quantify each rivers' performance.
+Note that the y axis is normalized by maximum observed values to compare across rivers.
+In general, the temporal dynamics of $k_{600}$ are reproduced quite well by BIKER, with the highs and lows of evasion correctly inferred by BIKER.
+Notably, there is sometimes positive bias in the estimates (e.g.
+Seine River).
+Some rivers are inferred nearly perfectly (e.g.
+Kanawha, Ohio Section 2, Wabash rivers) while others perform more poorly (e.g.
+Ash Slough, Ohio), though again much of this error stems from errors in inferring the magnitude of $k_{600}$ rather than its temporal dynamics.
+For some rivers, (e.g.
+Mississippi Intermediate), the first timestep is *THIS MIGHT BE DUE TO TO CERTAINTY OF THE PRIOR ETHIER ON K600 OR THE LIKELIHOOD???/*
 
@@
 River timeseries plots of $k_{600}$.
-Orange is calculated via observed hydraulics, while green is BIKER.
-The green ribbon indicates the 95% CIs for the predicted values.
+Orange is calculated via observed hydraulics, while green is BIKER and uses SWOT measurements as its sole input.
+The green ribbon indicates the 95% CIs for the BIKER predicted values.
 Y axis is normalized by maximum observed values to compare visually.](cache/validation/timeseries_noerr.jpg)
@@
 We calculate performance metrics following section 3.2.2 and Table S2.
-These are presented in Figure 5a, which plots validation metrics calculated for each river with and without SWOT measurement error (green and purple boxplots, respectively).
-SWOT measurement errors negligibly influence BIKER's performance across all four error metrics (Figure 5a), though caution should be used in over-interpreting boxplots with a sample size of only `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])`.
-Given Figure 5a, we deem that SWOT measurement error does not exert a significant influence on BIKER and so the results presented for the rest of the manuscript assume no measurement error in order to use all `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers.
+These are presented in Figure 5a, which plots the scores for rivers with and without SWOT measurement error (green and purple boxplots, respectively).
+For the no-error scenario, median river $KGE$ is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$KGE, na.rm=T),2)` and median river *r* is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$r, na.rm=T),2)`.
+This indicates very strong inference of each river's temporal $k_{600}$ dynamics.
+These scores are excellent given that absolutely no in situ information is being used to predict $k_{600}$.
+NRMSE has a median score of `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$NRMSE, na.rm=T),2)`, which is reasonable for a completely ungauged method.
+These are comparable to the best NRMSE scores achieved by recent versions of the SWOT RSQ algorithms [@frassonExploringFactorsControlling2021].
+Median rBIAS is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$rBIAS, na.rm=T),2)`, though the shape of the rBIAS boxplot in Figure 5a highlights a large portion of rivers which have a positive bias (Figure 4 also confirms this).
+Taken in aggregate, Figures 4-5 indicate that BIKER is quite good at capturing temporal dynamics in $k_600$, however there is sometimes positive bias in its estimates.
 
@@
 See Table S2 for metric definitions.
-b: *r* versus the number of SWOT observations used in the Bayesian inversion.
-c) *r* versus the mean river width.
-As expected, performance improves with more data.
 Dashed lines denote scores of 0, 0.50, and 1.
-CIs.](cache/validation/validation_by_river.jpg)
+CIs.
+b-c: Empirical cummulative density distributions of the *KGE* (b) and *r* (c) scores for both error and no-error rivers.](cache/validation/validation_by_river.jpg){width=1000px}
 
-Median river-specific $KGE$ is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$KGE, na.rm=T),2)` and median river-specific *r* is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$r, na.rm=T),2)`.
-This indicates very strong inference of each river's temporal $k_{600}$ dynamics.
-These scores are excellent given that absolutely no in situ information is being used to predict $k_{600}$.
-NRMSE has a median score of `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$NRMSE, na.rm=T),2)`, which is reasonable for a completely ungauged method.
-These are comparable to the best NRMSE scores achieved by recent versions of the SWOT RSQ algorithms [@frassonExploringFactorsControlling2021].
-Median rBIAS is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$rBIAS, na.rm=T),2)`, which is again excellent.
-However, the shape of the rBIAS boxplot in Figure 3a highlights a large portion of rivers which have a positive bias (Figure 4 also confirms this).
-Taken in aggregate, Figures 4-5 indicate that BIKER is quite good at capturing temporal dynamics in $k_600$, however there is often positive bias in its estimates.
+Figure 5 also compares BIKER results under a 'no-error' and 'error' scenario: the green and purple respectively in Figure 5a-c, where b-c plot the empirical cummulative density distributions (eCDFs) for *KGE* and *r* scores.
+There appears to be neglibile differences between results in Figures 5a-c, suggesting that SWOT measurement errors negligibly influence BIKER's performance across all four error metrics (Figure 5a).
+In fact, for some metrics, the error scenario appears to improve results.
+For example, median KGE without error is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$KGE),2)` while with-error median KGE is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 1,]$KGE),2)`.
+However, caution should be used here as these purple boxplots are composed of only `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` of the rivers in the green boxplots and many of the most poorly performing rivers are not in this set.
+The eCDFs more clearly indicate a marginal difference in performance when measurement errors are used to corrupt the BIKER inputs (with nearly identical performance when BIKER performs well for either metric).
+Given these results, we deem that SWOT measurement error does not exert a significant influence on BIKER performance and so the results presented for the rest of the manuscript assume no measurement error in order to use all `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers.
 
-Figure 5b highlights one benefit of using Bayesian inference to estimate $k_{600}$: because the posterior is conditional on the SWOT observations, performance should improve with more data.
-Figure 6b plots by-river *r* scores versus the number of SWOT observations.
-While performance does not directly increase from having more data, very poor performances disappear as the amount of available data increases.
-The vast majority of the extremely poorly performing rivers are limited to those with only 12 days of SWOT observations.
-This bodes well for future SWOT implementation, as SWOT has a nominal life-cycle of three years.
-A similar relationship also exists between mean observed river width and *r* (Figure 5c), where wider rivers have better BIKER performance.
-However, it is impossible to say whether this is a genuine trend or an artifact of the specific rivers being tested, where the larger rivers often have longer timeseries.
+Given the promising results in Figure 5a and 5c, we further explore BIKER's ability to infer temporal trends in $k_{600}$.
+Figure 6a plots a histogram of the slopes of power-law regressions between BIKER-predicted $k_{600}$ and observed $k_{600}$.
+A slope of 1 indicates that BIKER correctly infers the daily flucuations in gas exchange, while a slope well above or below 1 indicates incorrect inference of temporal dynamics in $k_{600}$.
+Figure 6a shows that the majority of rivers' slopes approximate 1 (mean slope: `r round(median(results_dynamics$Slope),2)`).
 
-Given the promising results *r* results in Figure 5a, we further explored BIKER's ability to predict temporal trends in $k_{600}$.
-Figure 6a plots the regression lines between BIKER $k_{600}$ and observed $k_{600}$ for each river, making clear that some rivers have notable bias (regression lines sitting above and below the 1:1 dashed blue line).However regardless of coefficient error, nearly all rivers scale at a slope of approximately 1 (dashed blue line), indicating that temporal trends are strongly captured.
-This is explicitly confirmed in Figure 6b, where we plot a histogram of the slopes of Figure 6a's regression lines, showing that vast majority of by-river regression slopes approximate 1 (median slope: `r round(median(results_dynamics$Slope),2)`).
+![Figure 6: a: Histogram of the slopes of power-law regressions between BIKER estimates and observed $k_{600}$.
+Slopes equal 1 indicate correct inference of the temporal trends in daily $k_{600} (regardless of bias in the estimates).
+b: .](cache/validation/validation_temporal.jpg){width=1000px}
 
-![Figure 6: a: Temporal scaling of BIKER predictions for the 47 rivers, compared to observed k600 and plotted as regression lines to aid in visualization.
-b: Histogram of the slopes of these by-river regression lines.
-Slopes equal 1 indicate perfect predicition of the temporal trends in daily k600 (regardless of bias in the estimates).](cache/validation/dynamics.jpg)
+Finally, we explore whether BIKER improves upon the prior with respect to temporal variance in $k_{600}$, here defined using the coefficient of variation ($CV=\frac{SD}{mean}$).
+Figure 6b plots eCDFs of observed, prior, and posterior $k_{600}$ *CV*, strongly showing that BIKER improves upon the prior information available to us about $k_{600}$, with posterior $k_{600}$ CV almost entirely overlapping the observed eCDF and sufficiently 'closing the gap' between the prior and observed eCDFs.
+It must be stressed that this validation was performed using a relatively uninformed prior on $k_{600}$ (Text S3) and that this gap may not be as large in many settings.
+However, Figure 6b certainly confirms that Bayesian technqiues can be used to infer temporal variation in $k_{600}$ when in situ work is not feasible, effectively reconciling bias between our prior knowledge of $k_{600}$ temporal variation and the observed variation.
 
 #### 3.3.2 Carbon efflux
-Next, we compare BIKER's estimates of carbon fluxes against established in situ techniques.
-It is important to remember that only BIKER is completely globally implementable, while the other three models presented necessarily rely on having a streamflow record readily available.
+We now compare BIKER's estimates of carbon fluxes against established in situ techniques.
+It is important to remember that only BIKER is completely globally implementable, while the other three models necessarily rely on having a streamflow record readily available.
 Therefore, the HG methods reflect their 'best-case scenarios' while BIKER reflects its worst case scenario, where priors are set entirely from SWOT observations and are generally the least informative they will ever be.
 
-Figure 7a evaluates the $FCO_2$ estimates using the same 4 error metrics as Figure 6a.
+**UPdate this paragraph**
+Figure 7a evaluates the $FCO_2$ estimates using the same 4 error metrics as Figure 5a.
 BIKER, the sole remote technique, is in red while the three in situ techniques are plotted as different shades of purple.
-We see that BIKER performs slightly worse than all three in situ methods across all metrics except *r*, where it performs similarly to the 'Raymond 2013' model.
-**NEED ACTUAL NUMBERS HERE**.
-However, across all 47 rivers BIKER has the least overall bias in its estimates (**add number here**).
-In summary, BIKER performs slightly worse than the in situ tools, as expected.
-However, the drop in performance is not substantial and is promising given that the in situ models represent their best case scenarios (perfect in situ discharge records) while BIKER represents its worst-case scenario (relatively uninformed priors).
+We see that BIKER performs slightly worse than all three in situ methods across all metrics except *r*, where it performs similarly (median BIKER *KGE* of `r round(median(results_fco2_rivs[results_fco2_rivs$key == 'FCO2_BIKER',]$KGE),2)` and median *r* of `r round(median(results_fco2_rivs[results_fco2_rivs$key == 'FCO2_BIKER',]$r),2)`).
+BIKER is also more biased (median *rBIAS* of `r round(median(results_fco2_rivs[results_fco2_rivs$key == 'FCO2_BIKER',]$rBIAS),2)`) than either 'Brinkerhoff 2019' or 'Raymond 2012' models and similarly biased to the 'Raymond 2013' model (median *rBIAS* of `r round(median(results_fco2_rivs[results_fco2_rivs$key == 'FCO2_brinkerhoff2019',]$rBIAS),2)`, `r round(median(results_fco2_rivs[results_fco2_rivs$key == 'FCO2_raymond2012',]$rBIAS),2)`, `r round(median(results_fco2_rivs[results_fco2_rivs$key == 'FCO2_raymond2013',]$rBIAS),2)`, respectively).
+
+**POTENTIAL PROBLEM: The swot rivers are so long (reach lengths of 100+km sometimes) that our total surface area is actually greater than Brian and I have for the northeast....
+and so the total C flux is greater than we model in the northeastern US.
+We don't actually care about these raw numbers so does this matter?**
 
@@
 b) Average carbon emissions via CO2 evasion across all rivers and timesteps, accounting for surface area.
-Completely remotely-sensed methods are colored in red, in situ methods in purple, and the observed values in orange.](cache/FCO2/FCO2_models.jpg)
+Completely remotely-sensed methods are colored in red, in situ methods in purple, and the observed values in orange.](cache/FCO2/FCO2_models.jpg){width=1000px}
 
+Finally, we carry these $FCO_2$ calculations all the way to annual carbon emissions rates.
 Figures 7b compares the average carbon emissions rate (via $F_{CO_2}$) from the `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers using BIKER posterior means and the three streamgauge-based HG models.
-All estimates are positively biased, with BIKER (`r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_BIKER',]$sumFCO2, 2)` gG-C/yr) and 'Raymond 2013' (`r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_raymond2013',]$sumFCO2, 2)` gG-C/yr) the most overestimated relative to the observed flux (`r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_obs',]$sumFCO2, 2)` gG-C/yr).
-*MORE* Thus, despite BIKER using absolutely no streamgauge or other in situ data like the other 3 methods do, it provides a reasonable upscaling estimate of the bulk carbon efflux across many rivers (Figure 6c).
+All estimates are positively biased, with BIKER (`r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_BIKER',]$yearlyCFux, 2)` gG-C/yr) and 'Raymond 2013' (`r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_raymond2013',]$yearlyCFux, 2)` gG-C/yr) the most overestimated relative to the observed flux (`r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_obs',]$yearlyCFux, 2)` gG-C/yr).
+*MORE* Thus, despite BIKER using absolutely no streamgauge or other in situ data like the other 3 methods do, it provides a reasonable (if positively biased) upscaling estimate of the bulk carbon efflux across many rivers (Figure 6c).
+As expected, BIKER performs equivalently or slightly worse than the in situ tools, but the drop in performance is not substantial and is promising given that the in situ models represent their best case scenarios (perfect in situ discharge records) while BIKER represents its worst-case scenario (relatively uninformed priors).
 
-## 4 Discussion
+## 4 Discussion NEEDS UPDATING READ AT OWN RISK!!
 ### 4.1 Gas exchange in hydraulically-wide rivers
@@
 This is empirically validated in Figure 3d.
-Scaling *k* this way (via both the turbulent dissipation rate and the shear Reynold's number) is frequently done in oceanic gas exchange studies [*EXAMPLES*] and has been occasionally explored in open-channel flows [@moogAirWaterGasTransfer1999; @wangPhysicallyBasedScaling2021].
+Scaling *k* this way (via a shear Reynold's number) is commonly done to parameterize wave-related gas exchange models in the open ocean [@brumerWaveRelatedReynoldsNumber2017; @zhaoDependenceWhitecapCoverage2001; @zhaoEffectWindWaves2003] and has been occasionally explored in open-channel flows [@moogAirWaterGasTransfer1999; @wangPhysicallyBasedScaling2021].
 However to our knowledge, @moogAirWaterGasTransfer1999's specific model has never been empirically validated in rivers.

### 2022-01-21 generateMS.R

@@ -1,5 +1,6 @@
 #################
-#This will grab and store all results in your current RStudio environment so that you can knit the manuscript correctly. It then knits the manuscript too.
+#This will grab and store all results in your current RStudio environment so that you can knit the manuscript correctly. It then knits the manuscript to a
+  #word doc that, via the redoc package, can be tracked changed and convert back to Rmd!
 #################
 
 #setwd("C:/Users/cbrinkerhoff/OneDrive - University of Massachusetts/Ongoing Projects/RSK600")
@@ -12,7 +13,7 @@ HG_swot <- read.csv('cache\\k600_theory\\HG_swot.csv')
 
 results_dynamics <- read.csv('cache/validation/results_dynamics.csv')
 results_k600_rivs <- read.csv('cache//validation//results_by_riv.csv')
-results_fco2_rivs <- read.csv('cache//FCO2//CO2_results.csv')
+results_fco2_rivs <- read.csv('cache//FCO2//results_by_riv.csv')
 results_fco2_bulk <- read.csv('cache//FCO2//massFluxes.csv')
 results_fco2_bulk$key <- as.character(results_fco2_bulk$key)
 
@@ -22,3 +23,15 @@ print('Manuscript generated.')
 
 rmarkdown::render("supp_v7.Rmd")
 print('Supplemental Information generated.')
+
+
+#To convert tracked changes edited manuscripts back to rmarkdown! #https://noamross.github.io/redoc/articles/mixed-workflows-with-redoc.html
+#NOTES:
+  #Can also be set to automatically accept tracked changes and comments.
+  #This package is no longer maintained...
+  #for tracked changes:
+    #{-- --}
+  #comments:
+    #comments in {== highlighted text  ==}{>> comment <<}
+
+#redoc::dedoc('manuscript_v7_e.docx', track_changes = 'criticmarkup', overwrite=TRUE, wrap=getOption("redoc.wrap", NA))

### 2022-01-21 lib/inputs.R

@@ -14,4 +14,4 @@ uncertainity <- 0.30 #see regression model for actual uncertainty
 #Implementation settings
 ###############
 cores <- 47 #num cores to run parallel processes on (depends on HPC session setup)
-samplingRate <- 11 #sampling frequency used with the pCO2 data from Beaulieu et al. (2012)
+samplingRate <- 14 #sampling frequency used with the pCO2 data from Beaulieu et al. (2012)

### 2022-01-21 src/figures_FCO2.R

@@ -18,41 +18,38 @@ output$river <- as.character(output$river)
 
 #Calculate total wetted surface area over all rivers
 total_sa <- output[!duplicated(output$sa), ] #[m2]
-total_sa <- sum(total_sa$sa)
+total_sa <-sum(total_sa$sa)
 
 #CO2 data is only for 29 samples, so ignore k values beyond that (applies to a few rivers)
 output <- drop_na(output)
 
 #Calculate total mass fluxes of CO2 from rivers------------------------------------------------
 massFluxes <- gather(output, key=key, value=value, c(FCO2_obs, FCO2_BIKER, FCO2_BIKER_low, FCO2_BIKER_high, FCO2_raymond2012, FCO2_raymond2013, FCO2_brinkerhoff2019)) %>%
+  group_by(river, key) %>%
+  summarise(yearlyCFux_byriv = mean(value*1e-12*sa_m2, na.rm=T)) %>% #tG-C/yr from all rivers
   group_by(key) %>%
-  summarise(sumFCO2 = mean(value * 1e-12 * 365, na.rm=T)*total_sa, #tG-C/yr from all rivers
-            medianFCO2 = median(value * 1e-12 * 365, na.rm=T)*total_sa) #tG-C/yr from all rivers
+  summarise(yearlyCFux = sum(yearlyCFux_byriv, na.rm=T))
 write.csv(massFluxes, 'cache/FCO2/massFluxes.csv')
 
 ##########################
 ##PLOT MEDIAN/MEAN EFFLUX AND BY-RIVER METRICS PLOT----------------------------------------------------------------------
 ###########################
 #bar plots of mass fluxes----------------------------------------
-lowCI_sum <- massFluxes[massFluxes$key == 'FCO2_BIKER_low',]$sumFCO2
-lowCI_med <- massFluxes[massFluxes$key == 'FCO2_BIKER_low',]$medianFCO2
-highCI_sum <- massFluxes[massFluxes$key == 'FCO2_BIKER_high',]$sumFCO2
-highCI_med <- massFluxes[massFluxes$key == 'FCO2_BIKER_high',]$medianFCO2
+lowCI_sum <- massFluxes[massFluxes$key == 'FCO2_BIKER_low',]$yearlyCFux
[…]

### 2022-01-21 src/figures_k600.R

@@ -24,14 +24,14 @@ stats_by_river <- group_by(results, river, errFlag) %>%
             NRMSE = sqrt(mean((kobs - kest_mean)^2, na.rm=T)) / mean(kobs, na.rm=T),
             rBIAS =   mean(kest_mean- kobs, na.rm=T) / mean(kobs, na.rm=T),
             KGE = KGE(kest_mean, kobs),
-            CV = sd(kobs)/mean(kobs),
+            obsCV = sd(kobs)/mean(kobs),
             meanKobs = mean(kobs, na.rm=T),
             meanWobs = mean(Wobs, na.rm=T),
             meanSobs = mean(Sobs, na.rm=T),
             meanDobs = mean(Dobs, na.rm=T),
             n_data=n(),
-            priorKbias = (mean(kprior, na.rm=T)-mean(kobs, na.rm=T))/mean(kobs, na.rm=T),
-            BIKERbias = (sum(kest_mean - kobs, na.rm=T)/n())/(mean(kobs, na.rm=T)))
+            priorCV = sd(kprior)/mean(kprior),
+            posteriorCV = sd(kest_mean)/mean(kest_mean))
 
 plot_stats <- gather(stats_by_river, key=key, value=value, c('NRMSE', 'rBIAS', 'KGE', 'r'))
 plot_stats <- filter(plot_stats, key %in% c('NRMSE', 'KGE', 'rBIAS', 'r'))
@@ -58,6 +58,32 @@ plotSWOTreaches <- ggplot(plot_stats, aes(x=key, y=value, fill=factor(errFlag)))
         legend.text = element_text(size=17),
         legend.title = element_text(size=17, face='bold'))
 
+plotSWOTreaches_cdfs_kge <- ggplot(plot_stats[plot_stats$key=='KGE',], aes(x=value, color=factor(errFlag))) +
+  stat_ecdf(size=2) +
+  xlab('KGE') +
+  ylab('Density') +
+  scale_color_brewer(palette = 'Accent', name='', labels=c('No Error', 'SWOT Measurement Error')) +
+  theme(legend.position='none',
+        axis.text=element_text(size=20),
+        axis.title=element_text(size=24,face="bold"),
+        legend.text = element_text(size=17),
+        legend.title = element_text(size=17, face='bold'))
+
[…]

### 2022-01-21 src/runFCO2.R

@@ -10,6 +10,9 @@ print('starting FCO2...')
 legend_labels <- c('BIKER',
                    'Model using observed \ndepth')
 
+#This model specifically predicts w and d from HG and then obtains velocity by continuity. The model fit for velocity was notably worse.
+Brinkerhoff2019_AHG <- read.csv('cache/AHG.csv')
+
 ###################
 ##FUNCTION TO GENERATE FCO2 FROM ALL DEPTH MODELS AND ALL SWOT RIVERS--------------------------------
 ####################
@@ -37,11 +40,15 @@ fun_FCO2_analysis <- function(river) {
   area[area<0]=NA
   Q_obs[Q_obs<0]=NA
 
-  #Sample every 14 days because CO2 data is biweekly (convienantly, average SWOT overpass frequency is 11 days, so close)
+  #Sample every 14 days because CO2 data is biweekly (conveniently, average SWOT overpass frequency is 11 days, so close)
   W_obs = W_obs[,seq(river_start_date, ncol(W_obs), samplingRate)]
   S_obs = S_obs[,seq(river_start_date, ncol(S_obs), samplingRate)]
   Q_obs = Q_obs[,seq(river_start_date, ncol(Q_obs), samplingRate)]
   area = area[,seq(river_start_date, ncol(area), samplingRate)]
+  
+  if(is.null(dim(W_obs))==1) {
+    return(NA)
+  }
 
   #Some NA handling in slopes (borrowed from Frassion etal 2021 BAM runs)-----------------------------------------------------------
   if (any(apply(S_obs,2,sum,na.rm=TRUE) ==0)){ #removes timesteps with all NA slopes
@@ -78,7 +85,7 @@ fun_FCO2_analysis <- function(river) {
     }
   }
   D_raymond2013 <- d
-  D_brinkerhoff2019 <- 0.265191*Q_obs^(0.39496) #Brinkerhoff etal 2019, estimated in CONUS_CO2 code
+  D_brinkerhoff2019 <- Brinkerhoff2019_AHG[2,]$coef*Q_obs^(Brinkerhoff2019_AHG[2,]$slope) #Brinkerhoff etal 2019, estimated in CONUS_CO2 code
 
[…]

### 2022-01-21 src/swot_k_model_v3.R

@@ -96,6 +96,13 @@ data$k600 <- NA
 data$study <- 'Brinkerhoff_etal_2019'
 data <- rbind(data, ulseth_data_s, usgs_data_s)
 
+#########QUICK DETOUR TO CALCULATE HG MODELS FOR FCO2 CALCULATIONS------------------------------
+num <- nrow(data)
+wid <- lm(log(width)~log(Qm3s), data=data)
+dep <- lm(log(depth)~log(Qm3s), data=data)
+models <- data.frame('model' = c('width', 'depth'), 'r2'=c(summary(wid)$r.squared, summary(dep)$r.squared), 'coef'=c(exp(wid$coefficient[1]), exp(dep$coefficient[1])), 'slope'=c(wid$coefficient[2], dep$coefficient[2]))
+write.csv(models, 'cache\\AHG.csv')
+
 #######CALCULATE CHANNEL GEOMETRY---------------------------------------------------------
 data$Rh <- (data$depth*data$width)/(data$width + 2*data$depth) #hydraulic radius, assuming rectangular channel [m]
 data$ustar <- sqrt(g*data$Rh*data$slope) #friction/shear velocity [m/s]

### 2022-01-24 manuscript_v7.Rmd
manuscript v7

@@
 - BIKER is robust to measurement errors implicit in SWOT river measurements
-- BIKER will allow for unprecedented study of gas exchange spatiotemporal dynamics after SWOT's launch
+- BIKER will allow for novel study of gas exchange spatiotemporal dynamics after SWOT's launch
 
@@
 These models usually predict $k_{600}$, or *k* normalized by a Schmidt number of 600.
-This is to remove the effect of water temperature and gas type from predictive models, as warmer waters and lower *Sc* numbers both increase gas exchange rates [@hallGasExchangeStreams2020], i.e.e $k_{600}$ reflects the $k_{CO_{2}}$ at 20 degrees Celsius.
+This is to remove the effect of water temperature and gas type from predictive models, as warmer waters and lower *Sc* numbers both increase gas exchange rates [@hallGasExchangeStreams2020].
+Specifically, $k_{600}$ reflects the $k_{CO_{2}}$ at 20 degrees Celsius.
 Through this normalization, these models exclusively perturb the geophysical controls on gas exchange [@hallGasExchangeStreams2020].
@@
 It has been shown at the field-scale that temporal dynamics of gas exchange can vary widely from site to site [@wallinSpatiotemporalVariabilityGas2011], but it has remained impractical to obtain temporally explicit $k_{CO_2}$ at continental-to-global scales.
-This knowledge gap is contrasted agaist the robust existing literature exploring spatiotemporal patterns in $CO_2_{water}$ [e.g.
+More recently, *Liu etal accepted* performed a first assessment of monthly temporal dynamics in the global river $CO_2$ flux, though their focus was not on *k* and they relied on modeled streamflow to achieve this.
+This knowledge gap is contrasted against the robust existing literature exploring spatiotemporal patterns in $CO_2_{water}$ [e.g.
 @ahoDistinctConcentrationdischargeDynamics2021; @ahoDifferentialResponseGreenhouse2019; @crawfordCO2TimeSeries2017; @peterScalesDriversTemporal2014; @ranDynamicsRiverineCO22017; @rocher-rosLandscapeProcessDomains2019; @raymondAtmosphericCO2Evasion2000; @liuHydrologicControlsPCO22018a].
@@
 However, they were still limited by a lack of direct hydraulic measurements and had to rely on scaling equations to estimate river depth and velocity.
-Even though approaches like this are incredibly useful for expanding our mechanistic understanding of gas exchange, it is less useful for global upscaling purposes as it relies on highly detailed in situ DO data.
+Even though approaches like this are incredibly useful for expanding our mechanistic understanding of gas exchange, it is less useful for global upscaling purposes as it relies on highly detailed in situ DO data for every river [@hallGasExchangeStreams2020].
 
@@
 
-To answer this first question, we use the hydraulic geometry [the fundamental geomorphic relationships between streamflow and channel shape- @leopoldHydraulicGeometryStream1953] of large, SWOT-observable rivers to develop a process-based model for large-river gas exchange.
+To answer this first question, we use hydraulic geometry [the fundamental geomorphic relationships between streamflow and channel shape- @leopoldHydraulicGeometryStream1953] to develop a process-based model for large-river gas exchange.
 We then take these findings and explore question 2 by implementing the model within an RS of $k_{600}$ algorithm named BIKER ('Bayesian Inference of the $k_{600}$ Evasion Rate').
@@
 ## 2 Hydraulic geometry in large rivers and predicting fluvial gas exchange velocity
-To predict *k* in the very large rivers that SWOT will observe, we start from established process-based models for *k*, impose hydraulic assumptions that are valid in SWOT-observable rivers, and then obtain a suite of models we empirically test.
+To predict *k* in the very large rivers that SWOT will observe, we start from established process-based models for *k*, impose hydraulic assumptions that are valid in SWOT-observable rivers, and obtain a suite of models we empirically test.
 Following a description of the data used (section 2.1), we outline these established models (section 2.2) and then step through our hydraulic assumptions to arrive at equations that can be ingested by SWOT (section 2.2).
@@
 
-To quantify how many SWOT-observable rivers meet our hydraulic assumptions, we expand our set of stream hydraulics (but not $k_{600}$) using a previously published compilation of in situ hydraulic measurements [@brinkerhoffReconcilingAtaStationAtManyStations2019].
+To quantify how many SWOT-observable rivers meet our hydraulic assumptions, we expand our dataset of stream hydraulics (but not $k_{600}$) using a previously published compilation of in situ hydraulic measurements [@brinkerhoffReconcilingAtaStationAtManyStations2019].
 That dataset contains over 530,000 unique measurements of river channel velocity, width, and discharge from across the continental United States, originally made to calibrate United States Geological Survey (USGS) streamgauge rating curves and made public by the USGS.
@@
 The Reynolds model is hypothetically useful in low-turbulence flows where a relative lack of large-scale eddies effectively 'filter out' the number of small-eddies that actually reach the interface and initiate gas exchange [@talkeTurbulentKineticEnergy2013].
-Scaling *k* via a shear Reynold's formulation is sometimes done to parameterize wave-breaking gas exchange models in the open ocean  [@brumerWaveRelatedReynoldsNumber2017; @zhaoDependenceWhitecapCoverage2001; @zhaoEffectWindWaves2003], though is infrequently done in rivers.
+While scaling *k* via a shear Reynold's formulation is sometimes done to parameterize wave-breaking gas exchange models in the open ocean  [@brumerWaveRelatedReynoldsNumber2017; @zhaoDependenceWhitecapCoverage2001; @zhaoEffectWindWaves2003], it is infrequently done in rivers.
 In the context of BIKER, we chose to test this model because large, SWOT-observable rivers are generally the smoothest, least-turbulent flows along the stream-to-ocean continuum.
@@
 ### 2.3 Imposing large-river hydraulic geometry on *k* models
-Given the theoretical context privded in Section 2.2, we know turn to SWOT-observable systems specifically.
+Given the theoretical context provided in Section 2.2, we know turn to SWOT-observable systems specifically.
 Rivers and streams change predictably along their longitudinal profile from headwater to ocean, and we can exploit the hydraulic geometry of large rivers at the end of this continumn to estimate *k* in SWOT-observable systems.
@@
 Assuming that all SWOT-observable rivers are hydraulically-wide, we test the performance of four models for predicting *k* in hydraulically-wide channels: the four unique combinations of equations 3-4 and the two $\epsilon$ models outlined above.
-While the complete model derivations for the small-eddy models (section 2.3.1) and the reynolds-extension models (section 2.3.2) are provided in Text S2, the final equations are provided in Table 1.
+While the complete model derivations for the small-eddy models (section 2.3.1) and the Reynolds-extension models (section 2.3.2) are provided in Text S2, the final equations are provided in Table 1.
 The overall goal was to reduce the equations down to their fundamental parameters, identifying which terms are SWOT observable and limiting the number of terms not directly measurable via SWOT.
@@
 *Table 1: Four physically-based models for $k_{600}$ each derived under a different set of mass transport assumptions but all valid in hydraulically-wide rivers.
-Their derivations are outlined in sections 3.2.1 and 3.2.2 and their empirical tests are shown in section 3.3.*
+Their derivations are outlined in sections 2.3.1 and 2.3.2 and their empirical tests are shown in section 2.4.*
 
@@
 (d) Reynolds-extension model with form-drag dissipation.
-Note that axes are plotted in logarithmic space just for visualization: model fit and validation were calculated in linear space.](cache/k600_theory/k600Plot.jpg){width=1000px}
+Note that axes are plotted in logarithmic space just for visualization: model fit and validation were calculated in linear space.](cache/k600_theory/k600Plot.jpg)
 
@@
 They systematically underestimate lower gas exchange velocities and overestimate higher ones, with the log-law-of-the-wall model (Figure 3a) performing marginally worse both visually and via $r^2$ (`r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eS',]$r2, 2)` versus `r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eD',]$r2, 2)` for Figure 3c).
-However, when this model is expanded to retard the dissipation rate in the smoothest flows via the Reynolds extension model [equations 6 and 7- @moogAirWaterGasTransfer1999], the dynamics are more accurately captured (Figures 3b, 3d) and $r^2$ scores go up appreciably (`r round(modelsWIDE[modelsWIDE$name == 'reynolds-eS',]$r2, 2)` and `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2, 2)` for the log-law-of-the-wall and form-drag dissipation, respectively).
+However, when this model is expanded to further retard the dissipation rate via the Reynolds extension model [equations 6 and 7- @moogAirWaterGasTransfer1999], the dynamics are more accurately captured (Figures 3b, 3d) and $r^2$ scores increase (`r round(modelsWIDE[modelsWIDE$name == 'reynolds-eS',]$r2, 2)` and `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2, 2)` for the log-law-of-the-wall and form-drag dissipation, respectively).
 Modeling dissipation via total form-drag rather than just via bed shear improves model performance regardless of $k_{600}$ equation used.
@@
 Note that this $CO_2$ data is for the Ohio River only but was applied to all rivers (which includes multiple sections of the Ohio River).
-Because we are exclusively interested in the relative differences between $FCO_2$ estimates and not the raw fluxes themselves, any $CO_2$ data representative of SWOT-observable rivers was deemed acceptable for this validation exercise.
+Therefore, the raw carbon emissions estimates presented in this paper are largely meaningless in the context of actually measured carbon emissions from these rivers.
+This approach was deemed acceptable for this validation exercise because we are exclusively interested in the relative differences between methods tested and not the raw fluxes themselves.
 
@@
 All of this algebra simplifies to equation 10.
-Conveniently, $k_{600}$ as measured by a tracer addition to a stream is inherently a reach-scale quantity.
+Conveniently, $k_{600}$ as measured by tracer additions to a stream is inherently a reach-scale quantity (in a mass-conserved reach).
 Therefore, equations 8 and 10 both yield a reach-scale $k_{600}$ (i.e.
@@
 
-With the sampling model defined ($f(x|\Theta)$ = equation 11) and hyperparameters $p(x$) described (Texts S3 and S4), a joint posterior distribution conditional on the SWOT observations ($p(\Theta|x) \propto f(x|\Theta)p(x)$) is therefore also specified.
+With the sampling model described ($f(x|\Theta)$ = equation 11) and hyperparameters $p(x$) specified (Texts S3 and S4), a joint posterior distribution conditional on the SWOT observations ($p(\Theta|x) \propto f(x|\Theta)p(x)$) is therefore also specified.
 To approximate this distribution, we use a Markov Chain Monte Carlo (MCMC) algorithm implemented using the Stan probabilistic programming language.
@@
 
-With the setup described by equations 11 and 12, validation is performed using the BIKER posterior means.
-Because we are most interested in BIKER's ability to correctly infer temporal dynamics in $k_{600}$, and each river's priors are described specifically for that river, we validate by river.
-These metrics are defined in Table S2 and consist of the following:
+With the setup described by equations 11 and 12, validation is performed for each river using the BIKER posterior means. These metrics are defined in Table S2 and consist of the following:
+
 - Correlation coefficient *r* to quantify accuracy of BIKER's temporal dynamics
@@
 #### 3.2.3 Carbon efflux validation
-It is one thing to accurately model the temporal dynamics of $k_{600}$, but researchers are often most interested in the actual gas fluxes from rivers and ultimately the carbon emitted from river to atmosphere.
+It is one thing to accurately model the temporal dynamics of $k_{600}$, but researchers are often most interested in the actual carbon emitted from river to atmosphere.
 Therefore, we also explore BIKER's ability to reproduce $FCO_2$ (equation 1) and the annual carbon emission rate from these `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers.
@@
 Equation 13 is used to obtain a temperature and $CO_2$ specific gas exchange velocity from BIKER's output, where the Schmidt number *Sc* was estimated following @raymondScalingGasTransfer2012.
-We then estimate a yearly average carbon emission rate (via $CO_2$ evasion) by applying each river's mean $FCO_2$ over the river's surface area and then summing all of these emission rates to obtain an emissions rate in carbon per year across all rivers.
-
+We then estimate a yearly average carbon emission rate (via $CO_2$ evasion) by applying each river's mean $FCO_2$ over the river's surface area and then summing all of these emission rates.
 $$k_{CO2, temp}=k_{600}*(Sc/600)^{-1/2} \mathbf{(13)}$$
 Finally, we also compare BIKER's outputs with established in situ techniques, whose specific details and parameters are provided in Table S3.
-Streamflow is more readily modeled at the global scale than river channel geometry, and so these in situ models are used to predict $\bar{U}$ and *H* as functions of streamflow.
-Their outputs are in turn used to estimate *k* in upscaling workflows.
+Streamflow is more readily modeled at the global scale than river channel geometry, and so these in situ models are used to predict $\bar{U}$ and *H* as functions of streamflow using hydraulic geometry (HG).
+Per equation 2, their outputs are in turn used to estimate *k*.
 Therefore, they rely on in situ streamflow records or high-quality streamflow routing outputs (unlike BIKER, which only requires a river be SWOT-observable).
@@
 For some rivers, (e.g.
-Mississippi Intermediate), the first timestep is *THIS MIGHT BE DUE TO TO CERTAINTY OF THE PRIOR ETHIER ON K600 OR THE LIKELIHOOD???/*
+Mississippi Intermediate), the first timestep is **THIS MIGHT BE DUE TO TO CERTAINTY OF THE PRIOR ETHIER ON K600 OR THE LIKELIHOOD?**
 
@@
 These scores are excellent given that absolutely no in situ information is being used to predict $k_{600}$.
-NRMSE has a median score of `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$NRMSE, na.rm=T),2)`, which is reasonable for a completely ungauged method.
-These are comparable to the best NRMSE scores achieved by recent versions of the SWOT RSQ algorithms [@frassonExploringFactorsControlling2021].
-Median rBIAS is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$rBIAS, na.rm=T),2)`, though the shape of the rBIAS boxplot in Figure 5a highlights a large portion of rivers which have a positive bias (Figure 4 also confirms this).
+NRMSE has a median score of `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$NRMSE, na.rm=T),2)`, which is comparable to the better NRMSE scores achieved by recent versions of the SWOT RSQ algorithms [@frassonExploringFactorsControlling2021].
+Median rBIAS is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$rBIAS, na.rm=T),2)`, highlighting a large portion of rivers which have a significant positive bias (Figure 4 also confirms this).
 Taken in aggregate, Figures 4-5 indicate that BIKER is quite good at capturing temporal dynamics in $k_600$, however there is sometimes positive bias in its estimates.
@@
 ![Figure 5.
-a: Performance metrics by river.
+a) Performance metrics by river.
 See Table S2 for metric definitions.
 Dashed lines denote scores of 0, 0.50, and 1.
-CIs.
-b-c: Empirical cummulative density distributions of the *KGE* (b) and *r* (c) scores for both error and no-error rivers.](cache/validation/validation_by_river.jpg){width=1000px}
+b-c) Empirical cummulative density distributions of the *KGE* (b) and *r* (c) scores for both error and no-error rivers.](cache/validation/validation_by_river.jpg)
 
@@
 However, caution should be used here as these purple boxplots are composed of only `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` of the rivers in the green boxplots and many of the most poorly performing rivers are not in this set.
-The eCDFs more clearly indicate a marginal difference in performance when measurement errors are used to corrupt the BIKER inputs (with nearly identical performance when BIKER performs well for either metric).
+The eCDFs more clearly indicate a marginal difference in performance when measurement errors are used to corrupt the BIKER inputs (with nearly identical performance in the best-performing rivers).
 Given these results, we deem that SWOT measurement error does not exert a significant influence on BIKER performance and so the results presented for the rest of the manuscript assume no measurement error in order to use all `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers.
@@
 Figure 6a plots a histogram of the slopes of power-law regressions between BIKER-predicted $k_{600}$ and observed $k_{600}$.
-A slope of 1 indicates that BIKER correctly infers the daily flucuations in gas exchange, while a slope well above or below 1 indicates incorrect inference of temporal dynamics in $k_{600}$.
-Figure 6a shows that the majority of rivers' slopes approximate 1 (mean slope: `r round(median(results_dynamics$Slope),2)`).
+A slope of 1 indicates that BIKER correctly infers the daily flucuations in gas exchange (even if there is bias baked into the estimates), while a slope well above or below 1 indicates incorrect inference of the temporal dynamics of $k_{600}$.
+Figure 6a shows that the majority of rivers' slopes approximate 1 (mean slope: `r round(median(results_dynamics$Slope),2)`), indicating strong inference of temporal gas exchange trends.
+A handful of rivers to have slopes well above or below 1, though the vast majority are at or just below 1.
 
-![Figure 6: a: Histogram of the slopes of power-law regressions between BIKER estimates and observed $k_{600}$.
-Slopes equal 1 indicate correct inference of the temporal trends in daily $k_{600} (regardless of bias in the estimates).
-b: .](cache/validation/validation_temporal.jpg){width=1000px}
+![Figure 6: a) Histogram of the slopes of power-law regressions between BIKER estimates and observed $k_{600}$.
+Slopes equaling 1 indicate correct inference of the temporal trends in daily $k_{600}$ (regardless of bias in the estimates).
+b) Empirical cummulative density distributions of the coefficient of varation of prior, posterior, and observed $k_{600}$ per river.
+If BIKER is correctly inferring temporal variation in $k_{600}$, the posterior eCDF will approximate the obersved eCDF (as it does).](cache/validation/validation_temporal.jpg)
 
-Finally, we explore whether BIKER improves upon the prior with respect to temporal variance in $k_{600}$, here defined using the coefficient of variation ($CV=\frac{SD}{mean}$).
-Figure 6b plots eCDFs of observed, prior, and posterior $k_{600}$ *CV*, strongly showing that BIKER improves upon the prior information available to us about $k_{600}$, with posterior $k_{600}$ CV almost entirely overlapping the observed eCDF and sufficiently 'closing the gap' between the prior and observed eCDFs.
+Finally, we explore whether BIKER improves upon the prior with respect to temporal variation in $k_{600}$, here defined using the coefficient of variation ($CV=\frac{SD}{mean}$).
+Figure 6b plots eCDFs of observed, prior, and posterior $k_{600}$ *CV*, strongly showing that BIKER improves upon the prior information available to us about $k_{600}$.
+Posterior $k_{600}$ *CV* approximates the observed eCDF and sufficiently 'closes the gap' between our prior knowledge on $k_{600}$ temporal variation and the actually observed variation.
 It must be stressed that this validation was performed using a relatively uninformed prior on $k_{600}$ (Text S3) and that this gap may not be as large in many settings.
-However, Figure 6b certainly confirms that Bayesian technqiues can be used to infer temporal variation in $k_{600}$ when in situ work is not feasible, effectively reconciling bias between our prior knowledge of $k_{600}$ temporal variation and the observed variation.
+However, Figure 6b certainly confirms that Bayesian techniques can be used to infer temporal variation in $k_{600}$ when in situ work is not feasible and effectively reconcile any bias between our prior knowledge of $k_{600}$ temporal variation and the observed variation.
 
 #### 3.3.2 Carbon efflux
-We now compare BIKER's estimates of carbon fluxes against established in situ techniques.
+Finally, we carry these $FCO_2$ calculations all the way to annual carbon emissions rates.
 It is important to remember that only BIKER is completely globally implementable, while the other three models necessarily rely on having a streamflow record readily available.
 Therefore, the HG methods reflect their 'best-case scenarios' while BIKER reflects its worst case scenario, where priors are set entirely from SWOT observations and are generally the least informative they will ever be.
+We also stress again that the raw emissions rates here are largely meaningless as they are calculated from an identical $CO_{2_water}$ timeseries applied to every river.
+We are principally interested in relative differences between methods employed.
 
-**UPdate this paragraph**
-Figure 7a evaluates the $FCO_2$ estimates using the same 4 error metrics as Figure 5a.
-BIKER, the sole remote technique, is in red while the three in situ techniques are plotted as different shades of purple.
-We see that BIKER performs slightly worse than all three in situ methods across all metrics except *r*, where it performs similarly (median BIKER *KGE* of `r round(median(results_fco2_rivs[results_fco2_rivs$key == 'FCO2_BIKER',]$KGE),2)` and median *r* of `r round(median(results_fco2_rivs[results_fco2_rivs$key == 'FCO2_BIKER',]$r),2)`).
-BIKER is also more biased (median *rBIAS* of `r round(median(results_fco2_rivs[results_fco2_rivs$key == 'FCO2_BIKER',]$rBIAS),2)`) than either 'Brinkerhoff 2019' or 'Raymond 2012' models and similarly biased to the 'Raymond 2013' model (median *rBIAS* of `r round(median(results_fco2_rivs[results_fco2_rivs$key == 'FCO2_brinkerhoff2019',]$rBIAS),2)`, `r round(median(results_fco2_rivs[results_fco2_rivs$key == 'FCO2_raymond2012',]$rBIAS),2)`, `r round(median(results_fco2_rivs[results_fco2_rivs$key == 'FCO2_raymond2013',]$rBIAS),2)`, respectively).
+Figures 7 compares the annual carbon emissions rate (via $F_{CO_2}$) from the rivers using BIKER posterior means and the three streamgauge-based HG models.
+Surprisingly, BIKER outperforms the gauge-based approaches, nearly correctly inferring the annual carbon emissions rate (`r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_BIKER',]$yearlyCFux, 2)` gG-C/yr for BIKER versus `r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_obs',]$yearlyCFux, 2)` gG-C/yr observed).
+The three HG models overestimate this emissions rate (`r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_raymond2013',]$yearlyCFux, 2)`, `r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_raymond2012',]$yearlyCFux, 2)`, and `r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_brinkerhoff2019',]$yearlyCFux, 2)`) for 'Raymond 2013', 'Raymond 2012', and 'Brinkerhoff 2019' respectively.
+Both 'Brinkerhoff 2019' and 'Raymond 2012' fall within the BIKER CIs.
+**I need to add CIs for Brink 2019.
+Can't calc for raymond models as only some of the parameter CIs are reported**.
+BIKER's realtively stronger performance in FIgure 7 compared to Figures 4-6 is likely due to the temporal averaging performed at each river, which effecively reduces the influence of positive bias sometimes seen in BIKER results.
 
-**POTENTIAL PROBLEM: The swot rivers are so long (reach lengths of 100+km sometimes) that our total surface area is actually greater than Brian and I have for the northeast....
+Thus, despite BIKER using absolutely no streamgauge or other in situ data like the other 3 methods do, it provides a strong upscaling estimate of the bulk carbon efflux across many rivers (Figure 7).
+Interestingly, the in situ models represent their best case scenarios (perfect in situ discharge records) while BIKER represents its worst-case scenario (relatively uninformed priors).
+This means that BIKER's annual carbon emission estimate can only improve.
+
+![Figure 7: Yearly carbon emissions via CO2 evasion across all rivers and accounting for surface area.
+Completely remotely-sensed methods are colored in red, in situ methods in purple, and the observed in orange.](cache/FCO2/FCO2_models.jpg)
+
+**NOTE: The swot rivers are so long (reach lengths of 100+km sometimes) that our total surface area is actually greater than Brian and I have for the northeast....
 and so the total C flux is greater than we model in the northeastern US.
-We don't actually care about these raw numbers so does this matter?**
-
-![Figure 7: a) FCO2 validation for BIKER and the three in situ HG models.
-b) Average carbon emissions via CO2 evasion across all rivers and timesteps, accounting for surface area.
-Completely remotely-sensed methods are colored in red, in situ methods in purple, and the observed values in orange.](cache/FCO2/FCO2_models.jpg){width=1000px}
-
-Finally, we carry these $FCO_2$ calculations all the way to annual carbon emissions rates.
-Figures 7b compares the average carbon emissions rate (via $F_{CO_2}$) from the `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers using BIKER posterior means and the three streamgauge-based HG models.
-All estimates are positively biased, with BIKER (`r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_BIKER',]$yearlyCFux, 2)` gG-C/yr) and 'Raymond 2013' (`r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_raymond2013',]$yearlyCFux, 2)` gG-C/yr) the most overestimated relative to the observed flux (`r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_obs',]$yearlyCFux, 2)` gG-C/yr).
-*MORE* Thus, despite BIKER using absolutely no streamgauge or other in situ data like the other 3 methods do, it provides a reasonable (if positively biased) upscaling estimate of the bulk carbon efflux across many rivers (Figure 6c).
-As expected, BIKER performs equivalently or slightly worse than the in situ tools, but the drop in performance is not substantial and is promising given that the in situ models represent their best case scenarios (perfect in situ discharge records) while BIKER represents its worst-case scenario (relatively uninformed priors).
+We don't actually care about these raw numbers and I mention that earlier in the paper but worth noting**
 
@@
 ## 7 Apendix A
-*Table A1: Variable description and notation for this study. ADD THE BAYES PARAMETERS YET*
+*Table A1: Variable description and notation for this study*
 

### 2022-01-24 scraps.Rmd
manuscript v7

@@
+Figure 7a evaluates the $FCO_2$ estimates using the same 4 error metrics as Figure 5a.
+BIKER, the sole remote technique, is in red while the three in situ techniques are plotted as different shades of purple.
+We see that BIKER performs slightly worse than all three in situ methods across all metrics except *r*, where it performs similarly (median BIKER *KGE* of `r round(median(results_fco2_rivs[results_fco2_rivs$key == 'FCO2_BIKER',]$KGE),2)` and median *r* of `r round(median(results_fco2_rivs[results_fco2_rivs$key == 'FCO2_BIKER',]$r),2)`).
+BIKER is also more biased (median *rBIAS* of `r round(median(results_fco2_rivs[results_fco2_rivs$key == 'FCO2_BIKER',]$rBIAS),2)`) than either 'Brinkerhoff 2019' or 'Raymond 2012' models and similarly biased to the 'Raymond 2013' model (median *rBIAS* of `r round(median(results_fco2_rivs[results_fco2_rivs$key == 'FCO2_brinkerhoff2019',]$rBIAS),2)`, `r round(median(results_fco2_rivs[results_fco2_rivs$key == 'FCO2_raymond2012',]$rBIAS),2)`, `r round(median(results_fco2_rivs[results_fco2_rivs$key == 'FCO2_raymond2013',]$rBIAS),2)`, respectively).
+
 Beyond $CO_2$ emissions, *k* for various dissolved gases are of significant importance to aquatic ecologists modeling stream metabolism [e.g.

### 2022-01-24 src/figures_FCO2.R

@@ -93,6 +93,6 @@ boxPlots <- ggplot(plot_stats, aes(x=key2, y=value2, fill=key))+
 #######################
 #BRING VARIOUS FCO2 FIGURES TOGETHER INTO SINGLE FILE AND WRITE TO DISK-----------------------------
 #######################
-FCO2_models <- plot_grid(boxPlots, barPlot, ncol=2, labels=c('a', 'b'), label_size=18)
-ggsave('cache/FCO2/FCO2_models.jpg', FCO2_models, width=12, height=7)
+#FCO2_models <- plot_grid(boxPlots, barPlot, ncol=2, labels=c('a', 'b'), label_size=18)
+ggsave('cache/FCO2/FCO2_models.jpg', barPlot, width=8, height=8)
 write.csv(stats, 'cache/FCO2/results_by_riv.csv')

### 2022-01-28 docs/supp_v7.Rmd
manuscript v8 initial

--- 2022-01-14 supp_v7.Rmd
@@
 
-|**Study**|**Workers**|**Number of measurements**|
+|**Study**|**Workers**|**Additional studies**|
 |---------|-------------|--------------------------|
-|@ulsethDistinctAirWater2019|@ulsethDistinctAirWater2019|
-|@ulsethDistinctAirWater2019|@hallUseArgonMeasure2018|
-|@ulsethDistinctAirWater2019|@schelkerCO2EvasionSteep2016|
-|@ulsethDistinctAirWater2019|@mauriceInfluenceFlowBed2017|
-|@ulsethDistinctAirWater2019|@raymondScalingGasTransfer2012|
-|@churchillPredictionStreamReaeration1964|@churchillPredictionStreamReaeration1964|
-|@owensReaerationStudiesStreams1964|@owensReaerationStudiesStreams1964|
+|@ulsethDistinctAirWater2019|@ulsethDistinctAirWater2019|NA|
+|@ulsethDistinctAirWater2019|@hallUseArgonMeasure2018|NA|
+|@ulsethDistinctAirWater2019|@schelkerCO2EvasionSteep2016|NA|
+|@ulsethDistinctAirWater2019|@mauriceInfluenceFlowBed2017|NA|
+|@ulsethDistinctAirWater2019|@raymondScalingGasTransfer2012|@melchingReaerationEquationsDerived1999; @bottEcosystemMetabolismStreams2006; @bottEcosystemMetabolismPiedmont2006; @mulhollandInterbiomeComparisonFactors2001; @bernotInterregionalComparisonLanduse2010; @tsivoglouCharacterizationStreamReaeration1972|
+|@churchillPredictionStreamReaeration1964|@churchillPredictionStreamReaeration1964|NA|
+|@owensReaerationStudiesStreams1964|@owensReaerationStudiesStreams1964|NA|
 

### 2022-01-28 manuscript_v8.Rmd
manuscript v8 initial

---
title: "Using hydraulic geometry and the SWOT satellite to remotely sense river gas exchange velocity"
author:
  - Craig B Brinkerhoff:
      email: cbrinkerhoff@umass.edu
      institute: UMass
      correspondence: true
  - Colin J Gleason:
      institute: UMass
  - Christopher J Zappa:
      institute: Lamont
  - Peter A Raymond:
      institute: Yale
  - Merritt H Harlan:
      institute: UMass
institute:
  - UMass: Department of Civil & Environmental Engineering, University of Massachusetts-Amherst
  - Lamont: Lamont-Doherty Earth Observatory, Columbia University
  - Yale: School of the Environment, Yale University
bibliography: References.bib
csl: agu_citation_style.csl
output:
  word_document:
      pandoc_args:
        - '--lua-filter=scholarly-metadata.lua'
        - '--lua-filter=author-info-blocks.lua'
      reference_docx: word-styles-reference.docx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Key Points
- The BIKER (Bayesian Inference of the $k_{600}$ Exchange Rate) algorithm predicts gas exchange velocity and $CO_2$ fluxes solely from simulated SWOT data without calibration
- BIKER is robust to measurement errors implicit in SWOT river measurements
- BIKER will allow for novel study of gas exchange spatiotemporal dynamics after SWOT's launch

## Keywords
gas exchange, fluvial geomorphology, remote sensing, open-channel flow, SWOT, biogeochemistry

## Abstract
Extensive research over the past two decades has shown that the global river network emits significant amounts of greenhouse gas via gas exchange. Despite much progress, there is still large uncertainty in the temporal dynamics of gas exchange and thus carbon emissions to the atmosphere. Much of this uncertainty stems from uncertainty in gas exchange velocity $k_{600}$, the rate of this diffusion. We propose that the NASA/CNES/USKA/CSA SWOT satellite, set to usher in a new age of hydrology research at the global scale, can provide similarly transformative insights to fluvial gas exchange modeling upon launch in 2022. Here, we present work inferring $k_{600}$ from future SWOT observations without in situ calibration. We exploit the unique geomorphology of SWOT-observable rivers to develop a physical model of gas exchange that is nearly entirely remotely sensible and explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2,2)*100`% of variation in $k_{600}$ We then couple this model with established Bayesian inference techniques to develop BIKER, the 'Bayesian Inference of the $k_{600}$ Exchange Rate' algorithm. We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers (SWOT has not yet launched) and 166 discrete measurements of river gas exchange, yielding an algorithm that predicts the $k_{600}$ timeseries solely from SWOT observations with a by-river median Kling-Gupta Efficiency of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$KGE), 2)`. Like similar Bayesian remote sensing algorithms, BIKER is better at reproducing the temporal dynamics of gas exchange (median correlation coefficient of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$r), 2)`), than the absolute rates of exchange (median relative bias of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$rBIAS), 2)*100`%). Finally, we show BIKER robust to measurement errors implicit in the SWOT data. With SWOT set to launch in late 2022, we suggest that BIKER will be useful in mapping the global-scale spatiotemporal dynamics of fluvial gas exchange in large rivers.

## 1 Introduction
Natural systems play a critical role in the budgeting and accounting of the global carbon cycle under climate change. Following @colePlumbingGlobalCarbon2007, the global river network is recognized to emit substantial amounts of carbon to the atmosphere via evasion (gas exchange driven by a diffusion gradient and water turbulence), in addition to their long understood role in transporting carbon to the oceans via downstream advection. Current estimates of total carbon dioxide evasion ($FCO_2$) to the atmosphere from the global river network vary from 650-2000 Tg C/yr [*Liu accepted*; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013], with 167 Tg-C/yr coming from mountain streams alone [@horgbyUnexpectedLargeEvasion2019]. Despite rivers' incredibly small percentage of the global land surface [0.67%- *Liu accepted*], this C flux is on par with the total oceanic $CO_2$ uptake rate [@gruberOceanicSinkAnthropogenic2019; @horgbyUnexpectedLargeEvasion2019] and the global forest carbon uptake rate [@panLargePersistentCarbon2011].

River $CO_2$ evasion is increasingly better constrained and is clearly a critical component of the global carbon cycle. Equation 1 represents this riverine flux given $\delta CO_2$ (the gas concentration gradient between the water $CO_{2_{water}}$ and the air $CO_{2_{air}}$) and the gas exchange velocity *k*. Consult Appendix A for variable nomenclature used throughout this study.

$$FCO_2=k* \delta CO_2 \mathbf{(1)}$$

The structure of equation 1 necessitates that calculations of this flux are highly sensitive to measurements/estimates of *k*. However, *k* can only be directly calculated via a known gas concentration gradient, eddy-covariance measurements, or tracer additions to the stream [@hallGasExchangeStreams2020]. In trying to constrain the global fluvial $FCO_2$ flux across millions of rivers, this calculation is impossible, and necessitates the use of predictive models for *k* that are based on easily obtained river hydraulic properties. In that vein, there have been over 20 empirical models developed to predict *k* from river hydraulics, generally using some combination of mean velocity $\bar{U}$, shear velocity $U_*$, width $W$, depth $H$, and slope $S$ as predictors [@wangPhysicallyBasedScaling2021]. These models usually predict $k_{600}$, or *k* normalized by a Schmidt number *Sc* of 600. This is to remove the effect of water temperature and gas type from predictive models, as warmer waters and lower *Sc* numbers both increase gas exchange rates [@hallGasExchangeStreams2020]. Specifically, $k_{600}$ reflects the $k_{CO_{2}}$ at 20 degrees Celsius. Through this normalization, these models exclusively perturb the geophysical controls on gas exchange [@hallGasExchangeStreams2020].

These empirical models enable estimating a global $FCO_2$ flux ($FCO_2_{upscaled}$) from millions of rivers, but they also change the base parameters that ultimately control that aggregate estimate. That is, by making *k* a function of hydraulics, $FCO_2_{upscaled}$ is now a direct function of river hydraulics. This functional relationship is described in equation 2. It suggests that $FCO_2$ estimates are not only at the mercy of the accuracy and spatiotemporal resolution of $\delta CO_2$ as discussed previously, but also the accuracy and resolution of our river hydraulics estimates.

$$FCO_{2_{upscaled}}=f(k_{CO_2}, \delta CO_2)=f(\bar{U}, H, S, W, \delta CO_2) \mathbf{(2)}$$
There is a robust existing literature exploring spatiotemporal patterns in $CO_2_{water}$ [e.g. @ahoDistinctConcentrationdischargeDynamics2021; @ahoDifferentialResponseGreenhouse2019; @crawfordCO2TimeSeries2017; @peterScalesDriversTemporal2014; @ranDynamicsRiverineCO22017; @rocher-rosLandscapeProcessDomains2019; @raymondAtmosphericCO2Evasion2000; @liuHydrologicControlsPCO22018a]. This work has lead to recent river-reach explicit modeling of $\delta CO_2$ using global hydrography datasets at up to monthlytemporal resolutions [@horgbyUnexpectedLargeEvasion2019; @brinkerhoffLakeMorphometryRiver2021; @saccardiImprovingPredictionsStream2021; *Liu accepetd*], but an equivalently sophisticated representation of $k_{CO_2}$ is still lacking. As equations 1 and 2 dictate, $CO_2_{water}$ and $k_{CO_2}$ share the burden of calculating $FCO_2$ and therefore these next-generation models will have to contend with this discrepancy, which is directly dictated by the lack of direct measurements of global river hydraulics.

For example, to upscale $FCO_2$ globally, @raymondGlobalCarbonDioxide2013, @lauerwaldSpatialPatternsCO22015, and @horgbyUnexpectedLargeEvasion2019 all relied on $k_{CO_2}$ values indirectly estimated using mean annual streamflow models and scaling equations to predict hydraulic terms, while @borgesGloballySignificantGreenhousegas2015 used a combination of the above method and a constant $k_{CO_2}$ in space and time to upscale over Africa. In all of these foundational studies, the temporal dynamics of the gas exchange velocity (and thus $FCO_2$) were ignored because of hydraulic data limitations. It has been shown at the field-scale that temporal dynamics of gas exchange can vary widely from site to site [@wallinSpatiotemporalVariabilityGas2011], but it has remained impractical to obtain temporally explicit $k_{CO_2}$ at continental-to-global scales. More recently, *Liu etal accepted* performed a first assessment of monthly temporal dynamics in the global river $CO_2$ flux, though they relied on modeled streamflow and used the same model for $k_{600}$ as previous studies [@raymondGlobalCarbonDioxide2013] to achieve this.

@wangPhysicallyBasedScaling2021 recently attempted to address this global *k* problem by simulating $k_{O_2}$ in 35 rivers of many sizes (widths ranging from 0.23–349m) using a stream metabolism model [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets to infer what *k* must have been to produce their observations. They then compared this simulated dataset against direct measurements of *k*, finding similar performance and parameter values for process-based models of gas evasion. However, they were still limited by a lack of direct hydraulic measurements and had to rely on scaling equations to estimate river depth and velocity. Even though approaches like this are incredibly useful for expanding our mechanistic understanding of gas exchange, it is less useful for global upscaling purposes as it relies on highly detailed in situ DO data for every river [@hallGasExchangeStreams2020].

In theory, the discrepancy between the quality of our $CO_2_{water}$ and $k_{CO_2}$ estimates could be alleviated if direct hydraulics measurements (and in turn $k_{CO_2}$ via equation 2) were available at the global scale at a sufficient temporal resolution. This is turn would also address the uncertainty regarding continental-to-global scale temporal dynamics of gas exchange noted earlier. Conveniently, these data will soon be available via the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission. SWOT is expected to launch in 2022 and provide the world's first direct measurements of global water surface extent and elevation (and therefore water surface slope) at novel temporal resolutions. SWOT is a wide swath radar interferometer and will sample rivers every 1 to 7 days per 21 day repeat cycle, measuring rivers wider than 100m with a goal of expanding this to rivers at least 50m wide [@biancamariaSWOTMissionIts2016]. Via its direct hydraulic measurements, SWOT is expected to usher in a sea change in global-scale hydrology, and could similarly transform fluvial biogeochemsitry if techniques are developed to ingest SWOT data and infer *k*. In this context, we borrow tools from fluvial geomorphology and existing SWOT algorithms to answer the following two questions:

  - Can we develop a physically-based hydraulic model for $k_{600}$ unique to SWOT-observed rivers?

  - Can we exploit such a model to infer $k_{600}$ (and its uncertainty) solely from SWOT observations?

To answer this first question, we use hydraulic geometry [the fundamental geomorphic relationships between streamflow and channel shape- @leopoldHydraulicGeometryStream1953; @gleasonHydraulicGeometryNatural2015] to develop a process-based model for large-river (here defined a wider than 50m) gas exchange. We then take these findings and explore the second question by implementing this hydraulic model, which in turn defines $k_{600}$, within an algorithm named BIKER ('Bayesian Inference of the $k_{600}$ Evasion Rate'). The goal of BIKER is to require no in situ inputs of any kind (although in situ data could be ingested and should improve results) such that it is globally implementable on any SWOT-observable river. We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers (as SWOT has not yet launched) under both a 'no-measurement-error' and a 'measurement-error' scenario to explore algorithm robustness to the error tolerances expected in the satellite's measurements. Finally, we also couple BIKER's $k_{600}$ estimates with $\delta CO_2$ to predict gas fluxes and compare these against established literature methods that are reliant on in situ hydraulic measurements.

This paper is split into two distinct parts: gas-exchange theory/model development (section 2) and BIKER algorithm development/validation (section 3). Section 3 is fundamentally dependent on the results presented in section 2, therefore section 2 presents both theory and results. Both sections detail the data used. We conclude with a discussion (section 4) that encompasses both gas exchange theory and remote sensing. We also provide a flowchart detailing the entire study as Figure 1.

![Figure 1: Flowchart detailing the BIKER algorithm. Dashed lines indicate workflow to calculate the observed data that we validate against. See Appendix A for variable nomenclature used in this study.](cache/flowchart_2.jpg)

## 2 Prediciting $k_{600}$ from large-river hydraulic geometry
To predict $k_{600}$ in the large rivers that SWOT will observe, we start from established process-based models for *k*, impose hydraulic assumptions that are valid in SWOT-observable rivers, and obtain a suite of models we empirically test. Following a description of the data (section 2.1), we outline established models (section 2.2) and then step through our hydraulic assumptions to arrive at equations that are compatible with SWOT measurements (section 2.2). We then validate the model (section 2.3)

### 2.1 Data
We develop our hydraulic model for $k_{600}$ using multiple datasets of measured $k_{600}$ and stream hydraulics collected from the literature. In total, there are 763 measurements of $k_{600}$, with 701 of these measurements previously gathered by @raymondScalingGasTransfer2012 and @ulsethDistinctAirWater2019. The remaining measurements come from @churchillPredictionStreamReaeration1964 and @owensReaerationStudiesStreams1964. See Table S1 for a complete list of the studies that collected these measurements. All $k_{600}$ measurements come from tracer studies and thus define $k_{600}$ and the rivers' mean hydraulic properties at the reach scale.

In addition to hydraulics measured alongside $k_{600}$ and reported above, we expand our dataset of stream hydraulics using a previously published compilation of in situ hydraulic measurements [@brinkerhoffReconcilingAtaStationAtManyStations2019]. That dataset contains over 530,000 unique measurements of river channel velocity, width, and discharge from across the continental United States, originally made to calibrate United States Geological Survey (USGS) streamgauge rating curves and made public by the USGS.

### 2.2 Process-based hydraulic modeling of river *k*
*k* scales with near-surface turbulence in turbulent streamflows [@hallGasExchangeStreams2020], and extensive field and laboratory experiments have converged on the 'small-eddy model' as derived by @lamontEddyCellModel1970 and empirically anticipated by @calderbankContinuousPhaseHeat1961. This model scales *k* via the smallest-scale turbulent eddies and has been repeatedly empirically validated in freshwater systems [e.g. @katulStructureFunctionModel2018; @lorkeUnifiedScalingRelation2006; @zappaEnvironmentalTurbulentMixing2007; @zappaVariationSurfaceTurbulence2003; @moogStreamReaerationNonuniform1999; @vachonRelationshipNearsurfaceTurbulence2010; @tokoroHighGastransferVelocity2008; @wangPhysicallyBasedScaling2021]. The small-eddy model is provided as equation 3, where $\epsilon$ is the dissipation rate of near-surface turbulence, $v$ is the kinematic viscosity, and *Sc* is the Schmidt number.

$$k\propto Sc^{-1/2}(v\epsilon)^{1/4} \mathbf{(3)}$$
Some laboratory and field observations additionally suggest that open channel flows with small bed roughness do not exhibit homogeneous surface dissipation at their air-water interface [@moogAirWaterGasTransfer1999; @talkeTurbulentKineticEnergy2013]. Given this observation, @moogAirWaterGasTransfer1999 proposed an extension to the small-eddy model, additionally scaling $k_{600}$ using a shear Reynold's number formulation. This is equation 4 and is referred to here as the 'Reynolds extension' model. The Reynolds model is hypothetically useful in low-turbulence flows where a relative lack of large-scale eddies effectively 'filter out' the number of small-eddies that actually reach the interface and initiate gas exchange [@talkeTurbulentKineticEnergy2013]. While scaling *k* via a shear Reynold's formulation is sometimes done to parameterize wave-breaking gas exchange models in the open ocean  [@brumerWaveRelatedReynoldsNumber2017; @zhaoDependenceWhitecapCoverage2001; @zhaoEffectWindWaves2003], it is infrequently done in rivers. In the context of BIKER, we chose to test this model because large, SWOT-observable rivers are generally the smoothest, least-turbulent flows along the stream-to-ocean continuum where small eddies might not reach the surface. Further, to our knowledge this Reynolds extension model has never been empirically tested in predicting river *k*, aside from confirming that large-scale eddies differentially move turbulence to the surface in a large river [@talkeTurbulentKineticEnergy2013].

$$k \propto Sc^{-1/2}(v\epsilon)^{1/4}(\frac{U_*H}{v})^{3/8} \mathbf{(4)}$$
Equations 3 and 4 both rely on $\epsilon$, which is non-trivial to measure. When working at large scales, a commonly used $\epsilon$ model assumes that all turbulence is generated at the bed and transported to the air-water interface via the log-law-of-the-wall via equation 5 [@lorkeUnifiedScalingRelation2006; @nezuTurbulenceOpenChannel1993]. Another approach models *k* via 'form-drag dissipation' (equation 6) which is equivalently the total stream power per unit mass water. This normalized stream power captures the bulk frictional resistance (and thus energy dissipation) via channel banks, meanders, bars, etc. that is unique to fluvial systems [@moogStreamReaerationNonuniform1999]. Authors have since shown that equation 6l can reasonably predict *k* in rivers and streams [@raymondScalingGasTransfer2012; @ulsethDistinctAirWater2019; @wangPhysicallyBasedScaling2021].

$$\epsilon = \epsilon_S = \frac{U_*^3}{H} \mathbf{(5)}$$
$$\epsilon = \epsilon_D = gS\bar{U} \mathbf{(6)}$$

### 2.3 Deriving a large-river $k_{600}$ model
Given the theoretical context provided in Section 2.2, we now turn to SWOT-observable systems specifically. Rivers and streams change predictably along their longitudinal profile from headwater to ocean, and we can exploit the hydraulic geometry of large rivers at the end of this continumn to estimate *k* in SWOT-observable systems. In general, as river size increases, channels become more rectangular, their shapes elongate (becoming wider quicker than they become deeper) and their hydraulic radii begin to approximate their mean flow depth, i.e. $R_h \approx H$. This is a common assumption in hydraulic and geomorphic modeling of large rivers, and the average flow in a SWOT-observable river flow has a $\frac{R_h}{H}$ ratio of `r HG_swot$mean` (n = `r HG_swot$n`; see Text S1 for how we built this dataset). We refer to these rivers as 'hydraulically-wide'.

We therefore assume that all SWOT-observable rivers are hydraulically-wide. We then test the performance of four models for predicting *k* in hydraulically-wide channels: the four unique combinations of equations 3-4 and equations 5-6 outlined above. The overall goal was to reduce the equations down to their fundamental parameters, identifying which terms are SWOT observable and limiting the number of terms not directly measurable via SWOT. While the complete model derivations and results for the four models are provided in Text S2 and Figure S1, the final and best-performing model is presented below.

We impose $\epsilon=\epsilon_D$ and $R_H=H$ on the Reynolds extension model and arrive at equation 7 (with statistical coefficient $\beta_1$). Equation 7 thus define gas exchange velocity solely as a function of slope, mean flow depth, and mean flow velocity. This is theoretically valid only in a hydraulically-wide channel.

$$k_{600}=\beta_1(gS)^{7/16}\bar{U}^{1/4}H^{9/16} \mathbf{(7)}$$
### 2.4 Model validation
With equation 7 derived, we now empirically its strength of fit for hydraulically-wide river flows. We validate on the dataset of in situ measurements of $k_{600}$, after filtering for measurements made in hydraulically-wide channels, which was operationally defined as flows whose hydraulic radius was within 1% of their mean flow depth. All told, this amounts to `r n` direct measurements of hydraulically-wide $k_{600}$ and stream hydraulics to test with. The model is assessed via the coefficient of determination ($r^2$) and plotted in Figure 2. Note that Figure 2 axes are plotted in logarithmic space only for visualization: model fit and validation (via $r^2$) were calculated in linear space as their models dictate.

![Figure 2: Empirical testing of our large-river $k_{600}$ model on 166 measurements made in hydraulically-wide rivers. Note that axes are plotted in logarithmic space just for visualization: model fit and validation (via the coefficient of determination $r^2$) were calculated in linear space.](cache/k600_theory/k600Plot.jpg)

Our model explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2, 2) * 100`% of variation in observed $k_{600}$ in hydraulically-wide rivers and accurately captures the scaling dynamics of observed $k_{600}$ as well. While the individual residuals can be quite large, the overall scaling of $k_{600}$ with river hydraulics is strongly captured. Compared to the other models tested (Figure S1), there is also less bias in the estimates thanks to the addition of the Reynold's number scaling for low-turbulent flows. The success of this model in hydraulically-wide channels provides us with a strong physical-model for gas evasion. The river hydraulics terms in equation 7 can either be directly measured or reasonably inferred from SWOT measurements, effectively opening the door for remotely sensing the gas exchange velocity. This is explored next.

## 3 Exploiting equation 8 to remotely sense gas exchange velocity
We have shown that scaling $k_{600}$ via equation 8 explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2,2)*100`% of the variation in $k_{600}$ in hydraulically-wide rivers, which includes nearly all SWOT-observable rivers. Further, equation 8 has only three non-remotely-sensible terms: $k_{600}$, mean flow depth, and mean flow velocity. Conveniently, techniques to simultaneously infer mean flow depth and velocity from SWOT data (among other parameters) have been established over the last decade to infer streamflow from SWOT's measurements [e.g. @andreadisConstrainingAssimilationSWOT2020; @brinkerhoffConstrainingRemoteRiver2020; @durandEstimatingReachaveragedDischarge2014; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014a; @garamboisVariationalEstimationEffective2020; @brissetAssimilationAltimetricData2018; @hagemannBAMBayesianAMHGManning2017; @larnierRiverDischargeBathymetry2020a; @oubanasDischargeEstimationUngauged2018]. For BIKER, we follow the work developed by @hagemannBAMBayesianAMHGManning2017, @brinkerhoffConstrainingRemoteRiver2020, and @durandEstimatingReachaveragedDischarge2014 to infer $k_{600}$, chanel depth and velocity from SWOT observations using a modified form of Manning's equation. Following a description of the data used (section 3.1), we detail algorithm development and experimental design (section 3.2) and then we present the validation results (section 3.3).

### 3.1 Data
![Figure 3: Map of the 47 hydraulic models and 1 timeseries of CO2 samples used in this study. Note that hydraulic model locations are approximate as some of the models are not geo-referenced.](cache/maps/map3.jpg)

To validate BIKER, we cannot use actual SWOT measurements as SWOT has yet to launch. In the hydrology literature, it is therefore standard practice to benchmark SWOT-related algorithms on "SWOT-like" data [@durandIntercomparisonRemoteSensing2016]. We use 47 SWOT-simulated rivers for validation, where these simulated rivers are simply reach-averaged hydraulic model outputs where the water surface heights, slopes, and widths are labelled as RS observations and are used as the sole inputs to BIKER. These data were published by @frassonExploringFactorsControlling2021 and @durandIntercomparisonRemoteSensing2016 as benchmarking datasets to explore remote sensing of discharge (RSQ) algorithm performance for the SWOT mission. These datasets are created using standard hydraulic models forced with known inflows and measured bathymetry to model the hydraulic response of the rivers. These `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers represent the 'no-measurement-error' scenario that reflects a best-case scenario that is used here as a first validation of BIKER. `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` of these rivers are then further prepared by @frassonExploringFactorsControlling2021 to more closely mimic SWOT by adding realistic radar errors and sampling along the satellite’s future ground track. Specifically, they account for 1) the SWOT error budget itself [CITE}], 2) thermal noise, and 3) radar layover error, witht the latter two being modeled by [@durandHowWillRadar2020].

*COLIN: @frassonExploringFactorsControlling2021 also developed a sophisticated error model that mimics the effects of the atmosphere, topography, and water surface geometry on SWOT measurements (16/47 rivers) to add realistic radar layover errors that depend on the time and place of observation. The resulting data are the most realistic synthetic SWOT river data in the literature. The approximate locations of these rivers are plotted in Figure 1.*

**According to Durand 2016 and Frasson, there is no measurement error added to the non-error scenario rivers. Mike had a small bit of Gaussian noise to pepsi 1 but he doesn't seem to count it has true measurement error.**

For the $CO_2$ evasion and carbon emissions calculations (section 3.1.3), we use 26 bi-weekly dissolved $CO_2$ samples collected by @beaulieuControlsGasTransfer2012 at one location in the Ohio River for one calender year from 2008-2009 (Figure S2). Note that this $CO_2$ data is for the Ohio River only but was applied to all rivers (which includes multiple sections of the Ohio River). Therefore, the raw carbon emissions estimates presented in this paper are meaningless in the context of actually measured carbon emissions from these rivers. However, we are interested in the effect of BIKER $k_{600}$ errors on eventual fluxes, and therefore applying these 'unit' $CO_2$ values allows such a compariosn by providing a realtistic timeseries.

### 3.2 Section 3 methods
#### 3.2.1 BIKER
BIKER, and Bayesian inference in general, starts from Bayes rule (equation 8), where $\Theta$ is some set of non-remotely-sensible parameters we want to solve for (including $k_{600}$), *x* is the observed data, $f(x|\Theta)$ is the sampling model where data are conditional on the parameters, and $p(\Theta)$ is the joint prior distribution of the parameters. Therefore, we are interested in solving for $p(\Theta|x)$, or the 'posterior' distribution. Note that $p(x)$ is usually computationally intractable to integrate exactly, but Bayesian inference tools require only the proportionality to be specified: $p(\Theta|x) \propto f(x|\Theta)p(x)$. Sampling algorithms are then used to approximate the posterior distribution, as is done in BIKER.

$$p(\Theta|x) = \frac{f(x|\Theta)p(\Theta)}{p(x)} \mathbf{(8)}$$
The heart of BIKER is its reformulation of the $k_{600}$ model (equation 9) as a Bayesian sampling model that is conditional on the non-remotely-sensible parameters (i.e. $f(x|\Theta)$). This is similar to the 'McFLI' ('Mass-Conserved Flow Law Inversion') logic used in some SWOT RSQ algorithms [@gleasonTrackingRiverFlows2017]. To start, we write $k_{600}$ as a function of SWOT-observables $W$ and $S$. This algebra is carried out using equation 7, the fitted value for $\beta_1$ from Figure 2 (`r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$coef,2)`), and Manning's equation for mean flow velocity ($\bar{U}=\frac{1}{n}R_h^{2/3}S^{1/2}$). Following section 2.3, we continue to assume that the channel is hydraulically-wide ($R_h = H =\frac{A}{W}$). To leverage additional SWOT data, the wetted channel area *A* is further split into the the SWOT-unobservable portion $A_0$ and SWOT-observable portion $dA$ following @durandEstimatingReachaveragedDischarge2014 and @hagemannBAMBayesianAMHGManning2017 where $dA_{it} = \sum_{t':W_{t'}\le W_t}{W_{t'} \delta H_{t'}}$ for cross-section *i* and timestep *t*.

All of this algebra simplifies to equation 9. Conveniently, $k_{600}$ as measured by tracer additions to a stream is inherently a reach-scale quantity (in a mass-conserved reach). Therefore, equations 7 and 9 both yield a reach-scale $k_{600}$ (i.e. $k_{600_i}=k_{600}\forall i$). This lowers the number of parameters BIKER must infer and makes the problem much better constrained.

$$\begin{eqnarray} k_{600_t} = 62.82 g^{7/16}S_{it}^{9/16}\frac{1}{n_i}^{1/4}(\frac{A_0_i+dA_{it}}{W_{it}})^{35/48} \end{eqnarray} \mathbf{(9)}$$
Next, equation 9 is written as a Bayesian sampling model, in which all of the SWOT observations are sampled from the unknown model parameters ($A_0_i$, $k_{600_t}$, and $n_i$). This is equation 10 after describing everything as normal distributions of the log-transformed terms. $\sigma_{k_{600}}^2$ refers to the total uncertainty implicit in equation 9 This uncertainty arises from parameter uncertainty in equation 9, Manning's equation, and the rectangular channel assumption.

$$\left[log(62.82) + \frac{7}{16}log(g) + \frac{9}{16}log(S_{i,t}) - \frac{35}{48}log(W_{i,t})\right] \sim normal\left(\left[log(k_{600_t})+\frac{1}{4}log(n_i)-\frac{35}{48}log(A_0_i+dA_{i,t})\right], \sigma^2_{k_{600}}\right) \mathbf{(10)}$$
Equations 8 and 10 also necessitate that we specify prior distributions for the parameters $A_0_i$, $k_{600_t}$, and $n_i$. Prior distributions, described by their hyperparameters, formalize the a priori estimates and uncertainties for the non-remotely-sensed terms. More intuitively, BIKER priors represent our 'prior river knowledge' of what $A_0_i$, $k_{600_t}$, and $n_i$ probably are for some river since they cannot be directly remotely sensed. This is equivalent to the 'empirical Bayes approach' to Bayesian inference [@hoffFirstCourseBayesian2009]. Our goal in prior specification was to rely on absolutely no in situ information such that we could run this method on any river on Earth solely using SWOT observations. In theory, more informed priors via various a priori information about a specific river would improve BIKER performance, but here we chose to test the fully generalized algorithm. Therefore, the validation presented here is a 'worst-case scenario', wherein BIKER performance should improve with better prior information on the river. In that context, we used a variation of the prior specification method developed by @brinkerhoffConstrainingRemoteRiver2020, who developed geomorphic 'river types' with distinct prior sets for $A_0_i$ and $n_i$. These priors are assigned to a river solely using SWOT data *W* and *S*, therefore meeting our needs for complete global implementability. Prior assignment for $k_{600_t}$ was developed similarly (all prior specifications are elaborated on in Texts S3 and S4).

With the sampling model described ($f(x|\Theta)$ = equation 10) and priors $p(x$) specified (Texts S3 and S4), a joint posterior distribution conditional on the SWOT observations ($p(\Theta|x) \propto f(x|\Theta)p(x)$) is therefore also specified. To approximate this distribution, we use a Markov Chain Monte Carlo (MCMC) algorithm implemented using the Stan probabilistic programming language. Specifically, Stan uses a Hamiltonian Monte Carlo sampler which reduces computation time relative to other sampling algorithms [@hagemannBAMBayesianAMHGManning2017].

#### 3.2.2 BIKER validation
We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers (section 3.1) using daily simulated hydraulics. We also re-validate BIKER on the `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` rivers with [@frassonExploringFactorsControlling2021]'s SWOT error model to corrupt the hydraulics to mimic realistic SWOT measurements (widths, heights, and slopes). BIKER is unique in that it can provide a timeseries of $k_{600}$- for each SWOT observation, it yields a unique $k_{600}$. There are, to our knowledge, no datasets of $k_{600}$ over time approaching the temporal density of our simulated SWOT rivers. We therefore apply equation 7 as validated in Figure 2 to specify $k_{600}$ given the true hydraulics of each case. Remember that SWOT cannot observe below the water surface and therefore cannot measure $\bar{U}$ or *H* (hence the need for equation 9), and that all SWOT observations contain errors in both space and time (hence equation 10). We acknowledge that there is error in equation 7 as shown in Figure 2, but this error can be explicitly parameterized in our Bayesian system. Therefore, the BIKER validation presented here is an exercise to see how well the imperfect and partial SWOT observations can infer $k_{600}$ given hydraulic assumptions in equation 9 and uncertainty in the data itself.

We validate BIKER as a timeseries of $k_{600}$ for each river using the BIKER posterior means. Our error metrics consider the timeseries nature of the problem, are defined in Table S2, and consist of the following:

- Correlation coefficient *r* to quantify accuracy of BIKER's temporal dynamics
- Root mean square error normalized by the observed mean (NRMSE) and prediction bias normalized by the observed mean (rBIAS) to assess bias
- Kling-Gupta Efficiency (KGE). KGE is frequently used to assess streamflow prediction and simultaneously assesses accuracy in both bias and dynamics. While a value greater than -0.41 means the model outperforms a uniform prediction of the mean [@knobenTechnicalNoteInherent2019], generally KGE scores are interpreted as being meaningful in ungauged settings if > 0.

#### 3.2.3 Carbon emissions validation
It is one thing to accurately model the temporal dynamics of $k_{600}$ as above, but researchers are often most interested in the actual carbon emitted from river to atmosphere. Per equation 2, this is done using river hydraulic models to estimate $k_{CO_2}$ and in turn $FCO_{2_{upscaled}}$. However, streamflow data and/or model outputs are more readily modeled at the global scale than river channel geometry, and so upscaling models usually predict $\bar{U}$ and *H* as functions of streamflow (*Q*) using hydraulic geometry (HG) scaling relationships. This effectively reduces equation 2 down to equation 11. It is worth stressing that these upscaling workflows rely on in situ streamflow records and/or high-quality streamflow routing outputs (unlike BIKER, which only requires a river be SWOT-observable).

$$FCO_{2_{upscaled}}=f(k_{CO_2}, \delta CO_2)=f(Q, \delta CO_2) \mathbf{(11)}$$

In that context, we sought to compare BIKER against established upscaling methods from the literature for upscaling carbon emissions from our 47 rivers. So, we pair the 26 biweekly $CO_2$ and water temperature samples from @beaulieuControlsGasTransfer2012 (section 3.1, figure S2) with every 14th set of daily SWOT observations (as the $CO_2$ data is bi-weekly). We then calculate $FCO_2$ using equation 1, an atmospheric $CO_{2_{air}}$ of 390 uatm, and four different models for $k_{600}$, which are explained in full in Table S3. In brief, the four methods are : 1) BIKER-inferred $K_{600}$, 2-3) $k_{600}$ calculated using equation 7 and HG models for depth and velocity from @raymondGlobalCarbonDioxide2013 and @raymondScalingGasTransfer2012, and 4) HG models fit to our hydraulics data from section 2.2.1. These models are referred to as 'BIKER', 'Raymond 2013', 'Raymond 2012', and 'Brinkerhoff 2019', respectively. These latter three approaches use the in situ streamflow record as their input, while BIKER uses only SWOT observations. This experiment lets us perturb two things: 1) how sensitive upscaled emissions estimates are to the HG models that predict river hydraulics from streamflow (as everything else is held constant) and 2) how a fully-remote technique like BIKER compares to in situ techniques like the other three.

Next, equation 12 is used to obtain a temperature and $CO_2$ specific gas exchange velocity from the flour flavors of $k_{600}$, where the Schmidt number *Sc* was estimated following @raymondScalingGasTransfer2012. We then estimate a yearly total carbon emission rate (via $CO_2$ evasion) by applying each river's mean $FCO_2$ over the river's surface area and summing all of these emission rates.
$$k_{CO2, temp}=k_{600}*(Sc/600)^{-1/2} \mathbf{(12)}$$

### 3.3 Section 3 results
#### 3.3.1 BIKER
Figure 4 plots a representative set of the `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` timeseries plots of predicted and observed $k_{600}$, assuming no SWOT measurement error. Two rivers each rivers were randomly sampled from the three tertiles of river *KGE* scores (Table S2). Consult Figure S3 for all 47 timeseries plots and Figure S4 for the 16 rivers with the [@frassonExploringFactorsControlling2021] error model. Note that the y axis is normalized by maximum observed values to compare across rivers. In general, the temporal dynamics of $k_{600}$ are reproduced quite well by BIKER, with the highs and lows of evasion correctly inferred by BIKER. Notably, there is sometimes positive bias in the estimates (e.g. Seine River). Some rivers are inferred nearly perfectly (e.g. Kanawha, Ohio Section 2, Wabash rivers) while others perform more poorly (e.g. Ash Slough, Ohio), though again much of this error stems from errors in inferring the magnitude of $k_{600}$ rather than its temporal dynamics.

![Figure 4. Representative river timeseries plots of $k_{600}$. Orange is calculated via observed hydraulics, while green is BIKER and uses SWOT measurements as its sole input. The green ribbon indicates the 95% CIs for the BIKER predicted values. Rivers randomly sampled from the three tertiles of KGE scores. (a-b) are poorest performing rivers, (b-c) are in the middle, and (e-f) are the best performing rivers. Y axis is normalized by maximum observed values to compare visually.](cache/validation/timeseries_noerr_short.jpg)

We calculate performance metrics following section 3.2.2 and Table S2. These are presented in Figure 5, which plots the scores for the 47 rivers with no measurement errors as empirical cumulative density functions (eCDFs). Median river $KGE$ is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$KGE, na.rm=T),2)` and median river *r* is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$r, na.rm=T),2)`. This indicates very strong inference of each river's temporal $k_{600}$ dynamics given that absolutely no in situ information is being used to predict $k_{600}$. NRMSE has a median score of `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$NRMSE, na.rm=T),2)` and median rBIAS is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$rBIAS, na.rm=T),2)`, highlighting that many rivers which have a significant positive bias (Figure 4 also confirms this). Taken in aggregate, Figures 4-5 indicate that BIKER is quite good at capturing temporal dynamics in $k_600$, however there is sometimes positive bias in its estimates.

![Figure 5. Performance metrics by river, plooted as empirical cummulative density functions (eCDFs). Each subpanel is labelled by its performance metric (Table S2). Dashed lines denote medians.](cache/validation/validation_by_river.jpg)

Figure 6 compares BIKER results under the 'no-error' and 'error' scenario for the 16 rivers that have the @frassonExploringFactorsControlling2021 imposed on their SWOT observations (section **something**). For *r*, *rBIAS*, and *NRMSE*, BIKER performs nearly identically, regardless of inherent measurement error in the SWOT observations: points are scattered along the 1:1 line and their regressions also mirror this 1:1 line. For *KGE*, performance is also largely similar except for one river which get significantly worse and skews the regression line (Seine Upstream river). Overall, these results strongly suggest that BIKER will be robust to the measurement errors that will be implicit in SWOT's observations of river width and slope. Given these results, we deem that SWOT measurement error does not exert a significant influence on BIKER performance and so the results presented for the rest of the manuscript assume no measurement error in order to use all `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers.

![Figure 6: Comparison of BIKER performance when considering measurement error in the SWOT observations for 16 rivers. Each subpanel is labelled by its performance metric (Table S2). Black line denotes linear regression while dashed grey line is the 1:1 line.](cache/validation/errorPlot.jpg)

Given the promising *r* scores across all 47 rivers in Figure 5, we further explore BIKER's ability to infer temporal trends in $k_{600}$. Figure 7a plots a histogram of the slopes of linear regressions between BIKER-predicted $k_{600}$ and observed $k_{600}$. A slope of 1 indicates that BIKER correctly infers the daily flucuations in gas exchange (even if there is bias baked into the estimates), while a slope well above or below 1 indicates incorrect inference of the temporal dynamics of $k_{600}$. Figure 6a shows that the majority of rivers' slopes approximate 1 (median slope: `r round(median(results_dynamics$Slope),2)`), indicating strong inference of temporal gas exchange trends. A handful of rivers to have slopes well above or below 1, though the vast majority are at or just below 1.

![Figure 7: a) Histogram of the slopes of linear regressions between BIKER estimates and observed $k_{600}$. Slopes equaling 1 indicate correct inference of the temporal trends in daily $k_{600}$ (regardless of bias in the estimates). b) Empirical cummulative density distributions of the coefficient of varation of prior, posterior, and observed $k_{600}$ per river. If BIKER is correctly inferring temporal variation in $k_{600}$, the posterior eCDF will approximate the obersved eCDF (as it does).](cache/validation/validation_temporal.jpg)

Finally, we explore whether BIKER improves upon the prior with respect to temporal variation in $k_{600}$, here defined using the coefficient of variation ($CV=\frac{SD}{mean}$). Figure 6b plots eCDFs of observed, prior, and posterior $k_{600}$ *CV*, strongly showing that BIKER improves upon the prior information available to us about $k_{600}$. Posterior $k_{600}$ *CV* approximates the observed eCDF and sufficiently 'closes the gap' between our prior knowledge on $k_{600}$ temporal variation and the actually observed variation in $k_{600}$. It must be stressed that this validation was performed using a relatively uninformed prior on $k_{600}$ (Text S3) and that this gap may not be as large in many settings with better prior knowledge. However, Figure 6b confirms that Bayesian techniques can be used to infer temporal variation in $k_{600}$ when good prior or in situ data are not available.

#### 3.3.2 Carbon Emissions
Finally, we carry these $FCO_2$ calculations all the way to annual carbon emissions rates and compare BIKER against established in situ techniques. It is important to remember that only BIKER is completely globally implementable, while the other three models necessarily rely on having a streamflow record ( or high-quality routed streamflow) readily available. Therefore, the HG methods reflect their 'best-case scenarios' while BIKER reflects its worst case scenario, where priors are set entirely from SWOT observations and are generally the least informative they will ever be. We also stress again that the raw emissions rates here are largely meaningless as they are calculated from an identical $CO_{2_water}$ timeseries applied to every river. We are principally interested in relative differences between techniques employed.

Figures 8 compares the annual carbon emissions rate (via $F_{CO_2}$) from the rivers using BIKER posterior means and the three streamgauge-based HG models. Surprisingly, BIKER outperforms the gauge-based approaches, nearly correctly inferring the annual carbon emissions rate (`r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_BIKER',]$yearlyCFux, 2)` gG-C/yr for BIKER versus `r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_obs',]$yearlyCFux, 2)` gG-C/yr observed). The three HG models overestimate this emissions rate (`r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_raymond2013',]$yearlyCFux, 2)`, `r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_raymond2012',]$yearlyCFux, 2)`, and `r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_brinkerhoff2019',]$yearlyCFux, 2)`) for 'Raymond 2013', 'Raymond 2012', and 'Brinkerhoff 2019' respectively. Both 'Brinkerhoff 2019' and 'Raymond 2012' fall within the BIKER CIs and are reasonably close to the observed value, while 'Raymond 2013' overestimates the emission rate. BIKER's relatively stronger performance than the in situ HG models is elaborated on in section 4.3.

Thus, despite BIKER using absolutely no streamgauge or other in situ data like the other 3 methods do, it provides a strong upscaling estimate of the annual carbon emission rate across 47 rivers (Figure 8).  The in situ models represent their best case scenarios (perfect in situ discharge records) while BIKER represents its worst-case scenario (relatively uninformed priors). This means that BIKER's annual carbon emission estimate can only improve.

![Figure 8: Yearly carbon emissions via CO2 evasion across all rivers and accounting for surface area. Completely remotely-sensed methods are colored in red, in situ methods in purple, and the observed in orange.](cache/FCO2/FCO2_models.jpg)

## 4 Discussion
### 4.1 Gas exchange in hydraulically-wide rivers
Field studies of gas exchange in wide rivers have suggested that $k_{600}$ behaves differently than in steeper and smaller rivers [@ulsethDistinctAirWater2019; @beaulieuControlsGasTransfer2012; @alinPhysicalControlsCarbon2011; @wangPhysicallyBasedScaling2021]. While much work has focused on the small-stream side of the stream-to-river continuum, comparatively less work has been done in large systems. Here, we focus on the 'smooth-channel' end of the continuum, using a model for gas exchange that scales $k_{600}$ by both $\epsilon_D$ and the Shear Reynold's number that is specifically adapted for 'hydraulically-wide' rivers. This is empirically validated in Figure 3. Specifically, Figure S1 confirms that scaling $k_{600}$ with a combined small-eddy/Shear Reynold's formulation (equation 7) reduces positive bias in the model's predictions of the smallest $k_{600}$ values [where the relative decrease in turbulence reaching the surface is greater than the small-eddy model alone suggests, per equation 7's theoretical basis- @moogAirWaterGasTransfer1999]. Scaling $k_{600}$ via a shear Reynold's number alone is commonly done to parameterize wave-related gas exchange models in the open ocean [@brumerWaveRelatedReynoldsNumber2017; @zhaoDependenceWhitecapCoverage2001; @zhaoEffectWindWaves2003]. However to our knowledge, @moogAirWaterGasTransfer1999's specific model (imposing a differential surface turbulence theory on the small-eddy model) has never been empirically validated in rivers. Figure 3 provides this empirical verification for hydraulically wide channels, where it's theoretical basis should generally hold. Using our full dataset of $k_{600}$, we also observed this model breaks down when including non-hydraulically-wide rivers (as the theory dictates). Future tests should also explore other shear Reynold's scaling relations for gas exchange in rivers.

Crucially, we are not accounting for wind-driven gas exchange, which can exerts a significant influence on lake and ocean gas exchange and has been suggested to play an important role in wide-river gas exchange [@beaulieuControlsGasTransfer2012; @wangPhysicallyBasedScaling2021]. Similarly, none of the existing hydraulics-driven gas exchange models account for wind-driven gas exchange either. In that vein, BIKER's outputs can be interpreted as the 'gas exchange velocity in low-wind conditions', when surface turbulence is dominated by hydraulics rather than wind. That said, BIKER's flexible implementation is a good start towards eventually coupling hydraulics-driven gas exchange with wind-driven gas exchange. Such a model would likely take a form similar to the conceptual models proposed by @wangPhysicallyBasedScaling2021, chuWindStreamFlow2003, an/.or @plateReaerationOpenChannel1984. All three generally propose calculating both a wind-driven *k* and a hydraulics-driven *k* and then weighting each term via parameters. In the case of BIKER, this would necessitate additional parameters that would need to be known a priori for specific rivers. This is left to future work.

### 4.2 Towards remote sensing of global spatiotemporal dynamics of $k_{600}$ in large rivers
To date, the studies exploring the spatiotemporal dynamics of riverine gas exchange have arguably been held back by a lack of data. A few studies have investigated these dynamics, but they have been limited to individual rivers and/or limited field seasons [@hallAirWaterOxygen2012; @sand-jensenCO2DynamicsDanish2012]. For example, @wallinSpatiotemporalVariabilityGas2011 performed a preliminary analysis in northern Sweden relating cross-section specific temporal variability in gas exchange with channel slope, but they were limited to a mean of only 8 measurements per river in a single watershed. While a good start, this is obviously not sufficient for developing process-level understandings of gas exchange at the global-scale.

Therefore, inferring $k_{600}$ from SWOT data is an attractive option to address this. While BIKER will not directly measure $k_{600}$, it does robustly infer temporal trends in $k_{600}$ (Figures 4-7) which will no doubt be useful in large-scale analyses. **MOre go here**.

With that said, Figures 4, 5, S3, and S4 all highlight a substantial range of algorithm performances across rivers. These differences in performance are likely due to the representativeness of the priors used for that river. This makes sense as Section 2 has effectively reduced $k_{600}$ to a function of hydraulics that are nearly all directly measurable by SWOT. Any resulting bias in BIKER's predictions is likely attributable to bias in the priors used for the non-remotely sensed terms (equations 9-10). For SWOT discharge algorithms, authors have repeatedly shown that the 'quality' of prior hyperparameters plays a large role in the success of discharge inversions [@frassonExploringFactorsControlling2021; @andreadisConstrainingAssimilationSWOT2020; @brinkerhoffConstrainingRemoteRiver2020; @tuozzoloEstimatingRiverDischarge2019] and our results here further corroborate this finding. It should be stressed that a substantial portion of rivers from [@frassonExploringFactorsControlling2021] are canal-shaped in nature with different hydraulic properties than a natural river channel. Nearly universally, these rivers underperformed (Figures 4, S3), however it is impossible to isolate whether that is due to channel geomorphology or the fact that these canals also usually had limited data for the Bayesian inference (only around 12 days).

### 4.3 Coupling BIKER with upscaling workflows
Figure 8 confirms that BIKER is quite successful, when coupled with $CO_2$ data, at predicting annual upscaled carbon emissions from the river network. This encouraging result has two main implications for future work coupling remote sensing via SWOT with in situ data. First, figure 8 directly implies that BIKER will be useful when coupled with large-scale $CO_{2_water}$ models [e.g. **Liu etal in review**; @saccardiImprovingPredictionsStream2021]. These two models robustly predict reach-scale dissolved $CO_2$ concentrations using two different approaches: machine learning [**Liu et al in review**] and process-based reactive transport modeling [@saccardiImprovingPredictionsStream2021] but both models yield spatially and temporally consistent $CO_2$ estimates. Our promising results suggest that BIKER could provide additional (and directly observed) measurements of $k_{CO_2}$ to these models, thereby better informing model results through direct observations. This is likely accomplished via data assimilation which has proven useful for similar problems across geophysical modeling applications [@carrassiDataAssimilationGeosciences2018] and specifically for assimilating remotely sensed information into streamflow routing models [@ishitsukaCombiningOpticalRemote2020], of which @saccardiImprovingPredictionsStream2021 takes an approximate form of.

Figure 8 highlights a potentially large source of uncetainty on current river $CO_2$ upscaling estimates: the HG models employed to scale river channel hydraulics with streamflow. As previously stressed, the only difference between the three in situ HG models and the observed estimate in Figure 8 is the specific HG model employed to predict river depth and velocity. This means that drammatically different carbon emission estimates are obtainable depending on the training data used for these HG models. While BIKER performs similarly to these methods, care should be taken to use the HG models most approriate for your specific river geomorphology.

Finally, at the field-scale, Fgure 8 also confirms that we can couple BIKER with in situ gas concentration loggers to produce $FCO_2$ estimates at novel temporal resolutions in SWOT-observable rivers. High temporal fidelity datasets of dissolved $CO_2$ in SWOT-observable rivers now exist [e.g. @ahoHighestRatesGross2021] but no such similar datasets for $k_{CO_2}$ at equivalent temporal resolutions exist. For rivers unobservable by SWOT, we further suggest that BIKER could be run at the field scale (rather than via satellite-based altimeters like SWOT) using arrays of in situ pressure transducers to estimate water surface slope following recent similar work for estimating streamflow [@harlanDischargeEstimationDense2021]. Regardless, both approaches would produce $CO_2_{water}$ and $k_{CO_2}$ datasets at equivalent temporal resolution and allow us to directly calculate daily to sub-daily $CO_2$ fluxes from river reaches.

## 5 Conclusions
The $CO_2$ flux from the global river network is a major natural component of the global carbon cycle, on par with total forest uptake of $CO_2$ [@panLargePersistentCarbon2011]. Despite much interest and progress, there is still large uncertainty in the temporal dynamics of gas exchange and thus carbon emissions to the atmosphere. Much of this uncertainty stems from uncertainty in gas exchange velocity $k_{600}$ at the global scale. In this paper, we propose that the soon-to-launch SWOT satellite will provide enough hydraulic measurements to analyse the temporal dynamics of $k_{600}$ using SWOT measurements, and therefore allow for a global-scale analysis of spatiotemporal trends in large-river $k_{600}$ once SWOT launches. In that context, we develop 1) a hydraulic model for $k_{600}$ that is nearly SWOT observable and explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2,2)*100`% of variation in $k_{600}$, and 2) the BIKER algorithm to infer $k_{600}$ using no on-the-ground information. Validating on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers, we show show strong recovery of rivers' temporal $k_{600}$ dynamics and the the total annual carbon emission rate across all `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers. These results suggest BIKER can be used to infer global-scale, near daily estimates of fluvial gas exchange velocity once SWOT launches in 2022. This in turn will be useful in mapping the global-scale spatiotemporal dynamics of fluvial gas exchange in large rivers.

## 6 Acknowledgements
C.B. Brinkerhoff was funded on **FINESST**. The version of BIKER used in this study will be available upon acception. BIKER remains in active development and is available at https://github.com/craigbrinkerhoff/BIKER. All code to build and generate results, figures, and the manuscript is available at https://github.com/craigbrinkerhoff/RSK600. We thank Renato Frasson, Michael Durand, Amber Ulseth, and Jake Beaulieu for generously making their data available for this study. We also thank the entire SWOT discharge working group for their decade-plus body of work which inspired this study.

## 7 Apendix A
*Table A1: Variable description and notation for this study*

|**Notation**|**Description**|**Calculation (if not directly measured)**|**Units**|
|------------|---------------|-------------------------------|---------|
|$A$|Channel cross-sectional area|NA|$[L^2]$|
|$A_0$|Non-SWOT-observable cross-sectional area|NA|$[L^2]$|
|$\alpha$|Statistical parameter for equation 5 scaling relation|NA|$\frac{m^{1/2}s^{3/4}}{dy}$|
|$\alpha_1$|Statistical parameter for equation 6 scaling relation|NA|$\frac{m^{1/2}s^{3/4}}{dy}$|
|*Bulk carbon efflux*|carbon mass transport rate from river to atmosphere|NA|$[\frac{M}{T}]$|
|$\beta$|Statistical parameter for equation 7 scaling relation|NA|$\frac{m^{-1/4}s^{9/8}}{dy}$|
|$\beta_1$|Statistical parameter for equation 8 scaling relation|NA|$\frac{m^{1/2}s^{3/4}}{dy}$|
|$CO_{2_{water}}$|Water-side $CO_2$ concentration|NA|$[L]$|
|$CO_{2_{air}}$|Atmospheric-side $CO_2$ concentration|NA|$[L]$|
|$dA$|change in cross-sectional area|$\delta H_e*W$|$[L]$|
|$D_m$|Molecular diffusion coefficient|NA|$[\frac{L^2}{T}]$|
|$\epsilon$|Dissipation rate of near-surface turbulence|NA|$[\frac{E}{M*T}]$|
|$\epsilon_S$|log-law-of-the-wall model for $\epsilon$|$\frac{U_*^3}{H}$|$[\frac{E}{M*T}]$|
|$\epsilon_D$|Form-drag model for $\epsilon$|$gS\bar{U}$|$[\frac{E}{M*T}]$
|$FCO_2$|$CO_2$ evasion flux from river to atmosphere|NA|$[\frac{M}{L^2T}]$|
|$FCO_{2_{upscaled}}$|Upscaling estimate of the  global $CO_2$ evasion flux from river to atmosphere|NA|$[\frac{M}{L^2T}]$|
|$g$|gravitational acceleration|9.8|$[\frac{L}{T^2}]$|
|$H$|Mean flow depth|$A/W$|$[L]$|
|$H_e$|Water surface elevation|NA|$[L]$|
|$i$|Cross-section discretization within river reach|NA|NA|
|$k$|gas exchange velocity|NA|$[\frac{L}{T}]$|
|$k_{600}$|gas exchange velocity normalized to $Sc=600$|$(\frac{600}{Sc_k})^{-0.5}k$|$[\frac{L}{T}]$|
|$k_{600_{obs}}$|"observed" gas exchange velocity normalized to $Sc=600$|$k_{600_{obs}}=62.82(gS)^{7/16}\bar{U}^{1/4}H^{9/16}$|$[\frac{L}{T}]$|
|$n$|Manning's roughness coefficient|$\frac{R_h^{2/3}S^{1/2}}{\bar{U}}$|$[\frac{T}{L^3}]$|
|$\uprho$|Density of water|NA|$[\frac{M}{L^3}]$|
|$Q$|River discharge|NA|$\frac{L^3}{T}$|
|$R_h$|Hydraulic radius|$\frac{AW}{2H+W}$|$[L]$|
|$S$|River slope|NA|$unitless$|
|$Sc$|Schmidt number|$\frac{v}{D_m}$|NA|
|$\sigma^2_{k_{600}}$|BIKER uncertainty (posterior distribution standard deviation)|NA|$[L/T]$|
|$t$|timestep discretization within river reach|NA|NA|
|$\Theta$|Bayesian parameter set|NA|NA
|$\bar{U}$|Cross-sectional average velocity|$\frac{Q}{A}$|$[\frac{L}{T}]$|
|$U_*$|Shear velocity|$\sqrt{gSR_h}$|$[\frac{L}{T}]$|
|$\mu$|Viscosity|NA|$[P*T]$|
|$v$|kinematic viscosity|$\frac{\mu}{\uprho}$|$[\frac{L^2}{T}]$|
|$W$|Flow width|NA|$[L]$|
|$x$|Bayesian data set|NA|NA|

## References

### 2022-01-28 scraps.Rmd
manuscript v8 initial

@@
 BIKER is also more biased (median *rBIAS* of `r round(median(results_fco2_rivs[results_fco2_rivs$key == 'FCO2_BIKER',]$rBIAS),2)`) than either 'Brinkerhoff 2019' or 'Raymond 2012' models and similarly biased to the 'Raymond 2013' model (median *rBIAS* of `r round(median(results_fco2_rivs[results_fco2_rivs$key == 'FCO2_brinkerhoff2019',]$rBIAS),2)`, `r round(median(results_fco2_rivs[results_fco2_rivs$key == 'FCO2_raymond2012',]$rBIAS),2)`, `r round(median(results_fco2_rivs[results_fco2_rivs$key == 'FCO2_raymond2013',]$rBIAS),2)`, respectively).
+
+*Table 1: Four physically-based models for $k_{600}$ each derived under a different set of mass transport assumptions but all valid in hydraulically-wide rivers.
+Their derivations are outlined in sections 2.3.1 and 2.3.2 and their empirical tests are shown in section 2.4.*
+
+|**$k_{600}$ model**|**$\epsilon$ model**|**Final equation**|**Empirical test**|**Empirical fit ($r^2$)**|
+|-----------|--------------------|------------------|-----------------|----------------------------|
+|Small-eddy|Log-law-of-the-wall|$k_{600}=\alpha(gS)^{3/8}H^{1/8}$|Figure 3a|`r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eS',]$r2, 2)`|
+|Small-eddy|Form-drag dissipation|$k_{600}=\alpha_1(gS\bar{U})^{1/4}$|Figure 3c|`r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eD',]$r2, 2)`|
+|Reynolds-extension|Log-law-of-the-wall|$k_{600}=\beta(gS)^{9/16}H^{11/16}$|Figure 3b|`r round(modelsWIDE[modelsWIDE$name == 'reynolds-eS',]$r2, 2)`|
+|Reynolds-extension|Form-drag dissipation|$k_{600}=\beta_1(gS)^{7/16}\bar{U}^{1/4}H^{9/16}$|Figure 3d|`r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2, 2)`|
+
+Despite reasonable $r^2$ scores, both small-eddy models produce clearly biased estimates (Figures 3a, 3c).
+They systematically underestimate lower gas exchange velocities and overestimate higher ones, with the log-law-of-the-wall model (Figure 3a) performing marginally worse both visually and via $r^2$ (`r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eS',]$r2, 2)` versus `r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eD',]$r2, 2)` for Figure 3c).
+However, when this model is expanded to further retard the dissipation rate via the Reynolds extension model [equations 6 and 7- @moogAirWaterGasTransfer1999], the dynamics are more accurately captured (Figures 3b, 3d) and $r^2$ scores increase (`r round(modelsWIDE[modelsWIDE$name == 'reynolds-eS',]$r2, 2)` and `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2, 2)` for the log-law-of-the-wall and form-drag dissipation, respectively).
+Modeling dissipation via total form-drag rather than just via bed shear improves model performance regardless of $k_{600}$ equation used.
+This is likely because the majority of dissipation occurring in rivers comes not just from the streambed, but from bulk form drag via channel walls, bars, and meanders [@raymondScalingGasTransfer2012; @moogStreamReaerationNonuniform1999; @wangPhysicallyBasedScaling2021; @ulsethDistinctAirWater2019].
 

### 2022-01-28 supp_v8.Rmd
manuscript v8 initial

--- 2022-01-28 docs/supp_v7.Rmd
@@
 ## Contents
-This supplementary information contains 4 texts, 1 figure and 3 tables.
+This supplementary information contains 4 texts, 2 figures and 3 tables.
 Please consult https://github.com/craigbrinkerhoff/RSK600 for all code to build and generate results, figures, and the manuscript.
@@
 ## Text S2: Gas exchange model derivations
-In this text we provide the full algebra to arrive at the four physically-based gas exchange models in the main text (equations 3-6).
-Equation 3 is derived via equations S1-S5.
-Equation 4 is derived via equations S6-S7.
-Equation 5 is derived via equations S8-S13.
-Equation 6 is derived via equations S14-S18.
+In this text we provide the full algebra to arrive at the four physically-based gas exchange models we tested in this study.
 Consult Appendix A for all variable definitions.
@@
 ## Figure S1
-![Figure S1 Timeseries of the biweekly CO2 data from @beaulieuControlsGasTransfer2012.
+![Figure S1: Four $k_{600}$~hydraulics models tested in this study, with model assumptions noted in the sub-panel titles.
+Panel (d) is also presented in the main text as Figure 2.](cache/k600_theory/k600Plot_SI.jpg)
+
+## Figure S2
+![Figure S2 Timeseries of the biweekly CO2 data from @beaulieuControlsGasTransfer2012.
 Sampling took place 2008-2009 in the Ohio River (upstream of Cincinnati, Ohio, United States).
-Each point here was joined to the 11-day SWOT observations used in this study (section 2.4).](cache/FCO2/Beaulieu_timeseries.jpg)
+Dashed black line denotes atmospheric CO2 at 400 uatm.](cache/FCO2/Beaulieu_timeseries.jpg)
+
+## Figure S3
+![Figure s3: Observed (orange) versus BIKER-inferred (green) timerseries of daily $k_{600}$ for the 47 SWOT-simulated rivers with no additional SWOT errors.
+Note that the y axis is normalized by maximum observed values to compare across rivers.](cache/validation/timeseries_noerr.jpg)
+
+## Figure S4
+![Figure s3: Observed (orange) versus BIKER-inferred (green) timerseries of daily $k_{600}$ for the 16 SWOT-simulated rivers with the Frasson et al.
+(2021) error model.
+Note that the y axis is normalized by maximum observed values to compare across rivers.](cache/validation/timeseries_err.jpg)
 

### 2022-01-28 src/runFCO2.R

@@ -11,7 +11,8 @@ legend_labels <- c('BIKER',
                    'Model using observed \ndepth')
 
 #This model specifically predicts w and d from HG and then obtains velocity by continuity. The model fit for velocity was notably worse.
-Brinkerhoff2019_AHG <- read.csv('cache/AHG.csv')
+widAHG <- read_rds('cache/widAHG.rds')
+depAHG <- read_rds('cache/depAHG.rds')
 
 ###################
 ##FUNCTION TO GENERATE FCO2 FROM ALL DEPTH MODELS AND ALL SWOT RIVERS--------------------------------
@@ -85,7 +86,8 @@ fun_FCO2_analysis <- function(river) {
     }
   }
   D_raymond2013 <- d
-  D_brinkerhoff2019 <- Brinkerhoff2019_AHG[2,]$coef*Q_obs^(Brinkerhoff2019_AHG[2,]$slope) #Brinkerhoff etal 2019, estimated in CONUS_CO2 code
+  D_brinkerhoff2019 <- exp(summary(depAHG)$coefficient[1])*Q_obs^(summary(depAHG)$coefficient[2]) #Brinkerhoff etal 2019, estimated in CONUS_CO2 code
+  D_brinkerhoff2019 <- predict(depAHG, newdata=)
 
   #Calculate velocity rating curves following suite of rating curve models [m/s]
   V_raymond2012 <- exp(-1.64)*Q_obs^0.285
@@ -98,7 +100,7 @@ fun_FCO2_analysis <- function(river) {
   }
 
   V_raymond2013 <- v
-  V_brinkerhoff2019 <- (1/(Brinkerhoff2019_AHG[2,]$coef*Brinkerhoff2019_AHG[1,]$coef))*Q_obs^(1-Brinkerhoff2019_AHG[2,]$slope-Brinkerhoff2019_AHG[1,]$slope) #Brinkerhoff etal 2019, estimated in CONUS_CO2 code
+  V_brinkerhoff2019 <- exp(1/(summary(widAHG)$coefficient[1]*summary(depAHG)$coefficient[1]))*Q_obs^(1-summary(widAHG)$coefficient[2]-summary(depAHG)$coefficient[2]) #Brinkerhoff etal 2019, estimated in CONUS_CO2 code
 
   #Generate k600 estimates [m/dy]---------------------------
   k600_obs <- k600_model(D_obs, S_obs, V_obs)

### 2022-01-28 src/swot_k_model_v3.R

@@ -15,10 +15,8 @@ library(tidyr)
 library(ggplot2)
 library(colorspace)
 library(cowplot)
+library(readr)
 theme_set(theme_classic())
-
-#setwd('C:\\Users\\cbrinkerhoff\\OneDrive - University of Massachusetts\\Ongoing Projects\\RSK600')
-setwd('C:\\Users\\craig\\Documents\\OneDrive - University of Massachusetts\\Ongoing Projects\\RSK600')
 
 #some constants
 g <- 9.8 #gravitational acceleration [m/s2]
@@ -100,8 +98,8 @@ data <- rbind(data, ulseth_data_s, usgs_data_s)
 num <- nrow(data)
 wid <- lm(log(width)~log(Qm3s), data=data)
 dep <- lm(log(depth)~log(Qm3s), data=data)
-models <- data.frame('model' = c('width', 'depth'), 'r2'=c(summary(wid)$r.squared, summary(dep)$r.squared), 'coef'=c(exp(wid$coefficient[1]), exp(dep$coefficient[1])), 'slope'=c(wid$coefficient[2], dep$coefficient[2]))
-write.csv(models, 'cache\\AHG.csv')
+write_rds(wid, 'cache/widAHG.rds')
+write_rds(dep, 'cache/depAHG.rds')
 
 #######CALCULATE CHANNEL GEOMETRY---------------------------------------------------------
 data$Rh <- (data$depth*data$width)/(data$width + 2*data$depth) #hydraulic radius, assuming rectangular channel [m]
@@ -133,9 +131,9 @@ hydraulicallyWide$k600_pred_wideHydraulics <- predict(lm_hydraulicallyWide_small
 
 #plot model
 plot_smallEddy_eS <- ggplot(hydraulicallyWide, aes(x=k600_pred_wideHydraulics, y=k600)) +
-  geom_point(size=5, color='#bebada') +
+  geom_point(size=5, color='#377eb8') +
   geom_abline(linetype='dashed', color='darkgrey', size=1.5)+ #1:1 line
-  annotate("text", label = paste0('r2: ', round(summary(lm_hydraulicallyWide_smallEddy_eS)$r.squared,2)), x = 1, y = 100, size = 8, colour = "purple")+
[…]

### 2022-02-14 docs/manuscript_v8.Rmd
some ms and prepping for final validation

--- 2022-01-28 manuscript_v8.Rmd
@@
 Despite much progress, there is still large uncertainty in the temporal dynamics of gas exchange and thus carbon emissions to the atmosphere.
-Much of this uncertainty stems from uncertainty in gas exchange velocity $k_{600}$, the rate of this diffusion.
+Much of this uncertainty stems from uncertainty in gas exchange velocity $k_{600}$ (the rate of this diffusive transport).
 We propose that the NASA/CNES/USKA/CSA SWOT satellite, set to usher in a new age of hydrology research at the global scale, can provide similarly transformative insights to fluvial gas exchange modeling upon launch in 2022.
@@
 Natural systems play a critical role in the budgeting and accounting of the global carbon cycle under climate change.
-Following @colePlumbingGlobalCarbon2007, the global river network is recognized to emit substantial amounts of carbon to the atmosphere via evasion (gas exchange driven by a diffusion gradient and water turbulence), in addition to their long understood role in transporting carbon to the oceans via downstream advection.
+Following @colePlumbingGlobalCarbon2007, the global river network is recognized to emit substantial amounts of carbon to the atmosphere via evasion (gas exchange driven by a diffusion gradient and near-surface turbulence), in addition to their long understood role in transporting carbon to the oceans via downstream advection.
 Current estimates of total carbon dioxide evasion ($FCO_2$) to the atmosphere from the global river network vary from 650-2000 Tg C/yr [*Liu accepted*; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013], with 167 Tg-C/yr coming from mountain streams alone [@horgbyUnexpectedLargeEvasion2019].
@@
 @ahoDistinctConcentrationdischargeDynamics2021; @ahoDifferentialResponseGreenhouse2019; @crawfordCO2TimeSeries2017; @peterScalesDriversTemporal2014; @ranDynamicsRiverineCO22017; @rocher-rosLandscapeProcessDomains2019; @raymondAtmosphericCO2Evasion2000; @liuHydrologicControlsPCO22018a].
-This work has lead to recent river-reach explicit modeling of $\delta CO_2$ using global hydrography datasets at up to monthlytemporal resolutions [@horgbyUnexpectedLargeEvasion2019; @brinkerhoffLakeMorphometryRiver2021; @saccardiImprovingPredictionsStream2021; *Liu accepetd*], but an equivalently sophisticated representation of $k_{CO_2}$ is still lacking.
+This work has lead to recent river-reach explicit modeling of $\delta CO_2$ using global hydrography datasets at up to monthly temporal resolutions [@horgbyUnexpectedLargeEvasion2019; @brinkerhoffLakeMorphometryRiver2021; @saccardiImprovingPredictionsStream2021; *Liu accepetd*], but an equivalently sophisticated representation of $k_{CO_2}$ is still lacking.
 As equations 1 and 2 dictate, $CO_2_{water}$ and $k_{CO_2}$ share the burden of calculating $FCO_2$ and therefore these next-generation models will have to contend with this discrepancy, which is directly dictated by the lack of direct measurements of global river hydraulics.
@@
 
-@wangPhysicallyBasedScaling2021 recently attempted to address this global *k* problem by simulating $k_{O_2}$ in 35 rivers of many sizes (widths ranging from 0.23–349m) using a stream metabolism model [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets to infer what *k* must have been to produce their observations.
+@wangPhysicallyBasedScaling2021 recently attempted to address this global *k* problem by simulating $k_{O_2}$ in 35 rivers of many sizes (widths ranging from 0.23–349m) using a stream metabolism model [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets to infer what *k* must have been to produce their 'observations'.
 They then compared this simulated dataset against direct measurements of *k*, finding similar performance and parameter values for process-based models of gas evasion.
@@
 
-To answer this first question, we use hydraulic geometry [the fundamental geomorphic relationships between streamflow and channel shape- @leopoldHydraulicGeometryStream1953; @gleasonHydraulicGeometryNatural2015] to develop a process-based model for large-river (here defined a wider than 50m) gas exchange.
+To answer this first question, we use hydraulic geometry- the fundamental geomorphic relationships between streamflow and channel shape [@leopoldHydraulicGeometryStream1953; @gleasonHydraulicGeometryNatural2015] to develop a process-based model for large-river (here defined as wider than 50m) gas exchange.
 We then take these findings and explore the second question by implementing this hydraulic model, which in turn defines $k_{600}$, within an algorithm named BIKER ('Bayesian Inference of the $k_{600}$ Evasion Rate').
 The goal of BIKER is to require no in situ inputs of any kind (although in situ data could be ingested and should improve results) such that it is globally implementable on any SWOT-observable river.
-We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers (as SWOT has not yet launched) under both a 'no-measurement-error' and a 'measurement-error' scenario to explore algorithm robustness to the error tolerances expected in the satellite's measurements.
+We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers (as SWOT has not yet launched) under two 'measurement-error' scenarios to explore algorithm robustness to the expected measurement errors implicit in the satellite's observations.
 Finally, we also couple BIKER's $k_{600}$ estimates with $\delta CO_2$ to predict gas fluxes and compare these against established literature methods that are reliant on in situ hydraulic measurements.
@@
 ## 2 Prediciting $k_{600}$ from large-river hydraulic geometry
-To predict $k_{600}$ in the large rivers that SWOT will observe, we start from established process-based models for *k*, impose hydraulic assumptions that are valid in SWOT-observable rivers, and obtain a suite of models we empirically test.
-Following a description of the data (section 2.1), we outline established models (section 2.2) and then step through our hydraulic assumptions to arrive at equations that are compatible with SWOT measurements (section 2.2).
-We then validate the model (section 2.3)
+To predict $k_{600}$ in the large rivers that SWOT will observe, we start from an established process-based model for *k*, impose hydraulic assumptions that are valid in SWOT-observable rivers, and obtain a model we empirically test.
+Following a description of the data (section 2.1), we outline established models (section 2.2) and then step through our hydraulic assumptions to arrive at an equation that is compatible with SWOT measurements (section 2.3).
+We then empirically validate the model (section 2.3)
 
@@
 This is equation 4 and is referred to here as the 'Reynolds extension' model.
-The Reynolds model is hypothetically useful in low-turbulence flows where a relative lack of large-scale eddies effectively 'filter out' the number of small-eddies that actually reach the interface and initiate gas exchange [@talkeTurbulentKineticEnergy2013].
+The Reynolds model is hypothetically useful in low-turbulence scenarios where a relative lack of large-scale eddies effectively 'filter out' the number of small-eddies that actually reach the interface and initiate gas exchange [@talkeTurbulentKineticEnergy2013].
 While scaling *k* via a shear Reynold's formulation is sometimes done to parameterize wave-breaking gas exchange models in the open ocean  [@brumerWaveRelatedReynoldsNumber2017; @zhaoDependenceWhitecapCoverage2001; @zhaoEffectWindWaves2003], it is infrequently done in rivers.
@@
 Equations 3 and 4 both rely on $\epsilon$, which is non-trivial to measure.
-When working at large scales, a commonly used $\epsilon$ model assumes that all turbulence is generated at the bed and transported to the air-water interface via the log-law-of-the-wall via equation 5 [@lorkeUnifiedScalingRelation2006; @nezuTurbulenceOpenChannel1993].
-Another approach models *k* via 'form-drag dissipation' (equation 6) which is equivalently the total stream power per unit mass water.
+When working at large scales, a commonly used $\epsilon$ model assumes that all turbulence is generated at the bed and transported to the air-water interface via the log-law-of-the-wall [equation 5- @lorkeUnifiedScalingRelation2006; @nezuTurbulenceOpenChannel1993].
+Another approach (specific to fluvial settings) models *k* via 'form-drag dissipation' (equation 6) which is equivalently the total stream power per unit mass water.
 This normalized stream power captures the bulk frictional resistance (and thus energy dissipation) via channel banks, meanders, bars, etc.
 that is unique to fluvial systems [@moogStreamReaerationNonuniform1999].
-Authors have since shown that equation 6l can reasonably predict *k* in rivers and streams [@raymondScalingGasTransfer2012; @ulsethDistinctAirWater2019; @wangPhysicallyBasedScaling2021].
+Authors have since shown that equation 6 can reasonably predict *k* in rivers and streams [@raymondScalingGasTransfer2012; @ulsethDistinctAirWater2019; @wangPhysicallyBasedScaling2021].
 
@@
 $R_h \approx H$.
-This is a common assumption in hydraulic and geomorphic modeling of large rivers, and the average flow in a SWOT-observable river flow has a $\frac{R_h}{H}$ ratio of `r HG_swot$mean` (n = `r HG_swot$n`; see Text S1 for how we built this dataset).
+This is a common assumption in hydraulic and geomorphic modeling of large rivers, where the average flow in a SWOT-observable river flow has a $\frac{R_h}{H}$ ratio of `r HG_swot$mean` (n = `r HG_swot$n`; see Text S1 for how we built this dataset).
 We refer to these rivers as 'hydraulically-wide'.
 
-We therefore assume that all SWOT-observable rivers are hydraulically-wide.
-We then test the performance of four models for predicting *k* in hydraulically-wide channels: the four unique combinations of equations 3-4 and equations 5-6 outlined above.
+We therefore assume that all SWOT-observable rivers are hydraulically-wide and derive a model for gas exchange.
 The overall goal was to reduce the equations down to their fundamental parameters, identifying which terms are SWOT observable and limiting the number of terms not directly measurable via SWOT.
-While the complete model derivations and results for the four models are provided in Text S2 and Figure S1, the final and best-performing model is presented below.
-
-We impose $\epsilon=\epsilon_D$ and $R_H=H$ on the Reynolds extension model and arrive at equation 7 (with statistical coefficient $\beta_1$).
+To do this, we impose $\epsilon=\epsilon_D$ and $R_H=H$ on the Reynolds extension model and arrive at equation 7 (with statistical coefficient $\beta_1$).
 Equation 7 thus define gas exchange velocity solely as a function of slope, mean flow depth, and mean flow velocity.
@@
 $$k_{600}=\beta_1(gS)^{7/16}\bar{U}^{1/4}H^{9/16} \mathbf{(7)}$$
+
+we also test the performance of three other models for predicting *k* in hydraulically-wide channels via the other three unique combinations of equations 3-4 and equations 5-6.
+T While the complete model derivations and results for the four models are provided in Text S2 and Figure S1, the final and best-performing model (equation 7) is presented below.
+
 ### 2.4 Model validation
@@
 The success of this model in hydraulically-wide channels provides us with a strong physical-model for gas evasion.
-The river hydraulics terms in equation 7 can either be directly measured or reasonably inferred from SWOT measurements, effectively opening the door for remotely sensing the gas exchange velocity.
+The river hydraulics terms in equation 7 ($\bar{U}$, $H$, and $S$) can either be directly measured or reasonably inferred from SWOT measurements, effectively opening the door for remotely sensing the gas exchange velocity.
 This is explored next.
 
-## 3 Exploiting equation 8 to remotely sense gas exchange velocity
-We have shown that scaling $k_{600}$ via equation 8 explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2,2)*100`% of the variation in $k_{600}$ in hydraulically-wide rivers, which includes nearly all SWOT-observable rivers.
-Further, equation 8 has only three non-remotely-sensible terms: $k_{600}$, mean flow depth, and mean flow velocity.
+## 3 Exploiting equation 7 to remotely sense gas exchange velocity
+We have shown that scaling $k_{600}$ via equation 7 explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2,2)*100`% of the variation in $k_{600}$ in hydraulically-wide rivers, which includes nearly all SWOT-observable rivers.
+Further, equation 7 has only three non-remotely-sensible terms: $k_{600}$, mean flow depth, and mean flow velocity.
 Conveniently, techniques to simultaneously infer mean flow depth and velocity from SWOT data (among other parameters) have been established over the last decade to infer streamflow from SWOT's measurements [e.g.
 @andreadisConstrainingAssimilationSWOT2020; @brinkerhoffConstrainingRemoteRiver2020; @durandEstimatingReachaveragedDischarge2014; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014a; @garamboisVariationalEstimationEffective2020; @brissetAssimilationAltimetricData2018; @hagemannBAMBayesianAMHGManning2017; @larnierRiverDischargeBathymetry2020a; @oubanasDischargeEstimationUngauged2018].
-For BIKER, we follow the work developed by @hagemannBAMBayesianAMHGManning2017, @brinkerhoffConstrainingRemoteRiver2020, and @durandEstimatingReachaveragedDischarge2014 to infer $k_{600}$, chanel depth and velocity from SWOT observations using a modified form of Manning's equation.
+For BIKER, we follow the work developed by @hagemannBAMBayesianAMHGManning2017, @brinkerhoffConstrainingRemoteRiver2020, and @durandEstimatingReachaveragedDischarge2014 to infer $k_{600}$, channel depth and velocity from SWOT observations using a modified form of Manning's equation.
 Following a description of the data used (section 3.1), we detail algorithm development and experimental design (section 3.2) and then we present the validation results (section 3.3).
@@
 ### 3.1 Data
-![Figure 3: Map of the 47 hydraulic models and 1 timeseries of CO2 samples used in this study.
+![Figure 3: Map of the 47 hydraulic models and 1 timeseries of $CO_2$ samples used in this study.
 Note that hydraulic model locations are approximate as some of the models are not geo-referenced.](cache/maps/map3.jpg)
@@
 These datasets are created using standard hydraulic models forced with known inflows and measured bathymetry to model the hydraulic response of the rivers.
-These `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers represent the 'no-measurement-error' scenario that reflects a best-case scenario that is used here as a first validation of BIKER.
-`r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` of these rivers are then further prepared by @frassonExploringFactorsControlling2021 to more closely mimic SWOT by adding realistic radar errors and sampling along the satellite’s future ground track.
-Specifically, they account for 1) the SWOT error budget itself [CITE}], 2) thermal noise, and 3) radar layover error, witht the latter two being modeled by [@durandHowWillRadar2020].
 
-*COLIN: @frassonExploringFactorsControlling2021 also developed a sophisticated error model that mimics the effects of the atmosphere, topography, and water surface geometry on SWOT measurements (16/47 rivers) to add realistic radar layover errors that depend on the time and place of observation.
-The resulting data are the most realistic synthetic SWOT river data in the literature.
-The approximate locations of these rivers are plotted in Figure 1.*
-
-**According to Durand 2016 and Frasson, there is no measurement error added to the non-error scenario rivers.
-Mike had a small bit of Gaussian noise to pepsi 1 but he doesn't seem to count it has true measurement error.**
+Recall that we validate BIKER under two different 'error scenarios'.
+While SWOT will provide river surface measurements of novel quality and resolution, as with all remote sensing products there are expected errors that will be implicit in these measurements.
+Here, we validate BIKER under a 'no-measurement-error' scenario that reflects a best-case situation and is used as a first test of algorithm validity.
+`r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` of these rivers are then additionally validated under a 'measurement-error' scenario to more closely mimic SWOT by adding realistic radar errors and sampling along the satellite’s future ground track.
+SWOT river error modeling was developed by @durandHowWillRadar2020 and then applied to these `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` rivers by @frassonExploringFactorsControlling2021.
 
@@
 Therefore, the raw carbon emissions estimates presented in this paper are meaningless in the context of actually measured carbon emissions from these rivers.
-However, we are interested in the effect of BIKER $k_{600}$ errors on eventual fluxes, and therefore applying these 'unit' $CO_2$ values allows such a compariosn by providing a realtistic timeseries.
+However, we are interested in the effect of BIKER $k_{600}$ errors on eventual fluxes, and therefore applying these 'unit' $CO_2$ values allows for such a comparison by providing a realistic timeseries.
 
@@
 #### 3.2.1 BIKER
+**You made a potential note about moving this to the SI**
 BIKER, and Bayesian inference in general, starts from Bayes rule (equation 8), where $\Theta$ is some set of non-remotely-sensible parameters we want to solve for (including $k_{600}$), *x* is the observed data, $f(x|\Theta)$ is the sampling model where data are conditional on the parameters, and $p(\Theta)$ is the joint prior distribution of the parameters.
@@
 $$p(\Theta|x) = \frac{f(x|\Theta)p(\Theta)}{p(x)} \mathbf{(8)}$$
-The heart of BIKER is its reformulation of the $k_{600}$ model (equation 9) as a Bayesian sampling model that is conditional on the non-remotely-sensible parameters (i.e.
+The heart of BIKER is its reformulation of the $k_{600}$ model (equation 7) as a Bayesian sampling model that is conditional on the non-remotely-sensible parameters (i.e.
 $f(x|\Theta)$).
@@
 This is equation 10 after describing everything as normal distributions of the log-transformed terms.
-$\sigma_{k_{600}}^2$ refers to the total uncertainty implicit in equation 9 This uncertainty arises from parameter uncertainty in equation 9, Manning's equation, and the rectangular channel assumption.
+$\sigma_{k_{600}}^2$ refers to the total uncertainty implicit in equation 9.
+This uncertainty arises from 1) parameter uncertainty in equation 7, 2) Manning's equation, and 3) the rectangular channel assumption.
 
@@
 Equations 8 and 10 also necessitate that we specify prior distributions for the parameters $A_0_i$, $k_{600_t}$, and $n_i$.
-Prior distributions, described by their hyperparameters, formalize the a priori estimates and uncertainties for the non-remotely-sensed terms.
+Prior distributions formalize the a priori estimates and uncertainties for the non-remotely-sensed terms.
 More intuitively, BIKER priors represent our 'prior river knowledge' of what $A_0_i$, $k_{600_t}$, and $n_i$ probably are for some river since they cannot be directly remotely sensed.
-This is equivalent to the 'empirical Bayes approach' to Bayesian inference [@hoffFirstCourseBayesian2009].
+This is analogous to the 'empirical Bayes approach' to Bayesian inference [@hoffFirstCourseBayesian2009].
 Our goal in prior specification was to rely on absolutely no in situ information such that we could run this method on any river on Earth solely using SWOT observations.
@@
 These priors are assigned to a river solely using SWOT data *W* and *S*, therefore meeting our needs for complete global implementability.
-Prior assignment for $k_{600_t}$ was developed similarly (all prior specifications are elaborated on in Texts S3 and S4).
+Prior assignment for $k_{600_t}$ was developed similarly (all prior specifications are elaborated on in Text S3).
 
-With the sampling model described ($f(x|\Theta)$ = equation 10) and priors $p(x$) specified (Texts S3 and S4), a joint posterior distribution conditional on the SWOT observations ($p(\Theta|x) \propto f(x|\Theta)p(x)$) is therefore also specified.
+With the sampling model described ($f(x|\Theta)$ = equation 10) and priors $p(x$) specified (Text S3), a joint posterior distribution conditional on the SWOT observations ($p(\Theta|x) \propto f(x|\Theta)p(x)$) is therefore also specified.
 To approximate this distribution, we use a Markov Chain Monte Carlo (MCMC) algorithm implemented using the Stan probabilistic programming language.
@@
 #### 3.2.2 BIKER validation
-We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers (section 3.1) using daily simulated hydraulics.
-We also re-validate BIKER on the `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` rivers with [@frassonExploringFactorsControlling2021]'s SWOT error model to corrupt the hydraulics to mimic realistic SWOT measurements (widths, heights, and slopes).
+We validate BIKER, assuming no measurement error, on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers using daily simulated hydraulics.
+We also re-validate BIKER under the 'measurement-error' scenario using the `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` rivers with [@frassonExploringFactorsControlling2021]'s SWOT error model to corrupt the hydraulics to mimic realistic SWOT measurements (widths, heights, and slopes).
+These data were outlined in section 3.1.
+
 BIKER is unique in that it can provide a timeseries of $k_{600}$- for each SWOT observation, it yields a unique $k_{600}$.
@@
 We acknowledge that there is error in equation 7 as shown in Figure 2, but this error can be explicitly parameterized in our Bayesian system.
-Therefore, the BIKER validation presented here is an exercise to see how well the imperfect and partial SWOT observations can infer $k_{600}$ given hydraulic assumptions in equation 9 and uncertainty in the data itself.
+This is elaborated on in section 4.2.
+Therefore, the BIKER validation presented here is an exercise to see how well the imperfect and partial SWOT observations can infer $k_{600}$ given the hydraulic assumptions in equation 9 and uncertainty in the data itself.
 
-We validate BIKER as a timeseries of $k_{600}$ for each river using the BIKER posterior means.
-Our error metrics consider the timeseries nature of the problem, are defined in Table S2, and consist of the following:
+Therefore, we validate BIKER as a timeseries of $k_{600}$ for each river using the BIKER posterior means.
+Our error metrics consider the timeseries nature of the problem and are formally defined in Table S2.
+They consist of the following:
 
@@
 However, streamflow data and/or model outputs are more readily modeled at the global scale than river channel geometry, and so upscaling models usually predict $\bar{U}$ and *H* as functions of streamflow (*Q*) using hydraulic geometry (HG) scaling relationships.
-This effectively reduces equation 2 down to equation 11.
+This effectively reduces equation 2 to equation 11.
 It is worth stressing that these upscaling workflows rely on in situ streamflow records and/or high-quality streamflow routing outputs (unlike BIKER, which only requires a river be SWOT-observable).
@@
 
-In that context, we sought to compare BIKER against established upscaling methods from the literature for upscaling carbon emissions from our 47 rivers.
-So, we pair the 26 biweekly $CO_2$ and water temperature samples from @beaulieuControlsGasTransfer2012 (section 3.1, figure S2) with every 14th set of daily SWOT observations (as the $CO_2$ data is bi-weekly).
+In that context, we sought to compare BIKER against established methods from the literature for upscaling carbon emissions from river networks (in this case, our 47 rivers) So, we pair the 26 biweekly $CO_2$ and water temperature samples from @beaulieuControlsGasTransfer2012 (section 3.1, figure S2) with every 14th set of daily SWOT observations (as the $CO_2$ data is bi-weekly).
 We then calculate $FCO_2$ using equation 1, an atmospheric $CO_{2_{air}}$ of 390 uatm, and four different models for $k_{600}$, which are explained in full in Table S3.
-In brief, the four methods are : 1) BIKER-inferred $K_{600}$, 2-3) $k_{600}$ calculated using equation 7 and HG models for depth and velocity from @raymondGlobalCarbonDioxide2013 and @raymondScalingGasTransfer2012, and 4) HG models fit to our hydraulics data from section 2.2.1.
+In brief, the four methods are : 1) BIKER-inferred $K_{600}$, 2) $k_{600}$ calculated using equation 7 and HG models for depth and velocity from @raymondGlobalCarbonDioxide2013, 3) the same using HG models from @raymondScalingGasTransfer2012, and 4) the same using HG models fit to our hydraulics data from section 2.2.1.
 These models are referred to as 'BIKER', 'Raymond 2013', 'Raymond 2012', and 'Brinkerhoff 2019', respectively.
-These latter three approaches use the in situ streamflow record as their input, while BIKER uses only SWOT observations.
-This experiment lets us perturb two things: 1) how sensitive upscaled emissions estimates are to the HG models that predict river hydraulics from streamflow (as everything else is held constant) and 2) how a fully-remote technique like BIKER compares to in situ techniques like the other three.
+These latter three approaches all use the in situ streamflow record as their input, while BIKER uses only SWOT observations.
+This experiment lets us perturb two things: 1) how sensitive the eventual upscaled emissions estimates are to the HG models used (as everything else is held constant) and 2) how a fully-remote technique like BIKER compares to in situ techniques like the other three.
 
-Next, equation 12 is used to obtain a temperature and $CO_2$ specific gas exchange velocity from the flour flavors of $k_{600}$, where the Schmidt number *Sc* was estimated following @raymondScalingGasTransfer2012.
-We then estimate a yearly total carbon emission rate (via $CO_2$ evasion) by applying each river's mean $FCO_2$ over the river's surface area and summing all of these emission rates.
+Equation 12 is used to obtain a temperature and $CO_2$ specific gas exchange velocity from the four $k_{600}$ models, where the Schmidt number *Sc* was estimated following @raymondScalingGasTransfer2012.
+Finally, we estimate a yearly total carbon emission rate (via $CO_2$ evasion) by applying each river's mean $FCO_2$ over the river's surface area and summing all rates across rivers.
 $$k_{CO2, temp}=k_{600}*(Sc/600)^{-1/2} \mathbf{(12)}$$
@@
 #### 3.3.1 BIKER
-Figure 4 plots a representative set of the `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` timeseries plots of predicted and observed $k_{600}$, assuming no SWOT measurement error.
-Two rivers each rivers were randomly sampled from the three tertiles of river *KGE* scores (Table S2).
-Consult Figure S3 for all 47 timeseries plots and Figure S4 for the 16 rivers with the [@frassonExploringFactorsControlling2021] error model.
+Figure 4 plots a representative set of the `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` timeseries of predicted and observed $k_{600}$, assuming no SWOT measurement error.
+Two rivers each were randomly sampled from the three tertiles of river *KGE* scores (Table S2).
+Consult Figure S3 for all 47 timeseries plots (assuming no measurement error) and Figure S4 for the 16 rivers with measurement errors.
 Note that the y axis is normalized by maximum observed values to compare across rivers.
-In general, the temporal dynamics of $k_{600}$ are reproduced quite well by BIKER, with the highs and lows of evasion correctly inferred by BIKER.
+In general, the temporal dynamics of $k_{600}$ are reproduced quite well by BIKER, with the highs and lows of evasion correctly inferred by BIKER in the better-performing rivers.
 Notably, there is sometimes positive bias in the estimates (e.g.
-Seine River).
-Some rivers are inferred nearly perfectly (e.g.
-Kanawha, Ohio Section 2, Wabash rivers) while others perform more poorly (e.g.
-Ash Slough, Ohio), though again much of this error stems from errors in inferring the magnitude of $k_{600}$ rather than its temporal dynamics.
+both 'middle KGE' rivers).
+Some rivers yield the correct temporal dynamics, but the magnitude of these values is stretched relative to the observed (e.g.
+Ohio River).
+In this example, temporal trends are still reasonably inferred even though the magnitude of the estimates is quite off.
 
@@
 Representative river timeseries plots of $k_{600}$.
-Orange is calculated via observed hydraulics, while green is BIKER and uses SWOT measurements as its sole input.
-The green ribbon indicates the 95% CIs for the BIKER predicted values.
-Rivers randomly sampled from the three tertiles of KGE scores.
+Orange is observed, while green is BIKER and uses SWOT measurements as its sole input.
+The green ribbon indicates the 95% CIs for the BIKER posteriors.
+Rivers are randomly sampled from the three tertiles of KGE scores.
 (a-b) are poorest performing rivers, (b-c) are in the middle, and (e-f) are the best performing rivers.
@@
 
-We calculate performance metrics following section 3.2.2 and Table S2.
+Next, we calculate performance metrics following section 3.2.2 and Table S2.
 These are presented in Figure 5, which plots the scores for the 47 rivers with no measurement errors as empirical cumulative density functions (eCDFs).
@@
 This indicates very strong inference of each river's temporal $k_{600}$ dynamics given that absolutely no in situ information is being used to predict $k_{600}$.
-NRMSE has a median score of `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$NRMSE, na.rm=T),2)` and median rBIAS is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$rBIAS, na.rm=T),2)`, highlighting that many rivers which have a significant positive bias (Figure 4 also confirms this).
-Taken in aggregate, Figures 4-5 indicate that BIKER is quite good at capturing temporal dynamics in $k_600$, however there is sometimes positive bias in its estimates.
+NRMSE has a median score of `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$NRMSE, na.rm=T),2)` and median rBIAS is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$rBIAS, na.rm=T),2)`, highlighting that many rivers which have significant positive biases (Figure 4 also confirms this visually).
+Taken in aggregate, Figures 4-5 indicate that BIKER is quite good at capturing temporal dynamics in $k_600$, however there is often positive bias in its estimates.
 
 ![Figure 5.
-Performance metrics by river, plooted as empirical cummulative density functions (eCDFs).
-Each subpanel is labelled by its performance metric (Table S2).
-Dashed lines denote medians.](cache/validation/validation_by_river.jpg)
+Performance metrics by river, ploted as empirical cummulative density functions (eCDFs).
+Each subpanel is labelled by its performance metric (defined in Table S2).
+Dashed lines denote median scores.](cache/validation/validation_by_river.jpg)
 
-Figure 6 compares BIKER results under the 'no-error' and 'error' scenario for the 16 rivers that have the @frassonExploringFactorsControlling2021 imposed on their SWOT observations (section **something**).
-For *r*, *rBIAS*, and *NRMSE*, BIKER performs nearly identically, regardless of inherent measurement error in the SWOT observations: points are scattered along the 1:1 line and their regressions also mirror this 1:1 line.
-For *KGE*, performance is also largely similar except for one river which get significantly worse and skews the regression line (Seine Upstream river).
+Figure 6 compares BIKER results under the 'no-measurement-error' and 'measurement-error' scenarios for the `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` rivers.
+For *r*, *rBIAS*, and *NRMSE*, BIKER performs nearly identically, regardless of implicit measurement errors in the inputs to BIKER: points are scattered along the 1:1 line and their regressions also mirror this 1:1 line.
+For *KGE*, performance is also largely similar except for one river which gets significantly worse and skews the regression line (Seine Upstream river).
 Overall, these results strongly suggest that BIKER will be robust to the measurement errors that will be implicit in SWOT's observations of river width and slope.
@@
 b) Empirical cummulative density distributions of the coefficient of varation of prior, posterior, and observed $k_{600}$ per river.
-If BIKER is correctly inferring temporal variation in $k_{600}$, the posterior eCDF will approximate the obersved eCDF (as it does).](cache/validation/validation_temporal.jpg)
+If BIKER is correctly inferring temporal variation in $k_{600}$, the posterior eCDF will approximate the observed eCDF (as it does).](cache/validation/validation_temporal.jpg)
 
@@
 Finally, we carry these $FCO_2$ calculations all the way to annual carbon emissions rates and compare BIKER against established in situ techniques.
-It is important to remember that only BIKER is completely globally implementable, while the other three models necessarily rely on having a streamflow record ( or high-quality routed streamflow) readily available.
+It is important to remember that only BIKER is completely globally implementable, while the other three models necessarily rely on having a streamflow record or high-quality routed streamflow readily available.
 Therefore, the HG methods reflect their 'best-case scenarios' while BIKER reflects its worst case scenario, where priors are set entirely from SWOT observations and are generally the least informative they will ever be.
+This means that BIKER's annual carbon emission estimate can only improve from what is presented here.
 We also stress again that the raw emissions rates here are largely meaningless as they are calculated from an identical $CO_{2_water}$ timeseries applied to every river.
@@
 
-Figures 8 compares the annual carbon emissions rate (via $F_{CO_2}$) from the rivers using BIKER posterior means and the three streamgauge-based HG models.
-Surprisingly, BIKER outperforms the gauge-based approaches, nearly correctly inferring the annual carbon emissions rate (`r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_BIKER',]$yearlyCFux, 2)` gG-C/yr for BIKER versus `r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_obs',]$yearlyCFux, 2)` gG-C/yr observed).
+Figure 8 compares the annual carbon emissions rate (via $F_{CO_2}$) from the rivers using BIKER posterior means and the three streamgauge-based HG models.
+Surprisingly, BIKER outperforms the gauge-based approaches, nearly correctly inferring the annual carbon emissions rate (`r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_BIKER',]$yearlyCFux, 2)` Tg-C/yr for BIKER versus `r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_obs',]$yearlyCFux, 2)` Tg-C/yr observed).
 The three HG models overestimate this emissions rate (`r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_raymond2013',]$yearlyCFux, 2)`, `r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_raymond2012',]$yearlyCFux, 2)`, and `r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_brinkerhoff2019',]$yearlyCFux, 2)`) for 'Raymond 2013', 'Raymond 2012', and 'Brinkerhoff 2019' respectively.
@@
 BIKER's relatively stronger performance than the in situ HG models is elaborated on in section 4.3.
+Thus, despite BIKER using absolutely no streamgauge or other in situ data like the other 3 methods do, it provides a strong upscaling estimate of the annual carbon emission rate for the rivers (Figure 8).
 
-Thus, despite BIKER using absolutely no streamgauge or other in situ data like the other 3 methods do, it provides a strong upscaling estimate of the annual carbon emission rate across 47 rivers (Figure 8).
-The in situ models represent their best case scenarios (perfect in situ discharge records) while BIKER represents its worst-case scenario (relatively uninformed priors).
-This means that BIKER's annual carbon emission estimate can only improve.
-
-![Figure 8: Yearly carbon emissions via CO2 evasion across all rivers and accounting for surface area.
+![Figure 8: Yearly carbon emissions rate via $CO_2$ evasion across all rivers.
 Completely remotely-sensed methods are colored in red, in situ methods in purple, and the observed in orange.](cache/FCO2/FCO2_models.jpg)
@@
 ### 4.1 Gas exchange in hydraulically-wide rivers
-Field studies of gas exchange in wide rivers have suggested that $k_{600}$ behaves differently than in steeper and smaller rivers [@ulsethDistinctAirWater2019; @beaulieuControlsGasTransfer2012; @alinPhysicalControlsCarbon2011; @wangPhysicallyBasedScaling2021].
+Field studies of gas exchange in wide rivers have suggested that $k_{600}$ behaves differently than in steeper and smaller rivers [@ulsethDistinctAirWater2019; @beaulieuControlsGasTransfer2012; @alinPhysicalControlsCarbon2011; @wangPhysicallyBasedScaling2021; @raymondGasExchangeRivers2001a].
 While much work has focused on the small-stream side of the stream-to-river continuum, comparatively less work has been done in large systems.
-Here, we focus on the 'smooth-channel' end of the continuum, using a model for gas exchange that scales $k_{600}$ by both $\epsilon_D$ and the Shear Reynold's number that is specifically adapted for 'hydraulically-wide' rivers.
-This is empirically validated in Figure 3.
-Specifically, Figure S1 confirms that scaling $k_{600}$ with a combined small-eddy/Shear Reynold's formulation (equation 7) reduces positive bias in the model's predictions of the smallest $k_{600}$ values [where the relative decrease in turbulence reaching the surface is greater than the small-eddy model alone suggests, per equation 7's theoretical basis- @moogAirWaterGasTransfer1999].
-Scaling $k_{600}$ via a shear Reynold's number alone is commonly done to parameterize wave-related gas exchange models in the open ocean [@brumerWaveRelatedReynoldsNumber2017; @zhaoDependenceWhitecapCoverage2001; @zhaoEffectWindWaves2003].
-However to our knowledge, @moogAirWaterGasTransfer1999's specific model (imposing a differential surface turbulence theory on the small-eddy model) has never been empirically validated in rivers.
+Here, we focus on the larger, 'smooth-channel' end of the continuum, using a model for gas exchange that scales $k_{600}$ by both $\epsilon_D$ and a Shear Reynold's number.
+This model is empirically validated in Figure 3.
+Specifically, Figure S1 confirms that scaling $k_{600}$ with a Shear Reynold's adaption of the small-eddy model (equation 7) reduces positive bias in the model's predictions of the smallest $k_{600}$ values [where the relative decrease in turbulence reaching the surface is greater than the small-eddy model alone suggests, per equation 7's theoretical basis- @moogAirWaterGasTransfer1999].
+Scaling $k_{600}$ via a shear Reynold's number is commonly done to parameterize breaking-wave gas exchange models in the open ocean [@brumerWaveRelatedReynoldsNumber2017; @zhaoDependenceWhitecapCoverage2001; @zhaoEffectWindWaves2003], though this specific to high wind speeds in open ocean.
+To our knowledge, @moogAirWaterGasTransfer1999's specific setup, which imposes a 'differentially-experienced' surface turbulence theory on the small-eddy model, has never been empirically validated in rivers.
 Figure 3 provides this empirical verification for hydraulically wide channels, where it's theoretical basis should generally hold.
-Using our full dataset of $k_{600}$, we also observed this model breaks down when including non-hydraulically-wide rivers (as the theory dictates).
+Using our full dataset of $k_{600}$, we also observed this model breaks down when including non-hydraulically-wide rivers (as the theory would suggest).
 Future tests should also explore other shear Reynold's scaling relations for gas exchange in rivers.
 
-Crucially, we are not accounting for wind-driven gas exchange, which can exerts a significant influence on lake and ocean gas exchange and has been suggested to play an important role in wide-river gas exchange [@beaulieuControlsGasTransfer2012; @wangPhysicallyBasedScaling2021].
-Similarly, none of the existing hydraulics-driven gas exchange models account for wind-driven gas exchange either.
-In that vein, BIKER's outputs can be interpreted as the 'gas exchange velocity in low-wind conditions', when surface turbulence is dominated by hydraulics rather than wind.
+Crucially, we are not accounting for wind-driven gas exchange, which is suggested to play an important role in wide rivers because river surface area is sufficiently large that wind can drive gas exchange [@beaulieuControlsGasTransfer2012; @wangPhysicallyBasedScaling2021; @raymondGasExchangeRivers2001a].
+Similarly, none of the existing hydraulics-driven fluvial $k_{600}$ models account for wind-driven gas exchange either.
+BIKER's outputs can be interpreted as the '$k_{600}$ under low-wind conditions', when surface turbulence is dominated by hydraulics rather than wind.
 That said, BIKER's flexible implementation is a good start towards eventually coupling hydraulics-driven gas exchange with wind-driven gas exchange.
-Such a model would likely take a form similar to the conceptual models proposed by @wangPhysicallyBasedScaling2021, chuWindStreamFlow2003, an/.or @plateReaerationOpenChannel1984.
+Such a model would likely take a form similar to the conceptual models proposed by @wangPhysicallyBasedScaling2021, chuWindStreamFlow2003, and/or @plateReaerationOpenChannel1984.
 All three generally propose calculating both a wind-driven *k* and a hydraulics-driven *k* and then weighting each term via parameters.
@@
 For example, @wallinSpatiotemporalVariabilityGas2011 performed a preliminary analysis in northern Sweden relating cross-section specific temporal variability in gas exchange with channel slope, but they were limited to a mean of only 8 measurements per river in a single watershed.
-While a good start, this is obviously not sufficient for developing process-level understandings of gas exchange at the global-scale.
+While this is a good start, this is obviously not sufficient for further developing process-level understandings of gas exchange at the global-scale.
 
-Therefore, inferring $k_{600}$ from SWOT data is an attractive option to address this.
-While BIKER will not directly measure $k_{600}$, it does robustly infer temporal trends in $k_{600}$ (Figures 4-7) which will no doubt be useful in large-scale analyses.
-**MOre go here**.
+Therefore, inferring $k_{600}$ from SWOT data is an attractive option to address this problem of limited data.
+For reference, 95% of the SWOT-visible rivers globally (202,811) will have sufficient SWOT observations along the river to run BIKER at least once every 21 days, with most of the temperate and Arctic rivers having 3+ observations per 21-day cycle [@altenauSurfaceWaterOcean2021].
+While BIKER will not directly measure $k_{600}$, it does robustly infer temporal trends in $k_{600}$ (Figures 4-7) and when coupled with this much data will provide a novel dataset of $k_{600}$ never before possible.
 
@@
 This makes sense as Section 2 has effectively reduced $k_{600}$ to a function of hydraulics that are nearly all directly measurable by SWOT.
-Any resulting bias in BIKER's predictions is likely attributable to bias in the priors used for the non-remotely sensed terms (equations 9-10).
-For SWOT discharge algorithms, authors have repeatedly shown that the 'quality' of prior hyperparameters plays a large role in the success of discharge inversions [@frassonExploringFactorsControlling2021; @andreadisConstrainingAssimilationSWOT2020; @brinkerhoffConstrainingRemoteRiver2020; @tuozzoloEstimatingRiverDischarge2019] and our results here further corroborate this finding.
+Any resulting bias in BIKER's predictions is likely attributable to either bias in the priors used for the non-remotely sensed terms (equations 9-10) or in the $k_{600}$ model itself (the scaling coefficient in equation 7).
+For SWOT discharge algorithms, authors have repeatedly shown that the 'quality' of prior river knowledge plays a large role in the success of discharge inversions [@frassonExploringFactorsControlling2021; @andreadisConstrainingAssimilationSWOT2020; @brinkerhoffConstrainingRemoteRiver2020; @tuozzoloEstimatingRiverDischarge2019] and our results here further corroborate this finding.
 It should be stressed that a substantial portion of rivers from [@frassonExploringFactorsControlling2021] are canal-shaped in nature with different hydraulic properties than a natural river channel.
@@
 Figure 8 confirms that BIKER is quite successful, when coupled with $CO_2$ data, at predicting annual upscaled carbon emissions from the river network.
-This encouraging result has two main implications for future work coupling remote sensing via SWOT with in situ data.
+This encouraging result has three main implications for future work coupling remote sensing via SWOT with in situ data.
+These are outlined below.
+
 First, figure 8 directly implies that BIKER will be useful when coupled with large-scale $CO_{2_water}$ models [e.g.
 **Liu etal in review**; @saccardiImprovingPredictionsStream2021].
-These two models robustly predict reach-scale dissolved $CO_2$ concentrations using two different approaches: machine learning [**Liu et al in review**] and process-based reactive transport modeling [@saccardiImprovingPredictionsStream2021] but both models yield spatially and temporally consistent $CO_2$ estimates.
+These two models robustly predict reach-scale dissolved $CO_2$ concentrations using two different approaches: machine learning [**Liu et al in review**] and process-based reactive transport modeling [@saccardiImprovingPredictionsStream2021] but both models yield $CO_2$ estimates that would be spatially and temporally consistent with BIKER's output.
 Our promising results suggest that BIKER could provide additional (and directly observed) measurements of $k_{CO_2}$ to these models, thereby better informing model results through direct observations.
-This is likely accomplished via data assimilation which has proven useful for similar problems across geophysical modeling applications [@carrassiDataAssimilationGeosciences2018] and specifically for assimilating remotely sensed information into streamflow routing models [@ishitsukaCombiningOpticalRemote2020], of which @saccardiImprovingPredictionsStream2021 takes an approximate form of.
+This is likely to be accomplished via data assimilation which has proven useful in using remotly-sensed discharge to improve streamflow routing models [@fengRecentChangesArctic2021a; @ishitsukaCombiningOpticalRemote2020], and of which the @saccardiImprovingPredictionsStream2021 $CO_2$ model takes a similar form.
 
-Figure 8 highlights a potentially large source of uncetainty on current river $CO_2$ upscaling estimates: the HG models employed to scale river channel hydraulics with streamflow.
-As previously stressed, the only difference between the three in situ HG models and the observed estimate in Figure 8 is the specific HG model employed to predict river depth and velocity.
-This means that drammatically different carbon emission estimates are obtainable depending on the training data used for these HG models.
-While BIKER performs similarly to these methods, care should be taken to use the HG models most approriate for your specific river geomorphology.
-
-Finally, at the field-scale, Fgure 8 also confirms that we can couple BIKER with in situ gas concentration loggers to produce $FCO_2$ estimates at novel temporal resolutions in SWOT-observable rivers.
+Second, at the field-scale Figure 8 confirms that we can couple BIKER with in situ gas concentration loggers to produce $FCO_2$ estimates at novel temporal resolutions in SWOT-observable rivers.
 High temporal fidelity datasets of dissolved $CO_2$ in SWOT-observable rivers now exist [e.g.
@@
 Regardless, both approaches would produce $CO_2_{water}$ and $k_{CO_2}$ datasets at equivalent temporal resolution and allow us to directly calculate daily to sub-daily $CO_2$ fluxes from river reaches.
+
+Finally, Figure 8 also uniquely allows us to directly compare the influence of one's HG model on total carbon emission rates from river networks, as all other calculations and parameters were held constant.
+Therefore, figure 8 highlights a potentially large source of uncertainty in current river $CO_2$ upscaling estimates: the HG models employed to scale river channel hydraulics with streamflow.
+As previously stressed, the only difference between the three in situ HG models and the observed estimate in Figure 8 is the specific HG model employed to predict river depth and velocity.
+This means that dramatically different carbon emission estimates are obtainable depending on the training data used for these HG models.
+While BIKER performs similarly to these methods, it must be stressed that the global representativeness of one's training data for HG models is of paramount importance.
 
@@
 In this paper, we propose that the soon-to-launch SWOT satellite will provide enough hydraulic measurements to analyse the temporal dynamics of $k_{600}$ using SWOT measurements, and therefore allow for a global-scale analysis of spatiotemporal trends in large-river $k_{600}$ once SWOT launches.
-In that context, we develop 1) a hydraulic model for $k_{600}$ that is nearly SWOT observable and explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2,2)*100`% of variation in $k_{600}$, and 2) the BIKER algorithm to infer $k_{600}$ using no on-the-ground information.
-Validating on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers, we show show strong recovery of rivers' temporal $k_{600}$ dynamics and the the total annual carbon emission rate across all `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers.
+In that context, we develop 1) a hydraulic model for $k_{600}$ that is nearly entirely SWOT observable and explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2,2)*100`% of variation in $k_{600}$, and 2) the BIKER algorithm to infer $k_{600}$ using no on-the-ground information.
+Validating on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers, we show show strong recovery of rivers' temporal $k_{600}$ dynamics and the total annual carbon emission rate across all `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers.
 These results suggest BIKER can be used to infer global-scale, near daily estimates of fluvial gas exchange velocity once SWOT launches in 2022.

### 2022-02-14 docs/supp_v8.Rmd
some ms and prepping for final validation

--- 2022-01-28 supp_v8.Rmd
@@
 ## Contents
-This supplementary information contains 4 texts, 2 figures and 3 tables.
+This supplementary information contains 3 texts, 4 figures and 3 tables.
 Please consult https://github.com/craigbrinkerhoff/RSK600 for all code to build and generate results, figures, and the manuscript.
 
-## Test S1: Estimating how many SWOT rivers are hydraulically wide
-To quantify how frequently SWOT-observable rivers are also hydraulically wide river channels (section 3 of main text), we used the dataset of field-measured river hydraulics in @brinkerhoffReconcilingAtaStationAtManyStations2019.
+## Test S1: Estimating how hydraulically wide SWOT rivers are
+To quantify the prevalence of hydraulically-wide, SWOT-observable rivers, we used the dataset of field-measured river hydraulics outlined in Section 2.1.
 That dataset has over 500,000 discrete measurements of river width, velocity, area, and discharge that were made by the United States Geological Survey (USGS) to calibrate streamgauge rating curves.
@@
 ## Text S3: BIKER hyperparameterization
-In this text we explain in detail how BIKER's hyperparameter values were set for each river.
+In this text we explain in detail how BIKER's prior distributions were set for each river.
 
-We assign prior hyperparameters using SWOT data only.
-All priors are formalized within the model as truncated normal distributions of the log-transformed terms such that $log(X) \sim  \mathcal{N}(\mu, \sigma^2)$ for $\lambda < log(X) < \gamma$, using prior hyperparameters mean ($\mu$), standard deviation ($\sigma$), and upper ($\gamma$) and lower bounds ($\lambda$) for any parameter *X*.
+Prior distributions are defined by their hyperparameters.
+For BIKER, prior distriubtions are formalized as truncated normal distributions of the log-transformed terms such that $log(X) \sim  \mathcal{N}(\mu, \sigma^2)$ for $\lambda < log(X) < \gamma$, using prior hyperparameters mean ($\mu$), standard deviation ($\sigma$), and upper ($\gamma$) and lower bounds ($\lambda$) for any parameter *X*.
+It is important to again stress, as we do in the main text, that BIKER prior hyperparameters are described using only SWOT data to be completely globally implementanble.
+This is analogous to the 'empirical Bayes method' [@hoffFirstCourseBayesian2009].
 
@@
 
-Finally, $k_{600_{\sigma}}$ is set to 0.30 (log-space).
+$k_{600_{\sigma}}$ is set to 0.30 (log-space).
 This corresponds to a coefficient of variation of approximately 0.30, which we took to reflect a reasonably strong agreement between the prior and the observed values.
@@
 
-## Text S4: Specifiying the $\sigma_{k_{600}}$ parameter
-Here, we detail how the $\sigma_{k_{600}}$ parameter is specified in this study, as well as how it should be specified once SWOT launches.
-
-Recall that $\sigma_{k_{600}}$ refers to the total uncertainty inherent in equation 8, i.e.
-stemming from both the Reynolds/form-drag dissipation model and the Manning's model for $\bar{U}$.
-For the purposes of this study, we are validating BIKER against equation 10 in the main text and so all uncertainties associated with the $\beta_1$ parameter are ignored and we only need to reflect the Manning's uncertainty in our specification of $\sigma_{k_{600}}^2$.
+Finally, recall that $\sigma_{k_{600}}$ refers to the total uncertainty inherent in equation 9, i.e.
+stemming from the Reynolds-extension model (equation 7) and Manning's equation for $\bar{U}$.
+For the purposes of this study, we are validating BIKER against equation 7 in the main text and so all uncertainties associated with the $\beta_1$ parameter are ignored and we only need to reflect the Manning's uncertainty in our specification of $\sigma_{k_{600}}^2$.
 Therefore, we take @hagemannBAMBayesianAMHGManning2017's estimated uncertainty from Manning's equation to infer streamflow from SWOT observations (0.25) and inflate it slightly to also account for the hydraulically-wide channel assumption and arrive at 0.30.
@@
 *Table S1: Studies that gas exchange velocity measurements come from.
-'Study' refers to the paper from which these measurements were taken.
+'Study' refers to the paper from which these measurements were obtained.
 Any data wrangling was done by those authors.
-'Field Workers' refers to who actually made the measurements.
+'Workers' refers to who actually made the measurements.
 The Raymond et al.
 (2012) dataset is itself a meta-analysis.
-Please see that paper for how those measurements were collected.
-We used the set of measurements ultimately published by Ulseth et al.
-2019.*
+Please see that paper for how those measurements were collected, and see the 'Additional Studies' for the workers who actually collected the measurements.*
 

### 2022-02-14 manuscript_v9.Rmd
some ms and prepping for final validation

--- 2022-02-14 docs/manuscript_v8.Rmd
@@
 ---
-title: "Using hydraulic geometry and the SWOT satellite to remotely sense river gas exchange velocity"
+title: "Remotely sensing river greenhouse gas exchange velocity using the SWOT satellite"
 author:
@@
 institute:
-  - UMass: Department of Civil & Environmental Engineering, University of Massachusetts-Amherst
-  - Lamont: Lamont-Doherty Earth Observatory, Columbia University
-  - Yale: School of the Environment, Yale University
+  - UMass: Department of Civil & Environmental Engineering, University of Massachusetts, Amherst, MA
+  - Lamont: Lamont-Doherty Earth Observatory, Columbia University, Palisades, NY
+  - Yale: School of the Environment, Yale University, New Haven, CT
 bibliography: References.bib
@@
 We propose that the NASA/CNES/USKA/CSA SWOT satellite, set to usher in a new age of hydrology research at the global scale, can provide similarly transformative insights to fluvial gas exchange modeling upon launch in 2022.
-Here, we present work inferring $k_{600}$ from future SWOT observations without in situ calibration.
-We exploit the unique geomorphology of SWOT-observable rivers to develop a physical model of gas exchange that is nearly entirely remotely sensible and explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2,2)*100`% of variation in $k_{600}$ We then couple this model with established Bayesian inference techniques to develop BIKER, the 'Bayesian Inference of the $k_{600}$ Exchange Rate' algorithm.
-We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers (SWOT has not yet launched) and 166 discrete measurements of river gas exchange, yielding an algorithm that predicts the $k_{600}$ timeseries solely from SWOT observations with a by-river median Kling-Gupta Efficiency of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$KGE), 2)`.
-Like similar Bayesian remote sensing algorithms, BIKER is better at reproducing the temporal dynamics of gas exchange (median correlation coefficient of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$r), 2)`), than the absolute rates of exchange (median relative bias of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$rBIAS), 2)*100`%).
-Finally, we show BIKER robust to measurement errors implicit in the SWOT data.
+Here, we present work inferring $k_{600}$ from synthetic future SWOT observations without in situ calibration.
+We exploit the unique geomorphology of SWOT-observable rivers to develop a physical model of gas exchange that is remotely sensible and explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2,2)*100`% of variation in $k_{600}$ We then couple this model with established Bayesian inference techniques to develop BIKER, the 'Bayesian Inference of the $k_{600}$ Exchange Rate' algorithm.
+We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers and 166 discrete measurements of river gas exchange, yielding an algorithm that predicts the $k_{600}$ timeseries solely from SWOT observations with a by-river median Kling-Gupta Efficiency of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$KGE), 2)`.
+**get more impressive skill score than KGE** Similar to Bayesian remote sensing algorithms for SWOT, BIKER is better at reproducing the temporal dynamics of gas exchange (median correlation coefficient of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$r), 2)`), than the absolute rates of exchange (median relative bias of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$rBIAS), 2)*100`%).
+Finally, we show BIKER is robust to measurement errors implicit in the SWOT data.
 With SWOT set to launch in late 2022, we suggest that BIKER will be useful in mapping the global-scale spatiotemporal dynamics of fluvial gas exchange in large rivers.
@@
 
-The structure of equation 1 necessitates that calculations of this flux are highly sensitive to measurements/estimates of *k*.
+There is a robust existing literature exploring spatiotemporal patterns in $\delta CO_2_{water}$ [e.g.
+@ahoDistinctConcentrationdischargeDynamics2021; @ahoDifferentialResponseGreenhouse2019; @crawfordCO2TimeSeries2017; @peterScalesDriversTemporal2014; @ranDynamicsRiverineCO22017; @rocher-rosLandscapeProcessDomains2019; @raymondAtmosphericCO2Evasion2000; @liuHydrologicControlsPCO22018a].
+This work has lead to recent river-reach explicit modeling of $\delta CO_2$ using global hydrography datasets at up to monthly temporal resolutions [@horgbyUnexpectedLargeEvasion2019; @brinkerhoffLakeMorphometryRiver2021; @saccardiImprovingPredictionsStream2021; *Liu accepetd*], but an equivalently sophisticated representation of $k_{CO_2}$ is still lacking.
+As equations 1 and 2 dictate, $\delta CO_2_{water}$ and $k_{CO_2}$ share the burden of calculating $FCO_2$ and therefore these next-generation models will have to contend with this discrepancy, which is directly dictated by the lack of direct measurements of global river hydraulics.
+
+The structure of equation 1 necessitates that calculations of $FCO_2$ are highly sensitive to measurements/estimates of *k*.
 However, *k* can only be directly calculated via a known gas concentration gradient, eddy-covariance measurements, or tracer additions to the stream [@hallGasExchangeStreams2020].
 In trying to constrain the global fluvial $FCO_2$ flux across millions of rivers, this calculation is impossible, and necessitates the use of predictive models for *k* that are based on easily obtained river hydraulic properties.
-In that vein, there have been over 20 empirical models developed to predict *k* from river hydraulics, generally using some combination of mean velocity $\bar{U}$, shear velocity $U_*$, width $W$, depth $H$, and slope $S$ as predictors [@wangPhysicallyBasedScaling2021].
-These models usually predict $k_{600}$, or *k* normalized by a Schmidt number *Sc* of 600.
+In that vein, there have been over 20 empirical models developed to predict *k* from river hydraulics, generally using some combination of mean velocity $\bar{U}$, shear velocity $U_*$, width $W$, depth $H$, and slope $S$ as predictors [@hallGasExchangeStreams2020; @wangPhysicallyBasedScaling2021].
+These models usually predict $k_{600}$, or *k* normalized by a Schmidt number (*Sc*) of 600.
 This is to remove the effect of water temperature and gas type from predictive models, as warmer waters and lower *Sc* numbers both increase gas exchange rates [@hallGasExchangeStreams2020].
 Specifically, $k_{600}$ reflects the $k_{CO_{2}}$ at 20 degrees Celsius.
-Through this normalization, these models exclusively perturb the geophysical controls on gas exchange [@hallGasExchangeStreams2020].
+Through this normalization, these models focus solely on geormorphic explanantions for variation in *k* [@hallGasExchangeStreams2020].
 
-These empirical models enable estimating a global $FCO_2$ flux ($FCO_2_{upscaled}$) from millions of rivers, but they also change the base parameters that ultimately control that aggregate estimate.
+Applying these *k* models across watersheds, regions, or continents is called ‘upscaling.’ This upscaling gives the impossible-to-measure *k* term in Equation 1 for any arbitrary number of rivers, but also changes the base parameters that ultimately control the final estimate of $FCO_2$.
 That is, by making *k* a function of hydraulics, $FCO_2_{upscaled}$ is now a direct function of river hydraulics.
 This functional relationship is described in equation 2.
-It suggests that $FCO_2$ estimates are not only at the mercy of the accuracy and spatiotemporal resolution of $\delta CO_2$ as discussed previously, but also the accuracy and resolution of our river hydraulics estimates.
+It suggests that $FCO_2$ estimates are not only at the mercy of the accuracy and spatiotemporal resolution of $\delta CO_2$, but also the accuracy and resolution of our river hydraulics estimates.
 
 $$FCO_{2_{upscaled}}=f(k_{CO_2}, \delta CO_2)=f(\bar{U}, H, S, W, \delta CO_2) \mathbf{(2)}$$
-There is a robust existing literature exploring spatiotemporal patterns in $CO_2_{water}$ [e.g.
-@ahoDistinctConcentrationdischargeDynamics2021; @ahoDifferentialResponseGreenhouse2019; @crawfordCO2TimeSeries2017; @peterScalesDriversTemporal2014; @ranDynamicsRiverineCO22017; @rocher-rosLandscapeProcessDomains2019; @raymondAtmosphericCO2Evasion2000; @liuHydrologicControlsPCO22018a].
-This work has lead to recent river-reach explicit modeling of $\delta CO_2$ using global hydrography datasets at up to monthly temporal resolutions [@horgbyUnexpectedLargeEvasion2019; @brinkerhoffLakeMorphometryRiver2021; @saccardiImprovingPredictionsStream2021; *Liu accepetd*], but an equivalently sophisticated representation of $k_{CO_2}$ is still lacking.
-As equations 1 and 2 dictate, $CO_2_{water}$ and $k_{CO_2}$ share the burden of calculating $FCO_2$ and therefore these next-generation models will have to contend with this discrepancy, which is directly dictated by the lack of direct measurements of global river hydraulics.
 
-For example, to upscale $FCO_2$ globally, @raymondGlobalCarbonDioxide2013, @lauerwaldSpatialPatternsCO22015, and @horgbyUnexpectedLargeEvasion2019 all relied on $k_{CO_2}$ values indirectly estimated using mean annual streamflow models and scaling equations to predict hydraulic terms, while @borgesGloballySignificantGreenhousegas2015 used a combination of the above method and a constant $k_{CO_2}$ in space and time to upscale over Africa.
-In all of these foundational studies, the temporal dynamics of the gas exchange velocity (and thus $FCO_2$) were ignored because of hydraulic data limitations.
+Global upscalng has been attempted inthe literature.
+@raymondGlobalCarbonDioxide2013, @lauerwaldSpatialPatternsCO22015, and @horgbyUnexpectedLargeEvasion2019 all relied on $k_{CO_2}$ values indirectly estimated using mean annual streamflow models and hydraulic scaling equations to predict the hydraulic terms to in turn predict $k_{CO_2}$, while @borgesGloballySignificantGreenhousegas2015 used a combination of the above method and a constant $k_{CO_2}$ in space and time to upscale over Africa.
+In all of these foundational studies, the temporal dynamics of $k_{CO_2}$ (and thus dynamics in $FCO_2$) were ignored because of hydraulic data limitations.
 It has been shown at the field-scale that temporal dynamics of gas exchange can vary widely from site to site [@wallinSpatiotemporalVariabilityGas2011], but it has remained impractical to obtain temporally explicit $k_{CO_2}$ at continental-to-global scales.
@@
 
-@wangPhysicallyBasedScaling2021 recently attempted to address this global *k* problem by simulating $k_{O_2}$ in 35 rivers of many sizes (widths ranging from 0.23–349m) using a stream metabolism model [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets to infer what *k* must have been to produce their 'observations'.
-They then compared this simulated dataset against direct measurements of *k*, finding similar performance and parameter values for process-based models of gas evasion.
-However, they were still limited by a lack of direct hydraulic measurements and had to rely on scaling equations to estimate river depth and velocity.
-Even though approaches like this are incredibly useful for expanding our mechanistic understanding of gas exchange, it is less useful for global upscaling purposes as it relies on highly detailed in situ DO data for every river [@hallGasExchangeStreams2020].
+@wangPhysicallyBasedScaling2021 recently attempted to address this global *k* problem by simulating $k_{O_2}$ in 35 rivers of many sizes (widths ranging from 0.23–349m) using a stream metabolism model [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets to infer what $K_{O_2}$ must have been to produce their 'observations'.
+They then compared this simulated dataset against direct measurements of *k*, finding similar performance and parameter values against process-based models of gas evasion.
+However, they were still limited by a lack of direct hydraulic measurements and had to rely on hydraulic scaling equations to estimate river depth and velocity.
+Even though approaches like @wangPhysicallyBasedScaling2021's are incredibly useful for expanding our mechanistic understanding of gas exchange, they are less useful for global upscaling purposes as they rely on highly detailed in situ DO data for every river [@hallGasExchangeStreams2020].
 
-In theory, the discrepancy between the quality of our $CO_2_{water}$ and $k_{CO_2}$ estimates could be alleviated if direct hydraulics measurements (and in turn $k_{CO_2}$ via equation 2) were available at the global scale at a sufficient temporal resolution.
+We have established that literature has a reasonably good understanding of $\delta CO_2$ and a poor understanding in $k_{600}$ (and therefore $k_{CO_2}$) across large areas and in time.
+In theory, the discrepancy between the quality of our $\delta CO_2_{water}$ and $k_{CO_2}$ estimates could be alleviated if direct hydraulics measurements (and in turn $k_{CO_2}$ via equation 2) were available at the global scale at a sufficient temporal resolution.
 This is turn would also address the uncertainty regarding continental-to-global scale temporal dynamics of gas exchange noted earlier.
-Conveniently, these data will soon be available via the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission.
+Conveniently, these hydraulic data will soon be available via the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission.
 SWOT is expected to launch in 2022 and provide the world's first direct measurements of global water surface extent and elevation (and therefore water surface slope) at novel temporal resolutions.
@@
 
-  - Can we develop a physically-based hydraulic model for $k_{600}$ unique to SWOT-observed rivers?
+1) Can we develop a physically-based hydraulic model for $k_{600}$ unique to SWOT-observed rivers?
 
-  - Can we exploit such a model to infer $k_{600}$ (and its uncertainty) solely from SWOT observations?
+2) Can we exploit such a model to infer $k_{600}$ (and its uncertainty) solely from SWOT observations?
 
-To answer this first question, we use hydraulic geometry- the fundamental geomorphic relationships between streamflow and channel shape [@leopoldHydraulicGeometryStream1953; @gleasonHydraulicGeometryNatural2015] to develop a process-based model for large-river (here defined as wider than 50m) gas exchange.
-We then take these findings and explore the second question by implementing this hydraulic model, which in turn defines $k_{600}$, within an algorithm named BIKER ('Bayesian Inference of the $k_{600}$ Evasion Rate').
-The goal of BIKER is to require no in situ inputs of any kind (although in situ data could be ingested and should improve results) such that it is globally implementable on any SWOT-observable river.
-We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers (as SWOT has not yet launched) under two 'measurement-error' scenarios to explore algorithm robustness to the expected measurement errors implicit in the satellite's observations.
+To answer this first question, we use hydraulic geometry- the fundamental geomorphic relationships between streamflow and channel shape [@leopoldHydraulicGeometryStream1953; @gleasonHydraulicGeometryNatural2015] to develop a process-based model for large-river (here defined as wider than 50m to align with SWOT) $k_{600}$.
+We then take these findings and explore the second question by implementing this hydraulic model, which in turn defines $k_{600}$, within an algorithm named BIKER ('Bayesian Inference of the $k_{600}$ Evasion Rate') to infer $k_{600$ solely from SWOT measurements.
+The goal of BIKER is to require no in situ inputs of any kind (although in situ data could be ingested and would improve results) such that it is globally implementable on any SWOT-observable river.
+We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers (as SWOT has not yet launched) and explore BIKER's robustness to the expected measurement errors implicit in the satellite's observations.
 Finally, we also couple BIKER's $k_{600}$ estimates with $\delta CO_2$ to predict gas fluxes and compare these against established literature methods that are reliant on in situ hydraulic measurements.
@@
 ## 2 Prediciting $k_{600}$ from large-river hydraulic geometry
-To predict $k_{600}$ in the large rivers that SWOT will observe, we start from an established process-based model for *k*, impose hydraulic assumptions that are valid in SWOT-observable rivers, and obtain a model we empirically test.
+To predict $k_{600}$ in the large rivers that SWOT will observe, we start from an established process-based model for $k_{600}$, impose hydraulic assumptions that are valid in SWOT-observable rivers, and obtain a model we empirically test.
 Following a description of the data (section 2.1), we outline established models (section 2.2) and then step through our hydraulic assumptions to arrive at an equation that is compatible with SWOT measurements (section 2.3).
@@
 We develop our hydraulic model for $k_{600}$ using multiple datasets of measured $k_{600}$ and stream hydraulics collected from the literature.
-In total, there are 763 measurements of $k_{600}$, with 701 of these measurements previously gathered by @raymondScalingGasTransfer2012 and @ulsethDistinctAirWater2019.
+In total, we obtain 763 measurements of $k_{600}$, with 701 of these measurements previously gathered by @raymondScalingGasTransfer2012 and @ulsethDistinctAirWater2019.
 The remaining measurements come from @churchillPredictionStreamReaeration1964 and @owensReaerationStudiesStreams1964.
 See Table S1 for a complete list of the studies that collected these measurements.
-All $k_{600}$ measurements come from tracer studies and thus define $k_{600}$ and the rivers' mean hydraulic properties at the reach scale.
+All $k_{600}$ measurements come from tracer studies and thus define $k_{600}$ at the reach scale.
 
@@
 That dataset contains over 530,000 unique measurements of river channel velocity, width, and discharge from across the continental United States, originally made to calibrate United States Geological Survey (USGS) streamgauge rating curves and made public by the USGS.
+This dataset is used to calculate how frequently SWOT observable rivers meet our large-river hydraulic assumptions (section 2.2).
 
@@
 In the context of BIKER, we chose to test this model because large, SWOT-observable rivers are generally the smoothest, least-turbulent flows along the stream-to-ocean continuum where small eddies might not reach the surface.
-Further, to our knowledge this Reynolds extension model has never been empirically tested in predicting river *k*, aside from confirming that large-scale eddies differentially move turbulence to the surface in a large river [@talkeTurbulentKineticEnergy2013].
+Further, to our knowledge, this Reynolds extension model has never been empirically tested in predicting river *k*, aside from confirming that large-scale eddies differentially move turbulence to the surface in a large river [@talkeTurbulentKineticEnergy2013].
 
@@
 When working at large scales, a commonly used $\epsilon$ model assumes that all turbulence is generated at the bed and transported to the air-water interface via the log-law-of-the-wall [equation 5- @lorkeUnifiedScalingRelation2006; @nezuTurbulenceOpenChannel1993].
-Another approach (specific to fluvial settings) models *k* via 'form-drag dissipation' (equation 6) which is equivalently the total stream power per unit mass water.
+Another approach specific to fluvial settings models *k* via 'form-drag dissipation' (equation 6) which is equivalently the total stream power per unit mass water.
 This normalized stream power captures the bulk frictional resistance (and thus energy dissipation) via channel banks, meanders, bars, etc.
@@
 In general, as river size increases, channels become more rectangular, their shapes elongate (becoming wider quicker than they become deeper) and their hydraulic radii begin to approximate their mean flow depth, i.e.
-$R_h \approx H$.
-This is a common assumption in hydraulic and geomorphic modeling of large rivers, where the average flow in a SWOT-observable river flow has a $\frac{R_h}{H}$ ratio of `r HG_swot$mean` (n = `r HG_swot$n`; see Text S1 for how we built this dataset).
+$R_h \approx H$ **[CITE]**.
+This is a common assumption in hydraulic and geomorphic modeling of large rivers, and flows in a SWOT-observable river flow have an average $\frac{R_h}{H}$ ratio of `r HG_swot$mean` and standard deviation of `r HG_swot$sd` (n = `r HG_swot$n`; see Text S1 for how we built this dataset).
 We refer to these rivers as 'hydraulically-wide'.
 
-We therefore assume that all SWOT-observable rivers are hydraulically-wide and derive a model for gas exchange.
-The overall goal was to reduce the equations down to their fundamental parameters, identifying which terms are SWOT observable and limiting the number of terms not directly measurable via SWOT.
+We therefore assume that all SWOT-observable rivers are hydraulically-wide to derive a model for gas exchange.
+The overall goal is to reduce the equations down to their fundamental parameters, identifying which terms are SWOT observable and limiting the number of terms not directly measurable via SWOT.
 To do this, we impose $\epsilon=\epsilon_D$ and $R_H=H$ on the Reynolds extension model and arrive at equation 7 (with statistical coefficient $\beta_1$).
@@
 
-we also test the performance of three other models for predicting *k* in hydraulically-wide channels via the other three unique combinations of equations 3-4 and equations 5-6.
-T While the complete model derivations and results for the four models are provided in Text S2 and Figure S1, the final and best-performing model (equation 7) is presented below.
+We also test the performance of three other models for predicting *k* in hydraulically-wide channels via the other three unique combinations of equations 3-4 and equations 5-6.
+T While the complete model derivations and results for the four models are provided in Text S2 and Figure S1, the final and best-performing model (equation 7) is presented and used below.
 
@@
 With equation 7 derived, we now empirically its strength of fit for hydraulically-wide river flows.
-We validate on the dataset of in situ measurements of $k_{600}$, after filtering for measurements made in hydraulically-wide channels, which was operationally defined as flows whose hydraulic radius was within 1% of their mean flow depth.
-All told, this amounts to `r n` direct measurements of hydraulically-wide $k_{600}$ and stream hydraulics to test with.
-The model is assessed via the coefficient of determination ($r^2$) and plotted in Figure 2.
+We validate on the dataset of in situ measurements of $k_{600}$, after filtering for measurements made in hydraulically-wide channels, which was defined as flows whose hydraulic radius was within 1% of their mean flow depth.
+All told, this amounts to `r n` direct measurements of hydraulically-wide $k_{600}$.
+Equation 7 is assessed via the coefficient of determination ($r^2$) and plotted in Figure 2.
 Note that Figure 2 axes are plotted in logarithmic space only for visualization: model fit and validation (via $r^2$) were calculated in linear space as their models dictate.
@@
 ![Figure 2: Empirical testing of our large-river $k_{600}$ model on 166 measurements made in hydraulically-wide rivers.
+Dashed grey line denotes the 1:1 line.
 Note that axes are plotted in logarithmic space just for visualization: model fit and validation (via the coefficient of determination $r^2$) were calculated in linear space.](cache/k600_theory/k600Plot.jpg)
 
-Our model explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2, 2) * 100`% of variation in observed $k_{600}$ in hydraulically-wide rivers and accurately captures the scaling dynamics of observed $k_{600}$ as well.
-While the individual residuals can be quite large, the overall scaling of $k_{600}$ with river hydraulics is strongly captured.
-Compared to the other models tested (Figure S1), there is also less bias in the estimates thanks to the addition of the Reynold's number scaling for low-turbulent flows.
-The success of this model in hydraulically-wide channels provides us with a strong physical-model for gas evasion.
+Figure 2 shows that equation 7 explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2, 2) * 100`% of variation in observed $k_{600}$ in hydraulically-wide rivers and accurately captures the scaling dynamics of observed $k_{600}$ as well.
+While the individual residuals can be quite large, the overall scaling of $k_{600}$ with river hydraulics is strongly captured, remembering that Figure 2 is plotted in logarithmic space.
+Compared to other combinations of equations 3-6 (Figure S1), there is also less bias in the estimates thanks to the addition of the Reynold's number scaling for low-turbulent flows (as expected).
+The success of this model in hydraulically-wide channels provides us with a strong physical-model for gas evasion built with SWOT in mind.
 The river hydraulics terms in equation 7 ($\bar{U}$, $H$, and $S$) can either be directly measured or reasonably inferred from SWOT measurements, effectively opening the door for remotely sensing the gas exchange velocity.
@@
 ## 3 Exploiting equation 7 to remotely sense gas exchange velocity
-We have shown that scaling $k_{600}$ via equation 7 explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2,2)*100`% of the variation in $k_{600}$ in hydraulically-wide rivers, which includes nearly all SWOT-observable rivers.
-Further, equation 7 has only three non-remotely-sensible terms: $k_{600}$, mean flow depth, and mean flow velocity.
+We have shown that scaling $k_{600}$ via equation 7 should provide a reliable basis for hydraulically-wide rivers with a defined error structure.
+Further, equation 7 has only three non-directly-remotely-sensible terms: $k_{600}$, mean flow depth, and mean flow velocity.
 Conveniently, techniques to simultaneously infer mean flow depth and velocity from SWOT data (among other parameters) have been established over the last decade to infer streamflow from SWOT's measurements [e.g.
-@andreadisConstrainingAssimilationSWOT2020; @brinkerhoffConstrainingRemoteRiver2020; @durandEstimatingReachaveragedDischarge2014; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014a; @garamboisVariationalEstimationEffective2020; @brissetAssimilationAltimetricData2018; @hagemannBAMBayesianAMHGManning2017; @larnierRiverDischargeBathymetry2020a; @oubanasDischargeEstimationUngauged2018].
+**GLEASON & SMITH 2014**; @andreadisConstrainingAssimilationSWOT2020; @brinkerhoffConstrainingRemoteRiver2020; @durandEstimatingReachaveragedDischarge2014; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014a; @garamboisVariationalEstimationEffective2020; @brissetAssimilationAltimetricData2018; @hagemannBAMBayesianAMHGManning2017; @larnierRiverDischargeBathymetry2020a; @oubanasDischargeEstimationUngauged2018].
 For BIKER, we follow the work developed by @hagemannBAMBayesianAMHGManning2017, @brinkerhoffConstrainingRemoteRiver2020, and @durandEstimatingReachaveragedDischarge2014 to infer $k_{600}$, channel depth and velocity from SWOT observations using a modified form of Manning's equation.
@@
 To validate BIKER, we cannot use actual SWOT measurements as SWOT has yet to launch.
-In the hydrology literature, it is therefore standard practice to benchmark SWOT-related algorithms on "SWOT-like" data [@durandIntercomparisonRemoteSensing2016].
+In the hydrology literature, it has become standard practice to benchmark SWOT-related algorithms on "SWOT-like" data [@durandIntercomparisonRemoteSensing2016].
 We use 47 SWOT-simulated rivers for validation, where these simulated rivers are simply reach-averaged hydraulic model outputs where the water surface heights, slopes, and widths are labelled as RS observations and are used as the sole inputs to BIKER.
+These datasets are created using standard hydraulic models forced with known inflows and measured bathymetry to model the hydraulic response of the rivers, and then those terms visible to SWOT were extracted to produce hydraulically realistic synthetic observations.
 These data were published by @frassonExploringFactorsControlling2021 and @durandIntercomparisonRemoteSensing2016 as benchmarking datasets to explore remote sensing of discharge (RSQ) algorithm performance for the SWOT mission.
-These datasets are created using standard hydraulic models forced with known inflows and measured bathymetry to model the hydraulic response of the rivers.
 
-Recall that we validate BIKER under two different 'error scenarios'.
+We validate BIKER under two different 'error scenarios'.
 While SWOT will provide river surface measurements of novel quality and resolution, as with all remote sensing products there are expected errors that will be implicit in these measurements.
-Here, we validate BIKER under a 'no-measurement-error' scenario that reflects a best-case situation and is used as a first test of algorithm validity.
-`r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` of these rivers are then additionally validated under a 'measurement-error' scenario to more closely mimic SWOT by adding realistic radar errors and sampling along the satellite’s future ground track.
+Here, we validate BIKER under a 'no-measurement-error' scenario that reflects an unrealistic measurement as if SWOT has perfect accuracy and precision: we use the hydraulic model output directly as a first test of algorithm validity.
+`r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` of these rivers are then additionally validated under a 'measurement-error' scenario that more closely mimics expected SWOT by adding realistic radar errors and sampling along the satellite’s future ground track.
 SWOT river error modeling was developed by @durandHowWillRadar2020 and then applied to these `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` rivers by @frassonExploringFactorsControlling2021.
+This error modelling is non-trivial and computationally expensive, and thus we rely on these previously published data.
 
-For the $CO_2$ evasion and carbon emissions calculations (section 3.1.3), we use 26 bi-weekly dissolved $CO_2$ samples collected by @beaulieuControlsGasTransfer2012 at one location in the Ohio River for one calender year from 2008-2009 (Figure S2).
-Note that this $CO_2$ data is for the Ohio River only but was applied to all rivers (which includes multiple sections of the Ohio River).
-Therefore, the raw carbon emissions estimates presented in this paper are meaningless in the context of actually measured carbon emissions from these rivers.
-However, we are interested in the effect of BIKER $k_{600}$ errors on eventual fluxes, and therefore applying these 'unit' $CO_2$ values allows for such a comparison by providing a realistic timeseries.
+For $CO_2$ evasion and carbon emissions calculations, we use 26 bi-weekly dissolved $CO_2$ samples collected by @beaulieuControlsGasTransfer2012 at one location in the Ohio River for one calender year from 2008-2009 (Figure S2).
+Note that this $CO_2$ data is for the Ohio River only but was applied to all rivers to provide a physically realistic signal for $CO_2$ fluxes with meaningful seasonality and dynamics.
+Therefore, the raw carbon emissions estimates presented in this paper are completely meaningless in the context of actually measured carbon emissions from these rivers, but are better than specifying $CO_2$ concentrations devoid of context.
+These data are necessary as we are interested in the effect of BIKER $k_{600}$ errors on eventual fluxes and comparing these fluxes with published methods.
+Therefore, applying these 'unit' $CO_2$ values allows for such a comparison by providing a realistic timeseries.
 
@@
 #### 3.2.1 BIKER
-**You made a potential note about moving this to the SI**
 BIKER, and Bayesian inference in general, starts from Bayes rule (equation 8), where $\Theta$ is some set of non-remotely-sensible parameters we want to solve for (including $k_{600}$), *x* is the observed data, $f(x|\Theta)$ is the sampling model where data are conditional on the parameters, and $p(\Theta)$ is the joint prior distribution of the parameters.
@@
 Our goal in prior specification was to rely on absolutely no in situ information such that we could run this method on any river on Earth solely using SWOT observations.
-In theory, more informed priors via various a priori information about a specific river would improve BIKER performance, but here we chose to test the fully generalized algorithm.
+In theory, more informed priors via various a priori information about a specific river will improve BIKER performance, but here we chose to test the fully generalized algorithm.
 Therefore, the validation presented here is a 'worst-case scenario', wherein BIKER performance should improve with better prior information on the river.
@@
 These priors are assigned to a river solely using SWOT data *W* and *S*, therefore meeting our needs for complete global implementability.
-Prior assignment for $k_{600_t}$ was developed similarly (all prior specifications are elaborated on in Text S3).
+Prior assignment for $k_{600_t}$ was developed similarly (all prior specifications are elaborated in Text S3).
 
@@
 
-BIKER is unique in that it can provide a timeseries of $k_{600}$- for each SWOT observation, it yields a unique $k_{600}$.
+BIKER is unique in that it can provide a timeseries of $k_{600}$: for each SWOT observation, it yields a unique $k_{600}$.
 There are, to our knowledge, no datasets of $k_{600}$ over time approaching the temporal density of our simulated SWOT rivers.
-We therefore apply equation 7 as validated in Figure 2 to specify $k_{600}$ given the true hydraulics of each case.
+We therefore apply equation 7 as validated in Figure 2 to specify $k_{600}$ given the true hydraulics of each case and compare BIKER’s inversion to that value: given observed hydraulics, ‘observed’ $k_{600}$ comes from equation 7.
 Remember that SWOT cannot observe below the water surface and therefore cannot measure $\bar{U}$ or *H* (hence the need for equation 9), and that all SWOT observations contain errors in both space and time (hence equation 10).
@@
 #### 3.2.3 Carbon emissions validation
+**THis is colin's edits. Might need some reorg**
 It is one thing to accurately model the temporal dynamics of $k_{600}$ as above, but researchers are often most interested in the actual carbon emitted from river to atmosphere.
@@
 This effectively reduces equation 2 to equation 11.
-It is worth stressing that these upscaling workflows rely on in situ streamflow records and/or high-quality streamflow routing outputs (unlike BIKER, which only requires a river be SWOT-observable).
+It is worth stressing that these literature upscaling workflows rely on in situ streamflow records and/or high-quality streamflow routing models (unlike BIKER, which only requires a river be SWOT-observable).
 
@@
 
-In that context, we sought to compare BIKER against established methods from the literature for upscaling carbon emissions from river networks (in this case, our 47 rivers) So, we pair the 26 biweekly $CO_2$ and water temperature samples from @beaulieuControlsGasTransfer2012 (section 3.1, figure S2) with every 14th set of daily SWOT observations (as the $CO_2$ data is bi-weekly).
-We then calculate $FCO_2$ using equation 1, an atmospheric $CO_{2_{air}}$ of 390 uatm, and four different models for $k_{600}$, which are explained in full in Table S3.
-In brief, the four methods are : 1) BIKER-inferred $K_{600}$, 2) $k_{600}$ calculated using equation 7 and HG models for depth and velocity from @raymondGlobalCarbonDioxide2013, 3) the same using HG models from @raymondScalingGasTransfer2012, and 4) the same using HG models fit to our hydraulics data from section 2.2.1.
-These models are referred to as 'BIKER', 'Raymond 2013', 'Raymond 2012', and 'Brinkerhoff 2019', respectively.
-These latter three approaches all use the in situ streamflow record as their input, while BIKER uses only SWOT observations.
-This experiment lets us perturb two things: 1) how sensitive the eventual upscaled emissions estimates are to the HG models used (as everything else is held constant) and 2) how a fully-remote technique like BIKER compares to in situ techniques like the other three.
+In that context, we sought to compare BIKER against established methods from the literature for upscaling carbon emissions from river networks (in this case, our `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])`  rivers) So, we pair the 26 biweekly $CO_2$ and water temperature samples from @beaulieuControlsGasTransfer2012 (section 3.1, figure S2) with every 14th set of daily SWOT observations (as the $CO_2$ data is bi-weekly).
+We then calculate $FCO_2$ using equation 1, an atmospheric $CO_{2_{air}}$ of 390 uatm, and a temperature dependent *Sc* was estimated following Raymond et al.
+(2012).
+Finally, we estimate a yearly total carbon emission rate (via $CO_2$ evasion) by applying each river’s mean $FCO_2$ over the river’s surface area and summing all rates across rivers.
 
-Equation 12 is used to obtain a temperature and $CO_2$ specific gas exchange velocity from the four $k_{600}$ models, where the Schmidt number *Sc* was estimated following @raymondScalingGasTransfer2012.
-Finally, we estimate a yearly total carbon emission rate (via $CO_2$ evasion) by applying each river's mean $FCO_2$ over the river's surface area and summing all rates across rivers.
-$$k_{CO2, temp}=k_{600}*(Sc/600)^{-1/2} \mathbf{(12)}$$
+BIKER represents a new way of approaching the problem against existing GHG scaling literature.
+We are therefore interested in how the final fluxes that result from BIKER (Eq 11) compare to literature methods to gauge BIKER’s utility: BIKER is novel, but is it useful?
+We have the data to test four different models for fluxes: 'BIKER', 'Raymond 2013', 'Raymond 2012', and 'Brinkerhoff 2019'.
+These latter three approaches all use the same philosophy for $k_{600}$: making hydraulic and geomorphic assumptions to associate $k_{600}$ with observed hydraulics before using the $CO_2$ data as a realistic concentration timeseries to yield fluxes per equation 11.
+In all three approaches, these observed hydraulics are streamflow, while BIKER uses only SWOT observations.
+Therefore, the advantages of BIKER in ease of application are clear as SWOT will observe all global rivers wider than 50m while streamflow observations are extremely geographically limited, but BIKER is only attractive if it can produce fluxes with similar errors to published methods.
+Text S4 and Table S3 fully describe these three literature models.
 
@@
 Figure 4 plots a representative set of the `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` timeseries of predicted and observed $k_{600}$, assuming no SWOT measurement error.
-Two rivers each were randomly sampled from the three tertiles of river *KGE* scores (Table S2).
+Two rivers each were sampled from the three tertiles of river *KGE* scores (Table S2) for display.
 Consult Figure S3 for all 47 timeseries plots (assuming no measurement error) and Figure S4 for the 16 rivers with measurement errors.
@@
 
-Figure 6 compares BIKER results under the 'no-measurement-error' and 'measurement-error' scenarios for the `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` rivers.
-For *r*, *rBIAS*, and *NRMSE*, BIKER performs nearly identically, regardless of implicit measurement errors in the inputs to BIKER: points are scattered along the 1:1 line and their regressions also mirror this 1:1 line.
-For *KGE*, performance is also largely similar except for one river which gets significantly worse and skews the regression line (Seine Upstream river).
+Figure 6 compares BIKER results under the 'no-measurement-error' and 'measurement-error' scenarios for the `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` rivers for which @frassonExploringFactorsControlling2021 provide an error model.
+Rivers that fall within the purple zone get worse when accounting for measurement error, while rivers in the green get better.
+Note also that axes are flipped in order to visualize all 'better performances' in the upper-right-corner of each sub-plot.
+BIKER performs nearly identically, regardless of implicit measurement errors in the inputs to BIKER: points are scattered along the 1:1 line.
 Overall, these results strongly suggest that BIKER will be robust to the measurement errors that will be implicit in SWOT's observations of river width and slope.
@@
 Each subpanel is labelled by its performance metric (Table S2).
+Rivers that fall within the purple zone get worse when accounting for measurement error, while rivers in the green get better.
+Note also that axes are flipped in order to visualize all 'better performances' in the upper-right-corner of each sub-plot.
 Black line denotes linear regression while dashed grey line is the 1:1 line.](cache/validation/errorPlot.jpg)
+
+**something something temporal variability**
+You need to flesh this out a bit better and stop leaning so heavily on temporality.
+Feels almost like we're hiding something, which we aren't Colin was not a fan of the prior vs posterior CV eCDFs and also suggested that this should still hold even if we are doing just magntiude.
 
@@
 Figure 7a plots a histogram of the slopes of linear regressions between BIKER-predicted $k_{600}$ and observed $k_{600}$.
-A slope of 1 indicates that BIKER correctly infers the daily flucuations in gas exchange (even if there is bias baked into the estimates), while a slope well above or below 1 indicates incorrect inference of the temporal dynamics of $k_{600}$.
+A slope of 1 indicates that BIKER correctly infers the daily flucuations in gas exchange (even if there is bias implicit in these estimates), while a slope well above or below 1 indicates incorrect inference of the temporal dynamics of $k_{600}$.
 Figure 6a shows that the majority of rivers' slopes approximate 1 (median slope: `r round(median(results_dynamics$Slope),2)`), indicating strong inference of temporal gas exchange trends.
-A handful of rivers to have slopes well above or below 1, though the vast majority are at or just below 1.
 
@@
 Finally, we explore whether BIKER improves upon the prior with respect to temporal variation in $k_{600}$, here defined using the coefficient of variation ($CV=\frac{SD}{mean}$).
+This is an essential test of a Bayesian method to determine if SWOT observations have improved our knowledge or if our initial guess proved statistically most useful.
 Figure 6b plots eCDFs of observed, prior, and posterior $k_{600}$ *CV*, strongly showing that BIKER improves upon the prior information available to us about $k_{600}$.
@@
 ### 4.1 Gas exchange in hydraulically-wide rivers
-Field studies of gas exchange in wide rivers have suggested that $k_{600}$ behaves differently than in steeper and smaller rivers [@ulsethDistinctAirWater2019; @beaulieuControlsGasTransfer2012; @alinPhysicalControlsCarbon2011; @wangPhysicallyBasedScaling2021; @raymondGasExchangeRivers2001a].
+Field studies of gas exchange in wide rivers have suggested that $k_{600}$ behaves differently in these rivers than in steeper and smaller rivers [@ulsethDistinctAirWater2019; @beaulieuControlsGasTransfer2012; @alinPhysicalControlsCarbon2011; @wangPhysicallyBasedScaling2021; @raymondGasExchangeRivers2001a].
 While much work has focused on the small-stream side of the stream-to-river continuum, comparatively less work has been done in large systems.
@@
 Similarly, none of the existing hydraulics-driven fluvial $k_{600}$ models account for wind-driven gas exchange either.
-BIKER's outputs can be interpreted as the '$k_{600}$ under low-wind conditions', when surface turbulence is dominated by hydraulics rather than wind.
+BIKER's outputs can therefore be interpreted as the '$k_{600}$ under low-wind conditions', when surface turbulence is dominated by hydraulics rather than wind.
 That said, BIKER's flexible implementation is a good start towards eventually coupling hydraulics-driven gas exchange with wind-driven gas exchange.
@@
 For reference, 95% of the SWOT-visible rivers globally (202,811) will have sufficient SWOT observations along the river to run BIKER at least once every 21 days, with most of the temperate and Arctic rivers having 3+ observations per 21-day cycle [@altenauSurfaceWaterOcean2021].
-While BIKER will not directly measure $k_{600}$, it does robustly infer temporal trends in $k_{600}$ (Figures 4-7) and when coupled with this much data will provide a novel dataset of $k_{600}$ never before possible.
+While BIKER will not directly measure $k_{600}$, it does robustly infer temporal trends in $k_{600}$ and less robustly the absolute magnitude of $k_{600}$ (Figures 4-7) and when coupled with SWOT this much data will provide a novel dataset of $k_{600}$ on a scale never before possible.
 
@@
 This makes sense as Section 2 has effectively reduced $k_{600}$ to a function of hydraulics that are nearly all directly measurable by SWOT.
-Any resulting bias in BIKER's predictions is likely attributable to either bias in the priors used for the non-remotely sensed terms (equations 9-10) or in the $k_{600}$ model itself (the scaling coefficient in equation 7).
+Any resulting bias in BIKER's predictions is likely attributable to either bias in the priors used for the non-remotely sensed terms (equations 9-10) or in the $k_{600}$ model itself (equation 7, including the aforementioned wind errors).
 For SWOT discharge algorithms, authors have repeatedly shown that the 'quality' of prior river knowledge plays a large role in the success of discharge inversions [@frassonExploringFactorsControlling2021; @andreadisConstrainingAssimilationSWOT2020; @brinkerhoffConstrainingRemoteRiver2020; @tuozzoloEstimatingRiverDischarge2019] and our results here further corroborate this finding.
-It should be stressed that a substantial portion of rivers from [@frassonExploringFactorsControlling2021] are canal-shaped in nature with different hydraulic properties than a natural river channel.
-Nearly universally, these rivers underperformed (Figures 4, S3), however it is impossible to isolate whether that is due to channel geomorphology or the fact that these canals also usually had limited data for the Bayesian inference (only around 12 days).
 
 ### 4.3 Coupling BIKER with upscaling workflows
-Figure 8 confirms that BIKER is quite successful, when coupled with $CO_2$ data, at predicting annual upscaled carbon emissions from the river network.
-This encouraging result has three main implications for future work coupling remote sensing via SWOT with in situ data.
-These are outlined below.
+Figure 8 confirms that BIKER is quite successful at predicting annual upscaled carbon emissions from the river network when coupled with $CO_2$ data.
+This encouraging result has three main implications for future work.
 
-First, figure 8 directly implies that BIKER will be useful when coupled with large-scale $CO_{2_water}$ models [e.g.
-**Liu etal in review**; @saccardiImprovingPredictionsStream2021].
-These two models robustly predict reach-scale dissolved $CO_2$ concentrations using two different approaches: machine learning [**Liu et al in review**] and process-based reactive transport modeling [@saccardiImprovingPredictionsStream2021] but both models yield $CO_2$ estimates that would be spatially and temporally consistent with BIKER's output.
-Our promising results suggest that BIKER could provide additional (and directly observed) measurements of $k_{CO_2}$ to these models, thereby better informing model results through direct observations.
+First, figure 8 directly implies that BIKER will be useful when coupled with large-scale $CO_{2_water}$ models, provided these models are accurate.
+The models would give time and spcae varying gas exchange.
+**Liu etal in review** and @saccardiImprovingPredictionsStream2021 each propose models that robustly predict reach-scale dissolved $CO_2$ concentrations using two different approaches: machine learning [**Liu et al in review**] and process-based reactive transport modeling [@saccardiImprovingPredictionsStream2021], but both models yield $CO_2$ estimates that would be spatially and temporally consistent with BIKER's output.
+Our promising results suggest that BIKER could provide additional (and directly inferred) measurements of $k_{CO_2}$ to these models, thereby better informing model results through direct observations.
 This is likely to be accomplished via data assimilation which has proven useful in using remotly-sensed discharge to improve streamflow routing models [@fengRecentChangesArctic2021a; @ishitsukaCombiningOpticalRemote2020], and of which the @saccardiImprovingPredictionsStream2021 $CO_2$ model takes a similar form.
@@
 
-Finally, Figure 8 also uniquely allows us to directly compare the influence of one's HG model on total carbon emission rates from river networks, as all other calculations and parameters were held constant.
-Therefore, figure 8 highlights a potentially large source of uncertainty in current river $CO_2$ upscaling estimates: the HG models employed to scale river channel hydraulics with streamflow.
-As previously stressed, the only difference between the three in situ HG models and the observed estimate in Figure 8 is the specific HG model employed to predict river depth and velocity.
-This means that dramatically different carbon emission estimates are obtainable depending on the training data used for these HG models.
-While BIKER performs similarly to these methods, it must be stressed that the global representativeness of one's training data for HG models is of paramount importance.
+Finally, Figure 8 also uniquely allows us to directly compare the influence of geomorphic assumptions on total carbon emission rates from river networks, as all other calculations and parameters were held constant across our four tested models (Text S4).
+Therefore, figure 8 highlights a potentially large source of uncertainty in current river $CO_2$ upscaling estimates: the geomorphic models employed to scale river channel hydraulics with streamflow.
+In this case, the only difference between the three literature models and the observed estimate in Figure 8 is the specific HG model employed to predict river depth and velocity (Text S4, Table S3.
+**put some best case/worse case syle writing from the results here**.
 
@@
 Much of this uncertainty stems from uncertainty in gas exchange velocity $k_{600}$ at the global scale.
-In this paper, we propose that the soon-to-launch SWOT satellite will provide enough hydraulic measurements to analyse the temporal dynamics of $k_{600}$ using SWOT measurements, and therefore allow for a global-scale analysis of spatiotemporal trends in large-river $k_{600}$ once SWOT launches.
-In that context, we develop 1) a hydraulic model for $k_{600}$ that is nearly entirely SWOT observable and explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2,2)*100`% of variation in $k_{600}$, and 2) the BIKER algorithm to infer $k_{600}$ using no on-the-ground information.
-Validating on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers, we show show strong recovery of rivers' temporal $k_{600}$ dynamics and the total annual carbon emission rate across all `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers.
+In this paper, we propose that the soon-to-launch SWOT satellite will provide enough hydraulic measurements to analyse the temporal dynamics of $k_{600}$, and therefore allow for a global-scale analysis of spatiotemporal trends in large-river $k_{600}$ once SWOT launches.
+In that context, we developed 1) a hydraulic model for $k_{600}$ that is nearly entirely SWOT observable and explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2,2)*100`% of variation in $k_{600}$, and 2) the BIKER algorithm to infer $k_{600}$ using no on-the-ground information.
+Validating on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers, we show show strong recovery of rivers' temporal $k_{600}$ dynamics and a hypothetical total annual carbon emission rate across all `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers.
 These results suggest BIKER can be used to infer global-scale, near daily estimates of fluvial gas exchange velocity once SWOT launches in 2022.
@@
 C.B.
-Brinkerhoff was funded on **FINESST**.
+Brinkerhoff was funded by a NASA Future Investigators in Earth and Space Science Fellowship (#xxxxxx).
+C.J.
+Gleason was supported by NASA SWOT Science Team grant #yyyyyy.
 The version of BIKER used in this study will be available upon acception.

### 2022-02-14 supp_v9.Rmd
some ms and prepping for final validation

--- 2022-02-14 docs/supp_v8.Rmd
@@
 ---
-title: "Supplemental Information: Using hydraulic geometry and the SWOT satellite to remotely sense river gas exchange velocity"
+title: "Supplemental Information: Remotely sensing river greenhouse gas exchange velocity using the SWOT satellite"
 author:
@@
 
+## Text S4: Literature $CO_2$ flux models
+Here, we describe in more detail the $CO_2$ flux models used to compare against BIKER in section 3.2.2. The specific equations used are outlined in Table S3.
+
+In brief, gas fluxes ($FCO_2$) are only obtainable at the global scale via predictive equations for $k_{CO_2}$, as outlined in the main text.
+Here, we obtain $FCO_2$ using the in situ $CO_2$ data described in section 3.1 and $k_{CO_2}$ estimates calculated using 1) equation 7, 2) in situ streamflow data, and 3) hydraulic scaling equations used in the literature to predict river velocity and depth from discharge.
+Specifically, these literature equations include the 'Raymond 2012', 'Raymond 2013', and 'Brinkerhoff 2019' models.
+
+For example, the 'Raymond 2012' model uses the hydraulic scaling equations from @raymondScalingGasTransfer2012 to predict velocity and depth, which are in turn used to predict $k_{CO_2}$, which finally is used to predict $FCO_2$.
+Similarly, the 'Raymond 2013' model uses the equations outlined in @raymondGlobalCarbonDioxide2013 and the 'Brinkerhoff 2019' model uses new equations fit to the hydraulics dataset from @brinkerhoffReconcilingAtaStationAtManyStations2019.
+We chose to include the 'Brinkerhoff 2019' model as the training dataset is far larger than those used in either of the previous two models (Table S3: 104,624 versus 10,837 versus 1,026, respectively).
+Finally, we converted from $k_600$ to $k_{CO_2}$ via @raymondScalingGasTransfer2012.
+
 ## Figure S1
@@
 ## Table S3
-*Table S3: Details on the 3 depth hydraulic geometry models used to estimate FCO2 from the SWOT rivers (section 2.4).*
+*Table S3: Details on the 3 literature models for $CO_2$ fluxes. See Text S4 for more details.*
 
@@
 |--------|------------------|---------------------|---------------|-------------|
-|Brinkerhoff 2019|$D=0.265Q^{0.395}$|$\bar{U}=0.284Q^{0.191}$|530,945 measurements made across the United States at streamgauges|this study; @brinkerhoffReconcilingAtaStationAtManyStations2019|
+|Brinkerhoff 2019|$D=0.265Q^{0.395}$|$\bar{U}=0.284Q^{0.191}$|104,624 measurements made across the United States at streamgauges|brinkerhoffReconcilingAtaStationAtManyStations2019|
 |Raymond 2012|$D=0.409Q^{0.294}$|$\bar{U}=0.194Q^{0.285}$||1,026 measurements across the United States|@raymondScalingGasTransfer2012|

### 2022-02-14 src/swot_k_model_v3.R

@@ -109,9 +109,10 @@ data$ustar <- sqrt(g*data$Rh*data$slope) #friction/shear velocity [m/s]
 data$flag_swot <- ifelse(data$width >= 100, 'SWOT', 'Small')
 
 #########HYDRAULIC GEOMETRY OF SWOT-OBSERVABLE FLOWS
-HG_swot <- round(mean((data[data$flag_swot == 'SWOT',]$Rh/data[data$flag_swot == 'SWOT',]$depth)),2)
+HG_swot_mean <- round(mean((data[data$flag_swot == 'SWOT',]$Rh/data[data$flag_swot == 'SWOT',]$depth)),2)
+HG_swot_sd <- round(sd((data[data$flag_swot == 'SWOT',]$Rh/data[data$flag_swot == 'SWOT',]$depth)),2)
 n <- nrow(data[data$flag_swot == 'SWOT',])
-df <- data.frame('mean'=HG_swot, 'n'=n)
+df <- data.frame('mean'=HG_swot_mean, 'sd'=HG_swot_sd, 'n'=n)
 write.csv(df, 'cache/k600_theory/HG_swot.csv')
 
 #######LOG TRANSFORM SOME VARIABLES-----------------------------------------------------------
@@ -237,7 +238,7 @@ plot_reynolds_eD <- ggplot(hydraulicallyWide, aes(x=k600_pred_wideHydraulics, y=
   geom_point(size=5, color='#377eb8') +
   geom_abline(linetype='dashed', color='darkgrey', size=1.5)+ #1:1 line
   annotate("text", label = expression(paste(r^2, ': 0.70')), x = 1, y = 100, size = 8, colour = "#377eb8")+
-  labs(x = expression(bold(paste(beta[1]*(gS)^{7/16}*U^{1/4}*H^{9/16}, ' [', m, '/', dy, ']'))),
+  labs(x = expression(bold(paste(beta[1]*(gS)^{7/16}*U^{1/4}*H^{9/16}, ' [', m, '/', dy, ']', ' (eq. 7)'))),
        y = expression(bold(paste(k[600], ' [', m, '/', dy, ']'))))+
   scale_y_log10(limits=c(10^-1,10^2),
                 breaks=c(0.1, 1, 10, 100),

### 2022-02-24 manuscript_v9.Rmd
prepping to rerun with wider k600

@@
 We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers and 166 discrete measurements of river gas exchange, yielding an algorithm that predicts the $k_{600}$ timeseries solely from SWOT observations with a by-river median Kling-Gupta Efficiency of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$KGE), 2)`.
-**get more impressive skill score than KGE** Similar to Bayesian remote sensing algorithms for SWOT, BIKER is better at reproducing the temporal dynamics of gas exchange (median correlation coefficient of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$r), 2)`), than the absolute rates of exchange (median relative bias of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$rBIAS), 2)*100`%).
+Similar to Bayesian remote sensing algorithms for SWOT, BIKER is better at reproducing the temporal dynamics of gas exchange (median correlation coefficient of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$r), 2)`), than the absolute rates of exchange (median relative bias of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$MAE), 2)*100`%).
 Finally, we show BIKER is robust to measurement errors implicit in the SWOT data.
@@
 To answer this first question, we use hydraulic geometry- the fundamental geomorphic relationships between streamflow and channel shape [@leopoldHydraulicGeometryStream1953; @gleasonHydraulicGeometryNatural2015] to develop a process-based model for large-river (here defined as wider than 50m to align with SWOT) $k_{600}$.
-We then take these findings and explore the second question by implementing this hydraulic model, which in turn defines $k_{600}$, within an algorithm named BIKER ('Bayesian Inference of the $k_{600}$ Evasion Rate') to infer $k_{600$ solely from SWOT measurements.
+We then take these findings and explore the second question by implementing this hydraulic model, which in turn defines $k_{600}$, within an algorithm named BIKER ('Bayesian Inference of the $k_{600}$ Evasion Rate') to infer $k_{600}$ solely from SWOT measurements.
 The goal of BIKER is to require no in situ inputs of any kind (although in situ data could be ingested and would improve results) such that it is globally implementable on any SWOT-observable river.
@@
 Dashed grey line denotes the 1:1 line.
-Note that axes are plotted in logarithmic space just for visualization: model fit and validation (via the coefficient of determination $r^2$) were calculated in linear space.](cache/k600_theory/k600Plot.jpg)
+Note that axes are plotted in logarithmic space just for visualization: model fit and validation (via the coefficient of determination $r^2$) were calculated in linear space.](cache/k600_theory/fig2.jpg)
 
@@
 - Correlation coefficient *r* to quantify accuracy of BIKER's temporal dynamics
-- Root mean square error normalized by the observed mean (NRMSE) and prediction bias normalized by the observed mean (rBIAS) to assess bias
+- Root mean square error normalized by the observed mean (NRMSE) to assess bias
+- mean absolute residual error (MAE) to assess bias
 - Kling-Gupta Efficiency (KGE).
@@
 In that context, we sought to compare BIKER against established methods from the literature for upscaling carbon emissions from river networks (in this case, our `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])`  rivers) So, we pair the 26 biweekly $CO_2$ and water temperature samples from @beaulieuControlsGasTransfer2012 (section 3.1, figure S2) with every 14th set of daily SWOT observations (as the $CO_2$ data is bi-weekly).
-We then calculate $FCO_2$ using equation 1, an atmospheric $CO_{2_{air}}$ of 390 uatm, and a temperature dependent *Sc* was estimated following Raymond et al.
+We then calculate $FCO_2$ using equation 1, an atmospheric $CO_{2_{air}}$ of 400 uatm, and a temperature dependent *Sc* was estimated following Raymond et al.
 (2012).
@@
 (a-b) are poorest performing rivers, (b-c) are in the middle, and (e-f) are the best performing rivers.
-Y axis is normalized by maximum observed values to compare visually.](cache/validation/timeseries_noerr_short.jpg)
+Y axis is normalized by maximum observed values to compare visually.](cache/validation/fig4.jpg)
 
@@
 This indicates very strong inference of each river's temporal $k_{600}$ dynamics given that absolutely no in situ information is being used to predict $k_{600}$.
-NRMSE has a median score of `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$NRMSE, na.rm=T),2)` and median rBIAS is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$rBIAS, na.rm=T),2)`, highlighting that many rivers which have significant positive biases (Figure 4 also confirms this visually).
+NRMSE has a median score of `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$NRMSE, na.rm=T),2)`, highlighting that many rivers which have significant positive biases (Figure 4 also confirms this visually).
+Median MAE, however, is only `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$MAE, na.rm=T),2)` m/day.
 Taken in aggregate, Figures 4-5 indicate that BIKER is quite good at capturing temporal dynamics in $k_600$, however there is often positive bias in its estimates.
@@
 Each subpanel is labelled by its performance metric (defined in Table S2).
-Dashed lines denote median scores.](cache/validation/validation_by_river.jpg)
+Dashed lines denote median scores.
+Note the log10 scale for MAE.](cache/validation/fig5.jpg)
 
@@
 Note also that axes are flipped in order to visualize all 'better performances' in the upper-right-corner of each sub-plot.
-Black line denotes linear regression while dashed grey line is the 1:1 line.](cache/validation/errorPlot.jpg)
+Black line denotes linear regression to aid in visualization.](cache/validation/fig6.jpg)
 
-**something something temporal variability**
-You need to flesh this out a bit better and stop leaning so heavily on temporality.
-Feels almost like we're hiding something, which we aren't Colin was not a fan of the prior vs posterior CV eCDFs and also suggested that this should still hold even if we are doing just magntiude.
+Finally, we sought to further explore both BIKER's ability to infer temporal trends in $k_{600}$, as well as the influence of prior error on $k_{600}$.
 
-Given the promising *r* scores across all 47 rivers in Figure 5, we further explore BIKER's ability to infer temporal trends in $k_{600}$.
 Figure 7a plots a histogram of the slopes of linear regressions between BIKER-predicted $k_{600}$ and observed $k_{600}$.
-A slope of 1 indicates that BIKER correctly infers the daily flucuations in gas exchange (even if there is bias implicit in these estimates), while a slope well above or below 1 indicates incorrect inference of the temporal dynamics of $k_{600}$.
+A slope of 1 indicates that BIKER correctly infers the daily fluctuations in gas exchange (even if there is bias implicit in these estimates), while a slope well above or below 1 indicates incorrect inference of the temporal dynamics of $k_{600}$.
 Figure 6a shows that the majority of rivers' slopes approximate 1 (median slope: `r round(median(results_dynamics$Slope),2)`), indicating strong inference of temporal gas exchange trends.
@@
 Slopes equaling 1 indicate correct inference of the temporal trends in daily $k_{600}$ (regardless of bias in the estimates).
-b) Empirical cummulative density distributions of the coefficient of varation of prior, posterior, and observed $k_{600}$ per river.
-If BIKER is correctly inferring temporal variation in $k_{600}$, the posterior eCDF will approximate the observed eCDF (as it does).](cache/validation/validation_temporal.jpg)
+b) Empirical cummulative density distributions of prior and posterior mean absolute error (MAE).](cache/validation/fig7.jpg)
 
-Finally, we explore whether BIKER improves upon the prior with respect to temporal variation in $k_{600}$, here defined using the coefficient of variation ($CV=\frac{SD}{mean}$).
-This is an essential test of a Bayesian method to determine if SWOT observations have improved our knowledge or if our initial guess proved statistically most useful.
-Figure 6b plots eCDFs of observed, prior, and posterior $k_{600}$ *CV*, strongly showing that BIKER improves upon the prior information available to us about $k_{600}$.
-Posterior $k_{600}$ *CV* approximates the observed eCDF and sufficiently 'closes the gap' between our prior knowledge on $k_{600}$ temporal variation and the actually observed variation in $k_{600}$.
-It must be stressed that this validation was performed using a relatively uninformed prior on $k_{600}$ (Text S3) and that this gap may not be as large in many settings with better prior knowledge.
-However, Figure 6b confirms that Bayesian techniques can be used to infer temporal variation in $k_{600}$ when good prior or in situ data are not available.
+Figure 7b comes prior error for $k_{600}$ against posterior error in $k_{600}$.
+We used the mean absolute error (Table S2), which is a metric that captures the average residual error in predictions, to assess this and plotted the results as eCDFS across all `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers in the 'no-error' scenario.
+In Figure 7b, we see that BIKER reduces MAE, when the prior estimate was already quite good (approximately less than 3 m/day), while BIKER increases MAE if the prior estimate was already quite poor (again, approximately > 3 m/day).
+This indicates that our prior estimate of $k_{600}$ is too certain, i.e.
+BIKER is convinveed that the prior estimate is reasonable and, when combined with the hydraulics inference also in the model, yields a net worse performance.
+TO perturb this, we reran our BIKER validation using a less certain $k_{600}$ prior.
+SHOULD THIS BE SI OR JUST BE THE ACTUAL STUDY
 
@@
 ![Figure 8: Yearly carbon emissions rate via $CO_2$ evasion across all rivers.
-Completely remotely-sensed methods are colored in red, in situ methods in purple, and the observed in orange.](cache/FCO2/FCO2_models.jpg)
+Completely remotely-sensed methods are colored in red, in situ methods in purple, and the observed in orange.
+Note that confidence intervals were not calculated for the remaining two models because their uncertainties are not available.](cache/FCO2/fig8.jpg)
 

### 2022-02-24 supp_v9.Rmd
prepping to rerun with wider k600

@@
 That dataset has over 500,000 discrete measurements of river width, velocity, area, and discharge that were made by the United States Geological Survey (USGS) to calibrate streamgauge rating curves.
-Here, we describe how this dataset was filtered down to 171,553 measurements and how we quantified what was 'hydraulically wide'.
+Here, we describe how this dataset was filtered down to `r HG_swot$n` rivers.
 
@@
 Consult Appendix A for all variable definitions.
-Note that $\alpha$ and $\beta$ are statistical parameters obtained via least squares regression for the small-eddy and Reynolds models, respectively.
-Subscripts denote variants of the same parameter, depending on which $\epsilon$ model is employed.
+Note that $\alpha$ and $\beta$ are statistical parameters obtained via least squares regression for the small-eddy and Reynolds-extension models, respectively.
+Subscripts denote variants of the same parameter, depending on which $\epsilon$ model is employed (see Main Text).
 
@@
 
-*log-law-of-the-wall dissipation*
+$\epsilon = \epsilon_S$
+
 $$k_{600}=\alpha(\epsilon_S)^{1/4} \mathbf{(S1)}$$
@@
 
-*form-drag dissipation*
+$\epsilon = \epsilon_D$
+
 $$k_{600}=\alpha_1(\epsilon_D)^{1/4} \mathbf{(S6)}$$
@@
 
-*log-law-of-the-wall dissipation*
+$\epsilon = \epsilon_S$
+
 $$k_{600}=\beta_0(\epsilon_S)^{1/4}A_p^{3/4} \mathbf{(S8)}$$
@@
 
-*form-drag dissipation*
+$\epsilon = \epsilon_D$
+
 $$k_{600}=\beta_0(\epsilon_D)^{1/4}A_p^{3/4} \mathbf{(S14)}$$
@@
 ## Text S3: BIKER hyperparameterization
-In this text we explain in detail how BIKER's prior distributions were set for each river.
+In this text we explain in detail how BIKER's prior distributions were determined for a given river.
 
@@
 For BIKER, prior distriubtions are formalized as truncated normal distributions of the log-transformed terms such that $log(X) \sim  \mathcal{N}(\mu, \sigma^2)$ for $\lambda < log(X) < \gamma$, using prior hyperparameters mean ($\mu$), standard deviation ($\sigma$), and upper ($\gamma$) and lower bounds ($\lambda$) for any parameter *X*.
-It is important to again stress, as we do in the main text, that BIKER prior hyperparameters are described using only SWOT data to be completely globally implementanble.
-This is analogous to the 'empirical Bayes method' [@hoffFirstCourseBayesian2009].
+It is important to again stress, as we do in the main text, that BIKER prior hyperparameters are described using only SWOT data to be completely globally implementable.
+This use of the data to describe the priors is analogous to the 'empirical Bayes method' [@hoffFirstCourseBayesian2009].
 
-$A_0_i$ and $n_i$ prior hyperparameters were assigned following an updated version of the method developed by @brinkerhoffConstrainingRemoteRiver2020.
+$A_0_i$ and $n_i$ prior hyperparameters were assigned following @brinkerhoffConstrainingRemoteRiver2020.
 They developed a set of river channel prior hyperparameters for McFLI algorithms that are entirely RS-able and reflect differential channel hydraulics as a function of river geomorphology.
@@
 This leaves the $k_{600_t}$ hyperparameters to be defined.
-$k_{600_{\mu}}$ is set by invoking the hydraulic geometry (HG) relationships developed in section 3.2.3 of the main text using the data from @brinkerhoffReconcilingAtaStationAtManyStations2019.
-We replaced both depth and velocity terms from our gas exchange model (equation 6 in the main text) with these HG models, resulting in equation S19 where Q is the mass-conserved streamflow for the river reach.
+$k_{600_{\mu}}$ is set by invoking the hydraulic geometry (HG) relationships developed in section 2.2.1 of the main text using the data from @brinkerhoffReconcilingAtaStationAtManyStations2019.
+We replaced both depth and velocity terms from our gas exchange model (equation 7 in the Main Text) with these HG models, resulting in equation S19 where Q is the mass-conserved streamflow for the river reach.
 
@@
 Using a temporally-invariant estimate of streamflow is the worst case scenario and BIKER's performance will improve with a more informed prior on streamflow (and therefore $k_{600}$).
-This was recently shown on many of these rivers for SWOT discharge algorithms [@frassonExploringFactorsControlling2021].
-However as noted in the main text, our primary goal with this initial validation is to benchmark BIKER's worst case scenario for performance and so we do that here.
+However as noted in the Main Text, our primary goal with this initial validation is to benchmark BIKER's worst case scenario for performance and so we do that here.
 
@@
 
-In the scenario that BIKER is run on real SWOT data, $\sigma_k_{600}$ must reflect the full uncertainty implicit in equation 8 in the main text.
+In the scenario that BIKER is run on real SWOT data, $\sigma_k_{600}$ must reflect the full uncertainty implicit in equation 9 in the main text.
 This means we must also account for uncertainty from the $\beta_1$ parameter.
@@
 
-In brief, gas fluxes ($FCO_2$) are only obtainable at the global scale via predictive equations for $k_{CO_2}$, as outlined in the main text.
-Here, we obtain $FCO_2$ using the in situ $CO_2$ data described in section 3.1 and $k_{CO_2}$ estimates calculated using 1) equation 7, 2) in situ streamflow data, and 3) hydraulic scaling equations used in the literature to predict river velocity and depth from discharge.
-Specifically, these literature equations include the 'Raymond 2012', 'Raymond 2013', and 'Brinkerhoff 2019' models.
+In brief, gas fluxes ($FCO_2$) are only obtainable at the global scale via predictive equations for $k_{CO_2}$ per equation 2.
+Here, we obtain $FCO_2$ using the in situ $CO_2$ data described in section 3.1 and $k_{CO_2}$ estimates.
+The $k_{CO_2}$ estimates are calculated using 1) equation 7, 2) in situ streamflow data, and 3) hydraulic scaling equations from the literature that predict river velocity and depth from streamflow.
+Specifically, these literature equations include the 'Raymond 2012', 'Raymond 2013', and 'Brinkerhoff 2019' models (Table S3).
 
-For example, the 'Raymond 2012' model uses the hydraulic scaling equations from @raymondScalingGasTransfer2012 to predict velocity and depth, which are in turn used to predict $k_{CO_2}$, which finally is used to predict $FCO_2$.
-Similarly, the 'Raymond 2013' model uses the equations outlined in @raymondGlobalCarbonDioxide2013 and the 'Brinkerhoff 2019' model uses new equations fit to the hydraulics dataset from @brinkerhoffReconcilingAtaStationAtManyStations2019.
+For example, the 'Raymond 2012' model uses the hydraulic scaling equations from @raymondScalingGasTransfer2012 to predict velocity and depth, which are in turn used to calculate a $k_{CO_2}$, which finally is used to calculate $FCO_2$.
+Similarly, the 'Raymond 2013' model uses the equations outlined in @raymondGlobalCarbonDioxide2013 and the 'Brinkerhoff 2019' model uses new equations fit to the hydraulics dataset from @brinkerhoffReconcilingAtaStationAtManyStations2019 following Text S1.
 We chose to include the 'Brinkerhoff 2019' model as the training dataset is far larger than those used in either of the previous two models (Table S3: 104,624 versus 10,837 versus 1,026, respectively).
-Finally, we converted from $k_600$ to $k_{CO_2}$ via @raymondScalingGasTransfer2012.
+Finally, we converted from $k_600$ to $k_{CO_2}$ following @raymondScalingGasTransfer2012.
 
@@
 ![Figure S1: Four $k_{600}$~hydraulics models tested in this study, with model assumptions noted in the sub-panel titles.
-Panel (d) is also presented in the main text as Figure 2.](cache/k600_theory/k600Plot_SI.jpg)
+Panel (d) is also presented in the main text as Figure 2 and equation 7.](cache/k600_theory/figS1.jpg)
 
 ## Figure S2
-![Figure S2 Timeseries of the biweekly CO2 data from @beaulieuControlsGasTransfer2012.
+![Figure S2 Timeseries of the biweekly $CO_2$ data from @beaulieuControlsGasTransfer2012.
 Sampling took place 2008-2009 in the Ohio River (upstream of Cincinnati, Ohio, United States).
-Dashed black line denotes atmospheric CO2 at 400 uatm.](cache/FCO2/Beaulieu_timeseries.jpg)
+Dashed black line denotes atmospheric $CO_2$ at 400 uatm.](cache/FCO2/figS2.jpg)
 
@@
 ![Figure s3: Observed (orange) versus BIKER-inferred (green) timerseries of daily $k_{600}$ for the 47 SWOT-simulated rivers with no additional SWOT errors.
-Note that the y axis is normalized by maximum observed values to compare across rivers.](cache/validation/timeseries_noerr.jpg)
+Note that the y axis is normalized by maximum observed values to compare across rivers.](cache/validation/figS3.jpg)
 
@@
 (2021) error model.
-Note that the y axis is normalized by maximum observed values to compare across rivers.](cache/validation/timeseries_err.jpg)
+Note that the y axis is normalized by maximum observed values to compare across rivers.](cache/validation/figS4.jpg)
 
@@
 Any data wrangling was done by those authors.
-'Workers' refers to who actually made the measurements.
+'Workers' refers to who actually made the measurements in the field.
 The Raymond et al.
@@
 σ refers to the variance of the sample and μ refers to the mean of the sample.
-As is standard, a carrot accent indicates the predicted value.*
+Carrot accents indicate the predicted value.*
 
-|**Description**|**Acronym**|**Definition**|**Ideal Score**|**Possible Range**|**Validation Scheme**|
-|-----------|-------|----------|-----------|--------------|-----------------|
-|Correlation Coefficient|$r$| $\sqrt{1-(\frac{\sum\limits_{i=1}^{N_t}{(k_{600, i}-\hat{k_{600}})^2}}{\sum{(k_{600, i}-\bar{k_{600}})^2}})}$ |1|-1 to 1|	All rivers and all timesteps|
-|Root-mean-square-error|RMSE| $\sqrt{\frac{1}{N_t}\sum\limits_{i=1}^{N_t}(\hat{k_{600, i}}-k_{600, i})^2}$ |0|0 to ∞|All rivers and all timesteps|
-|Relative root-mean-square error|RRMSE| $\sqrt{\frac{1}{N_t}\sum\limits_{i=1}^{N_t}(\frac{\hat{k_{600, i}}-k_{600, i}}{k_{600, i}})^2}$ |0|0 to ∞|By river|
-|Normalized root-mean-square error|NRMSE| $\sqrt{\frac{1}{N_t}\sum\limits_{i=1}^{N_t}(\frac{\hat{k_{600, i}}-k_{600, i}}{\bar{k_{600, i}}})^2}$ |0|0 to ∞|By river|
-|Relative bias|rBIAS| $\frac{1}{N_t}\sum\limits_{i=1}^{N_t}(\frac{\hat{k_{600, i}}-k_{600, i}}{\bar{k_{600, i}}})$ |0|-∞ to ∞|By river|
+|**Description**|**Acronym**|**Definition**|**Ideal Score**|**Possible Range**|
+|---------------|-----------|--------------|---------------|------------------|
+|Correlation Coefficient|$r$| $\sqrt{1-(\frac{\sum\limits_{i=1}^{N_t}{(k_{600, i}-\hat{k_{600_i}})^2}}{\sum{(k_{600, i}-\bar{k_{600}})^2}})}$ |1|-1 to 1|
+|Normalized root-mean-square error|NRMSE| $\sqrt{\frac{1}{N_t}\sum\limits_{i=1}^{N_t}(\frac{\hat{k_{600, i}}-k_{600, i}}{\bar{k_{600, i}}})^2}$ |0|0 to ∞|
+|Kling-Gupta Efficiency|KGE|$1 - \sqrt{ (r(k_{600_i}, \hat{k_{600_i}}) - 1)^2 + (\frac{\mu_{\hat{k_{600_i}}} }{\mu_{k_{600_i}}} - 1)^2 + (\frac{ \sigma _{\hat{k_{600_i}}} / \mu _{\hat{k_{600_i}}} }{ \sigma_{k_{600_i}} / \mu_{k_{600_i}}} - 1)^2 }$|1|-∞ to 1|
+|Mean Absolute Error|MAE| $\frac{\sum\limits_{i=1}^{N_t}(|\hat{k_{600_i}}-k_{600_i})|)}{n})$ |0|0 to ∞|
 
@@
 |--------|------------------|---------------------|---------------|-------------|
-|Brinkerhoff 2019|$D=0.265Q^{0.395}$|$\bar{U}=0.284Q^{0.191}$|104,624 measurements made across the United States at streamgauges|brinkerhoffReconcilingAtaStationAtManyStations2019|
+|Brinkerhoff 2019|$D=0.265Q^{0.395}$|$\bar{U}=0.284Q^{0.191}$|104,624 measurements made across the United States at streamgauges|@brinkerhoffReconcilingAtaStationAtManyStations2019|
 |Raymond 2012|$D=0.409Q^{0.294}$|$\bar{U}=0.194Q^{0.285}$||1,026 measurements across the United States|@raymondScalingGasTransfer2012|

### 2022-02-24 src/runFCO2.R

@@ -216,7 +216,7 @@ beaulieu <- ggplot(data=Data, aes(y=CO2_uatm, x=Date)) +
         axis.title=element_text(size=24,face="bold"),
         legend.text = element_text(size=17),
         axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))
-ggsave('cache/FCO2/Beaulieu_timeseries.jpg', beaulieu, width=15, height=6)
+ggsave('cache/FCO2/figS2.jpg', beaulieu, width=15, height=6)
 
 #BIKER results for the SWOT rivers (only those with no measurement error)--------------------------------------------
 BIKER_results <- read.csv('cache/validation/BIKER_validation_results.csv')

### 2022-02-24 src/swot_k_model_v3.R

@@ -231,7 +231,7 @@ plot_reynolds_eD <- ggplot(hydraulicallyWide, aes(x=k600_pred_wideHydraulics, y=
 
 k600_modelPlot_SI <- plot_grid(plot_smallEddy_eS, plot_reynolds_eS, plot_smallEddy_eD, plot_reynolds_eD, ncol=2, label_size = 18, labels=c('a', 'b', 'c', 'd'))
 
-ggsave('cache\\k600_theory\\k600Plot_SI.jpg', k600_modelPlot_SI, height=9, width=10)
+ggsave('cache\\k600_theory\\figS1.jpg', k600_modelPlot_SI, height=9, width=10)
 
 #### SAVE JUST THE FINAL MODEL---------------------------------------
 plot_reynolds_eD <- ggplot(hydraulicallyWide, aes(x=k600_pred_wideHydraulics, y=k600)) +
@@ -252,7 +252,7 @@ plot_reynolds_eD <- ggplot(hydraulicallyWide, aes(x=k600_pred_wideHydraulics, y=
         legend.text = element_text(size=17),
         legend.title = element_text(size=17, face='bold'),
         legend.position = 'none')
-ggsave('cache\\k600_theory\\k600Plot.jpg', plot_reynolds_eD, height=6, width=6)
+ggsave('cache\\k600_theory\\fig2.jpg', plot_reynolds_eD, height=6, width=6)
 
 #######WRITE reynolds MODEL TO FILE-------------------------------
 models <- data.frame('name'=c('reynolds-eS', 'Small-eddy-eS', 'reynolds-eD', 'Small-eddy-eD'),

### 2022-03-09 docs/manuscript_v9.Rmd
manuscript v10

--- 2022-02-24 manuscript_v9.Rmd
@@
 Much of this uncertainty stems from uncertainty in gas exchange velocity $k_{600}$ (the rate of this diffusive transport).
-We propose that the NASA/CNES/USKA/CSA SWOT satellite, set to usher in a new age of hydrology research at the global scale, can provide similarly transformative insights to fluvial gas exchange modeling upon launch in 2022.
+We propose that the NASA/CNES/USKA/CSA SWOT satellite can provide new insights to fluvial gas exchange modeling upon launch in 2022.
 Here, we present work inferring $k_{600}$ from synthetic future SWOT observations without in situ calibration.
-We exploit the unique geomorphology of SWOT-observable rivers to develop a physical model of gas exchange that is remotely sensible and explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2,2)*100`% of variation in $k_{600}$ We then couple this model with established Bayesian inference techniques to develop BIKER, the 'Bayesian Inference of the $k_{600}$ Exchange Rate' algorithm.
-We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers and 166 discrete measurements of river gas exchange, yielding an algorithm that predicts the $k_{600}$ timeseries solely from SWOT observations with a by-river median Kling-Gupta Efficiency of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$KGE), 2)`.
-Similar to Bayesian remote sensing algorithms for SWOT, BIKER is better at reproducing the temporal dynamics of gas exchange (median correlation coefficient of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$r), 2)`), than the absolute rates of exchange (median relative bias of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$MAE), 2)*100`%).
-Finally, we show BIKER is robust to measurement errors implicit in the SWOT data.
-With SWOT set to launch in late 2022, we suggest that BIKER will be useful in mapping the global-scale spatiotemporal dynamics of fluvial gas exchange in large rivers.
+We exploit the unique geomorphology of SWOT-observable rivers to develop a physical model of gas exchange that is remotely sensible and explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2,2)*100`% of variation in $k_{600}$ We then couple this model with established techniques to develop BIKER, the 'Bayesian Inference of the $k_{600}$ Exchange Rate' algorithm.
+We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers and 166 measurements of river gas exchange, yielding an algorithm that predicts the $k_{600}$ timeseries solely from SWOT observations with a by-river median Kling-Gupta Efficiency of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$KGE), 2)`.
+Similar to remote sensing algorithms for SWOT, BIKER is better at reproducing the temporal dynamics of gas exchange (median correlation coefficient of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$r), 2)`), than the absolute rates of exchange (median normalized RMSE of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$NRMSE), 2)*100`%).
+Finally, BIKER is robust to measurement errors implicit in the SWOT data.
+Upon SWOT's launch, we suggest that BIKER will be useful in mapping the global-scale spatiotemporal dynamics of fluvial gas exchange in large rivers.
 
@@
 This work has lead to recent river-reach explicit modeling of $\delta CO_2$ using global hydrography datasets at up to monthly temporal resolutions [@horgbyUnexpectedLargeEvasion2019; @brinkerhoffLakeMorphometryRiver2021; @saccardiImprovingPredictionsStream2021; *Liu accepetd*], but an equivalently sophisticated representation of $k_{CO_2}$ is still lacking.
-As equations 1 and 2 dictate, $\delta CO_2_{water}$ and $k_{CO_2}$ share the burden of calculating $FCO_2$ and therefore these next-generation models will have to contend with this discrepancy, which is directly dictated by the lack of direct measurements of global river hydraulics.
 
@@
 
-Applying these *k* models across watersheds, regions, or continents is called ‘upscaling.’ This upscaling gives the impossible-to-measure *k* term in Equation 1 for any arbitrary number of rivers, but also changes the base parameters that ultimately control the final estimate of $FCO_2$.
+Applying these *k* models across watersheds, regions, or continents is called ‘upscaling.’ This upscaling allows for estimating the impossible-to-measure *k* term in Equation 1 for any arbitrary number of rivers, but also changes the base parameters that ultimately control the final estimate of $FCO_2$.
 That is, by making *k* a function of hydraulics, $FCO_2_{upscaled}$ is now a direct function of river hydraulics.
@@
 
-Global upscalng has been attempted inthe literature.
+Global upscalng has been performed in the literature.
 @raymondGlobalCarbonDioxide2013, @lauerwaldSpatialPatternsCO22015, and @horgbyUnexpectedLargeEvasion2019 all relied on $k_{CO_2}$ values indirectly estimated using mean annual streamflow models and hydraulic scaling equations to predict the hydraulic terms to in turn predict $k_{CO_2}$, while @borgesGloballySignificantGreenhousegas2015 used a combination of the above method and a constant $k_{CO_2}$ in space and time to upscale over Africa.
@@
 
-@wangPhysicallyBasedScaling2021 recently attempted to address this global *k* problem by simulating $k_{O_2}$ in 35 rivers of many sizes (widths ranging from 0.23–349m) using a stream metabolism model [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets to infer what $K_{O_2}$ must have been to produce their 'observations'.
+@wangPhysicallyBasedScaling2021 recently attempted to address this global *k* problem by simulating $k_{O_2}$ in 35 rivers of many sizes (widths ranging from 0.23–349m) using a stream metabolism model [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets to infer what $k_{O_2}$ must have been to produce their 'observations'.
 They then compared this simulated dataset against direct measurements of *k*, finding similar performance and parameter values against process-based models of gas evasion.
@@
 
-We have established that literature has a reasonably good understanding of $\delta CO_2$ and a poor understanding in $k_{600}$ (and therefore $k_{CO_2}$) across large areas and in time.
+We have established that literature has a reasonably good understanding of $\delta CO_2$ and a relatively poorer understanding of $k_{600}$ (and therefore $k_{CO_2}$) across large areas and in time.
 In theory, the discrepancy between the quality of our $\delta CO_2_{water}$ and $k_{CO_2}$ estimates could be alleviated if direct hydraulics measurements (and in turn $k_{CO_2}$ via equation 2) were available at the global scale at a sufficient temporal resolution.
@@
 To answer this first question, we use hydraulic geometry- the fundamental geomorphic relationships between streamflow and channel shape [@leopoldHydraulicGeometryStream1953; @gleasonHydraulicGeometryNatural2015] to develop a process-based model for large-river (here defined as wider than 50m to align with SWOT) $k_{600}$.
-We then take these findings and explore the second question by implementing this hydraulic model, which in turn defines $k_{600}$, within an algorithm named BIKER ('Bayesian Inference of the $k_{600}$ Evasion Rate') to infer $k_{600}$ solely from SWOT measurements.
+We then take these findings and explore the second question by implementing this hydraulic model, which defines $k_{600}$, within an algorithm named BIKER ('Bayesian Inference of the $k_{600}$ Evasion Rate') to infer $k_{600}$ solely from SWOT measurements.
 The goal of BIKER is to require no in situ inputs of any kind (although in situ data could be ingested and would improve results) such that it is globally implementable on any SWOT-observable river.
 We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers (as SWOT has not yet launched) and explore BIKER's robustness to the expected measurement errors implicit in the satellite's observations.
-Finally, we also couple BIKER's $k_{600}$ estimates with $\delta CO_2$ to predict gas fluxes and compare these against established literature methods that are reliant on in situ hydraulic measurements.
+Finally, we also couple BIKER's $k_{600}$ estimates with $\delta CO_2$ to predict gas fluxes and compare these against established literature methods that rely on in situ hydraulic measurements.
 
@@
 We conclude with a discussion (section 4) that encompasses both gas exchange theory and remote sensing.
-We also provide a flowchart detailing the entire study as Figure 1.
+Figure 1 conceptually maps out the algorithm's approach to inferring $k_{600}$ from SWOT data.
 
-![Figure 1: Flowchart detailing the BIKER algorithm.
-Dashed lines indicate workflow to calculate the observed data that we validate against.
-See Appendix A for variable nomenclature used in this study.](cache/flowchart_2.jpg)
+![Figure 1: Conceptual overview of the BIKER algorithm.
+SWOT will directly measure water surface width and elevation (and thus slope), and the BIKER algorithm uses those observations to infer gas exchange velocity $k_{600}$ via a process-based model (section 2).](cache/flowchart3.jpg)
 
@@
 Given the theoretical context provided in Section 2.2, we now turn to SWOT-observable systems specifically.
-Rivers and streams change predictably along their longitudinal profile from headwater to ocean, and we can exploit the hydraulic geometry of large rivers at the end of this continumn to estimate *k* in SWOT-observable systems.
+Rivers and streams change predictably along their longitudinal profile from source to sea, and we can exploit the hydraulic geometry of large rivers at the end of this continumn to estimate *k* in SWOT-observable systems.
 In general, as river size increases, channels become more rectangular, their shapes elongate (becoming wider quicker than they become deeper) and their hydraulic radii begin to approximate their mean flow depth, i.e.
-$R_h \approx H$ **[CITE]**.
-This is a common assumption in hydraulic and geomorphic modeling of large rivers, and flows in a SWOT-observable river flow have an average $\frac{R_h}{H}$ ratio of `r HG_swot$mean` and standard deviation of `r HG_swot$sd` (n = `r HG_swot$n`; see Text S1 for how we built this dataset).
+$R_h \approx H$ [@leopoldHydraulicGeometryStream1953].
+This is a common assumption in hydraulic and geomorphic modeling of large rivers.
+For example, SWOT-observable river flows have an average $\frac{R_h}{H}$ ratio of `r HG_swot$mean` and standard deviation of `r HG_swot$sd` (n = `r HG_swot$n`; see Text S1 for how we built this dataset).
 We refer to these rivers as 'hydraulically-wide'.
@@
 To do this, we impose $\epsilon=\epsilon_D$ and $R_H=H$ on the Reynolds extension model and arrive at equation 7 (with statistical coefficient $\beta_1$).
-Equation 7 thus define gas exchange velocity solely as a function of slope, mean flow depth, and mean flow velocity.
+Equation 7 thus defines gas exchange velocity solely as a function of slope, mean flow depth, and mean flow velocity.
 This is theoretically valid only in a hydraulically-wide channel.
@@
 We also test the performance of three other models for predicting *k* in hydraulically-wide channels via the other three unique combinations of equations 3-4 and equations 5-6.
-T While the complete model derivations and results for the four models are provided in Text S2 and Figure S1, the final and best-performing model (equation 7) is presented and used below.
+While the complete model derivations and results for all four models are provided in Text S2 and Figure S1, the final and best-performing model (equation 7) is presented and used below.
 
 ### 2.4 Model validation
-With equation 7 derived, we now empirically its strength of fit for hydraulically-wide river flows.
+With equation 7 derived, we now empirically test its strength of fit in hydraulically-wide river flows.
 We validate on the dataset of in situ measurements of $k_{600}$, after filtering for measurements made in hydraulically-wide channels, which was defined as flows whose hydraulic radius was within 1% of their mean flow depth.
@@
 ## 3 Exploiting equation 7 to remotely sense gas exchange velocity
-We have shown that scaling $k_{600}$ via equation 7 should provide a reliable basis for hydraulically-wide rivers with a defined error structure.
+We have shown that scaling $k_{600}$ via equation 7 is useful in hydraulically-wide rivers.
 Further, equation 7 has only three non-directly-remotely-sensible terms: $k_{600}$, mean flow depth, and mean flow velocity.
 Conveniently, techniques to simultaneously infer mean flow depth and velocity from SWOT data (among other parameters) have been established over the last decade to infer streamflow from SWOT's measurements [e.g.
-**GLEASON & SMITH 2014**; @andreadisConstrainingAssimilationSWOT2020; @brinkerhoffConstrainingRemoteRiver2020; @durandEstimatingReachaveragedDischarge2014; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014a; @garamboisVariationalEstimationEffective2020; @brissetAssimilationAltimetricData2018; @hagemannBAMBayesianAMHGManning2017; @larnierRiverDischargeBathymetry2020a; @oubanasDischargeEstimationUngauged2018].
+@gleasonGlobalMappingRiver2014a; @andreadisConstrainingAssimilationSWOT2020; @brinkerhoffConstrainingRemoteRiver2020; @durandEstimatingReachaveragedDischarge2014; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014a; @garamboisVariationalEstimationEffective2020; @brissetAssimilationAltimetricData2018; @hagemannBAMBayesianAMHGManning2017; @larnierRiverDischargeBathymetry2020a; @oubanasDischargeEstimationUngauged2018].
 For BIKER, we follow the work developed by @hagemannBAMBayesianAMHGManning2017, @brinkerhoffConstrainingRemoteRiver2020, and @durandEstimatingReachaveragedDischarge2014 to infer $k_{600}$, channel depth and velocity from SWOT observations using a modified form of Manning's equation.
@@
 ### 3.1 Data
-![Figure 3: Map of the 47 hydraulic models and 1 timeseries of $CO_2$ samples used in this study.
-Note that hydraulic model locations are approximate as some of the models are not geo-referenced.](cache/maps/map3.jpg)
-
 To validate BIKER, we cannot use actual SWOT measurements as SWOT has yet to launch.
@@
 We use 47 SWOT-simulated rivers for validation, where these simulated rivers are simply reach-averaged hydraulic model outputs where the water surface heights, slopes, and widths are labelled as RS observations and are used as the sole inputs to BIKER.
-These datasets are created using standard hydraulic models forced with known inflows and measured bathymetry to model the hydraulic response of the rivers, and then those terms visible to SWOT were extracted to produce hydraulically realistic synthetic observations.
-These data were published by @frassonExploringFactorsControlling2021 and @durandIntercomparisonRemoteSensing2016 as benchmarking datasets to explore remote sensing of discharge (RSQ) algorithm performance for the SWOT mission.
+These datasets are created using standard hydraulic models forced with known inflows and measured bathymetry to model the hydraulic response of the rivers, and then those terms visible to SWOT are extracted to produce hydraulically realistic synthetic observations.
+These data were published by @frassonExploringFactorsControlling2021 and @durandIntercomparisonRemoteSensing2016.
+See Figure 3 for a map of these river's locations.
+
+![Figure 3: Map of the 47 hydraulic models and 1 timeseries of $CO_2$ samples used in this study.
+Note that hydraulic model locations are approximate as some of the models are not geo-referenced.](cache/maps/map3.jpg)
 
@@
 
-For $CO_2$ evasion and carbon emissions calculations, we use 26 bi-weekly dissolved $CO_2$ samples collected by @beaulieuControlsGasTransfer2012 at one location in the Ohio River for one calender year from 2008-2009 (Figure S2).
+For $CO_2$ evasion and carbon emissions calculations, we use 26 bi-weekly dissolved $CO_2$ samples collected by @beaulieuControlsGasTransfer2012 at one location in the Ohio River (Figure 3) for one calender year from 2008-2009 (Figure S2).
 Note that this $CO_2$ data is for the Ohio River only but was applied to all rivers to provide a physically realistic signal for $CO_2$ fluxes with meaningful seasonality and dynamics.
@@
 Following section 2.3, we continue to assume that the channel is hydraulically-wide ($R_h = H =\frac{A}{W}$).
-To leverage additional SWOT data, the wetted channel area *A* is further split into the the SWOT-unobservable portion $A_0$ and SWOT-observable portion $dA$ following @durandEstimatingReachaveragedDischarge2014 and @hagemannBAMBayesianAMHGManning2017 where $dA_{it} = \sum_{t':W_{t'}\le W_t}{W_{t'} \delta H_{t'}}$ for cross-section *i* and timestep *t*.
+To leverage additional SWOT data, the wetted channel area *A* is further split into the the SWOT-unobservable portion $A_0$ and SWOT-observable portion $dA$ following @durandEstimatingReachaveragedDischarge2014 and @hagemannBAMBayesianAMHGManning2017 where $dA_{it} = \sum_{t':W_{t'}\le W_t}{W_{t'} \delta H_e_{t'}}$ for cross-section *i* and timestep *t* within a mass-conserved river reach.
 
@@
 Therefore, the validation presented here is a 'worst-case scenario', wherein BIKER performance should improve with better prior information on the river.
-In that context, we used a variation of the prior specification method developed by @brinkerhoffConstrainingRemoteRiver2020, who developed geomorphic 'river types' with distinct prior sets for $A_0_i$ and $n_i$.
+In that context, we used a variation of the prior specification method developed by @brinkerhoffConstrainingRemoteRiver2020, who developed 'geomorphic river types' with distinct prior sets for $A_0_i$ and $n_i$.
 These priors are assigned to a river solely using SWOT data *W* and *S*, therefore meeting our needs for complete global implementability.
@@
 We validate BIKER, assuming no measurement error, on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers using daily simulated hydraulics.
-We also re-validate BIKER under the 'measurement-error' scenario using the `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` rivers with [@frassonExploringFactorsControlling2021]'s SWOT error model to corrupt the hydraulics to mimic realistic SWOT measurements (widths, heights, and slopes).
+We also re-validate BIKER under the 'measurement-error' scenario using the `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` rivers with @frassonExploringFactorsControlling2021's SWOT error model to corrupt the hydraulics to mimic realistic SWOT measurements (widths, heights, and slopes).
 These data were outlined in section 3.1.
@@
 Remember that SWOT cannot observe below the water surface and therefore cannot measure $\bar{U}$ or *H* (hence the need for equation 9), and that all SWOT observations contain errors in both space and time (hence equation 10).
-We acknowledge that there is error in equation 7 as shown in Figure 2, but this error can be explicitly parameterized in our Bayesian system.
-This is elaborated on in section 4.2.
+We acknowledge that there is error in equation 7 as shown in Figure 2, but this error can be explicitly parameterized in our Bayesian system (this is elaborated on in Text S4).
 Therefore, the BIKER validation presented here is an exercise to see how well the imperfect and partial SWOT observations can infer $k_{600}$ given the hydraulic assumptions in equation 9 and uncertainty in the data itself.
+Note also that we have already empirically validated equation 7 in Figure 2.
 
-Therefore, we validate BIKER as a timeseries of $k_{600}$ for each river using the BIKER posterior means.
+We validate BIKER as a timeseries of $k_{600}$ for each river using the BIKER posterior means.
 Our error metrics consider the timeseries nature of the problem and are formally defined in Table S2.
@@
 - Root mean square error normalized by the observed mean (NRMSE) to assess bias
-- mean absolute residual error (MAE) to assess bias
+- Normalized mean absolute residual error (NMAE) to assess bias
 - Kling-Gupta Efficiency (KGE).
@@
 #### 3.2.3 Carbon emissions validation
-**THis is colin's edits. Might need some reorg**
 It is one thing to accurately model the temporal dynamics of $k_{600}$ as above, but researchers are often most interested in the actual carbon emitted from river to atmosphere.
@@
 This effectively reduces equation 2 to equation 11.
-It is worth stressing that these literature upscaling workflows rely on in situ streamflow records and/or high-quality streamflow routing models (unlike BIKER, which only requires a river be SWOT-observable).
+It is worth stressing that these literature upscaling workflows rely on in situ streamflow records and/or high-quality streamflow routing models.
 
@@
 
-In that context, we sought to compare BIKER against established methods from the literature for upscaling carbon emissions from river networks (in this case, our `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])`  rivers) So, we pair the 26 biweekly $CO_2$ and water temperature samples from @beaulieuControlsGasTransfer2012 (section 3.1, figure S2) with every 14th set of daily SWOT observations (as the $CO_2$ data is bi-weekly).
-We then calculate $FCO_2$ using equation 1, an atmospheric $CO_{2_{air}}$ of 400 uatm, and a temperature dependent *Sc* was estimated following Raymond et al.
+Conversely, BIKER represents a new way of approaching equation 11 compared to existing literature models: BIKER has no reliance on a streaflow model nor HG scaling relationships and only requires that a river is SWOT-observable.
+We are therefore interested in how the final carbon emissions that result from BIKER compare against literature methods that use equation 11.
+We have the data to test four different models for fluxes: 'BIKER', 'Raymond 2013', 'Raymond 2012', and 'Brinkerhoff 2019'.
+These latter three approaches all use the same philosophy for $k_{600}$: making hydraulic and geomorphic assumptions to associate $k_{600}$ with observed hydraulics before using the $CO_2$ data as a realistic timeseries to yield fluxes per equation 11.
+In all three approaches, these observed hydraulics are streamflow, while BIKER uses only SWOT observations.
+Therefore, the advantage of BIKER is in its ease of application, as SWOT will observe all global rivers wider than 50m while streamflow observations are extremely geographically limited.
+But, BIKER is only attractive if it can produce fluxes with similar errors to published methods.
+Text S5 and Table S3 fully describe these three literature models.
+
+To benchmark BIKER against these literature methods, we pair the 26 biweekly $CO_2$ and water temperature samples from @beaulieuControlsGasTransfer2012 (section 3.1, figure S2) with every 14th set of daily SWOT observations (as the $CO_2$ data is bi-weekly).
+We then calculate $FCO_2$ using equation 1, an atmospheric $CO_{2_{air}}$ of 400 uatm, and a *Sc* estimated following Raymond et al.
 (2012).
+The *k* in equation 1 is obatained using BIKER or the three literature models (Table S3 and Text S5).
 Finally, we estimate a yearly total carbon emission rate (via $CO_2$ evasion) by applying each river’s mean $FCO_2$ over the river’s surface area and summing all rates across rivers.
-
-BIKER represents a new way of approaching the problem against existing GHG scaling literature.
-We are therefore interested in how the final fluxes that result from BIKER (Eq 11) compare to literature methods to gauge BIKER’s utility: BIKER is novel, but is it useful?
-We have the data to test four different models for fluxes: 'BIKER', 'Raymond 2013', 'Raymond 2012', and 'Brinkerhoff 2019'.
-These latter three approaches all use the same philosophy for $k_{600}$: making hydraulic and geomorphic assumptions to associate $k_{600}$ with observed hydraulics before using the $CO_2$ data as a realistic concentration timeseries to yield fluxes per equation 11.
-In all three approaches, these observed hydraulics are streamflow, while BIKER uses only SWOT observations.
-Therefore, the advantages of BIKER in ease of application are clear as SWOT will observe all global rivers wider than 50m while streamflow observations are extremely geographically limited, but BIKER is only attractive if it can produce fluxes with similar errors to published methods.
-Text S4 and Table S3 fully describe these three literature models.
 
@@
 Notably, there is sometimes positive bias in the estimates (e.g.
-both 'middle KGE' rivers).
+both 'Middle Skill' rivers).
 Some rivers yield the correct temporal dynamics, but the magnitude of these values is stretched relative to the observed (e.g.
@@
 The green ribbon indicates the 95% CIs for the BIKER posteriors.
-Rivers are randomly sampled from the three tertiles of KGE scores.
+Rivers are sampled from the three tertiles of KGE scores.
 (a-b) are poorest performing rivers, (b-c) are in the middle, and (e-f) are the best performing rivers.
@@
 This indicates very strong inference of each river's temporal $k_{600}$ dynamics given that absolutely no in situ information is being used to predict $k_{600}$.
-NRMSE has a median score of `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$NRMSE, na.rm=T),2)`, highlighting that many rivers which have significant positive biases (Figure 4 also confirms this visually).
-Median MAE, however, is only `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$MAE, na.rm=T),2)` m/day.
+NRMSE has a median score of `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$NRMSE, na.rm=T),2)`, highlighting many rivers which have notable positive biases (Figure 4 also confirms this visually).
+Median NMAE, however, is only `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$NMAE, na.rm=T),2)` m/day.
 Taken in aggregate, Figures 4-5 indicate that BIKER is quite good at capturing temporal dynamics in $k_600$, however there is often positive bias in its estimates.
@@
 Each subpanel is labelled by its performance metric (defined in Table S2).
-Dashed lines denote median scores.
-Note the log10 scale for MAE.](cache/validation/fig5.jpg)
+Dashed lines denote median scores.](cache/validation/fig5.jpg)
 
@@
 Note also that axes are flipped in order to visualize all 'better performances' in the upper-right-corner of each sub-plot.
-BIKER performs nearly identically, regardless of implicit measurement errors in the inputs to BIKER: points are scattered along the 1:1 line.
+BIKER performs nearly identically, regardless of implicit measurement errors in the inputs to the algorithm, i.e.
+points are scattered along the 1:1 line.
 Overall, these results strongly suggest that BIKER will be robust to the measurement errors that will be implicit in SWOT's observations of river width and slope.
@@
 
-Finally, we sought to further explore both BIKER's ability to infer temporal trends in $k_{600}$, as well as the influence of prior error on $k_{600}$.
+Finally, we sought to explore the overall influence of prior error/bias on $k_{600}$.
+We plot eCDFs of prior and posterior NMAE (see Table S2 for metric definition) across all `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers in Figure 7a.
+Generally, errors/biases associated with the prior $k_{600}$ are propagated through the posterior in an approximately 1:1 manner (which is expected), except for a subset of rivers in which posterior error actually increases relative to the prior (approximately percentiles 0.6-0.9).
+We explore why this happens below.
 
-Figure 7a plots a histogram of the slopes of linear regressions between BIKER-predicted $k_{600}$ and observed $k_{600}$.
-A slope of 1 indicates that BIKER correctly infers the daily fluctuations in gas exchange (even if there is bias implicit in these estimates), while a slope well above or below 1 indicates incorrect inference of the temporal dynamics of $k_{600}$.
-Figure 6a shows that the majority of rivers' slopes approximate 1 (median slope: `r round(median(results_dynamics$Slope),2)`), indicating strong inference of temporal gas exchange trends.
+![Figure 7: Empirical cummulative density functions for prior and posterior $k_{600}$ NMAE.
+See Table S2 for metric definition.
+a) all rivers.
+b) Only rivers with a coefficient of varaiton (CV) of observed $k_{600}$ > 0.10.
+c) Only rivers with CV > 0.20.
+d) Only rivers with CV > 0.30.
+Overall we see that rivers with greater temporal variation in $k_{600}$ behave better for BIKER, i.e.
+Bayesian inference reduces bias from prior to posterior.](cache/validation/fig7.jpg)
 
-![Figure 7: a) Histogram of the slopes of linear regressions between BIKER estimates and observed $k_{600}$.
-Slopes equaling 1 indicate correct inference of the temporal trends in daily $k_{600}$ (regardless of bias in the estimates).
-b) Empirical cummulative density distributions of prior and posterior mean absolute error (MAE).](cache/validation/fig7.jpg)
-
-Figure 7b comes prior error for $k_{600}$ against posterior error in $k_{600}$.
-We used the mean absolute error (Table S2), which is a metric that captures the average residual error in predictions, to assess this and plotted the results as eCDFS across all `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers in the 'no-error' scenario.
-In Figure 7b, we see that BIKER reduces MAE, when the prior estimate was already quite good (approximately less than 3 m/day), while BIKER increases MAE if the prior estimate was already quite poor (again, approximately > 3 m/day).
-This indicates that our prior estimate of $k_{600}$ is too certain, i.e.
-BIKER is convinveed that the prior estimate is reasonable and, when combined with the hydraulics inference also in the model, yields a net worse performance.
-TO perturb this, we reran our BIKER validation using a less certain $k_{600}$ prior.
-SHOULD THIS BE SI OR JUST BE THE ACTUAL STUDY
+Recall that BIKER relies on a timeseries of SWOT data, and that these timeseries may not be representative of the full spectrum of $k_{600}$ values that are actually experienced in the river, therefore potentially biasing both the prior estimation methods (Text S3) and the actual Bayesian inference.
+Put another way, we suggest that a SWOT timeseries that does not sufficiently capture a river's temporal dynamics will introduce additional error to the inference results.
+To test this, we subset our validation dataset by progressively higher coefficients of variation (*CV*) (> 10%, >20%, and >30%).
+These filtered datasets are plotted in Figures 7b-7d (for > 10%, >20%, and >30% *CV*, respectively), where our hypothesis is confirmed.
+Posterior bias systematically drops as the temporal variability of the SWOT data increases, with BIKER posterior error the smallest, and much less than prior error, in Figure 7d.
+This result is elaborated on in section 4.2.
 
@@
 This means that BIKER's annual carbon emission estimate can only improve from what is presented here.
-We also stress again that the raw emissions rates here are largely meaningless as they are calculated from an identical $CO_{2_water}$ timeseries applied to every river.
+We also stress again that the raw emissions rates here are largely meaningless as they are calculated from an identical $CO_{2_{water}}$ timeseries applied to every river.
 We are principally interested in relative differences between techniques employed.
@@
 Surprisingly, BIKER outperforms the gauge-based approaches, nearly correctly inferring the annual carbon emissions rate (`r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_BIKER',]$yearlyCFux, 2)` Tg-C/yr for BIKER versus `r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_obs',]$yearlyCFux, 2)` Tg-C/yr observed).
-The three HG models overestimate this emissions rate (`r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_raymond2013',]$yearlyCFux, 2)`, `r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_raymond2012',]$yearlyCFux, 2)`, and `r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_brinkerhoff2019',]$yearlyCFux, 2)`) for 'Raymond 2013', 'Raymond 2012', and 'Brinkerhoff 2019' respectively.
-Both 'Brinkerhoff 2019' and 'Raymond 2012' fall within the BIKER CIs and are reasonably close to the observed value, while 'Raymond 2013' overestimates the emission rate.
+The three HG models overestimate this emissions rate: `r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_raymond2013',]$yearlyCFux, 2)`, `r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_raymond2012',]$yearlyCFux, 2)`, and `r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_brinkerhoff2019',]$yearlyCFux, 2)` for 'Raymond 2013', 'Raymond 2012', and 'Brinkerhoff 2019' respectively.
+'Raymond 2012' falls within the BIKER CIs and is reasonably close to the observed value, while 'Raymond 2013' overestimates the emission rate.
+'Brinkerhoff 2019''s lower CI is comparable to BIKER's higher CI.
 BIKER's relatively stronger performance than the in situ HG models is elaborated on in section 4.3.
@@
 While much work has focused on the small-stream side of the stream-to-river continuum, comparatively less work has been done in large systems.
-Here, we focus on the larger, 'smooth-channel' end of the continuum, using a model for gas exchange that scales $k_{600}$ by both $\epsilon_D$ and a Shear Reynold's number.
+Here, we focus on the larger, 'smooth-channel' end of the continuum, using a model for gas exchange that scales $k_{600}$ by both $\epsilon_D$ and a shear Reynold's number.
 This model is empirically validated in Figure 3.
-Specifically, Figure S1 confirms that scaling $k_{600}$ with a Shear Reynold's adaption of the small-eddy model (equation 7) reduces positive bias in the model's predictions of the smallest $k_{600}$ values [where the relative decrease in turbulence reaching the surface is greater than the small-eddy model alone suggests, per equation 7's theoretical basis- @moogAirWaterGasTransfer1999].
-Scaling $k_{600}$ via a shear Reynold's number is commonly done to parameterize breaking-wave gas exchange models in the open ocean [@brumerWaveRelatedReynoldsNumber2017; @zhaoDependenceWhitecapCoverage2001; @zhaoEffectWindWaves2003], though this specific to high wind speeds in open ocean.
+Specifically, Figure S1 confirms that scaling $k_{600}$ with a shear Reynold's adaption of the small-eddy model (equation 7) reduces positive bias in the model's predictions of the smallest $k_{600}$ values [where the relative decrease in turbulence reaching the surface is greater than the small-eddy model alone suggests, per equation 7's theoretical basis- @moogAirWaterGasTransfer1999].
+Scaling $k_{600}$ via a shear Reynold's number is often done to parameterize breaking-wave gas exchange models in the open ocean [@brumerWaveRelatedReynoldsNumber2017; @zhaoDependenceWhitecapCoverage2001; @zhaoEffectWindWaves2003], though this is specific to high wind speeds in open ocean.
 To our knowledge, @moogAirWaterGasTransfer1999's specific setup, which imposes a 'differentially-experienced' surface turbulence theory on the small-eddy model, has never been empirically validated in rivers.
@@
 That said, BIKER's flexible implementation is a good start towards eventually coupling hydraulics-driven gas exchange with wind-driven gas exchange.
-Such a model would likely take a form similar to the conceptual models proposed by @wangPhysicallyBasedScaling2021, chuWindStreamFlow2003, and/or @plateReaerationOpenChannel1984.
+Such a model would likely take a form similar to the conceptual models proposed by @wangPhysicallyBasedScaling2021, @chuWindStreamFlow2003, and/or @plateReaerationOpenChannel1984.
 All three generally propose calculating both a wind-driven *k* and a hydraulics-driven *k* and then weighting each term via parameters.
@@
 
+Furthermore, Figure 7 implies that a sufficient variability in SWOT observations is necessary to strongly infer a $k_{600}$ timeseries.
+'Hydraulic visibility', i.e.
+the ability of a remote sensor to identify a hydrological response in the river [@garamboisHydraulicVisibilityUsing2017a] is applicable here.
+If we apply hydraulic visibility to a sensor's ability to identify temporal variations in $k_{600}$, Figure 7 suggests that a 'minimally sufficient' hydraulic variability in SWOT measurements (in Figure 7b, suggested to be > 10% *CV*) is needed to uniformally improve upon the prior.
+This will be important once SWOT launches and BIKER is implemented at the global-scale.
+
 ### 4.3 Coupling BIKER with upscaling workflows
@@
 
-First, figure 8 directly implies that BIKER will be useful when coupled with large-scale $CO_{2_water}$ models, provided these models are accurate.
+First, figure 8 directly implies that BIKER will be useful when coupled with large-scale $CO_{2_{water}}$ models, provided these models are accurate.
 The models would give time and spcae varying gas exchange.
@@
 
-Finally, Figure 8 also uniquely allows us to directly compare the influence of geomorphic assumptions on total carbon emission rates from river networks, as all other calculations and parameters were held constant across our four tested models (Text S4).
+Finally, Figure 8 also uniquely allows us to directly compare the influence of geomorphic assumptions on total carbon emission rates from river networks, as all other calculations and parameters were held constant across our four tested models (Text S5).
 Therefore, figure 8 highlights a potentially large source of uncertainty in current river $CO_2$ upscaling estimates: the geomorphic models employed to scale river channel hydraulics with streamflow.
-In this case, the only difference between the three literature models and the observed estimate in Figure 8 is the specific HG model employed to predict river depth and velocity (Text S4, Table S3.
-**put some best case/worse case syle writing from the results here**.
+In this case, the only difference between the three literature models and the observed estimate in Figure 8 is the specific HG model employed to predict river depth and velocity (Text S5, Table S3), and yet the eventual carbon emissions estimates (Figure 8) are quite different.
+Further, recall that the BIKER results in Figure 8 reflect a worst-case scenario (relatively uniformed priors), while the three in-situ methods represent  best case scenarios (perfect streamflow records).
+We suggest future work should perform a formal sensitivity analysis for these HG parameters.
 
 ## 5 Conclusions
-The $CO_2$ flux from the global river network is a major natural component of the global carbon cycle, on par with total forest uptake of $CO_2$ [@panLargePersistentCarbon2011].
+The $CO_2$ flux from the global river network is a major natural component of the global carbon cycle, on par with total forest carbon uptake [@panLargePersistentCarbon2011].
 Despite much interest and progress, there is still large uncertainty in the temporal dynamics of gas exchange and thus carbon emissions to the atmosphere.
-Much of this uncertainty stems from uncertainty in gas exchange velocity $k_{600}$ at the global scale.
+Much of this uncertainty stems from uncertainty in $k_{600}$ at the global scale.
 In this paper, we propose that the soon-to-launch SWOT satellite will provide enough hydraulic measurements to analyse the temporal dynamics of $k_{600}$, and therefore allow for a global-scale analysis of spatiotemporal trends in large-river $k_{600}$ once SWOT launches.
 In that context, we developed 1) a hydraulic model for $k_{600}$ that is nearly entirely SWOT observable and explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2,2)*100`% of variation in $k_{600}$, and 2) the BIKER algorithm to infer $k_{600}$ using no on-the-ground information.
-Validating on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers, we show show strong recovery of rivers' temporal $k_{600}$ dynamics and a hypothetical total annual carbon emission rate across all `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers.
+Validating on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers, we show strong recovery of rivers' temporal $k_{600}$ dynamics and a hypothetical total annual carbon emission rate across all `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers.
 These results suggest BIKER can be used to infer global-scale, near daily estimates of fluvial gas exchange velocity once SWOT launches in 2022.
@@
 Gleason was supported by NASA SWOT Science Team grant #yyyyyy.
-The version of BIKER used in this study will be available upon acception.
+The version of BIKER used in this study will be available upon acceptance.
 BIKER remains in active development and is available at https://github.com/craigbrinkerhoff/BIKER.

### 2022-03-09 manuscript_v10.Rmd
manuscript v10

--- 2022-03-09 docs/manuscript_v9.Rmd
@@
       institute: Yale
-  - Merritt H Harlan:
+  - Merritt E Harlan:
       institute: UMass
@@
   word_document:
-      pandoc_args:
-        - '--lua-filter=scholarly-metadata.lua'
-        - '--lua-filter=author-info-blocks.lua'
+    #  pandoc_args:
+    #    - '--lua-filter=scholarly-metadata.lua'
+    #    - '--lua-filter=author-info-blocks.lua'
       reference_docx: word-styles-reference.docx
@@
 ## Key Points
+
 -   The BIKER (Bayesian Inference of the $k_{600}$ Exchange Rate) algorithm predicts gas exchange velocity and $CO_2$ fluxes solely from simulated SWOT data without calibration
-- BIKER is robust to measurement errors implicit in SWOT river measurements
+-   BIKER is robust to SWOT measurement errors
 -   BIKER will allow for novel study of gas exchange spatiotemporal dynamics after SWOT's launch
@@
 ## Keywords
+
 gas exchange, fluvial geomorphology, remote sensing, open-channel flow, SWOT, biogeochemistry
@@
 ## Abstract
-Extensive research over the past two decades has shown that the global river network emits significant amounts of greenhouse gas via gas exchange.
+
+Extensive research over the past two decades has shown that the global river network emits significant amounts of greenhouse gas.
 Despite much progress, there is still large uncertainty in the temporal dynamics of gas exchange and thus carbon emissions to the atmosphere.
@@
 Here, we present work inferring $k_{600}$ from synthetic future SWOT observations without in situ calibration.
-We exploit the unique geomorphology of SWOT-observable rivers to develop a physical model of gas exchange that is remotely sensible and explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2,2)*100`% of variation in $k_{600}$ We then couple this model with established techniques to develop BIKER, the 'Bayesian Inference of the $k_{600}$ Exchange Rate' algorithm.
+We exploit the unique geomorphology of SWOT-observable rivers to develop a physical model of gas exchange that is remotely sensible and explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2,2)*100`% of variation in $k_{600}$ We then couple this model with established inversion techniques to develop BIKER, the 'Bayesian Inference of the $k_{600}$ Exchange Rate' algorithm.
 We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers and 166 measurements of river gas exchange, yielding an algorithm that predicts the $k_{600}$ timeseries solely from SWOT observations with a by-river median Kling-Gupta Efficiency of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$KGE), 2)`.
@@
 Finally, BIKER is robust to measurement errors implicit in the SWOT data.
-Upon SWOT's launch, we suggest that BIKER will be useful in mapping the global-scale spatiotemporal dynamics of fluvial gas exchange in large rivers.
+Upon SWOT’s launch, we suggest that BIKER will be useful in mapping global-scale fluvial gas exchange and improving CO2 emissions estimates when coupled with river CO2 models.
 
 ## 1 Introduction
+
 Natural systems play a critical role in the budgeting and accounting of the global carbon cycle under climate change.
 Following @colePlumbingGlobalCarbon2007, the global river network is recognized to emit substantial amounts of carbon to the atmosphere via evasion (gas exchange driven by a diffusion gradient and near-surface turbulence), in addition to their long understood role in transporting carbon to the oceans via downstream advection.
-Current estimates of total carbon dioxide evasion ($FCO_2$) to the atmosphere from the global river network vary from 650-2000 Tg C/yr [*Liu accepted*; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013], with 167 Tg-C/yr coming from mountain streams alone [@horgbyUnexpectedLargeEvasion2019].
-Despite rivers' incredibly small percentage of the global land surface [0.67%- *Liu accepted*], this C flux is on par with the total oceanic $CO_2$ uptake rate [@gruberOceanicSinkAnthropogenic2019; @horgbyUnexpectedLargeEvasion2019] and the global forest carbon uptake rate [@panLargePersistentCarbon2011].
+Current estimates of total carbon dioxide evasion ($FCO_2$) to the atmosphere from the global river network vary from 650-2000 Tg C/yr [@liuImportanceHydrologyRouting2022; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013], with 167 Tg-C/yr coming from mountain streams alone [@horgbyUnexpectedLargeEvasion2019].
+Despite rivers' incredibly small percentage of the global land surface [0.67%- @liuImportanceHydrologyRouting2022], this C flux is on par with the total oceanic $CO_2$ uptake rate [@gruberOceanicSinkAnthropogenic2019; @horgbyUnexpectedLargeEvasion2019] and the global forest carbon uptake rate [@panLargePersistentCarbon2011].
 
@@
 There is a robust existing literature exploring spatiotemporal patterns in $\delta CO_2_{water}$ [e.g.
-@ahoDistinctConcentrationdischargeDynamics2021; @ahoDifferentialResponseGreenhouse2019; @crawfordCO2TimeSeries2017; @peterScalesDriversTemporal2014; @ranDynamicsRiverineCO22017; @rocher-rosLandscapeProcessDomains2019; @raymondAtmosphericCO2Evasion2000; @liuHydrologicControlsPCO22018a].
-This work has lead to recent river-reach explicit modeling of $\delta CO_2$ using global hydrography datasets at up to monthly temporal resolutions [@horgbyUnexpectedLargeEvasion2019; @brinkerhoffLakeMorphometryRiver2021; @saccardiImprovingPredictionsStream2021; *Liu accepetd*], but an equivalently sophisticated representation of $k_{CO_2}$ is still lacking.
+@ahoDistinctConcentrationdischargeDynamics2021; @ahoDifferentialResponseGreenhouse2019; @crawfordCO2TimeSeries2017; @peterScalesDriversTemporal2014; @ranDynamicsRiverineCO22017; @rocher-rosLandscapeProcessDomains2019; @raymondAtmosphericCO2Evasion2000; @liuHydrologicControlsPCO22018].
+This work has lead to recent river-reach explicit modeling of $\delta CO_2$ using global hydrography datasets at up to monthly temporal resolutions [@horgbyUnexpectedLargeEvasion2019; @brinkerhoffLakeMorphometryRiver2021; @saccardiImprovingPredictionsStream2021; @liuImportanceHydrologyRouting2022], but an equivalently sophisticated representation of $k_{CO_2}$ is still lacking.
 
@@
 These models usually predict $k_{600}$, or *k* normalized by a Schmidt number (*Sc*) of 600.
-This is to remove the effect of water temperature and gas type from predictive models, as warmer waters and lower *Sc* numbers both increase gas exchange rates [@hallGasExchangeStreams2020].
+This is to remove the effect of water temperature and gas type from predictive models, as warmer waters and lower *Sc* numbers each increase gas exchange rates [@hallGasExchangeStreams2020].
 Specifically, $k_{600}$ reflects the $k_{CO_{2}}$ at 20 degrees Celsius.
-Through this normalization, these models focus solely on geormorphic explanantions for variation in *k* [@hallGasExchangeStreams2020].
+Through this normalization, these models focus solely on physical explanations for variation in *k* [@hallGasExchangeStreams2020].
 
-Applying these *k* models across watersheds, regions, or continents is called ‘upscaling.’ This upscaling allows for estimating the impossible-to-measure *k* term in Equation 1 for any arbitrary number of rivers, but also changes the base parameters that ultimately control the final estimate of $FCO_2$.
+Applying these *k* models across watersheds, regions, or continents is called 'upscaling.' This upscaling allows for estimating the impossible-to-measure *k* term in Equation 1 for any arbitrary number of rivers, but also changes the base parameters that ultimately control the final estimate of $FCO_2$.
 That is, by making *k* a function of hydraulics, $FCO_2_{upscaled}$ is now a direct function of river hydraulics.
@@
 @raymondGlobalCarbonDioxide2013, @lauerwaldSpatialPatternsCO22015, and @horgbyUnexpectedLargeEvasion2019 all relied on $k_{CO_2}$ values indirectly estimated using mean annual streamflow models and hydraulic scaling equations to predict the hydraulic terms to in turn predict $k_{CO_2}$, while @borgesGloballySignificantGreenhousegas2015 used a combination of the above method and a constant $k_{CO_2}$ in space and time to upscale over Africa.
+More recently, @liuImportanceHydrologyRouting2022 performed a first upscaling assessment of monthly temporal dynamics in global river $FCO_2$, though they relied on monthly modeled streamflow and used the same model for $k_{600}$ as previous studies [@raymondGlobalCarbonDioxide2013] to achieve this.
 In all of these foundational studies, the temporal dynamics of $k_{CO_2}$ (and thus dynamics in $FCO_2$) were ignored because of hydraulic data limitations.
 It has been shown at the field-scale that temporal dynamics of gas exchange can vary widely from site to site [@wallinSpatiotemporalVariabilityGas2011], but it has remained impractical to obtain temporally explicit $k_{CO_2}$ at continental-to-global scales.
-More recently, *Liu etal accepted* performed a first assessment of monthly temporal dynamics in the global river $CO_2$ flux, though they relied on modeled streamflow and used the same model for $k_{600}$ as previous studies [@raymondGlobalCarbonDioxide2013] to achieve this.
 
-@wangPhysicallyBasedScaling2021 recently attempted to address this global *k* problem by simulating $k_{O_2}$ in 35 rivers of many sizes (widths ranging from 0.23–349m) using a stream metabolism model [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets to infer what $k_{O_2}$ must have been to produce their 'observations'.
-They then compared this simulated dataset against direct measurements of *k*, finding similar performance and parameter values against process-based models of gas evasion.
+@wangPhysicallyBasedScaling2021 recently attempted to address this global *k* problem by simulating $k_{O_2}$ in 35 rivers of many sizes (widths ranging from 0.23--349m) using a stream metabolism model [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets to infer what $k_{O_2}$ must have been to produce their 'observations'.
+They then compared this simulated dataset against direct measurements of *k*, finding similar performance and parameter values for process-based models of gas evasion.
 However, they were still limited by a lack of direct hydraulic measurements and had to rely on hydraulic scaling equations to estimate river depth and velocity.
@@
 SWOT is a wide swath radar interferometer and will sample rivers every 1 to 7 days per 21 day repeat cycle, measuring rivers wider than 100m with a goal of expanding this to rivers at least 50m wide [@biancamariaSWOTMissionIts2016].
-Via its direct hydraulic measurements, SWOT is expected to usher in a sea change in global-scale hydrology, and could similarly transform fluvial biogeochemsitry if techniques are developed to ingest SWOT data and infer *k*.
+Via its direct hydraulic measurements, SWOT is expected to usher in a sea change in global-scale hydrology, and could similarly influence fluvial biogeochemsitry if techniques are developed to ingest SWOT data and infer *k*.
 In this context, we borrow tools from fluvial geomorphology and existing SWOT algorithms to answer the following two questions:
@@
 
-To answer this first question, we use hydraulic geometry- the fundamental geomorphic relationships between streamflow and channel shape [@leopoldHydraulicGeometryStream1953; @gleasonHydraulicGeometryNatural2015] to develop a process-based model for large-river (here defined as wider than 50m to align with SWOT) $k_{600}$.
+To answer this first question, we use hydraulic geometry- the fundamental geomorphic relationships between streamflow and channel shape [@leopoldHydraulicGeometryStream1953; @gleasonHydraulicGeometryNatural2015] to develop a process-based model for large-river $k_{600}$ (here defined as wider than 50m to align with SWOT).
 We then take these findings and explore the second question by implementing this hydraulic model, which defines $k_{600}$, within an algorithm named BIKER ('Bayesian Inference of the $k_{600}$ Evasion Rate') to infer $k_{600}$ solely from SWOT measurements.
@@
 ![Figure 1: Conceptual overview of the BIKER algorithm.
-SWOT will directly measure water surface width and elevation (and thus slope), and the BIKER algorithm uses those observations to infer gas exchange velocity $k_{600}$ via a process-based model (section 2).](cache/flowchart3.jpg)
+1: SWOT will directly measure water surface width and elevation (and thus slope) at a series of cross-sections within mass-conserved river reaches.
+2: These hydraulics measurements are used to calculate turbulent dissipation in the river channel (section 3).
+3: Turbulent dissipation is used to infer $k_{600}$ via a process-based model (section 2).](cache/maps/diagram.tif)
 
-## 2 Prediciting $k_{600}$ from large-river hydraulic geometry
+## 2 Predicting $k_{600}$ from large-river hydraulic geometry
+
 To predict $k_{600}$ in the large rivers that SWOT will observe, we start from an established process-based model for $k_{600}$, impose hydraulic assumptions that are valid in SWOT-observable rivers, and obtain a model we empirically test.
@@
 ### 2.1 Data
+
 We develop our hydraulic model for $k_{600}$ using multiple datasets of measured $k_{600}$ and stream hydraulics collected from the literature.
@@
 ### 2.2 Process-based hydraulic modeling of river *k*
-*k* scales with near-surface turbulence in turbulent streamflows [@hallGasExchangeStreams2020], and extensive field and laboratory experiments have converged on the 'small-eddy model' as derived by @lamontEddyCellModel1970 and empirically anticipated by @calderbankContinuousPhaseHeat1961.
+
+*k* scales with near-surface turbulence in turbulent streamflows [@hallGasExchangeStreams2020], and extensive field and laboratory experiments have converged on the 'small-eddy model' for *k* as derived by @lamontEddyCellModel1970 and empirically anticipated by @calderbankContinuousPhaseHeat1961.
 This model scales *k* via the smallest-scale turbulent eddies and has been repeatedly empirically validated in freshwater systems [e.g.
-@katulStructureFunctionModel2018; @lorkeUnifiedScalingRelation2006; @zappaEnvironmentalTurbulentMixing2007; @zappaVariationSurfaceTurbulence2003; @moogStreamReaerationNonuniform1999; @vachonRelationshipNearsurfaceTurbulence2010; @tokoroHighGastransferVelocity2008; @wangPhysicallyBasedScaling2021].
+@katulStructureFunctionModel2018; @lorkeUnifiedScalingRelation2006a; @zappaEnvironmentalTurbulentMixing2007; @zappaVariationSurfaceTurbulence2003; @moogStreamReaerationNonuniform1999; @vachonRelationshipNearsurfaceTurbulence2010; @tokoroHighGastransferVelocity2008; @wangPhysicallyBasedScaling2021].
 The small-eddy model is provided as equation 3, where $\epsilon$ is the dissipation rate of near-surface turbulence, $v$ is the kinematic viscosity, and *Sc* is the Schmidt number.
 
-$$k\propto Sc^{-1/2}(v\epsilon)^{1/4} \mathbf{(3)}$$
-Some laboratory and field observations additionally suggest that open channel flows with small bed roughness do not exhibit homogeneous surface dissipation at their air-water interface [@moogAirWaterGasTransfer1999; @talkeTurbulentKineticEnergy2013].
+$$k\propto Sc^{-1/2}(v\epsilon)^{1/4} \mathbf{(3)}$$ Some laboratory and field observations additionally suggest that open channel flows with small bed roughness do not exhibit homogeneous surface dissipation at their air-water interface [@moogAirWaterGasTransfer1999; @talkeTurbulentKineticEnergy2013].
 Given this observation, @moogAirWaterGasTransfer1999 proposed an extension to the small-eddy model, additionally scaling $k_{600}$ using a shear Reynold's number formulation.
@@
 The Reynolds model is hypothetically useful in low-turbulence scenarios where a relative lack of large-scale eddies effectively 'filter out' the number of small-eddies that actually reach the interface and initiate gas exchange [@talkeTurbulentKineticEnergy2013].
-While scaling *k* via a shear Reynold's formulation is sometimes done to parameterize wave-breaking gas exchange models in the open ocean  [@brumerWaveRelatedReynoldsNumber2017; @zhaoDependenceWhitecapCoverage2001; @zhaoEffectWindWaves2003], it is infrequently done in rivers.
+While scaling *k* via a shear Reynold's formulation is sometimes done to parameterize wave-breaking gas exchange models in the open ocean [e.g.
+@brumerWaveRelatedReynoldsNumber2017; @zhaoDependenceWhitecapCoverage2001; @zhaoEffectWindWaves2003], it is infrequently done in rivers.
 In the context of BIKER, we chose to test this model because large, SWOT-observable rivers are generally the smoothest, least-turbulent flows along the stream-to-ocean continuum where small eddies might not reach the surface.
@@
 
-$$k \propto Sc^{-1/2}(v\epsilon)^{1/4}(\frac{U_*H}{v})^{3/8} \mathbf{(4)}$$
-Equations 3 and 4 both rely on $\epsilon$, which is non-trivial to measure.
-When working at large scales, a commonly used $\epsilon$ model assumes that all turbulence is generated at the bed and transported to the air-water interface via the log-law-of-the-wall [equation 5- @lorkeUnifiedScalingRelation2006; @nezuTurbulenceOpenChannel1993].
+$$k \propto Sc^{-1/2}(v\epsilon)^{1/4}(\frac{U_*H}{v})^{3/8} \mathbf{(4)}$$ Equations 3 and 4 both rely on $\epsilon$, which is non-trivial to measure.
+When working at large scales, a commonly used $\epsilon$ model assumes that all turbulence is generated at the bed and transported to the air-water interface via the log-law-of-the-wall (equation 5).
 Another approach specific to fluvial settings models *k* via 'form-drag dissipation' (equation 6) which is equivalently the total stream power per unit mass water.
@@
 
-$$\epsilon = \epsilon_S = \frac{U_*^3}{H} \mathbf{(5)}$$
-$$\epsilon = \epsilon_D = gS\bar{U} \mathbf{(6)}$$
+$$\epsilon = \epsilon_S = \frac{U_*^3}{H} \mathbf{(5)}$$ $$\epsilon = \epsilon_D = gS\bar{U} \mathbf{(6)}$$
 
 ### 2.3 Deriving a large-river $k_{600}$ model
+
 Given the theoretical context provided in Section 2.2, we now turn to SWOT-observable systems specifically.
@@
 This is a common assumption in hydraulic and geomorphic modeling of large rivers.
-For example, SWOT-observable river flows have an average $\frac{R_h}{H}$ ratio of `r HG_swot$mean` and standard deviation of `r HG_swot$sd` (n = `r HG_swot$n`; see Text S1 for how we built this dataset).
+For example, SWOT-observable river flows have an average $\frac{R_h}{H}$ ratio of `r HG_swot$mean` and standard deviation of only `r HG_swot$sd` (n = `r HG_swot$n`; see Text S1 for how we built this dataset).
 We refer to these rivers as 'hydraulically-wide'.
@@
 ### 2.4 Model validation
+
 With equation 7 derived, we now empirically test its strength of fit in hydraulically-wide river flows.
@@
 ## 3 Exploiting equation 7 to remotely sense gas exchange velocity
+
 We have shown that scaling $k_{600}$ via equation 7 is useful in hydraulically-wide rivers.
@@
 Conveniently, techniques to simultaneously infer mean flow depth and velocity from SWOT data (among other parameters) have been established over the last decade to infer streamflow from SWOT's measurements [e.g.
-@gleasonGlobalMappingRiver2014a; @andreadisConstrainingAssimilationSWOT2020; @brinkerhoffConstrainingRemoteRiver2020; @durandEstimatingReachaveragedDischarge2014; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014a; @garamboisVariationalEstimationEffective2020; @brissetAssimilationAltimetricData2018; @hagemannBAMBayesianAMHGManning2017; @larnierRiverDischargeBathymetry2020a; @oubanasDischargeEstimationUngauged2018].
+@andreadisConstrainingAssimilationSWOT2020; @gleasonGlobalMappingRiver2014; @brinkerhoffConstrainingRemoteRiver2020; @durandEstimatingReachaveragedDischarge2014; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014; @garamboisVariationalEstimationEffective2020; @brissetAssimilationAltimetricData2018; @hagemannBAMBayesianAMHGManning2017; @larnierRiverDischargeBathymetry2020; @oubanasDischargeEstimationUngauged2018].
 For BIKER, we follow the work developed by @hagemannBAMBayesianAMHGManning2017, @brinkerhoffConstrainingRemoteRiver2020, and @durandEstimatingReachaveragedDischarge2014 to infer $k_{600}$, channel depth and velocity from SWOT observations using a modified form of Manning's equation.
@@
 ### 3.1 Data
+
 To validate BIKER, we cannot use actual SWOT measurements as SWOT has yet to launch.
@@
 These data were published by @frassonExploringFactorsControlling2021 and @durandIntercomparisonRemoteSensing2016.
-See Figure 3 for a map of these river's locations.
-
-![Figure 3: Map of the 47 hydraulic models and 1 timeseries of $CO_2$ samples used in this study.
-Note that hydraulic model locations are approximate as some of the models are not geo-referenced.](cache/maps/map3.jpg)
+See Figure S2 for a map of these river's locations.
 
@@
 Here, we validate BIKER under a 'no-measurement-error' scenario that reflects an unrealistic measurement as if SWOT has perfect accuracy and precision: we use the hydraulic model output directly as a first test of algorithm validity.
-`r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` of these rivers are then additionally validated under a 'measurement-error' scenario that more closely mimics expected SWOT by adding realistic radar errors and sampling along the satellite’s future ground track.
+`r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` of these rivers are then additionally validated under a 'measurement-error' scenario that more closely mimics expected SWOT by adding realistic radar errors and sampling along the satellite's future ground track.
 SWOT river error modeling was developed by @durandHowWillRadar2020 and then applied to these `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` rivers by @frassonExploringFactorsControlling2021.
@@
 
-For $CO_2$ evasion and carbon emissions calculations, we use 26 bi-weekly dissolved $CO_2$ samples collected by @beaulieuControlsGasTransfer2012 at one location in the Ohio River (Figure 3) for one calender year from 2008-2009 (Figure S2).
+For $CO_2$ evasion and carbon emissions calculations, we use 26 bi-weekly dissolved $CO_2$ samples collected by @beaulieuControlsGasTransfer2012 at one location in the Ohio River (Figure S2) for one calender year from 2008-2009 (Figure S3).
 Note that this $CO_2$ data is for the Ohio River only but was applied to all rivers to provide a physically realistic signal for $CO_2$ fluxes with meaningful seasonality and dynamics.
@@
 ### 3.2 Section 3 methods
+
 #### 3.2.1 BIKER
+
 BIKER, and Bayesian inference in general, starts from Bayes rule (equation 8), where $\Theta$ is some set of non-remotely-sensible parameters we want to solve for (including $k_{600}$), *x* is the observed data, $f(x|\Theta)$ is the sampling model where data are conditional on the parameters, and $p(\Theta)$ is the joint prior distribution of the parameters.
@@
 
-$$p(\Theta|x) = \frac{f(x|\Theta)p(\Theta)}{p(x)} \mathbf{(8)}$$
-The heart of BIKER is its reformulation of the $k_{600}$ model (equation 7) as a Bayesian sampling model that is conditional on the non-remotely-sensible parameters (i.e.
+$$p(\Theta|x) = \frac{f(x|\Theta)p(\Theta)}{p(x)} \mathbf{(8)}$$ The heart of BIKER is its reformulation of the $k_{600}$ model (equation 7) as a Bayesian sampling model that is conditional on the non-remotely-sensible parameters (i.e.
 $f(x|\Theta)$).
-This is similar to the 'McFLI' ('Mass-Conserved Flow Law Inversion') logic used in some SWOT RSQ algorithms [@gleasonTrackingRiverFlows2017].
+This is similar to the 'McFLI' (Mass-Conserved Flow Law Inversion) logic used in some SWOT RSQ algorithms [@gleasonTrackingRiverFlows2017].
 To start, we write $k_{600}$ as a function of SWOT-observables $W$ and $S$.
@@
 
-$$\begin{eqnarray} k_{600_t} = 62.82 g^{7/16}S_{it}^{9/16}\frac{1}{n_i}^{1/4}(\frac{A_0_i+dA_{it}}{W_{it}})^{35/48} \end{eqnarray} \mathbf{(9)}$$
-Next, equation 9 is written as a Bayesian sampling model, in which all of the SWOT observations are sampled from the unknown model parameters ($A_0_i$, $k_{600_t}$, and $n_i$).
+$$\begin{eqnarray} k_{600_t} = 62.82 g^{7/16}S_{it}^{9/16}\frac{1}{n_i}^{1/4}(\frac{A_0_i+dA_{it}}{W_{it}})^{35/48} \end{eqnarray} \mathbf{(9)}$$ Next, equation 9 is written as a Bayesian sampling model, in which all of the SWOT observations are sampled from the unknown model parameters ($A_0_i$, $k_{600_t}$, and $n_i$).
 This is equation 10 after describing everything as normal distributions of the log-transformed terms.
@@
 
-$$\left[log(62.82) + \frac{7}{16}log(g) + \frac{9}{16}log(S_{i,t}) - \frac{35}{48}log(W_{i,t})\right] \sim normal\left(\left[log(k_{600_t})+\frac{1}{4}log(n_i)-\frac{35}{48}log(A_0_i+dA_{i,t})\right], \sigma^2_{k_{600}}\right) \mathbf{(10)}$$
-Equations 8 and 10 also necessitate that we specify prior distributions for the parameters $A_0_i$, $k_{600_t}$, and $n_i$.
+$$\left[log(62.82) + \frac{7}{16}log(g) + \frac{9}{16}log(S_{i,t}) - \frac{35}{48}log(W_{i,t})\right] \sim normal\left(\left[log(k_{600_t})+\frac{1}{4}log(n_i)-\frac{35}{48}log(A_0_i+dA_{i,t})\right], \sigma^2_{k_{600}}\right) \mathbf{(10)}$$ Equations 8 and 10 also necessitate that we specify prior distributions for the parameters $A_0_i$, $k_{600_t}$, and $n_i$.
 Prior distributions formalize the a priori estimates and uncertainties for the non-remotely-sensed terms.
@@
 These priors are assigned to a river solely using SWOT data *W* and *S*, therefore meeting our needs for complete global implementability.
-Prior assignment for $k_{600_t}$ was developed similarly (all prior specifications are elaborated in Text S3).
+Prior assignment for $k_{600_t}$ was developed similarly (all prior specifications are elaborated on in Text S3).
 
@@
 #### 3.2.2 BIKER validation
+
 We validate BIKER, assuming no measurement error, on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers using daily simulated hydraulics.
@@
 There are, to our knowledge, no datasets of $k_{600}$ over time approaching the temporal density of our simulated SWOT rivers.
-We therefore apply equation 7 as validated in Figure 2 to specify $k_{600}$ given the true hydraulics of each case and compare BIKER’s inversion to that value: given observed hydraulics, ‘observed’ $k_{600}$ comes from equation 7.
+We therefore apply equation 7 as validated in Figure 2 to specify $k_{600}$ given the true hydraulics of each case and compare BIKER's inversion to that value: given observed hydraulics, 'observed' $k_{600}$ comes from equation 7.
 Remember that SWOT cannot observe below the water surface and therefore cannot measure $\bar{U}$ or *H* (hence the need for equation 9), and that all SWOT observations contain errors in both space and time (hence equation 10).
@@
 We validate BIKER as a timeseries of $k_{600}$ for each river using the BIKER posterior means.
-Our error metrics consider the timeseries nature of the problem and are formally defined in Table S2.
-They consist of the following:
+Our error metrics consider the timeseries nature of the problem and are formally defined in Table 1.
+They consist of the correlation coefficient *r* to quantify accuracy of BIKER's temporal dynamics, the root mean square error normalized by the observed mean (NRMSE) and normalized mean absolute residual error (NMAE) to assess bias, and the Kling-Gupta Efficiency (KGE).
+KGE is frequently used to assess streamflow prediction and simultaneously assesses accuracy in both bias and dynamics.
+While a value greater than -0.41 means the model outperforms a uniform prediction of the mean [@knobenTechnicalNoteInherent2019], generally KGE scores are interpreted as being meaningful in ungauged settings if \> 0.
 
-- Correlation coefficient *r* to quantify accuracy of BIKER's temporal dynamics
-- Root mean square error normalized by the observed mean (NRMSE) to assess bias
-- Normalized mean absolute residual error (NMAE) to assess bias
-- Kling-Gupta Efficiency (KGE).
-KGE is frequently used to assess streamflow prediction and simultaneously assesses accuracy in both bias and dynamics.
-While a value greater than -0.41 means the model outperforms a uniform prediction of the mean [@knobenTechnicalNoteInherent2019], generally KGE scores are interpreted as being meaningful in ungauged settings if > 0.
+*Table 1: Validation metrics used in this study, where Nt is number of observations and i is the specific observation.
+σ refers to the variance of the sample and μ refers to the mean of the sample.
+Carrot accents indicate the predicted value.*
+
+| **Description**                   | **Acronym** | **Definition**                                                                                                                                                                                                 | **Ideal Score** | **Possible Range** |
+|-----------------------------------|-------------|----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|-----------------|--------------------|
+| Correlation Coefficient           | $r$         | $\sqrt{1-(\frac{\sum\limits_{i=1}^{N_t}{(k_{600, i}-\hat{k_{600_i}})^2}}{\sum{(k_{600, i}-\mu_{k_{600}})^2}})}$                                                                                                | 1               | -1 to 1            |
+| Normalized root-mean-square error | NRMSE       | $\sqrt{\frac{1}{N_t}\sum\limits_{i=1}^{N_t}(\frac{\hat{k_{600, i}}-k_{600, i}}{\mu_{k_{600, i}}})^2}$                                                                                                          | 0               | 0 to ∞             |
+| Kling-Gupta Efficiency            | KGE         | $1 - \sqrt{ (r(k_{600_i}, \hat{k_{600_i}}) - 1)^2 + (\frac{\mu_{\hat{k_{600}}} }{\mu_{k_{600}}} - 1)^2 + (\frac{ \sigma _{\hat{k_{600}}} / \mu _{\hat{k_{600}}} }{ \sigma_{k_{600}} / \mu_{k_{600}}} - 1)^2 }$ | 1               | -∞ to 1            |
+| Normalized Mean Absolute Error    | MAE         | $\frac{\frac{\sum\limits_{i=1}^{N_t}(|\hat{k_{600_i}}-k_{600_i})|)}{n})}{\mu_{k_{600}}}$                                                                                                                       | 0               | 0 to ∞             |
 
 #### 3.2.3 Carbon emissions validation
+
 It is one thing to accurately model the temporal dynamics of $k_{600}$ as above, but researchers are often most interested in the actual carbon emitted from river to atmosphere.
@@
 
-Conversely, BIKER represents a new way of approaching equation 11 compared to existing literature models: BIKER has no reliance on a streaflow model nor HG scaling relationships and only requires that a river is SWOT-observable.
+Conversely, BIKER represents a new way of approaching equation 11 compared to existing literature models: BIKER has no reliance on a streamflow model nor HG scaling relationships and only requires that a river is SWOT-observable.
 We are therefore interested in how the final carbon emissions that result from BIKER compare against literature methods that use equation 11.
@@
 But, BIKER is only attractive if it can produce fluxes with similar errors to published methods.
-Text S5 and Table S3 fully describe these three literature models.
+Text S5 and Table S2 fully describe these three literature models.
 
-To benchmark BIKER against these literature methods, we pair the 26 biweekly $CO_2$ and water temperature samples from @beaulieuControlsGasTransfer2012 (section 3.1, figure S2) with every 14th set of daily SWOT observations (as the $CO_2$ data is bi-weekly).
+To benchmark BIKER against these literature methods, we pair the 26 biweekly $CO_2$ and water temperature samples from @beaulieuControlsGasTransfer2012 (section 3.1, Figure S3) with every 14th set of daily SWOT observations (as the $CO_2$ data is bi-weekly).
 We then calculate $FCO_2$ using equation 1, an atmospheric $CO_{2_{air}}$ of 400 uatm, and a *Sc* estimated following Raymond et al.
 (2012).
-The *k* in equation 1 is obatained using BIKER or the three literature models (Table S3 and Text S5).
-Finally, we estimate a yearly total carbon emission rate (via $CO_2$ evasion) by applying each river’s mean $FCO_2$ over the river’s surface area and summing all rates across rivers.
+The *k* in equation 1 is obtained using BIKER or the three literature models (Table S2 and Text S5).
+Finally, we estimate a pseudo yearly total carbon emission rate (via $CO_2$ evasion) by applying each river's mean $FCO_2$ over the river's surface area and summing all rates across rivers, remembering that we are applying 'unit' $CO_2$ data to all rivers.
 
 ### 3.3 Section 3 results
+
 #### 3.3.1 BIKER
-Figure 4 plots a representative set of the `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` timeseries of predicted and observed $k_{600}$, assuming no SWOT measurement error.
-Two rivers each were sampled from the three tertiles of river *KGE* scores (Table S2) for display.
-Consult Figure S3 for all 47 timeseries plots (assuming no measurement error) and Figure S4 for the 16 rivers with measurement errors.
+
+Figure 3 plots a representative set of the `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` timeseries of predicted and observed $k_{600}$, assuming no SWOT measurement error.
+Two rivers each were sampled from the three tertiles of river *KGE* scores (Table 1) for display.
+Consult Figure S4 for all 47 timeseries plots (assuming no measurement error) and Figure S5 for the 16 rivers with measurement errors.
 Note that the y axis is normalized by maximum observed values to compare across rivers.
@@
 Notably, there is sometimes positive bias in the estimates (e.g.
-both 'Middle Skill' rivers).
+the Connecticut and Missouri Midsection rivers).
 Some rivers yield the correct temporal dynamics, but the magnitude of these values is stretched relative to the observed (e.g.
@@
 
-![Figure 4.
+![Figure 3.
 Representative river timeseries plots of $k_{600}$.
@@
 Rivers are sampled from the three tertiles of KGE scores.
-(a-b) are poorest performing rivers, (b-c) are in the middle, and (e-f) are the best performing rivers.
+(a-b) are poorest performing rivers, (c-d) are in the middle, and (e-f) are the best performing rivers.
 Y axis is normalized by maximum observed values to compare visually.](cache/validation/fig4.jpg)
 
-Next, we calculate performance metrics following section 3.2.2 and Table S2.
-These are presented in Figure 5, which plots the scores for the 47 rivers with no measurement errors as empirical cumulative density functions (eCDFs).
+Next, we calculate performance metrics following section 3.2.2 and Table 1.
+These are presented in Figure 4, which plots the scores for the 47 rivers with no measurement errors as empirical cumulative density functions (eCDFs).
 Median river $KGE$ is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$KGE, na.rm=T),2)` and median river *r* is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$r, na.rm=T),2)`.
 This indicates very strong inference of each river's temporal $k_{600}$ dynamics given that absolutely no in situ information is being used to predict $k_{600}$.
-NRMSE has a median score of `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$NRMSE, na.rm=T),2)`, highlighting many rivers which have notable positive biases (Figure 4 also confirms this visually).
-Median NMAE, however, is only `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$NMAE, na.rm=T),2)` m/day.
+NRMSE has a median score of `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$NRMSE, na.rm=T),2)`, highlighting many rivers which have notable positive biases (Figure 3 also confirms this visually).
+Median NMAE is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$NMAE, na.rm=T),2)*100`%.
 Taken in aggregate, Figures 4-5 indicate that BIKER is quite good at capturing temporal dynamics in $k_600$, however there is often positive bias in its estimates.
 
-![Figure 5.
+![Figure 4.
 Performance metrics by river, ploted as empirical cummulative density functions (eCDFs).
-Each subpanel is labelled by its performance metric (defined in Table S2).
+Each subpanel is labelled by its performance metric (defined in Table 1).
 Dashed lines denote median scores.](cache/validation/fig5.jpg)
 
-Figure 6 compares BIKER results under the 'no-measurement-error' and 'measurement-error' scenarios for the `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` rivers for which @frassonExploringFactorsControlling2021 provide an error model.
+Figure 5 compares BIKER results under the 'no-measurement-error' and 'measurement-error' scenarios for the `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` rivers for which @frassonExploringFactorsControlling2021 provide an error model.
 Rivers that fall within the purple zone get worse when accounting for measurement error, while rivers in the green get better.
@@
 
-![Figure 6: Comparison of BIKER performance when considering measurement error in the SWOT observations for 16 rivers.
-Each subpanel is labelled by its performance metric (Table S2).
+![Figure 5: Comparison of BIKER performance when considering measurement error in the SWOT observations for 16 rivers.
+Each subpanel is labelled by its performance metric (Table 1).
 Rivers that fall within the purple zone get worse when accounting for measurement error, while rivers in the green get better.
-Note also that axes are flipped in order to visualize all 'better performances' in the upper-right-corner of each sub-plot.
+Note that some axes are flipped in order to visualize all 'better performances' in the upper-right-corner of each sub-plot.
 Black line denotes linear regression to aid in visualization.](cache/validation/fig6.jpg)
@@
 Finally, we sought to explore the overall influence of prior error/bias on $k_{600}$.
-We plot eCDFs of prior and posterior NMAE (see Table S2 for metric definition) across all `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers in Figure 7a.
+We plot eCDFs of prior and posterior NMAE (see Table 1 for metric definition) across all `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers in Figure 6a.
 Generally, errors/biases associated with the prior $k_{600}$ are propagated through the posterior in an approximately 1:1 manner (which is expected), except for a subset of rivers in which posterior error actually increases relative to the prior (approximately percentiles 0.6-0.9).
@@
 
-![Figure 7: Empirical cummulative density functions for prior and posterior $k_{600}$ NMAE.
-See Table S2 for metric definition.
+![Figure 6: Empirical cummulative density functions for prior and posterior $k_{600}$ NMAE.
+See Table 1 for metric definition.
 a) all rivers.
-b) Only rivers with a coefficient of varaiton (CV) of observed $k_{600}$ > 0.10.
-c) Only rivers with CV > 0.20.
-d) Only rivers with CV > 0.30.
+b) Only rivers with a coefficient of varaiton (CV) of observed $k_{600}$ \> 0.10.
+c) Only rivers with CV \> 0.20.
+d) Only rivers with CV \> 0.30.
 Overall we see that rivers with greater temporal variation in $k_{600}$ behave better for BIKER, i.e.
@@
 Recall that BIKER relies on a timeseries of SWOT data, and that these timeseries may not be representative of the full spectrum of $k_{600}$ values that are actually experienced in the river, therefore potentially biasing both the prior estimation methods (Text S3) and the actual Bayesian inference.
-Put another way, we suggest that a SWOT timeseries that does not sufficiently capture a river's temporal dynamics will introduce additional error to the inference results.
-To test this, we subset our validation dataset by progressively higher coefficients of variation (*CV*) (> 10%, >20%, and >30%).
-These filtered datasets are plotted in Figures 7b-7d (for > 10%, >20%, and >30% *CV*, respectively), where our hypothesis is confirmed.
-Posterior bias systematically drops as the temporal variability of the SWOT data increases, with BIKER posterior error the smallest, and much less than prior error, in Figure 7d.
+Put another way, we suggest that if a SWOT timeseries does not sufficiently capture a river's temporal dynamics, it will introduce additional error to the inference results.
+To test this, we subset our validation dataset by progressively higher coefficients of variation (*CV*) (\> 10%, \>20%, and \>30%).
+These filtered datasets are plotted in Figures 6b-6d (for \> 10%, \>20%, and \>30% *CV*, respectively), where our hypothesis is confirmed.
+Posterior bias systematically drops as the temporal variability of the SWOT data increases, with BIKER posterior error the smallest, and much less than prior error, in Figure 6d.
 This result is elaborated on in section 4.2.
@@
 #### 3.3.2 Carbon Emissions
+
 Finally, we carry these $FCO_2$ calculations all the way to annual carbon emissions rates and compare BIKER against established in situ techniques.
 It is important to remember that only BIKER is completely globally implementable, while the other three models necessarily rely on having a streamflow record or high-quality routed streamflow readily available.
-Therefore, the HG methods reflect their 'best-case scenarios' while BIKER reflects its worst case scenario, where priors are set entirely from SWOT observations and are generally the least informative they will ever be.
+Therefore, the in situ methods reflect their 'best-case scenarios' while BIKER reflects its worst case scenario, where priors are set entirely from SWOT observations and are generally the least informative they will ever be.
 This means that BIKER's annual carbon emission estimate can only improve from what is presented here.
@@
 
-Figure 8 compares the annual carbon emissions rate (via $F_{CO_2}$) from the rivers using BIKER posterior means and the three streamgauge-based HG models.
+Figure 7 compares the annual carbon emissions rate (via $F_{CO_2}$) from the rivers using BIKER posterior means and the three streamgauge-based HG models.
 Surprisingly, BIKER outperforms the gauge-based approaches, nearly correctly inferring the annual carbon emissions rate (`r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_BIKER',]$yearlyCFux, 2)` Tg-C/yr for BIKER versus `r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_obs',]$yearlyCFux, 2)` Tg-C/yr observed).
@@
 'Brinkerhoff 2019''s lower CI is comparable to BIKER's higher CI.
-BIKER's relatively stronger performance than the in situ HG models is elaborated on in section 4.3.
-Thus, despite BIKER using absolutely no streamgauge or other in situ data like the other 3 methods do, it provides a strong upscaling estimate of the annual carbon emission rate for the rivers (Figure 8).
+BIKER's relatively stronger performance than the in situ models is elaborated on in section 4.3.
+Despite BIKER using absolutely no streamgauge or other in situ data like the other 3 methods do, it provides a strong upscaling estimate of the annual carbon emission rate for the rivers (Figure 7).
 
-![Figure 8: Yearly carbon emissions rate via $CO_2$ evasion across all rivers.
+![Figure 7: Yearly carbon emissions rate via $CO_2$ evasion across all rivers.
 Completely remotely-sensed methods are colored in red, in situ methods in purple, and the observed in orange.
@@
 ## 4 Discussion
+
 ### 4.1 Gas exchange in hydraulically-wide rivers
-Field studies of gas exchange in wide rivers have suggested that $k_{600}$ behaves differently in these rivers than in steeper and smaller rivers [@ulsethDistinctAirWater2019; @beaulieuControlsGasTransfer2012; @alinPhysicalControlsCarbon2011; @wangPhysicallyBasedScaling2021; @raymondGasExchangeRivers2001a].
+
+Field studies of gas exchange in wide rivers have suggested that $k_{600}$ behaves differently in these rivers than in steeper and smaller rivers [@ulsethDistinctAirWater2019; @beaulieuControlsGasTransfer2012; @alinPhysicalControlsCarbon2011; @wangPhysicallyBasedScaling2021; @raymondGasExchangeRivers2001].
 While much work has focused on the small-stream side of the stream-to-river continuum, comparatively less work has been done in large systems.
 Here, we focus on the larger, 'smooth-channel' end of the continuum, using a model for gas exchange that scales $k_{600}$ by both $\epsilon_D$ and a shear Reynold's number.
-This model is empirically validated in Figure 3.
+This model is empirically validated in Figure 2.
 Specifically, Figure S1 confirms that scaling $k_{600}$ with a shear Reynold's adaption of the small-eddy model (equation 7) reduces positive bias in the model's predictions of the smallest $k_{600}$ values [where the relative decrease in turbulence reaching the surface is greater than the small-eddy model alone suggests, per equation 7's theoretical basis- @moogAirWaterGasTransfer1999].
@@
 To our knowledge, @moogAirWaterGasTransfer1999's specific setup, which imposes a 'differentially-experienced' surface turbulence theory on the small-eddy model, has never been empirically validated in rivers.
-Figure 3 provides this empirical verification for hydraulically wide channels, where it's theoretical basis should generally hold.
+Figure 2 provides this empirical verification for hydraulically wide channels, where it's theoretical basis should generally hold.
 Using our full dataset of $k_{600}$, we also observed this model breaks down when including non-hydraulically-wide rivers (as the theory would suggest).
@@
 
-Crucially, we are not accounting for wind-driven gas exchange, which is suggested to play an important role in wide rivers because river surface area is sufficiently large that wind can drive gas exchange [@beaulieuControlsGasTransfer2012; @wangPhysicallyBasedScaling2021; @raymondGasExchangeRivers2001a].
+Crucially, we are not accounting for wind-driven gas exchange, which is suggested to play an important role in wide rivers because river surface area is sufficiently large that wind can drive gas exchange [@beaulieuControlsGasTransfer2012; @wangPhysicallyBasedScaling2021; @raymondGasExchangeRivers2001].
 Similarly, none of the existing hydraulics-driven fluvial $k_{600}$ models account for wind-driven gas exchange either.
@@
 Such a model would likely take a form similar to the conceptual models proposed by @wangPhysicallyBasedScaling2021, @chuWindStreamFlow2003, and/or @plateReaerationOpenChannel1984.
-All three generally propose calculating both a wind-driven *k* and a hydraulics-driven *k* and then weighting each term via parameters.
+All three generally propose calculating both a wind-driven *k* and a hydraulics-driven *k* and then weighting each term.
 In the case of BIKER, this would necessitate additional parameters that would need to be known a priori for specific rivers.
@@
 ### 4.2 Towards remote sensing of global spatiotemporal dynamics of $k_{600}$ in large rivers
+
 To date, the studies exploring the spatiotemporal dynamics of riverine gas exchange have arguably been held back by a lack of data.
@@
 For reference, 95% of the SWOT-visible rivers globally (202,811) will have sufficient SWOT observations along the river to run BIKER at least once every 21 days, with most of the temperate and Arctic rivers having 3+ observations per 21-day cycle [@altenauSurfaceWaterOcean2021].
-While BIKER will not directly measure $k_{600}$, it does robustly infer temporal trends in $k_{600}$ and less robustly the absolute magnitude of $k_{600}$ (Figures 4-7) and when coupled with SWOT this much data will provide a novel dataset of $k_{600}$ on a scale never before possible.
+While BIKER will not directly measure $k_{600}$, it does robustly infer temporal trends in $k_{600}$ and less robustly the absolute magnitude of $k_{600}$ (Figures 4-6) and when coupled with SWOT this much data will provide a novel dataset of $k_{600}$ on a scale never before possible.
 
-With that said, Figures 4, 5, S3, and S4 all highlight a substantial range of algorithm performances across rivers.
+With that said, Figures 4, 5, S4, and S5 all highlight a substantial range of BIKER performance across rivers.
 These differences in performance are likely due to the representativeness of the priors used for that river.
@@
 
-Furthermore, Figure 7 implies that a sufficient variability in SWOT observations is necessary to strongly infer a $k_{600}$ timeseries.
+Furthermore, Figure 6 implies that a sufficient variability in SWOT observations is necessary to strongly infer a $k_{600}$ timeseries.
 'Hydraulic visibility', i.e.
-the ability of a remote sensor to identify a hydrological response in the river [@garamboisHydraulicVisibilityUsing2017a] is applicable here.
-If we apply hydraulic visibility to a sensor's ability to identify temporal variations in $k_{600}$, Figure 7 suggests that a 'minimally sufficient' hydraulic variability in SWOT measurements (in Figure 7b, suggested to be > 10% *CV*) is needed to uniformally improve upon the prior.
+the ability of a remote sensor to identify a hydrological response in the river [@garamboisHydraulicVisibilityUsing2017] is applicable here.
+If we apply hydraulic visibility to a sensor's ability to identify temporal variations in $k_{600}$, Figure 6 suggests that a 'minimally sufficient' hydraulic variability in SWOT measurements (in Figure 6b, suggested to be \> 10% *CV*) is needed to uniformally improve upon the prior.
 This will be important once SWOT launches and BIKER is implemented at the global-scale.
@@
 ### 4.3 Coupling BIKER with upscaling workflows
-Figure 8 confirms that BIKER is quite successful at predicting annual upscaled carbon emissions from the river network when coupled with $CO_2$ data.
+
+Figure 7 confirms that BIKER is quite successful at predicting annual upscaled carbon emissions from the river network when coupled with $CO_2$ data.
 This encouraging result has three main implications for future work.
 
-First, figure 8 directly implies that BIKER will be useful when coupled with large-scale $CO_{2_{water}}$ models, provided these models are accurate.
-The models would give time and spcae varying gas exchange.
-**Liu etal in review** and @saccardiImprovingPredictionsStream2021 each propose models that robustly predict reach-scale dissolved $CO_2$ concentrations using two different approaches: machine learning [**Liu et al in review**] and process-based reactive transport modeling [@saccardiImprovingPredictionsStream2021], but both models yield $CO_2$ estimates that would be spatially and temporally consistent with BIKER's output.
+First, Figure 7 directly implies that BIKER will be useful when coupled with large-scale $CO_{2_{water}}$ models, provided these models are accurate.
+The models would give time and space varying gas exchange.
+@liuImportanceHydrologyRouting2022 and @saccardiImprovingPredictionsStream2021 each propose models that robustly predict reach-scale dissolved $CO_2$ concentrations using two different approaches: machine learning [@liuImportanceHydrologyRouting2022] and process-based reactive transport modeling [@saccardiImprovingPredictionsStream2021], but both models yield $CO_2$ estimates that would be spatially and temporally consistent with BIKER's output.
 Our promising results suggest that BIKER could provide additional (and directly inferred) measurements of $k_{CO_2}$ to these models, thereby better informing model results through direct observations.
-This is likely to be accomplished via data assimilation which has proven useful in using remotly-sensed discharge to improve streamflow routing models [@fengRecentChangesArctic2021a; @ishitsukaCombiningOpticalRemote2020], and of which the @saccardiImprovingPredictionsStream2021 $CO_2$ model takes a similar form.
+This is likely to be accomplished via data assimilation which has proven useful in using remotly-sensed discharge to improve streamflow routing models [@fengRecentChangesArctic2021; @ishitsukaCombiningOpticalRemote2020], and of which the @saccardiImprovingPredictionsStream2021 $CO_2$ model takes a similar form.
 
-Second, at the field-scale Figure 8 confirms that we can couple BIKER with in situ gas concentration loggers to produce $FCO_2$ estimates at novel temporal resolutions in SWOT-observable rivers.
+Second, at the field-scale Figure 7 confirms that we can couple BIKER with in situ gas concentration loggers to produce $FCO_2$ estimates at novel temporal resolutions in SWOT-observable rivers.
 High temporal fidelity datasets of dissolved $CO_2$ in SWOT-observable rivers now exist [e.g.
@@
 
-Finally, Figure 8 also uniquely allows us to directly compare the influence of geomorphic assumptions on total carbon emission rates from river networks, as all other calculations and parameters were held constant across our four tested models (Text S5).
-Therefore, figure 8 highlights a potentially large source of uncertainty in current river $CO_2$ upscaling estimates: the geomorphic models employed to scale river channel hydraulics with streamflow.
-In this case, the only difference between the three literature models and the observed estimate in Figure 8 is the specific HG model employed to predict river depth and velocity (Text S5, Table S3), and yet the eventual carbon emissions estimates (Figure 8) are quite different.
-Further, recall that the BIKER results in Figure 8 reflect a worst-case scenario (relatively uniformed priors), while the three in-situ methods represent  best case scenarios (perfect streamflow records).
+Finally, Figure 7 also uniquely allows us to directly compare the influence of geomorphic assumptions on total carbon emission rates from river networks, as all other calculations and parameters were held constant across our four tested models (Text S5).
+Therefore, Figure 7 highlights a potentially large source of uncertainty in current river $CO_2$ upscaling estimates: the geomorphic models employed to scale river channel hydraulics with streamflow.
+In this case, the only difference between the three literature models and the observed estimate in Figure 7 is the specific HG model employed to predict river depth and velocity (Text S5, Table S2), and yet the eventual carbon emissions estimates (Figure 7) are quite different.
+Further, recall that the BIKER results in Figure 7 reflect a worst-case scenario (relatively uniformed priors), while the three in-situ methods represent best case scenarios (perfect streamflow records).
 We suggest future work should perform a formal sensitivity analysis for these HG parameters.
@@
 ## 5 Conclusions
+
 The $CO_2$ flux from the global river network is a major natural component of the global carbon cycle, on par with total forest carbon uptake [@panLargePersistentCarbon2011].
@@
 ## 6 Acknowledgements
+
 C.B.
-Brinkerhoff was funded by a NASA Future Investigators in Earth and Space Science Fellowship (#xxxxxx).
+Brinkerhoff was funded by a NASA Future Investigators in Earth and Space Science Fellowship (80NSSC21K1591).
 C.J.
-Gleason was supported by NASA SWOT Science Team grant #yyyyyy.
-The version of BIKER used in this study will be available upon acceptance.
-BIKER remains in active development and is available at https://github.com/craigbrinkerhoff/BIKER.
-All code to build and generate results, figures, and the manuscript is available at https://github.com/craigbrinkerhoff/RSK600.
+Gleason was supported by a NASA SWOT Science Team grant (80NSSC20K1141).
 We thank Renato Frasson, Michael Durand, Amber Ulseth, and Jake Beaulieu for generously making their data available for this study.
@@
 
-## 7 Apendix A
+## 7 Data availability statement
+The version of BIKER used in this study will be archived upon acceptance.
+BIKER remains in active development and is available at <https://github.com/craigbrinkerhoff/BIKER>.
+All code to build and generate results, figures, and the manuscript is available at <https://github.com/craigbrinkerhoff/RSK600> and will be archived upon acceptance.
+
+## 8 Apendix A
+
 *Table A1: Variable description and notation for this study*
 
-|**Notation**|**Description**|**Calculation (if not directly measured)**|**Units**|
-|------------|---------------|-------------------------------|---------|
-|$A$|Channel cross-sectional area|NA|$[L^2]$|
-|$A_0$|Non-SWOT-observable cross-sectional area|NA|$[L^2]$|
-|$\alpha$|Statistical parameter for equation 5 scaling relation|NA|$\frac{m^{1/2}s^{3/4}}{dy}$|
-|$\alpha_1$|Statistical parameter for equation 6 scaling relation|NA|$\frac{m^{1/2}s^{3/4}}{dy}$|
-|*Bulk carbon efflux*|carbon mass transport rate from river to atmosphere|NA|$[\frac{M}{T}]$|
-|$\beta$|Statistical parameter for equation 7 scaling relation|NA|$\frac{m^{-1/4}s^{9/8}}{dy}$|
-|$\beta_1$|Statistical parameter for equation 8 scaling relation|NA|$\frac{m^{1/2}s^{3/4}}{dy}$|
-|$CO_{2_{water}}$|Water-side $CO_2$ concentration|NA|$[L]$|
-|$CO_{2_{air}}$|Atmospheric-side $CO_2$ concentration|NA|$[L]$|
-|$dA$|change in cross-sectional area|$\delta H_e*W$|$[L]$|
-|$D_m$|Molecular diffusion coefficient|NA|$[\frac{L^2}{T}]$|
-|$\epsilon$|Dissipation rate of near-surface turbulence|NA|$[\frac{E}{M*T}]$|
-|$\epsilon_S$|log-law-of-the-wall model for $\epsilon$|$\frac{U_*^3}{H}$|$[\frac{E}{M*T}]$|
-|$\epsilon_D$|Form-drag model for $\epsilon$|$gS\bar{U}$|$[\frac{E}{M*T}]$
-|$FCO_2$|$CO_2$ evasion flux from river to atmosphere|NA|$[\frac{M}{L^2T}]$|
-|$FCO_{2_{upscaled}}$|Upscaling estimate of the  global $CO_2$ evasion flux from river to atmosphere|NA|$[\frac{M}{L^2T}]$|
-|$g$|gravitational acceleration|9.8|$[\frac{L}{T^2}]$|
-|$H$|Mean flow depth|$A/W$|$[L]$|
-|$H_e$|Water surface elevation|NA|$[L]$|
-|$i$|Cross-section discretization within river reach|NA|NA|
-|$k$|gas exchange velocity|NA|$[\frac{L}{T}]$|
-|$k_{600}$|gas exchange velocity normalized to $Sc=600$|$(\frac{600}{Sc_k})^{-0.5}k$|$[\frac{L}{T}]$|
-|$k_{600_{obs}}$|"observed" gas exchange velocity normalized to $Sc=600$|$k_{600_{obs}}=62.82(gS)^{7/16}\bar{U}^{1/4}H^{9/16}$|$[\frac{L}{T}]$|
-|$n$|Manning's roughness coefficient|$\frac{R_h^{2/3}S^{1/2}}{\bar{U}}$|$[\frac{T}{L^3}]$|
-|$\uprho$|Density of water|NA|$[\frac{M}{L^3}]$|
-|$Q$|River discharge|NA|$\frac{L^3}{T}$|
-|$R_h$|Hydraulic radius|$\frac{AW}{2H+W}$|$[L]$|
-|$S$|River slope|NA|$unitless$|
-|$Sc$|Schmidt number|$\frac{v}{D_m}$|NA|
-|$\sigma^2_{k_{600}}$|BIKER uncertainty (posterior distribution standard deviation)|NA|$[L/T]$|
-|$t$|timestep discretization within river reach|NA|NA|
-|$\Theta$|Bayesian parameter set|NA|NA
-|$\bar{U}$|Cross-sectional average velocity|$\frac{Q}{A}$|$[\frac{L}{T}]$|
-|$U_*$|Shear velocity|$\sqrt{gSR_h}$|$[\frac{L}{T}]$|
-|$\mu$|Viscosity|NA|$[P*T]$|
-|$v$|kinematic viscosity|$\frac{\mu}{\uprho}$|$[\frac{L^2}{T}]$|
-|$W$|Flow width|NA|$[L]$|
-|$x$|Bayesian data set|NA|NA|
+| **Notation**         | **Description**                                                               | **Calculation (if not directly measured)** | **Units**                    |
+|----------------------|-------------------------------------------------------------------------------|--------------------------------------------|------------------------------|
+| $A$                  | Channel cross-sectional area                                                  | NA                                         | $[L^2]$                      |
+| $A_0$                | Non-SWOT-observable cross-sectional area                                      | NA                                         | $[L^2]$                      |
+| $\alpha$             | Statistical parameter for equation S5 scaling relation                        | NA                                         | $\frac{m^{1/2}s^{3/4}}{dy}$  |
+| $\alpha_1$           | Statistical parameter for equation S7 scaling relation                        | NA                                         | $\frac{m^{1/2}s^{3/4}}{dy}$  |
+| $\beta$              | Statistical parameter for equation S13 scaling relation                       | NA                                         | $\frac{m^{-1/4}s^{9/8}}{dy}$ |
+| $\beta_1$            | Statistical parameter for equation 7 scaling relation                         | NA                                         | $\frac{m^{1/2}s^{3/4}}{dy}$  |
+| $CO_{2_{water}}$     | Water-side $CO_2$ concentration                                               | NA                                         | $[\frac{M}{L^3}]$            |
+| $CO_{2_{air}}$       | Atmospheric-side $CO_2$ concentration                                         | NA                                         | $[\frac{M}{L^3}]$            |
+| $dA$                 | change in cross-sectional area                                                | $\delta H_e*W$                             | $[L]$                        |
+| $D_m$                | Molecular diffusion coefficient                                               | NA                                         | $[\frac{L^2}{T}]$            |
+| $\epsilon$           | Dissipation rate of near-surface turbulence                                   | NA                                         | $[\frac{E}{M*T}]$            |
+| $\epsilon_S$         | log-law-of-the-wall model for $\epsilon$                                      | $\frac{U_*^3}{H}$                          | $[\frac{E}{M*T}]$            |
+| $\epsilon_D$         | Form-drag model for $\epsilon$                                                | $gS\bar{U}$                                | $[\frac{E}{M*T}]$            |
+| $FCO_2$              | $CO_2$ evasion flux from river to atmosphere                                  | NA                                         | $[\frac{M}{L^2T}]$           |
+| $FCO_{2_{upscaled}}$ | Upscaling estimate of the global $CO_2$ evasion flux from river to atmosphere | NA                                         | $[\frac{M}{L^2T}]$           |
+| $g$                  | gravitational acceleration                                                    | 9.8                                        | $[\frac{L}{T^2}]$            |
+| $H$                  | Mean flow depth                                                               | $\frac{A}{W}$                              | $[L]$                        |
+| $H_e$                | Water surface elevation                                                       | NA                                         | $[L]$                        |
+| $i$                  | Cross-section discretization within a mass-conserved river reach              | NA                                         | NA                           |
+| $k$                  | gas exchange velocity                                                         | NA                                         | $[\frac{L}{T}]$              |
+| $k_{600}$            | gas exchange velocity normalized to $Sc=600$                                  | $(\frac{600}{Sc})^{-0.5}k$                 | $[\frac{L}{T}]$              |
+| $n$                  | Manning's roughness coefficient                                               | $\frac{R_h^{2/3}S^{1/2}}{\bar{U}}$         | $[\frac{T}{L^3}]$            |
+| $\uprho$             | Density of water                                                              | NA                                         | $[\frac{M}{L^{1/3}}]$        |
+| $Q$                  | River discharge                                                               | $WH\bar{U}$                                | $\frac{L^3}{T}$              |
+| $R_h$                | Hydraulic radius                                                              | $\frac{AW}{2H+W}$                          | $[L]$                        |
+| $S$                  | River slope                                                                   | NA                                         | $unitless$                   |
+| $Sc$                 | Schmidt number                                                                | $\frac{v}{D_m}$                            | NA                           |
+| $t$                  | timestep discretization within river reach                                    | NA                                         | NA                           |
+| $\Theta$             | Bayesian parameter set                                                        | NA                                         | NA                           |
+| $\bar{U}$            | Cross-sectional average velocity                                              | $\frac{Q}{A}$                              | $[\frac{L}{T}]$              |
+| $U_*$                | Shear velocity                                                                | $\sqrt{gSR_h}$                             | $[\frac{L}{T}]$              |
+| $\mu$                | Viscosity                                                                     | NA                                         | $[P*T]$                      |
+| $v$                  | kinematic viscosity                                                           | $\frac{\mu}{\uprho}$                       | $[\frac{L^2}{T}]$            |
+| $W$                  | Flow width                                                                    | NA                                         | $[L]$                        |
+| $x$                  | Bayesian data set                                                             | NA                                         | NA                           |
 

### 2022-03-09 src/figures_k600.R

@@ -23,16 +23,17 @@ stats_by_river <- group_by(results, river, errFlag) %>%
   summarise(r = sqrt(summary(lm(kest_mean~kobs))$r.squared),
             NRMSE = sqrt(mean((kest_mean - kobs)^2, na.rm=T)) / mean(kobs, na.rm=T),
             KGE = KGE(kest_mean, kobs),
-            MAE = mean(abs(kest_mean-kobs), na.rm=T),
-            MAE_prior = mean(abs(kprior - kobs), na.rm=T),
+            NMAE = mean(abs(kest_mean-kobs)/mean(kobs, na.rm=T), na.rm=T),
+            NMAE_prior = mean(abs(kprior - kobs), na.rm=T)/mean(kobs, na.rm=T),
             meanKobs = mean(kobs, na.rm=T),
             meanWobs = mean(Wobs, na.rm=T),
             meanSobs = mean(Sobs, na.rm=T),
             meanDobs = mean(Dobs, na.rm=T),
-            n_data=n())
+            n_data=n(),
+            cvObs = sd(kobs, na.rm=T)/mean(kobs))
 
-plot_stats <- gather(stats_by_river, key=key, value=value, c('NRMSE', 'MAE', 'KGE', 'r'))
-plot_stats <- filter(plot_stats, key %in% c('NRMSE', 'KGE', 'MAE', 'r'))
+plot_stats <- gather(stats_by_river, key=key, value=value, c('NRMSE', 'NMAE', 'KGE', 'r'))
+plot_stats <- filter(plot_stats, key %in% c('NRMSE', 'KGE', 'NMAE', 'r'))
 
 ########################
 ##PLOT BY-RIVER ERROR METRICS w/ eCDFs---------------------------------------
@@ -70,68 +71,27 @@ nrmse_cdf <- ggplot(stats_by_river[stats_by_river$errFlag==0,], aes(x=NRMSE)) +
         legend.text = element_text(size=17),
         plot.title = element_text(size = 30, face = "bold"))
 
-mae_cdf <- ggplot(stats_by_river[stats_by_river$errFlag==0,], aes(x=MAE)) +
+nmae_cdf <- ggplot(stats_by_river[stats_by_river$errFlag==0,], aes(x=NMAE)) +
   stat_ecdf(size=3, color='#1c9099') +
   scale_x_log10()+
   geom_hline(yintercept = 0.5, linetype='dashed', size=1) +
[…]

### 2022-03-09 src/old/roughness_k600.R

@@ -1,315 +1,0 @@
-###########################
-#Description: Modeling k600 from hydraulics/roughness. Using open-channel flow to explain k600 scaling dyanmics with surface turbulence
-#Creator: Craig Brinkerhoff
-#Spring 2021
-###########################
-
-print('starting k600 theory...')
-
-options(scipen = 999)
-
-#Raymond 2012 morphology stuff----------------------------------------------------
-raymond2012 <- read.csv('inputs/k600/raymond2012.csv')
-raymond2012 <- filter(raymond2012, is.na(k600)==0 & k600 > 0)
-raymond2012$Reach.Control <- as.character(raymond2012$Reach.Control)
-raymond2012$Bottom.Description <- as.character(raymond2012$Bottom.Description)
-
-raymond2012$Reach.Control <- ifelse(raymond2012$Reach.Control != 'Channel Control' & raymond2012$Reach.Control != 'Pool and Riffle', 'Hybrid', raymond2012$Reach.Control)
-
-raymond_summary <- group_by(raymond2012, Reach.Control) %>%
-  tally()
-
-qual_plot <- ggplot(raymond2012, aes(x=Reach.Control, fill=Reach.Control, y=k600)) +
-  geom_violin(size=1.2, color='black', draw_quantiles=c(0.50)) +
-  scale_fill_brewer(palette = 'Set2', name='') +
-  geom_text(data = raymond_summary, aes(Reach.Control, Inf, label = n), vjust = 1, size=8)+
-  scale_y_log10(
-    breaks = scales::trans_breaks("log10", function(x) 10^x),
-    labels = scales::trans_format("log10", scales::math_format(10^.x))
-  ) +
-  xlab('') +
-  ylab('K600 [m/day]') +
-  theme(legend.position = 'none',
-        axis.text=element_text(size=20),
-        axis.title=element_text(size=24,face="bold"),
-        legend.text = element_text(size=17),
-        legend.title = element_text(size=17, face='bold'))
-ggsave('outputs//k600//roughness_theory//qualPlot.jpg', qual_plot, width=9, height=7)
-
-#Ulseth etal 2019------------------------------------------------------------
-ulseth_data <- read.csv('data/k600/Ulseth_etal_2019.csv', fileEncoding="UTF-8-BOM") #contains data from 5 studies
[…]

### 2022-03-09 src/old/scraps.R

@@ -1,177 +1,0 @@
-###########
-##SCRAPS-----------------------------------
-######
-#test plot------------------------
-#prepare ulseth data for this plot...
-#ulseth_data <- read.csv('data/k600/Ulseth_etal_2019.csv', fileEncoding="UTF-8-BOM") #contains data from 5 studies
-#ulseth_data <- ulseth_data[,-8]
-#ulseth_data$data <- ifelse(ulseth_data$data == 'This Study', 'Ulseth et al. 2019', as.character(ulseth_data$data)) #fix some labeling
-#ulseth_data$data <- ifelse(ulseth_data$data == 'Raymond et al. 2012', 'Raymond et al. 2012 [5 studies]', as.character(ulseth_data$data))
-
-#data <- ulseth_data
-#data <- filter(data, is.na(width)==0) #filter out no width measurements
-
-#calculate eD and other hydraulics---------------------------------------------------------
-#data$eD <- g * data$slope * data$Vms #dissipation rate of surface turbulence originating from depth-scale form drag [W/kg]
-#data$ustar <- sqrt(g*data$slope*data$depth)
-#data$eS <- (sqrt(data$depth*g*data$slope))^3/data$depth #dissipation rate of surface turbulence originating from bed friction [W/kg]
-#data$Rh <- (data$depth*data$width)/(data$width + 2*data$depth) #hydraulic radius [m]
-#data$area <- data$width*data$depth #channel area [m2]
-#data$keulegan <- 11*data$depth*(1/exp(data$Vms/(2.5*(9.8*data$depth*data$slope)^(1/2)))) #[m]
-
-#log transform some variables
-#data$log_eD <- log(data$eD)
-#data$log_k600 <- log(data$k600)
-#data$log_slope <- log(data$slope)
-#data$log_area <- log(data$area)
-
-#dummy values with regression models
-#lseq <- function(from, to, length.out) {
-  # logarithmic spaced sequence
-  # blatantly stolen from library("emdbook"), because need only this
-#  exp(seq(log(from), log(to), length.out = length.out))
-#}
-
-#test <- data.frame('ustar'=seq(0,0.8, 0.1)) #data.frame('eD'=lseq(10^-5, 10^1, 100))
-#test$k600_1d <- 48*test$ustar #exp(4.44383)*test$eD^0.59957
-#test$k600_2d <- ifelse(test$eD < exp(-6.494), exp(0.6277)*test$eD^(0.030075),
[…]

### 2022-03-09 src/old/swot_k_model.R

@@ -1,468 +1,0 @@
-#################
-##Description: Turbulence controls on k~turbulence dissipation relationship
-##Creator: Craig Brinkerhoff
-##Date: Summer 2021
-#################
-
-#######SETUP-----------------------------
-library(tidyverse)
-library(colorspace)
-library(cowplot)
-library(segmented)
-theme_set(theme_cowplot())
-
-setwd('C:\\Users\\craig\\Documents\\OneDrive - University of Massachusetts\\Ongoing Projects\\RSK600')
-
-#some constants
-g <- 9.8
-
-#########READ IN HYDRAULICS DATA FROM BRINKERHOFF ETAL 2019------------------------------
-data <- read.csv('data\\Brinkerhoff_etal_2019\\field_measurements.csv')
-nhd_med <- read.csv('data\\Brinkerhoff_etal_2019\\NHD_join_table.csv')
-data <- left_join(data, select(nhd_med, SLOPE, SOURCE_FEA), by = c("site_no" = "SOURCE_FEA"))
-
-data <- filter(data, is.finite(chan_width) ==1) %>%
-  filter(is.finite(chan_velocity)==1) %>%
-  filter(is.finite(chan_discharge)==1) %>%
-  filter(is.finite(chan_area)==1) %>%
-  filter(chan_width > 0) %>%
-  filter(chan_velocity > 0) %>%
-  filter(chan_discharge > 0) %>%
-  filter(chan_area > 0) %>%
-  filter(measured_rating_diff %in% c('Excellent', 'EXCL', 'GOOD', 'Good', 'Fair', 'FAIR')) %>%
-  filter(is.finite(SLOPE)==1)
-
-#imperial to metric
-data$area <- data$chan_area * 0.092903 #ft2 to m2
-data$width <- data$chan_width*0.305 #m
-data$Vms <- data$chan_velocity*0.305 #m/s
-data$depth <- data$chan_area / data$width
-data$slope <- data$SLOPE
-data$Qm3s <- data$chan_discharge * 0.0283 #ft3/s to m3/s
-
-######READ IN FIELD DATA WITH K600-------------------
-#Ulseth etal 2019
-ulseth_data <- read.csv('data/Ulseth_etal_2019.csv', fileEncoding="UTF-8-BOM")
-ulseth_data <- ulseth_data[,-8]
-ulseth_data$dataset <- ifelse(ulseth_data$data == 'This Study', 'Ulseth et al. 2019', as.character(ulseth_data$data)) #fix some labeling
[…]

### 2022-03-09 src/old/swot_k_model_v2.R

@@ -1,375 +1,0 @@
-#################
-##Description: Developing gas exchange model for SWOT-observable rivers
-##Creator: Craig Brinkerhoff
-##Date: Fall 2021
-#################
-
-#######SETUP-----------------------------
-library(tidyverse)
-library(colorspace)
-library(cowplot)
-library(segmented)
-theme_set(theme_cowplot())
-
-setwd('C:\\Users\\cbrinkerhoff\\OneDrive - University of Massachusetts\\Ongoing Projects\\RSK600')
-
-#some constants
-g <- 9.8 #gravitational acceleration [m/s2]
-
-#######READ IN HYDRAULICS DATA FROM BRINKERHOFF ETAL 2019 TO GET SWOT VERSUS INEFFICIENT CHANNELS------------------------------
-data <- read.csv('data\\Brinkerhoff_etal_2019\\field_measurements.csv')
-
-#some necessary filtering
-data <- filter(data, is.finite(chan_width) ==1) %>%
-  filter(is.finite(chan_velocity)==1) %>%
-  filter(is.finite(chan_discharge)==1) %>%
-  filter(is.finite(chan_area)==1) %>%
-  filter(chan_width > 0) %>%
-  filter(chan_velocity > 0) %>%
-  filter(chan_discharge > 0) %>%
-  filter(chan_area > 0) %>%
-  filter(measured_rating_diff %in% c('Excellent', 'EXCL', 'GOOD', 'Good', 'Fair', 'FAIR'))
-
-#imperial to metric
-data$area <- data$chan_area * 0.092903 #ft2 to m2
-data$width <- data$chan_width*0.305 #m
-data$Vms <- data$chan_velocity*0.305 #m/s
-data$depth <- data$area / data$width
-data$slope <- NA #data$SLOPE
-data$Qm3s <- data$chan_discharge * 0.0283 #ft3/s to m3/s
-
-#calculate bankfull width so we can throw out out-of-bank events
-bankfullW = group_by(data, site_no) %>%
-  filter(n() >= 20) %>%
-  mutate(rank = rank(width, ties.method="first")) %>%
-  mutate(n = n()) %>%
-  mutate(desiredRank = round((n+1)/2, digits=0)) %>% #2 for a two year return period
-  mutate(temp = which(rank == desiredRank)) %>%
-  summarise(bankful_width = max(width[temp]))
-data <- left_join(data, bankfullW, 'site_no')
-
-#calculate bankfull depth so we can throw out out-of-bank events
[…]

### 2022-04-05 docs/supp_v10.Rmd
final ms edits

--- 2022-02-24 supp_v9.Rmd
@@
   word_document:
-      pandoc_args:
-        - '--lua-filter=scholarly-metadata.lua'
-        - '--lua-filter=author-info-blocks.lua'
+   #   pandoc_args:
+  #      - '--lua-filter=scholarly-metadata.lua'
+  #      - '--lua-filter=author-info-blocks.lua'
       reference_docx: word-styles-reference.docx
@@
 ## Contents
-This supplementary information contains 3 texts, 4 figures and 3 tables.
-Please consult https://github.com/craigbrinkerhoff/RSK600 for all code to build and generate results, figures, and the manuscript.
+This supplementary information contains 5 texts, 5 figures and 2 tables.
 
@@
 
-$\epsilon = \epsilon_S$
+*Log-law-of-the-wall model for the turbulent dissipation rate*
 
@@
 
-$\epsilon = \epsilon_D$
+*Form-drag model for the turbulent dissipation rate*
 
@@
 
-$\epsilon = \epsilon_S$
+*Log-law-of-the-wall model for the turbulent dissipation rate*
 
-$$k_{600}=\beta_0(\epsilon_S)^{1/4}A_p^{3/4} \mathbf{(S8)}$$
+$$k_{600}=\beta(\epsilon_S)^{1/4}A_p^{3/4} \mathbf{(S8)}$$
 $$k_{600}=\beta (\frac{U_*^3}{H})^{1/4}((U_*H)^{1/2})^{3/4} \mathbf{(S9)}$$
@@
 
-$\epsilon = \epsilon_D$
+*Form-drag model for the turbulent dissipation rate*
 
-$$k_{600}=\beta_0(\epsilon_D)^{1/4}A_p^{3/4} \mathbf{(S14)}$$
+$$k_{600}=\beta_1(\epsilon_D)^{1/4}A_p^{3/4} \mathbf{(S14)}$$
 $$k_{600}=\beta_1(gS\bar{U})^{1/4}((U_*H)^{1/2})^{3/4} \mathbf{(S15)}$$
@@
 They developed a set of river channel prior hyperparameters for McFLI algorithms that are entirely RS-able and reflect differential channel hydraulics as a function of river geomorphology.
-They used an extensive database of field measurements and machine learning to identify patterns that associate river width with the hydraulic priors needed to run McFLIs so that prior hyperparameters are assigned to rivers using only the remotely sensed measurements.
+They used an extensive database of field measurements and statistical learning to identify patterns that associate river width with the hydraulic priors needed to run McFLIs so that prior hyperparameters are assigned to rivers using only the remotely sensed measurements.
 For this study, we extracted $A_0_{\lambda}$ and $A_0_{\gamma}$ as the 5th and 95th percentile values rather than the absolute maximum and minimum values to avoid physically impossible bounds on $A_0$.
@@
 This leaves the $k_{600_t}$ hyperparameters to be defined.
-$k_{600_{\mu}}$ is set by invoking the hydraulic geometry (HG) relationships developed in section 2.2.1 of the main text using the data from @brinkerhoffReconcilingAtaStationAtManyStations2019.
+$k_{600_{\mu}}$ is set by invoking the hydraulic geometry (HG) relationships developed in section 3.2.3 of the main text and Table S2 using the data from @brinkerhoffReconcilingAtaStationAtManyStations2019.
 We replaced both depth and velocity terms from our gas exchange model (equation 7 in the Main Text) with these HG models, resulting in equation S19 where Q is the mass-conserved streamflow for the river reach.
 
-$$k_{600_{\mu}} = 62.82(gS_{i,t})^{7/16}(0.265Q_{t}^{0.395})^{9/16}*(0.284Q_{t}^{0.191})^{1/4} \mathbf{(S19)}$$
+$$k_{600_{\mu}} = 62.82(gS_{i,t})^{7/16}(0.252Q^{0.388})^{9/16}*(0.276Q^{0.164})^{1/4} \mathbf{(S19)}$$
 
@@
 So, we use the Q prior that will be globally available when SWOT launches: a mean annual estimate from a water balance model.
-This is provided with the simulated SWOT data by both @durandIntercomparisonRemoteSensing2016 and @frassonExploringFactorsControlling2021.
+This is the standard prior information used to benchmark the SWOT discharge algroithms and is provided with the simulated SWOT data by both @durandIntercomparisonRemoteSensing2016 and @frassonExploringFactorsControlling2021.
 Using a temporally-invariant estimate of streamflow is the worst case scenario and BIKER's performance will improve with a more informed prior on streamflow (and therefore $k_{600}$).
@@
 $k_{600_{\sigma}}$ is set to 0.30 (log-space).
-This corresponds to a coefficient of variation of approximately 0.30, which we took to reflect a reasonably strong agreement between the prior and the observed values.
+This corresponds to a coefficient of variation of approximately 30%, which we took to reflect a reasonably strong agreement between the prior and the observed values.
 $k_{600_{\lambda}}$ was set to log(0.001) m/day.
@@
 
-Finally, recall that $\sigma_{k_{600}}$ refers to the total uncertainty inherent in equation 9, i.e.
+## Text S4: Determining complete BIKER model uncertainty
+Recall that $\sigma_{k_{600}}$ refers to the total uncertainty inherent in equation 9, i.e.
 stemming from the Reynolds-extension model (equation 7) and Manning's equation for $\bar{U}$.
@@
 This means we must also account for uncertainty from the $\beta_1$ parameter.
-Assuming perfect, no-error measurements are made by the SWOT satellite, the $k_{600}$ relative uncertainty is expressed for some set of hydraulic observations as equation S20.
-We use the 166 hydraulically-wide measurements in our field-measured $k_{600}$ dataset and run 166 different Monte Carlo simulations (each 10,000 runs) to obtain 166 different $\sigma_k_{600}$ terms.
-We then take the average of those to be a reasonable estimate of $\sigma_k_{600}$.
-This ultimately provided a value of 1.1 for the log-transformed $\sigma_k_{600}$ that should be used once SWOT launches.
+Assuming perfect, no-error measurements are made by the SWOT satellite, the full equation 7 uncertainty is expressed for some set of hydraulic observations as equation S20.
+To quantify total uncertainty in $k_{600}$, we push Monte Carlo simulations through equation S20 and take the mean uncertainty as a reasonable value for $\sigma_{k_{600}}$ once SWOT launches.
+
+More specifically, we use the 166 hydraulically-wide measurements in our field-measured $k_{600}$ dataset (section 2 in the Main Text) and push 166 different Monte Carlo simulations through equation S20.
+Note that each Monte Carlo simulation is itself 10,000 runs in order to obtain a distribution of $k_{600}$ estimates.
+For each of the 166 distributions of estimated $l_{600}$, we extract the standard deviation as the uncertainty.
+We then take the average of those standard deviations to be a reasonable estimate of $\sigma_k_{600}$.
+This ultimately provided a value of 1.1 for the log-transformed $\sigma_k_{600}$ and should be used when running BIKER on real SWOT data.
 
@@
 
-## Text S4: Literature $CO_2$ flux models
-Here, we describe in more detail the $CO_2$ flux models used to compare against BIKER in section 3.2.2. The specific equations used are outlined in Table S3.
+## Text S5: Literature $CO_2$ flux models
+Here, we describe in more detail the $CO_2$ flux models used to compare against BIKER in section 3.2.3. The specific equations used are outlined in Table S2.
 
@@
 The $k_{CO_2}$ estimates are calculated using 1) equation 7, 2) in situ streamflow data, and 3) hydraulic scaling equations from the literature that predict river velocity and depth from streamflow.
-Specifically, these literature equations include the 'Raymond 2012', 'Raymond 2013', and 'Brinkerhoff 2019' models (Table S3).
+Specifically, these literature equations include the 'Raymond 2012', 'Raymond 2013', and 'Brinkerhoff 2019' models (Table S2).
 
 For example, the 'Raymond 2012' model uses the hydraulic scaling equations from @raymondScalingGasTransfer2012 to predict velocity and depth, which are in turn used to calculate a $k_{CO_2}$, which finally is used to calculate $FCO_2$.
-Similarly, the 'Raymond 2013' model uses the equations outlined in @raymondGlobalCarbonDioxide2013 and the 'Brinkerhoff 2019' model uses new equations fit to the hydraulics dataset from @brinkerhoffReconcilingAtaStationAtManyStations2019 following Text S1.
+Similarly, the 'Raymond 2013' model uses the equations outlined in @raymondGlobalCarbonDioxide2013 and the 'Brinkerhoff 2019' model uses new equations fit to the hydraulics dataset from @brinkerhoffReconcilingAtaStationAtManyStations2019 after the filtering descirbed in Text S1.
 We chose to include the 'Brinkerhoff 2019' model as the training dataset is far larger than those used in either of the previous two models (Table S3: 104,624 versus 10,837 versus 1,026, respectively).
@@
 ## Figure S2
-![Figure S2 Timeseries of the biweekly $CO_2$ data from @beaulieuControlsGasTransfer2012.
-Sampling took place 2008-2009 in the Ohio River (upstream of Cincinnati, Ohio, United States).
-Dashed black line denotes atmospheric $CO_2$ at 400 uatm.](cache/FCO2/figS2.jpg)
+![Figure S2: Map of simulated SWOT river locations around the world.
+Also includes approximate location of the $CO_2$ timeseries used in this study (Figure S3).
+Note that river locations are approximate as some hydraulic models are not geo-referenced.](cache/maps/figS2.jpg)
 
 ## Figure S3
-![Figure s3: Observed (orange) versus BIKER-inferred (green) timerseries of daily $k_{600}$ for the 47 SWOT-simulated rivers with no additional SWOT errors.
-Note that the y axis is normalized by maximum observed values to compare across rivers.](cache/validation/figS3.jpg)
+![Figure S3 Timeseries of the biweekly $CO_2$ data from @beaulieuControlsGasTransfer2012.
+Sampling took place 2008-2009 in the Ohio River (upstream of Cincinnati, Ohio, United States).
+Dashed black line denotes atmospheric $CO_2$ at 400 uatm.](cache/FCO2/figS3.jpg)
 
 ## Figure S4
-![Figure s3: Observed (orange) versus BIKER-inferred (green) timerseries of daily $k_{600}$ for the 16 SWOT-simulated rivers with the Frasson et al.
-(2021) error model.
+![Figure S4: Observed (orange) versus BIKER-inferred (green) timerseries of daily $k_{600}$ for the 47 SWOT-simulated rivers under the 'no-measurement-error' scenario.
 Note that the y axis is normalized by maximum observed values to compare across rivers.](cache/validation/figS4.jpg)
+
+## Figure S5
+![Figure S5: Observed (orange) versus BIKER-inferred (green) timerseries of daily $k_{600}$ for 16 rivers under the 'measurement-error' scenario.
+Note that the y axis is normalized by maximum observed values to compare across rivers.](cache/validation/figS5.jpg)
 
@@
 *Table S1: Studies that gas exchange velocity measurements come from.
-'Study' refers to the paper from which these measurements were obtained.
+'Study' refers to the paper from which these measurements were obtained by us.
 Any data wrangling was done by those authors.
@@
 ## Table S2
-*Table S2: Validation metrics used in this study, where Nt is number of observations and i is the specific observation.
-σ refers to the variance of the sample and μ refers to the mean of the sample.
-Carrot accents indicate the predicted value.*
-
-|**Description**|**Acronym**|**Definition**|**Ideal Score**|**Possible Range**|
-|---------------|-----------|--------------|---------------|------------------|
-|Correlation Coefficient|$r$| $\sqrt{1-(\frac{\sum\limits_{i=1}^{N_t}{(k_{600, i}-\hat{k_{600_i}})^2}}{\sum{(k_{600, i}-\bar{k_{600}})^2}})}$ |1|-1 to 1|
-|Normalized root-mean-square error|NRMSE| $\sqrt{\frac{1}{N_t}\sum\limits_{i=1}^{N_t}(\frac{\hat{k_{600, i}}-k_{600, i}}{\bar{k_{600, i}}})^2}$ |0|0 to ∞|
-|Kling-Gupta Efficiency|KGE|$1 - \sqrt{ (r(k_{600_i}, \hat{k_{600_i}}) - 1)^2 + (\frac{\mu_{\hat{k_{600_i}}} }{\mu_{k_{600_i}}} - 1)^2 + (\frac{ \sigma _{\hat{k_{600_i}}} / \mu _{\hat{k_{600_i}}} }{ \sigma_{k_{600_i}} / \mu_{k_{600_i}}} - 1)^2 }$|1|-∞ to 1|
-|Mean Absolute Error|MAE| $\frac{\sum\limits_{i=1}^{N_t}(|\hat{k_{600_i}}-k_{600_i})|)}{n})$ |0|0 to ∞|
-
-## Table S3
-*Table S3: Details on the 3 literature models for $CO_2$ fluxes. See Text S4 for more details.*
+*Table S2: Details on the 3 literature models for $CO_2$ fluxes. See Text S5 for more details.*
 
@@
 |--------|------------------|---------------------|---------------|-------------|
-|Brinkerhoff 2019|$D=0.265Q^{0.395}$|$\bar{U}=0.284Q^{0.191}$|104,624 measurements made across the United States at streamgauges|@brinkerhoffReconcilingAtaStationAtManyStations2019|
-|Raymond 2012|$D=0.409Q^{0.294}$|$\bar{U}=0.194Q^{0.285}$||1,026 measurements across the United States|@raymondScalingGasTransfer2012|
+|Brinkerhoff 2019|$D=0.252Q^{0.388}$|$\bar{U}=0.276Q^{0.164}$|104,624 measurements made across the United States at streamgauges|@brinkerhoffReconcilingAtaStationAtManyStations2019|
+|Raymond 2012|$D=0.409Q^{0.294}$|$\bar{U}=0.194Q^{0.285}$|@raymondScalingGasTransfer2012|1,026 measurements across the United States|@raymondScalingGasTransfer2012|
 |Raymond 2013|$D=0.409Q^{0.294},  D=0.449Q^{0.37}$|$\bar{U}=0.194Q^{0.285}, \bar{U}=0.346Q^{0.120}$|Average of the Raymond 2012 equation and one using 9,811 measurements at US streamgauges|@raymondGlobalCarbonDioxide2013|

### 2022-04-05 manuscript_v11.Rmd
final ms edits

--- 2022-03-09 manuscript_v10.Rmd
@@
 
--   The BIKER (Bayesian Inference of the $k_{600}$ Exchange Rate) algorithm predicts gas exchange velocity and $CO_2$ fluxes solely from simulated SWOT data without calibration
+-   The BIKER algorithm predicts gas exchange velocity and $CO_2$ fluxes solely from simulated SWOT data without calibration
 -   BIKER is robust to SWOT measurement errors
@@
 Here, we present work inferring $k_{600}$ from synthetic future SWOT observations without in situ calibration.
-We exploit the unique geomorphology of SWOT-observable rivers to develop a physical model of gas exchange that is remotely sensible and explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2,2)*100`% of variation in $k_{600}$ We then couple this model with established inversion techniques to develop BIKER, the 'Bayesian Inference of the $k_{600}$ Exchange Rate' algorithm.
-We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers and 166 measurements of river gas exchange, yielding an algorithm that predicts the $k_{600}$ timeseries solely from SWOT observations with a by-river median Kling-Gupta Efficiency of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$KGE), 2)`.
-Similar to remote sensing algorithms for SWOT, BIKER is better at reproducing the temporal dynamics of gas exchange (median correlation coefficient of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$r), 2)`), than the absolute rates of exchange (median normalized RMSE of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$NRMSE), 2)*100`%).
+We exploit the unique geomorphology of SWOT-observable rivers to develop a physical model of gas exchange that is remotely sensible and explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2,2)*100`% of variation across 166 field measurement of $k_{600}$.
+We then couple this model with established inversion techniques to develop BIKER, the 'Bayesian Inference of the $k_{600}$ Exchange Rate' algorithm.
+We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers, yielding an algorithm that predicts the $k_{600}$ timeseries solely from SWOT observations with a by-river median Kling-Gupta Efficiency of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$KGE), 2)`.
+Similar to remote sensing algorithms for SWOT, BIKER is better at explaining the temporal variation of gas exchange (median correlation coefficient of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$r), 2)`), than reproducing the absolute rates of exchange (median normalized RMSE of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$NRMSE), 2)*100`%).
 Finally, BIKER is robust to measurement errors implicit in the SWOT data.
-Upon SWOT’s launch, we suggest that BIKER will be useful in mapping global-scale fluvial gas exchange and improving CO2 emissions estimates when coupled with river CO2 models.
+Upon SWOT’s launch, we suggest that BIKER will be useful in mapping global-scale fluvial gas exchange and improving $CO_2$ emissions estimates when coupled with river $CO_2$ models.
 
@@
 Natural systems play a critical role in the budgeting and accounting of the global carbon cycle under climate change.
-Following @colePlumbingGlobalCarbon2007, the global river network is recognized to emit substantial amounts of carbon to the atmosphere via evasion (gas exchange driven by a diffusion gradient and near-surface turbulence), in addition to their long understood role in transporting carbon to the oceans via downstream advection.
+Following @colePlumbingGlobalCarbon2007, the global river network is recognized to emit substantial amounts of carbon to the atmosphere via evasion (gas exchange driven by diffusion and near-surface turbulence at th air/water interface), in addition to their long understood role in transporting carbon to the oceans via downstream advection.
 Current estimates of total carbon dioxide evasion ($FCO_2$) to the atmosphere from the global river network vary from 650-2000 Tg C/yr [@liuImportanceHydrologyRouting2022; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013], with 167 Tg-C/yr coming from mountain streams alone [@horgbyUnexpectedLargeEvasion2019].
-Despite rivers' incredibly small percentage of the global land surface [0.67%- @liuImportanceHydrologyRouting2022], this C flux is on par with the total oceanic $CO_2$ uptake rate [@gruberOceanicSinkAnthropogenic2019; @horgbyUnexpectedLargeEvasion2019] and the global forest carbon uptake rate [@panLargePersistentCarbon2011].
+Despite rivers' small percentage of the global land surface [0.67%- @liuImportanceHydrologyRouting2022], this carbon flux is on par with the total oceanic $CO_2$ uptake rate [@gruberOceanicSinkAnthropogenic2019; @horgbyUnexpectedLargeEvasion2019] and the global forest carbon uptake rate [@panLargePersistentCarbon2011].
 
 River $CO_2$ evasion is increasingly better constrained and is clearly a critical component of the global carbon cycle.
-Equation 1 represents this riverine flux given $\delta CO_2$ (the gas concentration gradient between the water $CO_{2_{water}}$ and the air $CO_{2_{air}}$) and the gas exchange velocity *k*.
+Equation 1 represents this riverine flux given $\delta CO_2$ (the air-water flux between the water $CO_{2_{water}}$ and the air $CO_{2_{air}}$) and the gas exchange velocity *k*.
 Consult Appendix A for variable nomenclature used throughout this study.
 
-$$FCO_2=k* \delta CO_2 \mathbf{(1)}$$
+$$FCO_2=k \delta CO_2 \mathbf{(1)}$$
 
@@
 
-The structure of equation 1 necessitates that calculations of $FCO_2$ are highly sensitive to measurements/estimates of *k*.
-However, *k* can only be directly calculated via a known gas concentration gradient, eddy-covariance measurements, or tracer additions to the stream [@hallGasExchangeStreams2020].
-In trying to constrain the global fluvial $FCO_2$ flux across millions of rivers, this calculation is impossible, and necessitates the use of predictive models for *k* that are based on easily obtained river hydraulic properties.
+The structure of Equation 1 necessitates that calculations of $FCO_2$ are highly sensitive to measurements/estimates of *k*.
+However, *k* can only be directly calculated via a known gas flux, eddy-covariance measurements, or tracer additions to the stream [@hallGasExchangeStreams2020].
+In trying to constrain the global fluvial $FCO_2$ flux across millions of rivers, this calculation is impossible and necessitates the use of predictive models for *k* that are based on easily obtained river hydraulic properties.
 In that vein, there have been over 20 empirical models developed to predict *k* from river hydraulics, generally using some combination of mean velocity $\bar{U}$, shear velocity $U_*$, width $W$, depth $H$, and slope $S$ as predictors [@hallGasExchangeStreams2020; @wangPhysicallyBasedScaling2021].
@@
 
-Applying these *k* models across watersheds, regions, or continents is called 'upscaling.' This upscaling allows for estimating the impossible-to-measure *k* term in Equation 1 for any arbitrary number of rivers, but also changes the base parameters that ultimately control the final estimate of $FCO_2$.
+Applying these *k* models across watersheds, regions, or continents is called 'upscaling.' This upscaling allows for estimating the difficult-to-measure *k* term in Equation 1 for any arbitrary number of rivers, but also changes the base parameters that ultimately control the final estimate of $FCO_2$.
 That is, by making *k* a function of hydraulics, $FCO_2_{upscaled}$ is now a direct function of river hydraulics.
@@
 
-Global upscalng has been performed in the literature.
+Global upscaling has been performed in the literature.
 @raymondGlobalCarbonDioxide2013, @lauerwaldSpatialPatternsCO22015, and @horgbyUnexpectedLargeEvasion2019 all relied on $k_{CO_2}$ values indirectly estimated using mean annual streamflow models and hydraulic scaling equations to predict the hydraulic terms to in turn predict $k_{CO_2}$, while @borgesGloballySignificantGreenhousegas2015 used a combination of the above method and a constant $k_{CO_2}$ in space and time to upscale over Africa.
@@
 However, they were still limited by a lack of direct hydraulic measurements and had to rely on hydraulic scaling equations to estimate river depth and velocity.
-Even though approaches like @wangPhysicallyBasedScaling2021's are incredibly useful for expanding our mechanistic understanding of gas exchange, they are less useful for global upscaling purposes as they rely on highly detailed in situ DO data for every river [@hallGasExchangeStreams2020].
+Even though approaches like @wangPhysicallyBasedScaling2021's are incredibly useful for expanding our mechanistic understanding of gas exchange, they are less useful for global upscaling purposes as they rely on high fidelity in situ DO data for every river [@hallGasExchangeStreams2020].
 
 We have established that literature has a reasonably good understanding of $\delta CO_2$ and a relatively poorer understanding of $k_{600}$ (and therefore $k_{CO_2}$) across large areas and in time.
-In theory, the discrepancy between the quality of our $\delta CO_2_{water}$ and $k_{CO_2}$ estimates could be alleviated if direct hydraulics measurements (and in turn $k_{CO_2}$ via equation 2) were available at the global scale at a sufficient temporal resolution.
-This is turn would also address the uncertainty regarding continental-to-global scale temporal dynamics of gas exchange noted earlier.
+In theory, the discrepancy between the quality of our $\delta CO_2_{water}$ and $k_{CO_2}$ estimates could be alleviated if direct hydraulics measurements (and in turn $k_{CO_2}$ via Equation 2) were available at the global scale at a sufficient temporal resolution.
+Spatially and temporally dynamic hydraulic measurements in turn would also address the uncertainty regarding continental-to-global scale temporal dynamics of gas exchange noted earlier.
+
 Conveniently, these hydraulic data will soon be available via the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission.
-SWOT is expected to launch in 2022 and provide the world's first direct measurements of global water surface extent and elevation (and therefore water surface slope) at novel temporal resolutions.
+SWOT is expected to launch in late 2022 and provide the world's first direct measurements of global water surface extent and elevation (and therefore water surface slope) at novel temporal resolutions.
 SWOT is a wide swath radar interferometer and will sample rivers every 1 to 7 days per 21 day repeat cycle, measuring rivers wider than 100m with a goal of expanding this to rivers at least 50m wide [@biancamariaSWOTMissionIts2016].
@@
 
-This paper is split into two distinct parts: gas-exchange theory/model development (section 2) and BIKER algorithm development/validation (section 3).
-Section 3 is fundamentally dependent on the results presented in section 2, therefore section 2 presents both theory and results.
+This paper is split into two distinct parts: gas-exchange theory/model development (Section 2) and BIKER algorithm development/validation (Section 3).
+Section 3 is fundamentally dependent on the results presented in Section 2, therefore Section 2 presents both theory and results.
 Both sections detail the data used.
-We conclude with a discussion (section 4) that encompasses both gas exchange theory and remote sensing.
+We conclude with a discussion (Section 4) that encompasses both gas exchange theory and remote sensing.
 Figure 1 conceptually maps out the algorithm's approach to inferring $k_{600}$ from SWOT data.
@@
 1: SWOT will directly measure water surface width and elevation (and thus slope) at a series of cross-sections within mass-conserved river reaches.
-2: These hydraulics measurements are used to calculate turbulent dissipation in the river channel (section 3).
-3: Turbulent dissipation is used to infer $k_{600}$ via a process-based model (section 2).](cache/maps/diagram.tif)
+2: These hydraulics measurements are used to calculate turbulent dissipation in the river channel (Section 3).
+3: Turbulent dissipation is used to infer $k_{600}$ via a process-based model (Section 2).](cache/maps/diagram.tif)
 
@@
 To predict $k_{600}$ in the large rivers that SWOT will observe, we start from an established process-based model for $k_{600}$, impose hydraulic assumptions that are valid in SWOT-observable rivers, and obtain a model we empirically test.
-Following a description of the data (section 2.1), we outline established models (section 2.2) and then step through our hydraulic assumptions to arrive at an equation that is compatible with SWOT measurements (section 2.3).
-We then empirically validate the model (section 2.3)
+Following a description of the data (Section 2.1), we outline established models (Section 2.2) and then step through our hydraulic assumptions to arrive at an equation that is compatible with SWOT measurements (Section 2.3).
+We then empirically validate the model (Section 2.4)
 
@@
 In addition to hydraulics measured alongside $k_{600}$ and reported above, we expand our dataset of stream hydraulics using a previously published compilation of in situ hydraulic measurements [@brinkerhoffReconcilingAtaStationAtManyStations2019].
-That dataset contains over 530,000 unique measurements of river channel velocity, width, and discharge from across the continental United States, originally made to calibrate United States Geological Survey (USGS) streamgauge rating curves and made public by the USGS.
-This dataset is used to calculate how frequently SWOT observable rivers meet our large-river hydraulic assumptions (section 2.2).
+That dataset contains over 530,000 unique measurements of river channel velocity, width, and discharge from across the continental United States, originally made to calibrate United States Geological Survey (USGS) stream gauge rating curves and made public by the USGS.
+This dataset is used to calculate how frequently SWOT observable rivers meet our large-river hydraulic assumptions (Section 2.3).
 
@@
 @katulStructureFunctionModel2018; @lorkeUnifiedScalingRelation2006a; @zappaEnvironmentalTurbulentMixing2007; @zappaVariationSurfaceTurbulence2003; @moogStreamReaerationNonuniform1999; @vachonRelationshipNearsurfaceTurbulence2010; @tokoroHighGastransferVelocity2008; @wangPhysicallyBasedScaling2021].
-The small-eddy model is provided as equation 3, where $\epsilon$ is the dissipation rate of near-surface turbulence, $v$ is the kinematic viscosity, and *Sc* is the Schmidt number.
+The small-eddy model is provided as Equation 3, where $\epsilon$ is the dissipation rate of near-surface turbulence, $v$ is the kinematic viscosity, and *Sc* is the Schmidt number.
 
-$$k\propto Sc^{-1/2}(v\epsilon)^{1/4} \mathbf{(3)}$$ Some laboratory and field observations additionally suggest that open channel flows with small bed roughness do not exhibit homogeneous surface dissipation at their air-water interface [@moogAirWaterGasTransfer1999; @talkeTurbulentKineticEnergy2013].
-Given this observation, @moogAirWaterGasTransfer1999 proposed an extension to the small-eddy model, additionally scaling $k_{600}$ using a shear Reynold's number formulation.
-This is equation 4 and is referred to here as the 'Reynolds extension' model.
+$$k\propto Sc^{-1/2}(v\epsilon)^{1/4} \mathbf{(3)}$$ Some laboratory and field observations additionally suggest that open channel flows with small bed roughness do not exhibit homogeneous surface dissipation across the entire reach's air-water interface [@moogAirWaterGasTransfer1999; @talkeTurbulentKineticEnergy2013].
+Given this observation, @moogAirWaterGasTransfer1999 proposed an extension to the small-eddy model, additionally scaling $k_{600}$ using a shear Reynold's number $Re_*$ formulation.
+This is Equation 4 and is referred to here as the 'Reynolds extension' model.
 The Reynolds model is hypothetically useful in low-turbulence scenarios where a relative lack of large-scale eddies effectively 'filter out' the number of small-eddies that actually reach the interface and initiate gas exchange [@talkeTurbulentKineticEnergy2013].
@@
 
-$$k \propto Sc^{-1/2}(v\epsilon)^{1/4}(\frac{U_*H}{v})^{3/8} \mathbf{(4)}$$ Equations 3 and 4 both rely on $\epsilon$, which is non-trivial to measure.
-When working at large scales, a commonly used $\epsilon$ model assumes that all turbulence is generated at the bed and transported to the air-water interface via the log-law-of-the-wall (equation 5).
-Another approach specific to fluvial settings models *k* via 'form-drag dissipation' (equation 6) which is equivalently the total stream power per unit mass water.
+$$k \propto Sc^{-1/2}(v\epsilon)^{1/4}(Re_*)^{3/8} \mathbf{(4)}$$
+
+Equations 3 and 4 both rely on $\epsilon$, which is non-trivial to measure.
+When working at large scales, a commonly used $\epsilon$ model assumes that all turbulence is generated at the bed and transported to the air-water interface via the log-law-of-the-wall (Equation 5).
+Another approach specific to fluvial settings models *k* via 'form-drag dissipation' (Equation 6) which is equivalently the total stream power per unit mass water.
 This normalized stream power captures the bulk frictional resistance (and thus energy dissipation) via channel banks, meanders, bars, etc.
 that is unique to fluvial systems [@moogStreamReaerationNonuniform1999].
-Authors have since shown that equation 6 can reasonably predict *k* in rivers and streams [@raymondScalingGasTransfer2012; @ulsethDistinctAirWater2019; @wangPhysicallyBasedScaling2021].
+Authors have since shown that Equation 6 can reasonably predict *k* in rivers and streams [@raymondScalingGasTransfer2012; @ulsethDistinctAirWater2019; @wangPhysicallyBasedScaling2021].
 
@@
 Given the theoretical context provided in Section 2.2, we now turn to SWOT-observable systems specifically.
-Rivers and streams change predictably along their longitudinal profile from source to sea, and we can exploit the hydraulic geometry of large rivers at the end of this continumn to estimate *k* in SWOT-observable systems.
+Rivers and streams change predictably along their longitudinal profile from source to sea, and we can exploit the hydraulic geometry of large rivers at the end of this continumn to estimate *k* in SWOT rivers.
 In general, as river size increases, channels become more rectangular, their shapes elongate (becoming wider quicker than they become deeper) and their hydraulic radii begin to approximate their mean flow depth, i.e.
@@
 The overall goal is to reduce the equations down to their fundamental parameters, identifying which terms are SWOT observable and limiting the number of terms not directly measurable via SWOT.
-To do this, we impose $\epsilon=\epsilon_D$ and $R_H=H$ on the Reynolds extension model and arrive at equation 7 (with statistical coefficient $\beta_1$).
+To do this, we impose $\epsilon=\epsilon_D$ and $R_H=H$ on the Reynolds extension model (Equation 4) and arrive at Equation 7 (with statistical coefficient $\beta_1$).
 Equation 7 thus defines gas exchange velocity solely as a function of slope, mean flow depth, and mean flow velocity.
@@
 
-We also test the performance of three other models for predicting *k* in hydraulically-wide channels via the other three unique combinations of equations 3-4 and equations 5-6.
-While the complete model derivations and results for all four models are provided in Text S2 and Figure S1, the final and best-performing model (equation 7) is presented and used below.
+We also test the performance of three other models for predicting *k* in hydraulically-wide channels via the other three unique combinations of Equations 3-4 and Equations 5-6.
+While the complete model derivations and results for all four models are provided in Text S2 and Figure S1, the final and best-performing model (Equation 7) is presented and used below.
 
@@
 
-With equation 7 derived, we now empirically test its strength of fit in hydraulically-wide river flows.
+With Equation 7 derived, we now empirically test its strength of fit in hydraulically-wide river flows.
 We validate on the dataset of in situ measurements of $k_{600}$, after filtering for measurements made in hydraulically-wide channels, which was defined as flows whose hydraulic radius was within 1% of their mean flow depth.
-All told, this amounts to `r n` direct measurements of hydraulically-wide $k_{600}$.
+All told, this amounts to `r n` measurements of hydraulically-wide $k_{600}$.
 Equation 7 is assessed via the coefficient of determination ($r^2$) and plotted in Figure 2.
@@
 
-Figure 2 shows that equation 7 explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2, 2) * 100`% of variation in observed $k_{600}$ in hydraulically-wide rivers and accurately captures the scaling dynamics of observed $k_{600}$ as well.
+Figure 2 shows that Equation 7 explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2, 2) * 100`% of variation in observed $k_{600}$ in hydraulically-wide rivers and accurately captures the scaling dynamics of observed $k_{600}$ as well.
+The value for $\beta_1$ is 62.8, which is similar the fitted parameters in other $Re_*$ and $U_*$ models for river *k* [@wangPhysicallyBasedScaling2021].
 While the individual residuals can be quite large, the overall scaling of $k_{600}$ with river hydraulics is strongly captured, remembering that Figure 2 is plotted in logarithmic space.
-Compared to other combinations of equations 3-6 (Figure S1), there is also less bias in the estimates thanks to the addition of the Reynold's number scaling for low-turbulent flows (as expected).
+Compared to other combinations of Equations 3-6 (Figure S1), there is also a better fit with the estimates thanks to the addition of the Reynold's number scaling for low-turbulent flows (as expected).
 The success of this model in hydraulically-wide channels provides us with a strong physical-model for gas evasion built with SWOT in mind.
-The river hydraulics terms in equation 7 ($\bar{U}$, $H$, and $S$) can either be directly measured or reasonably inferred from SWOT measurements, effectively opening the door for remotely sensing the gas exchange velocity.
-This is explored next.
+The river hydraulics terms in Equation 7 ($\bar{U}$, $H$, and $S$) can either be directly measured or reasonably inferred from SWOT measurements, effectively opening the door for remotely sensing the gas exchange velocity.
 
-## 3 Exploiting equation 7 to remotely sense gas exchange velocity
+## 3 Exploiting Equation 7 to remotely sense gas exchange velocity
 
-We have shown that scaling $k_{600}$ via equation 7 is useful in hydraulically-wide rivers.
-Further, equation 7 has only three non-directly-remotely-sensible terms: $k_{600}$, mean flow depth, and mean flow velocity.
+We have shown that scaling $k_{600}$ via Equation 7 is useful in hydraulically-wide rivers.
+Further, Equation 7 has only three non-directly-remotely-sensible terms: $k_{600}$, mean flow depth, and mean flow velocity.
 Conveniently, techniques to simultaneously infer mean flow depth and velocity from SWOT data (among other parameters) have been established over the last decade to infer streamflow from SWOT's measurements [e.g.
@@
 For BIKER, we follow the work developed by @hagemannBAMBayesianAMHGManning2017, @brinkerhoffConstrainingRemoteRiver2020, and @durandEstimatingReachaveragedDischarge2014 to infer $k_{600}$, channel depth and velocity from SWOT observations using a modified form of Manning's equation.
-Following a description of the data used (section 3.1), we detail algorithm development and experimental design (section 3.2) and then we present the validation results (section 3.3).
+Following a description of the data used (Section 3.1), we detail algorithm development and experimental design (Section 3.2) and then we present the validation results (Section 3.3).
 
@@
 These data were published by @frassonExploringFactorsControlling2021 and @durandIntercomparisonRemoteSensing2016.
-See Figure S2 for a map of these river's locations.
+See Figure S2 for a map of these river's locations along the global SWOT river network [@altenauSurfaceWaterOcean2021].
+This river network corresponds to every river that BIKER will be run on once SWOT launches.
 
@@
 Note that this $CO_2$ data is for the Ohio River only but was applied to all rivers to provide a physically realistic signal for $CO_2$ fluxes with meaningful seasonality and dynamics.
-Therefore, the raw carbon emissions estimates presented in this paper are completely meaningless in the context of actually measured carbon emissions from these rivers, but are better than specifying $CO_2$ concentrations devoid of context.
+Therefore, the raw carbon emissions estimates presented in this paper are meaningless in the context of actually measured carbon emissions from these rivers, but are better than specifying $CO_2$ concentrations devoid of context.
 These data are necessary as we are interested in the effect of BIKER $k_{600}$ errors on eventual fluxes and comparing these fluxes with published methods.
@@
 
-BIKER, and Bayesian inference in general, starts from Bayes rule (equation 8), where $\Theta$ is some set of non-remotely-sensible parameters we want to solve for (including $k_{600}$), *x* is the observed data, $f(x|\Theta)$ is the sampling model where data are conditional on the parameters, and $p(\Theta)$ is the joint prior distribution of the parameters.
+BIKER, and Bayesian inference in general, starts from Bayes rule (Equation 8), where $\Theta$ is some set of non-remotely-sensible parameters we want to solve for (including $k_{600}$), *x* is the observed data, $f(x|\Theta)$ is the sampling model where data are conditional on the parameters, and $p(\Theta)$ is the joint prior distribution of the parameters.
 Therefore, we are interested in solving for $p(\Theta|x)$, or the 'posterior' distribution.
-Note that $p(x)$ is usually computationally intractable to integrate exactly, but Bayesian inference tools require only the proportionality to be specified: $p(\Theta|x) \propto f(x|\Theta)p(x)$.
-Sampling algorithms are then used to approximate the posterior distribution, as is done in BIKER.
+Note that $p(x)$ is usually computationally intractable to integrate exactly, but Bayesian inference requires only the proportionality to be specified: $p(\Theta|x) \propto f(x|\Theta)p(x)$.
+Sampling algorithms are then used to approximate the actual posterior distribution, as is done in BIKER.
 
-$$p(\Theta|x) = \frac{f(x|\Theta)p(\Theta)}{p(x)} \mathbf{(8)}$$ The heart of BIKER is its reformulation of the $k_{600}$ model (equation 7) as a Bayesian sampling model that is conditional on the non-remotely-sensible parameters (i.e.
+$$p(\Theta|x) = \frac{f(x|\Theta)p(\Theta)}{p(x)} \mathbf{(8)}$$ The heart of BIKER is its reformulation of the $k_{600}$ model (Equation 7) as a Bayesian sampling model that is conditional on the non-remotely-sensible parameters (i.e.
 $f(x|\Theta)$).
@@
 To start, we write $k_{600}$ as a function of SWOT-observables $W$ and $S$.
-This algebra is carried out using equation 7, the fitted value for $\beta_1$ from Figure 2 (`r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$coef,2)`), and Manning's equation for mean flow velocity ($\bar{U}=\frac{1}{n}R_h^{2/3}S^{1/2}$).
-Following section 2.3, we continue to assume that the channel is hydraulically-wide ($R_h = H =\frac{A}{W}$).
+This algebra is carried out using Equation 7, the fitted value for $\beta_1$ from Figure 2 (`r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$coef,2)`), and Manning's equation for mean flow velocity ($\bar{U}=\frac{1}{n}R_h^{2/3}S^{1/2}$).
+Following Section 2.3, we continue to assume that the channel is hydraulically-wide ($R_h = H =\frac{A}{W}$).
 To leverage additional SWOT data, the wetted channel area *A* is further split into the the SWOT-unobservable portion $A_0$ and SWOT-observable portion $dA$ following @durandEstimatingReachaveragedDischarge2014 and @hagemannBAMBayesianAMHGManning2017 where $dA_{it} = \sum_{t':W_{t'}\le W_t}{W_{t'} \delta H_e_{t'}}$ for cross-section *i* and timestep *t* within a mass-conserved river reach.
 
-All of this algebra simplifies to equation 9.
+All of this algebra simplifies to Equation 9.
 Conveniently, $k_{600}$ as measured by tracer additions to a stream is inherently a reach-scale quantity (in a mass-conserved reach).
-Therefore, equations 7 and 9 both yield a reach-scale $k_{600}$ (i.e.
+Therefore, Equations 7 and 9 both yield a reach-scale $k_{600}$ (i.e.
 $k_{600_i}=k_{600}\forall i$).
@@
 
-$$\begin{eqnarray} k_{600_t} = 62.82 g^{7/16}S_{it}^{9/16}\frac{1}{n_i}^{1/4}(\frac{A_0_i+dA_{it}}{W_{it}})^{35/48} \end{eqnarray} \mathbf{(9)}$$ Next, equation 9 is written as a Bayesian sampling model, in which all of the SWOT observations are sampled from the unknown model parameters ($A_0_i$, $k_{600_t}$, and $n_i$).
-This is equation 10 after describing everything as normal distributions of the log-transformed terms.
-$\sigma_{k_{600}}^2$ refers to the total uncertainty implicit in equation 9.
-This uncertainty arises from 1) parameter uncertainty in equation 7, 2) Manning's equation, and 3) the rectangular channel assumption.
+$$\begin{eqnarray} k_{600_t} = 62.82 g^{7/16}S_{it}^{9/16}\frac{1}{n_i}^{1/4}(\frac{A_0_i+dA_{it}}{W_{it}})^{35/48} \end{eqnarray} \mathbf{(9)}$$ Next, Equation 9 is written as a Bayesian sampling model, in which all of the SWOT observations are sampled from the unknown model parameters ($A_0_i$, $k_{600_t}$, and $n_i$).
+This is Equation 10 after describing everything as normal distributions of the log-transformed terms.
+$\sigma_{k_{600}}^2$ refers to the total uncertainty implicit in Equation 9.
+This uncertainty arises from 1) parameter uncertainty in Equation 7, 2) Manning's equation, and 3) the rectangular channel assumption.
 
@@
 
-With the sampling model described ($f(x|\Theta)$ = equation 10) and priors $p(x$) specified (Text S3), a joint posterior distribution conditional on the SWOT observations ($p(\Theta|x) \propto f(x|\Theta)p(x)$) is therefore also specified.
+With the sampling model described ($f(x|\Theta)$ = Equation 10) and priors $p(x$) specified (Text S3), a joint posterior distribution conditional on the SWOT observations ($p(\Theta|x) \propto f(x|\Theta)p(x)$) is therefore also specified.
 To approximate this distribution, we use a Markov Chain Monte Carlo (MCMC) algorithm implemented using the Stan probabilistic programming language.
@@
 We also re-validate BIKER under the 'measurement-error' scenario using the `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` rivers with @frassonExploringFactorsControlling2021's SWOT error model to corrupt the hydraulics to mimic realistic SWOT measurements (widths, heights, and slopes).
-These data were outlined in section 3.1.
+These data were outlined in Section 3.1.
 
@@
 There are, to our knowledge, no datasets of $k_{600}$ over time approaching the temporal density of our simulated SWOT rivers.
-We therefore apply equation 7 as validated in Figure 2 to specify $k_{600}$ given the true hydraulics of each case and compare BIKER's inversion to that value: given observed hydraulics, 'observed' $k_{600}$ comes from equation 7.
-Remember that SWOT cannot observe below the water surface and therefore cannot measure $\bar{U}$ or *H* (hence the need for equation 9), and that all SWOT observations contain errors in both space and time (hence equation 10).
-We acknowledge that there is error in equation 7 as shown in Figure 2, but this error can be explicitly parameterized in our Bayesian system (this is elaborated on in Text S4).
-Therefore, the BIKER validation presented here is an exercise to see how well the imperfect and partial SWOT observations can infer $k_{600}$ given the hydraulic assumptions in equation 9 and uncertainty in the data itself.
-Note also that we have already empirically validated equation 7 in Figure 2.
+We therefore apply Equation 7 as validated in Figure 2 to specify $k_{600}$ given the true hydraulics of each case and compare BIKER's inversion to that value: given observed hydraulics, 'observed' $k_{600}$ comes from Equation 7.
+Remember that SWOT cannot observe below the water surface and therefore cannot measure $\bar{U}$ or *H* (hence the need for Equation 9), and that all SWOT observations contain errors in both space and time (hence Equation 10).
+We acknowledge that there is error in Equation 7 as shown in Figure 2, but this error can be explicitly parameterized in our Bayesian system (this is elaborated on in Text S4).
+Therefore, the BIKER validation presented here is an exercise to see how well the imperfect and partial SWOT observations can infer $k_{600}$ given the hydraulic assumptions in Equation 9 and uncertainty in the data itself.
+Note also that we have already empirically validated Equation 7 in Figure 2.
 
@@
 It is one thing to accurately model the temporal dynamics of $k_{600}$ as above, but researchers are often most interested in the actual carbon emitted from river to atmosphere.
-Per equation 2, this is done using river hydraulic models to estimate $k_{CO_2}$ and in turn $FCO_{2_{upscaled}}$.
-However, streamflow data and/or model outputs are more readily modeled at the global scale than river channel geometry, and so upscaling models usually predict $\bar{U}$ and *H* as functions of streamflow (*Q*) using hydraulic geometry (HG) scaling relationships.
-This effectively reduces equation 2 to equation 11.
-It is worth stressing that these literature upscaling workflows rely on in situ streamflow records and/or high-quality streamflow routing models.
+Per Equation 2, this is done using river hydraulic models to estimate $k_{CO_2}$ and in turn $FCO_{2_{upscaled}}$.
+However, streamflow data and/or model outputs are more readily modeled at the global scale than river channel geometry, and so upscaling models usually predict $\bar{U}$ and *H* as functions of streamflow (*Q*) using hydraulic geometry scaling relationships.
+This effectively reduces Equation 2 to Equation 11.
+It is worth stressing that these literature upscaling workflows rely on in situ streamflow records and/or high-quality streamflow models.
 
@@
 
-Conversely, BIKER represents a new way of approaching equation 11 compared to existing literature models: BIKER has no reliance on a streamflow model nor HG scaling relationships and only requires that a river is SWOT-observable.
-We are therefore interested in how the final carbon emissions that result from BIKER compare against literature methods that use equation 11.
+Conversely, BIKER represents a new way of approaching Equation 11 compared to existing literature models: BIKER has no reliance on a streamflow model nor hydraulic geometry scaling relationships and only requires that a river is SWOT-observable.
+We are therefore interested in how the final carbon emissions that result from BIKER compare against literature methods that use Equation 11.
 We have the data to test four different models for fluxes: 'BIKER', 'Raymond 2013', 'Raymond 2012', and 'Brinkerhoff 2019'.
-These latter three approaches all use the same philosophy for $k_{600}$: making hydraulic and geomorphic assumptions to associate $k_{600}$ with observed hydraulics before using the $CO_2$ data as a realistic timeseries to yield fluxes per equation 11.
+These latter three approaches all use the same philosophy for $k_{600}$: making hydraulic and geomorphic assumptions to associate $k_{600}$ with observed hydraulics before using the $CO_2$ data as a realistic timeseries to yield fluxes per Equation 11.
 In all three approaches, these observed hydraulics are streamflow, while BIKER uses only SWOT observations.
@@
 
-To benchmark BIKER against these literature methods, we pair the 26 biweekly $CO_2$ and water temperature samples from @beaulieuControlsGasTransfer2012 (section 3.1, Figure S3) with every 14th set of daily SWOT observations (as the $CO_2$ data is bi-weekly).
-We then calculate $FCO_2$ using equation 1, an atmospheric $CO_{2_{air}}$ of 400 uatm, and a *Sc* estimated following Raymond et al.
+To benchmark BIKER against these literature methods, we pair the 26 biweekly $CO_2$ and water temperature samples from @beaulieuControlsGasTransfer2012 (Section 3.1, Figure S3) with every 14th set of daily SWOT observations (as the $CO_2$ data is bi-weekly).
+We then calculate $FCO_2$ using Equation 1, an atmospheric $CO_{2_{air}}$ of 400 uatm, and a *Sc* estimated following Raymond et al.
 (2012).
-The *k* in equation 1 is obtained using BIKER or the three literature models (Table S2 and Text S5).
+The *k* in Equation 1 is obtained using BIKER or the three literature models (Table S2 and Text S5).
 Finally, we estimate a pseudo yearly total carbon emission rate (via $CO_2$ evasion) by applying each river's mean $FCO_2$ over the river's surface area and summing all rates across rivers, remembering that we are applying 'unit' $CO_2$ data to all rivers.
@@
 (a-b) are poorest performing rivers, (c-d) are in the middle, and (e-f) are the best performing rivers.
-Y axis is normalized by maximum observed values to compare visually.](cache/validation/fig4.jpg)
+Y axis is normalized by maximum observed values to compare visually.](cache/validation/fig3.jpg)
 
-Next, we calculate performance metrics following section 3.2.2 and Table 1.
+Next, we calculate performance metrics following Section 3.2.2 and Table 1.
 These are presented in Figure 4, which plots the scores for the 47 rivers with no measurement errors as empirical cumulative density functions (eCDFs).
 Median river $KGE$ is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$KGE, na.rm=T),2)` and median river *r* is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$r, na.rm=T),2)`.
-This indicates very strong inference of each river's temporal $k_{600}$ dynamics given that absolutely no in situ information is being used to predict $k_{600}$.
+Further, `r sum(results_k600_rivs[results_k600_rivs$errFlag==0,]$KGE > -0.41)`/`r nrow(results_k600_rivs[results_k600_rivs$errFlag==0,])` rivers outperform a uniform prediction of the mean (KGE = -0.41).
+This indicates strong inference of each river's temporal $k_{600}$ dynamics given that absolutely no in situ information is being used to predict $k_{600}$.
 NRMSE has a median score of `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$NRMSE, na.rm=T),2)`, highlighting many rivers which have notable positive biases (Figure 3 also confirms this visually).
 Median NMAE is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$NMAE, na.rm=T),2)*100`%.
-Taken in aggregate, Figures 4-5 indicate that BIKER is quite good at capturing temporal dynamics in $k_600$, however there is often positive bias in its estimates.
+Taken in aggregate, Figures 3-4 indicate that BIKER is quite good at capturing temporal dynamics in $k_600$, however there is often positive bias in its estimates.
 
@@
 Each subpanel is labelled by its performance metric (defined in Table 1).
-Dashed lines denote median scores.](cache/validation/fig5.jpg)
+Dashed lines denote median scores.](cache/validation/fig4.jpg)
 
@@
 Note that some axes are flipped in order to visualize all 'better performances' in the upper-right-corner of each sub-plot.
-Black line denotes linear regression to aid in visualization.](cache/validation/fig6.jpg)
+Black line denotes linear regression to aid in visualization.](cache/validation/fig5.jpg)
 
@@
 a) all rivers.
-b) Only rivers with a coefficient of varaiton (CV) of observed $k_{600}$ \> 0.10.
-c) Only rivers with CV \> 0.20.
-d) Only rivers with CV \> 0.30.
+b) Only rivers with a coefficient of varaiton (CV) of observed $k_{600}$ > 0.10.
+c) Only rivers with CV > 0.20.
+d) Only rivers with CV > 0.30.
 Overall we see that rivers with greater temporal variation in $k_{600}$ behave better for BIKER, i.e.
-Bayesian inference reduces bias from prior to posterior.](cache/validation/fig7.jpg)
+Bayesian inference reduces bias from prior to posterior.](cache/validation/fig6.jpg)
 
@@
 Put another way, we suggest that if a SWOT timeseries does not sufficiently capture a river's temporal dynamics, it will introduce additional error to the inference results.
-To test this, we subset our validation dataset by progressively higher coefficients of variation (*CV*) (\> 10%, \>20%, and \>30%).
-These filtered datasets are plotted in Figures 6b-6d (for \> 10%, \>20%, and \>30% *CV*, respectively), where our hypothesis is confirmed.
-Posterior bias systematically drops as the temporal variability of the SWOT data increases, with BIKER posterior error the smallest, and much less than prior error, in Figure 6d.
-This result is elaborated on in section 4.2.
+To test this, we subset our validation dataset by progressively higher coefficients of variation (*CV*) (> 10%, >20%, and >30%).
+These filtered datasets are plotted in Figures 6b-6d (for > 10%, >20%, and >30% *CV*, respectively), where our hypothesis is confirmed.
+Posterior bias drops once the temporal variability of the SWOT data is sufficiently high, with BIKER posterior error the smallest, and much less than prior error, in Figure 6d.
+This result is elaborated on in Section 4.2.
 
@@
 
-Figure 7 compares the annual carbon emissions rate (via $F_{CO_2}$) from the rivers using BIKER posterior means and the three streamgauge-based HG models.
+Figure 7 compares the annual carbon emissions rate (via $F_{CO_2}$) from the rivers using BIKER posterior means and the three stream gauge-based HG models.
 Surprisingly, BIKER outperforms the gauge-based approaches, nearly correctly inferring the annual carbon emissions rate (`r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_BIKER',]$yearlyCFux, 2)` Tg-C/yr for BIKER versus `r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_obs',]$yearlyCFux, 2)` Tg-C/yr observed).
@@
 'Brinkerhoff 2019''s lower CI is comparable to BIKER's higher CI.
-BIKER's relatively stronger performance than the in situ models is elaborated on in section 4.3.
-Despite BIKER using absolutely no streamgauge or other in situ data like the other 3 methods do, it provides a strong upscaling estimate of the annual carbon emission rate for the rivers (Figure 7).
+BIKER's relatively stronger performance than the in situ models is elaborated on in Section 4.3.
+Finally, BIKER's uncertainity is on par with the in situ technique, despite being obtained solely from SWOT data.
+Taken in aggregate, BIKER provides a strong upscaling estimate of the annual carbon emission rate for the rivers and is either comparable or better than established in situ techniques (Figure 7).
 
@@
 Completely remotely-sensed methods are colored in red, in situ methods in purple, and the observed in orange.
-Note that confidence intervals were not calculated for the remaining two models because their uncertainties are not available.](cache/FCO2/fig8.jpg)
+Note that confidence intervals were not calculated for the remaining two models because their uncertainties are not available.](cache/FCO2/fig7.jpg)
 
@@
 This model is empirically validated in Figure 2.
-Specifically, Figure S1 confirms that scaling $k_{600}$ with a shear Reynold's adaption of the small-eddy model (equation 7) reduces positive bias in the model's predictions of the smallest $k_{600}$ values [where the relative decrease in turbulence reaching the surface is greater than the small-eddy model alone suggests, per equation 7's theoretical basis- @moogAirWaterGasTransfer1999].
+Specifically, Figure S1 confirms that scaling $k_{600}$ with a shear Reynold's adaption of the small-eddy model (Equation 7) improves the model's predictions of the smallest $k_{600}$ values [where the relative decrease in turbulence reaching the surface is greater than the small-eddy model alone suggests, per Equation 7's theoretical basis- @moogAirWaterGasTransfer1999].
 Scaling $k_{600}$ via a shear Reynold's number is often done to parameterize breaking-wave gas exchange models in the open ocean [@brumerWaveRelatedReynoldsNumber2017; @zhaoDependenceWhitecapCoverage2001; @zhaoEffectWindWaves2003], though this is specific to high wind speeds in open ocean.
-To our knowledge, @moogAirWaterGasTransfer1999's specific setup, which imposes a 'differentially-experienced' surface turbulence theory on the small-eddy model, has never been empirically validated in rivers.
+To our knowledge, @moogAirWaterGasTransfer1999's specific setup, which imposes a space-and-time varying, fractional area surface turbulence theory on the small-eddy model, has never been empirically validated in rivers.
 Figure 2 provides this empirical verification for hydraulically wide channels, where it's theoretical basis should generally hold.
@@
 
-Crucially, we are not accounting for wind-driven gas exchange, which is suggested to play an important role in wide rivers because river surface area is sufficiently large that wind can drive gas exchange [@beaulieuControlsGasTransfer2012; @wangPhysicallyBasedScaling2021; @raymondGasExchangeRivers2001].
-Similarly, none of the existing hydraulics-driven fluvial $k_{600}$ models account for wind-driven gas exchange either.
+Crucially, we are not accounting for wind-driven gas exchange, which is suggested to play an important role in wide rivers because river surface area is sufficiently large that sheltering no longer limits the influence of wind-derived turbulence [@beaulieuControlsGasTransfer2012; @wangPhysicallyBasedScaling2021; @raymondGasExchangeRivers2001].
+None of the existing hydraulics-driven fluvial $k_{600}$ models account for wind-driven gas exchange either.
+Additionally, under higher-wind scenarios the turbulent regime will switch from hydraulically-driven to wind-driven turbulence [@zappaEnvironmentalTurbulentMixing2007] and the assumptions under-pinning BIKER will likely break down.
 BIKER's outputs can therefore be interpreted as the '$k_{600}$ under low-wind conditions', when surface turbulence is dominated by hydraulics rather than wind.
-That said, BIKER's flexible implementation is a good start towards eventually coupling hydraulics-driven gas exchange with wind-driven gas exchange.
-Such a model would likely take a form similar to the conceptual models proposed by @wangPhysicallyBasedScaling2021, @chuWindStreamFlow2003, and/or @plateReaerationOpenChannel1984.
-All three generally propose calculating both a wind-driven *k* and a hydraulics-driven *k* and then weighting each term.
-In the case of BIKER, this would necessitate additional parameters that would need to be known a priori for specific rivers.
+That said, BIKER's flexible implementation is a good start towards eventually coupling hydraulics-driven gas exchange with wind-driven gas exchange under these moderate-to-high wind scenarios.
 This is left to future work.
@@
 For example, @wallinSpatiotemporalVariabilityGas2011 performed a preliminary analysis in northern Sweden relating cross-section specific temporal variability in gas exchange with channel slope, but they were limited to a mean of only 8 measurements per river in a single watershed.
-While this is a good start, this is obviously not sufficient for further developing process-level understandings of gas exchange at the global-scale.
+While this is a good start, this is insufficient for further developing process-level understandings of gas exchange at the global-scale.
 
@@
 For reference, 95% of the SWOT-visible rivers globally (202,811) will have sufficient SWOT observations along the river to run BIKER at least once every 21 days, with most of the temperate and Arctic rivers having 3+ observations per 21-day cycle [@altenauSurfaceWaterOcean2021].
-While BIKER will not directly measure $k_{600}$, it does robustly infer temporal trends in $k_{600}$ and less robustly the absolute magnitude of $k_{600}$ (Figures 4-6) and when coupled with SWOT this much data will provide a novel dataset of $k_{600}$ on a scale never before possible.
+While BIKER will not directly measure $k_{600}$, it does robustly infer temporal trends in $k_{600}$ and reasonably infers the absolute magnitude of $k_{600}$ (Figures 4-6).
+This much data will provide a novel dataset of $k_{600}$ on a scale never before possible.
 
@@
 This makes sense as Section 2 has effectively reduced $k_{600}$ to a function of hydraulics that are nearly all directly measurable by SWOT.
-Any resulting bias in BIKER's predictions is likely attributable to either bias in the priors used for the non-remotely sensed terms (equations 9-10) or in the $k_{600}$ model itself (equation 7, including the aforementioned wind errors).
+Any resulting bias in BIKER's predictions is likely attributable to either bias in the priors used for the non-remotely sensed terms (Equations 9-10) or in the $k_{600}$ model itself (Equation 7, including the aforementioned wind errors).
 For SWOT discharge algorithms, authors have repeatedly shown that the 'quality' of prior river knowledge plays a large role in the success of discharge inversions [@frassonExploringFactorsControlling2021; @andreadisConstrainingAssimilationSWOT2020; @brinkerhoffConstrainingRemoteRiver2020; @tuozzoloEstimatingRiverDischarge2019] and our results here further corroborate this finding.
@@
 the ability of a remote sensor to identify a hydrological response in the river [@garamboisHydraulicVisibilityUsing2017] is applicable here.
-If we apply hydraulic visibility to a sensor's ability to identify temporal variations in $k_{600}$, Figure 6 suggests that a 'minimally sufficient' hydraulic variability in SWOT measurements (in Figure 6b, suggested to be \> 10% *CV*) is needed to uniformally improve upon the prior.
+If we apply hydraulic visibility to a sensor's ability to identify temporal variations in $k_{600}$, Figure 6 suggests that a 'minimally sufficient' hydraulic variability in SWOT measurements (in Figure 6b, suggested to be approximately >20% *CV*) is needed to improve upon the prior.
 This will be important once SWOT launches and BIKER is implemented at the global-scale.
@@
 The models would give time and space varying gas exchange.
-@liuImportanceHydrologyRouting2022 and @saccardiImprovingPredictionsStream2021 each propose models that robustly predict reach-scale dissolved $CO_2$ concentrations using two different approaches: machine learning [@liuImportanceHydrologyRouting2022] and process-based reactive transport modeling [@saccardiImprovingPredictionsStream2021], but both models yield $CO_2$ estimates that would be spatially and temporally consistent with BIKER's output.
+@liuImportanceHydrologyRouting2022 and @saccardiImprovingPredictionsStream2021 each propose models that robustly predict reach-scale dissolved $CO_2$ concentrations using two different approaches [machine learning- @liuImportanceHydrologyRouting2022 and process-based reactive transport modeling- @saccardiImprovingPredictionsStream2021], but both models yield $CO_2$ estimates that would be spatially and temporally consistent with BIKER's output.
 Our promising results suggest that BIKER could provide additional (and directly inferred) measurements of $k_{CO_2}$ to these models, thereby better informing model results through direct observations.
@@
 Validating on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers, we show strong recovery of rivers' temporal $k_{600}$ dynamics and a hypothetical total annual carbon emission rate across all `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers.
-These results suggest BIKER can be used to infer global-scale, near daily estimates of fluvial gas exchange velocity once SWOT launches in 2022.
+These results suggest BIKER can be used to infer global-scale, near daily estimates of fluvial gas exchange velocity once SWOT launches in late 2022.
 This in turn will be useful in mapping the global-scale spatiotemporal dynamics of fluvial gas exchange in large rivers.
@@
 ## 6 Acknowledgements
-
 C.B.
@@
 ## 7 Data availability statement
-The version of BIKER used in this study will be archived upon acceptance.
-BIKER remains in active development and is available at <https://github.com/craigbrinkerhoff/BIKER>.
-All code to build and generate results, figures, and the manuscript is available at <https://github.com/craigbrinkerhoff/RSK600> and will be archived upon acceptance.
+Datasets required for this research are available from @ulsethDistinctAirWater2019 (<https://doi.org/10.1038/s41561-019-0324-8>), @brinkerhoffReconcilingAtaStationAtManyStations2019 (<https://doi.org/10.1029/2019GL084529>), @frassonExploringFactorsControlling2021 (<https://zenodo.org/record/3817817>), @churchillPredictionStreamReaeration1964 (<https://pubs.usgs.gov/pp/0737/report.pdf>), @owensReaerationStudiesStreams1964 (<https://pubs.usgs.gov/pp/0737/report.pdf>), @beaulieuControlsGasTransfer2012 (<https://doi.org/10.1029/2011JG001794>), and @durandIntercomparisonRemoteSensing2016 (<https://doi.org/10.1002/2015WR018434>).
+All code to build and generate results and figures is archived at <https://zenodo.org/record/6416069>.
 
-## 8 Apendix A
+## 8 Software availability statement
+The BIKER algorithm remains in active development and is available at <https://github.com/craigbrinkerhoff/BIKER>.
+The specific version of BIKER used in this study is archived at <https://zenodo.org/record/6415434>.
+
+## 9 Apendix A
 
@@
 |----------------------|-------------------------------------------------------------------------------|--------------------------------------------|------------------------------|
-| $A$                  | Channel cross-sectional area                                                  | NA                                         | $[L^2]$                      |
-| $A_0$                | Non-SWOT-observable cross-sectional area                                      | NA                                         | $[L^2]$                      |
-| $\alpha$             | Statistical parameter for equation S5 scaling relation                        | NA                                         | $\frac{m^{1/2}s^{3/4}}{dy}$  |
-| $\alpha_1$           | Statistical parameter for equation S7 scaling relation                        | NA                                         | $\frac{m^{1/2}s^{3/4}}{dy}$  |
-| $\beta$              | Statistical parameter for equation S13 scaling relation                       | NA                                         | $\frac{m^{-1/4}s^{9/8}}{dy}$ |
-| $\beta_1$            | Statistical parameter for equation 7 scaling relation                         | NA                                         | $\frac{m^{1/2}s^{3/4}}{dy}$  |
-| $CO_{2_{water}}$     | Water-side $CO_2$ concentration                                               | NA                                         | $[\frac{M}{L^3}]$            |
-| $CO_{2_{air}}$       | Atmospheric-side $CO_2$ concentration                                         | NA                                         | $[\frac{M}{L^3}]$            |
+| $A$                  | Channel cross-sectional area                                                  |                                            | $[L^2]$                      |
+| $A_0$                | Non-SWOT-observable cross-sectional area                                      |                                            | $[L^2]$                      |
+| $A_p$                | Active zone fraction (Moog & Jirka 1999)                                      |$A_p \propto Re_*^{1/2}$(Moog & Jirka 1999) | $dimensionless$              |
+| $\alpha$             | Statistical parameter for Equation S5 scaling relation                        |                                            | $\frac{m^{1/2}s^{3/4}}{dy}$  |
+| $\alpha_1$           | Statistical parameter for Equation S7 scaling relation                        |                                            | $\frac{m^{1/2}s^{3/4}}{dy}$  |
+| $\beta$              | Statistical parameter for Equation S13 scaling relation                       |                                            | $\frac{m^{-1/4}s^{9/8}}{dy}$ |
+| $\beta_1$            | Statistical parameter for Equation 7 scaling relation                         |                                            | $\frac{m^{1/2}s^{3/4}}{dy}$  |
+| $CO_{2_{water}}$     | Water-side $CO_2$ concentration                                               |                                            | $[\frac{M}{L^3}]$            |
+| $CO_{2_{air}}$       | Atmospheric-side $CO_2$ concentration                                         |                                            | $[\frac{M}{L^3}]$            |
 | $dA$                 | change in cross-sectional area                                                | $\delta H_e*W$                             | $[L]$                        |
-| $D_m$                | Molecular diffusion coefficient                                               | NA                                         | $[\frac{L^2}{T}]$            |
-| $\epsilon$           | Dissipation rate of near-surface turbulence                                   | NA                                         | $[\frac{E}{M*T}]$            |
+| $D_m$                | Molecular diffusion coefficient                                               |                                            | $[\frac{L^2}{T}]$            |
+| $\epsilon$           | Dissipation rate of near-surface turbulence                                   |                                            | $[\frac{E}{M*T}]$            |
 | $\epsilon_S$         | log-law-of-the-wall model for $\epsilon$                                      | $\frac{U_*^3}{H}$                          | $[\frac{E}{M*T}]$            |
 | $\epsilon_D$         | Form-drag model for $\epsilon$                                                | $gS\bar{U}$                                | $[\frac{E}{M*T}]$            |
-| $FCO_2$              | $CO_2$ evasion flux from river to atmosphere                                  | NA                                         | $[\frac{M}{L^2T}]$           |
-| $FCO_{2_{upscaled}}$ | Upscaling estimate of the global $CO_2$ evasion flux from river to atmosphere | NA                                         | $[\frac{M}{L^2T}]$           |
+| $FCO_2$              | $CO_2$ evasion flux from river to atmosphere                                  |                                            | $[\frac{M}{L^2T}]$           |
+| $FCO_{2_{upscaled}}$ | Upscaling estimate of the global $CO_2$ evasion flux from river to atmosphere |                                            | $[\frac{M}{L^2T}]$           |
 | $g$                  | gravitational acceleration                                                    | 9.8                                        | $[\frac{L}{T^2}]$            |
 | $H$                  | Mean flow depth                                                               | $\frac{A}{W}$                              | $[L]$                        |
-| $H_e$                | Water surface elevation                                                       | NA                                         | $[L]$                        |
-| $i$                  | Cross-section discretization within a mass-conserved river reach              | NA                                         | NA                           |
-| $k$                  | gas exchange velocity                                                         | NA                                         | $[\frac{L}{T}]$              |
+| $H_e$                | Water surface elevation                                                       |                                            | $[L]$                        |
+| $i$                  | Cross-section discretization within a mass-conserved river reach              |                                            |                              |
+| $k$                  | gas exchange velocity                                                         |                                            | $[\frac{L}{T}]$              |
 | $k_{600}$            | gas exchange velocity normalized to $Sc=600$                                  | $(\frac{600}{Sc})^{-0.5}k$                 | $[\frac{L}{T}]$              |
 | $n$                  | Manning's roughness coefficient                                               | $\frac{R_h^{2/3}S^{1/2}}{\bar{U}}$         | $[\frac{T}{L^3}]$            |
-| $\uprho$             | Density of water                                                              | NA                                         | $[\frac{M}{L^{1/3}}]$        |
+| $\uprho$             | Density of water                                                              |                                            | $[\frac{M}{L^{1/3}}]$        |
 | $Q$                  | River discharge                                                               | $WH\bar{U}$                                | $\frac{L^3}{T}$              |
 | $R_h$                | Hydraulic radius                                                              | $\frac{AW}{2H+W}$                          | $[L]$                        |
-| $S$                  | River slope                                                                   | NA                                         | $unitless$                   |
-| $Sc$                 | Schmidt number                                                                | $\frac{v}{D_m}$                            | NA                           |
-| $t$                  | timestep discretization within river reach                                    | NA                                         | NA                           |
-| $\Theta$             | Bayesian parameter set                                                        | NA                                         | NA                           |
+| $Re_8$               | Shear Reynold's numbers                                                       | $\frac{U_*H}{v}$                           | $dimensionless$              |
+| $S$                  | River slope                                                                   |                                            | $dimensionless$              |
+| $Sc$                 | Schmidt number                                                                | $\frac{v}{D_m}$                            |                              |
+| $t$                  | timestep discretization within river reach                                    |                                            |                              |
+| $\Theta$             | Bayesian parameter set                                                        |                                            |                              |
 | $\bar{U}$            | Cross-sectional average velocity                                              | $\frac{Q}{A}$                              | $[\frac{L}{T}]$              |
 | $U_*$                | Shear velocity                                                                | $\sqrt{gSR_h}$                             | $[\frac{L}{T}]$              |
-| $\mu$                | Viscosity                                                                     | NA                                         | $[P*T]$                      |
+| $\mu$                | Viscosity                                                                     |                                            | $[P*T]$                      |
 | $v$                  | kinematic viscosity                                                           | $\frac{\mu}{\uprho}$                       | $[\frac{L^2}{T}]$            |
-| $W$                  | Flow width                                                                    | NA                                         | $[L]$                        |
-| $x$                  | Bayesian data set                                                             | NA                                         | NA                           |
+| $W$                  | Flow width                                                                    |                                            | $[L]$                        |
+| $x$                  | Bayesian data set                                                             |                                            |                              |
 

### 2022-04-05 supp_v11.Rmd
final ms edits

--- 2022-04-05 docs/supp_v10.Rmd
@@
 
-First, we removed all measurements tagged by the USGS as 'poor', measurements with impossible values, or measurements of 0.
+First, we removed all measurements tagged by the USGS as 'poor', measurements with non-finite values, or measurements of 0.
 While this would indicate a dry channel, our hydraulic geometry model necessitates within-bank flow.
@@
 While the only true way to calculate bankfull hydraulics is manually in the field, this is obviously impractical here.
-A two year return period is a standard approximation for determining out-of-bank flow in single-channel meandering rivers and was the method used by @brinkerhoffReconcilingAtaStationAtManyStations2019.
+A two-year return period is a standard approximation for determining out-of-bank flow in single-channel meandering rivers and was the method used by @brinkerhoffReconcilingAtaStationAtManyStations2019.
 We then removed all measurements with a width or depth beyond their respective at-a-station 2 year values.
@@
 
-**Small-eddy models**
+**Small-eddy models (Equation 5 from Main Test)**
 
-*Log-law-of-the-wall model for the turbulent dissipation rate*
+*Log-law-of-the-wall model for the turbulent dissipation rate (assume* $\epsilon=\epsilon_S$ *)*
 
@@
 
-*Form-drag model for the turbulent dissipation rate*
+*Form-drag model for the turbulent dissipation rate (assume* $\epsilon=\epsilon_D$ *)*
 
@@
 
-**Reynolds extension models**
+**Reynolds extension models (Equation 6 from Main Text)**
 
-*Log-law-of-the-wall model for the turbulent dissipation rate*
+*Log-law-of-the-wall model for the turbulent dissipation rate (assume* $\epsilon=\epsilon_S$ *)*
 
@@
 Prior distributions are defined by their hyperparameters.
-For BIKER, prior distriubtions are formalized as truncated normal distributions of the log-transformed terms such that $log(X) \sim  \mathcal{N}(\mu, \sigma^2)$ for $\lambda < log(X) < \gamma$, using prior hyperparameters mean ($\mu$), standard deviation ($\sigma$), and upper ($\gamma$) and lower bounds ($\lambda$) for any parameter *X*.
+For BIKER, prior distributions are formalized as truncated normal distributions of the log-transformed terms such that $log(X) \sim  \mathcal{N}(\mu, \sigma^2)$ for $\lambda < log(X) < \gamma$, using prior hyperparameters mean ($\mu$), standard deviation ($\sigma$), and upper ($\gamma$) and lower bounds ($\lambda$) for any parameter *X*.
 It is important to again stress, as we do in the main text, that BIKER prior hyperparameters are described using only SWOT data to be completely globally implementable.
@@
 This leaves the $k_{600_t}$ hyperparameters to be defined.
-$k_{600_{\mu}}$ is set by invoking the hydraulic geometry (HG) relationships developed in section 3.2.3 of the main text and Table S2 using the data from @brinkerhoffReconcilingAtaStationAtManyStations2019.
-We replaced both depth and velocity terms from our gas exchange model (equation 7 in the Main Text) with these HG models, resulting in equation S19 where Q is the mass-conserved streamflow for the river reach.
+$k_{600_{\mu}}$ is set by invoking the hydraulic geometry (HG) relationships developed in Section 3.2.3 of the main text and Table S2 using the data from @brinkerhoffReconcilingAtaStationAtManyStations2019.
+We replaced both depth and velocity terms from our gas exchange model (Equation 7 in the Main Text) with these HG models, resulting in Equation S19 where Q is the mass-conserved streamflow for the river reach.
 
@@
 ## Text S4: Determining complete BIKER model uncertainty
-Recall that $\sigma_{k_{600}}$ refers to the total uncertainty inherent in equation 9, i.e.
-stemming from the Reynolds-extension model (equation 7) and Manning's equation for $\bar{U}$.
-For the purposes of this study, we are validating BIKER against equation 7 in the main text and so all uncertainties associated with the $\beta_1$ parameter are ignored and we only need to reflect the Manning's uncertainty in our specification of $\sigma_{k_{600}}^2$.
+Recall that $\sigma_{k_{600}}$ refers to the total uncertainty inherent in Equation 9, i.e.
+stemming from the Reynolds-extension model (Equation 7) and Manning's equation for $\bar{U}$.
+For the purposes of this study, we are validating BIKER against Equation 7 in the main text and so all uncertainties associated with the $\beta_1$ parameter are ignored and we only need to reflect the Manning's uncertainty in our specification of $\sigma_{k_{600}}^2$.
 Therefore, we take @hagemannBAMBayesianAMHGManning2017's estimated uncertainty from Manning's equation to infer streamflow from SWOT observations (0.25) and inflate it slightly to also account for the hydraulically-wide channel assumption and arrive at 0.30.
@@
 
-In the scenario that BIKER is run on real SWOT data, $\sigma_k_{600}$ must reflect the full uncertainty implicit in equation 9 in the main text.
+In the scenario that BIKER is run on real SWOT data, $\sigma_k_{600}$ must reflect the full uncertainty implicit in Equation 9 in the main text.
 This means we must also account for uncertainty from the $\beta_1$ parameter.
-Assuming perfect, no-error measurements are made by the SWOT satellite, the full equation 7 uncertainty is expressed for some set of hydraulic observations as equation S20.
-To quantify total uncertainty in $k_{600}$, we push Monte Carlo simulations through equation S20 and take the mean uncertainty as a reasonable value for $\sigma_{k_{600}}$ once SWOT launches.
+Assuming perfect, no-error measurements are made by the SWOT satellite, the full Equation 7 uncertainty is expressed for some set of hydraulic observations as Equation S20.
+To quantify total uncertainty in $k_{600}$, we push Monte Carlo simulations through Equation S20 and take the mean uncertainty as a reasonable value for $\sigma_{k_{600}}$ once SWOT launches.
 
-More specifically, we use the 166 hydraulically-wide measurements in our field-measured $k_{600}$ dataset (section 2 in the Main Text) and push 166 different Monte Carlo simulations through equation S20.
+More specifically, we use the 166 hydraulically-wide measurements in our field-measured $k_{600}$ dataset (Section 2 in the Main Text) and push 166 different Monte Carlo simulations through Equation S20.
 Note that each Monte Carlo simulation is itself 10,000 runs in order to obtain a distribution of $k_{600}$ estimates.
@@
 ## Text S5: Literature $CO_2$ flux models
-Here, we describe in more detail the $CO_2$ flux models used to compare against BIKER in section 3.2.3. The specific equations used are outlined in Table S2.
+Here, we describe in more detail the three $CO_2$ flux models used to compare against BIKER in Section 3.2.3. The specific equations used are outlined in Table S2.
 
-In brief, gas fluxes ($FCO_2$) are only obtainable at the global scale via predictive equations for $k_{CO_2}$ per equation 2.
-Here, we obtain $FCO_2$ using the in situ $CO_2$ data described in section 3.1 and $k_{CO_2}$ estimates.
-The $k_{CO_2}$ estimates are calculated using 1) equation 7, 2) in situ streamflow data, and 3) hydraulic scaling equations from the literature that predict river velocity and depth from streamflow.
-Specifically, these literature equations include the 'Raymond 2012', 'Raymond 2013', and 'Brinkerhoff 2019' models (Table S2).
+In brief, gas fluxes ($FCO_2$) are only obtainable at the global scale via predictive equations for $k_{CO_2}$ per Equation 2.
+Here, we obtain $FCO_2$ using the in situ $CO_2$ data described in Section 3.1 and $k_{CO_2}$ estimates.
+The $k_{CO_2}$ estimates are calculated using Equation 7.
+Equation 7 is calculated using 1) in situ streamflow data, and 2) hydraulic scaling equations from the literature that predict river velocity and depth from streamflow.
+Specifically, these literature equations are the 'Raymond 2012', 'Raymond 2013', and 'Brinkerhoff 2019' models (Table S2).
 
-For example, the 'Raymond 2012' model uses the hydraulic scaling equations from @raymondScalingGasTransfer2012 to predict velocity and depth, which are in turn used to calculate a $k_{CO_2}$, which finally is used to calculate $FCO_2$.
-Similarly, the 'Raymond 2013' model uses the equations outlined in @raymondGlobalCarbonDioxide2013 and the 'Brinkerhoff 2019' model uses new equations fit to the hydraulics dataset from @brinkerhoffReconcilingAtaStationAtManyStations2019 after the filtering descirbed in Text S1.
+For example, the 'Raymond 2012' model uses the hydraulic scaling equations from @raymondScalingGasTransfer2012 to predict velocity and depth, which are in turn used to calculate a $k_{CO_2}$ via equation 7, which finally is used to calculate $FCO_2$ via equation 1.
+Similarly, the 'Raymond 2013' model uses the equations outlined in @raymondGlobalCarbonDioxide2013 and the 'Brinkerhoff 2019' model uses new equations fit to the hydraulics dataset from @brinkerhoffReconcilingAtaStationAtManyStations2019 after the filtering described in Text S1.
 We chose to include the 'Brinkerhoff 2019' model as the training dataset is far larger than those used in either of the previous two models (Table S3: 104,624 versus 10,837 versus 1,026, respectively).
@@
 ![Figure S1: Four $k_{600}$~hydraulics models tested in this study, with model assumptions noted in the sub-panel titles.
-Panel (d) is also presented in the main text as Figure 2 and equation 7.](cache/k600_theory/figS1.jpg)
+Panel (d) is also presented in the main text as Figure 2 and Equation 7.](cache/k600_theory/figS1.jpg)
 
 ## Figure S2
-![Figure S2: Map of simulated SWOT river locations around the world.
+![Figure S2: Map of simulated SWOT river locations along the global SWOT river network [@altenauSurfaceWaterOcean2021].
+This corresponds to every river that BIKER could be run on once SWOT launches.
 Also includes approximate location of the $CO_2$ timeseries used in this study (Figure S3).
@@
 ## Figure S4
-![Figure S4: Observed (orange) versus BIKER-inferred (green) timerseries of daily $k_{600}$ for the 47 SWOT-simulated rivers under the 'no-measurement-error' scenario.
+![Figure S4: Observed (orange) versus BIKER-inferred (green) timeseries of daily $k_{600}$ for the 47 SWOT-simulated rivers under the 'no-measurement-error' scenario.
+Green ribbon is BIKER posterior 95% confidence intervals.
 Note that the y axis is normalized by maximum observed values to compare across rivers.](cache/validation/figS4.jpg)
@@
 ## Figure S5
-![Figure S5: Observed (orange) versus BIKER-inferred (green) timerseries of daily $k_{600}$ for 16 rivers under the 'measurement-error' scenario.
+![Figure S5: Observed (orange) versus BIKER-inferred (green) timeseries of daily $k_{600}$ for 16 rivers under the 'measurement-error' scenario.
+Green ribbon is BIKER posterior 95% confidence intervals.
 Note that the y axis is normalized by maximum observed values to compare across rivers.](cache/validation/figS5.jpg)
@@
 |Brinkerhoff 2019|$D=0.252Q^{0.388}$|$\bar{U}=0.276Q^{0.164}$|104,624 measurements made across the United States at streamgauges|@brinkerhoffReconcilingAtaStationAtManyStations2019|
-|Raymond 2012|$D=0.409Q^{0.294}$|$\bar{U}=0.194Q^{0.285}$|@raymondScalingGasTransfer2012|1,026 measurements across the United States|@raymondScalingGasTransfer2012|
+|Raymond 2012|$D=0.409Q^{0.294}$|$\bar{U}=0.194Q^{0.285}$|1,026 measurements across the United States|@raymondScalingGasTransfer2012|
 |Raymond 2013|$D=0.409Q^{0.294},  D=0.449Q^{0.37}$|$\bar{U}=0.194Q^{0.285}, \bar{U}=0.346Q^{0.120}$|Average of the Raymond 2012 equation and one using 9,811 measurements at US streamgauges|@raymondGlobalCarbonDioxide2013|

### 2022-04-05 generateMS.R

@@ -7,14 +7,13 @@ modelsWIDE <- read.csv('cache\\k600_theory\\hydraulicWide_models.csv')
 n <- 166 #number of hydraulically wide measurements used in k600 validation
 HG_swot <- read.csv('cache\\k600_theory\\HG_swot.csv')
 
-results_dynamics <- read.csv('cache/validation/results_dynamics.csv')
 results_k600_rivs <- read.csv('cache//validation//results_by_riv.csv')
 results_fco2_bulk <- read.csv('cache//FCO2//massFluxes.csv')
 results_fco2_bulk$key <- as.character(results_fco2_bulk$key)
 
 #knit manuscript
-rmarkdown::render("manuscript_v10.Rmd")
+rmarkdown::render("manuscript_v11.Rmd")
 print('Manuscript generated.')
 
-rmarkdown::render("supp_v10.Rmd")
+rmarkdown::render("supp_v11.Rmd")
 print('Supplemental Information generated.')

### 2022-04-05 src/figures_k600.R

@@ -14,13 +14,13 @@ theme_set(theme_classic())
 ##READ IN RESULTS----------------------------------
 ###############
 results <- read.csv('cache/validation/BIKER_validation_results.csv')
-full_output <- filter(results, errFlag == 0) #remove results with SWOT measurement error
+full_output <- filter(results, errFlag == 0) #remove results with SWOT measurement error (for now)
 
 ########################
 ##CALCULATE BY-RIVER ERROR METRICS---------------------------------------------
 ########################
 stats_by_river <- group_by(results, river, errFlag) %>%
-  summarise(r = sqrt(summary(lm(kest_mean~kobs))$r.squared),
+  summarise(r = cor(kobs, kest_mean, method='pearson'),
             NRMSE = sqrt(mean((kest_mean - kobs)^2, na.rm=T)) / mean(kobs, na.rm=T),
             KGE = KGE(kest_mean, kobs),
             NMAE = mean(abs(kest_mean-kobs)/mean(kobs, na.rm=T), na.rm=T),
@@ -84,7 +84,7 @@ nmae_cdf <- ggplot(stats_by_river[stats_by_river$errFlag==0,], aes(x=NMAE)) +
         plot.title = element_text(size = 30, face = "bold"))
 
 statsPlot <- plot_grid(kge_cdf, nrmse_cdf, r_cdf, nmae_cdf, ncol=2, labels = NA)
-ggsave('cache/validation/fig5.jpg', statsPlot, width=10, height=10)
+ggsave('cache/validation/fig4.jpg', statsPlot, width=10, height=10)
 
 ########################
 ##ERROR/NO ERROR COMPARISON------------------------------------
@@ -121,13 +121,13 @@ forError_r <- pivot_wider(forError, names_from=errFlag, values_from = r) %>%
             r_err = sum(`1`, na.rm=T))
 
 r_scatter <- ggplot(forError_r, aes(x=r_no_err, y=r_err)) +
-  geom_polygon(data = data.frame(x = c(0, 1, 0), y = c(0,1,1)), aes(x = x, y = y), fill = "#8dd3c7" )+  # better triangle
-  geom_polygon(data = data.frame(x = c(0, 1, 1), y = c(0,1,0)), aes(x = x, y = y), fill = "#bebada" )+  # worse triangle
[…]

### 2022-04-05 src/main.R

@@ -1,6 +1,9 @@
 ###############
-#Set up and run project
+## Set up and run project
+## Creator: Craig Brinkerhoff
+## Winter 2022
 ###############
+
 .libPaths('/nas/cee-water/cjgleason/r-lib')
 library(ProjectTemplate, lib.loc = "/nas/cee-water/cjgleason/r-lib/")
 load.project()

### 2022-04-05 src/old/debugBIKER.R

@@ -1,0 +1,123 @@
+#packages--------------------------------------------------------------------
+library(BIKER)
+library(ncdf4)
+library(dplyr)
+
+g <- 9.8
+uncertainity <- 0.30
+
+#load in rivers----------------------------------------------------------------
+files <- list.files('C:/Users/craig/Documents/OneDrive - University of Massachusetts/Ongoing Projects/RSK600/data/Frasson_etal_2021/IdealDataxxxxxx', pattern="*.nc", full.names = TRUE) #pepsi 2
+files2 <- list.files('C:/Users/craig/Documents/OneDrive - University of Massachusetts/Ongoing Projects/RSK600/data/Durand_etal_2016/xxxxxxxxxxxxxxxx', pattern="*.nc", full.names = TRUE) #pepsi 1
+files <- c(files, files2)
+
+#grab a dummy river
+currPepsi <- files[2]
+
+#extract river name-----------------------------------------------------------
+name <- substr(currPepsi, 40, nchar(currPepsi))
+name <- substr(name,1,nchar(name)-3)
+set.seed(12)
+data_in =nc_open(currPepsi)
+
+#extract reach-scale river hydraulics-----------------------------------------------------------
+W_obs=ncvar_get(data_in,'Reach_Timeseries/W')
+H_obs=ncvar_get(data_in,'Reach_Timeseries/H')
+S_obs = ncvar_get(data_in, 'Reach_Timeseries/S')
+area = ncvar_get(data_in, 'Reach_Timeseries/A')
+Q_obs=ncvar_get(data_in,'Reach_Timeseries/Q')
+priorQ <- ncvar_get(data_in, 'River_Info/QWBM')
+
+W_obs=W_obs[1:3,1:4]
+H_obs=H_obs[1:3,1:4]
+S_obs = S_obs[1:3,1:4]
+area = area[1:3,1:4]
+Q_obs= Q_obs[1:3,1:4]
+
+#prep river hydraulics-----------------------------------------------------------
+S_obs[S_obs<=0]=NA
+S_obs[is.na(S_obs)] = 0.000017 #min obs SWOT slope Biancarma etal 2016
+W_obs[W_obs<0]=NA
+H_obs[H_obs<0]=NA
+Q_obs[Q_obs<0]=NA
+area[area<0]=NA
+
[…]

### 2022-04-05 src/runFCO2.R

@@ -6,11 +6,16 @@
 
 print('starting FCO2...')
 
-#legend labels------------------
+######################
+#LEGEND LABELS------------------
 legend_labels <- c('BIKER',
                    'Model using observed \ndepth')
+######################
 
-#This model specifically predicts w and d from HG and then obtains velocity by continuity. The model fit for velocity was notably worse.
+#######################
+#READ IN HYDRAULICS MODELS FOR 'BRINKERHOFF 2019' MODEL--------------
+########################
+  #This model specifically predicts w and d from HG and then obtains velocity by continuity. The model fit for velocity was notably worse.
 widAHG <- read_rds('cache/widAHG.rds')
 depAHG <- read_rds('cache/depAHG.rds')
 
@@ -51,7 +56,7 @@ fun_FCO2_analysis <- function(river) {
     return(NA)
   }
 
-  #Some NA handling in slopes (borrowed from Frassion etal 2021 BAM runs)-----------------------------------------------------------
+  #Some NA handling in slopes (borrowed from Frasson etal 2021 BAM runs)
   if (any(apply(S_obs,2,sum,na.rm=TRUE) ==0)){ #removes timesteps with all NA slopes
     remove_index =  which((apply(S_obs,2,sum,na.rm=TRUE) ==0) ==TRUE)
 
@@ -61,7 +66,7 @@ fun_FCO2_analysis <- function(river) {
     area=area[,-remove_index]
   }
 
-  if (any(apply(S_obs,1,sum,na.rm=TRUE) ==0)){ #removes sptial steps with all NA slopes
+  if (any(apply(S_obs,1,sum,na.rm=TRUE) ==0)){ #removes spatial steps with all NA slopes
     remove_index =  which((apply(S_obs,1,sum,na.rm=TRUE) ==0) ==TRUE)
 
     W_obs=W_obs[-remove_index,]
@@ -77,9 +82,9 @@ fun_FCO2_analysis <- function(river) {
   D_obs <- area / W_obs #[m]
 
   #Calculate depth rating curves following suite of rating curve models [m/s]
[…]

### 2022-06-08 docs/intial_drafts/manuscript_v11.Rmd
first round of review, setting up for rerun on HPC

--- 2022-04-05 manuscript_v11.Rmd
@@
 River $CO_2$ evasion is increasingly better constrained and is clearly a critical component of the global carbon cycle.
-Equation 1 represents this riverine flux given $\delta CO_2$ (the air-water flux between the water $CO_{2_{water}}$ and the air $CO_{2_{air}}$) and the gas exchange velocity *k*.
+Equation 1 represents this riverine flux given $\delta CO_2$ (the air-water difference between the water $CO_{2_{water}}$ and the air $CO_{2_{air}}$) and the gas exchange velocity *k*.
 Consult Appendix A for variable nomenclature used throughout this study.
@@
 Figure 2 shows that Equation 7 explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2, 2) * 100`% of variation in observed $k_{600}$ in hydraulically-wide rivers and accurately captures the scaling dynamics of observed $k_{600}$ as well.
-The value for $\beta_1$ is 62.8, which is similar the fitted parameters in other $Re_*$ and $U_*$ models for river *k* [@wangPhysicallyBasedScaling2021].
+The value for $\beta_1$ is 62.8, which is similar to the fitted parameters in other $Re_*$ and $U_*$ models for river *k* [@wangPhysicallyBasedScaling2021].
 While the individual residuals can be quite large, the overall scaling of $k_{600}$ with river hydraulics is strongly captured, remembering that Figure 2 is plotted in logarithmic space.
@@
 
-We validate BIKER under two different 'error scenarios'.
+We validate BIKER under two different 'error scenarios' (section 3.2.2).
 While SWOT will provide river surface measurements of novel quality and resolution, as with all remote sensing products there are expected errors that will be implicit in these measurements.
@@
 
-$$\left[log(62.82) + \frac{7}{16}log(g) + \frac{9}{16}log(S_{i,t}) - \frac{35}{48}log(W_{i,t})\right] \sim normal\left(\left[log(k_{600_t})+\frac{1}{4}log(n_i)-\frac{35}{48}log(A_0_i+dA_{i,t})\right], \sigma^2_{k_{600}}\right) \mathbf{(10)}$$ Equations 8 and 10 also necessitate that we specify prior distributions for the parameters $A_0_i$, $k_{600_t}$, and $n_i$.
+$$\left[log(62.82) + \frac{7}{16}log(g) + \frac{9}{16}log(S_{i,t}) - \frac{35}{48}log(W_{i,t})\right] \sim \mathcal{N}\left(\left[log(k_{600_t})+\frac{1}{4}log(n_i)-\frac{35}{48}log(A_0_i+dA_{i,t})\right], \sigma^2_{k_{600}}\right) \mathbf{(10)}$$ Equations 8 and 10 also necessitate that we specify prior distributions for the parameters $A_0_i$, $k_{600_t}$, and $n_i$.
 Prior distributions formalize the a priori estimates and uncertainties for the non-remotely-sensed terms.
@@
 BIKER's relatively stronger performance than the in situ models is elaborated on in Section 4.3.
-Finally, BIKER's uncertainity is on par with the in situ technique, despite being obtained solely from SWOT data.
+Finally, BIKER's uncertainity is on par with the in situ technique (‘Brinkerhoff 2019’), despite being obtained solely from SWOT data.
 Taken in aggregate, BIKER provides a strong upscaling estimate of the annual carbon emission rate for the rivers and is either comparable or better than established in situ techniques (Figure 7).
@@
 The models would give time and space varying gas exchange.
-@liuImportanceHydrologyRouting2022 and @saccardiImprovingPredictionsStream2021 each propose models that robustly predict reach-scale dissolved $CO_2$ concentrations using two different approaches [machine learning- @liuImportanceHydrologyRouting2022 and process-based reactive transport modeling- @saccardiImprovingPredictionsStream2021], but both models yield $CO_2$ estimates that would be spatially and temporally consistent with BIKER's output.
+@liuImportanceHydrologyRouting2022 and @saccardiImprovingPredictionsStream2021 each propose models that robustly predict reach-scale dissolved $CO_2$ concentrations using two different approaches- machine learning for @liuImportanceHydrologyRouting2022 and process-based reactive transport modeling for @saccardiImprovingPredictionsStream2021- but both models yield $CO_2$ estimates that would be spatially and temporally consistent with BIKER's output.
 Our promising results suggest that BIKER could provide additional (and directly inferred) measurements of $k_{CO_2}$ to these models, thereby better informing model results through direct observations.

### 2022-06-08 docs/intial_drafts/manuscript_v3.Rmd
first round of review, setting up for rerun on HPC

--- 2021-06-04 manuscript_v3.Rmd
@@
 ##1 Introduction
-**I don't like this opening paragraph** The Earth's carbon cycle moves carbon between the atmosphere, geosphere, hydrosphere, and biosphere via a complex set of interconnected biogeochemical processes.
+**I don't like this opening paragraph** The Earth moves carbon between the atmosphere, geosphere, hydrosphere, and biosphere via a complex set of interconnected biogeochemical processes.
 The primary connection between the terrestrial and oceanic biospheres is the global river network, which exports a significant amount of carbon from the continents to the oceans [@colePlumbingGlobalCarbon2007a], while additionally transporting and transforming sediments, nutrients, organic matter, dissolved gases, and other constituents.
@@
 Equation 1 represents this flux for any sparingly soluble gas $[M/L^2T]$ given the gas concentration gradient between the water $C_{water}$ $[M/L^3]$ and the air $C_{air}$ $[M/L^3]$ and the gas exchange velocity *k [L/T]*.
-Given the structure of equation 1, calculations of this flux are very sensitive to one's measurements/estimates of *k*.
+Given the structure of equation 1, calculations of this flux are highly sensitive to one's measurements/estimates of *k*.
 Note that *k* is different for different gases and water temperatures [see @hallGasExchangeStreams2020 for a thorough review of gas exchange in rivers].
@@
 *k* is also vital to understanding freshwater ecology through bulk river metabolism [more specifically, gross primary production and ecological respiration- @bernhardtMetabolicRegimesFlowing2018].
-The general mass-balance equation for river metabolism (*M* $[M/L^2T]$) is provided as equation 2, where *DO* is the dissolved oxygen concentration $[M/L^3]$, *D* is river depth [L], and $k_{o_2}$ is *k* for oxygen at some temperature.
+The general mass-balance equation for river metabolism (*M* $[M/L^2T]$) is provided as equation 2, where *DO* is the dissolved oxygen concentration $[M/L^3]$, *Z* is river depth [L], and $k_{o_2}$ is *k* for oxygen at some temperature.
 When viewing equations 1 and 2 in aggregate, it is clear that *k* plays a critical role in two of the core, fundamental biogeochemical processes that occur in river networks.
 Both are reliant on accurate and high-fidelity assessments of *k*, as well as a rigorous understanding of the physical mechanisms driving *k*.
-$$\mathbf{(2)} \frac{dDO}{dt}=\frac{M}{D}+\frac{k_{o_2}}{D}(DO_{air}-DO_{water})$$
+$$\mathbf{(2)} \frac{dDO}{dt}=\frac{M}{Z}+\frac{k_{o_2}}{Z}(DO_{air}-DO_{water})$$
 Researchers use a range of methods to measure *k*, from floating domes to eddy covariance towers to measure at a specific point in the river.
@@
 Second, it yields a simple linear relationship across all rivers, regardless of their size or steepeness.
-This is not necessairly true of other *k* models, for example those based on *eD* (section 1).
-The relationship between *eD* and *k* is not necessairly linear [@raymondScalingGasTransfer2012a], suggesting poorer performance for small *k* values.
+This is not true of other *k* models, for example those based on *eD* (section 1).
+The relationship between *eD* and *k* is not necessairly linear, with reasonable performance from both linear and power-law models and generally poorer predictive performance for small *k* values [@raymondScalingGasTransfer2012a].
 Further, recent work has shown that *k* does not scale with *eD* via the same statistical parameters across all river geomorphologies [@ulsethDistinctAirWater2019a].
-We further explored the relationship presentedby [@ulsethDistinctAirWater2019a] and found that the $k \sim eD$ relationship fundamentally breaks down in rivers with low *eD*, which corresponds to nearly all rivers that SWOT will observe (Figure S2).
+We further explored the relationship presented by [@ulsethDistinctAirWater2019a] in the SWOT context, finding that the $k \sim eD$ relationship fundamentally breaks down in rivers with low *eD*, which corresponds to nearly all rivers that SWOT will observe (Figure S2).
 Therefore, an *eD* based predictive model is less useful in SWOT rivers.
@@
 However, our goal of using BIKER on any river that is observable by SWOT necessitates that we pick an equation that is applicable across as many rivers as possible.
-Future work should look at using a river-specific *k* equation to pseudo-calibrate BIKER to a specific river.
+Future work should look at using a river-specific *k* equation to pseudo-calibrate BIKER to a specific river, as well as the influence of wind-induced gas exchange (see section 5.2).
 
@@
 
-With the likelihood function (equation 6), prior distributions, and the $\sigma_{k_{o_2}}^2$ parameter described (section 3.2.2), a joint posterior distribution conditional on the SWOT observations is specified.
+With the likelihood function (section 3.2.1), prior distributions (section 3.2.2), and the $\sigma_{k_{o_2}}^2$ parameter described (section 3.2.2), a joint posterior distribution conditional on the SWOT observations is specified.
 To approximate this distribution, we use a Markov Chain Monte Carlo (MCMC) algorithm implemented using the Stan probabilistic programming language.
-Specifically, BIKER uses a Hamiltonian Monte Carlo sampler which reduces computation time relative to other sampling algorithms [@hagemannBAMBayesianAMHGManning2017a].
+Specifically, Stan uses a Hamiltonian Monte Carlo sampler which reduces computation time relative to other sampling algorithms [@hagemannBAMBayesianAMHGManning2017a].
 
@@
 Further, we are principally interested in BIKER's ability to reproduce $k_{o_2}$ from SWOT observations and are less concerned with the actual accuracy of the predictive model itself, which can be validated using existing datasets as has been previously done [e.g.
-@wangPhysicallyBasedScaling2021; @ulsethDistinctAirWater2019a; @raymondScalingGasTransfer2012a].
+@raymondScalingGasTransfer2012a; @ulsethDistinctAirWater2019a; @wangPhysicallyBasedScaling2021].
 Therefore, we take the model outlined in section 3.1 and use that to calculate the observed $k_{o_2}$ that BIKER is validated against.
@@
 It is one thing to accurately predict $k_{o_2}$, but researchers are often most interested in the actual gas fluxes from rivers and ultimately the carbon emitted from river to atmosphere.
-Therefore, we explore 1) the ability of BIKER to reproduce $CO_2$ evasion fluxes ($FCO_2$ $[M/L^2T]$) from these 49 rivers, and 2) the sensitivity of the carbon efflux $[M/T]$ due to the chosen method to approximate $R_h$ and ultiamtely $U_\star$.
+Therefore, we explore 1) the ability of BIKER to reproduce $CO_2$ evasion fluxes ($FCO_2$ $[M/L^2T]$) from these 49 rivers, and 2) the sensitivity of the carbon efflux $[M/T]$ due to the chosen method to approximate $R_h$ and ultimately $U_\star$.
 
@@
 To do this, we pair the biweekly 26 $CO_2$ samples (section 2) with every 11th SWOT observation by date, ignoring the timesteps beyond 26 (only ~15% of the SWOT observations were ignored here and we deemed this acceptable).
-We chose to sample every 11 days as this is the average sampling resolution for SWOT and the $CO_2$ data is only every 14 days.
+We chose to sample every 11 days as this is the average sampling resolution for SWOT and the $CO_2$ data is approximately the same at a 14 day resolution.
 Not all of the SWOT rivers have observations for a full year, and when simulation dates were not available they were assumed to start on January 1st.
@@
 In equation 9, 530 is the Schmidt number for oxygen at 20 degrees Celsius (which is the temperature at which BIKER predicts *k*) and *Sc* is the Schmidt number at one's desired temperature for some dissolved gas.
+This was calculated following @raymondScalingGasTransfer2012a and @wanninkhofRelationshipWindSpeed1992a.
 Atmospheric $CO_2$ was assumed 390 uatm.
-The Schmidt number, used in equation 9, was calculated following @raymondScalingGasTransfer2012a and @wanninkhofRelationshipWindSpeed1992a.
 $FCO_2$ validation was performed using the same metrics as $k_{o_2}$ validation (Table 1).
@@
 Next, we estimate bulk carbon efflux using five models for average channel depth (used to calculate $U_\star$):  BIKER's posterior means and four gauge-based HG models previously used for $CO_2$ upscaling [@horgbyUnexpectedLargeEvasion2019; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a; @raymondScalingGasTransfer2012a].
-See  for their definitions.
+See Table S1 for their definitions.
 It is worth stressing that the @lauerwaldSpatialPatternsCO22015 model is actually one of two components of the @raymondGlobalCarbonDioxide2013a model, and so was developed by the latter's authors.

### 2022-06-08 docs/intial_drafts/manuscript_v5.Rmd
first round of review, setting up for rerun on HPC

---
title: "Gas exchange in large rivers influenced by multiple turbulence scales: implications for remotely sensing gas exchange via the SWOT satellite"
author:
  - Craig B Brinkerhoff:
      email: cbrinkerhoff@umass.edu
      institute: UMass
      correspondence: true
  - Colin J Gleason:
      institute: UMass
  - Peter A Raymond:
      institute: Yale
  - Christopher J Zappa:
      institute: Lamont
  - Merritt H Harlan:
      institute: UMass
institute:
  - UMass: Department of Civil & Environmental Engineering, University of Massachusetts-Amherst
  - Yale: School of the Environment, Yale University
  - Lamont: Lamont-Doherty Earth Observatory, Columbia University
bibliography: References.bib
csl: agu_citation_style.csl
output:
  word_document:
      pandoc_args:
        - '--lua-filter=scholarly-metadata.lua'
        - '--lua-filter=author-info-blocks.lua'
      reference_docx: word-styles-reference.docx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Key Points
- Multiple scales of turbulence influence gas exchange in the large rivers that the SWOT satellite will measure
- BIKER algorithm exploits this theory, predicting gas exchange velocity and $CO_2$ fluxes from simulated SWOT data and in situ $CO_2$ data
- BIKER is also robust to measurement errors implicit in SWOT river measurements

## Keywords
gas exchange, fluvial geomorphology, remote sensing, open-channel flow, SWOT, biogeochemistry, ungauged basin

## Abstract
Extensive research over the past two decades has shown that the global river network emits significant amounts of greenhouse gases to the atmosphere via evasion. Despite much progress, there is still significant uncertainty in the temporal dynamics of gas exchange velocity (the rate of this evasion), with most models relying on temporally-coarse estimates of this velocity. We propose that the NASA/CNES/USKA/CSA SWOT satellite, set to usher in a new age of river hydrology research at the global scale, will provide similarly transformative insights to fluvial gas exchange modeling. Here, we present work inferring gas exchange velocity from SWOT observations. We exploit the unique geomorphology of SWOT-observable rivers to develop a physical model that is nearly entirely remotely sensible. We then couple this model with established Bayesian techniques to develop BIKER, or the 'Bayesian Inference of the $k_{600}$ Exchange Rate' algorithm. We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers and 763 discrete measurements of river gas exchange, yielding 1) a theoretical model that explains `r round(modelsWIDE[modelsWIDE$name == 'Chainsaw-eD',]$r2,2)*100`% of variation in gas exchange in SWOT-like rivers and 2) an algorithm that predicts this theoretical model solely from SWOT observations with a by-river median normalized RMSE of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$NRMSE), 2)*100`%. BIKER is most robust at reproducing the temporal dynamics of gas exchange (by-river median correlation coefficient of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$r), 2)`), rather than estimating the absolute rates of exchange. Finally, it is robust to measurement errors implicit in the SWOT data. With SWOT set to launch in 2022, we suggest that BIKER will prove a fundamental asset for exploring global-scale, spatiotemporal dynamics of fluvial gas exchange.

## 1 Introduction
Natural systems play a critical role in the budgeting and accounting of the global carbon cycle under climate change. Since the publication of @colePlumbingGlobalCarbon2007, the global river network is recognized to emit substantial amounts of carbon to the atmosphere, in addition to exporting it to the oceans. Current estimates of total carbon dioxide evasion ($FCO_2$) to the atmosphere from the global river network vary from 650-1800 Tg C/yr [@lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013] (*Liu etal in review- add number is paper is accepted in time*), with 167 Tg-C/yr coming from mountain streams alone [@horgbyUnexpectedLargeEvasion2019]. Despite its incredibly small percentage of the global land surface [0.`r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])`%- @raymondGlobalCarbonDioxide2013], this flux is on par with the total oceanic $CO_2$ uptake rate [@gruberOceanicSinkAnthropogenic2019; @horgbyUnexpectedLargeEvasion2019] and the global forest carbon uptake rate [@panLargePersistentCarbon2011]. It is still relatively poorly constrained (as evident by these studies' reported uncertainties) but is clearly a critical component of the global carbon cycle. Equation 1 represents this flux given the gas concentration gradient between the water $CO_{2_{water}}$ and the air $CO_{2_{air}}$ and the gas exchange velocity *k*. Consult Appendix A for variable nomenclature used throughout this study.

$$FCO_2=k(CO_{2_{water}}-CO_{2_{air}}) \mathbf{(1)}$$

The structure of equation 1 necessitates that calculations of this flux are highly sensitive to one's measurements/estimates of *k* [@hallGasExchangeStreams2020]. Even beyond $CO_2$ emissions, fluvial *k* for various dissolved gases are of significant importance to aquatic ecologists modeling stream metabolism [e.g. @bernhardtMetabolicRegimesFlowing2018] and water quality engineers modeling river responses to pollutant loadings [among other things- @chapraSurfaceWaterQualityModeling2008]. Because of this far-reaching importance, decades of work have focused on elucidating the physical mechanisms behind *k* in oceans and lakes [@wanninkhofAdvancesQuantifyingAirSea2009 for a review] and, to a lesser extent, fluvial gas exchange [@hallGasExchangeStreams2020 for a review].

Given that fluvial *k* should scale with near-surface turbulence in a turbulent flow [@hallGasExchangeStreams2020], extensive field and laboratory experiments have converged on the 'small-eddy model' as first proposed by @lamontEddyCellModel1970 which scales *k* via the smallest-scale turbulent eddies [e.g. @lorkeUnifiedScalingRelation2006; @zappaEnvironmentalTurbulentMixing2007; @zappaVariationSurfaceTurbulence2003; @moogStreamReaerationNonuniform1999; @vachonRelationshipNearsurfaceTurbulence2010; @tokoroHighGastransferVelocity2008; @katulStructureFunctionModel2018]. This model is provided as equation 2, where $\epsilon$ is the dissipation rate of near-surface turbulence, $v$ is the kinematic viscosity, and *Sc* is the Schmidt number. To compare across gases and water temperatures, *k* is often reported as $k_{600}$, where *k* is normalized to a Schmidt number of 600 [reflecting a $k_{CO_2}$ at 20 degrees Celsius- @hallGasExchangeStreams2020]. Note that directly measuring $\epsilon$ is non-trivial; therefore, a frequently used model for $\epsilon$ assumes that all turbulence is generated at the bed and transported to the air-water interface via the log-law-of-the-wall [@lorkeUnifiedScalingRelation2006; @nezuTurbulenceOpenChannel1993].

$$k\propto Sc^{-1/2}(v\epsilon)^{1/4} \mathbf{(2)}$$

While this model works reasonably well in non-fluvial environments where turbulence production is mostly limited to bottom shear [e.g. @lorkeUnifiedScalingRelation2006; @zappaEnvironmentalTurbulentMixing2007], there is considerable uncertainty in how it applies to fluvial systems. In particular, there is uncertainty as to which models are most appropriate at different locations along the stream-to-river continuum given that river hydraulics change predictably along a river's longitudinal profile [@wangPhysicallyBasedScaling2021]. In small and steep headwaters, 'bubble-mediated gas exchange' in whitewater might lead to substantially elevated *k* where bedload is large and general channel roughness is extremely high [@mauriceInfluenceFlowBed2017; @ulsethDistinctAirWater2019]. At the other end of the stream-to-river continuum (the largest rivers draining the continents), relatively little attention has been paid to gas exchange. The handful of existing field studies of large-river *k* suggest that *k* begins to be influenced by wind shear once the water surface is less protected, though little else is well-established [@alinPhysicalControlsCarbon2011; @beaulieuControlsGasTransfer2012; @wangPhysicallyBasedScaling2021]. Other workers have noted that laboratory and field observations suggest small-roughness open channel flows (generally big rivers) do not exhibit homogeneous surface dissipation at their air-water interface [@moogAirWaterGasTransfer1999; @talkeTurbulentKineticEnergy2013]. @moogAirWaterGasTransfer1999 proposed a conceptual 'chainsaw model', where large, depth-scale eddies transport small eddies to the interface, but the small eddies are responsible for the actual gas evasion that occurs at this interface. Finally, another approach proposed regardless of one's location along the stream-to-river continuum is to model *k* via 'form-drag shear' (i.e. total stream power per unit mass water). This normalized stream power captures the bulk frictional resistance via channel banks, meanders, bars, etc. that is unique to fluvial systems [@moogStreamReaerationNonuniform1999]. Authors have since shown that this model can reasonably predict *k* in rivers [@raymondScalingGasTransfer2012; @ulsethDistinctAirWater2019; @wangPhysicallyBasedScaling2021], however it requires estimates of the mean flow velocity and therefore can introduce additional uncertainties.

Taken in aggregate, our theoretical understanding of fluvial gas exchange is likely limited by a dearth of field-measured fluvial *k*, particularly in large rivers where direct measurements are difficult. @wangPhysicallyBasedScaling2021 attempted to address this by simulating *k* in 35 reasonably sized rivers (widths ranging from 0.23–349m) using a stream metabolism model [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets to implicitly solve for *k*. They found that equation 2, coupled with $\epsilon$ modeled via the log-law-of-the-wall, is valid in their simulated rivers and that *k*~streamflow relationships break down in large rivers. However, their simulated rivers were limited by a lack of direct hydraulic measurements and had to rely on models to estimate river depth and velocity, introducing additional uncertainty. Further, they did not assess model fit in very steep systems [@ulsethDistinctAirWater2019] and stopped short of parsing out hydraulic explanations for their results.

All of these mechanistic uncertainties are then propagated through upscaling workflows when biogeochemists predict *k* across thousands of rivers [e.g @borgesGloballySignificantGreenhousegas2015; @horgbyUnexpectedLargeEvasion2019; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013] via equation 1 coupled with hydraulic geometry (HG). HG is the scaling relationships between streamflow and river channel hydraulics [@leopoldHydraulicGeometryStream1953; @gleasonHydraulicGeometryNatural2015] that allow biogeochemists to predict river depth and velocity from streamflow models and/or streamgauges . It is currently not well understood how sensitive global estimates of fluvial gas evasion are to the specific HG model that is employed by the worker. Further, these approaches rely on either in situ discharge records or modeled streamflow which further contributes uncertainty. This is all exacerbated in ungauged basins that cover large areas, especially in the carbon-rich Arctic inland waters, where little in situ information is available and fieldwork is impractical [@gleasonRemoteSensingRiver2020].

A potential alternative to this upscaling approach is to directly estimate a river's hydraulic properties (and thus *k*) from remote sensing (RS) data. Remote sensing of river hydraulics is a burgeoning subfield within remote sensing of hydrology, often in service of remote sensing of river discharge [RSQ- @gleasonRemoteSensingRiver2020]. This is accomplished via two general approaches: 'gauged' methods which rely on in situ river data to calibrate one's method to the river(s) at hand and 'ungauged' techniques which focus on hydraulic generalizability in the service of merely improving existing knowledge in data-poor domains [@gleasonRemoteSensingRiver2020]. Many, but not all, of these ungauged approaches are developed in the context of the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission. SWOT is expected to launch in 2022 and provide the world's first global measurements of water surface extent and elevation at novel temporal resolutions. SWOT is a wide swath radar interferometer and will sample rivers every 1 to 7 days per 21 day repeat cycle and will measure rivers wider than 100m with a goal of expanding this to rivers at least 50m wide [@biancamariaSWOTMissionIts2016]. A decade of SWOT work has explored the multi-parameter problem of estimating the river hydraulic parameters of roughness and bathymetry from remote sensing to produce the SWOT discharge product [e.g. @andreadisConstrainingAssimilationSWOT2020; @brinkerhoffConstrainingRemoteRiver2020; @durandEstimatingReachaveragedDischarge2014; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014a; @garamboisVariationalEstimationEffective2020; @brissetAssimilationAltimetricData2018; @hagemannBAMBayesianAMHGManning2017; @larnierRiverDischargeBathymetry2020a; @oubanasDischargeEstimationUngauged2018].

Once SWOT launches, there will be a massive trove of high-temporal-resolution hydraulics measurements of the world's largest river surfaces, as well as established techniques to infer streamflow from those measurements. Given the uncertainties and limited field measurements of river *k* as described above, the SWOT measurements will hopefully also prove transformative in elucidating the spatiotemporal dynamics of gas exchange. However, in order to make sense of SWOT data for these purposes, a method must first be developed that can ingest SWOT data and return reasonable estimates of the gas exchange velocity solely from those data. In this context, we borrow tools from fluvial geomorphology and the SWOT discharge algorithms to answer the following questions: 1) can we leverage the unique geomorphology of the large rivers that SWOT will observe to develop a physically-based gas evasion model unique to these rivers? And 2) can we then exploit the findings from this analysis to develop a novel methodology that predicts this physically-based $k_{600}$ model (and its explicit uncertainty) solely using SWOT observations?

We name the RS of $k_{600}$ algorithm BIKER, or the 'Bayesian Inference of the $k_{600}$ Evasion Rate' and it requires no in situ inputs of any kind (although in situ data can be ingested and will improve results). We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-observable rivers from around the world using hydraulic models to produce SWOT-like data (as SWOT has not yet launched). We also quantify BIKER's sensitivity to the expected SWOT measurement errors on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` of those rivers: while SWOT data represent a sea change in inland water monitoring, it is expected to have an approximately 10cm error in water surface elevation [@biancamariaSWOTMissionIts2016] as well as river width errors [@frassonExploringFactorsControlling2021]. Finally, we use previously published dissolved $CO_2$ data to represent a hypothetical in situ $CO_2$ sensor and compare the bulk carbon efflux from the `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers as calculated using BIKER and previously published in situ techniques for predicting $k_{600}$ from streamflow.

## 2 Paper overview
Following the description of the data used (section 3), this paper is split into two distinct parts: gas-exchange model development and validation (section 4) and BIKER algorithm development and validation (section 5). Each section is further split into both methods and results sections, with section 5 fundamentally dependent on section 4. We conclude with a Discussion regarding both sections 4 and 5 (section 6). We also provide a flowchart detailing the entire study as Figure 1.

![Figure 1: Flowchart detailing entire study. Dashed lines indicate workflow to calculate the observed data that we validate against. See Appendix A for variable nomenclature used in this study.](cache/flowchart.jpg)

## 3 Data
Numerous datasets were used in this study. Please see Figure 2 for a map of the approximate locations for the data used in this study.

![Figure 2: Map of the `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` hydraulic models and 1 timeseries of CO2 samples used in this study. Note that hydraulic model locations are approximate as some of the models are not geo-referenced. Not mapped here, but also used, are over 170,000 discrete measurements of river channel hydraulics from across the continental United States (Brinkerhoff et al. 2019) and 763 gas exchange velocity measurements (Table S1).](cache/maps/map_fin.jpg)

Gas exchange theory (section 4) is explored using multiple datasets where $k_{600}$ was measured directly and/or collected from the literature. In total, there are 763 measurements of stream hydraulics and $k_{600}$. This is, to our knowledge, the largest collection of field-measured river and stream $k_{600}$. 701 of these measurements were previously collected from other studies by @raymondScalingGasTransfer2012 and @ulsethDistinctAirWater2019. The remaining measurements were added by us but measured by @churchillPredictionStreamReaeration1964 and @owensReaerationStudiesStreams1964. See Table S1 for a complete list of the studies that collected these measurements.

To quantify our geomorphic assumptions for SWOT rivers (section 4.1), we additionally use a previously published compilation of field hydraulics measurements [@brinkerhoffReconcilingAtaStationAtManyStations2019]. These were originally made to calibrate United States Geological Survey (USGS) streamgauge rating curves. That dataset contains over 530,000 unique measurements of river channel velocity, width, and discharge from across the continental United States. Ultimately, this dataset was filtered down to `r sum(percs$n)` discrete measurements for this study (see Text S1 for our filtering protocol).

BIKER validation (section 4) was performed on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers. Because SWOT has yet to launch, it is standard practice to benchmark SWOT-related algorithms on "SWOT-like" data. These simulated rivers are simply river-reach-averaged hydraulic model outputs where the water surface heights, slopes, and widths are labelled as RS observations and are used as the sole inputs to BIKER. Here, we use `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])`/51 rivers collected by @frassonExploringFactorsControlling2021 and @durandIntercomparisonRemoteSensing2016. These are the two benchmarking studies that have explored RSQ algorithm performance for the SWOT mission. The approximate locations of these rivers are plotted in Figure 1, and please consult both of those papers for all of the hydraulic model specifications. Ultimately, the `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers are spread across the United States, France, Italy, the United Kingdom, and Bangladesh. We omit three models from @durandIntercomparisonRemoteSensing2016 because they lack enough hydraulic information to calculate the mean flow depth, which is necessary for algorithm validation (section 5.1.2). These are two models for different sections of the Saint Lawrence River in Canada and one for the Tanana River in Alaska, United States.

To assess the influence of SWOT's implicit measurement error on BIKER performance, we use the error model developed by @durandHowWillRadar2020 and implemented on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])`/`r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` of the rivers by @frassonExploringFactorsControlling2021. Error in SWOT measurements will come from both the error tolerances intrinsic in the satellite data product as well as radar layover error. Layover error is the phenomenon when radar returns from different places arrive at the sensor at the same time, leading to taller landscape features appearing closer to the sensor than shorter landscape features that are the same horizontal distance from the sensor [@durandHowWillRadar2020]. Width errors due to poor water classification are ignored as they were in @frassonExploringFactorsControlling2021. @frassonExploringFactorsControlling2021 tested a suite of measurement error scenarios, however here we test only their 'full uncertainty scenario' that most closely represents the measurement errors likely to be implicit in SWOT data.

For the $CO_2$ evasion and carbon efflux calculations (section 5.1.3), we use 26 bi-weekly dissolved $CO_2$ samples made by @beaulieuControlsGasTransfer2012 at one location in the Ohio River for one calender year from 2008-2009 (Figure S1). Note that this $CO_2$ data is for the Ohio River only but was applied to all rivers (which includes multiple sections of the Ohio River). Because we are exclusively interested in the relative differences between $FCO_2$ estimates and not the raw fluxes themselves, any $CO_2$ data representative of SWOT-observable rivers was deemed acceptable for this validation exercise.

## 4 Hydraulic geometry in large rivers and predicting fluvial gas exchange velocity
Given that *k* can only be directly measured via a known gas concentration gradient, eddy-covariance measurements, or tracer additions to the stream [@hallGasExchangeStreams2020], SWOT data will not provide the necessary parameters to directly measure *k*. Further, these techniques are entirely infeasible at the global scale where our goal with BIKER is primarily to better understand spatiotemporal patterns of *k* rather than obtain absolute measurements of it. Therefore, we need a simple yet physically-based model for *k* to directly relate channel hydraulics to gas exchange in the large rivers that SWOT will measure. We hypothesize that this can be done by leveraging the unique geomorphology of the very wide rivers that SWOT will observe. In this section, we develop and test process-based models for *k*, selecting one that is then implemented within BIKER (section 5).

### 4.1 Section 4 methods
Rivers and streams change predictably along their longitudinal profile from headwater to ocean. In general, as river size increases, channels become more rectangular, their shapes elongate (becoming wider quicker than they become deeper) and their hydraulic radii begin to approximate their mean flow depth, i.e. $R_h \approx H$ [@leopoldHydraulicGeometryStream1953]. This is a common assumption in hydraulic and geomorphic modeling of large rivers, and occurs in `r round(percs[percs$flag_swot == 'SWOT' & percs$flag_hydraulicWide == 'Rh=H',]$perc,0)`% of our SWOT-observable river measurements but only `r round(percs_overall[percs_overall$flag_hydraulicWide == 'Rh=H',]$perc,0)`% of non-SWOT observable river measurements (see Text S1 for how we arrived at these numbers). These channels are often referred to as 'hydraulically-wide'.

Assuming that all SWOT-observable rivers are hydraulically-wide, we test the performance of four models for predicting *k* in hydraulically-wide channels. We test two conceptual models for gas evasion (both descrbied in the Introduction): the 'small-eddy' model [@lamontEddyCellModel1970- section 4.1.1] and the 'chainsaw' model [@moogAirWaterGasTransfer1999- section 4.1.2]. We also test two different models for the turbulent dissipation rate (also described in the Introduction): the log-law-of-the-wall ($\epsilon=\epsilon_S$) and form-drag dissipation [@moogStreamReaerationNonuniform1999] ($\epsilon=\epsilon_D$). While the complete model derivations for sections 4.1.1 and 4.1.2 are provided in Text S2, the broad goal was to reduce the equations down to their fundamental parameters, identifying which terms are SWOT observable and limiting the number of terms not directly measurable via SWOT.

Next, all models are empirically tested on the dataset described in Section 3 after filtering for measurements made in hydraulically-wide channels (section 4.2). This was defined here as those whose hydraulic radius was within 1% of their mean flow depth. Note that, to account for different water temperatures and gases used in measuring *k*, all data are normalized to a Schmidt number of 600 (i.e. $k_{600}$), as is standard practice. All told, this amounts to 166 direct measurements of hydraulically-wide $k_{600}$ to test with.

#### 4.1.1 'Small-eddy' model for gas evasion
First, we derive equations starting from the small-eddy model (equation 2) for hydraulically-wide channels. If we assume the log-law-of-the-wall model for $\epsilon$ (see Appendix A for its calculation), simplify equation 2 using $R_H = H$, and finally solve for $k_{600}$ via statistical parameter $\alpha$, we eventually arrive at equation 3. If we do the same but via the form-drag model for dissipation and statistical parameter $\alpha_1$, we arrive at equation 4. $\bar{U}$ refers to mean flow velocity.

$$k_{600}=\alpha(gS)^{3/8}H^{1/8} \mathbf{(3)}$$

$$k_{600}=\alpha_1(gS\bar{U})^{1/4} \mathbf{(4)}$$

Equations 3 and 4 thus define gas exchange velocity solely as a function of water surface slope and mean flow depth or mean flow velocity. This is theoretically valid only in a hydraulically-wide channel.

#### 4.1.2 'Chainsaw' model for gas evasion
As mentioned in the Introduction, @moogAirWaterGasTransfer1999 noted that laboratory observations suggest small-roughness open-channel flows do not exhibit homogeneous surface dissipation at their air-water interfaces. Under this guise, their chainsaw model extends the small eddy model by limiting surface dissipation to only 'active zones' within which a homogeneous dissipation rate occurs and the small-eddy model is valid. The fraction of the interface that are active zones is proportional to the larger, depth-scale eddies via the so-called 'Reynolds shear number'. This means that if the entire interface is treated as an active zone, the model would simply reduce to the small-eddy model. This is hypothetically useful in low-turbulence flows where a relative lack of large-scale eddies effectively 'filter out' the number of small-eddies that actually reach the interface and lead to gas exchange [@talkeTurbulentKineticEnergy2013].

In the context of BIKER, we chose to test this model because large, SWOT-observable rivers are generally the smoothest, least-turbulent flows along the stream-to-river continuum. Further, to our knowledge the chainsaw model has never been empirically tested in predicting *k*, aside from confirming that large-scale eddies differentially move turbulence to the surface in a hydraulically-wide river [@talkeTurbulentKineticEnergy2013]. To do this, we impose a hydraulically-wide channel on the chainsaw model just like we did in section 4.1.1 and arrive at equation 5 (with statistical coefficient $\beta$). The result is identical to the hydraulically-wide small-eddy model (equation 3), except for the exponents which are both steeper. Equation 6 reflects the same derivation, but via the form-drag model for dissipation and statistical coefficient $\beta_1$.

$$k_{600}=\beta(gS)^{9/16}H^{11/16} \mathbf{(5)}$$
$$k_{600}=\beta_1(gS)^{7/16}\bar{U}^{1/4}H^{9/16} \mathbf{(6)}$$

We now have four different models for $k_{600}$ in hydraulically-wide rivers, each with a different set of assumptions and parameters but all defined as functions of the water surface slope and other hydraulic terms that are not directly measurable via SWOT. All four models are summarized in Table 1.

*Table 1: Four physically-based models for $k_{600}$ each derived under a different set of mass transport assumptions but all valid in hydraulically-wide rivers. Their derivations are outlined in sections 4.1.1 and 4.1.2 and their empirical tests are shown in section 4.2.*

|**Description**|**Equation**|**References**|**Empirical test**|**Empirical fit via $r^2$**|
|---------------|------------|--------------|------------------|-----------------|
|Small-eddy, log-law-of-the-wall|$k_{600}=\alpha(gs)^{3/8}H^{1/8}$|@lamontEddyCellModel1970; log-law-of-the-wall|Figure 3a|`r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eS',]$r2, 2)`|
|Small-eddy, form-drag dissipation|$k_{600}=\alpha_1(gS\bar{U})^{1/4}$|@lamontEddyCellModel1970; @moogStreamReaerationNonuniform1999|Figure 3c|`r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eD',]$r2, 2)`|
|Chainsaw, log-law-of-the-wall|$k_{600}=\beta(gs)^{9/16}H^{11/16}$|@moogAirWaterGasTransfer1999; log-law-of-the-wall|Figure 3b|`r round(modelsWIDE[modelsWIDE$name == 'Chainsaw-eS',]$r2, 2)`|
|Chainsaw, form-drag dissipation|$k_{600}=\beta_1(gs)^{7/16}\bar{U}^{1/4}H^{9/16}$|@moogAirWaterGasTransfer1999; @moogStreamReaerationNonuniform1999|Figure 3d|`r round(modelsWIDE[modelsWIDE$name == 'Chainsaw-eD',]$r2, 2)`|

### 4.2 Section 3 results
With equations 3-6 defined, we now empirically test their strength of fit for hydraulically-wide channels via the coefficient of determination ($r^2$). These results are plotted in Figure 3. Note that Figure 3 axes are plotted in logarithmic space just for visualization: model fit and validation (via $r^2$) were calculated in linear space as their models dictate.

![Figure 3: Empirical testing of four models for predicting gas exchange velocity in hydraulically-wide channels: (a) Small-eddy model with log-law-of-the-wall dissipation. (b) Chainsaw model with log-law-of-the-wall dissipation. (c) Small-eddy model with form-drag dissipation. (d) Chainsaw model with form-drag dissipation. Note that axes are plotted in logarithmic space just for visualization: model fit and validation were calculated in linear space.](cache/k600_theory/k600Plot.jpg)

Despite reasonable $r^2$ scores, both small-eddy models produce clearly biased estimates (Figures 3a, 3c). They systematically underestimate lower gas exchange velocities and overestimate higher ones, with the log-law-of-the-wall model (Figure 3a) performing marginally worse both visually and via $r^2$ (`r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eS',]$r2, 2)` versus `r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eD',]$r2, 2)` for Figure 3c). However, when this model is expanded to account for a differential dissipation rate at the interface via the chainsaw model [equations 5 and 6- @moogAirWaterGasTransfer1999], the dynamics are more accurately captured (Figures 3b, 3d) and $r^2$ scores go up appreciably (`r round(modelsWIDE[modelsWIDE$name == 'Chainsaw-eS',]$r2, 2)` and `r round(modelsWIDE[modelsWIDE$name == 'Chainsaw-eD',]$r2, 2)` for the log-law-of-the-wall and form-drag dissipation, respectively). Modeling dissipation via total form-drag rather than just via bed shear and the log-law-of-the-wall improves model performance. This is likely because the majority of dissipation occurring in rivers comes not just from the streambed, but from bulk form drag via channel walls, bars, and meanders [@raymondScalingGasTransfer2012; @moogStreamReaerationNonuniform1999; @wangPhysicallyBasedScaling2021; @ulsethDistinctAirWater2019].

Given the algebraic structure of the small-eddy and chainsaw models, the performance differences observed in Figure 3 are largely due to the exponents applied to the river hydraulic terms (which are described in the subplot x axis labels). While the use of the form-drag dissipation model includes a mean flow velocity term and improves performance, the primary difference between the small-eddy and chainsaw equations is the exponents. Intuitively, this exponent discrepancy indicates that in hydraulically-wide rivers where turbulence is relatively low, large-scale eddies exert a noteworthy influence on surface dissipation rates and that the small-eddy model is not sufficient. It is also worth noting that this is, to our knowledge, the first empirical confirmation that the chainsaw model is valid in rivers (at least hydraulically-wide ones). This is elaborated on in section 6.1.

The success of the chainsaw model in hydraulically-wide channels, and in particular when coupled with the form-drag dissipation model (equation 6, Figure 3d), provides us with a strong physical-model for gas evasion. The river hydraulics terms in equation 6 can either be directly measured or reasonably inferred from SWOT measurements, effectively opening the door for remotely sensing gas exchange. This is explored next.

## 5 Exploiting equation 6 to remotely sense gas exchange velocity
We have shown that scaling $k_{600}$ via equation 6 explains `r round(modelsWIDE[modelsWIDE$name == 'Chainsaw-eD',]$r2,2)*100`% of the variation in $k_{600}$ in hydraulically-wide rivers, which includes nearly all SWOT-observable rivers. Further, equation 6 has only three non-remotely-sensible terms: $k_{600}$, mean flow depth, and mean flow velocity. Conveniently, inferring mean flow velocity and mean flow depth from SWOT measurements is a well-established domain for workers estimating streamflow from SWOT. For BIKER, we follow the work developed by @durandEstimatingReachaveragedDischarge2014, @hagemannBAMBayesianAMHGManning2017, and @brinkerhoffConstrainingRemoteRiver2020 to infer mean flow velocity and depth from SWOT observations using a modified form of the Manning's equation for open-channel flow velocity. These papers conceptualize streamflow as a Bayesian remote sensing problem, which we largely follow here to conceptualize $k_{600}$ as a Bayesian remote sensing problem that can be solved using SWOT data. First, we detail algorithm development and experimental design (section 5.1) and then we present the validation results (section 5.2).

### 5.1 Section 5 methods
#### 5.1.1 BIKER
BIKER, and Bayesian inference in general, starts from Bayes rule (equation 7), where $\Theta$ is some set of non-remotely-sensible parameters we want to solve for (including $k_{600}$), *x* is the observed data, $f(x|\Theta)$ is the sampling model where data are conditional on the parameters, and $p(\Theta)$ is the joint prior distribution of the parameters. For BIKER, *x* is the SWOT-observables: river width *W* and water surface elevation $H_e$ (which is used to calculate the water surface slope *S*), while $\Theta$ are the parameters $k_{600}$, *n*, and $A_0$. Therefore, we are interested in solving for $p(\Theta|x)$, or the 'posterior' distribution. Note that $p(x)$ is usually computationally intractable to integrate exactly, but Bayesian inference tools require only the proportionality to be specified: $p(\Theta|x) \propto f(x|\Theta)p(x)$. Sampling algorithms are then used to approximate the posterior distribution, as is done in BIKER.

$$p(\Theta|x) = \frac{f(x|\Theta)p(\Theta)}{p(x)} \mathbf{(7)}$$

The heart of BIKER is its reformulation of the $k_{600}$ model (equation 6) as a Bayesian sampling model that is conditional on the non-remotely-sensible parameters (i.e. $f(x|\Theta)$). To do this, we first conceptualize a mass-conserved river reach, where hydraulic properties vary along cross-sections *i* within the reach but the entire mass of water moving through is conserved and thus only varies with timestep *t* [@gleasonTrackingRiverFlows2017]. Consult Figure S2 for a scematic of this model. This means that the reach-averaged $k_{600}$, which is to some degree a function of the mass-conserved streamflow [@wangPhysicallyBasedScaling2021], will only vary temporally as well. Under this conceptual river reach model, BIKER predicts a time-varying, reach-averaged $k_{600_t}$. Despite recent discussion on the importance of within-reach heterogeneity of *k* [@botterHeterogeneityMattersAggregation2021], reach-averaged values are suitable for the global scale where a coarser resolution must be adopted to facilitate computation.

Under this conceptual river reach, we write $k_{600_t}$ as a function of SWOT-observables $W_{i,t}$ and $S_{i,t}$. This algebra is carried out using equation 6, the fitted value for $\beta_1$ from Figure 3d (`r round(modelsWIDE[modelsWIDE$name == 'Chainsaw-eD',]$coef,2)`), and Manning's equation for mean flow velocity ($\bar{U}=\frac{1}{n}R_h^{2/3}S^{1/2}$). This algebra yields equation 8, assuming that the channel is hydraulically-wide ($R_h = H =\frac{A}{W}$). To leverage additional SWOT data, the wetted channel area *A* is further split into the SWOT-observable portion $dA_{i,t}$ and the unobservable portion $A_0_i$ following @durandEstimatingReachaveragedDischarge2014. *dA* is also readily approximated using SWOT measurements if assuming a rectangular river channel ($dA_{i,t} \approx \delta H_e_{i,t}*W_{i,t}$), which is again valid in a hydraulically-wide channel.

$$\begin{eqnarray} k_{600_t} = 62.82 g^{7/16}S_{i,t}^{9/16}\frac{1}{n_i}^{1/4}(\frac{A_0_i+dA_{i,t}}{W_{i,t}})^{35/48} \end{eqnarray} \mathbf{(8)}$$

Next, equation 8 is written as a Bayesian sampling model, in which all of the SWOT observations are sampled from the unknown model parameters ($A_0_i$, $k_{600_t}$, and $n_i$). This is equation 9 after describing them as log-normal distributions. $\sigma_{k_{600}}^2$ refers to the total uncertainty implicit in equation 8. This uncertainty is due to parameter uncertainty from both the chainsaw/form-drag dissipation model and the Manning's model for $\bar{U}$.

$$\left[log(62.82) + \frac{7}{16}log(g) + \frac{9}{16}log(S_{i,t}) - \frac{35}{48}log(W_{i,t})\right] \sim normal\left(\left[log(k_{600_t})+\frac{1}{4}log(n_i)-\frac{35}{48}log(A_0_i+dA_{i,t})\right], \sigma^2_{k_{600}}\right) \mathbf{(9)}$$

Equations 7 and 9 also necessitate that we specify prior distributions for the parameters $A_0_i$, $k_{600_t}$, and $n_i$. Prior distributions, described by their hyperparameters, formalize the a priori estimates (and uncertainties) for the non-remotely-sensed terms. More intuitively, BIKER priors represent our 'prior river knowledge' of what $A_0_i$, $k_{600_t}$, and $n_i$ probably are for some river since they cannot be directly remotely sensed, but educated guesses can be made using the SWOT data available to us. This is sometimes referred to as the 'empirical Bayes method'. Our goal in prior specification was to rely on absolutely no in situ information such that we could run this method on any river on Earth solely using SWOT observations. In theory, more informed priors via various a priori information about a specific river would improve BIKER performance, but we wanted to test the fully generalized algorithm in this paper. Therefore, the validation presented here is a 'worst-case scenario', wherein BIKER performance will improve with better prior information on the river. In that context, we used a variation of the prior specification method developed by @brinkerhoffConstrainingRemoteRiver2020, who developed geomorphic 'river types' with distinct prior hyperparameter sets for $A_0_i$ and $n_i$. These hyperparameters are assigned to a river solely using SWOT observables, therefore meeting our needs for complete global implementability. Hyperparameter assignment for $k_{600_t}$ was developed similarly (all hyperparameter specifications are elaborated on in Texts S3 and S4).

With the sampling model (equation 9) and hyperparameters described (Text S3), a joint posterior distribution conditional on the SWOT observations ($p(\Theta|x) \propto f(x|\Theta)p(x)$) is necessarily specified. To approximate this distribution, we use a Markov Chain Monte Carlo (MCMC) algorithm implemented using the Stan probabilistic programming language. Specifically, Stan uses a Hamiltonian Monte Carlo sampler which reduces computation time relative to other sampling algorithms [@hagemannBAMBayesianAMHGManning2017].

#### 5.1.2 BIKER validation
We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers (section 2) using daily observed hydraulics and the metrics described in Table S2. We also re-validate BIKER on the `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` rivers with the SWOT error model which corrupts the SWOT measurements (widths, heights, and slopes).

Regardless of the validation setup, we do not have observed $k_{600_t}$ data for these rivers and, to our knowledge, no field dataset of $k_{600}$ exists in the type of temporal and spatial frequency that SWOT (and therefore the BIKER algorithm) provides. Thus, we take the model outlined in equation 6 and Figure 3d and use that to calculate the observed $k_{600_t}$ that BIKER is validated against (equation 10). Note that equation 9 and 10 both treat $\beta_1$ (62.82) as known and not a parameter in order to directly explore BIKER's ability to infer observed $A_0_i$, $k_{600_t}$, and $n_i$ simultaneously from $W_{i,t}$ and $H_{i,t}$ alone. This was deemed acceptable given that the $k_{600}$ model (equation 6) has already been successfully validated in Figure 3d.

$$k_{600_{obs}}=62.82(gS_{obs})^{7/16}\bar{U}_{obs}^{1/4}H_{obs}^{9/16} \mathbf{(10)}$$

It also means that, for a fair validation scheme, $\sigma^2_{k_{600}}$ in equation 9 must reflect only error from using Manning's equation and our calculation of *dA* and not the parameter uncertainty inherent in $\beta_1$. Because this model is only theoretically valid when $R_h=H$, we conveniently do not need to account for uncertainties associated with assuming that $R_h = H$. So for this validation, $\sigma^2_{k_{600}}$ is set to 0.30 (natural-log space) to reflect uncertainty from Manning's equation and assuming a rectangular channel (Text S4). However, if BIKER is run on real SWOT observations after the satellite launches, $\sigma^2_{k_{600}}$ should reflect the total uncertainty, which is described in Text S4. Note that when run on real SWOT observations, we could also treat $\beta_1$ as a parameter to solve for. This would complicate the model to a degree, but would also allow for a robust posterior estimate of $\beta_1$'s value. This is left to future work once SWOT launches.

With the setup described by equations 9 and 10, validation is performed using the BIKER posterior means. Because we are most interested in BIKER's ability to correctly infer temporal dynamics in $k_{600}$, and each river's priors are described specifically for that river, we validate by river. These metrics are defined in Table S2 and consist of the correlation coefficient *r* to quantify accuracy of BIKER's temporal dynamics, the root mean square error normalized by the observed mean (NRMSE) and prediction bias normalized by the observed mean (rBIAS) to assess bias, and the Kling-Gupta Efficiency (KGE). KGE is a commonly used metric to assess streamflow prediction and simultaneously assesses accuracy in predicting both bias and dynamics. While a value greater than -0.41 means the model outperforms a uniform prediction of the mean [@knobenTechnicalNoteInherent2019], generally KGE scores are interpreted as being meaningful in ungauged settings if > 0.

#### 5.1.3 Carbon efflux validation
It is one thing to accurately model the temporal dynamics of $k_{600}$, but researchers are often most interested in the actual gas fluxes from rivers and ultimately the carbon emitted from river to atmosphere. Therefore, we also explore BIKER's ability to reproduce $FCO_2$ (equation 1) and the bulk carbon efflux from these `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers. To do this, we pair the 26 biweekly $CO_2$ and water temperature samples from @beaulieuControlsGasTransfer2012 (section 2, figure S1) with every 14th set of daily SWOT observations (as the $CO_2$ data is bi-weekly). Conveniently, this is very close to the nominal average sampling resolution for SWOT (11 days). Not all of the SWOT rivers have observations for a full year, and when simulation dates are not available they are assumed to start on January 1st. We then calculate $FCO_2$ using equation 1 and assuming atmospheric $CO_{2_{air}}$ is 390 uatm. Equation 11 is used to obtain a temperature and $CO_2$ specific gas exchange velocity from BIKER's output, where the Schmidt number *Sc* was estimated following @raymondScalingGasTransfer2012. We then estimate median and mean bulk carbon effluxes by multiplying the median or mean $FCO_2$ value across all rivers and timesteps (in Tg-C/yr) by the total average surface area of all 47 rivers.

$$\mathbf{(11)} k_{CO2, temp}=k_{600}*(Sc/600)^{-1/2}$$

Finally, we also compare BIKER's outputs with established in situ methods, whose specific details and parameters are provided in Table S3. Because streamflow measurements and models are vastly more frequent than measurements of river channel geometry, these in situ models are used to predict reach-averaged flow depth and flow velocity as functions of streamflow. Their outputs are in turn used to estimate *k* in upscaling workflows. Therefore, they rely on in situ streamflow records or high-quality streamflow routing outputs (unlike BIKER, which only requires a river be SWOT-observable). This allows us to assess whether BIKER's estimates (wholly ungauged) are comparable to gauged methods (all three HG models). Of the three models tested here, one is trained on the @brinkerhoffReconcilingAtaStationAtManyStations2019 dataset, and two are previously published models [@raymondGlobalCarbonDioxide2013; @raymondScalingGasTransfer2012].

### 5.2 Section 5 results
#### 5.2.1 BIKER
First, we explore whether general trends are correctly captured by BIKER, assuming no SWOT measurement error. Figure 4 plots BIKER-inferred $k_{600}$ against the $k_{600}$ calculated using in situ hydraulics across all  rivers and all timesteps. The points are the posterior means while the black lines are the 95% confidence intervals (CIs) for the predictions. Points are also colored by river. Using absolutely no in situ information, BIKER reasonably captures the magnitude of the predictions and most points fall on or near the 1:1 line across all  `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers. However, it is clear that some rivers perform better than others. Next, we look at individual river predictions, which we are most interested in.

![Figure 4. Validation of BIKER. Black bars are 95% CIs for the modeled values. The dashed grey line is 1:1 line. Colors correspond to rivers.](cache/validation/validation.jpg)

Figure 5 plots all `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` timeseries plots of predicted and observed $k_{600}$, again assuming no SWOT measurement error. Note that the y axis is normalized by maximum observed values to compare rivers. In general, temporal dynamics of $k_{600}$ are reproduced quite well by BIKER, however there is often positive bias in the estimates (e.g. Olentangy River). Some rivers are predicted nearly perfectly (e.g. Brahmaputra, Ohio Section 3, Wabash) while others perform more poorly (e.g. San Joaquin, Stanislaus), though again much of this error stems from errors in predicting the magnitude of $k_{600}$ rather than the temporal dynamics. Rivers where the temporal dynamics are more poorly inferred tend to have shorter timeseries (e.g. Merced River, Tuolumne River). **Maybe add something here comparing the k600 prior to these results?**. Next, we explicitly quantify each rivers' performance.

![Figure 5. River timeseries plots of $k_{600}$. Orange is calculated via observed hydraulics, while green is BIKER. The green ribbon indicates the 95% CIs for the predicted values. Y axis is normalized by maximum observed values to compare visually.](cache/validation/timeseries_noerr.jpg)

We calculate performance metrics following section 5.1.2 and Table S2. These are presented in Figure 6a, which plots validation metrics calculated for each river with and without SWOT measurement error (green and purple, respectively). SWOT measurement errors negligibly influence BIKER's performance across all four error metrics (Figure 6a), though caution should be used in over-interpreting boxplots with a sample size of only `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])`. Given Figure 6a, we deem that SWOT measurement error does not exert a significant influence on BIKER and so the results presented for the rest of the manuscript assume no measurement error in order to use all `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers.

![Figure 6. a: Performance metrics by river. See Table S2 for metric definitions. b: *r* versus the number of SWOT observations used in the Bayesian inversion. c) *r* versus the mean river width. As expected, performance improves with more data. Dashed lines denote scores of 0, 0.50, and 1. CIs.](cache/validation/validation_by_river.jpg)

Median river-specific $KGE$ is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$KGE, na.rm=T),2)` and median river-specific *r* is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$r, na.rm=T),2)`. This indicates very strong inference of each river's temporal $k_{600}$ dynamics. These scores are excellent given that absolutely no in situ information is being used to predict $k_{600}$. NRMSE has a median score of `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$NRMSE, na.rm=T),2)`, which is reasonable for a completely ungauged method. These are comparable to the best NRMSE scores achieved by recent versions of the SWOT RSQ algorithms [@frassonExploringFactorsControlling2021]. Median rBIAS is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$rBIAS, na.rm=T),2)`, which is again excellent. However, the shape of the rBIAS boxplot in Figure 3a highlights a large portion of rivers which have a positive bias (Figure 5 also confirms this). Taken in aggregate, Figures 4-6 indicate that BIKER is quite good at capturing temporal dynamics in $k_600$, however there is often positive bias in its estimates. This necessarily increases the NRMSE scores to be relatively worse than *r* or KGE. This is discussed in more detail in section 6.

Finally, Figure 6b highlights one benefit of using Bayesian inference to estimate $k_{600}$: because the posterior is conditional on the SWOT observations, performance should improve with more data. Figure 6b plots by-river *r* scores versus the number of SWOT observations. While performance does not directly increase from having more data, very poor performances disappearas the amount of availble data increases. The vast majority of the extremely poorly performing rivers are limited to those with only 12 days of SWOT observations. This bodes well for future SWOT implementation, which has a nominal life-cycle of three years. This relationship also exists between mean observed river width and *r* (Figure 6c), where wider rivers have better BIKER performance. However, it is impossible to say whether this is a genuine trend or an artifact of the specific rivers being tested, where the larger rivers often have longer timeseries.

#### 5.2.2 Carbon efflux
Next, we compare BIKER's estimates of carbon fluxes against established in situ techniques. It is important to remember that only BIKER is completely globally implementable, while the other three models presented necessarily rely on having a streamflow record readily available. Therefore, the HG methods reflect their 'best-case scenarios' while BIKER reflects its worst case scenario, where priors are set entirely from SWOT observations and are generally the least informative they will ever be.

Figure 7a evaluates the $FCO_2$ estimates using the same 4 error metrics as Figure 6a. BIKER, the sole remote technique, is in red while the three in situ techniques are plotted as different shades of purple. We see that BIKER performs slightly worse than all three in situ methods across all metrics except *r*, where it performs similarly to the 'Raymond 2013' model. **NEED ACTUAL NUMBERS HERE**. However, across all 47 rivers BIKER has the least overall bias in its estimates (**add number here**). In summary, BIKER performs slightly worse than the in situ tools, as expected. However, the drop in performance is not substantial and promising given that the in situ models represent their best case scenarios (perfect in situ discharge records) and these BIKER results are the worst-case scenario (relatively uninformed priors).

![Figure 7: a) FCO2 validation for BIKER and the three in situ HG models. b) Mean carbon efflux across all rivers and timesteps, accounting for surface area. d) Median carbon efflux across all rivers and timesteps, accounting for surface area. Completely remotely-sensed methods are colored in red, in situ methods in purple, and the obserevd values in orange.](cache/FCO2/FCO2_models.jpg)

Figures 7b and 7c compares the bulk carbon efflux (via $F_{CO_2}$) from the `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers using BIKER posterior means and the three streamgauge-based HG models. For the means, all estimates are positively biased, with BIKER (`r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_BIKER',]$sumFCO2, 2)` gG-C/yr) and 'Raymond 2013' (`r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_raymond2013',]$sumFCO2, 2)` gG-C/yr) the most overestimated relative to the observed flux (`r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_obs',]$sumFCO2, 2)` gG-C/yr). However, this is evidently due to one or a few significantly biased outliers, as median fluxes are nearly perfectly captured by BIKER (`r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_BIKER',]$medianFCO2, 2)` gG-C/yr) and 'Raymond 2012' (`r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_raymond2012',]$medianFCO2, 2)` gG-C/yr) versus the observed flux (`r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_obs',]$medianFCO2, 2)` gG-C/yr). The two remaining models are still overestimated, though less so than for the mean flux. Thus, despite BIKER using absolutely no streamgauge or other in situ data like the other 3 methods do, it provides a reasonable upscaling estimate of the bulk carbon efflux across many rivers (Figure 6c). This is promising for BIKER's implementation within future upscaling workflows.

## 6 Discussion
### 6.1 Gas exchange in hydraulically-wide rivers
**REWRITE THIS**
Field studies of gas exchange in wide rivers have suggested that their *k* properties behave differently than in steeper and smaller rivers [@ulsethDistinctAirWater2019; @beaulieuControlsGasTransfer2012; @alinPhysicalControlsCarbon2011; @wangPhysicallyBasedScaling2021]. While much work has focused on the small-stream side of the stream-to-river continuum, @moogAirWaterGasTransfer1999 proposed a conceptual model for gas exchange in low turbulence, small roughness flows (which coincides with large rivers). Figure 3d is to our knowledge the first empirical confirmation that the chainsaw model is valid in rivers (at least hydraulically-wide ones). Future tests should be performed to explore it's feasibility in smaller streams and/or other aquatic environments where turbulence is also generally low. We are not accounting for dissipation generated via wind shear, as none of these hydraulics-driven gas exchange models do. We suspect that this model will be less representative in less-sheletered rivers where wind becomes a driving factor of gas exchange rates, however this model setup is a good start towards coupling hydraulics-driven gas exchange with wind-driven gas exchange. Such a model would likely take a form similar to the conceptual model proposed by @wangPhysicallyBasedScaling2021.

### 6.2 Towards remote sensing of global spatiotemporal dynamics of *k* in large rivers
To date, the studies exploring the spatiotemporal dynamics of riverine gas exchange have arguably been held back by a lack of data. A few studies have investigated these dynamics, but they have been limited to individual rivers and/or limited field seasons [@hallAirWaterOxygen2012; @sand-jensenCO2DynamicsDanish2012]. For example, @wallinSpatiotemporalVariabilityGas2011 performed a preliminary analysis in northern Sweden relating cross-section specific temporal variability in gas exchange with channel slope, but they were limited to a mean of only 8 measurements per river in a single watershed.

Therefore, estimating *k* from SWOT data is an attractive option, at least for exploring its spatiotemporal dynamics at fine temporal resolution and at the global-scale. Section 5 shows that BIKER successfully infers the temporal dynamics of $k_{600}$ using simulated SWOT data over a wide range of rivers and with implicit measurement errors accounted for (Figures 4-6). This bodes well for BIKER's eventual implementation on real SWOT observations, as does BIKER's improved performance with longer timeseries (Figure 6b). The nominal three-year SWOT campaign will provide hundreds of observations for most rivers, suggesting even better BIKER inversions than those presented here. All of this suggests that near-daily estimation of global riverine gas exchange and its temporal dynamics could be possible once SWOT launches.

With that said, Figures 4-7 highlight a substantial range of algorithm performances across rivers. These differences in performance are likely due to the representativeness of the priors used for that river. For SWOT discharge algorithms, authors have repeatedly shown that the 'quality' of prior hyperparameters plays a large role in the success of one's discharge inversions [@frassonExploringFactorsControlling2021; @andreadisConstrainingAssimilationSWOT2020; @brinkerhoffConstrainingRemoteRiver2020; @tuozzoloEstimatingRiverDischarge2019] and our results here further corroborate this finding. It should be stressed that a substantial portion of rivers from [@frassonExploringFactorsControlling2021] are canal-shaped in nature, suggesting different hydraulic properties than a natural river channel. Nearly universally, these rivers underperformed (Figure 5), however it is impossible to isolate whether that is due to channel geomorphology or the fact that these canals also usually had short timeseries (usually only 12 days).

### 6.3 Towards remotely sensing riverine carbon emissions using SWOT
Figure 7 confirms that BIKER is quite successful, without any in situ information aside from a $CO_2$ logger, at predicting both $FCO_2$ (Figure 6a) and 2) the bulk carbon efflux (Figure 7b-7c). This encouraging result has two main implications for future work coupling remote sensing via SWOT with in situ data. First, it confirms that we can couple BIKER and SWOT with in situ gas concentration loggers to produce reasonable $FCO_2$ estimates at novel temporal resolutions in SWOT-observable rivers. This is particularly useful given recent advances in in situ gas concentration loggers [e.g. @ahoHighestRatesGross2021] but no such similar advances in modeling *k* at equivalent temporal resolutions. BIKER can likely also be ran at the field scale using arrays of pressure transducers to estimate water surface slope (rather than using satellite-based altimeters like SWOT) following recent similar work for estimating streamflow [@harlanDischargeEstimationDense2021].

Secondly, it is important to again stress that unlike BIKER, the HG models in Figure 7 rely on an in situ streamgauge. This means that Figure 7 represents the best performance that those models could ever have, while BIKER represents the worse performance it could ever have. If the HG models are ran using modeled discharge, their accuracy would necessarily decrease. Thus, these results suggest that BIKER will be useful in not only assessing gas exchange velocities, but also in providing additional measurements to upscaling workflows that rely on modeled streamflow [e.g. **Liu etal in review**].

## 7 Conclusions
Gas exchange from aquatic systems has been studied for nearly a century and dozens of predictive equations have been developed to obtain *k* values from easily measured hydraulic properties rather than via direct measurement. Likewise, equation 2 has been repeatedly verified across many non-fluvial environments. However. despite renewed interest in fluvial gas exchange in the last decade or so, there are considerable uncertainties in how fluvial *k* varies with respect to time. This is generally due to a lack of direct measurements of $k_{600}$. In this paper, we propose that the soon-to-launch SWOT satellite will provide enough hydraulic measurements to analyse the temporal dynamics of $k_{600}$, given that there is a robust way to ingest SWOT measurements and infer gas exchange velocities from those measurements.

In that context, we develop the BIKER algorithm to infer $k_{600}$ from simulated SWOT measurements, showing strong recovery of rivers' temporal $k_{600}$ dynamics and relying on absolutely no on-the-ground information. We also pair BIKER with an in situ $CO_2$ logger and show strong performance in reproducing $CO_2$ evasion fluxes and bulk carbon efflux from the rivers' combined surfaces. These strong results functionally open the door for global-scale, near daily estimates of fluvial gas exchange velocity once SWOT launches in 2022. This unprecedented amount of data should allow for significant insights into the temporal dynamics of fluvial gas exchange in large rivers around the world.

## 8 Acknowledgements
C.B. Brinkerhoff was funded on **FINESST**. The version of BIKER used in this study is available at **citable code here**. BIKER remains in active development and is available for download at https://github.com/craigbrinkerhoff/BIKER. All code to build and generate results, figures, and the manuscript is available at https://github.com/craigbrinkerhoff/RSK600. We thank Renato Frasson, Michael Durand, Amber Ulseth, and Jake Beaulieu for generously making their data available for this study. We also thank the entire SWOT discharge working group for their decade-plus body of work which inspired this study.

## 9 Apendix A
*Table A1: Variable description and notation for this study. ADD THE BAYES PARAMETERS YET*

|**Notation**|**Description**|**Calculation (if applicable)**|**Units**|
|------------|---------------|-------------------------------|---------|
|$A$|Channel cross-sectional area|NA|$[L^2]$|
|$A_0$|Median cross-sectional area|NA|$[L^2]$|
|$A_p$|Active-zone fraction|$A_p \propto (\frac{U_*H}{v})^{1/2}$ [@moogAirWaterGasTransfer1999]|unitless|
|*Bulk carbon efflux*|carbon mass transport rate from river to atmosphere|NA|$[\frac{M}{T}]$|
|$CO_{2_{water}}$|Water-side $CO_2$ concentration|NA|$[L]$|
|$CO_{2_{air}}$|Atmospheric-side $CO_2$ concentration|NA|$[L]$|
|$dA$|change in cross-sectional area|$\delta H_e*W$|$[L]$|
|$D_m$|Molecular diffusion coefficient|NA|$[\frac{L^2}{T}]$|
|$\epsilon$|Dissipation rate of near-surface turbulence|NA|$[\frac{E}{M*T}]$|
|$\epsilon_S$|Dissipation rate of bottom-generated turbulence|log-law-of-the-wall model: $\frac{U_*^3}{H}$|$[\frac{E}{M*T}]$|
|$\epsilon_D$|Dissipation rate of form-drag turbulence|form-drag model: $gS\bar{U}$ [@moogStreamReaerationNonuniform1999]||$[\frac{E}{M*T}]$
|$FCO_2$|$CO_2$ evasion flux from river to atmosphere|NA|$[\frac{M}{L^2T}]$|
|$g$|gravitational acceleration|9.8|$[\frac{L}{T^2}]$|
|$H$|Mean flow depth|$A/W$|$[L]$|
|$H_e$|Water surface elevation|NA|$[L]$|
|$i$|Spatial discretization step in mass-conserved river reach|NA|NA|
|$k$|gas exchange velocity|NA|$[\frac{L}{T}]$|
|$k_{600}$|gas exchange velocity normalized to $Sc=600$|$(\frac{600}{Sc_k})^{-0.5}k$ [@raymondScalingGasTransfer2012]|$[\frac{L}{T}]$|
|$n$|Manning's roughness coefficient|$\frac{R_h^{2/3}S^{1/2}}{\bar{U}}$|$[\frac{T}{L^3}]$|
|$\uprho$|Density of water|NA|$[\frac{M}{L^3}]$|
|$R_h$|Hydraulic radius|$\frac{AW}{2H+W}$|$[L]$|
|$S$|River slope|NA|$unitless$|
|$Sc$|Schmidt number|$\frac{v}{D_m}$|$unitless$|
|$t$|temporal discretization step in mass-conserved river reach|NA|NA|
|$\bar{U}$|Cross-sectional average velocity|NA|$[\frac{L}{T}]$|
|$U_*$|Shear velocity|$\sqrt{gSR_h}$|$[\frac{L}{T}]$|
|$\mu$|Viscosity|NA|$[P*T]$|
|$v$|kinematic viscosity|$\frac{\mu}{\uprho}$|$[\frac{L^2}{T}]$|
|$W$|Flow width|NA|$[L]$|

## References

### 2022-06-08 docs/intial_drafts/manuscript_v6.Rmd
first round of review, setting up for rerun on HPC

--- 2022-06-08 docs/intial_drafts/manuscript_v5.Rmd
@@
 ---
-title: "Gas exchange in large rivers influenced by multiple turbulence scales: implications for remotely sensing gas exchange via the SWOT satellite"
+title: "Exploiting hydraulic geometry to remotely sense gas exchange velocity via the SWOT satellite"
 author:
@@
 ## Key Points
-- Multiple scales of turbulence influence gas exchange in the large rivers that the SWOT satellite will measure
-- BIKER algorithm exploits this theory, predicting gas exchange velocity and $CO_2$ fluxes from simulated SWOT data and in situ $CO_2$ data
-- BIKER is also robust to measurement errors implicit in SWOT river measurements
+- BIKER algorithm predicts gas exchange velocity and $CO_2$ fluxes from simulated SWOT data
+- BIKER is robust to measurement errors implicit in SWOT river measurements
+- BIKER will allow for unprecedented study of gas exchange spatiotemporal dynamics upon SWOT's launch
 
 ## Keywords
-gas exchange, fluvial geomorphology, remote sensing, open-channel flow, SWOT, biogeochemistry, ungauged basin
+gas exchange, fluvial geomorphology, remote sensing, open-channel flow, SWOT, biogeochemistry
 
 ## Abstract
-Extensive research over the past two decades has shown that the global river network emits significant amounts of greenhouse gases to the atmosphere via evasion.
-Despite much progress, there is still significant uncertainty in the temporal dynamics of gas exchange velocity (the rate of this evasion), with most models relying on temporally-coarse estimates of this velocity.
-We propose that the NASA/CNES/USKA/CSA SWOT satellite, set to usher in a new age of river hydrology research at the global scale, will provide similarly transformative insights to fluvial gas exchange modeling.
-Here, we present work inferring gas exchange velocity from SWOT observations.
-We exploit the unique geomorphology of SWOT-observable rivers to develop a physical model that is nearly entirely remotely sensible.
+Extensive research over the past two decades has shown that the global river network emits significant amounts of greenhouse gas via evasion.
+Despite much progress, there is still significant uncertainty in the temporal dynamics of gas exchange velocity (the rate of evasion) and thus carbon emissions to the atmosphere.
+We propose that the NASA/CNES/USKA/CSA SWOT satellite, set to usher in a new age of hydrology research at the global scale, can provide similarly transformative insights to fluvial gas exchange modeling upon launch in 2022.
+Here, we present work inferring gas exchange velocity from future SWOT observations without an in situ calibration.
+We exploit the unique geomorphology of SWOT-observable rivers to develop a physical model of gas exchange that is nearly entirely remotely sensible and explains `r round(modelsWIDE[modelsWIDE$name == 'Chainsaw-eD',]$r2,2)*100`% of variation in gas exchange.
 We then couple this model with established Bayesian techniques to develop BIKER, or the 'Bayesian Inference of the $k_{600}$ Exchange Rate' algorithm.
-We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers and 763 discrete measurements of river gas exchange, yielding 1) a theoretical model that explains `r round(modelsWIDE[modelsWIDE$name == 'Chainsaw-eD',]$r2,2)*100`% of variation in gas exchange in SWOT-like rivers and 2) an algorithm that predicts this theoretical model solely from SWOT observations with a by-river median normalized RMSE of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$NRMSE), 2)*100`%.
-BIKER is most robust at reproducing the temporal dynamics of gas exchange (by-river median correlation coefficient of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$r), 2)`), rather than estimating the absolute rates of exchange.
-Finally, it is robust to measurement errors implicit in the SWOT data.
-With SWOT set to launch in 2022, we suggest that BIKER will prove a fundamental asset for exploring global-scale, spatiotemporal dynamics of fluvial gas exchange.
+We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers and 763 discrete measurements of river gas exchange, yielding an algorithm that predicts this theoretical model solely from SWOT observations with a by-river median normalized RMSE of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$NRMSE), 2)*100`%.
+Like many fluvial Bayesian inference algorithms, BIKER is better at reproducing the temporal dynamics of gas exchange (median correlation coefficient of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$r), 2)`), rather than estimating the absolute rates of exchange.
+Finally, we show BIKER robust to measurement errors implicit in the SWOT data.
+With SWOT set to launch in 2022, we suggest that BIKER can be deployed to map the global-scale spatiotemporal dynamics of fluvial gas exchange in large rivers.
 
@@
 Natural systems play a critical role in the budgeting and accounting of the global carbon cycle under climate change.
-Since the publication of @colePlumbingGlobalCarbon2007, the global river network is recognized to emit substantial amounts of carbon to the atmosphere, in addition to exporting it to the oceans.
+Following @colePlumbingGlobalCarbon2007, the global river network is recognized to emit substantial amounts of carbon to the atmosphere, in addition to their long understood role of exporting carbon to the oceans.
 Current estimates of total carbon dioxide evasion ($FCO_2$) to the atmosphere from the global river network vary from 650-1800 Tg C/yr [@lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013] (*Liu etal in review- add number is paper is accepted in time*), with 167 Tg-C/yr coming from mountain streams alone [@horgbyUnexpectedLargeEvasion2019].
-Despite its incredibly small percentage of the global land surface [0.`r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])`%- @raymondGlobalCarbonDioxide2013], this flux is on par with the total oceanic $CO_2$ uptake rate [@gruberOceanicSinkAnthropogenic2019; @horgbyUnexpectedLargeEvasion2019] and the global forest carbon uptake rate [@panLargePersistentCarbon2011].
-It is still relatively poorly constrained (as evident by these studies' reported uncertainties) but is clearly a critical component of the global carbon cycle.
-Equation 1 represents this flux given the gas concentration gradient between the water $CO_{2_{water}}$ and the air $CO_{2_{air}}$ and the gas exchange velocity *k*.
+Despite its incredibly small percentage of the global land surface [0.`r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])`%- @raymondGlobalCarbonDioxide2013], this C flux is on par with the total oceanic $CO_2$ uptake rate [@gruberOceanicSinkAnthropogenic2019; @horgbyUnexpectedLargeEvasion2019] and the global forest carbon uptake rate [@panLargePersistentCarbon2011].
+River $CO_2$ evasion is increasingly better constrained (as evident by these studies' reported uncertainties) and is clearly a critical component of the global carbon cycle.
+Equation 1 represents this flux given $\delta CO_2$ (the gas concentration gradient between the water $CO_{2_{water}}$ and the air $CO_{2_{air}}$) and the gas exchange velocity *k*.
 Consult Appendix A for variable nomenclature used throughout this study.
 
-$$FCO_2=k(CO_{2_{water}}-CO_{2_{air}}) \mathbf{(1)}$$
+$$FCO_2=k* \delta CO_2 \mathbf{(1)}$$
+The structure of equation 1 necessitates that calculations of this flux are highly sensitive to measurements/estimates of *k*.
+However, *k* can only be directly measured via a known gas concentration gradient, eddy-covariance measurements, or tracer additions to the stream [@hallGasExchangeStreams2020].
+When trying to constrain the global fluvial $FCO_2$ flux across millions of rivers, this limitation necessitates the use of predictive models for fluvial *k* that are based on easily estimated river hydraulic properties.
+In that vein, there have been over 20 empirical models developed for this and similar purposes, generally using some combination of mean velocity $\bar{U}$, shear velocity $U_*$, width $W$, depth $H$, and slope $S$ as predictors [@wangPhysicallyBasedScaling2021].
+Note that these models usually predict $k_{600}$, or *k* normalized by a Schmidt number of 600, to make use of additional measurements taken at different water temperatures and with different gases.
+For example, $k_{600}$ reflects the $k_{CO_{2}}$ at 20 degrees Celsius.
 
-The structure of equation 1 necessitates that calculations of this flux are highly sensitive to one's measurements/estimates of *k* [@hallGasExchangeStreams2020].
-Even beyond $CO_2$ emissions, fluvial *k* for various dissolved gases are of significant importance to aquatic ecologists modeling stream metabolism [e.g.
-@bernhardtMetabolicRegimesFlowing2018] and water quality engineers modeling river responses to pollutant loadings [among other things- @chapraSurfaceWaterQualityModeling2008].
-Because of this far-reaching importance, decades of work have focused on elucidating the physical mechanisms behind *k* in oceans and lakes [@wanninkhofAdvancesQuantifyingAirSea2009 for a review] and, to a lesser extent, fluvial gas exchange [@hallGasExchangeStreams2020 for a review].
+While this workflow enables estimating the global $FCO_2$ flux ($FCO_2_{upscaled}$) across millions of rivers, it also changes the base parameters that ultimately control that estimate.
+This functional relationship is described in equation 2.
+It suggests that $FCO_2$ estimates are not only at the mercy of the accuracy and spatiotemporal resolution of $\delta CO_2$ [of which an extensive literature has explored- e.g.
+@crawfordCO2TimeSeries2017; @peterScalesDriversTemporal2014; @ranDynamicsRiverineCO22017; @rocher-rosLandscapeProcessDomains2019; @raymondAtmosphericCO2Evasion2000], but also the accuracy and resolution of our river hydraulics estimates.
 
-Given that fluvial *k* should scale with near-surface turbulence in a turbulent flow [@hallGasExchangeStreams2020], extensive field and laboratory experiments have converged on the 'small-eddy model' as first proposed by @lamontEddyCellModel1970 which scales *k* via the smallest-scale turbulent eddies [e.g.
-@lorkeUnifiedScalingRelation2006; @zappaEnvironmentalTurbulentMixing2007; @zappaVariationSurfaceTurbulence2003; @moogStreamReaerationNonuniform1999; @vachonRelationshipNearsurfaceTurbulence2010; @tokoroHighGastransferVelocity2008; @katulStructureFunctionModel2018].
-This model is provided as equation 2, where $\epsilon$ is the dissipation rate of near-surface turbulence, $v$ is the kinematic viscosity, and *Sc* is the Schmidt number.
-To compare across gases and water temperatures, *k* is often reported as $k_{600}$, where *k* is normalized to a Schmidt number of 600 [reflecting a $k_{CO_2}$ at 20 degrees Celsius- @hallGasExchangeStreams2020].
-Note that directly measuring $\epsilon$ is non-trivial; therefore, a frequently used model for $\epsilon$ assumes that all turbulence is generated at the bed and transported to the air-water interface via the log-law-of-the-wall [@lorkeUnifiedScalingRelation2006; @nezuTurbulenceOpenChannel1993].
+$$FCO_{2_{upscaled}}=f(k_{CO_2}, \delta CO_2)=f(\bar{U}, H, S, W, \delta CO_2) \mathbf{(2)}$$
+For example, the @raymondGlobalCarbonDioxide2013, @lauerwaldSpatialPatternsCO22015, and @horgbyUnexpectedLargeEvasion2019 $FCO_2$ global upscaling studies all relied on $k_{CO_2}$ values calculated using mean annual hydraulic conditions, while @borgesGloballySignificantGreenhousegas2015 used a combination of mean annual values and a constant $k_{CO_2}$ in space and time to upscale over Africa.
+In all of these key studies, the temporal dynamics of the gas exchange velocity (and thus $FCO_2$) were ignored because of data limitations.
+As workers move towards river-reach-explicit modeling of gas exchange at the global scale with more and more elaborate models for $CO_{water}$ [@horgbyUnexpectedLargeEvasion2019; @brinkerhoffLakeMorphometryRiver2021; *Liu etal in review; Saccardi & Winnick et al in review*], they will also have to contend with the lack of direct measurements of global river hydraulics.
 
-$$k\propto Sc^{-1/2}(v\epsilon)^{1/4} \mathbf{(2)}$$
+This lack of direct measurements also inhibits our process-level understandings of the spatiotemporal dynamics of $k_{600}$.
+A few field-scale studies have directly investigated these dynamics but they have been limited to individual rivers and/or limited field seasons [@hallAirWaterOxygen2012; @sand-jensenCO2DynamicsDanish2012].
+For example, @wallinSpatiotemporalVariabilityGas2011 performed a preliminary analysis in northern Sweden relating cross-section specific temporal variability in gas exchange with channel slope, but they were limited to a mean of only 8 measurements per river in a single watershed.
+They also often focus on small rivers.
+More recently, @wangPhysicallyBasedScaling2021 attempted to address this problem by simulating *k* in 35 reasonably sized rivers (widths ranging from 0.23–349m) using a stream metabolism model [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets to implicitly solve for *k*.
+They found that process-based models for *k* generally work well in these simulated rivers and that *k*~streamflow relationships change with river size, though they too were limited by a lack of direct hydraulic measurements and had to rely on scaling equations to estimate river depth and velocity.
 
-While this model works reasonably well in non-fluvial environments where turbulence production is mostly limited to bottom shear [e.g.
-@lorkeUnifiedScalingRelation2006; @zappaEnvironmentalTurbulentMixing2007], there is considerable uncertainty in how it applies to fluvial systems.
-In particular, there is uncertainty as to which models are most appropriate at different locations along the stream-to-river continuum given that river hydraulics change predictably along a river's longitudinal profile [@wangPhysicallyBasedScaling2021].
-In small and steep headwaters, 'bubble-mediated gas exchange' in whitewater might lead to substantially elevated *k* where bedload is large and general channel roughness is extremely high [@mauriceInfluenceFlowBed2017; @ulsethDistinctAirWater2019].
-At the other end of the stream-to-river continuum (the largest rivers draining the continents), relatively little attention has been paid to gas exchange.
-The handful of existing field studies of large-river *k* suggest that *k* begins to be influenced by wind shear once the water surface is less protected, though little else is well-established [@alinPhysicalControlsCarbon2011; @beaulieuControlsGasTransfer2012; @wangPhysicallyBasedScaling2021].
-Other workers have noted that laboratory and field observations suggest small-roughness open channel flows (generally big rivers) do not exhibit homogeneous surface dissipation at their air-water interface [@moogAirWaterGasTransfer1999; @talkeTurbulentKineticEnergy2013].
-@moogAirWaterGasTransfer1999 proposed a conceptual 'chainsaw model', where large, depth-scale eddies transport small eddies to the interface, but the small eddies are responsible for the actual gas evasion that occurs at this interface.
-Finally, another approach proposed regardless of one's location along the stream-to-river continuum is to model *k* via 'form-drag shear' (i.e.
-total stream power per unit mass water).
-This normalized stream power captures the bulk frictional resistance via channel banks, meanders, bars, etc.
-that is unique to fluvial systems [@moogStreamReaerationNonuniform1999].
-Authors have since shown that this model can reasonably predict *k* in rivers [@raymondScalingGasTransfer2012; @ulsethDistinctAirWater2019; @wangPhysicallyBasedScaling2021], however it requires estimates of the mean flow velocity and therefore can introduce additional uncertainties.
+In theory, the above problems could be alleviated and a global analysis of *k* spatiotemporal trends would be possible if direct hydraulics measurements were available at the global scale at a sufficient temporal resolution.
+Conveniently, near-daily hydraulics measurements will soon be available via the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission.
+SWOT is expected to launch in 2022 and provide the world's first direct measurements of global water surface extent and elevation at novel temporal resolutions.
+SWOT is a wide swath radar interferometer and will sample rivers every 1 to 7 days per 21 day repeat cycle, measuring rivers wider than 100m with a goal of expanding this to rivers at least 50m wide [@biancamariaSWOTMissionIts2016].
+Via its direct hydraulics measurements, SWOT is expected to usher in a sea change in global-scale hydrology, and fluvial biogeochemsitry should be no different.
+We suggest that SWOT data will prove transformative to global-scale biogeochemistry if techniques are developed that can ingest SWOT data and infer the parameters of interest.
+Here, we focus specifically on $k_{600}$, borrowing tools from fluvial geomorphology and existing SWOT algorithms to answer the following two questions:
 
-Taken in aggregate, our theoretical understanding of fluvial gas exchange is likely limited by a dearth of field-measured fluvial *k*, particularly in large rivers where direct measurements are difficult.
-@wangPhysicallyBasedScaling2021 attempted to address this by simulating *k* in 35 reasonably sized rivers (widths ranging from 0.23–349m) using a stream metabolism model [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets to implicitly solve for *k*.
-They found that equation 2, coupled with $\epsilon$ modeled via the log-law-of-the-wall, is valid in their simulated rivers and that *k*~streamflow relationships break down in large rivers.
-However, their simulated rivers were limited by a lack of direct hydraulic measurements and had to rely on models to estimate river depth and velocity, introducing additional uncertainty.
-Further, they did not assess model fit in very steep systems [@ulsethDistinctAirWater2019] and stopped short of parsing out hydraulic explanations for their results.
+  - Can we leverage the unique geomorphology of the large rivers that SWOT will observe to develop a physically-based gas evasion model unique to these rivers?
 
-All of these mechanistic uncertainties are then propagated through upscaling workflows when biogeochemists predict *k* across thousands of rivers [e.g @borgesGloballySignificantGreenhousegas2015; @horgbyUnexpectedLargeEvasion2019; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013] via equation 1 coupled with hydraulic geometry (HG).
-HG is the scaling relationships between streamflow and river channel hydraulics [@leopoldHydraulicGeometryStream1953; @gleasonHydraulicGeometryNatural2015] that allow biogeochemists to predict river depth and velocity from streamflow models and/or streamgauges .
-It is currently not well understood how sensitive global estimates of fluvial gas evasion are to the specific HG model that is employed by the worker.
-Further, these approaches rely on either in situ discharge records or modeled streamflow which further contributes uncertainty.
-This is all exacerbated in ungauged basins that cover large areas, especially in the carbon-rich Arctic inland waters, where little in situ information is available and fieldwork is impractical [@gleasonRemoteSensingRiver2020].
-
-A potential alternative to this upscaling approach is to directly estimate a river's hydraulic properties (and thus *k*) from remote sensing (RS) data.
-Remote sensing of river hydraulics is a burgeoning subfield within remote sensing of hydrology, often in service of remote sensing of river discharge [RSQ- @gleasonRemoteSensingRiver2020].
-This is accomplished via two general approaches: 'gauged' methods which rely on in situ river data to calibrate one's method to the river(s) at hand and 'ungauged' techniques which focus on hydraulic generalizability in the service of merely improving existing knowledge in data-poor domains [@gleasonRemoteSensingRiver2020].
-Many, but not all, of these ungauged approaches are developed in the context of the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission.
-SWOT is expected to launch in 2022 and provide the world's first global measurements of water surface extent and elevation at novel temporal resolutions.
-SWOT is a wide swath radar interferometer and will sample rivers every 1 to 7 days per 21 day repeat cycle and will measure rivers wider than 100m with a goal of expanding this to rivers at least 50m wide [@biancamariaSWOTMissionIts2016].
-A decade of SWOT work has explored the multi-parameter problem of estimating the river hydraulic parameters of roughness and bathymetry from remote sensing to produce the SWOT discharge product [e.g.
-@andreadisConstrainingAssimilationSWOT2020; @brinkerhoffConstrainingRemoteRiver2020; @durandEstimatingReachaveragedDischarge2014; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014a; @garamboisVariationalEstimationEffective2020; @brissetAssimilationAltimetricData2018; @hagemannBAMBayesianAMHGManning2017; @larnierRiverDischargeBathymetry2020a; @oubanasDischargeEstimationUngauged2018].
-
-Once SWOT launches, there will be a massive trove of high-temporal-resolution hydraulics measurements of the world's largest river surfaces, as well as established techniques to infer streamflow from those measurements.
-Given the uncertainties and limited field measurements of river *k* as described above, the SWOT measurements will hopefully also prove transformative in elucidating the spatiotemporal dynamics of gas exchange.
-However, in order to make sense of SWOT data for these purposes, a method must first be developed that can ingest SWOT data and return reasonable estimates of the gas exchange velocity solely from those data.
-In this context, we borrow tools from fluvial geomorphology and the SWOT discharge algorithms to answer the following questions: 1) can we leverage the unique geomorphology of the large rivers that SWOT will observe to develop a physically-based gas evasion model unique to these rivers?
-And 2) can we then exploit the findings from this analysis to develop a novel methodology that predicts this physically-based $k_{600}$ model (and its explicit uncertainty) solely using SWOT observations?
+  - Can we then exploit the findings from this analysis to develop a novel methodology that infers this physically-based $k_{600}$ model (and its uncertainty) solely using SWOT observations?
 
@@
 We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-observable rivers from around the world using hydraulic models to produce SWOT-like data (as SWOT has not yet launched).
-We also quantify BIKER's sensitivity to the expected SWOT measurement errors on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` of those rivers: while SWOT data represent a sea change in inland water monitoring, it is expected to have an approximately 10cm error in water surface elevation [@biancamariaSWOTMissionIts2016] as well as river width errors [@frassonExploringFactorsControlling2021].
+We also quantify BIKER's sensitivity to the expected SWOT measurement errors on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` of those rivers: SWOT data is expected to have an approximately 10cm error in water surface elevation [@biancamariaSWOTMissionIts2016] as well as river width errors as a function of the cross-track distance [@frassonExploringFactorsControlling2021].
 Finally, we use previously published dissolved $CO_2$ data to represent a hypothetical in situ $CO_2$ sensor and compare the bulk carbon efflux from the `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers as calculated using BIKER and previously published in situ techniques for predicting $k_{600}$ from streamflow.
 
-## 2 Paper overview
-Following the description of the data used (section 3), this paper is split into two distinct parts: gas-exchange model development and validation (section 4) and BIKER algorithm development and validation (section 5).
-Each section is further split into both methods and results sections, with section 5 fundamentally dependent on section 4.
-We conclude with a Discussion regarding both sections 4 and 5 (section 6).
+Following the description of the data used (section 2), this paper is split into two distinct parts: gas-exchange theory and model development (section 3) and BIKER algorithm development and validation (section 4).
+Section 4 is fundamentally dependent on the results presented in section 3.
+We conclude with a Discussion regarding both sections 3 and 4 (section 5).
 We also provide a flowchart detailing the entire study as Figure 1.
@@
 
-## 3 Data
+## 2 Data
 Numerous datasets were used in this study. Please see Figure 2 for a map of the approximate locations for the data used in this study.
@@
 
-Gas exchange theory (section 4) is explored using multiple datasets where $k_{600}$ was measured directly and/or collected from the literature.
+Gas exchange theory (section 3) is explored using multiple datasets where $k_{600}$ was measured directly and/or collected from the literature.
 In total, there are 763 measurements of stream hydraulics and $k_{600}$.
@@
 
-To quantify our geomorphic assumptions for SWOT rivers (section 4.1), we additionally use a previously published compilation of field hydraulics measurements [@brinkerhoffReconcilingAtaStationAtManyStations2019].
+To quantify our geomorphic assumptions for SWOT rivers (section 3.1), we additionally use a previously published compilation of field hydraulics measurements [@brinkerhoffReconcilingAtaStationAtManyStations2019].
 These were originally made to calibrate United States Geological Survey (USGS) streamgauge rating curves.
@@
 Ultimately, the `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers are spread across the United States, France, Italy, the United Kingdom, and Bangladesh.
-We omit three models from @durandIntercomparisonRemoteSensing2016 because they lack enough hydraulic information to calculate the mean flow depth, which is necessary for algorithm validation (section 5.1.2).
+We omit three models from @durandIntercomparisonRemoteSensing2016 because they lack enough hydraulic information to calculate the mean flow depth, which is necessary for algorithm validation (section 4.1.2).
 These are two models for different sections of the Saint Lawrence River in Canada and one for the Tanana River in Alaska, United States.
 
-To assess the influence of SWOT's implicit measurement error on BIKER performance, we use the error model developed by @durandHowWillRadar2020 and implemented on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])`/`r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` of the rivers by @frassonExploringFactorsControlling2021.
-Error in SWOT measurements will come from both the error tolerances intrinsic in the satellite data product as well as radar layover error.
+To assess the influence of SWOT's implicit measurement error on BIKER performance, we use the error models developed by @durandHowWillRadar2020 and @frassonExploringFactorsControlling2021, which the latter implemented on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])`/`r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` of the rivers.
+Error in SWOT measurements will come from both the error tolerances intrinsic in the satellite data product as well as radar layover error and river width error.
 Layover error is the phenomenon when radar returns from different places arrive at the sensor at the same time, leading to taller landscape features appearing closer to the sensor than shorter landscape features that are the same horizontal distance from the sensor [@durandHowWillRadar2020].
-Width errors due to poor water classification are ignored as they were in @frassonExploringFactorsControlling2021.
-@frassonExploringFactorsControlling2021 tested a suite of measurement error scenarios, however here we test only their 'full uncertainty scenario' that most closely represents the measurement errors likely to be implicit in SWOT data.
+Width errors due to poor water classification are ignored but width errors due to the cross-track distance are included [@frassonExploringFactorsControlling2021].
+The cross-track distance, reflective of the amount of the river cross-section that falls within SWOT's swath, controls the size of the pixels **help, I on't understand cross-track distance**.
+@frassonExploringFactorsControlling2021 tested a suite of measurement error scenarios, however here we test only their 'full uncertainty scenario' that most closely represents the measurement errors that will be implicit in SWOT data.
 
-For the $CO_2$ evasion and carbon efflux calculations (section 5.1.3), we use 26 bi-weekly dissolved $CO_2$ samples made by @beaulieuControlsGasTransfer2012 at one location in the Ohio River for one calender year from 2008-2009 (Figure S1).
+For the $CO_2$ evasion and carbon efflux calculations (section 4.1.3), we use 26 bi-weekly dissolved $CO_2$ samples collected by @beaulieuControlsGasTransfer2012 at one location in the Ohio River for one calender year from 2008-2009 (Figure S1).
 Note that this $CO_2$ data is for the Ohio River only but was applied to all rivers (which includes multiple sections of the Ohio River).
@@
 
-## 4 Hydraulic geometry in large rivers and predicting fluvial gas exchange velocity
-Given that *k* can only be directly measured via a known gas concentration gradient, eddy-covariance measurements, or tracer additions to the stream [@hallGasExchangeStreams2020], SWOT data will not provide the necessary parameters to directly measure *k*.
-Further, these techniques are entirely infeasible at the global scale where our goal with BIKER is primarily to better understand spatiotemporal patterns of *k* rather than obtain absolute measurements of it.
-Therefore, we need a simple yet physically-based model for *k* to directly relate channel hydraulics to gas exchange in the large rivers that SWOT will measure.
-We hypothesize that this can be done by leveraging the unique geomorphology of the very wide rivers that SWOT will observe.
-In this section, we develop and test process-based models for *k*, selecting one that is then implemented within BIKER (section 5).
+## 3 Hydraulic geometry in large rivers and predicting fluvial gas exchange velocity
+To predict *k* in the very large rivers that SWOT will observe, we start from established process-based models for *k*, impose hydraulic assumptions that are valid in large rivers, and then obtain a suite of models we can test.
+First, we outline these established models (section 3.1) and then step through our hydraulic assumptions to arrive at equations that can be ingested by SWOT (section 3.2).
+Following this model development, all models are empirically tested on the dataset described in Section 2 after filtering for measurements made in hydraulically-wide channels (section 3.3).
+This was defined here as those whose hydraulic radius was within 1% of their mean flow depth.
+All told, this amounts to 166 direct measurements of hydraulically-wide $k_{600}$ to test with.
 
-### 4.1 Section 4 methods
-Rivers and streams change predictably along their longitudinal profile from headwater to ocean.
+### 3.1 Process-based modeling of river *k*
+*k* should scale with near-surface turbulence in a turbulent flow [@hallGasExchangeStreams2020] and thus extensive field and laboratory experiments have converged on the 'small-eddy model' as first proposed by @lamontEddyCellModel1970 to scale *k* via the smallest-scale turbulent eddies [e.g.
+@lorkeUnifiedScalingRelation2006; @zappaEnvironmentalTurbulentMixing2007; @zappaVariationSurfaceTurbulence2003; @moogStreamReaerationNonuniform1999; @vachonRelationshipNearsurfaceTurbulence2010; @tokoroHighGastransferVelocity2008; @katulStructureFunctionModel2018; @wangPhysicallyBasedScaling2021].
+This model is provided as equation 3, where $\epsilon$ is the dissipation rate of near-surface turbulence, $v$ is the kinematic viscosity, and *Sc* is the Schmidt number.
+
+$$k\propto Sc^{-1/2}(v\epsilon)^{1/4} \mathbf{(3)}$$
+Further, some workers have noted that laboratory and field observations suggest open channel flows with small bed roughness (generally big rivers) do not exhibit homogeneous surface dissipation at their air-water interface [@moogAirWaterGasTransfer1999; @talkeTurbulentKineticEnergy2013].
+Given this observation, @moogAirWaterGasTransfer1999 proposed a 'chainsaw model' which extends the small-eddy model by limiting surface dissipation to only 'active zones' within which a homogeneous dissipation rate occurs and the small-eddy model is valid.
+This is equation 4 where $A_p$ is the fraction of the interface that are active zones and is proportional to the larger, depth-scale eddies [@moogAirWaterGasTransfer1999].
+This means that if the entire interface is treated as an active zone, the model would simply reduce to the small-eddy model.
+This is hypothetically useful in low-turbulence flows where a relative lack of large-scale eddies effectively 'filter out' the number of small-eddies that actually reach the interface and lead to gas exchange [@talkeTurbulentKineticEnergy2013].
+In the context of BIKER, we chose to test this model because large, SWOT-observable rivers are generally the smoothest, least-turbulent flows along the stream-to-river continuum.
+Further, to our knowledge the chainsaw model has never been empirically tested in predicting *k*, aside from confirming that large-scale eddies differentially move turbulence to the surface in a hydraulically-wide river [@talkeTurbulentKineticEnergy2013].
+
+$$k \propto Sc^{-1/2}(v\epsilon)^{1/4}A_p^{3/4} \mathbf{(4)}$$
+Finally, both equations rely on measurements of $\epsilon$, of which directly measuring is non-trivial.
+When working at large scales, a commonly used $\epsilon$ model assumes that all turbulence is generated at the bed and transported to the air-water interface via the log-law-of-the-wall such that $\epsilon = \epsilon_S = \frac{U_*^3}{H}$ [@lorkeUnifiedScalingRelation2006; @nezuTurbulenceOpenChannel1993].
+Another approach models *k* via 'form-drag dissipation' ($\epsilon = \epsilon_D = gS\bar{U}$) which is equivalently the total stream power per unit mass water.
+This normalized stream power captures the bulk frictional resistance (and thus energy dissipation) via channel banks, meanders, bars, etc.
+that is unique to fluvial systems [@moogStreamReaerationNonuniform1999].
+Authors have since shown that this model can reasonably predict *k* [@raymondScalingGasTransfer2012; @ulsethDistinctAirWater2019; @wangPhysicallyBasedScaling2021], however it requires estimates of mean flow velocity and does not necessarily reflect the surface dissipation rate.
+
+### 3.2 Imposing large-river hydraulic geometry
+Rivers and streams change predictably along their longitudinal profile from headwater to ocean, and we can exploit the distinct hydraulic geometry of large rivers to estimate *k* in SWOT-observable systems.
 In general, as river size increases, channels become more rectangular, their shapes elongate (becoming wider quicker than they become deeper) and their hydraulic radii begin to approximate their mean flow depth, i.e.
@@
 
-Assuming that all SWOT-observable rivers are hydraulically-wide, we test the performance of four models for predicting *k* in hydraulically-wide channels.
-We test two conceptual models for gas evasion (both descrbied in the Introduction): the 'small-eddy' model [@lamontEddyCellModel1970- section 4.1.1] and the 'chainsaw' model [@moogAirWaterGasTransfer1999- section 4.1.2].
-We also test two different models for the turbulent dissipation rate (also described in the Introduction): the log-law-of-the-wall ($\epsilon=\epsilon_S$) and form-drag dissipation [@moogStreamReaerationNonuniform1999] ($\epsilon=\epsilon_D$).
-While the complete model derivations for sections 4.1.1 and 4.1.2 are provided in Text S2, the broad goal was to reduce the equations down to their fundamental parameters, identifying which terms are SWOT observable and limiting the number of terms not directly measurable via SWOT.
+Assuming that all SWOT-observable rivers are hydraulically-wide, we test the performance of four models for predicting *k* in hydraulically-wide channels: the four unique combinations of equations 3-4 and the two $\epsilon$ models already outlined.
+While the complete model derivations for the small-eddy models (section 3.2.1) and the chainsaw models (section 3.2.2) are provided in Text S2, the broad goal was to reduce the equations down to their fundamental parameters, identifying which terms are SWOT observable and limiting the number of terms not directly measurable via SWOT.
 
-Next, all models are empirically tested on the dataset described in Section 3 after filtering for measurements made in hydraulically-wide channels (section 4.2).
-This was defined here as those whose hydraulic radius was within 1% of their mean flow depth.
-Note that, to account for different water temperatures and gases used in measuring *k*, all data are normalized to a Schmidt number of 600 (i.e.
-$k_{600}$), as is standard practice.
-All told, this amounts to 166 direct measurements of hydraulically-wide $k_{600}$ to test with.
+#### 3.2.1 'Small-eddy' model for gas evasion
+We derive equations starting from the small-eddy model (equation 3) for hydraulically-wide channels.
+If we assume the log-law-of-the-wall model for $\epsilon$, simplify equation 2 using $R_H = H$, and finally solve for $k_{600}$ via statistical parameter $\alpha$, we eventually arrive at equation 5.
+If we do the same but via the form-drag model for dissipation and statistical parameter $\alpha_1$, we arrive at equation 6.
 
-#### 4.1.1 'Small-eddy' model for gas evasion
-First, we derive equations starting from the small-eddy model (equation 2) for hydraulically-wide channels.
-If we assume the log-law-of-the-wall model for $\epsilon$ (see Appendix A for its calculation), simplify equation 2 using $R_H = H$, and finally solve for $k_{600}$ via statistical parameter $\alpha$, we eventually arrive at equation 3.
-If we do the same but via the form-drag model for dissipation and statistical parameter $\alpha_1$, we arrive at equation 4.
-$\bar{U}$ refers to mean flow velocity.
+$$k_{600}=\alpha(gS)^{3/8}H^{1/8} \mathbf{(5)}$$
 
-$$k_{600}=\alpha(gS)^{3/8}H^{1/8} \mathbf{(3)}$$
+$$k_{600}=\alpha_1(gS\bar{U})^{1/4} \mathbf{(6)}$$
+Equations 5 and 6 thus define gas exchange velocity solely as a function of slope and mean flow depth or mean flow velocity. This is theoretically valid only in a hydraulically-wide channel.
 
-$$k_{600}=\alpha_1(gS\bar{U})^{1/4} \mathbf{(4)}$$
+#### 3.2.2 'Chainsaw' model for gas evasion
+We also impose a hydraulically-wide channel and $\epsilon_S$ on the chainsaw model just like we did in section 3.1.1 and arrive at equation 7 (with statistical coefficient $\beta$).
+The result is identical to the hydraulically-wide small-eddy model (equation 5), except for the exponents which are both steeper.
+Equation 8 reflects the same derivation, but via the form-drag model for dissipation and statistical coefficient $\beta_1$.
 
-Equations 3 and 4 thus define gas exchange velocity solely as a function of water surface slope and mean flow depth or mean flow velocity.
-This is theoretically valid only in a hydraulically-wide channel.
-
-#### 4.1.2 'Chainsaw' model for gas evasion
-As mentioned in the Introduction, @moogAirWaterGasTransfer1999 noted that laboratory observations suggest small-roughness open-channel flows do not exhibit homogeneous surface dissipation at their air-water interfaces.
-Under this guise, their chainsaw model extends the small eddy model by limiting surface dissipation to only 'active zones' within which a homogeneous dissipation rate occurs and the small-eddy model is valid.
-The fraction of the interface that are active zones is proportional to the larger, depth-scale eddies via the so-called 'Reynolds shear number'.
-This means that if the entire interface is treated as an active zone, the model would simply reduce to the small-eddy model.
-This is hypothetically useful in low-turbulence flows where a relative lack of large-scale eddies effectively 'filter out' the number of small-eddies that actually reach the interface and lead to gas exchange [@talkeTurbulentKineticEnergy2013].
-
-In the context of BIKER, we chose to test this model because large, SWOT-observable rivers are generally the smoothest, least-turbulent flows along the stream-to-river continuum.
-Further, to our knowledge the chainsaw model has never been empirically tested in predicting *k*, aside from confirming that large-scale eddies differentially move turbulence to the surface in a hydraulically-wide river [@talkeTurbulentKineticEnergy2013].
-To do this, we impose a hydraulically-wide channel on the chainsaw model just like we did in section 4.1.1 and arrive at equation 5 (with statistical coefficient $\beta$).
-The result is identical to the hydraulically-wide small-eddy model (equation 3), except for the exponents which are both steeper.
-Equation 6 reflects the same derivation, but via the form-drag model for dissipation and statistical coefficient $\beta_1$.
-
-$$k_{600}=\beta(gS)^{9/16}H^{11/16} \mathbf{(5)}$$
-$$k_{600}=\beta_1(gS)^{7/16}\bar{U}^{1/4}H^{9/16} \mathbf{(6)}$$
-
+$$k_{600}=\beta(gS)^{9/16}H^{11/16} \mathbf{(7)}$$
+$$k_{600}=\beta_1(gS)^{7/16}\bar{U}^{1/4}H^{9/16} \mathbf{(8)}$$
 We now have four different models for $k_{600}$ in hydraulically-wide rivers, each with a different set of assumptions and parameters but all defined as functions of the water surface slope and other hydraulic terms that are not directly measurable via SWOT.
@@
 *Table 1: Four physically-based models for $k_{600}$ each derived under a different set of mass transport assumptions but all valid in hydraulically-wide rivers.
-Their derivations are outlined in sections 4.1.1 and 4.1.2 and their empirical tests are shown in section 4.2.*
+Their derivations are outlined in sections 3.2.1 and 3.2.2 and their empirical tests are shown in section 3.3.*
 
@@
 
-### 4.2 Section 3 results
-With equations 3-6 defined, we now empirically test their strength of fit for hydraulically-wide channels via the coefficient of determination ($r^2$).
+### 3.3 Model validation
+With equations 5-8 defined, we now empirically test their strength of fit for hydraulically-wide channels via the coefficient of determination ($r^2$).
 These results are plotted in Figure 3.
@@
 They systematically underestimate lower gas exchange velocities and overestimate higher ones, with the log-law-of-the-wall model (Figure 3a) performing marginally worse both visually and via $r^2$ (`r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eS',]$r2, 2)` versus `r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eD',]$r2, 2)` for Figure 3c).
-However, when this model is expanded to account for a differential dissipation rate at the interface via the chainsaw model [equations 5 and 6- @moogAirWaterGasTransfer1999], the dynamics are more accurately captured (Figures 3b, 3d) and $r^2$ scores go up appreciably (`r round(modelsWIDE[modelsWIDE$name == 'Chainsaw-eS',]$r2, 2)` and `r round(modelsWIDE[modelsWIDE$name == 'Chainsaw-eD',]$r2, 2)` for the log-law-of-the-wall and form-drag dissipation, respectively).
+However, when this model is expanded to account for a differential dissipation rate at the interface via the chainsaw model [equations 6 and 7- @moogAirWaterGasTransfer1999], the dynamics are more accurately captured (Figures 3b, 3d) and $r^2$ scores go up appreciably (`r round(modelsWIDE[modelsWIDE$name == 'Chainsaw-eS',]$r2, 2)` and `r round(modelsWIDE[modelsWIDE$name == 'Chainsaw-eD',]$r2, 2)` for the log-law-of-the-wall and form-drag dissipation, respectively).
 Modeling dissipation via total form-drag rather than just via bed shear and the log-law-of-the-wall improves model performance.
@@
 It is also worth noting that this is, to our knowledge, the first empirical confirmation that the chainsaw model is valid in rivers (at least hydraulically-wide ones).
-This is elaborated on in section 6.1.
+This is elaborated on in section 5.1.
 
-The success of the chainsaw model in hydraulically-wide channels, and in particular when coupled with the form-drag dissipation model (equation 6, Figure 3d), provides us with a strong physical-model for gas evasion.
-The river hydraulics terms in equation 6 can either be directly measured or reasonably inferred from SWOT measurements, effectively opening the door for remotely sensing gas exchange.
+The success of the chainsaw model in hydraulically-wide channels, and in particular when coupled with the form-drag dissipation model (equation 8, Figure 3d), provides us with a strong physical-model for gas evasion.
+The river hydraulics terms in equation 8 can either be directly measured or reasonably inferred from SWOT measurements, effectively opening the door for remotely sensing gas exchange.
 This is explored next.
 
-## 5 Exploiting equation 6 to remotely sense gas exchange velocity
-We have shown that scaling $k_{600}$ via equation 6 explains `r round(modelsWIDE[modelsWIDE$name == 'Chainsaw-eD',]$r2,2)*100`% of the variation in $k_{600}$ in hydraulically-wide rivers, which includes nearly all SWOT-observable rivers.
-Further, equation 6 has only three non-remotely-sensible terms: $k_{600}$, mean flow depth, and mean flow velocity.
-Conveniently, inferring mean flow velocity and mean flow depth from SWOT measurements is a well-established domain for workers estimating streamflow from SWOT.
-For BIKER, we follow the work developed by @durandEstimatingReachaveragedDischarge2014, @hagemannBAMBayesianAMHGManning2017, and @brinkerhoffConstrainingRemoteRiver2020 to infer mean flow velocity and depth from SWOT observations using a modified form of the Manning's equation for open-channel flow velocity.
-These papers conceptualize streamflow as a Bayesian remote sensing problem, which we largely follow here to conceptualize $k_{600}$ as a Bayesian remote sensing problem that can be solved using SWOT data.
-First, we detail algorithm development and experimental design (section 5.1) and then we present the validation results (section 5.2).
+## 4 Exploiting equation 8 to remotely sense gas exchange velocity
+We have shown that scaling $k_{600}$ via equation 8 explains `r round(modelsWIDE[modelsWIDE$name == 'Chainsaw-eD',]$r2,2)*100`% of the variation in $k_{600}$ in hydraulically-wide rivers, which includes nearly all SWOT-observable rivers.
+Further, equation 8 has only three non-remotely-sensible terms: $k_{600}$, mean flow depth, and mean flow velocity.
+Conveniently, techniques to simultaneously infer hydraulic parameters from SWOT dta have been established over the last decade to infer streamflow from SWOT's measurements [e.g.
+@andreadisConstrainingAssimilationSWOT2020; @brinkerhoffConstrainingRemoteRiver2020; @durandEstimatingReachaveragedDischarge2014; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014a; @garamboisVariationalEstimationEffective2020; @brissetAssimilationAltimetricData2018; @hagemannBAMBayesianAMHGManning2017; @larnierRiverDischargeBathymetry2020a; @oubanasDischargeEstimationUngauged2018].
+For BIKER, we follow the work developed by @durandEstimatingReachaveragedDischarge2014, @hagemannBAMBayesianAMHGManning2017, and @brinkerhoffConstrainingRemoteRiver2020 to infer mean flow velocity and depth from SWOT observations using a modified form of the Manning's equation for mean flow velocity.
+First, we detail algorithm development and experimental design (section 4.1) and then we present the validation results (section 4.2).
 
-### 5.1 Section 5 methods
-#### 5.1.1 BIKER
-BIKER, and Bayesian inference in general, starts from Bayes rule (equation 7), where $\Theta$ is some set of non-remotely-sensible parameters we want to solve for (including $k_{600}$), *x* is the observed data, $f(x|\Theta)$ is the sampling model where data are conditional on the parameters, and $p(\Theta)$ is the joint prior distribution of the parameters.
+### 4.1 Section 4 methods
+#### 4.1.1 BIKER
+BIKER, and Bayesian inference in general, starts from Bayes rule (equation 9), where $\Theta$ is some set of non-remotely-sensible parameters we want to solve for (including $k_{600}$), *x* is the observed data, $f(x|\Theta)$ is the sampling model where data are conditional on the parameters, and $p(\Theta)$ is the joint prior distribution of the parameters.
 For BIKER, *x* is the SWOT-observables: river width *W* and water surface elevation $H_e$ (which is used to calculate the water surface slope *S*), while $\Theta$ are the parameters $k_{600}$, *n*, and $A_0$.
@@
 
-$$p(\Theta|x) = \frac{f(x|\Theta)p(\Theta)}{p(x)} \mathbf{(7)}$$
-
-The heart of BIKER is its reformulation of the $k_{600}$ model (equation 6) as a Bayesian sampling model that is conditional on the non-remotely-sensible parameters (i.e.
+$$p(\Theta|x) = \frac{f(x|\Theta)p(\Theta)}{p(x)} \mathbf{(9)}$$
+The heart of BIKER is its reformulation of the $k_{600}$ model (equation 8) as a Bayesian sampling model that is conditional on the non-remotely-sensible parameters (i.e.
 $f(x|\Theta)$).
 To do this, we first conceptualize a mass-conserved river reach, where hydraulic properties vary along cross-sections *i* within the reach but the entire mass of water moving through is conserved and thus only varies with timestep *t* [@gleasonTrackingRiverFlows2017].
-Consult Figure S2 for a scematic of this model.
-This means that the reach-averaged $k_{600}$, which is to some degree a function of the mass-conserved streamflow [@wangPhysicallyBasedScaling2021], will only vary temporally as well.
+This means that the reach-averaged $k_{600}$, which is a function of the channel hydraulics per equations 5-8 [@wangPhysicallyBasedScaling2021], will only vary temporally as well.
 Under this conceptual river reach model, BIKER predicts a time-varying, reach-averaged $k_{600_t}$.
@@
 Under this conceptual river reach, we write $k_{600_t}$ as a function of SWOT-observables $W_{i,t}$ and $S_{i,t}$.
-This algebra is carried out using equation 6, the fitted value for $\beta_1$ from Figure 3d (`r round(modelsWIDE[modelsWIDE$name == 'Chainsaw-eD',]$coef,2)`), and Manning's equation for mean flow velocity ($\bar{U}=\frac{1}{n}R_h^{2/3}S^{1/2}$).
-This algebra yields equation 8, assuming that the channel is hydraulically-wide ($R_h = H =\frac{A}{W}$).
+This algebra is carried out using equation 8, the fitted value for $\beta_1$ from Figure 3d (`r round(modelsWIDE[modelsWIDE$name == 'Chainsaw-eD',]$coef,2)`), and Manning's equation for mean flow velocity ($\bar{U}=\frac{1}{n}R_h^{2/3}S^{1/2}$).
+This algebra yields equation 10, assuming that the channel is hydraulically-wide ($R_h = H =\frac{A}{W}$).
 To leverage additional SWOT data, the wetted channel area *A* is further split into the SWOT-observable portion $dA_{i,t}$ and the unobservable portion $A_0_i$ following @durandEstimatingReachaveragedDischarge2014.
@@
 
-$$\begin{eqnarray} k_{600_t} = 62.82 g^{7/16}S_{i,t}^{9/16}\frac{1}{n_i}^{1/4}(\frac{A_0_i+dA_{i,t}}{W_{i,t}})^{35/48} \end{eqnarray} \mathbf{(8)}$$
+$$\begin{eqnarray} k_{600_t} = 62.82 g^{7/16}S_{i,t}^{9/16}\frac{1}{n_i}^{1/4}(\frac{A_0_i+dA_{i,t}}{W_{i,t}})^{35/48} \end{eqnarray} \mathbf{(10)}$$
+Next, equation 10 is written as a Bayesian sampling model, in which all of the SWOT observations are sampled from the unknown model parameters ($A_0_i$, $k_{600_t}$, and $n_i$).
+This is equation 11 after describing them as log-normal distributions.
+$\sigma_{k_{600}}^2$ refers to the total uncertainty implicit in equation 10.
+This uncertainty is due to parameter uncertainty from equation 8, Manning's equation, and the rectangular channel assumption.
 
-Next, equation 8 is written as a Bayesian sampling model, in which all of the SWOT observations are sampled from the unknown model parameters ($A_0_i$, $k_{600_t}$, and $n_i$).
-This is equation 9 after describing them as log-normal distributions.
-$\sigma_{k_{600}}^2$ refers to the total uncertainty implicit in equation 8.
-This uncertainty is due to parameter uncertainty from both the chainsaw/form-drag dissipation model and the Manning's model for $\bar{U}$.
-
-$$\left[log(62.82) + \frac{7}{16}log(g) + \frac{9}{16}log(S_{i,t}) - \frac{35}{48}log(W_{i,t})\right] \sim normal\left(\left[log(k_{600_t})+\frac{1}{4}log(n_i)-\frac{35}{48}log(A_0_i+dA_{i,t})\right], \sigma^2_{k_{600}}\right) \mathbf{(9)}$$
-
-Equations 7 and 9 also necessitate that we specify prior distributions for the parameters $A_0_i$, $k_{600_t}$, and $n_i$.
+$$\left[log(62.82) + \frac{7}{16}log(g) + \frac{9}{16}log(S_{i,t}) - \frac{35}{48}log(W_{i,t})\right] \sim normal\left(\left[log(k_{600_t})+\frac{1}{4}log(n_i)-\frac{35}{48}log(A_0_i+dA_{i,t})\right], \sigma^2_{k_{600}}\right) \mathbf{(11)}$$
+Equations 9 and 11 also necessitate that we specify prior distributions for the parameters $A_0_i$, $k_{600_t}$, and $n_i$.
 Prior distributions, described by their hyperparameters, formalize the a priori estimates (and uncertainties) for the non-remotely-sensed terms.
@@
 Our goal in prior specification was to rely on absolutely no in situ information such that we could run this method on any river on Earth solely using SWOT observations.
-In theory, more informed priors via various a priori information about a specific river would improve BIKER performance, but we wanted to test the fully generalized algorithm in this paper.
+In theory, more informed priors via various a priori information about a specific river would improve BIKER performance, but here we chose to test only the fully generalized algorithm.
 Therefore, the validation presented here is a 'worst-case scenario', wherein BIKER performance will improve with better prior information on the river.
@@
 
-With the sampling model (equation 9) and hyperparameters described (Text S3), a joint posterior distribution conditional on the SWOT observations ($p(\Theta|x) \propto f(x|\Theta)p(x)$) is necessarily specified.
+With the sampling model defined ($f(x|\Theta)$ = equation 11) and hyperparameters $p(x0$) described (Texts S3 and S4), a joint posterior distribution conditional on the SWOT observations ($p(\Theta|x) \propto f(x|\Theta)p(x)$) is specified.
 To approximate this distribution, we use a Markov Chain Monte Carlo (MCMC) algorithm implemented using the Stan probabilistic programming language.
@@
 
-#### 5.1.2 BIKER validation
-We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers (section 2) using daily observed hydraulics and the metrics described in Table S2.
+#### 4.1.2 BIKER validation
+We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers (section 2) using daily observed hydraulics.
 We also re-validate BIKER on the `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` rivers with the SWOT error model which corrupts the SWOT measurements (widths, heights, and slopes).
 
-Regardless of the validation setup, we do not have observed $k_{600_t}$ data for these rivers and, to our knowledge, no field dataset of $k_{600}$ exists in the type of temporal and spatial frequency that SWOT (and therefore the BIKER algorithm) provides.
-Thus, we take the model outlined in equation 6 and Figure 3d and use that to calculate the observed $k_{600_t}$ that BIKER is validated against (equation 10).
-Note that equation 9 and 10 both treat $\beta_1$ (62.82) as known and not a parameter in order to directly explore BIKER's ability to infer observed $A_0_i$, $k_{600_t}$, and $n_i$ simultaneously from $W_{i,t}$ and $H_{i,t}$ alone.
-This was deemed acceptable given that the $k_{600}$ model (equation 6) has already been successfully validated in Figure 3d.
+We do not have observed $k_{600_t}$ data for these rivers and, to our knowledge, no field dataset of $k_{600}$ exists in the type of temporal and spatial frequency that SWOT (and therefore the BIKER algorithm) provides.
+Thus, we take the model outlined in equation 7 and Figure 3d and use that to calculate the observed $k_{600_t}$ that BIKER is validated against (equation 12).
+This directly explores BIKER's ability to infer observed $A_0_i$, $k_{600_t}$, and $n_i$ simultaneously from $W_{i,t}$ and $H_{i,t}$ alone and was deemed acceptable given that the $k_{600}$ model (equation 7) has already been successfully validated in Figure 3d.
 
-$$k_{600_{obs}}=62.82(gS_{obs})^{7/16}\bar{U}_{obs}^{1/4}H_{obs}^{9/16} \mathbf{(10)}$$
+$$k_{600_{obs}}=62.82(gS_{obs})^{7/16}\bar{U}_{obs}^{1/4}H_{obs}^{9/16} \mathbf{(12)}$$
+It also means that, for a fair validation scheme, the parameter specification for $\sigma^2_{k_{600}}$ in equation 11 should not include the uncertainty from equation 8.
+So for this validation only, $\sigma^2_{k_{600}}$ is set to 0.30 (natural-log space) to reflect uncertainty from Manning's equation and assuming a rectangular channel (Text S4).
+However, if BIKER is run on real SWOT observations after the satellite launches, $\sigma^2_{k_{600}}$ should reflect the total uncertainty, which is described in Text S4.
 
-It also means that, for a fair validation scheme, $\sigma^2_{k_{600}}$ in equation 9 must reflect only error from using Manning's equation and our calculation of *dA* and not the parameter uncertainty inherent in $\beta_1$.
-Because this model is only theoretically valid when $R_h=H$, we conveniently do not need to account for uncertainties associated with assuming that $R_h = H$.
-So for this validation, $\sigma^2_{k_{600}}$ is set to 0.30 (natural-log space) to reflect uncertainty from Manning's equation and assuming a rectangular channel (Text S4).
-However, if BIKER is run on real SWOT observations after the satellite launches, $\sigma^2_{k_{600}}$ should reflect the total uncertainty, which is described in Text S4.
-Note that when run on real SWOT observations, we could also treat $\beta_1$ as a parameter to solve for.
-This would complicate the model to a degree, but would also allow for a robust posterior estimate of $\beta_1$'s value.
-This is left to future work once SWOT launches.
-
-With the setup described by equations 9 and 10, validation is performed using the BIKER posterior means.
+With the setup described by equations 11 and 12, validation is performed using the BIKER posterior means.
 Because we are most interested in BIKER's ability to correctly infer temporal dynamics in $k_{600}$, and each river's priors are described specifically for that river, we validate by river.
 These metrics are defined in Table S2 and consist of the correlation coefficient *r* to quantify accuracy of BIKER's temporal dynamics, the root mean square error normalized by the observed mean (NRMSE) and prediction bias normalized by the observed mean (rBIAS) to assess bias, and the Kling-Gupta Efficiency (KGE).
-KGE is a commonly used metric to assess streamflow prediction and simultaneously assesses accuracy in predicting both bias and dynamics.
+KGE is frequently used to assess streamflow prediction and simultaneously assesses accuracy in both bias and dynamics.
 While a value greater than -0.41 means the model outperforms a uniform prediction of the mean [@knobenTechnicalNoteInherent2019], generally KGE scores are interpreted as being meaningful in ungauged settings if > 0.
 
-#### 5.1.3 Carbon efflux validation
+#### 4.1.3 Carbon efflux validation
 It is one thing to accurately model the temporal dynamics of $k_{600}$, but researchers are often most interested in the actual gas fluxes from rivers and ultimately the carbon emitted from river to atmosphere.
@@
 We then calculate $FCO_2$ using equation 1 and assuming atmospheric $CO_{2_{air}}$ is 390 uatm.
-Equation 11 is used to obtain a temperature and $CO_2$ specific gas exchange velocity from BIKER's output, where the Schmidt number *Sc* was estimated following @raymondScalingGasTransfer2012.
+Equation 13 is used to obtain a temperature and $CO_2$ specific gas exchange velocity from BIKER's output, where the Schmidt number *Sc* was estimated following @raymondScalingGasTransfer2012.
 We then estimate median and mean bulk carbon effluxes by multiplying the median or mean $FCO_2$ value across all rivers and timesteps (in Tg-C/yr) by the total average surface area of all 47 rivers.
 
-$$\mathbf{(11)} k_{CO2, temp}=k_{600}*(Sc/600)^{-1/2}$$
-
+$$k_{CO2, temp}=k_{600}*(Sc/600)^{-1/2} \mathbf{(13)}$$
 Finally, we also compare BIKER's outputs with established in situ methods, whose specific details and parameters are provided in Table S3.
@@
 
-### 5.2 Section 5 results
-#### 5.2.1 BIKER
+### 4.2 Section 4 results
+#### 4.2.1 BIKER
 First, we explore whether general trends are correctly captured by BIKER, assuming no SWOT measurement error.
@@
 
-We calculate performance metrics following section 5.1.2 and Table S2.
+We calculate performance metrics following section 4.1.2 and Table S2.
 These are presented in Figure 6a, which plots validation metrics calculated for each river with and without SWOT measurement error (green and purple, respectively).
@@
 
-#### 5.2.2 Carbon efflux
+#### 4.2.2 Carbon efflux
 Next, we compare BIKER's estimates of carbon fluxes against established in situ techniques.
@@
 
-## 6 Discussion
-### 6.1 Gas exchange in hydraulically-wide rivers
+## 5 Discussion
+### 5.1 Gas exchange in hydraulically-wide rivers
 **REWRITE THIS**
@@
 
-### 6.2 Towards remote sensing of global spatiotemporal dynamics of *k* in large rivers
+### 5.2 Towards remote sensing of global spatiotemporal dynamics of *k* in large rivers
 To date, the studies exploring the spatiotemporal dynamics of riverine gas exchange have arguably been held back by a lack of data.
@@
 These differences in performance are likely due to the representativeness of the priors used for that river.
-For SWOT discharge algorithms, authors have repeatedly shown that the 'quality' of prior hyperparameters plays a large role in the success of one's discharge inversions [@frassonExploringFactorsControlling2021; @andreadisConstrainingAssimilationSWOT2020; @brinkerhoffConstrainingRemoteRiver2020; @tuozzoloEstimatingRiverDischarge2019] and our results here further corroborate this finding.
+For SWOT discharge algorithms, authors have repeatedly shown that the 'quality' of prior hyperparameters plays a large role in the success of discharge inversions [@frassonExploringFactorsControlling2021; @andreadisConstrainingAssimilationSWOT2020; @brinkerhoffConstrainingRemoteRiver2020; @tuozzoloEstimatingRiverDischarge2019] and our results here further corroborate this finding.
 It should be stressed that a substantial portion of rivers from [@frassonExploringFactorsControlling2021] are canal-shaped in nature, suggesting different hydraulic properties than a natural river channel.
@@
 
-### 6.3 Towards remotely sensing riverine carbon emissions using SWOT
+### 5.3 Towards remotely sensing riverine carbon emissions using SWOT
 Figure 7 confirms that BIKER is quite successful, without any in situ information aside from a $CO_2$ logger, at predicting both $FCO_2$ (Figure 6a) and 2) the bulk carbon efflux (Figure 7b-7c).
@@
 
-## 7 Conclusions
+## 6 Conclusions
 Gas exchange from aquatic systems has been studied for nearly a century and dozens of predictive equations have been developed to obtain *k* values from easily measured hydraulic properties rather than via direct measurement.
-Likewise, equation 2 has been repeatedly verified across many non-fluvial environments.
+Likewise, equation 3 has been repeatedly verified across many non-fluvial environments.
 However.
@@
 
-## 8 Acknowledgements
+## 7 Acknowledgements
 C.B.
@@
 
-## 9 Apendix A
+## 8 Apendix A
 *Table A1: Variable description and notation for this study. ADD THE BAYES PARAMETERS YET*
@@
 |$k_{600}$|gas exchange velocity normalized to $Sc=600$|$(\frac{600}{Sc_k})^{-0.5}k$ [@raymondScalingGasTransfer2012]|$[\frac{L}{T}]$|
+|*$k_{model}$|riverine gas exchange velocity estimated by predictive models|NA|$[\frac{L}{T}]$|
 |$n$|Manning's roughness coefficient|$\frac{R_h^{2/3}S^{1/2}}{\bar{U}}$|$[\frac{T}{L^3}]$|

### 2022-06-08 docs/intial_drafts/supp_v11.Rmd
first round of review, setting up for rerun on HPC

--- 2022-04-05 supp_v11.Rmd
@@
 That dataset has over 500,000 discrete measurements of river width, velocity, area, and discharge that were made by the United States Geological Survey (USGS) to calibrate streamgauge rating curves.
-Here, we describe how this dataset was filtered down to `r HG_swot$n` rivers.
+Here, we describe how this dataset was filtered down to `r HG_swot$n` SWOT-observable rivers.
 
@@
 
-**Small-eddy models (Equation 5 from Main Test)**
+**Small-eddy models (Equation 3 from Main Test)**
 
@@
 
-**Reynolds extension models (Equation 6 from Main Text)**
+**Reynolds extension models (Equation 4 from Main Text)**
 
@@
 For BIKER, prior distributions are formalized as truncated normal distributions of the log-transformed terms such that $log(X) \sim  \mathcal{N}(\mu, \sigma^2)$ for $\lambda < log(X) < \gamma$, using prior hyperparameters mean ($\mu$), standard deviation ($\sigma$), and upper ($\gamma$) and lower bounds ($\lambda$) for any parameter *X*.
-It is important to again stress, as we do in the main text, that BIKER prior hyperparameters are described using only SWOT data to be completely globally implementable.
+It is important to again stress, as we do in the Main Text, that BIKER prior hyperparameters are described using only SWOT data to be completely globally implementable.
 This use of the data to describe the priors is analogous to the 'empirical Bayes method' [@hoffFirstCourseBayesian2009].
@@
 This leaves the $k_{600_t}$ hyperparameters to be defined.
-$k_{600_{\mu}}$ is set by invoking the hydraulic geometry (HG) relationships developed in Section 3.2.3 of the main text and Table S2 using the data from @brinkerhoffReconcilingAtaStationAtManyStations2019.
+$k_{600_{\mu}}$ is set by invoking the hydraulic geometry (HG) relationships described in Section 3.2.3 of the main text and Table S2 using the data from @brinkerhoffReconcilingAtaStationAtManyStations2019.
 We replaced both depth and velocity terms from our gas exchange model (Equation 7 in the Main Text) with these HG models, resulting in Equation S19 where Q is the mass-conserved streamflow for the river reach.
@@
 Note that each Monte Carlo simulation is itself 10,000 runs in order to obtain a distribution of $k_{600}$ estimates.
-For each of the 166 distributions of estimated $l_{600}$, we extract the standard deviation as the uncertainty.
+For each of the 166 distributions of estimated $k_{600}$, we extract the standard deviation as the uncertainty.
 We then take the average of those standard deviations to be a reasonable estimate of $\sigma_k_{600}$.
@@
 
-$$log_{k_{600}} \sim normal \left[ log(\beta_1), log(\sigma_{\beta_1}) \right] + \frac{1}{4}normal\left[log(\bar{U}), log(\sigma_{mannings})\right] \frac{7}{16}log(gS) + \frac{9}{16}log(H)  \mathbf{(S20)}$$
+$$log_{k_{600}} \sim \mathcal{N} \left[ log(\beta_1), log(\sigma_{\beta_1}) \right] + \frac{1}{4}\mathcal{N}\left[log(\bar{U}), log(\sigma_{mannings})\right] \frac{7}{16}log(gS) + \frac{9}{16}log(H)  \mathbf{(S20)}$$
 

### 2022-06-08 docs/intial_drafts/supp_v5.Rmd
first round of review, setting up for rerun on HPC

--- 2022-01-07 docs/supp_v6.Rmd
@@
 ---
-title: "Supplemental Information: Exploiting hydraulic geometry to remotely sense gas exchange velocity via the SWOT satellite"
+title: "Supplemental Information: Gas exchange in large rivers influenced by multiple turbulence scales: implications for remotely sensing gas exchange via the SWOT satellite"
 author:
@@
 
+## Figure S2
+![Figure S2 Conceptual river reach model used in this study following McFLI logic [@gleasonTrackingRiverFlows2017] DOOOOOOOOO](cache/FCO2/Beaulieu_timeseries.jpg)
+
 ## Table S1

### 2022-06-08 docs/intial_drafts/supp_v9.Rmd
first round of review, setting up for rerun on HPC

--- 2022-02-24 supp_v9.Rmd
@@
 **Small-eddy models**
-
-$\epsilon = \epsilon_S$
+**Log-law-of-the-wall model for the turbulent dissipation rate**
 
@@
 
-$\epsilon = \epsilon_D$
+**Form-drag model for the turbulent dissipation rate**
 
@@
 **Reynolds extension models**
-
-$\epsilon = \epsilon_S$
+**Log-law-of-the-wall model for the turbulent dissipation rate**
 
@@
 
-Finally, recall that $\sigma_{k_{600}}$ refers to the total uncertainty inherent in equation 9, i.e.
+## Text S4: Detemrining complete BIKER model uncertainty
+Recall that $\sigma_{k_{600}}$ refers to the total uncertainty inherent in equation 9, i.e.
 stemming from the Reynolds-extension model (equation 7) and Manning's equation for $\bar{U}$.
@@
 This means we must also account for uncertainty from the $\beta_1$ parameter.
-Assuming perfect, no-error measurements are made by the SWOT satellite, the $k_{600}$ relative uncertainty is expressed for some set of hydraulic observations as equation S20.
-We use the 166 hydraulically-wide measurements in our field-measured $k_{600}$ dataset and run 166 different Monte Carlo simulations (each 10,000 runs) to obtain 166 different $\sigma_k_{600}$ terms.
-We then take the average of those to be a reasonable estimate of $\sigma_k_{600}$.
+Assuming perfect, no-error measurements are made by the SWOT satellite, the full equation 7 uncertainty is expressed for some set of hydraulic observations as equation S20.
+To quantify total uncertainty in $k_{600}$, we then push 166 different Monte Carlo simulations through equation S20 and take the mean uncertainty as a reasonable value for $\sigma_{k_{600}}$ once SWOT launches.
+
+To do this, we use the 166 hydraulically-wide measurements in our field-measured $k_{600}$ dataset (section 2 in the Main Text) and push 166 different Monte Carlo simulations through equation S20.
+NOte that each Monte Carlo simulation is itself 10,000 runs in order to obtain a distribution of $k_{600}$ estimates.
+For each of the 166 distributions of estimated $l_{600}$, we extract the standard deviation.
+We then take the average of those standard deviations to be a reasonable estimate of $\sigma_k_{600}$.
 This ultimately provided a value of 1.1 for the log-transformed $\sigma_k_{600}$ that should be used once SWOT launches.
@@
 
-## Text S4: Literature $CO_2$ flux models
+## Text S5: Literature $CO_2$ flux models
 Here, we describe in more detail the $CO_2$ flux models used to compare against BIKER in section 3.2.2. The specific equations used are outlined in Table S3.
@@
 For example, the 'Raymond 2012' model uses the hydraulic scaling equations from @raymondScalingGasTransfer2012 to predict velocity and depth, which are in turn used to calculate a $k_{CO_2}$, which finally is used to calculate $FCO_2$.
-Similarly, the 'Raymond 2013' model uses the equations outlined in @raymondGlobalCarbonDioxide2013 and the 'Brinkerhoff 2019' model uses new equations fit to the hydraulics dataset from @brinkerhoffReconcilingAtaStationAtManyStations2019 following Text S1.
+Similarly, the 'Raymond 2013' model uses the equations outlined in @raymondGlobalCarbonDioxide2013 and the 'Brinkerhoff 2019' model uses new equations fit to the hydraulics dataset from @brinkerhoffReconcilingAtaStationAtManyStations2019 after the filtering descirbed in Text S1.
 We chose to include the 'Brinkerhoff 2019' model as the training dataset is far larger than those used in either of the previous two models (Table S3: 104,624 versus 10,837 versus 1,026, respectively).
@@
 |---------------|-----------|--------------|---------------|------------------|
-|Correlation Coefficient|$r$| $\sqrt{1-(\frac{\sum\limits_{i=1}^{N_t}{(k_{600, i}-\hat{k_{600_i}})^2}}{\sum{(k_{600, i}-\bar{k_{600}})^2}})}$ |1|-1 to 1|
-|Normalized root-mean-square error|NRMSE| $\sqrt{\frac{1}{N_t}\sum\limits_{i=1}^{N_t}(\frac{\hat{k_{600, i}}-k_{600, i}}{\bar{k_{600, i}}})^2}$ |0|0 to ∞|
-|Kling-Gupta Efficiency|KGE|$1 - \sqrt{ (r(k_{600_i}, \hat{k_{600_i}}) - 1)^2 + (\frac{\mu_{\hat{k_{600_i}}} }{\mu_{k_{600_i}}} - 1)^2 + (\frac{ \sigma _{\hat{k_{600_i}}} / \mu _{\hat{k_{600_i}}} }{ \sigma_{k_{600_i}} / \mu_{k_{600_i}}} - 1)^2 }$|1|-∞ to 1|
-|Mean Absolute Error|MAE| $\frac{\sum\limits_{i=1}^{N_t}(|\hat{k_{600_i}}-k_{600_i})|)}{n})$ |0|0 to ∞|
+|Correlation Coefficient|$r$| $\sqrt{1-(\frac{\sum\limits_{i=1}^{N_t}{(k_{600, i}-\hat{k_{600_i}})^2}}{\sum{(k_{600, i}-\mu_{k_{600}})^2}})}$ |1|-1 to 1|
+|Normalized root-mean-square error|NRMSE| $\sqrt{\frac{1}{N_t}\sum\limits_{i=1}^{N_t}(\frac{\hat{k_{600, i}}-k_{600, i}}{\mu_{k_{600, i}}})^2}$ |0|0 to ∞|
+|Kling-Gupta Efficiency|KGE|$1 - \sqrt{ (r(k_{600_i}, \hat{k_{600_i}}) - 1)^2 + (\frac{\mu_{\hat{k_{600}}} }{\mu_{k_{600}}} - 1)^2 + (\frac{ \sigma _{\hat{k_{600}}} / \mu _{\hat{k_{600}}} }{ \sigma_{k_{600}} / \mu_{k_{600}}} - 1)^2 }$|1|-∞ to 1|
+|Normalized Mean Absolute Error|MAE| $\frac{\frac{\sum\limits_{i=1}^{N_t}(|\hat{k_{600_i}}-k_{600_i})|)}{n})}{\mu_{k_{600}}}$ |0|0 to ∞|
 

### 2022-06-08 manuscript_r_v1.Rmd
first round of review, setting up for rerun on HPC

--- 2022-06-08 docs/intial_drafts/manuscript_v11.Rmd
@@
 csl: agu_citation_style.csl
-output:
-  word_document:
+output:redoc::redoc
+  #word_document:
     #  pandoc_args:
@@
     #    - '--lua-filter=author-info-blocks.lua'
-      reference_docx: word-styles-reference.docx
+  #    reference_docx: word-styles-reference.docx
 ---

### 2022-06-08 supp_r_v1.Rmd
first round of review, setting up for rerun on HPC

--- 2022-06-08 docs/intial_drafts/supp_v11.Rmd
@@
 csl: agu_citation_style.csl
-output:
-  word_document:
+output:redoc::redoc
+  #word_document:
   #    pandoc_args:
@@
   #      - '--lua-filter=author-info-blocks.lua'
-      reference_docx: word-styles-reference.docx
+  #    reference_docx: word-styles-reference.docx
 

### 2022-06-08 docs/review_round2/k_model_for_reviewer1.R

@@ -16,6 +16,8 @@ library(ggplot2)
 library(colorspace)
 library(cowplot)
 library(readr)
+library(rstanarm)
+library(grid)
 theme_set(theme_classic())
 
 #some constants
@@ -121,181 +123,169 @@ data$log_slope <- log(data$slope)
 ######## K600 MODELS--------------------
 #hydraulically-wide rivers
 data <- filter(data, study %in% c('Ulseth_etal_2019', 'Churchill_etal_1962', 'Owens_etal_1964'))
+out <- data.frame('width_min'=min(data$width),
+                  'width_max'=max(data$width),
+                  'width_mean'=mean(data$width),
+                  'Q_min'=min(data$Qm3s),
+                  'Q_max'=max(data$Qm3s),
+                  'Q_mean'=mean(data$Qm3s),
+                  'k600_min'=min(data$k600),
+                  'k600_max'=max(data$k600),
+                  'k600_mean'=mean(data$k600))
+write.csv(out, 'cache/k600_dataStats.csv') #save dataset stats to file
 
 #determine if hydraulically wide
 data$flag_hydraulicWide <- ifelse(data$Rh/data$depth >= 0.99, 'Rh=H', 'Rh=/=H')
 
 hydraulicallyWide <- filter(data, flag_hydraulicWide == 'Rh=H')
 
-#fit small-eddy model with log-law-of-the-wall dissipation
-hydraulicallyWide$hydraulicallyWideModel <- g^(3/8)*hydraulicallyWide$slope^(3/8)*hydraulicallyWide$depth^(1/8)
-lm_hydraulicallyWide_smallEddy_eS <- lm(k600~hydraulicallyWideModel+0, data=hydraulicallyWide)
-hydraulicallyWide$k600_pred_wideHydraulics <- predict(lm_hydraulicallyWide_smallEddy_eS, hydraulicallyWide)
+#tests for reviewer 1############################################
+#test 1
+hydraulicallyWide$log_hydraulicallyWideModel <- (7/16)*log(g*hydraulicallyWide$slope) + (1/4)*log(hydraulicallyWide$Vms) + (9/16)*log(hydraulicallyWide$depth)
+beta_1 <- mean(c(log(hydraulicallyWide$k600) - hydraulicallyWide$log_hydraulicallyWideModel))
+hydraulicallyWide$logk600_pred_wideHydraulics <- beta_1+hydraulicallyWide$log_hydraulicallyWideModel
[…]

### 2022-06-08 generateMS.R

@@ -12,8 +12,8 @@ results_fco2_bulk <- read.csv('cache//FCO2//massFluxes.csv')
 results_fco2_bulk$key <- as.character(results_fco2_bulk$key)
 
 #knit manuscript
-rmarkdown::render("manuscript_v11.Rmd")
+rmarkdown::render("manuscript_r_v1.Rmd")
 print('Manuscript generated.')
 
-rmarkdown::render("supp_v11.Rmd")
+rmarkdown::render("supp_r_v1.Rmd")
 print('Supplemental Information generated.')

### 2022-06-08 src/calcSWOT_SA.R

@@ -1,0 +1,50 @@
+##################
+## Calculates SWORD/SWOT mean annual surface area (widths come from GRWL and /or MERIT Hydro) and network length
+## Does eachcontinent seperately to spare holding the global river network in memory...
+## Craig Brinkerhoff
+## Summer 2022
+###################
+
+library(sf)
+
+#path to SWORD v6
+setwd('C:\\Users\\cbrinkerhoff\\OneDrive - University of Massachusetts\\Datasets\\SWORD_v06\\Reaches_Nodes\\shp')
+
+########################################SWOT/SWORD########################################################################
+#get files
+files <- list.files(pattern = '*.shp', recursive = TRUE)
+files <- files[substr(files, 22, 28) == 'reaches']
+
+#run per continent, concatenating SA
+SA_fin <- 0
+length_fin <- 0
+continent <- unique(substr(files, 1, 2))
+for (k in continent){
+  files2 <- files[substr(files, 1, 2) == k]
+  
+  #concatenate shapefiles
+  sword <- st_read(files2[1], quiet=TRUE)
+  files2 <- files2[-1]
+  for(i in files2){
+    temp_sf <- st_read(i, quiet=TRUE)
+    sword <- rbind(sword, temp_sf)
+  }
+  
+  #estimate SA
+  SA <- sum(sword$width * sword$reach_len, na.rm=T) #[m2]
+  SA <- SA*1e-6 #[km2]
+  SA_fin <- SA_fin + SA
+  print(paste0(k, ' SA: ', SA))
+  print(paste0('total SA: ', SA_fin))
+  
+  #estimate length
+  length <- sum(sword$reach_len) #[m]
+  length <- length / 1000 #[km]
+  length_fin <- length + length_fin
+  print(paste0(k, ' Length: ', length))
+  print(paste0('total Length: ', length_fin))
+}
+
+#calc percent SWOT observable
+SA_fin/811000 #Liu et al 2022
+length_fin/23300000 #riverATLAS Messager et al 2021

### 2022-06-08 src/runBIKER.R

@@ -102,7 +102,7 @@ run_BIKER <- function(currPepsi, errFlag) {
   #run BIKER
   data <- biker_data(w=W_obs, s=S_obs, dA=dA_obs, priorQ=as.matrix(priorQ))
   priors <- biker_priors(data)
-  priors$river_type_priors$logk_sd <- rep(0.30, ncol(W_obs))
+  priors$river_type_priors$logk_sd <- rep(0.25, ncol(W_obs)) #obtained from Bayesian linear regression model
   priors$sigma_model$sigma_post = matrix(uncertainity, nrow=nrow(W_obs), ncol=ncol(W_obs)) #For this validation, we only want Rh uncertainty. Real implementation would use full model uncertainty (calculate in '~src\swot_k_model.R')
   kest <- biker_estimate(bikerdata = data, bikerpriors = priors, meas_err=F,iter = 3000L) #measurement error is currently in devolpment, don't use for this study
 

### 2022-07-27 manuscript_r2.Rmd
riewew round 2 big push

---
title: "Remotely sensing river greenhouse gas exchange velocity using the SWOT satellite"
author:
  - Craig B Brinkerhoff:
      email: cbrinkerhoff@umass.edu
      institute: UMass
      correspondence: true
  - Colin J Gleason:
      institute: UMass
  - Christopher J Zappa:
      institute: Lamont
  - Peter A Raymond:
      institute: Yale
  - Merritt E Harlan:
      institute: UMass
institute:
  - UMass: Department of Civil & Environmental Engineering, University of Massachusetts, Amherst, MA
  - Lamont: Lamont-Doherty Earth Observatory, Columbia University, Palisades, NY
  - Yale: School of the Environment, Yale University, New Haven, CT
bibliography: References.bib
csl: agu_citation_style.csl
output:
  word_document:
      pandoc_args:
        - '--lua-filter=scholarly-metadata.lua'
        - '--lua-filter=author-info-blocks.lua'
      reference_docx: word-styles-reference.docx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Key Points

-   BIKER  (Bayesian Inference of the $K_{600}$ Evasion Rate) algorithm predicts gas exchange velocity from simulated SWOT data without calibration
-   BIKER is marginally influenced by expected SWOT measurement errors
-   BIKER and near-daily SWOT data will allow for the study of gas exchange spatiotemporal dynamics at novel temporal resolutions

## Keywords

gas exchange, fluvial geomorphology, remote sensing, open-channel flow, SWOT, biogeochemistry

## Abstract

Two decades of research has shown that the global river network emits significant amounts of greenhouse gas. Despite much progress, there is still large uncertainty in the temporal dynamics of gas exchange and thus carbon emissions to the atmosphere. Much of this uncertainty stems from a lack of existing tools for studying the spatiotemporal dynamics of gas exchange velocity $k_{600}$ (the rate of this diffusive transport). We propose that the NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite can provide new insights to fluvial gas exchange modeling upon launch and subsequent data collection in 2022. Here, we exploit the distinct geomorphology of SWOT-observable rivers (> 50m wide) to develop a physical model of gas exchange that is remotely sensible and explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2,2)*100`% of log-transformed variation across `r n` field measurements of $k_{600}$. We then couple this model with established inversion techniques to develop BIKER, the 'Bayesian Inference of the $k_{600}$ Exchange Rate' algorithm. We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers without an in-situ calibration, yielding an algorithm that predicts the $k_{600}$ timeseries solely from SWOT observations with a by-river median Kling-Gupta Efficiency of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$KGE), 2)`. BIKER is better at inferring the temporal variation of gas exchange (median correlation coefficient of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$r), 2)`), than reproducing the absolute rates of exchange (median normalized RMSE of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$NRMSE), 2)*100`%). Finally, BIKER is robust to measurement errors implicit in the SWOT data. We suggest that BIKER will be useful in mapping global-scale fluvial gas exchange and improving $[CO_2]$ emissions estimates when coupled with river $[CO_2]$ models.

## 1 Introduction

Rivers play a critical role in the budgeting and accounting of the global carbon cycle under climate change. Following @colePlumbingGlobalCarbon2007, the global river network is recognized to emit substantial amounts of carbon to the atmosphere via evasion (gas exchange driven by diffusion and near-surface turbulence at the air/water interface), in addition to their long understood role in transporting carbon to the oceans via downstream advection. Current estimates of total carbon dioxide evasion ($[FCO_2]$) to the atmosphere from the global river network vary from 1800-2000 Tg C/yr [@liuImportanceHydrologyRouting2022; @raymondGlobalCarbonDioxide2013], with 650 Tg-C/yr and 167 Tg-C/yr coming from mid-to-large rivers [@lauerwaldSpatialPatternsCO22015] and mountain streams [@horgbyUnexpectedLargeEvasion2019], respectively. Despite rivers' small percentage of the global land surface [0.67%- @liuImportanceHydrologyRouting2022], this carbon flux is on par with the total oceanic $[CO_2]$ uptake rate [2600 Tg-C/yr- @gruberOceanicSinkAnthropogenic2019; @horgbyUnexpectedLargeEvasion2019] and the global forest carbon uptake rate [2400 Tg-C/yr- @panLargePersistentCarbon2011].

River $[CO_2]$ evasion is increasingly better constrained and is clearly a critical component of the global carbon cycle. Equation 1 represents this riverine flux given $\Delta [CO_2]$ ($[CO_{2_{water}}] - [CO_{2_{air}}]$) and the gas exchange velocity *k*. Consult Appendix A for notation used throughout this study.

$$[FCO_2]=k \Delta [CO_2] \mathbf{(1)}$$

There is a robust existing literature exploring spatiotemporal patterns in $\Delta [CO_2_{water}]$ [e.g. @ahoDistinctConcentrationdischargeDynamics2021; @ahoDifferentialResponseGreenhouse2019; @crawfordCO2TimeSeries2017; @peterScalesDriversTemporal2014; @ranDynamicsRiverineCO22017; @rocher-rosLandscapeProcessDomains2019; @raymondAtmosphericCO2Evasion2000; @liuHydrologicControlsPCO22018]. This work has lead to recent river-reach explicit modeling of $\Delta [CO_2]$ using global hydrography datasets at up to monthly temporal resolutions [@horgbyUnexpectedLargeEvasion2019; @brinkerhoffLakeMorphometryRiver2021; @saccardiImprovingPredictionsStream2021; @liuImportanceHydrologyRouting2022], but an equivalently sophisticated representation of $k_{CO_2}$ is still lacking.

The structure of Equation 1 necessitates that calculations of $[FCO_2]$ are highly sensitive to estimates of *k*. However, *k* can only be estimated via a known gas flux, eddy-covariance measurements, or tracer additions to the stream [@hallGasExchangeStreams2020]. In trying to constrain the global fluvial $[FCO_2]$ flux across millions of rivers, this calculation is impossible and necessitates the use of predictive models for *k* that are based on easily obtained river hydraulic properties. In that vein, over 20 empirical models exist to predict *k* from river hydraulics, generally using some combination of mean velocity $\bar{U}$, shear velocity $U_*$, width $W$, depth $H$, and slope $S$ as predictors [@hallGasExchangeStreams2020; @wangPhysicallyBasedScaling2021]. These models usually predict $k_{600}$, or *k* normalized by a Schmidt number (*Sc*) of 600 to remove the effect of water temperature and gas type from predictive models, as warmer waters and lower *Sc* numbers each increase gas exchange rates [@hallGasExchangeStreams2020]. Specifically, $k_{600}$ reflects the $k_{CO_{2}}$ at 20 degrees Celsius. Through this normalization, these models focus solely on physical explanations for variation in *k* [@hallGasExchangeStreams2020].

Applying these *k* models across watersheds, regions, or continents is called 'upscaling.' This upscaling allows for quantifying the difficult-to-estimate *k* term in Equation 1 for any arbitrary number of rivers, but also changes the base parameters that ultimately control the final estimate of $[FCO_2]$. That is, by making *k* a function of hydraulics, $[FCO_2_{upscaled}]$ is now a direct function of river hydraulics. This functional relationship is described in equation 2. It suggests that global $[FCO_2]$ estimates are not only at the mercy of the accuracy and spatiotemporal resolution of $\Delta [CO_2]$, but also the accuracy and resolution of our river hydraulics estimates.

$$[FCO_{2_{upscaled}}]=f(k_{CO_2}, \Delta [CO_2])=f(\bar{U}, H, S, W, \Delta [CO_2]) \mathbf{(2)}$$

Global upscaling has been performed using various techniques. @raymondGlobalCarbonDioxide2013, @lauerwaldSpatialPatternsCO22015, and @horgbyUnexpectedLargeEvasion2019 all relied on $k_{CO_2}$ values indirectly estimated using mean annual streamflow models and hydraulic scaling equations to predict the hydraulic terms used to in turn predict $k_{CO_2}$, while @borgesGloballySignificantGreenhousegas2015 used a combination of the above method and a constant $k_{CO_2}$ in space and time to upscale over Africa. More recently, @liuImportanceHydrologyRouting2022 performed a first upscaling assessment of monthly temporal dynamics in global river $[FCO_2]$, though they relied on monthly modeled streamflow and used the same model for $k_{600}$ as previous studies [@raymondGlobalCarbonDioxide2013]. In all of these foundational studies, the temporal dynamics of $k_{CO_2}$ (and thus dynamics in $[FCO_2]$) were ignored because of hydraulic data limitations. It has been shown at the field-scale that temporal dynamics of gas exchange can vary widely from site to site [@wallinSpatiotemporalVariabilityGas2011], but it has remained impractical to obtain temporally explicit $k_{CO_2}$ at continental-to-global scales.

@wangPhysicallyBasedScaling2021 recently attempted to address this global *k* problem by simulating the gas exchange velocity of dissolved oxygen ($k_{O_2}$) in 35 rivers of many sizes (widths ranging from 0.23-349m) using a stream metabolism inverse model [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen datasets to infer what $k_{O_2}$ must have been to produce their 'observations'. They then compared this simulated dataset against direct estimates of *k*, finding similar performance and parameter values for process-based models of gas evasion. However, they were still limited by a lack of direct hydraulic measurements and had to rely on hydraulic scaling equations to estimate river depth and velocity. Even though approaches like @wangPhysicallyBasedScaling2021's are incredibly useful for expanding our mechanistic understanding of gas exchange, they are less useful for global upscaling purposes as they rely on high fidelity in situ dissolved oxygen data for every river [@hallGasExchangeStreams2020].

We have established that literature has a reasonably good understanding of $\Delta [CO_2]$ and a relatively poorer understanding of $k_{600}$ (and therefore $k_{CO_2}$) across large areas and in time. In theory, the discrepancy between the quality of our $\Delta [CO_2_{water}]$ and $k_{CO_2}$ estimates could be alleviated if direct hydraulics measurements (and in turn $k_{CO_2}$ via Equation 2) were available at the global scale at a sufficient temporal resolution. Spatially and temporally dynamic hydraulic measurements in turn would also address the uncertainty regarding continental-to-global scale temporal dynamics of gas exchange noted earlier.

Conveniently, these hydraulic data will soon be available via the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission. SWOT is expected to launch in late 2022 and provide the world's first direct measurements of global water surface extent and elevation (and therefore water surface slope) at novel temporal resolutions. SWOT is a wide swath radar interferometer and will sample rivers every 1 to 7 days per 21 day repeat cycle, measuring rivers wider than 100m with a goal of expanding to rivers at least 50m wide [@biancamariaSWOTMissionIts2016]. Via its direct hydraulic measurements, SWOT is expected to usher in a sea change in global-scale hydrology, and could similarly influence fluvial biogeochemsitry if techniques are developed to ingest SWOT data and infer *k*. In this context, we borrow tools from fluvial geomorphology and existing SWOT algorithms to begin addressing the current knowledge gap in the spatiotemporal dynamics of gas exchange velocity. More specifically, we seek to answer the following two questions:

1)  What is the performance of a physically-based hydraulic model for $k_{600}$ unique to SWOT-observed rivers?

2)  How well can we exploit such a model to infer $k_{600}$ (and its uncertainty) solely from SWOT observations?

To answer the first question, we use hydraulic geometry- the fundamental geomorphic relationships between streamflow and channel shape [@leopoldHydraulicGeometryStream1953; @gleasonHydraulicGeometryNatural2015] to develop a process-based model for large-river $k_{600}$ (here defined as wider than 50m to align with SWOT). We then take these findings and explore the second question by implementing this hydraulic model, which defines $k_{600}$, within an algorithm named BIKER ('Bayesian Inference of the $k_{600}$ Evasion Rate') to infer $k_{600}$ solely from SWOT measurements. The goal of BIKER is to require no in situ inputs of any kind (although in situ data could be ingested and would improve results) such that it is globally implementable on any SWOT-observable river. We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers (as SWOT has not yet launched) and explore BIKER's robustness to the expected measurement errors implicit in the satellite's observations. Finally, we also couple BIKER's $k_{600}$ estimates with $\Delta [CO_2]$ to predict gas fluxes and compare these against established literature methods that rely on hydraulic scaling equations.

This paper is split into two distinct parts: gas-exchange theory/model development (Section 2) and BIKER algorithm development/validation (Section 3). Section 3 is fundamentally dependent on the results presented in Section 2, therefore Section 2 presents both theory and results. Both sections detail the data used. We conclude with a discussion (Section 4) that encompasses both gas exchange theory and remote sensing. Figure 1 conceptually maps out the algorithm's approach to inferring $k_{600}$ from SWOT data, while Figure S1 details the entire study's workflow.

![Figure 1: Conceptual overview of the BIKER algorithm. 1: SWOT will directly measure water surface width and elevation (and thus slope) at a series of cross-sections within mass-conserved river reaches. 2: These hydraulics measurements are used to calculate turbulent dissipation in the river channel (Section 3). 3: Turbulent dissipation is used to infer $k_{600}$ via a process-based model (Section 2).](cache/maps/diagram.tif)

## 2 Predicting $k_{600}$ from large-river hydraulic geometry

### 2.1 Data

We develop our hydraulic model for $k_{600}$ using multiple datasets of field-estimated $k_{600}$ and measured stream hydraulics collected from the literature. In total, we obtain 763 estimates of $k_{600}$, with 701 of these estimates previously gathered by @raymondScalingGasTransfer2012 and @ulsethDistinctAirWater2019. The remaining estimates come from @churchillPredictionStreamReaeration1964 and @owensReaerationStudiesStreams1964. See Table S1 for a complete list of the studies that collected these data. All $k_{600}$ estimates come from tracer studies and thus define $k_{600}$ at the reach scale.

The 763 estimates cover different times of year and hydrological events. They include both individual estimates and repeat estimates in over 500 river reaches across the United States, Wales, Switzerland, and Austria. They span a wide variety of environments from temperate higher-order rivers to small mountain streams and represent a full range of river flows (width ranges from 0.26m to 1,742m, discharge ranges from 8e-4 m3/s to 489 m3/s, and $k_{600}$ ranges from 0.1 m/dy to 4,118 m/dy). While there are still geographic and hydrologic biases in this dataset, it is to our knowledge the largest such dataset of field-estimated, reach-scale $k_{600}$ where hydraulic data is concurrently available.

In addition to hydraulics measured alongside $k_{600}$ and reported above, we expand our dataset of stream hydraulics using a previously published compilation of in situ hydraulic measurements [@brinkerhoffReconcilingAtaStationAtManyStations2019]. That dataset contains over 530,000 unique measurements of river channel velocity, width, and discharge from across the continental United States, originally made to calibrate United States Geological Survey (USGS) streamgauge rating curves and made public by the USGS. This dataset is used to calculate how frequently SWOT observable rivers meet our large-river hydraulic assumptions (Section 2.3).

### 2.2 Process-based hydraulic modeling of river *k*

*k* scales with near-surface turbulence in turbulent streamflows [@hallGasExchangeStreams2020], and extensive field and laboratory experiments have converged on the 'small-eddy model' for *k* as derived by @lamontEddyCellModel1970 and empirically anticipated by @calderbankContinuousPhaseHeat1961. This model scales *k* via the smallest-scale turbulent eddies and has been repeatedly empirically validated in freshwater systems [e.g. @katulStructureFunctionModel2018; @lorkeUnifiedScalingRelation2006a; @zappaEnvironmentalTurbulentMixing2007; @zappaVariationSurfaceTurbulence2003; @moogStreamReaerationNonuniform1999; @vachonRelationshipNearsurfaceTurbulence2010; @tokoroHighGastransferVelocity2008; @wangPhysicallyBasedScaling2021]. The small-eddy model is provided as Equation 3, where $\epsilon$ is the dissipation rate of near-surface turbulence, $v$ is the kinematic viscosity, and *Sc* is the Schmidt number.

$$k\propto Sc^{-1/2}(v\epsilon)^{1/4} \mathbf{(3)}$$

Some laboratory and field observations additionally suggest that open channel flows with small bed roughness do not exhibit homogeneous surface dissipation across the entire reach's air-water interface [@moogAirWaterGasTransfer1999; @talkeTurbulentKineticEnergy2013]. Given this observation, @moogAirWaterGasTransfer1999 proposed an extension to the small-eddy model, additionally scaling $k_{600}$ using a shear Reynold's number $Re_*$ formulation (equation 4, referred to as the 'Reynolds extension' model). The Reynolds extension model is hypothetically useful in low-turbulence scenarios where a relative lack of large-scale eddies effectively 'filter out' the number of small-eddies that actually reach the interface and initiate gas exchange [@talkeTurbulentKineticEnergy2013]. While scaling *k* via a shear Reynold's formulation is sometimes done to parameterize wave-breaking gas exchange models in the open ocean [e.g. @brumerWaveRelatedReynoldsNumber2017; @zhaoDependenceWhitecapCoverage2001; @zhaoEffectWindWaves2003], it is infrequently done in rivers. In the context of BIKER, we chose to test this model because large, SWOT-observable rivers are generally the smoothest, least-turbulent flows along the stream-to-ocean continuum where small eddies might not reach the surface. Further, to our knowledge, the Reynolds extension model has never been empirically tested in predicting river *k*, aside from confirming that large-scale eddies differentially move turbulence to the surface in a large river [@talkeTurbulentKineticEnergy2013].

$$k \propto Sc^{-1/2}(v\epsilon)^{1/4}(Re_*)^{3/8} \mathbf{(4)}$$

Equations 3 and 4 both rely on $\epsilon$, which is difficult to measure. When working at large scales, a commonly used $\epsilon$ model assumes that all turbulence is generated at the bed and transported to the air-water interface via the log-law-of-the-wall (Equation 5). Another approach specific to fluvial settings models *k* via 'form-drag dissipation' (Equation 6) which is equivalently the total stream power per unit mass water. This normalized stream power captures the bulk frictional resistance (and thus energy dissipation) via channel banks, meanders, bars, etc. that is unique to rivers [@moogStreamReaerationNonuniform1999]. Authors have since shown that Equation 6 can reasonably predict *k* in rivers and streams [@raymondScalingGasTransfer2012; @ulsethDistinctAirWater2019; @wangPhysicallyBasedScaling2021].

$$\epsilon = \epsilon_S = \frac{U_*^3}{H} \mathbf{(5)}$$ $$\epsilon = \epsilon_D = gS\bar{U} \mathbf{(6)}$$

### 2.3 Deriving a large-river $k_{600}$ model

Given the theoretical context provided in Section 2.2, we now turn to SWOT-observable rivers specifically. Rivers and streams change predictably along their longitudinal profile from source to sea, and we can exploit the hydraulic geometry of large rivers at the end of this continuum to estimate *k* in SWOT rivers. In general, as river size increases, channels become more rectangular, their shapes elongate (becoming wider quicker than they become deeper) and their hydraulic radii begin to approximate their mean flow depth, i.e. $R_h \approx H$ [@leopoldHydraulicGeometryStream1953], a common assumption in hydraulic and geomorphic modeling of large rivers. For example, SWOT-observable river flows have an average $\frac{R_h}{H}$ ratio of `r HG_swot$mean` and standard deviation of only `r HG_swot$sd` (n = `r HG_swot$n`; see Text S1 for how we built this dataset). We refer to these rivers as 'hydraulically-wide'.

We therefore assume that all SWOT-observable rivers are hydraulically-wide to derive a model for gas exchange. The overall goal is to reduce the equations down to their fundamental parameters, identifying which terms are SWOT observable and limiting the number of terms not directly measurable via SWOT. We first impose $\epsilon=\epsilon_D$ and $R_H=H$ on the Reynolds extension model (Equation 4) and then log-transform to maintain homogeneous variance across the orders of magnitude of $k_{600}$ to arrive at Equation 7 (with statistical coefficient $\beta_1$). Equation 7 thus defines gas exchange velocity solely as a function of slope, mean flow depth, and mean flow velocity and is theoretically valid only in a hydraulically-wide channel.

$$log(k_{600})=\beta_1 + \frac{7}{16}log(gS) + \frac{1}{4}log(\bar{U}) + \frac{9}{16}log(H) \mathbf{(7)}$$

We also test the performance of three other models for predicting *k* in hydraulically-wide channels via the other three unique combinations of Equations 3-4 and Equations 5-6. While the complete model derivations and results for all four models are provided in Text S2 and Figure S2, the final and best-performing model (Equation 7) is presented and used below.

### 2.4 Model validation

With Equation 7 derived, we now test its strength of fit in hydraulically-wide river flows. We validate on the dataset of in situ estimates of $k_{600}$, after filtering for those made in hydraulically-wide channels, which was defined as flows whose hydraulic radius was within 1% of their mean flow depth. All told, there are `r n` estimates of hydraulically-wide $k_{600}$. Equation 7 is assessed via a Bayesian linear regression model, where the theoretically-derived coefficients from equation 7 are used as reasonably informative priors for the regression coefficients. The intercept and model uncertainty both use uninformed priors. See Text S3 for the model specification. We assess the goodness-of-fit using the posterior mean coefficient of determination ($r^2$) following [@gelmanRsquaredBayesianRegression2019].

![Figure 2: Empirical testing of our large-river $k_{600}$ model on 166 measurements made in hydraulically-wide rivers. Dashed grey line denotes the 1:1 line. For the model, points reflect posterior means while lines reflect the 95% credible intervals.](cache/k600_theory/fig2.jpg)

Equation 7 explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2, 2) * 100`% of variation in observed (log-transformed) $k_{600}$ in hydraulically-wide rivers and accurately captures the scaling dynamics of observed $k_{600}$ as well (Figure 2). The posterior mean in natural log space for $\beta_1$ is `r round(log(10^(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$coef)),2)` (Table 1). While the individual residuals can be quite large (indicating that additional processes and/or field estimation error are presumably controlling the 50% unexplained variation in log-transformed $k_{600}$), the general scaling of $k_{600}$ with river hydraulics is strongly captured. We also tested two non-Bayes implementations of equation 7 using 1) the theoretically-derived coefficients and 2) a fully empirical line fitting approach (Table 1). Interestingly, the data exerted little-to-no influence on the theoretically-derived priors, and both the empirical and posterior coefficients were virtually identical to those obtained theoretically (Table 1), indicating that equation 7 is a good theoretical model for gas exchange in hydraulically wide flows.

*Table 1: Comparison of three regression models for $k_{600}$ model for hydraulically-wide rivers. ‘Theoretical’ refers to the model coefficients derived by imposing a hydraulically wide channel on the ‘Reynolds-extension’ model for gas exchange (section 2.3). ‘Empirical’ refers to the result of fitting the coefficients by least-squares regression. ‘Bayesian’ refers to the posterior means of the coefficients, conditional on priors informed by the ‘Theoretical’ coefficients (Text S3, Figure 2).*

| **Equation 7 results**          | **'Theoretical'**                                               | **'Empirical'**                                                 | **'Bayesian'**                                                  |
|---------------------------------|-----------------------------------------------------------------|-----------------------------------------------------------------|-----------------------------------------------------------------|
| $\alpha_1$                      |$\frac{7}{16} \approx 0.44$                                      |`r round(modelsSTATS[modelsSTATS$name == 'empirical',]$slope1,2)`|`r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$slope1,2)`|
| $\alpha_2$                      |$\frac{1}{4} \approx 0.25$                                       |`r round(modelsSTATS[modelsSTATS$name == 'empirical',]$slope2,2)`|`r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$slope2,2)`|
| $\alpha_3$                      |$\frac{9}{16} \approx 0.56$                                      |`r round(modelsSTATS[modelsSTATS$name == 'empirical',]$slope3,2)`|`r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$slope3,2)`|
| $\beta_1$ (natural log space)   |`r round(modelsSTATS[modelsSTATS$name == 'theoretical',]$coef,2)`|`r round(modelsSTATS[modelsSTATS$name == 'empirical',]$coef,2)`  |`r round(log(10^(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$coef)),2)`  |
| $r^2$                           |`r round(modelsSTATS[modelsSTATS$name == 'theoretical',]$r2,2)`  |`r round(modelsSTATS[modelsSTATS$name == 'empirical',]$r2,2)`    |`r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2,2)`    |

Compared to other combinations of Equations 3-6 (Figure S2), there is also a better fit using equation 7 thanks to the addition of the Reynold's number scaling for low-turbulent flows (as expected). The success of the final model in hydraulically-wide channels (provided as equation 8 using the posterior means) provides us with a physical-model for gas evasion built with SWOT in mind. The river hydraulics terms in Equation 8 ($\bar{U}$, $H$, and $S$) can either be directly measured or reasonably inferred from SWOT measurements, effectively opening the door for remotely sensing the gas exchange velocity.

$$log(k_{600})=3.89 + (0.43)log(gS) + (0.31)log(\bar{U}) + (0.52)log(H) \mathbf{(8)}$$

## 3 BIKER algorithm development and validation

### 3.1 Data

To validate BIKER, we cannot use actual SWOT measurements as SWOT has yet to launch. In the hydrology literature, it has become standard practice to benchmark SWOT-related algorithms on "SWOT-like" data [@durandIntercomparisonRemoteSensing2016; @frassonExploringFactorsControlling2021]. We use `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers for validation, where these simulated rivers are simply reach-averaged hydraulic model outputs where the water surface heights, slopes, and widths are labelled as RS observations and are used as the sole inputs to BIKER (Figure S1). These datasets are created using standard hydraulic models forced with known inflows and measured bathymetry to model the hydraulic response of the rivers, and then those terms visible to SWOT are extracted to produce hydraulically realistic synthetic observations. These data were published by @frassonExploringFactorsControlling2021 and @durandIntercomparisonRemoteSensing2016. See Figure S3 for a map of these river's locations along the global SWOT river network [@altenauSurfaceWaterOcean2021].

There is considerable geographic bias in our validation rivers, with rivers only present in North America, Western Europe, and Bangladesh. Further, no Arctic rivers are included. We acknowledge that this bias limits our ability to validate BIKER across many environments ahead of SWOT’s launch. However, it is a sufficient validation set for a first proof-of-concept study consistent with the hydrology literature for SWOT. Further, the data requirements to create these test cases are strict and the processing time is enormous.  Also note that SWOT water surface slope measurements will have a lower detection limit of 1.7 cm/km [@biancamariaSWOTMissionIts2016], and therefore any slope measurement in our data less than this threshold was reassigned this minimum value.

We validate BIKER under two different 'error scenarios' (section 3.2.2). While SWOT will provide river surface measurements of novel quality and resolution, as with all remote sensing products there are expected errors that will be implicit in these measurements. Here, we validate BIKER under a 'no-measurement-error' scenario that reflects an unrealistic measurement as if SWOT has perfect accuracy and precision: we use the hydraulic model of `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers directly as a first test of algorithm validity. `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` of these rivers are then additionally validated under a 'measurement-error' scenario that more closely mimics expected SWOT by adding realistic radar errors and sampling along the satellite's future ground track. SWOT river error modeling was developed by @durandHowWillRadar2020 and then applied to these `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` rivers by @frassonExploringFactorsControlling2021. This error modelling is non-trivial and computationally expensive, and so @frassonExploringFactorsControlling2021 were limited to only `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` test cases with SWOT measurement errors. Likewise, we stick to these `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` rivers for the same reason. These rivers are detailed in Figure S3. Given that it is only `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 1,])` synthetic rivers, caution should be used in over generalizing our results beyond our proof of concept.

For $[CO_2]$ evasion and carbon emissions calculations, we use 26 bi-weekly dissolved $[CO_2]$ samples collected by @beaulieuControlsGasTransfer2012 at one location in the Ohio River (Figure S3) for one calender year from 2008-2009 (Figure S4). Note that this $[CO_2]$ data is for the Ohio River only but was applied to all rivers to provide a physically realistic signal for $[CO_2]$ fluxes with meaningful seasonality and dynamics. Therefore, the raw carbon emissions estimates presented in this paper are meaningless in the context of actually measured carbon emissions from these rivers but are better than specifying $[CO_2]$ concentrations devoid of context. These data are necessary as we are interested in the effect of BIKER $k_{600}$ errors on eventual fluxes and comparing these fluxes with published methods. Therefore, applying these 'unit' $[CO_2]$ values allows for such a comparison by providing a realistic timeseries.

### 3.2 Section 3 methods

#### 3.2.1 BIKER

To develop BIKER, we follow the work of @hagemannBAMBayesianAMHGManning2017, @brinkerhoffConstrainingRemoteRiver2020, and @durandEstimatingReachaveragedDischarge2014 to infer $k_{600}$, *H*, and $\bar{U}$ from SWOT observations. These Bayesian approaches to inverse modeling via SWOT data start from Bayes rule (Equation 9), where $\Theta$ is some set of non-remotely-sensible parameters we want to solve for (including $k_{600}$), *x* is the observed data, $f(x|\Theta)$ is the sampling model where data are conditional on the parameters, and $p(\Theta)$ is the joint prior distribution of the parameters. Therefore, we are interested in solving for $p(\Theta|x)$, or the 'posterior' distribution. Note that $p(x)$ is usually computationally intractable to integrate exactly, but Bayesian inference requires only the proportionality to be specified: $p(\Theta|x) \propto f(x|\Theta)p(\Theta)$. Sampling algorithms are then used to approximate the actual posterior distribution, as is done in BIKER.

$$p(\Theta|x) = \frac{f(x|\Theta)p(\Theta)}{p(x)} \mathbf{(9)}$$

The heart of BIKER is its reformulation of the $k_{600}$ model (Equation 8) as a Bayesian sampling model that is conditional on the non-remotely-sensed parameters (i.e. $f(x|\Theta)$). This approach is similar to the 'McFLI' (Mass-Conserved Flow Law Inversion) logic used in some SWOT remote sensing of discharge algorithms [@gleasonTrackingRiverFlows2017]. To start, we write $k_{600}$ as a function of SWOT-observables $W$ and $S$. This algebra is carried out using Equation 8 and Manning's equation for mean flow velocity ($\bar{U}=\frac{1}{n}R_h^{2/3}S^{1/2}$). Following Section 2.3, we continue to assume that the channel is hydraulically-wide ($R_h = H =\frac{A}{W}$). To leverage additional SWOT data, we use the "Durand transform" originally published by [@durandEstimatingReachaveragedDischarge2014]: the wetted channel area *A* is further split into the SWOT-unobservable portion $A_0$ and SWOT-observable portion $dA$ where $dA_{it} = \sum_{t':W_{t'}\le W_t}{W_{t'} \delta H_e_{t'}}$ for cross-section *i* and timestep *t* within a mass-conserved river reach.

All of the above algebra simplifies to Equation 10. Conveniently, $k_{600}$ as estimated by tracer additions to a stream is inherently a reach-scale quantity (in a mass-conserved reach). Therefore, Equations 7, 8, and 10 all yield a reach-scale $k_{600}$ (i.e. $k_{600_i}=k_{600}\forall i$), thus lowering the number of parameters BIKER must infer and making the problem much better constrained.

$$log(k_{600_t})=3.89+(0.4320)log(g)+(0.5862)log(S_{i,t})+(0.3084)log(\frac{1}{n_i})+(0.7282)log(\frac{A_{0_i}+dA_{i,t}}{W_{i,t}}) \mathbf{(10)}$$

Next, Equation 10 is re-written as a Bayesian sampling model $f(x|\Theta)$, in which the joint data distribution *x* (i.e. SWOT observations) is sampled from the joint parameter distribution ($\Theta$). We first rearrange equation 10 to isolate *x* from $\Theta$ (equation 11). Then, equation 12a-c re-expresses equation 11 as a normal distribution with standard deviation $\sigma_{k_600}$. The uncertainty expressed by $\sigma_{k_{600}}$ arises from uncertainties in 1) parameters in Equation 8, 2) Manning's equation, and 3) the hydraulically wide channel assumption.

$$3.89+(0.4320)log(g)+(0.5862)log(S_{i,t})-(0.7282)log(W_{i,t})=log(K_{600_t})+(0.3084)log(n_i)-(0.7282)log(A_{0_i}+dA_{i,t})\mathbf{(11)}$$

$$x=[3.89+(0.4320)log(g)+(0.5862)log(S_{i,t})-(0.7282)log(W_{i,t})]\mathbf{12a}$$
$$\Theta = [log(K_{600_t})+(0.3084)log(n_i)-(0.7282)log(A_{0_i}+dA_{i,t})]\mathbf{(12b)}$$
$$x ~ N(\Theta, \sigma_{k_600}^2)\mathbf{(12c)}$$

Equations 9 and 12 necessitate that we specify prior distributions for the parameters $A_0_i$, $k_{600_t}$, and $n_i$. Prior distributions formalize the a priori estimates and uncertainties for the non-remotely-sensed terms. More intuitively, BIKER priors represent our 'prior river knowledge' of what $A_0_i$, $k_{600_t}$, and $n_i$ probably are for some river since they cannot be directly remotely sensed. This method of hyperparameter assignment is analogous to the 'empirical Bayes approach' to Bayesian inference [@hoffFirstCourseBayesian2009]. Our goal in prior specification was to rely on absolutely no in situ information such that we could run this method on any river on Earth solely using SWOT observations. In theory, more informed priors via various a priori information about a specific river will improve BIKER performance, but here we chose to test the fully generalized algorithm. Therefore, the validation presented here is a 'worst-case scenario', wherein BIKER performance should improve with better prior information on the river. In that context, we used a variation of the prior specification method developed by @brinkerhoffConstrainingRemoteRiver2020, who developed 'geomorphic river types' with distinct prior sets for $A_0_i$ and $n_i$. These priors are assigned to a river solely using SWOT data *W* and *S*, therefore meeting our needs for complete global implementability. Prior assignment for $k_{600_t}$ was developed similarly (all prior specifications are elaborated on in Text S4).

With the sampling model described ($f(x|\Theta)$ = Equation 12a-c) and priors $p(\Theta)$ specified (Text S4), a joint posterior distribution conditional on the SWOT observations ($p(\Theta|x) \propto f(x|\Theta)p(\Theta)$) is therefore also specified. To approximate this distribution, we use a Markov Chain Monte Carlo (MCMC) algorithm implemented using the Stan probabilistic programming language. Specifically, Stan uses a Hamiltonian Monte Carlo sampler which reduces computation time relative to other sampling algorithms [@hagemannBAMBayesianAMHGManning2017].

#### 3.2.2 BIKER validation

BIKER provides a timeseries of $k_{600}$: for each SWOT observation, it yields a unique $k_{600}$. There are, to our knowledge, no datasets of $k_{600}$ over time approaching the temporal density of our simulated SWOT rivers. We therefore apply Equation 8 as validated in Figure 2 to specify $k_{600}$ given the true hydraulics of each case and compare BIKER's inversion to that value: given observed hydraulics, 'observed' $k_{600}$ comes from Equation 8. Remember that SWOT cannot observe below the water surface and therefore cannot measure $\bar{U}$ or *H* (hence the need for Equation 10), and that all SWOT observations contain errors in both space and time (hence Equation 12). We acknowledge that there is error in Equation 8 as shown in Figure 2, but this error can be explicitly parameterized in our Bayesian system (elaborated on in Text S5). Therefore, the BIKER validation presented here is an exercise to see how well the imperfect and partial SWOT observations can infer $k_{600}$ given the hydraulic assumptions in Equation 10 and uncertainty in the data itself. Note also that we have already validated Equation 7 in Figure 2 and Table 1.

We validate BIKER as a timeseries of $k_{600}$ for each river using the BIKER posterior means. Our error metrics consider the timeseries nature of the problem and are formally defined in Table S2. They consist of the correlation coefficient *r* to quantify accuracy of BIKER's temporal dynamics, the root mean square error normalized by the observed mean (NRMSE) and normalized mean absolute residual error (NMAE) to assess bias, and the Kling-Gupta Efficiency (KGE). KGE is frequently used to assess streamflow prediction and simultaneously assesses accuracy in both bias and dynamics. While a value greater than -0.41 means the model outperforms a uniform prediction of the mean [@knobenTechnicalNoteInherent2019], generally KGE scores are interpreted as being meaningful in ungauged settings if \> 0.

#### 3.2.3 Carbon emissions validation

It is one thing to accurately model the temporal dynamics of $k_{600}$ as above, but researchers are often most interested in the actual carbon emitted from river to atmosphere. Per Equation 2, emissions upscaling is done using river hydraulic models to estimate $k_{CO_2}$ and in turn $[FCO_{2_{upscaled}}]$. However, streamflow data and/or model outputs are more readily modeled at the global scale than river channel geometry, and so upscaling models usually predict $\bar{U}$ and *H* as functions of streamflow (*Q*) using hydraulic geometry scaling relationships. This workflow effectively reduces Equation 2 to Equation 13. It is worth stressing that these literature upscaling workflows rely on in situ streamflow records and/or high-quality streamflow models.

$$[FCO_{2_{upscaled}}]=f(k_{CO_2}, \Delta [CO_2])=f(Q, \Delta [CO_2]) \mathbf{(13)}$$

Conversely, BIKER represents a new way of approaching Equation 13 compared to existing literature models: BIKER has no reliance on a streamflow model nor hydraulic geometry scaling relationships and only requires that a river is SWOT-observable. We are therefore interested in how the final carbon emissions that result from BIKER compare against literature methods that use Equation 13. We have the data to test four different models for fluxes: 'BIKER', 'Raymond 2013', 'Raymond 2012', and 'Brinkerhoff 2019'. These latter three approaches all use the same philosophy for $k_{600}$: making hydraulic and geomorphic assumptions to associate $k_{600}$ with observed hydraulics before using the $[CO_2]$ data as a realistic timeseries to yield fluxes per Equation 13. In all three approaches, these observed hydraulics are streamflow, while BIKER uses only SWOT observations. Therefore, the advantage of BIKER is in its ease of application, as SWOT will observe all global rivers wider than 50m while streamflow observations are extremely geographically limited. But, BIKER is only attractive if it can produce fluxes with similar errors to published methods. Text S6 and Table S3 fully describe these three literature models.

To benchmark BIKER against these literature methods, we pair the 26 biweekly $[CO_2]$ and water temperature samples from @beaulieuControlsGasTransfer2012 (Section 3.1, Figure S4) with every 14th set of daily SWOT observations (as the $[CO_2]$ data is bi-weekly). We then calculate $[FCO_2]$ using Equation 1, an atmospheric $[CO_{2_{air}}]$ of 400 $\mu$atm, and a *Sc* estimated following Raymond et al. (2012). The *k* in Equation 1 is obtained using BIKER or the three literature models (Table S3 and Text S6). Finally, we estimate a pseudo yearly total carbon emission rate (via $[CO_2]$ evasion) by applying each river's mean $[FCO_2]$ over the river's surface area and summing all rates across rivers, remembering that we are applying 'unit' $[CO_2]$ data to all rivers.

### 3.3 Section 3 results

#### 3.3.1 BIKER

In general, the temporal dynamics of $k_{600}$ are reproduced quite well by BIKER (Figure 3). In the best performing rivers, both bias and temporal dynamics are strongly captured (Figure 3e-f). Notably, there is sometimes positive bias in the estimates (e.g. the Connecticut and Iowa rivers, Figure 3c-d). Some rivers yield the correct temporal dynamics, but the magnitude of these values is stretched relative to the observed (e.g. Ohio River and Seine River, Figure 3a-b). In these two examples, temporal trends are still reasonably inferred even though the magnitude of the estimates is quite wrong. The size of posterior uncertainty in $k_{600}$ does not appear to be associated with overall algorithm performance, with both certain and uncertain results spread across the rivers, regardless of their KGE (Figure 3, Figure S5).

![Figure 3. Representative (6 out of 47) river timeseries plots of $k_{600}$. Orange is observed, while green is BIKER posterior means and uses SWOT measurements as its sole input. The green ribbon indicates the 95% credible intervals for the BIKER posteriors. Rivers are sampled from the three tertiles of KGE scores. (a-b) are poorest performing rivers, (c-d) are in the middle, and (e-f) are the best performing rivers. Y axis is normalized by maximum observed values to compare visually. Consult Figure S5 for all 47 timeseries plots (assuming no measurement error) and Figure S6 for the 16 rivers with measurement errors.](cache/validation/fig3.jpg)

Overall, river performance across error metrics (Table S2) is reasonable given the strict validation setup we have employed (Figure 4). Median river $KGE$ is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$KGE, na.rm=T),2)` and median river *r* is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$r, na.rm=T),2)` (Figure 4). Further, `r sum(results_k600_rivs[results_k600_rivs$errFlag==0,]$KGE > -0.41)`/`r nrow(results_k600_rivs[results_k600_rivs$errFlag==0,])` rivers outperform a uniform prediction of the mean (KGE = -0.41). The correlation coefficient *r* out-performs the other metrics which assess bias (Figure 4). This result indicates strong inference of each river's temporal $k_{600}$ dynamics given that absolutely no in situ information is being used to predict $k_{600}$. NRMSE has a median score of `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$NRMSE, na.rm=T),2)` (Figure 4), highlighting many rivers which have notable positive biases (Figure 3 also confirms this result visually). Median NMAE is `r round(median(results_k600_rivs[results_k600_rivs$errFlag == 0,]$NMAE, na.rm=T),2)*100`% (Figure 4). Taken in aggregate, Figures 3-4 indicate that BIKER is quite good at capturing temporal dynamics in $k_600$, however there is often positive bias in its estimates.

![Figure 4. Performance metrics across all 47 rivers, ploted as empirical cumulative density functions (eCDFs). Each subpanel is labelled by its performance metric (defined in Table S2). Dashed lines denote median scores.](cache/validation/fig4.jpg)

BIKER is robust to the measurement errors that will be implicit in SWOT's observations of river width and slope (Figure 5). BIKER performs nearly identically across all four error metrics, particularity with respect to the metrics that singularly assess bias or correlation errors (Figure 5b-d). The three very poor *KGE* rivers actually improve under the error scenario (Figure 5a), though this is likely an artifact of BIKER already not working in these rivers. Given these results, we deem that SWOT measurement error does not exert a significant influence on BIKER performance and so the results presented for the rest of the manuscript assume no measurement error in order to use all `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers.

![Figure 5: Comparison of BIKER performance when considering measurement error in the SWOT observations for 16 rivers. Each subpanel is labelled by its performance metric (Table S2). Rivers that fall within the purple zone get worse when accounting for measurement error, while rivers in the green get better. Note that some axes are flipped in order to visualize the 'best performances' in the upper-right-corner of each sub-plot. Black line denotes linear regression to aid in visualization.](cache/validation/fig5.jpg)

Finally, errors/biases associated with the prior $k_{600}$ are correctly propagated through the posterior in an approximately 1:1 manner (which is expected), except for a subset of rivers in which posterior error actually increases relative to the prior (Figure 6a). We explore why this phenomenon happens below.

![Figure 6: Empirical cumulative density functions for prior and posterior $k_{600}$ NMAE. See Table S2 for metric definitions. a) all rivers. b) Only rivers with a coefficient of variation (CV) of observed $k_{600}$ > 0.10. c) Only rivers with CV > 0.20. d) Only rivers with CV > 0.30. Overall we see that rivers with greater temporal variation in $k_{600}$ behave better for BIKER, i.e. Bayesian inference reduces bias from prior to posterior.](cache/validation/fig6.jpg)

Recall that BIKER relies on a timeseries of SWOT data, and that these timeseries may not be representative of the full spectrum of $k_{600}$ values that are actually experienced in the river, therefore potentially biasing both the prior estimation methods (Text S4) and the actual Bayesian inference. Put another way, we suggest that if a SWOT timeseries does not sufficiently capture a river's temporal dynamics, it will introduce additional error to the inference results. To test this hypothesis, we subset our validation dataset by progressively higher coefficients of variation (*CV*) (> 10%, >20%, and >30%). Posterior bias drops once the temporal variability of the SWOT data is sufficiently high (Figure 6b-d), with BIKER posterior error the smallest, and much less than prior error, in Figure 6d. This result is elaborated on in Section 4.2.

#### 3.3.2 Carbon Emissions

Finally, we carry these $k_{600}$ estimates all the way to annual carbon emissions rates and compare BIKER against three established in situ techniques in the literature. It is important to remember that only BIKER is completely globally implementable, while the other three models necessarily rely on having a streamflow record or high-quality routed streamflow readily available. Therefore, the in situ methods reflect their 'best-case scenarios' while BIKER reflects its worst case scenario, where priors are set entirely from SWOT observations and are generally the least informative they will ever be. This validation setup means that BIKER's annual carbon emission estimate can only improve from what is presented here. We also stress again that the raw emissions rates here are meaningless as they are calculated from an identical $[CO_{2_{water}}]$ timeseries applied to every river. We are principally interested in relative differences between techniques employed.

Figure 7 compares the annual carbon emissions rate (via $[F_{CO_2}]$) from the rivers using BIKER posterior means and the three stream gauge-based HG models. BIKER outperforms the gauge-based approaches, nearly correctly inferring the annual carbon emissions rate (`r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_BIKER',]$yearlyCFux, 2)` Tg-C/yr for BIKER versus `r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_obs',]$yearlyCFux, 2)` Tg-C/yr observed). The three HG models overestimate this emissions rate: `r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_raymond2013',]$yearlyCFux, 2)` Tg-C/yr, `r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_raymond2012',]$yearlyCFux, 2)` Tg-C/yr, and `r round(results_fco2_bulk[results_fco2_bulk$key == 'FCO2_brinkerhoff2019',]$yearlyCFux, 2)` Tg-C/yr for 'Raymond 2013', 'Raymond 2012', and 'Brinkerhoff 2019' respectively. 'Raymond 2012' falls within the BIKER credible intervals and is reasonably close to the observed value, while 'Raymond 2013' overestimates the emission rate. 'Brinkerhoff 2019''s lower confidence interval is comparable to BIKER's higher credible interval. BIKER's relatively stronger performance than the in situ models is elaborated on in Section 4.3. Finally, BIKER's uncertainty is on par with the in situ technique (‘Brinkerhoff 2019’), despite being obtained solely from SWOT data. Taken in aggregate, BIKER provides a strong upscaling estimate of the annual carbon emission rate for the rivers and is either similar or better than established in situ techniques (Figure 7).

![Figure 7: Yearly carbon emissions rate via $[CO_2]$ evasion across all rivers. Completely remotely-sensed methods are colored in red (with 95% credible intervals), in situ methods in purple (with 95% confidence intervals when available), and the observed in orange. Consult section 3.2.3 for details on the 'observed' flux.](cache/FCO2/fig7.jpg)

## 4 Discussion

In this paper, we propose that the soon-to-launch SWOT satellite will provide enough hydraulic measurements to analyze the temporal dynamics of $k_{600}$, and therefore allow for a global-scale analysis of spatiotemporal trends in large-river $k_{600}$ once SWOT launches. In preparation for SWOT’s launch, we developed 1) a wide-river-specific hydraulic model for $k_{600}$ that explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2,2)*100`% of variation in $k_{600}$ and 2) the BIKER algorithm to infer $k_{600}$ using no on-the-ground information. Validating on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers, we show strong recovery of rivers’ temporal $k_{600}$ dynamics and a hypothetical total annual carbon emission rate across all `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` rivers (section 3.3).

### 4.1 Gas exchange in hydraulically-wide rivers

Field studies of gas exchange in wide rivers have suggested that $k_{600}$ behaves differently in these rivers than in steeper and smaller rivers [@ulsethDistinctAirWater2019; @beaulieuControlsGasTransfer2012; @alinPhysicalControlsCarbon2011; @wangPhysicallyBasedScaling2021; @raymondGasExchangeRivers2001]. While much work has focused on the small-stream side of the stream-to-river continuum, comparatively less work has been done in large rivers. Here, we focus on the larger, 'smooth-channel' end of the continuum, using a model for gas exchange that scales $k_{600}$ by both $\epsilon_D$ and a shear Reynold's number. Scaling $k_{600}$ via a shear Reynold's number is often done to parameterize breaking-wave gas exchange models in the open ocean [@brumerWaveRelatedReynoldsNumber2017; @zhaoDependenceWhitecapCoverage2001; @zhaoEffectWindWaves2003], though these models are specific to high wind speeds in open ocean. To our knowledge, @moogAirWaterGasTransfer1999's specific setup, which imposes a space-and-time varying, fractional area surface turbulence theory on the small-eddy model, has not been empirically validated in rivers until now (Figure 2 for hydraulically wide channels). Using our full dataset of $k_{600}$, we also observe that this model breaks down when including non-hydraulically-wide rivers (as the theory would suggest). Future tests should also explore other shear Reynold's scaling relations for gas exchange in rivers.

Crucially, we are not accounting for wind-driven gas exchange, which is suggested to play an important role in wide rivers because river surface area is sufficiently large that sheltering no longer limits the influence of wind-derived turbulence [@beaulieuControlsGasTransfer2012; @wangPhysicallyBasedScaling2021; @raymondGasExchangeRivers2001]. None of the existing hydraulics-driven fluvial $k_{600}$ models account for wind-driven gas exchange either. Additionally, under higher-wind scenarios the turbulent regime will switch from hydraulically-driven to wind-driven turbulence [@zappaEnvironmentalTurbulentMixing2007] and the assumptions under-pinning BIKER will likely break down. BIKER's outputs can therefore be interpreted as the '$k_{600}$ under low-wind conditions' when surface turbulence is dominated by hydraulics rather than wind. That said, BIKER's flexible implementation is a good start towards eventually coupling hydraulics-driven gas exchange with wind-driven gas exchange under moderate-to-high wind scenarios.

### 4.2 Towards remote sensing of global spatiotemporal dynamics of $k_{600}$ in large rivers

To date, the studies exploring the spatiotemporal dynamics of riverine gas exchange have arguably been held back by a lack of data. A few studies have investigated these dynamics, but they have been limited to individual rivers and/or limited field seasons [@hallAirWaterOxygen2012; @sand-jensenCO2DynamicsDanish2012]. For example, @wallinSpatiotemporalVariabilityGas2011 performed a preliminary analysis in northern Sweden relating cross-section specific temporal variability in gas exchange with channel slope, but they were limited to a mean of only 8 measurements per river in a single watershed. While these studies are a good start, they are insufficient for further developing process-level understandings of gas exchange at the global-scale.

Therefore, inferring $k_{600}$ from SWOT data is an attractive option to address this problem of limited data. For reference, 95% of the SWOT-visible rivers globally (202,811) will have sufficient SWOT observations along the river to run BIKER at least once every 21 days, with most of the temperate and Arctic rivers having 3+ observations per 21-day cycle [@altenauSurfaceWaterOcean2021]. While BIKER will not directly measure $k_{600}$, it does robustly infer temporal trends in $k_{600}$ and reasonably infers the absolute magnitude of $k_{600}$ (Figures 4-6). This many data will provide a novel dataset of $k_{600}$ on a scale never before possible.

However, SWOT’s relatively coarse spatial resolution limits BIKER’s use to large rivers. This limitation means SWOT cannot see the vast majority of the global river network (which are too narrow for SWOT), though it is likely to observe much of its air/water interface at which gas exchange occurs (rivers wide enough for SWOT to observe). To confirm this hypothesis, we obtained the global estimates for SWOT-observable surface area and length [at mean annual streamflow- @altenauSurfaceWaterOcean2021] and compared them to the most recent estimates of global river surface area and length [@liuImportanceHydrologyRouting2022- Table S4]. We found that `r round(global_swot_coverage[global_swot_coverage$model == 'swot',]$sa_km2 / global_swot_coverage[global_swot_coverage$model == 'total',]$sa_km2 * 100,0)`% of the global riverine surface area is SWOT-observable, while only `r round(global_swot_coverage[global_swot_coverage$model == 'swot',]$length_km / global_swot_coverage[global_swot_coverage$model == 'total',]$length_km * 100,2)`% of the network length is SWOT-observable. While small streams in aggregate exert a significant influence on GHG emissions from river networks [@liuImportanceHydrologyRouting2022]; @raymondGlobalCarbonDioxide2013], BIKER will still be capable of inferring k600 for much of the global freshwater air/water interface.

With that said, there is a substantial range of BIKER performance across rivers (Figures 3, 4, S4, S5). These differences in performance are likely due to the representativeness of the priors, which makes sense as Section 2.3 and section 3.2.1 have effectively reduced $k_{600}$ to a function of hydraulics that are nearly all directly measurable by SWOT. Any resulting bias in BIKER's predictions is likely attributable to either bias in the priors used for the non-remotely sensed terms (Equation 12b) or in the $k_{600}$ model itself (Equation 8, including the aforementioned wind errors). For SWOT discharge algorithms, authors have repeatedly shown that the 'quality' of prior river knowledge plays a large role in the success of discharge inversions [@frassonExploringFactorsControlling2021; @andreadisConstrainingAssimilationSWOT2020; @brinkerhoffConstrainingRemoteRiver2020; @tuozzoloEstimatingRiverDischarge2019] and our results here further corroborate this finding.

Furthermore, there is likely a minimum sufficient variability in SWOT observations that is necessary to strongly infer a $k_{600}$ timeseries (Figure 6). 'Hydraulic visibility', i.e. the ability of a remote sensor to identify a hydrological response in the river [@garamboisHydraulicVisibilityUsing2017] is applicable here. If we apply hydraulic visibility to a sensor's ability to identify temporal variations in $k_{600}$, our results suggest a 'minimally sufficient' hydraulic variability in SWOT measurements is needed to improve upon the prior (in Figure 6b, suggested to be approximately >20% *CV* for this sample of rivers). This finding will be important once SWOT launches and BIKER is implemented at the global-scale. Although it is presently impossible to know whether SWOT will achieve ‘sufficient hydraulic visibility’ over its lifetime, recent similar work using the Landsat archive suggests that most rivers’ full flow regime will be observed by the SWOT satellite: @allenTimingLandsatOverpasses2020 found that the Landsat record observed 97% of streamflow percentiles in 90% of United States streamgauges. Landsat has an average temporal resolution of 16 days, which is approximately similar to the repeat cycle for SWOT. Further, SWOT will penetrate clouds and provide even more data on cloudy days (unlike Landsat's optical sensor). With that said, the nominal lifespan of SWOT is only three years, within which certain streamflow magnitudes may not be experienced and reduce the chance that ‘sufficient hydraulic visibility’ is achieved.

Finally, because of its reliance on Manning’s equation and hydraulic geometry (section 3.2.1.), BIKER cannot invert overbank flow events, similar to many SWOT discharge algorithms. This concept is an important distinction that must be accounted for when BIKER is run on actual SWOT data, though future work should also look to couple floodplain flow laws with BIKER to capture gas exchange in seasonally-inundated floodplains.

### 4.3 Coupling BIKER with upscaling workflows

BIKER will likely be useful for informing annual upscaled carbon emissions estimates from river networks when coupled with $[CO_2]$ data (Figure 7). This encouraging result has two main implications for future work.

First, it implies that BIKER will be useful when coupled with large-scale $[CO_{2_{water}}]$ models, provided these models are accurate. The models would give time and space varying gas exchange. @liuImportanceHydrologyRouting2022 and @saccardiImprovingPredictionsStream2021 each propose models that robustly predict reach-scale dissolved $[CO_2]$ concentrations using two different approaches- machine learning for @liuImportanceHydrologyRouting2022 and process-based reactive transport modeling for @saccardiImprovingPredictionsStream2021- but both models yield $[CO_2]$ estimates that would be spatially and temporally consistent with BIKER's output. Our promising results suggest that BIKER could provide additional (and directly inferred) measurements of $k_{CO_2}$ to these models, thereby better informing model results through direct observations. This modeling would likely be accomplished via data assimilation which has proven useful in using remotely-sensed discharge to improve streamflow routing models [@fengRecentChangesArctic2021; @ishitsukaCombiningOpticalRemote2020], and of which the @saccardiImprovingPredictionsStream2021 $[CO_2]$ model takes a similar form.

Second, our experiments allow us to directly compare the influence of geomorphic assumptions on total carbon emission rates from river networks (Figure 7), as all other calculations and parameters were held constant across our four tested models (Text S6). Therefore, we highlight a potentially large source of uncertainty in current river $[CO_2]$ upscaling estimates: the geomorphic models employed to scale river channel hydraulics with streamflow (Figure 7). In this case, the only difference between the three literature models and the observed estimate is the specific HG model employed to predict river depth and velocity (Text S6, Table S3), and yet the eventual carbon emissions estimates are quite different (Figure 7). Further, recall that the BIKER results in Figure 7 reflect a worst-case scenario (relatively uniformed priors), while the three in-situ methods represent best case scenarios (perfect streamflow records). We suggest future work should perform a formal sensitivity analysis for these HG parameters.

## 5 Conclusions

This proof-of-concept study verifies that BIKER can provide meaningful information on the spatiotemporal dynamics of gas exchange solely from SWOT data and functionally opens the door for a global-scale analysis of riverine gas exchange upon SWOT’s launch (and data collection). Although BIKER results are often biased in magnitude, they strongly capture the temporal dynamics of gas exchange velocity and will provide an unprecedented amount of new information on global riverine gas exchange that should be essential for better constraining existing river $[CO_2]$ models.

## 6 Acknowledgements
C.B. Brinkerhoff was funded by a NASA Future Investigators in Earth and Space Science Fellowship (80NSSC21K1591). C.J. Gleason was supported by a NASA SWOT Science Team grant (80NSSC20K1141). We thank Renato Frasson, Michael Durand, Amber Ulseth, and Jake Beaulieu for generously making their data available for this study. We also thank the entire SWOT discharge algorithm working group for their decade-plus body of work which inspired this study.

## 7 Data availability statement
Datasets required for this research are available from @ulsethDistinctAirWater2019 (<https://doi.org/10.1038/s41561-019-0324-8>), @brinkerhoffReconcilingAtaStationAtManyStations2019 (<https://doi.org/10.1029/2019GL084529>), @frassonExploringFactorsControlling2021 (<https://zenodo.org/record/3817817>), @churchillPredictionStreamReaeration1964 (<https://pubs.usgs.gov/pp/0737/report.pdf>), @owensReaerationStudiesStreams1964 (<https://pubs.usgs.gov/pp/0737/report.pdf>), @beaulieuControlsGasTransfer2012 (<https://doi.org/10.1029/2011JG001794>), and @durandIntercomparisonRemoteSensing2016 (<https://doi.org/10.1002/2015WR018434>). All code to generate results and figures is archived at <https://zenodo.org/record/6914119>.

## 8 Software availability statement
The BIKER algorithm remains in active development and is available at <https://github.com/craigbrinkerhoff/BIKER>. The specific version of BIKER used in this study is archived at <https://zenodo.org/record/6914064>.

## 9 Appendix A

*Table A1: Variable description and notation for this study. Unit quantities are as follows: M for mass, L for length, T for time, P for pressure, and E for energy.*

| **Notation**           | **Description**                                                                 | **Calculation (if necessary)**                              | **Units**                    |
|------------------------|---------------------------------------------------------------------------------|-------------------------------------------------------------|------------------------------|
| $A$                    | Channel cross-sectional area                                                    |                                                             | $[L^2]$                      |
| $A_0$                  | Non-SWOT-observable cross-sectional area                                        |                                                             | $[L^2]$                      |
| $A_p$                  | Active zone fraction (Moog & Jirka 1999)                                        |$A_p \propto Re_*^{1/2}$(Moog & Jirka 1999)                  | $dimensionless$              |
| $[CO_2]$               | Carbon dioxide concentration                                                    |                                                             | $[\frac{M}{L^3}]$            |
| $[CO_{2_{water}}]$     | Water-side $[CO_2]$ concentration                                               |                                                             | $[\frac{M}{L^3}]$            |
| $[CO_{2_{air}}]$       | Atmospheric-side $[CO_2]$ concentration                                         |                                                             | $[\frac{M}{L^3}]$            |
| $dA$                   | change in cross-sectional area                                                  | $\sum_{t':W_{t'}\le W_t}{W_{t'} \delta H_e_{t'}}$           | $[L^2]$                      |
| $D_m$                  | Molecular diffusion coefficient                                                 |                                                             | $[\frac{L^2}{T}]$            |
| $\epsilon$             | Dissipation rate of near-surface turbulence                                     |                                                             | $[\frac{E}{M*T}]$            |
| $\epsilon_S$           | log-law-of-the-wall model for $\epsilon$                                        | $\frac{U_*^3}{H}$                                           | $[\frac{E}{M*T}]$            |
| $\epsilon_D$           | Form-drag model for $\epsilon$                                                  | $gS\bar{U}$                                                 | $[\frac{E}{M*T}]$            |
| $[FCO_2]$              | $[CO_2]$ evasion flux from river to atmosphere                                  |                                                             | $[\frac{M}{L^2T}]$           |
| $[FCO_{2_{upscaled}}]$ | Upscaling estimate of the global $[CO_2]$ evasion flux from river to atmosphere |                                                             | $[\frac{M}{L^2T}]$           |
| $g$                    | gravitational acceleration                                                      | 9.8                                                         | $[\frac{L}{T^2}]$            |
| $H$                    | Mean flow depth                                                                 | $\frac{A}{W}$                                               | $[L]$                        |
| $H_e$                  | Water surface elevation                                                         |                                                             | $[L]$                        |
| $i$                    | Cross-section discretization within a mass-conserved river reach                |                                                             |                              |
| $k_z$                  | gas exchange velocity for gas *z*                                               |                                                             | $[\frac{L}{T}]$              |
| $k_{600}$              | gas exchange velocity normalized to $Sc=600$                                    | $(\frac{600}{Sc})^{-0.5}k$                                  | $[\frac{L}{T}]$              |
| $n$                    | Manning's roughness coefficient                                                 | $\frac{R_h^{2/3}S^{1/2}}{\bar{U}}$                          | $[\frac{T}{L^3}]$            |
| $\uprho$               | Density of water                                                                |                                                             | $[\frac{M}{L^{1/3}}]$        |
| $Q$                    | River discharge                                                                 | $WH\bar{U}$                                                 | $\frac{L^3}{T}$              |
| $R_h$                  | Hydraulic radius                                                                | $\frac{AW}{2H+W}$                                           | $[L]$                        |
| $Re_*$                 | Shear Reynold's number                                                          | $\frac{U_*H}{v}$                                            | $dimensionless$              |
| $S$                    | River slope                                                                     |                                                             | $dimensionless$              |
| $Sc$                   | Schmidt number                                                                  | $\frac{v}{D_m}$                                             |                              |
| $t$                    | timestep discretization within river reach                                      |                                                             |                              |
| $\Theta$               | Bayesian parameter set                                                          |                                                             |                              |
| $\bar{U}$              | Cross-sectional average velocity                                                | $\frac{Q}{A}$                                               | $[\frac{L}{T}]$              |
| $U_*$                  | Shear velocity                                                                  | $\sqrt{gSR_h}$                                              | $[\frac{L}{T}]$              |
| $\mu$                  | Viscosity                                                                       |                                                             | $[P*T]$                      |
| $v$                    | kinematic viscosity                                                             | $\frac{\mu}{\uprho}$                                        | $[\frac{L^2}{T}]$            |
| $W$                    | Flow width                                                                      |                                                             | $[L]$                        |
| $x$                    | Bayesian data set                                                               |                                                             |                              |

*Table A2: Bayesian parameter and hyperparameter notation for this study.*

| **Notation**       | **Description**                                                                                       |
|--------------------|-------------------------------------------------------------------------------------------------------|
| *x*                | Joint data distribution for BIKER                                                                     |
| $\Theta$           | Joint parameter distribution for BIKER                                                                |
| $\Theta_\mu$       | Mean hyperparameter for BIKER prior distributions                                                     |
| $\Theta_\sigma$    | Standard deviation hyperparameter for BIKER prior distributions                                       |
| $\Theta_\gamma$    | Upper bound hyperparameter for BIKER prior distributions                                              |
| $\Theta_\lambda$   | Lower bound hyperparameter for BIKER prior distributions                                              |
| $\sigma_{k_{600}}$ | 'Uncertainty' hyperparameter for BIKER likelihood distribution                                        |
| $\alpha_n$         | Coefficient parameter distributions for k600 scaling relations, where n = independent variable number |
| $\beta_1$          | Intercept parameter distribution for k600 scaling relations                                           |
| $\sigma_{LM}$      | Uncertainty parameter distribution for k600 scaling relations                                         |

## References

### 2022-07-27 supp_r2.Rmd
riewew round 2 big push

--- 2022-06-08 docs/intial_drafts/supp_v11.Rmd
@@
   word_document:
-  #    pandoc_args:
-  #      - '--lua-filter=scholarly-metadata.lua'
-  #      - '--lua-filter=author-info-blocks.lua'
+      pandoc_args:
+        - '--lua-filter=scholarly-metadata.lua'
+        - '--lua-filter=author-info-blocks.lua'
       reference_docx: word-styles-reference.docx
@@
 knitr::opts_chunk$set(echo = TRUE)
+options(scipen=999)
 ```
@@
 ## Contents
-This supplementary information contains 5 texts, 5 figures and 2 tables.
+This supplementary information contains 6 texts, 6 figures and 4 tables.
 
@@
 Consult Appendix A for all variable definitions.
-Note that $\alpha$ and $\beta$ are statistical parameters obtained via least squares regression for the small-eddy and Reynolds-extension models, respectively.
 Subscripts denote variants of the same parameter, depending on which $\epsilon$ model is employed (see Main Text).
@@
 
-$$k_{600}=\alpha(\epsilon_S)^{1/4} \mathbf{(S1)}$$
-$$k_{600}=\alpha(\frac{U_*^3}{H})^{1/4} \mathbf{(S2)}$$
-$$k_{600}=\alpha(\frac{\sqrt{gR_hS}^3}{H})^{1/4} \mathbf{(S3)}$$
-$$k_{600}=\alpha(\frac{(gHS)^{3/2}}{H})^{1/4} \mathbf{(S4)}$$
-$$k_{600}=\alpha(gS)^{3/8}H^{1/8} \mathbf{(S5)}$$
+$$k_{600}=\beta_1(\epsilon_S)^{1/4} \mathbf{(S1)}$$
+$$k_{600}=\beta_1(\frac{U_*^3}{H})^{1/4} \mathbf{(S2)}$$
+$$k_{600}=\beta_1(\frac{\sqrt{gR_hS}^3}{H})^{1/4} \mathbf{(S3)}$$
+$$k_{600}=\beta_1(\frac{(gHS)^{3/2}}{H})^{1/4} \mathbf{(S4)}$$
+$$log(k_{600})=\beta_1+\frac{3}{8}log(gS)+\frac{1}{8}log(H)\mathbf{(S5)}$$
 
@@
 
-$$k_{600}=\alpha_1(\epsilon_D)^{1/4} \mathbf{(S6)}$$
-$$k_{600}=\alpha_1(gS\bar{U})^{1/4} \mathbf{(S7)}$$
+$$k_{600}=\beta_1(\epsilon_D)^{1/4} \mathbf{(S6)}$$
+$$log(k_{600})=\beta_1+\frac{1}{4}log(gS\bar{U}) \mathbf{(S7)}$$
 
@@
 
-$$k_{600}=\beta(\epsilon_S)^{1/4}A_p^{3/4} \mathbf{(S8)}$$
-$$k_{600}=\beta (\frac{U_*^3}{H})^{1/4}((U_*H)^{1/2})^{3/4} \mathbf{(S9)}$$
-$$k_{600}=\beta U_*^{9/8}H^{1/8} \mathbf{(S10)}$$
-$$k_{600}=\beta (gR_hS)^{9/16}H^{1/8} \mathbf{(S11)}$$
+$$k_{600}=\beta_1(\epsilon_S)^{1/4}A_p^{3/4} \mathbf{(S8)}$$
+$$k_{600}=\beta_1(\frac{U_*^3}{H})^{1/4}((U_*H)^{1/2})^{3/4} \mathbf{(S9)}$$
+$$k_{600}=\beta_1 U_*^{9/8}H^{1/8} \mathbf{(S10)}$$
+$$k_{600}=\beta_1 (gR_hS)^{9/16}H^{1/8} \mathbf{(S11)}$$
 $$k_{600}=\beta (gHS)^{9/16}H^{1/8} \mathbf{(S12)}$$
-$$k_{600}=\beta (gS)^{9/16}H^{11/16} \mathbf{(S13)}$$
+
+$$log(k_{600})=\beta_1+\frac{9}{16}log(gS)+\frac{11}{16}log(H) \mathbf{(S13)}$$
 
@@
 $$k_{600}=\beta_1(gS\bar{U})^{1/4}(gHS)^{3/16}H^{3/8} \mathbf{(S17)}$$
-$$k_{600}=\beta_1(gS)^{7/16}\bar{U}^{1/4}H^{9/16} \mathbf{(S18)}$$
+$$log(k_{600})=\beta_1+\frac{7}{16}log(gS)+\frac{1}{4}log(\bar{U})+\frac{9}{16}log(H) \mathbf{(S18)}$$
 
-## Text S3: BIKER hyperparameterization
+## Text S3: Bayesian linear regression hyperparameterization
+As detailed in section 2.4, we fit a Bayesian linear regression model using equation 7 and our `r n` estimates of hydraulically-wide $k_{600}$ Here, we provide the model specification.
+Equation S19 is the regression equation and the prior hyperparameterizations are equations S20-S24.
+Priors are specified as normal distributions $N(\mu, \sigma^2)$, save $\sigma_{LM}$ which is characteristically specified as an exponential distribution.
+
+$$log(k_{600})=\beta_1+\alpha_1log(gS)+\alpha_2log(\bar{U})+\alpha_3log(H) \mathbf{(S19)}$$
+$$\alpha_1 \sim N(\frac{7}{16}, \frac{1}{8}^2) \mathbf{(S20)}$$
+$$\alpha_2 \sim N(\frac{1}{4}, \frac{1}{8}^2) \mathbf{(S21)}$$
+$$\alpha_3 \sim N(\frac{9}{16}, \frac{1}{8}^2) \mathbf{(S22)}$$
+$$\beta_1 \sim N(0, 1^2) \mathbf{(S23)}$$
+$$\sigma_{LM} \sim exp(1) \mathbf{(S24)}$$
+
+The equation 7 coefficients, derived from geomorphic assumptions of the SWOT-observable river channel, were used to assign reasonably informative priors for $\alpha_1$, $\alpha_2$, and $\alpha_3$.
+For the other models in Figure S1, the prior distribution means for $\alpha_1$, $\alpha_2$, and $\alpha_3$ were changed to their respective coefficients derived in Text S2.
+$\beta_1$ and the model uncertainty $\sigma_{LM}$ were left as uninformative given that we lack any prior knowledge of what they might be.
+
+## Text S4: BIKER hyperparameterization
 In this text we explain in detail how BIKER's prior distributions were determined for a given river.
@@
 This leaves the $k_{600_t}$ hyperparameters to be defined.
-$k_{600_{\mu}}$ is set by invoking the hydraulic geometry (HG) relationships described in Section 3.2.3 of the main text and Table S2 using the data from @brinkerhoffReconcilingAtaStationAtManyStations2019.
-We replaced both depth and velocity terms from our gas exchange model (Equation 7 in the Main Text) with these HG models, resulting in Equation S19 where Q is the mass-conserved streamflow for the river reach.
+$k_{600_{\mu}}$ is set by invoking the hydraulic geometry (HG) relationships described in Section 3.2.3 of the main text and Table S3 using the data from @brinkerhoffReconcilingAtaStationAtManyStations2019.
+We replaced both depth and velocity terms from our gas exchange model (Equation 7 in the Main Text) with these HG models, resulting in Equation S25 where *Q* is the mass-conserved streamflow for the river reach.
 
-$$k_{600_{\mu}} = 62.82(gS_{i,t})^{7/16}(0.252Q^{0.388})^{9/16}*(0.276Q^{0.164})^{1/4} \mathbf{(S19)}$$
+$$k_{600_{\mu}} = 62.82(gS_{i,t})^{7/16}(0.252Q^{0.388})^{9/16}*(0.276Q^{0.164})^{1/4} \mathbf{(S25)}$$
 
-Obviously, we have no a priori information about Q.
-So, we use the Q prior that will be globally available when SWOT launches: a mean annual estimate from a water balance model.
-This is the standard prior information used to benchmark the SWOT discharge algroithms and is provided with the simulated SWOT data by both @durandIntercomparisonRemoteSensing2016 and @frassonExploringFactorsControlling2021.
+Obviously, we have no a priori information about *Q*.
+So, we use the *Q* prior that will be globally available when SWOT launches: a mean annual estimate from a water balance model.
+This is the standard prior information used to benchmark the SWOT discharge algorithms and is provided with the simulated SWOT data by both @durandIntercomparisonRemoteSensing2016 and @frassonExploringFactorsControlling2021.
 Using a temporally-invariant estimate of streamflow is the worst case scenario and BIKER's performance will improve with a more informed prior on streamflow (and therefore $k_{600}$).
@@
 
-$k_{600_{\sigma}}$ is set to 0.30 (log-space).
-This corresponds to a coefficient of variation of approximately 30%, which we took to reflect a reasonably strong agreement between the prior and the observed values.
+$k_{600_{\sigma}}$ is set to 0.58 (log-space).
+This was provided by posterior mean $\sigma_{LM}$ from the Bayesian linear regression (equation 8 and Table 1) in the main text (Text S3 as well).
 $k_{600_{\lambda}}$ was set to log(0.001) m/day.
@@
 
-## Text S4: Determining complete BIKER model uncertainty
-Recall that $\sigma_{k_{600}}$ refers to the total uncertainty inherent in Equation 9, i.e.
-stemming from the Reynolds-extension model (Equation 7) and Manning's equation for $\bar{U}$.
-For the purposes of this study, we are validating BIKER against Equation 7 in the main text and so all uncertainties associated with the $\beta_1$ parameter are ignored and we only need to reflect the Manning's uncertainty in our specification of $\sigma_{k_{600}}^2$.
+## Text S5: Determining complete BIKER model uncertainty
+Recall that $\sigma_{k_{600}}$ refers to the total uncertainty inherent in Equation 10, i.e.
+stemming from the Reynolds-extension model (Equation 8) and Manning's equation for $\bar{U}$.
+For the purposes of this study, we are validating BIKER against Equation 8 in the main text and so all uncertainties associated with the $\beta_1$ parameter are ignored and we only need to reflect the Manning's uncertainty in our specification of $\sigma_{k_{600}}^2$.
 Therefore, we take @hagemannBAMBayesianAMHGManning2017's estimated uncertainty from Manning's equation to infer streamflow from SWOT observations (0.25) and inflate it slightly to also account for the hydraulically-wide channel assumption and arrive at 0.30.
@@
 
-In the scenario that BIKER is run on real SWOT data, $\sigma_k_{600}$ must reflect the full uncertainty implicit in Equation 9 in the main text.
+In the scenario that BIKER is run on real SWOT data, $\sigma_k_{600}$ must reflect the full uncertainty implicit in Equations 10 and 11 in the main text.
 This means we must also account for uncertainty from the $\beta_1$ parameter.
-Assuming perfect, no-error measurements are made by the SWOT satellite, the full Equation 7 uncertainty is expressed for some set of hydraulic observations as Equation S20.
-To quantify total uncertainty in $k_{600}$, we push Monte Carlo simulations through Equation S20 and take the mean uncertainty as a reasonable value for $\sigma_{k_{600}}$ once SWOT launches.
+Assuming perfect, no-error measurements are made by the SWOT satellite, the full Equation 8 uncertainty is expressed for some set of hydraulic observations as Equation S25.
+To quantify total uncertainty in $k_{600}$, we push Monte Carlo simulations through Equation S25 and take the mean uncertainty as a reasonable value for $\sigma_{k_{600}}$ once SWOT launches.
 
-More specifically, we use the 166 hydraulically-wide measurements in our field-measured $k_{600}$ dataset (Section 2 in the Main Text) and push 166 different Monte Carlo simulations through Equation S20.
+More specifically, we use the 166 hydraulically-wide measurements in our field-measured $k_{600}$ dataset (Section 2 in the Main Text) and run `r n` different Monte Carlo simulations using the distributions for $\alpha_1$, $\alpha_2$, $\alpha_3$, $\beta_1$, $k_{600_\sigma}$, and the Manning's equation uncertainty (0.30 in log-space).
 Note that each Monte Carlo simulation is itself 10,000 runs in order to obtain a distribution of $k_{600}$ estimates.
-For each of the 166 distributions of estimated $k_{600}$, we extract the standard deviation as the uncertainty.
+For each of the `r n` distributions of estimated $k_{600}$, we extract the standard deviation as the uncertainty.
 We then take the average of those standard deviations to be a reasonable estimate of $\sigma_k_{600}$.
-This ultimately provided a value of 1.1 for the log-transformed $\sigma_k_{600}$ and should be used when running BIKER on real SWOT data.
+This ultimately provided a value of 0.71 and should be used when running BIKER on real SWOT data.
 
-$$log_{k_{600}} \sim \mathcal{N} \left[ log(\beta_1), log(\sigma_{\beta_1}) \right] + \frac{1}{4}\mathcal{N}\left[log(\bar{U}), log(\sigma_{mannings})\right] \frac{7}{16}log(gS) + \frac{9}{16}log(H)  \mathbf{(S20)}$$
+## Text S6: Literature $[CO_2]$ flux models
+Here, we describe in more detail the three $[CO_2]$ flux models used to compare against BIKER in Section 3.2.3. The specific equations used are outlined in Table S3.
 
-## Text S5: Literature $CO_2$ flux models
-Here, we describe in more detail the three $CO_2$ flux models used to compare against BIKER in Section 3.2.3. The specific equations used are outlined in Table S2.
+In brief, gas fluxes ($[FCO_2]$) are only obtainable at the global scale via predictive equations for $k_{CO_2}$ per Equation 2 and 13.
+Here, we obtain $[FCO_2]$ using the in situ $[CO_2]$ data described in Section 3.1 and $k_{CO_2}$ estimates.
+The $k_{CO_2}$ estimates are calculated using Equation 8.
+Equation 8 is calculated using 1) in situ streamflow data, and 2) hydraulic scaling equations from the literature that predict river velocity and depth from streamflow.
+Specifically, these literature equations are the 'Raymond 2012', 'Raymond 2013', and 'Brinkerhoff 2019' models (Table S3).
 
-In brief, gas fluxes ($FCO_2$) are only obtainable at the global scale via predictive equations for $k_{CO_2}$ per Equation 2.
-Here, we obtain $FCO_2$ using the in situ $CO_2$ data described in Section 3.1 and $k_{CO_2}$ estimates.
-The $k_{CO_2}$ estimates are calculated using Equation 7.
-Equation 7 is calculated using 1) in situ streamflow data, and 2) hydraulic scaling equations from the literature that predict river velocity and depth from streamflow.
-Specifically, these literature equations are the 'Raymond 2012', 'Raymond 2013', and 'Brinkerhoff 2019' models (Table S2).
-
-For example, the 'Raymond 2012' model uses the hydraulic scaling equations from @raymondScalingGasTransfer2012 to predict velocity and depth, which are in turn used to calculate a $k_{CO_2}$ via equation 7, which finally is used to calculate $FCO_2$ via equation 1.
+For example, the 'Raymond 2012' model uses the hydraulic scaling equations from @raymondScalingGasTransfer2012 to predict velocity and depth, which are in turn used to calculate a $k_{CO_2}$ via equation 8, which finally is used to calculate $[FCO_2]$ via equation 1.
 Similarly, the 'Raymond 2013' model uses the equations outlined in @raymondGlobalCarbonDioxide2013 and the 'Brinkerhoff 2019' model uses new equations fit to the hydraulics dataset from @brinkerhoffReconcilingAtaStationAtManyStations2019 after the filtering described in Text S1.
@@
 ## Figure S1
-![Figure S1: Four $k_{600}$~hydraulics models tested in this study, with model assumptions noted in the sub-panel titles.
-Panel (d) is also presented in the main text as Figure 2 and Equation 7.](cache/k600_theory/figS1.jpg)
+![Figure S1: Flowchart detailing entire study workflow.](cache/maps/workflow.jpg)
 
 ## Figure S2
-![Figure S2: Map of simulated SWOT river locations along the global SWOT river network [@altenauSurfaceWaterOcean2021].
+![Figure S2: Four $k_{600}$~hydraulics models tested in this study, with model assumptions noted in the sub-panel titles.
+Panel (d) is also presented in the main text as Figure 2 and Equation 8.](cache/k600_theory/figS1.jpg)
+
+## Figure S3
+![Figure S3: Map of simulated SWOT river locations along the global SWOT river network [@altenauSurfaceWaterOcean2021].
 This corresponds to every river that BIKER could be run on once SWOT launches.
-Also includes approximate location of the $CO_2$ timeseries used in this study (Figure S3).
+Also includes approximate location of the $[CO_2]$ timeseries used in this study (Figure S4).
 Note that river locations are approximate as some hydraulic models are not geo-referenced.](cache/maps/figS2.jpg)
 
-## Figure S3
-![Figure S3 Timeseries of the biweekly $CO_2$ data from @beaulieuControlsGasTransfer2012.
+## Figure S4
+![Figure S4 Timeseries of the biweekly $[CO_2]$ data from @beaulieuControlsGasTransfer2012.
 Sampling took place 2008-2009 in the Ohio River (upstream of Cincinnati, Ohio, United States).
-Dashed black line denotes atmospheric $CO_2$ at 400 uatm.](cache/FCO2/figS3.jpg)
+Dashed black line denotes atmospheric $[CO_2]$ at 400 uatm.](cache/FCO2/figS3.jpg)
 
-## Figure S4
-![Figure S4: Observed (orange) versus BIKER-inferred (green) timeseries of daily $k_{600}$ for the 47 SWOT-simulated rivers under the 'no-measurement-error' scenario.
-Green ribbon is BIKER posterior 95% confidence intervals.
+## Figure S5
+![Figure S5: Observed (orange) versus BIKER-inferred (green) timeseries of daily $k_{600}$ for the 47 SWOT-simulated rivers under the 'no-measurement-error' scenario.
+Green ribbon is BIKER posterior 95% credible intervals.
 Note that the y axis is normalized by maximum observed values to compare across rivers.](cache/validation/figS4.jpg)
 
-## Figure S5
-![Figure S5: Observed (orange) versus BIKER-inferred (green) timeseries of daily $k_{600}$ for 16 rivers under the 'measurement-error' scenario.
-Green ribbon is BIKER posterior 95% confidence intervals.
+## Figure S6
+![Figure S6: Observed (orange) versus BIKER-inferred (green) timeseries of daily $k_{600}$ for 16 rivers under the 'measurement-error' scenario.
+Green ribbon is BIKER posterior 95% credible intervals.
 Note that the y axis is normalized by maximum observed values to compare across rivers.](cache/validation/figS5.jpg)
@@
 ## Table S2
-*Table S2: Details on the 3 literature models for $CO_2$ fluxes. See Text S5 for more details.*
+*Table S2: Validation metrics used in this study, where Nt is number of observations and i is the specific observation.
+$\sigma$ refers to the variance of the sample and $\mu$ refers to the mean of the sample.
+Carrot accents indicate the predicted value.*
+
+| **Description**                   | **Acronym** | **Definition**                                                                                                                                                                                                 | **Ideal Score** | **Possible Range** |
+|-----------------------------------|-------------|----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|-----------------|--------------------|
+| Correlation Coefficient           | $r$         | $\sqrt{1-(\frac{\sum\limits_{i=1}^{N_t}{(k_{600, i}-\hat{k_{600_i}})^2}}{\sum{(k_{600, i}-\mu_{k_{600}})^2}})}$                                                                                                | 1               | -1 to 1            |
+| Normalized root-mean-square error | NRMSE       | $\sqrt{\frac{1}{N_t}\sum\limits_{i=1}^{N_t}(\frac{\hat{k_{600, i}}-k_{600, i}}{\mu_{k_{600, i}}})^2}$                                                                                                          | 0               | 0 to ∞             |
+| Kling-Gupta Efficiency            | KGE         | $1 - \sqrt{ (r(k_{600_i}, \hat{k_{600_i}}) - 1)^2 + (\frac{\mu_{\hat{k_{600}}} }{\mu_{k_{600}}} - 1)^2 + (\frac{ \sigma _{\hat{k_{600}}} / \mu _{\hat{k_{600}}} }{ \sigma_{k_{600}} / \mu_{k_{600}}} - 1)^2 }$ | 1               | -∞ to 1            |
+| Normalized Mean Absolute Error    | MAE         | $\frac{\frac{\sum\limits_{i=1}^{N_t}(|\hat{k_{600_i}}-k_{600_i})|)}{n})}{\mu_{k_{600}}}$                                                                                                                       | 0               | 0 to ∞             |
+
+## Table S3
+*Table S3: Details on the 3 literature models for $CO_2$ fluxes. See Text S6 for more details.*
 
@@
 
+## Table S4
+*Table S4: Comparison of SWOT-observable hydrography and global hydrography (at mean annual streamflow conditions).*
+
+| **Hydrography Model** | **Surface Area $\mathbf{[km^2]}$** | **Length [km]** |
+|--|--|--|
+| SWOT River Network [@altenauSurfaceWaterOcean2021] | `r round(global_swot_coverage[global_swot_coverage$model == 'swot',]$sa_km2,0)` | `r round(global_swot_coverage[global_swot_coverage$model == 'swot',]$length_km,0)` |
+| Most recent global hydrography estimate [@liuImportanceHydrologyRouting2022] | `r round(global_swot_coverage[global_swot_coverage$model == 'total',]$sa_km2,0)` | `r round(global_swot_coverage[global_swot_coverage$model == 'total',]$length_km,0)` |
+| % that SWOT observes [%] | `r round((global_swot_coverage[global_swot_coverage$model == 'swot',]$sa_km2 / global_swot_coverage[global_swot_coverage$model == 'total',]$sa_km2)*100, 1)` | `r round((global_swot_coverage[global_swot_coverage$model == 'swot',]$length_km / global_swot_coverage[global_swot_coverage$model == 'total',]$length_km)*100, 1)` |
 ## References

### 2022-07-27 docs/review_round2/k_model_for_reviewer1.R

@@ -1,291 +1,0 @@
-#################
-##Description: Development of gas exchange velocity model for 'hydraulically wide channels'
-              #It also estimates how many SWOT-observable rivers meet this hydraulically-wide condition
-##Creator: Craig Brinkerhoff
-##Date: Fall 2021
-
-##TO RUN:
-#  1) set the working directory below to the '~/RSK600' directory
-#  2) run script!
-#################
-
-#######SETUP-----------------------------
-library(dplyr)
-library(tidyr)
-library(ggplot2)
-library(colorspace)
-library(cowplot)
-library(readr)
-library(rstanarm)
-library(grid)
-theme_set(theme_classic())
-
-#some constants
-g <- 9.8 #gravitational acceleration [m/s2]
-
-#######READ IN HYDRAULICS DATA FROM BRINKERHOFF ETAL 2019 TO GET LARGER HYDRAULICS DATASET------------------------------
-data <- read.csv('data\\Brinkerhoff_etal_2019\\field_measurements.csv')
-
-#some necessary filtering
-data <- filter(data, is.finite(chan_width) ==1) %>%
-  filter(is.finite(chan_velocity)==1) %>%
-  filter(is.finite(chan_discharge)==1) %>%
-  filter(is.finite(chan_area)==1) %>%
-  filter(chan_width > 0) %>%
-  filter(chan_velocity > 0) %>%
-  filter(chan_discharge > 0) %>%
-  filter(chan_area > 0) %>%
-  filter(measured_rating_diff %in% c('Excellent', 'EXCL', 'GOOD', 'Good'))
-
-#imperial to metric
-data$area <- data$chan_area * 0.092903 #ft2 to m2
-data$width <- data$chan_width*0.305 #m
-data$Vms <- data$chan_velocity*0.305 #m/s
-data$depth <- data$area / data$width
-data$slope <- NA 
-data$Qm3s <- data$chan_discharge * 0.0283 #ft3/s to m3/s
-
-#calculate bankfull width so we can throw out out-of-bank events
-bankfullW = group_by(data, site_no) %>%
-  filter(n() >= 20) %>%
-  mutate(rank = rank(width, ties.method="first")) %>%
-  mutate(n = n()) %>%
-  mutate(desiredRank = round((n+1)/2, digits=0)) %>% #2 for a two year return period
-  mutate(temp = which(rank == desiredRank)) %>%
[…]

### 2022-07-27 generateMS.R

@@ -4,16 +4,18 @@
 
 #Move all results to global environment
 modelsWIDE <- read.csv('cache\\k600_theory\\hydraulicWide_models.csv')
+modelsSTATS <- read.csv('cache\\k600_theory\\k_stats_models.csv')
 n <- 166 #number of hydraulically wide measurements used in k600 validation
 HG_swot <- read.csv('cache\\k600_theory\\HG_swot.csv')
+global_swot_coverage <- read.csv('cache\\swot_total_comparison.csv')
 
 results_k600_rivs <- read.csv('cache//validation//results_by_riv.csv')
 results_fco2_bulk <- read.csv('cache//FCO2//massFluxes.csv')
 results_fco2_bulk$key <- as.character(results_fco2_bulk$key)
 
 #knit manuscript
-rmarkdown::render("manuscript_r_v1.Rmd")
+rmarkdown::render("manuscript_r2.Rmd")
 print('Manuscript generated.')
 
-rmarkdown::render("supp_r_v1.Rmd")
+rmarkdown::render("supp_r2.Rmd")
 print('Supplemental Information generated.')

### 2022-07-27 src/calcSWOT_SA.R

@@ -1,6 +1,6 @@
 ##################
 ## Calculates SWORD/SWOT mean annual surface area (widths come from GRWL and /or MERIT Hydro) and network length
-## Does eachcontinent seperately to spare holding the global river network in memory...
+## Does each continent separately to spare holding the global river network in memory...
 ## Craig Brinkerhoff
 ## Summer 2022
 ###################
@@ -8,12 +8,13 @@
 library(sf)
 
 #path to SWORD v6
-setwd('C:\\Users\\cbrinkerhoff\\OneDrive - University of Massachusetts\\Datasets\\SWORD_v06\\Reaches_Nodes\\shp')
+setwd('C:\\Users\\craig\\OneDrive - University of Massachusetts\\Datasets\\SWORD_shp_v2\\shp')
 
 ########################################SWOT/SWORD########################################################################
 #get files
-files <- list.files(pattern = '*.shp', recursive = TRUE)
-files <- files[substr(files, 22, 28) == 'reaches']
+files <- list.files(recursive = TRUE)
+files <- files[substr(files, nchar(files)-3, nchar(files)) == '.shp']
+files <- files[substr(files, 13, 19) == 'reaches']
 
 #run per continent, concatenating SA
 SA_fin <- 0
@@ -30,9 +31,11 @@ for (k in continent){
     sword <- rbind(sword, temp_sf)
   }
   
+  sword <- sword[sword$width >= 50 & substr(sword$reach_id, 11,11) == '1',] #grab only rivers, ignore
+  
   #estimate SA
-  SA <- sum(sword$width * sword$reach_len, na.rm=T) #[m2]
-  SA <- SA*1e-6 #[km2]
+  SA <- sum((sword$width * sword$reach_len)*1e-6, na.rm=T) #[m2]
+  #SA <- SA/1000000 #[km2]
   SA_fin <- SA_fin + SA
   print(paste0(k, ' SA: ', SA))
   print(paste0('total SA: ', SA_fin))
@@ -47,4 +50,10 @@ for (k in continent){
 
 #calc percent SWOT observable
 SA_fin/811000 #Liu et al 2022
-length_fin/23300000 #riverATLAS Messager et al 2021
+length_fin/443509286 #Liu et al 2022
+
+df <- data.frame('sa_km2'=c(SA_fin, 811000),
+                 'length_km' = c(length_fin, 443509286),
+                 'model'=c('swot', 'total'))
+
[…]

### 2022-07-27 src/figures_FCO2.R

@@ -54,7 +54,7 @@ barPlot <- ggplot(massFluxes, aes(y=yearlyCFux, x=key, fill=key)) +
   geom_errorbar(aes(x=key, ymin=lowCI_sum, ymax=highCI_sum), width=0.4, colour="black", size=1.3)+
   geom_hline(yintercept = massFluxes[massFluxes$key=='FCO2_obs', ]$yearlyCFux, size=1.2, linetype='dashed', color='#ff7f00') +
   ylab('CO2 Flux [Tg-C/yr]') +
-  scale_fill_manual(values=c('#e31a1c', '#3f007d', '#6a51a3', '#9e9ac8', '#ff7f00'), name='', labels=c('BIKER \n ', 'Brinkerhoff \n2019', 'Raymond \n2012', 'Raymond \n2013', 'Observed \n ')) +
+  scale_fill_manual(values=c('#e31a1c', '#3f007d', '#6a51a3', '#9e9ac8', '#ff7f00'), name='', labels=c('BIKER \n ', 'Brinkerhoff \n2019', 'Raymond \n2012', 'Raymond \n2013', '"Observed" \n ')) +
   theme(legend.position = "right",
         axis.title.x=element_blank(),
         axis.text.x=element_blank(),

### 2022-07-27 src/figures_k600.R

@@ -73,7 +73,7 @@ nrmse_cdf <- ggplot(stats_by_river[stats_by_river$errFlag==0,], aes(x=NRMSE)) +
 
 nmae_cdf <- ggplot(stats_by_river[stats_by_river$errFlag==0,], aes(x=NMAE)) +
   stat_ecdf(size=3, color='#1c9099') +
-  scale_x_log10()+
+  scale_x_log10(limits=c(min(stats_by_river[stats_by_river$errFlag==0,]$NMAE),max(stats_by_river[stats_by_river$errFlag==0,]$NMAE)+0.3))+ #kluge to force label to not go off of screen...
   geom_hline(yintercept = 0.5, linetype='dashed', size=1) +
   xlab('Value')+
   ylab('')+
@@ -181,7 +181,7 @@ nmae_scatter <- ggplot(forError_nmae, aes(x=abs(nmae_no_err), y=abs(nmae_err))) 
         legend.title = element_text(size=17, face='bold'),
         plot.title = element_text(size = 30, face = "bold"))
 
-fig5 <- plot_grid(kge_scatter, NRMSE_scatter, r_scatter, nmae_scatter, ncol=2, labels=NA)
+fig5 <- plot_grid(kge_scatter, NRMSE_scatter, r_scatter, nmae_scatter, ncol=2, labels='auto')
 ggsave('cache/validation/fig5.jpg', fig5, width=10, height=10)
 
 ###########################
@@ -209,7 +209,7 @@ priorPosterPlot_2 <- ggplot(forPlot2[forPlot2$cvObs >= 0.10,], aes(x=value, colo
   xlab('')+
   ylab('')+
   xlim(0,2.2)+
-  annotate("text", x = 1.25, y = 0.5, label = 'Observed CV > 0.10\n(n = 38 rivers)', size=8, color='darkblue')+
+  annotate("text", x = 1.25, y = 0.5, label = 'Observed CV > 0.10\n(n = 39 rivers)', size=8, color='darkblue')+
   theme(axis.text=element_text(size=22),
         axis.title=element_text(size=26,face="bold"),
         plot.title = element_text(size = 30, face = "bold"),
@@ -221,7 +221,7 @@ priorPosterPlot_3 <- ggplot(forPlot2[forPlot2$cvObs >= 0.20,], aes(x=value, colo
   xlab('NMAE')+
   ylab('Percentile')+
   xlim(0,2.2)+
[…]

### 2022-07-27 src/old/swot_k_model_v3.R

@@ -1,0 +1,301 @@
+#################
+##Description: Development of gas exchange velocity model for 'hydraulically wide channels'
+              #It also estimates how many SWOT-observable rivers meet this hydraulically-wide condition
+##Creator: Craig Brinkerhoff
+##Date: Fall 2021
+
+##TO RUN:
+#  1) set the working directory below to the '~/RSK600' directory
+#  2) run script!
+#################
+
+#######SETUP-----------------------------
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+library(colorspace)
+library(cowplot)
+library(readr)
+theme_set(theme_classic())
+
+#some constants
+g <- 9.8 #gravitational acceleration [m/s2]
+
+#######READ IN HYDRAULICS DATA FROM BRINKERHOFF ETAL 2019 TO GET LARGER HYDRAULICS DATASET------------------------------
+data <- read.csv('data\\Brinkerhoff_etal_2019\\field_measurements.csv')
+
+#some necessary filtering
+data <- filter(data, is.finite(chan_width) ==1) %>%
+  filter(is.finite(chan_velocity)==1) %>%
+  filter(is.finite(chan_discharge)==1) %>%
+  filter(is.finite(chan_area)==1) %>%
+  filter(chan_width > 0) %>%
+  filter(chan_velocity > 0) %>%
+  filter(chan_discharge > 0) %>%
+  filter(chan_area > 0) %>%
+  filter(measured_rating_diff %in% c('Excellent', 'EXCL', 'GOOD', 'Good'))
+
+#imperial to metric
+data$area <- data$chan_area * 0.092903 #ft2 to m2
+data$width <- data$chan_width*0.305 #m
+data$Vms <- data$chan_velocity*0.305 #m/s
+data$depth <- data$area / data$width
+data$slope <- NA 
+data$Qm3s <- data$chan_discharge * 0.0283 #ft3/s to m3/s
+
+#calculate bankfull width so we can throw out out-of-bank events
+bankfullW = group_by(data, site_no) %>%
+  filter(n() >= 20) %>%
+  mutate(rank = rank(width, ties.method="first")) %>%
+  mutate(n = n()) %>%
+  mutate(desiredRank = round((n+1)/2, digits=0)) %>% #2 for a two year return period
[…]

### 2022-07-27 src/swot_k_model_v3.R

@@ -1,301 +1,0 @@
-#################
-##Description: Development of gas exchange velocity model for 'hydraulically wide channels'
-              #It also estimates how many SWOT-observable rivers meet this hydraulically-wide condition
-##Creator: Craig Brinkerhoff
-##Date: Fall 2021
-
-##TO RUN:
-#  1) set the working directory below to the '~/RSK600' directory
-#  2) run script!
-#################
-
-#######SETUP-----------------------------
-library(dplyr)
-library(tidyr)
-library(ggplot2)
-library(colorspace)
-library(cowplot)
-library(readr)
-theme_set(theme_classic())
-
-#some constants
-g <- 9.8 #gravitational acceleration [m/s2]
-
-#######READ IN HYDRAULICS DATA FROM BRINKERHOFF ETAL 2019 TO GET LARGER HYDRAULICS DATASET------------------------------
-data <- read.csv('data\\Brinkerhoff_etal_2019\\field_measurements.csv')
-
-#some necessary filtering
-data <- filter(data, is.finite(chan_width) ==1) %>%
-  filter(is.finite(chan_velocity)==1) %>%
-  filter(is.finite(chan_discharge)==1) %>%
-  filter(is.finite(chan_area)==1) %>%
-  filter(chan_width > 0) %>%
-  filter(chan_velocity > 0) %>%
-  filter(chan_discharge > 0) %>%
-  filter(chan_area > 0) %>%
-  filter(measured_rating_diff %in% c('Excellent', 'EXCL', 'GOOD', 'Good'))
-
-#imperial to metric
-data$area <- data$chan_area * 0.092903 #ft2 to m2
-data$width <- data$chan_width*0.305 #m
-data$Vms <- data$chan_velocity*0.305 #m/s
-data$depth <- data$area / data$width
-data$slope <- NA 
-data$Qm3s <- data$chan_discharge * 0.0283 #ft3/s to m3/s
-
-#calculate bankfull width so we can throw out out-of-bank events
-bankfullW = group_by(data, site_no) %>%
-  filter(n() >= 20) %>%
-  mutate(rank = rank(width, ties.method="first")) %>%
-  mutate(n = n()) %>%
-  mutate(desiredRank = round((n+1)/2, digits=0)) %>% #2 for a two year return period
[…]

### 2022-07-27 src/swot_k_model_v4.R

@@ -134,9 +134,8 @@ hydraulicallyWide$logDepth <- log10(hydraulicallyWide$depth)
 bayeslm_log_hydraulicallyWide_smallEddy_eS <- stan_glm(log10(k600)~logGravitySlope+logDepth, data=hydraulicallyWide,
                                                       family=gaussian(link='identity'),
                                                       prior = normal(location=c(3/8, 1/8), scale=c(1/8,1/8)),
-                                                      prior_intercept = normal(location=0.47, scale=1),#location=0 after internal recentering
-                                                      #intercept prior is left weakly informed using default prior, i.e. normal(0.47, 2.5)
-                                                      #sigma prior is left weakly informed using default prior, i.e. exponential(rate=1)
+                                                      prior_intercept = normal(location=0, scale=1),
+                                                      prior_aux = exponential(rate = 1),
                                                       seed=12345)
 
 #calc 'bayesian r2': https://doi.org/10.1080/00031305.2018.1549100
@@ -179,9 +178,8 @@ hydraulicallyWide$logDepth <- log10(hydraulicallyWide$depth)
 bayeslm_log_hydraulicallyWide_reynolds_eS <- stan_glm(log10(k600)~logGravitySlope+logDepth, data=hydraulicallyWide,
                                                        family=gaussian(link='identity'),
                                                        prior = normal(location=c(9/16, 11/16), scale=c(1/8,1/8)),
-                                                       prior_intercept = normal(location=0.47, scale=1),#location=0 after internal recentering
-                                                       #intercept prior is left weakly informed using default prior, i.e. normal(0.47, 2.5)
-                                                       #sigma prior is left weakly informed using default prior, i.e. exponential(rate=1)
+                                                       prior_intercept = normal(location=0, scale=1),
+                                                       prior_aux = exponential(rate = 1),
                                                        seed=12345)
 
 #calc 'bayesian r2': https://doi.org/10.1080/00031305.2018.1549100
@@ -224,9 +222,8 @@ hydraulicallyWide$logVms <- log10(hydraulicallyWide$Vms)
 bayeslm_log_hydraulicallyWide_smallEddy_eD <- stan_glm(log10(k600)~logGravitySlope+logVms, data=hydraulicallyWide,
                                                       family=gaussian(link='identity'),
[…]
